### 2005-03-29 paper/metamonads.tex
Commit of outline.txt (as sent by email) Create version of metamonads.tex, which currently is essentially just outline.txt which can be passed through LaTeX.

\documentclass[11pt]{llncs}

\title{Declarative Assembler}
\author{ Jacques Carette\inst{1}
 \and
 Oleg Kyselyov\inst{2}
} 
\institute{McMaster University,
1280 Main St. West, Hamilton, Ontario Canada L8S 4K1
\and
Monterey, CA 93943}

\usepackage{amsmath}

\begin{document}
\maketitle

\begin{abstract}
We use monads and Ocaml's advanced module system to
demonstrate that, in a generative context, it is 
possible to totally eliminate the overhead of abstraction.
This lets one use extreme forms of \textit{information hiding}
at no run-time cost.  Furthermore the typed nature of
the generative context provide static guarantees about the 
generated code.  The various extensions
(aspects) can be made completely orthogonal and compositional,
even in the presence of name-generation for temporaries and
other bindings and ``interleaving'' of aspects.  We also
show how to encode some domain-specific knowledge so that
``clearly wrong'' compositions can be statically rejected by the
compiler.
\end{abstract}

\section{Introduction}

	- problem of high-performance and numeric computing:
performance versus abstraction. Manual inlining and optimizations is
error-prone and makes code unreadable. Furthermore, the current
architectures demand more and more frequent tweaks (which, in general,
cannot be done by the compiler because the tweaking often involves domain
knowledge).

	- Gaussian elimination as a representative example.

	- Generating code. MetaOCaml. Static guarantees: the generated
code is well-typed.

	- Problems: generativity: generated code cannot be examined
and optimized at all. It must be generated just right. The FFT paper
showed one way: abstract interpretation. But more problems remain
(cite Padua et al paper at Haskell Workshop04): generating names in
the presence of `if', 'while' etc. control operators (simple problem
of generating names solved in the FFT paper); making CPS code clear
(note that many authors are afraid of CPS code because it quickly
becomes unreadable; we need CPS for name generation); compositionality
of various (often interleaving: explain: determinant computation
requires attention at several places) aspects and expressing
dependencies among them; expressing more static guarantees about the
generated code (e.g., not attempting full (fractional) division for
the integer domain; generating full-pivoting code for the rational
domain without `better\_than' operation). In short: can we achieve all
of the abstractions needed for the flexible Gaussian Elimination? Can
we eliminate all of that abstraction's overhead in the generated code
(without using intensional code analysis -- that is, maintaining the
guarantees of MetaOCaml and using pure generative approach).

[Note why it is important to maintain generativity: stronger
equational theory: Walid's paper. Also, intensional code analysis
(e.g., SPIRAL project) essentially requires one to insert an
optimizing compiler into the code generating system. That is extremely
complex, quite error-prone, and very difficult to ascertain the
correctness.]

\section{CPS and the problem of generating names}

[Oleg to write.]

\section{Monadic notation, making CPS code clear}

What is our monad: what is the action and what is the value.
ifM and ifL: show example. Mention 'ifM' and `reset'. Investigating
the connection with the delimited continuations: for future work.

What is state and why we need it: making aspects interleave: use det
as an example.

\section{Functors}

Simple parameterization: by the domain, by the container

Functors and the generated return type: the Out structure.

Making state modular and composable: several aspects can contribute
to the state, without interference (show: Rank, Det, Permutation
aspects).

Expressing domain constraints in functors (Div and integer domain).
Show the type error that occurs in improper instantiation.

Show several examples (e.g., code with NullPivot and nodet: no traces
of either det nor pivoting). Then the code with RowPivoting, Det and
float domain.

\section{Future work}
connections with delimited continuations: making notation
more direct-style and potentially clearer.

More aspects that can be handled: Input variations (augmented 
matrices).  Fit into the larger program family, where there are
more aspects still.

\bibliography{metamonads} 
\bibliographystyle{plain}

\end{document}

### 2005-03-30 paper/metamonads.tex
sections, packages
Check in for Oleg

@@
 
-\title{Declarative Assembler}
+% Reduce display spacing
+\divide\abovedisplayskip 2
+\divide\belowdisplayskip 2
+\divide\abovedisplayshortskip 2
+\divide\belowdisplayshortskip 2
+
+\usepackage{hyphenat}
+\usepackage{comment}
+\usepackage{amsmath}
+\usepackage{amstext}
+\usepackage{url}
+
+\usepackage{ifpdf}
+\ifpdf
+    \pdfpageheight=11in
+    \pdfpagewidth=8.5in
+\fi
+
+%\usepackage{refrange}
+
+\usepackage{fancyvrb}
+\DefineShortVerb{\|}
+\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\mathindent}
+\setlength{\parskip}{0pt}
+\newlength{\mathindent}
+\setlength{\mathindent}{1em}
+
+% Reduce list spacing
+%% \makeatletter
+%% \renewcommand\@@listI{\leftmargin\leftmargini
+%% \parsep \z@@
+%% \topsep 3\p@@ \@@plus\p@@ \@@minus 2\p@@
+%% \itemsep 2\p@@ \@@plus\p@@ \@@minus\p@@}
+%% \let\@@listi\@@listI
+%% \@@listi
+%% \makeatother
+
+\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
+
+\begin{document}
+\title{What is our title?}
 \author{ Jacques Carette\inst{1}
@@
 
-\usepackage{amsmath}
-
-\begin{document}
 \maketitle
@@
 the integer domain; generating full-pivoting code for the rational
-domain without `better\_than' operation). In short: can we achieve all
+domain without |better_than| operation). In short: can we achieve all
 of the abstractions needed for the flexible Gaussian Elimination? Can
@@
 
-[Oleg to write.]
+In this paper, we are concerned with building and composing
+generators. We start with the code expressing the algorithm (in
+OCaml), and, adding MetaOCaml annotations, turn that code into a
+\emph{code expression} that will generate the code when runs. We then
+modify that code expression to add various parameters and aspects, to
+yield the complete generator.
+
+Here we illustrate the MetaOCaml constrcutions and code generators.
+
+Turning code into the code expression is relatively straightforward
+until we come to binding constructs such as |let|. For example, here
+is a piece of code that defines a function that searches a given array
+|arr| for the maximal (in absolute value) element, and the index of
+that element and the found absolute value. This code is representative
+of pivot determination. The code uses mutations and destructive
+updates -- partly for illustrative purposes and mainly because many
+linear algebra algorithms are expressed via an imperative code (so it
+is less error-prone to translate the ``canonical'' algorithms with
+minimal changes).
+
+\begin{code}
+let fp = fun arr n ->
+   let pv = ref (0,arr.(0)) in
+   for i = 1 to n-1 do
+     if abs_float arr.(i) > snd !pv then
+        pv := (i, abs_float arr.(i))
+   done;
+   !pv
+\end{code}
+
+This function has the type |float array -> int -> int * float|. By
+enclosing the code into MetaOCaml brackets, we obtain the simplest
+generator expression.
+
+\oleg{later on, we can write the following just let fpc = .<...>.}
+
+\begin{code}
+let fpc = .<fun arr n ->
+   let pv = ref (0,arr.(0)) in
+   for i = 1 to n-1 do
+     if abs_float arr.(i) > snd !pv then
+        pv := (i, abs_float arr.(i))
+   done;
+   !pv>.
+\end{code}
+
+This value has the type |('a, float array -> int -> int * float) code|.
+Here |'a| type parameter is needed (ask Walid for the reference).
+The main difference between |fp| and |fpc| is that |fp| is a function,
+which we can immediately use as \verb^fp [|1.0; -4.; 3.|] 3^. In
+contrast, |fpc| is not a function: it is a \emph{code} value that denotes
+that function. We would say that the code expression is the
+``second-stage'' value. The ability to manipulate code expressions as
+regular values is the distinguished fature of MetaOCaml.
+We can ``compile'' the code by using MetaOCaml |run| command, or |.!|:
+|.! fpc|. The result is equivalent is |fp|. Although |fpc| doesn't
+seem to do much, it is already useful as we can use offshoring (cite)
+to translate |fpc| into C or Fortran code (which we can dynamically
+link into the running MetaOCaml code -- or use separately).
+
+The code |fpc| obviously lacks generality: the name |abs_float|
+implies that that code deals with |float| arrays. We wish to handle
+arrays in various domains, integer, rationals, etc. We can make the
+code generic by parametrizing over the |abs| function:
+
+\begin{code}
+let fpca absf = .<fun arr n ->
+   let pv = ref (0,arr.(0)) in
+   for i = 1 to n-1 do
+     if absf arr.(i) > snd !pv then
+        pv := (i, absf arr.(i))
+   done;
+   !pv>.
+\end{code}
+
+and so |fpca abs_float| will manipulate |float array| and |fpca abs|
+will do for the integer arrays. Here we used the value from the first
+stage, |absf|, to incorporate into the code at the second stage:
+croos-stage persistent values.
+
+We should now consider the function |fpca| in a browder context: of a
+function that computes the pivot and then swaps the pivot withe 0-th
+element of the array:
+
+\begin{code}
+let upv = .<fun arr n ->
+   let pv = .~(fpca abs_float) arr n in
+   let i = fst pv in
+   (if not (i = 0) then
+      let t = arr.(0) in
+      begin arr.(0) <- arr.(i); arr.(i) <- t end);
+   arr>.
+\end{code}
+
+Here, we see the third MetaOCaml operator, escape |.~|. It does ...
+MetaOcaml can print out the code, which looks like
+|.<fun arr n -> let pv = (fun arr n -> fpcs code) arr n in ... >.|
+That is, the compiler inlined the code for fpca -- which is a
+functional expression. The |upv| code shows the functional
+application, |(fun arr n -> ...)| applied to arguments |arr n|. That
+application will be executed at run-time, unless a sufficiently smart
+compiler can inline it. As we stated, our aim is to not to rely on
+post-generation optimizations and sufficiently smart compiler. We
+strive to generate the efficient code ourselves. We can do that by
+modifying the definition of |fpca| as
+
+\begin{code}
+let fpca absf arr n = .<
+   let pv = ref (0,(.~arr).(0)) in
+   for i = 1 to .~n-1 do
+     if absf (.~arr).(i) > snd !pv then
+        pv := (i, absf (.~arr).(i))
+   done;
+   !pv>.
+\end{code}
+\begin{code}
+let upv = .<fun arr n ->
+   let pv = .~(fpca abs_float .<arr>. .<n>.) in
+   let i = fst pv in
+   (if not (i = 0) then
+      let t = arr.(0) in
+      begin arr.(0) <- arr.(i); arr.(i) <- t end);
+   arr>.
+\end{code}
+% (.! upv) [|1.0; -4.; 2.|] 3;;
+We see that we pass code fragment |.<arr>.| as arguments to the
+function. Now when we look at the code for |upv|, we see
+|.<fun arr n -> let pv1 = let pv = ref (0,arr.(0)) in ...>.|
+So the fpca code got really inlined.
+
+The inlining is not perfect however. We would like that our code
+expression to look like
+\begin{code}
+ fun arr n ->
+   let pv = ref (0,arr.(0)) in ... the rest of fpca ...
+   in let i = fst !pv in ...
+\end{code}
+
+We would like that the inlining of |.~(fpca abs_float .<arr>. .<n>.)|
+generated us the |let pv = ref ... | statement. But this seems to
+quite a challenge. We wish the resulting code to look
+|let pv = ... in body| where |pv = ...| part come from |fpca| and
+|body| come from |upv|. At first blush, this seems impossible. A code
+expression may contain only complete expressions, such as
+|let v =... in ...|. A code expression cannot contain only the binding part of
+|let|, which is not an expression per se.
+
+The solution: CPS. And here where I go to sleep.
 
@@
 
+\section{Related work}
+
 \section{Future work}
@@
 
+\section{Conclusion}
+
 \bibliography{metamonads}

### 2005-03-30 paper/metamonads.tex
Add a .bib file.  Start work on the introduction.

@@
 \begin{document}
-\title{What is our title?}
-\author{ Jacques Carette\inst{1}
+\title{Functors, CPS and monads, or how to generate efficient
+code from abstract designs}
+\author{Jacques Carette\inst{1}
  \and
@@
 
-- problem of high-performance and numeric computing:
-performance versus abstraction. Manual inlining and optimizations is
-error-prone and makes code unreadable. Furthermore, the current
-architectures demand more and more frequent tweaks (which, in general,
-cannot be done by the compiler because the tweaking often involves domain
-knowledge).
+In high-performance and numeric computing, there is a well-known
+issue of balancing between maximal performance and the level of
+abstraction at which code is written.  Furthermore, already in
+linear algebra, there is a wealth of different concerns that 
+\emph{may} need to be addressed.  While it is possible to manually
+inline and simplify such code, this is labour-intensive and 
+error-prone.  Furthermore the resulting code is frequently unreadable,
+and thus unmaintainable.
 
-- Gaussian elimination as a representative example.
+There are a number of projects that have attempted to address these issues
+\cite{Czarnecki,Veldhuizen:1998:ISCOPE,Veldhuizen:1998:OO98,musser94algorithmoriented,BOOST,POOMA,ATLAS}, 
+and they have generally succeeded in generating
+efficient code, but the cost was frequently the loss of modularity
+of the design.  In other words, to allow the various aspects to
+weave with each other \emph{efficiently}, information which should
+have been kept hidden leaked from module to module.  This seriously
+affects maintainability and scalability.  Furthermore, current architectures
+demand more and more frequent tweaks which, in general, cannot be done by the
+compiler because the tweaking often involves domain knowledge.
 
-- Generating code. MetaOCaml. Static guarantees: the generated
-code is well-typed.
+Our goal was to implement a completely modular, textbook-style 
+modular design \cite{journals/cacm/parnas72a}, yet 
+have our code be very efficient.  Gaussian elimination, as it is 
+well-known, already contains many different aspects \cite{carette04},
+was chosen as a representative example.
 
-- Problems: generativity: generated code cannot be examined
-and optimized at all. It must be generated just right. The FFT paper
+Whether one prefers to code in an imperative, object-oriented, functional, 
+or even in a mixed style, all abstraction mechanisms in current
+languages have a non-zero run-time cost.  Whether it is procedure,
+method or function call, these have a cost; worse still, higher level
+and/or higher order abstractions (like Functors or AspectJ's aspects)
+frequently have a larger cost still.  Of course, if one is faced with
+writing 60 different implementations of the ``same'' algorithm, this
+abstraction cost starts to look cheap!  Although bad programmers have
+been known to resort to cut\&paste programming, no self-respecting
+programmer would be caught doing that\footnote{Sadly, there are still
+many programmers who do not see the harm of cut\&paste programming}.
+The common solution is to turn to code generation.
+
+While many people still make do with string-based code generation,
+this is not much different than writing code in a dynamically typed
+interpreted scripting language.  It works, it is convenient, but
+it does not scale well, it is a debugging nightmare, and usually
+suffers most from Software Aging \cite{parnas_aging}.  As with
+other tasks in programming, the answer is to turn to statically
+typed languages, which provide strong guarantees about the soundness
+of the resulting code.  MetaOCaml in particular stands out: it
+guarantees that both the code generator and the generated
+code are well-typed.  While Template programming in C++ and 
+Template Haskell do offer some amount of static typing 
+\cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml\footnote{and MetaML.}
+offers stronger guarantees.  As well, since MetaOCaml is implemented
+on top of the very solid Ocaml language, one accrues additional benefits.
+
+While it is theoretically feasible in a strings-based
+generative system to re-parse the generative code to inspect it, such
+an approach rapidly degenerates into complete chaos.  In languages
+that natively offer reflection and reification (either at no cost, 
+such as in LISP and Scheme, or at a modest cost in languages like 
+Maple), this is actually feasible, even easy.  But this is fragile
+and error-prone, since the languages offer little support for catching
+errors early.
+
+In MetaOCaml, generated code cannot be examined and optimized at all. It must
+be generated just right.
+
+- The FFT paper
 showed one way: abstract interpretation. But more problems remain

### 2005-03-31 paper/metamonads.tex
a better draft of the CPS section

@@
 
-The inlining is not perfect however. We would like that our code
-expression to look like
+The generator |fpca| shows one particular way of searching for pivot.
+In the domain of exact rational numbers, for example, we would be
+satisfied with the first non-zero element as a pivot, and so we would
+write a different pivot-searching generator. The body of |upv| moves
+the found pivot to the ``canonical'' location. Again, depending on the
+container (a dense matrix, a sparse matrix, etc), there are different
+ways of doing that. So, we would like to be able to ``compose'' pivot
+generating function with the swap generating function. We would like
+to write something like this
+
 \begin{code}
- fun arr n ->
-   let pv = ref (0,arr.(0)) in ... the rest of fpca ...
-   in let i = fst !pv in ...
+let fpca absf arr n pv = .<
+   for i = 1 to .~n-1 do
+     if absf (.~arr).(i) > snd ! (.~pv) then
+        (.~pv) := (i, absf (.~arr).(i))
+   done>.
+let swapper arr n pv = .<
+   let i = fst !(.~pv) in
+   if not (i = 0) then
+      let t = (.~arr).(0) in
+      begin (.~arr).(0) <- (.~arr).(i); (.~arr).(i) <- t end>.
+let seqv a b c = .<begin .~a; .~b; .~c end>.
+let pv_swap absf arr n =
+   let pv = .<ref (0,(.~arr).(0))>. in
+   seqv (fpca absf arr n pv)
+        (swapper arr n pv)
+        arr
+let top =  .<fun arr n -> .~(pv_swap abs_float .<arr>. .<n>.)>.
 \end{code}
 
-We would like that the inlining of |.~(fpca abs_float .<arr>. .<n>.)|
-generated us the |let pv = ref ... | statement. But this seems to
-quite a challenge. We wish the resulting code to look
-|let pv = ... in body| where |pv = ...| part come from |fpca| and
-|body| come from |upv|. At first blush, this seems impossible. A code
-expression may contain only complete expressions, such as
-|let v =... in ...|. A code expression cannot contain only the binding part of
-|let|, which is not an expression per se.
+The function |pv_swap| shows our ideal -- composing code gerenators
+|fpca| and |swapper|, using a higher-order combinator `seqv'
+(sequence of code fragments). We lifted the pivot declaration out of
+old version of |fpca| so that both |fpca| and |swapper| could be
+parameterized by the pivot (and |fpca| and similar function will deal
+only with pivot determination). |top| puts it all together. The above
+code represents the pattern we will be following for the rest of the
+paper.
 
-The solution: CPS. And here where I go to sleep.
+And it \emph{almost} works. It is type correct.  When we look at the
+generated code for |top|, we notice the `(ref (0, arr_1.(0)))' is
+inlined wherever pivot was expected. For example, the swapping part of
+the code begins with |let i_3 = (fst (! (ref (0, arr_1.(0))))) in
+...|. That is certainly not what we want. We meant for pivot searching
+and swapping parts of the code to communicate via a shared mutable
+state, variable |pv|. In the generated code, each piece of code gets
+its own copy of |pv|. Rather then inline (ref (0, arr_1.(0)))
+everywhere, we should have generated |let pv = ref (0,(.~arr).(0)) in ...|
+and inlined |pv|. Incidentally, even in a pure functional code we
+often wish to generate such a |let| statement so we can compute a
+complex expression only once and use it value in several other
+places. Again, we would rather not leave the job of detecting such
+common subexpressions to the compiler (because in the complex code the
+compiler may fail to detect the common subexpressions). The problem
+with generating such |let| expressions is that we wish our geenrator
+to return just |.<let pv = ref (0,(.~arr).(0)) in>.|, without the `body'
+of the |let|, which we would fill in later. In languages which
+generate code in the string form this is easy. We generate the string
+|''let pv = ref (0,(.~arr).(0)) in''| and then append to it the string
+representting the body, wcich would give us the complete |let|
+statement. In MetaOCaml, we can't generate incomplete expressions, for
+a good reason.
+
+To solve the problem of writing a generator for a |let| expression 
+when the generated body is not yet known, we use CPS.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+And now, a better -- and shorter -- write-up.
+
+The essence of our approach is assembling the code of the program by combining
+instances of various aspects together. In the case of Gaussian
+Elimination, the aspects are pivot determination, swapping rows and
+columns, determinant tracking, etc. A primitive code generator yields a
+code expression 
+A combinator takes (complete and well-typed) code expressions combines
+them into a composite expression (which is too statically guaranteed
+to be well-formed and well-typed).
+
+For example, the following is probably the simplest code generators:
+\begin{code}
+let one = .<1>.
+\end{code}
+
+(in the floating-poiint instance of that aspect, we would have
+|.<1.0>.| instead). We use MetaOCaml brackets |.<...>.| to generate
+code fragments (perhaps say one more sentence about the brackets).
+The simplest combination is inlining:
+\begin{code}
+let plus x y = .<.~x + .~y>.
+\end{code}
+Here we see the second component of MetaOCaml -- escapes, which are
+...
+
+We can see the result in
+\begin{code}
+let simplest_code =
+  let gen x y = plus x (plus y one) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+
+MetaOCaml helpfully prints out the result of the latter, as
+|.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|
+
+We must point out that 'gen x y' looks like a combinator; the body of
+'gen x y' involves several apoplications of the function 'plus'. When
+we look at the generated code however, we see no traces of function
+'plus' and no such administrative applications. Instead, we see the
+fully inlined expression. These administative applications are done at
+the code generation stage.
+
+The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
+execute the above code: |.! simplest_code| is a function of two
+integers, which we can apply: |(.! simplest_code) 1 2|. The original
+simplest_code is not the function on integers -- it's a code
+expression.
+
+To see the benefit of code generation, we notice that we can easily
+parameterize our code:
+
+\begin{code}
+let simplest_param_code plus one =
+  let gen x y = plus x (plus y one) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+
+and use it to generate code that operates on integers, floating pint
+numbers or boolean -- in general, any domain that implements |plus|
+and |one|:
+\begin{code}
+let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
+  simplest_param_code plus one
+let plus x y = .<.~x || .~y>. and one = .<true>. in
+  simplest_param_code plus one
+\end{code}
+Running the former expression yields the function on |float|s, whereas
+the latter expression is the code expression for a bollean function.
+
+We can see the separation of concerns, implementation of domain
+operations.
+
+Let us consider a more complex expression:
+\begin{code}
+let param_code1 plus one =
+  let gen x y = plus (plus y one) (plus x (plus y one)) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+
+We see that expression `(plus y one)' appears twice. That may be quite
+a complex computation (depending on the implementation of `plus'), and
+so we would rather not do it twice. We may be tempted to rely on the
+compiler's built-in common-subexpression elimination algorithm, which
+is a typical part of the optimizer. We would rather not relied on the
+sufficiently smart compiler: when the generated code is very complex,
+the compiler may overlook common subexpressions (or the subexpression
+occur in such an imperative context that the compiler might not be
+able to decide if lifting the subexpression is sound. So, being
+conservative, the optimizer will leave the duplicates). We may attempt
+to do
+\begin{code}
+let param_code1 plus one =
+  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+
+However, examinine the code for |param_code1 plus one|
+shows that |.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|
+we might have saved the computation on the code generation
+stage. However, the generated code still exhibits the
+duplication (which makes the code larger and may take longer to
+execute, if the optimizer fails to eliminate the duplication).
+So, we need a combinator that inserts the |let| expression in the
+generat\emph{ed} code. We need a combinator |letgen| so we could use
+that as |let ce = letgen (plus y one) in plus ce (plus x ce)| 
+and get the code that looks like .<... let t = y + 1 in t + (x + t)>.
+But that seems impossible because |letgen exp| needs to generate
+the let expression .<let t = exp in body>. but |letgen| doesn't have
+the body yet. The body needs that temporary identifier .<t>. that will
+be generated by |letgen| itself. Certainly |letgen| cannot generate
+only a part of a let-expression. All generated expressions in
+MetaOCaml are well-formed and complete.
+
+The key is to use CPS. It was first pointed out by XXX in the context
+of partial evaluation, and extensively used by (FFT) for code
+generation.
+
+\begin{code}
+let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
+let param_code2 plus one =
+  let gen x y k = letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
+  and k0 x = x
+  in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
+\end{code}
+
+Now, |param_code2 plus one| gives us the desired code
+|.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.|.
 

### 2005-03-31 paper/metamonads.tex
some corrections

@@
 this is not much different than writing code in a dynamically typed
-interpreted scripting language.  It works, it is convenient, but
-it does not scale well, it is a debugging nightmare, and usually
-suffers most from Software Aging \cite{parnas_aging}.  As with
-other tasks in programming, the answer is to turn to statically
-typed languages, which provide strong guarantees about the soundness
-of the resulting code.  MetaOCaml in particular stands out: it
-guarantees that both the code generator and the generated
-code are well-typed.  While Template programming in C++ and 
-Template Haskell do offer some amount of static typing 
-\cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml\footnote{and MetaML.}
-offers stronger guarantees.  As well, since MetaOCaml is implemented
-on top of the very solid Ocaml language, one accrues additional benefits.
+interpreted scripting language.  It works, it is convenient, but it
+does not scale well, it is a debugging nightmare, and usually suffers
+most from Software Aging \cite{parnas_aging}.  As with other tasks in
+programming, the answer is to turn to statically typed languages,
+which provide strong guarantees about the soundness of the resulting
+code.  MetaOCaml (cite website?) \cite{CTHL03} in particular stands
+out: not only it guarantees the absense of problems such as
+accidental variable capture -- so-called hygiene cite? \oleg{should we
+  also mention automatic well-formedness}, MetaOcaml also
+statically ensures that both the code generator and the generated code
+are well-typed \cite{TahaSheard97,TahaThesis}.  While Template
+programming in C++ and Template Haskell do offer some amount of static
+typing \cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml\footnote{and
+  MetaML.}  offers stronger guarantees.  As well, since MetaOCaml is
+implemented on top of the very solid Ocaml language, one accrues
+additional benefits.
 
@@
 And it \emph{almost} works. It is type correct.  When we look at the
-generated code for |top|, we notice the `(ref (0, arr_1.(0)))' is
+generated code for |top|, we notice the |(ref (0, arr_1.(0)))| is
 inlined wherever pivot was expected. For example, the swapping part of
-the code begins with |let i_3 = (fst (! (ref (0, arr_1.(0))))) in
-...|. That is certainly not what we want. We meant for pivot searching
+the code begins with |let i_3 = (fst (! (ref (0, arr_1.(0))))) in ...|. 
+That is certainly not what we want. We meant for pivot searching
 and swapping parts of the code to communicate via a shared mutable
 state, variable |pv|. In the generated code, each piece of code gets
-its own copy of |pv|. Rather then inline (ref (0, arr_1.(0)))
+its own copy of |pv|. Rather then inline |(ref (0, arr_1.(0)))|
 everywhere, we should have generated |let pv = ref (0,(.~arr).(0)) in ...|
@@
 
-The essence of our approach is assembling the code of the program by combining
-instances of various aspects together. In the case of Gaussian
-Elimination, the aspects are pivot determination, swapping rows and
-columns, determinant tracking, etc. A primitive code generator yields a
-code expression 
-A combinator takes (complete and well-typed) code expressions combines
-them into a composite expression (which is too statically guaranteed
-to be well-formed and well-typed).
+The essence of our approach is assembling the code of the program by
+combining instances of various aspects together. In the case of
+Gaussian Elimination, the aspects are pivot determination, swapping of
+rows and columns, determinant tracking, etc. A primitive code
+generator yields a code expression.  A combinator takes (complete and
+well-typed) code expressions and combines them into a composite expression
+(which is too statically guaranteed to be well-formed and well-typed).
 
-For example, the following is probably the simplest code generators:
+We will be using MetaOcaml, which, as the instance of a multi-stage
+programming system \cite{TahaThesis}, provides exactly the needed
+features: to construct a code expression, to combine them, and to
+execute them
+
+For example, the following is probably the simplest code generator:
 \begin{code}
@@
 
-(in the floating-poiint instance of that aspect, we would have
+(in the floating-point instance of that aspect, we would have
 |.<1.0>.| instead). We use MetaOCaml brackets |.<...>.| to generate
-code fragments (perhaps say one more sentence about the brackets).
+code expressions (perhaps say one more sentence about the brackets).
 The simplest combination is inlining:
@@
 integers, which we can apply: |(.! simplest_code) 1 2|. The original
-simplest_code is not the function on integers -- it's a code
+|simplest_code| is not the function on integers -- it's a code
 expression.
@@
 
-However, examinine the code for |param_code1 plus one|
+However, examinining the code for |param_code1 plus one|
 shows that |.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|

### 2005-03-31 paper/metamonads.tex
section about monads is half-finished.

@@
 correctness.]
+
+\oleg{explictly enumerate our contributions, not foiregtting the mdo
+  notation. The subsubsection that outlines what the other sections
+  are for.}
 
@@
 features: to construct a code expression, to combine them, and to
-execute them
+execute them.
 
@@
 |.<1.0>.| instead). We use MetaOCaml brackets |.<...>.| to generate
-code expressions (perhaps say one more sentence about the brackets).
-The simplest combination is inlining:
+code expressions -- in other words, to construct future-stage
+computations. The simplest combination is inlining:
 \begin{code}
@@
 \end{code}
-Here we see the second component of MetaOCaml -- escapes, which are
-...
+Here we see the second component of MetaOCaml -- escapes |.~|. They
+let us perform an immediate code generating computation while we are
+building the future-stage computation (code expression). In the
+example above, the immediate computation |.~x| is trivial -- obtaining
+the value of the variable |x|, which should be a code expression, and
+inlining it. The following shows an example of using the simple
+generators:
 
-We can see the result in
 \begin{code}
@@
 
-MetaOCaml helpfully prints out the result of the latter, as
+Here, the immediate computation |gen .<x>. .<y>.| is less trivial --
+it includes the evaluation of the function |gen|, which in turn
+applies the function |plus|. The function receives code expressions
+|.<x>.| and |.<y>.| as arguments. At the generating stage, we can
+manipulate code expressions as values. The function |gen| returns a
+code expression, which will be inlined in the place of the
+escape. MetaOCaml can print out code expressions, so we can see the
+final generated code:\\
   |.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|
 
-We must point out that 'gen x y' looks like a combinator; the body of
-'gen x y' involves several apoplications of the function 'plus'. When
+We must point out that |gen x y| looks like a combinator; its body 
+involves several applications of the function |plus|. When
 we look at the generated code however, we see no traces of function
-'plus' and no such administrative applications. Instead, we see the
+|plus| and no such administrative applications. Instead, we see the
 fully inlined expression. These administative applications are done at
@@
 The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
-execute the above code: |.! simplest_code| is a function of two
+execute the code expression: |.! simplest_code| is a function of two
 integers, which we can apply: |(.! simplest_code) 1 2|. The original
-|simplest_code| is not the function on integers -- it's a code
+|simplest_code| is not the function on integers -- it is a code
 expression.
@@
 
-and use it to generate code that operates on integers, floating pint
-numbers or boolean -- in general, any domain that implements |plus|
+and use it to generate code that operates on integers, floating point
+numbers or booleans -- in general, any domain that implements |plus|
 and |one|:
@@
 
-We can see the separation of concerns, implementation of domain
+We can see the separation of concerns, namely of that for domain
 operations.
@@
 
-We see that expression `(plus y one)' appears twice. That may be quite
-a complex computation (depending on the implementation of `plus'), and
+We notice two occurrences of |plus y one|. Depending on the
+implementation of |plus| that may be quite a complex computation, and
 so we would rather not do it twice. We may be tempted to rely on the
-compiler's built-in common-subexpression elimination algorithm, which
-is a typical part of the optimizer. We would rather not relied on the
-sufficiently smart compiler: when the generated code is very complex,
-the compiler may overlook common subexpressions (or the subexpression
-occur in such an imperative context that the compiler might not be
-able to decide if lifting the subexpression is sound. So, being
-conservative, the optimizer will leave the duplicates). We may attempt
-to do
+compiler's common-subexpression elimination optimization. We would
+rather not relied on a ``sufficiently smart compiler'': when the
+generated code is very complex, the compiler may overlook common
+subexpressions. The subexpressions may occur in such an imperative
+context that the compiler might not be able to determine if lifting
+them is sound. So, being conservative, the optimizer will leave the
+duplicates as they are. We may attempt to write
+
 \begin{code}
 let param_code1 plus one =
-  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
+  let gen x y = 
+     let ce = (plus y one) in  plus ce (plus x ce) in
   .<fun x y -> .~(gen .<x>. .<y>.)>.
@@
 
-However, examinining the code for |param_code1 plus one|
-shows that |.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|
-we might have saved the computation on the code generation
-stage. However, the generated code still exhibits the
-duplication (which makes the code larger and may take longer to
-execute, if the optimizer fails to eliminate the duplication).
-So, we need a combinator that inserts the |let| expression in the
-generat\emph{ed} code. We need a combinator |letgen| so we could use
-that as |let ce = letgen (plus y one) in plus ce (plus x ce)| 
-and get the code that looks like .<... let t = y + 1 in t + (x + t)>.
-But that seems impossible because |letgen exp| needs to generate
-the let expression .<let t = exp in body>. but |letgen| doesn't have
-the body yet. The body needs that temporary identifier .<t>. that will
-be generated by |letgen| itself. Certainly |letgen| cannot generate
-only a part of a let-expression. All generated expressions in
+However, the result of |param_code1 plus one|, which prints as\\
+|.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|,
+still exhibits duplicate sub-expressions. Our |let|-insertion
+optimization saved the computation on the generating stage. 
+We need a combinator that inserts the |let| expression in the
+generat\emph{ed} code. We need a combinator |letgen| to be used
+as\\|let ce = letgen (plus y one) in plus ce (plus x ce)| 
+yielding the code like |.<let t = y + 1 in t + (x + t)>.|
+But that seems impossible because |letgen exp| has to generate
+the expression\\|.<let t = exp in body>.| but |letgen| does not have
+the |body| yet. The body needs that temporary identifier |.<t>.| that
+is supposed to be the result of |letgen| itself. 
+Certainly |letgen| cannot generate only a part of a let-expression,
+without the |body|.  All generated expressions in
 MetaOCaml are well-formed and complete.
 
-The key is to use CPS. It was first pointed out by XXX in the context
-of partial evaluation, and extensively used by (FFT) for code
+The key is to use continuation-passing style (CPS). It was first
+pointed out by \cite{Bondorf:92} in the context
+of partial evaluation, and extensively used by \cite{KiselyovTaha} for code
 generation.
@@
 
-Now, |param_code2 plus one| gives us the desired code
+Now, |param_code2 plus one| gives us the desired code\\
 |.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.|.
@@
 
-What is our monad: what is the action and what is the value.
+Comparing the code that did let-insertion at the generating stage\\
+|let ce = (plus y one) in  plus ce (plus x ce)|\\
+with the corresponding code inserting let at the generated code stage\\
+|letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))|\\
+clearly shows the difference between the direct-style and CPS code.
+What was |let ce = init in ...| in direct style became
+|init' (fun ce -> ...)| in CPS. For one thing, |let| became
+inverted. For another, what used to be an expression that yields
+value, |init|, now became an expression that takes an extra argument,
+the continuation, and invokes it. The differences look negligible in
+the above example. In larger expressions with many let-forms, the
+number of paranetheses around |fun| increases, the need to add and
+then invoke the |k|, continuation argument, become annoying. The
+inconvenience is great enough for some people to explicitly avoid CPS
+or claim that numerical programmers (our users) cannot or will not
+program in CPS. Clearly a better notation is needed.
+
+The |do|-notation of Haskell (cire report?) shows that it is possible
+to write essentially CPS code in the conventional-looking style. The
+|do| notation is the notation for writing monadic code (cite Moggi?).
+The benefit of monadic notation is not that it can represent CPS (cite
+Filinski), but it helps in composability by offering to add different
+layers of effects (state, exception, non-determinism, etc) to the
+basic monad (cite Hudak, Peyton-Jones).
+
+In the present paper, we use the monad that is the CPS monad with
+another layer of state. We will discuss the need for the state in
+Section XXX. The type of the monad is:
+\begin{code}
+type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
+\end{code}
+
+The monad is just the function of two arguments: the state (of type
+|'s|) and the continuation. The continuation receives the current
+state, the value (of type |v|) and yields the answer of the type |w|.
+The monad is polymorphic over all these three type parameters. Except
+a few primitime monad operations below, all other code treat the monad
+as an abstract data type.
+
+Monads represent computations that can have ``effects''. In our case,
+we can have ``state effects'' (update of state) and ``control effect''
+(which occurs whenever the computation does something else but
+invoking its continuation once). As we will see, ``control effects''
+in our case manifest themselves as ``code generation effects''.
+
+\begin{code}
+let ret a = fun s k -> k s a
+let bind a f = fun s k -> a s (fun s' b -> f b s' k)
+\end{code}
+
+The two monad operations, |ret| and |bind| are conventional: |ret a|
+returns a computation that takes the current state and the
+continuation and immediately invokes the continuation passing it the
+unmodified state and the value. Obviously, the computation |ret a| has
+no control effects. The operation |bind a f| combines a
+monad |a| of the type |(s,v,w) monad| and the function |f| of the
+type |v -> (s,v1,w) monad| into a bigger |(s,v1,w) monad|. The latter
+will execute computation |a|, pass it result to |f| and execute the
+latter's computation. The state facet of our monad is handled with two
+operations:
+
+\begin{code}
+let fetch s k = k s s
+let store v s k = k (v::s) ()
+\end{code}
+
+|fetch| is the computation that yields the current state and |store v|
+is the computation that updates the current state and yields |()|. We
+model the state as a list (which is actually a list of polymorphic
+variants, as we shall see later).
+
+\begin{code}
+let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
+\end{code}
+
+The operation |retN a| is the let-insertion operation (whose simpler 
+version was called |letgen| before). It is the first computation with
+a control effect: indeed, the result of |retN a| is \emph{not} the
+result of invoking its continuation |k|. Rather, its result is a |let|
+code expression. Mention similarity with reset?
+
+Finally, we should be able to run our monad -- that is, perform the
+computation of teh monad and get the result -- which in our case is
+the code expression. We run the monad by passing it initial state (the
+empty list) and the initial continuation |let k0 s v = v|.
+
+We can now re-write our |param_code2| example of the previous section
+as
+
+\begin{code}
+let param_code3 plus one =
+  let gen x y = bind (retN (plus y one)) (fun ce -> 
+                ret (plus ce (plus x ce)))
+  and run m = m [] k0 
+  in .<fun x y -> .~(run (gen .<x>. .<y>.))>.
+\end{code}
+
+That does not seem like the improvement. But here we can introduce the
+do notation, which is patterned after the do-namtion of Haskell.
+\oleg{Explain our mdo briefly? Refer to the message on the
+  Ocaml-list?. Refer to camlp4 and note the implementation}
+
+\begin{code}
+let param_code4 plus one =
+  let gen x y = mdo { ce <-- retN (plus y one);
+                      ret (plus ce (plus x ce)) }
+  and run m = m [] k0 
+  in .<fun x y -> .~(run (gen .<x>. .<y>.))>.
+\end{code}
+
+The function |param_code4| is totally equivalent to |param_code3| --
+the |mdo|-notation is just the syntactic sugar. In fact, the camlp4
+preprocessor (with our extension) will transform the |param_code4|
+into |param_code3|. And yet, |param_code4| looks far more
+conventional, as if it were indeed in direct style.
+
+\oleg{Should we rename lif into ifL?}
+Now we can build more complex monadic operations.
+\begin{code}
+let lif test th el = ret .< if .~test then .~th else .~el >.
+\end{code}
+
 ifM and ifL: show example. Mention 'ifM' and `reset'. Investigating
@@
 
+Mention the mdo notation. Mention it into the introduction.
+
 \bibliography{metamonads}

### 2005-03-31 paper/metamonads.tex
some re-writing

@@
 
+\newcommand{\omitnow}[1]{}
 \newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
@@
 \section{CPS and the problem of generating names}
-
+\omitnow{
 In this paper, we are concerned with building and composing
@@
 when the generated body is not yet known, we use CPS.
+}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-And now, a better -- and shorter -- write-up.
+%% And now, a better -- and shorter -- write-up.
 
@@
 
-The |do|-notation of Haskell (cire report?) shows that it is possible
+The |do|-notation of Haskell (cite report?) shows that it is possible
 to write essentially CPS code in the conventional-looking style. The
@@
 
-In the present paper, we use the monad that is the CPS monad with
-another layer of state. We will discuss the need for the state in
-Section XXX. The type of the monad is:
+A monad (cite Moggi?) is an abstract datatype representing
+computations that yield a value and may have an \emph{effect}.
+The datatype must have at least two operations, |return| to build
+trivial effect-less computations and |bind| for combining
+computations. These operations must satisfy \emph{monadic laws}:
+|return| being the left and the right unit of |bind| and |bind| being
+associative. Figure XXX defines the monad in the present paper
+and shows its implementation.
+
+\begin{figure*}
 \begin{code}
 type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
-\end{code}
-
-The monad is just the function of two arguments: the state (of type
-|'s|) and the continuation. The continuation receives the current
-state, the value (of type |v|) and yields the answer of the type |w|.
-The monad is polymorphic over all these three type parameters. Except
-a few primitime monad operations below, all other code treat the monad
-as an abstract data type.
-
-Monads represent computations that can have ``effects''. In our case,
-we can have ``state effects'' (update of state) and ``control effect''
-(which occurs whenever the computation does something else but
-invoking its continuation once). As we will see, ``control effects''
-in our case manifest themselves as ``code generation effects''.
-
-\begin{code}
 let ret a = fun s k -> k s a
 let bind a f = fun s k -> a s (fun s' b -> f b s' k)
-\end{code}
-
-The two monad operations, |ret| and |bind| are conventional: |ret a|
-returns a computation that takes the current state and the
-continuation and immediately invokes the continuation passing it the
-unmodified state and the value. Obviously, the computation |ret a| has
-no control effects. The operation |bind a f| combines a
-monad |a| of the type |(s,v,w) monad| and the function |f| of the
-type |v -> (s,v1,w) monad| into a bigger |(s,v1,w) monad|. The latter
-will execute computation |a|, pass it result to |f| and execute the
-latter's computation. The state facet of our monad is handled with two
-operations:
-
-\begin{code}
 let fetch s k = k s s
 let store v s k = k (v::s) ()
-\end{code}
 
-|fetch| is the computation that yields the current state and |store v|
-is the computation that updates the current state and yields |()|. We
-model the state as a list (which is actually a list of polymorphic
-variants, as we shall see later).
+let k0 s v = v
+let runM m = m [] k0
 
-\begin{code}
 let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
 \end{code}
+\caption{Our monad}
+\end{figure*}
+
+Our monad represents two kinds of computational effects: reading and
+writing a computation-wide state, and control effects. The latter are
+normally associated with exceptions, forking of computations, etc. --
+in general, whenever a computation ends with something other than
+invoking its natural continuation in the tail position. In our case
+however the control effects manifest themselves as code generation.
+
+In Figure XXX, the monad is implemented as a function of two
+arguments: the state (of type |s|) and the continuation. The
+continuation receives the current state, the value (of the type |v|) and
+yields the answer of the type |w|.  The monad is polymorphic over all
+these three type parameters. Other implementations are
+possible. Except the code in Figure XXX, the rest of our code treats
+the monad as a truly abstract data type. The implementation of basic
+monadic operations |ret| and |bind| are conventional. It is easy to
+see that the monadic laws are satisfied. Other monadic operations
+construct computations that do have effects. Operations |fetch| and
+|store v| construct computations that read and write the state. In our
+case the state is a list (of polymorphic variants), which models an
+open discriminated union, as we shall see later.
 
@@
 
-Finally, we should be able to run our monad -- that is, perform the
-computation of teh monad and get the result -- which in our case is
-the code expression. We run the monad by passing it initial state (the
-empty list) and the initial continuation |let k0 s v = v|.
-
-We can now re-write our |param_code2| example of the previous section
-as
+Finally, |runM| runs our monad, that is, performs the computation of
+the monad and get the result, which in our case is the code
+expression. We run the monad by passing it the initial state and the
+initial continuation |k0|. We can now re-write our |param_code2|
+example of the previous section as
 
@@
                 ret (plus ce (plus x ce)))
-  and run m = m [] k0 
-  in .<fun x y -> .~(run (gen .<x>. .<y>.))>.
+  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
 \end{code}
+% param_code3 plus one;;
 
 That does not seem like the improvement. But here we can introduce the
-do notation, which is patterned after the do-namtion of Haskell.
-\oleg{Explain our mdo briefly? Refer to the message on the
-  Ocaml-list?. Refer to camlp4 and note the implementation}
+do notation (Appendix A), which is patterned after the do-notation of Haskell.
 
@@
                       ret (plus ce (plus x ce)) }
-  and run m = m [] k0 
-  in .<fun x y -> .~(run (gen .<x>. .<y>.))>.
+  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
 \end{code}
@@
 
+Note that the monad is similar to the one used in \cite{KiselyovTaha}.
+However, the latter work used only |retN| of all monadic operations,
+and used fixpoints (for performing iterations at the generation time).
+In this paper we do not use monadic fixpoints (because the generator
+is not recursive) but we make extensive use of monadic operations for
+generating conditional and looping operations.
+
 \section{Future work}
@@
 
+\section{Appendix A}
+The do notation for Ocaml.
+We use camlp4. Show the grammar. Explain the difficulty: although
+nice, the do notation cannot actually be expressed in a LR(n) grammar.
+Perhaps give the command line how to load the camlp4 module into
+ocaml. Point to the code.
 \end{document}

### 2005-04-01 paper/metamonads.tex
some re-writing

@@
 
+let l1 f = fun x -> mdo { t <-- x; f t}
+let l2 f = fun x y -> mdo { tx <-- x; ty <-- y; f tx ty}
+
 let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
+
+let lif test th el = ret .< if .~test then .~th else .~el >.
+let ifM test th el = fun s k ->
+  k s .< if .~(test s k0) then .~(th s k0) else .~(el s k0) >.
 \end{code}
@@
 \oleg{Should we rename lif into ifL?}
-Now we can build more complex monadic operations.
+We can write operations that generate code other than let-statements,
+e.g., conditionals: |lif| (Figure XXX). The function |lif|, albeit
+straightforward, is not as general as we wish: its argument are
+already generated pieces of code rather than monadic values. So, we
+may want to ``lift it'', see |ifM'|:
+
 \begin{code}
-let lif test th el = ret .< if .~test then .~th else .~el >.
+let ifM' test th el = mdo {
+  testc <-- test;
+  thc   <-- th;
+  elc   <-- el;
+  lif testc thc elc}
 \end{code}
 
-ifM and ifL: show example. Mention 'ifM' and `reset'. Investigating
-the connection with the delimited continuations: for future work.
+\omitnow{
+\begin{code}
+let ifM' test th el = bind test (fun testc ->
+                      bind th   (fun thc ->
+                      bind el   (fun elc ->
+		      lif testc thc elc)));;
+\end{code}
+}
 
-What is state and why we need it: making aspects interleave: use det
-as an example.
+We can define functions |l1|, |l2|, |l3| (analogues of |liftM|,
+|liftM2|, |liftM3| of Haskell) to make such a lifting generic.
+We soon notice the need in another |ifM| function, with tha same
+interface (see ifM in Fig. XXX). The difference between them is
+apparent from the following:
+
+\begin{code}
+let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
+                           (ret .<None>.)
+ in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
+;;
+.<fun a_1 ->
+   fun i_2 ->
+    let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.
+\end{code}
+
+whereas if we replace |ifM'| with |ifM|, the result will be
+
+\omitnow{
+\begin{code}
+let gen a i = ifM (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
+                           (ret .<None>.)
+ in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
+;;
+\end{code}
+}
+\begin{code}
+.<fun a_1 ->
+   fun i_2 ->
+    if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.
+\end{code}
+
+The difference is apparent: in the code with |ifM'|, the let-insertion
+happend \emph{before} the if-expression, that is, before the test that
+the index |i| is positive. That is, if |i| turned out
+negative, |a.(i)| will generate the out-of-bound array access
+error. On the other hand, the code with |ifM| accesses the array only
+when we have verified that the index is non-negative. This example
+makes it clear that the code generation (such as the one in |retN|) is 
+truly an effect and we have to be clear about the sequencing of
+effects, when generating control constrcutions such as conditionals.
+The form |ifM| handles the such effects correctly. In our code, we
+need similar operators for other Ocaml control forms: for generating
+case-matching statements and for- and while-loops. To make writing
+such code general and independent of the implementation of the monad,
+we define a function XXX. As the name implies, we do use control
+effect implemented via essentially delimited continuations (cite?).
 
@@
 aspects).
+What is state and why we need it: making aspects interleave: use det
+as an example. State as an open record. List of polymorphic variants
+appeared to be the easiest way to implement such a union, in a pure
+functional way.
 

### 2005-04-01 paper/metamonads.tex
Update the introduction.  Almost done first draft of the intro.

@@
-\documentclass[11pt]{llncs}
+\documentclass{llncs}
 
@@
 \usepackage{comment}
-\usepackage{amsmath}
+\usepackage{amsmath,amssymb}
 \usepackage{amstext}
@@
 
+% \usepackage[backref,colorlinks,bookmarks=true]{hyperref}
+
 % Reduce list spacing
@@
 \newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
+\newcommand{\jacques}[1]{{\it [Jacques says: #1]}}
 
@@
 modular design \cite{journals/cacm/parnas72a}, yet 
-have our code be very efficient.  Gaussian elimination, as it is 
-well-known, already contains many different aspects \cite{carette04},
-was chosen as a representative example.
+have our code be very efficient.  Gaussian elimination, as it is a  
+well-known algorithm, already contains many different aspects 
+\cite{carette04}, was chosen as a representative example.
 
@@
 which provide strong guarantees about the soundness of the resulting
-code.  MetaOCaml (cite website?) \cite{CTHL03} in particular stands
-out: not only it guarantees the absense of problems such as
-accidental variable capture -- so-called hygiene cite? \oleg{should we
-  also mention automatic well-formedness}, MetaOcaml also
+code.  MetaOCaml \cite{CTHL03,metaocaml-org} in particular stands
+out: not only it guarantees well-formedness and the absense of problems 
+such as accidental variable capture \cite{HygienicMacros}
+\jacques{is this the citation you meant?}
+MetaOcaml also
 statically ensures that both the code generator and the generated code
@@
 In MetaOCaml, generated code cannot be examined and optimized at all. It must
-be generated just right.
-
-- The FFT paper
-showed one way: abstract interpretation. But more problems remain
-(cite Padua et al paper at Haskell Workshop04): generating names in
-the presence of `if', 'while' etc. control operators (simple problem
-of generating names solved in the FFT paper); making CPS code clear
-(note that many authors are afraid of CPS code because it quickly
-becomes unreadable; we need CPS for name generation); compositionality
-of various (often interleaving: explain: determinant computation
-requires attention at several places) aspects and expressing
-dependencies among them; expressing more static guarantees about the
-generated code (e.g., not attempting full (fractional) division for
-the integer domain; generating full-pivoting code for the rational
-domain without |better_than| operation). In short: can we achieve all
-of the abstractions needed for the flexible Gaussian Elimination? Can
-we eliminate all of that abstraction's overhead in the generated code
-(without using intensional code analysis -- that is, maintaining the
-guarantees of MetaOCaml and using pure generative approach).
+be generated just right.  For simple code, this is relatively easy
+(see \cite{TahaThesis} for many examples).  For more complex examples,
+new techniques are necessary.  \cite{KiselyovTaha} introduces one
+such tool: abstract interpretation.  But more problems remain:
+\jacques{can't find this 'Padua' paper at the 2004 Haskell Workshop 
+you refer to}: 
+\begin{itemize}
+	\item Generating names in the presence of ``if'', ``while'', and other
+		control operators. \cite{KiselyovTaha} showed how to solve
+		this problem in simple cases.
+	\item Making continuation-passing style (CPS) code clear.  Many
+		authors understandably shy away from CPS code as it quickly
+		becomes unreadable.  But this is needed for proper name
+		generation.
+	\item Compositionality of various interleaving aspects, as well
+		as expressing dependencies among them.  For example, if we
+		choose to compute the determinant of a matrix while performing
+		Gaussian Elimination, this requires inserting code in 
+		several specific places in the algorithm.  The method to
+		compute this determinant crucially depends on whether 
+		a division-based or a fraction-free Gaussian Elimination is
+		used.
+	\item As mentionned above, there are dependencies between
+		certain algorithmic choices and properties of the
+		underlying domain, and we would like to statically guarantee
+		that inappropriate combinations cannot be chosen.  For instance,
+		one should not attempt to use full division when dealing
+		with matrices of exact integers, nor is it worthwhile to use
+		full pivoting on a matrix of rational integers ($\mathbb Q $).
+\end{itemize}
+The question we want to answer is:
+Can we achieve all of the abstractions needed for a flexible 
+Gaussian Elimination, and can
+we eliminate all of that abstraction's overhead in the generated code?
+Of course, we want to do this without using
+intensional code analysis -- that is, maintaining the
+guarantees of MetaOCaml and using a purely generative approach.
 
@@
 
-\oleg{explictly enumerate our contributions, not foiregtting the mdo
-  notation. The subsubsection that outlines what the other sections
+This work makes the following contributions:
+\begin{itemize}
+	\item mdo notation
+	\item use of monads for meta-programming
+	\item use of generative functors
+	\item use of type constraints to ensure correctness
+\end{itemize}
+\oleg{The subsubsection that outlines what the other sections
   are for.}

### 2005-04-01 paper/metamonads.tex
Finish first draft of complete introduction.

@@
 There are a number of projects that have attempted to address these issues
-\cite{Czarnecki,Veldhuizen:1998:ISCOPE,Veldhuizen:1998:OO98,musser94algorithmoriented,BOOST,POOMA,ATLAS}, 
+\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser94algorithmoriented,BOOST,POOMA,ATLAS}, 
 and they have generally succeeded in generating
@@
 
-[Note why it is important to maintain generativity: stronger
-equational theory: Walid's paper. Also, intensional code analysis
-(e.g., SPIRAL project) essentially requires one to insert an
-optimizing compiler into the code generating system. That is extremely
-complex, quite error-prone, and very difficult to ascertain the
-correctness.]
+Maintaining generativity is not just an interesting exercise:
+one gets a stronger equational theory \cite{Taha2000}, and avoids
+the danger of creating unsoundness \cite{TahaThesis}.  It should
+also be remarked that in general, intensional code analysis
+\cite{Pueschel:05,Kennedy01Telescoping,dongarra7,Veldhuizen:2004} 
+essentially requires
+one to insert both an optimizing compiler and an automated theorem
+proving system into the code generating system.  While this is
+potentially extremely powerful and an exciting area of research,
+it is also extremely complex, which means that it is currently more
+error-prone and difficult to ascertain the correctness of the 
+resulting code.
 
@@
 \begin{itemize}
+	\item orderly aspect weaving
 	\item mdo notation
@@
 \end{itemize}
-\oleg{The subsubsection that outlines what the other sections
-  are for.}
 
-\section{CPS and the problem of generating names}
+The rest of this paper is structured as follows:
+The next section introduces code generation in MetaOCaml, the problem
+of name generation, and how continuation-passing style is a general
+tool for solving that problem.  Section~\ref{monadicnotation} introduces
+the monad that greatly simplifies the task of dealing with
+possibly effect-inducing CPS code combinators.  Section~\ref{functors}
+shows how one can use the advanced module system of OCaml, namely its
+parametrized modules, to encode all of the aspects of the Gaussian
+Elimination algorithm family in completely separate, independent modules.
+We then briefly discuss related work in section~\ref{related}, outline
+some work still to do in section~\ref{future}, and finally draw some 
+conclusions in section~\ref{conclusion}.  As an appendix, we show
+the details of the syntax extension for monadic programming in Ocaml
+and MetaOCaml.
+
+The first author wishes to thank Cristiano Calgano for his help in
+adapting camlp4 for use with MetaOCaml.
+
+\section{CPS and the problem of generating names}\label{CPS}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% And now, a better -- and shorter -- write-up.
+
+The essence of our approach is assembling the code of the program by
+combining instances of various aspects together. In the case of
+Gaussian Elimination, the aspects are pivot determination, swapping of
+rows and columns, determinant tracking, etc. A primitive code
+generator yields a code expression.  A combinator takes (complete and
+well-typed) code expressions and combines them into a composite expression
+(which is too statically guaranteed to be well-formed and well-typed).
+
+We will be using MetaOcaml, which, as the instance of a multi-stage
+programming system \cite{TahaThesis}, provides exactly the needed
+features: to construct a code expression, to combine them, and to
+execute them.
+
+For example, the following is probably the simplest code generator:
+\begin{code}
+let one = .<1>.
+\end{code}
+
+(in the floating-point instance of that aspect, we would have
+|.<1.0>.| instead). We use MetaOCaml brackets |.<...>.| to generate
+code expressions -- in other words, to construct future-stage
+computations. The simplest combination is inlining:
+\begin{code}
+let plus x y = .<.~x + .~y>.
+\end{code}
+Here we see the second component of MetaOCaml -- escapes |.~|. They
+let us perform an immediate code generating computation while we are
+building the future-stage computation (code expression). In the
+example above, the immediate computation |.~x| is trivial -- obtaining
+the value of the variable |x|, which should be a code expression, and
+inlining it. The following shows an example of using the simple
+generators:
+
+\begin{code}
+let simplest_code =
+  let gen x y = plus x (plus y one) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+
+Here, the immediate computation |gen .<x>. .<y>.| is less trivial --
+it includes the evaluation of the function |gen|, which in turn
+applies the function |plus|. The function receives code expressions
+|.<x>.| and |.<y>.| as arguments. At the generating stage, we can
+manipulate code expressions as values. The function |gen| returns a
+code expression, which will be inlined in the place of the
+escape. MetaOCaml can print out code expressions, so we can see the
+final generated code:\\
+  |.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|
+
+We must point out that |gen x y| looks like a combinator; its body 
+involves several applications of the function |plus|. When
+we look at the generated code however, we see no traces of function
+|plus| and no such administrative applications. Instead, we see the
+fully inlined expression. These administative applications are done at
+the code generation stage.
+
+The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
+execute the code expression: |.! simplest_code| is a function of two
+integers, which we can apply: |(.! simplest_code) 1 2|. The original
+|simplest_code| is not the function on integers -- it is a code
+expression.
+
+To see the benefit of code generation, we notice that we can easily
+parameterize our code:
+
+\begin{code}
+let simplest_param_code plus one =
+  let gen x y = plus x (plus y one) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+
+and use it to generate code that operates on integers, floating point
+numbers or booleans -- in general, any domain that implements |plus|
+and |one|:
+\begin{code}
+let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
+  simplest_param_code plus one
+let plus x y = .<.~x || .~y>. and one = .<true>. in
+  simplest_param_code plus one
+\end{code}
+Running the former expression yields the function on |float|s, whereas
+the latter expression is the code expression for a bollean function.
+
+We can see the separation of concerns, namely of that for domain
+operations.
+
+Let us consider a more complex expression:
+\begin{code}
+let param_code1 plus one =
+  let gen x y = plus (plus y one) (plus x (plus y one)) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+
+We notice two occurrences of |plus y one|. Depending on the
+implementation of |plus| that may be quite a complex computation, and
+so we would rather not do it twice. We may be tempted to rely on the
+compiler's common-subexpression elimination optimization. We would
+rather not relied on a ``sufficiently smart compiler'': when the
+generated code is very complex, the compiler may overlook common
+subexpressions. The subexpressions may occur in such an imperative
+context that the compiler might not be able to determine if lifting
+them is sound. So, being conservative, the optimizer will leave the
+duplicates as they are. We may attempt to write
+
+\begin{code}
+let param_code1 plus one =
+  let gen x y = 
+     let ce = (plus y one) in  plus ce (plus x ce) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+
+However, the result of |param_code1 plus one|, which prints as\\
+|.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|,
+still exhibits duplicate sub-expressions. Our |let|-insertion
+optimization saved the computation on the generating stage. 
+We need a combinator that inserts the |let| expression in the
+generat\emph{ed} code. We need a combinator |letgen| to be used
+as\\|let ce = letgen (plus y one) in plus ce (plus x ce)| 
+yielding the code like |.<let t = y + 1 in t + (x + t)>.|
+But that seems impossible because |letgen exp| has to generate
+the expression\\|.<let t = exp in body>.| but |letgen| does not have
+the |body| yet. The body needs that temporary identifier |.<t>.| that
+is supposed to be the result of |letgen| itself. 
+Certainly |letgen| cannot generate only a part of a let-expression,
+without the |body|.  All generated expressions in
+MetaOCaml are well-formed and complete.
+
+The key is to use continuation-passing style (CPS). It was first
+pointed out by \cite{Bondorf:92} in the context
+of partial evaluation, and extensively used by \cite{KiselyovTaha} for code
+generation.
+
+\begin{code}
+let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
+let param_code2 plus one =
+  let gen x y k = letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
+  and k0 x = x
+  in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
+\end{code}
+
+Now, |param_code2 plus one| gives us the desired code\\
+|.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.|.
+
+\section{Monadic notation, making CPS code clear}\label{monadicnotation}
+
+Comparing the code that did let-insertion at the generating stage\\
+|let ce = (plus y one) in  plus ce (plus x ce)|\\
+with the corresponding code inserting let at the generated code stage\\
+|letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))|\\
+clearly shows the difference between the direct-style and CPS code.
+What was |let ce = init in ...| in direct style became
+|init' (fun ce -> ...)| in CPS. For one thing, |let| became
+inverted. For another, what used to be an expression that yields
+value, |init|, now became an expression that takes an extra argument,
+the continuation, and invokes it. The differences look negligible in
+the above example. In larger expressions with many let-forms, the
+number of paranetheses around |fun| increases, the need to add and
+then invoke the |k|, continuation argument, become annoying. The
+inconvenience is great enough for some people to explicitly avoid CPS
+or claim that numerical programmers (our users) cannot or will not
+program in CPS. Clearly a better notation is needed.
+
+The |do|-notation of Haskell (cite report?) shows that it is possible
+to write essentially CPS code in the conventional-looking style. The
+|do| notation is the notation for writing monadic code (cite Moggi?).
+The benefit of monadic notation is not that it can represent CPS (cite
+Filinski), but it helps in composability by offering to add different
+layers of effects (state, exception, non-determinism, etc) to the
+basic monad (cite Hudak, Peyton-Jones).
+
+A monad (cite Moggi?) is an abstract datatype representing
+computations that yield a value and may have an \emph{effect}.
+The datatype must have at least two operations, |return| to build
+trivial effect-less computations and |bind| for combining
+computations. These operations must satisfy \emph{monadic laws}:
+|return| being the left and the right unit of |bind| and |bind| being
+associative. Figure XXX defines the monad in the present paper
+and shows its implementation.
+
+\begin{figure*}
+\begin{code}
+type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
+let ret a = fun s k -> k s a
+let bind a f = fun s k -> a s (fun s' b -> f b s' k)
+let fetch s k = k s s
+let store v s k = k (v::s) ()
+
+let k0 s v = v
+let runM m = m [] k0
+
+let l1 f = fun x -> mdo { t <-- x; f t}
+let l2 f = fun x y -> mdo { tx <-- x; ty <-- y; f tx ty}
+
+let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
+
+let lif test th el = ret .< if .~test then .~th else .~el >.
+let ifM test th el = fun s k ->
+  k s .< if .~(test s k0) then .~(th s k0) else .~(el s k0) >.
+\end{code}
+\caption{Our monad}
+\end{figure*}
+
+Our monad represents two kinds of computational effects: reading and
+writing a computation-wide state, and control effects. The latter are
+normally associated with exceptions, forking of computations, etc. --
+in general, whenever a computation ends with something other than
+invoking its natural continuation in the tail position. In our case
+however the control effects manifest themselves as code generation.
+
+In Figure XXX, the monad is implemented as a function of two
+arguments: the state (of type |s|) and the continuation. The
+continuation receives the current state, the value (of the type |v|) and
+yields the answer of the type |w|.  The monad is polymorphic over all
+these three type parameters. Other implementations are
+possible. Except the code in Figure XXX, the rest of our code treats
+the monad as a truly abstract data type. The implementation of basic
+monadic operations |ret| and |bind| are conventional. It is easy to
+see that the monadic laws are satisfied. Other monadic operations
+construct computations that do have effects. Operations |fetch| and
+|store v| construct computations that read and write the state. In our
+case the state is a list (of polymorphic variants), which models an
+open discriminated union, as we shall see later.
+
+The operation |retN a| is the let-insertion operation (whose simpler
+version was called |letgen| before). It is the first computation with
+a control effect: indeed, the result of |retN a| is \emph{not} the
+result of invoking its continuation |k|. Rather, its result is a |let|
+code expression. Mention similarity with reset?
+
+Finally, |runM| runs our monad, that is, performs the computation of
+the monad and get the result, which in our case is the code
+expression. We run the monad by passing it the initial state and the
+initial continuation |k0|. We can now re-write our |param_code2|
+example of the previous section as
+
+\begin{code}
+let param_code3 plus one =
+  let gen x y = bind (retN (plus y one)) (fun ce -> 
+                ret (plus ce (plus x ce)))
+  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
+\end{code}
+% param_code3 plus one;;
+
+That does not seem like the improvement. But here we can introduce the
+do notation (Appendix A), which is patterned after the do-notation of Haskell.
+
+\begin{code}
+let param_code4 plus one =
+  let gen x y = mdo { ce <-- retN (plus y one);
+                      ret (plus ce (plus x ce)) }
+  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
+\end{code}
+
+The function |param_code4| is totally equivalent to |param_code3| --
+the |mdo|-notation is just the syntactic sugar. In fact, the camlp4
+preprocessor (with our extension) will transform the |param_code4|
+into |param_code3|. And yet, |param_code4| looks far more
+conventional, as if it were indeed in direct style.
+
+\oleg{Should we rename lif into ifL?}
+We can write operations that generate code other than let-statements,
+e.g., conditionals: |lif| (Figure XXX). The function |lif|, albeit
+straightforward, is not as general as we wish: its argument are
+already generated pieces of code rather than monadic values. So, we
+may want to ``lift it'', see |ifM'|:
+
+\begin{code}
+let ifM' test th el = mdo {
+  testc <-- test;
+  thc   <-- th;
+  elc   <-- el;
+  lif testc thc elc}
+\end{code}
+
+\omitnow{
+\begin{code}
+let ifM' test th el = bind test (fun testc ->
+                      bind th   (fun thc ->
+                      bind el   (fun elc ->
+		      lif testc thc elc)));;
+\end{code}
+}
+
+We can define functions |l1|, |l2|, |l3| (analogues of |liftM|,
+|liftM2|, |liftM3| of Haskell) to make such a lifting generic.
+We soon notice the need in another |ifM| function, with tha same
+interface (see ifM in Fig. XXX). The difference between them is
+apparent from the following:
+
+\begin{code}
+let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
+                           (ret .<None>.)
+ in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
+;;
+.<fun a_1 ->
+   fun i_2 ->
+    let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.
+\end{code}
+
+whereas if we replace |ifM'| with |ifM|, the result will be
+
+\omitnow{
+\begin{code}
+let gen a i = ifM (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
+                           (ret .<None>.)
+ in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
+;;
+\end{code}
+}
+\begin{code}
+.<fun a_1 ->
+   fun i_2 ->
+    if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.
+\end{code}
+
+The difference is apparent: in the code with |ifM'|, the let-insertion
+happend \emph{before} the if-expression, that is, before the test that
+the index |i| is positive. That is, if |i| turned out
+negative, |a.(i)| will generate the out-of-bound array access
+error. On the other hand, the code with |ifM| accesses the array only
+when we have verified that the index is non-negative. This example
+makes it clear that the code generation (such as the one in |retN|) is 
+truly an effect and we have to be clear about the sequencing of
+effects, when generating control constrcutions such as conditionals.
+The form |ifM| handles the such effects correctly. In our code, we
+need similar operators for other Ocaml control forms: for generating
+case-matching statements and for- and while-loops. To make writing
+such code general and independent of the implementation of the monad,
+we define a function XXX. As the name implies, we do use control
+effect implemented via essentially delimited continuations (cite?).
+
+\section{Functors}\label{functors}
+
+Simple parameterization: by the domain, by the container
+
+Functors and the generated return type: the Out structure.
+
+Making state modular and composable: several aspects can contribute
+to the state, without interference (show: Rank, Det, Permutation
+aspects).
+What is state and why we need it: making aspects interleave: use det
+as an example. State as an open record. List of polymorphic variants
+appeared to be the easiest way to implement such a union, in a pure
+functional way.
+
+Expressing domain constraints in functors (Div and integer domain).
+Show the type error that occurs in improper instantiation.
+
+Show several examples (e.g., code with NullPivot and nodet: no traces
+of either det nor pivoting). Then the code with RowPivoting, Det and
+float domain.
+
+\section{Related work}\label{related}
+
+Note that the monad is similar to the one used in \cite{KiselyovTaha}.
+However, the latter work used only |retN| of all monadic operations,
+and used fixpoints (for performing iterations at the generation time).
+In this paper we do not use monadic fixpoints (because the generator
+is not recursive) but we make extensive use of monadic operations for
+generating conditional and looping operations.
+
+\section{Future work}\label{future}
+connections with delimited continuations: making notation
+more direct-style and potentially clearer.
+
+More aspects that can be handled: Input variations (augmented
+matrices). Fit into the larger program family, where there are
+more aspects still.
+
+\section{Conclusion}\label{conclusion}
+
+Mention the mdo notation. Mention it into the introduction.
+
+\bibliography{metamonads}
+\bibliographystyle{plain}
+
+\section{Appendix A}
+The do notation for Ocaml.
+We use camlp4. Show the grammar. Explain the difficulty: although
+nice, the do notation cannot actually be expressed in a LR(n) grammar.
+Perhaps give the command line how to load the camlp4 module into
+ocaml. Point to the code.
+\end{document}
+
 \omitnow{
@@
 }
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% And now, a better -- and shorter -- write-up.
-
-The essence of our approach is assembling the code of the program by
-combining instances of various aspects together. In the case of
-Gaussian Elimination, the aspects are pivot determination, swapping of
-rows and columns, determinant tracking, etc. A primitive code
-generator yields a code expression.  A combinator takes (complete and
-well-typed) code expressions and combines them into a composite expression
-(which is too statically guaranteed to be well-formed and well-typed).
-
-We will be using MetaOcaml, which, as the instance of a multi-stage
-programming system \cite{TahaThesis}, provides exactly the needed
-features: to construct a code expression, to combine them, and to
-execute them.
-
-For example, the following is probably the simplest code generator:
-\begin{code}
-let one = .<1>.
-\end{code}
-
-(in the floating-point instance of that aspect, we would have
-|.<1.0>.| instead). We use MetaOCaml brackets |.<...>.| to generate
-code expressions -- in other words, to construct future-stage
-computations. The simplest combination is inlining:
-\begin{code}
-let plus x y = .<.~x + .~y>.
-\end{code}
-Here we see the second component of MetaOCaml -- escapes |.~|. They
-let us perform an immediate code generating computation while we are
-building the future-stage computation (code expression). In the
-example above, the immediate computation |.~x| is trivial -- obtaining
-the value of the variable |x|, which should be a code expression, and
-inlining it. The following shows an example of using the simple
-generators:
-
-\begin{code}
-let simplest_code =
-  let gen x y = plus x (plus y one) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-\end{code}
-
-Here, the immediate computation |gen .<x>. .<y>.| is less trivial --
-it includes the evaluation of the function |gen|, which in turn
-applies the function |plus|. The function receives code expressions
-|.<x>.| and |.<y>.| as arguments. At the generating stage, we can
-manipulate code expressions as values. The function |gen| returns a
-code expression, which will be inlined in the place of the
-escape. MetaOCaml can print out code expressions, so we can see the
-final generated code:\\
-  |.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|
-
-We must point out that |gen x y| looks like a combinator; its body 
-involves several applications of the function |plus|. When
-we look at the generated code however, we see no traces of function
-|plus| and no such administrative applications. Instead, we see the
-fully inlined expression. These administative applications are done at
-the code generation stage.
-
-The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
-execute the code expression: |.! simplest_code| is a function of two
-integers, which we can apply: |(.! simplest_code) 1 2|. The original
-|simplest_code| is not the function on integers -- it is a code
-expression.
-
-To see the benefit of code generation, we notice that we can easily
-parameterize our code:
-
-\begin{code}
-let simplest_param_code plus one =
-  let gen x y = plus x (plus y one) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-\end{code}
-
-and use it to generate code that operates on integers, floating point
-numbers or booleans -- in general, any domain that implements |plus|
-and |one|:
-\begin{code}
-let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
-  simplest_param_code plus one
-let plus x y = .<.~x || .~y>. and one = .<true>. in
-  simplest_param_code plus one
-\end{code}
-Running the former expression yields the function on |float|s, whereas
-the latter expression is the code expression for a bollean function.
-
-We can see the separation of concerns, namely of that for domain
-operations.
-
-Let us consider a more complex expression:
-\begin{code}
-let param_code1 plus one =
-  let gen x y = plus (plus y one) (plus x (plus y one)) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-\end{code}
-
-We notice two occurrences of |plus y one|. Depending on the
-implementation of |plus| that may be quite a complex computation, and
-so we would rather not do it twice. We may be tempted to rely on the
-compiler's common-subexpression elimination optimization. We would
-rather not relied on a ``sufficiently smart compiler'': when the
-generated code is very complex, the compiler may overlook common
-subexpressions. The subexpressions may occur in such an imperative
-context that the compiler might not be able to determine if lifting
-them is sound. So, being conservative, the optimizer will leave the
-duplicates as they are. We may attempt to write
-
-\begin{code}
-let param_code1 plus one =
-  let gen x y = 
-     let ce = (plus y one) in  plus ce (plus x ce) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-\end{code}
-
-However, the result of |param_code1 plus one|, which prints as\\
-|.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|,
-still exhibits duplicate sub-expressions. Our |let|-insertion
-optimization saved the computation on the generating stage. 
-We need a combinator that inserts the |let| expression in the
-generat\emph{ed} code. We need a combinator |letgen| to be used
-as\\|let ce = letgen (plus y one) in plus ce (plus x ce)| 
-yielding the code like |.<let t = y + 1 in t + (x + t)>.|
-But that seems impossible because |letgen exp| has to generate
-the expression\\|.<let t = exp in body>.| but |letgen| does not have
-the |body| yet. The body needs that temporary identifier |.<t>.| that
-is supposed to be the result of |letgen| itself. 
-Certainly |letgen| cannot generate only a part of a let-expression,
-without the |body|.  All generated expressions in
-MetaOCaml are well-formed and complete.
-
-The key is to use continuation-passing style (CPS). It was first
-pointed out by \cite{Bondorf:92} in the context
-of partial evaluation, and extensively used by \cite{KiselyovTaha} for code
-generation.
-
-\begin{code}
-let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
-let param_code2 plus one =
-  let gen x y k = letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
-  and k0 x = x
-  in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
-\end{code}
-
-Now, |param_code2 plus one| gives us the desired code\\
-|.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.|.
-
-\section{Monadic notation, making CPS code clear}
-
-Comparing the code that did let-insertion at the generating stage\\
-|let ce = (plus y one) in  plus ce (plus x ce)|\\
-with the corresponding code inserting let at the generated code stage\\
-|letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))|\\
-clearly shows the difference between the direct-style and CPS code.
-What was |let ce = init in ...| in direct style became
-|init' (fun ce -> ...)| in CPS. For one thing, |let| became
-inverted. For another, what used to be an expression that yields
-value, |init|, now became an expression that takes an extra argument,
-the continuation, and invokes it. The differences look negligible in
-the above example. In larger expressions with many let-forms, the
-number of paranetheses around |fun| increases, the need to add and
-then invoke the |k|, continuation argument, become annoying. The
-inconvenience is great enough for some people to explicitly avoid CPS
-or claim that numerical programmers (our users) cannot or will not
-program in CPS. Clearly a better notation is needed.
-
-The |do|-notation of Haskell (cite report?) shows that it is possible
-to write essentially CPS code in the conventional-looking style. The
-|do| notation is the notation for writing monadic code (cite Moggi?).
-The benefit of monadic notation is not that it can represent CPS (cite
-Filinski), but it helps in composability by offering to add different
-layers of effects (state, exception, non-determinism, etc) to the
-basic monad (cite Hudak, Peyton-Jones).
-
-A monad (cite Moggi?) is an abstract datatype representing
-computations that yield a value and may have an \emph{effect}.
-The datatype must have at least two operations, |return| to build
-trivial effect-less computations and |bind| for combining
-computations. These operations must satisfy \emph{monadic laws}:
-|return| being the left and the right unit of |bind| and |bind| being
-associative. Figure XXX defines the monad in the present paper
-and shows its implementation.
-
-\begin{figure*}
-\begin{code}
-type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
-let ret a = fun s k -> k s a
-let bind a f = fun s k -> a s (fun s' b -> f b s' k)
-let fetch s k = k s s
-let store v s k = k (v::s) ()
-
-let k0 s v = v
-let runM m = m [] k0
-
-let l1 f = fun x -> mdo { t <-- x; f t}
-let l2 f = fun x y -> mdo { tx <-- x; ty <-- y; f tx ty}
-
-let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
-
-let lif test th el = ret .< if .~test then .~th else .~el >.
-let ifM test th el = fun s k ->
-  k s .< if .~(test s k0) then .~(th s k0) else .~(el s k0) >.
-\end{code}
-\caption{Our monad}
-\end{figure*}
-
-Our monad represents two kinds of computational effects: reading and
-writing a computation-wide state, and control effects. The latter are
-normally associated with exceptions, forking of computations, etc. --
-in general, whenever a computation ends with something other than
-invoking its natural continuation in the tail position. In our case
-however the control effects manifest themselves as code generation.
-
-In Figure XXX, the monad is implemented as a function of two
-arguments: the state (of type |s|) and the continuation. The
-continuation receives the current state, the value (of the type |v|) and
-yields the answer of the type |w|.  The monad is polymorphic over all
-these three type parameters. Other implementations are
-possible. Except the code in Figure XXX, the rest of our code treats
-the monad as a truly abstract data type. The implementation of basic
-monadic operations |ret| and |bind| are conventional. It is easy to
-see that the monadic laws are satisfied. Other monadic operations
-construct computations that do have effects. Operations |fetch| and
-|store v| construct computations that read and write the state. In our
-case the state is a list (of polymorphic variants), which models an
-open discriminated union, as we shall see later.
-
-The operation |retN a| is the let-insertion operation (whose simpler
-version was called |letgen| before). It is the first computation with
-a control effect: indeed, the result of |retN a| is \emph{not} the
-result of invoking its continuation |k|. Rather, its result is a |let|
-code expression. Mention similarity with reset?
-
-Finally, |runM| runs our monad, that is, performs the computation of
-the monad and get the result, which in our case is the code
-expression. We run the monad by passing it the initial state and the
-initial continuation |k0|. We can now re-write our |param_code2|
-example of the previous section as
-
-\begin{code}
-let param_code3 plus one =
-  let gen x y = bind (retN (plus y one)) (fun ce -> 
-                ret (plus ce (plus x ce)))
-  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
-\end{code}
-% param_code3 plus one;;
-
-That does not seem like the improvement. But here we can introduce the
-do notation (Appendix A), which is patterned after the do-notation of Haskell.
-
-\begin{code}
-let param_code4 plus one =
-  let gen x y = mdo { ce <-- retN (plus y one);
-                      ret (plus ce (plus x ce)) }
-  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
-\end{code}
-
-The function |param_code4| is totally equivalent to |param_code3| --
-the |mdo|-notation is just the syntactic sugar. In fact, the camlp4
-preprocessor (with our extension) will transform the |param_code4|
-into |param_code3|. And yet, |param_code4| looks far more
-conventional, as if it were indeed in direct style.
-
-\oleg{Should we rename lif into ifL?}
-We can write operations that generate code other than let-statements,
-e.g., conditionals: |lif| (Figure XXX). The function |lif|, albeit
-straightforward, is not as general as we wish: its argument are
-already generated pieces of code rather than monadic values. So, we
-may want to ``lift it'', see |ifM'|:
-
-\begin{code}
-let ifM' test th el = mdo {
-  testc <-- test;
-  thc   <-- th;
-  elc   <-- el;
-  lif testc thc elc}
-\end{code}
-
-\omitnow{
-\begin{code}
-let ifM' test th el = bind test (fun testc ->
-                      bind th   (fun thc ->
-                      bind el   (fun elc ->
-		      lif testc thc elc)));;
-\end{code}
-}
-
-We can define functions |l1|, |l2|, |l3| (analogues of |liftM|,
-|liftM2|, |liftM3| of Haskell) to make such a lifting generic.
-We soon notice the need in another |ifM| function, with tha same
-interface (see ifM in Fig. XXX). The difference between them is
-apparent from the following:
-
-\begin{code}
-let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
-                           (ret .<None>.)
- in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
-;;
-.<fun a_1 ->
-   fun i_2 ->
-    let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.
-\end{code}
-
-whereas if we replace |ifM'| with |ifM|, the result will be
-
-\omitnow{
-\begin{code}
-let gen a i = ifM (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
-                           (ret .<None>.)
- in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
-;;
-\end{code}
-}
-\begin{code}
-.<fun a_1 ->
-   fun i_2 ->
-    if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.
-\end{code}
-
-The difference is apparent: in the code with |ifM'|, the let-insertion
-happend \emph{before} the if-expression, that is, before the test that
-the index |i| is positive. That is, if |i| turned out
-negative, |a.(i)| will generate the out-of-bound array access
-error. On the other hand, the code with |ifM| accesses the array only
-when we have verified that the index is non-negative. This example
-makes it clear that the code generation (such as the one in |retN|) is 
-truly an effect and we have to be clear about the sequencing of
-effects, when generating control constrcutions such as conditionals.
-The form |ifM| handles the such effects correctly. In our code, we
-need similar operators for other Ocaml control forms: for generating
-case-matching statements and for- and while-loops. To make writing
-such code general and independent of the implementation of the monad,
-we define a function XXX. As the name implies, we do use control
-effect implemented via essentially delimited continuations (cite?).
-
-\section{Functors}
-
-Simple parameterization: by the domain, by the container
-
-Functors and the generated return type: the Out structure.
-
-Making state modular and composable: several aspects can contribute
-to the state, without interference (show: Rank, Det, Permutation
-aspects).
-What is state and why we need it: making aspects interleave: use det
-as an example. State as an open record. List of polymorphic variants
-appeared to be the easiest way to implement such a union, in a pure
-functional way.
-
-Expressing domain constraints in functors (Div and integer domain).
-Show the type error that occurs in improper instantiation.
-
-Show several examples (e.g., code with NullPivot and nodet: no traces
-of either det nor pivoting). Then the code with RowPivoting, Det and
-float domain.
-
-\section{Related work}
-
-Note that the monad is similar to the one used in \cite{KiselyovTaha}.
-However, the latter work used only |retN| of all monadic operations,
-and used fixpoints (for performing iterations at the generation time).
-In this paper we do not use monadic fixpoints (because the generator
-is not recursive) but we make extensive use of monadic operations for
-generating conditional and looping operations.
-
-\section{Future work}
-connections with delimited continuations: making notation
-more direct-style and potentially clearer.
-
-More aspects that can be handled: Input variations (augmented
-matrices). Fit into the larger program family, where there are
-more aspects still.
-
-\section{Conclusion}
-
-Mention the mdo notation. Mention it into the introduction.
-
-\bibliography{metamonads}
-\bibliographystyle{plain}
-
-\section{Appendix A}
-The do notation for Ocaml.
-We use camlp4. Show the grammar. Explain the difficulty: although
-nice, the do notation cannot actually be expressed in a LR(n) grammar.
-Perhaps give the command line how to load the camlp4 module into
-ocaml. Point to the code.
-\end{document}

### 2005-04-01 paper/metamonads.tex
J's minor edits in O's sections

@@
 combining instances of various aspects together. In the case of
-Gaussian Elimination, the aspects are pivot determination, swapping of
+Gaussian Elimination, the algorithmic aspects are pivot determination, 
+swapping of
 rows and columns, determinant tracking, etc. A primitive code
@@
 well-typed) code expressions and combines them into a composite expression
-(which is too statically guaranteed to be well-formed and well-typed).
+(which is also statically guaranteed to be well-formed and well-typed).
 
-We will be using MetaOcaml, which, as the instance of a multi-stage
+We will be using MetaOcaml which, as an instance of a multi-stage
 programming system \cite{TahaThesis}, provides exactly the needed
@@
 
-(in the floating-point instance of that aspect, we would have
-|.<1.0>.| instead). We use MetaOCaml brackets |.<...>.| to generate
+In the floating-point instance of that aspect, we would have
+|.<1.0>.| instead. We use MetaOCaml brackets |.<...>.| to generate
 code expressions -- in other words, to construct future-stage
@@
 |.<x>.| and |.<y>.| as arguments. At the generating stage, we can
-manipulate code expressions as values. The function |gen| returns a
+manipulate code expressions as (opaque) values. The function |gen| returns a
 code expression, which will be inlined in the place of the
@@
 involves several applications of the function |plus|. When
-we look at the generated code however, we see no traces of function
+we look at the generated code however, we see no traces of the function
 |plus| and no such administrative applications. Instead, we see the
@@
 
-and use it to generate code that operates on integers, floating point
+\noindent and use it to generate code that operates on integers, floating point
 numbers or booleans -- in general, any domain that implements |plus|
@@
 Running the former expression yields the function on |float|s, whereas
-the latter expression is the code expression for a bollean function.
+the latter expression is the code expression for a boolean function.
 
-We can see the separation of concerns, namely of that for domain
+This clearly shows the separation of concerns, namely of that for domain
 operations.
@@
 compiler's common-subexpression elimination optimization. We would
-rather not relied on a ``sufficiently smart compiler'': when the
+rather not rely on a ``sufficiently smart compiler'': when the
 generated code is very complex, the compiler may overlook common
-subexpressions. The subexpressions may occur in such an imperative
-context that the compiler might not be able to determine if lifting
+subexpressions.  Or the subexpressions may occur in an imperative
+context where the compiler might not be able to determine if lifting
 them is sound. So, being conservative, the optimizer will leave the
@@
 still exhibits duplicate sub-expressions. Our |let|-insertion
-optimization saved the computation on the generating stage. 
+optimization saved the computation at the generating stage. 
 We need a combinator that inserts the |let| expression in the
@@
 But that seems impossible because |letgen exp| has to generate
-the expression\\|.<let t = exp in body>.| but |letgen| does not have
-the |body| yet. The body needs that temporary identifier |.<t>.| that
+the expression\\|.<let t = exp in body>.| although |letgen| does not have
+the |body| yet. The body needs a temporary identifier |.<t>.| that
 is supposed to be the result of |letgen| itself. 
-Certainly |letgen| cannot generate only a part of a let-expression,
-without the |body|.  All generated expressions in
+Certainly |letgen| cannot generate only part of a let-expression,
+without the |body|,  as all generated expressions in
 MetaOCaml are well-formed and complete.
 
-The key is to use continuation-passing style (CPS). It was first
+The key is to use continuation-passing style (CPS). Its benefits were first
 pointed out by \cite{Bondorf:92} in the context
@@
 inverted. For another, what used to be an expression that yields
-value, |init|, now became an expression that takes an extra argument,
+a value, |init|, now became an expression that takes an extra argument,
 the continuation, and invokes it. The differences look negligible in
 the above example. In larger expressions with many let-forms, the
-number of paranetheses around |fun| increases, the need to add and
-then invoke the |k|, continuation argument, become annoying. The
+number of parentheses around |fun| increases, the need to add and
+then invoke the |k| continuation argument become increasingly annoying. The
 inconvenience is great enough for some people to explicitly avoid CPS
@@
 
-The |do|-notation of Haskell (cite report?) shows that it is possible
-to write essentially CPS code in the conventional-looking style. The
+The |do|-notation of Haskell \cite{Haskell98Report} shows that it is possible
+to write essentially CPS code in conventional-looking style. The
 |do| notation is the notation for writing monadic code (cite Moggi?).
@@
 layers of effects (state, exception, non-determinism, etc) to the
-basic monad (cite Hudak, Peyton-Jones).
+basic monad (cite Hudak, Peyton-Jones) in a controlled way.
 
@@
 |return| being the left and the right unit of |bind| and |bind| being
-associative. Figure XXX defines the monad in the present paper
-and shows its implementation.
+associative. Figure~\ref{ourmonad} defines the monad used throughout
+the present paper and shows its implementation.
 
-\begin{figure*}
+\begin{figure*}\label{ourmonad}
 \begin{code}
@@
 
-let lif test th el = ret .< if .~test then .~th else .~el >.
+let ifL test th el = ret .< if .~test then .~th else .~el >.
 let ifM test th el = fun s k ->
@@
 
-In Figure XXX, the monad is implemented as a function of two
+In Figure~\ref{ourmonad}, the monad is implemented as a function of two
 arguments: the state (of type |s|) and the continuation. The
@@
 these three type parameters.  Other implementations are
-possible. Except the code in Figure XXX, the rest of our code treats
-the monad as a truly abstract data type. The implementation of basic
+possible. Except for the code in Figure~\ref{ourmonad}, the rest of our code
+treats the monad as a truly abstract data type. The implementation of basic
 monadic operations |ret| and |bind| are conventional. It is easy to
@@
 
-The operation |retN a| is the let-insertion operation (whose simpler
-version was called |letgen| before). It is the first computation with
+The operation |retN a| is the let-insertion operation, whose simpler
+version we called |letgen| earlier. It is the first computation with
 a control effect: indeed, the result of |retN a| is \emph{not} the
 result of invoking its continuation |k|. Rather, its result is a |let|
-code expression. Mention similarity with reset?
+code expression. \oleg{Mention similarity with reset?}
+\jacques{yes}
 
 Finally, |runM| runs our monad, that is, performs the computation of
-the monad and get the result, which in our case is the code
+the monad and gets the result, which in our case is the code
 expression. We run the monad by passing it the initial state and the
@@
 
-That does not seem like the improvement. But here we can introduce the
+That does not seem like much of an improvement. But here we can introduce the
 do notation (Appendix A), which is patterned after the do-notation of Haskell.
@@
 The function |param_code4| is totally equivalent to |param_code3| --
-the |mdo|-notation is just the syntactic sugar. In fact, the camlp4
-preprocessor (with our extension) will transform the |param_code4|
-into |param_code3|. And yet, |param_code4| looks far more
+the |mdo|-notation is just syntactic sugar. In fact, the camlp4
+preprocessor (with our extension) will transform |param_code4|
+into exactly |param_code3|. And yet, |param_code4| looks far more
 conventional, as if it were indeed in direct style.
 
-\oleg{Should we rename lif into ifL?}
 We can write operations that generate code other than let-statements,
-e.g., conditionals: |lif| (Figure XXX). The function |lif|, albeit
-straightforward, is not as general as we wish: its argument are
+e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
+albeit straightforward, is not as general as we wish: its argument are
 already generated pieces of code rather than monadic values. So, we
-may want to ``lift it'', see |ifM'|:
-
+may want to ``lift it'':
 \begin{code}
@@
   elc   <-- el;
-  lif testc thc elc}
+  ifL testc thc elc}
 \end{code}
 
-\omitnow{
-\begin{code}
-let ifM' test th el = bind test (fun testc ->
-                      bind th   (fun thc ->
-                      bind el   (fun elc ->
-		      lif testc thc elc)));;
-\end{code}
-}
-
-We can define functions |l1|, |l2|, |l3| (analogues of |liftM|,
+\noindent We can define functions |l1|, |l2|, |l3| (analogues of |liftM|,
 |liftM2|, |liftM3| of Haskell) to make such a lifting generic.
-We soon notice the need in another |ifM| function, with tha same
-interface (see ifM in Fig. XXX). The difference between them is
+We soon notice the need for another |ifM| function, with the same
+interface (see ifM in Figure~\ref{ourmonad}). The difference between them is
 apparent from the following:
-
 \begin{code}
@@
 ;;
-.<fun a_1 ->
-   fun i_2 ->
+.<fun a_1 -> fun i_2 ->
     let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.
 \end{code}
-
-whereas if we replace |ifM'| with |ifM|, the result will be
-
-\omitnow{
+\noindent whereas if we replace |ifM'| with |ifM|, the result will be
 \begin{code}
-let gen a i = ifM (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
-                           (ret .<None>.)
- in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
-;;
-\end{code}
-}
-\begin{code}
-.<fun a_1 ->
-   fun i_2 ->
+.<fun a_1 -> fun i_2 ->
     if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.
@@
 the index |i| is positive. That is, if |i| turned out
-negative, |a.(i)| will generate the out-of-bound array access
+negative, |a.(i)| will generate an out-of-bound array access
 error. On the other hand, the code with |ifM| accesses the array only
@@
 truly an effect and we have to be clear about the sequencing of
-effects, when generating control constrcutions such as conditionals.
+effects, when generating control constructions such as conditionals.
 The form |ifM| handles the such effects correctly. In our code, we
@@
 effect implemented via essentially delimited continuations (cite?).
+\jacques{yes!}
 

### 2005-04-03 paper/metamonads.tex
part of section 4 written.  slow going...

@@
 
+Now that we know that we have the appropriate basic techniques for
+writing our code generators, we need to worry about modularizing them
+appropriately.  Here we can really take advantage of the fact that
+MetaOCaml is a staged language: it is not really relevant how much
+computation happens at staging time, as long as the generated code
+is correct and efficient.  In other words, we can use any abstraction
+mechanisms we want to structure our code generators, as long as none
+of those abstractions infiltrate the generated code.
+
+While the Object-Oriented Design community has acquired an extensive
+vocabulary for describing modularity ideas, the guiding principles for
+modular designs has not changed since they were first articulated by
+Parnas~\cite{journals/cacm/parnas72a} and Dijkstra~
+\cite{EWD:EWD447}: information hiding and separation of concerns.  To
+apply these principles to the study of Gaussian Elimination, we need
+to understand what changes between different implementations, and 
+what concerns need to be addressed.  We also need to study the degree
+to which these concerns are independent.
+
+A careful study of Gaussian Elimination \cite{carette04} shows that
+(at least) the following variations occur:
+\begin{enumerate}
+	\item \textbf{Domain}: In which (algebraic) domain do the matrix
+		elements belong to.  Some implementations were very specific
+		($\mathbb{Z}, \mathbb{Q}, \mathbb{Z_p}, 
+		\mathbb{Z}_p\left[\alpha_1,\ldots,\alpha_n\right], 
+		\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$, 
+		$\mathbb{Q}\left[\alpha\right]$, and floating point numbers for 
+		example), while others were generic for elements of a field,
+		multivariate polynomials over a field, elements of a formal Ring
+		with possibly undecidable zero-equivalence, or elements of a 
+		Division Ring.  In the roughly 85 pieces of code surveyed,
+		20 different domains were encountered.
+	\item \textbf{Container}: Whether the matrix
+		was represented as an array of arrays, a one-dimensional array,
+		a hash table, a sparse matrix, etc., and
+		whether indexing was done in C or Fortran style.  Additionally,
+		if a particular representation had a special mechanism for efficient
+		row exchanges, this was sometimes used.  
+	\item \textbf{Output choices}:  Whether just the reduced matrix, or
+		additionally the rank, the determinant, and the
+		pivoting matrix were to be returned.
+		It is worthwhile noting that in the larger algorithm family,
+		routines like Maple's \texttt{LinearAlgebra:-LUDecomposition} have
+		up to $2^6 + 2^5 + 2^2 = 100$ possible outputs.
+	\item \textbf{Fraction-free}: Whether the Gaussian Elimination
+		algorithm is allowed to use unrestricted division, or only
+		exact (remainder-free) division.
+	\item \textbf{Pivoting}: Whether to use no pivoting at all, only
+		column-wise pivoting, or full pivoting.
+	\item \textbf{Augmented Matrices}: Whether we are doing GE on
+		a full matrix, or only a restricted number of columns, while
+		doing elimination on the full matrix.  Note that we currently
+		do not treat this aspect in our code.
+\end{enumerate}
+\noindent In addition to the above variations, there are two aspects that 
+recur frequently:
+\begin{enumerate}
+	\item \textbf{Length measure}:  For stability reasons
+		(whether numerical or coefficient growth), if a domain possesses
+		an appropriate length measure, this was sometimes used to choose
+		an ``optimal'' pivot for row exchanges.
+	\item \textbf{Normalization and zero-equivalence}: Whether the 
+		arithmetic operations of the domain at hand gives results in 
+		normalized form, and whether a specialized zero-equivalence 
+		routine needs to be used.
+\end{enumerate}
+\noindent These are separated out from the others are they are cross-cutting
+concerns: in the case of the length measure, then a property of the domain
+will influence the method used to do pivoting \textbf{if} pivoting is to be
+performed.
+
 Simple parameterization: by the domain, by the container

### 2005-04-03 paper/metamonads.tex
Finish comments on modularization design. Show DOMAIN and CONTAINER in full.

@@
 	\item \textbf{Domain}: In which (algebraic) domain do the matrix
-		elements belong to.  Some implementations were very specific
-		($\mathbb{Z}, \mathbb{Q}, \mathbb{Z_p}, 
+		elements belong to.  Sometimes the domains are very specific
+		($\mathbb{Z}, \mathbb{Q}, \mathbb{Z}_p, 
 		\mathbb{Z}_p\left[\alpha_1,\ldots,\alpha_n\right], 
@@
 		$\mathbb{Q}\left[\alpha\right]$, and floating point numbers for 
-		example), while others were generic for elements of a field,
+		example), while in other cases the domains were left generic,
+		namely for elements of a field,
 		multivariate polynomials over a field, elements of a formal Ring
@@
 	\item \textbf{Container}: Whether the matrix
-		was represented as an array of arrays, a one-dimensional array,
+		is represented as an array of arrays, a one-dimensional array,
 		a hash table, a sparse matrix, etc., and
-		whether indexing was done in C or Fortran style.  Additionally,
+		whether indexing is done in C or Fortran style.  Additionally,
 		if a particular representation had a special mechanism for efficient
-		row exchanges, this was sometimes used.  
+		row exchanges, this is sometimes used.  
 	\item \textbf{Output choices}:  Whether just the reduced matrix, or
 		additionally the rank, the determinant, and the
-		pivoting matrix were to be returned.
+		pivoting matrix are to be returned.
 		It is worthwhile noting that in the larger algorithm family,
@@
 		(whether numerical or coefficient growth), if a domain possesses
-		an appropriate length measure, this was sometimes used to choose
+		an appropriate length measure, this is sometimes used to choose
 		an ``optimal'' pivot for row exchanges.
@@
 
-Simple parameterization: by the domain, by the container
+The simplest parametrization is to make the domain abstract.  As it 
+turns out, we need the following to exist in our domains:  $0$, $1$,
+$+$, $*$, (unary and binary) $-$, some kind of division, normalization,
+and potentially a relative size measure.  We definitely need at least
+\emph{exact} division to exist, but if full division is available, than
+can be used as well.  Normalization is frequently trivial for domains
+where elements have unique representations, but in many applications
+we have domains (like polynomials in sum-of-product form) where this
+is not the case.  This leads us to defining a |DOMAIN| type as
+\begin{small}
+\begin{code}
+module type DOMAIN = sig
+  type v
+  type kind (* Field or Ring ? *)
+  type 'a vc = ('a,v) code
+  val zero : 'a vc
+  val one : 'a vc
+  val plus : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
+  val times : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
+  val minus : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
+  val uminus : 'a vc -> ('a vc, 's, 'w) monad
+  val div : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
+  val better_than : ('a vc -> 'a vc -> 
+      (('a,bool) code, 's, 'w) monad) option
+  val normalizerf : (('a,v -> v) code ) option
+  val normalizerg : 'a vc -> 'a vc
+end 
+\end{code}
+\end{small}
+\noindent  The first item to notice is that the types above are
+generally lifted twice: once from the value domain |v| to the code
+domain |'a vc|, and once more from values to monadic computations
+|('a vc, 's, 'w) monad|.  Since there are times where we don't want
+to redundantly walk a whole matrix to apply normalization when 
+normalization is not necessary, |normalizerf| is optional.
+
+For example, when lifted to a |DOMAIN|, the type |int| becomes
+\begin{small}
+\begin{code}
+module IntegerDomain = struct
+    type v = int
+    type kind = domain_is_ring
+    type 'a vc = ('a,v) code
+    let zero = .< 0 >.  
+    let one = .< 1 >. 
+    let plus x y = ret .<.~x + .~y>. 
+    let times x y = ret .<.~x * .~y>.
+    let minus x y = ret .<.~x - .~y>.
+    let uminus x = ret .< - .~x>.
+    let div x y = ret .<.~x / .~y>. 
+    let better_than = Some (fun x y -> ret .<abs .~x > abs .~y >. )
+    let normalizerf = None 
+    let normalizerg = fun x -> x
+end
+\end{code}
+\end{small}
+\noindent While the |DOMAIN| type may have looked daunting to some, 
+the implementation is quite straightforward.  It is equally simple to 
+make implementations for |float| and for |Num.num| (arbitrary precision
+exact rational numbers).  With scarcely more work (as these are not 
+built-in and the arithmetic must be implemented), $\mathbb{Z}_p$,
+$\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$ and more
+complex domains can be done as well.
+
+Parametrizing by the kind of container used to represent a matrix is
+almost as straightforward.  We choose to make the container parametric
+over a |DOMAIN| by making containers functors from a |DOMAIN| module
+to the actual implementation of a container.  Our containers are
+explicitly 2-dimensional, and we provide functions |dim1| and |dim2|
+to extract each.  We also need |get| and |set| functions -- these come
+in both a monadic flavour as well as a direct code-generation version.
+For convenience and efficiency, several additional functions are also
+required: a way to |copy| a container, a method (|mapper|) to map an
+optional function over the elements of a container, as well as methods
+to swap rows and columns.  The inclusion of these functions in the 
+signature of all containers makes it simpler to optimize the relevant
+functions depending on the actual representation of the container
+while not burdening the users of containers with efficiency details.
+As an example, below is an implementation of a matrix as an
+array of rows (themselves arrays).  This implementation clearly shows
+how swapping rows is more efficient than swapping columns in this
+representation.  We have not (yet) implemented the further potential
+optimization wherein if it is known that part of two rows/columns are already
+known to be equal (in fact to |Dom.zero| in the case of GE) and
+row/column exchange needs to be performed element-wise, then certain
+swaps are unecessary.
+
+\begin{small}
+\begin{code}
+module GenericArrayContainer(Dom:DOMAIN) =
+  struct
+  type contr = Dom.v array array (* Array of rows *)
+  type 'a vc = ('a,contr) code
+  type 'a vo = ('a,Dom.v) code
+  let get' x n m = .< (.~x).(.~n).(.~m) >.
+  let get x n m = ret (get' x n m)
+  let set' x n m y = .< (.~x).(.~n).(.~m) <- .~y >.
+  let set x n m y = ret (set' x n m y)
+  let dim2 x = .< Array.length .~x >.       (* number of rows *)
+  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
+  let mapper g a = match g with
+      | Some f -> .< Array.map (fun x -> Array.map .~f x) .~a >.
+      | None   -> a
+  let copy = (fun a -> .<Array.map (fun x -> Array.copy x) 
+                       (Array.copy .~a) >. )
+  (* this can be optimized with a swap_rows_from if it is known that
+     everything before that is already = Dom.zero *)
+  let swap_rows_stmt a r1 r2 =
+      .< let t = (.~a).(.~r1) in
+         begin 
+             (.~a).(.~r1) <- (.~a).(.~r2);
+             (.~a).(.~r2) <- t
+         end >.
+  let swap_cols_stmt a c1 c2 = .< 
+      for r = 0 to .~(dim2 a)-1 do
+          let t = (.~a).(r).(.~c1) in
+          begin 
+              (.~a).(r).(.~c1) <- (.~a).(r).(.~c2);
+              (.~a).(r).(.~c2) <- t
+          end
+      done  >.
+end
+\end{code}
+\end{small}
+\noindent  Implementing other containers based on flattened vectors,
+Fortran-style access, |Bigarray| or |Bigarray.Array2| is also possible.
+Naturally, for proper use of |Bigarray| with specific types, it is 
+important that the |DOMAIN| type matches -- but luckily the compiler 
+will prevent us from making such a mistake.
+
+The use of a |functor| for making a container parametric is fairly 
+straightforward.  It gets more interesting when one turns to the aspect
+of what to return from the GE algorithm.
 

### 2005-04-04 paper/metamonads.tex
continue with write up of section 4. add some comments in later sections too.

@@
 straightforward.  It gets more interesting when one turns to the aspect
-of what to return from the GE algorithm.
-
-Functors and the generated return type: the Out structure.
+of what to return from the GE algorithm.  One could create an algebraic
+data type (as was done in \cite{carette04}) to encode the various
+choices; but this is wholly unsatisfying as we know that for any single
+use, only one of the choices is ever possible, yet any routine which
+calls the generated code must deal with these unreachable options.
+Instead we use a module type with an \emph{abstract} type |res| for the
+result type; the actual result type is then made by instantiating
+this module type with different modules that decide what will be 
+output: the matrix, the matrix and the rank, the matrix and the
+determinant, the matrix, rank and determinant, and so on.  More
+precisely, below we show this type and one instantiation.
+
+\begin{small}
+\begin{code}
+module type OUTPUT = sig
+  type contr
+  type res
+  module D : DETERMINANT
+  module R : RANK
+  module P : TRACKPIVOT
+  val make_result : ('a,contr) code -> 
+    (('a,res) code,
+     [> `TDet of 'a D.lstate | `TRan of 'a R.lstate | `TPivot of 'a P.lstate]
+       list,
+     ('a,'w) code) monad
+end
+
+module OutDetRank(Dom:DOMAIN)(C: CONTAINER2D)
+    (Det : DETERMINANT with type indet = Dom.v and type outdet = Dom.v)
+    (Rank : RANK) =
+  struct
+  module Ctr = C(Dom)
+  type contr = Ctr.contr
+  type res = contr * Det.outdet * int
+  module D = Det
+  module R = Rank
+  module P = DiscardPivot
+  let make_result b = mdo {
+    det  <-- D.fin ();
+    rank <-- R.fin ();
+    ret .< ( .~b, .~det, .~rank ) >. }
+end
+\end{code}
+\end{small}
+\noindent It is worthwhile noting that while the |OUTPUT| type is
+relatively simple, instantiating it requires a rather complex functor
+with some additional type constraints.  While we will explain some of
+the details of the above code later, right now it suffices to notice
+that |res| above is a 3-tuple with type |contr * Det.outdet * int|,
+where in another instance this type is simply |contr|, and these are
+formally incomparable types.
+
+As is apparent from the output choices, several different quantities
+\emph{may} need to be tracked in a particular GE implementation.
+We therefore need to be able to conditionally generate code-variables,
+and weave in corresponding tracking code for each of these quantities.
+To be precise, we may need to keep track (independently) of the rank,
+the determinant and the permutation list.  These pieces of data then
+become part of the \emph{state} that is tracked by our monad.  To have
+all this choice when needed, and yet have our code be modular and
+composable as well as ensuring that the generated code does not 
+contain any abstraction artifacts, it is important to make this 
+state modular.  As this discussion is rather abstract, an example
+will likely illustrate our point more clearly: we will show the 
+code for tracking the determinant.
+
+\begin{small}
+\begin{code}
+module type DETERMINANT = sig
+  type indet
+  type outdet
+  type tdet = outdet ref
+  type 'a lstate
+  val decl : unit -> 
+    (unit, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
+  val upd_sign : unit -> 
+    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
+  val zero_sign : unit -> 
+    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
+  val acc : ('a,indet) code -> 
+    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
+  val get : unit ->
+    (('a,tdet) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
+  val set : ('a,indet) code -> 
+    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
+  val fin : unit -> 
+    (('a,outdet) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
+end
+\end{code}
+\end{small}
+\noindent  Determinant tracking is quite complex: we need to 
+declare that we are tracking the determinant (|decl|), have some functions
+for tracking the sign of the determinant (|upd_sign|,|zero_sign|),
+|set|, |get|, accumulate (|acc|) unsigned values for the determinant, as 
+well as returning a final value (|fin|) for the determinant.  In some cases,
+like for float matrices where we do not return the determinant, all of
+the above should generate no code at all, while for integer matrices where
+we want the determinant as part of the output, each of these need to be
+implemented.
+
+\jacques{I am stopping my description of state here -- what I have
+written above is quite dull, and not making the point very well.
+I will need to revisit this}
 
@@
 of either det nor pivoting). Then the code with RowPivoting, Det and
-float domain.
+float domain. \jacques{which would really be great if we had a lot 
+more space, but I think we're already long, so something is going to
+have to be cut.  Maybe not this, but something!}
 
@@
 
+\jacques{What do we consider \emph{related work} here?  Do we essentially gloss 
+over related work in numerical software, code generation in general,
+aspect-oriented stuff?  We don't want this section to be too long, but
+not so short that we make it seem like we're inventing more than we
+really are}
+
 \section{Future work}\label{future}
 connections with delimited continuations: making notation
-more direct-style and potentially clearer.
+more direct-style and potentially clearer.  More syntactic sugar
+would help a lot here.
 
@@
 
+Making all of the algorithmic code be completely independent of 
+MetaOCaml markup and isolate all the code generation to a select
+few 'base' modules and functors.  This should allow for simpler
+debugging, where one can choose to instantiate either a direct-style
+set of base modules for debugging, or a code-generation style set
+of base modules for generating optimized code.
+
 \section{Conclusion}\label{conclusion}
@@
 Mention the mdo notation. Mention it into the introduction.
+
+Relation with aspect-oriented code: in AspectJ, aspects are (comparatively)
+lightly typed, and are post-facto extensions of an existing piece of
+code.  Here aspects are weaved together ``from scracth'' to make up a
+piece of code/functionality.  One can understand previous work to be
+more akin to dynamically typed aspect weaving, while we have started
+investigating statically typed aspect weaving.
 

### 2005-04-04 paper/metamonads.tex
misc changes

@@
 frequently have a larger cost still.  Of course, if one is faced with
-writing 60 different implementations of the ``same'' algorithm, this
+writing 60 different implementations of the ``same'' algorithm
+\oleg{citation?}, this
 abstraction cost starts to look cheap!  Although bad programmers have
@@
 out: not only it guarantees well-formedness and the absense of problems 
-such as accidental variable capture \cite{HygienicMacros}
-\jacques{is this the citation you meant?}
-MetaOcaml also
+such as accidental variable capture \cite{HygienicMacros}. MetaOcaml also
 statically ensures that both the code generator and the generated code
@@
 generative system to re-parse the generative code to inspect it, such
-an approach rapidly degenerates into complete chaos.  In languages
+an approach rapidly degenerates into complete chaos \oleg{is there a
+  good citation?}.  In languages
 that natively offer reflection and reification (either at no cost, 
@@
 new techniques are necessary.  \cite{KiselyovTaha} introduces one
-such tool: abstract interpretation.  But more problems remain:
-\jacques{can't find this 'Padua' paper at the 2004 Haskell Workshop 
-you refer to}: 
+such tool: abstract interpretation.  But more problems remain
+\cite{Padua:MetaOcaml:04}:
 \begin{itemize}
@@
 to write essentially CPS code in conventional-looking style. The
-|do| notation is the notation for writing monadic code (cite Moggi?).
-The benefit of monadic notation is not that it can represent CPS (cite
-Filinski), but it helps in composability by offering to add different
+|do| notation is the notation for writing monadic code \cite{moggi-notions}.
+The benefit of monadic notation is not that it can represent CPS \cite{Filinski:Representing}, but it helps in composability by offering to add different
 layers of effects (state, exception, non-determinism, etc) to the
-basic monad (cite Hudak, Peyton-Jones) in a controlled way.
-
-A monad (cite Moggi?) is an abstract datatype representing
+basic monad \cite{liang-interpreter} in a controlled way.
+
+A monad \cite{moggi-notions} is an abstract datatype representing
 computations that yield a value and may have an \emph{effect}.
@@
 		with possibly undecidable zero-equivalence, or elements of a 
-		Division Ring.  In the roughly 85 pieces of code surveyed,
+		Division Ring.  In the roughly 85 pieces of code
+		surveyed \oleg{citation?},
 		20 different domains were encountered.
@@
 we have domains (like polynomials in sum-of-product form) where this
-is not the case.  This leads us to defining a |DOMAIN| type as
+is not the case.  The simplest case of such an abstraction was given
+in Section XXX, |param_code1|. In the latter, code-generators such as
+|plus| and |one| were passed as arguments. We see however that we 
+need far more than two parameters, so we have to group them. Instead
+of the grouping offered by regular records, we use Ocaml
+\emph{structures} (i.e., modules) (cite Ocaml module system tutorial?)
+so we can take advantage extensions, type abstraction and constraints,
+and especially parameterized structures (i.e., \emph{functors}).
+We first define the type of the domain, the signature |DOMAIN| to which
+different domain must satisfy:
 \begin{small}
@@
 domain |'a vc|, and once more from values to monadic computations
-|('a vc, 's, 'w) monad|.  Since there are times where we don't want
+|('a vc, 's, 'w) monad|.  Since there are times where we do not want
 to redundantly walk a whole matrix to apply normalization when 
@@
 
-For example, when lifted to a |DOMAIN|, the type |int| becomes
+One particular instantiation of the domain (a particular structure of
+|DOMAIN| signature) is the |int| domain \oleg{should we put the following?}
+|IntegerDomain : DOMAIN| to say that the compiler will verify that our
+|IntegerDomain| is indeed a |DOMAIN|, that is, satisfies the required
+signature. The constraint |DOMAIN| may be omitted; in that case, 
+the compiler will verify the type when we try to use that structure as
+a |DOMAIN|. In any case, the errors such as missing ``methods'' or 
+methods of wrong types will be caught statically, even \emph{before} any code
+generation takes place.
 \begin{small}

### 2005-04-04 paper/metamonads.tex
a few more corrections

@@
 result of invoking its continuation |k|. Rather, its result is a |let|
-code expression. \oleg{Mention similarity with reset?}
-\jacques{yes}
+code expression. Such a behavior is symptomatic of cntrol operators
+(in particular, |abort|).
 
@@
 We soon notice the need for another |ifM| function, with the same
-interface (see ifM in Figure~\ref{ourmonad}). The difference between them is
+interface (see |ifM| in Figure~\ref{ourmonad}). The difference between them is
 apparent from the following:
@@
 need similar operators for other Ocaml control forms: for generating
-case-matching statements and for- and while-loops. To make writing
-such code general and independent of the implementation of the monad,
-we define a function XXX. As the name implies, we do use control
-effect implemented via essentially delimited continuations (cite?).
-\jacques{yes!}
+case-matching statements and for- and while-loops.
 
@@
 really are}
+\oleg{Yes, let's gloss over ATLAS, SPIRAL, Axiom. I wonder what would
+  be the use of functors? Something by David McQueen or R. Harper.
+  In any way, mention that nobody before used functors to abstract
+  code generators -- or mix functors and multi-stage programming.}
 
@@
 \section{Conclusion}\label{conclusion}
-
 Mention the mdo notation. Mention it into the introduction.
@@
 The do notation for Ocaml.
-We use camlp4. Show the grammar. Explain the difficulty: although
-nice, the do notation cannot actually be expressed in a LR(n) grammar.
+To implement the |mdo| notation, we use the Ocaml pre-processor,
+camplp4 (cite?). Informally, our notation is
+\begin{code}
+mdo { exp }
+mdo { exp1; exp2 }
+mdo { x <-- exp; exp }
+mdo { let x = foo in; exp }
+\end{code}
+which matches closely the Haskell |do| notation, modulo |do|/|mdo| and 
+|<-|/|<--|.
+
+Grammar formally:
+\begin{code}
+mdo { <do-body> }
+<do-body> :: =
+    "let" var = EXP ("and" var = EXP)* "in" ";" <do-body>
+    | EXP
+    | (pat <--)? EXP ";" <do-body>
+\end{code}
+The semantics is given by the following rules re-writing the |mdo|
+form into the regular Ocaml code. We assume that the variable |bind|
+must be in scope.
+\begin{code}
+mdo { exp } ===> exp
+mdo { pat <-- exp; rest } ===> bind exp (fun pat -> mdo { rest })
+mdo { exp; rest } ===> bind exp (fun _ -> mdo { rest })
+mdo { let pat = exp in; rest } ===> let pat = exp in mdo { rest }
+\end{code}
+We also introduce |odo| form, which is equivalent to |mdo| but uses 
+a bind `method' rather than a bind `value'.
+
+The major difficulty with the |do| notation is that it cannot truly be
+parsed by an LR(n)-grammar for any finite |n| because many |pat| can also
+be parsed as an |exp|.
+
 Perhaps give the command line how to load the camlp4 module into

### 2005-04-05 paper/metamonads.tex
added a bit of related work

@@
 
+Mention |Blitz++| and template meta-programming. Can do similar
+eliminating of abstraction. Can even encode some domain-specific
+knowledge in the form of concepts. But inlining critically depends on
+the compiler's fully inlining of all methods. Furthermore, all errors
+(such as type errors and concept violation errors, that is,
+composition errors) are detected only when compiling the generated
+code. It is immensely difficult to correlate errors (e.g., line numbers) 
+to the ones in the generator itself. Furthermore, I think templates
+can't be local, right? So, a lot of code generation is spread around
+file-level entities.
+
+SPIRAL does the intentional code analysis. Probably ATLAS as well.
+
 \jacques{What do we consider \emph{related work} here?  Do we essentially gloss 

### 2005-04-06 paper/metamonads.tex
making text more concise

@@
 \usepackage{url}
+\usepackage[dvips]{color}
 
@@
 \DefineShortVerb{\|}
-\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\mathindent}
+\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\mathindent,commandchars=\\\{\}}
+\newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
 \setlength{\parskip}{0pt}
@@
 of name generation, and how continuation-passing style is a general
-tool for solving that problem.  Section~\ref{monadicnotation} introduces
+tool for solving that problem.  We also introduce
 the monad that greatly simplifies the task of dealing with
@@
 
-\section{CPS and the problem of generating names}\label{CPS}
+\section{The problem of generating names, CPS, and monad}\label{CPS}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 features: to construct a code expression, to combine them, and to
-execute them.
-
-For example, the following is probably the simplest code generator:
-\begin{code}
-let one = .<1>.
-\end{code}
-
-In the floating-point instance of that aspect, we would have
-|.<1.0>.| instead. We use MetaOCaml brackets |.<...>.| to generate
-code expressions -- in other words, to construct future-stage
-computations. The simplest combination is inlining:
-\begin{code}
-let plus x y = .<.~x + .~y>.
-\end{code}
-Here we see the second component of MetaOCaml -- escapes |.~|. They
-let us perform an immediate code generating computation while we are
-building the future-stage computation (code expression). In the
-example above, the immediate computation |.~x| is trivial -- obtaining
-the value of the variable |x|, which should be a code expression, and
-inlining it. The following shows an example of using the simple
-generators:
-
-\begin{code}
-let simplest_code =
-  let gen x y = plus x (plus y one) in
+execute them. The figure XXX shows the simplest code generator |one|,
+and more complex generators.
+
+\begin{figure*}
+\begin{code}
+let one = .<1>. and plus x y = .<.~x + .~y>.
+let simplest_code = let gen x y = plus x (plus y one) in
   .<fun x y -> .~(gen .<x>. .<y>.)>.
-\end{code}
-
-Here, the immediate computation |gen .<x>. .<y>.| is less trivial --
-it includes the evaluation of the function |gen|, which in turn
-applies the function |plus|. The function receives code expressions
-|.<x>.| and |.<y>.| as arguments. At the generating stage, we can
-manipulate code expressions as (opaque) values. The function |gen| returns a
-code expression, which will be inlined in the place of the
-escape. MetaOCaml can print out code expressions, so we can see the
-final generated code:\\
-  |.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|
-
-We must point out that |gen x y| looks like a combinator; its body 
-involves several applications of the function |plus|. When
-we look at the generated code however, we see no traces of the function
-|plus| and no such administrative applications. Instead, we see the
-fully inlined expression. These administative applications are done at
-the code generation stage.
-
-The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
-execute the code expression: |.! simplest_code| is a function of two
+\evalresult{.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.}
+let simplest_param_code plus one = let gen x y = plus x (plus y one) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
+  simplest_param_code plus one
+let param_code1 plus one =
+  let gen x y = plus (plus y one) (plus x (plus y one)) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+let param_code1' plus one =
+  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+param_code1' plus one
+\evalresult{.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.}
+\end{code}
+\caption{Code generation and combinators. $\Longrightarrow$ shows the
+  result of the evaluation of the preceeding expression}
+\end{figure*}
+
+We use MetaOCaml brackets |.<...>.| to generate code expressions -- in
+other words, to construct future-stage computations. We use escapes
+|.~| to perform an immediate code generating computation while we are
+building the future-stage computation (code expression). The immediate
+computation in |simplest_code| includes the evaluation of the function
+|gen|, which in turn applies the function |plus|. The function
+receives code expressions |.<x>.| and |.<y>.| as arguments. At the
+generating stage, we can manipulate code expressions as (opaque)
+values. The function |gen| returns a code expression, which will be
+inlined in the place of the escape. MetaOCaml can print out code
+expressions, so we can see the final generated code, It has
+no traces of the functions |gen| and |plus|: their applications are
+done at the code generation stage.
+
+The final MetaOCaml feature, |.!| (pronounced ``run'') 
+executes the code expression: |.! simplest_code| is a function of two
 integers, which we can apply: |(.! simplest_code) 1 2|. The original
@@
 To see the benefit of code generation, we notice that we can easily
-parameterize our code:
-
-\begin{code}
-let simplest_param_code plus one =
-  let gen x y = plus x (plus y one) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-\end{code}
-
-\noindent and use it to generate code that operates on integers, floating point
-numbers or booleans -- in general, any domain that implements |plus|
-and |one|:
-\begin{code}
-let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
-  simplest_param_code plus one
-let plus x y = .<.~x || .~y>. and one = .<true>. in
-  simplest_param_code plus one
-\end{code}
-Running the former expression yields the function on |float|s, whereas
-the latter expression is the code expression for a boolean function.
-
-This clearly shows the separation of concerns, namely of that for domain
-operations.
-
-Let us consider a more complex expression:
-\begin{code}
-let param_code1 plus one =
-  let gen x y = plus (plus y one) (plus x (plus y one)) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-\end{code}
-
-We notice two occurrences of |plus y one|. Depending on the
-implementation of |plus| that may be quite a complex computation, and
-so we would rather not do it twice. We may be tempted to rely on the
-compiler's common-subexpression elimination optimization. We would
-rather not rely on a ``sufficiently smart compiler'': when the
-generated code is very complex, the compiler may overlook common
-subexpressions.  Or the subexpressions may occur in an imperative
-context where the compiler might not be able to determine if lifting
-them is sound. So, being conservative, the optimizer will leave the
-duplicates as they are. We may attempt to write
-
-\begin{code}
-let param_code1 plus one =
-  let gen x y = 
-     let ce = (plus y one) in  plus ce (plus x ce) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-\end{code}
-
-However, the result of |param_code1 plus one|, which prints as\\
-|.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|,
-still exhibits duplicate sub-expressions. Our |let|-insertion
+parameterize our code, |simplest_param_code|, and use it to generate
+code that operates on integers, floating point numbers or booleans --
+in general, any domain that implements |plus| and |one|.
+
+A more complex expression |param_code1| has two occurrences of 
+|plus y one|. Depending on the implementation of |plus| that may be quite a
+complex computation, and so we would rather not do it twice. We may be
+tempted to rely on the compiler's common-subexpression elimination
+optimization. We would rather not rely on a ``sufficiently smart
+compiler'': when the generated code is very complex, the compiler may
+overlook common subexpressions.  Or the subexpressions may occur in an
+imperative context where the compiler might not be able to determine
+if lifting them is sound. So, being conservative, the optimizer will
+leave the duplicates as they are. We may attempt to eliminate
+subexpressions as in |param_code1'|. However, the result of
+|param_code1' plus one| still exhibits duplicate sub-expressions. 
+Our |let|-insertion
 optimization saved the computation at the generating stage. 
@@
 of partial evaluation, and extensively used by \cite{KiselyovTaha} for code
-generation.
-
+generation. Now, |param_code2 plus one| gives us the desired code.
+
+\begin{small}
 \begin{code}
@@
   in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
-\end{code}
-
-Now, |param_code2 plus one| gives us the desired code\\
-|.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.|.
-
-\section{Monadic notation, making CPS code clear}\label{monadicnotation}
+param_code2 plus one
+\evalresult{.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.}
+\end{code}
+\end{small}
 
@@
 initial continuation |k0|. We can now re-write our |param_code2|
-example of the previous section as
-
+example of the previous section as |param_code3|, see Fig. XXX.
+
+\begin{figure*}
 \begin{code}
@@
   in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
-\end{code}
+let param_code4 plus one =
+  let gen x y = mdo \{ ce <-- retN (plus y one);
+                      ret (plus ce (plus x ce)) \}
+  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
+let ifM' test th el = mdo \{
+  testc <-- test; thc <-- th; elc <-- el;
+  ifL testc thc elc\}
+let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
+ in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
+\evalresult{.<fun a_1 -> fun i_2 -> 
+    let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.}
+let gen a i = ifM (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
+ in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
+\evalresult{.<fun a_1 -> fun i_2 ->
+    if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.}
+\end{code}
+\end{figure*}
 % param_code3 plus one;;
 
-That does not seem like much of an improvement. But here we can introduce the
-do notation (Appendix A), which is patterned after the do-notation of Haskell.
-
-\begin{code}
-let param_code4 plus one =
-  let gen x y = mdo { ce <-- retN (plus y one);
-                      ret (plus ce (plus x ce)) }
-  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
-\end{code}
-
-The function |param_code4| is totally equivalent to |param_code3| --
-the |mdo|-notation is just syntactic sugar. In fact, the camlp4
-preprocessor (with our extension) will transform |param_code4|
-into exactly |param_code3|. And yet, |param_code4| looks far more
-conventional, as if it were indeed in direct style.
+That does not seem like much of an improvement. With the help of
+camplp4 pre-processor (cite), we introduce the do notation (give URL
+or cite the message on the MetaOCaml list), patterned after the
+do-notation of Haskell (cite). The function |param_code4|, written in
+the do-notation, is equivalent to |param_code3| -- in fact, the camlp4
+preprocessor will convert the former into the latter. And yet,
+|param_code4| looks far more conventional, as if it were indeed in
+direct style.
 
@@
 already generated pieces of code rather than monadic values. So, we
-may want to ``lift it'':
-\begin{code}
-let ifM' test th el = mdo {
-  testc <-- test;
-  thc   <-- th;
-  elc   <-- el;
-  ifL testc thc elc}
-\end{code}
-
-\noindent We can define functions |l1|, |l2|, |l3| (analogues of |liftM|,
-|liftM2|, |liftM3| of Haskell) to make such a lifting generic.
+may want to ``lift it'', see |ifM'|. We can define functions |l1|,
+|l2|, |l3| (analogues of |liftM|, |liftM2|, |liftM3| of Haskell) 
+to make such a lifting generic.
 We soon notice the need for another |ifM| function, with the same
 interface (see |ifM| in Figure~\ref{ourmonad}). The difference between them is
-apparent from the following:
-\begin{code}
-let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
-                           (ret .<None>.)
- in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
-;;
-.<fun a_1 -> fun i_2 ->
-    let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.
-\end{code}
-\noindent whereas if we replace |ifM'| with |ifM|, the result will be
-\begin{code}
-.<fun a_1 -> fun i_2 ->
-    if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.
-\end{code}
-
-The difference is apparent: in the code with |ifM'|, the let-insertion
+apparent, see Fig. XXX: in the code with |ifM'|, the let-insertion
 happend \emph{before} the if-expression, that is, before the test that
@@
   type 'a vc = ('a,v) code
-  val zero : 'a vc
-  val one : 'a vc
+  val zero : 'a vc   val one : 'a vc
   val plus : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
@@
 \begin{code}
-module IntegerDomain = struct
-    type v = int
-    type kind = domain_is_ring
+module IntegerDomain : DOMAIN = struct
+  type v = int  type kind = domain_is_ring
   type 'a vc = ('a,v) code
-    let zero = .< 0 >.  
-    let one = .< 1 >. 
+  let zero = .< 0 >.  and one = .< 1 >. 
   let plus x y = ret .<.~x + .~y>. 
-    let times x y = ret .<.~x * .~y>.
-    let minus x y = ret .<.~x - .~y>.
-    let uminus x = ret .< - .~x>.
-    let div x y = ret .<.~x / .~y>. 
   let better_than = Some (fun x y -> ret .<abs .~x > abs .~y >. )
   let normalizerf = None
-    let normalizerg = fun x -> x
+  ...
 end
@@
 \end{small}
-\noindent While the |DOMAIN| type may have looked daunting to some, 
+\noindent Here |domain_is_ring| is an abstract data type to encode semantic constraint
+-- the kind of the domain.
+While the |DOMAIN| type may have looked daunting to some, 
 the implementation is quite straightforward.  It is equally simple to 
@@
 module type OUTPUT = sig
-  type contr
-  type res
+  type contr  type res
   module D : DETERMINANT
@@
     (Det : DETERMINANT with type indet = Dom.v and type outdet = Dom.v)
-    (Rank : RANK) =
-  struct
+    (Rank : RANK) = struct
   module Ctr = C(Dom)
@@
 module type DETERMINANT = sig
-  type indet
-  type outdet
+  type indet  type outdet  type 'a lstate
   type tdet = outdet ref   
-  type 'a lstate
   val decl : unit -> 
@@
     (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
-  val zero_sign : unit -> 
-    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
-  val acc : ('a,indet) code -> 
-    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
-  val get : unit ->
-    (('a,tdet) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
-  val set : ('a,indet) code -> 
-    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
-  val fin : unit -> 
-    (('a,outdet) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
+  ...
 end
@@
 \bibliographystyle{plain}
+\end{document}
+
+% It seems no appendices are allowed. Check with Walid?
 
@@
 ocaml. Point to the code.
-\end{document}
 
@@
 }
+
+%%% Old, detailed version is here
+
+\section{CPS and the problem of generating names}\label{CPS}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% And now, a better -- and shorter -- write-up.
+
+The essence of our approach is assembling the code of the program by
+combining instances of various aspects together. In the case of
+Gaussian Elimination, the algorithmic aspects are pivot determination, 
+swapping of
+rows and columns, determinant tracking, etc. A primitive code
+generator yields a code expression.  A combinator takes (complete and
+well-typed) code expressions and combines them into a composite expression
+(which is also statically guaranteed to be well-formed and well-typed).
+
+We will be using MetaOcaml which, as an instance of a multi-stage
+programming system \cite{TahaThesis}, provides exactly the needed
+features: to construct a code expression, to combine them, and to
+execute them.
+
+For example, the following is probably the simplest code generator:
+\begin{code}
+let one = .<1>.
+\end{code}
+
+In the floating-point instance of that aspect, we would have
+|.<1.0>.| instead. We use MetaOCaml brackets |.<...>.| to generate
+code expressions -- in other words, to construct future-stage
+computations. The simplest combination is inlining:
+\begin{code}
+let plus x y = .<.~x + .~y>.
+\end{code}
+Here we see the second component of MetaOCaml -- escapes |.~|. They
+let us perform an immediate code generating computation while we are
+building the future-stage computation (code expression). In the
+example above, the immediate computation |.~x| is trivial -- obtaining
+the value of the variable |x|, which should be a code expression, and
+inlining it. The following shows an example of using the simple
+generators:
+
+\begin{code}
+let simplest_code =
+  let gen x y = plus x (plus y one) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+
+Here, the immediate computation |gen .<x>. .<y>.| is less trivial --
+it includes the evaluation of the function |gen|, which in turn
+applies the function |plus|. The function receives code expressions
+|.<x>.| and |.<y>.| as arguments. At the generating stage, we can
+manipulate code expressions as (opaque) values. The function |gen| returns a
+code expression, which will be inlined in the place of the
+escape. MetaOCaml can print out code expressions, so we can see the
+final generated code:\\
+  |.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|
+
+We must point out that |gen x y| looks like a combinator; its body 
+involves several applications of the function |plus|. When
+we look at the generated code however, we see no traces of the function
+|plus| and no such administrative applications. Instead, we see the
+fully inlined expression. These administative applications are done at
+the code generation stage.
+
+The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
+execute the code expression: |.! simplest_code| is a function of two
+integers, which we can apply: |(.! simplest_code) 1 2|. The original
+|simplest_code| is not the function on integers -- it is a code
+expression.
+
+To see the benefit of code generation, we notice that we can easily
+parameterize our code:
+
+\begin{code}
+let simplest_param_code plus one =
+  let gen x y = plus x (plus y one) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+
+\noindent and use it to generate code that operates on integers, floating point
+numbers or booleans -- in general, any domain that implements |plus|
+and |one|:
+\begin{code}
+let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
+  simplest_param_code plus one
+let plus x y = .<.~x || .~y>. and one = .<true>. in
+  simplest_param_code plus one
+\end{code}
+Running the former expression yields the function on |float|s, whereas
+the latter expression is the code expression for a boolean function.
+
+This clearly shows the separation of concerns, namely of that for domain
+operations.
+
+Let us consider a more complex expression:
+\begin{code}
+let param_code1 plus one =
+  let gen x y = plus (plus y one) (plus x (plus y one)) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+
+We notice two occurrences of |plus y one|. Depending on the
+implementation of |plus| that may be quite a complex computation, and
+so we would rather not do it twice. We may be tempted to rely on the
+compiler's common-subexpression elimination optimization. We would
+rather not rely on a ``sufficiently smart compiler'': when the
+generated code is very complex, the compiler may overlook common
+subexpressions.  Or the subexpressions may occur in an imperative
+context where the compiler might not be able to determine if lifting
+them is sound. So, being conservative, the optimizer will leave the
+duplicates as they are. We may attempt to write
+
+\begin{code}
+let param_code1 plus one =
+  let gen x y = 
+     let ce = (plus y one) in  plus ce (plus x ce) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+
+However, the result of |param_code1 plus one|, which prints as\\
+|.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|,
+still exhibits duplicate sub-expressions. Our |let|-insertion
+optimization saved the computation at the generating stage. 
+We need a combinator that inserts the |let| expression in the
+generat\emph{ed} code. We need a combinator |letgen| to be used
+as\\|let ce = letgen (plus y one) in plus ce (plus x ce)| 
+yielding the code like |.<let t = y + 1 in t + (x + t)>.|
+But that seems impossible because |letgen exp| has to generate
+the expression\\|.<let t = exp in body>.| although |letgen| does not have
+the |body| yet. The body needs a temporary identifier |.<t>.| that
+is supposed to be the result of |letgen| itself. 
+Certainly |letgen| cannot generate only part of a let-expression,
+without the |body|,  as all generated expressions in
+MetaOCaml are well-formed and complete.
+
+The key is to use continuation-passing style (CPS). Its benefits were first
+pointed out by \cite{Bondorf:92} in the context
+of partial evaluation, and extensively used by \cite{KiselyovTaha} for code
+generation.
+
+\begin{code}
+let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
+let param_code2 plus one =
+  let gen x y k = letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
+  and k0 x = x
+  in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
+\end{code}
+
+Now, |param_code2 plus one| gives us the desired code\\
+|.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.|.
+
+\section{Monadic notation, making CPS code clear}\label{monadicnotation}
+
+Comparing the code that did let-insertion at the generating stage\\
+|let ce = (plus y one) in  plus ce (plus x ce)|\\
+with the corresponding code inserting let at the generated code stage\\
+|letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))|\\
+clearly shows the difference between the direct-style and CPS code.
+What was |let ce = init in ...| in direct style became
+|init' (fun ce -> ...)| in CPS. For one thing, |let| became
+inverted. For another, what used to be an expression that yields
+a value, |init|, now became an expression that takes an extra argument,
+the continuation, and invokes it. The differences look negligible in
+the above example. In larger expressions with many let-forms, the
+number of parentheses around |fun| increases, the need to add and
+then invoke the |k| continuation argument become increasingly annoying. The
+inconvenience is great enough for some people to explicitly avoid CPS
+or claim that numerical programmers (our users) cannot or will not
+program in CPS. Clearly a better notation is needed.
+
+The |do|-notation of Haskell \cite{Haskell98Report} shows that it is possible
+to write essentially CPS code in conventional-looking style. The
+|do| notation is the notation for writing monadic code \cite{moggi-notions}.
+The benefit of monadic notation is not that it can represent CPS \cite{Filinski:Representing}, but it helps in composability by offering to add different
+layers of effects (state, exception, non-determinism, etc) to the
+basic monad \cite{liang-interpreter} in a controlled way.
+
+A monad \cite{moggi-notions} is an abstract datatype representing
+computations that yield a value and may have an \emph{effect}.
+The datatype must have at least two operations, |return| to build
+trivial effect-less computations and |bind| for combining
+computations. These operations must satisfy \emph{monadic laws}:
+|return| being the left and the right unit of |bind| and |bind| being
+associative. Figure~\ref{ourmonad} defines the monad used throughout
+the present paper and shows its implementation.
+
+\begin{figure*}\label{ourmonad}
+\begin{code}
+type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
+let ret a = fun s k -> k s a
+let bind a f = fun s k -> a s (fun s' b -> f b s' k)
+let fetch s k = k s s
+let store v s k = k (v::s) ()
+
+let k0 s v = v
+let runM m = m [] k0
+
+let l1 f = fun x -> mdo { t <-- x; f t}
+let l2 f = fun x y -> mdo { tx <-- x; ty <-- y; f tx ty}
+
+let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
+
+let ifL test th el = ret .< if .~test then .~th else .~el >.
+let ifM test th el = fun s k ->
+  k s .< if .~(test s k0) then .~(th s k0) else .~(el s k0) >.
+\end{code}
+\caption{Our monad}
+\end{figure*}
+
+Our monad represents two kinds of computational effects: reading and
+writing a computation-wide state, and control effects. The latter are
+normally associated with exceptions, forking of computations, etc. --
+in general, whenever a computation ends with something other than
+invoking its natural continuation in the tail position. In our case
+however the control effects manifest themselves as code generation.
+
+In Figure~\ref{ourmonad}, the monad is implemented as a function of two
+arguments: the state (of type |s|) and the continuation. The
+continuation receives the current state, the value (of the type |v|) and
+yields the answer of the type |w|.  The monad is polymorphic over all
+these three type parameters.  Other implementations are
+possible. Except for the code in Figure~\ref{ourmonad}, the rest of our code
+treats the monad as a truly abstract data type. The implementation of basic
+monadic operations |ret| and |bind| are conventional. It is easy to
+see that the monadic laws are satisfied.  Other monadic operations
+construct computations that do have effects.  Operations |fetch| and
+|store v| construct computations that read and write the state. In our
+case the state is a list (of polymorphic variants), which models an
+open discriminated union, as we shall see later.
+
+The operation |retN a| is the let-insertion operation, whose simpler
+version we called |letgen| earlier. It is the first computation with
+a control effect: indeed, the result of |retN a| is \emph{not} the
+result of invoking its continuation |k|. Rather, its result is a |let|
+code expression. Such a behavior is symptomatic of cntrol operators
+(in particular, |abort|).
+
+Finally, |runM| runs our monad, that is, performs the computation of
+the monad and gets the result, which in our case is the code
+expression. We run the monad by passing it the initial state and the
+initial continuation |k0|. We can now re-write our |param_code2|
+example of the previous section as
+
+\begin{code}
+let param_code3 plus one =
+  let gen x y = bind (retN (plus y one)) (fun ce -> 
+                ret (plus ce (plus x ce)))
+  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
+\end{code}
+% param_code3 plus one;;
+
+That does not seem like much of an improvement. But here we can introduce the
+do notation (Appendix A), which is patterned after the do-notation of Haskell.
+
+\begin{code}
+let param_code4 plus one =
+  let gen x y = mdo { ce <-- retN (plus y one);
+                      ret (plus ce (plus x ce)) }
+  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
+\end{code}
+
+The function |param_code4| is totally equivalent to |param_code3| --
+the |mdo|-notation is just syntactic sugar. In fact, the camlp4
+preprocessor (with our extension) will transform |param_code4|
+into exactly |param_code3|. And yet, |param_code4| looks far more
+conventional, as if it were indeed in direct style.
+
+We can write operations that generate code other than let-statements,
+e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
+albeit straightforward, is not as general as we wish: its argument are
+already generated pieces of code rather than monadic values. So, we
+may want to ``lift it'':
+\begin{code}
+let ifM' test th el = mdo {
+  testc <-- test;
+  thc   <-- th;
+  elc   <-- el;
+  ifL testc thc elc}
+\end{code}
+
+\noindent We can define functions |l1|, |l2|, |l3| (analogues of |liftM|,
+|liftM2|, |liftM3| of Haskell) to make such a lifting generic.
+We soon notice the need for another |ifM| function, with the same
+interface (see |ifM| in Figure~\ref{ourmonad}). The difference between them is
+apparent from the following:
+\begin{code}
+let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
+                           (ret .<None>.)
+ in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
+;;
+.<fun a_1 -> fun i_2 ->
+    let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.
+\end{code}
+\noindent whereas if we replace |ifM'| with |ifM|, the result will be
+\begin{code}
+.<fun a_1 -> fun i_2 ->
+    if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.
+\end{code}
+
+The difference is apparent: in the code with |ifM'|, the let-insertion
+happend \emph{before} the if-expression, that is, before the test that
+the index |i| is positive. That is, if |i| turned out
+negative, |a.(i)| will generate an out-of-bound array access
+error. On the other hand, the code with |ifM| accesses the array only
+when we have verified that the index is non-negative. This example
+makes it clear that the code generation (such as the one in |retN|) is 
+truly an effect and we have to be clear about the sequencing of
+effects, when generating control constructions such as conditionals.
+The form |ifM| handles the such effects correctly. In our code, we
+need similar operators for other Ocaml control forms: for generating
+case-matching statements and for- and while-loops.

### 2005-04-06 paper/metamonads.tex
more changes

@@
 	\item \textbf{Length measure}:  For stability reasons
-		(whether numerical or coefficient growth), if a domain possesses
+		(whether numerical or coefficient growth), if a domain
+	  possesses
 		an appropriate length measure, this is sometimes used to choose
@@
 different domain must satisfy:
-\begin{small}
+\begin{figure*}
 \begin{code}
 module type DOMAIN = sig
-  type v
+  type v    type 'a vc = ('a,v) code
   type kind (* Field or Ring ? *)
-  type 'a vc = ('a,v) code
   val zero : 'a vc   val one : 'a vc
   val plus : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
-  val times : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
-  val minus : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
-  val uminus : 'a vc -> ('a vc, 's, 'w) monad
-  val div : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
+  (* times, minus, uminus, div elided for brevity *)
   val better_than : ('a vc -> 'a vc -> 
@@
   val normalizerf : (('a,v -> v) code ) option
-  val normalizerg : 'a vc -> 'a vc
 end 
-\end{code}
-\end{small}
-\noindent  The first item to notice is that the types above are
-generally lifted twice: once from the value domain |v| to the code
-domain |'a vc|, and once more from values to monadic computations
-|('a vc, 's, 'w) monad|.  Since there are times where we do not want
-to redundantly walk a whole matrix to apply normalization when 
-normalization is not necessary, |normalizerf| is optional.
-
-One particular instantiation of the domain (a particular structure of
-|DOMAIN| signature) is the |int| domain \oleg{should we put the following?}
-|IntegerDomain : DOMAIN| to say that the compiler will verify that our
-|IntegerDomain| is indeed a |DOMAIN|, that is, satisfies the required
-signature. The constraint |DOMAIN| may be omitted; in that case, 
-the compiler will verify the type when we try to use that structure as
-a |DOMAIN|. In any case, the errors such as missing ``methods'' or 
-methods of wrong types will be caught statically, even \emph{before} any code
-generation takes place.
-\begin{small}
-\begin{code}
 module IntegerDomain : DOMAIN = struct
@@
 \end{code}
-\end{small}
-\noindent Here |domain_is_ring| is an abstract data type to encode semantic constraint
--- the kind of the domain.
-While the |DOMAIN| type may have looked daunting to some, 
-the implementation is quite straightforward.  It is equally simple to 
-make implementations for |float| and for |Num.num| (arbitrary precision
-exact rational numbers).  With scarcely more work (as these are not 
-built-in and the arithmetic must be implemented), $\mathbb{Z}_p$,
-$\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$ and more
-complex domains can be done as well.
+\end{figure*}
+\noindent  The first item to notice is that the types above are
+generally lifted twice: once from the value domain |v| to the code
+domain |'a vc|, and once more from values to monadic computations
+|('a vc, 's, 'w) monad|.  Since there are times where we do not want
+to redundantly walk a whole matrix to apply normalization when 
+normalization is not necessary, |normalizerf| is optional.
+
+One particular instantiation of the domain (a particular structure of
+|DOMAIN| signature) is the |int| domain. The notattion
+|module IntegerDomain : DOMAIN| makes the compiler
+verify that our |IntegerDomain| is indeed a |DOMAIN|, that is,
+satisfies the required signature. The constraint |DOMAIN| may be
+omitted; in that case, the compiler will verify the type when we try
+to use that structure as a |DOMAIN|. In any case, the errors such as
+missing ``methods'' or methods of wrong types will be caught
+statically, even \emph{before} any code generation takes place. The
+abstract type |domain_is_ring| encodes a semantic constraint that the
+full division is not available. While the |DOMAIN| type may have
+looked daunting to some, the implementation is quite straightforward.
+It is equally simple to make implementations for |float| and for
+|Num.num| (arbitrary precision exact rational numbers).  With scarcely
+more work (as these are not built-in and the arithmetic must be
+implemented), $\mathbb{Z}_p$, $\mathbb{Z}\left[x\right]$,
+$\mathbb{Q}\left(x\right)$ and more complex domains can be done as
+well.
 
@@
 over a |DOMAIN| by making containers functors from a |DOMAIN| module
-to the actual implementation of a container.  Our containers are
-explicitly 2-dimensional, and we provide functions |dim1| and |dim2|
-to extract each.  We also need |get| and |set| functions -- these come
-in both a monadic flavour as well as a direct code-generation version.
-For convenience and efficiency, several additional functions are also
-required: a way to |copy| a container, a method (|mapper|) to map an
-optional function over the elements of a container, as well as methods
-to swap rows and columns.  The inclusion of these functions in the 
-signature of all containers makes it simpler to optimize the relevant
-functions depending on the actual representation of the container
-while not burdening the users of containers with efficiency details.
-As an example, below is an implementation of a matrix as an
-array of rows (themselves arrays).  This implementation clearly shows
-how swapping rows is more efficient than swapping columns in this
-representation.  We have not (yet) implemented the further potential
-optimization wherein if it is known that part of two rows/columns are already
+to the actual implementation of a container. The functor signature
+|CONTAINER2D| specifies that a container must provide functions |dim1|
+and |dim2| to extract the dimensions, functions |get| and |set| to
+generate container getter and setters, the cloning generator |copy|
+and functions that generate code for row and column swapping. The
+inclusion of these functions in the signature of all containers makes
+it simpler to optimize the relevant functions depending on the actual
+representation of the container while not burdening the users of
+containers with efficiency details. 
+\omitnow{ As an example, below is
+  an implementation of a matrix as an array of rows (themselves
+  arrays).  This implementation clearly shows how swapping rows is
+  more efficient than swapping columns in this representation.  We
+  have not (yet) implemented the further potential optimization
+  wherein if it is known that part of two rows/columns are already
   known to be equal (in fact to |Dom.zero| in the case of GE) and
@@
 important that the |DOMAIN| type matches -- but luckily the compiler 
-will prevent us from making such a mistake.
+will prevent us from making such a mistake.}
 
@@
 determinant, the matrix, rank and determinant, and so on.  More
-precisely, below we show this type and one instantiation.
+precisely, below we show this module type and one instantiation,
+which specifies the output of a  GE algorithm as a 3-tuple
+|contr * Det.outdet * int| of the U-factor, the determinant and the rank.
 
@@
   type contr  type res
-  module D : DETERMINANT
-  module R : RANK
+  module D : DETERMINANT   module R : RANK
   module P : TRACKPIVOT
@@
      [> `TDet of 'a D.lstate | `TRan of 'a R.lstate | `TPivot of 'a P.lstate]
-       list,
-     ('a,'w) code) monad
+       list, ('a,'w) code) monad
 end
-
 module OutDetRank(Dom:DOMAIN)(C: CONTAINER2D)
@@
   type res = contr * Det.outdet * int
-  module D = Det
-  module R = Rank
+  module D = Det   module R = Rank
   module P = DiscardPivot
-  let make_result b = mdo {
-    det  <-- D.fin ();
-    rank <-- R.fin ();
-    ret .< ( .~b, .~det, .~rank ) >. }
+  let make_result b = mdo \{ det  <-- D.fin ();  rank <-- R.fin ();
+    ret .< ( .~b, .~det, .~rank ) >. \}
 end
@@
 \end{small}
-\noindent It is worthwhile noting that while the |OUTPUT| type is
-relatively simple, instantiating it requires a rather complex functor
-with some additional type constraints.  While we will explain some of
-the details of the above code later, right now it suffices to notice
-that |res| above is a 3-tuple with type |contr * Det.outdet * int|,
-where in another instance this type is simply |contr|, and these are
-formally incomparable types.
 
@@
 \end{small}
-\noindent  Determinant tracking is quite complex: we need to 
-declare that we are tracking the determinant (|decl|), have some functions
-for tracking the sign of the determinant (|upd_sign|,|zero_sign|),
-|set|, |get|, accumulate (|acc|) unsigned values for the determinant, as 
-well as returning a final value (|fin|) for the determinant.  In some cases,
-like for float matrices where we do not return the determinant, all of
-the above should generate no code at all, while for integer matrices where
-we want the determinant as part of the output, each of these need to be
-implemented.
-
-\jacques{I am stopping my description of state here -- what I have
-written above is quite dull, and not making the point very well.
-I will need to revisit this}
-
-Making state modular and composable: several aspects can contribute
-to the state, without interference (show: Rank, Det, Permutation
-aspects).
-What is state and why we need it: making aspects interleave: use det
-as an example. State as an open record. List of polymorphic variants
+\noindent  Determinant tracking aspects involves being able to
+generate code that defines variables used for tracking (|decl|),
+generate code that updates the sign or the absolute
+value of the determinant, and finally, converting the tracking state
+to the final determinant value of the type |outdet|. GE of a
+floating-point matrix with no determinant tracking uses the
+instantiation of |DETERMINANT| where |outdet| is |unit| and all the
+functions of that module generate no code. For integer matrices, we
+have to track some aspects of the determinant even if we don't output
+it. The determinant tracking aspect is complex becasue tracking
+variables, if any, are to be declared at the beginning of GE; the sign
+of the determinant has to be updated on each row or column
+permutation; the value of the determinant should be updated per each
+pivoting. We use |lstate| to pass the tracking state, which may be, for
+example, a piece of code for the value of the type |float ref| among
+various determinant-tracking functions. The |lstate| is a part of the
+overall monadic state. Other aspects, e.g., rank tracking, may use the
+monadic state for passing of rank tracking variables. To be able to
+compose determinant and rank tracking functors -- each of which may
+(or may not) use the monadic state for passing its own data -- we make
+an extensive use of open records (list of polymorphic variants
 appeared to be the easiest way to implement such a union, in a pure
-functional way.
+functional way). This lets us freely compose determinant-tracking,
+rank-tracking, and other aspects.
 
@@
 of either det nor pivoting). Then the code with RowPivoting, Det and
-float domain. \jacques{which would really be great if we had a lot 
-more space, but I think we're already long, so something is going to
-have to be cut.  Maybe not this, but something!}
+float domain.
 

### 2005-04-07 paper/metamonads.tex
node changes

@@
 \DefineShortVerb{\|}
-\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\mathindent,commandchars=\\\{\}}
+\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\mathindent,commandchars=\\\{\},fontsize=\small}
 \newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
@@
 
-\begin{small}
 \begin{code}
@@
 \end{code}
-\end{small}
 
@@
 \end{code}
-\end{small}
 \noindent  Implementing other containers based on flattened vectors,
@@
 \end{code}
-\end{small}
 \noindent  Determinant tracking aspects involves being able to
@@
 
-Expressing domain constraints in functors (Div and integer domain).
-Show the type error that occurs in improper instantiation.
-
-Show several examples (e.g., code with NullPivot and nodet: no traces
-of either det nor pivoting). Then the code with RowPivoting, Det and
-float domain.
+The GE generator functor is as follows. It is
+parameterized by the domain, container, pivoting policy (full, row,
+nonzero, nopivoting), update policy (that use either ``fraction-less'
+or full division), and by what to yield as the result. Some of the
+argument modules such as |PIVOT| are functors themselves (parameterized
+by the domain, the container, and the determinant functor). The sharing
+constraints express obvious constraints on the instantiation of |Gen|,
+for example, pivoting, determinant etc. components all use the same
+domain. It must be stressed that all structures (i.e., module
+instances) are stateless, and so we never have to worry that different
+aspect functors (such as |CONTAINER2D| and |PIVOT|) are instantiated
+with different but type-compatible instances of |DOMAIN|. That is, we
+are not concerened at all about the value sharing. Aspects such as
+determinant tracking may be stateful so that the determinant update
+code have access to the determinant tracking variables declared
+previsouly. But that state is handled via the monadic state. As we
+have shown, open unions makes the overall monadic state compositional
+with respect to the state of various aspects.
+
+\begin{figure*}
+\begin{code}
+module Gen(Dom: DOMAIN)(C: CONTAINER2D)(PivotF: PIVOT)
+          (Update: UPDATE with type baseobj = Dom.v and type ctr = C(Dom).contr)
+          (Out: OUTPUT with type contr = C(Dom).contr and type D.indet = Dom.v 
+                        and type 'a D.lstate = 'a Update.D.lstate) =
+   struct
+    module Ctr = C(Dom)
+    module Pivot = PivotF(Dom)(C)(Out.D)
+    type v = Dom.v
+    let gen =
+      let zerobelow b r c m n brc =
+        let innerbody i = mdo \{
+            bic <-- Ctr.get b i c;
+            whenM (l1 LogicCode.not (LogicCode.equal bic Dom.zero ))
+                (seqM (retLoopM (Idx.succ c) (Idx.pred m)
+                          (fun k -> Update.update b r c i k) )
+                      (Ctr.set b i c Dom.zero)) \} in 
+        mdo \{
+              seqM (retLoopM (Idx.succ r) (Idx.pred n) innerbody) 
+                   (Update.update_det brc) \} in
+      let dogen a = mdo \{
+          r <-- Out.R.decl ();
+          c <-- retN (liftRef Idx.zero);
+          b <-- retN (Ctr.mapper Dom.normalizerf (Ctr.copy a));
+          m <-- retN (Ctr.dim1 a);
+          n <-- retN (Ctr.dim2 a);
+          () <-- Update.D.decl ();
+          () <-- Out.P.decl ();
+          seqM 
+            (retWhileM (LogicCode.and_ (Idx.less (liftGet c) m)
+                                       (Idx.less (liftGet r) n) )
+               ( mdo \{
+               rr <-- retN (liftGet r);
+               cc <-- retN (liftGet c);
+               pivot <-- l1 retN (Pivot.findpivot b rr m cc n);
+               seqM (retMatchM pivot (fun pv -> 
+                        seqM (zerobelow b rr cc m n pv)
+                             (Out.R.succ ()) )
+                        (Update.D.zero_sign () ))
+                    (Code.update c Idx.succ) \} ))
+            (Out.make_result b) \} 
+    in
+    .<fun a -> .~(runM (dogen .<a>.)) >.
+end
+\end{code}
+\end{figure*}
+
+In addition to the ``regular'' type sharing constraints shown in the
+|Gen| functor, there are also ``semantic'' sharing constraints, shown
+in the following structure of the |UPDATE| signature:
+
+\begin{code}
+module DivisionUpdate
+  (Dom:DOMAIN with type kind = domain_is_field)
+  (C:CONTAINER2D)
+  (Det:DETERMINANT with type indet=Dom.v) = struct ... end
+\end{code}
+
+The structure implements update policy that can use |Dom.div| operation without
+restrictions. That is possible only if the domain has such an
+operation. A domain such as integer domain may still provide |Dom.div|
+of the same type, but that operation can only be used when we are sure
+that the division is exact. Using the integer domain with
+|DivisionUpdate| will lead to the compile-time error, when compiling
+the \emph{generator} code. Thus, in some cases we can use module types
+to express ``semantic'' constraints that cannot normally be expressed
+via the types of module members.
+
+\begin{code}
+module GenFV4 = Gen(FloatDomain)
+   (GenericVectorContainer)(FullPivot)
+   (DivisionUpdate(FloatDomain)(GenericVectorContainer)(FDet))
+   (OutDetRank(FloatDomain)(GenericVectorContainer)(FDet)(Rank))
+module GenIA1 = Gen(IntegerDomain)
+   (GenericArrayContainer)(RowPivot)
+   (FractionFreeUpdate(IntegerDomain)(GenericArrayContainer)(IDet))
+   (OutJustMatrix(IntegerDomain)(GenericArrayContainer)(IDet))
+\end{code}
+
+We can instantiate the |Gen| functor as shown above and inspect the
+generated code, e.g., by printing |GenIA1.gen|. The code can then be
+``compiled'' as |!. GenIA1.gen| or with off-shiring (cite). The code
+for |GenFV4| shows full pivoting, determinant and rank tracking. The
+code for all these aspects is fully inlined; no extra functions are
+invoked and no tests other than those needed by the GE algorithm
+itself are performed. The GE function returns a triple 
+|float array array * float * int| of the U-factor, determinant and 
+the rank. The
+code generated by |GenIA1| shows absolutely no traces of determinant
+tracking: no declaratrion of spurious variables, no extra tests,
+etc. The code appears as if determinant tracking aspect did not exist
+at all. The generated code for the above and other instantiations of
+|Gen| can be examined at (cite).
 
@@
 Mention the mdo notation. Mention it into the introduction.
+
+Combination of stateless modules (functors and structures) with
+the monad with the compositional state greatly makes the aspects
+freely composable without having to worry about presence or absense of
+value aliasing. The only constraints to compositionality are the
+typing constraints plus the constraints we specifically impose,
+including semantic constraints (e.g., rings do not have the full
+division).
 

### 2005-04-07 paper/metamonads.tex
abstract

@@
 \begin{abstract}
+With Gaussian Elimination as a representative family of numerical
+algorithms, we use multi-stage programming, monads and Ocaml's
+advanced module system to demonstrate the complete elimination of the
+abstraction overhead while avoiding any inspection of the generated
+code.  We parameterize our Gaussian Elimination code to great extent
+(over domain, matrix representations, determinant tracking and
+pivoting policies, result types, etc) at no run-time cost.  Because
+the resulting code is generated just right and not changed afterwards,
+we enjoy MetaOCaml guarantees that the generated code is well-typed.
+We further demonstrate that various abstraction parameters (aspects)
+can be made orthogonal and compositional, even in the presence of
+name-generation for temporaries and other bindings and
+``interleaving'' of aspects.  We also show how to encode some
+domain-specific knowledge so that ``clearly wrong'' compositions can
+be statically rejected by the compiler when processing the generator
+rather than the generated code.
+
 We use monads and Ocaml's advanced module system to

### 2005-04-08 paper/metamonads.tex
minor changes to abstract. changes for spacing.

--- 2005-04-07 paper/metamonads.tex
@@
 %\usepackage{refrange}
+
+\usepackage[medium,compact]{titlesec}
 
@@
 %% \makeatother
+
+\parskip 0pt plus 1pt minus 2pt
+\textfloatsep 4pt plus 2pt minus 4pt % Less space around figures
+\abovecaptionskip 0pt plus 0pt minus 2pt
+\belowcaptionskip 0pt plus 0pt minus 2pt
+
+\renewcommand\floatpagefraction{.9}
+\renewcommand\topfraction{.9}
+\renewcommand\bottomfraction{.9}
+\renewcommand\textfraction{.1}   
+\setcounter{totalnumber}{50}
+\setcounter{topnumber}{50}
+\setcounter{bottomnumber}{50}
 
@@
 \begin{abstract}
-We use monads and Ocaml's advanced module system to
-demonstrate that, in a generative context, it is
-possible to totally eliminate the overhead of abstraction.
-This lets one use extreme forms of \textit{information hiding}
-at no run-time cost.  Furthermore the typed nature of
-the generative context provide static guarantees about the
-generated code.  The various extensions
-(aspects) can be made completely orthogonal and compositional,
-even in the presence of name-generation for temporaries and
-other bindings and ``interleaving'' of aspects.  We also
-show how to encode some domain-specific knowledge so that
-``clearly wrong'' compositions can be statically rejected by the
-compiler.
+With Gaussian Elimination as a representative family of numerical
+and symbolic
+algorithms, we use multi-stage programming, monads and Ocaml's
+advanced module system to demonstrate the complete elimination of the
+abstraction overhead while avoiding any inspection of the generated
+code.  We parameterize our Gaussian Elimination code to a great extent
+(over domain, matrix representations, determinant tracking, 
+pivoting policies, result types, etc) at no run-time cost.  Because
+the resulting code is generated just right and not changed afterwards,
+we enjoy MetaOCaml's guaranty that the generated code is well-typed.
+We further demonstrate that various abstraction parameters (aspects)
+can be made orthogonal and compositional, even in the presence of
+name-generation for temporaries and other bindings and
+``interleaving'' of aspects.  We also show how to encode some
+domain-specific knowledge so that ``clearly wrong'' compositions can
+be statically rejected by the compiler when processing the generator
+rather than the generated code.
 \end{abstract}
@@
 
-In high-performance and numeric computing, there is a well-known
+In high-performance, symbolic, and numeric computing, there is a well-known
 issue of balancing between maximal performance and the level of
@@
 been known to resort to cut\&paste programming, no self-respecting
-programmer would be caught doing that\footnote{Sadly, there are still
-many programmers who do not see the harm of cut\&paste programming}.
+programmer would be caught doing that.
 The common solution is to turn to code generation.
@@
 programming in C++ and Template Haskell do offer some amount of static
-typing \cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml\footnote{and
-  MetaML.} offers stronger guarantees.  As well, since MetaOCaml is
+typing \cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml
+offers stronger guarantees.  As well, since MetaOCaml is
 implemented on top of the very solid Ocaml language, one accrues
@@
           (Out: OUTPUT with type contr = C(Dom).contr and type D.indet = Dom.v 
-                        and type 'a D.lstate = 'a Update.D.lstate) =
-   struct
+                        and type 'a D.lstate = 'a Update.D.lstate) = struct
     module Ctr = C(Dom)
     module Pivot = PivotF(Dom)(C)(Out.D)
-    type v = Dom.v
     let gen =
@@
                     (Code.update c Idx.succ) \} ))
-            (Out.make_result b) \} 
-    in
+            (Out.make_result b) \} in
     .<fun a -> .~(runM (dogen .<a>.)) >.
@@
 generator expression.
-
-\oleg{later on, we can write the following just let fpc = .<...>.}
 

### 2005-04-08 paper/metamonads.tex
edit everything from p.1 to p.7.  checkpointing.

@@
 \newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
-\setlength{\parskip}{0pt}
+% \setlength{\parskip}{0pt}
 \newlength{\mathindent}
@@
 \parskip 0pt plus 1pt minus 2pt
-\textfloatsep 4pt plus 2pt minus 4pt % Less space around figures
+\textfloatsep 4pt plus 2pt minus 3pt % Less space around figures
 \abovecaptionskip 0pt plus 0pt minus 2pt
 \belowcaptionskip 0pt plus 0pt minus 2pt
+
+\intextsep 2pt plus 0pt minus 1pt
 
@@
 inline and simplify such code, this is labour-intensive and 
-error-prone.  Furthermore the resulting code is frequently unreadable,
+error-prone.  The resulting code is frequently unreadable,
 and thus unmaintainable.
@@
 
-Our goal was to implement a completely modular, textbook-style 
-modular design \cite{journals/cacm/parnas72a}, yet 
-have our code be very efficient.  Gaussian elimination, as it is a  
-well-known algorithm, already contains many different aspects 
+Our goal was to implement a textbook-style modular design 
+\cite{journals/cacm/parnas72a}, yet 
+have our code be very efficient.  Gaussian elimination, as a  
+well-known algorithm with many different aspects 
 \cite{carette04}, was chosen as a representative example.
@@
 frequently have a larger cost still.  Of course, if one is faced with
-writing 60 different implementations of the ``same'' algorithm
-\oleg{citation?}, this
+writing 35 different implementations of the ``same'' algorithm
+\cite{carette04}, this
 abstraction cost starts to look cheap!  Although bad programmers have
@@
 generative system to re-parse the generative code to inspect it, such
-an approach rapidly degenerates into complete chaos \oleg{is there a
-  good citation?}.  In languages
+an approach rapidly degenerates into complete chaos.
+% \oleg{is there a good citation?}
+% \jacquesP{none that I am aware of}
+In languages
 that natively offer reflection and reification (either at no cost, 
 such as in LISP and Scheme, or at a modest cost in languages like 
-Maple), this is actually feasible, even easy.  But this is fragile
+Maple), inspection is actually feasible, even easy.  But this is fragile
 and error-prone, since the languages offer little support for catching
@@
 \end{itemize}
-The question we want to answer is:
+The questions we want to answer is:
 Can we achieve all of the abstractions needed for a flexible 
@@
 also be remarked that in general, intensional code analysis
-\cite{Pueschel:05,Kennedy01Telescoping,dongarra7,Veldhuizen:2004} 
-essentially requires
-one to insert both an optimizing compiler and an automated theorem
-proving system into the code generating system.  While this is
+essentially requires one to insert both an optimizing compiler and 
+an automated theorem proving system into the code generating system
+\cite{Pueschel:05,Kennedy01Telescoping,dongarra7,Veldhuizen:2004}.
+While this is
 potentially extremely powerful and an exciting area of research,
@@
 This work makes the following contributions:
+\vspace*{-2pt}
 \begin{itemize}
+    \item use of generative functors for meta-programming
     \item orderly aspect weaving
-    \item mdo notation
+    \item mdo notation for greater monadic code clarity
     \item use of monads for meta-programming
-    \item use of generative functors
-    \item use of type constraints to ensure correctness
+    \item use of type constraints to encode domain knowledge and 
+		ensure correctness
 \end{itemize}
@@
 features: to construct a code expression, to combine them, and to
-execute them. The figure XXX shows the simplest code generator |one|,
-and more complex generators.
-
-\begin{figure*}
+execute them. Figure~\ref{easycode} shows the simplest code generator |one|,
+as well as more complex generators.
+
+\begin{figure}\label{easycode}
 \begin{code}
@@
 \caption{Code generation and combinators. $\Longrightarrow$ shows the
-  result of the evaluation of the preceeding expression}
-\end{figure*}
+  result of evaluating the preceeding expression}
+\end{figure}
 
@@
 A more complex expression |param_code1| has two occurrences of 
-|plus y one|. Depending on the implementation of |plus| that may be quite a
+|plus y one|. Depending on the implementation of |plus|, this may be quite a
 complex computation, and so we would rather not do it twice. We may be
@@
 But that seems impossible because |letgen exp| has to generate
-the expression\\|.<let t = exp in body>.| although |letgen| does not have
+the expression |.<let t = exp in body>.| although |letgen| does not have
 the |body| yet. The body needs a temporary identifier |.<t>.| that
@@
 
-Comparing the code that did let-insertion at the generating stage\\
+\noindent Comparing the code that did let-insertion at the generating stage\\
 |let ce = (plus y one) in  plus ce (plus x ce)|\\
@@
 The |do|-notation of Haskell \cite{Haskell98Report} shows that it is possible
-to write essentially CPS code in conventional-looking style. The
+to write CPS code in conventional-looking style. The
 |do| notation is the notation for writing monadic code \cite{moggi-notions}.
@@
 
-\begin{figure*}\label{ourmonad}
+\begin{figure}\label{ourmonad}
 \begin{code}
@@
 \caption{Our monad}
-\end{figure*}
+\end{figure}
 
@@
 invoking its natural continuation in the tail position. In our case
-however the control effects manifest themselves as code generation.
+the control effects manifest themselves as code generation.
 
@@
 result of invoking its continuation |k|. Rather, its result is a |let|
-code expression. Such a behavior is symptomatic of cntrol operators
+code expression. Such a behavior is symptomatic of control operators
 (in particular, |abort|).
@@
 initial continuation |k0|. We can now re-write our |param_code2|
-example of the previous section as |param_code3|, see Fig. XXX.
-
-\begin{figure*}
+example of the previous section as |param_code3|.
+
+\begin{figure}
 \begin{code}
@@
   ifL testc thc elc\}
-let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
+let gen a i = ifM' (ret .<(.~i) >= 0>.) 
+                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
  in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
-\evalresult{.<fun a_1 -> fun i_2 -> 
+\evalresult{.<fun a_1 i_2 ->  
 let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.}
-let gen a i = ifM (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
+let gen a i = ifM (ret .<(.~i) >= 0>.) 
+                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
  in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
-\evalresult{.<fun a_1 -> fun i_2 ->
+\evalresult{.<fun a_1 i_2 -> 
 if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.}
 \end{code}
-\end{figure*}
+\end{figure}
 % param_code3 plus one;;
@@
 That does not seem like much of an improvement. With the help of
-camplp4 pre-processor (cite), we introduce the do notation (give URL
-or cite the message on the MetaOCaml list), patterned after the
-do-notation of Haskell (cite). The function |param_code4|, written in
+camplp4 pre-processor, we introduce the do notation (see
+Appendix A), patterned after the
+do-notation of Haskell. The function |param_code4|, written in
 the do-notation, is equivalent to |param_code3| -- in fact, the camlp4
@@
 We soon notice the need for another |ifM| function, with the same
-interface (see |ifM| in Figure~\ref{ourmonad}). The difference between them is
-apparent, see Fig. XXX: in the code with |ifM'|, the let-insertion
-happend \emph{before} the if-expression, that is, before the test that
-the index |i| is positive. That is, if |i| turned out
-negative, |a.(i)| will generate an out-of-bound array access
+interface (see Figure~\ref{ourmonad}). The difference between them is
+apparent: in the code above with |ifM'|, the let-insertion
+happened \emph{before} the if-expression, that is, before the test that
+the index |i| is positive. If |i| turned out
+negative, |a.(i)| would generate an out-of-bound array access
 error. On the other hand, the code with |ifM| accesses the array only
@@
 truly an effect and we have to be clear about the sequencing of
-effects, when generating control constructions such as conditionals.
-The form |ifM| handles the such effects correctly. In our code, we
+effects when generating control constructions such as conditionals.
+The form |ifM| handles such effects correctly. We
 need similar operators for other Ocaml control forms: for generating
@@
 different domain must satisfy:
-\begin{figure*}
+\begin{figure}
 \begin{code}
@@
 \end{code}
-\end{figure*}
+\end{figure}
 \noindent  The first item to notice is that the types above are
@@
 
-\begin{figure*}
+\begin{figure}
 \begin{code}
@@
 \end{code}
-\end{figure*}
+\end{figure}
 

### 2005-04-08 paper/metamonads.tex
spelling corrections
Mostly tricks to cut down on length.

@@
  \and
-Oleg Kyselyov\inst{2}
+Oleg Kiselyov\inst{2}
 }
@@
 \and
-Monterey, CA 93943}
+FNMOC, Monterey, CA 93943}
 
@@
 This work makes the following contributions:
-\vspace*{-2pt}
+\vspace*{-6pt}
 \begin{itemize}
@@
 Elimination algorithm family in completely separate, independent modules.
-We then briefly discuss related work in section~\ref{related}, outline
-some work still to do in section~\ref{future}, and finally draw some 
+We then briefly discuss related work in section~\ref{related} and outline
+some work still to do, and finally draw some 
 conclusions in section~\ref{conclusion}.  As an appendix, we show
@@
 
-Now that we know that we have the appropriate basic techniques for
-writing our code generators, we need to worry about modularizing them
-appropriately.  Here we can really take advantage of the fact that
-MetaOCaml is a staged language: it is not really relevant how much
+Now that we have the appropriate basic techniques for
+writing our code generators, we need to worry about modularization.
+Here we can really take advantage of the fact that
+MetaOCaml is a staged language: it is not particularly relevant how much
 computation happens at staging time, as long as the generated code
@@
 apply these principles to the study of Gaussian Elimination, we need
-to understand what changes between different implementations, and 
+to understand what are the changes between different implementations, and 
 what concerns need to be addressed.  We also need to study the degree
 to which these concerns are independent.
-
-A careful study of Gaussian Elimination \cite{carette04} shows that
-(at least) the following variations occur:
+A study of Gaussian Elimination \cite{carette04} shows that
+the following variations occur:
 \begin{enumerate}
+\vspace*{-6pt}
 	\item \textbf{Domain}: In which (algebraic) domain do the matrix
@@
 		Division Ring.  In the roughly 85 pieces of code
-		surveyed \oleg{citation?},
-		20 different domains were encountered.
+		surveyed 20 different domains were encountered.
+		% \oleg{citation?} \jacques no citation needed, carette04 is
+		% already cited above.
 	\item \textbf{Container}: Whether the matrix
@@
 		routines like Maple's \texttt{LinearAlgebra:-LUDecomposition} have
-		up to $2^6 + 2^5 + 2^2 = 100$ possible outputs.
+		up to $2^6 + 2^5 + 2^2 = 100$ outputs.
 	\item \textbf{Fraction-free}: Whether the Gaussian Elimination
@@
 		exact (remainder-free) division.
-	\item \textbf{Pivoting}: Whether to use no pivoting at all, only
+	\item \textbf{Pivoting}: Whether to use no pivoting, 
 		column-wise pivoting, or full pivoting.
@@
 recur frequently:
+\vspace*{-6pt}
 \begin{enumerate}
@@
 $+$, $*$, (unary and binary) $-$, some kind of division, normalization,
-and potentially a relative size measure.  We definitely need at least
-\emph{exact} division to exist, but if full division is available, than
+and potentially a relative size measure.  We need at least
+\emph{exact} division to exist, but if full division is available, that
 can be used as well.  Normalization is frequently trivial for domains
-where elements have unique representations, but in many applications
-we have domains (like polynomials in sum-of-product form) where this
+where elements have unique representations, but for some
+domains (like polynomials in sum-of-product form) this
 is not the case.  The simplest case of such an abstraction was given
-in Section XXX, |param_code1|. In the latter, code-generators such as
+in Fig.~\ref{easycode}, |param_code1|. In the latter, code-generators such as
 |plus| and |one| were passed as arguments. We see however that we 
@@
 of the grouping offered by regular records, we use Ocaml
-\emph{structures} (i.e., modules) (cite Ocaml module system tutorial?)
-so we can take advantage extensions, type abstraction and constraints,
-and especially parameterized structures (i.e., \emph{functors}).
-We first define the type of the domain, the signature |DOMAIN| to which
-different domain must satisfy:
-\begin{figure}
+\emph{structures} (i.e., modules) 
+% (cite Ocaml module system tutorial?)
+% normally yes, but our references are what's making this long right now!
+so we can take advantage of extensions, type abstraction and constraints,
+and especially parameterized structures (\emph{functors}).
+We first define the type of the domain, the signature |DOMAIN| which
+different domains must satisfy:
+\begin{figure}[h]
 \begin{code}
@@
 \end{figure}
+
 \noindent  The first item to notice is that the types above are
@@
 One particular instantiation of the domain (a particular structure of
-|DOMAIN| signature) is the |int| domain. The notattion
+|DOMAIN| signature) is the |int| domain. The notation
 |module IntegerDomain : DOMAIN| makes the compiler
@@
 to use that structure as a |DOMAIN|. In any case, the errors such as
-missing ``methods'' or methods of wrong types will be caught
+missing ``methods'' or methods with incorrect types will be caught
 statically, even \emph{before} any code generation takes place. The
@@
 looked daunting to some, the implementation is quite straightforward.
-It is equally simple to make implementations for |float| and for
-|Num.num| (arbitrary precision exact rational numbers).  With scarcely
-more work (as these are not built-in and the arithmetic must be
-implemented), $\mathbb{Z}_p$, $\mathbb{Z}\left[x\right]$,
-$\mathbb{Q}\left(x\right)$ and more complex domains can be done as
-well.
+|float| and |Num.num| (arbitrary precision exact rational numbers)
+are equally simple, and with scarcely
+more work, $\mathbb{Z}_p$, $\mathbb{Z}\left[x\right]$,
+$\mathbb{Q}\left(x\right)$ as well more complex domains can be implemented.
 
@@
   type contr  type res
-  module D : DETERMINANT   module R : RANK
-  module P : TRACKPIVOT
+  module D : DETERMINANT   module R : RANK   module P : TRACKPIVOT
   val make_result : ('a,contr) code -> 
@@
   type res = contr * Det.outdet * int
-  module D = Det   module R = Rank
-  module P = DiscardPivot
+  module D = Det   module R = Rank   module P = DiscardPivot
   let make_result b = mdo \{ det  <-- D.fin ();  rank <-- R.fin ();
@@
 functions of that module generate no code. For integer matrices, we
-have to track some aspects of the determinant even if we don't output
-it. The determinant tracking aspect is complex becasue tracking
+have to track some aspects of the determinant, even if we don't output
+it. The determinant tracking aspect is complex because tracking
 variables, if any, are to be declared at the beginning of GE; the sign
@@
 pivoting. We use |lstate| to pass the tracking state, which may be, for
-example, a piece of code for the value of the type |float ref| among
+example, a piece of code for the value of the type |Dom.v ref| among
 various determinant-tracking functions. The |lstate| is a part of the
@@
 (or may not) use the monadic state for passing its own data -- we make
-an extensive use of open records (list of polymorphic variants
-appeared to be the easiest way to implement such a union, in a pure
+extensive use of open records (a list of polymorphic variants
+appeared to be the easiest way to implement such a union, in a purely
 functional way). This lets us freely compose determinant-tracking,
 rank-tracking, and other aspects.
-
-The GE generator functor is as follows. It is
-parameterized by the domain, container, pivoting policy (full, row,
-nonzero, nopivoting), update policy (that use either ``fraction-less'
-or full division), and by what to yield as the result. Some of the
-argument modules such as |PIVOT| are functors themselves (parameterized
-by the domain, the container, and the determinant functor). The sharing
-constraints express obvious constraints on the instantiation of |Gen|,
-for example, pivoting, determinant etc. components all use the same
-domain. It must be stressed that all structures (i.e., module
-instances) are stateless, and so we never have to worry that different
-aspect functors (such as |CONTAINER2D| and |PIVOT|) are instantiated
-with different but type-compatible instances of |DOMAIN|. That is, we
-are not concerened at all about the value sharing. Aspects such as
-determinant tracking may be stateful so that the determinant update
-code have access to the determinant tracking variables declared
-previsouly. But that state is handled via the monadic state. As we
-have shown, open unions makes the overall monadic state compositional
-with respect to the state of various aspects.
 
@@
 
+The GE generator functor itself is 
+parameterized by the domain, container, pivoting policy (full, row,
+nonzero, nopivoting), update policy (that use either ``fraction-less'
+or full division), and by what to yield as the result. Some of the
+argument modules such as |PIVOT| are functors themselves (parameterized
+by the domain, the container, and the determinant functor). The sharing
+constraints express obvious constraints on the instantiation of |Gen|,
+for example, pivoting, determinant etc. components all use the same
+domain. It must be stressed that all structures (i.e., module
+instances) are stateless, and so we never have to worry that different
+aspect functors (such as |CONTAINER2D| and |PIVOT|) are instantiated
+with different but type-compatible instances of |DOMAIN|. That is, we
+are not concerened at all about the value sharing. Aspects such as
+determinant tracking may be stateful so that the determinant update
+code have access to the determinant tracking variables declared
+previously. But that state is handled via the monadic state. As we
+have shown, open unions makes the overall monadic state compositional
+with respect to the state of various aspects.
+
 In addition to the ``regular'' type sharing constraints shown in the
@@
 in the following structure of the |UPDATE| signature:
-
-\begin{code}
+\vspace*{-5pt}\begin{code}
 module DivisionUpdate
@@
 \end{code}
-
-The structure implements update policy that can use |Dom.div| operation without
-restrictions. That is possible only if the domain has such an
-operation. A domain such as integer domain may still provide |Dom.div|
+\vspace*{-4pt}
+This structure implements an update policy that can use the 
+|Dom.div| operation without
+restrictions -- which is possible only if the domain has such an
+operation. A domain such as the integer domain may still provide |Dom.div|
 of the same type, but that operation can only be used when we are sure
-that the division is exact. Using the integer domain with
-|DivisionUpdate| will lead to the compile-time error, when compiling
+that the division is exact. But using it with
+|DivisionUpdate| will lead to a compile-time error, when compiling
 the \emph{generator} code. Thus, in some cases we can use module types
@@
 via the types of module members.
-
+\vspace*{-5pt}
 \begin{code}
@@
 \end{code}
-
+\vspace*{-5pt}
 We can instantiate the |Gen| functor as shown above and inspect the
 generated code, e.g., by printing |GenIA1.gen|. The code can then be
-``compiled'' as |!. GenIA1.gen| or with off-shiring (cite). The code
+``compiled'' as |!. GenIA1.gen| or with off-shoring. The code
+% (cite).  No space!
 for |GenFV4| shows full pivoting, determinant and rank tracking. The
@@
 tracking: no declaratrion of spurious variables, no extra tests,
-etc. The code appears as if determinant tracking aspect did not exist
+etc. The code appears as if the determinant tracking aspect did not exist
 at all. The generated code for the above and other instantiations of
-|Gen| can be examined at (cite).
-
-\section{Related work}\label{related}
+|Gen| can be examined at \cite{metamonadsURL}.
+
+\section{Related and future work}\label{related}
 
@@
 However, the latter work used only |retN| of all monadic operations,
-and used fixpoints (for performing iterations at the generation time).
+and used fixpoints (for performing iterations at generation time).
 In this paper we do not use monadic fixpoints (because the generator
@@
 
-Mention |Blitz++| and template meta-programming. Can do similar
-eliminating of abstraction. Can even encode some domain-specific
-knowledge in the form of concepts. But inlining critically depends on
-the compiler's fully inlining of all methods. Furthermore, all errors
+|Blitz++| \cite{Veldhuizen:1998:ISCOPE} and in general template 
+meta-programming in C++ achieve similar results of eliminating levels
+of abstraction.  Using traits and concepts, some domain-specific
+knowledge can also be encoded.  However inlining critically depends on
+the compiler's fully inlining of all methods, and as has been reported
+in the literature, this can be challenging to insure. Furthermore, all errors
 (such as type errors and concept violation errors, that is,
@@
 code. It is immensely difficult to correlate errors (e.g., line numbers) 
-to the ones in the generator itself. Furthermore, I think templates
-can't be local, right? So, a lot of code generation is spread around
-file-level entities.
-
-SPIRAL does the intentional code analysis. Probably ATLAS as well.
-
-\jacques{What do we consider \emph{related work} here?  Do we essentially gloss 
-over related work in numerical software, code generation in general,
-aspect-oriented stuff?  We don't want this section to be too long, but
-not so short that we make it seem like we're inventing more than we
-really are}
-\oleg{Yes, let's gloss over ATLAS, SPIRAL, Axiom. I wonder what would
-  be the use of functors? Something by David McQueen or R. Harper.
-  In any way, mention that nobody before used functors to abstract
-  code generators -- or mix functors and multi-stage programming.}
-
-\section{Future work}\label{future}
+to the ones in the generator itself. Furthermore, as templates
+cannot be local, a lot of code generation is scattered around instead
+of being modularized.
+
+ATLAS \cite{ATLAS} is another successful project in this area.  However
+they use much simpler weaving technology, which leads them to note
+that \emph{generator complexity tends to go up along with flexibility, 
+so that these routines become almost insurmountable barriers to 
+outside contribution}.  Our results show how to surmount this barrier,
+by building modular, composable generators.
+SPIRAL \cite{Pueschel:05} is another such even more ambitious project.
+But SPIRAL does intentional code analysis, relying on a set of code
+transformation ``rules'' which make sense, but which are not proved 
+to be either complete or confluent.  The strength of both of these
+project relies on their platform-specific optimizations performed
+via search techniques, something we have not attempted here.
+
+To the best of our knowledge, nobody has yet used functors to
+abstract code generators, or even mixed functors and 
+multi-stage programming.
+% \oleg{Yes, let's gloss over ATLAS, SPIRAL, Axiom. I wonder what would
+%   be the use of functors? Something by David McQueen or R. Harper.
+%   Mention that nobody before used functors to abstract
+%   code generators -- or mix functors and multi-stage programming.}
+
 connections with delimited continuations: making notation

### 2005-04-08 paper/metamonads.tex
new intro

@@
 
-\usepackage[medium,compact]{titlesec}
+%\usepackage[medium,compact]{titlesec}
 
@@
 
+% previous introduction is saved at the end
 \section{Introduction}
 
-In high-performance, symbolic, and numeric computing, there is a well-known
-issue of balancing between maximal performance and the level of
-abstraction at which code is written.  Furthermore, already in
-linear algebra, there is a wealth of different concerns that 
-\emph{may} need to be addressed.  While it is possible to manually
-inline and simplify such code, this is labour-intensive and 
-error-prone.  The resulting code is frequently unreadable,
-and thus unmaintainable.
-
-There are a number of projects that have attempted to address these issues
-\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser94algorithmoriented,BOOST,POOMA,ATLAS}, 
-and they have generally succeeded in generating
-efficient code, but the cost was frequently the loss of modularity
-of the design.  In other words, to allow the various aspects to
-weave with each other \emph{efficiently}, information which should
-have been kept hidden leaked from module to module.  This seriously
-affects maintainability and scalability.  Furthermore, current architectures
+In high-performance, symbolic, and numeric computing, there is a
+well-known issue of balancing between maximal performance and the
+level of abstraction at which code is written.  Furthermore, already
+in linear algebra, there is a wealth of different aspects that
+\emph{may} need to be addressed. For example, implementations of the
+widely used Gaussian Elimination (GE) algorithm --- the running
+example of our paper --- may need to account for the representation of
+the matrix, whether to compute and return the detrminant or rank, how
+and whether search for pivot, etc. Furthermore, current architectures
 demand more and more frequent tweaks which, in general, cannot be done by the
 compiler because the tweaking often involves domain knowledge. 
-
-Our goal was to implement a textbook-style modular design 
-\cite{journals/cacm/parnas72a}, yet 
-have our code be very efficient.  Gaussian elimination, as a  
-well-known algorithm with many different aspects 
-\cite{carette04}, was chosen as a representative example.
-
-Whether one prefers to code in an imperative, object-oriented, functional, 
-or even in a mixed style, all abstraction mechanisms in current
-languages have a non-zero run-time cost.  Whether it is procedure,
-method or function call, these have a cost; worse still, higher level
-and/or higher order abstractions (like Functors or AspectJ's aspects)
-frequently have a larger cost still.  Of course, if one is faced with
-writing 35 different implementations of the ``same'' algorithm
-\cite{carette04}, this
-abstraction cost starts to look cheap!  Although bad programmers have
-been known to resort to cut\&paste programming, no self-respecting
-programmer would be caught doing that.
-The common solution is to turn to code generation.
-
-While many people still make do with string-based code generation,
-this is not much different than writing code in a dynamically typed
-interpreted scripting language.  It works, it is convenient, but it
-does not scale well, it is a debugging nightmare, and usually suffers
-most from Software Aging \cite{parnas_aging}.  As with other tasks in
-programming, the answer is to turn to statically typed languages,
-which provide strong guarantees about the soundness of the resulting
-code.  MetaOCaml \cite{CTHL03,metaocaml-org} in particular stands
-out: not only it guarantees well-formedness and the absense of problems 
-such as accidental variable capture \cite{HygienicMacros}. MetaOcaml also
-statically ensures that both the code generator and the generated code
-are well-typed \cite{TahaSheard97,TahaThesis}.  While Template
-programming in C++ and Template Haskell do offer some amount of static
-typing \cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml
-offers stronger guarantees.  As well, since MetaOCaml is
-implemented on top of the very solid Ocaml language, one accrues
-additional benefits.
-
-While it is theoretically feasible in a strings-based
-generative system to re-parse the generative code to inspect it, such
-an approach rapidly degenerates into complete chaos.
-% \oleg{is there a good citation?}
-% \jacquesP{none that I am aware of}
-In languages
-that natively offer reflection and reification (either at no cost, 
-such as in LISP and Scheme, or at a modest cost in languages like 
-Maple), inspection is actually feasible, even easy.  But this is fragile
-and error-prone, since the languages offer little support for catching
-errors early.
-
-In MetaOCaml, generated code cannot be examined and optimized at all. It must
-be generated just right.  For simple code, this is relatively easy
-(see \cite{TahaThesis} for many examples).  For more complex examples,
-new techniques are necessary.  \cite{KiselyovTaha} introduces one
-such tool: abstract interpretation.  But more problems remain
-\cite{Padua:MetaOcaml:04}:
-\begin{itemize}
-    \item Generating names in the presence of ``if'', ``while'', and other
-        control operators. \cite{KiselyovTaha} showed how to solve
-        this problem in simple cases.
-    \item Making continuation-passing style (CPS) code clear.  Many
-        authors understandably shy away from CPS code as it quickly
-        becomes unreadable.  But this is needed for proper name
-        generation.
-    \item Compositionality of various interleaving aspects, as well
-        as expressing dependencies among them.  For example, if we
-        choose to compute the determinant of a matrix while performing
-        Gaussian Elimination, this requires inserting code in 
-        several specific places in the algorithm.  The method to
-        compute this determinant crucially depends on whether 
-        a division-based or a fraction-free Gaussian Elimination is
-        used.
-    \item As mentionned above, there are dependencies between
-        certain algorithmic choices and properties of the
-        underlying domain, and we would like to statically guarantee
-        that inappropriate combinations cannot be chosen.  For instance,
-        one should not attempt to use full division when dealing
-        with matrices of exact integers, nor is it worthwhile to use
-        full pivoting on a matrix of rational integers ($\mathbb Q $).
-\end{itemize}
-The questions we want to answer is:
-Can we achieve all of the abstractions needed for a flexible 
-Gaussian Elimination, and can
-we eliminate all of that abstraction's overhead in the generated code?
-Of course, we want to do this without using
-intensional code analysis -- that is, maintaining the
-guarantees of MetaOCaml and using a purely generative approach.
-
-Maintaining generativity is not just an interesting exercise:
-one gets a stronger equational theory \cite{Taha2000}, and avoids
-the danger of creating unsoundness \cite{TahaThesis}.  It should
-also be remarked that in general, intensional code analysis
-essentially requires one to insert both an optimizing compiler and 
-an automated theorem proving system into the code generating system
+The survey \cite{carette04} of
+Gaussian elimination implementation in an industrial package Maple
+found XXX various aspects and XXX different implementations of the
+algorithm.  We can manually write each of these implementations
+optimizing for particular aspects and using cut-and-paste to ``share''
+similar pieces of code. \oleg{I don't think we need to say much here
+-- the shear number of various implementations should be enough of a
+clue.} We can write a very generic GE procedure that accounts for
+all the aspects with appropriate abstractions (cite axiom?). The
+abstraction mechanisms however -- be they procedure, method or a
+function call -- have a significant cost, especially for
+high-performance numerical computing (cite? At least carette04).
+
+A more appealing (acceptable? indispensable? need a better word)
+approach is generative programming
+\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser94algorithmoriented,BOOST,POOMA,ATLAS}.
+The approach is not without problems, e.g., making sure that the
+generated code is well-formed. This is a challenge in string-based
+generation systems, which generally do not offer such guarantees and
+therefore make it very difficult to determine which part of the
+generator is at fault when the generated code cannot be parsed. Other
+problems is preventing accidental variable capture (so-called hygiene
+\cite{HygienicMacros}) and ensuring the generated code is
+well-typed. Lisp-style macros, Scheme hygienic macros, camlp4
+preprocessor (cite XXX), C++ template meta-programming, Template
+Haskell \cite{conf/dagstuhl/CzarneckiOST03} solve some of the above
+problems. Of the widely available maintainable languages, only
+MetaOCaml \cite{CTHL03,metaocaml-org}  solves all the above problems
+including the well-typing of the generated code \cite{TahaSheard97,TahaThesis}.
+
+But more difficult problems remain. Is the generated code optimal? Do
+we still need post-processing to eliminate common subexpressions and
+fold constants, remove redundant bindings? Is the generator readable,
+resembling the original algorithm, and extensible? Are the aspects
+truly modular? Can we add another aspect to it or another instance of
+the existing aspect without affecting the existing ones? Finally, can
+we express domain-specific knowledge, e.g., one should not attempt to
+use full division when dealing with matrices of exact integers, nor is
+it worthwhile to use full pivoting on a matrix of rational integers
+($\mathbb Q $).
+
+MetaOCaml is \emph{generative}: generated code can only be treated as
+a black box: it cannot be inspected and it cannot be post-processed
+(i.e., no intensional analysis). This approach gives a stronger
+equational theory \cite{Taha2000}, and avoids the danger of creating
+unsoundness \cite{TahaThesis}. Furthermore, intensional code analysis
+essentially requires one to insert both an optimizing compiler and an
+automated theorem proving system into the code generating system
 \cite{Pueschel:05,Kennedy01Telescoping,dongarra7,Veldhuizen:2004}.
-While this is
-potentially extremely powerful and an exciting area of research,
-it is also extremely complex, which means that it is currently more
-error-prone and difficult to ascertain the correctness of the 
-resulting code.
-
-This work makes the following contributions:
+While this is potentially extremely powerful and an exciting area of
+research, it is also extremely complex, which means that it is
+currently more error-prone and difficult to ascertain the correctness
+of the resulting code.
+
+Therefore, in MetaOCaml, code must be generated just right (see
+\cite{TahaThesis} for many simple examples).  For more complex
+examples, new techniques are necessary, e.g., abstract interpretation
+\cite{KiselyovTaha}.  But more problems remain
+\cite{Padua:MetaOcaml:04}: generating binding statements (``names''),
+especially when generating loop bodies or conditional branches; making
+continuation-passing style (CPS) code clear.  Many authors
+understandably shy away from CPS code as it quickly becomes
+unreadable.  But this is needed for proper name generation.
+The problems of compositionality of code generators, expressing
+dependencies among them and domain-specific knowledge remain.
+
+In this paper we report on progress of solving these problems using GE
+as our running example. Specifically, our contributions:
 \vspace*{-6pt}
 \begin{itemize}
-    \item use of generative functors for meta-programming
-    \item orderly aspect weaving
-    \item mdo notation for greater monadic code clarity
-    \item use of monads for meta-programming
-    \item use of type constraints to encode domain knowledge and 
-		ensure correctness
+    \item Extending monad of \cite{KiselyovTaha} for generating
+      binding statements when generating control structures such as
+      loops and conditionals. We argue that code generation is an
+      effect and so requires something like a monad for proper
+      serialization.
+    \item Implementation of the |mdo|-notation (patterned after
+      |do|-notation of Haskell) to make monadic code readable
+    \item Use of functors (including higher-order functors) to
+      modularize the generator, express aspects (including results of
+      various types) and \emph{assure composability of aspects} even
+      for aspects that use state and have to be accounted in many
+      places in the generated code
+    \item use functor type sharing constraints to encode domain-specific
+      knowledge
 \end{itemize}
 
-The rest of this paper is structured as follows:
-The next section introduces code generation in MetaOCaml, the problem
-of name generation, and how continuation-passing style is a general
-tool for solving that problem.  We also introduce
-the monad that greatly simplifies the task of dealing with
-possibly effect-inducing CPS code combinators.  Section~\ref{functors}
-shows how one can use the advanced module system of OCaml, namely its
-parametrized modules, to encode all of the aspects of the Gaussian
-Elimination algorithm family in completely separate, independent modules.
-We then briefly discuss related work in section~\ref{related} and outline
-some work still to do, and finally draw some 
-conclusions in section~\ref{conclusion}.  As an appendix, we show
-the details of the syntax extension for monadic programming in Ocaml
-and MetaOCaml.
-
-The first author wishes to thank Cristiano Calgano for his help in
-adapting camlp4 for use with MetaOCaml.
-
-\section{The problem of generating names, CPS, and monad}\label{CPS}
+The rest of this paper is structured as follows: The next section
+introduces code generation in MetaOCaml, the problem of name
+generation, and continuation-passing style (CPS) as a general
+solution.  We also introduce the monad and the issues of generating
+control statements. Section~\ref{functors} describes the use of
+parametrized modules of OCaml to encode all of the aspects of the
+Gaussian Elimination algorithm family in completely separate,
+independent modules.  We briefly discuss related work in
+section~\ref{related}. We then outline the future work and conclude.
+Appendices give samples of the generated code (which is availablke in
+full at XXX).
+
+We wish to thank Cristiano Calgano for his help in adapting camlp4 for
+use with MetaOCaml.
+
+\section{Generating binding statements, CPS, and monad}\label{CPS}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% And now, a better -- and shorter -- write-up.
-
-The essence of our approach is assembling the code of the program by
-combining instances of various aspects together. In the case of
-Gaussian Elimination, the algorithmic aspects are pivot determination, 
-swapping of
-rows and columns, determinant tracking, etc. A primitive code
-generator yields a code expression.  A combinator takes (complete and
-well-typed) code expressions and combines them into a composite expression
-(which is also statically guaranteed to be well-formed and well-typed).
-
-We will be using MetaOcaml which, as an instance of a multi-stage
+
+We build code generators out of primitive ones using code generation 
+combinators. MetaOcaml, as an instance of a multi-stage
 programming system \cite{TahaThesis}, provides exactly the needed
@@
 case-matching statements and for- and while-loops.
+
+% old intro
+\section{Introduction}
+
+In high-performance, symbolic, and numeric computing, there is a well-known
+issue of balancing between maximal performance and the level of
+abstraction at which code is written.  Furthermore, already in
+linear algebra, there is a wealth of different concerns that 
+\emph{may} need to be addressed.  While it is possible to manually
+inline and simplify such code, this is labour-intensive and 
+error-prone.  The resulting code is frequently unreadable,
+and thus unmaintainable.
+
+There are a number of projects that have attempted to address these issues
+\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser94algorithmoriented,BOOST,POOMA,ATLAS}, 
+and they have generally succeeded in generating
+efficient code, but the cost was frequently the loss of modularity
+of the design.  In other words, to allow the various aspects to
+weave with each other \emph{efficiently}, information which should
+have been kept hidden leaked from module to module.  This seriously
+affects maintainability and scalability.  Furthermore, current architectures
+demand more and more frequent tweaks which, in general, cannot be done by the
+compiler because the tweaking often involves domain knowledge.
+
+Our goal was to implement a textbook-style modular design 
+\cite{journals/cacm/parnas72a}, yet 
+have our code be very efficient.  Gaussian elimination, as a  
+well-known algorithm with many different aspects 
+\cite{carette04}, was chosen as a representative example.
+
+Whether one prefers to code in an imperative, object-oriented, functional, 
+or even in a mixed style, all abstraction mechanisms in current
+languages have a non-zero run-time cost.  Whether it is procedure,
+method or function call, these have a cost; worse still, higher level
+and/or higher order abstractions (like Functors or AspectJ's aspects)
+frequently have a larger cost still.  Of course, if one is faced with
+writing 35 different implementations of the ``same'' algorithm
+\cite{carette04}, this
+abstraction cost starts to look cheap!  Although bad programmers have
+been known to resort to cut\&paste programming, no self-respecting
+programmer would be caught doing that.
+The common solution is to turn to code generation.
+
+While many people still make do with string-based code generation,
+this is not much different than writing code in a dynamically typed
+interpreted scripting language.  It works, it is convenient, but it
+does not scale well, it is a debugging nightmare, and usually suffers
+most from Software Aging \cite{parnas_aging}.  As with other tasks in
+programming, the answer is to turn to statically typed languages,
+which provide strong guarantees about the soundness of the resulting
+code.  MetaOCaml \cite{CTHL03,metaocaml-org} in particular stands
+out: not only it guarantees well-formedness and the absense of problems 
+such as accidental variable capture \cite{HygienicMacros}. MetaOcaml also
+statically ensures that both the code generator and the generated code
+are well-typed \cite{TahaSheard97,TahaThesis}.  While Template
+programming in C++ and Template Haskell do offer some amount of static
+typing \cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml
+offers stronger guarantees.  As well, since MetaOCaml is
+implemented on top of the very solid Ocaml language, one accrues
+additional benefits.
+
+While it is theoretically feasible in a strings-based
+generative system to re-parse the generative code to inspect it, such
+an approach rapidly degenerates into complete chaos.
+% \oleg{is there a good citation?}
+% \jacquesP{none that I am aware of}
+In languages
+that natively offer reflection and reification (either at no cost, 
+such as in LISP and Scheme, or at a modest cost in languages like 
+Maple), inspection is actually feasible, even easy.  But this is fragile
+and error-prone, since the languages offer little support for catching
+errors early.
+
+In MetaOCaml, generated code cannot be examined and optimized at all. It must
+be generated just right.  For simple code, this is relatively easy
+(see \cite{TahaThesis} for many examples).  For more complex examples,
+new techniques are necessary.  \cite{KiselyovTaha} introduces one
+such tool: abstract interpretation.  But more problems remain
+\cite{Padua:MetaOcaml:04}:
+\begin{itemize}
+    \item Generating names in the presence of ``if'', ``while'', and other
+        control operators. \cite{KiselyovTaha} showed how to solve
+        this problem in simple cases.
+    \item Making continuation-passing style (CPS) code clear.  Many
+        authors understandably shy away from CPS code as it quickly
+        becomes unreadable.  But this is needed for proper name
+        generation.
+    \item Compositionality of various interleaving aspects, as well
+        as expressing dependencies among them.  For example, if we
+        choose to compute the determinant of a matrix while performing
+        Gaussian Elimination, this requires inserting code in 
+        several specific places in the algorithm.  The method to
+        compute this determinant crucially depends on whether 
+        a division-based or a fraction-free Gaussian Elimination is
+        used.
+    \item As mentionned above, there are dependencies between
+        certain algorithmic choices and properties of the
+        underlying domain, and we would like to statically guarantee
+        that inappropriate combinations cannot be chosen.  For instance,
+        one should not attempt to use full division when dealing
+        with matrices of exact integers, nor is it worthwhile to use
+        full pivoting on a matrix of rational integers ($\mathbb Q $).
+\end{itemize}
+The questions we want to answer is:
+Can we achieve all of the abstractions needed for a flexible 
+Gaussian Elimination, and can
+we eliminate all of that abstraction's overhead in the generated code?
+Of course, we want to do this without using
+intensional code analysis -- that is, maintaining the
+guarantees of MetaOCaml and using a purely generative approach.
+
+Maintaining generativity is not just an interesting exercise:
+one gets a stronger equational theory \cite{Taha2000}, and avoids
+the danger of creating unsoundness \cite{TahaThesis}.  It should
+also be remarked that in general, intensional code analysis
+essentially requires one to insert both an optimizing compiler and 
+an automated theorem proving system into the code generating system
+\cite{Pueschel:05,Kennedy01Telescoping,dongarra7,Veldhuizen:2004}.
+While this is
+potentially extremely powerful and an exciting area of research,
+it is also extremely complex, which means that it is currently more
+error-prone and difficult to ascertain the correctness of the 
+resulting code.
+
+This work makes the following contributions:
+\vspace*{-6pt}
+\begin{itemize}
+    \item use of generative functors for meta-programming
+    \item orderly aspect weaving
+    \item mdo notation for greater monadic code clarity
+    \item use of monads for meta-programming
+    \item use of type constraints to encode domain knowledge and 
+		ensure correctness
+\end{itemize}
+
+The rest of this paper is structured as follows:
+The next section introduces code generation in MetaOCaml, the problem
+of name generation, and how continuation-passing style is a general
+tool for solving that problem.  We also introduce
+the monad that greatly simplifies the task of dealing with
+possibly effect-inducing CPS code combinators.  Section~\ref{functors}
+shows how one can use the advanced module system of OCaml, namely its
+parametrized modules, to encode all of the aspects of the Gaussian
+Elimination algorithm family in completely separate, independent modules.
+We then briefly discuss related work in section~\ref{related} and outline
+some work still to do, and finally draw some 
+conclusions in section~\ref{conclusion}.  As an appendix, we show
+the details of the syntax extension for monadic programming in Ocaml
+and MetaOCaml.
+
+The first author wishes to thank Cristiano Calgano for his help in
+adapting camlp4 for use with MetaOCaml.

### 2005-04-08 paper/metamonads.tex
more changes

@@
   .<fun x y -> .~(gen .<x>. .<y>.)>.
-let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
-  simplest_param_code plus one
 let param_code1 plus one =
@@
 \end{code}
-\caption{Code generation and combinators. $\Longrightarrow$ shows the
-  result of evaluating the preceeding expression}
+\caption{Code generation and combinators. $\Longrightarrow$ under an
+  expression shows the result of its evaluation}
 \end{figure}
 
-We use MetaOCaml brackets |.<...>.| to generate code expressions -- in
-other words, to construct future-stage computations. We use escapes
-|.~| to perform an immediate code generating computation while we are
-building the future-stage computation (code expression). The immediate
-computation in |simplest_code| includes the evaluation of the function
-|gen|, which in turn applies the function |plus|. The function
-receives code expressions |.<x>.| and |.<y>.| as arguments. At the
-generating stage, we can manipulate code expressions as (opaque)
-values. The function |gen| returns a code expression, which will be
-inlined in the place of the escape. MetaOCaml can print out code
-expressions, so we can see the final generated code, It has
-no traces of the functions |gen| and |plus|: their applications are
-done at the code generation stage.
+We use MetaOCaml brackets |.<...>.| to generate code expressions,
+i.e., to construct future-stage computations. We use escapes |.~| to
+perform an immediate code generating computation \emph{while} we are
+building the future-stage computation. The immediate computation in
+|simplest_code| is the evaluation of the function |gen|, which in turn
+applies |plus|. The function |gen| receives code expressions |.<x>.|
+and |.<y>.| as arguments. At the generating stage, we can manipulate
+code expressions as (opaque) values. The function |gen| returns a code
+expression, which is inlined in the place of the escape. MetaOCaml can
+print out code expressions, so we can see the final generated code. It
+has no traces of |gen| and |plus|: their applications are done at the
+generation stage.
 
@@
 integers, which we can apply: |(.! simplest_code) 1 2|. The original
-|simplest_code| is not the function on integers -- it is a code
+|simplest_code| is not a function on integers -- it is a code
 expression.
@@
 
-A more complex expression |param_code1| has two occurrences of 
-|plus y one|. Depending on the implementation of |plus|, this may be quite a
-complex computation, and so we would rather not do it twice. We may be
-tempted to rely on the compiler's common-subexpression elimination
-optimization. We would rather not rely on a ``sufficiently smart
-compiler'': when the generated code is very complex, the compiler may
-overlook common subexpressions.  Or the subexpressions may occur in an
-imperative context where the compiler might not be able to determine
-if lifting them is sound. So, being conservative, the optimizer will
-leave the duplicates as they are. We may attempt to eliminate
-subexpressions as in |param_code1'|. However, the result of
-|param_code1' plus one| still exhibits duplicate sub-expressions. 
-Our |let|-insertion
-optimization saved the computation at the generating stage. 
-We need a combinator that inserts the |let| expression in the
-generat\emph{ed} code. We need a combinator |letgen| to be used
-as\\|let ce = letgen (plus y one) in plus ce (plus x ce)| 
+A more generator |param_code1| has two occurrences of |plus y one|,
+which may be quite a complex computation and so we would rather not do
+it twice. We may be tempted to rely on the compiler's
+common-subexpression elimination optimization. When the generated code is
+very complex, however, the compiler may overlook common subexpressions.  Or the
+subexpressions may occur in such an imperative context where the compiler
+might not be able to determine if lifting them is sound. So, being
+conservative, the optimizer will leave the duplicates as they are. We
+may attempt to eliminate subexpressions as in |param_code1'|. However,
+the result of |param_code1' plus one| still exhibits duplicate
+sub-expressions.  Our |let|-insertion optimization saved the
+computation at the generating stage.  We need a combinator that
+inserts the |let| expression in the generat\emph{ed} code. We need a
+combinator |letgen| to be used as\\
+|let ce = letgen (plus y one) in plus ce (plus x ce)|
 yielding the code like |.<let t = y + 1 in t + (x + t)>.| 
 But that seems impossible because |letgen exp| has to generate
-the expression |.<let t = exp in body>.| although |letgen| does not have
-the |body| yet. The body needs a temporary identifier |.<t>.| that
-is supposed to be the result of |letgen| itself. 
-Certainly |letgen| cannot generate only part of a let-expression,
-without the |body|,  as all generated expressions in
-MetaOCaml are well-formed and complete.
+the expression |.<let t = exp in body>.| but |letgen| does not
+have the |body| yet. The body needs a temporary identifier |.<t>.|
+that is supposed to be the result of |letgen| itself.  Certainly
+|letgen| cannot generate only part of a let-expression, without the
+|body|, as all generated expressions in MetaOCaml are well-formed and
+complete.
 
@@
 
-\noindent Comparing the code that did let-insertion at the generating stage\\
+\noindent Comparison of the code that did let-insertion at the generating stage\\
 |let ce = (plus y one) in  plus ce (plus x ce)|\\
@@
 |letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))|\\
-clearly shows the difference between the direct-style and CPS code.
+clearly shows the difference between  direct-style and CPS code.
 What was |let ce = init in ...| in direct style became
 |init' (fun ce -> ...)| in CPS. For one thing, |let| became
-inverted. For another, what used to be an expression that yields
-a value, |init|, now became an expression that takes an extra argument,
+``inverted''. For another, what used to be an expression that yields
+a value, |init|, became an expression that takes an extra argument,
 the continuation, and invokes it. The differences look negligible in
@@
 The |do|-notation of Haskell \cite{Haskell98Report} shows that it is possible
-to write CPS code in conventional-looking style. The
-|do| notation is the notation for writing monadic code \cite{moggi-notions}.
-The benefit of monadic notation is not that it can represent CPS \cite{Filinski:Representing}, but it helps in composability by offering to add different
+to write CPS code in a conventional-looking style. The
+|do|-notation is the notation for monadic code \cite{moggi-notions}.
+Not only can monadic code represent CPS \cite{Filinski:Representing},
+it also helps in composability by offering to add different
 layers of effects (state, exception, non-determinism, etc) to the
@@
 let bind a f = fun s k -> a s (fun s' b -> f b s' k)
-let fetch s k = k s s
-let store v s k = k (v::s) ()
+let fetch s k = k s s  and  store v s k = k (v::s) ()
 
@@
 
-let l1 f = fun x -> mdo { t <-- x; f t}
-let l2 f = fun x y -> mdo { tx <-- x; ty <-- y; f tx ty}
+let l1 f = fun x -> mdo \{ t <-- x; f t\}
+let l2 f = fun x y -> mdo \{ tx <-- x; ty <-- y; f tx ty\}
 
@@
 continuation receives the current state, the value (of the type |v|) and
-yields the answer of the type |w|.  The monad is polymorphic over all
-these three type parameters.  Other implementations are
+yields the answer of the type |w|.  The monad is polymorphic over the
+three type parameters.  Other implementations are
 possible. Except for the code in Figure~\ref{ourmonad}, the rest of our code
-treats the monad as a truly abstract data type. The implementation of basic
+treats the monad as a truly abstract data type. The implementation of
+the basic
 monadic operations |ret| and |bind| are conventional. It is easy to
 see that the monadic laws are satisfied.  Other monadic operations
-construct computations that do have effects.  Operations |fetch| and
+construct computations that do have specific effects.  Operations |fetch| and
 |store v| construct computations that read and write the state. In our
@@
 Finally, |runM| runs our monad, that is, performs the computation of
-the monad and gets the result, which in our case is the code
+the monad and returns its result, which in our case is the code
 expression. We run the monad by passing it the initial state and the
@@
 That does not seem like much of an improvement. With the help of
-camplp4 pre-processor, we introduce the do notation (see
-Appendix A), patterned after the
-do-notation of Haskell. The function |param_code4|, written in
-the do-notation, is equivalent to |param_code3| -- in fact, the camlp4
-preprocessor will convert the former into the latter. And yet,
-|param_code4| looks far more conventional, as if it were indeed in
-direct style.
+camlp4 pre-processor, we introduce the |mdo|-notation (cite XXX),
+patterned after the |do|-notation of Haskell. The function
+|param_code4|, written in the |mdo|-notation, is equivalent to
+|param_code3| -- in fact, the camlp4 preprocessor will convert the
+former into the latter. And yet, |param_code4| looks far more
+conventional, as if it were indeed in direct style.
 
@@
 e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
-albeit straightforward, is not as general as we wish: its argument are
-already generated pieces of code rather than monadic values. So, we
-may want to ``lift it'', see |ifM'|. We can define functions |l1|,
+albeit straightforward, is not as general as we wish: its arguments are
+already generated pieces of code rather than monadic values. We
+``lift it'', see |ifM'|. We define functions |l1|,
 |l2|, |l3| (analogues of |liftM|, |liftM2|, |liftM3| of Haskell) 
-to make such a lifting generic.
-We soon notice the need for another |ifM| function, with the same
+to make such a lifting generic. However we also need
+another |ifM| function, with the same
 interface (see Figure~\ref{ourmonad}). The difference between them is
@@
 need similar operators for other Ocaml control forms: for generating
-case-matching statements and for- and while-loops.
-
-\section{Functors}\label{functors}
-
-Now that we have the appropriate basic techniques for
-writing our code generators, we need to worry about modularization.
-Here we can really take advantage of the fact that
-MetaOCaml is a staged language: it is not particularly relevant how much
-computation happens at staging time, as long as the generated code
-is correct and efficient.  In other words, we can use any abstraction
+case-matching statements and |for|- and |while|-loops.
+
+\section{Aspects and Functors}\label{functors}
+
+The monad represents finer-scale code generation. We need tools for
+larger-scale modularization; we can use any abstraction
 mechanisms we want to structure our code generators, as long as none
@@
 \vspace*{-6pt}
-	\item \textbf{Domain}: In which (algebraic) domain do the matrix
-		elements belong to.  Sometimes the domains are very specific
-		($\mathbb{Z}, \mathbb{Q}, \mathbb{Z}_p, 
-		\mathbb{Z}_p\left[\alpha_1,\ldots,\alpha_n\right], 
-		\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$, 
-		$\mathbb{Q}\left[\alpha\right]$, and floating point numbers for 
-		example), while in other cases the domains were left generic,
-		namely for elements of a field,
-		multivariate polynomials over a field, elements of a formal Ring
-		with possibly undecidable zero-equivalence, or elements of a 
-		Division Ring.  In the roughly 85 pieces of code
-		surveyed 20 different domains were encountered.
-		% \oleg{citation?} \jacques no citation needed, carette04 is
-		% already cited above.
-	\item \textbf{Container}: Whether the matrix
-		is represented as an array of arrays, a one-dimensional array,
-		a hash table, a sparse matrix, etc., and
-		whether indexing is done in C or Fortran style.  Additionally,
-		if a particular representation had a special mechanism for efficient
-		row exchanges, this is sometimes used.  
-	\item \textbf{Output choices}:  Whether just the reduced matrix, or
-		additionally the rank, the determinant, and the
-		pivoting matrix are to be returned.
-		It is worthwhile noting that in the larger algorithm family,
-		routines like Maple's \texttt{LinearAlgebra:-LUDecomposition} have
-		up to $2^6 + 2^5 + 2^2 = 100$ outputs.
+	\item \textbf{Domain}: In which (algebraic) domain do the
+	  matrix elements belong to.  Sometimes the domains are very
+	  specific (e.g., $\mathbb{Z}, \mathbb{Q}, \mathbb{Z}_p$ and
+	  floating point numbers), while in other cases the domains
+	  were left generic, e.g., multivariate polynomials over a
+	  field.  In the roughly 85 pieces of code surveyed
+	  \cite{carette04} 20 different domains were encountered.
+	\item \textbf{Container}: Whether the matrix is represented as
+	  an array of arrays, a one-dimensional array, a hash table, a
+	  sparse matrix, etc., and whether indexing is done in C or
+	  Fortran style.  Additionally, if a particular representation
+	  had a special mechanism for efficient row exchanges.
+	\item \textbf{Output choices}: Whether just the reduced
+	  matrix, or additionally the rank, the determinant, and the
+	  pivoting matrix are to be returned. In the larger algorithm
+	  family, routines like Maple's
+	  \texttt{LinearAlgebra:-LUDecomposition} have up to $2^6 +
+	  2^5 + 2^2 = 100$ outputs.
 	\item \textbf{Fraction-free}: Whether the Gaussian Elimination
@@
 		a full matrix, or only a restricted number of columns, while
-		doing elimination on the full matrix.  Note that we currently
+		doing elimination on the full matrix.  We currently
 		do not treat this aspect in our code.
@@
 	\item \textbf{Length measure}:  For stability reasons
-		(whether numerical or coefficient growth), if a domain
+		(numerical or coefficient growth), if a domain
 	  possesses
 		an appropriate length measure, this is sometimes used to choose
-		an ``optimal'' pivot for row exchanges.
+		an ``optimal'' pivot.
 	\item \textbf{Normalization and zero-equivalence}: Whether the 
@@
 \noindent These are separated out from the others are they are cross-cutting
-concerns: in the case of the length measure, then a property of the domain
-will influence the method used to do pivoting \textbf{if} pivoting is to be
+concerns: in the case of the length measure, a property of the domain
+will influence the method used to do pivoting \emph{if} pivoting is to be
 performed.
@@
 turns out, we need the following to exist in our domains: $0$, $1$,
-$+$, $*$, (unary and binary) $-$, some kind of division, normalization,
-and potentially a relative size measure.  We need at least
-\emph{exact} division to exist, but if full division is available, that
-can be used as well.  Normalization is frequently trivial for domains
-where elements have unique representations, but for some
-domains (like polynomials in sum-of-product form) this
-is not the case.  The simplest case of such an abstraction was given
-in Fig.~\ref{easycode}, |param_code1|. In the latter, code-generators such as
-|plus| and |one| were passed as arguments. We see however that we 
-need far more than two parameters, so we have to group them. Instead
-of the grouping offered by regular records, we use Ocaml
-\emph{structures} (i.e., modules) 
+$+$, $*$, (unary and binary) $-$, at least \emph{exact} division,
+normalization, and potentially a relative size measure. The simplest
+case of such domain abstraction is |param_code1| in
+Fig.~\ref{easycode}. There, code-generators such as |plus| and |one|
+were passed as arguments. We need far more than
+two parameters, so we have to group them. Instead of the grouping
+offered by regular records, we use Ocaml \emph{structures} (i.e.,
+modules)
 % (cite Ocaml module system tutorial?)
 % normally yes, but our references are what's making this long right now!
-so we can take advantage of extensions, type abstraction and constraints,
+so we can take advantage of extensibility, type abstraction and constraints,
 and especially parameterized structures (\emph{functors}).
-We first define the type of the domain, the signature |DOMAIN| which
+We define the type of the domain, the signature |DOMAIN| which
 different domains must satisfy:
@@
 
-\noindent  The first item to notice is that the types above are
+\noindent  The types above are
 generally lifted twice: once from the value domain |v| to the code
 domain |'a vc|, and once more from values to monadic computations
-|('a vc, 's, 'w) monad|.  Since there are times where we do not want
-to redundantly walk a whole matrix to apply normalization when 
-normalization is not necessary, |normalizerf| is optional.
-
-One particular instantiation of the domain (a particular structure of
-|DOMAIN| signature) is the |int| domain. The notation
-|module IntegerDomain : DOMAIN| makes the compiler
-verify that our |IntegerDomain| is indeed a |DOMAIN|, that is,
-satisfies the required signature. The constraint |DOMAIN| may be
-omitted; in that case, the compiler will verify the type when we try
-to use that structure as a |DOMAIN|. In any case, the errors such as
-missing ``methods'' or methods with incorrect types will be caught
-statically, even \emph{before} any code generation takes place. The
-abstract type |domain_is_ring| encodes a semantic constraint that the
-full division is not available. While the |DOMAIN| type may have
-looked daunting to some, the implementation is quite straightforward.
-|float| and |Num.num| (arbitrary precision exact rational numbers)
-are equally simple, and with scarcely
-more work, $\mathbb{Z}_p$, $\mathbb{Z}\left[x\right]$,
-$\mathbb{Q}\left(x\right)$ as well more complex domains can be implemented.
-
-Parametrizing by the kind of container used to represent a matrix is
-almost as straightforward.  We choose to make the container parametric
-over a |DOMAIN| by making containers functors from a |DOMAIN| module
+|('a vc, 's, 'w) monad|.
+
+One particular domain instance is |IntegerDomain|. The notation
+|module IntegerDomain : DOMAIN| makes the compiler verify that our
+|IntegerDomain| is indeed a |DOMAIN|, that is, satisfies the required
+signature. The constraint |DOMAIN| may be omitted; in that case, the
+compiler will verify the type when we try to use that structure as a
+|DOMAIN|. In any case, the errors such as missing ``methods'' or
+methods with incorrect types will be caught statically, even
+\emph{before} any code generation takes place. The abstract type
+|domain_is_ring| encodes a semantic constraint that the full division
+is not available. While the |DOMAIN| type may have looked daunting to
+some, the implementation is quite straightforward.  Other domains such
+as |float| and arbitrary precision exact rational numbers |Num.num|
+are equally simple.
+
+Parametrizing by the kind of container representing a matrix is
+almost as straightforward.  Our containers are  parametric
+over a |DOMAIN|, i.e., functors from a |DOMAIN| module
 to the actual implementation of a container. The functor signature
@@
 The use of a |functor| for making a container parametric is fairly
-straightforward.  It gets more interesting when one turns to the aspect
-of what to return from the GE algorithm.  One could create an algebraic
-data type (as was done in \cite{carette04}) to encode the various
-choices; but this is wholly unsatisfying as we know that for any single
-use, only one of the choices is ever possible, yet any routine which
-calls the generated code must deal with these unreachable options.
-Instead we use a module type with an \emph{abstract} type |res| for the
-result type; the actual result type is then made by instantiating
-this module type with different modules that decide what will be 
-output: the matrix, the matrix and the rank, the matrix and the
-determinant, the matrix, rank and determinant, and so on.  More
-precisely, below we show this module type and one instantiation,
-which specifies the output of a  GE algorithm as a 3-tuple
-|contr * Det.outdet * int| of the U-factor, the determinant and the rank.
+straightforward.  More interesting is the aspect of what to return
+from the GE algorithm.  One could create an algebraic data type (as
+was done in \cite{carette04}) to encode the various choices: the
+matrix, the matrix and the rank, the matrix and the determinant, the
+matrix, rank and determinant, and so on. This is wholly unsatisfying
+as we know that for any single use, only one of the choices is ever
+possible, yet any routine which calls the generated code must deal
+with these unreachable options.  Instead we use a module type with an
+\emph{abstract} type |res| for the result type; different instances of
+the signature set the result type differently. Given below is this
+module type and one instantiation, which specifies the output of a GE
+algorithm as a 3-tuple |contr * Det.outdet * int| of the U-factor, the
+determinant and the rank.
 
@@
 As is apparent from the output choices, several different quantities
-\emph{may} need to be tracked in a particular GE implementation.
-We therefore need to be able to conditionally generate code-variables,
-and weave in corresponding tracking code for each of these quantities.
-To be precise, we may need to keep track (independently) of the rank,
-the determinant and the permutation list.  These pieces of data then
-become part of the \emph{state} that is tracked by our monad.  To have
-all this choice when needed, and yet have our code be modular and
+\emph{may} need to be tracked in a particular GE implementation.  We
+therefore need to be able to conditionally generate variables
+representing the tracking state, and weave in corresponding tracking
+code. We may need to (independently) keep track of the rank, the
+determinant and the permutation list.  The tracking state variables
+then become part of the \emph{state} that is tracked by our monad.  To
+have all this choice when needed, and yet have our code be modular and
 composable as well as ensuring that the generated code does not
-contain any abstraction artifacts, it is important to make this 
-state modular.  As this discussion is rather abstract, an example
-will likely illustrate our point more clearly: we will show the 
-code for tracking the determinant.
-
+contain any abstraction artifacts, it is important to make this state
+modular.  For example,
 \begin{code}
@@
 \end{code}
-\noindent  Determinant tracking aspects involves being able to
+\noindent  determinant tracking aspects involves being able to
 generate code that defines variables used for tracking (|decl|),
@@
 permutation; the value of the determinant should be updated per each
-pivoting. We use |lstate| to pass the tracking state, which may be, for
-example, a piece of code for the value of the type |Dom.v ref| among
+pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
+code for the value of the type |Dom.v ref|, among
 various determinant-tracking functions. The |lstate| is a part of the
@@
 \vspace*{-4pt} 
-This structure implements an update policy that can use the 
-|Dom.div| operation without
-restrictions -- which is possible only if the domain has such an
-operation. A domain such as the integer domain may still provide |Dom.div|
-of the same type, but that operation can only be used when we are sure
-that the division is exact. But using it with
-|DivisionUpdate| will lead to a compile-time error, when compiling
-the \emph{generator} code. Thus, in some cases we can use module types
-to express ``semantic'' constraints that cannot normally be expressed
-via the types of module members.
+This structure implements an update policy of using
+|Dom.div| operation without restrictions -- which is possible only if
+the domain has such unrestricted operation. A domain such as the integer domain
+may still provide |Dom.div| of the same type, but that operation may
+only be used when we are sure that the division is exact. Our type
+sharing constraint expresses such domain-specific knowledge:
+instantiating |DivisionUpdate| with |integerDomain| leads to a
+compile-time error, when compiling the \emph{generator} code. Thus, in
+some cases we can use module types for ``semantic'' constraints
+that cannot normally be expressed via the types of module members.
 \vspace*{-5pt}

### 2005-04-09 paper/metamonads.tex
change title as per email.  minor edits in intro.

@@
 \begin{document}
-\title{Functors, CPS and monads, or how to generate efficient
-code from abstract designs}
+%\title{Functors, CPS and monads, or how to generate efficient
+%code from abstract designs}
+\title{Multi-stage programming with functors and monads:
+eliminating abstraction overhead from generic code}
 \author{Jacques Carette\inst{1}
@@
 compiler because the tweaking often involves domain knowledge. 
-The survey \cite{carette04} of
-Gaussian elimination implementation in an industrial package Maple
-found XXX various aspects and XXX different implementations of the
-algorithm.  We can manually write each of these implementations
+A survey \cite{carette04} of
+Gaussian elimination implementations in an industrial package Maple
+found 6 different aspects and 35 different implementations of the
+algorithm, as well as 45 implementations of directly related
+algorithms.  We can manually write each of these implementations
 optimizing for particular aspects and using cut-and-paste to ``share''
-similar pieces of code. \oleg{I don't think we need to say much here
--- the shear number of various implementations should be enough of a
-clue.} We can write a very generic GE procedure that accounts for
-all the aspects with appropriate abstractions (cite axiom?). The
+similar pieces of code.
+We can write a very generic GE procedure that accounts for
+all the aspects with appropriate abstractions \cite{Axiom}. The
 abstraction mechanisms however -- be they procedure, method or a
@@
 
-A more appealing (acceptable? indispensable? need a better word)
-approach is generative programming
+A more appealing approach is generative programming
 \cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser94algorithmoriented,BOOST,POOMA,ATLAS}.
@@
 as our running example. Specifically, our contributions:
-\vspace*{-6pt}
+\vspace*{-2pt}
 \begin{itemize}
@@
 
-A more generator |param_code1| has two occurrences of |plus y one|,
+The generator |param_code1| has two occurrences of |plus y one|,
 which may be quite a complex computation and so we would rather not do

### 2005-04-10 paper/metamonads.tex
corrections to the code

@@
 \begin{code}
-module GenFV4 = Gen(FloatDomain)
+module GenIV5 = Gen(IntegerDomain)
    (GenericVectorContainer)(FullPivot)
-   (DivisionUpdate(FloatDomain)(GenericVectorContainer)(FDet))
-   (OutDetRank(FloatDomain)(GenericVectorContainer)(FDet)(Rank))
-module GenIA1 = Gen(IntegerDomain)
+   (FractionFreeUpdate(IntegerDomain)(GenericVectorContainer)(IDet))
+   (OutDetRank(IntegerDomain)(GenericVectorContainer)(IDet)(Rank))
+module GenFA1 = Gen(FloatDomain)
    (GenericArrayContainer)(RowPivot)
-   (FractionFreeUpdate(IntegerDomain)(GenericArrayContainer)(IDet))
-   (OutJustMatrix(IntegerDomain)(GenericArrayContainer)(IDet))
+   (DivisionUpdate(FloatDomain)(GenericArrayContainer)(NoDet(FloatDomain)))
+   (OutJustMatrix(FloatDomain)(GenericArrayContainer)(NoDet(FloatDomain)))
 \end{code}
@@
 We can instantiate the |Gen| functor as shown above and inspect the
-generated code, e.g., by printing |GenIA1.gen|. The code can then be
-``compiled'' as |!. GenIA1.gen| or with off-shoring. The code
-% (cite).  No space!
-for |GenFV4| shows full pivoting, determinant and rank tracking. The
+generated code, e.g., by printing |GenFA1.gen|. The code can then be
+``compiled'' as |!. GenFA1.gen| or with off-shoring. The code
+for |GenIV5| (Appendix A) shows full pivoting, determinant and rank tracking. The
 code for all these aspects is fully inlined; no extra functions are
@@
 itself are performed. The GE function returns a triple 
-|float array array * float * int| of the U-factor, determinant and 
+|int array * int * int| of the U-factor, determinant and 
 the rank. The
-code generated by |GenIA1| shows absolutely no traces of determinant
-tracking: no declaratrion of spurious variables, no extra tests,
+code generated by |GenFA1| (Appendix B) shows absolutely no traces of determinant
+tracking: no declaration of spurious variables, no extra tests,
 etc. The code appears as if the determinant tracking aspect did not exist
@@
 \bibliographystyle{plain}
+\section{Appendix A}
+The code generated for |GenIV5|, fraction-free GE of the integer matrix
+represented by a flat vector, full pivoting, returning the |U|-factor,
+the determinant and the rank.
+\begin{code}
+# val resIV5 :
+  ('a,
+   Funct4.GenIV5.Ctr.contr ->
+   Funct4.OutDetRank(Funct4.IntegerDomain)(Funct4.GenericVectorContainer)(Funct4.IDet)(Funct4.Rank).res)
+  code =
+  .<fun a_405 ->
+   let t_406 = (ref 0) in
+   let t_407 = (ref 0) in
+   let t_408 = {arr = (Array.copy a_405.arr)} (a_405) in
+   let t_409 = a_405.m in
+   let t_410 = a_405.n in
+   let t_411 = (ref 1) in
+   let t_412 = (ref 1) in
+   while (((! t_407) < t_409) && ((! t_406) < t_410)) do
+    let t_413 = (! t_406) in
+    let t_414 = (! t_407) in
+    let t_415 = (ref (None)) in
+    let t_435 =
+     begin
+      for j_431 = t_413 to (t_410 - 1) do
+       for j_432 = t_414 to (t_409 - 1) do
+        let t_433 = (t_408.arr).((j_431 * t_408.m) + j_432) in
+        if (not (t_433 = 0)) then
+         (match (! t_415) with
+          | Some (i_434) ->
+             if ((abs (snd i_434)) > (abs t_433)) then
+              (t_415 := (Some ((j_431, j_432), t_433)))
+             else ()
+          | None -> (t_415 := (Some ((j_431, j_432), t_433))))
+        else ()
+       done
+      done;
+      (match (! t_415) with
+       | Some (i_416) ->
+          if ((snd (fst i_416)) <> t_414) then begin
+           let a_424 = t_408.arr
+           and nm_425 = (t_408.n * t_408.m)
+           and m_426 = t_408.m in
+           let rec loop_427 =
+            fun i1_428 ->
+             fun i2_429 ->
+              if (i2_429 < nm_425) then
+               let t_430 = a_424.(i1_428) in
+               a_424.(i1_428) <- a_424.(i2_429);
+               a_424.(i2_429) <- t_430;
+               (loop_427 (i1_428 + m_426) (i2_429 + m_426))
+              else () in
+           (loop_427 t_414 (snd (fst i_416)));
+           (t_412 := (~- (! t_412)))
+          end else ();
+          if ((fst (fst i_416)) <> t_413) then begin
+           let a_417 = t_408.arr
+           and n_418 = t_408.n
+           and m_419 = t_408.m in
+           let i1_420 = (t_413 * m_419)
+           and i2_421 = ((fst (fst i_416)) * m_419) in
+           for i_422 = 0 to (m_419 - 1) do
+            let t_423 = a_417.(i1_420 + i_422) in
+            a_417.(i1_420 + i_422) <- a_417.(i2_421 + i_422);
+            a_417.(i2_421 + i_422) <- t_423
+           done;
+           (t_412 := (~- (! t_412)))
+          end else ();
+          (Some (snd i_416))
+       | None -> (None))
+     end in
+    (match t_435 with
+     | Some (i_436) ->
+        begin
+         for j_437 = (t_413 + 1) to (t_410 - 1) do
+          if (not ((t_408.arr).((j_437 * t_408.m) + t_414) = 0)) then begin
+           for j_438 = (t_414 + 1) to (t_409 - 1) do
+            (t_408.arr).((j_437 * t_408.m) + j_438) <-
+             ((((t_408.arr).((j_437 * t_408.m) + j_438) *
+                 (t_408.arr).((t_413 * t_408.m) + t_414)) -
+                ((t_408.arr).((t_413 * t_408.m) + j_438) *
+                  (t_408.arr).((j_437 * t_408.m) + t_413))) / (! t_411))
+           done;
+           (t_408.arr).((j_437 * t_408.m) + t_414) <- 0
+          end else ()
+         done;
+         (t_411 := i_436)
+        end;
+        (t_406 := ((! t_406) + 1))
+     | None -> (t_412 := 0));
+    (t_407 := ((! t_407) + 1))
+   done;
+   (t_408,
+    if ((! t_412) = 0) then 0
+    else if ((! t_412) = 1) then (! t_411)
+    else (~- (! t_411)), (! t_406))>.
+\end{code}
+\section{Appendix B}
+The code generated for |GenFA1|, GE of the floating point matrix
+represented by a 2D array, row pivoting, returning just the |U|-factor.
+\begin{code}
+#   val resFA1 :
+  ('a,
+   Funct4.GenFA1.Ctr.contr ->
+   Funct4.OutJustMatrix(Funct4.FloatDomain)(Funct4.GenericArrayContainer)(Funct4.NoDet(Funct4.FloatDomain)).res)
+  code =
+  .<fun a_1 ->
+   let t_2 = (ref 0) in
+   let t_3 = (ref 0) in
+   let t_5 = (Array.map (fun x_4 -> (Array.copy x_4)) (Array.copy a_1)) in
+   let t_6 = (Array.length a_1.(0)) in
+   let t_7 = (Array.length a_1) in
+   while (((! t_3) < t_6) && ((! t_2) < t_7)) do
+    let t_8 = (! t_2) in
+    let t_9 = (! t_3) in
+    let t_10 = (ref (None)) in
+    let t_16 =
+     begin
+      for j_13 = t_8 to (t_7 - 1) do
+       let t_14 = (t_5.(j_13)).(t_9) in
+       if (not (t_14 = 0.)) then
+        (match (! t_10) with
+         | Some (i_15) ->
+            if ((abs_float (snd i_15)) < (abs_float t_14)) then
+             (t_10 := (Some (j_13, t_14)))
+            else ()
+         | None -> (t_10 := (Some (j_13, t_14))))
+       else ()
+      done;
+      (match (! t_10) with
+       | Some (i_11) ->
+          if ((fst i_11) <> t_8) then begin
+           let t_12 = t_5.(t_8) in
+           t_5.(t_8) <- t_5.(fst i_11);
+           t_5.(fst i_11) <- t_12;
+           ()
+          end else ();
+          (Some (snd i_11))
+       | None -> (None))
+     end in
+    (match t_16 with
+     | Some (i_17) ->
+        begin
+         for j_18 = (t_8 + 1) to (t_7 - 1) do
+          if (not ((t_5.(j_18)).(t_9) = 0.)) then begin
+           for j_19 = (t_9 + 1) to (t_6 - 1) do
+            (t_5.(j_18)).(j_19) <-
+             ((t_5.(j_18)).(j_19) -.
+               (((t_5.(j_18)).(t_9) /. (t_5.(t_8)).(t_9)) *.
+                 (t_5.(t_8)).(j_19)))
+           done;
+           (t_5.(j_18)).(t_9) <- 0.
+          end else ()
+         done;
+         ()
+        end;
+        (t_2 := ((! t_2) + 1))
+     | None -> ());
+    (t_3 := ((! t_3) + 1))
+   done;
+   t_5>.
+\end{code}
 \end{document}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % It seems no appendices are allowed. Check with Walid?

### 2005-04-10 paper/metamonads.tex
Final tweaks - complete bibliography, conclusion, future work.
Also, move unused text to unused.tex to make editing simpler.

@@
 function call -- have a significant cost, especially for
-high-performance numerical computing (cite? At least carette04).
+high-performance numerical computing \cite{carette04}.
 
@@
 well-typed. Lisp-style macros, Scheme hygienic macros, camlp4
-preprocessor (cite XXX), C++ template meta-programming, Template
+preprocessor \cite{camlp4}, C++ template meta-programming, Template
 Haskell \cite{conf/dagstuhl/CzarneckiOST03} solve some of the above
@@
 use full division when dealing with matrices of exact integers, nor is
-it worthwhile to use full pivoting on a matrix of rational integers
-($\mathbb Q $).
+it worthwhile to use full pivoting on a matrix over $\mathbb Q $.
 
@@
     \item Implementation of the |mdo|-notation (patterned after
-      |do|-notation of Haskell) to make monadic code readable
+      |do|-notation of Haskell) to make monadic code readable.
     \item Use of functors (including higher-order functors) to
@@
       for aspects that use state and have to be accounted in many
-      places in the generated code
-    \item use functor type sharing constraints to encode domain-specific
-      knowledge
+      places in the generated code.
+    \item Use functor type sharing constraints to encode domain-specific
+      knowledge.
 \end{itemize}
@@
 section~\ref{related}. We then outline the future work and conclude.
-Appendices give samples of the generated code (which is availablke in
-full at XXX).
+Appendices give samples of the generated code (which is available in
+full at \cite{metamonadsURL}).
 
@@
 clearly shows the difference between  direct-style and CPS code.
-What was |let ce = init in ...| in direct style became
+What was\\ |let ce = init in ...| in direct style became
 |init' (fun ce -> ...)| in CPS. For one thing, |let| became
@@
 \evalresult{.<fun a_1 i_2 ->  
-let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.}
+let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
 let gen a i = ifM (ret .<(.~i) >= 0>.) 
@@
 \evalresult{.<fun a_1 i_2 -> 
-if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.}
+if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
 \end{code}
@@
 modules)
-% (cite Ocaml module system tutorial?)
-% normally yes, but our references are what's making this long right now!
 so we can take advantage of extensibility, type abstraction and constraints,
@@
 
-One particular domain instance is |IntegerDomain|. The notation
+One particular domain instance is |IntegerDomain|. The notation\\
 |module IntegerDomain : DOMAIN| makes the compiler verify that our
@@
 containers with efficiency details.
-\omitnow{ As an example, below is
-  an implementation of a matrix as an array of rows (themselves
-  arrays).  This implementation clearly shows how swapping rows is
-  more efficient than swapping columns in this representation.  We
-  have not (yet) implemented the further potential optimization
-  wherein if it is known that part of two rows/columns are already
-  known to be equal (in fact to |Dom.zero| in the case of GE) and
-  row/column exchange needs to be performed element-wise, then certain
-  swaps are unecessary.
-
-\begin{code}
-module GenericArrayContainer(Dom:DOMAIN) =
-  struct
-  type contr = Dom.v array array (* Array of rows *)
-  type 'a vc = ('a,contr) code
-  type 'a vo = ('a,Dom.v) code
-  let get' x n m = .< (.~x).(.~n).(.~m) >.
-  let get x n m = ret (get' x n m)
-  let set' x n m y = .< (.~x).(.~n).(.~m) <- .~y >.
-  let set x n m y = ret (set' x n m y)
-  let dim2 x = .< Array.length .~x >.       (* number of rows *)
-  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
-  let mapper g a = match g with
-      | Some f -> .< Array.map (fun x -> Array.map .~f x) .~a >.
-      | None   -> a
-  let copy = (fun a -> .<Array.map (fun x -> Array.copy x) 
-                       (Array.copy .~a) >. )
-  (* this can be optimized with a swap_rows_from if it is known that
-     everything before that is already = Dom.zero *)
-  let swap_rows_stmt a r1 r2 =
-      .< let t = (.~a).(.~r1) in
-         begin 
-             (.~a).(.~r1) <- (.~a).(.~r2);
-             (.~a).(.~r2) <- t
-         end >.
-  let swap_cols_stmt a c1 c2 = .< 
-      for r = 0 to .~(dim2 a)-1 do
-          let t = (.~a).(r).(.~c1) in
-          begin 
-              (.~a).(r).(.~c1) <- (.~a).(r).(.~c2);
-              (.~a).(r).(.~c2) <- t
-          end
-      done  >.
-end
-\end{code}
-\noindent  Implementing other containers based on flattened vectors,
-Fortran-style access, |Bigarray| or |Bigarray.Array2| is also possible.
-Naturally, for proper use of |Bigarray| with specific types, it is 
-important that the |DOMAIN| type matches -- but luckily the compiler 
-will prevent us from making such a mistake.}
 
@@
 \vspace*{-5pt}
-We can instantiate the |Gen| functor as shown above and inspect the
-generated code, e.g., by printing |GenFA1.gen|. The code can then be
-``compiled'' as |!. GenFA1.gen| or with off-shoring. The code
-for |GenIV5| (Appendix A) shows full pivoting, determinant and rank tracking. The
-code for all these aspects is fully inlined; no extra functions are
-invoked and no tests other than those needed by the GE algorithm
-itself are performed. The GE function returns a triple 
-|int array * int * int| of the U-factor, determinant and 
-the rank. The
-code generated by |GenFA1| (Appendix B) shows absolutely no traces of determinant
-tracking: no declaration of spurious variables, no extra tests,
+We can instantiate the |Gen| functor as shown above and inspect the generated
+code, e.g., by printing |GenFA1.gen|. The code can then be ``compiled'' as 
+|!. GenFA1.gen| or with off-shoring. The code for |GenIV5| (Appendix A) shows
+full pivoting, determinant and rank tracking. The code for all these aspects is
+fully inlined; no extra functions are invoked and no tests other than those
+needed by the GE algorithm itself are performed. The GE function returns a
+triple |int array * int * int| of the U-factor, determinant and the rank. The
+code generated by |GenFA1| (Appendix B) shows absolutely no traces of
+determinant tracking: no declaration of spurious variables, no extra tests,
 etc. The code appears as if the determinant tracking aspect did not exist
@@
 
+The highly parametric version of our Gaussian Elimination is directly
+influenced by the generic implementations available in Axiom
+\cite{Axiom} and Aldor \cite{Watt:2002:HCA}.  Even though the Aldor
+compiler frequently can optimize away a lot of abstraction overhead, 
+it does not provide any guarantees that it will do so, unlike our
+approach.
+
 To the best of our knowledge, nobody has yet used functors to
@@
 multi-stage programming.
-% \oleg{Yes, let's gloss over ATLAS, SPIRAL, Axiom. I wonder what would
-%   be the use of functors? Something by David McQueen or R. Harper.
-%   Mention that nobody before used functors to abstract
-%   code generators -- or mix functors and multi-stage programming.}
-
-connections with delimited continuations: making notation
-more direct-style and potentially clearer.  More syntactic sugar
-would help a lot here.
-
-More aspects that can be handled: Input variations (augmented
-matrices). Fit into the larger program family, where there are
-more aspects still.
-
-Making all of the algorithmic code be completely independent of 
-MetaOCaml markup and isolate all the code generation to a select
-few 'base' modules and functors.  This should allow for simpler
-debugging, where one can choose to instantiate either a direct-style
-set of base modules for debugging, or a code-generation style set
-of base modules for generating optimized code.
+
+We plan to further investigate the connection between delimited
+continuations and our implementations of code generators like
+|ifM|.  As well, by using some additional syntactic sugar
+(for |ifM|, |whileM|, etc.), the available notation should be
+even more direct-style, and potentially clearer.
+
+There are many more aspects which can also be handled
+Input variations (augmented
+matrices), error reporting (i.e. asking for the determinant of a 
+non-square matrix), memory hierarchy issues, loop-unrolling
+\cite{Padua:MetaOcaml:04},
+warnings when zero-testing is undecidable and
+a value is only probabilistically non-zero, etc.  The larger program
+family of LU decompositions contains more aspects still.
+
+Finally, we would like to be able to make the |Gen| functor applicable
+to runnable functors in Ocaml, as well as to code generation functors
+in MetaOCaml - for ease of debugging.  The current use of abstract
+types should make this straightforward.
 
 \section{Conclusion}\label{conclusion}
-Mention the mdo notation. Mention it into the introduction.
-
-Combination of stateless modules (functors and structures) with
-the monad with the compositional state greatly makes the aspects
-freely composable without having to worry about presence or absense of
+The combination of stateless modules (functors and structures), and our
+monad with the compositional state makes aspects
+freely composable without having to worry about the presence or absense of
 value aliasing. The only constraints to compositionality are the
 typing constraints plus the constraints we specifically impose,
-including semantic constraints (e.g., rings do not have the full
+including semantic constraints (e.g., rings do not have full
 division).
 
-Relation with aspect-oriented code: in AspectJ, aspects are (comparatively)
-lightly typed, and are post-facto extensions of an existing piece of
-code.  Here aspects are weaved together ``from scracth'' to make up a
-piece of code/functionality.  One can understand previous work to be
+There is an interesting relation with aspect-oriented code
+\cite{kiczales97aspectoriented}: in AspectJ, 
+aspects are (comparatively) lightly typed, and are post-facto extensions of an
+existing piece of code.  Here aspects are weaved together ``from scracth'' to
+make up a piece of code/functionality.  One can understand previous work to be
 more akin to dynamically typed aspect weaving, while we have started
@@
    Funct4.GenIV5.Ctr.contr ->
-   Funct4.OutDetRank(Funct4.IntegerDomain)(Funct4.GenericVectorContainer)(Funct4.IDet)(Funct4.Rank).res)
+   Funct4.OutDetRank(Funct4.IntegerDomain)(Funct4.GenericVectorContainer)
+                    (Funct4.IDet)(Funct4.Rank).res)
   code =
@@
    Funct4.GenFA1.Ctr.contr ->
-   Funct4.OutJustMatrix(Funct4.FloatDomain)(Funct4.GenericArrayContainer)(Funct4.NoDet(Funct4.FloatDomain)).res)
+   Funct4.OutJustMatrix(Funct4.FloatDomain)(Funct4.GenericArrayContainer)
+                       (Funct4.NoDet(Funct4.FloatDomain)).res)
   code =
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% It seems no appendices are allowed. Check with Walid?
-
-\section{Appendix A}
-The do notation for Ocaml.
-To implement the |mdo| notation, we use the Ocaml pre-processor,
-camplp4 (cite?). Informally, our notation is
-\begin{code}
-mdo { exp }
-mdo { exp1; exp2 }
-mdo { x <-- exp; exp }
-mdo { let x = foo in; exp }
-\end{code}
-which matches closely the Haskell |do| notation, modulo |do|/|mdo| and 
-|<-|/|<--|.
-
-Grammar formally:
-\begin{code}
-mdo { <do-body> }
-<do-body> :: =
-    "let" var = EXP ("and" var = EXP)* "in" ";" <do-body>
-    | EXP
-    | (pat <--)? EXP ";" <do-body>
-\end{code}
-The semantics is given by the following rules re-writing the |mdo|
-form into the regular Ocaml code. We assume that the variable |bind|
-must be in scope.
-\begin{code}
-mdo { exp } ===> exp
-mdo { pat <-- exp; rest } ===> bind exp (fun pat -> mdo { rest })
-mdo { exp; rest } ===> bind exp (fun _ -> mdo { rest })
-mdo { let pat = exp in; rest } ===> let pat = exp in mdo { rest }
-\end{code}
-We also introduce |odo| form, which is equivalent to |mdo| but uses 
-a bind `method' rather than a bind `value'.
-
-The major difficulty with the |do| notation is that it cannot truly be
-parsed by an LR(n)-grammar for any finite |n| because many |pat| can also
-be parsed as an |exp|.
-
-Perhaps give the command line how to load the camlp4 module into
-ocaml. Point to the code.
-
-\omitnow{
-In this paper, we are concerned with building and composing
-generators. We start with the code expressing the algorithm (in
-OCaml), and, adding MetaOCaml annotations, turn that code into a
-\emph{code expression} that will generate the code when runs. We then
-modify that code expression to add various parameters and aspects, to
-yield the complete generator.
-
-Here we illustrate the MetaOCaml constrcutions and code generators.
-
-Turning code into the code expression is relatively straightforward
-until we come to binding constructs such as |let|. For example, here
-is a piece of code that defines a function that searches a given array
-|arr| for the maximal (in absolute value) element, and the index of
-that element and the found absolute value. This code is representative
-of pivot determination. The code uses mutations and destructive
-updates -- partly for illustrative purposes and mainly because many
-linear algebra algorithms are expressed via an imperative code (so it
-is less error-prone to translate the ``canonical'' algorithms with
-minimal changes).
-
-\begin{code}
-let fp = fun arr n ->
-   let pv = ref (0,arr.(0)) in
-   for i = 1 to n-1 do
-     if abs_float arr.(i) > snd !pv then
-        pv := (i, abs_float arr.(i))
-   done;
-   !pv
-\end{code}
-
-This function has the type |float array -> int -> int * float|. By
-enclosing the code into MetaOCaml brackets, we obtain the simplest
-generator expression.
-
-\begin{code}
-let fpc = .<fun arr n ->
-   let pv = ref (0,arr.(0)) in
-   for i = 1 to n-1 do
-     if abs_float arr.(i) > snd !pv then
-        pv := (i, abs_float arr.(i))
-   done;
-   !pv>.
-\end{code}
-
-This value has the type |('a, float array -> int -> int * float) code|.
-Here |'a| type parameter is needed (ask Walid for the reference).
-The main difference between |fp| and |fpc| is that |fp| is a function,
-which we can immediately use as \verb^fp [|1.0; -4.; 3.|] 3^. In
-contrast, |fpc| is not a function: it is a \emph{code} value that denotes
-that function. We would say that the code expression is the
-``second-stage'' value. The ability to manipulate code expressions as
-regular values is the distinguished fature of MetaOCaml.
-We can ``compile'' the code by using MetaOCaml |run| command, or |.!|:
-|.! fpc|. The result is equivalent is |fp|. Although |fpc| doesn't
-seem to do much, it is already useful as we can use offshoring (cite)
-to translate |fpc| into C or Fortran code (which we can dynamically
-link into the running MetaOCaml code -- or use separately).
-
-The code |fpc| obviously lacks generality: the name |abs_float|
-implies that that code deals with |float| arrays. We wish to handle
-arrays in various domains, integer, rationals, etc. We can make the
-code generic by parametrizing over the |abs| function:
-
-\begin{code}
-let fpca absf = .<fun arr n ->
-   let pv = ref (0,arr.(0)) in
-   for i = 1 to n-1 do
-     if absf arr.(i) > snd !pv then
-        pv := (i, absf arr.(i))
-   done;
-   !pv>.
-\end{code}
-
-and so |fpca abs_float| will manipulate |float array| and |fpca abs|
-will do for the integer arrays. Here we used the value from the first
-stage, |absf|, to incorporate into the code at the second stage:
-croos-stage persistent values.
-
-We should now consider the function |fpca| in a browder context: of a
-function that computes the pivot and then swaps the pivot withe 0-th
-element of the array:
-
-\begin{code}
-let upv = .<fun arr n ->
-   let pv = .~(fpca abs_float) arr n in
-   let i = fst pv in
-   (if not (i = 0) then
-      let t = arr.(0) in
-      begin arr.(0) <- arr.(i); arr.(i) <- t end);
-   arr>.
-\end{code}
-
-Here, we see the third MetaOCaml operator, escape |.~|. It does ...
-MetaOcaml can print out the code, which looks like
-|.<fun arr n -> let pv = (fun arr n -> fpcs code) arr n in ... >.|
-That is, the compiler inlined the code for fpca -- which is a
-functional expression. The |upv| code shows the functional
-application, |(fun arr n -> ...)| applied to arguments |arr n|. That
-application will be executed at run-time, unless a sufficiently smart
-compiler can inline it. As we stated, our aim is to not to rely on
-post-generation optimizations and sufficiently smart compiler. We
-strive to generate the efficient code ourselves. We can do that by
-modifying the definition of |fpca| as
-
-\begin{code}
-let fpca absf arr n = .<
-   let pv = ref (0,(.~arr).(0)) in
-   for i = 1 to .~n-1 do
-     if absf (.~arr).(i) > snd !pv then
-        pv := (i, absf (.~arr).(i))
-   done;
-   !pv>.
-\end{code}
-\begin{code}
-let upv = .<fun arr n ->
-   let pv = .~(fpca abs_float .<arr>. .<n>.) in
-   let i = fst pv in
-   (if not (i = 0) then
-      let t = arr.(0) in
-      begin arr.(0) <- arr.(i); arr.(i) <- t end);
-   arr>.
-\end{code}
-% (.! upv) [|1.0; -4.; 2.|] 3;;
-We see that we pass code fragment |.<arr>.| as arguments to the
-function. Now when we look at the code for |upv|, we see
-|.<fun arr n -> let pv1 = let pv = ref (0,arr.(0)) in ...>.|
-So the fpca code got really inlined.
-
-The generator |fpca| shows one particular way of searching for pivot.
-In the domain of exact rational numbers, for example, we would be
-satisfied with the first non-zero element as a pivot, and so we would
-write a different pivot-searching generator. The body of |upv| moves
-the found pivot to the ``canonical'' location. Again, depending on the
-container (a dense matrix, a sparse matrix, etc), there are different
-ways of doing that. So, we would like to be able to ``compose'' pivot
-generating function with the swap generating function. We would like
-to write something like this
-
-\begin{code}
-let fpca absf arr n pv = .<
-   for i = 1 to .~n-1 do
-     if absf (.~arr).(i) > snd ! (.~pv) then
-        (.~pv) := (i, absf (.~arr).(i))
-   done>.
-let swapper arr n pv = .<
-   let i = fst !(.~pv) in
-   if not (i = 0) then
-      let t = (.~arr).(0) in
-      begin (.~arr).(0) <- (.~arr).(i); (.~arr).(i) <- t end>.
-let seqv a b c = .<begin .~a; .~b; .~c end>.
-let pv_swap absf arr n =
-   let pv = .<ref (0,(.~arr).(0))>. in
-   seqv (fpca absf arr n pv)
-        (swapper arr n pv)
-        arr
-let top =  .<fun arr n -> .~(pv_swap abs_float .<arr>. .<n>.)>.
-\end{code}
-
-The function |pv_swap| shows our ideal -- composing code gerenators
-|fpca| and |swapper|, using a higher-order combinator `seqv'
-(sequence of code fragments). We lifted the pivot declaration out of
-old version of |fpca| so that both |fpca| and |swapper| could be
-parameterized by the pivot (and |fpca| and similar function will deal
-only with pivot determination). |top| puts it all together. The above
-code represents the pattern we will be following for the rest of the
-paper.
-
-And it \emph{almost} works. It is type correct.  When we look at the
-generated code for |top|, we notice the |(ref (0, arr_1.(0)))| is
-inlined wherever pivot was expected. For example, the swapping part of
-the code begins with |let i_3 = (fst (! (ref (0, arr_1.(0))))) in ...|. 
-That is certainly not what we want. We meant for pivot searching
-and swapping parts of the code to communicate via a shared mutable
-state, variable |pv|. In the generated code, each piece of code gets
-its own copy of |pv|. Rather then inline |(ref (0, arr_1.(0)))|
-everywhere, we should have generated |let pv = ref (0,(.~arr).(0)) in ...|
-and inlined |pv|. Incidentally, even in a pure functional code we
-often wish to generate such a |let| statement so we can compute a
-complex expression only once and use it value in several other
-places. Again, we would rather not leave the job of detecting such
-common subexpressions to the compiler (because in the complex code the
-compiler may fail to detect the common subexpressions). The problem
-with generating such |let| expressions is that we wish our geenrator
-to return just |.<let pv = ref (0,(.~arr).(0)) in>.|, without the `body'
-of the |let|, which we would fill in later. In languages which
-generate code in the string form this is easy. We generate the string
-|''let pv = ref (0,(.~arr).(0)) in''| and then append to it the string
-representting the body, wcich would give us the complete |let|
-statement. In MetaOCaml, we can't generate incomplete expressions, for
-a good reason.
-
-To solve the problem of writing a generator for a |let| expression 
-when the generated body is not yet known, we use CPS.
-}
-
-%%% Old, detailed version is here
-
-\section{CPS and the problem of generating names}\label{CPS}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% And now, a better -- and shorter -- write-up.
-
-The essence of our approach is assembling the code of the program by
-combining instances of various aspects together. In the case of
-Gaussian Elimination, the algorithmic aspects are pivot determination, 
-swapping of
-rows and columns, determinant tracking, etc. A primitive code
-generator yields a code expression.  A combinator takes (complete and
-well-typed) code expressions and combines them into a composite expression
-(which is also statically guaranteed to be well-formed and well-typed).
-
-We will be using MetaOcaml which, as an instance of a multi-stage
-programming system \cite{TahaThesis}, provides exactly the needed
-features: to construct a code expression, to combine them, and to
-execute them.
-
-For example, the following is probably the simplest code generator:
-\begin{code}
-let one = .<1>.
-\end{code}
-
-In the floating-point instance of that aspect, we would have
-|.<1.0>.| instead. We use MetaOCaml brackets |.<...>.| to generate
-code expressions -- in other words, to construct future-stage
-computations. The simplest combination is inlining:
-\begin{code}
-let plus x y = .<.~x + .~y>.
-\end{code}
-Here we see the second component of MetaOCaml -- escapes |.~|. They
-let us perform an immediate code generating computation while we are
-building the future-stage computation (code expression). In the
-example above, the immediate computation |.~x| is trivial -- obtaining
-the value of the variable |x|, which should be a code expression, and
-inlining it. The following shows an example of using the simple
-generators:
-
-\begin{code}
-let simplest_code =
-  let gen x y = plus x (plus y one) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-\end{code}
-
-Here, the immediate computation |gen .<x>. .<y>.| is less trivial --
-it includes the evaluation of the function |gen|, which in turn
-applies the function |plus|. The function receives code expressions
-|.<x>.| and |.<y>.| as arguments. At the generating stage, we can
-manipulate code expressions as (opaque) values. The function |gen| returns a
-code expression, which will be inlined in the place of the
-escape. MetaOCaml can print out code expressions, so we can see the
-final generated code:\\
-  |.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|
-
-We must point out that |gen x y| looks like a combinator; its body 
-involves several applications of the function |plus|. When
-we look at the generated code however, we see no traces of the function
-|plus| and no such administrative applications. Instead, we see the
-fully inlined expression. These administative applications are done at
-the code generation stage.
-
-The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
-execute the code expression: |.! simplest_code| is a function of two
-integers, which we can apply: |(.! simplest_code) 1 2|. The original
-|simplest_code| is not the function on integers -- it is a code
-expression.
-
-To see the benefit of code generation, we notice that we can easily
-parameterize our code:
-
-\begin{code}
-let simplest_param_code plus one =
-  let gen x y = plus x (plus y one) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-\end{code}
-
-\noindent and use it to generate code that operates on integers, floating point
-numbers or booleans -- in general, any domain that implements |plus|
-and |one|:
-\begin{code}
-let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
-  simplest_param_code plus one
-let plus x y = .<.~x || .~y>. and one = .<true>. in
-  simplest_param_code plus one
-\end{code}
-Running the former expression yields the function on |float|s, whereas
-the latter expression is the code expression for a boolean function.
-
-This clearly shows the separation of concerns, namely of that for domain
-operations.
-
-Let us consider a more complex expression:
-\begin{code}
-let param_code1 plus one =
-  let gen x y = plus (plus y one) (plus x (plus y one)) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-\end{code}
-
-We notice two occurrences of |plus y one|. Depending on the
-implementation of |plus| that may be quite a complex computation, and
-so we would rather not do it twice. We may be tempted to rely on the
-compiler's common-subexpression elimination optimization. We would
-rather not rely on a ``sufficiently smart compiler'': when the
-generated code is very complex, the compiler may overlook common
-subexpressions.  Or the subexpressions may occur in an imperative
-context where the compiler might not be able to determine if lifting
-them is sound. So, being conservative, the optimizer will leave the
-duplicates as they are. We may attempt to write
-
-\begin{code}
-let param_code1 plus one =
-  let gen x y = 
-     let ce = (plus y one) in  plus ce (plus x ce) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-\end{code}
-
-However, the result of |param_code1 plus one|, which prints as\\
-|.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|,
-still exhibits duplicate sub-expressions. Our |let|-insertion
-optimization saved the computation at the generating stage. 
-We need a combinator that inserts the |let| expression in the
-generat\emph{ed} code. We need a combinator |letgen| to be used
-as\\|let ce = letgen (plus y one) in plus ce (plus x ce)| 
-yielding the code like |.<let t = y + 1 in t + (x + t)>.|
-But that seems impossible because |letgen exp| has to generate
-the expression\\|.<let t = exp in body>.| although |letgen| does not have
-the |body| yet. The body needs a temporary identifier |.<t>.| that
-is supposed to be the result of |letgen| itself. 
-Certainly |letgen| cannot generate only part of a let-expression,
-without the |body|,  as all generated expressions in
-MetaOCaml are well-formed and complete.
-
-The key is to use continuation-passing style (CPS). Its benefits were first
-pointed out by \cite{Bondorf:92} in the context
-of partial evaluation, and extensively used by \cite{KiselyovTaha} for code
-generation.
-
-\begin{code}
-let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
-let param_code2 plus one =
-  let gen x y k = letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
-  and k0 x = x
-  in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
-\end{code}
-
-Now, |param_code2 plus one| gives us the desired code\\
-|.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.|.
-
-\section{Monadic notation, making CPS code clear}\label{monadicnotation}
-
-Comparing the code that did let-insertion at the generating stage\\
-|let ce = (plus y one) in  plus ce (plus x ce)|\\
-with the corresponding code inserting let at the generated code stage\\
-|letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))|\\
-clearly shows the difference between the direct-style and CPS code.
-What was |let ce = init in ...| in direct style became
-|init' (fun ce -> ...)| in CPS. For one thing, |let| became
-inverted. For another, what used to be an expression that yields
-a value, |init|, now became an expression that takes an extra argument,
-the continuation, and invokes it. The differences look negligible in
-the above example. In larger expressions with many let-forms, the
-number of parentheses around |fun| increases, the need to add and
-then invoke the |k| continuation argument become increasingly annoying. The
-inconvenience is great enough for some people to explicitly avoid CPS
-or claim that numerical programmers (our users) cannot or will not
-program in CPS. Clearly a better notation is needed.
-
-The |do|-notation of Haskell \cite{Haskell98Report} shows that it is possible
-to write essentially CPS code in conventional-looking style. The
-|do| notation is the notation for writing monadic code \cite{moggi-notions}.
-The benefit of monadic notation is not that it can represent CPS \cite{Filinski:Representing}, but it helps in composability by offering to add different
-layers of effects (state, exception, non-determinism, etc) to the
-basic monad \cite{liang-interpreter} in a controlled way.
-
-A monad \cite{moggi-notions} is an abstract datatype representing
-computations that yield a value and may have an \emph{effect}.
-The datatype must have at least two operations, |return| to build
-trivial effect-less computations and |bind| for combining
-computations. These operations must satisfy \emph{monadic laws}:
-|return| being the left and the right unit of |bind| and |bind| being
-associative. Figure~\ref{ourmonad} defines the monad used throughout
-the present paper and shows its implementation.
-
-\begin{figure}\label{ourmonad}
-\begin{code}
-type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
-let ret a = fun s k -> k s a
-let bind a f = fun s k -> a s (fun s' b -> f b s' k)
-let fetch s k = k s s
-let store v s k = k (v::s) ()
-
-let k0 s v = v
-let runM m = m [] k0
-
-let l1 f = fun x -> mdo { t <-- x; f t}
-let l2 f = fun x y -> mdo { tx <-- x; ty <-- y; f tx ty}
-
-let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
-
-let ifL test th el = ret .< if .~test then .~th else .~el >.
-let ifM test th el = fun s k ->
-  k s .< if .~(test s k0) then .~(th s k0) else .~(el s k0) >.
-\end{code}
-\caption{Our monad}
-\end{figure}
-
-Our monad represents two kinds of computational effects: reading and
-writing a computation-wide state, and control effects. The latter are
-normally associated with exceptions, forking of computations, etc. --
-in general, whenever a computation ends with something other than
-invoking its natural continuation in the tail position. In our case
-however the control effects manifest themselves as code generation.
-
-In Figure~\ref{ourmonad}, the monad is implemented as a function of two
-arguments: the state (of type |s|) and the continuation. The
-continuation receives the current state, the value (of the type |v|) and
-yields the answer of the type |w|.  The monad is polymorphic over all
-these three type parameters.  Other implementations are
-possible. Except for the code in Figure~\ref{ourmonad}, the rest of our code
-treats the monad as a truly abstract data type. The implementation of basic
-monadic operations |ret| and |bind| are conventional. It is easy to
-see that the monadic laws are satisfied.  Other monadic operations
-construct computations that do have effects.  Operations |fetch| and
-|store v| construct computations that read and write the state. In our
-case the state is a list (of polymorphic variants), which models an
-open discriminated union, as we shall see later.
-
-The operation |retN a| is the let-insertion operation, whose simpler
-version we called |letgen| earlier. It is the first computation with
-a control effect: indeed, the result of |retN a| is \emph{not} the
-result of invoking its continuation |k|. Rather, its result is a |let|
-code expression. Such a behavior is symptomatic of cntrol operators
-(in particular, |abort|).
-
-Finally, |runM| runs our monad, that is, performs the computation of
-the monad and gets the result, which in our case is the code
-expression. We run the monad by passing it the initial state and the
-initial continuation |k0|. We can now re-write our |param_code2|
-example of the previous section as
-
-\begin{code}
-let param_code3 plus one =
-  let gen x y = bind (retN (plus y one)) (fun ce -> 
-                ret (plus ce (plus x ce)))
-  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
-\end{code}
-% param_code3 plus one;;
-
-That does not seem like much of an improvement. But here we can introduce the
-do notation (Appendix A), which is patterned after the do-notation of Haskell.
-
-\begin{code}
-let param_code4 plus one =
-  let gen x y = mdo { ce <-- retN (plus y one);
-                      ret (plus ce (plus x ce)) }
-  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
-\end{code}
-
-The function |param_code4| is totally equivalent to |param_code3| --
-the |mdo|-notation is just syntactic sugar. In fact, the camlp4
-preprocessor (with our extension) will transform |param_code4|
-into exactly |param_code3|. And yet, |param_code4| looks far more
-conventional, as if it were indeed in direct style.
-
-We can write operations that generate code other than let-statements,
-e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
-albeit straightforward, is not as general as we wish: its argument are
-already generated pieces of code rather than monadic values. So, we
-may want to ``lift it'':
-\begin{code}
-let ifM' test th el = mdo {
-  testc <-- test;
-  thc   <-- th;
-  elc   <-- el;
-  ifL testc thc elc}
-\end{code}
-
-\noindent We can define functions |l1|, |l2|, |l3| (analogues of |liftM|,
-|liftM2|, |liftM3| of Haskell) to make such a lifting generic.
-We soon notice the need for another |ifM| function, with the same
-interface (see |ifM| in Figure~\ref{ourmonad}). The difference between them is
-apparent from the following:
-\begin{code}
-let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
-                           (ret .<None>.)
- in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
-;;
-.<fun a_1 -> fun i_2 ->
-    let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.
-\end{code}
-\noindent whereas if we replace |ifM'| with |ifM|, the result will be
-\begin{code}
-.<fun a_1 -> fun i_2 ->
-    if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.
-\end{code}
-
-The difference is apparent: in the code with |ifM'|, the let-insertion
-happend \emph{before} the if-expression, that is, before the test that
-the index |i| is positive. That is, if |i| turned out
-negative, |a.(i)| will generate an out-of-bound array access
-error. On the other hand, the code with |ifM| accesses the array only
-when we have verified that the index is non-negative. This example
-makes it clear that the code generation (such as the one in |retN|) is 
-truly an effect and we have to be clear about the sequencing of
-effects, when generating control constructions such as conditionals.
-The form |ifM| handles the such effects correctly. In our code, we
-need similar operators for other Ocaml control forms: for generating
-case-matching statements and for- and while-loops.
-
-% old intro
-\section{Introduction}
-
-In high-performance, symbolic, and numeric computing, there is a well-known
-issue of balancing between maximal performance and the level of
-abstraction at which code is written.  Furthermore, already in
-linear algebra, there is a wealth of different concerns that 
-\emph{may} need to be addressed.  While it is possible to manually
-inline and simplify such code, this is labour-intensive and 
-error-prone.  The resulting code is frequently unreadable,
-and thus unmaintainable.
-
-There are a number of projects that have attempted to address these issues
-\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser94algorithmoriented,BOOST,POOMA,ATLAS}, 
-and they have generally succeeded in generating
-efficient code, but the cost was frequently the loss of modularity
-of the design.  In other words, to allow the various aspects to
-weave with each other \emph{efficiently}, information which should
-have been kept hidden leaked from module to module.  This seriously
-affects maintainability and scalability.  Furthermore, current architectures
-demand more and more frequent tweaks which, in general, cannot be done by the
-compiler because the tweaking often involves domain knowledge.
-
-Our goal was to implement a textbook-style modular design 
-\cite{journals/cacm/parnas72a}, yet 
-have our code be very efficient.  Gaussian elimination, as a  
-well-known algorithm with many different aspects 
-\cite{carette04}, was chosen as a representative example.
-
-Whether one prefers to code in an imperative, object-oriented, functional, 
-or even in a mixed style, all abstraction mechanisms in current
-languages have a non-zero run-time cost.  Whether it is procedure,
-method or function call, these have a cost; worse still, higher level
-and/or higher order abstractions (like Functors or AspectJ's aspects)
-frequently have a larger cost still.  Of course, if one is faced with
-writing 35 different implementations of the ``same'' algorithm
-\cite{carette04}, this
-abstraction cost starts to look cheap!  Although bad programmers have
-been known to resort to cut\&paste programming, no self-respecting
-programmer would be caught doing that.
-The common solution is to turn to code generation.
-
-While many people still make do with string-based code generation,
-this is not much different than writing code in a dynamically typed
-interpreted scripting language.  It works, it is convenient, but it
-does not scale well, it is a debugging nightmare, and usually suffers
-most from Software Aging \cite{parnas_aging}.  As with other tasks in
-programming, the answer is to turn to statically typed languages,
-which provide strong guarantees about the soundness of the resulting
-code.  MetaOCaml \cite{CTHL03,metaocaml-org} in particular stands
-out: not only it guarantees well-formedness and the absense of problems 
-such as accidental variable capture \cite{HygienicMacros}. MetaOcaml also
-statically ensures that both the code generator and the generated code
-are well-typed \cite{TahaSheard97,TahaThesis}.  While Template
-programming in C++ and Template Haskell do offer some amount of static
-typing \cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml
-offers stronger guarantees.  As well, since MetaOCaml is
-implemented on top of the very solid Ocaml language, one accrues
-additional benefits.
-
-While it is theoretically feasible in a strings-based
-generative system to re-parse the generative code to inspect it, such
-an approach rapidly degenerates into complete chaos.
-% \oleg{is there a good citation?}
-% \jacquesP{none that I am aware of}
-In languages
-that natively offer reflection and reification (either at no cost, 
-such as in LISP and Scheme, or at a modest cost in languages like 
-Maple), inspection is actually feasible, even easy.  But this is fragile
-and error-prone, since the languages offer little support for catching
-errors early.
-
-In MetaOCaml, generated code cannot be examined and optimized at all. It must
-be generated just right.  For simple code, this is relatively easy
-(see \cite{TahaThesis} for many examples).  For more complex examples,
-new techniques are necessary.  \cite{KiselyovTaha} introduces one
-such tool: abstract interpretation.  But more problems remain
-\cite{Padua:MetaOcaml:04}:
-\begin{itemize}
-    \item Generating names in the presence of ``if'', ``while'', and other
-        control operators. \cite{KiselyovTaha} showed how to solve
-        this problem in simple cases.
-    \item Making continuation-passing style (CPS) code clear.  Many
-        authors understandably shy away from CPS code as it quickly
-        becomes unreadable.  But this is needed for proper name
-        generation.
-    \item Compositionality of various interleaving aspects, as well
-        as expressing dependencies among them.  For example, if we
-        choose to compute the determinant of a matrix while performing
-        Gaussian Elimination, this requires inserting code in 
-        several specific places in the algorithm.  The method to
-        compute this determinant crucially depends on whether 
-        a division-based or a fraction-free Gaussian Elimination is
-        used.
-    \item As mentionned above, there are dependencies between
-        certain algorithmic choices and properties of the
-        underlying domain, and we would like to statically guarantee
-        that inappropriate combinations cannot be chosen.  For instance,
-        one should not attempt to use full division when dealing
-        with matrices of exact integers, nor is it worthwhile to use
-        full pivoting on a matrix of rational integers ($\mathbb Q $).
-\end{itemize}
-The questions we want to answer is:
-Can we achieve all of the abstractions needed for a flexible 
-Gaussian Elimination, and can
-we eliminate all of that abstraction's overhead in the generated code?
-Of course, we want to do this without using
-intensional code analysis -- that is, maintaining the
-guarantees of MetaOCaml and using a purely generative approach.
-
-Maintaining generativity is not just an interesting exercise:
-one gets a stronger equational theory \cite{Taha2000}, and avoids
-the danger of creating unsoundness \cite{TahaThesis}.  It should
-also be remarked that in general, intensional code analysis
-essentially requires one to insert both an optimizing compiler and 
-an automated theorem proving system into the code generating system
-\cite{Pueschel:05,Kennedy01Telescoping,dongarra7,Veldhuizen:2004}.
-While this is
-potentially extremely powerful and an exciting area of research,
-it is also extremely complex, which means that it is currently more
-error-prone and difficult to ascertain the correctness of the 
-resulting code.
-
-This work makes the following contributions:
-\vspace*{-6pt}
-\begin{itemize}
-    \item use of generative functors for meta-programming
-    \item orderly aspect weaving
-    \item mdo notation for greater monadic code clarity
-    \item use of monads for meta-programming
-    \item use of type constraints to encode domain knowledge and 
-		ensure correctness
-\end{itemize}
-
-The rest of this paper is structured as follows:
-The next section introduces code generation in MetaOCaml, the problem
-of name generation, and how continuation-passing style is a general
-tool for solving that problem.  We also introduce
-the monad that greatly simplifies the task of dealing with
-possibly effect-inducing CPS code combinators.  Section~\ref{functors}
-shows how one can use the advanced module system of OCaml, namely its
-parametrized modules, to encode all of the aspects of the Gaussian
-Elimination algorithm family in completely separate, independent modules.
-We then briefly discuss related work in section~\ref{related} and outline
-some work still to do, and finally draw some 
-conclusions in section~\ref{conclusion}.  As an appendix, we show
-the details of the syntax extension for monadic programming in Ocaml
-and MetaOCaml.
-
-The first author wishes to thank Cristiano Calgano for his help in
-adapting camlp4 for use with MetaOCaml.
+% all the text that used to be here is now in unused.tex
+% same with any text in an \omitnow

### 2005-04-10 paper/unused.tex
Final tweaks - complete bibliography, conclusion, future work.
Also, move unused text to unused.tex to make editing simpler.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% These are old sections from the paper that
% did not end up being used, for various reasons

\section{Appendix A}
The do notation for Ocaml.
To implement the |mdo| notation, we use the Ocaml pre-processor,
camplp4 (cite?). Informally, our notation is
\begin{code}
mdo { exp }
mdo { exp1; exp2 }
mdo { x <-- exp; exp }
mdo { let x = foo in; exp }
\end{code}
which matches closely the Haskell |do| notation, modulo |do|/|mdo| and 
|<-|/|<--|.

Grammar formally:
\begin{code}
mdo { <do-body> }
<do-body> :: =
    "let" var = EXP ("and" var = EXP)* "in" ";" <do-body>
    | EXP
    | (pat <--)? EXP ";" <do-body>
\end{code}
The semantics is given by the following rules re-writing the |mdo|
form into the regular Ocaml code. We assume that the variable |bind|
must be in scope.
\begin{code}
mdo { exp } ===> exp
mdo { pat <-- exp; rest } ===> bind exp (fun pat -> mdo { rest })
mdo { exp; rest } ===> bind exp (fun _ -> mdo { rest })
mdo { let pat = exp in; rest } ===> let pat = exp in mdo { rest }
\end{code}
We also introduce |odo| form, which is equivalent to |mdo| but uses 
a bind `method' rather than a bind `value'.

The major difficulty with the |do| notation is that it cannot truly be
parsed by an LR(n)-grammar for any finite |n| because many |pat| can also
be parsed as an |exp|.

Perhaps give the command line how to load the camlp4 module into
ocaml. Point to the code.

\omitnow{
In this paper, we are concerned with building and composing
generators. We start with the code expressing the algorithm (in
OCaml), and, adding MetaOCaml annotations, turn that code into a
\emph{code expression} that will generate the code when runs. We then
modify that code expression to add various parameters and aspects, to
yield the complete generator.

Here we illustrate the MetaOCaml constrcutions and code generators.

Turning code into the code expression is relatively straightforward
until we come to binding constructs such as |let|. For example, here
is a piece of code that defines a function that searches a given array
|arr| for the maximal (in absolute value) element, and the index of
that element and the found absolute value. This code is representative
of pivot determination. The code uses mutations and destructive
updates -- partly for illustrative purposes and mainly because many
linear algebra algorithms are expressed via an imperative code (so it
is less error-prone to translate the ``canonical'' algorithms with
minimal changes).

\begin{code}
let fp = fun arr n ->
   let pv = ref (0,arr.(0)) in
   for i = 1 to n-1 do
     if abs_float arr.(i) > snd !pv then
        pv := (i, abs_float arr.(i))
   done;
   !pv
\end{code}

This function has the type |float array -> int -> int * float|. By
enclosing the code into MetaOCaml brackets, we obtain the simplest
generator expression.

\begin{code}
let fpc = .<fun arr n ->
   let pv = ref (0,arr.(0)) in
   for i = 1 to n-1 do
     if abs_float arr.(i) > snd !pv then
        pv := (i, abs_float arr.(i))
   done;
   !pv>.
\end{code}

This value has the type |('a, float array -> int -> int * float) code|.
Here |'a| type parameter is needed (ask Walid for the reference).
The main difference between |fp| and |fpc| is that |fp| is a function,
which we can immediately use as \verb^fp [|1.0; -4.; 3.|] 3^. In
contrast, |fpc| is not a function: it is a \emph{code} value that denotes
that function. We would say that the code expression is the
``second-stage'' value. The ability to manipulate code expressions as
regular values is the distinguished fature of MetaOCaml.
We can ``compile'' the code by using MetaOCaml |run| command, or |.!|:
|.! fpc|. The result is equivalent is |fp|. Although |fpc| doesn't
seem to do much, it is already useful as we can use offshoring (cite)
to translate |fpc| into C or Fortran code (which we can dynamically
link into the running MetaOCaml code -- or use separately).

The code |fpc| obviously lacks generality: the name |abs_float|
implies that that code deals with |float| arrays. We wish to handle
arrays in various domains, integer, rationals, etc. We can make the
code generic by parametrizing over the |abs| function:

\begin{code}
let fpca absf = .<fun arr n ->
   let pv = ref (0,arr.(0)) in
   for i = 1 to n-1 do
     if absf arr.(i) > snd !pv then
        pv := (i, absf arr.(i))
   done;
   !pv>.
\end{code}

and so |fpca abs_float| will manipulate |float array| and |fpca abs|
will do for the integer arrays. Here we used the value from the first
stage, |absf|, to incorporate into the code at the second stage:
croos-stage persistent values.

We should now consider the function |fpca| in a browder context: of a
function that computes the pivot and then swaps the pivot withe 0-th
element of the array:

\begin{code}
let upv = .<fun arr n ->
   let pv = .~(fpca abs_float) arr n in
   let i = fst pv in
   (if not (i = 0) then
      let t = arr.(0) in
      begin arr.(0) <- arr.(i); arr.(i) <- t end);
   arr>.
\end{code}

Here, we see the third MetaOCaml operator, escape |.~|. It does ...
MetaOcaml can print out the code, which looks like
|.<fun arr n -> let pv = (fun arr n -> fpcs code) arr n in ... >.|
That is, the compiler inlined the code for fpca -- which is a
functional expression. The |upv| code shows the functional
application, |(fun arr n -> ...)| applied to arguments |arr n|. That
application will be executed at run-time, unless a sufficiently smart
compiler can inline it. As we stated, our aim is to not to rely on
post-generation optimizations and sufficiently smart compiler. We
strive to generate the efficient code ourselves. We can do that by
modifying the definition of |fpca| as

\begin{code}
let fpca absf arr n = .<
   let pv = ref (0,(.~arr).(0)) in
   for i = 1 to .~n-1 do
     if absf (.~arr).(i) > snd !pv then
        pv := (i, absf (.~arr).(i))
   done;
   !pv>.
\end{code}
\begin{code}
let upv = .<fun arr n ->
   let pv = .~(fpca abs_float .<arr>. .<n>.) in
   let i = fst pv in
   (if not (i = 0) then
      let t = arr.(0) in
      begin arr.(0) <- arr.(i); arr.(i) <- t end);
   arr>.
\end{code}
% (.! upv) [|1.0; -4.; 2.|] 3;;
We see that we pass code fragment |.<arr>.| as arguments to the
function. Now when we look at the code for |upv|, we see
|.<fun arr n -> let pv1 = let pv = ref (0,arr.(0)) in ...>.|
So the fpca code got really inlined.

The generator |fpca| shows one particular way of searching for pivot.
In the domain of exact rational numbers, for example, we would be
satisfied with the first non-zero element as a pivot, and so we would
write a different pivot-searching generator. The body of |upv| moves
the found pivot to the ``canonical'' location. Again, depending on the
container (a dense matrix, a sparse matrix, etc), there are different
ways of doing that. So, we would like to be able to ``compose'' pivot
generating function with the swap generating function. We would like
to write something like this

\begin{code}
let fpca absf arr n pv = .<
   for i = 1 to .~n-1 do
     if absf (.~arr).(i) > snd ! (.~pv) then
        (.~pv) := (i, absf (.~arr).(i))
   done>.
let swapper arr n pv = .<
   let i = fst !(.~pv) in
   if not (i = 0) then
      let t = (.~arr).(0) in
      begin (.~arr).(0) <- (.~arr).(i); (.~arr).(i) <- t end>.
let seqv a b c = .<begin .~a; .~b; .~c end>.
let pv_swap absf arr n =
   let pv = .<ref (0,(.~arr).(0))>. in
   seqv (fpca absf arr n pv)
        (swapper arr n pv)
        arr
let top =  .<fun arr n -> .~(pv_swap abs_float .<arr>. .<n>.)>.
\end{code}

The function |pv_swap| shows our ideal -- composing code gerenators
|fpca| and |swapper|, using a higher-order combinator `seqv'
(sequence of code fragments). We lifted the pivot declaration out of
old version of |fpca| so that both |fpca| and |swapper| could be
parameterized by the pivot (and |fpca| and similar function will deal
only with pivot determination). |top| puts it all together. The above
code represents the pattern we will be following for the rest of the
paper.

And it \emph{almost} works. It is type correct.  When we look at the
generated code for |top|, we notice the |(ref (0, arr_1.(0)))| is
inlined wherever pivot was expected. For example, the swapping part of
the code begins with |let i_3 = (fst (! (ref (0, arr_1.(0))))) in ...|. 
That is certainly not what we want. We meant for pivot searching
and swapping parts of the code to communicate via a shared mutable
state, variable |pv|. In the generated code, each piece of code gets
its own copy of |pv|. Rather then inline |(ref (0, arr_1.(0)))|
everywhere, we should have generated |let pv = ref (0,(.~arr).(0)) in ...|
and inlined |pv|. Incidentally, even in a pure functional code we
often wish to generate such a |let| statement so we can compute a
complex expression only once and use it value in several other
places. Again, we would rather not leave the job of detecting such
common subexpressions to the compiler (because in the complex code the
compiler may fail to detect the common subexpressions). The problem
with generating such |let| expressions is that we wish our geenrator
to return just |.<let pv = ref (0,(.~arr).(0)) in>.|, without the `body'
of the |let|, which we would fill in later. In languages which
generate code in the string form this is easy. We generate the string
|''let pv = ref (0,(.~arr).(0)) in''| and then append to it the string
representting the body, wcich would give us the complete |let|
statement. In MetaOCaml, we can't generate incomplete expressions, for
a good reason.

To solve the problem of writing a generator for a |let| expression 
when the generated body is not yet known, we use CPS.
}

%%% Old, detailed version is here

\section{CPS and the problem of generating names}\label{CPS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% And now, a better -- and shorter -- write-up.

The essence of our approach is assembling the code of the program by
combining instances of various aspects together. In the case of
Gaussian Elimination, the algorithmic aspects are pivot determination, 
swapping of
rows and columns, determinant tracking, etc. A primitive code
generator yields a code expression.  A combinator takes (complete and
well-typed) code expressions and combines them into a composite expression
(which is also statically guaranteed to be well-formed and well-typed).

We will be using MetaOcaml which, as an instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the needed
features: to construct a code expression, to combine them, and to
execute them.

For example, the following is probably the simplest code generator:
\begin{code}
let one = .<1>.
\end{code}

In the floating-point instance of that aspect, we would have
|.<1.0>.| instead. We use MetaOCaml brackets |.<...>.| to generate
code expressions -- in other words, to construct future-stage
computations. The simplest combination is inlining:
\begin{code}
let plus x y = .<.~x + .~y>.
\end{code}
Here we see the second component of MetaOCaml -- escapes |.~|. They
let us perform an immediate code generating computation while we are
building the future-stage computation (code expression). In the
example above, the immediate computation |.~x| is trivial -- obtaining
the value of the variable |x|, which should be a code expression, and
inlining it. The following shows an example of using the simple
generators:

\begin{code}
let simplest_code =
  let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}

Here, the immediate computation |gen .<x>. .<y>.| is less trivial --
it includes the evaluation of the function |gen|, which in turn
applies the function |plus|. The function receives code expressions
|.<x>.| and |.<y>.| as arguments. At the generating stage, we can
manipulate code expressions as (opaque) values. The function |gen| returns a
code expression, which will be inlined in the place of the
escape. MetaOCaml can print out code expressions, so we can see the
final generated code:\\
  |.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.|

We must point out that |gen x y| looks like a combinator; its body 
involves several applications of the function |plus|. When
we look at the generated code however, we see no traces of the function
|plus| and no such administrative applications. Instead, we see the
fully inlined expression. These administative applications are done at
the code generation stage.

The final MetaOCaml feature, |.!| (pronounced ``run'') can be used to
execute the code expression: |.! simplest_code| is a function of two
integers, which we can apply: |(.! simplest_code) 1 2|. The original
|simplest_code| is not the function on integers -- it is a code
expression.

To see the benefit of code generation, we notice that we can easily
parameterize our code:

\begin{code}
let simplest_param_code plus one =
  let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}

\noindent and use it to generate code that operates on integers, floating point
numbers or booleans -- in general, any domain that implements |plus|
and |one|:
\begin{code}
let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
  simplest_param_code plus one
let plus x y = .<.~x || .~y>. and one = .<true>. in
  simplest_param_code plus one
\end{code}
Running the former expression yields the function on |float|s, whereas
the latter expression is the code expression for a boolean function.

This clearly shows the separation of concerns, namely of that for domain
operations.

Let us consider a more complex expression:
\begin{code}
let param_code1 plus one =
  let gen x y = plus (plus y one) (plus x (plus y one)) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}

We notice two occurrences of |plus y one|. Depending on the
implementation of |plus| that may be quite a complex computation, and
so we would rather not do it twice. We may be tempted to rely on the
compiler's common-subexpression elimination optimization. We would
rather not rely on a ``sufficiently smart compiler'': when the
generated code is very complex, the compiler may overlook common
subexpressions.  Or the subexpressions may occur in an imperative
context where the compiler might not be able to determine if lifting
them is sound. So, being conservative, the optimizer will leave the
duplicates as they are. We may attempt to write

\begin{code}
let param_code1 plus one =
  let gen x y = 
     let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}

However, the result of |param_code1 plus one|, which prints as\\
|.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.|,
still exhibits duplicate sub-expressions. Our |let|-insertion
optimization saved the computation at the generating stage. 
We need a combinator that inserts the |let| expression in the
generat\emph{ed} code. We need a combinator |letgen| to be used
as\\|let ce = letgen (plus y one) in plus ce (plus x ce)| 
yielding the code like |.<let t = y + 1 in t + (x + t)>.|
But that seems impossible because |letgen exp| has to generate
the expression\\|.<let t = exp in body>.| although |letgen| does not have
the |body| yet. The body needs a temporary identifier |.<t>.| that
is supposed to be the result of |letgen| itself. 
Certainly |letgen| cannot generate only part of a let-expression,
without the |body|,  as all generated expressions in
MetaOCaml are well-formed and complete.

The key is to use continuation-passing style (CPS). Its benefits were first
pointed out by \cite{Bondorf:92} in the context
of partial evaluation, and extensively used by \cite{KiselyovTaha} for code
generation.

\begin{code}
let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
let param_code2 plus one =
  let gen x y k = letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
  and k0 x = x
  in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
\end{code}

Now, |param_code2 plus one| gives us the desired code\\
|.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.|.

\section{Monadic notation, making CPS code clear}\label{monadicnotation}

Comparing the code that did let-insertion at the generating stage\\
|let ce = (plus y one) in  plus ce (plus x ce)|\\
with the corresponding code inserting let at the generated code stage\\
|letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))|\\
clearly shows the difference between the direct-style and CPS code.
What was |let ce = init in ...| in direct style became
|init' (fun ce -> ...)| in CPS. For one thing, |let| became
inverted. For another, what used to be an expression that yields
a value, |init|, now became an expression that takes an extra argument,
the continuation, and invokes it. The differences look negligible in
the above example. In larger expressions with many let-forms, the
number of parentheses around |fun| increases, the need to add and
then invoke the |k| continuation argument become increasingly annoying. The
inconvenience is great enough for some people to explicitly avoid CPS
or claim that numerical programmers (our users) cannot or will not
program in CPS. Clearly a better notation is needed.

The |do|-notation of Haskell \cite{Haskell98Report} shows that it is possible
to write essentially CPS code in conventional-looking style. The
|do| notation is the notation for writing monadic code \cite{moggi-notions}.
The benefit of monadic notation is not that it can represent CPS \cite{Filinski:Representing}, but it helps in composability by offering to add different
layers of effects (state, exception, non-determinism, etc) to the
basic monad \cite{liang-interpreter} in a controlled way.

A monad \cite{moggi-notions} is an abstract datatype representing
computations that yield a value and may have an \emph{effect}.
The datatype must have at least two operations, |return| to build
trivial effect-less computations and |bind| for combining
computations. These operations must satisfy \emph{monadic laws}:
|return| being the left and the right unit of |bind| and |bind| being
associative. Figure~\ref{ourmonad} defines the monad used throughout
the present paper and shows its implementation.

\begin{figure}\label{ourmonad}
\begin{code}
type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
let ret a = fun s k -> k s a
let bind a f = fun s k -> a s (fun s' b -> f b s' k)
let fetch s k = k s s
let store v s k = k (v::s) ()

let k0 s v = v
let runM m = m [] k0

let l1 f = fun x -> mdo { t <-- x; f t}
let l2 f = fun x y -> mdo { tx <-- x; ty <-- y; f tx ty}

let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.

let ifL test th el = ret .< if .~test then .~th else .~el >.
let ifM test th el = fun s k ->
  k s .< if .~(test s k0) then .~(th s k0) else .~(el s k0) >.
\end{code}
\caption{Our monad}
\end{figure}

Our monad represents two kinds of computational effects: reading and
writing a computation-wide state, and control effects. The latter are
normally associated with exceptions, forking of computations, etc. --
in general, whenever a computation ends with something other than
invoking its natural continuation in the tail position. In our case
however the control effects manifest themselves as code generation.

In Figure~\ref{ourmonad}, the monad is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state, the value (of the type |v|) and
yields the answer of the type |w|.  The monad is polymorphic over all
these three type parameters.  Other implementations are
possible. Except for the code in Figure~\ref{ourmonad}, the rest of our code
treats the monad as a truly abstract data type. The implementation of basic
monadic operations |ret| and |bind| are conventional. It is easy to
see that the monadic laws are satisfied.  Other monadic operations
construct computations that do have effects.  Operations |fetch| and
|store v| construct computations that read and write the state. In our
case the state is a list (of polymorphic variants), which models an
open discriminated union, as we shall see later.

The operation |retN a| is the let-insertion operation, whose simpler
version we called |letgen| earlier. It is the first computation with
a control effect: indeed, the result of |retN a| is \emph{not} the
result of invoking its continuation |k|. Rather, its result is a |let|
code expression. Such a behavior is symptomatic of cntrol operators
(in particular, |abort|).

Finally, |runM| runs our monad, that is, performs the computation of
the monad and gets the result, which in our case is the code
expression. We run the monad by passing it the initial state and the
initial continuation |k0|. We can now re-write our |param_code2|
example of the previous section as

\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
\end{code}
% param_code3 plus one;;

That does not seem like much of an improvement. But here we can introduce the
do notation (Appendix A), which is patterned after the do-notation of Haskell.

\begin{code}
let param_code4 plus one =
  let gen x y = mdo { ce <-- retN (plus y one);
                      ret (plus ce (plus x ce)) }
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
\end{code}

The function |param_code4| is totally equivalent to |param_code3| --
the |mdo|-notation is just syntactic sugar. In fact, the camlp4
preprocessor (with our extension) will transform |param_code4|
into exactly |param_code3|. And yet, |param_code4| looks far more
conventional, as if it were indeed in direct style.

We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its argument are
already generated pieces of code rather than monadic values. So, we
may want to ``lift it'':
\begin{code}
let ifM' test th el = mdo {
  testc <-- test;
  thc   <-- th;
  elc   <-- el;
  ifL testc thc elc}
\end{code}

\noindent We can define functions |l1|, |l2|, |l3| (analogues of |liftM|,
|liftM2|, |liftM3| of Haskell) to make such a lifting generic.
We soon notice the need for another |ifM| function, with the same
interface (see |ifM| in Figure~\ref{ourmonad}). The difference between them is
apparent from the following:
\begin{code}
let gen a i = ifM' (ret .<(.~i) >= 0>.) (retN .<Some (.~a).(.~i)>.)
                           (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
;;
.<fun a_1 -> fun i_2 ->
    let t_3 = (Some (a_1.(i_2))) in if (i_2 >= 0) then t_3 else (None)>.
\end{code}
\noindent whereas if we replace |ifM'| with |ifM|, the result will be
\begin{code}
.<fun a_1 -> fun i_2 ->
    if (i_2 >= 0) then let t_3 = (Some (a_1.(i_2))) in t_3 else (None)>.
\end{code}

The difference is apparent: in the code with |ifM'|, the let-insertion
happend \emph{before} the if-expression, that is, before the test that
the index |i| is positive. That is, if |i| turned out
negative, |a.(i)| will generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
when we have verified that the index is non-negative. This example
makes it clear that the code generation (such as the one in |retN|) is 
truly an effect and we have to be clear about the sequencing of
effects, when generating control constructions such as conditionals.
The form |ifM| handles the such effects correctly. In our code, we
need similar operators for other Ocaml control forms: for generating
case-matching statements and for- and while-loops.

% old intro
\section{Introduction}

In high-performance, symbolic, and numeric computing, there is a well-known
issue of balancing between maximal performance and the level of
abstraction at which code is written.  Furthermore, already in
linear algebra, there is a wealth of different concerns that 
\emph{may} need to be addressed.  While it is possible to manually
inline and simplify such code, this is labour-intensive and 
error-prone.  The resulting code is frequently unreadable,
and thus unmaintainable.

There are a number of projects that have attempted to address these issues
\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser94algorithmoriented,BOOST,POOMA,ATLAS}, 
and they have generally succeeded in generating
efficient code, but the cost was frequently the loss of modularity
of the design.  In other words, to allow the various aspects to
weave with each other \emph{efficiently}, information which should
have been kept hidden leaked from module to module.  This seriously
affects maintainability and scalability.  Furthermore, current architectures
demand more and more frequent tweaks which, in general, cannot be done by the
compiler because the tweaking often involves domain knowledge.

Our goal was to implement a textbook-style modular design 
\cite{journals/cacm/parnas72a}, yet 
have our code be very efficient.  Gaussian elimination, as a  
well-known algorithm with many different aspects 
\cite{carette04}, was chosen as a representative example.

Whether one prefers to code in an imperative, object-oriented, functional, 
or even in a mixed style, all abstraction mechanisms in current
languages have a non-zero run-time cost.  Whether it is procedure,
method or function call, these have a cost; worse still, higher level
and/or higher order abstractions (like Functors or AspectJ's aspects)
frequently have a larger cost still.  Of course, if one is faced with
writing 35 different implementations of the ``same'' algorithm
\cite{carette04}, this
abstraction cost starts to look cheap!  Although bad programmers have
been known to resort to cut\&paste programming, no self-respecting
programmer would be caught doing that.
The common solution is to turn to code generation.

While many people still make do with string-based code generation,
this is not much different than writing code in a dynamically typed
interpreted scripting language.  It works, it is convenient, but it
does not scale well, it is a debugging nightmare, and usually suffers
most from Software Aging \cite{parnas_aging}.  As with other tasks in
programming, the answer is to turn to statically typed languages,
which provide strong guarantees about the soundness of the resulting
code.  MetaOCaml \cite{CTHL03,metaocaml-org} in particular stands
out: not only it guarantees well-formedness and the absense of problems 
such as accidental variable capture \cite{HygienicMacros}. MetaOcaml also
statically ensures that both the code generator and the generated code
are well-typed \cite{TahaSheard97,TahaThesis}.  While Template
programming in C++ and Template Haskell do offer some amount of static
typing \cite{conf/dagstuhl/CzarneckiOST03}, MetaOCaml
offers stronger guarantees.  As well, since MetaOCaml is
implemented on top of the very solid Ocaml language, one accrues
additional benefits.

While it is theoretically feasible in a strings-based
generative system to re-parse the generative code to inspect it, such
an approach rapidly degenerates into complete chaos.
% \oleg{is there a good citation?}
% \jacquesP{none that I am aware of}
In languages
that natively offer reflection and reification (either at no cost, 
such as in LISP and Scheme, or at a modest cost in languages like 
Maple), inspection is actually feasible, even easy.  But this is fragile
and error-prone, since the languages offer little support for catching
errors early.

In MetaOCaml, generated code cannot be examined and optimized at all. It must
be generated just right.  For simple code, this is relatively easy
(see \cite{TahaThesis} for many examples).  For more complex examples,
new techniques are necessary.  \cite{KiselyovTaha} introduces one
such tool: abstract interpretation.  But more problems remain
\cite{Padua:MetaOcaml:04}:
\begin{itemize}
    \item Generating names in the presence of ``if'', ``while'', and other
        control operators. \cite{KiselyovTaha} showed how to solve
        this problem in simple cases.
    \item Making continuation-passing style (CPS) code clear.  Many
        authors understandably shy away from CPS code as it quickly
        becomes unreadable.  But this is needed for proper name
        generation.
    \item Compositionality of various interleaving aspects, as well
        as expressing dependencies among them.  For example, if we
        choose to compute the determinant of a matrix while performing
        Gaussian Elimination, this requires inserting code in 
        several specific places in the algorithm.  The method to
        compute this determinant crucially depends on whether 
        a division-based or a fraction-free Gaussian Elimination is
        used.
    \item As mentionned above, there are dependencies between
        certain algorithmic choices and properties of the
        underlying domain, and we would like to statically guarantee
        that inappropriate combinations cannot be chosen.  For instance,
        one should not attempt to use full division when dealing
        with matrices of exact integers, nor is it worthwhile to use
        full pivoting on a matrix of rational integers ($\mathbb Q $).
\end{itemize}
The questions we want to answer is:
Can we achieve all of the abstractions needed for a flexible 
Gaussian Elimination, and can
we eliminate all of that abstraction's overhead in the generated code?
Of course, we want to do this without using
intensional code analysis -- that is, maintaining the
guarantees of MetaOCaml and using a purely generative approach.

Maintaining generativity is not just an interesting exercise:
one gets a stronger equational theory \cite{Taha2000}, and avoids
the danger of creating unsoundness \cite{TahaThesis}.  It should
also be remarked that in general, intensional code analysis
essentially requires one to insert both an optimizing compiler and 
an automated theorem proving system into the code generating system
\cite{Pueschel:05,Kennedy01Telescoping,dongarra7,Veldhuizen:2004}.
While this is
potentially extremely powerful and an exciting area of research,
it is also extremely complex, which means that it is currently more
error-prone and difficult to ascertain the correctness of the 
resulting code.

This work makes the following contributions:
\vspace*{-6pt}
\begin{itemize}
    \item use of generative functors for meta-programming
    \item orderly aspect weaving
    \item mdo notation for greater monadic code clarity
    \item use of monads for meta-programming
    \item use of type constraints to encode domain knowledge and 
		ensure correctness
\end{itemize}

The rest of this paper is structured as follows:
The next section introduces code generation in MetaOCaml, the problem
of name generation, and how continuation-passing style is a general
tool for solving that problem.  We also introduce
the monad that greatly simplifies the task of dealing with
possibly effect-inducing CPS code combinators.  Section~\ref{functors}
shows how one can use the advanced module system of OCaml, namely its
parametrized modules, to encode all of the aspects of the Gaussian
Elimination algorithm family in completely separate, independent modules.
We then briefly discuss related work in section~\ref{related} and outline
some work still to do, and finally draw some 
conclusions in section~\ref{conclusion}.  As an appendix, we show
the details of the syntax extension for monadic programming in Ocaml
and MetaOCaml.

The first author wishes to thank Cristiano Calgano for his help in
adapting camlp4 for use with MetaOCaml.

As an example, below is
  an implementation of a matrix as an array of rows (themselves
  arrays).  This implementation clearly shows how swapping rows is
  more efficient than swapping columns in this representation.  We
  have not (yet) implemented the further potential optimization
  wherein if it is known that part of two rows/columns are already
  known to be equal (in fact to |Dom.zero| in the case of GE) and
  row/column exchange needs to be performed element-wise, then certain
  swaps are unecessary.

\begin{code}
module GenericArrayContainer(Dom:DOMAIN) =
  struct
  type contr = Dom.v array array (* Array of rows *)
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let get' x n m = .< (.~x).(.~n).(.~m) >.
  let get x n m = ret (get' x n m)
  let set' x n m y = .< (.~x).(.~n).(.~m) <- .~y >.
  let set x n m y = ret (set' x n m y)
  let dim2 x = .< Array.length .~x >.       (* number of rows *)
  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
  let mapper g a = match g with
      | Some f -> .< Array.map (fun x -> Array.map .~f x) .~a >.
      | None   -> a
  let copy = (fun a -> .<Array.map (fun x -> Array.copy x) 
                       (Array.copy .~a) >. )
  (* this can be optimized with a swap_rows_from if it is known that
     everything before that is already = Dom.zero *)
  let swap_rows_stmt a r1 r2 =
      .< let t = (.~a).(.~r1) in
         begin 
             (.~a).(.~r1) <- (.~a).(.~r2);
             (.~a).(.~r2) <- t
         end >.
  let swap_cols_stmt a c1 c2 = .< 
      for r = 0 to .~(dim2 a)-1 do
          let t = (.~a).(r).(.~c1) in
          begin 
              (.~a).(r).(.~c1) <- (.~a).(r).(.~c2);
              (.~a).(r).(.~c2) <- t
          end
      done  >.
end
\end{code}
\noindent  Implementing other containers based on flattened vectors,
Fortran-style access, |Bigarray| or |Bigarray.Array2| is also possible.
Naturally, for proper use of |Bigarray| with specific types, it is 
important that the |DOMAIN| type matches -- but luckily the compiler 
will prevent us from making such a mistake.

### 2005-04-10 paper/metamonads.tex
full spell check

@@
 example of our paper --- may need to account for the representation of
-the matrix, whether to compute and return the detrminant or rank, how
+the matrix, whether to compute and return the determinant or rank, how
 and whether search for pivot, etc. Furthermore, current architectures
@@
 We build code generators out of primitive ones using code generation 
-combinators. MetaOcaml, as an instance of a multi-stage
+combinators. MetaOCaml, as an instance of a multi-stage
 programming system \cite{TahaThesis}, provides exactly the needed
@@
 
-A monad \cite{moggi-notions} is an abstract datatype representing
+A monad \cite{moggi-notions} is an abstract data type representing
 computations that yield a value and may have an \emph{effect}.
-The datatype must have at least two operations, |return| to build
+The data type must have at least two operations, |return| to build
 trivial effect-less computations and |bind| for combining
@@
 
-Parametrizing by the kind of container representing a matrix is
+Parameterizing by the kind of container representing a matrix is
 almost as straightforward.  Our containers are  parametric
@@
 parameterized by the domain, container, pivoting policy (full, row,
-nonzero, nopivoting), update policy (that use either ``fraction-less'
+nonzero, no pivoting), update policy (that use either ``fraction-less'
 or full division), and by what to yield as the result. Some of the
@@
 with different but type-compatible instances of |DOMAIN|. That is, we
-are not concerened at all about the value sharing. Aspects such as
+are not concerned at all about the value sharing. Aspects such as
 determinant tracking may be stateful so that the determinant update
@@
 sharing constraint expresses such domain-specific knowledge:
-instantiating |DivisionUpdate| with |integerDomain| leads to a
+instantiating |DivisionUpdate| with |IntegerDomain| leads to a
 compile-time error, when compiling the \emph{generator} code. Thus, in
@@
 monad with the compositional state makes aspects
-freely composable without having to worry about the presence or absense of
+freely composable without having to worry about the presence or absence of
 value aliasing. The only constraints to compositionality are the
@@
 aspects are (comparatively) lightly typed, and are post-facto extensions of an
-existing piece of code.  Here aspects are weaved together ``from scracth'' to
+existing piece of code.  Here aspects are weaved together ``from scratch'' to
 make up a piece of code/functionality.  One can understand previous work to be

### 2005-04-10 paper/metamonads.tex
simple correction

@@
 That does not seem like much of an improvement. With the help of
-camlp4 pre-processor, we introduce the |mdo|-notation (cite XXX),
+camlp4 pre-processor, we introduce the |mdo|-notation \cite{metamonadsURL},
 patterned after the |do|-notation of Haskell. The function
@@
 
-Note that the monad is similar to the one used in \cite{KiselyovTaha}.
+The monad in this paper is similar to the one used in \cite{KiselyovTaha}.
 However, the latter work used only |retN| of all monadic operations,

### 2005-04-11 paper/metamonads.tex
small corrections

@@
 However, the latter work used only |retN| of all monadic operations,
-and used fixpoints (for performing iterations at generation time).
-In this paper we do not use monadic fixpoints (because the generator
-is not recursive) but we make extensive use of monadic operations for
+and fixpoints (for performing iterations at generation time).
+This paper does not involve monadic fixpoints (because the generator
+is not recursive) makes extensive use of monadic operations for
 generating conditional and looping operations.
@@
 of abstraction.  Using traits and concepts, some domain-specific
-knowledge can also be encoded.  However inlining critically depends on
-the compiler's fully inlining of all methods, and as has been reported
-in the literature, this can be challenging to insure. Furthermore, all errors
-(such as type errors and concept violation errors, that is,
+knowledge can also be encoded.  However overhead elimination
+critically depends on the compiler's fully inlining of all methods,
+which has been reported to be challenging to insure. Furthermore, all
+errors (such as type errors and concept violation errors, i.e.,
 composition errors) are detected only when compiling the generated
-code. It is immensely difficult to correlate errors (e.g., line numbers) 
-to the ones in the generator itself. Furthermore, as templates
-cannot be local, a lot of code generation is scattered around instead
-of being modularized.
+code. It is immensely difficult to correlate errors (e.g., line
+numbers) to the ones in the generator itself.
 
@@
 
-There are many more aspects which can also be handled
+There are many more aspects which can also be handled:
 Input variations (augmented
@@
 \section{Conclusion}\label{conclusion}
-The combination of stateless modules (functors and structures), and our
-monad with the compositional state makes aspects
-freely composable without having to worry about the presence or absence of
-value aliasing. The only constraints to compositionality are the
-typing constraints plus the constraints we specifically impose,
-including semantic constraints (e.g., rings do not have full
+In this paper we demonstrate numerical code extensively parameterized
+by complex aspects at no run-time overhead.  The combination of
+stateless functors and structures, and our monad with the
+compositional state makes aspects freely composable without having to
+worry about value aliasing. The only constraints to compositionality
+are the typing constraints plus the constraints we specifically
+impose, including semantic constraints (e.g., rings do not have full
 division).

### 2005-06-16 paper/metamonads.tex
Fix font in bib Fix figures and their references in paper Try to make font of appendix code smaller Fix some typos

@@
 \DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\mathindent,commandchars=\\\{\},fontsize=\small}
+\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\mathindent,commandchars=\\\{\},fontsize=\footnotesize}
 \newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
@@
 
-\begin{figure}\label{easycode}
+\begin{figure}
 \begin{code}
@@
 \caption{Code generation and combinators. $\Longrightarrow$ under an
-  expression shows the result of its evaluation}
+  expression shows the result of its evaluation}\label{easycode}
 \end{figure}
@@
 
-\begin{figure}\label{ourmonad}
+\begin{figure}
 \begin{code}
@@
 \end{code}
-\caption{Our monad}
+\caption{Our monad}\label{ourmonad}
 \end{figure}
@@
 \end{enumerate}
-\noindent These are separated out from the others are they are cross-cutting
+\noindent These are separated out from the others as they are cross-cutting
 concerns: in the case of the length measure, a property of the domain
@@
 
-\begin{code}
+\begin{code2}
 module type OUTPUT = sig
@@
 end
-\end{code}
+\end{code2}
 
@@
 contain any abstraction artifacts, it is important to make this state
-modular.  For example,
+modular.  For example, in
 \begin{code}
@@
 \begin{figure}
-\begin{code}
+\begin{code2}
 module Gen(Dom: DOMAIN)(C: CONTAINER2D)(PivotF: PIVOT)
@@
 end
-\end{code}
+\end{code2}
 \end{figure}
@@
 with different but type-compatible instances of |DOMAIN|. That is, we
-are not concerned at all about the value sharing. Aspects such as
+are not concerned at all about value sharing. Aspects such as
 determinant tracking may be stateful so that the determinant update
@@
 |Dom.div| operation without restrictions -- which is possible only if
-the domain has such unrestricted operation. A domain such as the integer domain
-may still provide |Dom.div| of the same type, but that operation may
+the domain has such an unrestricted operation. A domain such as the integer
+domain may still provide |Dom.div| of the same type, but that operation may
 only be used when we are sure that the division is exact. Our type
@@
 This paper does not involve monadic fixpoints (because the generator
-is not recursive) makes extensive use of monadic operations for
+is not recursive) but makes extensive use of monadic operations for
 generating conditional and looping operations.
@@
 the determinant and the rank.
-\begin{code}
+\begin{code2}
 # val resIV5 :
@@
     else (~- (! t_411)), (! t_406))>.
-\end{code}
+\end{code2}
 \section{Appendix B}
@@
 represented by a 2D array, row pivoting, returning just the |U|-factor.
-\begin{code}
+\begin{code2}
 #   val resFA1 :
@@
    t_5>.
-\end{code}
+\end{code2}
 \end{document}

### 2005-06-16 paper/metamonads.tex
taking care of extra references; some changes

@@
 \begin{itemize}
-    \item Extending monad of \cite{KiselyovTaha} for generating
-      binding statements when generating control structures such as
-      loops and conditionals. We argue that code generation is an
-      effect and so requires something like a monad for proper
-      serialization.
+    \item Extending a let-insertion monad of
+      \cite{MSP:PADL04,KiselyovTaha} for generating control structures
+      such as loops and conditionals. The extension is non-trivial: e.g.,\\
+      |let x = exp in ...| has a different \emph{effect} within a
+      conditional branch.
     \item Implementation of the |mdo|-notation (patterned after
@@
 full at \cite{metamonadsURL}).
-
-We wish to thank Cristiano Calgano for his help in adapting camlp4 for
-use with MetaOCaml.
 
@@
 
-The key is to use continuation-passing style (CPS). Its benefits were first
-pointed out by \cite{Bondorf:92} in the context
-of partial evaluation, and extensively used by \cite{KiselyovTaha} for code
-generation. Now, |param_code2 plus one| gives us the desired code.
+The key is to use continuation-passing style (CPS). Its benefits were
+first pointed out by \cite{Bondorf:92} in the context of partial
+evaluation, and extensively used by \cite{MSP:PADL04,KiselyovTaha} for
+code generation. Now, |param_code2 plus one| gives us the desired
+code.
 
@@
 
-The monad in this paper is similar to the one used in \cite{KiselyovTaha}.
-However, the latter work used only |retN| of all monadic operations,
-and fixpoints (for performing iterations at generation time).
-This paper does not involve monadic fixpoints (because the generator
-is not recursive) but makes extensive use of monadic operations for
-generating conditional and looping operations.
+The monad in this paper is similar to the one used in
+\cite{MSP:PADL04,KiselyovTaha}.  However, the latter works used only
+|retN| of all monadic operations, and fixpoints (for performing
+iterations at generation time).  This paper does not involve monadic
+fixpoints (because the generator is not recursive) but makes extensive
+use of monadic operations for generating conditional and looping
+operations.
 
@@
 
+We should also mention early work \cite{Gluck95} on automatic
+specialization of mathematical algorithms. Although it can eliminate
+some overhead from a very generic implementation (e.g., by inlining
+aspects implemented as higher-order functions), specialization cannot
+change the type of the function and cannot efficiently handle aspects
+that communicate via a private shared state.
+
+The paper \cite{GluckJ98} describes early simple experiments in
+\emph{automatic} and manual staging, and the multi-level language
+based on an annotated subset of Scheme (which is untyped and has no
+imperative features). The generated code requires post-processing to
+attain efficiency.
+
 To the best of our knowledge, nobody has yet used functors to
@@
 even more direct-style, and potentially clearer.
+We also would like to extend our monad to a monad transformer.
 
@@
 investigating statically typed aspect weaving.
+
+\subsection*{Acknowledgments}
+We wish to thank Cristiano Calgano for his help in adapting camlp4 for
+use with MetaOCaml. We are grateful to anonymous reviewers for many
+helpful suggestions.
 

### 2005-06-17 paper/metamonads.tex
corrections and fixes...

@@
 \begin{itemize}
-    \item Extending a let-insertion monad of
+    \item Extending a let-insertion, memoizing monad of
       \cite{MSP:PADL04,KiselyovTaha} for generating control structures
-      such as loops and conditionals. The extension is non-trivial: e.g.,\\
+      such as loops and conditionals. The extension is non-trivial
+      because of control dependencies and because
+      let-insertion, as we argue, is a control effect on its own.
+\begin{comment}
+      : e.g.,\\
       |let x = exp in ...| has a different \emph{effect} within a
       conditional branch.
+\end{comment}
     \item Implementation of the |mdo|-notation (patterned after
@@
 type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
-let ret a = fun s k -> k s a
+let ret (a :'v) : ('v,'s,'w) monad = fun s k -> k s a
 let bind a f = fun s k -> a s (fun s' b -> f b s' k)
@@
 
-In Figure~\ref{ourmonad}, the monad is implemented as a function of two
+In Figure~\ref{ourmonad}, the monad (yielding values of the type |v|)
+is implemented as a function of two
 arguments: the state (of type |s|) and the continuation. The
-continuation receives the current state, the value (of the type |v|) and
+continuation receives the current state and the value, and
 yields the answer of the type |w|.  The monad is polymorphic over the
@@
 the basic
-monadic operations |ret| and |bind| are conventional. It is easy to
-see that the monadic laws are satisfied.  Other monadic operations
+monadic operations |ret| and |bind| is conventional and clearly
+satisfies the monadic laws. Other monadic operations
 construct computations that do have specific effects.  Operations |fetch| and
@@
  in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
-\evalresult{.<fun a_1 i_2 ->  
+\evalresult{.<fun a_1 i_2 -> \protect\newline
 let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
@@
         exact (remainder-free) division.
-    \item \textbf{Pivoting}: Whether to use no pivoting, 
-        column-wise pivoting, or full pivoting.
-    \item \textbf{Augmented Matrices}: Whether we are doing GE on
-        a full matrix, or only a restricted number of columns, while
-        doing elimination on the full matrix.  We currently
-        do not treat this aspect in our code.
+    \item \textbf{Pivoting}: Whether to use no, 
+        column-wise, or full pivoting.
+    \item \textbf{Augmented Matrices}: Whether all or only some
+      columns of the matrix participate in elimination. We currently
+        do not implement this aspect.
 \end{enumerate}
@@
     \item \textbf{Normalization and zero-equivalence}: Whether the
-        arithmetic operations of the domain at hand gives results in 
-        normalized form, and whether a specialized zero-equivalence 
-        routine needs to be used.
+      arithmetic operations of the domain give normalized results, and
+      whether a specialized zero-equivalence routine is to be used.
 \end{enumerate}
@@
 concerns: in the case of the length measure, a property of the domain
-will influence the method used to do pivoting \emph{if} pivoting is to be
+will influence the pivoting method \emph{if} pivoting is to be
 performed.
@@
 and especially parameterized structures (\emph{functors}).
-We define the type of the domain, the signature |DOMAIN| which
+We define the type of the domain, the signature |DOMAIN|, which
 different domains must satisfy:
@@
 contain any abstraction artifacts, it is important to make this state
-modular.  For example, in
+modular.  For example,
 \begin{code}
@@
 \end{code}
-\noindent  determinant tracking aspects involves being able to
-generate code that defines variables used for tracking (|decl|),
-generate code that updates the sign or the absolute
-value of the determinant, and finally, converting the tracking state
+\noindent  to track determinant we should be able to generate code
+for: defining variables used for tracking (|decl|),
+updating the sign or the absolute
+value of the determinant, converting the tracking state
 to the final determinant value of the type |outdet|. GE of a
@@
 parameterized by the domain, container, pivoting policy (full, row,
-nonzero, no pivoting), update policy (that use either ``fraction-less'
-or full division), and by what to yield as the result. Some of the
+nonzero, no pivoting), update policy (with either `fraction-less'
+or full division), and the result specification. Some of the
 argument modules such as |PIVOT| are functors themselves (parameterized
@@
 previously. But that state is handled via the monadic state. As we
-have shown, open unions makes the overall monadic state compositional
+have shown, open unions make the overall monadic state compositional
 with respect to the state of various aspects.
@@
 
-The monad in this paper is similar to the one used in
-\cite{MSP:PADL04,KiselyovTaha}.  However, the latter works used only
-|retN| of all monadic operations, and fixpoints (for performing
-iterations at generation time).  This paper does not involve monadic
-fixpoints (because the generator is not recursive) but makes extensive
-use of monadic operations for generating conditional and looping
-operations.
-
-|Blitz++| \cite{Veldhuizen:1998:ISCOPE} and in general template
-meta-programming in C++ achieve similar results of eliminating levels
-of abstraction.  Using traits and concepts, some domain-specific
+The monad in this paper is similar to the one described in
+\cite{MSP:PADL04,KiselyovTaha}.  However the latter papers used only
+|retN| and fixpoints (for generation-time iterations).  This paper
+does not involve monadic fixpoints because the generator is not
+recursive, but heavily relies on monadic operations for generating
+conditionals and loops.
+
+|Blitz++| \cite{Veldhuizen:1998:ISCOPE} and {C++} template
+meta-programming in general similarly eliminate levels
+of abstraction.  With traits and concepts, some domain-specific
 knowledge can also be encoded.  However overhead elimination
@@
 imperative features). The generated code requires post-processing to
-attain efficiency.
+attain efficiency. Obtaining a generating extension from annotated
+code is an exciting area of our future research.
 
@@
 to runnable functors in Ocaml, as well as to code generation functors
-in MetaOCaml - for ease of debugging.  The current use of abstract
+in MetaOCaml --- for ease of debugging.  The current use of abstract
 types should make this straightforward.
@@
 \section{Conclusion}\label{conclusion}
-In this paper we demonstrate numerical code extensively parameterized
+In this paper we have demonstrated numerical code extensively parameterized
 by complex aspects at no run-time overhead.  The combination of
@@
 worry about value aliasing. The only constraints to compositionality
-are the typing constraints plus the constraints we specifically
+are the typing ones plus the constraints we specifically
 impose, including semantic constraints (e.g., rings do not have full
@@
 more akin to dynamically typed aspect weaving, while we have started
-investigating statically typed aspect weaving.
+investigating statically typed one.
 
@@
   .<fun a_405 ->
-   let t_406 = (ref 0) in
-   let t_407 = (ref 0) in
+   let t_406 = (ref 0) in let t_407 = (ref 0) in
    let t_408 = {arr = (Array.copy a_405.arr)} (a_405) in
-   let t_409 = a_405.m in
-   let t_410 = a_405.n in
-   let t_411 = (ref 1) in
-   let t_412 = (ref 1) in
+   let t_409 = a_405.m in let t_410 = a_405.n in
+   let t_411 = (ref 1) in let t_412 = (ref 1) in
    while (((! t_407) < t_409) && ((! t_406) < t_410)) do
-    let t_413 = (! t_406) in
-    let t_414 = (! t_407) in
+    let t_413 = (! t_406) in let t_414 = (! t_407) in
     let t_415 = (ref (None)) in
     let t_435 =
-     begin
+     begin  (* full pivoting *)
       for j_431 = t_413 to (t_410 - 1) do
@@
              if ((abs (snd i_434)) > (abs t_433)) then
-              (t_415 := (Some ((j_431, j_432), t_433)))
-             else ()
+              (t_415 := (Some ((j_431, j_432), t_433))) else ()
           | None -> (t_415 := (Some ((j_431, j_432), t_433))))
@@
       (match (! t_415) with
-       | Some (i_416) ->
+       | Some (i_416) ->  (* swapping of columns *)
           if ((snd (fst i_416)) <> t_414) then begin
@@
            (loop_427 t_414 (snd (fst i_416)));
-           (t_412 := (~- (! t_412)))
-          end else ();
-          if ((fst (fst i_416)) <> t_413) then begin
-           let a_417 = t_408.arr
-           and n_418 = t_408.n
-           and m_419 = t_408.m in
-           let i1_420 = (t_413 * m_419)
-           and i2_421 = ((fst (fst i_416)) * m_419) in
-           for i_422 = 0 to (m_419 - 1) do
-            let t_423 = a_417.(i1_420 + i_422) in
-            a_417.(i1_420 + i_422) <- a_417.(i2_421 + i_422);
-            a_417.(i2_421 + i_422) <- t_423
-           done;
-           (t_412 := (~- (! t_412)))
-          end else ();
+           (t_412 := (~- (! t_412))) (* adjust the sign of det *)
+          end else (); (* swapping of rows elided *)
           (Some (snd i_416))
@@
      | Some (i_436) ->
-        begin
+        begin (* elimination loop *)
          for j_437 = (t_413 + 1) to (t_410 - 1) do
@@
             (t_408.arr).((j_437 * t_408.m) + j_438) <-
-             ((((t_408.arr).((j_437 * t_408.m) + j_438) *
-                 (t_408.arr).((t_413 * t_408.m) + t_414)) -
-                ((t_408.arr).((t_413 * t_408.m) + j_438) *
-                  (t_408.arr).((j_437 * t_408.m) + t_413))) / (! t_411))
+             ((((t_408.arr).((j_437 * t_408.m) + j_438) * (* elided *)
            done;
@@
           end else ()
-         done;
-         (t_411 := i_436)
+         done; (t_411 := i_436)
         end;
-        (t_406 := ((! t_406) + 1))
+        (t_406 := ((! t_406) + 1)) (* advance the rank *)
      | None -> (t_412 := 0));
@@
    (t_408, 
-    if ((! t_412) = 0) then 0
+    if ((! t_412) = 0) then 0 (* adjust the sign of the determinant *)
     else if ((! t_412) = 1) then (! t_411)
@@
   .<fun a_1 ->
-   let t_2 = (ref 0) in
-   let t_3 = (ref 0) in
+   let t_2 = (ref 0) in let t_3 = (ref 0) in
    let t_5 = (Array.map (fun x_4 -> (Array.copy x_4)) (Array.copy a_1)) in
@@
    while (((! t_3) < t_6) && ((! t_2) < t_7)) do
-    let t_8 = (! t_2) in
-    let t_9 = (! t_3) in
+    let t_8 = (! t_2) in let t_9 = (! t_3) in
     let t_10 = (ref (None)) in
     let t_16 =
-     begin
+     begin  (* row pivoting *)
       for j_13 = t_8 to (t_7 - 1) do
@@
       (match (! t_10) with
-       | Some (i_11) ->
+       | Some (i_11) -> (* swapping of rows *)
           if ((fst i_11) <> t_8) then begin
@@
            t_5.(t_8) <- t_5.(fst i_11);
-           t_5.(fst i_11) <- t_12;
-           ()
-          end else ();
+           t_5.(fst i_11) <- t_12; () end else ();
           (Some (snd i_11))
@@
      | Some (i_17) ->
-        begin
-         for j_18 = (t_8 + 1) to (t_7 - 1) do
-          if (not ((t_5.(j_18)).(t_9) = 0.)) then begin
-           for j_19 = (t_9 + 1) to (t_6 - 1) do
-            (t_5.(j_18)).(j_19) <-
-             ((t_5.(j_18)).(j_19) -.
-               (((t_5.(j_18)).(t_9) /. (t_5.(t_8)).(t_9)) *.
-                 (t_5.(t_8)).(j_19)))
-           done;
-           (t_5.(j_18)).(t_9) <- 0.
-          end else ()
-         done;
-         ()
-        end;
+        begin (* elimintation loop, elided *) end;
         (t_2 := ((! t_2) + 1))

### 2005-06-23 paper/metamonads.tex
try to save more space - not sure how effective this was; all these changes could be reverted
a few misc changes, including biblioghraphy
replacing mdo for doM

@@
 
-\renewcommand\floatpagefraction{.9}
-\renewcommand\topfraction{.9}
-\renewcommand\bottomfraction{.9}
-\renewcommand\textfraction{.1}   
+\renewcommand\floatpagefraction{.95}
+\renewcommand\topfraction{.95}
+\renewcommand\bottomfraction{.95}
+\renewcommand\textfraction{.05}   
 \setcounter{totalnumber}{50}
@@
 \end{comment}
-    \item Implementation of the |mdo|-notation (patterned after
+    \item Implementation of the |doM|-notation (patterned after
       |do|-notation of Haskell) to make monadic code readable.
@@
 
-let l1 f = fun x -> mdo \{ t <-- x; f t\}
-let l2 f = fun x y -> mdo \{ tx <-- x; ty <-- y; f tx ty\}
+let l1 f = fun x -> doM \{ t <-- x; f t\}
+let l2 f = fun x y -> doM \{ tx <-- x; ty <-- y; f tx ty\}
 
@@
 let param_code4 plus one =
-  let gen x y = mdo \{ ce <-- retN (plus y one);
+  let gen x y = doM \{ ce <-- retN (plus y one);
                       ret (plus ce (plus x ce)) \}
   in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
-let ifM' test th el = mdo \{
+let ifM' test th el = doM \{
   testc <-- test; thc <-- th; elc <-- el;
@@
 That does not seem like much of an improvement. With the help of
-camlp4 pre-processor, we introduce the |mdo|-notation \cite{metamonadsURL},
+camlp4 pre-processor, we introduce the |doM|-notation \cite{metamonadsURL},
 patterned after the |do|-notation of Haskell. The function
-|param_code4|, written in the |mdo|-notation, is equivalent to
+|param_code4|, written in the |doM|-notation, is equivalent to
 |param_code3| -- in fact, the camlp4 preprocessor will convert the
@@
   module D = Det   module R = Rank   module P = DiscardPivot
-  let make_result b = mdo \{ det  <-- D.fin ();  rank <-- R.fin ();
+  let make_result b = doM \{ det  <-- D.fin ();  rank <-- R.fin ();
     ret .< ( .~b, .~det, .~rank ) >. \}
@@
       let zerobelow b r c m n brc =
-        let innerbody i = mdo \{
+        let innerbody i = doM \{
             bic <-- Ctr.get b i c;
@@
                       (Ctr.set b i c Dom.zero)) \} in 
-        mdo \{
+        doM \{
               seqM (retLoopM (Idx.succ r) (Idx.pred n) innerbody) 
                    (Update.update_det brc) \} in
-      let dogen a = mdo \{
+      let dogen a = doM \{
           r <-- Out.R.decl ();
@@
                                        (Idx.less (liftGet r) n) )
-               ( mdo \{
+               ( doM \{
                rr <-- retN (liftGet r);
@@
 
-The paper \cite{GluckJ98} describes early simple experiments in
+The paper \cite{GluckJ97} describes early simple experiments in
 \emph{automatic} and manual staging, and the multi-level language
@@
 We wish to thank Cristiano Calgano for his help in adapting camlp4 for
-use with MetaOCaml. We are grateful to anonymous reviewers for many
+use with MetaOCaml. Many helpful discussions with Walid Taha are very
+appreciated. We are grateful to anonymous reviewers for many
 helpful suggestions.
@@
 \begin{code2}
-# val resIV5 :
-  ('a,
+# val resIV5 : ('a,
    Funct4.GenIV5.Ctr.contr ->
    Funct4.OutDetRank(Funct4.IntegerDomain)(Funct4.GenericVectorContainer)
-                    (Funct4.IDet)(Funct4.Rank).res)
-  code =
+                    (Funct4.IDet)(Funct4.Rank).res) code =
   .<fun a_405 ->
@@
           if ((snd (fst i_416)) <> t_414) then begin
-           let a_424 = t_408.arr
-           and nm_425 = (t_408.n * t_408.m)
+           let a_424 = t_408.arr and nm_425 = (t_408.n * t_408.m)
            and m_426 = t_408.m in
            let rec loop_427 =
-            fun i1_428 ->
-             fun i2_429 ->
+            fun i1_428 -> fun i2_429 ->
               if (i2_429 < nm_425) then
@@
 \begin{code2}
-#   val resFA1 :
-  ('a,
+# val resFA1 : ('a,
    Funct4.GenFA1.Ctr.contr ->
    Funct4.OutJustMatrix(Funct4.FloatDomain)(Funct4.GenericArrayContainer)
-                       (Funct4.NoDet(Funct4.FloatDomain)).res)
-  code =
+                       (Funct4.NoDet(Funct4.FloatDomain)).res) code =
   .<fun a_1 ->

### 2005-06-23 paper/metamonads.tex
a few corrections, mentioning of the timing results.

@@
 at all. The generated code for the above and other instantiations of
-|Gen| can be examined at \cite{metamonadsURL}.
+|Gen| can be examined at \cite{metamonadsURL}. The website also 
+contains benchmark code and timing comparisons.
 
@@
 imperative features). The generated code requires post-processing to
-attain efficiency. Obtaining a generating extension from annotated
-code is an exciting area of our future research.
+attain efficiency.  We are looking into encapsulating staging
+annotations in few functors, so that the rest of the code (in
+particular, the |Gen| functor that puts it all together) should be
+annotation-free and so can be used as it is in one-stage environment
+(pure OCaml) as well as multi-stage environment (generating
+extensions). The one-stage code is a good baseline for benchmarks and
+regression tests. Obtaining a generating extension from properly
+modularized OCaml code (along the lines of our |Gen|) is an exciting
+area of our future research.
 
@@
 family of LU decompositions contains more aspects still.
-
-Finally, we would like to be able to make the |Gen| functor applicable
-to runnable functors in Ocaml, as well as to code generation functors
-in MetaOCaml --- for ease of debugging.  The current use of abstract
-types should make this straightforward.
 

### 2005-06-25 paper/metamonads.tex
run spell check -- only 1 real typo found

@@
 The form |ifM| handles such effects correctly. We
-need similar operators for other Ocaml control forms: for generating
+need similar operators for other OCaml control forms: for generating
 case-matching statements and |for|- and |while|-loops.
@@
 two parameters, so we have to group them. Instead of the grouping
-offered by regular records, we use Ocaml \emph{structures} (i.e.,
+offered by regular records, we use OCaml \emph{structures} (i.e.,
 modules)
@@
      | Some (i_17) ->
-        begin (* elimintation loop, elided *) end;
+        begin (* elimination loop, elided *) end;
         (t_2 := ((! t_2) + 1))

### 2005-06-27 paper/metamonads.tex
minor fixes to english in paper.  add submitted mm.zip file

@@
 knowledge can also be encoded.  However overhead elimination
-critically depends on the compiler's fully inlining of all methods,
+critically depends on full inlining of all methods by the compiler,
 which has been reported to be challenging to insure. Furthermore, all
@@
 imperative features). The generated code requires post-processing to
-attain efficiency.  We are looking into encapsulating staging
-annotations in few functors, so that the rest of the code (in
+attain efficiency.
+
+We are looking into encapsulating staging
+annotations into just a few functors, so that the rest of the code (in
 particular, the |Gen| functor that puts it all together) should be
-annotation-free and so can be used as it is in one-stage environment
-(pure OCaml) as well as multi-stage environment (generating
+annotation-free and thus can be used as is in a one-stage environment
+(pure OCaml) as well as in a multi-stage environment (generating
 extensions). The one-stage code is a good baseline for benchmarks and

### 2006-09-14 TOMS/paper/paper.tex
style files and rough outline of the TOMS paper

\documentclass[acmtoms]{acmtrans2m}

%\acmVolume{V}
%\acmNumber{N}
%\acmYear{YY}
%\acmMonth{M|

\newtheorem{theorem}{Theorem}[section]
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newdef{definition}[theorem]{Definition}
\newdef{remark}[theorem]{Remark}

\markboth{J. Carette and O. Kiselyov}{Our grand title}

\title{So what will our title be?}

\author{JACQUES CARETTE\\ McMaster University
    \and OLEG KISELYOV\\ FNMOC}

\begin{abstract} 
We describe statically typed generative techniques that enable one to write
generic yet highly efficient software.  We demonstrate the effectiveness
of this for various common scientific and symbolic programming tasks.

Furthermore, we also show how these techniques allow for significantly
more modularization than previous work, at no loss of program efficiency.
We are thus able to extract even more design aspects out of linear algebra
codes than previous work.  We are thus able to seamlessly bridge the gap
between numerical and symbolic linear algebra.  The end result is that
the core abstract algorithm becomes the program text for generating all
its specialization, and the various design decisions required to generate
these are made completely explicit.
\end{abstract}

\category{G.4}{Mathematical Software}{???}

\terms{Design,Performance,Languages}

\keywords{Abstraction, genericity, MetaOCaml, metaprogramming}

\begin{document}

\begin{bottomstuff} 
To be filled in later
\end{bottomstuff}

\maketitle

\section{Introduction}

Talk about how, at no loss of efficiency, one can pull out very abstract
algorithms out of countless variants.  This even allows creating new
variants at essentially no cost to the author of the code.  It can pull
out the real core of an algorithm and make visible all the choices that
can be made.  Generalizations then become significantly easier.

In fact, in some cases efficiency can even be gained. 
The easiest way is by taking out dynamic choices and making them static.
Another is to allow more structure (like sparsity) to be taken advantage
of in a higly modular manner; thus authors are more likely to have the
time to code special versions.  Also, various optimization techniques
(like loop unrolling) can be applied at no maintenance cost.

The order of the sections in the paper is not fixed yet.

\section{Related Work}
Must mention Boost (and lots of other C++ work).  Revisit old emails to
find more references.  Not clear if this goes here or towards the end.

\section{Problem}
Describe the problem solved.  Basically a highly typed way to extract
all the design information from code variants into modular code generators
that compose correctly \emph{by construction}.  Requires encoding as
much information as possible in types.  But also, since MetaOCaml is
stage, some information can be encoded in values available at
``compile'' time, and this information can be checked (again at compile
time) for correctness and consistency before any code is generated.
This leads one down partway to dependent types, which will not be 
explored in this particular work.

There are multiple goals:
\begin{itemize}
\item Generate correct code
\item Generate efficient code
\item Modular generator
\item Design-points discovery
\end{itemize}

\section{MetaOCaml}
For TOMS people, an introduction to MetaOCaml will really be needed.  A
comparison to meta-programming in other languages, especially C++, will
also be necessary.

\section{Methodology}
Describe the tools and techniques.  Repeat some of the ideas from the 
GPCE paper, add the new ones.

\section{Case studies}
Go through all our examples, in detail.  Different algorithms have 
different numbers of easily identifiable design points that can be
extracted.  But the same mechanism works accross them all.

\subsection{Gaussian Elimination family}
Describe the members of the family, all the various bits that can be
extracted out.  Some pieces are incompatible --- show how one insures
they cannot successfully be composed with each other (by construction).

The family is much larger than just GE, as it also includes LU and
in general GE-based linear solving.

\subsection{Linear ODE solving family}
Need to put more variants in the code to be able to talk about the 
family some more.

\subsection{Cholesky family}
Obvious question: does this belong here or in GE?  We'll need to 
actually code this to find out I think!

\section{Conclusion}
Always the last thing written?

\bibliographystyle{acmtrans}
\bibliography{paper} % rename this!

\begin{received}
Received Month Year; revised Month Year; accepted Month Year
\end{received}
\end{document}

### 2007-03-16 matrix/incope.hs
play around with various ideas for self interpretation, getting a bit closer each time

@@ -48,7 +48,6 @@ class Functor repr => Symantics repr where
 
     box :: repr a -> repr (Box a)
     unbox :: repr (Box a) -> repr a
-    out :: RR repr a -> repr a; out = unRR
 
 -- The following `projection' function is specific to repr.
 -- It is like `run' of the monad
@@ -161,6 +160,8 @@ instance Symantics L where
     leq e1 e2 = L( unL e1 + unL e2 + 1 )
     if_ be et ee = L( unL be +  unL et + unL ee  + 1 )
 
+    box e = L . unL $ e -- count the box as 0
+    unbox (L e) = L e   -- ditto
 
 compL = unL
 
@@ -589,6 +590,160 @@ test_epr = compR (an_ep int bool add mul app leq if_ lam fix)
 test_epc = compC (an_ep int bool add mul app leq if_ lam fix)
 test_epp = compP (an_ep int bool add mul app leq if_ lam fix)
 
+{-
+  The self-interpreter seems to want this instance to exist
+  or at least it did at one point!   And that was usually 
+  a sign of a bug, so this should not be allowed in, even
+  though it is a rather fun instance.
+instance Functor ((->) t) where
+    fmap f = \y -> f . y
+
+instance Symantics ((->) t) where
+    int x = \y -> x
+    bool b = \y -> b
+
+    lam f = \t -> \a -> let ta = const a in (f ta) t
+    app e1 e2 = \t -> (e1 t) (e2 t)
+
+    add e1 e2 = \y -> (e1 y) + (e2 y)
+    mul e1 e2 = \y -> (e1 y) * (e2 y)
+    leq e1 e2 = \y -> (e1 y) <= (e2 y)
+    if_ be et ee = \y -> if (be y) then (et y) else (ee y)
+-}
+
+-- A plain interpreter, with the right type
+interp :: (Symantics repr) => 
+         ((Int -> repr Int)
+      -> (Bool -> repr Bool)
+      -> (repr Int -> repr Int -> repr Int)  -- add
+      -> (repr Int -> repr Int -> repr Int)  -- mul
+      -> (forall a b. repr (a->b) -> repr a -> repr b)             -- app
+      -> (repr Int -> repr Int -> repr Bool) -- leq
+      -> (forall a. repr Bool -> repr a  -> repr a  -> repr a)     -- if
[…]

### 2007-03-17 matrix/incope.hs
just remove box/unbox for clarity

@@ -29,8 +29,6 @@ module Incope where
 
 -}
 
-newtype Box a = Box a deriving (Eq, Ord, Show)
-
 -- This class defines syntax (and its instances, semantics) of our language
 -- This class is Haskell98!
 class Functor repr => Symantics repr where
@@ -45,9 +43,6 @@ class Functor repr => Symantics repr where
     mul :: repr Int -> repr Int -> repr Int
     if_ :: repr Bool -> repr a -> repr a -> repr a
     leq :: repr Int -> repr Int -> repr Bool
-
-    box :: repr a -> repr (Box a)
-    unbox :: repr (Box a) -> repr a
 
 -- The following `projection' function is specific to repr.
 -- It is like `run' of the monad
@@ -98,9 +93,6 @@ instance Symantics R where
     mul e1 e2 = R( (unR e1) * (unR e2) )
     leq e1 e2 = R( (unR e1) <= (unR e2) )
     if_ be et ee = R( if (unR be) then unR et else unR ee )
-
-    box (R a) = R (Box a)
-    unbox (R (Box a)) = R a
 
 compR = unR
 
@@ -160,9 +152,6 @@ instance Symantics L where
     leq e1 e2 = L( unL e1 + unL e2 + 1 )
     if_ be et ee = L( unL be +  unL et + unL ee  + 1 )
 
-    box e = L . unL $ e -- count the box as 0
-    unbox (L e) = L e   -- ditto
-
 compL = unL
 
 ltest1 = compL . test1 $ ()
@@ -208,8 +197,6 @@ data ByteCode t where
     Leq :: ByteCode t1 -> ByteCode t1 -> ByteCode Bool
     IF  :: ByteCode Bool -> ByteCode t -> ByteCode t -> ByteCode t
     LIFT :: t -> ByteCode t                 -- Used only for eval and fmap
-    BOX :: ByteCode t -> ByteCode (Box t)
-    UNBOX :: ByteCode (Box t) -> ByteCode t
 
 instance Show (ByteCode t) where
     show (Var n) = "V" ++ show n
@@ -224,8 +211,6 @@ instance Show (ByteCode t) where
     show (IF be et ee)
         = "(if " ++ show be ++ 
           " then " ++ show et ++ " else " ++ show ee ++ ")"
-    show (BOX e) = "(box " ++ show e ++ ")"
-    show (UNBOX e) = "(unbox " ++ show e ++ ")"
 
[…]

### 2007-03-17 matrix/incope.hs
Add examples of self-interpretation by evaluation context

@@ -535,6 +535,53 @@ htest2r = eval . test2 $ ()
 htestgib1  = compH . testgib1 $ ()
 htestgib1r = eval . testgib1 $ ()
 
+--------------------------------------------------------------------------------
+
+twice_inc_3_e () =
+    -- The evaluation context in the following four lines is a self-interpreter
+    -- of the object language, encoded in the metalanguage.
+    let lam_ = zlam (\f -> f) in
+    let app_ = zlam (\f -> zlam (\x -> zapp f x)) in
+    let add_ = zlam (\m -> zlam (\n -> zadd m n)) in
+    let int_ = zlam (\i -> i) in
+    -- The term in the following three lines is the object term
+    --      let twice_ f x = f (f x) in let inc_ n = n + 1 in twice_ inc_ 3
+    -- encoded in the object language then encoded in the metalanguage.
+    let twice_ = zapp lam_ (zlam (\f -> zapp lam_ (zlam (\x -> zapp (zapp app_ f) (zapp (zapp app_ f) x))))) in
+    let inc_ = zapp lam_ (zlam (\n -> zapp (zapp add_ n) (zapp int_ (zint 1)))) in
+    zapp (zapp (zapp app_ twice_) inc_) (zapp int_ (zint 3))
+-- *Incope> compC (dynamic (twice_inc_3_e ()))
+-- 5
+
+twice_inc_3_ee () =
+    -- The evaluation context in the following four lines is a self-interpreter
+    -- of the object language, encoded in the metalanguage.
+    let lam_ = zlam (\f -> f) in
+    let app_ = zlam (\f -> zlam (\x -> zapp f x)) in
+    let add_ = zlam (\m -> zlam (\n -> zadd m n)) in
+    let int_ = zlam (\i -> i) in
+    -- The evaluation context in the following four lines is a self-interpreters
+    -- of the object language (the same one as above), encoded in the object
+    -- language then encoded in the metalanguage.  We use two underscores in
+    -- the variable names only because Haskell's "let" is "letrec".
+    let lam__ = zapp lam_ (zlam (\f -> f)) in
+    let app__ = zapp lam_ (zlam (\f -> zapp lam_ (zlam (\x -> zapp (zapp app_ f) x)))) in
+    let add__ = zapp lam_ (zlam (\m -> zapp lam_ (zlam (\n -> zapp (zapp add_ m) n)))) in
[…]

### 2007-03-17 matrix/tagless-final.tex
Added outline

\begin{document}

\section{Introduction}

See Section 1 of Walid02. Our subject area is that of a definitional DSL
interpreter.
Define Staging, meta-language and object language. (Walid uses
`subject' language).

\subsection{The Tag Problem}

\comment{see tagless_interp1.ml, module Tagfull for the complete code.
If you change the code in here, please adjust the .ml file
accordingly. Let the paper and the accompanying code be in sync.}

Consider the sample language of abstraction, application and boolean
literals with deBruin indices as variables. This is the language used
in the body of Walid02.

The language can be described by the following grammar (using OCaml
syntax)

\begin{code}
  type var = VZ | VS of var
  type exp = V of var | B of bool | A of exp * exp | L of exp
\end{code}

and a sample term is
\begin{code}
  let test1 = A ((L (V VZ)),(B true))
\end{code}

The first attempt at the interpreter: take the object language term
such as test1 and give us the value
\begin{code}
  let rec eval0 env = function 
  | V VZ -> List.hd env
  | V (VS v) -> eval0 (List.tl env) (V v)
  | B b -> b 
  | L e -> fun x -> eval0 (x::env) e
  | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

If our meta-language (the language in which we wrote test1
and that interpreter) were untyped, the above code would be acceptable.
The |L e| line exhibits interpretative overhead (see descr in Walid02
paper): |eval0 env' e| will be executed every time the result of
evaluating |L e| is applied. Staging can be used to remove the
interpretative overhead. Cite some paper -- see Walid02, Sec 1.1 and
1.2. They say it well.

If we use OCaml or other typed language as the meta-language, 
the function |eval0| is ill-typed. The line |B b|
tells that the function returns a boolean, whereas the next line says
the result of |eval0| is a function. Clearly, the result of the
function can't be both of these types. 
A related problem is the type of the env: since variables
in the language may be either of boolean or function types, a regular
OCaml list cannot be used for holding the corresponding values.

The solution is to introduce the Universal type (see Walid02 Section
1.3)

\begin{code}
  type u = UB of bool | UA of (u -> u)
\end{code}

We can then write the interpreter, and it works
\begin{code}
  let rec eval env = function
  | V VZ -> List.hd env
  | V (VS v) -> eval (List.tl env) (V v)
  | B b -> UB b
  | L e -> UA (fun x -> eval (x::env) e)
  | A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
\end{code}
whose inferred type is |u list -> exp -> u|,

\begin{code}
  let test1r = eval [] test1
  val test1r : u = UB true 
\end{code}

Unfortunately, the result is tagged: |UB true| rather than just
|true|. We also note that the eval function is partial in two
respects: first there is an inexhaustive pattern-matching in the 
line |A (e1,e2)|.
Second, the occurrences of |List.hd| and |List.tl|.
The failure of the latter corresponds to the evaluation of `open'
terms. One can see that if all our terms are closed, List.hd and
List.tl above will never be applied to the empty list. The partial
pattern match in the |A (e1,e2)| line raises an exception when we try to
evaluate something like |A (B true, B false)|, that is, the first
operand is not a function.

Suppose our object language were typed, too. Then terms like
|A (B true, B false)| would be ill-typed. Thus, if our evaluator
accepted only typed terms, then the missing alternatives in the |A|
rule would not occur. Alas, what is obvious to us is not obvious to
the meta-language. There is still the pattern match, and we still have
to attach and eliminate the UB and UA tags, even when there is no
longer a need for them. See Walid02, Sec 1.4 for more detail.

Of course there is also the issue how to design a type system to type
terms like test1 (GADT seem necessary). Also, how to design the
typesystem to keep track of open/closed status of terms, so we can
statically assure that only closed terms are passed to eval and that
the Nil checking in List.hd and List.tl can be eliminated as well. The
latter is even more interesting problem requiring even more advanced
type systems (cite env classifiers, Nanevski, Rowan Davies).

The problem is also outlined in Simon Peyton-Jones GADT paper, but
only in the context of a first-order language

In short: (put into really introduction or the conclusions?)
It is not satisfactory
to use the Universal type because tagging hurts performance and
necessitates partial pattern matching in the metalanguage, even for a
well-typed object term.

\subsection{Other solutions}
The tag problem motivated GADTs (SPJ, Xi) and dependent types
(WalidICFP02, Concoqtion). Or, perhaps just refer to related work?
But we need to say something here, to give the reader the feeling of
how advanced solutions were proposed and how rarely they are implemented.

\subsection{The Final Proposal}
We have realized that, if we change the encoding of the term from data
constructors to functions, then we can encode a tagless interpreter of a
typed object language in a metalanguage with a very simple type system.

\comment{see tagless_interp1.ml, module Tagless for the complete code.
If you change the code in here, please adjust the .ml file
accordingly. Let the paper and the accompanying code be in sync.}

Now, let us define our object language (embedded DSL) in a different
way. Instead of using algebraic datatypes to encode the terms of the
language in the meta-language (reminiscent to the initial approach),
we will be using functions as term constructors (final approach):

\begin{code}
  let b (bv:bool) env = bv
  let varZ (vc,_) = vc
  let varS vp (_,envr) = vp envr
  let app e1 e2 env = (e1 env) (e2 env)
  let lam e env = fun x -> e (x,env)
\end{code}

That is all of the interpreter. The previous term |test1| now reads
\begin{code}
  let testf1 = app (lam varZ) (b true)
\end{code}

To evaluate the term, we merely need to apply that to the empty
environment

\begin{code}
  let testf1r = testf1 ()
  val testf1r : bool = true
\end{code}

Now we see that the result has no tags. The above functions |b|,
|varZ|, |app| constitute the evaluator. It is patent that the
interpreter has no tags and no pattern matching. A term |b true| is
evaluated to a boolean and a term |lam varZ| is evaluated to a
function, with no tags attached. The application rule than applies that
function without any need for pattern-matching. The evaluator has
another feature: evaluation of open terms leads to a type error rather
than a run-time error. Indeed, the following gives a type error

\begin{code}
  let testf1bad = app (lam (varS varZ)) (b true)
  let testf1badr = testf1bad ()
\end{code}
that the initial environment should be a tuple rather than () --
meaning the term requires non-empty initial environment, the term is
open.

Thus all the errors that previously reported at run-time (open terms,
applications of non-functions) are now reported at compile
time. Running of the interpreter assuredly gives no run-time errors
because its code uses no run-time error-raising constructs.

The interpreter uses no Universal type or pattern matching, and it
is obvious to the compiler that non-termination can only result from
interpreting fix.  If the source language is strongly normalizing, then
our interpreter is total.

We should emphasize that not only our code can't have pattern-match
errors, that fact is apparent to the compiler itself. It is one thing
to write a code with pattern-matching and claim (and prove) that
missing cases cannot occur. However true that may be, the fact the
missing cases cannot occur is not evident to the compiler, which still
has to generate code to report missing cases or compile in the error
generation. That's where efficiency is lost, and Walid's papers on tag
elimination have numbers showing that performance is lost indeed. 
In our code, there is simply no pattern-mismatch opportunity: the
proof of no bad match is expressed in the code itself.

In the above code, the interpreter is wired in, built into the
functions |b|, |app|, etc. We explain how to abstract the interpreter
below, to make it possible to `evaluate' the same term in many different
ways (e.g., interpret term, compile it, compute its size or depth,
etc.)

\subsection{Contributions}
The term `constructor' functions lam, app, etc. look like free
variables in the encoding of an object term.  Defining these functions
differently gives rise to different interpreters.  Given the same term
representation but varying the interpreters, we can
	- evaluate the term to a value in the metalanguage
	- determine the size or depth of a term
	- compute a _CBN_ CPS transformation of a term
	- `compile' the term. This requires staging, as in MetaOCaml.
	- Surprisingly, it is also possible to write a partial
	evaluator!

We have implemented the above in (Meta)OCaml and Haskell.

What are our contributions:
0) we interpret/compile/pe the typed source language
   [Ken says that Danvy's source language is untyped]
1) a much clearer implementation
2) a nicely comparable Haskell/MetaOCaml version
3) a functorial Symantics that deals with interp/comp/pe uniformly
4) clear and easy extensibility to more features
- only closed terms can be evaluated
Making sure that we interpret terms that are well-typed and closed --
and we can do that in far simpler type system (e.g., Haskell or
OCaml). It was thought that doing either requires quite advanced
extensions (GADT, Nanevski, Concoqtion).

\section{Our base language: Tagless interpreter}

Introduce the language, 
  The language is simply-typed lambda-calculus with fixpoint,
  integers, booleans and comparison.

  Lam hoas_fn | App e e | Fix hoas_fn |
  I Int | B Bool | Add ie1 ie2 | Mul ie1 ie2 | Leq ie1 ie2 |
  IF b e-then e-else

this time with HOAS rather than deBruin indices. The language is close 
to the language of Xi03, but without the polymorphic lift and more 
constants so we can write better examples, enough to express
Fibonacci, factorial, and power.

How to make encoding flexible: abstract the interpreter
Haskell: type classes, OCaml: functors (give the signature of
Symantics without the 'sv part)

We have implemented the above in (Meta)OCaml and Haskell.  In Haskell,
the term `constructor' functions are defined as methods in a type class
Symantics (the name means that the class interface gives the syntax for
the source language and its instances give the semantics).
	class Symantics r where
	  int :: Int -> r Int
	  add :: r Int -> r Int -> r Int
	  lam :: (r a -> r b) -> r (a->b)
	  app :: r (a->b) -> r a -> r b
	  fix :: (r a -> r a) -> r a
	  ...
So, Haskell infers the type "Symantics r => r Int" for the example term
encoding above.

encoding of a few terms.

\section{Tagless compiler (aka, staged interpreter)}
Staged code in OCaml... Don't show the Haskell code (too much)

\section{Tagless partial evaluator}

the sv type problem: the inductive type function 'dv -> 'sv
its GADT solution. 
its type-level function solution: incope1.hs 
  Need to thank Martin Sulzmann

How can we do without GADT
Now show the Symantics signature in full, with the 'sv types

It's relation to PE and CPS. Currently, our CPS is CBN. Need to
investigate CBV.

Surprisingly, it is also possible to write a partial evaluator!  The
partial evaluator does have tags: whether a value is static or dynamic.
The pattern-matching on this _phase_ tag is always exhaustive; we still
have no universal type and no tags for object types.  The partial
evaluator uses the previously mentioned compiler to `interpret' dynamic
terms.  In this sense, the partial evaluator is a modular extension to
the compiler.

\section{Variations and discussion}

state and imperative features

CPS transformation (interpreting/compiling the term to CBN CPS)

\subsection{Translucent types}
The crucial role of the higher-order type parameter r

The type constructor "r" above represents a particular interpreter.  The
meta-type "r tau" hides how the interpreter represents the object type
"tau" yet exposes enough of the type information so we can type-check
the encoding of an object term without knowing what "r" is.  The checked
term is then well-typed in any interpreter.  Each instance of Symantics
instantiates "r" to interpret terms in a particular way.

\section{Self-interpretation}

\section{Related work}
Walid's Jones optimality paper. See his own ICFP02 paper that
described what he achieved then and what he _did not_. He did not
eliminate all the tags.

Walid's ICFP02

Walid's PEPM07 (Concoqtion paper)? At leats at the Concoqtion
presentation at WG2.11, Emir used tagless interpreter as the main and
only example.

Danvy's self-interpreter
In sharp contrast to Danvy's paper, all the evaluators in our case
cannot produce any pattern-match error! Except for the partial
evaluator, there are NO patterns to match. And this is critical for
tag elimination -- pattern matching means there are no tags. And in
PE, the pattern-match deals with phases rather than with types. Also,
in our case the exhaustiveness of the pattern match is apparent to the
compiler -- that's why there are no tags.

Simon Peyton-Jones GADTs (the motivation for GADT is tagless
interpreter, but only for the first-order language)

Xi's POPL03 (He uses a higher-order language, one of the motivations
for his GRDT)

Typing Dep Typing plus two messages on Haskell-Cafe: typechecking 
from untyped to the typed form, ready to be interpreted. Walid's ICFP02
uses dept types

\section{Conclusions}

Our approach showed better the meaning of GADT and when they can be
avoided.

\end{document}

### 2007-03-17 matrix/incope.hs
excising the functor from the signature

@@ -31,7 +31,7 @@ module Incope where
 
 -- This class defines syntax (and its instances, semantics) of our language
 -- This class is Haskell98!
-class Functor repr => Symantics repr where
+class Symantics repr where
     int :: Int -> repr Int                -- int literal
     bool :: Bool -> repr Bool             -- bool literal
 
@@ -77,9 +77,6 @@ testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
 -- "Boxes go Bananas" by Washburn and Weirich (intended or otherwise)
 newtype R a = R a deriving Show
 unR (R x) = x
-
-instance Functor R where
-    fmap f (R x) = R (f x)
 
 instance Symantics R where
     int x = R x
@@ -135,9 +132,6 @@ No tags at all...
 
 newtype L a = L Int deriving Show
 unL (L x) = x
-
-instance Functor L where
-    fmap f (L x) = L x
 
 instance Symantics L where
     int x  = L 1
@@ -241,8 +235,10 @@ eval env (Add e1 e2) = eval env e1 + eval env e2
 newtype C t = C (Int -> (ByteCode t, Int)) 
 unC (C t) vc0 = t vc0
 
+{-
 instance Functor C where
     fmap f = app (C(\vc -> (LIFT f, vc)))
+-}
 
 instance Symantics C where
     int x  = C(\vc -> (INT x, vc))
@@ -315,8 +311,10 @@ abstr (VB b) = bool b
 abstr (VF f) = lam (abstr . f . E)
 abstr (E x) = x
 
+{-
 instance Functor P where
     fmap f = E . fmap f . abstr
+-}
 
 instance Symantics P where
     int x  = VI x
@@ -508,9 +506,6 @@ eval (HMul e1 e2) = eval e1 * eval e2
 eval (HLeq e1 e2) = eval e1 <= eval e2
 eval (HIF be et ee) = if (eval be) then eval et else eval ee
 
-instance Functor HByteCode where
-    fmap = HApp . HVar
-
 
 instance Symantics HByteCode where
     int  = HINT
@@ -552,6 +547,24 @@ twice_inc_3_e () =
     zapp (zapp (zapp app_ twice_) inc_) (zapp int_ (zint 3))
 -- *Incope> compC (dynamic (twice_inc_3_e ()))
 -- 5
+
+{-
+twice_inc_3_e () =
+    -- The evaluation context in the following four lines is a self-interpreter
[…]

### 2007-03-17 matrix/incope.hs
small adjustments; generalizing the types

@@ -530,57 +530,63 @@ htest2r = eval . test2 $ ()
 htestgib1  = compH . testgib1 $ ()
 htestgib1r = eval . testgib1 $ ()
 
---------------------------------------------------------------------------------
+-- ---------------------------------------------------------------------------
 
 twice_inc_3_e () =
     -- The evaluation context in the following four lines is a self-interpreter
     -- of the object language, encoded in the metalanguage.
-    let lam_ = zlam (\f -> f) in
-    let app_ = zlam (\f -> zlam (\x -> zapp f x)) in
-    let add_ = zlam (\m -> zlam (\n -> zadd m n)) in
-    let int_ = zlam (\i -> i) in
+    let lam_ = lam (\f -> f) in
+    let app_ = lam (\f -> lam (\x -> app f x)) in
+    let add_ = lam (\m -> lam (\n -> add m n)) in
+    let int_ = lam (\i -> i) in
     -- The term in the following three lines is the object term
     --      let twice_ f x = f (f x) in let inc_ n = n + 1 in twice_ inc_ 3
     -- encoded in the object language then encoded in the metalanguage.
-    let twice_ = zapp lam_ (zlam (\f -> zapp lam_ (zlam (\x -> zapp (zapp app_ f) (zapp (zapp app_ f) x))))) in
-    let inc_ = zapp lam_ (zlam (\n -> zapp (zapp add_ n) (zapp int_ (zint 1)))) in
-    zapp (zapp (zapp app_ twice_) inc_) (zapp int_ (zint 3))
--- *Incope> compC (dynamic (twice_inc_3_e ()))
+    let twice_ = app lam_ (lam (\f -> 
+	   app lam_ (lam (\x -> app (app app_ f) (app (app app_ f) x))))) in
+    let inc_ = app lam_ (lam (\n -> app (app add_ n) (app int_ (int 1)))) in
+    app (app (app app_ twice_) inc_) (app int_ (int 3))
+test_inc3e_r = compR $ twice_inc_3_e ()
+test_inc3e_c = compC $ twice_inc_3_e ()
+test_inc3e_p = compP $ twice_inc_3_e ()
 -- 5
 
 {-
-twice_inc_3_e () =
+twice_inc_3_se () =
     -- The evaluation context in the following four lines is a self-interpreter
[…]

### 2007-03-17 matrix/tagless-final.tex
small adjustments; generalizing the types
Initialize TeX template

@@
+\documentclass[preprint]{sigplanconf}
+\usepackage{amsmath}
+\usepackage{natbib1}
+
+\newcommand{\jacques}[1]{{\it [Jacques says: #1]}}
+\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
+\newcommand{\ccshan}[1]{{\it [Ken says: #1]}}
+
+\usepackage[compact]{fancyvrb1}
+\DefineShortVerb{\|}
+\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=1em,fontsize=\small}
+
 \begin{document}
+
+\conferenceinfo{ICFP '07}{Freiburg, Germany} 
+\copyrightyear{2007} 
+\copyrightdata{[to be supplied]}
+
+\titlebanner{banner above paper title}        % These are ignored unless
+\preprintfooter{short description of paper}   % 'preprint' option specified.
+
+\title{Title Text}
+\subtitle{Subtitle Text, if any}
+
+\authorinfo{Jacques Carette}
+           {McMaster University}
+           {carette@mcmaster.ca}
+\authorinfo{Oleg Kiselyov}
+           {FNMOC}
+           {oleg@pobox.com}
+\authorinfo{Chung-chieh Shan}
+           {Rutgers University}
+           {ccshan@cs.rutgers.edu}
+
+\maketitle
+
+\begin{abstract}
+This is the text of the abstract.
+\end{abstract}
+
+%\category{CR-number}{subcategory}{third-level}
+%\terms term1, term2
+%\keywords keyword1, keyword2
 
@@
 
-\comment{see tagless_interp1.ml, module Tagfull for the complete code.
+\oleg{see tagless\_interp1.ml, module Tagfull for the complete code.
 If you change the code in here, please adjust the .ml file
@@
 
-\comment{see tagless_interp1.ml, module Tagless for the complete code.
+\oleg{see tagless\_interp1.ml, module Tagless for the complete code.
 If you change the code in here, please adjust the .ml file
@@
 	- determine the size or depth of a term
-	- compute a _CBN_ CPS transformation of a term
+        - compute a \emph{CBN} CPS transformation of a term
 	- `compile' the term. This requires staging, as in MetaOCaml.
@@
 
+\begin{code}
   Lam hoas_fn | App e e | Fix hoas_fn |
@@
   IF b e-then e-else
+\end{code}
 
@@
 partial evaluator does have tags: whether a value is static or dynamic.
-The pattern-matching on this _phase_ tag is always exhaustive; we still
+The pattern-matching on this \emph{phase} tag is always exhaustive; we still
 have no universal type and no tags for object types.  The partial
@@
 Walid's Jones optimality paper. See his own ICFP02 paper that
-described what he achieved then and what he _did not_. He did not
+described what he achieved then and what he \emph{did not}. He did not
 eliminate all the tags.
@@
 
+Asai; Simii et al.
+
+"Boxes go Bananas" by Washburn and Weirich (intended or otherwise)
+
 Typing Dep Typing plus two messages on Haskell-Cafe: typechecking 

### 2007-03-17 matrix/incope.hs
Add missing metaness to self-interpretation examples

@@ -545,17 +545,16 @@ twice_inc_3_e () =
     let twice_ = app lam_ (lam (\f -> 
 	   app lam_ (lam (\x -> app (app app_ f) (app (app app_ f) x))))) in
     let inc_ = app lam_ (lam (\n -> app (app add_ n) (app int_ (int 1)))) in
-    app (app (app app_ twice_) inc_) (app int_ (int 3))
+    app (app app_ (app (app app_ twice_) inc_)) (app int_ (int 3))
 test_inc3e_r = compR $ twice_inc_3_e ()
 test_inc3e_c = compC $ twice_inc_3_e ()
 test_inc3e_p = compP $ twice_inc_3_e ()
 -- 5
 
-{-
 twice_inc_3_se () =
     -- The evaluation context in the following four lines is a self-interpreter
     -- of the object language, encoded in the metalanguage.
-    let lam_ = lam (\f -> add (int 1) (f (int 0)) ) in
+    let lam_ = lam (\f -> add (int 1) (app f (int 0)) ) in
     let app_ = lam (\f -> lam (\x -> add (int 1) (add f x))) in
     let add_ = lam (\m -> lam (\n -> add (int 1) (add m n))) in
     let int_ = lam (\i -> int 1) in
@@ -565,12 +564,11 @@ twice_inc_3_se () =
     let twice_ = app lam_ (lam (\f -> 
 	   app lam_ (lam (\x -> app (app app_ f) (app (app app_ f) x))))) in
     let inc_ = app lam_ (lam (\n -> app (app add_ n) (app int_ (int 1)))) in
-    app (app (app app_ twice_) inc_) (app int_ (int 3))
+    app (app app_ (app (app app_ twice_) inc_)) (app int_ (int 3))
 test_inc3se_r = compR $ twice_inc_3_se ()
--- test_inc3e_c = compC $ twice_inc_3_e ()
--- test_inc3e_p = compP $ twice_inc_3_e ()
--- 5
--}
+test_inc3se_c = compC $ twice_inc_3_se ()
+test_inc3se_p = compP $ twice_inc_3_se ()
+-- 10
 
 twice_inc_3_ee () =
     -- The evaluation context in the following four lines is a self-interpreter
@@ -597,7 +595,7 @@ twice_inc_3_ee () =
     -- encoded in the object language then encoded in the metalanguage.
[…]

### 2007-03-18 matrix/tagless-final.tex
small correction

@@
 
-\subsection{The Final Proposal}
+\subsection{The `Final' Proposal}
 We have realized that, if we change the encoding of the term from data
@@
 
+We must formulate some propositions: a typed term makes progress in
+any interpreter. The reason it is typed in any interpreter:
+|forall r. Symantics r => r tau| directly says that in every model the
+term is typed.
+
 \subsection{Translucent types}

### 2007-03-19 matrix/incope.hs
remove some useless stuff, and try a harder test

@@ -551,6 +551,7 @@ test_inc3e_c = compC $ twice_inc_3_e ()
 test_inc3e_p = compP $ twice_inc_3_e ()
 -- 5
 
+-- This is for computing *size*.
 twice_inc_3_se () =
     -- The evaluation context in the following four lines is a self-interpreter
     -- of the object language, encoded in the metalanguage.
@@ -605,6 +606,29 @@ test_inc3ee_p = compP $ twice_inc_3_ee ()
 
 -- test_inc3e_c and test_inc3ee_c are most illustrative.
 
+{- Try to implement a term that is a lambda term rather than a ground
+   term -- and fail!
+test3_e () =
+    -- The evaluation context in the following four lines is a self-interpreter
+    -- of the object language, encoded in the metalanguage.
+    let lam_ = lam (\f -> f) in
+    let app_ = lam (\f -> lam (\x -> app f x)) in
+    let add_ = lam (\m -> lam (\n -> add m n)) in
+    let int_ = lam (\i -> i) in
+    -- The term in the following three lines is the object term
+    -- let test3_ = \x -> (x 1) + 2
+    -- encoded in the object language then encoded in the metalanguage.
+    -- let test3_ = app lam_ (lam (\x ->
+    --              app app_ (
+    --                 app (app add_ (app app_ x (app int_ (int 1)))) 
+    --              (app int_ (int 2)))))
+    let test3_ = app lam_ (lam (\x -> app (app app_ x) (app int_ (int 1))))
+    in test3_ 
+test_t3_r = compR $ test3_e ()
+test_t3_c = compC $ test3_e ()
+test_t3_p = compP $ test3_e ()
+-- ??
+-}
 
 -- start encoding some of Ken's ideas on a self-interpreter
 an_ep :: (Int  -> repr Int)
@@ -720,26 +744,6 @@ sinterp prog =
        (lift (\be te ee -> if_ be te ee)))
        (lift (\f -> lam f)))
        (lift (\f -> fix f))
-{-
-This can't possibly have the type as above!
-
-Perhaps the issue is that the type of the constructors in
-Symantics and the types in the self-interpreter should be
-different?  [Related, but not identical]
-interp prog = 
[…]

### 2007-03-19 matrix/incope.hs
fixing the test3

@@ -606,8 +606,7 @@ test_inc3ee_p = compP $ twice_inc_3_ee ()
 
 -- test_inc3e_c and test_inc3ee_c are most illustrative.
 
-{- Try to implement a term that is a lambda term rather than a ground
-   term -- and fail!
+-- Try to implement a term that is a lambda term rather than a ground term
 test3_e () =
     -- The evaluation context in the following four lines is a self-interpreter
     -- of the object language, encoded in the metalanguage.
@@ -618,17 +617,13 @@ test3_e () =
     -- The term in the following three lines is the object term
     -- let test3_ = \x -> (x 1) + 2
     -- encoded in the object language then encoded in the metalanguage.
-    -- let test3_ = app lam_ (lam (\x ->
-    --              app app_ (
-    --                 app (app add_ (app app_ x (app int_ (int 1)))) 
-    --              (app int_ (int 2)))))
-    let test3_ = app lam_ (lam (\x -> app (app app_ x) (app int_ (int 1))))
-    in test3_ 
+    app lam_ (lam (\x -> (app 
+			   (app add_
+			    (app (app app_ x) (app int_ (int 1))))
+			   (app int_ (int 2)))))
 test_t3_r = compR $ test3_e ()
 test_t3_c = compC $ test3_e ()
 test_t3_p = compP $ test3_e ()
--- ??
--}
 
 -- start encoding some of Ken's ideas on a self-interpreter
 an_ep :: (Int  -> repr Int)

### 2007-03-20 matrix/incope.hs
add another test case, just to get used to the whole process.  I am now happy enough with this

@@ -606,7 +606,8 @@ test_inc3ee_p = compP $ twice_inc_3_ee ()
 
 -- test_inc3e_c and test_inc3ee_c are most illustrative.
 
--- Try to implement a term that is a lambda term rather than a ground term
+-- Implement a term that is a lambda term rather than a ground term
+-- this is test3 above.
 test3_e () =
     -- The evaluation context in the following four lines is a self-interpreter
     -- of the object language, encoded in the metalanguage.
@@ -625,7 +626,51 @@ test_t3_r = compR $ test3_e ()
 test_t3_c = compC $ test3_e ()
 test_t3_p = compP $ test3_e ()
 
+{- And now something harder: powfix
+testpowfix () = lam (\x ->
+                      fix (\self -> lam (\n ->
+                        if_ (leq n (int 0)) (int 1)
+                            (mul x (app self (add n (int (-1))))))))
+at the same time, do
+testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
+-}
+testpowfix_e () =
+    -- The evaluation context in the following lines is a self-interpreter
+    -- of the object language, encoded in the metalanguage.
+    let lam_ = lam (\f -> f) in
+    let app_ = lam (\f -> lam (\x -> app f x)) in
+    let add_ = lam (\m -> lam (\n -> add m n)) in
+    let mul_ = lam (\m -> lam (\n -> mul m n)) in
+    let leq_ = lam (\m -> lam (\n -> leq m n)) in
+    let if__ = lam (\be -> lam (\ee -> lam (\te -> if_ be ee te))) in
+    let int_ = lam (\i -> i) in
+    -- careful, trying to print something with fix_ can loop!
+    let fix_ = lam (\f -> app f (app fix_ f)) in
+    -- The term in the following three lines is the object term
+    -- testpowfix above
+    -- encoded in the object language then encoded in the metalanguage.
+    -- That is seriously not pretty, but it works!
+    let tpf = app lam_ (lam (\x ->
+          app fix_ (lam (\self -> app lam_ (lam (\n ->
+             (app (app (app if__ (app (app leq_ n) (app int_ (int 0))))
+                (app int_ (int 1)))
+                (app (app mul_ x) (app (app app_ self)
+                     (app (app add_ n) (app int_ (int (-1))))))))))))) in  
[…]

### 2007-03-22 matrix/tagless-final.tex
a bit more write-up

@@
 
-this time with HOAS rather than deBruin indices. The language is close 
+this time with HOAS rather than deBruijn indices. The language is close 
 to the language of Xi03, but without the polymorphic lift and more 
@@
 How to make encoding flexible: abstract the interpreter
-Haskell: type classes, OCaml: functors (give the signature of
-Symantics without the 'sv part)
-
 We have implemented the above in (Meta)OCaml and Haskell.  In Haskell,
@@
 the source language and its instances give the semantics).
-	class Symantics r where
-	  int :: Int -> r Int
-	  add :: r Int -> r Int -> r Int
-	  lam :: (r a -> r b) -> r (a->b)
-	  app :: r (a->b) -> r a -> r b
-	  fix :: (r a -> r a) -> r a
-	  ...
-So, Haskell infers the type "Symantics r => r Int" for the example term
-encoding above.
+
+\begin{code}
+class Symantics repr where
+    int :: Int -> repr Int                -- int literal
+    bool :: Bool -> repr Bool             -- bool literal
+
+    lam :: (repr a -> repr b) -> repr (a->b)
+    app :: repr (a->b) -> repr a -> repr b
+    fix :: (repr a -> repr a) -> repr a
+
+    add :: repr Int -> repr Int -> repr Int
+    mul :: repr Int -> repr Int -> repr Int
+    if_ :: repr Bool -> repr a -> repr a -> repr a
+    leq :: repr Int -> repr Int -> repr Bool
+\end{code}
+
+The previous test now reads |app (lam (\x -> x)) (bool True)| and the
+inferred type is |Symantics repr => repr Bool|.
 
 encoding of a few terms.
+
+\begin{code}
+testpowfix () = lam (\x ->
+                      fix (\self -> lam (\n ->
+                        if_ (leq n (int 0)) (int 1)
+                            (mul x (app self (add n (int (-1))))))))
+testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
+\end{code}
+
+OCaml: functors (give the signature of
+Symantics without the 'sv part)
+
+\begin{code}
+module type Symantics = sig
+  type ('c,'dv) repr
+  val int  : int  -> ('c,int) repr
+  val bool : bool -> ('c,bool) repr
+  val add  : ('c,int) repr -> ('c,int) repr -> ('c,int) repr
+  ...
+  val if_ : ('c,bool) repr ->
+             (unit -> ('c,'da) repr) ->
+             (unit -> ('c,'da) repr) -> ('c,'da) repr
+
+  val lam : (('c,'da) repr -> ('c,'db) repr) 
+          -> ('c,'da->'db) repr
+  val app : ('c,'da->'db) repr
+    -> ('c,'da) repr -> ('c,'db) repr
+  val fix : (('c,'da->'db) repr -> ('c,'da->'db) repr) 
+            -> ('c,'da->'db) repr
+end
+\end{code}
+
+Some differences are in the additional type parameter |c| (the
+environment classifier, Walid03) to be used later when we get to code
+generation (required by MetaOCaml, part of the type of the
+code). MetaOCaml is a call-by-value language, and thus we have to eta-expand
+the fix-point combinator and encapsulate the branches of the |if_|
+into thunks to prevent their evaluation.
+
+\begin{code}
+module EX(S: Symantics) = struct
+ open S
+ let testpowfix () = lam (fun x ->
+                      fix (fun self -> lam (fun n ->
+                        if_ (leq n (int 0)) (fun () -> int 1)
+                            (fun () -> mul x (app self (add n (int (-1))))))))
+
+ let testpowfix7 = 
+    lam (fun x -> app (app (testpowfix ()) x) (int 7))
+end
+\end{code}
+The unit in |testpowfix| is for the sake of value restrcition (which
+interferes with the environment classifier inference).
+
+Thus the DSL expression has the type of the functor, from |Symantics|
+to the value.
 
@@
 
+Self-interpretation in our framework is
+a bit tricky due to a need for polymorphism.  An encoded term should
+be polymorphic in the interpreter type-constructor "r".  Further, the
+functions "lam", "app", etc. above should be polymorphic in the object
+types "a" and "b".  Thus the object language seems to require rank-2
+polymorphism; and so type annotations will be inevitable.
+
+To neatly bypass the need for rank-2 polymorphism, we use the notion
+of let-polymorphism and a `hole', as follows.  A partial evaluator (or
+a compiler or interpreter) is an evaluation context that (let-)binds
+variables named "app", "lam", etc.  For the object language to
+take advantage of let-bound polymorphism in the metalanguage, it
+is crucial that we encode object "let" to meta "let".  We achieve
+self-interpretation in that the following evaluation context in the
+object language defines an interpreter for encoded object terms.
+\begin{code}
+    let lam = \f -> f in
+    let app = \f x -> f x in
+    let add = \m n -> m + n in
+    let int = \i -> i in
+    [hole]
+\end{code}
+The obvious notion of optimality in this setup is easy to achieve.  For
+example, given the context above, our partial evaluator turns
+\begin{code}
+    app (lam (\x -> add x (int 1))) (int 3)
+\end{code}
+into the number 4, as desired.  We can also encode the self-interpreter
+above as the following evaluation context in the object language.
+\begin{code}
+    let lam = lam (\f -> f) in
+    let app = lam (\f -> lam (\x -> app f x)) in
+    let add = lam (\m -> lam (\n -> add m n)) in
+    let int = lam (\i -> i) in
+    [hole]
+\end{code}
+The "let" and the hole "[hole]" are meta-constructions, which must map to
+themselves in any interpretation.
+
 \section{Related work}

### 2007-03-24 matrix/tagless-final.tex
misc corrections to the text

@@
 
-\title{Title Text}
+\title{Staged typed type-preserving tagless interpreters, \emph{finally}}
 \subtitle{Subtitle Text, if any}
@@
 
+We have programmed out interpreters in OCaml/MetaOCaml (cite
+www.metaocaml.org) and Haskell. The complete code is available 
+% connot give URL in the ICFP paper due to double-blind reviews
+in the supplemental material to the paper. We will use both
+(Meta)OCaml and Haskell for our examples throughout the paper.
+
 \subsection{The Tag Problem}
@@
 
-The first attempt at the interpreter: take the object language term
-such as test1 and give us the value
+The first attempt at the interpreter, the function eval. It takes the
+object language term such as test1 and gives us its value.
+The first argument to |eval| is the environment, initially empty,
+which is the sequence of values associated with free variables in the
+interpreted code.
 \begin{code}
@@
 
-If our meta-language (the language in which we wrote test1
-and that interpreter) were untyped, the above code would be acceptable.
+Suppose our our meta-language (the language in which we wrote test1
+and that interpreter) were untyped. The above code would be acceptable.
 The |L e| line exhibits interpretative overhead (see descr in Walid02
@@
 
-We can then write the interpreter, and it works
+We can then write the interpreter, which is accepted by the typecheker
 \begin{code}
@@
 \end{code}
-whose inferred type is |u list -> exp -> u|,
+and whose inferred type is |u list -> exp -> u|. Now we can evaluate
+our sample |test1| term:
 
@@
 The failure of the latter corresponds to the evaluation of `open'
-terms. One can see that if all our terms are closed, List.hd and
-List.tl above will never be applied to the empty list. The partial
+terms. One can see that if all our terms are closed, |List.hd| and
+|List.tl| above will never be applied to the empty list. The partial
 pattern match in the |A (e1,e2)| line raises an exception when we try to
@@
 accepted only typed terms, then the missing alternatives in the |A|
-rule would not occur. Alas, what is obvious to us is not obvious to
+line would not occur. Alas, what is obvious to us is not obvious to
 the meta-language. There is still the pattern match, and we still have
@@
 statically assure that only closed terms are passed to eval and that
-the Nil checking in List.hd and List.tl can be eliminated as well. The
-latter is even more interesting problem requiring even more advanced
-type systems (cite env classifiers, Nanevski, Rowan Davies).
+the empty list checking in |List.hd| and |List.tl| can be eliminated
+as well. The latter is even more interesting problem requiring even
+more advanced type systems (cite env classifiers, Nanevski, Rowan
+Davies).
 
@@
 language in the meta-language (reminiscent to the initial approach),
-we will be using functions as term constructors (final approach):
+we will be using functions as term constructors (the final approach):
 
@@
 
-That is all of the interpreter. The previous term |test1| now reads
+That is all of the interpreter. Our sample term now reads
 \begin{code}
@@
 \end{code}
-
-To evaluate the term, we merely need to apply that to the empty
-environment
+which is almost the same as before, only written with lower-case
+identifiers. To evaluate the term, we merely need to apply 
+the term to the empty environment
 
@@
 
-Now we see that the result has no tags. The above functions |b|,
+The result now has no tags. The above functions |b|,
 |varZ|, |app| constitute the evaluator. It is patent that the
-interpreter has no tags and no pattern matching. A term |b true| is
+interpreter has no tags and no pattern matching. The term |b true| is
 evaluated to a boolean and a term |lam varZ| is evaluated to a
-function, with no tags attached. The application rule than applies that
+function, with no tags attached. The application rule applies that
 function without any need for pattern-matching. The evaluator has
@@
 that the initial environment should be a tuple rather than () --
-meaning the term requires non-empty initial environment, the term is
+meaning the term requires non-empty initial environment: the term is
 open.
 
-Thus all the errors that previously reported at run-time (open terms,
+Thus all the errors that were previously reported at run-time (open terms,
 applications of non-functions) are now reported at compile
 time. Running of the interpreter assuredly gives no run-time errors
-because its code uses no run-time error-raising constructs.
+because its code uses no operations that could possibly raise a
+run-time error.
 
@@
 is obvious to the compiler that non-termination can only result from
-interpreting fix.  If the source language is strongly normalizing, then
+interpreting |fix|.  If the source language is strongly normalizing, then
 our interpreter is total.
@@
 
+The structure of our paper is as follows. Section XXX ...
+We finally conclude.
+
 \section{Our base language: Tagless interpreter}
 
-Introduce the language, 
-  The language is simply-typed lambda-calculus with fixpoint,
-  integers, booleans and comparison.
+Introduce the language, The language is simply-typed lambda-calculus
+with fixpoint, integers, booleans and comparison, whose BNF grammar is
 
 \begin{code}
-  Lam hoas_fn | App e e | Fix hoas_fn |
-  I Int | B Bool | Add ie1 ie2 | Mul ie1 ie2 | Leq ie1 ie2 |
+  e ::= Lam hoas_fn | App e e | Fix hoas_fn |
+  I int_literal | B bool_literal | Add e1 e2 | Mul e1 e2 | Leq e1 e2 |
   IF b e-then e-else
@@
 
-this time with HOAS rather than deBruijn indices. The language is close 
-to the language of Xi03, but without the polymorphic lift and more 
-constants so we can write better examples, enough to express
-Fibonacci, factorial, and power.
+This time we use the higher-order abstract syntax (cite: but not
+Elliott. Cite Dale Miller!) rather than deBruijn indices.  The
+language is close to the language of Xi03, without the polymorphic
+lift but with more constants so we can write better examples, enough to
+express Fibonacci, factorial, and power.
 
@@
 
-The previous test now reads |app (lam (\x -> x)) (bool True)| and the
-inferred type is |Symantics repr => repr Bool|.
+Our sample code test1 now reads |app (lam (\x -> x)) (bool True)|,
+whose inferred type is |Symantics repr => repr Bool|.
 
@@
 \begin{code}
-testpowfix () = lam (\x ->
-                      fix (\self -> lam (\n ->
+testpowfix () = 
+  lam (\x -> fix (\self -> lam (\n ->
         if_ (leq n (int 0)) (int 1)
             (mul x (app self (add n (int (-1))))))))
-testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
+testpowfix7 () = 
+  lam (\x -> app (app (testpowfix ()) x) (int 7))
 \end{code}
+The dummy unit |()| argument above is to avoid the monomorphism
+restriction and so that the type of |testpowfix| and |testpowfix7|
+would remain polymorphic in |repr|. Instead of supplying the dummy
+argument, we could have given the terms explicit polymorphic
+sugnature. We however prefer if the typechecker of the meta-language
+(Haskell) inferred the types of the object terms for us. We could also
+avoid the |()| argument by switching off the monomophism restriction
+with a compiler flag.
 
-OCaml: functors (give the signature of
-Symantics without the 'sv part)
+OCaml: functors.
 
@@
 
+The power example and our sample term |test1| are encoded as
 \begin{code}
@@
  open S
- let testpowfix () = lam (fun x ->
-                      fix (fun self -> lam (fun n ->
+ let test1 () = app (lam (fun x -> x)) (bool true)
+ let testpowfix () = 
+   lam (fun x ->fix (fun self -> lam (fun n ->
          if_ (leq n (int 0)) (fun () -> int 1)
@@
 \end{code}
-The unit in |testpowfix| is for the sake of value restrcition (which
-interferes with the environment classifier inference).
+The dummy unit argument in |testpowfix| is the artefact of MetaOCaml
+(which we will be using to compile our tests in Sections below). To be
+more precise, the environment qualifier (which appears as the |'c|
+type parameter in the signature of Symantic functions) must remain
+uninstantiated in order for us to `run' the code. That means, the type
+must be polymorhic. The value restriction then dictates 
+the definitions of our object terms must look syntactically like
+values. Alternatively to giving the |()| argument, we could have used
+rank-2 record types of OCaml to maintain the necessary polymorphism.
 
-Thus the DSL expression has the type of the functor, from |Symantics|
-to the value.
+Thus the object expressions (expressions of the embedded DSL) have in
+OCaml the type of the functor, from |Symantics| to the value. This
+essentially the same as |Symantics repr => | constraint in the case of
+the embedding of the DSL in Haskell.
 
@@
 its type-level function solution: incope1.hs
-  Need to thank Martin Sulzmann
 
@@
 
+\subsection*{Acknowledgments}
+We woul like to tank Martin Sulzmann for helpful discussions.
+
 \end{document}

### 2007-03-24 matrix/tagless-final.tex
finished drafting Sec 2

@@
 \usepackage{natbib1}
+\usepackage{comment}
 
@@
   val int  : int  -> ('c,int) repr
-  val bool : bool -> ('c,bool) repr
-  val add  : ('c,int) repr -> ('c,int) repr -> ('c,int) repr
-  ...
+  val bool: bool -> ('c,bool) repr
+  val add : ('c,int) repr-> ('c,int) repr-> ('c,int) repr
+  val mul : ('c,int) repr-> ('c,int) repr-> ('c,int) repr
+  val leq : ('c,int) repr-> ('c,int) repr-> ('c,bool) repr
   val if_ : ('c,bool) repr ->
@@
             -> ('c,'da->'db) repr
-end
+end;;
 \end{code}
@@
 generation (required by MetaOCaml, part of the type of the
-code). MetaOCaml is a call-by-value language, and thus we have to eta-expand
+code). MetaOCaml is a call-by-value language, and thus we have to 
+$\eta$-expand
 the fix-point combinator and encapsulate the branches of the |if_|
@@
          if_ (leq n (int 0)) (fun () -> int 1)
-             (fun () -> mul x (app self (add n (int (-1))))))))
-
+             (fun () -> mul x (app self 
+                                   (add n (int (-1))))))))
  let testpowfix7 = 
     lam (fun x -> app (app (testpowfix ()) x) (int 7))
-end
+end;;
 \end{code}
@@
 the embedding of the DSL in Haskell.
+
+The first instatiation of the Symantics structure is the interpreter:
+evaluating the object (DSL) term to its value in the meta-language. 
+The following module essentially interprets (gives one-to-one
+correspondence of) each operation of the object language as the
+corresponding operation in the meta-language.
+\begin{code}
+module R  = struct
+  type ('c,'dv) repr = 'dv    (* absolutely no wrappers *)
+  let int (x:int) = x
+  let bool (b:bool) = b
+  let add e1 e2 = e1 + e2
+  let mul e1 e2 = e1 * e2
+  let leq x y = x <= y
+  let if_ eb et ee = if eb then (et ()) else (ee ())
+
+  let lam f = f
+  let app e1 e2 = e1 e2
+  let fix f = let rec self n = f self n in self
+end;;
+\end{code}
+
+This interpreter is patently tagless: the operation |add| is really
+OCaml's addition, and |app| is OCaml's application. We can run our
+examples by instatiating the |EX| functor with |R|.
+\oleg{Jacques: you mentioned "Boxes go Bananas" by Washburn and
+  Weirich. Do you want to comment more about that paper here?}
+
+\begin{code}
+  module EXR = EX(R);;
+\end{code}
+
+and so |EXR.test1 ()| evaluates to untagged boolean value |true|.
+In Haskell, we define a |newtype R a = R a| and 
+|instance Symantics R|
+with the straightforward implementation. We must stress that |R| is
+a |newtype|. Although it may look like a tag, it has no run-time
+representation (erased during the compilation) and pattern-matching
+against |R| cannot ever fail and so is operationally identity and
+compiled that way.
+
+Proposition: If |e| is the term in the meta-language has the type
+|t|, evaluation of that term in the interpreter R will either continue
+idenfinetely or termninate with a value of the same type |t|.
+
+We'd like show a different interpreter: it interprets each terms of
+the object language to its size (defined as the number of term
+constructors). The following is a slightly abbreviated code (see the
+accompanying source code for the complete definition)
+
+\begin{code}
+module L = struct
+  type ('c,'dv) repr = int
+  let int (x:int) = 1
+  let mul e1 e2 = e1 + e2 + 1
+  let if_ eb et ee = eb + et () + ee () + 1
+
+  let lam f = (f 0) + 1
+  let app e1 e2 = e1 + e2 + 1
+  let fix f = (f 0) + 1
+end;;
+\end{code}
+
+Now the expression
+\begin{code}
+  let module E = EX(L) in E.test1 ()
+\end{code}
+evaluates to |3|. This interpreter is not only tagless, it is also
+total. It `evaluates' even seemingly divergent terms like\\
+|app (fix (fun self -> self)) (int 1)|.
+
+\begin{comment}
+module EX1(S: Symantics) = struct
+ open S
+ let tfix () = app (fix (fun self -> self)) (int 1)
+end;;
+let module E =EX1(R) in E.tfix ();;
+let module E =EX1(L) in E.tfix ();;
+\end{comment}
 

### 2007-03-25 matrix/tagless-final.tex
more content inserted

@@
 \begin{abstract}
-This is the text of the abstract.
+We built a tagless partial evaluator for a typed object language,
+placing minimal requirements on the type system of the metalanguage:
+we need let-bound polymorphism and staging, but not dependent
+types, generalized algebraic data types, or a tag elimination as a
+postprocessing step.
+The trick is the "final" approach: we encode the
+higher-order abstract syntax of object terms not using constructors for
+an algebraic data type but by invoking cogen functions, which occur as
+free variables in the encoded term.
 \end{abstract}
@@
 
+\oleg{should say somewhere in the Introduction that we are interested
+  in the typed subject language embedded in the typed
+  meta-language. Refer to Sec 2 (see the remark below) why this is
+  good. We assume the DSL setting: we enter EDSL terms in the
+  meta-language program (e.g., GHCi or (Meta)OCaml top-level). A
+  different approach is reading EDSL terms from a file. In that case, we
+  need to typecheck them first. We do not deal with that problem here,
+  because it has been solved. First we mention Walid02 where they use
+  dependent typed. However, these don't seem to be needed: refer to
+  `typing dependent typing', and the related work section.}
+
+\jacques{I quite agree that the point that needs to be revisited is
+  that a \emph{direct} interpreter for a (typed) language "cannot" be
+  done. see discussion in related work about the direct interpreter,
+  with regards to Walid's paper. Perhaps bring some of it here?}
+\jacques{I have looked at a fair bit of the literature, and I have not
+  seen this done, especially not in the typed 'online' setting.  The
+  one question that seems to be open is, how accurate can we make the
+  PE without a tough BTA?}
+Should we mention some of these points in our contributions subsection
+below?
+
+        Taha et al and Xi et al (both POPL2003) argued that writing an
+interpreter of a typed language in a typed language is an interesting
+problem. The known solutions include the use of the Universal type, or
+GADTs. The former is not quite satisfactory for performance reasons;
+and also because of the presence of partial pattern matching in the
+interpreter; the inexhaustiveness of the pattern matching should never
+be triggered when executing a well-typed term. Alas, this fact cannot
+be made obvious to the compiler. That also exlains unoptimality.
+
 \subsection{The Tag Problem}
@@
 literals with deBruin indices as variables. This is the language used
-in the body of Walid02.
+in the body of Walid02, from which we borrow the first part of 
+this example.
 
@@
 \subsection{Contributions}
+
+        We have realized that if we change the encoding of the term --
+from data constructors to functions, we can encode the tagless
+interpreter of the type language in a typed language with no GADTs and
+a very simple type system. The interpreter uses no universal type, no
+pattern matching and is patently obviopus to the compiler that the
+only non-termination can come from the interpreting fix. If the source
+language is strongly normalizing, our interpreter is total.
+
 The term `constructor' functions lam, app, etc. look like free
@@
 with a compiler flag.
+
+Note that the subject language (DSL) is typed. If we make a mistake
+(for example, replace |int 7| with |bool True| in |testpowfix7|, the
+compiler will highlight that term and complain that it couldn't match
+the expected |Int| against the inferred |Bool|). The type error
+also results when entering |lam (\x -> app x x)|, with the compiuler
+complaining about the infinite type. The type-checker of the
+meta-language not only flags meaningless DSL (subject level) terms --
+before any of their evaluation. If we forget, for example, |app| in
+some of the terms above, we again get an error message with a precise
+location. So the type-checker also flags syntactically invalid
+subject-level terms.\oleg{Should we perhaps move this up, in the
+  intro? To justify why typed-metalanguages and the typed DSL are
+  good. Or keep the explanation here but refer to it from the Introduction.}
 
@@
 
-the sv type problem: the inductive type function 'dv -> 'sv
-its GADT solution. 
-its type-level function solution: incope1.hs
-
-How can we do without GADT
-Now show the Symantics signature in full, with the 'sv types
-
-It's relation to PE and CPS. Currently, our CPS is CBN. Need to
-investigate CBV.
-
 Surprisingly, it is also possible to write a partial evaluator!  The
@@
 the compiler.
+
+Compare the output of some of the ctest with the corresponding
+ptest. Use power?
+
+One should note that the partial evaluator has fully reused the code
+of the compiler. It is truly the `composition' of the interpreter and
+the compiler.
+
+the sv type problem: the inductive type function 'dv -> 'sv
+its GADT solution. 
+its type-level function solution: incope1.hs
+
+We need GADT -- or, to be more precise,
+representation that varies with the type. We need a value of the type
+|P Int| to have a different representation than the value of the type 
+|P (Int->Int)|. GADT give you that (although we don't use the GADT in
+their intended meaning for that).
+
+Drawbacks of the GADT solution. GADTs, although work, are still not
+quite satsifactory: let us look at the following lines in incope.hs
+\begin{code}
+    app ef ea = case ef of
+                        VF f -> f ea
+                        E  f -> E (app f (abstr ea))
+\end{code}
+The datatype has four constructors, |VI|, |VB|, |VF| and |E|. How come we used
+only two of them in the above expression. It is obvious, one may say,
+that the constructors |VI| and |VB| just can't occur, because of the
+typing consideration. This is obvious to us -- but it is not obvious
+to the compiler. The case matching above does look \emph{partial}. That
+seems like a tenuous point -- but it is precisely the main point of
+the whole tag elimination approach. In a typed tagged interpreter
+there are lots of pattern-matching that looks partial but in reality
+total. The ultimate goal is to make this totality syntactically
+apparent.
+
+How can we do without GADT
+Now show the Symantics signature in full, with the 'sv types
+
+In the ideal world, the MetaOCaml signature (or Haskell type class)
+Symantics would start with
+\begin{code}
+    kind tp = int | bool | (->) tp tp
+    type ('c, 'dv:tp) repr : tp
+    val int : int -> ('c, int) repr
+    val bool : bool -> ('c, bool) repr
+    val lam : (('c, 'da) repr -> ('c, 'db) repr) -> ('c, 'da -> 'db) repr
+    val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
+\end{code}
+
+Here "repr" is a tp-to-tp function, defined inductively in the case of
+partial evaluation:
+
+\begin{code}
+    ('c, int     ) repr_pe = VI of int
+                           | E of ('c, int) code
+    ('c, bool    ) repr_pe = VB of bool
+                           | E of ('c, bool) code
+    ('c, 'a -> 'b) repr_pe = VF of ('c, 'a) repr -> ('c, 'b) repr
+                           | E of ('c, 'a -> 'b) code
+\end{code}
+
+Unfortunately, we can't define type-level functions inductively.
+Instead we bake |repr_pe| above into the signature Symantics, so repr
+takes three type arguments |('c,'sv,'dv)|, where the extra argument 
+|'sv| is equal to |('c,'dv) repr_pe.|
+It seems Ken has just showed how to define type-level functions
+inductively... This shows also how a limited form of GADTs (the ones
+used for the sake of type-dependent representations) can be
+implemented with just functions.
+
+So, the full signature of Symantics looks like (cf Section 2)
+
+\begin{code}
+module type Symantics = sig
+  type ('c,'sv,'dv) repr
+  val int  : int  -> ('c,int,int) repr
+  val bool : bool -> ('c,bool,bool) repr
+  val add  : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val mul  : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val leq  : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
+  val if_ : ('c,bool,bool) repr ->
+             (unit -> ('c,'sa,'da) repr) ->
+             (unit -> ('c,'sa,'da) repr) -> ('c,'sa,'da) repr
+
+  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
+    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),'da->'db) repr
+  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),'da->'db) repr
+    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
+  val fix : (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,'da->'db) repr 
+             -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
+end
+\end{code}
+
+Also, unlike the GADT
+solution, there is no longer any seemingly partial
+pattern-matching. The pattern-match only occurs in PE (to check if the
+code is static or dynamic), it is only a phase rather than a type tag
+match, and the pattern match is \emph{patently} exhaustive. That is,
+totality is assured and it is assured \emph{syntactically}, in the way that
+is apparent to the compiler.
+
+It's relation to PE and CPS. Currently, our CPS is CBN. Need to
+investigate CBV.
+The type-level function |'da -> 'sa| is important because it gives 
+the CPS transform. So, the type-function
+that enables the PE is exactly the same type function that enables
+CPS....
+
+Say that when doing PE on fix, we made a deliberate decision to `go
+all the way' -- so long as the computation can proceed at the PE stage
+(`static value'), we do that rather than residualizing. We have also
+an alternative where fix is unrolled only once and them we
+residualize. We have used the first approach, even it (unlike the
+latter) may cause non-termination at the PE stage -- but \emph{only} if the
+computation we are partially evaluating is non-terminating and enough
+static information is available to trigger this non-termination.
+Say that our treating of fix may lead to the code bloat in the
+residual program. That is an issue investigated elsewhere (cite our
+PEPM2006 paper).
 
@@
 
+The PE in incope is really a mixture of online and off-line; it looks 
+online, but it has a sort of "just in time" binding-time analysis built in.
+
+If we just add a method 'lift' to Symantics with a type
+|lift :: a -> repr a| (cf polymorphic lift of Xi03, of MetaML and
+MetaOCaml: CSP)
+then |Symantics repr| implies |Functor repr|. We can write then
+\begin{code}
+  instance Semantics repr => Functor repr where
+  -- fmap :: (a->b) -> repr a -> repr b
+  fmap f ra = app (lift f) ra
+\end{code}
+
 \subsection{Translucent types}
@@
 term is then well-typed in any interpreter.  Each instance of Symantics
-instantiates "r" to interpret terms in a particular way.
+instantiates "r" to interpret terms in a particular way. The L
+interpreter is quite illustrative. We need the above semantics to be
+able to represent both R and L (the latter returns only Int as the
+values) in the same framework.
+
+We encode a term like |add 1 2| as
+|app _add (app _int 1) (app _int 2)| where |_add| and |_int| are just
+`free variables'. Now, how to typecheck such a term? Some type should
+be assigned to these free variables. The goal is to complete the work
+without needing any type annotations (so we don't have to introduce any
+type language), with all types inferred and all terms typed. It seems
+the second-order type R neatly separates the typechecking part from
+the representation of R: it hides aspects that depend on the
+particular interpreter, and yet lets enough type information through
+(via its type argument) to permit the typechecking of terms, and infer
+all the types.
+
+The only approach that does seem to work
+is the one in incope.hs or incope.ml. If we de-sugar away records and
+type-classes, the type of a term L of the inferred type tau is
+
+\begin{code}
+  (Int -> r Int) ->
+  (Bool -> r Bool) ->
+  (forall alpha beta. (r alpha -> r beta) -> r (alpha->beta)) -> ... r tau
+\end{code}
+
+or, if we denote the sequence of initial arguments as S r, terms have
+the type |S r -> r tau|
+The interpreter has the type
+|(forall r. S r -> r tau) -> r' tau|
+
+The higher-order type (variable) of kind |*->*| seems essential. So, at
+least we need some fragment of Fw (somehow our OCaml code manages to
+avoid the full Fw; probably because the module language is separated
+from the term language). Thus, we seem to need a fragment of Fw. It
+seems the inference is possible, as our Haskell and OCaml code
+constructively illustrates. Perhaps we need to characterize our
+fragment.
 
 \section{Self-interpretation}
+
+Taha et al define a self-interpreter si to be an object-language term
+such that, for any object-language term e,
+
+\begin{code}
+  si "e"    is observationally equivalent to    e
+\end{code}
+
+where |"e"| is an encoding of |e|.  Under the co-approach, we want to encode
+each object-language term as a function from a record of cogen methods
+to a generated rep.  (Thanks to currying, the object language need not
+actually support records.)  In the case of a self-interpreter, that
+record is simply si!  So we should define a self-interpreter si to be an
+object-language record-term such that, for any object-language term e,
+
+\begin{code}
+  "e" si    is observationally equivalent to    e
+\end{code}
+
+The self-interpreter then is just like the interpreter
+|R| in Section 2, except written at the object rather than meta level.
+
+Note on terms and special forms. In our class Symantics, |add|
+is not a term -- because it does not have the type repr t. We could
+have introduced |add| to be of the type |repr (Int->Int->Int)| -- but we
+didn't. So, |add| is a special form. It is to be applied in
+meta-language (using the Haskell application -- white space) rather
+than in the object language (app).
+
+Mention the RR interpreter: for any term E, (RR E) is equivalent to E.
+RR is not an identity: it is an interpreter that encapsulates another
+interpreter. The RR interpreter can be made self if we treat RR as a
+special form and interpret it always with itself (similar to let and
+hole below).
 
@@
 take advantage of let-bound polymorphism in the metalanguage, it
-is crucial that we encode object "let" to meta "let".  We achieve
+is crucial that we encode object "let" to meta "let".
+
+We also (unfortunately need to) define a self-interpreter slightly
+differently.  A self-interpreter |SI[]| is an object expression with a
+hole |[]| waiting to be plugged with the encoding of an object expression.
+The self-interpreter binds the free variables |_lam|, |_app|, |_int|, |_bool|,
+etc. in the plugged encoding.  We require |SI["E"]| to be equivalent to |E|,
+where |"E"| is the object encoding of the object expression |E|.
+
+This strategy of course limits what kinds of processing on encoded
+object expressions can be expressed in the object language.  For
+example, unlike if we had simply defined |"E"| to be the `identity
+encoding' E, we \emph{can} write a size-measurer |SZ[]|, which like |SI[]| is
+an object expression with a hole |[]| waiting to be plugged with |"E"|.
+However, this notion of size cannot distinguish |let x = E1 in E2[x]|
+from |"E2[E1]"|.  Cf. one way to type-check polymorphic let:
+\begin{code}
+        E1:T    E2[E1]:T'
+    ------------------------
+    let x = E1 in E2[x] : T'
+\end{code}
+That analogy shows that the self-interpreter with a hole is a
+euphemism for a higher-rank abstraction. That makes it clear where
+the difficulty lies, and how to simulate the higher-rank application
+(that is, higher-rank arrow elimination) `by hand'.
+
+We also need to duplicate the encoding of an object expression if we
+want to process the same object expression in multiple ways (e.g., both
+measure its size and interpret it).  But all this is consistent with (in
+fact, amplifies) our `coalgebraic/final/cogen' spirit, so our revised
+notion of self-interpretation (and, some might say, of size measurement)
+may just squeak by the reader's skepticism as we explain our desire for
+kind polymorphism.
+
+I guess the let-and-hole argument works out if we don't typecheck the
+encoded terms. We can't typecheck the interpreter with the whole
+either. We insert the encoded term into the interpreter -- and then
+typecheck the whole thing. That'll work. Yet a bit unsatisfactory...
+
+We achieve
 self-interpretation in that the following evaluation context in the
@@
 
+The 'result' of applying an interpreter to itself, needs to be an 
+interpreter too!  In fact, if it is Jones-optimal, then you should be 
+able to apply it to itself again and get the same answer.  
+Our code does that. First, |twice_inc_3_e| is the
+metalanguage encoding of the object term |si["e"]|
+where si is the self-interpreter of the object language (actually, only
+lam,app,add,int), |""| means the object-to-object encoding, and e is the
+object term
+\begin{code}
+  let twice_ f x = f (f x) in let inc_ n = n + 1 in twice_ inc_ 3
+\end{code}
+
+Second, |twice_inc_3_ee| is the metalanguage encoding of the object term
+|si["si[e]"]|
+which is the same as
+|si["si"["e"]]|
+The partial evaluator yields 5 on both |twice_inc_3_e| and 
+|twice_inc_3_ee|,
+as desired.
+
+``But how 
+to you create, in either Haskell or MetaOCaml, an untypechecked 
+interpreter-with-a-hole [UIH] ?''
+Well, one can make an argument that we already have such an
+interpreter with polymorphic let and the hole: incope. In Haskell, the
+declaration of an instance of Symantics is like the sequence of
+polymorphic lets. We construct terms where lam, add, etc, are free
+variables. We apply the interpreter to the semantics (plug the hole)
+by instantiating these terms (binding the free variables lam, etc. to
+the particular instance of Symantics). The unRR construction does
+this plugging in explicitly.
+Again, what is different from the above is that we \emph{can}
+typecheck terms separately, without inserting them first within the
+hole of a particular interpreter. Rank-2 type of |repr| helps. It lets
+enough of the type information out so the typechecking can
+proceed. So, |repr| is the representation of the polymorphic
+interpreter context with the hole, which permits separately
+typecheckable terms (the evaluation still entails `duplication' so to
+speak -- which is one way how polymorphism is resolved).
+
 \section{Related work}
@@
 described what he achieved then and what he \emph{did not}. He did not
-eliminate all the tags.
+eliminate all the tags. Ken's and mine impression of it:
+(i) we can't write a \emph{direct} interpreter of the type language
+           -- the only known to them, and \emph{partial} solution, was 
+              that by Zhe Yang;
+(ii) we have to use indirect interpreter, which needs the universal
+           type and hence tagging;
+(iii) that means any self-interpreter must have tags
+(iv) that means not Jones-optimal.
+It seems we should attack the point1 (in which case we don't have to
+bother with self-interpreters). We say that we \emph{can} write the direct
+interpreter, and without GADT or their equivalents (typeclasses).
+Meta-language computes types, and the compiler and interpreter respect
+those types and needs no tags.
 
@@
 Danvy's self-interpreter
+Ken has read Danvy et al's paper and said that the paper
+doesn't seem to even make the claim that the source language is
+typed. The interpreter is typed, but the source language doesn't seem
+to be.
 In sharp contrast to Danvy's paper, all the evaluators in our case
@@
 
-Asai; Simii et al.
+Asai; 
+It [dynamic part when dealing with bool, int, lam in PE] is
+Asai's trick for reducing the amount of reify/reflect back-and-forth
+in a PE for an untyped language that allows you back-and-forth.  Or at
+least that's how I read those papers.
+Asai's approach is related to our PE (emulating typelevel functions):
+Ken: ``In some sense we're just observing that Asai's code type-checks
+-- in Hindley-Milner as soon as we deforest the static code
+representation.''
 
-"Boxes go Bananas" by Washburn and Weirich (intended or otherwise)
+ Simii et al.
+
+"Boxes go Bananas" by Washburn and Weirich (intended or otherwise) Ken
+wrote about this: ``but a glance reveals no type classes.  Do you mean
+then that if we hide the definition |newtype R a = R a| from a client,
+and export only the interpreter |instance Symantics R R|, then the
+client would see R as an HOAS datatype a la Washburn and Weirich and
+only be able to create parametric ("non-exotic") terms?''  Jacques
+commented: ``I mean the HOAS parts, and how to deal with it (like in
+the instance for |Symantics R R|).  There are other nice tricks for
+dealing with HOAS in that paper worth a closer look.  It's all about
+plain polymorphism, not type classes, but there are still some clever
+ideas there, and they are about issues of tagging (an lack thereof).''
 
@@
 from untyped to the typed form, ready to be interpreted. Walid's ICFP02
-uses dept types
+uses dependent types
+
+Regarding our implementation of the type-level function |'sv -> 'dv|.
+Gibbons et al paper on
+the implementations of typecase may be relevant. I have not read it,
+but I have the feeling some of their typecase designs may be
+relevant (at least to cite). 
+However, the typecase paper relies on having a 
+bijection between types -- our |'sv| and |'dv| in this case, which is not 
+our situation.  We have a one-sided view of things, where |'sv -> 'dv| is 
+easy, and |'dv| "perfectly reflects" an |'sv|, but you still can't get at 
+that |'sv|.
+
+Self-Interpretation and Reflection in a Statically Typed Language 
+\url{http://webpages.cs.luc.edu/~laufer/papers/reflection93.pdf}
+He uses the SK language and some typeclasses. Their interpreter is a
+meta-circular interpreter -- rather than self-interpreter. We can
+claim to have implemented a far better interpreter for a far richer
+language.
 

### 2007-03-25 matrix/tagless-final.tex
partial editing pass on paper

@@
 
+\newtheorem{prop}{Proposition}
+
 \begin{document}
@@
 
-\title{Staged typed type-preserving tagless interpreters, \emph{finally}}
+\title{Staged typed type-preserving tagless ``interpreters'', \emph{finally}}
 \subtitle{Subtitle Text, if any}
@@
 \begin{abstract}
-We built a tagless partial evaluator for a typed object language,
-placing minimal requirements on the type system of the metalanguage:
+We have built a (family of) tagless partial evaluator(s) for a typed object
+language, placing minimal requirements on the type system of the metalanguage:
 we need let-bound polymorphism and staging, but not dependent
-types, generalized algebraic data types, or a tag elimination as a
+types, generalized algebraic data types, nor a tag elimination 
 postprocessing step.
-The trick is the "final" approach: we encode the
-higher-order abstract syntax of object terms not using constructors for
-an algebraic data type but by invoking cogen functions, which occur as
-free variables in the encoded term.
+The main tool is to take functions seriously: we encode the
+higher-order abstract syntax of object terms not by using constructors for
+an algebraic data type but rather by invoking cogen functions, which occur as
+free variables in the encoded term.  In other words, we eschew initial
+algebras for the wilder world of implicit representations of final algebras.
+Our family of evaluators contains an interpreter, a compiler, a partial
+evaluator and a term-size evaluator.
 \end{abstract}
@@
 
-\section{Introduction}
+\section{Introduction}\label{intro}
 
@@
 www.metaocaml.org) and Haskell. The complete code is available 
+\footnote{the URL will appear in the final version of the paper.}
 % connot give URL in the ICFP paper due to double-blind reviews
@@
 (Meta)OCaml and Haskell for our examples throughout the paper.
+Each language has its strength and weaknesses; this dual implementation
+forced us to clarify our own ideas.  We will continually leverage the
+respective strengths of these languages, as certain of our claims are
+patently obvious in one implementation, while requiring more argumentation
+in the other.
 
@@
   `typing dependent typing', and the related work section.}
-
+\jacques{I think it is important to emphasize, probably later in the 
+paper, that we could write a camlp4 extension (or simply a parser in
+Haskell) that would transform a nice syntax for a lambda calculus into
+syntactic terms for what we present here.  Even the self-interpreter
+terms could be done this way, which would simplify things tremendously.}
 \jacques{I quite agree that the point that needs to be revisited is
@@
 
-\subsection{The Tag Problem}
+\subsection{The Tag Problem}\label{tagproblem}
 
@@
 Consider the sample language of abstraction, application and boolean
-literals with deBruin indices as variables. This is the language used
+literals with de Bruin indices as variables. This is the language used
 in the body of Walid02, from which we borrow the first part of 
@@
 
-Suppose our our meta-language (the language in which we wrote test1
+Suppose our meta-language (the language in which we wrote test1
 and that interpreter) were untyped. The above code would be acceptable.
@@
 
-If we use OCaml or other typed language as the meta-language, 
+If we use OCaml or some other typed language as the meta-language, 
 the function |eval0| is ill-typed. The line |B b|
@@
 
-The solution is to introduce the Universal type (see Walid02 Section
-1.3)
+The solution is to introduce a Universal type (see Walid02 Section
+1.3) which contains both of these:
 
@@
 line would not occur. Alas, what is obvious to us is not obvious to
-the meta-language. There is still the pattern match, and we still have
+the meta-language. There is still a pattern match, and we still have
 to attach and eliminate the UB and UA tags, even when there is no
@@
 Of course there is also the issue how to design a type system to type
-terms like test1 (GADT seem necessary). Also, how to design the
+terms like |test1| (GADT seem necessary). Also, how to design the
 typesystem to keep track of open/closed status of terms, so we can
@@
 more advanced type systems (cite env classifiers, Nanevski, Rowan
-Davies).
+Davies, Pitts).
 
@@
 
-In short: (put into really introduction or the conclusions?)
+In short: \oleg{ (put into really introduction or the conclusions?) }
 It is not satisfactory
@@
 
-\subsection{Other solutions}
-The tag problem motivated GADTs (SPJ, Xi) and dependent types
-(WalidICFP02, Concoqtion). Or, perhaps just refer to related work?
-But we need to say something here, to give the reader the feeling of
-how advanced solutions were proposed and how rarely they are implemented.
-
-\subsection{The `Final' Proposal}
-We have realized that, if we change the encoding of the term from data
-constructors to functions, then we can encode a tagless interpreter of a
-typed object language in a metalanguage with a very simple type system.
+% \subsection{Other solutions}
+% The tag problem motivated GADTs (SPJ, Xi) and dependent types
+% (WalidICFP02, Concoqtion). Or, perhaps just refer to related work?
+% But we need to say something here, to give the reader the feeling of
+% how advanced solutions were proposed and how rarely they are implemented.
+
+See the Related Work section (\ref{related}) for a discussion of the other
+solutions to this problem.
+
+\subsection{The `Final' Proposal}\label{ourapproach}
+
+We have realized that if we change the encoding of the term from data
+constructors to functions, we can encode a tagless interpreter for a
+typed object language into a metalanguage with a very simple type system.
 
@@
 accordingly. Let the paper and the accompanying code be in sync.}
+
+\jacques{do we want to start using a lot of little files and
+use includes to insure the code stays in synch?}
 
@@
 another feature: evaluation of open terms leads to a type error rather
-than a run-time error. Indeed, the following gives a type error
+than a run-time error. Indeed,
 
@@
 \end{code}
+gives a type error saying 
 that the initial environment should be a tuple rather than () --
-meaning the term requires non-empty initial environment: the term is
+meaning the term requires a non-empty initial environment: the term is
 open.
@@
 applications of non-functions) are now reported at compile
-time. Running of the interpreter assuredly gives no run-time errors
-because its code uses no operations that could possibly raise a
-run-time error.
-
-The interpreter uses no Universal type or pattern matching, and it
+time. Running of the interpreter will never result in run-time errors
+because its code uses no operations that could possibly raise such
+an error.
+
+Also, note that 
+the interpreter uses no Universal type nor pattern matching, and it
 is obvious to the compiler that non-termination can only result from
@@
 our interpreter is total.
-
+\jacques{the language above has no fix (yet), so we should move this 
+comment somewhere else}
+
+\oleg{
 We should emphasize that not only our code can't have pattern-match
@@
 In our code, there is simply no pattern-mismatch opportunity: the
-proof of no bad match is expressed in the code itself.
-
-In the above code, the interpreter is wired in, built into the
-functions |b|, |app|, etc. We explain how to abstract the interpreter
-below, to make it possible to `evaluate' the same term in many different
-ways (e.g., interpret term, compile it, compute its size or depth,
-etc.)
-
-\subsection{Contributions}
+proof of no bad match is expressed in the code itself. }
+
+In the above code, the interpreter is wired in, directly built into the
+functions |b|, |app|, etc.  In the remainder of this paper, 
+we explain how to abstract the interpreter,
+to make it possible to `evaluate' the same term in many different
+ways: interpretation, compilation, compute the size (or depth), etc.
+
+\subsection{Contributions}\label{contributions}
 
         We have realized that if we change the encoding of the term --
-from data constructors to functions, we can encode the tagless
-interpreter of the type language in a typed language with no GADTs and
+from data constructors to functions, we can encode a tagless
+interpreter of a typed language in a typed language with no GADTs and
 a very simple type system. The interpreter uses no universal type, no
-pattern matching and is patently obviopus to the compiler that the
-only non-termination can come from the interpreting fix. If the source
+pattern matching and it is patently obvious to the compiler that the
+only non-termination can come from interpreting |fix|. If the source
 language is strongly normalizing, our interpreter is total.
 
-The term `constructor' functions lam, app, etc. look like free
+The term ``constructor'' functions lam, app, etc. look like free
 variables in the encoding of an object term.  Defining these functions
@@
 	- `compile' the term. This requires staging, as in MetaOCaml.
-	- Surprisingly, it is also possible to write a partial
-	evaluator!
+	- Surprisingly, it is also possible to write a partial evaluator!
 
@@
 What are our contributions:
-0) we interpret/compile/pe the typed source language
-   [Ken says that Danvy's source language is untyped]
-1) a much clearer implementation
-2) a nicely comparable Haskell/MetaOCaml version
-3) a functorial Symantics that deals with interp/comp/pe uniformly
-4) clear and easy extensibility to more features
-- only closed terms can be evaluated
+\begin{enumerate}
+\item we interpret/compile/partially-evaluate a typed source language
+   from within a typed language.  The requirements of the host type
+   system are \jacques{not so onerous?  simple?},
+\item a simple, clean implementation,
+\item a nicely comparable Haskell/MetaOCaml version,
+\item a functorial Symantics that deals with uniformly across all
+evaluators, from interpretation to partial evaluation,
+\item clear and easy extensibility to more features,
+\item only closed terms can be evaluated.
+\end{enumerate}
+
 Making sure that we interpret terms that are well-typed and closed --
@@
 extensions (GADT, Nanevski, Concoqtion).
+\jacques{can we state fairly clearly what type system we need?  I mean,
+Haskell98(?) and MetaOCaml with (objects, polymorphic variants, ...).
+With our let-polymorphism trick for self-interpretation, what do we really
+need?}
 
@@
 
-\section{Our base language: Tagless interpreter}
+\section{Our base language: Tagless interpreter}\label{language}
 
@@
 
-This time we use the higher-order abstract syntax (cite: but not
-Elliott. Cite Dale Miller!) rather than deBruijn indices.  The
+This time we use higher-order abstract syntax (cite: but not
+Elliott. Cite Dale Miller!) rather than de Bruijn indices.  The
 language is close to the language of Xi03, without the polymorphic
@@
 
-How to make encoding flexible: abstract the interpreter
+\subsection{How to make encoding flexible: abstract the interpreter}
 We have implemented the above in (Meta)OCaml and Haskell.  In Haskell,
@@
 whose inferred type is |Symantics repr => repr Bool|.
-
-encoding of a few terms.
+For example, we can encode the classical |power| function, as well
+as the partial application |\x -> power x 7| as
 
@@
 restriction and so that the type of |testpowfix| and |testpowfix7|
-would remain polymorphic in |repr|. Instead of supplying the dummy
+will remain polymorphic in |repr|. Instead of supplying the dummy
 argument, we could have given the terms explicit polymorphic
-sugnature. We however prefer if the typechecker of the meta-language
-(Haskell) inferred the types of the object terms for us. We could also
+signature.  We however prefer if the typechecker of the meta-language
+(Haskell) infers the types of the object terms for us. We could also
 avoid the |()| argument by switching off the monomophism restriction
@@
 compiler will highlight that term and complain that it couldn't match
-the expected |Int| against the inferred |Bool|). The type error
-also results when entering |lam (\x -> app x x)|, with the compiuler
-complaining about the infinite type. The type-checker of the
-meta-language not only flags meaningless DSL (subject level) terms --
+the expected |Int| against the inferred |Bool|). A type error
+also results when entering |lam (\x -> app x x)|, with the compiler
+complaining about an infinite type. The type-checker of the
+meta-language will flag meaningless DSL (subject level) terms --
 before any of their evaluation. If we forget, for example, |app| in
-some of the terms above, we again get an error message with a precise
+some of the terms above, we get an error message with a precise
 location. So the type-checker also flags syntactically invalid
@@
 
-OCaml: functors.
+We can do the above in (Meta)OCaml as well, replacing type classes with
+functors, via a module type.
 
@@
              (unit -> ('c,'da) repr) -> ('c,'da) repr 
-
   val lam : (('c,'da) repr -> ('c,'db) repr) 
@@
 \end{code}
-The dummy unit argument in |testpowfix| is the artefact of MetaOCaml
-(which we will be using to compile our tests in Sections below). To be
+The dummy unit argument in |testpowfix| is an artefact of MetaOCaml
+(which we will be using to compile our tests in Sections below),
+again related to monorphism. To be
 more precise, the environment qualifier (which appears as the |'c|
@@
 the definitions of our object terms must look syntactically like
-values. Alternatively to giving the |()| argument, we could have used
+values. Alternatively to giving an |()| argument, we could have used
 rank-2 record types of OCaml to maintain the necessary polymorphism.
@@
 Thus the object expressions (expressions of the embedded DSL) have in
-OCaml the type of the functor, from |Symantics| to the value. This
-essentially the same as |Symantics repr => | constraint in the case of
+OCaml the type of a functor from |Symantics| to the desired value. This
+is essentially the same as the |Symantics repr => | constraint in the case of
 the embedding of the DSL in Haskell.
 
-The first instatiation of the Symantics structure is the interpreter:
+The first instantiation of the Symantics structure is an interpreter:
 evaluating the object (DSL) term to its value in the meta-language. 
 The following module essentially interprets (gives one-to-one
-correspondence of) each operation of the object language as the
+correspondence to) each operation of the object language as the
 corresponding operation in the meta-language.
@@
 
-and so |EXR.test1 ()| evaluates to untagged boolean value |true|.
-In Haskell, we define a |newtype R a = R a| and 
-|instance Symantics R|
+and so |EXR.test1 ()| evaluates to the untagged boolean value |true|.
+In Haskell, we define a\\
+|newtype R a = R a| and |instance Symantics R|
 with the straightforward implementation. We must stress that |R| is
 a |newtype|. Although it may look like a tag, it has no run-time
-representation (erased during the compilation) and pattern-matching
-against |R| cannot ever fail and so is operationally identity and
+representation (erased during compilation) and pattern-matching
+against |R| cannot ever fail and so is operationally an identity and
 compiled that way.
 
-Proposition: If |e| is the term in the meta-language has the type
+\begin{prop}
+If the term |e| in the meta-language has type
 |t|, evaluation of that term in the interpreter R will either continue
-idenfinetely or termninate with a value of the same type |t|.
+indefinitely or termninate with a value of the same type |t|.
+\end{prop}
 
@@
 the object language to its size (defined as the number of term
-constructors). The following is a slightly abbreviated code (see the
-accompanying source code for the complete definition)
+constructors). The following is slightly abbreviated code (see the
+accompanying source code for the complete definition).
 
@@
 
-Now the expression
+\noindent Now the expression
 \begin{code}
@@
 evaluates to |3|. This interpreter is not only tagless, it is also
-total. It `evaluates' even seemingly divergent terms like\\
+total. It ``evaluates'' even seemingly divergent terms like\\
 |app (fix (fun self -> self)) (int 1)|.
@@
 
-\section{Tagless compiler (aka, staged interpreter)}
+\section{Tagless compiler (aka, staged interpreter)}\label{compiler}
 Staged code in OCaml... Don't show the Haskell code (too much)
 
-\section{Tagless partial evaluator}
+\section{Tagless partial evaluator}\label{PE}
 
@@
 
-\section{Variations and discussion}
+\section{Variations and discussion}\label{discussion}
 
@@
 
-\section{Self-interpretation}
+\section{Self-interpretation}\cite{selfinterp}
 
@@
 
-\section{Related work}
+\section{Related work}\label{related}
 Walid's Jones optimality paper. See his own ICFP02 paper that
@@
 
- Simii et al.
+ Sumii et al.
 
@@
 
-\section{Conclusions}
+\section{Conclusions}\label{conclusion}
 
@@
 \subsection*{Acknowledgments}
-We woul like to tank Martin Sulzmann for helpful discussions.
+We would like to thank Martin Sulzmann for helpful discussions.
 

### 2007-03-26 tagless/tagless-final.tex
move tagless interp work to its own directory

--- 2007-03-25 matrix/tagless-final.tex
@@
 \documentclass[preprint]{sigplanconf}
-\usepackage{amsmath}
+\usepackage{amsmath,amssymb}
 \usepackage{natbib1}
@@
 \newtheorem{prop}{Proposition}
+
+\newcommand{\ZZ}{\mathbb{Z}}
+\newcommand{\BB}{\mathbb{B}}
 
@@
 \item only closed terms can be evaluated.
+\item an approach to self-interpretation which is compatible with the
+  above.
 \end{enumerate}
@@
   e ::= Lam hoas_fn | App e e | Fix hoas_fn |
-  I int_literal | B bool_literal | Add e1 e2 | Mul e1 e2 | Leq e1 e2 |
-  IF b e-then e-else
+  I int_literal | B bool_literal | Add e1 e2 | 
+  Mul e1 e2 | Leq e1 e2 | IF b e-then e-else
 \end{code}
@@
 \subsection{How to make encoding flexible: abstract the interpreter}
+\label{encoding}
 We have implemented the above in (Meta)OCaml and Haskell.  In Haskell,
@@
 
-Our sample code test1 now reads |app (lam (\x -> x)) (bool True)|,
+|test1| now reads |app (lam (\x -> x)) (bool True)|,
 whose inferred type is |Symantics repr => repr Bool|.
@@
 We can do the above in (Meta)OCaml as well, replacing type classes with
-functors, via a module type.
+functors, via a module type.  For this section, the following simple
+signature will be sufficient:
 
@@
 
+\noindent
 \begin{code}
@@
 
-and so |EXR.test1 ()| evaluates to the untagged boolean value |true|.
+\noindent and so |EXR.test1 ()| evaluates to the untagged boolean value |true|.
 In Haskell, we define a\\
@@
 \section{Tagless compiler (aka, staged interpreter)}\label{compiler}
-Staged code in OCaml... Don't show the Haskell code (too much)
-
+
+We can also write a ``compiler'' from our lambda calculus directly
+into OCaml code, using MetaOCaml's staging facilities.  This compiler
+attempts no optimizations at all, it simply produces residual code.
+This is a completely straightforward staging of |module R|.
+
+% taken from incope.ml -- should this be put into another file, to make
+% sure the paper does not get out of synch?
+\begin{code}
+module C = struct
+  type ('c,'sv,'dv) repr = ('c,'dv) code
+  let int (x:int) = .<x>.
+  let bool (b:bool) = .<b>.
+  let add e1 e2 = .<.~e1 + .~e2>.
+  let mul e1 e2 = .<.~e1 * .~e2>.
+  let leq x y = .< .~x <= .~y >.
+  let eql x y = .< .~x = .~y >.
+  let if_ eb et ee = 
+    .<if .~eb then .~(et () ) else .~(ee () )>.
+
+  let lam f = .<fun x -> .~(f .<x>.)>.
+  let app e1 e2 = .<.~e1 .~e2>.
+  let fix f = 
+     .<let rec self n = .~(f .<self>.) n in self>.
+  let dyn_id (x:('c,'sv,'dv) repr):('c,'sv1,'dv) repr = x
+end;;
+\end{code}
+
+\noindent It is readily obvious that this compiler does not suffer
+any interpretive overhead whatsoever, nor does the resulting code
+contain any tags.  We add one item to the language, |dyn_id|, as it
+is needed in the next section, where we will explain its purpose.
+
+\jacques{we need to say something about the Haskell version here too,
+even though we don't show it.  This is the first place we use GADT,
+so maybe the ByteCode data type should be shown?}
 \section{Tagless partial evaluator}\label{PE}
@@
 have no universal type and no tags for object types.  The partial
-evaluator uses the previously mentioned compiler to `interpret' dynamic
-terms.  In this sense, the partial evaluator is a modular extension to
-the compiler.
-
-Compare the output of some of the ctest with the corresponding
-ptest. Use power?
+evaluator uses the previously mentioned compiler to ``interpret'' dynamic
+terms, and it uses the interpreter to reduce static terms.  In this sense, 
+the partial evaluator is a modular extension to both the compiler and
+the interpreter.
+
+\subsection{MetaOCaml version}
+
+Here is the full code for the MetaOCaml version.
+\begin{code}
+module P = struct
+  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
+  let abstr {dy = x} = x
+  let pdyn x = {st = None; dy = x}
+
+  let int  (x:int)  = {st = Some (R.int x); dy = C.int x}
+  let bool (x:bool) = {st = Some (R.bool x); dy = C.bool x}
+
+  (* generic build - takes a repr constructor, an 
+     interpreter function and a compiler function (all 
+     binary) and builds a PE version *)
+  let build cast f1 f2 = fun e1 -> fun e2 -> 
+     match (e1,e2) with
+         ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
+        | _ -> pdyn (f2 (abstr e1) (abstr e2))
+  (* same as 'build' but takes care of the neutral element 
+     (e) simplification allowed via a monoid structure which is 
+     implicitly present *)
+  let buildsimp cast e f1 f2 = fun e1 e2 -> 
+     match (e1,e2) with
+        | ({st = Some e'}, _) when e = e' -> e2
+        | (_, {st = Some e'}) when e = e' -> e1
+        | ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
+        | _ -> pdyn (f2 (abstr e1) (abstr e2))
+
+  let add e1 e2 = buildsimp int 0 R.add C.add e1 e2
+  let mul e1 e2 = buildsimp int 1 R.mul C.mul e1 e2
+  let leq e1 e2 = build bool R.leq C.leq e1 e2
+  let eql e1 e2 = build bool R.eql C.eql e1 e2
+  let if_ eb et ee = match eb with
+           {st = Some b} -> if b then et () else ee ()
+         | _ -> pdyn (C.if_ (abstr eb) 
+                          (fun () -> abstr (et ()))
+                          (fun () -> abstr (ee ())))
+  let lam f = {st = Some f; 
+               dy = C.lam (fun x -> abstr (f (pdyn x)))}
+  let app ef ea = match ef with
+        {st = Some f} -> f ea
+      | _ -> pdyn (C.app (C.dyn_id (abstr ef)) (abstr ea))
+
+   (* For now, to avoid divergence at the PE stage,
+      we unroll the fixpoint exactly once, and then
+      residualize *)
+  let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
+end;;
+\end{code}
+
+\oleg{Compare the output of some of the ctest with the corresponding
+ptest. Use power?}
 
 One should note that the partial evaluator has fully reused the code
-of the compiler. It is truly the `composition' of the interpreter and
-the compiler.
+of the compiler and the interpreter. It is truly the `composition' of the
+interpreter and the compiler.  \jacques{Perhaps cite a paper of Thiemann
+here, cogen in 6 lines?}
+
+\jacques{Comment on the various nice generic things in the implementation,
+which are a little silly in this small version, but would really pay off
+in something bigger.  See the abstract interpretation hard at work!
+See the monoid-driven simplifications!}
+\jacques{Also need to note that the type of the above is not quite
+the original Symantics (although the code has the right operational
+semantics).  We need to make some adjustments, as per below.}
 
@@
   val bool : bool -> ('c, bool) repr
-    val lam : (('c, 'da) repr -> ('c, 'db) repr) -> ('c, 'da -> 'db) repr
-    val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
+  val lam : (('c, 'da) repr -> ('c, 'db) repr) -> 
+            ('c, 'da -> 'db) repr
+  val add : ('c, int) repr -> ('c, int) repr -> 
+            ('c, int) repr
 \end{code}
@@
                        | E of ('c, bool) code
-    ('c, 'a -> 'b) repr_pe = VF of ('c, 'a) repr -> ('c, 'b) repr
-                           | E of ('c, 'a -> 'b) code
+  ('c, 'a->'b) repr_pe = VF of ('c,'a) repr -> ('c,'b) repr
+                       | E of ('c, 'a->'b) code
 \end{code}
@@
 
-So, the full signature of Symantics looks like (cf Section 2)
-
+So we have to augment the full signature of Symantics (from that of
+sec.~\ref{encoding}) to
 \begin{code}
@@
   val bool : bool -> ('c,bool,bool) repr
-  val add  : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val mul  : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val leq  : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
+  val add  : ('c,int,int) repr -> 
+      ('c,int,int) repr -> ('c,int,int) repr
+  val mul  : ('c,int,int) repr -> 
+      ('c,int,int) repr -> ('c,int,int) repr
+  val leq  : ('c,int,int) repr -> 
+      ('c,int,int) repr -> ('c,bool,bool) repr
   val if_  : ('c,bool,bool) repr ->
@@
              (unit -> ('c,'sa,'da) repr) -> ('c,'sa,'da) repr 
-
   val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
@@
 \end{code}
+
+\jacques{We talk a lot about GADT, but show none.  I think we need to
+show a bit of Haskell, to explain what we are talking about.}
 
@@
 
-\begin{code}
-  (Int -> r Int) ->
-  (Bool -> r Bool) ->
-  (forall alpha beta. (r alpha -> r beta) -> r (alpha->beta)) -> ... r tau
-\end{code}
+$$ 
+  (\ZZ \rightarrow r \ZZ) \rightarrow
+  (\BB \rightarrow r \BB) \rightarrow
+  \forall \alpha \beta. (r \alpha \rightarrow r \beta)
+      \rightarrow r (\alpha\rightarrow\beta) \rightarrow ... r \tau
+$$
+% (Int -> r Int) ->
+% (Bool -> r Bool) ->
+% (forall alpha beta. (r alpha -> r beta) -> r (alpha->beta)) -> ... r tau
 
@@
 
-Taha et al define a self-interpreter si to be an object-language term
-such that, for any object-language term e,
+Taha et al (cite different paper) define a self-interpreter si to be an
+object-language term such that, for any object-language term e,
 
@@
 presentation at WG2.11, Emir used tagless interpreter as the main and
-only example.
+only example. \jacques{but at the PEPM07 presentation, Walid did not
+quite do it that way.  He presented Concoqtion more as an exploration
+of (an obvious but difficult) part of the design space for programming
+languages.}
 
@@
 easy, and |'dv| "perfectly reflects" an |'sv|, but you still can't get at 
-that |'sv|.
+that |'sv|.  \jacques{Could we say that we have a meta-level Galois
+injection?  We have one direction completely covered at the object
+level (and so easy to lift), but 'dv to 'sv can only be done by
+an outside mechanism}
 
@@
 
+We show how to do many things without tags, in a simple type system, 
+essentially by avoiding any (inductive) types.
+
+\jacques{Should we mention that algebraic type tags give us a partly
+intensional representation of terms.  This nicely allows pattern matching,
+but as sits inside an extensional theory, ensuring that all traces 
+of these tags disappear seems quite difficult without a very powerful
+type system (enter GADTs).}
+
 Our approach showed better the meaning of GADT and when they can be

### 2007-03-27 tagless/tagless-final.tex
Section 3 mostly done; added fodder for Sec 4

@@
 \usepackage{comment}
+\usepackage{url}
 
@@
 We have built a (family of) tagless partial evaluator(s) for a typed object
-language, placing minimal requirements on the type system of the metalanguage:
+language \oleg{``,or, embedded DSL''}, placing minimal requirements on
+the type system of the meta-language:
 we need let-bound polymorphism and staging, but not dependent
@@
 postprocessing step.
+
 The main tool is to take functions seriously: we encode the
@@
 Our family of evaluators contains an interpreter, a compiler, a partial
-evaluator and a term-size evaluator.
+evaluator and a term-size evaluator \oleg{and a CPS CBN transfromer?}
+Our approach highlights the importance of order-2 type constructors
+(order-2 polymorphism), which lets us encode DSL terms with
+translucent types: enough abstraction is available to interpret the
+terms in various ways, and yet enought type information is available 
+to typecheck the terms and statically assure their interpretation will
+never get stuck.
+\oleg{say something about type functions, GADTs and the way to get
+  around them?}
 \end{abstract}
@@
 The first attempt at the interpreter, the function eval. It takes the
-object language term such as test1 and gives us its value.
+object language term such as |test1| and gives us its value.
 The first argument to |eval| is the environment, initially empty,
@@
 solutions to this problem.
+\oleg{I feel we need to give some details of the other solutions here, 
+to give the reader the feeling of
+how advanced solutions were proposed and how rarely they are
+implemented.}
 
@@
 use includes to insure the code stays in synch?}
+\oleg{Because the fragements are interrelated, we can't. We have to
+  use sort of a `partial include', and I don't think that is easy. I
+  have checked all the fragments so far.}
 
@@
 because its code uses no operations that could possibly raise such
-an error.
-
-Also, note that 
-the interpreter uses no Universal type nor pattern matching, and it
-is obvious to the compiler that non-termination can only result from
-interpreting |fix|.  If the source language is strongly normalizing, then
-our interpreter is total.
-\jacques{the language above has no fix (yet), so we should move this 
-comment somewhere else}
+an error. In fact, our interpreter is total.
 
@@
 
+Also, note that the interpreter uses no Universal type nor pattern
+matching, and it is obvious to the compiler that non-termination can
+only result from interpreting |fix|.
+
 \begin{prop}
@@
 We can also write a ``compiler'' from our lambda calculus directly
-into OCaml code, using MetaOCaml's staging facilities.  This compiler
-attempts no optimizations at all, it simply produces residual code.
-This is a completely straightforward staging of |module R|.
-
-% taken from incope.ml -- should this be put into another file, to make
-% sure the paper does not get out of synch?
+into OCaml code, using MetaOCaml's staging facilities. Namely,
+future-stage expressions of the type |t| are represented at the
+present stage as values of the type |('a,t) code| where |'a| is the
+environment classifier (Walid03, ESOP04). Code values are created by
+a \emph{bracket} form: |.<e>.| specifies the expression |e| is to be
+evaluated at the future stage. The \emph{escape} |.~e|, which must occur
+within a bracket, specifies that the expression |e| must be evaluated
+at the present stage; its result, which must be a code value, is
+spliced in into the code being built. The \emph{run} form |.! e|
+evaluates the code value, which involves run-time compilation and
+linking of the future-stage code. The bracket, escape and run are akin to
+quasi-quotation, unquotation and |eval| of Lisp.
+
+% this code does not have the 'sv parameter. It shows up later.
 \begin{code}
 module C = struct
-  type ('c,'sv,'dv) repr = ('c,'dv) code
+  type ('c,'dv) repr = ('c,'dv) code
   let int (x:int) = .<x>.
@@
   let leq x y = .< .~x <= .~y >.
-  let eql x y = .< .~x = .~y >.
   let if_ eb et ee = 
@@
      .<let rec self n = .~(f .<self>.) n in self>.
-  let dyn_id (x:('c,'sv,'dv) repr):('c,'sv1,'dv) repr = x
 end;;
 \end{code}
+This is a completely straightforward staging of
+|module R|.
+This compiler attempts no optimizations at all, it simply produces
+residual code. For example, the evaluation of our |test1|
+\begin{code}
+  let module E = EX(C) in E.test1 ();;
+\end{code}
+gives the value of the inferred type |('a, bool) C.repr| which looks
+like |.<((fun x_6 -> x_6) (true))>.|. MetaOCaml can print the code
+values, that is, the corresponding future-stage code.
+
+\begin{comment}
+\oleg{one may wish to give the folloiwng example too; the code
+  expression can be somewhat abbreviated...}
+\begin{code}
+  let module E = EX(C) in E.testpowfix7;;
+# - : ('_a, int -> int) C.repr =
+.<fun x_1 ->
+   (((fun x_2 ->
+       let rec self_3 =
+        fun n_4 ->
+         ((fun x_5 -> if (x_5 <= 0) then 1 else (x_2 * (self_3 (x_5 + (-1)))))
+           n_4) in
+       self_3) x_1) 7)>.
+\end{code}
+\end{comment}
 
 \noindent It is readily obvious that this compiler does not suffer
-any interpretive overhead whatsoever, nor does the resulting code
-contain any tags.  We add one item to the language, |dyn_id|, as it
-is needed in the next section, where we will explain its purpose.
-
-\jacques{we need to say something about the Haskell version here too,
-even though we don't show it.  This is the first place we use GADT,
-so maybe the ByteCode data type should be shown?}
+any interpretive overhead whatsoever. Indeed, the
+  code produced for a function is simply |fun x_6 -> x_6| and has no 
+  embedded calls to the evaluator: cf. the evaluator of DSL functions
+  in Section~\ref{tagproblem}, the L-line.
+The resulting code obviously contains no tags and no pattern-matching.
+
+We have also implemented the ``compiler'' in Haskell. Since Haskell
+has no (convenient, and typed) staging facilities, we had to emulate
+them. To be precise, we defined a data type |ByteCode| with the
+constructors |Var|, |Lam|, |App|, |Fix|, |INT| etc. with the obvious
+meaning. Unlike our staging language (which is based on higher-order
+abstract syntax), our bytecode, to be realistic, uses integer-named
+variables. We then define 
+\begin{code}
+  newtype C t = C (Int -> (ByteCode t, Int)) 
+\end{code}
+where |Int| is the counter for the creation of fresh variable
+names. We define the compiler by making |C| to be the instance of the
+class |Symantics|. The implementation is quite similar (but more
+verbose) than the corresponding MetaOCaml code given above. The
+accompanying complete code gives more details.
+
+\oleg{I wanted to avoid saying anything about GADTs here. There
+  doesn't seem to be much point in getting into too long
+  explanations...}
+
 \section{Tagless partial evaluator}\label{PE}
@@
 
+\subsection{Non-functorialness of functions problem}
+the sv type problem: the inductive type function 'dv -> 'sv
+its GADT solution. 
+its type-level function solution: incope1.hs 
+\begin{code}
+data P t where
+    VI :: Int -> P Int
+    VB :: Bool -> P Bool
+    VF :: (P a -> P b) -> P (a->b)
+    E  :: C t -> P t
+
+abstr :: P t -> C t
+abstr (VI i) = int i
+abstr (VB b) = bool b
+abstr (VF f) = lam (abstr . f . E)
+abstr (E x) = x
+
+instance Symantics P where
+    int x  = VI x
+    bool b = VB b
+
+    -- lam :: (repr a -> repr b) -> repr (a->b)
+    lam = VF
+    -- app :: repr (a->b) -> repr a -> repr b
+    app ef ea = case ef of
+                        VF f -> f ea
+                        E  f -> E (app f (abstr ea))
+    add e1 e2 = case (e1,e2) of
+                 (VI n1,VI n2) -> VI nr where nr = n1 + n2
+                 _ -> E (add (abstr e1) (abstr e2))
+\end{code}
+
+GADT: |E|, Purely dynamic; the other alternatives: |static| code.
+
+We need GADT -- or, to be more precise,
+representation that varies with the type. We need a value of the type
+|P Int| to have a different representation than the value of the type 
+|P (Int->Int)|. GADT give you that (although we don't use the GADT in
+their intended meaning for that).
+
+Drawbacks of the GADT solution. GADTs, although work, are still not
+quite satisfactory: let us look at the following lines in incope.hs
+\begin{code}
+    app ef ea = case ef of
+                        VF f -> f ea
+                        E  f -> E (app f (abstr ea))
+\end{code}
+The datatype has four constructors, |VI|, |VB|, |VF| and |E|. How come we used
+only two of them in the above expression. It is obvious, one may say,
+that the constructors |VI| and |VB| just can't occur, because of the
+typing consideration. This is obvious to us -- but it is not obvious
+to the compiler. The case matching above does look \emph{partial}. That
+seems like a tenuous point -- but it is precisely the main point of
+the whole tag elimination approach. In a typed tagged interpreter
+there are lots of pattern-matching that looks partial but in reality
+total. The ultimate goal is to make this totality syntactically
+apparent.
+
+How can we do without GADT
+Now show the Symantics signature in full, with the 'sv types
+
+In the ideal world, the MetaOCaml signature (or Haskell type class)
+Symantics would start with
+\begin{code}
+  kind tp = int | bool | (->) tp tp
+  type ('c, 'dv:tp) repr : tp
+  val int : int -> ('c, int) repr
+  val bool : bool -> ('c, bool) repr
+  val lam : (('c, 'da) repr -> ('c, 'db) repr) -> 
+            ('c, 'da -> 'db) repr
+  val add : ('c, int) repr -> ('c, int) repr -> 
+            ('c, int) repr
+\end{code}
+
+Here "repr" is a tp-to-tp function, defined inductively in the case of
+partial evaluation:
+
+\begin{code}
+  ('c, int   ) repr_pe = VI of int
+                       | E of ('c, int) code
+  ('c, bool  ) repr_pe = VB of bool
+                       | E of ('c, bool) code
+  ('c, 'a->'b) repr_pe = VF of ('c,'a) repr -> ('c,'b) repr
+                       | E of ('c, 'a->'b) code
+\end{code}
+
+Unfortunately, we can't define type-level functions inductively.
+Instead we bake |repr_pe| above into the signature Symantics, so repr
+takes three type arguments |('c,'sv,'dv)|, where the extra argument 
+|'sv| is equal to |('c,'dv) repr_pe.|
+It seems Ken has just showed how to define type-level functions
+inductively... This shows also how a limited form of GADTs (the ones
+used for the sake of type-dependent representations) can be
+implemented with just functions.
+
+So we have to augment the full signature of Symantics (from that of
+sec.~\ref{encoding}) to
+\begin{code}
+module type Symantics = sig
+  type ('c,'sv,'dv) repr
+  val int  : int  -> ('c,int,int) repr
+  val bool : bool -> ('c,bool,bool) repr
+  val add  : ('c,int,int) repr -> 
+      ('c,int,int) repr -> ('c,int,int) repr
+  val mul  : ('c,int,int) repr -> 
+      ('c,int,int) repr -> ('c,int,int) repr
+  val leq  : ('c,int,int) repr -> 
+      ('c,int,int) repr -> ('c,bool,bool) repr
+  val if_  : ('c,bool,bool) repr ->
+             (unit -> ('c,'sa,'da) repr) ->
+             (unit -> ('c,'sa,'da) repr) -> 
+             ('c,'sa,'da) repr 
+  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
+    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
+           'da->'db) repr
+  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
+             'da->'db) repr
+    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
+  val fix : 
+    (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,
+         'da->'db) repr 
+    -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
+end
+\end{code}
+
 \subsection{MetaOCaml version}
 
+%%  let dyn_id (x:('c,'sv,'dv) repr):('c,'sv1,'dv) repr = x
+
+%% We add one item to the language, |dyn_id|, as it
+%% is needed in the next section, where we will explain its purpose.
+
 Here is the full code for the MetaOCaml version.
@@
 module P = struct
-  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
+  type ('c,'sv,'dv) repr = 
+        {st: 'sv option; dy: ('c,'dv) code}
   let abstr {dy = x} = x
@@
         | (_, {st = Some e'}) when e = e' -> e1
-        | ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
+        | ({st=Some n1}, {st=Some n2})-> cast (f1 n1 n2)
         | _ -> pdyn (f2 (abstr e1) (abstr e2))
@@
   let mul e1 e2 = buildsimp int 1 R.mul C.mul e1 e2
-  let leq e1 e2 = build bool R.leq C.leq e1 e2
-  let eql e1 e2 = build bool R.eql C.eql e1 e2
+  let leq e1 e2 = 
+     match (e1,e2) with
+         ({st=Some n1},{st=Some n2})-> bool (R.leq n1 n2)
+        | _ -> pdyn (C.leq (abstr e1) (abstr e2))
   let if_ eb et ee = match eb with
@@
 
-   (* For now, to avoid divergence at the PE stage,
-      we unroll the fixpoint exactly once, and then
-      residualize *)
-  let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
+  let fix f = 
+    let fdyn = C.fix (fun x -> abstr (f (pdyn x))) in
+    let fdynn e = pdyn (C.app fdyn e.dy) in
+    {st=Some (function {st = Some _} as e -> 
+               let rec self n = 
+                 (match n.st with Some _-> app (f (lam self)) n 
+                                 | _ -> fdynn n) in 
+               app (f (lam self)) e
+             | e  -> fdynn e);
+    dy = fdyn }
 end;;
@@
 semantics).  We need to make some adjustments, as per below.}
-
-the sv type problem: the inductive type function 'dv -> 'sv
-its GADT solution. 
-its type-level function solution: incope1.hs
-
-We need GADT -- or, to be more precise,
-representation that varies with the type. We need a value of the type
-|P Int| to have a different representation than the value of the type 
-|P (Int->Int)|. GADT give you that (although we don't use the GADT in
-their intended meaning for that).
-
-Drawbacks of the GADT solution. GADTs, although work, are still not
-quite satsifactory: let us look at the following lines in incope.hs
-\begin{code}
-    app ef ea = case ef of
-                        VF f -> f ea
-                        E  f -> E (app f (abstr ea))
-\end{code}
-The datatype has four constructors, |VI|, |VB|, |VF| and |E|. How come we used
-only two of them in the above expression. It is obvious, one may say,
-that the constructors |VI| and |VB| just can't occur, because of the
-typing consideration. This is obvious to us -- but it is not obvious
-to the compiler. The case matching above does look \emph{partial}. That
-seems like a tenuous point -- but it is precisely the main point of
-the whole tag elimination approach. In a typed tagged interpreter
-there are lots of pattern-matching that looks partial but in reality
-total. The ultimate goal is to make this totality syntactically
-apparent.
-
-How can we do without GADT
-Now show the Symantics signature in full, with the 'sv types
-
-In the ideal world, the MetaOCaml signature (or Haskell type class)
-Symantics would start with
-\begin{code}
-  kind tp = int | bool | (->) tp tp
-  type ('c, 'dv:tp) repr : tp
-  val int : int -> ('c, int) repr
-  val bool : bool -> ('c, bool) repr
-  val lam : (('c, 'da) repr -> ('c, 'db) repr) -> 
-            ('c, 'da -> 'db) repr
-  val add : ('c, int) repr -> ('c, int) repr -> 
-            ('c, int) repr
-\end{code}
-
-Here "repr" is a tp-to-tp function, defined inductively in the case of
-partial evaluation:
-
-\begin{code}
-  ('c, int   ) repr_pe = VI of int
-                       | E of ('c, int) code
-  ('c, bool  ) repr_pe = VB of bool
-                       | E of ('c, bool) code
-  ('c, 'a->'b) repr_pe = VF of ('c,'a) repr -> ('c,'b) repr
-                       | E of ('c, 'a->'b) code
-\end{code}
-
-Unfortunately, we can't define type-level functions inductively.
-Instead we bake |repr_pe| above into the signature Symantics, so repr
-takes three type arguments |('c,'sv,'dv)|, where the extra argument 
-|'sv| is equal to |('c,'dv) repr_pe.|
-It seems Ken has just showed how to define type-level functions
-inductively... This shows also how a limited form of GADTs (the ones
-used for the sake of type-dependent representations) can be
-implemented with just functions.
-
-So we have to augment the full signature of Symantics (from that of
-sec.~\ref{encoding}) to
-\begin{code}
-module type Symantics = sig
-  type ('c,'sv,'dv) repr
-  val int  : int  -> ('c,int,int) repr
-  val bool : bool -> ('c,bool,bool) repr
-  val add  : ('c,int,int) repr -> 
-      ('c,int,int) repr -> ('c,int,int) repr
-  val mul  : ('c,int,int) repr -> 
-      ('c,int,int) repr -> ('c,int,int) repr
-  val leq  : ('c,int,int) repr -> 
-      ('c,int,int) repr -> ('c,bool,bool) repr
-  val if_  : ('c,bool,bool) repr ->
-             (unit -> ('c,'sa,'da) repr) ->
-             (unit -> ('c,'sa,'da) repr) -> ('c,'sa,'da) repr 
-  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
-    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),'da->'db) repr
-  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),'da->'db) repr
-    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
-  val fix : (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,'da->'db) repr 
-             -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
-end
-\end{code}
-
-\jacques{We talk a lot about GADT, but show none.  I think we need to
-show a bit of Haskell, to explain what we are talking about.}
 
@@
 
+separate compilation: separate
+typechecking of terms without knowing which particular
+hole they will be plugged into. Formulate this as a prop?
+
 \subsection*{Acknowledgments}
-We would like to thank Martin Sulzmann for helpful discussions.
+We would like to thank Martin Sulzmann and Walid Taha 
+for helpful discussions.
 

### 2007-03-28 tagless/tagless-final.tex
Section 4 is mostly done

@@
 
-\section{Our base language: Tagless interpreter}\label{language}
-
-Introduce the language, The language is simply-typed lambda-calculus
+\section{Base language and its tagless interpreters}\label{language}
+
+Our DSL language to be embedded is simply-typed lambda-calculus
 with fixpoint, integers, booleans and comparison, whose BNF grammar is
@@
 lift but with more constants so we can write better examples, enough to
-express Fibonacci, factorial, and power.
+express Gibonacci, factorial, and power.
 
@@
 \label{encoding}
-We have implemented the above in (Meta)OCaml and Haskell.  In Haskell,
-the term `constructor' functions are defined as methods in a type class
-Symantics (the name means that the class interface gives the syntax for
-the source language and its instances give the semantics).
+We have embedded the above language in (Meta)OCaml and Haskell.  In Haskell,
+the constructors of DSL terms are functions defined as methods in a type class
+Symantics. The name implies that the class interface gives the syntax for
+the source language and its instances give the semantics.
 
@@
 
-We can do the above in (Meta)OCaml as well, replacing type classes with
+We also embedded our DSL in (Meta)OCaml, replacing type classes with
 functors, via a module type.  For this section, the following simple
@@
 the definitions of our object terms must look syntactically like
-values. Alternatively to giving an |()| argument, we could have used
+values. Alternatively to giving the |()| argument, we could have used
 rank-2 record types of OCaml to maintain the necessary polymorphism.
@@
 
+\subsection{Two tagless interpreters}
 The first instantiation of the Symantics structure is an interpreter:
@@
 names. We define the compiler by making |C| to be the instance of the
-class |Symantics|. The implementation is quite similar (but more
+class |Symantics|. The implementation is quite similar (but slightly more
 verbose) than the corresponding MetaOCaml code given above. The
@@
 \subsection{Non-functorialness of functions problem}
-the sv type problem: the inductive type function 'dv -> 'sv
-its GADT solution. 
-its type-level function solution: incope1.hs 
+\label{S:PE-problem}
+We want to combine the interpreter |R| and the compiler |C| to produce
+an (online) partial evaluator. \oleg{discuss if our PE online or
+  offline: here or below. In the latter case refer to the discussion
+  from here.} The partial evaluator (PE) is also an interpreter: it
+evaluates a term either to immediate (or, `static' or present-stage)
+value or a `dynamic', future-stage value, or `code' (of the sort
+produced by the compiler C). The interpreter attempts to perform all
+possible computations (i.e., computations whose operarands are
+present-stage values) as the present-stage. Our first attempt is
+straightforward: we introduce
+\begin{code}
+data P1 t = S1 (R t) (C t) | E1 (C t)
+\end{code}
+as a `direct sum' of the interpreter |R| and the compiler |C|. The
+datatype |P1| has two \emph{phase} (not type!) tags |S1| 
+(for static) and |E1| for dynamic. The static value has both the
+present-stage and future-stage alternatives. The dynamic value does
+not have the present-stage component. We define
+\begin{code}
+abstr1 :: P1 t -> C t
+abstr1 (S1 _ dyn) = dyn
+abstr1 (E1 dyn)   = dyn
+\end{code}
+to extract the future-stage component. We then attempt to define the
+interpreter |P1| compositionally -- and it works for the first-order
+fragment of our DSL language.
+
+\begin{code}
+instance Symantics P1 where
+    int  x  = S1 (int x) (int x)
+    bool b  = S1 (bool b) (bool b)
+    add (S1 n1 _) (S1 n2 _) = int (unR (add n1 n2))
+    add e1 e2 = E1 (add (abstr1 e1) (abstr1 e2))
+    -- mul and leq are analogous and elided
+    if_ (S1 s _) et ef = if unR s then et else ef
+    if_ eb et ef = 
+       E1 (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
+\end{code}
+Integer and boolean literals are obviously immediate, present-stage
+values. The result of adding two values is static only if both values
+are. If so, we use the evaluator |R| to add the values; the result is
+an essentially integer literal, which can be promoted to the future
+stage. If at least one of the operands to |add| is dynamic, we promote
+both operands to the future-stage and use the compiler |C| to
+`compile' the code for adding the operand expressions.  Because the
+static value includes as a component its future-stage variant, and
+because we build both components together, we are able to get by
+without the polymorphic `lift' operation (that is, lift the
+computation of any type from the present-stage to the future
+stage). We can do such a lifting only on type-by-type basis (for
+booleans and for integers). The problem with the polymorphic lift is
+that it inhibits true compilation, to assembler or other similar first
+order target language with no polymorphic (run-time reified) types.
+The representation of integers in the target language most likely
+differs from the representation of booleans or functions,
+etc. Therefore, without the intentional type analysis, the polymorphic
+lift is not compilable.
+
+When we come to the higher-order fragment of our language, however, we
+stumble. Let us consider |lam f|. Here, |f| is a function from a
+|P1 a| value to a |P1 b| value. The result of |lam f| must be either a
+static value or a dynamic value -- but which? Unlike |add| or |if_|, 
+|lam| cannot really examine its operand. The arguments of |add| are 
+of type |P1 int|, which we can examine by pattern-matching and so 
+determine if they are static or dynamic. The argument of |lam| is a 
+higher-order function, which is not introspectible. One may only 
+apply |f| to a particular static or dynamic value, and receive another 
+value in return. Alas, as in the case |lam (\x -> mul x dynval)|, 
+applying the same function to different static values may give 
+either static or dynamic results (in the example above, applying 
+the function to (int 0) gives the static value in return). One
+solution is to be conservative and always use the |C| compiler to
+compile |f| to the future-stage expression, which we can tag with
+E1. But that defeats the purpose of partial evaluation as we may miss
+many opportunities for static computations and simplifications, such
+as multiplication by zero in the example above. Another solution is to
+delay the binding-type-analysis (BTA) until the function |f| is
+applied to a particular argument (that is, until the result of |lam f|
+appears as the argument of |app|). That means, we have to find a way
+to incorporate |f| as it is into the P1 data structure. We would like
+to write that when |t| is a particular type |a->b|, that data type
+should be 
+\begin{code}
+data P1 (a->b) = S1 (R (P1 a -> P1 b)) (C (a->b)) 
+               | E1 (C (a->b))
+\end{code}
+That is, we need a non-parameteric data type, something like a
+typecase (ref? Gibbons?).
+
+The problem we encountered here is quite akin to typed CPS
+transform. When transforming abstractions, we come across
+non-factorialness. The connection with CPS is not an accident, as we
+shall see below (in Section XXX).
+
+\subsection{GADT -- not quite tagless PE}
+\label{S:PE-GADT}
+
+Two common ways of providing typecase in Haskell is by using either
+generalized algebraic data types (GADT) or type-class functions
+(implemented via typeclasses with functional dependencies). These
+methods are equivalent; here we show the GADT; please see |incope1.ml|
+in the accompanying source code for the latter.
+
+We introduce the following GADT
 \begin{code}
@@
     E  :: C t -> P t
-
+\end{code}
+with four data constructors. The constructor |VF| stands out as its
+operand is of the type |P a -> P b| rather than |a -> b| as one may
+have expected if |P t| were parametric in |t|. As before, we define a
+function to extract the future-stage computation from the |P t| value:
+\begin{code}
 abstr :: P t -> C t
@@
 abstr (E x)  = x
-
+\end{code}
+
+We use the method |lam| of the |C| intrepreter to compile the argument
+of |lam| into the code value. We may now make |P| the instance of
+Symantics and thus implement the partial evaluator (the
+implementations of |mul|, |leq|, |if_| and |fix| are elided).
+\begin{code}
 instance Symantics P where
@@
 
-    -- lam :: (repr a -> repr b) -> repr (a->b)
+    add e1 e2 = case (e1,e2) of
+                 (VI n1,VI n2) -> VI (n1 + n2)
+                 _ -> E (add (abstr e1) (abstr e2))
     lam = VF
-    -- app :: repr (a->b) -> repr a -> repr b
     app ef ea = case ef of
@@
                         E  f -> E (app f (abstr ea))
-    add e1 e2 = case (e1,e2) of
-                 (VI n1,VI n2) -> VI nr where nr = n1 + n2
-                 _ -> E (add (abstr e1) (abstr e2))
-\end{code}
-
-GADT: |E|, Purely dynamic; the other alternatives: |static| code.
-
-We need GADT -- or, to be more precise,
-representation that varies with the type. We need a value of the type
-|P Int| to have a different representation than the value of the type 
-|P (Int->Int)|. GADT give you that (although we don't use the GADT in
-their intended meaning for that).
-
-Drawbacks of the GADT solution. GADTs, although work, are still not
-quite satisfactory: let us look at the following lines in incope.hs
-\begin{code}
-    app ef ea = case ef of
-                        VF f -> f ea
-                        E  f -> E (app f (abstr ea))
-\end{code}
-The datatype has four constructors, |VI|, |VB|, |VF| and |E|. How come we used
-only two of them in the above expression. It is obvious, one may say,
-that the constructors |VI| and |VB| just can't occur, because of the
-typing consideration. This is obvious to us -- but it is not obvious
-to the compiler. The case matching above does look \emph{partial}. That
-seems like a tenuous point -- but it is precisely the main point of
-the whole tag elimination approach. In a typed tagged interpreter
-there are lots of pattern-matching that looks partial but in reality
-total. The ultimate goal is to make this totality syntactically
+\end{code}
+
+The implementations of |int|, |bool|, and |add| are similar to the
+above. The interpretation of |lam f| merely incorporates the
+HOAS function |f| into the result. At any point we can convert it to a
+code value, if we must. But delaying such a conversion, we get the
+opportunity to apply |f| to a concrete argument. When we evaluating
+|app ef ea|, we check to see of the first argument is so-delayed
+lambda-expression. If it is, we apply the incorporated |f| to the
+concrete argument |ea|, giving us a chance to perform the static
+computations (we see the example in the next subsection). If |ef| is a
+dynamic value, we residualize.
+
+Although GADT solve our problem, the solution is not quite
+satisfactory. First of all, it cannot be ported to MetaOCaml as GADTs
+are not supported in OCaml. Second, the tagging problem reappeared, in
+the implementation of |app| above. The datatype |P t| has four data
+constructors; the pattern-matching in |app| however used only two of
+them, |VF| and |E|. It is obvious, one may say, that the constructors
+|VI| and |VB| just cannot occur, because the corresponding values
+cannot have the arrow type. What is obvious to us is not obvious to
+the compiler. The case matching above does look \emph{partial}. The
+point may loop tenuous -- and yet it is precisely the main point of
+tag elimination. In a typed tagged interpreter there are many
+pattern-matching forms that looks partial but are total in
+reality. The goal is to make this totality \emph{syntactically}
 apparent.
 
-How can we do without GADT
-Now show the Symantics signature in full, with the 'sv types
-
-In the ideal world, the MetaOCaml signature (or Haskell type class)
-Symantics would start with
-\begin{code}
-  kind tp = int | bool | (->) tp tp
-  type ('c, 'dv:tp) repr : tp
-  val int : int -> ('c, int) repr
-  val bool : bool -> ('c, bool) repr
-  val lam : (('c, 'da) repr -> ('c, 'db) repr) -> 
-            ('c, 'da -> 'db) repr
-  val add : ('c, int) repr -> ('c, int) repr -> 
-            ('c, int) repr
-\end{code}
-
-Here "repr" is a tp-to-tp function, defined inductively in the case of
-partial evaluation:
-
-\begin{code}
-  ('c, int   ) repr_pe = VI of int
-                       | E of ('c, int) code
-  ('c, bool  ) repr_pe = VB of bool
-                       | E of ('c, bool) code
-  ('c, 'a->'b) repr_pe = VF of ('c,'a) repr -> ('c,'b) repr
-                       | E of ('c, 'a->'b) code
-\end{code}
-
-Unfortunately, we can't define type-level functions inductively.
-Instead we bake |repr_pe| above into the signature Symantics, so repr
-takes three type arguments |('c,'sv,'dv)|, where the extra argument 
-|'sv| is equal to |('c,'dv) repr_pe.|
-It seems Ken has just showed how to define type-level functions
-inductively... This shows also how a limited form of GADTs (the ones
-used for the sake of type-dependent representations) can be
-implemented with just functions.
-
-So we have to augment the full signature of Symantics (from that of
-sec.~\ref{encoding}) to
+\subsection{The `final' solution}
+\label{S:PE-solution}
+Let us re-examone the problem in Section~\ref{S:PE-problem}. What we
+would ideally like is to write
+\begin{code}
+data P t = S (repr_pe t) (C t) | E (C t)
+\end{code}
+where |repr_pe| is a type function, defined inductively as 
+\begin{code}
+repr_pe Int    = Int
+repr_pe Bool   = Bool
+repr_pe (a->b) = (repr_pr a) -> (repr_pe b)
+\end{code}
+Although we can use typeclasses to define that type function
+in Haskell, that is not portable to MetaOCaml. We notice however that
+the type case alternatives above are all associated with the existing
+methods of |Symantics| and so the case analysis is already present. 
+A simple and more portable solution thus emerges: we bake |repr_pe| 
+into the signature |Symantics|. Switching to MetaOCaml for
+portability, we recall from Section~\ref{encoding} that |repr| type
+had two arguments |type ('c,'dv) repr|. We add an extra argument,
+|'sv|, which is the result of applying |repr_pe| function above to the
+type |'dv|. The signature |Symantics| now reads
 \begin{code}
@@
     -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
-end
-\end{code}
-
-\subsection{MetaOCaml version}
-
-%%  let dyn_id (x:('c,'sv,'dv) repr):('c,'sv1,'dv) repr = x
-
-%% We add one item to the language, |dyn_id|, as it
-%% is needed in the next section, where we will explain its purpose.
-
-Here is the full code for the MetaOCaml version.
+end;;
+\end{code}
+The interpreters |R|, |L| and |C| above only relied on the last
+argument of the |repr| type, which is unchanged. And so is their
+implementation; the newly added argument |'sv| will be phantom.
+For example, we would have write, in the case of the |C| interpreter,
+\begin{code}
+module C = struct
+  type ('c,'sv,'dv) repr = ('c,'dv) code
+  (* the rest is the same *)
+\end{code}
+When writing the partial evaluator however, we make use of that
+type argument |'sv|. Here's the corresponding instance of the
+Symantics:
+
 \begin{code}
@@
 
-  (* generic build - takes a repr constructor, an 
-     interpreter function and a compiler function (all 
-     binary) and builds a PE version *)
-  let build cast f1 f2 = fun e1 -> fun e2 -> 
-     match (e1,e2) with
-         ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
-        | _ -> pdyn (f2 (abstr e1) (abstr e2))
-  (* same as 'build' but takes care of the neutral element 
-     (e) simplification allowed via a monoid structure which is 
-     implicitly present *)
   let buildsimp cast e f1 f2 = fun e1 e2 -> 
@@
         {st = Some f} -> f ea
-      | _ -> pdyn (C.app (C.dyn_id (abstr ef)) (abstr ea))
+      | _ -> pdyn (C.app (abstr ef) (abstr ea))
 
@@
                let rec self n = 
-                 (match n.st with Some _-> app (f (lam self)) n 
+                (match n.st with Some _->app (f (lam self)) n 
                                 | _ -> fdynn n) in 
@@
 
-\oleg{Compare the output of some of the ctest with the corresponding
-ptest. Use power?}
+The type |repr| here literally expresses the type equation at the
+beginning of this section. The function |abstr|, as in the previous
+subsection, extracts the future-stage computation from the result of
+PE. The function |pdyn| does the converse. As in
+section~\ref{S:PE-problem}, we build future-stage computations as we
+build present-stage ones, if any. Therefore, we avoid the need for the
+polymorphic `lift' function. The implementations of
+|add| and |mul| are a bit more sophisticated, taking advantage of the
+implicit monoid structure. The function |buildsimp| checks to see if
+its two arguments |e1| and |e2| are static expressions. If so, the
+corresponding static computation (with the help of the intrepreter
+|R|) is performed. The function also checks if one argument is static
+and is the neutral element of the monoid structure 
+\jacques{Comment on the various nice generic things in the implementation,
+which are a little silly in this small version, but would really pay off
+in something bigger.  See the abstract interpretation hard at work!
+See the monoid-driven simplifications!}
+\oleg{I'm inclined to ditch buildsimp. We miss an opportunity here:
+  when multiplying anything by 0, the result is zero!}
+
+When doing PE on fix, we made a deliberate decision to `go all the
+way' -- so long as the computation can proceed at the PE stage
+(`static value'), we do that rather than residualizing. In the
+accompanying source code, we have also implemented an alternative
+where fix is unrolled only once and them we residualize. We prefer the
+first approach, even though it (unlike the latter) may cause
+non-termination at the PE stage -- but \emph{only} if the computation
+we are partially evaluating is non-terminating and enough static
+information is available to trigger this non-termination.  Our
+treatment of fix may lead to the code bloat in the residual
+program. That is an issue investigated elsewhere (cite our PEPM2006
+paper).
+
+Given the above Symantics, our running test now evaluates
+\begin{code}
+  let module E = EX(P) in E.test1 ();;
+\end{code}
+to
+\begin{code}
+- : ('a, bool, bool) P.repr = 
+   {P.st = Some true; P.dy = .<(true)>.}
+\end{code}
+It is instructive to compare this result with that of the |C|
+evaluator. We observe the beta-redex that was apparent there is now
+performed. The result is the static value |true|.
+More interesting is the partial evaluation of testpowfix7. The |C|
+interpreter evaluates |testpowfix7| to the future-stage code with many
+apparent beta-redexes.
+\begin{code}
+  let module E = EX(C) in E.testpowfix7;;
+ - : ('_a, int -> int) C.repr =
+.<fun x_1 ->
+   (((fun x_2 ->
+       let rec self_3 =
+        fun n_4 ->
+         ((fun x_5 -> if (x_5 <= 0) then 1 
+                      else (x_2 * (self_3 (x_5 + (-1)))))
+           n_4) in
+       self_3) x_1) 7)>.
+\end{code}
+The result of the |P| interpreter is different:
+
+\begin{code}
+  let module E = EX(P) in E.testpowfix7;;
+ - : ('_a, ('_a, int, int) P.repr -> ('_a, int, int) P.repr, 
+      int -> int) P.repr = 
+ {P.st = Some <fun>;
+  P.dy = .<fun x_1 -> 
+     (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * x_1))))))>.}
+\end{code}
+
+which is what one would have expected from the power example.
+
+Unlike the GADT approach in the previous subsection, there is no
+longer any seemingly partial pattern-matching. The pattern-match only
+occurs in PE (to check if the code is static or dynamic), it is only a
+phase rather than a type tag match, and the pattern match is
+\emph{patently} exhaustive. That is, totality is assured and it is
+assured \emph{syntactically}, in the way that is apparent to the
+compiler.
 
@@
 
-\jacques{Comment on the various nice generic things in the implementation,
-which are a little silly in this small version, but would really pay off
-in something bigger.  See the abstract interpretation hard at work!
-See the monoid-driven simplifications!}
-\jacques{Also need to note that the type of the above is not quite
-the original Symantics (although the code has the right operational
-semantics).  We need to make some adjustments, as per below.}
-
-Also, unlike the GADT
-solution, there is no longer any seemingly partial
-pattern-matching. The pattern-match only occurs in PE (to check if the
-code is static or dynamic), it is only a phase rather than a type tag
-match, and the pattern match is \emph{patently} exhaustive. That is,
-totality is assured and it is assured \emph{syntactically}, in the way that
-is apparent to the compiler.
+\section{Variations and discussion}\label{discussion}
+
+state and imperative features
+
+CPS transformation (interpreting/compiling the term to CBN CPS)
 
@@
 CPS....
-
-Say that when doing PE on fix, we made a deliberate decision to `go
-all the way' -- so long as the computation can proceed at the PE stage
-(`static value'), we do that rather than residualizing. We have also
-an alternative where fix is unrolled only once and them we
-residualize. We have used the first approach, even it (unlike the
-latter) may cause non-termination at the PE stage -- but \emph{only} if the
-computation we are partially evaluating is non-terminating and enough
-static information is available to trigger this non-termination.
-Say that our treating of fix may lead to the code bloat in the
-residual program. That is an issue investigated elsewhere (cite our
-PEPM2006 paper).
-
-\section{Variations and discussion}\label{discussion}
-
-state and imperative features
-
-CPS transformation (interpreting/compiling the term to CBN CPS)
 
@@
 avoided.
+This shows also how a limited form of GADTs (the ones
+used for the sake of type-dependent representations) can be
+implemented with just functions.
 

### 2007-03-28 tagless/incope.hs

@@ -289,15 +289,40 @@ ctestpw7  = compC . testpowfix7 $ ()
 ctestpw72 = compC  (app (testpowfix7 ()) (int 2))
 
 -- ------------------------------------------------------------------------
--- The partial evaluator
-
+-- The partial evaluator: the combination of the interpreter (R) and
+-- the compiler (C).
 -- We need no Lift byte-code instruction: no parametric CSP. That is great!
 
+-- First attempt: works for the first-order fragment of our language,
+-- but stumbles on the higher-order fragment.
+
+data P1 t = S1 (R t) (C t) | E1 (C t)
+
+abstr1 :: P1 t -> C t
+abstr1 (S1 _ dyn) = dyn
+abstr1 (E1 dyn)   = dyn
+
+instance Symantics P1 where
+    int  x  = S1 (int x) (int x)
+    bool b  = S1 (bool b) (bool b)
+    add (S1 n1 _) (S1 n2 _) = int (unR (add n1 n2))
+    add e1 e2 = E1 (add (abstr1 e1) (abstr1 e2))
+    mul (S1 n1 _) (S1 n2 _) = int (unR (mul n1 n2))
+    mul e1 e2 = E1 (mul (abstr1 e1) (abstr1 e2))
+    leq (S1 n1 _) (S1 n2 _) = bool (unR (leq n1 n2))
+    leq e1 e2 = E1 (leq (abstr1 e1) (abstr1 e2))
+    if_ (S1 s _) et ef = if unR s then et else ef
+    if_ eb et ef = E1 (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
+
+-- But the problem occurs when we try to implement lam. The result
+-- of (lam f) must be either S1 _ _ or E1 _ value. Alas, we won't know
+-- which is which until we apply the function 'f' to a particular
+-- P1 value. 
+
+
 -- The code below is NOT parametric. We could have used type-classes
--- instead of GADTs. Ken noted that the code below could be re-functionalized,
--- and so could in fact be translated into MetaOCaml.
--- Or, if we observe the correspondence between GADT and MetaOCaml <code>,
--- perhaps we can translate to MetaOCaml with an additional layer of code?
+-- instead of GADTs (see incope1.hs, where we did just that). 
+-- See also the GADT-less solution below.
 
 data P t where
     VI :: Int -> P Int                -- Possibly static (base type)
@@ -340,15 +365,15 @@ instance Symantics P where
     fix f = pfix f -- need this charade for GADTs sake
 
[…]

### 2007-03-28 tagless/tagless-final.tex
just some preliminary bib work

@@
 
-We have programmed out interpreters in OCaml/MetaOCaml (cite
-www.metaocaml.org) and Haskell. The complete code is available 
+We have programmed out interpreters in OCaml/MetaOCaml (\cite{metaocaml})
+and Haskell. The complete code is available 
 \footnote{the URL will appear in the final version of the paper.}
@@
 
-        Taha et al and Xi et al (both POPL2003) argued that writing an
+    At POPL 2003, Taha et al~\cite{TahaPOPL03} and Xi et al~\cite{XiPOPL03}
+both
 interpreter of a typed language in a typed language is an interesting
@@
 Davies, Pitts).
+(\cite{WalidPOPLo3,NanevskiICFP02,NanevskiJFP05,DaviesJACM01}
+\jacques{I hope that is the right Davies paper}
 
@@
 implemented.}
+\jacques{Sure, we can refer to this section from the Related Work section
+for details.}
 

### 2007-03-28 tagless/tagless-final.tex
minor edits, many more to come

@@
 We have built a (family of) tagless partial evaluator(s) for a typed object
-language \oleg{``,or, embedded DSL''}, placing minimal requirements on
+language (embedded DSL), placing minimal requirements on
 the type system of the meta-language:
@@
 Our family of evaluators contains an interpreter, a compiler, a partial
-evaluator and a term-size evaluator \oleg{and a CPS CBN transfromer?}
+evaluator, a term-size evaluator and a CPS call-by-name transformer.
 Our approach highlights the importance of order-2 type constructors
-(order-2 polymorphism), which lets us encode DSL terms with
+(rank-2 polymorphism), which lets us encode DSL terms with
 translucent types: enough abstraction is available to interpret the
@@
 \oleg{say something about type functions, GADTs and the way to get
-  around them?}
+  around them?}\jacques{not in the abstract?}
 \end{abstract}
@@
 
-We have programmed out interpreters in OCaml/MetaOCaml (\cite{metaocaml})
+We have programmed our interpreters in OCaml/MetaOCaml (\cite{metaocaml})
 and Haskell. The complete code is available 
@@
 
-\oleg{see tagless\_interp1.ml, module Tagfull for the complete code.
-  If you change the code in here, please adjust the .ml file
-  accordingly. Let the paper and the accompanying code be in sync.}
-
-Consider the sample language of abstraction, application and boolean
+%\oleg{see tagless\_interp1.ml, module Tagfull for the complete code.
+%  If you change the code in here, please adjust the .ml file
+%  accordingly. Let the paper and the accompanying code be in sync.}
+
+Consider a language with abstraction, application and boolean
 literals with de Bruin indices as variables. This is the language used
-in the body of Walid02, from which we borrow the first part of 
+in the body of \cite{WalidIFCP02}, from which we borrow the first part of 
 this example.

### 2007-03-29 tagless/tagless-final.tex
make many edits to sections 1,2,3.  section 4 to come later.

@@
 \newcommand{\ccshan}[1]{{\it [Ken says: #1]}}
+
+\usepackage{natbib1}
+\bibpunct();A{},
+\let\cite=\citep
 
@@
 free variables in the encoded term.  In other words, we eschew initial
-algebras for the wilder world of implicit representations of final algebras.
+algebras for the wilder world of final algebras.
 Our family of evaluators contains an interpreter, a compiler, a partial
@@
 `subject' language).
+\jacques{\emph{subject} and \emph{object} are both used in the current
+write-up; should we define them to be equivalent, or just stick to one?}
 
@@
 
-The first attempt at the interpreter, the function eval. It takes the
-object language term such as |test1| and gives us its value.
+Let us attempt to implment an the interpreter, the function |eval0|. It takes
+an object language term such as |test1| and gives us its value.
 The first argument to |eval| is the environment, initially empty,
@@
 and that interpreter) were untyped. The above code would be acceptable.
-The |L e| line exhibits interpretative overhead (see descr in Walid02
-paper): |eval0 env' e| will be executed every time the result of
-evaluating |L e| is applied. Staging can be used to remove the
-interpretative overhead. Cite some paper -- see Walid02, Sec 1.1 and
-1.2. They say it well.
+The |L e| line exhibits interpretive overhead \cite{WalidICFP02}: 
+|eval0 env' e| will be executed every time (the result of
+evaluating) |L e| is applied. Staging can be used to remove this
+interpretive overhead (sections 1.1 and 1.2 of \cite{WalidICFP02} explain
+this well).
 
@@
 the function |eval0| is ill-typed. The line |B b|
-tells that the function returns a boolean, whereas the next line says
-the result of |eval0| is a function. Clearly, the result of the
-function can't be both of these types. 
-A related problem is the type of the env: since variables
+says that the function returns a boolean, whereas the next line says
+the result is a function. Clearly, the result of |eval0|
+can't be both of these types. 
+A related problem is the type of the environment |env|: since variables
 in the language may be either of boolean or function types, a regular
@@
 
-The solution is to introduce a Universal type (see Walid02 Section
-1.3) which contains both of these:
+The usual solution is to introduce a Universal type (section 1.3 of
+\cite{WalidICFP02}) which contains both of these:
 
@@
 
-We can then write the interpreter, which is accepted by the typecheker
+We can then write an interpreter which is accepted by the typecheker
 \begin{code}
@@
 pattern match in the |A (e1,e2)| line raises an exception when we try to
-evaluate something like |A (B true, B false)|, that is, the first
+evaluate something like |A (B true, B false)|, that is when the first
 operand is not a function.
 
-Suppose our object language were typed, too. Then terms like
+Suppose our object language were typed too. Then terms like
 |A (B true, B false)| would be ill-typed. Thus, if our evaluator
-accepted only typed terms, then the missing alternatives in the |A|
+accepted only typed terms, the missing alternatives in the |A|
 line would not occur. Alas, what is obvious to us is not obvious to
@@
 to attach and eliminate the UB and UA tags, even when there is no
-longer a need for them. See Walid02, Sec 1.4 for more detail.
-
-Of course there is also the issue how to design a type system to type
-terms like |test1| (GADT seem necessary). Also, how to design the
-typesystem to keep track of open/closed status of terms, so we can
-statically assure that only closed terms are passed to eval and that
-the empty list checking in |List.hd| and |List.tl| can be eliminated
-as well. The latter is even more interesting problem requiring even
-more advanced type systems (cite env classifiers, Nanevski, Rowan
-Davies, Pitts).
+longer a need for them. See \cite{WalidICFP02}, Section 1.4 for more details.
+
+In short: It is not satisfactory
+to use a Universal type because tagging hurts performance and
+necessitates partial pattern matching in the metalanguage, even for a
+well-typed object term.
+
+\subsection{Other solutions}
+
+But how should one design a type system to type
+terms like |test1|? GADTs seem necessary. Similarly, how does one design a
+typesystem that can keep track of the open/closed status of terms, so that we
+can statically insure that only closed terms are passed to |eval| and that
+the empty list checking in |List.hd| and |List.tl| can be eliminated.
+The latter is an even more interesting problem requiring even
+more advanced type systems 
 (\cite{WalidPOPLo3,NanevskiICFP02,NanevskiJFP05,DaviesJACM01}
@@
 
-The problem is also outlined in Simon Peyton-Jones GADT paper, but
-only in the context of a first-order language
-
-In short: \oleg{ (put into really introduction or the conclusions?) }
-It is not satisfactory
-to use the Universal type because tagging hurts performance and
-necessitates partial pattern matching in the metalanguage, even for a
-well-typed object term.
-
-% \subsection{Other solutions}
-% The tag problem motivated GADTs (SPJ, Xi) and dependent types
-% (WalidICFP02, Concoqtion). Or, perhaps just refer to related work?
-% But we need to say something here, to give the reader the feeling of
-% how advanced solutions were proposed and how rarely they are implemented.
+This is also the main motivating problem (albeit in the context of
+a first-order language) for the introduction of GADTs 
+into Haskell \cite{PeytonJonesICFP2006}.  This was also a motivating
+factor behind the work of Xi on GADTs \cite{XiGADTs}, and some of the
+recent work on dependent types \cite{WalidICFP02,WalidConcoqtion}.
 
@@
 implemented.}
-\jacques{Sure, we can refer to this section from the Related Work section
-for details.}
-
-\subsection{The `Final' Proposal}\label{ourapproach}
+\jacques{I have rearranged this section so that it should be clear how
+to extend it to explain these other solutions}
+
+\subsection{The ``Final'' Proposal}\label{ourapproach}
 
@@
 
-\oleg{see tagless\_interp1.ml, module Tagless for the complete code.
-If you change the code in here, please adjust the .ml file
-accordingly. Let the paper and the accompanying code be in sync.}
-
-\jacques{do we want to start using a lot of little files and
-use includes to insure the code stays in synch?}
-\oleg{Because the fragements are interrelated, we can't. We have to
-  use sort of a `partial include', and I don't think that is easy. I
-  have checked all the fragments so far.}
-
 Now, let us define our object language (embedded DSL) in a different
 way. Instead of using algebraic datatypes to encode the terms of the
-language in the meta-language (reminiscent to the initial approach),
-we will be using functions as term constructors (the final approach):
+language in the meta-language (reminiscent to the \emph{initial algebra} 
+approach), we will be using functions as term constructors (the \emph{final}
+approach):
 
@@
 
-That is all of the interpreter. Our sample term now reads
+\noindent That is all of the interpreter. Our sample term now reads
 \begin{code}
@@
 which is almost the same as before, only written with lower-case
-identifiers. To evaluate the term, we merely need to apply 
+identifiers. To evaluate a term, we merely need to apply 
 the term to the empty environment
@@
 
-\oleg{
-We should emphasize that not only our code can't have pattern-match
-errors, that fact is apparent to the compiler itself. It is one thing
-to write a code with pattern-matching and claim (and prove) that
-missing cases cannot occur. However true that may be, the fact the
-missing cases cannot occur is not evident to the compiler, which still
-has to generate code to report missing cases or compile in the error
-generation. That's where efficiency is lost, and Walid's papers on tag
-elimination have numbers showing that performance is lost indeed. 
-In our code, there is simply no pattern-mismatch opportunity: the
-proof of no bad match is expressed in the code itself. }
+Furthermore, it should be clear that we do not use any
+pattern-matching at all, and thus cannot have any pattern-matching
+errors.  This is clear not just to us, but also to the compiler.
+In other words in traditional approaches one can prove a meta-theorem
+that all pattern-matching is total, but this is nevertheless not
+apparent to the compiler, which thus still has to generate code for
+those cases.  This seriously hurts efficiency~\cite{WalidICFP02}.
 
@@
 we explain how to abstract the interpreter,
-to make it possible to `evaluate' the same term in many different
+to make it possible to ``evaluate'' the same term in many different
 ways: interpretation, compilation, compute the size (or depth), etc.
@@
 pattern matching and it is patently obvious to the compiler that the
-only non-termination can come from interpreting |fix|. If the source
+only non-termination can come from interpreting the fixpoint
+combinator |fix|. If the source
 language is strongly normalizing, our interpreter is total.
@@
 
-We have implemented the above in (Meta)OCaml and Haskell.
-
-What are our contributions:
+\noindent We have implemented the above in (Meta)OCaml and Haskell.
+We view our contributions as:
 \begin{enumerate}
@@
 \item a nicely comparable Haskell/MetaOCaml version,
-\item a functorial Symantics that deals with uniformly across all
-evaluators, from interpretation to partial evaluation,
+\item a functorial approach which deals with uniformly with all
+evaluators, from interpretation to partial evaluation to CBN CPS transform,
 \item clear and easy extensibility to more features,
-\item only closed terms can be evaluated.
-\item an approach to self-interpretation which is compatible with the
-  above.
+\item only well-typed and closed terms can be evaluated, and this is 
+checked statically,
+\item and an approach to self-interpretation which is compatible with the
+  above.  Self-interpretation turned out to be considerably harder than
+  we at first expected.
 \end{enumerate}
 
-Making sure that we interpret terms that are well-typed and closed --
-and we can do that in far simpler type system (e.g., Haskell or
-OCaml). It was thought that doing either requires quite advanced
-extensions (GADT, Nanevski, Concoqtion).
+We want to reiterate that it was previously thought that insuring
+that types are well-typed \emph{or} closed required advanced type
+extensions, and we show that this can be done in a simple subset
+of both OCaml and Haskell.
+
 \jacques{can we state fairly clearly what type system we need?  I mean,
@@
 
-Our DSL language to be embedded is simply-typed lambda-calculus
+The language we will embed is a simply-typed lambda-calculus
 with fixpoint, integers, booleans and comparison, whose BNF grammar is
@@
 
-This time we use higher-order abstract syntax (cite: but not
-Elliott. Cite Dale Miller!) rather than de Bruijn indices.  The
-language is close to the language of Xi03, without the polymorphic
+This time we use higher-order abstract syntax \cite{DaleMiller}
+rather than de Bruijn indices.  The
+language is close to the language of \cite{XiPOPL03}, without the polymorphic
 lift but with more constants so we can write better examples, enough to
-express Gibonacci, factorial, and power.
+express Fibonacci, factorial, and power.
 
@@
 
-|test1| now reads |app (lam (\x -> x)) (bool True)|,
+\noindent |test1| now reads |app (lam (\x -> x)) (bool True)|,
 whose inferred type is |Symantics repr => repr Bool|.
@@
 restriction and so that the type of |testpowfix| and |testpowfix7|
-will remain polymorphic in |repr|. Instead of supplying the dummy
+will remain polymorphic in |repr|. Instead of supplying this dummy
 argument, we could have given the terms explicit polymorphic
@@
 
-Note that the subject language (DSL) is typed. If we make a mistake
+Note that the subject language is typed. If we make a mistake
 (for example, replace |int 7| with |bool True| in |testpowfix7|, the
 compiler will highlight that term and complain that it couldn't match
-the expected |Int| against the inferred |Bool|). A type error
+the expected |Int| against the inferred |Bool|. A type error
 also results when entering |lam (\x -> app x x)|, with the compiler
 complaining about an infinite type. The type-checker of the
-meta-language will flag meaningless DSL (subject level) terms --
+meta-language will flag meaningless subject level terms --
 before any of their evaluation. If we forget, for example, |app| in
@@
 
-We also embedded our DSL in (Meta)OCaml, replacing type classes with
+We have also embedded our DSL in (Meta)OCaml, replacing type classes with
 functors, via a module type.  For this section, the following simple
@@
 Some differences are in the additional type parameter |c| (the
-environment classifier, Walid03) to be used later when we get to code
-generation (required by MetaOCaml, part of the type of the
+environment classifier from \cite{WalidPOPL03}) to be used later when we get to
+code generation (this is required by MetaOCaml as part of the type for the
 code). MetaOCaml is a call-by-value language, and thus we have to 
@@
 type parameter in the signature of Symantic functions) must remain
-uninstantiated in order for us to `run' the code. That means, the type
-must be polymorhic. The value restriction then dictates 
+uninstantiated in order for us to `run' the code. That means the type
+must be polymorhic. The value restriction then dictates that
 the definitions of our object terms must look syntactically like
 values. Alternatively to giving the |()| argument, we could have used
-rank-2 record types of OCaml to maintain the necessary polymorphism.
+the rank-2 record types of OCaml to maintain the necessary polymorphism.
 
@@
 examples by instatiating the |EX| functor with |R|.
-\oleg{Jacques: you mentioned "Boxes go Bananas" by Washburn and
-  Weirich. Do you want to comment more about that paper here?}
-
 \noindent
@@
 |newtype R a = R a| and |instance Symantics R|
-with the straightforward implementation. We must stress that |R| is
+with a straightforward implementation. We must stress that |R| is
 a |newtype|. Although it may look like a tag, it has no run-time
@@
 
-Also, note that the interpreter uses no Universal type nor pattern
+Note that the interpreter uses no Universal type nor pattern
 matching, and it is obvious to the compiler that non-termination can
@@
   let if_ eb et ee = eb + et () + ee () + 1
-
   let lam f        = (f 0) + 1
@@
 present stage as values of the type |('a,t) code| where |'a| is the
-environment classifier (Walid03, ESOP04). Code values are created by
-a \emph{bracket} form: |.<e>.| specifies the expression |e| is to be
-evaluated at the future stage. The \emph{escape} |.~e|, which must occur
+environment classifier \cite{WalidPOPL03,WalidESOP04}. Code values are created
+by a \emph{bracket} form: |.<e>.| specifies the expression |e| is to be
+evaluated at a future stage. The \emph{escape} |.~e|, which must occur
 within a bracket, specifies that the expression |e| must be evaluated
-at the present stage; its result, which must be a code value, is
+at the current stage; its result, which must be a code value, is
 spliced in into the code being built. The \emph{run} form |.! e|
@@
     .<if .~eb then .~(et () ) else .~(ee () )>.
-
   let lam f         = .<fun x -> .~(f .<x>.)>.
@@
 \end{code}
-gives the value of the inferred type |('a, bool) C.repr| which looks
+gives a value with inferred type |('a, bool) C.repr| which looks
 like |.<((fun x_6 -> x_6) (true))>.|. MetaOCaml can print the code
@@
 
-\begin{comment}
-\oleg{one may wish to give the folloiwng example too; the code
-  expression can be somewhat abbreviated...}
-\begin{code}
-  let module E = EX(C) in E.testpowfix7;;
-# - : ('_a, int -> int) C.repr =
-.<fun x_1 ->
-   (((fun x_2 ->
-       let rec self_3 =
-        fun n_4 ->
-         ((fun x_5 -> if (x_5 <= 0) then 1 else (x_2 * (self_3 (x_5 + (-1)))))
-           n_4) in
-       self_3) x_1) 7)>.
-\end{code}
-\end{comment}
-
-\noindent It is readily obvious that this compiler does not suffer
+It is readily obvious that this compiler does not suffer
 any interpretive overhead whatsoever. Indeed, the
@@
 
-We have also implemented the ``compiler'' in Haskell. Since Haskell
-has no (convenient, and typed) staging facilities, we had to emulate
-them. To be precise, we defined a data type |ByteCode| with the
+We have also implemented this ``compiler'' in Haskell. Since Haskell
+has no (convenient, typed) staging facilities, we had to emulate
+them. To be precise, we defined a data type |ByteCode| with
 constructors |Var|, |Lam|, |App|, |Fix|, |INT| etc. with the obvious
@@
 where |Int| is the counter for the creation of fresh variable
-names. We define the compiler by making |C| to be the instance of the
-class |Symantics|. The implementation is quite similar (but slightly more
+names. We define the compiler by making |C| to be an instance of the
+class |Symantics|. The implementation\footnote{The implementation uses
+GADTs because we also wanted to write a typed interpreter for 
+the \textsf{ByteCode} \emph{datatype}.} is quite similar (but slightly more
 verbose) than the corresponding MetaOCaml code given above. The
-accompanying complete code gives more details.
-
-\oleg{I wanted to avoid saying anything about GADTs here. There
-  doesn't seem to be much point in getting into too long
-  explanations...}
+accompanying code gives the full details.
 
@@
 Surprisingly, it is also possible to write a partial evaluator!  The
-partial evaluator does have tags: whether a value is static or dynamic.
-The pattern-matching on this \emph{phase} tag is always exhaustive; we still
+partial evaluator does have tags: \emph{binding time} or \emph{phase}
+tags, i.e. whether a value is static or dynamic.
+Pattern-matching on this phase tag is always exhaustive; we still
 have no universal type and no tags for object types.  The partial
 evaluator uses the previously mentioned compiler to ``interpret'' dynamic
-terms, and it uses the interpreter to reduce static terms.  In this sense, 
+terms, and uses the interpreter to reduce static terms.  In this sense, 
 the partial evaluator is a modular extension to both the compiler and
@@
 
-\subsection{Non-functorialness of functions problem}
+\subsection{The trouble with (opaque) functions}
 \label{S:PE-problem}
 We want to combine the interpreter |R| and the compiler |C| to produce
-an (online) partial evaluator. \oleg{discuss if our PE online or
-  offline: here or below. In the latter case refer to the discussion
-  from here.} The partial evaluator (PE) is also an interpreter: it
-evaluates a term either to immediate (or, `static' or present-stage)
-value or a `dynamic', future-stage value, or `code' (of the sort
+an (online) partial evaluator.  We defer the discussion of the 
+choice of online/offline partial evaluation until after we have presented our
+solution in~\ref{S:PE-solution}.
+The partial evaluator (PE) is also an interpreter: it
+evaluates a term either to an immediate (or, `static' or present-stage)
+value or to a `dynamic', future-stage value, or `code' (of the sort
 produced by the compiler C). The interpreter attempts to perform all
@@
 
+\jacques{Comment on online/offline here}
+
 \section{Variations and discussion}\label{discussion}
@@
 
+Boxes go Bananas by Weirich and Washburn.
+
 \section{Conclusions}\label{conclusion}
 
-We show how to do many things without tags, in a simple type system, 
-essentially by avoiding any (inductive) types.
+We were dissatisfied with current approaches to DSL embedding: either
+the performance was hurt by tagging/untagging and other interpretive
+overhead, or the offered solutions required too much expressive power
+from the host language's type system.  By shifting from an initial
+algebra approach to one which leverages the natural co-algebraic structure
+of the $\lambda$-calculus, we show that both of these issues can be 
+simultaneously dealt with.
+
+% We show how to do many things without tags, in a simple type system, 
+% essentially by avoiding any (inductive) types.
 
@@
 
+\bibliographystyle{plainnat}
+\bibliography{tagless}
 \end{document}

### 2007-03-30 tagless/tagless-final.tex
Starting to work

@@
 \usepackage{amsmath,amssymb}
-\usepackage{natbib1}
 \usepackage{comment}
+\usepackage{hyphenat}
 \usepackage{url}
+\usepackage{prooftree1}
 
@@
 \usepackage{natbib1}
-\bibpunct();A{},
 \let\cite=\citep
@@
 \newcommand{\BB}{\mathbb{B}}
+
+\newcommand{\fun}[1]{\mathopen{\lambda\mathord{#1}.\,}}
+\newcommand{\fix}[1]{\mathopen{\mathrm{fix\,}\mathord{#1}.\,}}
+\newcommand{\Forall}[1]{\mathopen{\forall\mathord{#1}.\,}}
+\newcommand{\Exists}[1]{\mathopen{\exists\mathord{#1}.\,}}
+\newcommand{\cond}[3]{\mathrm{if\,}#1\mathrm{\,then\,}#2\mathrm{\,else\,}#3}
+
+% Make space around section headings no less than space around theorems
+\makeatletter
+\@tempdima\topsep \advance\@tempdima\parsep
+\ifdim\@sectionbelowskip   <\topsep \@sectionbelowskip   \topsep \fi
+\ifdim\@subsectionbelowskip<\topsep \@subsectionbelowskip\topsep \fi
+\makeatother
+
+% Match the rule inserted by \@makecaption in sigplanconf.cls
+\setproofrulebreadth .33pt
+\newenvironment{floatrule}
+    {\hrule width \hsize height .33pt \vspace{.5pc}}
+    {\par\addvspace{1ex}}
+
+% Make some Rel symbols into Bin symbols instead
+\DeclareMathSymbol{\to}{\mathbin}{symbols}{"21}
+\DeclareMathSymbol{:}{\mathbin}{operators}{"3A}
 
@@
 
-See Section 1 of Walid02. Our subject area is that of a definitional DSL
-interpreter. 
-Define Staging, meta-language and object language. (Walid uses
-`subject' language).
-\jacques{\emph{subject} and \emph{object} are both used in the current
-write-up; should we define them to be equivalent, or just stick to one?}
+A popular way to define and implement a programming language is to embed it in
+another \citep{reynolds-definitional}.  The embedded language is called the
+\emph{object} or \emph{subject} language, whereas the embedding language is the
+\emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
+specific object languages because it supports rapid prototyping and integration
+with the host environment.  If the metalanguage supports \emph{staging}, then
+the embedding can compile object programs to the metalanguage and avoid the
+overhead of interpreting them on the fly \citep{WalidICFP02}.  A staged
+definitional interpreter is thus a promising way to build a domain\hyp specific
+language.
+
+\begin{figure}
+    \begin{floatrule}
+    \begin{proofrules}
+        \[ \[ [x:t_1] \proofoverdots e:t_2 \] \justifies \fun{x}e:t_1\to t_2 \]
+        \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
+        \[ e_1:t_1\to t_2 \quad e_2:t_1 \justifies e_1 e_2: t_2 \]
+        \[ \text{$n$ is an integer} \justifies n:\ZZ \]
+        \[ \text{$b$ is a Boolean} \justifies b:\BB \]
+        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
+        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \times e_2:\ZZ \]
+        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]
+        \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
+    \end{proofrules}
+    \end{floatrule}
+    \caption{Our typed object language}
+    \label{fig:object}
+\end{figure}
+
+We focus on embedding a typed object language into a typed metalanguage.
+\ccshan{Collect reasons to type here.}
 
@@
 \oleg{should say somewhere in the Introduction that we are interested
-  in the typed subject language embedded in the typed
+  in the typed object language embedded in the typed
   meta-language. Refer to Sec 2 (see the remark below) why this is
@@
 literals with de Bruin indices as variables. This is the language used
-in the body of \cite{WalidIFCP02}, from which we borrow the first part of 
+by \citet{WalidICFP02}, from which we borrow the first part of 
 this example.
@@
 
-Note that the subject language is typed. If we make a mistake
+Note that the object language is typed. If we make a mistake
 (for example, replace |int 7| with |bool True| in |testpowfix7|, the
@@
 complaining about an infinite type. The type-checker of the
-meta-language will flag meaningless subject level terms --
+meta-language will flag meaningless object level terms --
 before any of their evaluation. If we forget, for example, |app| in
@@
 location. So the type-checker also flags syntactically invalid
-subject-level terms.\oleg{Should we perhaps move this up, in the
+object-level terms.\oleg{Should we perhaps move this up, in the
   intro? To justify why typed-metalanguages and the typed DSL are
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{mcbride}
 \bibliography{tagless}

### 2007-03-30 tagless/tagless-final.tex
a few cosmetic changes

@@
 never get stuck.
-\oleg{say something about type functions, GADTs and the way to get
-  around them?}\jacques{not in the abstract?}
 \end{abstract}
@@
 be made obvious to the compiler. That also exlains unoptimality.
+\oleg{Should say somewhere here that our code is surprisingly simple,
+  and, in the hindsight, obvious. The problem of typed direct
+  intrepreter of a typed language is difficult and thought to require
+  GADT or dependent types. 
+  In fact, several papers introducing GADTs (Xi03 and SPJ) as well as 
+  (language in Walid03) and Concoqtion, written 2003-2007, used the 
+  typed tagless intrepreter as the motivating example.
+  Taha (in Jones optimality paper, I think) said that typed
+  interpreter of the typed language was not solved then.
+  Or, should we refer to Sec `Other solutions'}
 
@@
 
-and a sample term is
+with a sample term (application of the identity function to a boolean) as
 \begin{code}
@@
 
-Let us attempt to implment an the interpreter, the function |eval0|. It takes
+Let us attempt to implement an the interpreter, the function |eval0|. It takes
 an object language term such as |test1| and gives us its value.
@@
 
-Suppose our meta-language (the language in which we wrote test1
+Suppose our meta-language (the language in which we wrote |test1|
 and that interpreter) were untyped. The above code would be acceptable.
@@
 the function |eval0| is ill-typed. The line |B b|
-says that the function returns a boolean, whereas the next line says
-the result is a function. Clearly, the result of |eval0|
-can't be both of these types. 
+says that |eval0| returns a boolean, whereas the next line says
+the result is a function. All branches of a pattern-match form must
+yield values of the same type, however. 
 A related problem is the type of the environment |env|: since variables
@@
 |true|. We also note that the eval function is partial in two
-respects: first there is an inexhaustive pattern-matching in the 
+respects: first there is an inexhaustive pattern-matching 
+|match ... with UA f -> ...| in the 
 line |A (e1,e2)|.
@@
   let b (bv:bool) env  = bv
-  let varZ (vc,_)      = vc
-  let varS vp (_,envr) = vp envr
+  let varZ env         = fst env
+  let varS vp env      = vp (snd env)
   let app e1 e2 env    = (e1 env) (e2 env)
@@
 to make it possible to ``evaluate'' the same term in many different
-ways: interpretation, compilation, compute the size (or depth), etc.
+ways: interpretation, compilation, partial evaluation, size
+computation, etc.
 
@@
 
-We have also embedded our DSL in (Meta)OCaml, replacing type classes with
-functors, via a module type.  For this section, the following simple
-signature will be sufficient:
+We have also embedded our DSL in (Meta)OCaml, replacing type classes
+and instances with signatures and structures. For this and the 
+following sections, the following simple signature suffices:
 
@@
 
-This interpreter is patently tagless: the operation |add| is really
+This interpreter is patently tagless, using neither Universal type nor
+any pattern-matching: the operation |add| is really
 OCaml's addition, and |app| is OCaml's application. We can run our
@@
 In Haskell, we define a\\
-|newtype R a = R a| and |instance Symantics R|
+|newtype R a = R{unR::a}| and |instance Symantics R|
 with a straightforward implementation. We must stress that |R| is
@@
 against |R| cannot ever fail and so is operationally an identity and
-compiled that way.
-
-Note that the interpreter uses no Universal type nor pattern
-matching, and it is obvious to the compiler that non-termination can
-only result from interpreting |fix|.
+compiled that way. In both cases, the absense of pattern-match
+failures is obvious to the compiler, because there are no
+pattern-matching. The failure to yield a value
+may only result from interpreting |fix|.
 
 \begin{prop}
-If the term |e| in the meta-language has type
+If the term |e| encoded in the meta-language has type
 |t|, evaluation of that term in the interpreter R will either continue
@@
 solution in~\ref{S:PE-solution}.
-The partial evaluator (PE) is also an interpreter: it
+The partial evaluator (PE) is too an interpreter: it
 evaluates a term either to an immediate (or, `static' or present-stage)
@@
 possible computations (i.e., computations whose operarands are
-present-stage values) as the present-stage. Our first attempt is
-straightforward: we introduce
+present-stage values) as the present-stage. Our first attempt, in
+Haskell, is straightforward: we introduce
 \begin{code}
@@
 determine if they are static or dynamic. The argument of |lam| is a 
-higher-order function, which is not introspectible. One may only 
+function, which is not introspectible. One may only 
 apply |f| to a particular static or dynamic value, and receive another 
@@
 many opportunities for static computations and simplifications, such
-as multiplication by zero in the example above. Another solution is to
+as multiplication by zero in the example above. A better solution is to
 delay the binding-type-analysis (BTA) until the function |f| is
@@
 transform. When transforming abstractions, we come across
-non-factorialness. The connection with CPS is not an accident, as we
+non-functorialness. The connection with CPS is not an accident, as we
 shall see below (in Section XXX).
@@
 (implemented via typeclasses with functional dependencies). These
-methods are equivalent; here we show the GADT; please see |incope1.ml|
+methods are equivalent; here we show the GADT; please see |incope1.hs|
 in the accompanying source code for the latter.
@@
 \end{code}
-where |repr_pe| is a type function, defined inductively as 
+where |repr_pe| is a type function, defined inductively
+% inductively because P below depends on repr_pe
+as 
 \begin{code}
@@
 repr_pe Bool   = Bool
-repr_pe (a->b) = (repr_pr a) -> (repr_pe b)
+repr_pe (a->b) = P a -> P b
 \end{code}
@@
 implementation; the newly added argument |'sv| will be phantom.
-For example, we would have write, in the case of the |C| interpreter,
+For example, we would now write the interpreter |C| as
 \begin{code}
@@
 \end{code}
-When writing the partial evaluator however, we make use of that
+When writing the partial evaluator however, we make use of the
 type argument |'sv|. Here's the corresponding instance of the
@@
 
+The PE in incope is really a mixture of online and off-line; it looks 
+online, but it has a sort of "just in time" binding-time analysis built in.
+
+If we just add a method 'lift' to Symantics with a type
+|lift :: a -> repr a| (cf polymorphic lift of Xi03, of MetaML and
+MetaOCaml: CSP)
+then |Symantics repr| implies |Functor repr|. We can write then
+\begin{code}
+  instance Semantics repr => Functor repr where
+  -- fmap :: (a->b) -> repr a -> repr b
+  fmap f ra = app (lift f) ra
+\end{code}
+
+\subsection{Translucent types}
+
 We must formulate some propositions: a typed term makes progress in
@@
 
-The PE in incope is really a mixture of online and off-line; it looks 
-online, but it has a sort of "just in time" binding-time analysis built in.
-
-If we just add a method 'lift' to Symantics with a type
-|lift :: a -> repr a| (cf polymorphic lift of Xi03, of MetaML and
-MetaOCaml: CSP)
-then |Symantics repr| implies |Functor repr|. We can write then
-\begin{code}
-  instance Semantics repr => Functor repr where
-  -- fmap :: (a->b) -> repr a -> repr b
-  fmap f ra = app (lift f) ra
-\end{code}
-
-\subsection{Translucent types}
 The crucial role of the higher-order type parameter r
@@
 fragment.
+
+The attempt to encode self-interpreter brings the |*->*| polymorphism
+the the forefront, as we shall now see.
 

### 2007-03-30 tagless/tagless-final.tex
a few small changes

@@
 \begin{abstract}
-We have built a (family of) tagless partial evaluator(s) for a typed object
-language (embedded DSL), placing minimal requirements on
-the type system of the meta-language:
-we need let-bound polymorphism and staging, but not dependent
-types, generalized algebraic data types, nor a tag elimination 
-postprocessing step.
+We have built a family of tagless embeddings (interpretations) for a
+higher-order typed object
+language in a typed metalanguage (current Haskell and ML), 
+requiring neither dependent
+types, nor generalized algebraic data types, nor a tag elimination 
+postprocessing step. That has prevously considered an open problem
+(impossible). 
+Our family of evaluators (interpreters, embeddings?) contains 
+an interpreter, a compiler, a partial
+evaluator, a term-size evaluator and a CPS call-by-name transformer.
+We take advantage of staging when available.
 
 The main tool is to take functions seriously: we encode the
-higher-order abstract syntax of object terms not by using constructors for
+deBruijn or higher-order abstract syntax  of object terms 
+not by using constructors for
 an algebraic data type but rather by invoking cogen functions, which occur as
@@
 algebras for the wilder world of final algebras.
-Our family of evaluators contains an interpreter, a compiler, a partial
-evaluator, a term-size evaluator and a CPS call-by-name transformer.
+(Or: we expose the co-algebraic structure of the object language).
 Our approach highlights the importance of order-2 type constructors
-(rank-2 polymorphism), which lets us encode DSL terms with
+(rank-2 polymorphism), available with ML functors and
+Haskell98 constructor classes. This lets us encode object terms with
 translucent types: enough abstraction is available to interpret the
@@
 never get stuck.
+
+\oleg{If we talk about self-interpreter, mention let-bound polymorphism}
 \end{abstract}
@@
 another \citep{reynolds-definitional}.  The embedded language is called the
-\emph{object} or \emph{subject} language, whereas the embedding language is the
+\emph{object} or \emph{subject} language, whereas the host language is the
 \emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
 specific object languages because it supports rapid prototyping and integration
-with the host environment.  If the metalanguage supports \emph{staging}, then
+with the host environment \oleg{Cite Hudaks' survey and perhaps say a
+  couple of more words}.  If the metalanguage supports \emph{staging}, then
 the embedding can compile object programs to the metalanguage and avoid the
@@
 
-We have programmed our interpreters in OCaml/MetaOCaml (\cite{metaocaml})
-and Haskell. The complete code is available 
-\footnote{the URL will appear in the final version of the paper.}
-% connot give URL in the ICFP paper due to double-blind reviews
-in the supplemental material to the paper. We will use both
-(Meta)OCaml and Haskell for our examples throughout the paper.
-Each language has its strength and weaknesses; this dual implementation
-forced us to clarify our own ideas.  We will continually leverage the
-respective strengths of these languages, as certain of our claims are
-patently obvious in one implementation, while requiring more argumentation
-in the other.
+    At POPL 2003, Taha et al~\cite{TahaPOPL03} and Xi et al~\cite{XiPOPL03}
+both
+interpreter of a typed language in a typed language is an interesting
+problem. The known solutions include the use of the Universal type, or
+GADTs. The former is not quite satisfactory for performance reasons;
+and also because of the presence of partial pattern matching in the
+interpreter; the inexhaustiveness of the pattern matching should never
+be triggered when executing a well-typed term. Alas, this fact cannot
+be made obvious to the compiler. That also exlains unoptimality.
+
+ The problem of typed direct
+  intrepreter of a typed language is difficult and thought to require
+  GADT or dependent types. 
+  In fact, several papers introducing GADTs (Xi03 and SPJ) as well as 
+  (language in Walid03) and Concoqtion, written 2003-2007, used the 
+  typed tagless intrepreter as the motivating example.
+  Taha (in Jones optimality paper, I think) said that typed
+  interpreter of the typed language was not solved then.
+  Or, should we refer to Sec `Other solutions'.
 
@@
 
-    At POPL 2003, Taha et al~\cite{TahaPOPL03} and Xi et al~\cite{XiPOPL03}
-both
-interpreter of a typed language in a typed language is an interesting
-problem. The known solutions include the use of the Universal type, or
-GADTs. The former is not quite satisfactory for performance reasons;
-and also because of the presence of partial pattern matching in the
-interpreter; the inexhaustiveness of the pattern matching should never
-be triggered when executing a well-typed term. Alas, this fact cannot
-be made obvious to the compiler. That also exlains unoptimality.
-\oleg{Should say somewhere here that our code is surprisingly simple,
-  and, in the hindsight, obvious. The problem of typed direct
-  intrepreter of a typed language is difficult and thought to require
-  GADT or dependent types. 
-  In fact, several papers introducing GADTs (Xi03 and SPJ) as well as 
-  (language in Walid03) and Concoqtion, written 2003-2007, used the 
-  typed tagless intrepreter as the motivating example.
-  Taha (in Jones optimality paper, I think) said that typed
-  interpreter of the typed language was not solved then.
-  Or, should we refer to Sec `Other solutions'}
+We have programmed our interpreters in OCaml (and, for staging,
+MetaOCaml (\cite{metaocaml})) and Haskell. The complete code is available 
+\footnote{the URL will appear in the final version of the paper.}
+% connot give URL in the ICFP paper due to double-blind reviews
+in the supplemental material to the paper. We will use both
+(Meta)OCaml and Haskell for our examples throughout the paper.
+Each language has its strength and weaknesses; this dual implementation
+forced us to clarify our own ideas.  We will continually leverage the
+respective strengths of these languages, as certain of our claims are
+patently obvious in one implementation, while requiring more argumentation
+in the other.
+Our code is surprisingly simple, and, in the hindsight, obvious. We
+must emphasize however that direct typeful tagless interpretation of a typed
+language until now has thought to be impossible in plain ML and
+require extensions like GADT -- which we discuss in detail in Section
+XXX.
 

### 2007-03-30 tagless/tagless-final.tex
A bit more work on introduction

@@
 \usepackage{amsmath,amssymb}
+\usepackage{mathptmx}
+%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
 \usepackage{comment}
@@
 \DefineShortVerb{\|}
-\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=1em,fontsize=\small}
+\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\parindent}
 
@@
 \newcommand{\Exists}[1]{\mathopen{\exists\mathord{#1}.\,}}
-\newcommand{\cond}[3]{\mathrm{if\,}#1\mathrm{\,then\,}#2\mathrm{\,else\,}#3}
+\newcommand{\cond}[3]{\mathrm{if\ }#1\mathrm{\ then\ }#2\mathrm{\ else\ }#3}
+\newcommand{\True}{\mathinner{\mathrm{true}}}
+\newcommand{\False}{\mathinner{\mathrm{false}}}
 
@@
 A popular way to define and implement a programming language is to embed it in
-another \citep{reynolds-definitional}.  The embedded language is called the
-\emph{object} or \emph{subject} language, whereas the host language is the
+another \citep{reynolds-definitional}.  Embedding means to represent
+terms and values of the \emph{object language} as terms and values in the
 \emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
 specific object languages because it supports rapid prototyping and integration
-with the host environment \oleg{Cite Hudaks' survey and perhaps say a
-  couple of more words}.  If the metalanguage supports \emph{staging}, then
+with the host environment \citep{hudak-building} \oleg{perhaps say a
+  couple more words}.  If the metalanguage supports \emph{staging}, then
 the embedding can compile object programs to the metalanguage and avoid the
@@
         \[ \text{$b$ is a Boolean} \justifies b:\BB \]
+        \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
         \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
@@
         \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]
-        \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
     \end{proofrules}
@@
 We focus on embedding a typed object language into a typed metalanguage.
-\ccshan{Collect reasons to type here.}
-
-    At POPL 2003, Taha et al~\cite{TahaPOPL03} and Xi et al~\cite{XiPOPL03}
-both
-interpreter of a typed language in a typed language is an interesting
-problem. The known solutions include the use of the Universal type, or
-GADTs. The former is not quite satisfactory for performance reasons;
-and also because of the presence of partial pattern matching in the
-interpreter; the inexhaustiveness of the pattern matching should never
-be triggered when executing a well-typed term. Alas, this fact cannot
-be made obvious to the compiler. That also exlains unoptimality.
+Types are useful as usual because they document the programmer's
+intentions and can be checked early by machine.  \ccshan{Collect reasons
+to type here.}  To be concrete, we use the typed object language in
+Figure~\ref{fig:object} throughout this paper.  We aim not just to
+interpret the object language in the narrow sense but also for
+compilation, partial evaluation, and other processing.
+
+\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
+a typed object language in a typed metalanguage as an interesting
+problem.  The known solutions to this problem store object terms and
+values in the metalanguage in a universal type, a generalized algebraic
+datatype (GADT), or a dependent type.  In the remainder of this section,
+we discuss these solutions, identify their drawbacks, then summarize our
+contributions.  No matter how we represent the object language in the
+metalanguage, the representation can be performed by hand or more
+conveniently by a parser or preprocessor such as camlp4.  We leave aside
+the issue of whether to program a parser or preprocessor using dependent
+types \citep{WalidICFP02} or not \citep{baars-typing}.
+
+\subsection{The Tag Problem}\label{tagproblem}
+
+%\oleg{see tagless\_interp1.ml, module Tagfull for the complete code.
+%  If you change the code in here, please adjust the .ml file
+%  accordingly. Let the paper and the accompanying code be in sync.}
+
+It is straightforward to create an algebraic data type, say in OCaml, to
+represent object terms such as those in Figure~\ref{fig:object}.
+\begin{code}
+type var = VZ | VS of var
+type exp = V of var | B of bool
+         | L of exp | A of exp * exp | ...
+\end{code}
+We represent each variable using a unary de-Bruijn number.
+For example, we represent the object term $(\fun{x}x)\True$ as
+\begin{code}
+let test1 = A (L (V VZ), B true)
+\end{code}
+
+Following \citet{WalidICFP02},
+we try to implement an interpreter function |eval0|. It takes
+an object term such as |test1| above and gives us its value.
+The first argument to |eval0| is the environment, initially empty,
+which is the list of values bound to free variables in the
+interpreted code.
+\begin{code}
+let rec lookup (x::env) = function
+| VZ   -> x
+| VS v -> lookup env v
+
+let rec eval0 env = function 
+| V v       -> lookup env v
+| B b       -> b 
+| L e       -> fun x -> eval0 (x::env) e
+| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
+\end{code}
+If our OCaml-like metalanguage were untyped, the code above would be acceptable.
+The |L e| line exhibits interpretive overhead:
+|eval0| traverses the function body~|e| every time (the result of
+evaluating) |L e| is applied. Staging can be used to remove this
+interpretive overhead \cite[\S1.1--2]{WalidICFP02}.
+
+If we use OCaml or some other typed language as the meta-language, 
+the function |eval0| is ill-typed. The line |B b|
+says that |eval0| returns a boolean, whereas the next line says
+the result is a function. All branches of a pattern-match form must
+yield values of the same type, however. 
+A related problem is the type of the environment |env|: since variables
+in the language may be either of boolean or function types, a regular
+OCaml list cannot be used for holding the corresponding values.
+
+The usual solution is to introduce a Universal type (section 1.3 of
+\cite{WalidICFP02}) which contains both of these:
+
+\begin{code}
+  type u = UB of bool | UA of (u -> u)
+\end{code}
+
+We can then write an interpreter which is accepted by the typecheker
+\begin{code}
+let rec eval env = function
+  | V VZ -> List.hd env
+  | V (VS v) -> eval (List.tl env) (V v)
+  | B b -> UB b
+  | L e -> UA (fun x -> eval (x::env) e)
+  | A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
+\end{code}
+and whose inferred type is |u list -> exp -> u|. Now we can evaluate
+our sample |test1| term:
+
+\begin{code}
+let test1r = eval [] test1
+val test1r : u = UB true 
+\end{code}
+
+Unfortunately, the result is tagged: |UB true| rather than just
+|true|. We also note that the eval function is partial in two
+respects: first there is an inexhaustive pattern-matching 
+|match ... with UA f -> ...| in the 
+line |A (e1,e2)|.
+Second, the occurrences of |List.hd| and |List.tl|.
+The failure of the latter corresponds to the evaluation of `open'
+terms. One can see that if all our terms are closed, |List.hd| and
+|List.tl| above will never be applied to the empty list. The partial
+pattern match in the |A (e1,e2)| line raises an exception when we try to
+evaluate something like |A (B true, B false)|, that is when the first
+operand is not a function.
+
+Suppose our object language were typed too. Then terms like
+|A (B true, B false)| would be ill-typed. Thus, if our evaluator
+accepted only typed terms, the missing alternatives in the |A|
+line would not occur. Alas, what is obvious to us is not obvious to
+the meta-language. There is still a pattern match, and we still have
+to attach and eliminate the UB and UA tags, even when there is no
+longer a need for them. See \cite{WalidICFP02}, Section 1.4 for more details.
+
+In short: It is not satisfactory
+to use a Universal type because tagging hurts performance and
+necessitates partial pattern matching in the metalanguage, even for a
+well-typed object term.
+
+The universal type fails to express in the metalanguage
+that the object program is well-typed.  This failure gives rise to
+tagging and untagging (that is, apparently inexhaustive pattern\hyp
+matching) operations that incur a performance penalty in interpretation
+and impair optimality in partial evaluation.  In short, the
+universal\hyp type solution is not type\hyp preserving.
+
+\subsection{Other solutions}
 
@@
 
-\oleg{should say somewhere in the Introduction that we are interested
-  in the typed object language embedded in the typed
-  meta-language. Refer to Sec 2 (see the remark below) why this is
-  good. We assume the DSL setting: we enter EDSL terms in the
-  meta-language program (e.g., GHCi or (Meta)OCaml top-level). A
-  different approach is reading EDSL terms from a file. In that case, we
-  need to typecheck them first. We do not deal with that problem here,
-  because it has been solved. First we mention Walid02 where they use
-  dependent typed. However, these don't seem to be needed: refer to
-  `typing dependent typing', and the related work section.}
-\jacques{I think it is important to emphasize, probably later in the 
-paper, that we could write a camlp4 extension (or simply a parser in
-Haskell) that would transform a nice syntax for a lambda calculus into
-syntactic terms for what we present here.  Even the self-interpreter
-terms could be done this way, which would simplify things tremendously.}
+But how should one design a type system to type
+terms like |test1|? GADTs seem necessary. Similarly, how does one design a
+typesystem that can keep track of the open/closed status of terms, so that we
+can statically insure that only closed terms are passed to |eval| and that
+the empty list checking in |List.hd| and |List.tl| can be eliminated.
+The latter is an even more interesting problem requiring even
+more advanced type systems 
+(\cite{WalidPOPLo3,NanevskiICFP02,NanevskiJFP05,DaviesJACM01}
+\jacques{I hope that is the right Davies paper}
+
+This is also the main motivating problem (albeit in the context of
+a first-order language) for the introduction of GADTs 
+into Haskell \cite{PeytonJonesICFP2006}.  This was also a motivating
+factor behind the work of Xi on GADTs \cite{XiGADTs}, and some of the
+recent work on dependent types \cite{WalidICFP02,WalidConcoqtion}.
+
+See the Related Work section (\ref{related}) for a discussion of the other
+solutions to this problem.
+\oleg{I feel we need to give some details of the other solutions here, 
+to give the reader the feeling of
+how advanced solutions were proposed and how rarely they are
+implemented.}
+\jacques{I have rearranged this section so that it should be clear how
+to extend it to explain these other solutions}
+
+\subsection{The ``Final'' Proposal}\label{ourapproach}
+
 \jacques{I quite agree that the point that needs to be revisited is
@@
 XXX.
-
-\subsection{The Tag Problem}\label{tagproblem}
-
-%\oleg{see tagless\_interp1.ml, module Tagfull for the complete code.
-%  If you change the code in here, please adjust the .ml file
-%  accordingly. Let the paper and the accompanying code be in sync.}
-
-Consider a language with abstraction, application and boolean
-literals with de Bruin indices as variables. This is the language used
-by \citet{WalidICFP02}, from which we borrow the first part of 
-this example.
-
-The language can be described by the following grammar (using OCaml
-syntax)
-
-\begin{code}
-  type var = VZ | VS of var
-  type exp = V of var | B of bool | A of exp * exp | L of exp
-\end{code}
-
-with a sample term (application of the identity function to a boolean) as
-\begin{code}
-  let test1 = A ((L (V VZ)),(B true))
-\end{code}
-
-Let us attempt to implement an the interpreter, the function |eval0|. It takes
-an object language term such as |test1| and gives us its value.
-The first argument to |eval| is the environment, initially empty,
-which is the sequence of values associated with free variables in the
-interpreted code.
-\begin{code}
-  let rec eval0 env = function 
-  | V VZ      -> List.hd env
-  | V (VS v)  -> eval0 (List.tl env) (V v)
-  | B b       -> b 
-  | L e       -> fun x -> eval0 (x::env) e
-  | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
-\end{code}
-
-Suppose our meta-language (the language in which we wrote |test1|
-and that interpreter) were untyped. The above code would be acceptable.
-The |L e| line exhibits interpretive overhead \cite{WalidICFP02}: 
-|eval0 env' e| will be executed every time (the result of
-evaluating) |L e| is applied. Staging can be used to remove this
-interpretive overhead (sections 1.1 and 1.2 of \cite{WalidICFP02} explain
-this well).
-
-If we use OCaml or some other typed language as the meta-language, 
-the function |eval0| is ill-typed. The line |B b|
-says that |eval0| returns a boolean, whereas the next line says
-the result is a function. All branches of a pattern-match form must
-yield values of the same type, however. 
-A related problem is the type of the environment |env|: since variables
-in the language may be either of boolean or function types, a regular
-OCaml list cannot be used for holding the corresponding values.
-
-The usual solution is to introduce a Universal type (section 1.3 of
-\cite{WalidICFP02}) which contains both of these:
-
-\begin{code}
-  type u = UB of bool | UA of (u -> u)
-\end{code}
-
-We can then write an interpreter which is accepted by the typecheker
-\begin{code}
-let rec eval env = function
-  | V VZ -> List.hd env
-  | V (VS v) -> eval (List.tl env) (V v)
-  | B b -> UB b
-  | L e -> UA (fun x -> eval (x::env) e)
-  | A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
-\end{code}
-and whose inferred type is |u list -> exp -> u|. Now we can evaluate
-our sample |test1| term:
-
-\begin{code}
-let test1r = eval [] test1
-val test1r : u = UB true 
-\end{code}
-
-Unfortunately, the result is tagged: |UB true| rather than just
-|true|. We also note that the eval function is partial in two
-respects: first there is an inexhaustive pattern-matching 
-|match ... with UA f -> ...| in the 
-line |A (e1,e2)|.
-Second, the occurrences of |List.hd| and |List.tl|.
-The failure of the latter corresponds to the evaluation of `open'
-terms. One can see that if all our terms are closed, |List.hd| and
-|List.tl| above will never be applied to the empty list. The partial
-pattern match in the |A (e1,e2)| line raises an exception when we try to
-evaluate something like |A (B true, B false)|, that is when the first
-operand is not a function.
-
-Suppose our object language were typed too. Then terms like
-|A (B true, B false)| would be ill-typed. Thus, if our evaluator
-accepted only typed terms, the missing alternatives in the |A|
-line would not occur. Alas, what is obvious to us is not obvious to
-the meta-language. There is still a pattern match, and we still have
-to attach and eliminate the UB and UA tags, even when there is no
-longer a need for them. See \cite{WalidICFP02}, Section 1.4 for more details.
-
-In short: It is not satisfactory
-to use a Universal type because tagging hurts performance and
-necessitates partial pattern matching in the metalanguage, even for a
-well-typed object term.
-
-\subsection{Other solutions}
-
-But how should one design a type system to type
-terms like |test1|? GADTs seem necessary. Similarly, how does one design a
-typesystem that can keep track of the open/closed status of terms, so that we
-can statically insure that only closed terms are passed to |eval| and that
-the empty list checking in |List.hd| and |List.tl| can be eliminated.
-The latter is an even more interesting problem requiring even
-more advanced type systems 
-(\cite{WalidPOPLo3,NanevskiICFP02,NanevskiJFP05,DaviesJACM01}
-\jacques{I hope that is the right Davies paper}
-
-This is also the main motivating problem (albeit in the context of
-a first-order language) for the introduction of GADTs 
-into Haskell \cite{PeytonJonesICFP2006}.  This was also a motivating
-factor behind the work of Xi on GADTs \cite{XiGADTs}, and some of the
-recent work on dependent types \cite{WalidICFP02,WalidConcoqtion}.
-
-See the Related Work section (\ref{related}) for a discussion of the other
-solutions to this problem.
-\oleg{I feel we need to give some details of the other solutions here, 
-to give the reader the feeling of
-how advanced solutions were proposed and how rarely they are
-implemented.}
-\jacques{I have rearranged this section so that it should be clear how
-to extend it to explain these other solutions}
-
-\subsection{The ``Final'' Proposal}\label{ourapproach}
 

### 2007-03-31 tagless/tagless-final.tex
A little bit more work on the introduction

@@
 %\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
+\usepackage[T1]{fontenc}
 \usepackage{comment}
@@
         \[ \text{$n$ is an integer} \justifies n:\ZZ \]
-        \[ \text{$b$ is a Boolean} \justifies b:\BB \]
+        \[ \text{$b$ is a boolean} \justifies b:\BB \]
         \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
@@
 we discuss these solutions, identify their drawbacks, then summarize our
-contributions.  No matter how we represent the object language in the
+proposal and contributions.  No matter how we represent the object language in the
 metalanguage, the representation can be performed by hand or more
@@
 \end{code}
+
 If our OCaml-like metalanguage were untyped, the code above would be acceptable.
@@
 
-If we use OCaml or some other typed language as the meta-language, 
-the function |eval0| is ill-typed. The line |B b|
-says that |eval0| returns a boolean, whereas the next line says
-the result is a function. All branches of a pattern-match form must
-yield values of the same type, however. 
-A related problem is the type of the environment |env|: since variables
-in the language may be either of boolean or function types, a regular
-OCaml list cannot be used for holding the corresponding values.
-
-The usual solution is to introduce a Universal type (section 1.3 of
-\cite{WalidICFP02}) which contains both of these:
-
-\begin{code}
-  type u = UB of bool | UA of (u -> u)
-\end{code}
-
-We can then write an interpreter which is accepted by the typecheker
+However, the function |eval0| is ill-typed
+if we use OCaml or some other typed language as the metalanguage.
+The line |B b|
+says that |eval0| returns a boolean, whereas the next line |L e| says
+the result is a function, but all branches of a pattern-match form must
+yield values of the same type. 
+A related problem is the type of the environment |env|: a regular
+OCaml list cannot hold both boolean and function values.
+
+The usual solution is to introduce a universal type \citep[\S1.3]
+{WalidICFP02} containing both booleans and functions.
+\begin{code}
+type u = UB of bool | UA of (u -> u) | ...
+\end{code}
+We can then write a typed interpreter
 \begin{code}
 let rec eval env = function
-  | V VZ -> List.hd env
-  | V (VS v) -> eval (List.tl env) (V v)
+| V v       -> lookup env v
 | B b       -> UB b
 | L e       -> UA (fun x -> eval (x::env) e)
-  | A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
-\end{code}
-and whose inferred type is |u list -> exp -> u|. Now we can evaluate
-our sample |test1| term:
-
+| A (e1,e2) -> match eval env e1 with UA f ->
+               f (eval env e2)
+\end{code}
+whose inferred type is |u list -> exp -> u|. Now we can evaluate
+our sample |test1| term.
 \begin{code}
@@
 
-Unfortunately, the result is tagged: |UB true| rather than just
-|true|. We also note that the eval function is partial in two
-respects: first there is an inexhaustive pattern-matching 
-|match ... with UA f -> ...| in the 
-line |A (e1,e2)|.
-Second, the occurrences of |List.hd| and |List.tl|.
-The failure of the latter corresponds to the evaluation of `open'
-terms. One can see that if all our terms are closed, |List.hd| and
-|List.tl| above will never be applied to the empty list. The partial
-pattern match in the |A (e1,e2)| line raises an exception when we try to
-evaluate something like |A (B true, B false)|, that is when the first
-operand is not a function.
-
-Suppose our object language were typed too. Then terms like
-|A (B true, B false)| would be ill-typed. Thus, if our evaluator
-accepted only typed terms, the missing alternatives in the |A|
-line would not occur. Alas, what is obvious to us is not obvious to
-the meta-language. There is still a pattern match, and we still have
-to attach and eliminate the UB and UA tags, even when there is no
-longer a need for them. See \cite{WalidICFP02}, Section 1.4 for more details.
-
-In short: It is not satisfactory
-to use a Universal type because tagging hurts performance and
-necessitates partial pattern matching in the metalanguage, even for a
-well-typed object term.
-
-The universal type fails to express in the metalanguage
-that the object program is well-typed.  This failure gives rise to
-tagging and untagging (that is, apparently inexhaustive pattern\hyp
-matching) operations that incur a performance penalty in interpretation
-and impair optimality in partial evaluation.  In short, the
-universal\hyp type solution is not type\hyp preserving.
-
-\subsection{Other solutions}
-
- The problem of typed direct
-  intrepreter of a typed language is difficult and thought to require
-  GADT or dependent types. 
-  In fact, several papers introducing GADTs (Xi03 and SPJ) as well as 
-  (language in Walid03) and Concoqtion, written 2003-2007, used the 
-  typed tagless intrepreter as the motivating example.
-  Taha (in Jones optimality paper, I think) said that typed
-  interpreter of the typed language was not solved then.
-  Or, should we refer to Sec `Other solutions'.
-
-But how should one design a type system to type
-terms like |test1|? GADTs seem necessary. Similarly, how does one design a
-typesystem that can keep track of the open/closed status of terms, so that we
-can statically insure that only closed terms are passed to |eval| and that
-the empty list checking in |List.hd| and |List.tl| can be eliminated.
-The latter is an even more interesting problem requiring even
-more advanced type systems 
-(\cite{WalidPOPLo3,NanevskiICFP02,NanevskiJFP05,DaviesJACM01}
-\jacques{I hope that is the right Davies paper}
-
-This is also the main motivating problem (albeit in the context of
-a first-order language) for the introduction of GADTs 
-into Haskell \cite{PeytonJonesICFP2006}.  This was also a motivating
-factor behind the work of Xi on GADTs \cite{XiGADTs}, and some of the
-recent work on dependent types \cite{WalidICFP02,WalidConcoqtion}.
-
-See the Related Work section (\ref{related}) for a discussion of the other
-solutions to this problem.
-\oleg{I feel we need to give some details of the other solutions here, 
-to give the reader the feeling of
-how advanced solutions were proposed and how rarely they are
-implemented.}
-\jacques{I have rearranged this section so that it should be clear how
-to extend it to explain these other solutions}
-
-\subsection{The ``Final'' Proposal}\label{ourapproach}
+The unfortunate tag |UB| in the result reflects that |eval| is a partial
+function, in two ways.  First, the pattern match |with UA f| in the line
+|A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
+as in the ill-typed term |A (B true, B false)|.
+\begin{code}
+let test2 = A (B true, B false)
+let test2r = eval [] test2
+Exception: Match_failure in eval
+\end{code}
+Second, the |lookup|
+function assumes a nonempty environment, so |eval| can fail if we
+evaluate an open term.
+\begin{code}
+let test3 = V VZ
+let test3r = eval [] test3
+Exception: Match_failure in lookup
+\end{code}
+After all, the type |exp| represents object
+terms both well-typed and ill-typed, both open and closed.
+
+If we evaluate only closed terms that have been type-checked, then
+|eval| would never fail. Alas, this soundness is not obvious to the
+metalanguage, whose type system we must still appease with the
+nonexhaustive pattern matching in |lookup| and |eval| and the tags |UB|
+and |UA| \citep[\S1.4]{WalidICFP02}.  In other words, the algebraic data
+types above fail to express in the metalanguage that the object program
+is well-typed.  This failure necessitates tagging and nonexhaustive
+pattern\hyp matching operations that incur a performance penalty in
+interpretation and impair optimality in partial evaluation
+\citep{taha-tag}.  In short, the universal\hyp type solution is
+unsatisfactory because it does not preserve typing.
+
+\subsection{Solutions Using Fancier Types}
+
+It is commonly thought that to interpret a typed object language in
+a typed metalanguage while preserving types is difficult and requires
+GADTs or dependent types \citep{taha-tag}.  In fact, this problem
+motivated much work on GADTs \citep{xi-guarded,peyton-jones-simple} and
+on dependent types \citep{WalidICFP02,fogarty-concoqtion}.
+
+For a metalanguage's type system to allow the well-typed object term
+|test1| but disallow the ill-typed object term |test2|, fancier types
+such as GADTs or dependent types seem necessary.  Yet other type systems
+have been proposed to distinguish closed terms like |test1| from open
+terms like |test3|
+\citep{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual},
+so that |lookup| never receives an empty environment.  We discuss these
+proposals further in \S\ref{related}; here we just note that many
+advanced type systems have been devised to ensure statically that an
+object term is well-typed and closed.
+
+\subsection{Our Final Proposal}\label{ourapproach}
+
+If we change the encoding of the term from data
+constructors to functions, we can encode a tagless interpreter for a
+typed object language into a metalanguage with a very simple type system.
 
@@
 XXX.
-
-We have realized that if we change the encoding of the term from data
-constructors to functions, we can encode a tagless interpreter for a
-typed object language into a metalanguage with a very simple type system.
 
@@
 rather than de Bruijn indices.  The
-language is close to the language of \cite{XiPOPL03}, without the polymorphic
+language is close to the language of \citet{xi-guarded}, without the polymorphic
 lift but with more constants so we can write better examples, enough to

### 2007-03-31 tagless/tagless-final.tex
A bit more work on introduction

@@
 values in the metalanguage in a universal type, a generalized algebraic
-datatype (GADT), or a dependent type.  In the remainder of this section,
+data type (GADT), or a dependent type.  In the remainder of this section,
 we discuss these solutions, identify their drawbacks, then summarize our
@@
 \begin{code}
-let test3 = V VZ
+let test3 = A (L (V (VS VZ)), B true)
 let test3r = eval [] test3
@@
 pattern\hyp matching operations that incur a performance penalty in
-interpretation and impair optimality in partial evaluation
+interpretation \citep{WalidICFP02} and impair optimality in partial evaluation
 \citep{taha-tag}.  In short, the universal\hyp type solution is
@@
 
-If we change the encoding of the term from data
-constructors to functions, we can encode a tagless interpreter for a
-typed object language into a metalanguage with a very simple type system.
-
-\jacques{I quite agree that the point that needs to be revisited is
-  that a \emph{direct} interpreter for a (typed) language "cannot" be
-  done. see discussion in related work about the direct interpreter,
-  with regards to Walid's paper. Perhaps bring some of it here?}
-\jacques{I have looked at a fair bit of the literature, and I have not
-  seen this done, especially not in the typed 'online' setting.  The
-  one question that seems to be open is, how accurate can we make the
-  PE without a tough BTA?}
-Should we mention some of these points in our contributions subsection
-below?
+We represent object programs using ordinary functions rather than
+data constructors.  The interpreter below provides these functions.
+\begin{code}
+let varZ (x,env)    = x
+let varS vp (x,env) = vp env
+let b (bv:bool) env = bv
+let lam e env       = fun x -> e (x,env)
+let app e1 e2 env   = (e1 env) (e2 env)
+\end{code}
+These functions comprise the entire interpreter.
+We now represent our sample term $(\fun{x}x)\True$ as
+\begin{code}
+let testf1 = app (lam varZ) (b true)
+\end{code}
+This representation is almost the same as in \S\ref{tagproblem}, only
+written with lowercase identifiers. To evaluate an object term is to
+apply its representation to the empty environment.
+\begin{code}
+let testf1r = testf1 ()
+val testf1r : bool = true
+\end{code}
+The result has no tags---the interpreter patently uses no tags and no
+pattern matching. The term |b true| evaluates to a boolean and the term
+|lam varZ| evaluates to a function, both untagged. The |app| function
+applies |lam varZ| without pattern matching. What is more, evaluating an
+open term such as |testf3| below gives a type error rather than
+a run-time error.
+\begin{code}
+let testf3 = app (lam (varS varZ)) (b true)
+let testf3r = testf3 ()
+\end{code}
+The type error correctly complains
+that the initial environment should be a tuple rather than |()|.
+In other words, the term is open.
+
+In sum, by encoding terms using ordinary functions (the \emph{final}
+rather than \emph{initial} approach), we achieve a tagless interpreter
+for a typed object language, in a metalanguage with a very simple type
+system \citep{hindley-principal,milner-theory}.  All the run-time errors
+in \S\ref{tagproblem} (applying a nonfunction; evaluating an open term)
+are now reported at compile time. Because this new interpreter uses no
+pattern matching, it never results in a run-time error and is in fact
+total.  Because this totality is obvious not just to us but also to the
+metalanguage implementation, we avoid the serious performance penalty
+\citep{WalidICFP02} of error checking.
+
+The interpreter above is wired directly into the
+functions |b|, |lam|, and so on.  In the remainder of this paper, 
+we explain how to abstract the interpreter so as
+to ``evaluate'' the same term in many different
+ways: interpretation, compilation, partial evaluation, size
+computation, and so forth.
+
+\subsection{Contributions}\label{contributions}
 
 We have programmed our interpreters in OCaml (and, for staging,
-MetaOCaml (\cite{metaocaml})) and Haskell. The complete code is available 
+MetaOCaml \citep{metaocaml} and Haskell. The complete code is available 
 \footnote{the URL will appear in the final version of the paper.}
@@
 
-Now, let us define our object language (embedded DSL) in a different
-way. Instead of using algebraic datatypes to encode the terms of the
-language in the meta-language (reminiscent to the \emph{initial algebra} 
-approach), we will be using functions as term constructors (the \emph{final}
-approach):
-
-\begin{code}
-  let b (bv:bool) env  = bv
-  let varZ env         = fst env
-  let varS vp env      = vp (snd env)
-  let app e1 e2 env    = (e1 env) (e2 env)
-  let lam e env        = fun x -> e (x,env)
-\end{code}
-
-\noindent That is all of the interpreter. Our sample term now reads
-\begin{code}
-  let testf1 = app (lam varZ) (b true)
-\end{code}
-which is almost the same as before, only written with lower-case
-identifiers. To evaluate a term, we merely need to apply 
-the term to the empty environment
-
-\begin{code}
-  let testf1r = testf1 ()
-  val testf1r : bool = true
-\end{code}
-
-The result now has no tags. The above functions |b|,
-|varZ|, |app| constitute the evaluator. It is patent that the
-interpreter has no tags and no pattern matching. The term |b true| is
-evaluated to a boolean and a term |lam varZ| is evaluated to a
-function, with no tags attached. The application rule applies that
-function without any need for pattern-matching. The evaluator has
-another feature: evaluation of open terms leads to a type error rather
-than a run-time error. Indeed,
-
-\begin{code}
-  let testf1bad = app (lam (varS varZ)) (b true)
-  let testf1badr = testf1bad ()
-\end{code}
-gives a type error saying 
-that the initial environment should be a tuple rather than () --
-meaning the term requires a non-empty initial environment: the term is
-open.
-
-Thus all the errors that were previously reported at run-time (open terms,
-applications of non-functions) are now reported at compile
-time. Running of the interpreter will never result in run-time errors
-because its code uses no operations that could possibly raise such
-an error. In fact, our interpreter is total.
-
-Furthermore, it should be clear that we do not use any
-pattern-matching at all, and thus cannot have any pattern-matching
-errors.  This is clear not just to us, but also to the compiler.
-In other words in traditional approaches one can prove a meta-theorem
-that all pattern-matching is total, but this is nevertheless not
-apparent to the compiler, which thus still has to generate code for
-those cases.  This seriously hurts efficiency~\cite{WalidICFP02}.
-
-In the above code, the interpreter is wired in, directly built into the
-functions |b|, |app|, etc.  In the remainder of this paper, 
-we explain how to abstract the interpreter,
-to make it possible to ``evaluate'' the same term in many different
-ways: interpretation, compilation, partial evaluation, size
-computation, etc.
-
-\subsection{Contributions}\label{contributions}
+\jacques{I quite agree that the point that needs to be revisited is
+  that a \emph{direct} interpreter for a (typed) language "cannot" be
+  done. see discussion in related work about the direct interpreter,
+  with regards to Walid's paper. Perhaps bring some of it here?}
+\jacques{I have looked at a fair bit of the literature, and I have not
+  seen this done, especially not in the typed 'online' setting.  The
+  one question that seems to be open is, how accurate can we make the
+  PE without a tough BTA?}
+Should we mention some of these points in our contributions subsection
+below?
 

### 2007-03-31 tagless/tagless-final.tex
Gone over introduction

@@
+\hyphenation{meta-language meta-languages}
 \documentclass[preprint]{sigplanconf}
@@
 \usepackage{prooftree1}
+\usepackage{mdwlist}
 
@@
 \end{code}
-We represent each variable using a unary de-Bruijn number.
+We represent each variable using a unary de Bruijn index.
 For example, we represent the object term $(\fun{x}x)\True$ as
@@
 
-In sum, by encoding terms using ordinary functions (the \emph{final}
-rather than \emph{initial} approach), we achieve a tagless interpreter
-for a typed object language, in a metalanguage with a very simple type
-system \citep{hindley-principal,milner-theory}.  All the run-time errors
-in \S\ref{tagproblem} (applying a nonfunction; evaluating an open term)
-are now reported at compile time. Because this new interpreter uses no
+In sum, by Church\hyp encoding terms using ordinary functions,
+we achieve a tagless interpreter
+for a typed object language in a metalanguage with a simple type
+system \citep{hindley-principal,milner-theory}.  In this \emph{final}
+rather than \emph{initial} approach, both kinds of run-time errors
+in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
+term) are reported at compile time. Because the new interpreter uses no
+universal type or
 pattern matching, it never results in a run-time error and is in fact
@@
 \citep{WalidICFP02} of error checking.
+\ccshan{Need to cite Thiemann here and in related work?}
 
 The interpreter above is wired directly into the
-functions |b|, |lam|, and so on.  In the remainder of this paper, 
+functions |b|, |lam|, |app|, and so on.  In the remainder of this paper, 
 we explain how to abstract the interpreter so as
@@
 
+The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
+free in the encoding of an object term such as |testf1| above.  Defining
+these functions differently gives rise to different interpreters, that
+is, different folds on object programs.  Given the same term
+representation but varying the interpreter, we can
+\begin{itemize*}
+    \item evaluate the term to a value in the metalanguage;
+    \item measure the size or depth of the term;
+    \item compile the term, with staging support such as in MetaOCaml;
+    \item even partially evaluate the term, online; and
+    \item transform the term to continuation\hyp passing style (CPS),
+        including call-by-name (CBN).
+\end{itemize*}
 We have programmed our interpreters in OCaml (and, for staging,
-MetaOCaml \citep{metaocaml} and Haskell. The complete code is available 
-\footnote{the URL will appear in the final version of the paper.}
+\citet{metaocaml}) and standard Haskell. The complete code is available%
+\footnote{The URL will appear in the final version of the paper.}
 % connot give URL in the ICFP paper due to double-blind reviews
-in the supplemental material to the paper. We will use both
-(Meta)OCaml and Haskell for our examples throughout the paper.
-Each language has its strength and weaknesses; this dual implementation
-forced us to clarify our own ideas.  We will continually leverage the
-respective strengths of these languages, as certain of our claims are
-patently obvious in one implementation, while requiring more argumentation
-in the other.
-Our code is surprisingly simple, and, in the hindsight, obvious. We
-must emphasize however that direct typeful tagless interpretation of a typed
-language until now has thought to be impossible in plain ML and
-require extensions like GADT -- which we discuss in detail in Section
-XXX.
-
-\jacques{I quite agree that the point that needs to be revisited is
-  that a \emph{direct} interpreter for a (typed) language "cannot" be
-  done. see discussion in related work about the direct interpreter,
-  with regards to Walid's paper. Perhaps bring some of it here?}
-\jacques{I have looked at a fair bit of the literature, and I have not
-  seen this done, especially not in the typed 'online' setting.  The
-  one question that seems to be open is, how accurate can we make the
-  PE without a tough BTA?}
-Should we mention some of these points in our contributions subsection
-below?
-
-   We have realized that if we change the encoding of the term --
-from data constructors to functions, we can encode a tagless
-interpreter of a typed language in a typed language with no GADTs and
-a very simple type system. The interpreter uses no universal type, no
-pattern matching and it is patently obvious to the compiler that the
-only non-termination can come from interpreting the fixpoint
-combinator |fix|. If the source
-language is strongly normalizing, our interpreter is total.
-
-The term ``constructor'' functions lam, app, etc. look like free
-variables in the encoding of an object term.  Defining these functions
-differently gives rise to different interpreters.  Given the same term
-representation but varying the interpreters, we can
-	- evaluate the term to a value in the metalanguage
-	- determine the size or depth of a term
-    - compute a \emph{CBN} CPS transformation of a term
-	- `compile' the term. This requires staging, as in MetaOCaml.
-	- Surprisingly, it is also possible to write a partial evaluator!
-
-\noindent We have implemented the above in (Meta)OCaml and Haskell.
-We view our contributions as:
+to supplement the paper. Our examples below switch between (Meta)OCaml
+and Haskell even though we have implemented each example equivalently in
+both metalanguages, because some of our claims are more obvious in one
+metalanguage than the other.  For example, our OCaml evaluator obviously
+terminates except if the fixpoint combinator $\mathrm{fix}$ is present
+in the object program.
+
+Our code ensures statically that an object term is well-typed and
+closed, yet is surprisingly simple and obvious in hindsight.  We stress
+that until now our achievement has been thought impossible: to
+interpret a typed object language in a typed metalanguage without
+tagging or type\hyp system extensions.  For example, \citet{taha-tag}
+states that ``expressing such an interpreter in a statically typed
+programming language is a rather subtle matter. In fact, it is only
+recently that some work on programming type-indexed values in ML
+\citep{yang-encoding} has given a hint of how such a function can be
+expressed.''  We further discuss related work in~\S\ref{related}.
+
+Our contributions are
 \begin{enumerate}
-\item we interpret/compile/partially-evaluate a typed source language
-   from within a typed language.  The requirements of the host type
-   system are \jacques{not so onerous?  simple?},
-\item a simple, clean implementation,
-\item a nicely comparable Haskell/MetaOCaml version,
-\item a functorial approach which deals with uniformly with all
-evaluators, from interpretation to partial evaluation to CBN CPS transform,
-\item clear and easy extensibility to more features,
-\item only well-typed and closed terms can be evaluated, and this is 
-checked statically,
-\item and an approach to self-interpretation which is compatible with the
-  above.  Self-interpretation turned out to be considerably harder than
-  we at first expected.
+\item to evaluate, compile, and partially evaluate a typed object language
+   in a typed metalanguage with a simple type system;
+\item to check statically for well-typed and closed object programs;
+\item clean, comparable implementations in MetaOCaml and Haskell;
+\item a functor signature that encompasses all our interpreters, from
+    evaluation and compilation to partial evaluation and CBN CPS
+    transformation;
+\item a clear way to extend the object language with more features;
+\item and an approach to self-interpretation compatible with the
+  above.  Self-interpretation turned out to be harder than expected.
 \end{enumerate}
-
-We want to reiterate that it was previously thought that insuring
-that types are well-typed \emph{or} closed required advanced type
-extensions, and we show that this can be done in a simple subset
-of both OCaml and Haskell.
-
 \jacques{can we state fairly clearly what type system we need?  I mean,
@@
 
-\section{Base language and its tagless interpreters}\label{language}
-
-The language we will embed is a simply-typed lambda-calculus
-with fixpoint, integers, booleans and comparison, whose BNF grammar is
-
-\begin{code}
-  e ::= Lam hoas_fn | App e e | Fix hoas_fn |
-  I int_literal | B bool_literal | Add e1 e2 | 
-  Mul e1 e2 | Leq e1 e2 | IF b e-then e-else
-\end{code}
-
-This time we use higher-order abstract syntax \cite{DaleMiller}
-rather than de Bruijn indices.  The
-language is close to the language of \citet{xi-guarded}, without the polymorphic
-lift but with more constants so we can write better examples, enough to
-express Fibonacci, factorial, and power.
+\section{The base language and its tagless interpreters}\label{language}
+
+Figure~\ref{fig:object} shows our object language, a simply-typed
+$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
+The language is close to \citets{xi-guarded}, without their polymorphic
+lift but with more constants so as to express Fibonacci, factorial, and
+power.
+
+In the sequel, we encode binding in object programs using higher-order
+abstract syntax \citep{miller-manipulating,pfenning-higher-order}
+rather than de Bruijn indices.
 

### 2007-04-01 tagless/tagless-final.tex
CBN CPS done

@@
 \subsection{Two tagless interpreters}
+\label{S:interpreter-RL}
 The first instantiation of the Symantics structure is an interpreter:
@@
 \jacques{Comment on online/offline here}
-
-\section{Variations and discussion}\label{discussion}
-
-state and imperative features
-
-CPS transformation (interpreting/compiling the term to CBN CPS)
-
-It's relation to PE and CPS. Currently, our CPS is CBN. Need to
-investigate CBV.
-The type-level function |'da -> 'sa| is important because it gives 
-the CPS transform. So, the type-function
-that enables the PE is exactly the same type function that enables
-CPS....
-
 The PE in incope is really a mixture of online and off-line; it looks 
@@
 
+\section{Variations and discussion}\label{discussion}
+
+\oleg{where to put the following? Should we even mention this?
+If not, we should still keep this in comments.}
 If we just add a method 'lift' to Symantics with a type
@@
 \end{code}
+
+\subsection{Call-by-name CPS interpreters}
+
+The object language generally inherits the evaluation strategy from
+the metalanguage. Evaluating an application |e1 e2| in OCaml requires
+the evaluation of |e2| before performing the application. Therefore,
+when evaluating the encoding of the object language application 
+|app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
+possible then to represent a call-by-name (CBN) object language in a
+call-by-value (CBV) metalanguage. The answer, given back by Reynolds
+and Plotkin (cite) is CPS, which was specifically introduced by them
+to make the evaluation strategy of the definitional interpreter
+independent of the strategy of the metalanguage. We demonstrate the
+same independence in the typed setting \oleg{mention the quote from
+  Reynolds that typed setting would be nice?}. Specifically, we
+demonstrate a CBN evaluation of our object language in OCaml, which is
+CBV. We do so by building a CBN CPS interpreter for the object
+language.
+
+The CBN CPS interpretation was introduced by Plotkin (cite). As in the
+more familiar CBV CPS, the interpretation of a value is a function
+accepting the continuation argument |k| and yielding the answer
+returned by that continuation.
+\begin{code}
+let int (x:int) = fun k -> k x
+let add e1 e2 = 
+  fun k -> e1 (fun v1 -> e2 (fun v2 -> k (v1+v2)))
+\end{code}
+In the both cases, the returned value has the type 
+|(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.
+
+The interpretation of abstraction and application in CBN CPS is
+different from CBV CPS and is as follows:
+\begin{code}
+let lam f = fun k -> k f
+let app e1 e2 = fun k -> e1 (fun f -> (f e2) k)
+\end{code}
+In the tell-tale sign of CBN, in interpreting the application
+we do not `evaluate' the argument e2, that is, we do apply it to the
+continuation. Rather, we pass the unevaluated |e2| to the abstraction.
+The result of |lam (fun x -> add x (int 1))| has the type
+|((((int->'w1)->'w1) -> ((int->'w1)->'w1)) -> 'w2) -> 'w2|.
+We want to collect the above interpretation functions into a module
+with the Symantics signature, so to fit the CBN CPS interpreter within our
+general framework. Alas, we encounter the same problem we had in
+Section~\ref{S:PE-problem}, in writing a type equation for the |repr|
+type that can represent both the type of |int i| and the type of 
+|lam f|. We cannot say that the the result of interpreting an object
+expression of the type |t| is |(t->'w)->'w| because if |t| happens to
+be an arrow type such as |int->int| in the above example, the result
+type obviously does not fit the desired pattern. We need a type
+function with a type-case distinction. In Section~\ref{S:PE-solution} we have
+shown the way to encode such a function, using the additional type
+argument |'sv| of the |repr| type. Furthermore, we notice that the type 
+function |repr_pe| needed for the partial evaluator,
+Section~\ref{S:PE-solution}, is precisely the same type function we
+need for CBN CPS. This gives us for the following CBN CPS interpreter,
+of the Symantics signature:
+
+\begin{code}
+module RCN = struct
+  type ('c,'sv,'dv) repr = 
+     {ko: 'w. ('sv -> 'w) -> 'w}
+  let int (x:int) = {ko = fun k -> k x}
+  let add e1 e2 = 
+    {ko = fun k -> e1.ko (fun v1 -> 
+                   e2.ko (fun v2 -> k (v1+v2)))}
+  let if_ eb et ee = 
+    {ko = fun k -> eb.ko 
+         (fun vb -> if vb then (et ()).ko k 
+                    else (ee ()).ko k)}
+  ...
+  let lam f = {ko = fun k -> k f}
+  let app e1 e2 = 
+     {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
+  let fix f = 
+     let rec fx f n = app (f (lam (fx f))) n in 
+     lam(fx f)
+
+  let get_res x = x.ko (fun v -> v)
+end
+\end{code}
+
+We have made our interpreter fully polymorphic over the answer type.
+We could have just as well made |RCN| a functor, parameterized over
+the answer type -- as is common when writing CPS in SML. In the
+present case, we chose to use higher-rank polymorphism of OCaml record
+types.
+
+Because |RCN| has the Semantics signature, we can instnatiate with it and run
+all of our previous examples. A more interesting is the following
+example.
+\begin{code}
+module EXS(S: Symantics) = struct
+ open S
+ let diverg () = 
+   app (lam (fun x -> int 1)) 
+       (app (fix (fun self -> self)) (int 2))
+end
+\end{code}
+
+If we interpret |EXS| with the |R| interpreter of
+Section~\ref{S:interpreter-RL} realized in OCaml
+\begin{code}
+let module M = EXS(R) in M.diverg ();;
+\end{code}
+we never get any result because of looping. In contrast, if we use
+the CBN interpreter
+\begin{code}
+let module M = EXS(RCN) in RCN.get_res (M.diverg ())
+\end{code}
+we get the result |1|.
+
+\oleg{mention some formal results? Like indifference theorems by
+  Plotkin? And say we generalized their result to the type setting?
+  Mention this in the conclusions?}
+
+\subsection{CBV CPS transformers}
+
+We can obtain CBV from CBN?
+
+\subsection{State and imperative features}
 

### 2007-04-01 tagless/tagless-final.tex
CBV CPS done

@@
 
-We can obtain CBV from CBN?
+We can easily change our CBN CPS interpreter into a CBV CPS
+interpreter, by replacing only one definition:
+\begin{code}
+module RCV = struct
+  include RCN
+  let lam f = {ko = 
+    fun k -> k (fun e -> e.ko 
+         (fun v -> f {ko = fun k -> k v}))}
+end;;
+\end{code}
+Now when we apply an abstraction to an argument |e|, we always force
+the evaluation of that argument before proceeding. This approach is in
+line with Reynolds (cite) (albeit in a typed setting). The interpreter RCV is
+useful for achieving the CBV evaluation of the object language
+no matter whether the metalanguage is CBV or CBN.
+
+In this section however we show a more general approach to CBV CPS:
+CPS transformer. We take any implementation of Symantics (interpreter,
+compiler, partial evaluator) and make it evaluate the CPS language
+instead. We can view this transformation of interpreters as a
+transformation on the object language, from direct to
+continuation-passing style.
+
+The transformation is canonical, as the following (abbreviated) code
+shows
+\begin{code}
+module CPST(S: Symantics) = struct
+  let int i = S.lam (fun k -> S.app k (S.int i))
+  let add e1 e2 = S.lam (fun k ->
+    S.app e1 (S.lam (fun v1 ->
+    S.app e2 (S.lam (fun v2 -> S.app k (S.add v1 v2))))))
+  let if_ ec et ef = S.lam (fun k ->
+    S.app ec (S.lam (fun vc ->
+    S.if_ vc (fun () -> S.app (et ()) k) (fun () -> S.app (ef ())
+    k))))
+  ...
+  let lam f = S.lam (fun k -> S.app k (S.lam (fun x ->
+    f (S.lam (fun k -> S.app k x)))))
+  let app e1 e2 = S.lam (fun k -> 
+    S.app e1 (S.lam (fun f ->
+    S.app e2 (S.lam (fun v -> S.app (S.app f v) k)))))
+  let fix f  = S.fix (fun self -> (f self))
+end
+\end{code}
+The code literally establishes a mapping from CPS interpretations to
+(direct) interpretations, that latter performed by 
+the base interpreter |S|.
+
+A reader may notice that the above module does not define the type
+equation for |repr| and so it does not have the signature Symantics.
+The reason for that is again the result type of |lam f|. Whereas
+|int i| or |add e1 e2| have the (abbreviated) 
+type |('c,..., (int -> 'w) -> 'w) S.repr|,
+the result of |lam (fun x -> add x (int 1))| has the type
+|('c,...((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr|. 
+Hence, to write the type equation for |CPST.repr| we need a 
+type case-discriminating
+function similar to the type function |repr_pe| of
+Section~\ref{S:PE-solution}. Alas, the function we need here is
+similar but not identical to |repr_pe|. Thus we need to add 
+another type variable to the |repr| type and modify the Symantics
+signature accodringly. Again, the terms in previous Symantics module
+stay unchanged, but the |repr| type equations in those structures have to
+account for a new (phantom, in that case) type variable.
+
+To save space, we show a simplified and less principal approach that
+lets us use the module |CPST| as it is. Because the module does not
+have the signature Symantics, we cannot apply the |EX| functor to it.
+But we can write the tests nevertheless:
+\begin{code}
+module T = struct
+ module M = CPST(C)
+ open M
+ let test1 () = app (lam (fun x -> x)) (bool true)
+ let testpowfix () = ... (* the same as before *)
+ let testpowfix7 = (* the same as before *)
+    lam (fun x -> app (app (testpowfix ()) x) (int 7))
+end;;
+\end{code}
+That is, we instantiate |CPST| with the desired base intrepreter (the
+compiler |C| in the case above) so that we can use the result |M| to
+interpret object level terms. The sample terms are \emph{exactly}
+as we used before. We just copied them over. That copying step is the
+price we pay for the simplified treatment. We shall encounter this
+copying when discussing the self-interpreter in the next section. We
+will see then that the copying step has a profound meaning of
+`plugging the hole', which is one of the faces of polymorphism.
+
+It is instructive to see the result of evaluating a few examples. With 
+|CPST| instantiated with a straightforward compiler |C| as above, our
+running test |T.test1 ()| gives
+\begin{code}
+.<fun x_5 -> ((fun x_2 -> 
+        (x_2 (fun x_3 -> fun x_4 -> (x_4 x_3))))
+        (fun x_6 -> ((fun x_1 -> (x_1 (true))) 
+        (fun x_7 -> ((x_6 x_7) x_5)))))>.
+\end{code}
+
+which shows the naive CPS transform of the original code. The result is
+quite unoptimal with several apparent $\beta$-redexes. We can reduce
+them by instantiating |CPST| with the |P| evaluator instead. Our
+running test gives then
+\begin{code}
+  P.repr = {P.st = Some <fun>; P.dy = .<fun x_5 -> (x_5 (true))>.}
+\end{code}
 
 \subsection{State and imperative features}
+
+We can easily modify the CBV CPS transform to pass a piece of state
+along with the continuation. This lets us support mutable state. We
+can also use reference variables of the metalanguage to emulate
+reference variables in the object language.
+\oleg{Should we have an example in the code?}
 

### 2007-04-01 tagless/tagless-final.tex
Went over Section 2-3

@@
-\hyphenation{meta-language meta-languages}
+\hyphenation{meta-language meta-languages Meta-OCaml}
 \documentclass[preprint]{sigplanconf}
@@
 Types are useful as usual because they document the programmer's
-intentions and can be checked early by machine.  \ccshan{Collect reasons
-to type here.}  To be concrete, we use the typed object language in
+intentions and can be checked early by machine (see~\S\ref{encoding}).
+To be concrete, we use the typed object language in
 Figure~\ref{fig:object} throughout this paper.  We aim not just to
@@
 both metalanguages, because some of our claims are more obvious in one
-metalanguage than the other.  For example, our OCaml evaluator obviously
-terminates except if the fixpoint combinator $\mathrm{fix}$ is present
-in the object program.
+metalanguage than the other.  For example, MetaOCaml provides
+convenient, typed staging facilities.
 
@@
 
-\section{The base language and its tagless interpreters}\label{language}
+\section{The object language and its tagless interpreters}\label{language}
 
@@
 \label{encoding}
-We have embedded the above language in (Meta)OCaml and Haskell.  In Haskell,
-the constructors of DSL terms are functions defined as methods in a type class
-Symantics. The name implies that the class interface gives the syntax for
-the source language and its instances give the semantics.
-
+
+We embed this language in (Meta)OCaml and Haskell.  In Haskell,
+the functions that construct object terms are methods in a type class
+|Symantics|. The class is so named because its interface gives the syntax for
+the object language and its instances give the semantics.
+The parameter |repr| to the class has the kind |* -> *|.
 \begin{code}
 class Symantics repr where
-  int :: Int   -> repr Int          -- int literal
-  bool :: Bool -> repr Bool         -- bool literal
-
-  lam :: (repr a -> repr b) -> repr (a->b)
-  app :: repr (a->b) -> repr a -> repr b
+
+  int  :: Int  -> repr Int
+  bool :: Bool -> repr Bool
+
+  lam :: (repr a -> repr b) -> repr (a -> b)
+  app :: repr (a -> b) -> repr a -> repr b
   fix :: (repr a -> repr a) -> repr a
@@
   mul :: repr Int -> repr Int -> repr Int
+  leq :: repr Int -> repr Int -> repr Bool
   if_ :: repr Bool -> repr a -> repr a -> repr a
-  leq :: repr Int  -> repr Int -> repr Bool
-\end{code}
-
-\noindent |test1| now reads |app (lam (\x -> x)) (bool True)|,
+\end{code}
+For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
+\S\ref{tagproblem} above as |app (lam (\x -> x)) (bool True)|,
 whose inferred type is |Symantics repr => repr Bool|.
-For example, we can encode the classical |power| function, as well
-as the partial application |\x -> power x 7| as
-
+For another example, the classical $\mathit{power}$ function is
 \begin{code}
@@
       (mul x (app self (add n (int (-1))))))))
+\end{code}
+and the partial application $\fun{x} \mathit{power}\;x\;7$ is
+\begin{code}
 testpowfix7 () = 
-  lam (\x -> app (app (testpowfix ()) x) (int 7))
-\end{code}
-The dummy unit |()| argument above is to avoid the monomorphism
-restriction and so that the type of |testpowfix| and |testpowfix7|
-will remain polymorphic in |repr|. Instead of supplying this dummy
+  lam (\x -> app (app (testpowfix ()) x)
+                 (int 7))
+\end{code}
+The dummy argument |()| above is to avoid the monomorphism
+restriction, to keep the type of |testpowfix| and |testpowfix7|
+polymorphic in |repr|. Instead of supplying this dummy
 argument, we could have given the terms explicit polymorphic
-signature.  We however prefer if the typechecker of the meta-language
-(Haskell) infers the types of the object terms for us. We could also
-avoid the |()| argument by switching off the monomophism restriction
+signatures.  We however prefer for
+Haskell to infer the object types for us. We could also
+avoid the dummy argument by switching off the monomophism restriction
 with a compiler flag.
 
-Note that the object language is typed. If we make a mistake
-(for example, replace |int 7| with |bool True| in |testpowfix7|, the
-compiler will highlight that term and complain that it couldn't match
-the expected |Int| against the inferred |Bool|. A type error
-also results when entering |lam (\x -> app x x)|, with the compiler
-complaining about an infinite type. The type-checker of the
-meta-language will flag meaningless object level terms --
-before any of their evaluation. If we forget, for example, |app| in
-some of the terms above, we get an error message with a precise
-location. So the type-checker also flags syntactically invalid
-object-level terms.\oleg{Should we perhaps move this up, in the
-  intro? To justify why typed-metalanguages and the typed DSL are
-  good. Or keep the explanation here but refer to it from the Introduction.}
-
-We have also embedded our DSL in (Meta)OCaml, replacing type classes
-and instances with signatures and structures. For this and the 
-following sections, the following simple signature suffices:
-
+Because the types in the |Symantics| class reflect the object language's type
+system, this representation admits only well-typed object terms for evaluation.
+If we make a mistake somewhere, say replace |int 7| with |bool True| in
+|testpowfix7|, Haskell will complain there that the expected type |Int| does not
+match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
+encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
+Haskell's type checker also flags syntactically invalid object terms, such as if
+we forget |app| somewhere above.
+
+\begin{figure*}
+\begin{floatrule}
 \begin{code}
 module type Symantics = sig
-  type ('c,'dv) repr
-  val int : int  -> ('c,int) repr
-  val bool: bool -> ('c,bool) repr
-  val add : ('c,int) repr-> ('c,int) repr-> ('c,int) repr
-  val mul : ('c,int) repr-> ('c,int) repr-> ('c,int) repr
-  val leq : ('c,int) repr-> ('c,int) repr-> ('c,bool) repr
-  val if_ : ('c,bool) repr ->
-             (unit -> ('c,'da) repr) ->
-             (unit -> ('c,'da) repr) -> ('c,'da) repr 
-  val lam : (('c,'da) repr -> ('c,'db) repr) 
-          -> ('c,'da->'db) repr
-  val app : ('c,'da->'db) repr
-    -> ('c,'da) repr -> ('c,'db) repr
-  val fix : (('c,'da->'db) repr -> ('c,'da->'db) repr) 
-            -> ('c,'da->'db) repr
-end;;
-\end{code}
-
-Some differences are in the additional type parameter |c| (the
-environment classifier from \cite{WalidPOPL03}) to be used later when we get to
-code generation (this is required by MetaOCaml as part of the type for the
-code). MetaOCaml is a call-by-value language, and thus we have to 
-$\eta$-expand
-the fix-point combinator and encapsulate the branches of the |if_|
-into thunks to prevent their evaluation.
-
-The power example and our sample term |test1| are encoded as
+
+  type ('c, 'dv) repr
+
+  val int : int  -> ('c, int) repr
+  val bool: bool -> ('c, bool) repr
+
+  val lam : (('c, 'da) repr -> ('c, 'db) repr) -> ('c, 'da -> 'db) repr
+  val app : ('c, 'da -> 'db) repr -> ('c, 'da) repr -> ('c, 'db) repr
+  val fix : (('c, 'da -> 'db) repr -> ('c, 'da -> 'db) repr) -> ('c, 'da -> 'db) repr
+
+  val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
+  val mul : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
+  val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
+  val if_ : ('c, bool) repr -> (unit -> ('c, 'da) repr) -> (unit -> ('c, 'da) repr) -> ('c, 'da) repr
+
+end
+\end{code}
+\end{floatrule}
+\caption{A simple (Meta)OCaml embedding of our object language}
+\label{fig:ocaml-simple}
+\end{figure*}
+
+To embed the same object language in (Meta)OCaml, we replace the type
+class |Symantics| and its instances by a module signature |Symantics|
+and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
+signature that suffices for this and the following sections.
+Two differences between this signature and the |Symantics| class in
+Haskell are: the additional type parameter |c|, an
+environment classifier \citep{WalidPOPL03} required by MetaOCaml for
+code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
+|fix| and thunk types in |if_| due to OCaml's being a call-by-value
+language.
+
+The functor below encodes our examples |test1| and $\mathit{power}$.
 \begin{code}
@@
   open S
- let test1 () = app (lam (fun x -> x)) (bool true)
+  let test1 () =
+    app (lam (fun x -> x)) (bool true)
   let testpowfix () = 
-   lam (fun x ->fix (fun self -> lam (fun n ->
+    lam (fun x -> fix (fun self -> lam (fun n ->
     if_ (leq n (int 0)) (fun () -> int 1)
-         (fun () -> mul x (app self 
-                               (add n (int (-1))))))))
+        (fun () -> mul x
+          (app self (add n (int (-1))))))))
   let testpowfix7 = 
-    lam (fun x -> app (app (testpowfix ()) x) (int 7))
+     lam (fun x -> app (app (testpowfix ()) x)
+                       (int 7))
 end;;
 \end{code}
-The dummy unit argument in |testpowfix| is an artefact of MetaOCaml
-(which we will be using to compile our tests in Sections below),
-again related to monorphism. To be
-more precise, the environment qualifier (which appears as the |'c|
-type parameter in the signature of Symantic functions) must remain
-uninstantiated in order for us to `run' the code. That means the type
-must be polymorhic. The value restriction then dictates that
+The dummy argument to |test1| and |testpowfix| is an artefact of
+MetaOCaml, again related to monomorphism: in order for us to run a
+piece of generated code, it must be polymorphic in its environment
+classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
+The value restriction then dictates that
 the definitions of our object terms must look syntactically like
@@
 
-Thus the object expressions (expressions of the embedded DSL) have in
-OCaml the type of a functor from |Symantics| to the desired value. This
-is essentially the same as the |Symantics repr => | constraint in the case of
-the embedding of the DSL in Haskell.
+Thus we represent an object expression in
+OCaml as a functor from |Symantics|. This
+is essentially the same as the constraint |Symantics repr =>| in the
+Haskell embedding.
 
@@
 \label{S:interpreter-RL}
-The first instantiation of the Symantics structure is an interpreter:
-evaluating the object (DSL) term to its value in the meta-language. 
-The following module essentially interprets (gives one-to-one
-correspondence to) each operation of the object language as the
-corresponding operation in the meta-language.
+
+Having abstracted our term representation over the interpreter, we are
+now ready to present a series of interpreters.  Each interpreter is an
+instance of the |Symantics| class in Haskell and a module implementing
+the |Symantics| signature in MetaOCaml.
+
+The first interpreter evaluates the object term to its value in the
+metalanguage.  The module below interprets each object\hyp language
+operation as the corresponding metalanguage operation.
 \begin{code}
 module R = struct
-  type ('c,'dv) repr = 'dv (* absolutely no wrappers *)
+
+  (* absolutely no wrappers *)
+  type ('c,'dv) repr = 'dv
+
   let int  (x:int)  = x
   let bool (b:bool) = b
+
+  let lam  f        = f
+  let app  e1 e2    = e1 e2
+  let fix  f        =
+    let rec self n = f self n in self
+
   let add  e1 e2    = e1 + e2
@@
   let leq  x y      = x <= y
-  let if_  eb et ee = if eb then (et ()) else (ee ())
-
-  let lam f         = f
-  let app e1 e2     = e1 e2
-  let fix f         = let rec self n = f self n in self
-end;;
-\end{code}
-
-This interpreter is patently tagless, using neither Universal type nor
-any pattern-matching: the operation |add| is really
-OCaml's addition, and |app| is OCaml's application. We can run our
-examples by instatiating the |EX| functor with |R|.
-\noindent
-\begin{code}
-  module EXR = EX(R);;
-\end{code}
-
-\noindent and so |EXR.test1 ()| evaluates to the untagged boolean value |true|.
-In Haskell, we define a\\
-|newtype R a = R{unR::a}| and |instance Symantics R|
-with a straightforward implementation. We must stress that |R| is
-a |newtype|. Although it may look like a tag, it has no run-time
-representation (erased during compilation) and pattern-matching
-against |R| cannot ever fail and so is operationally an identity and
-compiled that way. In both cases, the absense of pattern-match
-failures is obvious to the compiler, because there are no
-pattern-matching. The failure to yield a value
-may only result from interpreting |fix|.
-
+  let if_  eb et ee = if eb then et () else ee ()
+
+end
+\end{code}
+
+As in~\S\ref{ourapproach},
+this interpreter is patently tagless, using neither a universal type nor
+any pattern matching: the operation |add| is really
+OCaml's addition, and |app| is OCaml's application. To run our
+examples, we instantiate the |EX| functor from~\S\ref{encoding} with this |R| module.
+\begin{code}
+module EXR = EX(R)
+\end{code}
+Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
+In Haskell, we define
+\begin{code}
+newtype R a = R {unR::a}
+instance Symantics R where ...
+\end{code}
+Although |R| looks like a tag, it is only
+a |newtype|.  The types |a| and |R a| are represented differently
+only at compile time, not at run time.  Pattern matching against~|R|
+cannot ever fail and is compiled to no-op.
+In both OCaml and Haskell, it is obvious to the compiler that
+pattern matching cannot fail, because there is no
+pattern matching. Evaluation can only fail to yield a value
+due to interpreting |fix|.
+That is, the proposition below follows immediately from the soundness of
+the metalanguage's type system.
 \begin{prop}
-If the term |e| encoded in the meta-language has type
-|t|, evaluation of that term in the interpreter R will either continue
-indefinitely or termninate with a value of the same type |t|.
+If the term |e| encoded in the metalanguage has type~|t|,
+then evaluating~|e| in the interpreter~|R| either continues
+indefinitely or terminates with a value of the same type~|t|.
 \end{prop}
 
-We'd like show a different interpreter: it interprets each terms of
-the object language to its size (defined as the number of term
-constructors). The following is slightly abbreviated code (see the
+For variety, we show another interpreter, which measures the \emph{size}
+of each object term, defined as the number of term
+constructors. The following is slightly abbreviated code (see the
 accompanying source code for the complete definition).
-
 \begin{code}
@@
   let int (x:int)  = 1
+  let lam f        = f 0 + 1
+  let app e1 e2    = e1 + e2 + 1
+  let fix f        = f 0 + 1
   let mul e1 e2    = e1 + e2 + 1
   let if_ eb et ee = eb + et () + ee () + 1
-  let lam f        = (f 0) + 1
-  let app e1 e2    = e1 + e2 + 1
-  let fix f        = (f 0) + 1
 end;;
 \end{code}
-
-\noindent Now the expression
+Now the expression
 \begin{code}
@@
 \end{code}
-evaluates to |3|. This interpreter is not only tagless, it is also
-total. It ``evaluates'' even seemingly divergent terms like\\
-|app (fix (fun self -> self)) (int 1)|.
+evaluates to |3|. This interpreter is not only tagless but also
+total. It ``evaluates'' even seemingly divergent terms like
+\begin{code}
+app (fix (fun self -> self)) (int 1)
+\end{code}
 
@@
 
-\section{Tagless compiler (aka, staged interpreter)}\label{compiler}
-
-We can also write a ``compiler'' from our lambda calculus directly
-into OCaml code, using MetaOCaml's staging facilities. Namely,
-future-stage expressions of the type |t| are represented at the
-present stage as values of the type |('a,t) code| where |'a| is the
-environment classifier \cite{WalidPOPL03,WalidESOP04}. Code values are created
-by a \emph{bracket} form: |.<e>.| specifies the expression |e| is to be
-evaluated at a future stage. The \emph{escape} |.~e|, which must occur
-within a bracket, specifies that the expression |e| must be evaluated
+\section{A tagless compiler (or, a staged interpreter)}\label{compiler}
+
+We can also compile our object language
+into OCaml code using MetaOCaml's staging facilities. MetaOCaml
+represents future-stage expressions of type |t| at the
+present stage as values of type |('c,t) code| where |'c| is the
+environment classifier \cite{WalidPOPL03,calcagno-ml-like}. Code values are created
+by a \emph{bracket} form |.<e>.|, which specifies that the expression |e| is to be
+evaluated at a future stage. The \emph{escape} |.~e| must occur
+within a bracket and specifies that the expression |e| must be evaluated
 at the current stage; its result, which must be a code value, is
-spliced in into the code being built. The \emph{run} form |.! e|
-evaluates the code value, which involves run-time compilation and
-linking of the future-stage code. The bracket, escape and run are akin to
-quasi-quotation, unquotation and |eval| of Lisp.
-
+spliced into the code being built by the enclosing bracket. The \emph{run} form |.! e| evaluates
+the future-stage code value~|e| by compiling and linking it at run-time.
+The bracket, escape, and run are akin to
+quasi-quotation, unquotation, and |eval| of Lisp.
+
+Inserting brackets and escapes appropriately into the evaluating
+interpreter~|R| above yields the compiling interpreter~|C| below.
 % this code does not have the 'sv parameter. It shows up later.
@@
 module C = struct
+
   type ('c,'dv) repr = ('c,'dv) code
+
   let int (x:int)   = .<x>.
   let bool (b:bool) = .<b>.
-  let add e1 e2     = .<.~e1 + .~e2>.
-  let mul e1 e2     = .<.~e1 * .~e2>.
-  let leq x y       = .< .~x <= .~y >.
-  let if_ eb et ee = 
-    .<if .~eb then .~(et () ) else .~(ee () )>.
+
   let lam f         = .<fun x -> .~(f .<x>.)>.
@@
     .<let rec self n = .~(f .<self>.) n in self>.
-end;;
+
+  let add e1 e2     = .<.~e1 + .~e2>.
+  let mul e1 e2     = .<.~e1 * .~e2>.
+  let leq x y       = .<.~x <= .~y>.
+  let if_ eb et ee = 
+    .<if .~eb then .~(et ()) else .~(ee ())>.
+
+end
 \end{code}
@@
 |module R|.
-This compiler attempts no optimizations at all, it simply produces
-residual code. For example, the evaluation of our |test1|
-\begin{code}
-  let module E = EX(C) in E.test1 ();;
-\end{code}
-gives a value with inferred type |('a, bool) C.repr| which looks
-like |.<((fun x_6 -> x_6) (true))>.|. MetaOCaml can print the code
-values, that is, the corresponding future-stage code.
-
-It is readily obvious that this compiler does not suffer
-any interpretive overhead whatsoever. Indeed, the
-  code produced for a function is simply |fun x_6 -> x_6| and has no 
-  embedded calls to the evaluator: cf. the evaluator of DSL functions
-  in Section~\ref{tagproblem}, the L-line.
-The resulting code obviously contains no tags and no pattern-matching.
-
-We have also implemented this ``compiler'' in Haskell. Since Haskell
+This compiler simply produces
+residual code with no optimization. For example, evaluating our |test1|
+\begin{code}
+let module E = EX(C) in E.test1 ()
+\end{code}
+gives the code value |.<(fun x_6 -> x_6) (true)>.|
+of inferred type |('c, bool) C.repr|.
+
+This compiler does not incur
+any interpretive overhead: the
+code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| and contains no 
+  call to the evaluator, unlike the recursive calls to |eval0| and
+  |eval| in the |L e| lines in \S\ref{tagproblem}.
+The resulting code obviously contains no tags and no pattern matching.
+
+We have also implemented this compiler in Haskell. Since Haskell
 has no (convenient, typed) staging facilities, we had to emulate
 them. To be precise, we defined a data type |ByteCode| with
-constructors |Var|, |Lam|, |App|, |Fix|, |INT| etc. with the obvious
-meaning. Unlike our staging language (which is based on higher-order
-abstract syntax), our bytecode, to be realistic, uses integer-named
-variables. We then define 
+constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
+Whereas our representation of object terms uses higher-order
+abstract syntax, our bytecode uses integer-named
+variables to be realistic. We then define 
 \begin{code}
@@
 \end{code}
-where |Int| is the counter for the creation of fresh variable
-names. We define the compiler by making |C| to be an instance of the
+where |Int| is the counter for creating fresh variable
+names. We define the compiler by making |C| an instance of the
 class |Symantics|. The implementation\footnote{The implementation uses
@@
 
-\section{Tagless partial evaluator}\label{PE}
+\section{A tagless partial evaluator}\label{PE}
 

### 2007-04-01 tagless/tagless-final.tex
Starting with Section 4

@@
 
+\begin{comment}
 \authorinfo{Jacques Carette}
@@
            {ccshan@cs.rutgers.edu}
+\end{comment}
+\authorinfo{}{}{}
 
@@
 To be concrete, we use the typed object language in
-Figure~\ref{fig:object} throughout this paper.  We aim not just to
-interpret the object language in the narrow sense but also for
+Figure~\ref{fig:object} throughout this paper.  We aim not just for
+evaluation of object programs but also for
 compilation, partial evaluation, and other processing.
@@
 In sum, by Church\hyp encoding terms using ordinary functions,
-we achieve a tagless interpreter
+we achieve a tagless evaluator
 for a typed object language in a metalanguage with a simple type
@@
 
-The interpreter above is wired directly into the
-functions |b|, |lam|, |app|, and so on.  In the remainder of this paper, 
+The evaluator above is wired directly into the
+functions |b|, |lam|, |app|, and so on.  In the rest of this paper, 
 we explain how to abstract the interpreter so as
-to ``evaluate'' the same term in many different
-ways: interpretation, compilation, partial evaluation, size
+to process the same term in many other
+ways: compilation, partial evaluation, size
 computation, and so forth.
@@
 \item a clear way to extend the object language with more features;
-\item and an approach to self-interpretation compatible with the
-  above.  Self-interpretation turned out to be harder than expected.
+\item and an approach to self\hyp interpretation compatible with the
+  above.  Self\hyp interpretation turned out to be harder than expected.
 \end{enumerate}
@@
 
-We can also compile our object language
+Besides immediate evaluation, we can compile our object language
 into OCaml code using MetaOCaml's staging facilities. MetaOCaml
@@
 at the current stage; its result, which must be a code value, is
-spliced into the code being built by the enclosing bracket. The \emph{run} form |.! e| evaluates
+spliced into the code being built by the enclosing bracket. The \emph{run} form |.!e| evaluates
 the future-stage code value~|e| by compiling and linking it at run-time.
@@
 
-Inserting brackets and escapes appropriately into the evaluating
-interpreter~|R| above yields the compiling interpreter~|C| below.
+Inserting brackets and escapes appropriately into the
+evaluator~|R| above yields the compiler~|C| below.
 % this code does not have the 'sv parameter. It shows up later.
@@
 This compiler simply produces
-residual code with no optimization. For example, evaluating our |test1|
+residual code with no optimization. For example, interpreting our |test1|
 \begin{code}
@@
 any interpretive overhead: the
-code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| and contains no 
-  call to the evaluator, unlike the recursive calls to |eval0| and
+code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| and does not
+  call the interpreter, unlike the recursive calls to |eval0| and
   |eval| in the |L e| lines in \S\ref{tagproblem}.
@@
 GADTs because we also wanted to write a typed interpreter for 
-the \textsf{ByteCode} \emph{datatype}.} is quite similar (but slightly more
+the \textsf{ByteCode} \emph{data type}.} is quite similar (but slightly more
 verbose) than the corresponding MetaOCaml code given above. The
@@
 
-Surprisingly, it is also possible to write a partial evaluator!  The
+Surprisingly, we can write a partial evaluator using the idea above,
+namely to build object terms using ordinary functions rather than data
+constructors.  We present this partial evaluator in a sequence of four
+attempts, the last of which finally uses no universal type and no tags
+for object types.  We then discuss residualization and binding-time
+analysis.  Our partial evaluator is a modular extension to the evaluator
+in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
+that it uses the former to reduce static terms and the latter to build
+dynamic terms.
+
+\ccshan{Move later:} The
 partial evaluator does have tags: \emph{binding time} or \emph{phase}
 tags, i.e. whether a value is static or dynamic.
-Pattern-matching on this phase tag is always exhaustive; we still
-have no universal type and no tags for object types.  The partial
-evaluator uses the previously mentioned compiler to ``interpret'' dynamic
-terms, and uses the interpreter to reduce static terms.  In this sense, 
-the partial evaluator is a modular extension to both the compiler and
-the interpreter.
-
-\subsection{The trouble with (opaque) functions}
+Pattern-matching on this phase tag is always exhaustive.
+
+\subsection{Avoiding polymorphic lift}
+
+Roughly speaking, a partial evaluator interprets each object term either
+to a static (present-stage) term, as produced by the evaluator~|R|, or
+to a dynamic (future-stage) term, as produced by the compiler~|C|.  The
+interpreter computes as much as possible statically.  Thus one might try
+to define partial evaluation as the Haskell data type
+\begin{code}
+data P0 t = S0 (R t) | E0 (C t)
+\end{code}
+To extract a dynamic term from this type, we create the functions
+\begin{code}
+abstrInt :: P0 Int -> C Int
+abstrInt (S0 r) = int (unR r)
+abstrInt (E0 c) = c
+
+abstrBool :: P0 Bool -> C Bool
+abstrBool (S0 r) = bool (unR r)
+abstrBool (E0 c) = c
+\end{code}
+We then start to define |P0| as an instance of the |Symantics| class.
+\begin{code}
+instance Symantics P0 where
+  int  x = S0 (int x) (int x)
+  bool x = S0 (bool x) (bool x)
+  add (S0 e1) (S0 e2) = S0 (add e1 e2)
+  add e1 e2 = E0 (add (abstrInt e1)
+                      (abstrInt e2))
+\end{code}
+Integer and boolean literals are obviously immediate, present-stage
+values. Addition yields a static term if and only if both operands
+are static. If so, we use the evaluator~|R| to add the operands.
+If not, we extract dynamic terms from the operands and add them
+using the compiler~|C|.
+
+Whereas |mul| and |leq| are as easy to define as |add|, we encounter
+a problem with |if_|.  If the first argument to |if_| is a dynamic term
+(type |C Bool|) and the second and third arguments are a static term
+(type |R a|) and a dynamic term (type |C a|), then we need to convert
+the static term to dynamic, but there is no polymorphic ``lift''
+function |a -> C a| to send a value to the future stage.
+
+Our |Symantics| class includes only separate lifting methods |bool| and
+|int|, not a parametrically polymorphic lifting method, for good reason:
+When compiling to a first-order target language such as machine code,
+booleans, integers, and functions may well be represented differently.
+Thus polymorphic lift cannot be compiled without intensional type
+analysis.  To avoid the need for polymorphic lift, we turn to
+a technique due to Asai \ccshan{cite?}: build a dynamic term alongside
+every static term.
+
+\subsection{Delaying binding-time analysis}
 \label{S:PE-problem}
-We want to combine the interpreter |R| and the compiler |C| to produce
-an (online) partial evaluator.  We defer the discussion of the 
-choice of online/offline partial evaluation until after we have presented our
-solution in~\ref{S:PE-solution}.
-The partial evaluator (PE) is too an interpreter: it
-evaluates a term either to an immediate (or, `static' or present-stage)
-value or to a `dynamic', future-stage value, or `code' (of the sort
-produced by the compiler C). The interpreter attempts to perform all
-possible computations (i.e., computations whose operarands are
-present-stage values) as the present-stage. Our first attempt, in
-Haskell, is straightforward: we introduce
-\begin{code}
-data P1 t = S1 (R t) (C t) | E1 (C t)
-\end{code}
-as a `direct sum' of the interpreter |R| and the compiler |C|. The
-datatype |P1| has two \emph{phase} (not type!) tags |S1| 
-(for static) and |E1| for dynamic. The static value has both the
-present-stage and future-stage alternatives. The dynamic value does
-not have the present-stage component. We define
+
+We switch to the Haskell data type
+\begin{code}
+data P1 t = P1 (Maybe (R t)) (C t)
+\end{code}
+so that a partially evaluated term |P1 t| always contains a dynamic
+component and sometimes contains a static component.  The two
+alternative constructors of a |Maybe| value, |Just| and |Nothing|,
+tag each partially evaluated term with a phase: either present or
+future.  This tag is not an object type tag: all pattern matching below
+is exhaustive.  Because the future-stage component is always present, we
+can now define the polymorphic function
 \begin{code}
 abstr1 :: P1 t -> C t
-abstr1 (S1 _ dyn) = dyn
-abstr1 (E1 dyn)   = dyn
-\end{code}
-to extract the future-stage component. We then attempt to define the
-interpreter |P1| compositionally -- and it works for the first-order
-fragment of our DSL language.
-
+abstr1 (P1 _ dyn) = dyn
+\end{code}
+to extract it without requiring polymorphic lift into~|C|.  We then try
+to define the interpreter |P1|---and get as far as the first-order
+constructs of our object language, including |if_|.
 \begin{code}
 instance Symantics P1 where
-    int  x  = S1 (int x) (int x)
-    bool b  = S1 (bool b) (bool b)
-    add (S1 n1 _) (S1 n2 _) = int (unR (add n1 n2))
-    add e1 e2 = E1 (add (abstr1 e1) (abstr1 e2))
+  int  x = P1 (Just (int x)) (int x)
+  bool b = P1 (Just (bool b)) (bool b)
+  add (P1 (Just n1) _) (P1 (Just n2) _)
+    = int (unR (add n1 n2))
+  add e1 e2 = P1 Nothing
+    (add (abstr1 e1) (abstr1 e2))
   -- mul and leq are analogous and elided
-    if_ (S1 s _) et ef = if unR s then et else ef
-    if_ eb et ef = 
-       E1 (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
-\end{code}
-Integer and boolean literals are obviously immediate, present-stage
-values. The result of adding two values is static only if both values
-are. If so, we use the evaluator |R| to add the values; the result is
-an essentially integer literal, which can be promoted to the future
-stage. If at least one of the operands to |add| is dynamic, we promote
-both operands to the future-stage and use the compiler |C| to
-`compile' the code for adding the operand expressions.  Because the
-static value includes as a component its future-stage variant, and
-because we build both components together, we are able to get by
-without the polymorphic `lift' operation (that is, lift the
-computation of any type from the present-stage to the future
-stage). We can do such a lifting only on type-by-type basis (for
-booleans and for integers). The problem with the polymorphic lift is
-that it inhibits true compilation, to assembler or other similar first
-order target language with no polymorphic (run-time reified) types.
-The representation of integers in the target language most likely
-differs from the representation of booleans or functions,
-etc. Therefore, without the intentional type analysis, the polymorphic
-lift is not compilable.
+  if_ (P1 (Just s) _) et ef
+    = if unR s then et else ef
+  if_ eb et ef = P1 Nothing
+    (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
+\end{code}
 
@@
 |'sv|, which is the result of applying |repr_pe| function above to the
-type |'dv|. The signature |Symantics| now reads
+type |'dv|. Figure~\ref{fig:ocaml} shows the new signature |Symantics|.
+\begin{figure*}
+\begin{floatrule}
 \begin{code}
 module type Symantics = sig
+
   type ('c,'sv,'dv) repr
+
   val int : int  -> ('c,int,int) repr
-  val bool : bool -> ('c,bool,bool) repr
-  val add  : ('c,int,int) repr -> 
-      ('c,int,int) repr -> ('c,int,int) repr
-  val mul  : ('c,int,int) repr -> 
-      ('c,int,int) repr -> ('c,int,int) repr
-  val leq  : ('c,int,int) repr -> 
-      ('c,int,int) repr -> ('c,bool,bool) repr
-  val if_  : ('c,bool,bool) repr ->
-             (unit -> ('c,'sa,'da) repr) ->
-             (unit -> ('c,'sa,'da) repr) -> 
-             ('c,'sa,'da) repr 
-  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr)
-    -> ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
-           'da->'db) repr
-  val app : ('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr),
-             'da->'db) repr
-    -> ('c,'sa,'da) repr -> ('c,'sb,'db) repr
-  val fix : 
-    (('c,(('c,'sa,'da) repr -> ('c,'sb,'db) repr) as 's,
-         'da->'db) repr 
-    -> ('c,'s,'da->'db) repr)  -> ('c,'s,'da->'db) repr
-end;;
-\end{code}
+  val bool: bool -> ('c,bool,bool) repr
+
+  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
+  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
+
+  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
+  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
+
+end
+\end{code}
+\end{floatrule}
+\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
+\label{fig:ocaml}
+\end{figure*}
+
 The interpreters |R|, |L| and |C| above only relied on the last

### 2007-04-01 tagless/incope.hs

@@ -296,26 +296,25 @@ ctestpw72 = compC  (app (testpowfix7 ()) (int 2))
 -- First attempt: works for the first-order fragment of our language,
 -- but stumbles on the higher-order fragment.
 
-data P1 t = S1 (R t) (C t) | E1 (C t)
+data P1 t = P1 (Maybe (R t)) (C t)
 
 abstr1 :: P1 t -> C t
-abstr1 (S1 _ dyn) = dyn
-abstr1 (E1 dyn)   = dyn
+abstr1 (P1 _ dyn) = dyn
 
 instance Symantics P1 where
-    int  x  = S1 (int x) (int x)
-    bool b  = S1 (bool b) (bool b)
-    add (S1 n1 _) (S1 n2 _) = int (unR (add n1 n2))
-    add e1 e2 = E1 (add (abstr1 e1) (abstr1 e2))
-    mul (S1 n1 _) (S1 n2 _) = int (unR (mul n1 n2))
-    mul e1 e2 = E1 (mul (abstr1 e1) (abstr1 e2))
-    leq (S1 n1 _) (S1 n2 _) = bool (unR (leq n1 n2))
-    leq e1 e2 = E1 (leq (abstr1 e1) (abstr1 e2))
-    if_ (S1 s _) et ef = if unR s then et else ef
-    if_ eb et ef = E1 (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
+    int  x = P1 (Just (int x)) (int x)
+    bool b = P1 (Just (bool b)) (bool b)
+    add (P1 (Just n1) _) (P1 (Just n2) _) = int (unR (add n1 n2))
+    add e1 e2 = P1 Nothing (add (abstr1 e1) (abstr1 e2))
+    mul (P1 (Just n1) _) (P1 (Just n2) _) = int (unR (mul n1 n2))
+    mul e1 e2 = P1 Nothing (mul (abstr1 e1) (abstr1 e2))
+    leq (P1 (Just n1) _) (P1 (Just n2) _) = bool (unR (leq n1 n2))
+    leq e1 e2 = P1 Nothing (leq (abstr1 e1) (abstr1 e2))
+    if_ (P1 (Just s) _) et ef = if unR s then et else ef
+    if_ eb et ef = P1 Nothing (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
 
 -- But the problem occurs when we try to implement lam. The result
[…]

### 2007-04-01 tagless/tagless-final.tex
checking in more small edits a bit all over

@@
 postprocessing step. That has prevously considered an open problem
-(impossible). 
-Our family of evaluators (interpreters, embeddings?) contains 
+(i.e. impossible). 
+Our family of evaluators contains 
 an interpreter, a compiler, a partial
@@
 The main tool is to take functions seriously: we encode the
-deBruijn or higher-order abstract syntax  of object terms 
+de Bruijn or higher-order abstract syntax  of object terms 
 not by using constructors for
@@
 Types are useful as usual because they document the programmer's
-intentions and can be checked early by machine (see~\S\ref{encoding}).
+intentions and can be checked early (see~\S\ref{encoding}).
 To be concrete, we use the typed object language in
@@
 data constructors.  The interpreter below provides these functions.
+\jacques{but now this smaller language does not match anymore what
+we wrote earlier.  I think it is better to use a small language here,
+but i don't quite know how to say so}
 \begin{code}
@@
 metalanguage implementation, we avoid the serious performance penalty
-\citep{WalidICFP02} of error checking.
-\ccshan{Need to cite Thiemann here and in related work?}
+\citep{WalidICFP02} of error checking.  See \citep{Gluck-jones-optimality}
+for deeper technical reasons which inevitably lead to these performance
+penalties.
 
@@
 \end{enumerate}
-\jacques{can we state fairly clearly what type system we need?  I mean,
-Haskell98(?) and MetaOCaml with (objects, polymorphic variants, ...).
-With our let-polymorphism trick for self-interpretation, what do we really
-need?}
 
@@
 We finally conclude.
+\jacques{TODO}
 
@@
 the object language and its instances give the semantics.
-The parameter |repr| to the class has the kind |* -> *|.
+The parameter |repr| to the class has kind |* -> *|.
 \begin{code}
@@
 Thus we represent an object expression in
-OCaml as a functor from |Symantics|. This
+OCaml as a functor from |Symantics| to an appropriate semantic domain. This
 is essentially the same as the constraint |Symantics repr =>| in the
@@
 
-The first interpreter evaluates the object term to its value in the
+The first interpreter evaluates an object term to its value in the
 metalanguage.  The module below interprets each object\hyp language
@@
 module R = struct
-
   (* absolutely no wrappers *)
@@
 OCaml's addition, and |app| is OCaml's application. To run our
-examples, we instantiate the |EX| functor from~\S\ref{encoding} with this |R| module.
+examples, we instantiate the |EX| functor from~\S\ref{encoding} with this |R|
+module.
 \begin{code}
@@
 only at compile time, not at run time.  Pattern matching against~|R|
-cannot ever fail and is compiled to no-op.
+cannot ever fail and is compiled away.
 In both OCaml and Haskell, it is obvious to the compiler that
@@
 analysis.  To avoid the need for polymorphic lift, we turn to
-a technique due to Asai \ccshan{cite?}: build a dynamic term alongside
-every static term.
+a technique due to Asai \ccshan{cite?}\jacques{yes}: build a dynamic term
+alongside every static term.
 
@@
 determine if they are static or dynamic. The argument of |lam| is a 
-function, which is not introspectible. One may only 
+function, whose structure cannot be observed. One may only 
 apply |f| to a particular static or dynamic value, and receive another 
@@
 either static or dynamic results (in the example above, applying 
-the function to (int 0) gives the static value in return). One
+the function to (int 0) gives a static value in return). One
 solution is to be conservative and always use the |C| compiler to
-compile |f| to the future-stage expression, which we can tag with
-E1. But that defeats the purpose of partial evaluation as we may miss
+compile |f| to a future-stage expression, which we can tag with
+E1. But this defeats the purpose of partial evaluation as we may miss
 many opportunities for static computations and simplifications, such
@@
 applied to a particular argument (that is, until the result of |lam f|
-appears as the argument of |app|). That means, we have to find a way
+appears as the argument of |app|). In other words, we have to find a way
 to incorporate |f| as it is into the P1 data structure. We would like
-to write that when |t| is a particular type |a->b|, that data type
+to write that when |t| is a particular type |a->b|, this data type
 should be 
@@
 \end{code}
-That is, we need a non-parameteric data type, something like a
-typecase (ref? Gibbons?).
-
-The problem we encountered here is quite akin to typed CPS
+That is, we need a non-parametric data type, something akin to
+type-indexed functions and type-indexed types, as used in 
+\citet{Oliveira2005:TypeCase}, where it is dubbed the \emph{typecase}
+design pattern.
+
+The problem we encountered here is quite akin to the typed CPS
 transform. When transforming abstractions, we come across
 non-functorialness. The connection with CPS is not an accident, as we
-shall see below (in Section XXX).
+shall see in~\S\ref{S:CPS}.
 
@@
 generalized algebraic data types (GADT) or type-class functions
-(implemented via typeclasses with functional dependencies). These
-methods are equivalent; here we show the GADT; please see |incope1.hs|
+(implemented via typeclasses with functional dependencies)
+\citet{Oliveira2005:TypeCase}.  These
+methods are equivalent, and here we show the GADT; please see |incope1.hs|
 in the accompanying source code for the latter.
@@
 with four data constructors. The constructor |VF| stands out as its
-operand is of the type |P a -> P b| rather than |a -> b| as one may
+operand has type |P a -> P b| rather than |a -> b| as one may
 have expected if |P t| were parametric in |t|. As before, we define a
@@
 We use the method |lam| of the |C| intrepreter to compile the argument
-of |lam| into the code value. We may now make |P| the instance of
-Symantics and thus implement the partial evaluator (the
-implementations of |mul|, |leq|, |if_| and |fix| are elided).
+of |lam| into a code value. We may now make |P| an instance of
+Symantics and implement the partial evaluator (the
+implementations of |mul|, |leq|, |if_| and |fix| are elided) as follows.
 \begin{code}
@@
 HOAS function |f| into the result. At any point we can convert it to a
-code value, if we must. But delaying such a conversion, we get the
-opportunity to apply |f| to a concrete argument. When we evaluating
-|app ef ea|, we check to see of the first argument is so-delayed
+code value, if we must. By delaying such a conversion, we get the
+opportunity to apply |f| to a concrete argument. When evaluating
+|app ef ea|, we check to see of the first argument is such a delayed
 lambda-expression. If it is, we apply the incorporated |f| to the
-concrete argument |ea|, giving us a chance to perform the static
-computations (we see the example in the next subsection). If |ef| is a
+concrete argument |ea|, giving us a chance to perform static
+computations (see example |testpowfix| in the next subsection). If |ef| is a
 dynamic value, we residualize.
 
-Although GADT solve our problem, the solution is not quite
+Although GADTs solve our problem, the solution is not quite
 satisfactory. First of all, it cannot be ported to MetaOCaml as GADTs
-are not supported in OCaml. Second, the tagging problem reappeared, in
+are not supported there. Second, the tagging problem reappeared, in
 the implementation of |app| above. The datatype |P t| has four data
-constructors; the pattern-matching in |app| however used only two of
+constructors; pattern-matching in |app| however uses only two of
 them, |VF| and |E|. It is obvious, one may say, that the constructors
 |VI| and |VB| just cannot occur, because the corresponding values
-cannot have the arrow type. What is obvious to us is not obvious to
-the compiler. The case matching above does look \emph{partial}. The
-point may loop tenuous -- and yet it is precisely the main point of
+cannot be function types. What is obvious to us is not obvious to
+the compiler. The case matching above looks \emph{partial}.  While this
+may seem like a minor point, 
+this is precisely the main point of
 tag elimination. In a typed tagged interpreter there are many
@@
 
-\subsection{The `final' solution}
+\subsection{The ``final'' solution}
 \label{S:PE-solution}
-Let us re-examone the problem in Section~\ref{S:PE-problem}. What we
+Let us re-examine the problem in Section~\ref{S:PE-problem}. What we
 would ideally like is to write
@@
 \end{code}
-Although we can use typeclasses to define that type function
+Although we can use typeclasses to define this type function
 in Haskell, that is not portable to MetaOCaml. We notice however that
-the type case alternatives above are all associated with the existing
-methods of |Symantics| and so the case analysis is already present. 
+the type case alternatives above are all associated with existing
+methods of |Symantics|, and so the case analysis is already present. 
 A simple and more portable solution thus emerges: we bake |repr_pe| 
 into the signature |Symantics|. Switching to MetaOCaml for
-portability, we recall from Section~\ref{encoding} that |repr| type
+portability, we recall from Section~\ref{encoding} that the |repr| type
 had two arguments |type ('c,'dv) repr|. We add an extra argument,
-|'sv|, which is the result of applying |repr_pe| function above to the
+|'sv|, which is the result of applying the |repr_pe| function above to the
 type |'dv|. Figure~\ref{fig:ocaml} shows the new signature |Symantics|.
@@
 When writing the partial evaluator however, we make use of the
-type argument |'sv|. Here's the corresponding instance of the
-Symantics:
+type argument |'sv|. Here is the corresponding instance of
+|Symantics|:
 
@@
 subsection, extracts the future-stage computation from the result of
-PE. The function |pdyn| does the converse. As in
-section~\ref{S:PE-problem}, we build future-stage computations as we
-build present-stage ones, if any. Therefore, we avoid the need for the
+partial evaluation.  The function |pdyn| does the converse. As in
+section~\ref{S:PE-problem}, we build future-stage computations at the
+same time as we
+build present-stage ones, if any. Therefore, we avoid the need for a
 polymorphic `lift' function. The implementations of
-|add| and |mul| are a bit more sophisticated, taking advantage of the
-implicit monoid structure. The function |buildsimp| checks to see if
+|add| and |mul| are a bit more sophisticated, taking advantage 
+respectively of the implicit monoid and ring structure of the 
+underlying domain.
+The function |buildsimp| checks to see if
 its two arguments |e1| and |e2| are static expressions. If so, the
@@
 
-\subsection{Call-by-name CPS interpreters}
+\subsection{Call-by-name CPS interpreters}\label{S:CPS}
 
@@
 
+Thiemann, cogen in six lines (for the interpreter/compiler work)?
+
 Boxes go Bananas by Weirich and Washburn.

### 2007-04-01 tagless/tagless-final.tex
3 edits that escaped the previous check in

@@
 
-When doing PE on fix, we made a deliberate decision to `go all the
-way' -- so long as the computation can proceed at the PE stage
-(`static value'), we do that rather than residualizing. In the
+When doing partial evaluation on |fix|, we made a deliberate decision to `go all
+the way' -- so long as the computation can proceed at the PE stage,
+we do that rather than residualizing. In the
 accompanying source code, we have also implemented an alternative
-where fix is unrolled only once and them we residualize. We prefer the
-first approach, even though it (unlike the latter) may cause
+where |fix| is unrolled only once and then we residualize. We prefer the
+first approach, even though it may cause
 non-termination at the PE stage -- but \emph{only} if the computation
-we are partially evaluating is non-terminating and enough static
+we are partially evaluating is non-terminating, and enough static
 information is available to trigger this non-termination.  Our
-treatment of fix may lead to the code bloat in the residual
+treatment of |fix| may lead to code bloat in the residual
 program. That is an issue investigated elsewhere (cite our PEPM2006

### 2007-04-01 tagless/incope.hs
add simplifications to Haskell code.  complete simplifications in ML code too

@@ -340,6 +340,7 @@ instance Functor P where
     fmap f = E . fmap f . abstr
 -}
 
+
 instance Symantics P where
     int x  = VI x
     bool b = VB b
@@ -363,11 +364,16 @@ instance Symantics P where
 
     fix f = pfix f -- need this charade for GADTs sake
 
-    add e1 e2 = case (e1,e2) of
-                 (VI n1,VI n2) -> VI (n1 + n2)
-                 _ -> E $ add (abstr e1) (abstr e2)
-
+    add e1 e2 = case (e1, e2) of
+                 (VI n1, _) | (n1 == 0) -> e2
+                 (_, VI n2) | (n2 == 0) -> e1
+                 (VI n1, VI n2) -> VI (n1 + n2)
+                 _              -> E $ add (abstr e1) (abstr e2)
     mul e1 e2 = case (e1,e2) of
+                 (VI n1, _) | (n1 == 0) -> e1
+                 (_, VI n2) | (n2 == 0) -> e2
+                 (VI n1, _) | (n1 == 1) -> e2
+                 (_, VI n2) | (n2 == 1) -> e1
                  (VI n1,VI n2) -> VI (n1 * n2)
                  _ -> E $ mul (abstr e1) (abstr e2)
 

### 2007-04-01 tagless/tagless-final.tex
Done with Section 4.2

@@
 \usepackage[T1]{fontenc}
+\usepackage[override]{cmtt}
 \usepackage{comment}
@@
 \begin{abstract}
-We have built a family of tagless embeddings (interpretations) for a
+We have built a family of tagless interpretations for a
 higher-order typed object
-language in a typed metalanguage (current Haskell and ML), 
+language in a typed metalanguage (Haskell and ML), 
 requiring neither dependent
 types, nor generalized algebraic data types, nor a tag elimination 
-postprocessing step. That has prevously considered an open problem
-(i.e. impossible). 
-Our family of evaluators contains 
-an interpreter, a compiler, a partial
+postprocessing step. It has been considered an open (impossible)
+problem to do so.
+Our family of interpreters contains 
+an evaluator, a compiler, a partial
 evaluator, a term-size evaluator and a CPS call-by-name transformer.
@@
 The main tool is to take functions seriously: we encode the
-de Bruijn or higher-order abstract syntax  of object terms 
+de-Bruijn or higher-order abstract syntax  of object terms 
 not by using constructors for
@@
 represent object terms such as those in Figure~\ref{fig:object}.
+For brevity, we elide treating integers, conditionals, and fixpoint in
+this section.
 \begin{code}
@@
 
-let rec eval0 env = function 
+let rec eval0 env = function ...
 | V v       -> lookup env v
@@
 \begin{code}
-let rec eval env = function
+let rec eval env = function ...
 | V v       -> lookup env v
@@
 metalanguage implementation, we avoid the serious performance penalty
-\citep{WalidICFP02} of error checking.  See \citep{Gluck-jones-optimality}
-for deeper technical reasons which inevitably lead to these performance
+\citep{WalidICFP02} of error checking.  \Citet{Gluck-jones-optimality}
+explains deeper technical reasons that inevitably lead to these performance
 penalties.
@@
 analysis.  To avoid the need for polymorphic lift, we turn to
-a technique due to Asai \ccshan{cite?}\jacques{yes}: build a dynamic term
+a technique due to Asai \ccshan{cite}: build a dynamic term
 alongside every static term.
@@
 
-When we come to the higher-order fragment of our language, however, we
-stumble. Let us consider |lam f|. Here, |f| is a function from a
-|P1 a| value to a |P1 b| value. The result of |lam f| must be either a
-static value or a dynamic value -- but which? Unlike |add| or |if_|, 
-|lam| cannot really examine its operand. The arguments of |add| are 
-of type |P1 int|, which we can examine by pattern-matching and so 
-determine if they are static or dynamic. The argument of |lam| is a 
-function, whose structure cannot be observed. One may only 
-apply |f| to a particular static or dynamic value, and receive another 
-value in return. Alas, as in the case |lam (\x -> mul x dynval)|, 
-applying the same function to different static values may give 
-either static or dynamic results (in the example above, applying 
-the function to (int 0) gives a static value in return). One
-solution is to be conservative and always use the |C| compiler to
-compile |f| to a future-stage expression, which we can tag with
-E1. But this defeats the purpose of partial evaluation as we may miss
-many opportunities for static computations and simplifications, such
-as multiplication by zero in the example above. A better solution is to
-delay the binding-type-analysis (BTA) until the function |f| is
-applied to a particular argument (that is, until the result of |lam f|
-appears as the argument of |app|). In other words, we have to find a way
-to incorporate |f| as it is into the P1 data structure. We would like
-to write that when |t| is a particular type |a->b|, this data type
-should be 
-\begin{code}
-data P1 (a->b) = S1 (R (P1 a -> P1 b)) (C (a->b)) 
+When we come to functions, however, we stumble.  According to our
+definition of~|P1|, a partially evaluated object function, such as the
+identity $\fun{x}x$ embedded in Haskell as |lam (\x -> x)|\texttt{ ::
+}|P1 (a -> a)|, consists of a dynamic part (type |C1 (a -> a)|) and
+maybe a static part (type |R1 (a -> a)|).  The dynamic part is useful
+when this function is passed to another function that is only
+dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
+useful when this function is applied to a static argument, as in
+$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
+evaluate the function, that is, compute as much as possible statically
+when it is applied to a mix of static and dynamic inputs.  For example,
+the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
+by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
+not statically known.  Even the same static function, applied to
+different static arguments, can give both static and dynamic results: we
+want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
+to~$x$.
+
+To enable these simplifications, we delay binding-time analysis (BTA)
+for a static function until it is applied, that is, until |lam f|
+appears as the argument of |app|.  To do so, we have to incorporate |f|
+as it is into the |P1| data structure: applying the type constructor
+|P1| to a function type |a -> b| should yield one of
+\begin{code}
+data P1 (a -> b) = S1 (P1 a -> P1 b)
                  | E1 (C (a->b))
-\end{code}
-That is, we need a non-parametric data type, something akin to
-type-indexed functions and type-indexed types, as used in 
-\citet{Oliveira2005:TypeCase}, where it is dubbed the \emph{typecase}
-design pattern.
-
-The problem we encountered here is quite akin to the typed CPS
-transform. When transforming abstractions, we come across
-non-functorialness. The connection with CPS is not an accident, as we
-shall see in~\S\ref{S:CPS}.
+
+data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b))
+                      (C (a -> b))
+\end{code}
+That is, we need a nonparametric data type, something akin to
+type-indexed functions and type-indexed types, which
+\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
+
+Thus typed partial evaluation, like typed CPS transformation,
+inductively defines a map from source types to target types that
+performs case distinction on the source type.  The connection with CPS
+is not an accident, as we shall see in~\S\ref{S:CPS}.
 
@@
 Two common ways of providing typecase in Haskell is by using either
-generalized algebraic data types (GADT) or type-class functions
+GADT or type-class functions
 (implemented via typeclasses with functional dependencies)
-\citet{Oliveira2005:TypeCase}.  These
+\citep{oliveira-typecase}.  These
 methods are equivalent, and here we show the GADT; please see |incope1.hs|

### 2007-04-01 tagless/tagless-final.tex
edits up to section 4, will start rest soon

@@
 \subsection{Avoiding polymorphic lift}
+\label{S:PE-lift}
 
@@
   val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
-
 end
@@
 
-  let buildsimp cast e f1 f2 = fun e1 e2 -> 
-     match (e1,e2) with
+  let build cast f1 f2 = fun e1 -> fun e2 -> match (e1,e2) with
+                   ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
+                 | _ -> pdyn (f2 (abstr e1) (abstr e2))
+  let buildsimp_monoid cast e f1 f2 = fun e1 e2 -> match (e1,e2) with
                  | ({st = Some e'}, _) when e = e' -> e2
                  | (_, {st = Some e'}) when e = e' -> e1
-        | ({st=Some n1}, {st=Some n2})-> cast (f1 n1 n2)
-        | _ -> pdyn (f2 (abstr e1) (abstr e2))
-
-  let add e1 e2 = buildsimp int 0 R.add C.add e1 e2
-  let mul e1 e2 = buildsimp int 1 R.mul C.mul e1 e2
-  let leq e1 e2 = 
-     match (e1,e2) with
-         ({st=Some n1},{st=Some n2})-> bool (R.leq n1 n2)
-        | _ -> pdyn (C.leq (abstr e1) (abstr e2))
+                 | _ -> build cast f1 f2 e1 e2
+  let buildsimp_ring cast zero one f1 f2 = fun e1 e2 -> match (e1,e2) with
+                 | ({st = Some e'}, _) when e' = zero -> e1
+                 | (_, {st = Some e'}) when e' = zero -> e2
+                 | _ -> buildsimp_monoid cast one f1 f2 e1 e2
+
+  let add e1 e2 = buildsimp_monoid int 0 R.add C.add e1 e2
+  let mul e1 e2 = buildsimp_ring int 0 1 R.mul C.mul e1 e2
+  let leq e1 e2 = build bool R.leq C.leq e1 e2
   let if_ eb et ee = match eb with
@@
 underlying domain.
-The function |buildsimp| checks to see if
+The function |build| checks to see if
 its two arguments |e1| and |e2| are static expressions. If so, the
 corresponding static computation (with the help of the intrepreter
-|R|) is performed. The function also checks if one argument is static
-and is the neutral element of the monoid structure 
-\jacques{Comment on the various nice generic things in the implementation,
-which are a little silly in this small version, but would really pay off
-in something bigger.  See the abstract interpretation hard at work!
-See the monoid-driven simplifications!}
-\oleg{I'm inclined to ditch buildsimp. We miss an opportunity here:
-  when multiplying anything by 0, the result is zero!}
-
-When doing partial evaluation on |fix|, we made a deliberate decision to `go all
-the way' -- so long as the computation can proceed at the PE stage,
+|R|) is performed. The function |buildsim_monoid| also checks if one argument
+is static and is the neutral element of the monoid structure.
+|buildsimp_ring| also worries about the full ring structure rather than
+just the monoid structure.  These are put here to illustrate how we 
+would go about writing general algebraic simplification routines that
+would mimic various compiler optimizations, and these would be quite
+beneficial in a large language with more base types.
+
+When doing partial evaluation on |fix|, we made a deliberate decision to `go
+all the way' -- so long as the computation can proceed at the PE stage,
 we do that rather than residualizing. In the
@@
 treatment of |fix| may lead to code bloat in the residual
-program. That is an issue investigated elsewhere (cite our PEPM2006
-paper).
+program. The techniques of~\citet{SwadiTahaKiselyovPasalic2006} can
+be used to deal with this issue.
 
@@
 It is instructive to compare this result with that of the |C|
-evaluator. We observe the beta-redex that was apparent there is now
+evaluator. We observe that the beta-redex has now been 
 performed. The result is the static value |true|.
@@
  - : ('_a, int -> int) C.repr =
-.<fun x_1 ->
-   (((fun x_2 ->
+.<fun x_1 -> (((fun x_2 ->
   let rec self_3 =
@@
 \end{code}
-
-which is what one would have expected from the power example.
+as expected from the power example.
 
 Unlike the GADT approach in the previous subsection, there is no
-longer any seemingly partial pattern-matching. The pattern-match only
-occurs in PE (to check if the code is static or dynamic), it is only a
-phase rather than a type tag match, and the pattern match is
+longer any seemingly partial pattern-matching. All pattern-matching
+occurs in the partial evaluator only (to check if the current value 
+is static or dynamic), and is 
+phase rather than a type tag match. Furthermore the matching
 \emph{patently} exhaustive. That is, totality is assured and it is
-assured \emph{syntactically}, in the way that is apparent to the
+assured \emph{syntactically}, in a way that is apparent to the
 compiler.
@@
 of the compiler and the interpreter. It is truly the `composition' of the
-interpreter and the compiler.  \jacques{Perhaps cite a paper of Thiemann
-here, cogen in 6 lines?}
-
-\jacques{Comment on online/offline here}
-The PE in incope is really a mixture of online and off-line; it looks 
-online, but it has a sort of "just in time" binding-time analysis built in.
+interpreter and the compiler.   This is a simpler situation than
+in \citet{SperberThiemann:TwoForOne} where a partial evaluator and
+a compiler are `composed', but the general ideas are similar.
+
+Lastly, we should comment on the approach taken in the partial evaluator.
+It is an online, typed, polyvariant partial evaluator.  It has no
+termination analysis, and as written may diverge (but only on divergent
+programs) as we have decided to aggressively unroll loops.
 
 \section{Variations and discussion}\label{discussion}
-
-\oleg{where to put the following? Should we even mention this?
-If not, we should still keep this in comments.}
-If we just add a method 'lift' to Symantics with a type
-|lift :: a -> repr a| (cf polymorphic lift of Xi03, of MetaML and
-MetaOCaml: CSP)
-then |Symantics repr| implies |Functor repr|. We can write then
-\begin{code}
-  instance Semantics repr => Functor repr where
-  -- fmap :: (a->b) -> repr a -> repr b
-  fmap f ra = app (lift f) ra
-\end{code}
 
@@
 
+\section{Lifting and Functor}
+
+It is perhaps worthwhile noting that 
+if we were to add a method 'lift' to |Symantics| with type
+|lift :: a -> repr a| (the polymorphic lift of \citet{xi-guarded}, 
+of MetaML and
+MetaOCaml: CSP)
+then |Symantics repr| implies |Functor repr|.  This can be seen from
+\begin{code}
+instance Semantics repr => Functor repr where
+  -- fmap :: (a->b) -> repr a -> repr b
+  fmap f ra = app (lift f) ra
+\end{code}
+Of course, as we have mentioned earlier in~\S\ref{S:PE-lift}, 
+addin such a polymorphic lift would appear to introduce more
+problems than it solves.
+
 \subsection{Translucent types}

### 2007-04-01 tagless/tagless-final.tex
re-re-re fix de Bruijn's name

@@
 The main tool is to take functions seriously: we encode the
-de-Bruijn or higher-order abstract syntax  of object terms 
+de Bruijn or higher-order abstract syntax  of object terms 
 not by using constructors for

### 2007-04-01 tagless/tagless-final.tex
fix up various obvious things in section 5

@@
 
+\jacques{we should probably be more explicit about the contributions of
+the discussion section}
+
 The structure of our paper is as follows. Section XXX ...
@@
 
+We show that our |Symantics| and generally our approach accomodates
+a number of variants without any (new) difficulties.  In particular,
+we show how a a call-by-name CPS interpreter, a call-by-value CPS
+program transformation, as well as how to deal with imperative features.
+Finally, we comment on the requirements on the type system to support 
+our approach.
+
 \subsection{Call-by-name CPS interpreters}\label{S:CPS}
@@
 |app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
-possible then to represent a call-by-name (CBN) object language in a
-call-by-value (CBV) metalanguage. The answer, given back by Reynolds
-and Plotkin (cite) is CPS, which was specifically introduced by them
-to make the evaluation strategy of the definitional interpreter
+possible to represent a call-by-name (CBN) object language in a
+call-by-value (CBV) metalanguage. The answer, given already by 
+\cite{ReynoldsPlotkin} is continuation-passing style (CPS), which they
+specifically introduced 
+to make the evaluation strategy of a definitional interpreter
 independent of the strategy of the metalanguage. We demonstrate the
 same independence in the typed setting \oleg{mention the quote from
-  Reynolds that typed setting would be nice?}. Specifically, we
-demonstrate a CBN evaluation of our object language in OCaml, which is
-CBV. We do so by building a CBN CPS interpreter for the object
+Reynolds that typed setting would be nice?}\jacques{I think so, but from
+which paper? The one ReynoldsPlotkin paper is typed}. Specifically, we
+demonstrate CBN evaluation for our object language in OCaml, which is
+call-by-value. We do so by building a CBN CPS interpreter for the object
 language.
 
-The CBN CPS interpretation was introduced by Plotkin (cite). As in the
-more familiar CBV CPS, the interpretation of a value is a function
-accepting the continuation argument |k| and yielding the answer
+The CBN CPS interpretation was introduced by Plotkin \cite{PlotkinCBN}. As in
+the more familiar CBV CPS, the interpretation of a value is a function
+accepting a continuation argument |k| and yielding the answer
 returned by that continuation.
@@
 \end{code}
-In the both cases, the returned value has the type 
+In both cases, the returned value has type 
 |(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.
@@
 We want to collect the above interpretation functions into a module
-with the Symantics signature, so to fit the CBN CPS interpreter within our
+with the |Symantics| signature, so as to fit the CBN CPS interpreter within our
 general framework. Alas, we encounter the same problem we had in
@@
 type that can represent both the type of |int i| and the type of 
-|lam f|. We cannot say that the the result of interpreting an object
-expression of the type |t| is |(t->'w)->'w| because if |t| happens to
-be an arrow type such as |int->int| in the above example, the result
+|lam f|. We cannot say that the result of interpreting an object
+expression of type |t| is |(t->'w)->'w| because if |t| happens to
+be an arrow type (such as |int->int| in the above example), the result
 type obviously does not fit the desired pattern. We need a type
-function with a type-case distinction. In Section~\ref{S:PE-solution} we have
-shown the way to encode such a function, using the additional type
+function with a typecase distinction. In Section~\ref{S:PE-solution} we have
+shown how to encode such a function, using the extra type
 argument |'sv| of the |repr| type. Furthermore, we notice that the type 
-function |repr_pe| needed for the partial evaluator,
-Section~\ref{S:PE-solution}, is precisely the same type function we
+function |repr_pe| needed for the partial evaluator 
+(in~\S\ref{S:PE-solution}) is precisely the same type function we
 need for CBN CPS. This gives us for the following CBN CPS interpreter,
-of the Symantics signature:
+with the |Symantics| signature:
 
@@
 We have made our interpreter fully polymorphic over the answer type.
-We could have just as well made |RCN| a functor, parameterized over
+We could have just as easily made |RCN| a functor, parameterized over
 the answer type -- as is common when writing CPS in SML. In the
-present case, we chose to use higher-rank polymorphism of OCaml record
-types.
-
-Because |RCN| has the Semantics signature, we can instnatiate with it and run
-all of our previous examples. A more interesting is the following
-example.
+present case, we chose to use the higher-rank polymorphism of offered by
+OCaml record types.
+
+Because |RCN| has the signature Symantics, we can instantiate our previous
+examples with it, and all works as expected.  More interesting
+is the following example.
 \begin{code}
@@
 \end{code}
-we never get any result because of looping. In contrast, if we use
+we never get any result because the above diverges. In contrast, if we use
 the CBN interpreter
@@
   Mention this in the conclusions?}
+\jacques{you have strayed outside what I know -- what Plotkin results?
+  I always consider mentionning formal results a good thing}
 
@@
 In this section however we show a more general approach to CBV CPS:
-CPS transformer. We take any implementation of Symantics (interpreter,
+a CPS transformer. We take any implementation of |Symantics| (interpreter,
 compiler, partial evaluator) and make it evaluate the CPS language
@@
 
-\section{Lifting and Functor}
+\subsection{Lifting and Functor}
 

### 2007-04-02 tagless/tagless-final.tex
more small scale edits

@@
 
-Because |RCN| has the signature Symantics, we can instantiate our previous
+Because |RCN| has the signature |Symantics|, we can instantiate our previous
 examples with it, and all works as expected.  More interesting
@@
 If we interpret |EXS| with the |R| interpreter of
-Section~\ref{S:interpreter-RL} realized in OCaml
+\S\ref{S:interpreter-RL} realized in OCaml
 \begin{code}
@@
 the evaluation of that argument before proceeding. This approach is in
-line with Reynolds (cite) (albeit in a typed setting). The interpreter RCV is
-useful for achieving the CBV evaluation of the object language
+line with \cite{ReynoldsXXX} (albeit in a typed setting). The interpreter RCV
+is useful for achieving CBV evaluation of the object language
 no matter whether the metalanguage is CBV or CBN.
@@
 compiler, partial evaluator) and make it evaluate the CPS language
-instead. We can view this transformation of interpreters as a
+instead. We can view interpreter transformation as a
 transformation on the object language, from direct to
@@
 The code literally establishes a mapping from CPS interpretations to
-(direct) interpretations, that latter performed by 
+(direct) interpretations, the latter performed by 
 the base interpreter |S|.
@@
 type |('c,..., (int -> 'w) -> 'w) S.repr|,
-the result of |lam (fun x -> add x (int 1))| has the type
+the result of |lam (fun x -> add x (int 1))| has type
 |('c,...((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr|. 
 Hence, to write the type equation for |CPST.repr| we need a 
-type case-discriminating
+type-discriminating
 function similar to the type function |repr_pe| of
-Section~\ref{S:PE-solution}. Alas, the function we need here is
+\S\ref{S:PE-solution}. Alas, the function we need here is
 similar but not identical to |repr_pe|. Thus we need to add 
@@
 stay unchanged, but the |repr| type equations in those structures have to
-account for a new (phantom, in that case) type variable.
-
-To save space, we show a simplified and less principal approach that
+account for a new (phantom) type variable.
+
+To save space, we show a simplified and less principled approach that
 lets us use the module |CPST| as it is. Because the module does not
@@
 as we used before. We just copied them over. That copying step is the
-price we pay for the simplified treatment. We shall encounter this
+price we pay for the simplified treatment. We shall also encounter this
 copying when discussing the self-interpreter in the next section. We
-will see then that the copying step has a profound meaning of
-`plugging the hole', which is one of the faces of polymorphism.
+will see then that the copying step is not frivolous, but represents
+a profound instance of 
+``plugging the hole'', which is one of the many faces of polymorphism.
 
@@
 which shows the naive CPS transform of the original code. The result is
-quite unoptimal with several apparent $\beta$-redexes. We can reduce
-them by instantiating |CPST| with the |P| evaluator instead. Our
-running test gives then
+quite unoptimal with several apparent $\beta$-redexes. 
+By instantiating |CPST| with the |P| evaluator instead, our
+running test gives the expected
 \begin{code}
@@
 \oleg{Should we have an example in the code?}
+\jacques{In the code, yes.  Probably not in the paper, it is getting 
+long as it is}
 
@@
 
-Taha et al (cite different paper) define a self-interpreter si to be an
+A self-interpreter |si| is an
 object-language term such that, for any object-language term e,
@@
 
-where |"e"| is an encoding of |e|.  Under the co-approach, we want to encode
-each object-language term as a function from a record of cogen methods
-to a generated rep.  (Thanks to currying, the object language need not
-actually support records.)  In the case of a self-interpreter, that
-record is simply si!  So we should define a self-interpreter si to be an
-object-language record-term such that, for any object-language term e,
+where |"e"| is an encoding of |e|.  In our approach, we want to encode
+each object-language term as a function from a record%
+\footnote{Thanks to currying, the object language need not actually support
+records.}  of interpreter methods to a generated representation.  
+In the case of a self-interpreter, that
+record is simply |si|!  So we should define a self-interpreter |si| to be an
+object-language term such that, for any object-language term |e|,
 
@@
 
-The self-interpreter then is just like the interpreter
-|R| in Section 2, except written at the object rather than meta level.
-
-Note on terms and special forms. In our class Symantics, |add|
-is not a term -- because it does not have the type repr t. We could
-have introduced |add| to be of the type |repr (Int->Int->Int)| -- but we
-didn't. So, |add| is a special form. It is to be applied in
-meta-language (using the Haskell application -- white space) rather
-than in the object language (app).
-
-Mention the RR interpreter: for any term E, (RR E) is equivalent to E.
+A self-interpreter is then just like the interpreter
+|R| in Section 2, except written at the object rather than the meta level.
+
+It is wortwhile noting that, in our class |Symantics|, |add|
+is not a term since it does not have type |repr t| for any |t|. We could
+have introduced |add| to be of type |repr (Int->Int->Int)| -- but we
+didn't. Thus |add| is a special form, which is to be applied in the
+meta-language using the meta-language application syntax (i.e. white
+space) rather than in the object language (i.e. via |app|).
+
+oleg{Mention the RR interpreter: for any term E, (RR E) is equivalent to E.
 RR is not an identity: it is an interpreter that encapsulates another
@@
 special form and interpret it always with itself (similar to let and
-hole below).
-
-Self-interpretation in our framework is
-a bit tricky due to a need for polymorphism.  An encoded term should
-be polymorphic in the interpreter type-constructor "r".  Further, the
-functions "lam", "app", etc. above should be polymorphic in the object
+hole below).}
+
+Self-interpretation in our framework is somewhat
+tricky due to the need for more polymorphism than is easily available
+in the meta-language.  An encoded term should
+be polymorphic in the interpreter type-constructor "r".  Furthermore, the
+functions "lam", "app", etc. should be polymorphic in the object
 types "a" and "b".  Thus the object language seems to require rank-2
@@
 take advantage of let-bound polymorphism in the metalanguage, it
-is crucial that we encode object "let" to meta "let".
+is crucial that we encode the object-level "let" using the meta "let".
 
@@
 object expressions can be expressed in the object language.  For
-example, unlike if we had simply defined |"E"| to be the `identity
-encoding' E, we \emph{can} write a size-measurer |SZ[]|, which like |SI[]| is
-an object expression with a hole |[]| waiting to be plugged with |"E"|.
+example, unlike if we had simply defined |"E"| to be the \emph{identity
+encoding} E, we \emph{can} write a size-measure interpreter |SZ[]|, 
+which like |SI[]| is
+an object expression with a hole |[]| waiting to be plugged with 
+an |"E"|.
 However, this notion of size cannot distinguish |let x = E1 in E2[x]|
-from |"E2[E1]"|.  Cf. one way to type-check polymorphic let:
+from |"E2[E1]"|.  Recall that one way to type-check polymorphic let is:
 \begin{code}
@@
 
-I guess the let-and-hole argument works out if we don't typecheck the
-encoded terms. We can't typecheck the interpreter with the whole
+\oleg{I guess the let-and-hole argument works out if we don't typecheck the
+encoded terms. We can't typecheck the interpreter with the hole
 either. We insert the encoded term into the interpreter -- and then
-typecheck the whole thing. That'll work. Yet a bit unsatisfactory...
+typecheck the whole thing. That'll work. Yet a bit unsatisfactory...}
 
@@
 into the number 4, as desired.  We can also encode the self-interpreter
-above as the following evaluation context in the object language.
+as the following evaluation context in the object language.
 \begin{code}
@@
 
-The 'result' of applying an interpreter to itself, needs to be an 
+The 'result' of applying an interpreter to itself needs to be an 
 interpreter too!  In fact, if it is Jones-optimal, then you should be 
 able to apply it to itself again and get the same answer.  
-Our code does that. First, |twice_inc_3_e| is the
-metalanguage encoding of the object term |si["e"]|
-where si is the self-interpreter of the object language (actually, only
-lam,app,add,int), |""| means the object-to-object encoding, and e is the
-object term
+Our code does that. First, let |twice_inc_3_e| be the
+metalanguage encoding of the object term |si["e"]|,
+where si is the self-interpreter for the needed components of the object
+language (only |lam|, |app|, |add|, and |int| in this case), |""| means the
+object-to-object encoding, and |e| is the object term
 \begin{code}
@@
 \end{code}
-
 Second, |twice_inc_3_ee| is the metalanguage encoding of the object term
@@
 which is the same as
-|si["si"["e"]]|
+|si["si"["e"]]|.
 The partial evaluator yields 5 on both |twice_inc_3_e| and 
@@
 
-``But how 
+\jacques{But how 
 to you create, in either Haskell or MetaOCaml, an untypechecked 
-interpreter-with-a-hole [UIH] ?''
-Well, one can make an argument that we already have such an
+interpreter-with-a-hole [UIH] ?}
+\oleg{Well, one can make an argument that we already have such an
 interpreter with polymorphic let and the hole: incope. In Haskell, the
@@
 the particular instance of Symantics). The unRR construction does
-this plugging in explicitly.
-Again, what is different from the above is that we \emph{can}
+this plugging in explicitly.}
+Again, what is different from the above is that we can
 typecheck terms separately, without inserting them first within the
@@
 
-Boxes go Bananas by Weirich and Washburn.
-
 \section{Conclusions}\label{conclusion}

### 2007-04-02 tagless/tagless-final.tex
Some simple BibTeX work

@@
-\hyphenation{meta-language meta-languages Meta-OCaml}
+\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar}
 \documentclass[preprint]{sigplanconf}
@@
 evaluating) |L e| is applied. Staging can be used to remove this
-interpretive overhead \cite[\S1.1--2]{WalidICFP02}.
+interpretive overhead \citep[\S1.1--2]{WalidICFP02}.
 
@@
 present stage as values of type |('c,t) code| where |'c| is the
-environment classifier \cite{WalidPOPL03,calcagno-ml-like}. Code values are created
+environment classifier \citep{WalidPOPL03,calcagno-ml-like}. Code values are created
 by a \emph{bracket} form |.<e>.|, which specifies that the expression |e| is to be
@@
 analysis.  To avoid the need for polymorphic lift, we turn to
-a technique due to Asai \ccshan{cite}: build a dynamic term
+\citearound{'s technique}\citet[see also \citealp{sumii-hybrid}]{asai-binding-time}:
+build a dynamic term
 alongside every static term.
@@
 call-by-value (CBV) metalanguage. The answer, given already by 
-\cite{ReynoldsPlotkin} is continuation-passing style (CPS), which they
+\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is continuation-passing style (CPS), which they
 specifically introduced 
@@
 Reynolds that typed setting would be nice?}\jacques{I think so, but from
-which paper? The one ReynoldsPlotkin paper is typed}. Specifically, we
+which paper? The one ReynoldsPlotkin paper is typed}\ccshan{I think Oleg
+meant ``It should also be possible to define languages with a highly
+ refined syntactic type structure. Ideally, such a treatment should be
+ metacircular, in
+ the sense that the type structure used in the defined language should be adequate for the
+ defining language'' \citep{reynolds-definitional}, but perhaps Jacques
+ meant another paper by ``ReynoldsPlotkin''?}. Specifically, we
 demonstrate CBN evaluation for our object language in OCaml, which is
@@
 
-The CBN CPS interpretation was introduced by Plotkin \cite{PlotkinCBN}. As in
+The CBN CPS interpretation was introduced by \citet{PlotkinCBN}. As in
 the more familiar CBV CPS, the interpretation of a value is a function
@@
 the evaluation of that argument before proceeding. This approach is in
-line with \cite{ReynoldsXXX} (albeit in a typed setting). The interpreter RCV
+line with \citet{reynolds-relation} (albeit in a typed setting). The interpreter RCV
 is useful for achieving CBV evaluation of the object language
@@
 
-\section{Self-interpretation}\cite{selfinterp}
+\section{Self-interpretation}\label{selfinterp}
 
@@
 
-oleg{Mention the RR interpreter: for any term E, (RR E) is equivalent to E.
+\oleg{Mention the RR interpreter: for any term E, (RR E) is equivalent to E.
 RR is not an identity: it is an interpreter that encapsulates another
@@
 \bibliographystyle{mcbride}
+\bibsep=0pt
 \bibliography{tagless}

### 2007-04-04 tagless/tagless-final.tex
Done with Section 4.3; started with Section 4.4 (simplified PE code shown)

@@
-\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar}
+\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar type-case}
 \documentclass[preprint]{sigplanconf}
@@
 type exp = V of var | B of bool
-         | L of exp | A of exp * exp | ...
+         | L of exp | A of exp * exp
 \end{code}
@@
 
-let rec eval0 env = function ...
+let rec eval0 env = function
 | V v       -> lookup env v
@@
 \begin{code}
-type u = UB of bool | UA of (u -> u) | ...
+type u = UB of bool | UA of (u -> u)
 \end{code}
@@
 \begin{code}
-let rec eval env = function ...
+let rec eval env = function
 | V v       -> lookup env v
@@
 \begin{code}
-let varZ (x,env)    = x
-let varS vp (x,env) = vp env
+let varZ env        = fst env
+let varS vp env     = vp (snd env)
 let b (bv:bool) env = bv
@@
 In the sequel, we encode binding in object programs using higher-order
-abstract syntax \citep{miller-manipulating,pfenning-higher-order}
+abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
 rather than de Bruijn indices.
@@
 constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
-Whereas our representation of object terms uses higher-order
-abstract syntax, our bytecode uses integer-named
+Whereas our representation of object terms uses HOAS,
+our bytecode uses integer-named
 variables to be realistic. We then define 
@@
 data P1 (a -> b) = S1 (P1 a -> P1 b)
-                 | E1 (C (a->b))
+                 | E1 (C (a -> b))
 
@@
 
-\subsection{GADT -- not quite tagless PE}
+\subsection{Eliminating tags from typecase}
 \label{S:PE-GADT}
 
-Two common ways of providing typecase in Haskell is by using either
-GADT or type-class functions
-(implemented via typeclasses with functional dependencies)
+Two common ways to provide typecase in Haskell are
+GADTs and type-class functional dependencies
 \citep{oliveira-typecase}.  These
-methods are equivalent, and here we show the GADT; please see |incope1.hs|
-in the accompanying source code for the latter.
-
-We introduce the following GADT
+methods are equivalent, and here we show the GADT way; |incope1.hs|
+in the accompanying source code shows the latter.
+
+We introduce a GADT with four data constructors.
 \begin{code}
@@
   VB :: Bool -> P Bool
-  VF :: (P a -> P b) -> P (a->b)
+  VF :: (P a -> P b) -> P (a -> b)
   E  :: C t -> P t
 \end{code}
-with four data constructors. The constructor |VF| stands out as its
-operand has type |P a -> P b| rather than |a -> b| as one may
-have expected if |P t| were parametric in |t|. As before, we define a
-function to extract the future-stage computation from the |P t| value:
+The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
+in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
+the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
+operand of type |P a -> P b| rather than |a -> b|. As before, we define a
+function to extract a future-stage computation from a value of type |P t|.
 \begin{code}
@@
 \end{code}
-
-We use the method |lam| of the |C| intrepreter to compile the argument
-of |lam| into a code value. We may now make |P| an instance of
-Symantics and implement the partial evaluator (the
-implementations of |mul|, |leq|, |if_| and |fix| are elided) as follows.
+The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
+case uses the method |lam| of the |C| interpreter to compile~|f|.
+
+We may now make |P| an instance of
+|Symantics| and implement the partial evaluator as follows. We elide
+|mul|, |leq|, |if_|, and |fix|.
 \begin{code}
@@
   bool b = VB b
-
-add e1 e2 = case (e1,e2) of
-             (VI n1,VI n2) -> VI (n1 + n2)
-             _ -> E (add (abstr e1) (abstr e2))
+  add (VI n1) (VI n2) = VI (n1 + n2)
+  add e1 e2 = E (add (abstr e1) (abstr e2))
   lam = VF
-app ef ea = case ef of
-             VF f -> f ea
-             E  f -> E (app f (abstr ea))
-\end{code}
-
-The implementations of |int|, |bool|, and |add| are similar to the
-above. The interpretation of |lam f| merely incorporates the
-HOAS function |f| into the result. At any point we can convert it to a
-code value, if we must. By delaying such a conversion, we get the
-opportunity to apply |f| to a concrete argument. When evaluating
-|app ef ea|, we check to see of the first argument is such a delayed
-lambda-expression. If it is, we apply the incorporated |f| to the
+  app (VF f) ea = f ea
+  app (E f)  ea = E (app f (abstr ea))
+\end{code}
+The implementations of |int|, |bool|, and |add| are like
+in~\S\ref{S:PE-problem}.
+The interpretation of |lam f| just wraps the
+HOAS function |f| into the result. We can always compile |f| to a code value,
+but we delay it to apply |f| to concrete arguments. The interpretation of
+|app ef ea| checks to see if |ef| is such a delayed
+HOAS function |VF f|. If it is, we apply |f| to the
 concrete argument |ea|, giving us a chance to perform static
-computations (see example |testpowfix| in the next subsection). If |ef| is a
-dynamic value, we residualize.
-
-Although GADTs solve our problem, the solution is not quite
-satisfactory. First of all, it cannot be ported to MetaOCaml as GADTs
-are not supported there. Second, the tagging problem reappeared, in
-the implementation of |app| above. The datatype |P t| has four data
-constructors; pattern-matching in |app| however uses only two of
-them, |VF| and |E|. It is obvious, one may say, that the constructors
-|VI| and |VB| just cannot occur, because the corresponding values
-cannot be function types. What is obvious to us is not obvious to
-the compiler. The case matching above looks \emph{partial}.  While this
-may seem like a minor point, 
-this is precisely the main point of
-tag elimination. In a typed tagged interpreter there are many
-pattern-matching forms that looks partial but are total in
-reality. The goal is to make this totality \emph{syntactically}
-apparent.
+computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
+dynamic value |E f|, we residualize.
+
+This solution using GADTs works but is not quite satisfactory. First, it
+cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
+the problem of nonexhaustive pattern\hyp matching reappears in |app|
+above: the type |P t| has four constructors, of which pattern\hyp
+matching in |app| uses only |VF| and~|E|. One may say that the
+constructors |VI| and |VB| obviously cannot occur because they do not
+construct values of type |P (a -> b)| as required by the type of |app|.
+Indeed the compiler could reason thus, but it may not; GHC for one
+issues warnings.  Although this point may seem minor, it is the heart of
+the tagging problem and the purpose of tag elimination. A typed tagged
+interpreter contains many pattern\hyp matching forms that look partial
+but are total in reality. The goal is to make this totality
+\emph{syntactically} apparent.
 
@@
 \label{S:PE-solution}
-Let us re-examine the problem in Section~\ref{S:PE-problem}. What we
+Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
 would ideally like is to write
 \begin{code}
-data P t = S (repr_pe t) (C t) | E (C t)
-\end{code}
-where |repr_pe| is a type function, defined inductively
+data P t = P (Maybe (repr_pe t)) (C t)
+\end{code}
+where |repr_pe| is the type function defined
 % inductively because P below depends on repr_pe
-as 
+by
 \begin{code}
@@
 repr_pe Bool     = Bool
-repr_pe (a->b) = P a -> P b
-\end{code}
-Although we can use typeclasses to define this type function
-in Haskell, that is not portable to MetaOCaml. We notice however that
-the type case alternatives above are all associated with existing
-methods of |Symantics|, and so the case analysis is already present. 
+repr_pe (a -> b) = P a -> P b
+\end{code}
+Although we can use type classes to define this type function
+in Haskell, that is not portable to MetaOCaml. However,
+these three typecase alternatives are already present in existing
+methods of |Symantics|.
 A simple and more portable solution thus emerges: we bake |repr_pe| 
-into the signature |Symantics|. Switching to MetaOCaml for
-portability, we recall from Section~\ref{encoding} that the |repr| type
-had two arguments |type ('c,'dv) repr|. We add an extra argument,
-|'sv|, which is the result of applying the |repr_pe| function above to the
-type |'dv|. Figure~\ref{fig:ocaml} shows the new signature |Symantics|.
+into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
+portability, we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
+constructor took two arguments |'c| and~|'dv|. We add an argument
+|'sv| for the result of applying |repr_pe| to~|'dv|.
+Figure~\ref{fig:ocaml} shows the new |Symantics| signature.
 \begin{figure*}
@@
   val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
+
 end
@@
 
-The interpreters |R|, |L| and |C| above only relied on the last
-argument of the |repr| type, which is unchanged. And so is their
-implementation; the newly added argument |'sv| will be phantom.
-For example, we would now write the interpreter |C| as
+\begin{figure}
+\begin{floatrule}
+\begin{code}
+module P = struct
+
+  type ('c,'sv,'dv) repr = {st: 'sv option;
+                            dy: ('c,'dv) code}
+
+  let abstr {dy = x} = x
+  let pdyn x = {st = None; dy = x}
+
+  let int  (x:int)  = {st = Some (R.int x);
+                       dy = C.int x}
+  let bool (x:bool) = {st = Some (R.bool x);
+                       dy = C.bool x}
+
+  let add e1 e2 = match e1, e2 with
+  | {st = Some 0}, e | e, {st = Some 0} -> e
+  | {st = Some m}, {st = Some n} -> int (R.add m n)
+  | _ -> pdyn (C.add (abstr e1) (abstr e2))
+
+  let if_ eb et ee = match eb with
+  | {st = Some b} -> if b then et () else ee ()
+  | _ -> pdyn (C.if_ (abstr eb) 
+                     (fun () -> abstr (et ()))
+                     (fun () -> abstr (ee ())))
+
+  let lam f =
+  {st = Some f; 
+   dy = C.lam (fun x -> abstr (f (pdyn x)))}
+
+  let app ef ea = match ef with
+  | {st = Some f} -> f ea
+  | _ -> pdyn (C.app (abstr ef) (abstr ea))
+
+  let fix f = 
+    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
+    in let rec self = function
+       | {st = Some _} as e -> app (f (lam self)) e
+       | e -> pdyn (C.app fdyn (abstr e))
+       in {st = Some self; dy = fdyn}
+
+end
+\end{code}
+\end{floatrule}
+\caption{Our partial evaluator (eliding \texttt{mul} and \texttt{leq})}
+\label{fig:pe}
+\end{figure}
+
+The interpreters |R|, |L| and~|C| above only use the old
+type arguments |'c| and~|'dv|, which are treated by the new signature
+in the same way.  Hence all that needs to change in these interpreters
+to match the new signature is to add a phantom type
+argument~|'sv| to~|repr|.
+For example, the compiler |C| now begins
 \begin{code}
@@
 \end{code}
-When writing the partial evaluator however, we make use of the
-type argument |'sv|. Here is the corresponding instance of
-|Symantics|:
-
-\begin{code}
-module P = struct
-  type ('c,'sv,'dv) repr = 
-        {st: 'sv option; dy: ('c,'dv) code}
-  let abstr {dy = x} = x
-  let pdyn x = {st = None; dy = x}
-
-  let int  (x:int)  = {st = Some (R.int x); dy = C.int x}
-  let bool (x:bool) = {st = Some (R.bool x); dy = C.bool x}
-
-  let build cast f1 f2 = fun e1 -> fun e2 -> match (e1,e2) with
-                   ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
-                 | _ -> pdyn (f2 (abstr e1) (abstr e2))
-  let buildsimp_monoid cast e f1 f2 = fun e1 e2 -> match (e1,e2) with
-                 | ({st = Some e'}, _) when e = e' -> e2
-                 | (_, {st = Some e'}) when e = e' -> e1
-                 | _ -> build cast f1 f2 e1 e2
-  let buildsimp_ring cast zero one f1 f2 = fun e1 e2 -> match (e1,e2) with
-                 | ({st = Some e'}, _) when e' = zero -> e1
-                 | (_, {st = Some e'}) when e' = zero -> e2
-                 | _ -> buildsimp_monoid cast one f1 f2 e1 e2
-
-  let add e1 e2 = buildsimp_monoid int 0 R.add C.add e1 e2
-  let mul e1 e2 = buildsimp_ring int 0 1 R.mul C.mul e1 e2
-  let leq e1 e2 = build bool R.leq C.leq e1 e2
-  let if_ eb et ee = match eb with
-           {st = Some b} -> if b then et () else ee ()
-         | _ -> pdyn (C.if_ (abstr eb) 
-                          (fun () -> abstr (et ()))
-                          (fun () -> abstr (ee ())))
-  let lam f = {st = Some f; 
-               dy = C.lam (fun x -> abstr (f (pdyn x)))}
-  let app ef ea = match ef with
-        {st = Some f} -> f ea
-      | _ -> pdyn (C.app (abstr ef) (abstr ea))
-
-  let fix f = 
-    let fdyn = C.fix (fun x -> abstr (f (pdyn x))) in
-    let fdynn e = pdyn (C.app fdyn e.dy) in
-    {st=Some (function {st = Some _} as e -> 
-               let rec self n = 
-                (match n.st with Some _->app (f (lam self)) n 
-                                | _ -> fdynn n) in 
-               app (f (lam self)) e
-             | e  -> fdynn e);
-    dy = fdyn }
-end;;
-\end{code}
-
-The type |repr| here literally expresses the type equation at the
-beginning of this section. The function |abstr|, as in the previous
-subsection, extracts the future-stage computation from the result of
-partial evaluation.  The function |pdyn| does the converse. As in
-section~\ref{S:PE-problem}, we build future-stage computations at the
-same time as we
-build present-stage ones, if any. Therefore, we avoid the need for a
-polymorphic `lift' function. The implementations of
+In contrast, the partial evaluator~|P| relies on the type argument |'sv|.
+
+Figure~\ref{fig:pe} shows the partial evaluator~|P|.
+Its type |repr| literally expresses the type equation for |repr_pe| above.
+The function |abstr|, as in~\S\ref{S:PE-GADT},
+extracts a future-stage code value from a result of
+partial evaluation.  Conversely, the function |pdyn| coerces a
+code value into a partial\hyp evaluation result. As
+in~\S\ref{S:PE-problem}, we build dynamic terms alongside
+static ones, if any, to avoid
+polymorphic lift. The implementations of
 |add| and |mul| are a bit more sophisticated, taking advantage 
@@
 is static and is the neutral element of the monoid structure.
-|buildsimp_ring| also worries about the full ring structure rather than
+|ring| also worries about the full ring structure rather than
 just the monoid structure.  These are put here to illustrate how we 

### 2007-04-04 tagless/incope.hs

@@ -1,4 +1,4 @@
-{-# OPTIONS -fglasgow-exts #-}
+{-# OPTIONS -fglasgow-exts -W #-}
 
 -- Interpreter, Compiler, Partial Evaluator
 
@@ -134,8 +134,8 @@ newtype L a = L Int deriving Show
 unL (L x) = x
 
 instance Symantics L where
-    int x  = L 1
-    bool b = L 1
+    int _  = L 1
+    bool _ = L 1
 
     lam f = L( unL (f (L 0)) + 1 )
     app e1 e2 = L( unL e1 + unL e2 + 1 )
@@ -348,9 +348,8 @@ instance Symantics P where
     -- lam :: (repr a -> repr b) -> repr (a->b)
     lam = VF
     -- app :: repr (a->b) -> repr a -> repr b
-    app ef ea = case ef of
-                        VF f -> f ea
-                        E  f -> E (app f (abstr ea))
+    app (VF f) ea = f ea
+    app (E f)  ea = E (app f (abstr ea))
 
     -- fix :: (repr a -> repr a) -> repr a
     {- use to:
@@ -364,26 +363,22 @@ instance Symantics P where
 
     fix f = pfix f -- need this charade for GADTs sake
 
-    add e1 e2 = case (e1, e2) of
-                 (VI n1, _) | (n1 == 0) -> e2
-                 (_, VI n2) | (n2 == 0) -> e1
-                 (VI n1, VI n2) -> VI (n1 + n2)
-                 _              -> E $ add (abstr e1) (abstr e2)
-    mul e1 e2 = case (e1,e2) of
-                 (VI n1, _) | (n1 == 0) -> e1
-                 (_, VI n2) | (n2 == 0) -> e2
-                 (VI n1, _) | (n1 == 1) -> e2
-                 (_, VI n2) | (n2 == 1) -> e1
-                 (VI n1,VI n2) -> VI (n1 * n2)
-                 _ -> E $ mul (abstr e1) (abstr e2)
+    add (VI 0) e = e
+    add e (VI 0) = e
+    add (VI n1) (VI n2) = VI (n1 + n2)
+    add e1 e2 = E (add (abstr e1) (abstr e2))
+    mul e@(VI 0) _ = e
+    mul _ e@(VI 0) = e
+    mul (VI 1) e = e
+    mul e (VI 1) = e
+    mul (VI n1) (VI n2) = VI (n1 * n2)
[…]

### 2007-04-04 tagless/tagless-final.tex
editorial changes to abstract and Sec 1. Most of the changes to the beginning of Sec 1.

@@
 types, nor generalized algebraic data types, nor a tag elimination 
-postprocessing step. It has been considered an open (impossible)
-problem to do so.
+postprocessing step. That has been an open problem.
 Our family of interpreters contains 
-an evaluator, a compiler, a partial
-evaluator, a term-size evaluator and a CPS call-by-name transformer.
-We take advantage of staging when available.
+an evaluator, a (byte-)compiler and a staged evaluator, a partial
+evaluator, a term-size evaluator and a CPS call-by-name and
+call-by-value transformers.
 
@@
 algebras for the wilder world of final algebras.
-(Or: we expose the co-algebraic structure of the object language).
+\oleg{Or: we expose the co-algebraic structure of the object language.}
 Our approach highlights the importance of order-2 type constructors
@@
 terms in various ways, and yet enought type information is available 
-to typecheck the terms and statically assure their interpretation will
-never get stuck.
-
-\oleg{If we talk about self-interpreter, mention let-bound polymorphism}
+to typecheck the terms and statically assure their interpretation
+never gets stuck.
+
+The encoding of a self-interpreter clarifies our technique 
+\oleg{and rank-2 polymorphism?} 
+as the
+interpretation \oleg{evaluation?} of a term plugged into the context
+of let-polymorphic definitions.
+\oleg{Discuss on Fri, along with the colclusions.}
 \end{abstract}
@@
 specific object languages because it supports rapid prototyping and integration
-with the host environment \citep{hudak-building} \oleg{perhaps say a
-  couple more words}.  If the metalanguage supports \emph{staging}, then
+with the host environment \citep{hudak-building}.
+If the metalanguage supports \emph{staging}, then
 the embedding can compile object programs to the metalanguage and avoid the
@@
 We focus on embedding a typed object language into a typed metalanguage.
-Types are useful as usual because they document the programmer's
-intentions and can be checked early (see~\S\ref{encoding}).
+The particular benefit is ruling out meaningless object terms, and
+so enabling faster interpretation. In addition, our interpreters
+assuredly do not get stuck.
 To be concrete, we use the typed object language in
@@
 we discuss these solutions, identify their drawbacks, then summarize our
-proposal and contributions.  No matter how we represent the object language in the
-metalanguage, the representation can be performed by hand or more
-conveniently by a parser or preprocessor such as camlp4.  We leave aside
-the issue of whether to program a parser or preprocessor using dependent
+proposal and contributions.  
+No matter how we represent the object language in the
+metalanguage, the representation can be performed by hand (e.g., by
+entering object terms at a metalanguage interpreter's prompt) or
+created by a parser/typechecker reading from a text string.
+We leave aside
+the issue of whether to program the parser/typechecker using dependent
 types \citep{WalidICFP02} or not \citep{baars-typing}.
+\oleg{perhaps say more affirmatively: the problem of writing such
+  a parser/interpreter has been solved, with dependent types (cite) and
+  without (cite), and so will not be discussed here.}
 
@@
 \end{code}
-We represent each variable using a unary de Bruijn index.
+We represent each variable using an unary de Bruijn index.
 For example, we represent the object term $(\fun{x}x)\True$ as
@@
 The unfortunate tag |UB| in the result reflects that |eval| is a partial
-function, in two ways.  First, the pattern match |with UA f| in the line
+function.  First, the pattern match |with UA f| in the line
 |A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
@@
 data constructors.  The interpreter below provides these functions.
-\jacques{but now this smaller language does not match anymore what
-we wrote earlier.  I think it is better to use a small language here,
-but i don't quite know how to say so}
 \begin{code}
@@
 \end{code}
-The result has no tags---the interpreter patently uses no tags and no
+The result has no tags: the interpreter patently uses no tags and no
 pattern matching. The term |b true| evaluates to a boolean and the term
@@
 in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
-term) are reported at compile time. Because the new interpreter uses no
+term) are reported at compile time. 
+\oleg{Jacques, do we use final and initial here correctly?}
+Because the new interpreter uses no
 universal type or
@@
     \item transform the term to continuation\hyp passing style (CPS),
-        including call-by-name (CBN).
+        including call-by-name (CBN) CPS, so to isolate the evaluation
+        order of the object language from that of the metalanguage.
 \end{itemize*}
@@
 closed, yet is surprisingly simple and obvious in hindsight.  We stress
-that until now our achievement has been thought impossible: to
+that until now it has been considered impossible to
 interpret a typed object language in a typed metalanguage without
@@
 \item a functor signature that encompasses all our interpreters, from
-    evaluation and compilation to partial evaluation and CBN CPS
-    transformation;
+    evaluation and compilation to partial evaluation and CBN CPS;
 \item a clear way to extend the object language with more features;
@@
 
-The structure of our paper is as follows. Section XXX ...
-We finally conclude.
-\jacques{TODO}
+The structure of our paper is as follows. In the next Section, we
+define the OCaml and Haskell interfaces of the interpreter for our
+language and describe two instances, evaluating a term to a
+metalanguage value and computing term's size. In
+Section~\ref{compiler}, we stage the evaluator, obtaining the tagless
+compiler. Section~\ref{PE} describes partial evaluation, in
+particular, the problem posed by the need to delay the 
+binding-time analysis of term abstractions and the apparent 
+need for a typecase. We then show two solutions. The first, imperfect
+one, uses GADT. The better and more portable solution bakes in the
+typecase idiom into the interpreter interface, eliminating the need for
+advanced type systems and permitting implementations in Haskell98 or OCaml.
+The discussion Section~\ref{discussion} presents, inter alia, CBN CPS
+interpreter and the CBV CPS transformer. We then describe 
+self-interpretation (Section~\label{selfinterp}),
+discuss related work (Section~\ref{related}) and then conclude.
 

### 2007-04-04 tagless/tagless-final.tex
start crafting some language on related work

@@
 \section{Related work}\label{related}
-Walid's Jones optimality paper. See his own ICFP02 paper that
-described what he achieved then and what he \emph{did not}. He did not
-eliminate all the tags. Ken's and mine impression of it:
-(i) we can't write a \emph{direct} interpreter of the type language
-           -- the only known to them, and \emph{partial} solution, was 
-              that by Zhe Yang;
-(ii) we have to use indirect interpreter, which needs the universal
-           type and hence tagging;
-(iii) that means any self-interpreter must have tags
-(iv) that means not Jones-optimal.
-It seems we should attack the point1 (in which case we don't have to
-bother with self-interpreters). We say that we \emph{can} write the direct
-interpreter, and without GADT or their equivalents (typeclasses).
-Meta-language computes types, and the compiler and interpreter respect
-those types and needs no tags.
-
-Walid's ICFP02
-
-Walid's PEPM07 (Concoqtion paper)? At leats at the Concoqtion
-presentation at WG2.11, Emir used tagless interpreter as the main and
-only example. \jacques{but at the PEPM07 presentation, Walid did not
-quite do it that way.  He presented Concoqtion more as an exploration
-of (an obvious but difficult) part of the design space for programming
-languages.}
-
-Danvy's self-interpreter
-Ken has read Danvy et al's paper and said that the paper
-doesn't seem to even make the claim that the source language is
-typed. The interpreter is typed, but the source language doesn't seem
-to be.
-In sharp contrast to Danvy's paper, all the evaluators in our case
-cannot produce any pattern-match error! Except for the partial
-evaluator, there are NO patterns to match. And this is critical for
-tag elimination -- pattern matching means there are no tags. And in
-PE, the pattern-match deals with phases rather than with types. Also,
-in our case the exhaustiveness of the pattern match is apparent to the
-compiler -- that's why there are no tags.
-
-Simon Peyton-Jones GADTs (the motivation for GADT is tagless
-interpreter, but only for the first-order language)
-
-Xi's POPL03 (He uses a higher-order language, one of the motivations
-for his GRDT)
+
+Our initial motivation came from several papers 
+\citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} which used
+embedded interpreters as a central example to justify looking 
+into advanced type systems.  While we admired all this technical work,
+there remained the issue that we were not convinced of the strength
+of their motivating example.
+
+\citet{WalidICFP02}, when referring to the work of \citet{taha-tag},
+claims that this shows that self-interpretation is possible in a
+\emph{simply typed $\lambda$-calculus}, but this is somewhat misleading
+as their type system does not differentiate between the various ground
+types.  In other words, various constants (like \textsf{nil} and 
+\textsf{true}) both inhabit the same type $D$.  This is not very satisfying.
+
+\oleg{He did not eliminate all the tags.}\jacques{can you show me?}
+From reading \citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}, our 
+impression is that the claim is:
+\begin{enumerate}
+\item One needs to encode a typed language in a typed language based on
+a sum type,
+\item It is not possible to write a \emph{direct} interpreter for an typed
+language encoded this way, in a typed language, without either using a
+very advanced type system or incurring tagging/untagging overhead.
+\item Thus we have to use an indirect interpreter, which needs a universal
+  type (and hence tagging),
+\item Thus any self-interpreter must have tags.
+\item Hence the self-interpreter cannot be Jones-optimal.
+\end{enumerate}
+While the logic is sound, we showed that the premise in the first step
+was not valid.
+
+\oleg{the only known to them, and \emph{partial} solution, was 
+  that by Zhe Yang;}\jacques{I cannot find that reference}
+
+\citet{Danvy-tagging-encoding} discusses Jones optimality at length.
+There, the authors demonstrate the usefulness of HOAS for typed
+self-interpretation.  However, no claim is made that the source program
+is itself typed -- just that the interpreter is.  Our reading is that
+the source language is not typed.  However, in sharp contrast with 
+their interpreter, none of our evaluators can product pattern-match errors
+(unlike the code that accompanies \citet{Danvy-tagging-encoding}).
+In other words, their language (and interpreter) does use tags, however
+the shift to HOAS allows the partial evaluator to remove all the tags.
+In contrast, we have no tags at all.
 

### 2007-04-04 tagless/tagless-final.tex
Done with Section 4

@@
 \usepackage{mdwlist}
+\usepackage{dblfloatfix}
 
@@
 Our family of interpreters contains 
-an evaluator, a (byte-)compiler and a staged evaluator, a partial
-evaluator, a term-size evaluator and a CPS call-by-name and
-call-by-value transformers.
+an evaluator, a (byte-)compiler and staged evaluator, a partial
+evaluator, a term-size evaluator, and call-by-name and
+call-by-value CPS transformers.
 
@@
 We focus on embedding a typed object language into a typed metalanguage.
-The particular benefit is ruling out meaningless object terms, and
-so enabling faster interpretation. In addition, our interpreters
-assuredly do not get stuck.
+The benefit of types in this setting is to rule out meaningless object terms,
+thus enabling faster interpretation and assuring that our interpreters
+do not get stuck.
 To be concrete, we use the typed object language in
@@
 No matter how we represent the object language in the
-metalanguage, the representation can be performed by hand (e.g., by
+metalanguage, the representation can be created either by hand (for example, by
 entering object terms at a metalanguage interpreter's prompt) or
-created by a parser/typechecker reading from a text string.
-We leave aside
-the issue of whether to program the parser/typechecker using dependent
-types \citep{WalidICFP02} or not \citep{baars-typing}.
-\oleg{perhaps say more affirmatively: the problem of writing such
-  a parser/interpreter has been solved, with dependent types (cite) and
-  without (cite), and so will not be discussed here.}
-
-\subsection{The Tag Problem}\label{tagproblem}
+by a parser or type checker reading from a text string.
+We leave aside the solved problem of writing such a parser or type checker,
+whether using dependent types \citep{WalidICFP02} or not \citep{baars-typing}.
+
+\subsection{The tag problem}\label{tagproblem}
 
@@
 \end{code}
-We represent each variable using an unary de Bruijn index.
+We represent each variable using a unary de Bruijn index.
 For example, we represent the object term $(\fun{x}x)\True$ as
@@
 
-\subsection{Solutions Using Fancier Types}
+\subsection{Solutions using fancier types}
 
@@
 
-\subsection{Our Final Proposal}\label{ourapproach}
+\subsection{Our final proposal}\label{ourapproach}
 
@@
 \item a functor signature that encompasses all our interpreters, from
-    evaluation and compilation to partial evaluation and CBN CPS;
+    evaluation and compilation to partial evaluation and CPS transformation;
 \item a clear way to extend the object language with more features;
@@
 interpreter and the CBV CPS transformer. We then describe 
-self-interpretation (Section~\label{selfinterp}),
+self-interpretation (Section~\ref{selfinterp}),
 discuss related work (Section~\ref{related}) and then conclude.
@@
 
-The functor below encodes our examples |test1| and $\mathit{power}$.
+The functor below encodes our running examples |test1| and $\mathit{power}$.
 \begin{code}
@@
 gives the code value |.<(fun x_6 -> x_6) (true)>.|
-of inferred type |('c, bool) C.repr|.
+of inferred type |('c, bool) C.repr|.  Similarly, interpreting |testpowfix7|
+with
+\begin{code}
+let module E = EX(C) in E.testpowfix7
+\end{code}
+gives the code value
+\begin{code}
+.<fun x_1 -> (fun x_2 ->
+  let rec self_3 = fun n_4 ->
+   (fun x_5 -> if x_5 <= 0 then 1 
+               else x_2 * (self_3 (x_5 + (-1))))
+   n_4
+  in self_3) x_1 7>.
+\end{code}
+with many apparent $\beta$-redexes.
 
@@
 dynamic terms.
-
-\ccshan{Move later:} The
-partial evaluator does have tags: \emph{binding time} or \emph{phase}
-tags, i.e. whether a value is static or dynamic.
-Pattern-matching on this phase tag is always exhaustive.
 
@@
 
+\begin{figure*}
+\begin{floatrule}
+\begin{code}
+module type Symantics = sig
+
+  type ('c,'sv,'dv) repr
+
+  val int : int  -> ('c,int,int) repr
+  val bool: bool -> ('c,bool,bool) repr
+
+  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
+  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
+
+  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
+  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
+
+end
+\end{code}
+\end{floatrule}
+\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
+\label{fig:ocaml}
+\end{figure*}
+
+\begin{figure}
+\begin{floatrule}
+\begin{code}
+module P = struct
+
+  type ('c,'sv,'dv) repr = {st: 'sv option;
+                            dy: ('c,'dv) code}
+
+  let abstr {dy = x} = x
+  let pdyn x = {st = None; dy = x}
+
+  let int  (x:int)  = {st = Some (R.int x);
+                       dy = C.int x}
+  let bool (x:bool) = {st = Some (R.bool x);
+                       dy = C.bool x}
+
+  let add e1 e2 = match e1, e2 with
+  | {st = Some 0}, e | e, {st = Some 0} -> e
+  | {st = Some m}, {st = Some n} -> int (R.add m n)
+  | _ -> pdyn (C.add (abstr e1) (abstr e2))
+
+  let if_ eb et ee = match eb with
+  | {st = Some b} -> if b then et () else ee ()
+  | _ -> pdyn (C.if_ (abstr eb) 
+                     (fun () -> abstr (et ()))
+                     (fun () -> abstr (ee ())))
+
+  let lam f =
+  {st = Some f; 
+   dy = C.lam (fun x -> abstr (f (pdyn x)))}
+
+  let app ef ea = match ef with
+  | {st = Some f} -> f ea
+  | _ -> pdyn (C.app (abstr ef) (abstr ea))
+
+  let fix f = 
+    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
+    in let rec self = function
+       | {st = Some _} as e -> app (f (lam self)) e
+       | e -> pdyn (C.app fdyn (abstr e))
+       in {st = Some self; dy = fdyn}
+
+end
+\end{code}
+\end{floatrule}
+\caption{Our partial evaluator (\texttt{mul} and \texttt{leq} are elided)}
+\label{fig:pe}
+\end{figure}
+
 \subsection{The ``final'' solution}
@@
 Figure~\ref{fig:ocaml} shows the new |Symantics| signature.
-\begin{figure*}
-\begin{floatrule}
-\begin{code}
-module type Symantics = sig
-
-  type ('c,'sv,'dv) repr
-
-  val int : int  -> ('c,int,int) repr
-  val bool: bool -> ('c,bool,bool) repr
-
-  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
-  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
-
-  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
-  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
-
-end
-\end{code}
-\end{floatrule}
-\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
-\label{fig:ocaml}
-\end{figure*}
-
-\begin{figure}
-\begin{floatrule}
-\begin{code}
-module P = struct
-
-  type ('c,'sv,'dv) repr = {st: 'sv option;
-                            dy: ('c,'dv) code}
-
-  let abstr {dy = x} = x
-  let pdyn x = {st = None; dy = x}
-
-  let int  (x:int)  = {st = Some (R.int x);
-                       dy = C.int x}
-  let bool (x:bool) = {st = Some (R.bool x);
-                       dy = C.bool x}
-
-  let add e1 e2 = match e1, e2 with
-  | {st = Some 0}, e | e, {st = Some 0} -> e
-  | {st = Some m}, {st = Some n} -> int (R.add m n)
-  | _ -> pdyn (C.add (abstr e1) (abstr e2))
-
-  let if_ eb et ee = match eb with
-  | {st = Some b} -> if b then et () else ee ()
-  | _ -> pdyn (C.if_ (abstr eb) 
-                     (fun () -> abstr (et ()))
-                     (fun () -> abstr (ee ())))
-
-  let lam f =
-  {st = Some f; 
-   dy = C.lam (fun x -> abstr (f (pdyn x)))}
-
-  let app ef ea = match ef with
-  | {st = Some f} -> f ea
-  | _ -> pdyn (C.app (abstr ef) (abstr ea))
-
-  let fix f = 
-    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
-    in let rec self = function
-       | {st = Some _} as e -> app (f (lam self)) e
-       | e -> pdyn (C.app fdyn (abstr e))
-       in {st = Some self; dy = fdyn}
-
-end
-\end{code}
-\end{floatrule}
-\caption{Our partial evaluator (eliding \texttt{mul} and \texttt{leq})}
-\label{fig:pe}
-\end{figure}
 
@@
 extracts a future-stage code value from a result of
-partial evaluation.  Conversely, the function |pdyn| coerces a
-code value into a partial\hyp evaluation result. As
+partial evaluation.  Conversely, the function |pdyn| injects a
+code value into partial evaluation. As
 in~\S\ref{S:PE-problem}, we build dynamic terms alongside
-static ones, if any, to avoid
-polymorphic lift. The implementations of
-|add| and |mul| are a bit more sophisticated, taking advantage 
-respectively of the implicit monoid and ring structure of the 
-underlying domain.
-The function |build| checks to see if
-its two arguments |e1| and |e2| are static expressions. If so, the
-corresponding static computation (with the help of the intrepreter
-|R|) is performed. The function |buildsim_monoid| also checks if one argument
-is static and is the neutral element of the monoid structure.
-|ring| also worries about the full ring structure rather than
-just the monoid structure.  These are put here to illustrate how we 
-would go about writing general algebraic simplification routines that
-would mimic various compiler optimizations, and these would be quite
-beneficial in a large language with more base types.
-
-When doing partial evaluation on |fix|, we made a deliberate decision to `go
-all the way' -- so long as the computation can proceed at the PE stage,
-we do that rather than residualizing. In the
-accompanying source code, we have also implemented an alternative
-where |fix| is unrolled only once and then we residualize. We prefer the
-first approach, even though it may cause
-non-termination at the PE stage -- but \emph{only} if the computation
-we are partially evaluating is non-terminating, and enough static
-information is available to trigger this non-termination.  Our
-treatment of |fix| may lead to code bloat in the residual
-program. The techniques of~\citet{SwadiTahaKiselyovPasalic2006} can
-be used to deal with this issue.
-
-Given the above Symantics, our running test now evaluates
-\begin{code}
-let module E = EX(P) in E.test1 ();;
-\end{code}
-to
-\begin{code}
-- : ('a, bool, bool) P.repr = 
-   {P.st = Some true; P.dy = .<(true)>.}
-\end{code}
-It is instructive to compare this result with that of the |C|
-evaluator. We observe that the beta-redex has now been 
-performed. The result is the static value |true|.
-More interesting is the partial evaluation of testpowfix7. The |C|
-interpreter evaluates |testpowfix7| to the future-stage code with many
-apparent beta-redexes.
-\begin{code}
-let module E = EX(C) in E.testpowfix7;;
- - : ('_a, int -> int) C.repr =
-.<fun x_1 -> (((fun x_2 ->
-  let rec self_3 =
-   fun n_4 ->
-    ((fun x_5 -> if (x_5 <= 0) then 1 
-                 else (x_2 * (self_3 (x_5 + (-1)))))
-      n_4) in
-  self_3) x_1) 7)>.
-\end{code}
-The result of the |P| interpreter is different:
-
-\begin{code}
-let module E = EX(P) in E.testpowfix7;;
- - : ('_a, ('_a, int, int) P.repr -> ('_a, int, int) P.repr, 
-      int -> int) P.repr = 
+any static ones to avoid polymorphic lift.
+
+To illustrate how to add optimizations, we improve |add| (and |mul|,
+elided) to simplify the generated code using the monoid (and ring)
+structure of~|int|: not only is addition performed statically
+(using~|R|) when both operands are statically known, but it is
+eliminated when one operand is statically~$0$; similarly for
+multiplication by~$0$ or~$1$.  Such algebraic simplifications are easy
+to abstract over the specific domain (such as monoid or ring) where they
+apply.  These simplifications and abstractions help a lot in a large
+language with more base types and primitive operations.
+
+Any partial evaluator must decide how much to unfold recursion
+statically: unfolding too little can degrade the residual code, whereas
+unfolding too much risks nontermination.  Our partial evaluator is no
+exception, because our object language includes |fix|.  The code in
+Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
+way'', that is, unfold |fix| rather than residualize whenever the
+argument is static.  In the accompanying source code is a conservative
+alternative |P.fix| that unrolls recursion only once, then residualizes.
+Many sophisticated approaches have been developed to make this trade-off
+\citep{jones-partial}, but this issue is orthogonal to our presentation.
+A separate concern in our treatment of |fix| is possible code bloat in
+the residual program, which calls for let-insertion
+\citep{SwadiTahaKiselyovPasalic2006}.
+
+Given this implementation of~|P|, our running example
+\begin{code}
+let module E = EX(P) in E.test1 ()
+\end{code}
+evaluates to
+\begin{code}
+{P.st = Some true; P.dy = .<true>.}
+\end{code}
+of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
+a $\beta$-reduction has been statically performed to yield |true|.  More
+interestingly, whereas |testpowfix7| compiles to a code value with many
+$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
+\begin{code}
+let module E = EX(P) in E.testpowfix7
+\end{code}
+gives
+\begin{code}
 {P.st = Some <fun>;
-  P.dy = .<fun x_1 -> 
-     (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * x_1))))))>.}
-\end{code}
-as expected from the power example.
-
-Unlike the GADT approach in the previous subsection, there is no
-longer any seemingly partial pattern-matching. All pattern-matching
-occurs in the partial evaluator only (to check if the current value 
-is static or dynamic), and is 
-phase rather than a type tag match. Furthermore the matching
-\emph{patently} exhaustive. That is, totality is assured and it is
-assured \emph{syntactically}, in a way that is apparent to the
-compiler.
-
-One should note that the partial evaluator has fully reused the code
-of the compiler and the interpreter. It is truly the `composition' of the
-interpreter and the compiler.   This is a simpler situation than
-in \citet{SperberThiemann:TwoForOne} where a partial evaluator and
-a compiler are `composed', but the general ideas are similar.
-
-Lastly, we should comment on the approach taken in the partial evaluator.
-It is an online, typed, polyvariant partial evaluator.  It has no
-termination analysis, and as written may diverge (but only on divergent
-programs) as we have decided to aggressively unroll loops.
+ P.dy = .<fun x_1 -> x_1 * (x_1 * (x_1 * (x_1 *
+                    (x_1 * (x_1 * x_1)))))>.}
+\end{code}
+as desired.
+
+Unlike the GADT approach in~\S\ref{S:PE-GADT}, all pattern\hyp matching
+in~|P| is \emph{syntactically} exhaustive, so it is \emph{patent} to the
+compiler that |P| is total.  Further, all pattern\hyp matching occurs
+during partial evaluation, only to check if a value is known statically,
+never what type it has.  In other words, our partial evaluator tags
+phases (with |Some| and |None|) but not object types.
+
+Our typed partial evaluator is online and polyvariant.  It reuses the
+compiler~|C| and the evaluator~|R| by composing them.  This situation is
+simpler than \citets{SperberThiemann:TwoForOne} composition of a partial
+evaluator and a compiler, but the general ideas are similar.
 
@@
 
-\subsection{Lifting and Functor}
+\subsection{Lifting and functor}
 

### 2007-04-04 tagless/tagless-final.tex
Move figures closer to exposition

@@
 
+\begin{figure*}
+\begin{floatrule}
+\begin{code}
+module type Symantics = sig
+
+  type ('c,'sv,'dv) repr
+
+  val int : int  -> ('c,int,int) repr
+  val bool: bool -> ('c,bool,bool) repr
+
+  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
+  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
+
+  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
+  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
+
+end
+\end{code}
+\end{floatrule}
+\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
+\label{fig:ocaml}
+\end{figure*}
+
+\begin{figure}
+\begin{floatrule}
+\begin{code}
+module P = struct
+
+  type ('c,'sv,'dv) repr = {st: 'sv option;
+                            dy: ('c,'dv) code}
+
+  let abstr {dy = x} = x
+  let pdyn x = {st = None; dy = x}
+
+  let int  (x:int)  = {st = Some (R.int x);
+                       dy = C.int x}
+  let bool (x:bool) = {st = Some (R.bool x);
+                       dy = C.bool x}
+
+  let add e1 e2 = match e1, e2 with
+  | {st = Some 0}, e | e, {st = Some 0} -> e
+  | {st = Some m}, {st = Some n} -> int (R.add m n)
+  | _ -> pdyn (C.add (abstr e1) (abstr e2))
+
+  let if_ eb et ee = match eb with
+  | {st = Some b} -> if b then et () else ee ()
+  | _ -> pdyn (C.if_ (abstr eb) 
+                     (fun () -> abstr (et ()))
+                     (fun () -> abstr (ee ())))
+
+  let lam f =
+  {st = Some f; 
+   dy = C.lam (fun x -> abstr (f (pdyn x)))}
+
+  let app ef ea = match ef with
+  | {st = Some f} -> f ea
+  | _ -> pdyn (C.app (abstr ef) (abstr ea))
+
+  let fix f = 
+    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
+    in let rec self = function
+       | {st = Some _} as e -> app (f (lam self)) e
+       | e -> pdyn (C.app fdyn (abstr e))
+       in {st = Some self; dy = fdyn}
+
+end
+\end{code}
+\end{floatrule}
+\caption{Our partial evaluator (\texttt{mul} and \texttt{leq} are elided)}
+\label{fig:pe}
+\end{figure}
+
 Two common ways to provide typecase in Haskell are
@@
 construct values of type |P (a -> b)| as required by the type of |app|.
-Indeed the compiler could reason thus, but it may not; GHC for one
+Indeed the metalanguage implementation could reason thus, but it may not; GHC for one
 issues warnings.  Although this point may seem minor, it is the heart of
@@
 \emph{syntactically} apparent.
-
-\begin{figure*}
-\begin{floatrule}
-\begin{code}
-module type Symantics = sig
-
-  type ('c,'sv,'dv) repr
-
-  val int : int  -> ('c,int,int) repr
-  val bool: bool -> ('c,bool,bool) repr
-
-  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
-  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
-
-  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
-  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
-
-end
-\end{code}
-\end{floatrule}
-\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
-\label{fig:ocaml}
-\end{figure*}
-
-\begin{figure}
-\begin{floatrule}
-\begin{code}
-module P = struct
-
-  type ('c,'sv,'dv) repr = {st: 'sv option;
-                            dy: ('c,'dv) code}
-
-  let abstr {dy = x} = x
-  let pdyn x = {st = None; dy = x}
-
-  let int  (x:int)  = {st = Some (R.int x);
-                       dy = C.int x}
-  let bool (x:bool) = {st = Some (R.bool x);
-                       dy = C.bool x}
-
-  let add e1 e2 = match e1, e2 with
-  | {st = Some 0}, e | e, {st = Some 0} -> e
-  | {st = Some m}, {st = Some n} -> int (R.add m n)
-  | _ -> pdyn (C.add (abstr e1) (abstr e2))
-
-  let if_ eb et ee = match eb with
-  | {st = Some b} -> if b then et () else ee ()
-  | _ -> pdyn (C.if_ (abstr eb) 
-                     (fun () -> abstr (et ()))
-                     (fun () -> abstr (ee ())))
-
-  let lam f =
-  {st = Some f; 
-   dy = C.lam (fun x -> abstr (f (pdyn x)))}
-
-  let app ef ea = match ef with
-  | {st = Some f} -> f ea
-  | _ -> pdyn (C.app (abstr ef) (abstr ea))
-
-  let fix f = 
-    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
-    in let rec self = function
-       | {st = Some _} as e -> app (f (lam self)) e
-       | e -> pdyn (C.app fdyn (abstr e))
-       in {st = Some self; dy = fdyn}
-
-end
-\end{code}
-\end{floatrule}
-\caption{Our partial evaluator (\texttt{mul} and \texttt{leq} are elided)}
-\label{fig:pe}
-\end{figure}
 
@@
 Unlike the GADT approach in~\S\ref{S:PE-GADT}, all pattern\hyp matching
-in~|P| is \emph{syntactically} exhaustive, so it is \emph{patent} to the
-compiler that |P| is total.  Further, all pattern\hyp matching occurs
+in~|P| is \emph{syntactically} exhaustive, so it is patent to the metalanguage
+implementation that |P| is total.  Further, all pattern\hyp matching occurs
 during partial evaluation, only to check if a value is known statically,

### 2007-04-05 tagless/tagless-final.tex
small changes to the contributions

@@
 tagging or type\hyp system extensions.  For example, \citet{taha-tag}
-states that ``expressing such an interpreter in a statically typed
+state that ``expressing such an interpreter in a statically typed
 programming language is a rather subtle matter. In fact, it is only
@@
 \begin{enumerate}
-\item to evaluate, compile, and partially evaluate a typed object language
-   in a typed metalanguage with a simple type system;
-\item to check statically for well-typed and closed object programs;
+\item to evaluate in direct or continuation-passing styles, 
+  compile, and partially evaluate a typed higher-order object language
+   in a typed metalanguage with a simple type system. All these
+   interpreters have no type tags and never get stuck;
+\item a Hindley-Milner--typed partial evaluator with no 
+  polymorphic lift operation and with
+  the delayed binding-time analysis of functions . The partial
+  evaluator needs no advanced type systems such as GADT or intentional
+  type analysis;
+\item to check statically for well-typed and closed object programs
+      using the type system \oleg{typechecker?} of the metalanguage;
 \item clean, comparable implementations in MetaOCaml and Haskell;
 \item a functor signature that encompasses all our interpreters, from
-    evaluation and compilation to partial evaluation and CPS transformation;
-\item a clear way to extend the object language with more features;
+    evaluation and compilation to partial evaluation and CPS;
+\item a clear way to extend the object language with more features
+  such as state;
 \item and an approach to self\hyp interpretation compatible with the
@@
 \end{enumerate}
-
-\jacques{we should probably be more explicit about the contributions of
-the discussion section}
 
@@
 
-In the sequel, we encode binding in object programs using higher-order
+In contrast to \S\ref{tagproblem}, we encode 
+binding in object programs using higher-order
 abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
-rather than de Bruijn indices.
+rather than de Bruijn indices. This makes the encoding convenient and
+also ensures that our object programs are closed.
 
@@
 and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
-signature that suffices for this and the following sections.
+signature that suffices until \S\ref{PE}.
 Two differences between this signature and the |Symantics| class in

### 2007-04-05 tagless/tagless-final.tex
Comments on Sec 2 and 3

@@
 
+The methods |add|, |mul|, and |leq| are quite similar, and so are
+|int| and |bool|. Therefore, we will frequently show implementations of
+only one method of each group and elide the rest, to save space. The
+accompanying code has the complete code.
+
 Because the types in the |Symantics| class reflect the object language's type
 system, this representation admits only well-typed object terms for evaluation.
+\oleg{replace the above with the following?}
+The comparison of |Symantics| with Figure~\ref{fig:object} clearly
+shows that every typed, closed object term can be represented in the
+metalanguage. Moreover,
+\begin{prop}
+If an object term has the object type |t|, its
+representation in the metalanguage has the type 
+|forall repr. Symantics repr => repr t|
+\oleg{or, has the type repr t for some repr. Meta-language embeds all
+  object types.}
+\end{prop}
 If we make a mistake somewhere, say replace |int 7| with |bool True| in
@@
 only at compile time, not at run time.  Pattern matching against~|R|
-cannot ever fail and is compiled away.
-In both OCaml and Haskell, it is obvious to the compiler that
+cannot ever fail and is assuredly compiled away.
+In OCaml, too, it is obvious to the compiler that
 pattern matching cannot fail, because there is no
@@
 indefinitely or terminates with a value of the same type~|t|.
+\end{prop}
+In general, 
+\begin{prop}
+  If the definitional interpreter implementing |Symantics| 
+  in the soundly typed metalanguage never gets stuck,
+  the type system of the object
+  language is sound with respect to the dynamic semantics defined by
+  the interpreter.
 \end{prop}
@@
 
+The overarching theme of our approach -- representation of object terms in 
+\S\ref{ourapproach}, future-stage code in \S\ref{S:PE-lift} and static
+values in \S\ref{S:PE-solution} and CPS values in \S\ref{S:CPS} --
+demonstrates that pushing the computation towards the producers
+\oleg{or, constructors. Ken said: functionalization} generally
+requires simpler type systems (such as mere HM). The same idea appears
+in many other areas (e.g., ammortized heap analysis, TFP/TLCA paper).
+\oleg{I was told that speculation and informality is permitted in
+  conclusions.}
+
 Our approach showed better the meaning of GADT and when they can be

### 2007-04-05 tagless/tagless-final.tex
Minor editorial changes

@@
 universal type or
-pattern matching, it never results in a run-time error and is in fact
-total.  Because this totality is obvious not just to us but also to the
+pattern matching, it never results in a run-time error (and is in fact
+total).  Because this safety is obvious not just to us but also to the
 metalanguage implementation, we avoid the serious performance penalty
@@
 
-Our contributions are
+Our contributions are as follows.
 \begin{enumerate}
-\item to evaluate in direct or continuation-passing styles, 
+\item We build interpreters that evaluate (in direct or continuation\hyp passing styles),
   compile, and partially evaluate a typed higher-order object language
-   in a typed metalanguage with a simple type system. All these
-   interpreters have no type tags and never get stuck;
-\item a Hindley-Milner--typed partial evaluator with no 
-  polymorphic lift operation and with
-  the delayed binding-time analysis of functions . The partial
-  evaluator needs no advanced type systems such as GADT or intentional
-  type analysis;
-\item to check statically for well-typed and closed object programs
-      using the type system \oleg{typechecker?} of the metalanguage;
-\item clean, comparable implementations in MetaOCaml and Haskell;
-\item a functor signature that encompasses all our interpreters, from
-    evaluation and compilation to partial evaluation and CPS;
-\item a clear way to extend the object language with more features
-  such as state;
-\item and an approach to self\hyp interpretation compatible with the
-  above.  Self\hyp interpretation turned out to be harder than expected.
+   in a typed metalanguage.
+\item All these interpreters use no type tags, patently never get stuck,
+    and need no advanced type-system features such as GADTs, dependent types,
+    or intentional type analysis.
+\item The partial evaluator avoids polymorphic lift and delays binding-time
+    analysis.
+\item We use the type system of the metalanguage
+    to check statically for well-typed and closed object programs.
+\item We show clean, comparable implementations in MetaOCaml and Haskell.
+\item We specify a functor signature that encompasses all our interpreters, from
+    evaluation and compilation to partial evaluation and CPS transformation.
+\item We point a clear way to extend the object language with more features
+  such as state.
+\item We describe an approach to self\hyp interpretation compatible with the
+  above.  Self\hyp interpretation turned out harder than expected.
 \end{enumerate}
@@
 
-In contrast to \S\ref{tagproblem}, we encode 
+In contrast to \S\ref{intro}, we henceforth encode 
 binding in object programs using higher-order
@@
 only one method of each group and elide the rest, to save space. The
-accompanying code has the complete code.
-
-Because the types in the |Symantics| class reflect the object language's type
-system, this representation admits only well-typed object terms for evaluation.
-\oleg{replace the above with the following?}
-The comparison of |Symantics| with Figure~\ref{fig:object} clearly
-shows that every typed, closed object term can be represented in the
-metalanguage. Moreover,
+accompanying code has the complete implementations.
+
+Comparing |Symantics| with Figure~\ref{fig:object}
+shows how to represent every typed, closed object term in the
+metalanguage. Moreover, the representation preserves types, in the following
+sense.
 \begin{prop}
-If an object term has the object type |t|, its
+If an object term has the object type |t|, then its
 representation in the metalanguage has the type 
-|forall repr. Symantics repr => repr t|
-\oleg{or, has the type repr t for some repr. Meta-language embeds all
-  object types.}
+|forall repr.| |Symantics repr => repr t|.
 \end{prop}
-If we make a mistake somewhere, say replace |int 7| with |bool True| in
+Conversely, the type system of the metalanguage checks statically that the
+represented object term is well-typed and closed.
+If we err, say replace |int 7| with |bool True| in
 |testpowfix7|, Haskell will complain there that the expected type |Int| does not
@@
 and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
-signature that suffices until \S\ref{PE}.
+signature that suffices until~\S\ref{PE}.
 Two differences between this signature and the |Symantics| class in
@@
                        (int 7))
-end;;
+end
 \end{code}
@@
 due to interpreting |fix|.
-That is, the proposition below follows immediately from the soundness of
-the metalanguage's type system.
 \begin{prop}
@@
 \end{prop}
-In general, 
+Generalizing from~|R| to all interpreters, we have the following
+broader and more useful, if also more obvious, proposition.
 \begin{prop}
-  If the definitional interpreter implementing |Symantics| 
-  in the soundly typed metalanguage never gets stuck,
+  If an implementation of |Symantics| never gets stuck, then
   the type system of the object
   language is sound with respect to the dynamic semantics defined by
-  the interpreter.
+  that implementation.
 \end{prop}
+These propositions follow immediately from the soundness of the
+metalanguage's type system.
 
@@
   let if_ eb et ee = eb + et () + ee () + 1
-end;;
+end
 \end{code}
@@
 constructors.  We present this partial evaluator in a sequence of four
-attempts, the last of which finally uses no universal type and no tags
+attempts. It uses no universal type and no tags
 for object types.  We then discuss residualization and binding-time
@@
 to a static (present-stage) term, as produced by the evaluator~|R|, or
-to a dynamic (future-stage) term, as produced by the compiler~|C|.  The
-interpreter computes as much as possible statically.  Thus one might try
+to a dynamic (future-stage) term, as produced by the compiler~|C|.  To
+distinguish between static and dynamic terms, one might try
 to define partial evaluation as the Haskell data type
@@
 instance Symantics P0 where
-  int  x = S0 (int x) (int x)
-  bool x = S0 (bool x) (bool x)
+  int  x = S0 (int x)
+  bool x = S0 (bool x)
   add (S0 e1) (S0 e2) = S0 (add e1 e2)
@@
 using the compiler~|C|.
+Thus we use the overloaded functions |int| and |bool| from the |Symantics|
+instances for |R| and~|C| to define those for~|P0|.
 
@@
 identity $\fun{x}x$ embedded in Haskell as |lam (\x -> x)|\texttt{ ::
-}|P1 (a -> a)|, consists of a dynamic part (type |C1 (a -> a)|) and
-maybe a static part (type |R1 (a -> a)|).  The dynamic part is useful
+}|P1 (a -> a)|, consists of a dynamic part (type |C (a -> a)|) and
+maybe a static part (type |R (a -> a)|).  The dynamic part is useful
 when this function is passed to another function that is only
@@
 interpreter contains many pattern\hyp matching forms that look partial
-but are total in reality. The goal is to make this totality
+but never fail in reality. The goal is to make this safety
 \emph{syntactically} apparent.
@@
 in~|P| is \emph{syntactically} exhaustive, so it is patent to the metalanguage
-implementation that |P| is total.  Further, all pattern\hyp matching occurs
+implementation that |P| never gets stuck.  Further, all pattern\hyp matching occurs
 during partial evaluation, only to check if a value is known statically,
@@
 a number of variants without any (new) difficulties.  In particular,
-we show how a a call-by-name CPS interpreter, a call-by-value CPS
+we show how a call-by-name CPS interpreter, a call-by-value CPS
 program transformation, as well as how to deal with imperative features.
@@
 call-by-value (CBV) metalanguage. The answer, given already by 
-\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is continuation-passing style (CPS), which they
+\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is yes thanks to CPS, which they
 specifically introduced 
@@
 The CBN CPS interpretation was introduced by \citet{PlotkinCBN}. As in
-the more familiar CBV CPS, the interpretation of a value is a function
-accepting a continuation argument |k| and yielding the answer
-returned by that continuation.
+CBV CPS, the interpretation of an object term is a function
+mapping a continuation~|k| to the answer
+returned by~|k|.
 \begin{code}
 let int (x:int) = fun k -> k x
-let add e1 e2 = 
-  fun k -> e1 (fun v1 -> e2 (fun v2 -> k (v1+v2)))
+let add e1 e2 = fun k ->
+  e1 (fun v1 -> e2 (fun v2 -> k (v1 + v2)))
 \end{code}
@@
 examples with it, and all works as expected.  More interesting
-is the following example.
+is the example $(\fun{x}1)\bigl((\fix{f}f)\mathinner2\bigr)$, which terminates
+under CBN but not CBV\@.
 \begin{code}
@@
    app (lam (fun x -> int 1)) 
-       (app (fix (fun self -> self)) (int 2))
+       (app (fix (fun f -> f)) (int 2))
 end
@@
 \begin{code}
-let module M = EXS(R) in M.diverg ();;
+let module M = EXS(R) in M.diverg ()
 \end{code}
@@
          (fun v -> f {ko = fun k -> k v}))}
-end;;
+end
 \end{code}
@@
     lam (fun x -> app (app (testpowfix ()) x) (int 7))
-end;;
+end
 \end{code}
@@
 \subsection{Translucent types}
-
-We must formulate some propositions: a typed term makes progress in
-any interpreter. The reason it is typed in any interpreter:
-|forall r. Symantics r => r tau| directly says that in every model the
-term is typed.
 
@@
 The attempt to encode self-interpreter brings the |*->*| polymorphism
-the the forefront, as we shall now see.
+the forefront, as we shall now see.
 

### 2007-04-05 tagless/tagless-final.tex
small remarks

@@
 The interpretation of |lam f| just wraps the
-HOAS function |f| into the result. We can always compile |f| to a code value,
+HOAS function |f|. We can always compile |f| to a code value,
 but we delay it to apply |f| to concrete arguments. The interpretation of
@@
 methods of |Symantics|.
-A simple and more portable solution thus emerges: we bake |repr_pe| 
+A simple and portable solution thus emerges: we bake |repr_pe| 
 into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
@@
 \oleg{He did not eliminate all the tags.}\jacques{can you show me?}
+\oleg{WalidICFP02, the very end of p1 and the beginning of p2, first
+  column}
 From reading \citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}, our 
@@
 
-\oleg{the only known to them, and \emph{partial} solution, was 
-  that by Zhe Yang;}\jacques{I cannot find that reference}
+From the above:
+For example, \citet{taha-tag}
+state that ``expressing such an interpreter in a statically typed
+programming language is a rather subtle matter. In fact, it is only
+recently that some work on programming type-indexed values in ML
+\citep{yang-encoding} has given a hint of how such a function can be
+expressed.''
 
@@
 uses dependent types
+\url{http://www.haskell.org/pipermail/haskell-cafe/2006-October/019160.html}
+\url{http://www.haskell.org/pipermail/haskell-cafe/2006-October/019161.html}
+\url{http://www.haskell.org/pipermail/haskell-cafe/2006-October/019168.html}
 
@@
 
-Our approach showed better the meaning of GADT and when they can be
-avoided.
-This shows also how a limited form of GADTs (the ones
-used for the sake of type-dependent representations) can be
-implemented with just functions.
-
 separate compilation: separate
 typechecking of terms without knowing which particular
-hole they will be plugged into. Formulate this as a prop?
+hole they will be plugged into.
 

### 2007-04-05 tagless/tagless-final.tex
Absorb "Lifting and functor" section into an earlier footnote

@@
 the static term to dynamic, but there is no polymorphic ``lift''
-function |a -> C a| to send a value to the future stage.
+function |a -> C a| to send a value to the future stage
+\citep{xi-guarded,WalidPOPL03}.\footnote{We note in passing that, if we
+were to add polymorphic \texttt{lift} to the type class
+\texttt{Symantics repr}, then \texttt{repr} would become an instance of
+\texttt{Applicative} and thus \texttt{Functor}:\quad\texttt{fmap
+f = app (lift f)}}
 
@@
 
-\subsection{Lifting and functor}
-
-It is perhaps worthwhile noting that 
-if we were to add a method 'lift' to |Symantics| with type
-|lift :: a -> repr a| (the polymorphic lift of \citet{xi-guarded}, 
-of MetaML and
-MetaOCaml: CSP)
-then |Symantics repr| implies |Functor repr|.  This can be seen from
-\begin{code}
-instance Semantics repr => Functor repr where
-  -- fmap :: (a->b) -> repr a -> repr b
-  fmap f ra = app (lift f) ra
-\end{code}
-Of course, as we have mentioned earlier in~\S\ref{S:PE-lift}, 
-addin such a polymorphic lift would appear to introduce more
-problems than it solves.
-
 \subsection{Translucent types}

### 2007-04-05 tagless/tagless-final.tex
more of the related work

@@
 Our initial motivation came from several papers 
-\citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} which used
+\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} which used
 embedded interpreters as a central example to justify looking 
-into advanced type systems.  While we admired all this technical work,
-there remained the issue that we were not convinced of the strength
-of their motivating example.
+into advanced type systems.  While we admire all this technical work,
+there remained the issue that we were not convinced that their 
+motivating example actually needed all this machinery.
 
@@
 \textsf{true}) both inhabit the same type $D$.  This is not very satisfying.
-
-\oleg{He did not eliminate all the tags.}\jacques{can you show me?}
-\oleg{WalidICFP02, the very end of p1 and the beginning of p2, first
-  column}
+Furthermore, as is pointed out in the introduction to \citet{WalidICFP02},
+\citet{taha-tag} does not \emph{statically} guarantee that all tags will
+be removed - tag elimination must be performed at runtime.
+
 From reading \citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}, our 
@@
 \item One needs to encode a typed language in a typed language based on
-a sum type,
-\item It is not possible to write a \emph{direct} interpreter for an typed
-language encoded this way, in a typed language, without either using a
+a sum type (at some level of the hierarchy),
+\item It is not possible to write a \emph{direct} interpreter 
+for such an encoding of a typed language
+in a typed language, without either using a
 very advanced type system or incurring tagging/untagging overhead.
-\item Thus we have to use an indirect interpreter, which needs a universal
+\item Thus an indirect interpreter is necessary, which needs a universal
   type (and hence tagging),
@@
 \end{enumerate}
-While the logic is sound, we showed that the premise in the first step
+While the logic is sound, we showed that the premise in the very first step
 was not valid.
 
-From the above:
-For example, \citet{taha-tag}
-state that ``expressing such an interpreter in a statically typed
+In particular, \citet[\S5]{taha-tag}
+states that ``expressing such an interpreter in a statically typed
 programming language is a rather subtle matter. In fact, it is only
@@
 
-Asai; 
-It [dynamic part when dealing with bool, int, lam in PE] is
-Asai's trick for reducing the amount of reify/reflect back-and-forth
-in a PE for an untyped language that allows you back-and-forth.  Or at
-least that's how I read those papers.
-Asai's approach is related to our PE (emulating typelevel functions):
-Ken: ``In some sense we're just observing that Asai's code type-checks
--- in Hindley-Milner as soon as we deforest the static code
-representation.''
-
-Sumii et al.
-
-"Boxes go Bananas" by Washburn and Weirich (intended or otherwise) Ken
-wrote about this: ``but a glance reveals no type classes.  Do you mean
-then that if we hide the definition |newtype R a = R a| from a client,
-and export only the interpreter |instance Symantics R R|, then the
-client would see R as an HOAS datatype a la Washburn and Weirich and
-only be able to create parametric ("non-exotic") terms?''  Jacques
-commented: ``I mean the HOAS parts, and how to deal with it (like in
-the instance for |Symantics R R|).  There are other nice tricks for
-dealing with HOAS in that paper worth a closer look.  It's all about
-plain polymorphism, not type classes, but there are still some clever
-ideas there, and they are about issues of tagging (an lack thereof).''
-
-Typing Dep Typing plus two messages on Haskell-Cafe: typechecking 
-from untyped to the typed form, ready to be interpreted. Walid's ICFP02
-uses dependent types
-\url{http://www.haskell.org/pipermail/haskell-cafe/2006-October/019160.html}
-\url{http://www.haskell.org/pipermail/haskell-cafe/2006-October/019161.html}
-\url{http://www.haskell.org/pipermail/haskell-cafe/2006-October/019168.html}
-
+We used a method from \citet{Asai-SAS99} to reduce the amount
+of back-and-forth between reify/reflect: in our PE we track both
+static and dynamic values in parallel (whenever possible).  In some sense,
+we are observing that this method type-checks in Hindley-Milner once we
+deforest the static code representation.  Put another way, we have
+manually applied type-level partial-evaluation to our typelevel 
+functions (see \S\ref{S:pe-solution}) to obtain simpler types 
+acceptable to MetaOCaml.  It is worthwhile to note that 
+\citet{Sumii-HOSC2001} also takes advanatage of this method
+\citep[of][]{Asai-SAS99} to get interesting results combining online
+and offline partial evaluation.  Like \citet{SperberThiemann:TwoForOne},
+we strive for modularity by reusing earlier stages.  It would be interesting
+to see if we could derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
+in the same manner.
+
+The approach to HOAS via catamorphism and anamorphism of
+\citet{Washburn-Weirich-boxes} seems related to what we are doing.
+We have not investigated the relationship to any serious depth, but we
+feel that one could fruitfully combine their ideas with ours.
+
+A lot of effort has gone into finding ways to give precise types to untyped
+terms (see for example \citet{baars-typing,Guillemette-Monier-PLPV}
+as well as the messages \citet{haskell-019160,haskell-019161}).  There 
+take advantage of the host language's types to varying extents.  All 
+these approaches are very successful for proper terms, but less so for
+ill-typed terms, where error messages can be quite obscure.  We use the
+host's typing facilities to do all the work, and thus inherit the strengths
+and weaknesses of the host language.  While that may not be optimal, it is
+certainly much more modular.
+
+\jacques{Got to here}.
 Regarding our implementation of the type-level function |'sv -> 'dv|.
@@
 
-Thiemann, cogen in six lines (for the interpreter/compiler work)?
+co-algebra references.  Many.
 

### 2007-04-05 tagless/tagless-final.tex
almost done the related work.

@@
 
-\jacques{Got to here}.
-Regarding our implementation of the type-level function |'sv -> 'dv|.
-Gibbons et al paper on
-the implementations of typecase may be relevant. I have not read it,
-but I have the feeling some of their typecase designs may be
-relevant (at least to cite). 
-However, the typecase paper relies on having a 
-bijection between types -- our |'sv| and |'dv| in this case, which is not 
-our situation.  We have a one-sided view of things, where |'sv -> 'dv| is 
-easy, and |'dv| "perfectly reflects" an |'sv|, but you still can't get at 
-that |'sv|.  \jacques{Could we say that we have a meta-level Galois
-injection?  We have one direction completely covered at the object
-level (and so easy to lift), but 'dv to 'sv can only be done by
-an outside mechanism}
-
-Self-Interpretation and Reflection in a Statically Typed Language 
-\url{http://webpages.cs.luc.edu/~laufer/papers/reflection93.pdf}
-He uses the SK language and some typeclasses. Their interpreter is a
-meta-circular interpreter -- rather than self-interpreter. We can
-claim to have implemented a far better interpreter for a far richer
-language.
-
-co-algebra references.  Many.
+When dealing with datatypes which are equivalent but have different
+representations, it is customary to use injection-projection pairs
+(used in the context of interpreters by \citet{Ramsey-ML-module-mania}
+and \citet{Benton-embedded-interpreters}, partial evaluation by
+\citet{Danvy-TDPE}, and type-level functions by
+\citet{oliveira-typecase}).  Na\"{\i}vely, this would appear to apply to
+our implementation of the type-level function |'sv -> 'dv|.  However, in most
+of these cases, it is important that this establishes a bijection (between
+|'sv| and |'dv| in our case).  Our approach is one-sided: we requite
+an operational definition of |'sv -> 'dv| (which is easy), and an
+implementation of |'dv| which ``perfectly reflects'' |'sv| without 
+however making this uniformly available.
+
+\citet{fiore:nbe-ppdp2002,balat:tdpe-popl2004} also deal with a 
+tagfree partial-evaluator using quite a lot of interesting 
+technology (amongst them powerful continuation-based control operators).
+While they use a reify-reflect injection-projection pair, this appears
+to be done ``lightly'', in the sense that the structure of compiled code
+is never examined.  The exact relation between their work and ours is 
+still a little mysterious to us, but we will continue to investigate it.
+
+We are not the first to implement a typed interpreter for a typed
+language -- \citet{laod93} use a typed version of the SK language
+(and typeclasses) to implement a meta-circular interpreter (rather than
+a self-interpreter).  Interestingly, their meta-circular interpreter
+appears to be tagfree!  However, they could not have implemented a
+compiler or a partial-evaluator in the same way since they rely quite
+heavily on having injection-projection pairs (through reflection and
+reification).
+
+We have been unable to find a direct reference that establishes that
+the typed $\lambda$-calculus possesses a final co-algebra structure.
+For the untyped $\lambda$-calculus, 
+\citet{HonsellLenisa,honsell99coinductive} establish several interesting
+results along this line.  Of particular interest is their use of
+contexts with a hole \citep[p.13]{honsell99coinductive} for the definition
+of \emph{observational equivalence}, which directly relates to our
+use of terms with a hole as a context for our self-interpreter 
+(\S\ref{selfinterp}).  The reader is directed to bibliography of
+\citet{honsell99coinductive} for references to the foundational work
+in this important area.
+
+\jacques{Dig through the last couple of emails I just got from the
+category theory list with references, and this section will be done.}
 

### 2007-04-06 tagless/tagless-final.tex
Self-interpretation

@@
 \usepackage{mdwlist}
+\usepackage{mdwtab}
 \usepackage{dblfloatfix}
@@
 \newcommand{\cond}[3]{\mathrm{if\ }#1\mathrm{\ then\ }#2\mathrm{\ else\ }#3}
+\newcommand{\be}[1]{\mathrm{let\ }#1\mathrm{\ in\ }}
 \newcommand{\True}{\mathinner{\mathrm{true}}}
 \newcommand{\False}{\mathinner{\mathrm{false}}}
+\newcommand{\Encode}[1]{\mathinner{\mathopen{\text{``}}#1\mathclose{\text{''}}}}
+\newcommand{\encode}[1]{\mathinner{\mathopen{\text{`}}#1\mathclose{\text{'}}}}
+\newcommand{\ident}[1]{\mathinner{\text{\textit{#1\/}}}}
+\newcommand{\subst}[3]{#1\left\{#2\mapsto#3\right\}}
+\DeclareMathOperator{\pe}{PE}
+\DeclareMathOperator{\si}{SI}
 
@@
 \begin{prop}
-If an object term has the object type |t|, then its
+If an object term has the object type~$t$, then its
 representation in the metalanguage has the type 
-|forall repr.| |Symantics repr => repr t|.
+|forall repr.| |Symantics repr => repr |$t$.
 \end{prop}
@@
 \begin{prop}
-If the term |e| encoded in the metalanguage has type~|t|,
-then evaluating~|e| in the interpreter~|R| either continues
-indefinitely or terminates with a value of the same type~|t|.
+If an object term~$e$ encoded in the metalanguage has type~$t$,
+then evaluating~$e$ in the interpreter~|R| either continues
+indefinitely or terminates with a value of the same type~$t$.
 \end{prop}
@@
 into OCaml code using MetaOCaml's staging facilities. MetaOCaml
-represents future-stage expressions of type |t| at the
-present stage as values of type |('c,t) code| where |'c| is the
+represents future-stage expressions of type~$t$ at the
+present stage as values of type |('c,|$t$|) code| where |'c| is the
 environment classifier \citep{WalidPOPL03,calcagno-ml-like}. Code values are created
-by a \emph{bracket} form |.<e>.|, which specifies that the expression |e| is to be
-evaluated at a future stage. The \emph{escape} |.~e| must occur
-within a bracket and specifies that the expression |e| must be evaluated
+by a \emph{bracket} form |.<|$e$|>.|, which specifies that the expression~$e$ is to be
+evaluated at a future stage. The \emph{escape} |.~|$e$ must occur
+within a bracket and specifies that the expression~$e$ must be evaluated
 at the current stage; its result, which must be a code value, is
-spliced into the code being built by the enclosing bracket. The \emph{run} form |.!e| evaluates
-the future-stage code value~|e| by compiling and linking it at run-time.
+spliced into the code being built by the enclosing bracket. The \emph{run} form |.!|$e$ evaluates
+the future-stage code value~$e$ by compiling and linking it at run-time.
 The bracket, escape, and run are akin to
@@
 \end{code}
-This is a completely straightforward staging of
+This is a straightforward staging of
 |module R|.
-This compiler simply produces
-residual code with no optimization. For example, interpreting our |test1|
+This compiler produces
+unoptimized code. For example, interpreting our |test1|
 \begin{code}
@@
 \end{code}
-gives the code value |.<(fun x_6 -> x_6) (true)>.|
+gives the code value |.<(fun x_6 -> x_6) true>.|
 of inferred type |('c, bool) C.repr|.  Similarly, interpreting |testpowfix7|
@@
 We have also implemented this compiler in Haskell. Since Haskell
-has no (convenient, typed) staging facilities, we had to emulate
-them. To be precise, we defined a data type |ByteCode| with
+has no (convenient, typed) staging facility, we had to emulate
+it. To be precise, we defined a data type |ByteCode| with
 constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
@@
 GADTs because we also wanted to write a typed interpreter for 
-the \textsf{ByteCode} \emph{data type}.} is quite similar (but slightly more
+the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
 verbose) than the corresponding MetaOCaml code given above. The
@@
 the static term to dynamic, but there is no polymorphic ``lift''
-function |a -> C a| to send a value to the future stage
+function, of type |a -> C a|, to send a value to the future stage
 \citep{xi-guarded,WalidPOPL03}.\footnote{We note in passing that, if we
@@
 
-\begin{figure*}
-\begin{floatrule}
-\begin{code}
-module type Symantics = sig
-
-  type ('c,'sv,'dv) repr
-
-  val int : int  -> ('c,int,int) repr
-  val bool: bool -> ('c,bool,bool) repr
-
-  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
-  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
-
-  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
-  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
-
-end
-\end{code}
-\end{floatrule}
-\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
-\label{fig:ocaml}
-\end{figure*}
-
-\begin{figure}
-\begin{floatrule}
-\begin{code}
-module P = struct
-
-  type ('c,'sv,'dv) repr = {st: 'sv option;
-                            dy: ('c,'dv) code}
-
-  let abstr {dy = x} = x
-  let pdyn x = {st = None; dy = x}
-
-  let int  (x:int)  = {st = Some (R.int x);
-                       dy = C.int x}
-  let bool (x:bool) = {st = Some (R.bool x);
-                       dy = C.bool x}
-
-  let add e1 e2 = match e1, e2 with
-  | {st = Some 0}, e | e, {st = Some 0} -> e
-  | {st = Some m}, {st = Some n} -> int (R.add m n)
-  | _ -> pdyn (C.add (abstr e1) (abstr e2))
-
-  let if_ eb et ee = match eb with
-  | {st = Some b} -> if b then et () else ee ()
-  | _ -> pdyn (C.if_ (abstr eb) 
-                     (fun () -> abstr (et ()))
-                     (fun () -> abstr (ee ())))
-
-  let lam f =
-  {st = Some f; 
-   dy = C.lam (fun x -> abstr (f (pdyn x)))}
-
-  let app ef ea = match ef with
-  | {st = Some f} -> f ea
-  | _ -> pdyn (C.app (abstr ef) (abstr ea))
-
-  let fix f = 
-    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
-    in let rec self = function
-       | {st = Some _} as e -> app (f (lam self)) e
-       | e -> pdyn (C.app fdyn (abstr e))
-       in {st = Some self; dy = fdyn}
-
-end
-\end{code}
-\end{floatrule}
-\caption{Our partial evaluator (\texttt{mul} and \texttt{leq} are elided)}
-\label{fig:pe}
-\end{figure}
-
 Two common ways to provide typecase in Haskell are
@@
 in the accompanying source code shows the latter.
+
+\begin{figure*}
+\begin{floatrule}
+\begin{code}
+module type Symantics = sig
+
+  type ('c,'sv,'dv) repr
+
+  val int : int  -> ('c,int,int) repr
+  val bool: bool -> ('c,bool,bool) repr
+
+  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
+  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
+
+  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
+  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
+
+end
+\end{code}
+\end{floatrule}
+\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
+\label{fig:ocaml}
+\end{figure*}
 
@@
 Figure~\ref{fig:ocaml} shows the new |Symantics| signature.
+
+\begin{figure}
+\begin{floatrule}
+\begin{code}
+module P = struct
+
+  type ('c,'sv,'dv) repr = {st: 'sv option;
+                            dy: ('c,'dv) code}
+
+  let abstr {dy = x} = x
+  let pdyn x = {st = None; dy = x}
+
+  let int  (x:int)  = {st = Some (R.int x);
+                       dy = C.int x}
+  let bool (x:bool) = {st = Some (R.bool x);
+                       dy = C.bool x}
+
+  let add e1 e2 = match e1, e2 with
+  | {st = Some 0}, e | e, {st = Some 0} -> e
+  | {st = Some m}, {st = Some n} -> int (R.add m n)
+  | _ -> pdyn (C.add (abstr e1) (abstr e2))
+
+  let if_ eb et ee = match eb with
+  | {st = Some b} -> if b then et () else ee ()
+  | _ -> pdyn (C.if_ (abstr eb) 
+                     (fun () -> abstr (et ()))
+                     (fun () -> abstr (ee ())))
+
+  let lam f =
+  {st = Some f; 
+   dy = C.lam (fun x -> abstr (f (pdyn x)))}
+
+  let app ef ea = match ef with
+  | {st = Some f} -> f ea
+  | _ -> pdyn (C.app (abstr ef) (abstr ea))
+
+  let fix f = 
+    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
+    in let rec self = function
+       | {st = Some _} as e -> app (f (lam self)) e
+       | e -> pdyn (C.app fdyn (abstr e))
+       in {st = Some self; dy = fdyn}
+
+end
+\end{code}
+\end{floatrule}
+\caption{Our partial evaluator (\texttt{mul} and \texttt{leq} are elided)}
+\label{fig:pe}
+\end{figure}
 
@@
 \end{code}
-Now when we apply an abstraction to an argument |e|, we always force
+Now when we apply an abstraction to an argument, we always force
 the evaluation of that argument before proceeding. This approach is in
@@
 \begin{code}
-  P.repr = {P.st = Some <fun>; P.dy = .<fun x_5 -> (x_5 (true))>.}
+  P.repr = {P.st = Some <fun>;
+            P.dy = .<fun x_5 -> (x_5 (true))>.}
 \end{code}
@@
 We encode a term like |add 1 2| as
-|app _add (app _int 1) (app _int 2)| where |_add| and |_int| are just
+\texttt{app \_add (app \_int 1) (app \_int 2)} where |_add| and |_int| are just
 `free variables'. Now, how to typecheck such a term? Some type should
@@
 type-classes, the type of a term L of the inferred type tau is
-
 $$ 
@@
 
-A self-interpreter |si| is an
-object-language term such that, for any object-language term e,
-
-\begin{code}
-  si "e"    is observationally equivalent to    e
-\end{code}
-
-where |"e"| is an encoding of |e|.  In our approach, we want to encode
-each object-language term as a function from a record%
-\footnote{Thanks to currying, the object language need not actually support
-records.}  of interpreter methods to a generated representation.  
-In the case of a self-interpreter, that
-record is simply |si|!  So we should define a self-interpreter |si| to be an
-object-language term such that, for any object-language term |e|,
-
-\begin{code}
-  "e" si    is observationally equivalent to    e
-\end{code}
-
-A self-interpreter is then just like the interpreter
-|R| in Section 2, except written at the object rather than the meta level.
-
-It is wortwhile noting that, in our class |Symantics|, |add|
-is not a term since it does not have type |repr t| for any |t|. We could
-have introduced |add| to be of type |repr (Int->Int->Int)| -- but we
-didn't. Thus |add| is a special form, which is to be applied in the
-meta-language using the meta-language application syntax (i.e. white
-space) rather than in the object language (i.e. via |app|).
-
-\oleg{Mention the RR interpreter: for any term E, (RR E) is equivalent to E.
-RR is not an identity: it is an interpreter that encapsulates another
-interpreter. The RR interpreter can be made self if we treat RR as a
-special form and interpret it always with itself (similar to let and
-hole below).}
-
-Self-interpretation in our framework is somewhat
-tricky due to the need for more polymorphism than is easily available
-in the meta-language.  An encoded term should
-be polymorphic in the interpreter type-constructor "r".  Furthermore, the
-functions "lam", "app", etc. should be polymorphic in the object
-types "a" and "b".  Thus the object language seems to require rank-2
-polymorphism; and so type annotations will be inevitable.
-
-To neatly bypass the need for rank-2 polymorphism, we use the notion
-of let-polymorphism and a `hole', as follows.  A partial evaluator (or
-a compiler or interpreter) is an evaluation context that (let-)binds
-variables named "app", "lam", etc.  For the object language to
-take advantage of let-bound polymorphism in the metalanguage, it
-is crucial that we encode the object-level "let" using the meta "let".
-
-We also (unfortunately need to) define a self-interpreter slightly
-differently.  A self-interpreter |SI[]| is an object expression with a
-hole |[]| waiting to be plugged with the encoding of an object expression.
-The self-interpreter binds the free variables |_lam|, |_app|, |_int|, |_bool|,
-etc. in the plugged encoding.  We require |SI["E"]| to be equivalent to |E|,
-where |"E"| is the object encoding of the object expression |E|.
-
-This strategy of course limits what kinds of processing on encoded
-object expressions can be expressed in the object language.  For
-example, unlike if we had simply defined |"E"| to be the \emph{identity
-encoding} E, we \emph{can} write a size-measure interpreter |SZ[]|, 
-which like |SI[]| is
-an object expression with a hole |[]| waiting to be plugged with 
-an |"E"|.
-However, this notion of size cannot distinguish |let x = E1 in E2[x]|
-from |"E2[E1]"|.  Recall that one way to type-check polymorphic let is:
-\begin{code}
-        E1:T    E2[E1]:T'
-    ------------------------
-    let x = E1 in E2[x] : T'
-\end{code}
-That analogy shows that the self-interpreter with a hole is a
-euphemism for a higher-rank abstraction. That makes it clear where
-the difficulty lies, and how to simulate the higher-rank application
-(that is, higher-rank arrow elimination) `by hand'.
-
-We also need to duplicate the encoding of an object expression if we
-want to process the same object expression in multiple ways (e.g., both
-measure its size and interpret it).  But all this is consistent with (in
-fact, amplifies) our `coalgebraic/final/cogen' spirit, so our revised
-notion of self-interpretation (and, some might say, of size measurement)
-may just squeak by the reader's skepticism as we explain our desire for
-kind polymorphism.
-
-\oleg{I guess the let-and-hole argument works out if we don't typecheck the
-encoded terms. We can't typecheck the interpreter with the hole
-either. We insert the encoded term into the interpreter -- and then
-typecheck the whole thing. That'll work. Yet a bit unsatisfactory...}
-
-We achieve
-self-interpretation in that the following evaluation context in the
-object language defines an interpreter for encoded object terms.
-\begin{code}
-  let lam = \f -> f in
-  let app = \f x -> f x in
-  let add = \m n -> m + n in
-  let int = \i -> i in
-  [hole]
-\end{code}
-The obvious notion of optimality in this setup is easy to achieve.  For
-example, given the context above, our partial evaluator turns
-\begin{code}
-  app (lam (\x -> add x (int 1))) (int 3)
-\end{code}
-into the number 4, as desired.  We can also encode the self-interpreter
-as the following evaluation context in the object language.
-\begin{code}
-  let lam = lam (\f -> f) in
-  let app = lam (\f -> lam (\x -> app f x)) in
-  let add = lam (\m -> lam (\n -> add m n)) in
-  let int = lam (\i -> i) in
-  [hole]
-\end{code}
-The "let" and the hole "[hole]" are meta-constructions, which must map to
-themselves in any interpretation.
-
-The 'result' of applying an interpreter to itself needs to be an 
-interpreter too!  In fact, if it is Jones-optimal, then you should be 
-able to apply it to itself again and get the same answer.  
-Our code does that. First, let |twice_inc_3_e| be the
-metalanguage encoding of the object term |si["e"]|,
-where si is the self-interpreter for the needed components of the object
-language (only |lam|, |app|, |add|, and |int| in this case), |""| means the
-object-to-object encoding, and |e| is the object term
-\begin{code}
-  let twice_ f x = f (f x) in let inc_ n = n + 1 in twice_ inc_ 3
-\end{code}
-Second, |twice_inc_3_ee| is the metalanguage encoding of the object term
-|si["si[e]"]|
-which is the same as
-|si["si"["e"]]|.
-The partial evaluator yields 5 on both |twice_inc_3_e| and 
-|twice_inc_3_ee|,
-as desired.
-
+We turn to interpreting the object language in the object language, to
+clarify how expressive our typed object language can be and to argue that our
+partial evaluator is Jones\hyp optimal.
+
+Given an \emph{encoding} of each object term~$e$ as an \emph{object} term~$\Encode{e}$,
+a \emph{self\hyp interpreter} is usually defined as an object
+function~$\si$ such that any object term~$e$ is observationally
+equivalent to the object application $\si\Encode{e}$
+\citep{jones-partial,taha-tag,Danvy-tagging-encoding}.
+A partial evaluator~$\pe$ maps object terms~$e$ to observationally
+equivalent object terms~$\pe(e)$.  It is said to be
+\emph{optimal} with respect to~|si| if $\pe(\si\Encode{e})$
+is equal to~$e$ (up to $\alpha$\hyp conversion, or in some accounts, no
+less efficient than~$e$).
+
+Intuitively, self\hyp interpretation is straightforward in our
+framework: the functions comprising the interpreters
+in~\S\ref{S:interpreter-RL} may just as well be written in our object
+language.  In particular, the following \emph{object} functions implement an
+evaluator.  (We use the number~$0$ in lieu of a unit value.)
+\begin{align*}
+    \ident{int} &= \fun{x} x &
+    \ident{lam} &= \fun{f} f \\
+    \ident{add} &= \fun{x} \fun{y} x+y &
+    \ident{app} &= \fun{f} \fun{x} fx \\
+    \ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0} &
+    \ident{fix} &= \fun{g} \fix{f} \fun{x} gfx
+\end{align*}
+We thus map each object terms~$e$ to an object term~$\encode{e}$ as follows.
+We call this mapping \emph{pre-encoding}.
+\begin{align*}
+    \encode{x} &= x &
+    \encode{\fun{x}e} &= \ident{lam} (\fun{x} \encode{e}) \\
+    \encode{n} &= \ident{int}\, n &
+    \encode{fx} &= \ident{app} \encode{f} \encode{x} \\
+    \encode{e_1 + e_2} &= \ident{add} \encode{e_1} \encode{e_2} &
+    \encode{\fix{f}e} &= \ident{fix} (\fun{f} \encode{e}) \\
+    \encode{\cond{b}{t}{e}} &= \rlap{$\ident{if\_} \encode{b}
+        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$}
+\end{align*}
+The metavariables $x$ and~$n$ stand for a variable and an integer,
+respectively.
+This pre-encoding is just like how we represent object terms in the
+metalanguage in the preceding sections, but it produces
+terms in the object language rather than metalanguage.
+
+To evaluate~$\encode{e}$, then, we
+instantiate the free variables in~$\encode{e}$ such as $\ident{int}$,
+$\ident{lam}$, and $\ident{add}$ by their definitions above.  For
+example, the familiar object term $(\fun{x}x)\True$ pre-encodes to
+\begin{equation*}
+    \encode{(\fun{x}x)\True} = \ident{app}
+    (\ident{lam} (\fun{x} x))\, (\ident{bool} \True),
+\end{equation*}
+and to evaluate this pre-encoded term is to evaluate the object term
+\begin{equation*}
+    (\fun{f} \fun{x} fx) \,
+    ((\fun{f} f) (\fun{x} x))\, ((\fun{b} b) \True).
+\end{equation*}
+Because the evaluator above mostly consists of glorified identity
+functions, we expect even our simple partial evaluator to reduce the
+result of this instantiation to~$e$.  In general, to
+interpret~$\encode{e}$ using an interpreter is to instantiate these free
+variables by that interpreter's definitions.
+
+\subsection{Avoiding higher-rank polymorphism}
+
+Any approach to self\hyp interpretation needs to spell out first how to
+encode object terms~$e$ to object terms~$\Encode{e}$, and then how to
+interpret~$\Encode{e}$ in the object language.  For our approach, we
+want to define encoding in terms of pre-encoding, and interpretation
+using some notion of instantiation.  Unfortunately, the simple type
+structure of our object language hinders both tasks.  To continue with
+the example term above, we could try to define
+\begin{equation*}
+    \Encode{e} =
+    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}} \encode{e},
+\end{equation*}
+in particular
+\begin{equation*}
+    \Encode{(\fun{x}x)\True} =
+    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
+    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool} \True).
+\end{equation*}
+To type-check this encoded term, we give the variable $\ident{lam}$ the
+simple type $(\BB\to\BB)\to\BB\to\BB$.
+We then define the self\hyp interpreter
+\begin{equation*}
+    \si = \fun{e} e
+    (\fun{f} \fun{x} fx)
+    (\fun{f} f)
+    (\fun{b} b)
+\end{equation*}
+and apply it to the encoded term.  The result is the object term
+\begin{multline*}
+    \bigl(\fun{e} e (\fun{f} \fun{x} fx) (\fun{f} f) (\fun{b} b)\bigr)
+\\
+    \bigl(
+    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
+    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool} \True)
+    \bigr),
+\end{multline*}
+which partially evaluates to~$\True$ easily.
+However, encoding fails on an term with multiple $\lambda$\hyp
+abstractions at different types.  For example, the pre-encoding
+\begin{equation*}
+    \encode{\fun{f}\fun{x}fx}
+    = \ident{lam} (\fun{f} \ident{lam} (\fun{x} \ident{app} f x))
+\end{equation*}
+does not type-check in any typing environment, because $\ident{lam}$ needs
+to take two incompatible types.  In sum, we need polymorphism in the
+object type system to type $\ident{lam}$, $\ident{app}$, $\ident{fix}$,
+and~$\ident{if\_}$ (and $\Encode{e}$ and~$\si$).
+(The polytypes in |Symantics| are where Haskell type classes and OCaml
+modules supply this polymorphism.)  Moreover, we need to encode any
+polymorphism in the object language into the object language to achieve
+self\hyp interpretation.
+
+\subsection{Introducing let-bound polymorphism}
+
+Instead of adding higher-rank polymorphism to our object language (along
+with polymorphism over kinds), we add let-bound polymorphism.  As usual,
+we can add a new typing rule
+\begin{equation*}
+    \begin{prooftree}
+        e_1:t_1 \quad \subst{e_2}{x}{e_1}:t_2
+        \justifies \be{x=e_1} e_2 : t_2
+    \end{prooftree}
+    .
+\end{equation*}
+The pre-encoding of a let\hyp expression is trivial.
+\begin{align*}
+    \encode{\be{x=e_1}e_2} \quad &= \quad \be{x=\encode{e_1}} \encode{e_2}
+\intertext{A \emph{context} is an object term with a hole~$[~]$.  The
+hole may occur under a binder, so plugging a term into the context may
+capture free variables of the term.  By pre-encoding a hole to a hole,
+we extend pre-encoding from a translation on terms to a translation on
+contexts.}
+    \encode{[~]} \quad &= \quad [~]
+\end{align*}
+We define an interpreter in the object language to be not a term but
+a context.  For example, the evaluator is the context
+\begin{align*}
+    \si[~] &=
+    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
+        \be{\ident{int} &= \fun{x} x&} \\
+        \be{\ident{add} &= \fun{x} \fun{y} x+y&} \\
+        \be{\ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}&} \\
+        \be{\ident{lam} &= \fun{f} f&} \\
+        \be{\ident{app} &= \fun{f} \fun{x} fx&} \\
+        \be{\ident{fix} &= \fun{g} \fix{f} \fun{x} gfx&} [~],
+    \end{tabular}
+\intertext{and the size\hyp measurer is the context}
+    \mathrm{SZ}[~] &=
+    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
+        \be{\ident{int} &= \fun{x} 1&} \\
+        \be{\ident{add} &= \fun{x} \fun{y} x+y+1&} \\
+        \be{\ident{if\_}&= b + t\,0 + e\,0&} \\
+        \be{\ident{lam} &= \fun{f} f\,0 + 1&} \\
+        \be{\ident{app} &= \fun{f} \fun{x} f + x + 1&} \\
+        \be{\ident{fix} &= \fun{g} g\,0 + 1&} [~].
+    \end{tabular}
+\end{align*}
+To interpret an object term~$e$ using an interpreter $I[~]$ is to
+evaluate the object term~$I[\encode{e}]$.  $\si$ is
+a self\hyp interpreter in the following sense.
+\begin{prop}
+    $\si[\encode{e}]$ is observationally equivalent to~$e$.
+\end{prop}
+As a corollary, we can pre-encode the self\hyp interpreter itself as
+a context: the term $\si[\encode{\si[\encode{e}]}]$ is observationally
+equivalent to $\si[\encode{e}]$, and in turn to~$e$.  In other words,
+$\si$ can interpret itself.
+
+Our partial evaluator is optimal with respect to the self\hyp
+interpreter~$\si$.
+\begin{prop}
+    Let $\pe$ be the partial evaluator~|P| in~\S\ref{S:PE-solution}.
+    Then the object terms $\pe(\si[\encode{e}])$ and~$\pe(e)$ are
+    either both undefined or both defined and
+    equal up to $\alpha$\hyp conversion.
+    \ccshan{Check this}
+\end{prop}
+
+\subsection{Contexts clarify polymorphism}
+
+We type-check a pre-encoded term~$\encode{e}$ and its interpreter~$I[~]$
+only together, never separately.  This treatment has the drawback that
+we must duplicate a pre-encoded term in order to interpret it in
+multiple ways.  The meta-notions of contexts and plugging may seem ad
+hoc, but in fact they are explications of the type-class and module
+machinery that we have been using in this paper all along.
+
+In the presence of let-bound polymorphism, we can understand a term
+waiting to be plugged into a context as a higher-rank abstraction over
+the context.  Whereas our object language does not support higher-rank
+abstraction, our metalanguages do, so they can type-check an object term
+separately from its interpreter---as a functor from a |Symantics| module
+(in OCaml) or a value with a |Symantics| constraint (in Haskell).  Thus
+``context'' is a euphemism for a polymorphic argument, and ``plugging''
+is a euphemism for application.
+
+\begin{comment}
 \jacques{But how 
@@
 
+\oleg{Mention the RR interpreter: for any term E, (RR E) is equivalent to E.
+RR is not an identity: it is an interpreter that encapsulates another
+interpreter. The RR interpreter can be made self if we treat RR as a
+special form and interpret it always with itself (similar to let and
+hole below).}
+\end{comment}
+
 \section{Related work}\label{related}

### 2007-04-06 tagless/tagless-final.tex
Translucent types

@@
 long as it is}
-
-\subsection{Translucent types}
-
-The crucial role of the higher-order type parameter r
-
-The type constructor "r" above represents a particular interpreter.  The
-meta-type "r tau" hides how the interpreter represents the object type
-"tau" yet exposes enough of the type information so we can type-check
-the encoding of an object term without knowing what "r" is.  The checked
-term is then well-typed in any interpreter.  Each instance of Symantics
-instantiates "r" to interpret terms in a particular way. The L
-interpreter is quite illustrative. We need the above semantics to be
-able to represent both R and L (the latter returns only Int as the
-values) in the same framework.
-
-We encode a term like |add 1 2| as
-\texttt{app \_add (app \_int 1) (app \_int 2)} where |_add| and |_int| are just
-`free variables'. Now, how to typecheck such a term? Some type should
-be assigned to these free variables. The goal is to complete the work
-without needing any type annotations (so we don't have to introduce any
-type language), with all types inferred and all terms typed. It seems
-the second-order type R neatly separates the typechecking part from
-the representation of R: it hides aspects that depend on the
-particular interpreter, and yet lets enough type information through
-(via its type argument) to permit the typechecking of terms, and infer
-all the types.
-
-The only approach that does seem to work
-is the one in incope.hs or incope.ml. If we de-sugar away records and
-type-classes, the type of a term L of the inferred type tau is
-$$ 
-  (\ZZ \rightarrow r \ZZ) \rightarrow
-  (\BB \rightarrow r \BB) \rightarrow
-  \forall \alpha \beta. (r \alpha \rightarrow r \beta)
-      \rightarrow r (\alpha\rightarrow\beta) \rightarrow ... r \tau
-$$
-% (Int -> r Int) ->
-% (Bool -> r Bool) ->
-% (forall alpha beta. (r alpha -> r beta) -> r (alpha->beta)) -> ... r tau
-
-or, if we denote the sequence of initial arguments as S r, terms have
-the type |S r -> r tau|
-The interpreter has the type
-|(forall r. S r -> r tau) -> r' tau|
-
-The higher-order type (variable) of kind |*->*| seems essential. So, at
-least we need some fragment of Fw (somehow our OCaml code manages to
-avoid the full Fw; probably because the module language is separated
-from the term language). Thus, we seem to need a fragment of Fw. It
-seems the inference is possible, as our Haskell and OCaml code
-constructively illustrates. Perhaps we need to characterize our
-fragment.
-
-The attempt to encode self-interpreter brings the |*->*| polymorphism
-the forefront, as we shall now see.
 
@@
 
-\subsection{Avoiding higher-rank polymorphism}
+\subsection{Avoiding higher polymorphism}
 
@@
 
-Instead of adding higher-rank polymorphism to our object language (along
+Instead of adding higher-rank and higher-kind polymorphism to our object language (along
 with polymorphism over kinds), we add let-bound polymorphism.  As usual,
@@
 In the presence of let-bound polymorphism, we can understand a term
-waiting to be plugged into a context as a higher-rank abstraction over
-the context.  Whereas our object language does not support higher-rank
+waiting to be plugged into a context as a higher-rank and higher-kind abstraction over
+the context.  Whereas our object language does not support higher
 abstraction, our metalanguages do, so they can type-check an object term
-separately from its interpreter---as a functor from a |Symantics| module
-(in OCaml) or a value with a |Symantics| constraint (in Haskell).  Thus
+separately from its interpreter---either as a functor from a |Symantics| module
+containing a type constructor
+(in OCaml), or a value with a |Symantics| constraint over a type
+constructor (in Haskell).  Thus
 ``context'' is a euphemism for a polymorphic argument, and ``plugging''
@@
 
+\begin{comment}
+
+The crucial role of the higher-order type parameter r
+
+The type constructor "r" above represents a particular interpreter.  The
+meta-type "r tau" hides how the interpreter represents the object type
+"tau" yet exposes enough of the type information so we can type-check
+the encoding of an object term without knowing what "r" is.  The checked
+term is then well-typed in any interpreter.  Each instance of Symantics
+instantiates "r" to interpret terms in a particular way. The L
+interpreter is quite illustrative. We need the above semantics to be
+able to represent both R and L (the latter returns only Int as the
+values) in the same framework.
+
+We encode a term like |add 1 2| as
+\texttt{app \_add (app \_int 1) (app \_int 2)} where |_add| and |_int| are just
+`free variables'. Now, how to typecheck such a term? Some type should
+be assigned to these free variables. The goal is to complete the work
+without needing any type annotations (so we don't have to introduce any
+type language), with all types inferred and all terms typed. It seems
+the second-order type R neatly separates the typechecking part from
+the representation of R: it hides aspects that depend on the
+particular interpreter, and yet lets enough type information through
+(via its type argument) to permit the typechecking of terms, and infer
+all the types.
+
+The only approach that does seem to work
+is the one in incope.hs or incope.ml. If we de-sugar away records and
+type-classes, the type of a term L of the inferred type tau is
+$$ 
+  (\ZZ \rightarrow r \ZZ) \rightarrow
+  (\BB \rightarrow r \BB) \rightarrow
+  \forall \alpha \beta. (r \alpha \rightarrow r \beta)
+      \rightarrow r (\alpha\rightarrow\beta) \rightarrow ... r \tau
+$$
+% (Int -> r Int) ->
+% (Bool -> r Bool) ->
+% (forall alpha beta. (r alpha -> r beta) -> r (alpha->beta)) -> ... r tau
+
+or, if we denote the sequence of initial arguments as S r, terms have
+the type |S r -> r tau|
+The interpreter has the type
+|(forall r. S r -> r tau) -> r' tau|
+
+The higher-order type (variable) of kind |*->*| seems essential. So, at
+least we need some fragment of Fw (somehow our OCaml code manages to
+avoid the full Fw; probably because the module language is separated
+from the term language). Thus, we seem to need a fragment of Fw. It
+seems the inference is possible, as our Haskell and OCaml code
+constructively illustrates. Perhaps we need to characterize our
+fragment.
+
+\end{comment}
+
 \section{Related work}\label{related}

### 2007-04-06 tagless/tagless-final.tex
Conclusion

@@
 
-which shows the naive CPS transform of the original code. The result is
+which shows the naive CPS transformation of the original code. The result is
 quite unoptimal with several apparent $\beta$-redexes. 
@@
 
-We can easily modify the CBV CPS transform to pass a piece of state
-along with the continuation. This lets us support mutable state. We
-can also use reference variables of the metalanguage to emulate
-reference variables in the object language.
-\oleg{Should we have an example in the code?}
-\jacques{In the code, yes.  Probably not in the paper, it is getting 
-long as it is}
+We can modify the CBV CPS transformation to pass a piece of state along
+with the continuation. This technique lets us support mutable state. We
+can also add mutable references to the object language using mutable
+references of the metalanguage.  The accompanying code illustrates these
+extensions.
 
@@
 
-We were dissatisfied with current approaches to DSL embedding: either
-the performance was hurt by tagging/untagging and other interpretive
-overhead, or the offered solutions required too much expressive power
-from the host language's type system.  By shifting from an initial
-algebra approach to one which leverages the natural co-algebraic structure
-of the $\lambda$-calculus, we show that both of these issues can be 
-simultaneously dealt with.
-
-% We show how to do many things without tags, in a simple type system, 
-% essentially by avoiding any (inductive) types.
-
-\jacques{Should we mention that algebraic type tags give us a partly
-intensional representation of terms.  This nicely allows pattern matching,
-but as sits inside an extensional theory, ensuring that all traces 
-of these tags disappear seems quite difficult without a very powerful
-type system (enter GADTs).}
-
-The overarching theme of our approach -- representation of object terms in 
-\S\ref{ourapproach}, future-stage code in \S\ref{S:PE-lift} and static
-values in \S\ref{S:PE-solution} and CPS values in \S\ref{S:CPS} --
-demonstrates that pushing the computation towards the producers
-\oleg{or, constructors. Ken said: functionalization} generally
-requires simpler type systems (such as mere HM). The same idea appears
-in many other areas (e.g., ammortized heap analysis, TFP/TLCA paper).
-\oleg{I was told that speculation and informality is permitted in
-  conclusions.}
-
-separate compilation: separate
-typechecking of terms without knowing which particular
-hole they will be plugged into.
-
-\subsection*{Acknowledgments}
-We would like to thank Martin Sulzmann and Walid Taha 
+We solve the problem of embedding a typed object language in a typed
+metalanguage without using GADTs, dependent types, or a universal type.
+Our family of interpreters include an evaluator, a compiler, a partial
+evaluator, and CPS transformers.  It is patent that they never get stuck,
+because we represent object types as metalanguage types.  This work
+makes it safer and more efficient to embed domain\hyp specific languages
+in practical metalanguages such as Haskell and ML\@.
+
+Our main idea is to represent object programs not in an initial algebra
+but using the existing coalgebraic structure of the $\lambda$-calculus.
+More generally, to squeeze more invariants out of a type system as
+simple as Hindley-Milner, we shift the burden of representation and
+computation from consumers to producers: encoding object terms as calls
+to metalanguage functions (\S\ref{ourapproach}); build dynamic terms
+alongside static ones (\S\ref{S:PE-lift}); simulating type functions for
+partial evaluation (\S\ref{S:PE-solution}) and CPS
+transformation~(\S\ref{S:CPS}).  This shift also underlies fusion,
+functionalization, and amortized complexity analysis.
+
+When the metalanguage does provide higher-rank and higher-kind
+polymorphism, we can type-check and compile an object term separately
+from any interpreters it may be plugged into.
+
+\acks%\subsection*{Acknowledgments}
+We thank Martin Sulzmann and Walid Taha 
 for helpful discussions.

### 2007-04-06 tagless/tagless-final.tex
Add bibliography entries

@@
 
-We used a method from \citet{Asai-SAS99} to reduce the amount
+We used a method from \citet{asai-binding-time} to reduce the amount
 of back-and-forth between reify/reflect: in our PE we track both
@@
 manually applied type-level partial-evaluation to our typelevel 
-functions (see \S\ref{S:pe-solution}) to obtain simpler types 
+functions (see \S\ref{S:PE-solution}) to obtain simpler types 
 acceptable to MetaOCaml.  It is worthwhile to note that 
-\citet{Sumii-HOSC2001} also takes advanatage of this method
-\citep[of][]{Asai-SAS99} to get interesting results combining online
+\citet{sumii-hybrid} also takes advantage of this method
+\citep[of][]{asai-binding-time} to get interesting results combining online
 and offline partial evaluation.  Like \citet{SperberThiemann:TwoForOne},
@@
 terms (see for example \citet{baars-typing,Guillemette-Monier-PLPV}
-as well as the messages \citet{haskell-019160,haskell-019161}).  There 
+as well as the messages \citet{haskell-list}).  There 
 take advantage of the host language's types to varying extents.  All 

### 2007-04-06 tagless/tagless-final.tex
Revise abstract (and small changes elsewhere)

@@
 \usepackage{url}
-\usepackage{prooftree1}
+\usepackage{prooftree1} \proofrulebaseline=1.5ex \proofdotseparation=.75ex
 \usepackage{mdwlist}
@@
 \begin{abstract}
-We have built a family of tagless interpretations for a
+We have built the first family of tagless interpretations for a
 higher-order typed object
-language in a typed metalanguage (Haskell and ML), 
-requiring neither dependent
-types, nor generalized algebraic data types, nor a tag elimination 
-postprocessing step. That has been an open problem.
-Our family of interpreters contains 
-an evaluator, a (byte-)compiler and staged evaluator, a partial
-evaluator, a term-size evaluator, and call-by-name and
+language in a typed metalanguage (Haskell or ML) that
+require no dependent
+types, generalized algebraic data types, or
+postprocessing to eliminate tags.
+The interpretations include
+an evaluator, a compiler (or staged evaluator), a partial
+evaluator, and call-by-name and
 call-by-value CPS transformers.
 
-The main tool is to take functions seriously: we encode the
-de Bruijn or higher-order abstract syntax  of object terms 
-not by using constructors for
-an algebraic data type but rather by invoking cogen functions, which occur as
-free variables in the encoded term.  In other words, we eschew initial
-algebras for the wilder world of final algebras.
-\oleg{Or: we expose the co-algebraic structure of the object language.}
-Our approach highlights the importance of order-2 type constructors
-(rank-2 polymorphism), available with ML functors and
-Haskell98 constructor classes. This lets us encode object terms with
-translucent types: enough abstraction is available to interpret the
-terms in various ways, and yet enought type information is available 
-to typecheck the terms and statically assure their interpretation
-never gets stuck.
-
-The encoding of a self-interpreter clarifies our technique 
-\oleg{and rank-2 polymorphism?} 
-as the
-interpretation \oleg{evaluation?} of a term plugged into the context
-of let-polymorphic definitions.
-\oleg{Discuss on Fri, along with the colclusions.}
+Our main idea is to encode
+de Bruijn or higher-order abstract syntax
+using cogen functions rather than data constructors.
+In other words, we represent object terms not in an initial algebra
+but using the coalgebraic structure of the $\lambda$-calculus.
+Our representation also simulates inductive maps from types to
+types, which are required for typed partial evaluation and CPS transformations.
+
+Our encoding of an object term abstracts over the various ways to
+interpret it yet statically assures that the interpreters never get
+stuck.  To achieve self\hyp interpretation and show Jones\hyp
+optimality, we relate this exemplar of higher-rank and higher-kind
+polymorphism (provided by ML functors and Haskell~98 constructor
+classes) to plugging a term into a context of let\hyp polymorphic
+bindings.
 \end{abstract}
@@
 
-A popular way to define and implement a programming language is to embed it in
+A popular way to define and implement a language is to embed it in
 another \citep{reynolds-definitional}.  Embedding means to represent
@@
 entering object terms at a metalanguage interpreter's prompt) or
-by a parser or type checker reading from a text string.
-We leave aside the solved problem of writing such a parser or type checker,
+by a parser\slash type\hyp checker reading from a text string.
+We leave aside the solved problem of writing such a parser\slash type\hyp checker,
 whether using dependent types \citep{WalidICFP02} or not \citep{baars-typing}.
@@
 closed, yet is surprisingly simple and obvious in hindsight.  We stress
-that until now it has been considered impossible to
+that it has been an open problem to
 interpret a typed object language in a typed metalanguage without
@@
 Our contributions are as follows.
-\begin{enumerate}
+\begin{enumerate*}
 \item We build interpreters that evaluate (in direct or continuation\hyp passing styles),
@@
   above.  Self\hyp interpretation turned out harder than expected.
-\end{enumerate}
-
-The structure of our paper is as follows. In the next Section, we
+\end{enumerate*}
+
+The structure of our paper is as follows. In~\S\ref{language}, we
 define the OCaml and Haskell interfaces of the interpreter for our
 language and describe two instances, evaluating a term to a
-metalanguage value and computing term's size. In
-Section~\ref{compiler}, we stage the evaluator, obtaining the tagless
-compiler. Section~\ref{PE} describes partial evaluation, in
+metalanguage value and computing term's size.
+In~\S\ref{compiler}, we stage the evaluator, obtaining the tagless
+compiler. \S\ref{PE} describes partial evaluation, in
 particular, the problem posed by the need to delay the 
@@
 advanced type systems and permitting implementations in Haskell98 or OCaml.
-The discussion Section~\ref{discussion} presents, inter alia, CBN CPS
+The discussion \S\ref{discussion} presents, inter alia, CBN CPS
 interpreter and the CBV CPS transformer. We then describe 
-self-interpretation (Section~\ref{selfinterp}),
-discuss related work (Section~\ref{related}) and then conclude.
+self-interpretation (\S\ref{selfinterp}),
+discuss related work (\S\ref{related}) and then conclude.
 
@@
 testpowfix7 () = 
-  lam (\x -> app (app (testpowfix ()) x)
-                 (int 7))
+ lam (\x -> app (app (testpowfix ()) x) (int 7))
 \end{code}

### 2007-04-06 tagless/tagless-final.tex
Shortening

@@
 
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{Staged typed type-preserving tagless ``interpreters'', \emph{finally}}
-\subtitle{Subtitle Text, if any}
+\makeatletter
+\titlebanner{Submitted to \@conferencename}        % These are ignored unless
+\preprintfooter{\rlap{\@titletext}}   % 'preprint' option specified.
+\def \@formatyear {\llap{\number\year/\number\month/\number\day}}
+\makeatother
+
+\title{Finally Tagless Staged Interpreters for Simpler Typed Languages}
+%Staged typed type-preserving tagless ``interpreters'', \emph{finally}
+%Finally tagless, simply typed, partially evaluated
+\subtitle{\raisebox{-2.5em}[0pt][0pt]{\normalsize\it\begin{minipage}{30em}
+    It should also be possible to define languages with a highly
+    refined syntactic type structure. Ideally, such a treatment should
+    be metacircular, in the sense that the type structure used in the
+    defined language should be adequate for the defining language.
+    \hfill\rm\citep{reynolds-definitional}
+\end{minipage}}}
 
@@
 
-Our code ensures statically that an object term is well-typed and
-closed, yet is surprisingly simple and obvious in hindsight.  We stress
-that it has been an open problem to
+Our contributions are as follows.
+\begin{enumerate*}
+\item We build interpreters that evaluate (\S\ref{language}),
+    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
+   in a typed metalanguage, in direct and continuation\hyp passing styles
+   (\S\ref{discussion}).
+\item All these interpreters use no type tags, patently never get stuck,
+    and need no advanced type-system features such as GADTs, dependent types,
+    or intentional type analysis.
+\item The partial evaluator avoids polymorphic lift and delays binding-time
+    analysis.  It bakes a type-to-type map into the interpreter
+    interface to eliminate the need for GADTs and thus remain portable
+    across Haskell 98 and ML.
+\item We use the type system of the metalanguage
+    to check statically that an object program is well-typed and closed.
+\item We show clean, comparable implementations in MetaOCaml and Haskell.
+\item We specify a functor signature that encompasses all our interpreters, from
+    evaluation and compilation (\S\ref{language}) to partial evaluation and CPS transformation (\S\ref{PE}).
+\item We point a clear way to extend the object language with more features
+    such as state (\S\ref{state}).
+\item We describe an approach to self\hyp interpretation compatible with the
+  above (\S\ref{selfinterp}).  Self\hyp interpretation turned out harder than expected.
+\end{enumerate*}
+Our code is surprisingly simple and obvious in hindsight, but
+it has been an open problem to
 interpret a typed object language in a typed metalanguage without
@@
 expressed.''  We further discuss related work in~\S\ref{related}.
-
-Our contributions are as follows.
-\begin{enumerate*}
-\item We build interpreters that evaluate (in direct or continuation\hyp passing styles),
-  compile, and partially evaluate a typed higher-order object language
-   in a typed metalanguage.
-\item All these interpreters use no type tags, patently never get stuck,
-    and need no advanced type-system features such as GADTs, dependent types,
-    or intentional type analysis.
-\item The partial evaluator avoids polymorphic lift and delays binding-time
-    analysis.
-\item We use the type system of the metalanguage
-    to check statically for well-typed and closed object programs.
-\item We show clean, comparable implementations in MetaOCaml and Haskell.
-\item We specify a functor signature that encompasses all our interpreters, from
-    evaluation and compilation to partial evaluation and CPS transformation.
-\item We point a clear way to extend the object language with more features
-  such as state.
-\item We describe an approach to self\hyp interpretation compatible with the
-  above.  Self\hyp interpretation turned out harder than expected.
-\end{enumerate*}
-
-The structure of our paper is as follows. In~\S\ref{language}, we
-define the OCaml and Haskell interfaces of the interpreter for our
-language and describe two instances, evaluating a term to a
-metalanguage value and computing term's size.
-In~\S\ref{compiler}, we stage the evaluator, obtaining the tagless
-compiler. \S\ref{PE} describes partial evaluation, in
-particular, the problem posed by the need to delay the 
-binding-time analysis of term abstractions and the apparent 
-need for a typecase. We then show two solutions. The first, imperfect
-one, uses GADT. The better and more portable solution bakes in the
-typecase idiom into the interpreter interface, eliminating the need for
-advanced type systems and permitting implementations in Haskell98 or OCaml.
-The discussion \S\ref{discussion} presents, inter alia, CBN CPS
-interpreter and the CBV CPS transformer. We then describe 
-self-interpretation (\S\ref{selfinterp}),
-discuss related work (\S\ref{related}) and then conclude.
 
@@
 \end{code}
-gives
+gives the desired result
 \begin{code}
@@
 \end{code}
-as desired.
 
@@
 independent of the strategy of the metalanguage. We demonstrate the
-same independence in the typed setting \oleg{mention the quote from
-Reynolds that typed setting would be nice?}\jacques{I think so, but from
-which paper? The one ReynoldsPlotkin paper is typed}\ccshan{I think Oleg
-meant ``It should also be possible to define languages with a highly
- refined syntactic type structure. Ideally, such a treatment should be
- metacircular, in
- the sense that the type structure used in the defined language should be adequate for the
- defining language'' \citep{reynolds-definitional}, but perhaps Jacques
- meant another paper by ``ReynoldsPlotkin''?}. Specifically, we
+same independence in the typed setting. Specifically, we
 demonstrate CBN evaluation for our object language in OCaml, which is
@@
                           else (ee ()).ko k)}
-  ...
   let lam f = {ko = fun k -> k f}
@@
     let rec fx f n = app (f (lam (fx f))) n in 
-     lam(fx f)
-
+    lam (fx f)
   let get_res x = x.ko (fun v -> v)
@@
  open S
- let diverg () = 
-   app (lam (fun x -> int 1)) 
+ let diverg () = app (lam (fun x -> int 1)) 
                      (app (fix (fun f -> f)) (int 2))
@@
 \end{code}
-
 If we interpret |EXS| with the |R| interpreter of
-\S\ref{S:interpreter-RL} realized in OCaml
+\S\ref{S:interpreter-RL}
 \begin{code}
@@
 
-\oleg{mention some formal results? Like indifference theorems by
-  Plotkin? And say we generalized their result to the type setting?
-  Mention this in the conclusions?}
-\jacques{you have strayed outside what I know -- what Plotkin results?
-  I always consider mentionning formal results a good thing}
-
 \subsection{CBV CPS transformers}
@@
   include RCN
-  let lam f = {ko = 
-    fun k -> k (fun e -> e.ko 
-         (fun v -> f {ko = fun k -> k v}))}
+  let lam f = {ko = fun k -> k (fun e ->
+    e.ko (fun v -> f {ko = fun k -> k v}))}
 end
@@
 is useful for achieving CBV evaluation of the object language
-no matter whether the metalanguage is CBV or CBN.
+whether the metalanguage is CBV or CBN.
 
@@
 
-It is instructive to see the result of evaluating a few examples. With 
-|CPST| instantiated with a straightforward compiler |C| as above, our
+It is instructive to evaluate a few examples. With 
+|CPST| instantiated with the compiler~|C| above, our
 running test |T.test1 ()| gives
@@
 \subsection{State and imperative features}
+\label{state}
 
@@
 capture free variables of the term.  By pre-encoding a hole to a hole,
-we extend pre-encoding from a translation on terms to a translation on
+we extend pre-encoding from a translation on terms to one on
 contexts.}
@@
 
-From reading \citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}, our 
-impression is that the claim is:
+\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
+\citet{peyton-jones-simple} seem to argue as follows that a self\hyp
+interpreter of a typed language cannot be Jones\hyp optimal.
 \begin{enumerate}
@@
 
-In particular, \citet[\S5]{taha-tag}
-states that ``expressing such an interpreter in a statically typed
-programming language is a rather subtle matter. In fact, it is only
-recently that some work on programming type-indexed values in ML
-\citep{yang-encoding} has given a hint of how such a function can be
-expressed.''
-
 \citet{Danvy-tagging-encoding} discusses Jones optimality at length.

### 2007-04-06 tagless/incope.hs
fixed the implementation of fix to fix the divergence. Added checks for Prop 5

@@ -558,6 +558,7 @@ htestgib1  = compH . testgib1 $ ()
 htestgib1r = eval . testgib1 $ ()
 
 -- ---------------------------------------------------------------------------
+-- Self-interpretation
 
 twice_inc_3_e () =
     -- The evaluation context in the following four lines is a self-interpreter
@@ -576,6 +577,12 @@ twice_inc_3_e () =
 test_inc3e_r = compR $ twice_inc_3_e ()
 test_inc3e_c = compC $ twice_inc_3_e ()
 test_inc3e_p = compP $ twice_inc_3_e ()
+-- to verify the Prop 5
+test_inc3p = compP $ (app (app
+                       (lam (\twice -> lam (\inc_ -> 
+                                              app (app twice inc_) (int 3))))
+                       (lam (\f -> lam (\x -> app f (app f x)))))
+                       (lam (\n -> add n (int 1))))
 -- 5
 
 -- This is for computing *size*.
@@ -652,6 +659,9 @@ test3_e () =
 test_t3_r = compR $ test3_e ()
 test_t3_c = compC $ test3_e ()
 test_t3_p = compP $ test3_e ()
+-- to verify the Prop 5
+test_t3p = compP $ lam (\x -> (add (app x (int 1)) (int 2)))
+
 
 {- And now something harder: powfix
 testpowfix () = lam (\x ->
@@ -671,8 +681,7 @@ testpowfix_e () =
     let leq_ = lam (\m -> lam (\n -> leq m n)) in
     let if__ = lam (\be -> lam (\ee -> lam (\te -> if_ be ee te))) in
     let int_ = lam (\i -> i) in
-    -- careful, trying to print something with fix_ can loop!
-    let fix_ = lam (\f -> app f (app fix_ f)) in
+    let fix_ = lam (\f -> fix (\self -> app f self)) in
     -- The term in the following three lines is the object term
     -- testpowfix above
     -- encoded in the object language then encoded in the metalanguage.
@@ -689,10 +698,13 @@ testpowfix_e () =
 
 test_pf_r = compR.fst $ testpowfix_e ()
 test_pf_c = compC.fst $ testpowfix_e ()
-test_pf_p = compP.fst $ testpowfix_e () -- don't print this one!
+test_pf_p = compP.fst $ testpowfix_e ()
 test_pf7_r = compR.snd $ testpowfix_e ()
 test_pf7_c = compC.snd $ testpowfix_e ()
[…]

### 2007-04-06 tagless/tagless-final.tex
related work is now ready to be edited

@@
 \makeatother
+
+% Allow things to flow a little more tightly around figures
+\renewcommand\floatpagefraction{.95}
+\renewcommand\topfraction{.95}
+\renewcommand\bottomfraction{.95}
+\renewcommand\textfraction{.05}
 
@@
 We have been unable to find a direct reference that establishes that
-the typed $\lambda$-calculus possesses a final co-algebra structure.
+the \emph{typed} $\lambda$-calculus possesses a final co-algebra structure.
 For the untyped $\lambda$-calculus, 
@@
 \citet{honsell99coinductive} for references to the foundational work
-in this important area.
-
-\jacques{Dig through the last couple of emails I just got from the
-category theory list with references, and this section will be done.}
+in this important area.  Particularly intriguing is the link to the
+coinductive aspects of B\"{o}hm trees, as pointed out by
+\citet{berarducci-models} and \citet[Example 4.3.4]{jacobs-coalgebra}.
+
+One way to interpret some of our results is that we eschew sums (at either the
+term or type, or even kind level) for our terms, but instead use a record type 
+(via Functors in OCaml and typeclasses in Haskell).
+For the self-interpreter, we then proceed to use a Church encoding for
+recursive data types \citep{bohm-automatic}.
 
@@
 We thank Martin Sulzmann and Walid Taha 
-for helpful discussions.
+for helpful discussions.  Sam Staton, Pieter Hofstra, and Bart Jacobs
+kindly provided some useful references on co-algebraic structures in the
+$\lambda$\hyp calculus
 

### 2007-04-06 tagless/tagless-final.tex
remove all comments.  make all coalgebra references use the same spelling.  check uses of initial/final (all ok)

@@
 term) are reported at compile time. 
-\oleg{Jacques, do we use final and initial here correctly?}
 Because the new interpreter uses no
@@
     equal up to $\alpha$\hyp conversion.
-    \ccshan{Check this}
+%    \ccshan{Check this}
+%    \jacques{seems correct.  If it said that the results are operationally
+%    equivalent to e itself, then that would be less clear.}
 \end{prop}
@@
 
-or, if we denote the sequence of initial arguments as S r, terms have
+or, if we denote the sequence of initial arguments as |S r|, terms have
 the type |S r -> r tau|
@@
 We have been unable to find a direct reference that establishes that
-the \emph{typed} $\lambda$-calculus possesses a final co-algebra structure.
+the \emph{typed} $\lambda$-calculus possesses a final coalgebra structure.
 For the untyped $\lambda$-calculus, 
@@
 for helpful discussions.  Sam Staton, Pieter Hofstra, and Bart Jacobs
-kindly provided some useful references on co-algebraic structures in the
+kindly provided some useful references on coalgebraic structures in the
 $\lambda$\hyp calculus

### 2007-04-06 tagless/tagless-final.tex
lots of little edits here and there, for style and length

@@
    in a typed metalanguage, in direct and continuation\hyp passing styles
-   (\S\ref{discussion}).
+   (\S\ref{variations}).
 \item All these interpreters use no type tags, patently never get stuck,
@@
 
-\section{Variations and discussion}\label{discussion}
+\section{Variations}\label{variations}
 
@@
 we show how a call-by-name CPS interpreter, a call-by-value CPS
-program transformation, as well as how to deal with imperative features.
-Finally, we comment on the requirements on the type system to support 
-our approach.
+program transformation.  Finally, we comment on the requirements on the type
+system to support our approach.
 
@@
 The object language generally inherits the evaluation strategy from
-the metalanguage. Evaluating an application |e1 e2| in OCaml requires
-the evaluation of |e2| before performing the application. Therefore,
-when evaluating the encoding of the object language application 
-|app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
-possible to represent a call-by-name (CBN) object language in a
-call-by-value (CBV) metalanguage. The answer, given already by 
-\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is yes thanks to CPS, which they
-specifically introduced 
+the metalanguage -- call-by-value (CBV) in OCaml, call-by-name (CBN) in Haskell.
+One may wonder if it is
+possible to represent a call-by-name object language in a
+call-by-value metalanguage. The answer, given already by 
+\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is yes
+thanks to CPS, which they specifically introduced 
 to make the evaluation strategy of a definitional interpreter
@@
 same independence in the typed setting. Specifically, we
-demonstrate CBN evaluation for our object language in OCaml, which is
-call-by-value. We do so by building a CBN CPS interpreter for the object
+demonstrate CBN evaluation for our object language in OCaml.
+We do so by building a CBN CPS interpreter for the object
 language.
 
-The CBN CPS interpretation was introduced by \citet{PlotkinCBN}. As in
-CBV CPS, the interpretation of an object term is a function
+The interpretation of an object term is a function
 mapping a continuation~|k| to the answer
@@
 \end{code}
-In the tell-tale sign of CBN, in interpreting the application
-we do not `evaluate' the argument e2, that is, we do apply it to the
+Charateristic of CBN, interpreting the application does not
+``evaluate'' the argument |e2|, that is, we do apply it to the
 continuation. Rather, we pass the unevaluated |e2| to the abstraction.
 The result of |lam (fun x -> add x (int 1))| has the type
-|((((int->'w1)->'w1) -> ((int->'w1)->'w1)) -> 'w2) -> 'w2|.
-We want to collect the above interpretation functions into a module
-with the |Symantics| signature, so as to fit the CBN CPS interpreter within our
+|((((int->'w1)->'w1) -> ((int->'w1)->'w1)) ->'w2)->'w2|.
+We would like to collect those interpretation functions into a module
+with signature |Symantics|, to include the CBN CPS interpreter within our
 general framework. Alas, we encounter the same problem we had in
-Section~\ref{S:PE-problem}, in writing a type equation for the |repr|
-type that can represent both the type of |int i| and the type of 
-|lam f|. We cannot say that the result of interpreting an object
+\S\ref{S:PE-problem}.
+We cannot say that the result of interpreting an object
 expression of type |t| is |(t->'w)->'w| because if |t| happens to
-be an arrow type (such as |int->int| in the above example), the result
+be an arrow type, the result
 type obviously does not fit the desired pattern. We need a type
-function with a typecase distinction. In Section~\ref{S:PE-solution} we have
-shown how to encode such a function, using the extra type
-argument |'sv| of the |repr| type. Furthermore, we notice that the type 
+function with a typecase distinction.  Our previous solution 
+(\S\ref{S:PE-solution}) where we use an extra type
+argument |'sv| of the |repr| type, to encode such type-level functions,
+also works here.  Furthermore, we notice that the type 
 function |repr_pe| needed for the partial evaluator 
 (in~\S\ref{S:PE-solution}) is precisely the same type function we
-need for CBN CPS. This gives us for the following CBN CPS interpreter,
-with the |Symantics| signature:
+need for CBN CPS.
 
@@
 We have made our interpreter fully polymorphic over the answer type.
-We could have just as easily made |RCN| a functor, parameterized over
-the answer type -- as is common when writing CPS in SML. In the
-present case, we chose to use the higher-rank polymorphism of offered by
+We could have also made |RCN| a functor, parameterized over
+the answer type -- as is common when writing CPS in SML. Here
+we chose to use the higher-rank polymorphism offered by
 OCaml record types.
@@
 \begin{code}
-module EXS(S: Symantics) = struct
- open S
+module EXS(S: Symantics) = struct open S
  let diverg () = app (lam (fun x -> int 1)) 
@@
 the evaluation of that argument before proceeding. This approach is in
-line with \citet{reynolds-relation} (albeit in a typed setting). The interpreter RCV
-is useful for achieving CBV evaluation of the object language
+line with \citet{reynolds-relation} (albeit in a typed setting). The
+interpreter RCV is useful for achieving CBV evaluation of the object language
 whether the metalanguage is CBV or CBN.
@@
     S.app e2 (S.lam (fun v2 -> S.app k (S.add v1 v2))))))
-  let if_ ec et ef = S.lam (fun k ->
-    S.app ec (S.lam (fun vc ->
-    S.if_ vc (fun () -> S.app (et ()) k) (fun () -> S.app (ef ())
-    k))))
   ...
@@
 \end{code}
-The code literally establishes a mapping from CPS interpretations to
-(direct) interpretations, the latter performed by 
+The code explicitly establishes a mapping from CPS interpretations to
+(direct) interpretations, as performed by 
 the base interpreter |S|.
 
-A reader may notice that the above module does not define the type
-equation for |repr| and so it does not have the signature Symantics.
+A reader may notice that the above module does not define |repr|
+and thus does not have signature |Symantics|.
 The reason for that is again the result type of |lam f|. Whereas
@@
 type |('c,..., (int -> 'w) -> 'w) S.repr|,
-the result of |lam (fun x -> add x (int 1))| has type
+the result of |lam (fun x -> add x (int 1))| has type\\
 |('c,...((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr|. 
-Hence, to write the type equation for |CPST.repr| we need a 
+Hence, to write the type equation for |CPST.repr| we again need a 
 type-discriminating
@@
 To save space, we show a simplified and less principled approach that
-lets us use the module |CPST| as it is. Because the module does not
-have the signature Symantics, we cannot apply the |EX| functor to it.
-But we can write the tests nevertheless:
+lets us use the module |CPST| as is. Because the module does not
+have the signature |Symantics|, we cannot apply the |EX| functor to it.
+But we can nevertheless write the tests:
 \begin{code}
@@
 compiler |C| in the case above) so that we can use the result |M| to
-interpret object level terms. The sample terms are \emph{exactly}
-as we used before. We just copied them over. That copying step is the
-price we pay for the simplified treatment. We shall also encounter this
-copying when discussing the self-interpreter in the next section. We
+interpret object level terms. Those terms are \emph{exactly} as before.
+Having to textually copy the terms is the
+price we pay for this simplified treatment. We shall also encounter this
+copying when discussing self-interpretation.  We
 will see then that the copying step is not frivolous, but represents
@@
 It is instructive to evaluate a few examples. With 
-|CPST| instantiated with the compiler~|C| above, our
-running test |T.test1 ()| gives
+|CPST| instantiated with the compiler~|C| above,
+|T.test1| gives
 \begin{code}
@@
 
-which shows the naive CPS transformation of the original code. The result is
-quite unoptimal with several apparent $\beta$-redexes. 
-By instantiating |CPST| with the |P| evaluator instead, our
-running test gives the expected
+\noindent which shows the na\"{\i}ve CPS transformation of the original code.
+The result is sub-optimal with several apparent $\beta$-redexes. 
+By instantiating |CPST| with |P| instead, we get the expected
 \begin{code}
@@
 metalanguage in the preceding sections, but it produces
-terms in the object language rather than metalanguage.
+terms in the object language rather than the metalanguage.
 
@@
 Because the evaluator above mostly consists of glorified identity
-functions, we expect even our simple partial evaluator to reduce the
+functions, we expect our simple partial evaluator to reduce the
 result of this instantiation to~$e$.  In general, to
-interpret~$\encode{e}$ using an interpreter is to instantiate these free
+interpret~$\encode{e}$ using an interpreter is to instantiate its free
 variables by that interpreter's definitions.
@@
 does not type-check in any typing environment, because $\ident{lam}$ needs
-to take two incompatible types.  In sum, we need polymorphism in the
+to take two incompatible types.  In sum, we need more polymorphism in the
 object type system to type $\ident{lam}$, $\ident{app}$, $\ident{fix}$,
 and~$\ident{if\_}$ (and $\Encode{e}$ and~$\si$).
-(The polytypes in |Symantics| are where Haskell type classes and OCaml
+(The polytypes in |Symantics| given by Haskell's type classes and OCaml's
 modules supply this polymorphism.)  Moreover, we need to encode any
-polymorphism in the object language into the object language to achieve
+polymorphism of the object language \emph{into} the object language to achieve
 self\hyp interpretation.
@@
 
-Instead of adding higher-rank and higher-kind polymorphism to our object language (along
-with polymorphism over kinds), we add let-bound polymorphism.  As usual,
+Instead of adding higher-rank and higher-kind polymorphism to our object
+language (along with polymorphism over kinds!), we add let-bound polymorphism.
+As usual,
 we can add a new typing rule
@@
 equivalent to $\si[\encode{e}]$, and in turn to~$e$.  In other words,
-$\si$ can interpret itself.
-
-Our partial evaluator is optimal with respect to the self\hyp
-interpreter~$\si$.
+$\si$ can interpret itself.  Our partial evaluator is optimal with respect to
+the self\hyp interpreter~$\si$.
 \begin{prop}
@@
     equal up to $\alpha$\hyp conversion.
-%    \ccshan{Check this}
-%    \jacques{seems correct.  If it said that the results are operationally
-%    equivalent to e itself, then that would be less clear.}
 \end{prop}
@@
 multiple ways.  The meta-notions of contexts and plugging may seem ad
-hoc, but in fact they are explications of the type-class and module
+hoc, but in fact they are just reflections of the type-class and module
 machinery that we have been using in this paper all along.
@@
 In the presence of let-bound polymorphism, we can understand a term
-waiting to be plugged into a context as a higher-rank and higher-kind abstraction over
-the context.  Whereas our object language does not support higher
-abstraction, our metalanguages do, so they can type-check an object term
+waiting to be plugged into a context as a higher-rank and higher-kind
+abstraction over the context.  Whereas our object language does not support
+higher abstraction, our metalanguages do, so they can type-check an object term
 separately from its interpreter---either as a functor from a |Symantics| module
@@
 
-\citet{Danvy-tagging-encoding} discusses Jones optimality at length.
+\citet{Danvy-tagging-encoding} discuss Jones optimality at length.
 There, the authors demonstrate the usefulness of HOAS for typed
@@
 terms (see for example \citet{baars-typing,Guillemette-Monier-PLPV}
-as well as the messages \citet{haskell-list}).  There 
-take advantage of the host language's types to varying extents.  All 
-these approaches are very successful for proper terms, but less so for
-ill-typed terms, where error messages can be quite obscure.  We use the
-host's typing facilities to do all the work, and thus inherit the strengths
+as well as the messages \citet{haskell-list}).  These 
+take advantage of the host language's types to varying extents.  While
+very successful for proper terms, results for
+ill-typed terms can produce rather obscure error messages.  We use the
+host's typing facilities for this, and thus inherit the strengths
 and weaknesses of the host language.  While that may not be optimal, it is
@@
 an operational definition of |'sv -> 'dv| (which is easy), and an
-implementation of |'dv| which ``perfectly reflects'' |'sv| without 
-however making this uniformly available.
+implementation of |'dv| which ``perfectly reflects'' |'sv| but without 
+requiring an implementation.
 

### 2007-04-06 tagless/tagless-final.tex
lots more little edits everywhere

@@
 \citep{yang-encoding} has given a hint of how such a function can be
-expressed.''  We further discuss related work in~\S\ref{related}.
+expressed.''  We discuss related work in~\S\ref{related}.
 
@@
 
-In contrast to \S\ref{intro}, we henceforth encode 
-binding in object programs using higher-order
+In contrast to \S\ref{intro}, we encode binding using higher-order
 abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
@@
 
-We embed this language in (Meta)OCaml and Haskell.  In Haskell,
+We embed our language in (Meta)OCaml and Haskell.  In Haskell,
 the functions that construct object terms are methods in a type class
-|Symantics|. The class is so named because its interface gives the syntax for
-the object language and its instances give the semantics.
-The parameter |repr| to the class has kind |* -> *|.
+|Symantics| (with a parameter |repr| of kind |* -> *|). The class is so named
+because its interface gives the syntax of the object language and its
+instances give the semantics.
 \begin{code}
 class Symantics repr where
-
   int  :: Int  -> repr Int
@@
 \end{prop}
-Conversely, the type system of the metalanguage checks statically that the
+Conversely, the type system of the metalanguage statically checks that the
 represented object term is well-typed and closed.
@@
 module type Symantics = sig
-
   type ('c, 'dv) repr
@@
   val if_ : ('c, bool) repr -> (unit -> ('c, 'da) repr) -> (unit -> ('c, 'da) repr) -> ('c, 'da) repr
-
 end
@@
 and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
-signature that suffices until~\S\ref{PE}.
-Two differences between this signature and the |Symantics| class in
-Haskell are: the additional type parameter |c|, an
-environment classifier \citep{WalidPOPL03} required by MetaOCaml for
-code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
-|fix| and thunk types in |if_| due to OCaml's being a call-by-value
+signature that suffices until~\S\ref{PE}.  The two differences are:
+the additional type parameter |c|, an
+environment classifier \citep{WalidPOPL03} (required by MetaOCaml for
+code generation in~\S\ref{compiler}); and the $\eta$-expanded type for
+|fix| and thunk types in |if_| since OCaml is a call-by-value
 language.
@@
 The dummy argument to |test1| and |testpowfix| is an artefact of
-MetaOCaml, again related to monomorphism: in order for us to run a
+MetaOCaml, related to monomorphism: in order for us to run a
 piece of generated code, it must be polymorphic in its environment
 classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
-The value restriction then dictates that
+The value restriction dictates that
 the definitions of our object terms must look syntactically like
-values. Alternatively to giving the |()| argument, we could have used
+values. Alternatively, we could have used
 the rank-2 record types of OCaml to maintain the necessary polymorphism.
@@
 module R = struct
-  (* absolutely no wrappers *)
-  type ('c,'dv) repr = 'dv
+  type ('c,'dv) repr = 'dv (* no wrappers *)
 
@@
   let if_  eb et ee = if eb then et () else ee ()
-
 end
@@
 OCaml's addition, and |app| is OCaml's application. To run our
-examples, we instantiate the |EX| functor from~\S\ref{encoding} with this |R|
-module.
+examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|.
 \begin{code}
@@
 Inserting brackets and escapes appropriately into the
-evaluator~|R| above yields the compiler~|C| below.
+evaluator~|R| above yields the simple compiler~|C| below.
 % this code does not have the 'sv parameter. It shows up later.
@@
 module C = struct
-
   type ('c,'dv) repr = ('c,'dv) code
@@
     .<let rec self n = .~(f .<self>.) n in self>.
-
   let add e1 e2     = .<.~e1 + .~e2>.
@@
     .<if .~eb then .~(et ()) else .~(ee ())>.
-
 end
@@
 gives the code value |.<(fun x_6 -> x_6) true>.|
-of inferred type |('c, bool) C.repr|.  Similarly, interpreting |testpowfix7|
+of inferred type |('c, bool) C.repr|.  Interpreting |testpowfix7|
 with
@@
 \end{code}
-gives the code value
+gives a code value with many apparent $\beta$-redexes:
 \begin{code}
@@
                else x_2 * (self_3 (x_5 + (-1))))
-   n_4
-  in self_3) x_1 7>.
-\end{code}
-with many apparent $\beta$-redexes.
+   n_4 in self_3) x_1 7>.
+\end{code}
 
@@
 the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
-verbose) than the corresponding MetaOCaml code given above. The
+verbose) than the corresponding MetaOCaml code above. The
 accompanying code gives the full details.
@@
 for object types.  We then discuss residualization and binding-time
-analysis.  Our partial evaluator is a modular extension to the evaluator
+analysis.  Our partial evaluator is a modular extension of the evaluator
 in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
@@
 
-Roughly speaking, a partial evaluator interprets each object term either
-to a static (present-stage) term, as produced by the evaluator~|R|, or
-to a dynamic (future-stage) term, as produced by the compiler~|C|.  To
+Roughly speaking, a partial evaluator interprets each object term as either
+a static (present-stage) term (using the evaluator~|R|), or
+a dynamic (future-stage) term, (using the compiler~|C|).  To
 distinguish between static and dynamic terms, one might try
-to define partial evaluation as the Haskell data type
+to define partial evaluation via the Haskell data type
 \begin{code}
@@
 \end{code}
-Integer and boolean literals are obviously immediate, present-stage
-values. Addition yields a static term if and only if both operands
-are static. If so, we use the evaluator~|R| to add the operands.
-If not, we extract dynamic terms from the operands and add them
-using the compiler~|C|.
-Thus we use the overloaded functions |int| and |bool| from the |Symantics|
+Integer and boolean literals are immediate, present-stage
+values. Addition yields a static term (via~|R|) if and only if both operands
+are static, otherwise we extract the dynamics terms from the operands and
+add them using the compiler~|C|.  Uniformly, we use
+the overloaded functions from the |Symantics|
 instances for |R| and~|C| to define those for~|P0|.
@@
 a problem with |if_|.  If the first argument to |if_| is a dynamic term
-(type |C Bool|) and the second and third arguments are a static term
+(type |C Bool| say) and the second and third arguments are a static term
 (type |R a|) and a dynamic term (type |C a|), then we need to convert
@@
 module type Symantics = sig
-
   type ('c,'sv,'dv) repr
@@
 module P = struct
-
   type ('c,'sv,'dv) repr = {st: 'sv option;
                             dy: ('c,'dv) code}
-
   let abstr {dy = x} = x
@@
                      (fun () -> abstr (ee ())))
-
   let lam f =
@@
        in {st = Some self; dy = fdyn}
-
 end
@@
 
+We are not the first to implement a typed interpreter for a typed
+language -- \citet{laod93} use a typed version of the SK language
+(and typeclasses) to implement a meta-circular interpreter (rather than
+a self-interpreter).  Interestingly, their meta-circular interpreter
+appears to be tagfree!  However, they could not have implemented a
+compiler or a partial-evaluator in the same way since they rely quite
+heavily on having injection-projection pairs.
+
 \citet{fiore:nbe-ppdp2002,balat:tdpe-popl2004} also deal with a 
@@
 
-We are not the first to implement a typed interpreter for a typed
-language -- \citet{laod93} use a typed version of the SK language
-(and typeclasses) to implement a meta-circular interpreter (rather than
-a self-interpreter).  Interestingly, their meta-circular interpreter
-appears to be tagfree!  However, they could not have implemented a
-compiler or a partial-evaluator in the same way since they rely quite
-heavily on having injection-projection pairs (through reflection and
-reification).
-
-We have been unable to find a direct reference that establishes that
+We have been unable to find work that establishes that
 the \emph{typed} $\lambda$-calculus possesses a final coalgebra structure.
@@
 contexts with a hole \citep[p.13]{honsell99coinductive} for the definition
-of \emph{observational equivalence}, which directly relates to our
-use of terms with a hole as a context for our self-interpreter 
-(\S\ref{selfinterp}).  The reader is directed to bibliography of
+of \emph{observational equivalence}
+(see \S\ref{selfinterp}).  The reader is directed to bibliography of
 \citet{honsell99coinductive} for references to the foundational work

### 2007-04-06 tagless/tagless-final.tex
Almost done with Section 5

@@
 tagging or type\hyp system extensions.  For example, \citet{taha-tag}
-state that ``expressing such an interpreter in a statically typed
+say that ``expressing such an interpreter in a statically typed
 programming language is a rather subtle matter. In fact, it is only
@@
 
-Roughly speaking, a partial evaluator interprets each object term as either
-a static (present-stage) term (using the evaluator~|R|), or
-a dynamic (future-stage) term, (using the compiler~|C|).  To
-distinguish between static and dynamic terms, one might try
-to define partial evaluation via the Haskell data type
+Roughly, a partial evaluator interprets each object term to either
+a static (present-stage) term (using the evaluator~|R|) or
+a dynamic (future-stage) term (using the compiler~|C|).  To
+distinguish between static and dynamic terms, one might define
+the Haskell data type
 \begin{code}
@@
 \begin{code}
-abstrInt :: P0 Int -> C Int
-abstrInt (S0 r) = int (unR r)
-abstrInt (E0 c) = c
-
-abstrBool :: P0 Bool -> C Bool
-abstrBool (S0 r) = bool (unR r)
-abstrBool (E0 c) = c
+abstrI :: P0 Int -> C Int
+abstrI (S0 r) = int (unR r)
+abstrI (E0 c) = c
+
+abstrB :: P0 Bool -> C Bool
+abstrB (S0 r) = bool (unR r)
+abstrB (E0 c) = c
 \end{code}
@@
   add (S0 e1) (S0 e2) = S0 (add e1 e2)
-  add e1 e2 = E0 (add (abstrInt e1)
-                      (abstrInt e2))
+  add e1 e2 = E0 (add (abstrI e1) (abstrI e2))
 \end{code}
 Integer and boolean literals are immediate, present-stage
-values. Addition yields a static term (via~|R|) if and only if both operands
-are static, otherwise we extract the dynamics terms from the operands and
-add them using the compiler~|C|.  Uniformly, we use
-the overloaded functions from the |Symantics|
-instances for |R| and~|C| to define those for~|P0|.
+values. Addition yields a static term (using~|R|) if and only if both operands
+are static; otherwise we extract the dynamic terms from the operands and
+add them using~|C|.  Thus the two uses of |add| above refer to
+|add| for~|R| and |add| for~|C|.
 
@@
 a problem with |if_|.  If the first argument to |if_| is a dynamic term
-(type |C Bool| say) and the second and third arguments are a static term
+(type |C Bool|) and the second and third arguments are a static term
 (type |R a|) and a dynamic term (type |C a|), then we need to convert
@@
 
-\section{Variations}\label{variations}
-
-We show that our |Symantics| and generally our approach accomodates
-a number of variants without any (new) difficulties.  In particular,
-we show how a call-by-name CPS interpreter, a call-by-value CPS
-program transformation.  Finally, we comment on the requirements on the type
-system to support our approach.
+\section{Continuation\hyp passing style}\label{variations}
+
+Our approach accommodates
+several variants, including
+a call-by-name CPS interpreter and a call-by-value CPS
+transformation.
 
@@
 The object language generally inherits the evaluation strategy from
-the metalanguage -- call-by-value (CBV) in OCaml, call-by-name (CBN) in Haskell.
-One may wonder if it is
-possible to represent a call-by-name object language in a
-call-by-value metalanguage. The answer, given already by 
-\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is yes
-thanks to CPS, which they specifically introduced 
-to make the evaluation strategy of a definitional interpreter
-independent of the strategy of the metalanguage. We demonstrate the
-same independence in the typed setting. Specifically, we
-demonstrate CBN evaluation for our object language in OCaml.
-We do so by building a CBN CPS interpreter for the object
-language.
+the metalanguage---call-by-value (CBV) in OCaml, call-by-name (CBN) in
+Haskell.  To represent a CBN object language in a CBV metalanguage,
+\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}
+introduce CPS to make the evaluation strategy of a definitional
+interpreter indifferent to that of the metalanguage. To achieve the same
+indifference in the typed setting, we build a CBN CPS interpreter for
+our object language in OCaml.
 
@@
 \end{code}
-In both cases, the returned value has type 
-|(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.
-
-The interpretation of abstraction and application in CBN CPS is
-different from CBV CPS and is as follows:
+In both |int| and |add|, the interpretation has type 
+|(int -> 'w)|\texttt{ }|-> 'w|, where |'w| is the (polymorphic) answer type.
+
+Unlike CBV CPS, the CBN CPS interprets
+abstraction and application as follows:
 \begin{code}
 let lam f = fun k -> k f
-let app e1 e2 = fun k -> e1 (fun f -> (f e2) k)
-\end{code}
-Charateristic of CBN, interpreting the application does not
-``evaluate'' the argument |e2|, that is, we do apply it to the
-continuation. Rather, we pass the unevaluated |e2| to the abstraction.
-The result of |lam (fun x -> add x (int 1))| has the type
-|((((int->'w1)->'w1) -> ((int->'w1)->'w1)) ->'w2)->'w2|.
+let app e1 e2 = fun k -> e1 (fun f -> f e2 k)
+\end{code}
+Charateristic of CBN, |app e1 e2|
+does not evaluate the argument~|e2| by applying it to the
+continuation~|k|. Rather, it passes |e2| unevaluated to the abstraction.
+Interpreting $\fun{x} x+1$ yields type
+\begin{code}
+((((int -> 'w1) -> 'w1) -> (int -> 'w1) -> 'w1)
+ -> 'w2) -> 'w2
+\end{code}
+
 We would like to collect those interpretation functions into a module
 with signature |Symantics|, to include the CBN CPS interpreter within our
-general framework. Alas, we encounter the same problem we had in
-\S\ref{S:PE-problem}.
-We cannot say that the result of interpreting an object
-expression of type |t| is |(t->'w)->'w| because if |t| happens to
-be an arrow type, the result
-type obviously does not fit the desired pattern. We need a type
-function with a typecase distinction.  Our previous solution 
-(\S\ref{S:PE-solution}) where we use an extra type
-argument |'sv| of the |repr| type, to encode such type-level functions,
-also works here.  Furthermore, we notice that the type 
-function |repr_pe| needed for the partial evaluator 
-(in~\S\ref{S:PE-solution}) is precisely the same type function we
-need for CBN CPS.
-
+general framework. Alas, as in~\S\ref{S:PE-problem}, the type of
+an object term inductively determines the type of its interpretation:
+the interpretation of an object term of type~$t$ may not have type
+|(|$t$|->'w)->'w|, because $t$ may be a function type.  Again we
+simulate a type function with a typecase distinction, using an extra
+type argument to |repr|. Happily, the type function |repr_pe| needed for
+the partial evaluator 
+in~\S\ref{S:PE-solution} is precisely the same type function we
+need for CBN CPS\@.
 \begin{code}
@@
     lam (fx f)
-  let get_res x = x.ko (fun v -> v)
+  let run x = x.ko (fun v -> v)
 end
@@
 
-We have made our interpreter fully polymorphic over the answer type.
-We could have also made |RCN| a functor, parameterized over
-the answer type -- as is common when writing CPS in SML. Here
-we chose to use the higher-rank polymorphism offered by
-OCaml record types.
+This interpreter~|RCN| is fully polymorphic over the answer type,
+using higher-rank polymorphism through OCaml record types.
+It could also be a functor parameterized over
+the answer type.
 
@@
  let diverg () = app (lam (fun x -> int 1)) 
-                     (app (fix (fun f -> f)) (int 2))
+                  (app (fix (fun f->f)) (int 2))
 end
 \end{code}
-If we interpret |EXS| with the |R| interpreter of
-\S\ref{S:interpreter-RL}
+Interpreting |EXS| with the |R| interpreter of
+\S\ref{S:interpreter-RL} does not terminate.
 \begin{code}
@@
 \end{code}
-we never get any result because the above diverges. In contrast, if we use
-the CBN interpreter
-\begin{code}
-let module M = EXS(RCN) in RCN.get_res (M.diverg ())
-\end{code}
-we get the result |1|.
+In contrast, the CBN interpreter gives the result~|1|.
+\begin{code}
+let module M = EXS(RCN) in RCN.run (M.diverg ())
+\end{code}
 
@@
 
-We can easily change our CBN CPS interpreter into a CBV CPS
-interpreter, by replacing only one definition:
-\begin{code}
-module RCV = struct
-  include RCN
+Changing one definition turns our CBN CPS interpreter into CBV\@.
+\begin{code}
+module RCV = struct include RCN
   let lam f = {ko = fun k -> k (fun e ->
@@
 \end{code}
-Now when we apply an abstraction to an argument, we always force
-the evaluation of that argument before proceeding. This approach is in
-line with \citet{reynolds-relation} (albeit in a typed setting). The
-interpreter RCV is useful for achieving CBV evaluation of the object language
-whether the metalanguage is CBV or CBN.
-
-In this section however we show a more general approach to CBV CPS:
-a CPS transformer. We take any implementation of |Symantics| (interpreter,
-compiler, partial evaluator) and make it evaluate the CPS language
-instead. We can view interpreter transformation as a
-transformation on the object language, from direct to
-continuation-passing style.
-
-The transformation is canonical, as the following (abbreviated) code
-shows
+Now an applied abstraction
+evaluates its argument before proceeding. This approach is in
+line with \citets{reynolds-relation}, albeit typed. The
+interpreter~|RCV| is useful for CBV evaluation of the object language
+whether the metalanguage is CBV or CBN\@.
+
+We turn to a more general approach to CBV CPS: a CPS transformer that
+turns any implementation of |Symantics| into a CPS interpreter, whether
+it is an evaluator.  This functor on interpreters performs a textbook
+CPS transformation on the object language.
 \begin{code}
@@
     S.app e1 (S.lam (fun v1 ->
-    S.app e2 (S.lam (fun v2 -> S.app k (S.add v1 v2))))))
-  ...
-  let lam f = S.lam (fun k -> S.app k (S.lam (fun x ->
-    f (S.lam (fun k -> S.app k x)))))
+    S.app e2 (S.lam (fun v2 ->
+    S.app k (S.add v1 v2))))))
+  let lam f = S.lam (fun k -> S.app k (S.lam
+    (fun x -> f (S.lam (fun k -> S.app k x)))))
   let app e1 e2 = S.lam (fun k -> 
     S.app e1 (S.lam (fun f ->
-    S.app e2 (S.lam (fun v -> S.app (S.app f v) k)))))
+    S.app e2 (S.lam (fun v ->
+    S.app (S.app f v) k)))))
   let fix f = S.fix (fun self -> (f self))
@@
 \end{code}
-The code explicitly establishes a mapping from CPS interpretations to
-(direct) interpretations, as performed by 
-the base interpreter |S|.
-
-A reader may notice that the above module does not define |repr|
+The (abbreviated) code explicitly maps CPS interpretations to
+(direct) interpretations performed by 
+the base interpreter~|S|.
+
+The module returned by |CPST| does not define |repr|
 and thus does not have signature |Symantics|.
-The reason for that is again the result type of |lam f|. Whereas
+The reason is again the result type of |lam f|. Whereas
 |int i| or |add e1 e2| have the (abbreviated) 
@@
 \end{code}
-
-\noindent which shows the na\"{\i}ve CPS transformation of the original code.
+which shows the na\"{\i}ve CPS transformation of the original code.
 The result is sub-optimal with several apparent $\beta$-redexes. 
@@
 multiple ways.  The meta-notions of contexts and plugging may seem ad
-hoc, but in fact they are just reflections of the type-class and module
+hoc, but in fact they just reflect the type-class and module
 machinery that we have been using in this paper all along.
@@
 interpreter of a typed language cannot be Jones\hyp optimal.
-\begin{enumerate}
+\begin{enumerate*}
 \item One needs to encode a typed language in a typed language based on
@@
 for such an encoding of a typed language
-in a typed language, without either using a
-very advanced type system or incurring tagging/untagging overhead.
+in a typed language, without either an
+advanced type system or the overhead of tagging and untagging.
 \item Thus an indirect interpreter is necessary, which needs a universal
@@
 \item Hence the self-interpreter cannot be Jones-optimal.
-\end{enumerate}
+\end{enumerate*}
 While the logic is sound, we showed that the premise in the very first step
@@
 A lot of effort has gone into finding ways to give precise types to untyped
-terms (see for example \citet{baars-typing,Guillemette-Monier-PLPV}
-as well as the messages \citet{haskell-list}).  These 
+terms \citep{baars-typing,Guillemette-Monier-PLPV,haskell-list}).  These 
 take advantage of the host language's types to varying extents.  While
@@
 kindly provided some useful references on coalgebraic structures in the
-$\lambda$\hyp calculus
+$\lambda$\hyp calculus.
 

### 2007-04-06 tagless/tagless-final.tex
Done with Section 5

@@
 and thus does not have signature |Symantics|.
-The reason is again the result type of |lam f|. Whereas
-|int i| or |add e1 e2| have the (abbreviated) 
-type |('c,..., (int -> 'w) -> 'w) S.repr|,
-the result of |lam (fun x -> add x (int 1))| has type\\
-|('c,...((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr|. 
-Hence, to write the type equation for |CPST.repr| we again need a 
-type-discriminating
-function similar to the type function |repr_pe| of
-\S\ref{S:PE-solution}. Alas, the function we need here is
-similar but not identical to |repr_pe|. Thus we need to add 
-another type variable to the |repr| type and modify the Symantics
-signature accodringly. Again, the terms in previous Symantics module
-stay unchanged, but the |repr| type equations in those structures have to
-account for a new (phantom) type variable.
-
-To save space, we show a simplified and less principled approach that
-lets us use the module |CPST| as is. Because the module does not
-have the signature |Symantics|, we cannot apply the |EX| functor to it.
-But we can nevertheless write the tests:
+The reason is again the type of |lam f|. Whereas
+|int| and |add| return the (abbreviated) type
+|('c, ..., (int -> 'w) -> 'w) S.repr|,
+the type of |lam (add (int 1))| is
+\begin{code}
+('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2)
+          -> 'w2) S.repr
+\end{code}
+Hence, to write the type equation defining |CPST.repr| we again need
+a type function with a typecase distinction, similar to |repr_pe|
+in~\S\ref{S:PE-solution}. Alas, the type function we need is not
+identical to |repr_pe|, so we need to add another type argument to
+|repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
+terms in previous implementations of |Symantics| stay unchanged, but the
+|repr| type equations in those implementations have to take a new
+(phantom) type argument.
+
+To save space, we just
+use the module returned by |CPST| as is. Because it does not
+match the signature |Symantics|, we cannot apply the |EX| functor to it.
+Nevertheless, we can write the tests.
 \begin{code}
@@
  open M
- let test1 () = app (lam (fun x -> x)) (bool true)
- let testpowfix () = ... (* the same as before *)
- let testpowfix7 = (* the same as before *)
-    lam (fun x -> app (app (testpowfix ()) x) (int 7))
+ let test1 () = (* same as before *)
+   app (lam (fun x -> x)) (bool true)
+ let testpowfix () = ... (* same as before *)
+ let testpowfix7 = (* same as before *)
+    lam (fun x -> app (app (testpowfix ()) x)
+                      (int 7))
 end
 \end{code}
-That is, we instantiate |CPST| with the desired base intrepreter (the
-compiler |C| in the case above) so that we can use the result |M| to
-interpret object level terms. Those terms are \emph{exactly} as before.
+We instantiate |CPST| with the desired base interpreter~|C|,
+then use the result |M| to
+interpret object terms. Those terms are \emph{exactly} as before.
 Having to textually copy the terms is the
-price we pay for this simplified treatment. We shall also encounter this
-copying when discussing self-interpretation.  We
-will see then that the copying step is not frivolous, but represents
-a profound instance of 
-``plugging the hole'', which is one of the many faces of polymorphism.
-
-It is instructive to evaluate a few examples. With 
-|CPST| instantiated with the compiler~|C| above,
+price we pay for this simplified treatment.
+Our discussion of self\hyp interpretation in~\S\ref{selfinterp} shows
+that this copying is not frivolous but represents plugging a term into
+a context, which is one of the many faces of polymorphism.
+
+With 
+|CPST| instantiated by the compiler~|C| above,
 |T.test1| gives
 \begin{code}
-.<fun x_5 -> ((fun x_2 -> 
-        (x_2 (fun x_3 -> fun x_4 -> (x_4 x_3))))
-        (fun x_6 -> ((fun x_1 -> (x_1 (true))) 
-        (fun x_7 -> ((x_6 x_7) x_5)))))>.
-\end{code}
-which shows the na\"{\i}ve CPS transformation of the original code.
-The result is sub-optimal with several apparent $\beta$-redexes. 
-By instantiating |CPST| with |P| instead, we get the expected
-\begin{code}
-  P.repr = {P.st = Some <fun>;
-            P.dy = .<fun x_5 -> (x_5 (true))>.}
+.<fun x_5 ->
+  (fun x_2 -> x_2 (fun x_3 x_4 -> x_4 x_3))
+  (fun x_6 -> (fun x_1 -> x_1 true)
+              (fun x_7 -> x_6 x_7 x_5))>.
+\end{code}
+This output is the na\"{\i}ve CPS transformation of $(\fun{x}x)\True$,
+containing several apparent $\beta$-redexes.  To eliminate these
+redexes, we simply change~|T| to instantiate |CPST| with |P| instead.
+\begin{code}
+{P.st = Some <fun>;
+ P.dy = .<fun x_5 -> x_5 true>.}
 \end{code}

### 2007-04-06 tagless/tagless-final.tex
a few small changes

@@
 into advanced type systems.  While we admire all this technical work,
-there remained the issue that we were not convinced that their 
+we were not convinced that their 
 motivating example actually needed all this machinery.
@@
   type (and hence tagging),
-\item Thus any self-interpreter must have tags.
-\item Hence the self-interpreter cannot be Jones-optimal.
+\item Thus any self-interpreter must have tags and hence cannot be 
+  Jones-optimal.
 \end{enumerate*}
@@
 
-\citet{Danvy-tagging-encoding} discuss Jones optimality at length.
-There, the authors demonstrate the usefulness of HOAS for typed
-self-interpretation.  However, no claim is made that the source program
-is itself typed -- just that the interpreter is.  Our reading is that
-the source language is not typed.  However, in sharp contrast with 
-their interpreter, none of our evaluators can product pattern-match errors
-(unlike the code that accompanies \citet{Danvy-tagging-encoding}).
-In other words, their language (and interpreter) does use tags, however
-the shift to HOAS allows the partial evaluator to remove all the tags.
-In contrast, we have no tags at all.
+\citet{Danvy-tagging-encoding} discuss Jones optimality at length and
+demonstrate the usefulness of HOAS for typed self-interpretation.
+However, no claim is made that the source program is itself typed --
+just that the interpreter is.  Our reading is that the source language
+is not typed. Therefore, object term encoding has tags and the
+interpreter (given in the code accompanying
+\citet{Danvy-tagging-encoding}) extensively uses partial
+pattern-matching. The shift to HOAS, however, lets the partial
+evaluator to remove all the tags. In contrast, our object encoding and
+evaluators do not have tags to start with and have no possibility of
+pattern-match errors.
 
@@
 of back-and-forth between reify/reflect: in our PE we track both
-static and dynamic values in parallel (whenever possible).  In some sense,
+static and dynamic values in parallel.  In some sense,
 we are observing that this method type-checks in Hindley-Milner once we

### 2007-04-06 tagless/tagless-final.tex
Mostly fiddling with the title

@@
 
-\title{Finally Tagless Staged Interpreters for Simpler Typed Languages}
+\title{Finally tagless, partially evaluated}
+\subtitle{Tagless Staged Interpreters for Simpler Typed Languages}
 %Staged typed type-preserving tagless ``interpreters'', \emph{finally}
-%Finally tagless, simply typed, partially evaluated
-\subtitle{\raisebox{-2.5em}[0pt][0pt]{\normalsize\it\begin{minipage}{30em}
-    It should also be possible to define languages with a highly
-    refined syntactic type structure. Ideally, such a treatment should
-    be metacircular, in the sense that the type structure used in the
-    defined language should be adequate for the defining language.
-    \hfill\rm\citep{reynolds-definitional}
-\end{minipage}}}
 
@@
 \end{comment}
-\authorinfo{}{}{}
+\authorinfo
+{\smash{\normalsize\it\begin{minipage}{30em}
+    It should also be possible to define languages with a highly
+    refined syntactic type structure. Ideally, such a treatment should
+    be metacircular, in the sense that the type structure used in the
+    defined language should be adequate for the defining language.
+    \hfill\rm\citep{reynolds-definitional}
+\end{minipage}}}{}{}
 
@@
 \end{code}
-The (abbreviated) code explicitly maps CPS interpretations to
+This (abbreviated) code explicitly maps CPS interpretations to
 (direct) interpretations performed by 
@@
 \end{code}
-This output is the na\"{\i}ve CPS transformation of $(\fun{x}x)\True$,
+This output is a na\"{\i}ve CPS transformation of $(\fun{x}x)\True$,
 containing several apparent $\beta$-redexes.  To eliminate these
-redexes, we simply change~|T| to instantiate |CPST| with |P| instead.
+redexes, we just change~|T| to instantiate |CPST| with |P| instead.
 \begin{code}
@@
 Our initial motivation came from several papers 
-\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} which used
-embedded interpreters as a central example to justify looking 
-into advanced type systems.  While we admire all this technical work,
+\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} that used
+embedded interpreters to justify
+advanced type systems.  While we admire all this technical work,
 we were not convinced that their 

### 2007-04-06 tagless/final_code/incope.hs
copies of the code to attach to paper

@@ -1,0 +1,951 @@
+{-# OPTIONS -fglasgow-exts -W #-}
+
+-- Interpreter, Compiler, Partial Evaluator
+
+module Incope where
+
+{-
+  The language is simply-typed lambda-calculus with fixpoint,
+  integers, booleans and comparison.
+
+  Lam hoas_fn | App e e | Fix hoas_fn |
+  I Int | B Bool | Add ie1 ie2 | Mul ie1 ie2 | Leq ie1 ie2 |
+  IF b e-then e-else
+  
+  The language is just expressive enough for the Gibonacci function
+  and the power function.
+
+  The compiler, the interpreter and the source and target languages
+  are *all* typed. The interpreter and the compiler use no tags.
+  There is no pattern-match failure possible: the evaluators never
+  get stuck.
+
+  The implementation below is *almost* possible in OCaml. Indeed,
+  one can emulate Haskell typeclasses using explicit dictionary passing.
+  The dictionary will by a polymorphic (rank-2) record -- but it is OK,
+  OCaml permits it. Alas, The typeclass below uses higher-order type
+  constructors: repr is of a kind * -> *. OCaml does not support
+  higher-order polymorphism.
+
+-}
+
+-- This class defines syntax (and its instances, semantics) of our language
+-- This class is Haskell98!
+class Symantics repr where
+    int :: Int -> repr Int                -- int literal
+    bool :: Bool -> repr Bool             -- bool literal
+
+    lam :: (repr a -> repr b) -> repr (a->b)
+    app :: repr (a->b) -> repr a -> repr b
+    fix :: (repr a -> repr a) -> repr a
+
+    add :: repr Int -> repr Int -> repr Int
+    mul :: repr Int -> repr Int -> repr Int
+    if_ :: repr Bool -> repr a -> repr a -> repr a
+    leq :: repr Int -> repr Int -> repr Bool
+
+-- The following `projection' function is specific to repr.
+-- It is like `run' of the monad
+--    comp :: repr a -> something a
+
+test1 () = add (int 1) (int 2)
+test2 () = lam (\x -> add x x)
+test3 () = lam (\x -> add (app x (int 1)) (int 2))
+
+testgib () = lam (\x -> lam (\y ->
+                  fix (\self -> lam (\n ->
[…]

### 2007-04-06 tagless/final_code/incope1.hs

@@ -1,0 +1,269 @@
+{-# OPTIONS -fglasgow-exts #-}
+{-# OPTIONS -fallow-undecidable-instances #-}
+
+-- Interpreter, Compiler, Partial Evaluator
+
+module Incope1 where
+
+{-
+  The language is simply-typed lambda-calculus with fixpoint,
+  integers and comparison.
+
+  Lam hoas_fn | App e e | Fix hoas_fn |
+  I Int | Add ie1 ie2 |
+  IFEQ ie1 ie2 e-then e-else
+  
+  The language is just expressive enough for the Gibonacci function.
+
+  The compiler, the interpreter and the source and target languages
+  are *all* typed. The interpreter and the compiler use no tags.
+  There is no pattern-match failure possible: the evaluators never
+  get stuck.
+
+  The implementation below is *almost* possible in OCaml. Indeed,
+  one can emulate Haskell typeclasses using explicit dictionary passing.
+  The dictionary will by a polymorphic (rank-2) record -- but it is OK,
+  OCaml permits it. Alas, The typeclass below uses higher-order type
+  constructors: repr and res are of a kind * -> *. OCaml does not support
+  higher-order polymorphism.
+
+-}
+
+-- This class defines syntax (and its instances, semantics) of our language
+
+class SymanticsInt ar ci | ar -> ci  where
+    int :: Int -> ar ci Int                -- int literal
+    add :: ar ci Int -> ar ci Int -> ar ci Int
+
+
+class SymanticsA ar a ci ca | ar -> ci, ar a -> ci ca  where
+    ifeq:: ar ci Int -> ar ci Int -> ar ca a -> ar ca a -> ar ca a
+    fix :: (ar ca a -> ar ca a) -> ar ca a
+
+class SymanticsAB ar a b ca cb cab | ar a->ca, ar b->cb, ar a b -> ca cb cab  where
+    lam :: (ar ca a -> ar cb b) -> ar cab (a->b)
+    app :: ar cab (a->b) -> ar ca a -> ar cb b
+
+
+-- The following `projection' function is specific to repr.
+-- It is like `run' of the monad
+--    comp :: repr a -> something a
+
+test1 () = add (int 1) (int 2)
+test2 () = lam (\x -> add x x)
+test3 () = lam (\x -> add (app x (int 1)) (int 2))
+
+testgib () = lam (\x -> lam (\y ->
[…]

### 2007-04-06 tagless/final_code/incope.hs
prelim cleanup of the code

@@ -73,8 +73,6 @@ testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
 -- It is a typed, tagless interpreter: R is not a tag. The interpreter
 -- never gets stuck, because it evaluates typed terms only
 
--- Note that everything going on in the interpreter is straight out of
--- "Boxes go Bananas" by Washburn and Weirich (intended or otherwise)
 newtype R a = R a deriving Show
 unR (R x) = x
 
@@ -161,23 +159,18 @@ ltestpw   = compL . testpowfix $ ()
 ltestpw7  = compL . testpowfix7 $ ()
 ltestpw72 = compL (app (testpowfix7 ()) (int 2))
 
-
-
 -- ------------------------------------------------------------------------
 -- The compiler
 -- We compile to GADT, to be understood as a typed assembly language
--- (typed bytecode). The GADT does _not_ use the higher-order abstract
+-- (typed bytecode). The GADT does _not_ use higher-order abstract
 -- syntax. We could have used template Haskell. Alas, its expressions
 -- are untyped.
 -- Note how ByteCode represents MetaOCaml's `code'. Thus the compiler
 -- below neatly maps to the MetaOCaml (with no GADTs).
 -- Also note, that like MetaOCaml `code', we never pattern-match on
 -- ByteCode!
--- Note how the compiler never raises any exception and matches no tags
+-- Furthermore the compiler never raises any exception and matches no tags
 -- (no generated code has any tags)
-
--- The LIFT bytecode operation is used only during evaluation and for fmap
--- it corresponds to a CSP in MetaOCaml.
 
 data ByteCode t where
     Var :: Int -> ByteCode t                -- variables identified by numbers
@@ -190,7 +183,6 @@ data ByteCode t where
     Mul :: ByteCode Int -> ByteCode Int -> ByteCode Int
     Leq :: ByteCode t1 -> ByteCode t1 -> ByteCode Bool
     IF  :: ByteCode Bool -> ByteCode t -> ByteCode t -> ByteCode t
-    LIFT :: t -> ByteCode t                 -- Used only for eval and fmap
 
 instance Show (ByteCode t) where
     show (Var n) = "V" ++ show n
@@ -236,6 +228,7 @@ newtype C t = C (Int -> (ByteCode t, Int))
[…]

### 2007-04-06 tagless/final_code/incope1.hs

@@ -13,20 +13,9 @@ module Incope1 where
   I Int | Add ie1 ie2 |
   IFEQ ie1 ie2 e-then e-else
   
-  The language is just expressive enough for the Gibonacci function.
+  The language is just expressive enough for the Fibonacci function.
 
-  The compiler, the interpreter and the source and target languages
-  are *all* typed. The interpreter and the compiler use no tags.
-  There is no pattern-match failure possible: the evaluators never
-  get stuck.
-
-  The implementation below is *almost* possible in OCaml. Indeed,
-  one can emulate Haskell typeclasses using explicit dictionary passing.
-  The dictionary will by a polymorphic (rank-2) record -- but it is OK,
-  OCaml permits it. Alas, The typeclass below uses higher-order type
-  constructors: repr and res are of a kind * -> *. OCaml does not support
-  higher-order polymorphism.
-
+  This is the typeclass version of the code in incope.hs
 -}
 
 -- This class defines syntax (and its instances, semantics) of our language
@@ -43,11 +32,6 @@ class SymanticsA ar a ci ca | ar -> ci, ar a -> ci ca  where
 class SymanticsAB ar a b ca cb cab | ar a->ca, ar b->cb, ar a b -> ca cb cab  where
     lam :: (ar ca a -> ar cb b) -> ar cab (a->b)
     app :: ar cab (a->b) -> ar ca a -> ar cb b
-
-
--- The following `projection' function is specific to repr.
--- It is like `run' of the monad
---    comp :: repr a -> something a
 
 test1 () = add (int 1) (int 2)
 test2 () = lam (\x -> add x x)
@@ -68,8 +52,6 @@ testgib1 () = app (app (app (testgib ()) (int 1)) (int 1)) (int 5)
 -- It is typed, tagless interpreter: R is not a tag. The interpreter
 -- never gets stuck, because it evaluates typed terms only
 
--- Note that everything going on here is straight out of
--- "Boxes go Bananas" by Washburn and Weirich (intended or otherwise)
 newtype R t a = R a deriving Show
 unR (R x) = x
 
@@ -113,6 +95,7 @@ data ByteCode t where
     Add :: ByteCode Int -> ByteCode Int -> ByteCode Int
[…]

### 2007-04-06 tagless/tagless-final.tex
Related work done

@@
+\listfiles
 \hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar type-case}
@@
 \title{Finally tagless, partially evaluated}
-\subtitle{Tagless Staged Interpreters for Simpler Typed Languages}
+\subtitle{Tagless staged interpreters for simpler typed languages}
 %Staged typed type-preserving tagless ``interpreters'', \emph{finally}
@@
 \subsection{Contexts clarify polymorphism}
+\label{S:clarify}
 
@@
 
-\citet{WalidICFP02}, when referring to the work of \citet{taha-tag},
-claims that this shows that self-interpretation is possible in a
-\emph{simply typed $\lambda$-calculus}, but this is somewhat misleading
-as their type system does not differentiate between the various ground
-types.  In other words, various constants (like \textsf{nil} and 
-\textsf{true}) both inhabit the same type $D$.  This is not very satisfying.
-Furthermore, as is pointed out in the introduction to \citet{WalidICFP02},
-\citet{taha-tag} does not \emph{statically} guarantee that all tags will
-be removed - tag elimination must be performed at runtime.
+Even a simply typed $\lambda$-calculus obviously supports self\hyp
+interpretation, provided we use universal types \citep{taha-tag}.  The
+ensuing tagging overhead motivated \citet{taha-tag} to propose tag
+elimination, which however does not statically guarantee that all tags
+will be removed \citep{WalidICFP02}.
 
@@
 \citet{peyton-jones-simple} seem to argue as follows that a self\hyp
-interpreter of a typed language cannot be Jones\hyp optimal.
+interpreter of a typed language cannot be tagless or Jones\hyp optimal.
 \begin{enumerate*}
 \item One needs to encode a typed language in a typed language based on
-a sum type (at some level of the hierarchy),
+a sum type (at some level of the hierarchy).
 \item It is not possible to write a \emph{direct} interpreter 
@@
 \item Thus an indirect interpreter is necessary, which needs a universal
-  type (and hence tagging),
-\item Thus any self-interpreter must have tags and hence cannot be 
+  type (and hence tagging).
+\item Thus any self-interpreter must have tags and cannot be 
   Jones-optimal.
@@
 \citet{Danvy-tagging-encoding} discuss Jones optimality at length and
-demonstrate the usefulness of HOAS for typed self-interpretation.
-However, no claim is made that the source program is itself typed --
-just that the interpreter is.  Our reading is that the source language
-is not typed. Therefore, object term encoding has tags and the
-interpreter (given in the code accompanying
-\citet{Danvy-tagging-encoding}) extensively uses partial
-pattern-matching. The shift to HOAS, however, lets the partial
-evaluator to remove all the tags. In contrast, our object encoding and
-evaluators do not have tags to start with and have no possibility of
-pattern-match errors.
-
-We used a method from \citet{asai-binding-time} to reduce the amount
-of back-and-forth between reify/reflect: in our PE we track both
-static and dynamic values in parallel.  In some sense,
-we are observing that this method type-checks in Hindley-Milner once we
-deforest the static code representation.  Put another way, we have
-manually applied type-level partial-evaluation to our typelevel 
+apply HOAS to typed self\hyp interpretation.  However, their source
+language is untyped.  Therefore, their object\hyp term encoding has
+tags, and their interpreter can raise run-time errors.
+Nevertheless, HOAS lets the partial
+evaluator remove all the tags. In contrast, our object encoding and
+interpreters do not have tags to start with and obviously cannot
+raise run-time errors.
+
+Our final approach seems related to \citearound{'s approach to HOAS via
+catamorphism and anamorphism}\citet{Washburn-Weirich-boxes}.
+
+A lot of effort has gone into ``typing dynamic typing'': to statically
+type-check dynamically\hyp typed values
+\citep{baars-typing,WalidICFP02,Guillemette-Monier-PLPV,haskell-list},
+using the host language's type system to varying extents.
+Our object terms are statically typed, so we would
+need one of these techniques to interpret dynamically\hyp typed
+terms such as those read from a file.
+
+Our partial evaluator establishes a bijection |repr_pe| between static
+and dynamic types (the valid values of |'sv| and |'dv|), and between
+static and dynamic terms.  It is customary to implement such a bijection
+using an injection\hyp projection pair, as done for interpreters by
+\citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters},
+partial evaluation by \citet{Danvy-TDPE}, and type-level functions by
+\citet{oliveira-typecase}.  As explained in~\S\ref{S:PE-solution}, we
+avoid injection and projection at the type level by adding an argument
+to |repr|.
+
+At the term level, we also avoid converting between static and dynamic
+terms by building them in parallel, using \citearound{'s
+method}\citet{asai-binding-time}.
+This method type-checks in Hindley-Milner once we
+deforest the object term representation.  Put another way, we
+manual apply type-level partial evaluation to our type
 functions (see \S\ref{S:PE-solution}) to obtain simpler types 
-acceptable to MetaOCaml.  It is worthwhile to note that 
-\citet{sumii-hybrid} also takes advantage of this method
-\citep[of][]{asai-binding-time} to get interesting results combining online
+acceptable to MetaOCaml.
+\Citet{sumii-hybrid} also use Asai's method,
+to combine online
 and offline partial evaluation.  Like \citet{SperberThiemann:TwoForOne},
-we strive for modularity by reusing earlier stages.  It would be interesting
-to see if we could derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
+we strive for modularity by reusing interpreters for individual stages.  It would be interesting
+to try to derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
 in the same manner.
 
-The approach to HOAS via catamorphism and anamorphism of
-\citet{Washburn-Weirich-boxes} seems related to what we are doing.
-We have not investigated the relationship to any serious depth, but we
-feel that one could fruitfully combine their ideas with ours.
-
-A lot of effort has gone into finding ways to give precise types to untyped
-terms \citep{baars-typing,Guillemette-Monier-PLPV,haskell-list}).  These 
-take advantage of the host language's types to varying extents.  While
-very successful for proper terms, results for
-ill-typed terms can produce rather obscure error messages.  We use the
-host's typing facilities for this, and thus inherit the strengths
-and weaknesses of the host language.  While that may not be optimal, it is
-certainly much more modular.
-
-When dealing with datatypes which are equivalent but have different
-representations, it is customary to use injection-projection pairs
-(used in the context of interpreters by \citet{Ramsey-ML-module-mania}
-and \citet{Benton-embedded-interpreters}, partial evaluation by
-\citet{Danvy-TDPE}, and type-level functions by
-\citet{oliveira-typecase}).  Na\"{\i}vely, this would appear to apply to
-our implementation of the type-level function |'sv -> 'dv|.  However, in most
-of these cases, it is important that this establishes a bijection (between
-|'sv| and |'dv| in our case).  Our approach is one-sided: we requite
-an operational definition of |'sv -> 'dv| (which is easy), and an
-implementation of |'dv| which ``perfectly reflects'' |'sv| but without 
-requiring an implementation.
-
 We are not the first to implement a typed interpreter for a typed
-language -- \citet{laod93} use a typed version of the SK language
-(and typeclasses) to implement a meta-circular interpreter (rather than
-a self-interpreter).  Interestingly, their meta-circular interpreter
-appears to be tagfree!  However, they could not have implemented a
-compiler or a partial-evaluator in the same way since they rely quite
-heavily on having injection-projection pairs.
-
-\citet{fiore:nbe-ppdp2002,balat:tdpe-popl2004} also deal with a 
-tagfree partial-evaluator using quite a lot of interesting 
-technology (amongst them powerful continuation-based control operators).
-While they use a reify-reflect injection-projection pair, this appears
-to be done ``lightly'', in the sense that the structure of compiled code
-is never examined.  The exact relation between their work and ours is 
-still a little mysterious to us, but we will continue to investigate it.
-
-We have been unable to find work that establishes that
+language.  \Citet{laod93} use type classes to implement a metacircular
+interpreter (rather than a self\hyp interpreter) of a
+typed version of the SK language, which is quite different from our
+object language.  Their interpreter
+appears to be tagless, but they could not have implemented a
+compiler or partial evaluator in the same way, since they rely
+heavily on injection\hyp projection pairs.
+
+\Citet{fiore:nbe-ppdp2002} and \citet{balat:tdpe-popl2004} also build
+a tagless partial evaluator, using delimited control operators.  It is
+type-directed, so the user must represent, as a term, the type of every
+term to be partially evaluated.  We shift this work to the type checker
+of the metalanguage.  By avoiding term-level type representations, our
+approach makes it easier to perform algebraic simplifications (as
+in~\S\ref{S:PE-solution}).
+
+We could not find work that establishes that
 the \emph{typed} $\lambda$-calculus possesses a final coalgebra structure.
-For the untyped $\lambda$-calculus, 
-\citet{HonsellLenisa,honsell99coinductive} establish several interesting
-results along this line.  Of particular interest is their use of
-contexts with a hole \citep[p.13]{honsell99coinductive} for the definition
-of \emph{observational equivalence}
-(see \S\ref{selfinterp}).  The reader is directed to bibliography of
-\citet{honsell99coinductive} for references to the foundational work
-in this important area.  Particularly intriguing is the link to the
+\Citet{HonsellLenisa,honsell99coinductive}
+investigate the untyped $\lambda$-calculus,
+along this line.  In particular, they use
+contexts with a hole \citep[p.\,13]{honsell99coinductive} to define
+\emph{observational equivalence}
+(see \S\ref{selfinterp}).
+\citearound{'s bibliography}\Citet{honsell99coinductive} refers to the
+foundational work in this important area.  Particularly intriguing is the link to the
 coinductive aspects of B\"{o}hm trees, as pointed out by
@@
 
-One way to interpret some of our results is that we eschew sums (at either the
-term or type, or even kind level) for our terms, but instead use a record type 
-(via Functors in OCaml and typeclasses in Haskell).
-For the self-interpreter, we then proceed to use a Church encoding for
+One way to understand our main idea is to eschew sum types and sum kinds
+for their dual, record types and record kinds.
+For the self\hyp interpreter, we then proceed to use a Church encoding for
 recursive data types \citep{bohm-automatic}.
+
+As \S\ref{S:clarify} observes, higher-rank and higher-kind
+polymorphism lets us type-check and compile object terms separately from
+interpreters.  This observation is consistent with the role of
+polymorphism in the separate compilation of modules
+\citep{shao-typed,shao-transparent}.
 

### 2007-04-06 tagless/final_code/incope.hs
small changes to the comments

@@ -19,14 +19,6 @@ module Incope where
   are *all* typed. The interpreter and the compiler use no tags.
   There is no pattern-match failure possible: the evaluators never
   get stuck.
-
-  The implementation below is *almost* possible in OCaml. Indeed,
-  one can emulate Haskell typeclasses using explicit dictionary passing.
-  The dictionary will by a polymorphic (rank-2) record -- but it is OK,
-  OCaml permits it. Alas, The typeclass below uses higher-order type
-  constructors: repr is of a kind * -> *. OCaml does not support
-  higher-order polymorphism.
-
 -}
 
 -- This class defines syntax (and its instances, semantics) of our language
@@ -197,30 +189,6 @@ instance Show (ByteCode t) where
     show (IF be et ee)
         = "(if " ++ show be ++ 
           " then " ++ show et ++ " else " ++ show ee ++ ")"
-
--- An evaluator for the ByteCode: the virtual machine
--- The evaluator is partial: if we attempt to evaluate an open code
--- (a variable not found in the env), we will get an error.
--- We use Dynamic to `assure' that variables are properly typed,
--- that is, that the variable name (counter) witnesses its type.
--- A better assurance is to use STRef. Strange correspondence:
--- mutation and lambda application. The essence of both is sharing.
-{-
-type BEnv = [(Int,Dynamic)]
-
-eval :: Typeable t => BEnv -> ByteCode t -> t
-eval env (Var n) | Just dv <- lookup n env,
-		   Just v  <- fromDynamic dv = v
-eval env v@Var{} = error $ "Open code? variable not found: " ++ show v
-eval env (Lam v b) = \x -> eval ((v,toDyn x):env) b
-eval env (App e1 e2) = (eval env e1) (eval env e2)
--- eval1 (Fix n b) e = 
-eval env (INT n) = n
-eval env (BOOL n) = n
-eval env (Add e1 e2) = eval env e1 + eval env e2
--- eval1 (IF be et ee) e = 
--}
-
 
 -- Int is the variable counter
 -- for allocation of fresh variables
@@ -400,7 +368,7 @@ ptestpw7  = compP . testpowfix7 $ ()
 ptestpw72 = compP  (app (testpowfix7 ()) (int 2))
 
 
-{- GADTs, although work, are still not quite satsifactory:
+{- GADTs, although work, are still not quite satisfactory:
 let us look at the following lines in incope.hs
 
     app ef ea = case ef of

### 2007-04-06 tagless/final_code/incope1.hs

@@ -156,8 +156,6 @@ ctestg = compC . testgib1 $ ()
 -- The code below is NOT parametric. We could have used type-classes
 -- instead of GADTs. The code below could be re-functionalized,
 -- and so could in fact be translated into MetaOCaml.
--- Or, if we observe the correspondence between GADT and MetaOCaml <code>,
--- perhaps we can translate to MetaOCaml with an additional layer of code?
 
 data P cr t = PV cr | PE (C () t)
 

### 2007-04-06 tagless/tagless-final.tex
Final changes before (re)submission

@@
 
-\newcommand{\jacques}[1]{{\it [Jacques says: #1]}}
-\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
-\newcommand{\ccshan}[1]{{\it [Ken says: #1]}}
+% Set pdftex paper size
+\ifx\pdfoutput\undefined\else \ifx\pdfoutput\relax\else \ifcase\pdfoutput\else
+    \setlength{\pdfpagewidth}{8.5in}
+    \setlength{\pdfpageheight}{11in}
+\fi \fi \fi
+
+%\newcommand{\jacques}[1]{{\it [Jacques says: #1]}}
+%\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
+%\newcommand{\ccshan}[1]{{\it [Ken says: #1]}}
 
@@
 
-\title{Finally tagless, partially evaluated}
-\subtitle{Tagless staged interpreters for simpler typed languages}
+\title{Finally Tagless, Partially Evaluated}
+\subtitle{Tagless Staged Interpreters for Simpler Typed Languages}
 %Staged typed type-preserving tagless ``interpreters'', \emph{finally}
@@
 \authorinfo
-{\smash{\normalsize\it\begin{minipage}{30em}
+{\hspace*{-15em}\smash{\normalsize\it\begin{minipage}[t]{30em}
     It should also be possible to define languages with a highly
@@
     \hfill\rm\citep{reynolds-definitional}
-\end{minipage}}}{}{}
+\end{minipage}}\hspace*{-15em}}{}{}
 
@@
 Haskell to infer the object types for us. We could also
-avoid the dummy argument by switching off the monomophism restriction
+avoid the dummy argument by switching off the monomorphism restriction
 with a compiler flag.
@@
 \end{code}
-The dummy argument to |test1| and |testpowfix| is an artefact of
+The dummy argument to |test1| and |testpowfix| is an artifact of
 MetaOCaml, related to monomorphism: in order for us to run a
@@
 \end{code}
-Charateristic of CBN, |app e1 e2|
+Characteristic of CBN, |app e1 e2|
 does not evaluate the argument~|e2| by applying it to the

### 2007-05-15 SCP2006/paper/ge.tex
rename SCP to be more precise

\documentclass[11pt]{elsart}

\usepackage{amssymb}
\usepackage{fancyvrb}
\usepackage{times}
\usepackage{stmaryrd}

\usepackage[backref,colorlinks,bookmarks=true]{hyperref}

\begin{document}
\begin{frontmatter}
\title{Gaussian Elimination: a case study in efficient genericity with
MetaOCaml}
% --- Author Metadata here ---
\author{Jacques Carette\thanksref{nserc}}
\thanks[nserc]{Work supported in part by NSERC Discovery Grant XXX}
\ead{carette@mcmaster.ca}
\ead[url]{http://www.cas.mcmaster.ca/\textasciitilde carette}

% --- End of Author Metadata ---

\begin{abstract}
The Gaussian Elimination algorithm is in fact an algorithm family -
common implementations contain at least 6 (mostly independent)
``design choices''.  A generic implementation can easily be parametrized
by all these design choices, but this usually leads to slow and bloated
code.  Using MetaOCaml's staging facilities, we show how we can
produce a natural and type-safe implementation of Gaussian Elimination 
which exposes its design choices at code-generation time, so that these choices
can effectively be specialized away, and where the resulting code is 
quite efficient.
\end{abstract}

\begin{keyword}
MetaOcaml, linear algebra, genericity, generative, staging
\PACS code
\end{keyword}
\end{frontmatter}

\section{Introduction}

Gaussian Elimination is a fundamental algorithm in computational 
mathematics.  As it finds uses in many areas of mathematics, quite a
number of variants on the basic algorithm have been invented over
the years, but the basic algorithm persists through all variants.  
An analysis of the source code \footnote{Note to reviewers: for 
completeness, this paper includes source code samples in Maple
and the complete MetaOCaml implementation in appendices.  The
code which is deemed inessential for a final print version will
instead be made available on the Web}
for the library of Maple\footnote{Maple
is a registered trademark of Waterloo Maple Inc.}, a large
commercial computer algebra system, found 35 different implementations
of Gaussian Elimination (GE), and at least as many more implementations of
highly related algorithms (LU factorization, Cholesky factorization,
reduction to Smith Normal Form, linear equation solving, etc).  There
was little actual code duplication between these implementations, as they all
embodied a certain set of design decisions that were hard-coded.  As
these design decisions have a measurable effect on the actual shape of the
code, there was little obvious opportunity for abstraction.  However,
since the authors of the code knew that the underlying algorithm was
all the same, it would have been possible to use higher-order functions
to parametrize these choices.  On the other hand,
a number of these variants are rather esoteric, or depend on quite subtle
properties of the underlying domain at hand - certainly an average user
who knew they needed to use Gaussian Elimination could not be expected to
understand all these choices, and furnish the appropriate functions.
Thus it still necessary to expose this algorithm via different 
specialized interfaces.  But even when
there were parametrization opportunities, the ensuing loss of efficiency 
between the generic code versus the specialized code (up to a factor of 30,
in part due to the fact that Maple is a dynamically typed 
interpreted language, but even in a statically typed language, 
the overhead of an (indirect) function call, a likely pipeline stall
and cache miss(es) would produce similar overhead) 
naturally
reinforced the idea that all these versions need to be coded seperately.
This clearly leads to a maintenance nightmare.
Thus we seek to answer the question:
\begin{center}
\emph{Is it possible to write a generic yet efficient Gaussian Elimination
algorithm?}
\end{center}

\noindent In Maple itself, this question can be answered positively:
Maple has very good reflection and reification capabilities
(see~\cite{JoGoSe93} for definitions), with
full introspection of all programs being quite easy; in fact, these
capabilities compare favorably to those of LISP/Scheme~\cite{Abelson:1998:RRA}.
Maple's reflection/reification are somewhat more type-safe than 
Scheme's; like in Scheme it is not possible to reify procedures 
which are syntactic nonsense, but also some mild type invariants are
also guaranteed to hold, although the result may still be obvious
semantic nonsense.  We are more interested in the following more precise
question:

\begin{center}
\emph{Is it possible to write a generic, type-safe yet efficient 
Gaussian Elimination algorithm?}
\end{center}

To ensure type safety, it is necessary to move to a typed language; to
allow for efficiency, it is necessary to remove all the overhead
introduced when making the algorithm generic.  There are 3 obvious choices
of languages to try~\cite{conf/dagstuhl/CzarneckiOST03}: MetaOcaml, 
Template Haskell, and
C++.  Because we wanted our Gaussian Elimination algorithm to ``look like''
the original algorithms in Maple as much as possible while at the same time
enforced as much type-safety as is currently possible, the choice of 
MetaOCaml was quite natural.

\subsection{Algorithm families}

The concept of a \emph{program family}~\cite{journals/tse/Parnas76} is
now an integral
part of software engineering, at least at the design level.  This involves
abstracting all the common parts of a family of (related) programs, and
designing both the common parts and the variation points right at the 
outset.  Knowing the variation points makes a modularization based on
the \emph{design for change}~\cite{journals/cacm/Parnas72a} principle 
that much easier.

This concept of program family can apply to very small programs -- even
programs made up of just one function, as in our case.  This principle
tells us that we should isolate each \emph{aspect}\footnote{as in
separation of concerns, terminology coined by 
Dijkstra~\cite{EWD:EWD447} published as~\cite{EWD:EWD447pub},
the ``other side of the coin'' of modularization according to Parnas.} 
into a separate entity
(a \emph{module} in the case of program families), and then to somehow
paste these aspects together to form a whole.

This is exactly the route we will follow.  However, instead of presenting
the results as a {\it fait accompli}, we will in part show how to 
perform stepwise abstractions (i.e. the inverse steps of stepwise 
refinement~\cite{journals/CACM/Wirth71})
to go from a particular implementation of Gaussian Elimination to a 
generic one.  We believe this to be more instructive.  This presentation
method allows us
to show that, at every step, the resulting source code is quite readable, 
is still
recognizable as Gaussian Elimination, and yet still specializes to the 
original version.

\subsection{Generic and generative programming}

While one can find many different reasons for generic programming
\cite{musser89generic,musser94algorithmoriented,conf/ac/2003gp}, there is one 
fundamentally pragmatic reason:
all decent programmers intensely dislike code duplication.  Code
duplication is sometimes tolerated for two reasons: a poor programming
language which does not have proper abstraction mechanisms to unify
certains language idioms, and efficiency.  These issues are not 
independent: many
abstraction mechanisms (higher-order functions, late-binding, 
ad hoc polymorphism, type classes, etc)
generally incur a run-time cost.  In the case of scientific computation
applications, this run-time cost is frequently unacceptable, at least
in the inner-loop of intensive computations.

Macros and pre-processors were common pseudo-solutions for both of these issues
(and are still heavily used -- many models in physical oceanography are written
in \texttt{Fortran 77} with \texttt{cpp} directives peppered throughout for
configuration choices! \cite{SPOM}).  More recently, generative programming
\cite{Czarnecki} has become an increasingly popular method in scientific
computation - as witnessed by (amongst others) 
Blitz++ \cite{Veldhuizen:1998:ISCOPE}, 
Active Libraries \cite{Veldhuizen:1998:OO98},
GMCL \cite{Czarnecki},
POOMA \cite{POOMA},
BOOST \cite{BOOST},
and ATLAS \cite{ATLAS}.  These have also been applied to product families
\cite{GenerativeProductLines}.  These solutions are mostly
\verb=C++=-centric, which brings problems of its own
\cite{conf/dagstuhl/CzarneckiOST03}.

In a typeful setting, what we want to do is to \emph{expose} the 
configuration parameters of our programs in such a way that the type system
will help rather than hinder, in other words, the type system will 
insure that we are configuring our programs consistently.  In particular, 
we are not interested in being \emph{too} generic: being able to 
use any kind of connector to connect arbitrary components together is
pointless, and will merely result in delaying problems to (at worst)
run-time.

For example, if we have a stateful code type \texttt{code}, the type
of the sequential composition operator of statements \texttt{;} is not
\begin{Verbatim}[commandchars=\\\{\}, 
                 codes={\catcode`$=3\catcode`^=7}]
(;) : code $\times$ code -> code
\end{Verbatim}
but rather
\begin{Verbatim}[commandchars=\\\{\}, 
                 codes={\catcode`$=3\catcode`^=7}]
(;) : (state -> unit) code $\times$ (state -> unit) code 
      -> (state -> unit) code
\end{Verbatim}
where \texttt{state} is a specific representation for a (closed) state type.
This ensures that only statements operating on the same state can be
composed, thereby increasing robustness and safety.

\subsection{Problem}

The Gaussian Elimination algorithm has many different incarnations
(see section~\ref{GEintro}),
which are frequently coded seperately (for efficiency or sometimes
source code clarity reasons).  This is clearly unsatisfactory, as well as
potentially leading to a maintenance nightmare.  On the other hand, 
even when it is possible to use a language's abstraction facility to 
natively express all the design points of GE, this usually has too
high an efficiency cost.

We want to be able to express a generic version of GE which abstracts
out, in the context of symbolic computation, the most common design
points: the domain of the matrix elements ($\mathbb Z$, 
$\mathbb Q$, $\mathbb Z_p$, $\mathbb Q(\alpha)\left[x\right]$, or even
$\mathbb Z_p(\alpha_1,\ldots,\alpha_n)\left[\left[ x_1,\ldots,x_m\right]\right]$
and floating point numbers), the container type used 
(arrays of arrays, flattened arrays, implicit representations, as well 
as variants for storage orientation) for the matrix, the exact 
outputs (matrix, rank, determinant), as well
as some domain-dependent aspects, like whether the elements of the domain
are stored in a normalized manner, or whether there exists a natural (and 
useful) size function on the elements of the domain.  Not only should
this be done in a type-safe manner, the resulting code should also be
recognizable as being Gaussian Elimination.  As well, it should be clear
that the technique could scale to the complete family of algorithms
related to GE, like LU and Cholesky factorization, back-elimination, as
well as more general matrix solving algorithms.

We want our solution to still contain a recognizable Gaussian Elimination
algorithm.  While it is possible to write GE functionally, an imperative
solution is what is generally expected.  The best compromise would be to write
this in something like Haskell's State Monad which, via the do notation,
would \emph{look} right, yet still be functional.  As we did not want to 
add additional complexity by creating some syntactic sugar (via campl4), 
we bit the bullet and dove in headlong into an imperative solution.

\subsection{Contributions}

We show that a number of techniques, 
namely combinators, abstract interpretation
\cite{conf/popl/CousotC77}, state-passing style, and 
dictionary-passing at staging
time can be successfuly combined to achieve our goals of 
writing a generic, efficient, type-safe implementation of the
Gaussian Elimination family of algorithms.  These techniques have
to be very carefully combined, so that whenever a new design point of 
Gaussian Elimination is exposed, 
the generated (source) code does not include new unecessary artifacts, but 
in fact stays
constant when this design point is specialized back to its ``default'' value.

The principal features and contributions of this article are
\begin{itemize}
    \item a thorough analysis of a real situation,
    \item the use of staging for code configuration and clone management,
    \item the analysis of imperative-style code instead of 
		functional, as is more customary,
    \item cleanly documenting and exposing all design-time choices at 
        staging time,
    \item the use of state-passing style (for staging),
    \item and the use of dictionary passing at staging-time.
\end{itemize}

\subsection{Organization}

In section~\ref{GEintro}, we quickly review the Gaussian Elimination algorithm
and examine 3 different variants, to get an idea of the design points
involved.  We then proceed with a more thorough analysis of the design
points that we will tackle in this paper, followed by an analysis of the
relations between these design points.  Section~\ref{staging} quickly
introduces staging, as applied to an abstract Gaussian Elimination 
implementation.  We carefully show how turning this algorithm into a 
sequence of code generators can be done in such a way as to not introduce
any staging artifacts.  We then show how to insert a particular design
choice in to the staged version, and preserve this ``no artifacts''
property.  Section~\ref{techniques} then introduces the various techniques
used in the complete staging of GE.  Only short examples are provided
to illustrate the concepts, but the complete source code is included in an 
appendix.  Related work is reviewed in section~\ref{related}, and 
conclusions are drawn in section~\ref{concl}.  There are 3 appendices, 
one for a fraction-free implementation of GE (in Maple), one for a
feature-rich implementation of GE over algebraic extensions of finite
fields (in Maple), and finally a complete listing of our final implementation
of a staged GE in metaocaml.

\section{Gaussian Elimination}\label{GEintro}

After a quick review of the Gaussian Elimination algorithm, this section
reviews some examples of GE implementations from Maple's library, to get 
an idea of some of the design points involved.  This is followed by a
more thorough analysis of the design points that were encountered, and
which of those will be explicitly treated in this paper.  We draw particular
attention to those design points which are not fully orthogonal to other
design points, as these need special handling to avoid producing
inconsistent or meaningless code.

\subsection{Review of GE algorithm}

Gaussian Elimination is generally understood to be an algorithm which reduces
(square) Matrices to upper-triangular form via a sequence of elementary Matrix
operations, namely row/column interchanges and addition of a row multiple to
another row.  The basic requirements on the the Matrix elements are that they
can be added, multiplied, and zero elements can be recognized.  Many versions
also require that division for non-zero Matrix elements also exists.  The
requirements on the Matrix are that all elements are accessible, and it 
is simple to do elementary row operations (swapping and adding a multiple).

Denote the elements of Matrix $A$ by $a_{ij}$ as usual.
This algorithm proceeds as follows: in the first column, a non-zero element
is chosen, say in row $k$.  If this element is not in the first row, 
exchange row $k$ and the first row.  For every row $j$ below the first one,
add $-\frac{a_{j1}}{a_{11}}$ times the first row to row $j$; this ensures 
that in the resulting row, $a_{j1}=0$.  This is then repeated recursively
on the submatrix obtained by deleting the first row and column.  The
algorithm can be modified to deal with columns with all zeroes, as well as
dealing with non-square matrices.  Details from a numerical analysis point
of view can be found in Chapter 6 of \cite{BurdenFaires}, and for
exact computations in Chapter 9 of \cite{Geddes:1992:ACA}.  The
latter also contains an exposition of the fraction-free version of 
Gaussian Elimination.  One can then show that this version of the 
algorithm works
(ie reduces an $n\times n$ Matrix of rank $n$ to an upper triangular
matrix with a non-zero diagonal) whenever the elements of the Matrix
are from an arbitrary \emph{integral domain}.  There are extensions to
rings which contain zero-divisors as well, but these are beyond the scope
of the current paper.

A frequent variation point is that for reasons of either numerical stability
or coefficient growth, instead of choosing a random row $k$ with a non-zero
element, pivoting is performed by choosing an element that optimizes
a particular metric.  In the case of numerical stability, the largest (in
absolute value) is chosen, whereas for coefficient growth cases, it is 
the smallest (according to some given size metric) that is chosen.  For
numerical stability, full pivoting is also frequently used, although
we will not deal with this here.

\subsection{Some examples}

First, we discuss 3 different implementations of Gaussian Elimination
from Maple's library.  The first one (shown in figure~\ref{gausselim}) 
is the most
straightforward implementation of Gaussian Elimination.  However, it
does include some explicit parametrization, namely the use of
\texttt{Normalizer} and \texttt{Testzero} to allow the use of this
algorithm with any arithmetic domain (in other words one where
the globally defined, polymorphic operations and constants
$+, *, \slash, -$, and $0$ are defined, as well as having a notion of
\texttt{length}).  This implementation  uses Maple's old \texttt{matrix} 
datatype which is in fact implemented via hash tables; while this makes
element access very slow, it does allow very efficient storage of 
sparse matrices, as well as allowing \emph{indexing functions} to be
used for structured matrices.  See the Maple documentation
\cite{Monagan:2001:M7P} for more details on these features.  It is important
to note that this implementation would \textbf{not} be appropriate for 
floating point numbers, as it would be numerically quite unstable.

\begin{figure}
    \begin{small}
\begin{verbatim}
gausselim := proc(A::matrix)
local B,n,m,i,j,c,t,r;
n := rowdim(A); m := coldim(A);
B := map(Normalizer, A); r := 1;
for c to m while r <= n do
    # Search for a provably non-zero pivot element
    i := 0;
    for j from r to n do
        if not Testzero(B[j,c]) then
            if i = 0 then i := j;
            elif length(B[j,c]) < length(B[i,c]) then 
                i := j
            end if; # otherwise B[i,c] is the best
        end if;
    end do;
    if i <> 0 then
        # interchange row i with row r is necessary
        if i <> r then
          for j from c to m do 
            t := B[i,j]; B[i,j] := B[r,j]; B[r,j] := t
          end do
        end if;
        for i from r+1 to n do
          if B[i,c] <> 0 then 
              t := Normalizer(B[i,c]/B[r,c]);
              for j from c+1 to m do 
                B[i,j] := Normalizer(B[i,j]-t*B[r,j]) 
              end do;
              B[i,c] := 0
          end if;
        end do;
        r := r + 1          # go to next row
    end if
end do;              # go to next column
end proc:
\end{verbatim}
\end{small}
\caption{Generic Gaussian Elimination in Maple}
\label{gausselim}
\end{figure}

The second version (see appendix~\ref{FractionFreeElimination}) is
significantly more parametrized: it is implemented via Maple's \texttt{Domains}
package which implements generic programming as originally found in 
Axiom \cite{Jenks:1992:ASC}.  These facilities are modelled on the structures
of classical Algebra like Rings, Fields, Monoids, Groups, etc.  While
the resulting program is very generic, and in Axiom (at least when compiled 
via Aldor \cite{Watt:2002:HCA})
is quite efficient, in an interpreted language like Maple, the results
are unacceptably slow.  In any case, this program implements the fraction-free
version of Gaussian Elimination, which works over any Division Ring.
If the Division Ring is also a Euclidean Domain, then the underlying
Euclidean Norm is used as the notion of \emph{size}, otherwise the 
underlying universally polymorphic \texttt{length} function, which works
for any Maple structure, is used.  Again, one should note that floating
point numbers do not form a Division Ring (as addition is not associative
and multiplication does not distribute over addition).  This implementation
works for any container $A$ (storing the matrix) which allows 2-dimensional
indexing for element retrieval and storage.  Unlike the previous version, 
neither \texttt{Normalizer} nor \texttt{Testzero} are used, which implies
that the underlying rings are assumed to store their elements in 
normalized form, that arithmetic operations return results in normalized
form, and thus zero-testing is straightforward.

The third variant (see Appendix~\ref{modular}) is also quite parametrized,
but along different dimensions: in this case the input domain is fixed
to be an arbitrary algebraic extension (possibly empty) of a finite
field $\mathbb{Z}_p$.  This particular implementation has potentially
multiple outputs: the actual reduced Matrix (always), the rank of the
Matrix (sometimes), and the determinant of the Matrix (sometimes).  It
works for non-square matrices, as well as for augmented matrices, where
one can pass in an integer \texttt{right\_margin} parameter to specify this.
The determinant can only be computed in the case where the principal 
matrix is square.  Elements are not assumed normalized, and for 
efficiency reasons, the normalization procedure used for a pure finite
field is different than the one used for algebraic extensions.  There
is also explicit code for keeping track of the determinant.  This 
version is optimized for the newer \texttt{rtable} data-structure representation
of Maple's newer \texttt{Matrix} constructor, which allows a one-line
expression
\begin{verbatim}
B[i,1..-1], B[row,1..-1] := B[row,1..-1], B[i,1..-1];
\end{verbatim}
to be used for exchanging 2 rows.

For reference, the straightforward implementation of Gaussian Elimination
for matrices of floating point numbers, in OCaml is shown in
figure~\ref{ocaml1}.  It has been coded to resemble the first
Maple version in figure~\ref{gausselim} as much as possible, where 
this makes sense; the most significant difference is that pivoting is
done on the largest number instead of the smallest.

\begin{figure}
    \begin{small}
\begin{verbatim}
let ge_float a =
  let b = copy(a) in
  let n = Array.length b and m = Array.length b.(0) in
  let r = ref 0 and c = ref 0 and i = ref 0 in
  while !c < m && !r < n do
    i := (-1);
    for j = !r to n-1 do
      if not ( b.(j).(!c) = 0.) then 
        if !i = (-1) || 
           abs_float b.(j).(!c) > abs_float( b.(!i).(!c) ) then
             i := j;
    done;
    if !i != (-1) then begin
      if !i <> !r then
        begin
          for j = !c to m-1 do
              let t = b.(!i).(j) in
              b.(!i).(j) <- b.(!r).(j);
              b.(!r).(j) <- t;
          done;
        end;
      for ii = !r+1 to n-1 do
        if not (b.(ii).(!c) = 0.)  then begin
          let t = b.(ii).(!c) /. b.(!r).(!c) in
          for j = !c+1 to m-1 do
              b.(ii).(j) <-  b.(ii).(j) -. t*.b.(!r).(j)
          done;
          b.(ii).(!c) <- 0.;
        end;
      done;
      r := !r + 1;
    end;
    c := !c + 1;
  done ;
  b
  \end{verbatim} 
  \end{small}
  \caption{Floating-point GE in Ocaml}\label{ocaml1}
\end{figure}

\subsection{Design points analysis}\label{designpoints}

A careful review of the rougly 35 implementations of GE we found, as
well as roughly 50 additional implementations of related algorithms, 
a number of implementation patterns became readily apparent.  The main 
variations found were:
\begin{enumerate}
	\item \textbf{Domain}: In which (algebraic) domain do the matrix
		elements belong to.  Some implementations were very specific
		($\mathbb{Z}, \mathbb{Q}, \mathbb{Z_p}, 
		\mathbb{Z}_p\left[\alpha_1,\ldots,\alpha_n\right], 
		\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$, 
		$\mathbb{Q}\left[\alpha\right]$, and floating point numbers for 
		example), while others were generic for elements of a field,
		multivariate polynomials over a field, elements of a formal Ring
		with possibly undecidable zero-equivalence, or elements of a 
		Division Ring.  In the roughly 85 pieces of code we surveyed,
		20 different domains were encountered.
	\item \textbf{Fraction-free}: Whether the Gaussian Elimination
		algorithm is allowed to use unrestricted division, or only
		exact (remainder-free) division.
	\item \textbf{Representation of the matrix}: Whether the matrix
		was represented as an array of arrays, a one-dimensional array,
		a hash table, 
		whether indexing was done in C or Fortran style.  Additionally,
		if a particular representation had a special mechanism for efficient
		row exchanges, this was sometimes used.  Other representations
		were also used.
	\item \textbf{Length measure (for pivoting)}:  For stability reasons
		(whether numerical or coefficient growth), if a domain possesses
		an appropriate length measure, this was sometimes used to choose
		an ``optimal'' pivot for row exchanges.
	\item \textbf{Output choices}:  Whether just the reduced matrix, or
		additional the rank and the determinant were to be returned.
		It is worthwhile noting that in the larger algorithm family,
		routines like \texttt{LinearAlgebra:-LUDecomposition} have
		up to $2^6 + 2^5 + 2^2 = 100$ possible outputs.
	\item \textbf{Normalization and zero-equivalence}: Whether the 
		arithmetic operations of the domain at hand gives results in 
		normalized form, and whether a specialized zero-equivalence 
		routine needs to be used.
\end{enumerate}

\noindent These choices are not completely independent:  
\begin{itemize}
	\item While it is always possible to use full Gaussian Elimination for
		matrices with elements in a Division Ring, the result will in general
		lie in the Field of fractions associated with this Ring.  On the other
		hand, fraction-free GE will always give results in the same domain as
		the input.  There are subtle coefficient growth reasons that come
		into play when deciding, for each domain, the ``best'' version of
		GE to use.
	\item Some domains do not have meaningful length measures.  In particular
		$\mathbb{Z}_p$ is such a field; but by the same token, there is
		no coefficient growth problem with $\mathbb{Z}_p$.  On the other
		hand, even though floating point numbers are of fixed syntactic
		size, numerical stability implies that the absolute value gives
		a meaningful length measure.
	\item While zero-testing is often done via normalization, there are
		times when full normalization is very expensive, and heuristic-driven
		zero-testing can be fruitfully used as a filter for ``easy''
		zero equivalence, before more expensive techniques are used.
\end{itemize}

\noindent A few additional design points were encountered in some cases:
\begin{itemize}
	\item If zero equivalence was known to be undecidable, and a case
		was encountered where a (provably) non-zero pivot could not be
		found, either a warning or an error were issued.
	\item Some implementations had embedded \texttt{userinfo}
		statements, which is Maple lingo for user-controlled directives 
		that give the user information about the execution state of 
		particular algorithms.
	\item Specification that the Matrix was in fact an augmented matrix,
		and that it possessed a natural ``right margin''.  Elimination
		is then performed on the full matrix, but column choices stop
		at this given right margin.  Curiously, while this was frequently
		implemented, this feature was very rarely documented, even though
		this functionality is frequently used in textbooks which 
		outline how Gaussian Elimination can be used on augmented matrices
		to implement solving of linear equations and matrix inversion.
\end{itemize}

\section{Staging Gaussian Elimination}\label{staging}

Using an abstract version of Gaussian Elimination, and assuming that
the reader has some knowledge of staging in MetaOCaml (as expounded in
the other papers in this volume), we show a first na\"{\i}ve staging
of Gaussian Elimination, based on an abstract version of the 
algorithm.

\subsection{Abstract Gaussian Elimination}
Looking at all the Gaussian Elimination algorithms seen so far,
one can extract a staging-free high-level Ocaml version as follows:
\begin{small}
\begin{verbatim}
let ge_high findpivot swap zerobelow a m n =
    let b = copy(a) in
    let r = ref 0 and c = ref 0 in
    while !c < m && !r < n do
        match findpivot b c r n with
        Some i -> begin
            if i <> !r then
                for j = !c to m-1 do
                    swap b i r j;
                done;
            zerobelow b r c n m;
            r := !r + 1;
            end;
        | None -> () ;
        c := !c + 1;
    done ;
    b
\end{verbatim} 
\end{small}

For a version of GE that works with arrays of arrays of floating 
point numbers, the 3 required routines can be given as
\begin{small}
\begin{verbatim}
let findpivot b c r n =
  let i = ref (-1) in begin 
    for j = !r to n-1 do
      if not ( b.(j).(!c) = 0.) then 
        if !i = (-1) || 
          abs_float b.(j).(!c) < abs_float b.(!i).(!c) then
            i := j;
    done;
    if !i== -1 then None else Some !i;
  end;

let swap b i r j =
  let t = b.(i).(j) in begin
    b.(i).(j) <- b.(!r).(j);
    b.(!r).(j) <- t;
  end;

let zerobelow b r c n m = 
  for ii = !r+1 to n-1 do
    if not (b.(ii).(!c) = 0.)  then
      begin
        let t = b.(ii).(!c) /. b.(!r).(!c) in
        for j = !c+1 to m-1 do
          b.(ii).(j) <-  b.(ii).(j) -. t*.b.(!r).(j)
        done;
        b.(ii).(!c) <- 0.;
      end;
    done;
\end{verbatim}
\end{small}

We can recover the previous version \texttt{ge\_float} as
\begin{small}
\begin{verbatim}
let ge_float a = 
    let n = Array.length a and m = Array.length a.(0) in
    ge_high findpivot swap zerobelow a m n
\end{verbatim}
\end{small}

It is however worthwhile to examine closely the type inferred
for \texttt{ge\_high}:
\begin{small}
\begin{verbatim}
# val ge_high : 
  ('a -> int ref -> int ref -> int -> int option) ->
  ('a -> int -> int ref -> int -> 'b) ->
  ('a -> int ref -> int ref -> int -> int -> 'c) -> 
  'a -> int -> int -> 'a = <fun>
\end{verbatim}
\end{small}
\noindent The ``matrix'' \texttt{a} is completely abstracted
out, and we have a fully abstract \verb+'a+ as its type.  Only
when \texttt{findpivot} and friends are instantiated do we get 
this particular type refined.  As well, since the matrix type is
abstract, this also abstracts out the type of elements.
So this abstraction exercise has
already gained us (some) representation independence.

\subsection{First staging}
\noindent  Of course, if one were either greedy for every single
microsecond of efficiency or, as in our case, we are preparing for
a thorough staging of the above routine, we would like
to modify the above code to completely remove all (run-time) traces of 
these abstractions.  This can be done as follows:
\begin{small}
\begin{verbatim}
let ge_gen findpivot swap zerobelow =
    .< fun a m n ->
    let b = copy(a) in
    let r = ref 0 and c = ref 0 in
    while !c < m && !r < n do
        match .~(findpivot_gen .<b>. .<c>. .<r>. .<n>.) with
        Some i -> begin
            if i <> !r then
                for j = !c to m-1 do
                    .~(swap_gen .<b>. .<i>. (lr .<r>.) .<j>.);
                done;
            .~(zerobelow_gen .<b>. .<r>. .<c>. .<n>. .<m>.);
            r := !r + 1;
            end;
        | None -> () ;
        c := !c + 1;
    done ;
    b 
\end{verbatim}
\end{small}
\noindent  While the source code above is not particularly attractive, 
the modifications necessary are certainly quite routine, and
the resulting code is quite readable.  To get
routine \texttt{swap\_gen} from routine \texttt{swap} is also
straightforward:
\begin{small}
\begin{verbatim}
let lr x = .< ! .~ x >.
let swap_gen b i r j =
    .< let b = .~b and i = .~i and j = .~j and r = .~r in
        let t = b.(i).(j) in  begin
        b.(i).(j) <- b.(r).(j);
        b.(r).(j) <- t;
    end >.
\end{verbatim}
\end{small}
\noindent  However, this produces some extra let bindings in the
resulting code (not shown here).  To get back to the original source code,
as intended, the modification is still simple, although the resulting
code (below) now looks quite ugly.  We will assume that these let
bindings are harmless (in other words completely removed by the 
compiler), and use them in our presentation, since we know we could
remove them if needed.
\begin{small}
\begin{verbatim}
let swap_gen2 b i r j =
    .< let t = (.~b).(.~i).(.~j) in  begin
        (.~b).(.~i).(.~j) <- (.~b).(.~r).(.~j);
        (.~b).(.~r).(.~j) <- t;
    end >.
\end{verbatim}
\end{small}
\noindent  Similar modifications to \texttt{zerobelow} and
\texttt{findpivot} are necessary as well.  Once performed, we can
obtain \texttt{ge\_high} verbatim via specialization:
\begin{small}
\begin{verbatim}
let spec_ge = (ge_gen findpivot_gen swap_gen2 zerobelow_gen);
let ge_float2 a = 
    let n = Array.length a and m = Array.length a.(0) in
    let ge = .! spec_ge in
    ge a m n ;
\end{verbatim}
\end{small}
\noindent Close inspection of the program \texttt{spec\_ge} shows that,
up to variable renaming, the resulting code is essentially that 
of \texttt{spec\_float}.
The only real change (and it is not strictly necessary) is the use of
the \texttt{option} type to pass information out from \texttt{findpivot}
instead of using a flag value of $-1$ for the integer reference variable
\texttt{i}.

\subsection{Optimized pivoting}

The first ``design point'' that we will stage is whether we simple use
the first non-zero element we encounter as a pivot, or whether we 
search in the complete column for the ``best'' pivot.  We will thus
assume that we are (potentially) given a function \texttt{better\_pivot}
which, given two values, will return \texttt{true} if the second value
would be a better pivot.  By using an option type, we can also choose
whether we want to use this function.  We parametrize 
\texttt{findpivot\_gen} by this new value, giving:
\begin{small}
\begin{verbatim}
let findpivot_gen better_pivot b c r n =
    let cond i x y = match better_pivot with
    | Some f -> .< !(.~i) = (-1) || .~(f x y) >.
    | None ->   .< !(.~i) = (-1) >.
    in
    .< let i = ref (-1) in begin 
        for j = !(.~r) to .~n-1 do
            if not ( (.~b).(j).(!(.~c)) = 0.) then 
                if .~(cond .<i>. .< (.~b).(j).(!(.~c)) >.
                           .< (.~b).(!i).(! .~c) >. ) then
                    i := j
        done;
        if !i == -1 then None else Some !i;
    end >. ;;
\end{verbatim}
\end{small}
\noindent where we have added a new (local) code-generating
function \texttt{cond} which generates either single condition
\verb+!i = -1+ or splices in the code of our passed-in
\texttt{pivot\_len} function.  Since this is a code-generator,
the body of \texttt{findpivot\_gen} has to be modified accordingly
to pass in the \emph{code} representations necessary for properly
constructing the conditional.

Naturally, we need to make a corresponding change to the instantiation
call.  To reproduce the previous program, this would be done as
\begin{small}
\begin{verbatim}
let spec_ge = 
  let fp = findpivot_gen (Some 
    (fun x y -> .<abs_float .~x < abs_float .~y >. )) in
  ge_gen fp swap_gen2 zerobelow_gen;
\end{verbatim}
\end{small}

\section{Staging techniques}
\label{techniques}

While it is possible to continue staging the code in the na\"{\i}ve
manner described in the previous section, eventually the resulting
source code becomes almost impossible to read.  In other words, this ad hoc
method of staging, as we found out first-hand, does not scale very
well, nor is it always straightforward.
Also, this suggest that 
maintenance would be very challenging -- yet one of the
explicit goals of this exercise to help maintainability.  We need to
find smarter ways to combine and stage code, to simultaneously obtain
readable and efficient programs.  This section presents various
techniques that we have used to successfully stage all of the 
principal design points described in subsection~\ref{designpoints}.

We will use a combination of four techniques: the use of basic
combinators for pieces of code, state passing, abstract interpretation, and 
dictionary passing.  It is important to note that we are staging
(deeply!) imperative code, to obtain imperative code.  Thus we are forced
to develop some techniques which differ from previous work
\cite{TahaThesis,KiselyovTaha} where they were staging purely functional code,
sometimes with the aim of obtaining functional residual code, and sometimes
aiming for imperative residual code.  Of course, some of the techniques
developped below have similar semantic underpinnings as the monadic
approach of \cite{KiselyovTaha}, although this is probably not 
readily apparent when syntactically comparing our results with theirs.

\subsection{Preliminaries}\label{prelim}

Probably the most important point to remember when staging code, especially
in a higher-order language like Ocaml, is that the usual abstraction
technique of using higher-order functions does not quite work as 
expected anymore.  The only object that can be effectively inlined
has to be of \texttt{code} type; in particular it cannot be of
a functional type.  More subtly, we cannot just ``lift'' functions in the
hopes of still being able to apply them before run-time, we instead have to
be careful to deal with (lifted) function bodies.  For example, compare
\begin{small}
\begin{verbatim}
let lift x = .< x >.
let star1 = lift ( * ) in
let f1 x y = .< .~star1 (x+1) (y+3) >. in
f1 2 3 ;;

let star2 x y = .< .~x * .~y >. in
  let f2 x y = star2 .<x+1>. .<y+3>. 
  and f3 x y = let z=x+1 and zz=y+3 in 
      star2 (lift z) (lift zz) 
  and f4 x y = let z=x+1 and zz=y+3 in 
      star2 .<z>. .<zz>. in
[ f2 2 3 ; f3 2 4 ; f4 2 3] ;;
\end{verbatim}
\end{small}
\noindent which gives as results
\begin{small}
\begin{verbatim}
# - : ('a, int) code =
.<(((* compiled code omitted (as id: x) *)) 
    (2 + 1) (3 + 3))>.
# - : ('a, int) code list =
[.<((2 + 1) * (3 + 3))>.;
 .<((* compiled code omitted (as id: x) *) *
    (* compiled code omitted (as id: x) *))>.;
 .<(3 * 6)>.]
\end{verbatim}
\end{small}

All of these are operationally equivalent!  However, only for
\texttt{f4} do we get what we were really after.
Careful reading of \cite{TahaThesis} explains all of these, but
they are still a bit disconcerting to the novice MetaOCaml programmer.
While the difference between \texttt{f2} and \texttt{f4} is easy to
explain in terms of binding-time, the difference between \texttt{f1}
and \texttt{f4} takes more getting used to for the seasoned functional
programmer, who is used to passing around functions.  It is very 
important to pass around code instead - which is the main difference
between \texttt{star1} and \texttt{star2}: the first is completely
closed code which can only applied at run-time, while the latter can
still be applied at stage 1.  This may be obvious to meta programming
gurus, but is not emphasized enough in practical introductions to
staging in MetaOCaml.

\subsection{Combinators}

It is very natural to try to develop \emph{code combinators}, to attempt
to do for code what worked so well for parsers \cite{ParserCombinators}.
While this works to a certain extent, the next technique in part 
destroys the applicability of pure code combinators to many cases of interest,
as we are here mainly dealing with combining imperative program fragments.

While we experimented with many pure code combinators, in the end 
the only one we actually used was a simple lift of sequencing:

\begin{small}
\begin{verbatim}
let seq a b = .< begin .~a ; .~b end >.
\end{verbatim}
\end{small}

\noindent It is worthwhile noting that the begin/end are needed, otherwise
MetaOCaml complains that this is syntactically invalid.
Another useful combinator (used in development) was composition:

\begin{small}
\begin{verbatim}
let compose a b = fun x -> .< .~a ( .~b x ) >.
\end{verbatim}
\end{small}

\noindent While it is possible to create many more combinators, like
\begin{small}
\begin{verbatim}
let _if_ cond body = .<if .~cond then .~body >. ;;
\end{verbatim}
\end{small}
\noindent in actual use we have found that this ends up making the code
much less readable than staging larger pieces of code.  But the reader
should remember that our aims are quite different than that of other authors,
so these conclusions a priori might only apply to our particular setting.

\subsection{State passing}

The first really new technique (in the context of staging) is that
of \emph{state passing style}.  This is taken simultaneously from two
different sources: the state-transformer paradigm of denotational
semantics for programming languages \cite{DenSem}, the StateMonad
and ST monads of Haskell, and Sumii's work
\cite{sumii01hybrid}
on hybrid online-offline partial evaluation, where a state-passing style
was used over a continuation-passing style, to great effect.

While in these general paradigms, the state space is usually 
open\footnote{in other words is potentially infinite}, we want much tighter
control on the state.  So we use an explicit record type to encode our
state configuration, with the finite
set of state variables that our complete program will use.  We can then
rely on the type-checker to make sure that this is the only state we use.
For example, for our needs we define
\begin{small}
\begin{verbatim}
type 'a state = 
    {b:'a array array; r:int ref; c:int ref; m:int; n:int}
\end{verbatim}
\end{small}
\noindent which then forces us to modify all our generators
to refer to this state configuration explicitly, and we have to add record
dereferences as appropriate.
\begin{small}
\begin{verbatim}
let swap_gen s i j =
    .< let s = .~s and i= .~i and j= .~j in
       let t = s.b.(i).(j) in  begin
        s.b.(i).(j) <- s.b.(!(s.r)).(j);
        s.b.(!(s.r)).(j) <- t;
    end >.
\end{verbatim}
\end{small}
\noindent  While this does make some code a bit harder to read, this
has (essentially) no run-time cost, and definitely cuts down on the 
number of parameters
needed for each generator.  The main generator then needs to be modified
to create this state configuration
\begin{small}
\begin{verbatim}
.< fun a -> 
   let s = {b=a; r=ref 0; c=ref 0; m=Array.length a.(0); 
         n=Array.length a} in
  ... >.
\end{verbatim}
\end{small}

Note that it is possible to remove even this (potential) overhead of
record selection, by defining our state configuration to contain
code generators instead of values.  See subsection~\ref{dict} for
an example of this.

\subsection{Abstract interpretation}

Programmers coming from systems that permit unbounded manipulation of 
white-box code (Scheme, Maple, Mathematica, etc) need to remember that
white-box code cannot exist
in MetaOCaml.  Thus any information about a piece of code that is necessary to
be re-used later must be explicitly exposed, since such information cannot be
seen once it is encapsulated in a code type.

Suppose we know a priori that the generic code contains 
\verb+ if !i = -1 then+, but that sometimes we would like to modify
this to\\
{\small\verb+if !i = -1 ||+\\
\verb+abs_float b.(j).(!c) > abs_float( b.(!i).(!c) ) then+},
if we had white-box code, this could be done by actually finding that code,
and explicitly modifying it.  In metaocaml, this technique is not available.
There are reasons for this (like voiding equational reasoning principles
on computations inside brackets \cite{TahaThesis}, as this essentially
reduces the brackets to syntactic questions \cite{SICP}).
Instead, one must explicitly insert a code generator\\
\verb+if .~(cond_gen .<i>. etc) then+ which will conditionally generate
the correct code.  The question then becomes, how do we indeed 
conditionally generate the correct code for each particular instance?

We use a simple case of the general technique of \emph{abstract interpretation} 
\cite{conf/popl/CousotC77} to solve this problem; see \cite{KiselyovTaha} for a
different use of the same technique.  The main idea is to take the fully
abstract code type, and to add some additional information to it, 
information which can be used at staging time to generate more precise
code.  This additional information is an approximation of the complete
information contained in the white-box code.
In our case, this translates to knowing whether our domain contains
the functions necessary to optimize the choice of a pivot.  Thus we 
define a type
\begin{small}
\begin{verbatim}
type pivoting_code_option =
  (('a,'b) code -> ('a,'b) code -> (('a,bool) code)) option;
\end{verbatim}
\end{small}
\noindent which says that we \textit{might} have a ``smaller than'' 
boolean code generator available.  This is used with a conditional
code generator
\begin{small}
\begin{verbatim}
let orcond_gen main_cond = function
    | Some alt_cond -> .< .~main_cond || .~alt_cond >.
    | None -> main_cond ;;
\end{verbatim}
\end{small}
\noindent to generate the required code depending on the situation.
A more complex use of the same technique is when a domain requires a 
normalization step; in that case, we have to insert the normalization
code into the abstract code for mapping over our generic two-dimensional
container.  We end up using code that looks like
\begin{small}
\begin{verbatim}
let mdapply1 mapper g c1 = match g with
    | Some f -> .< .~(mapper f c1) >.
    | None   -> c1 ;

.~(mdapply1 mapper normalizerf .<a>. )
\end{verbatim}
\end{small}
where \texttt{mapper} is a code generator for mapping a function \texttt{f}
(also given as code) over the code of an abstract container.  Also 
see section~\ref{concl} for a more substantial application of abstract
interpretation.

\subsection{Dictionaries}\label{dict}

While it is possible to use just the techniques of the previous sections
to stage a fairly generic GE, the resulting source code would be hideous.  In
particular, most generators would have between 15 to 20 parameters, 
since a typical domain has 10 different components, and a generic 
container has 5.  Instead, it makes a lot more sense to gather these
parameters together into a type.

To be more specific, let us look at domains.  Since they are essentially
Division Rings, we know that they must contain the constants zero and one, 
as well as the operations of addition, subtraction, multiplication
and (a restricted kind of) division.  It may optionally come with
a ``smaller than'' relation, a normalizer, and a zero-tester.  For
simplicity, we can also assume that we are given an explicit $-1$ value
as well.  So we could define a type corresponding to a domain over
some arbitrary base set:
\begin{small}
\begin{verbatim}
type 'b domain = {zero:'b; one:'b; minusone:'b;
    plus:'b -> 'b -> 'b; times:'b -> 'b -> 'b;
    minus:'b -> 'b -> 'b; div:'b -> 'b ->'b;
    smaller_than:('b -> 'b -> bool) option;
    normalizer:('b -> 'b ) option;
} ;;
\end{verbatim}
\end{small}
\noindent It should be remarked that what we are essentially 
building there is a type for a (static) dictionary for domains.
This is akin to (simple) type classes in Haskell, or indeed of
defining a module in OCaml over an abstract type.

If we lift this domain up one stage, then via partial application 
and inlining, we could hope to completely eliminate the overhead
of this level of genericity.  But doing things this way would run
into the problems already outlined in subsection~\ref{prelim}.
While we would get a program which is operationally correct, and which
Ocaml's underlying optimizer \emph{might} resolve to what we want,
we certainly have no guarantees of that.  To be quite sure of 
this, we need to provide not the values and functions that
correspond to a Division Ring, but rather the code (and code
generators) for each of these.  In other words, the type we really
need is
\begin{small}
\begin{verbatim}
type ('a,'b) domain = {zero:('a,'b) code; one:('a,'b) code;
  minusone:('a,'b) code;
  plus:('a,'b) code -> ('a,'b) code -> ('a,'b) code; 
  times:('a,'b) code -> ('a,'b) code -> ('a,'b) code;
  minus:('a,'b) code -> ('a,'b) code -> ('a,'b) code; 
  div:('a,'b) code -> ('a,'b) code ->('a,'b) code;
  smaller_than:(('a,'b) code -> ('a,'b) code -> 
               (('a,bool) code)) option;
  normalizerf:(('a,'b -> 'b) code) option;
  normalizerg:(('a,'b) code -> ('a,'b) code) option;
} ;;
\end{verbatim}
\end{small}
\noindent  We need two different versions of the normalizer function
because we might use the normalizer in two different ways: either
in an imperative setting, like in the body of a \texttt{for} loop, 
or in a functional setting, like as an argument to \texttt{Array.map}.

For example, one can define the domain of integers $\mathbb{Z}$ as
\begin{small}
\begin{verbatim}
let dom_int = { zero = .< 0 >. ; one = .< 1 >. ;
   minusone = .< -1 >. ;
   plus = (fun x y -> .<.~x + .~y>.) ;
   times = (fun x y -> .<.~x * .~y>.) ;
   minus = (fun x y -> .<.~x - .~y>.) ;
   div = (fun x y -> .<.~x / .~y>.) ;
   smaller_than = Some(fun x y -> .< abs .~x < abs .~y >.) ;
   normalizerf = None ;
   normalizerg = None }
\end{verbatim}
\end{small}
\noindent where we have been careful to define code-generating
functions rather than code-of-functions.  We have also completely
eschewed the use of \texttt{lift}, to ensure we get complete
inlining.  This might be ultimately be unecessary, but we would have 
to trust the MetaOCaml \texttt{run} function \texttt{.!} to 
systematically do this for us, which\footnote{for theoretically sound 
reasons!} we cannot rely on.

Using similar techniques, we can also completely abstract out
the matrix container as follows:
\begin{small}
\begin{verbatim}
type ('a,'c,'d) container2d = {
  get: ('a,'c) code -> ('a,int) code -> 
       ('a,int) code -> ('a,'d) code ;
  set: ('a,'c) code -> ('a,int) code -> 
       ('a,int) code -> ('a,'d) code -> ('a,unit) code ;
  dim1: ('a,'c) code -> ('a,int) code;
  dim2: ('a,'c) code -> ('a,int) code;
  mapper: ('a, 'd->'d) code -> ('a,'c) code -> ('a,'c) code
} 
\end{verbatim}
\end{small}
\noindent where \texttt{'a} is the extra type variable required by the
abstract code type, \texttt{'c} is the abstract container type, and
\texttt{'d} is the abstract element type.  Even though from 
\texttt{get} and \texttt{set}, we could derive the code for \texttt{mapper},
letting the user give an explicit \texttt{mapper} functions allows for
extra efficiency.  This could certainly be considered to be an
additional design point.

It is worthwhile noting that these 2 types are completely independent, 
and only when are they actually used together will the type system
unify the 3rd type variable of \texttt{container2d}
with the 2nd type variable from \texttt{domain}.  This is quite pleasing,
as it reassures us that we have an appropriate separation of concerns
between the domain of elements and the abstract container.

These 2 types have values that are fully available at staging time.  
Thus any overhead that could be associated with these abstractions 
is in fact completely eliminated, and the generated code is free
of these abstractions.  In other words, we were able to use staging
instead of defunctionalisation \cite{defunc} 
to remove a complete layer of abstraction.

\section{Staging Gaussian Elimination, continued}

The techniques of the previous section allow us to take care of all
of the major design points that were outlined in subsection~\ref{designpoints}.
For example, by using a combination of abstract interpretation and a
code combinator, we can track which information is needed in the final
code, allowing the final code generator for the swap routine to look
like
\begin{small}\begin{verbatim}
let swapr_gen dom contr track s i j =
  let swap =
  .< let b = (.~s).b and r = !((.~s).r)
      and i= .~i and j= .~j in
     let t = .~(contr.get .<b>. .<i>. .<j>. ) in  begin
       .~(seq
           (contr.set .<b>. .<i>. .<j>. 
		             (contr.get .<b>. .<r>. .<j>. ))
           (contr.set .<b>. .<r>. .<j>. .<t>. ) );
  end >. in
  let tds = 
     .< (.~s).detsign := 
	   .~(dom.times .<!((.~s).detsign)>. dom.minusone) >. in
  match track with
  | TrackNothing -> swap
  | TrackDet     -> seq swap tds
\end{verbatim}\end{small}
\noindent where we see the use of the abstract container \texttt{cont},
the abstract domain \texttt{dom}, the tracking information
\texttt{track}, and the combinator \texttt{seq}.

The one area that was slightly less satisfying was in staging the
choice of outputs.  While any given instantiation of GE will produce
a specific output, the family of algorithms exposes a choice.  Since
typing in MetaOCaml is done once, this choice persists in the
\emph{type}
of the final program, even though it should be readily apparent that
this choice does not in fact persist in the actual generated code.
More precisely, we use a type
\begin{small}\begin{verbatim}
type ('b,'c) outputs = 
      Matrix of 'b
    | MatrixRank of 'b * int
    | MatrixDet of 'b * 'c
    | MatrixRankDet of 'b * int * 'c
\end{verbatim}\end{small}
\noindent for the output of GE.  Clearly any generated code will only
use one of these choices, but this will not be apparent in the 
type of the generated code.

In the end, the type of our function \texttt{specializer} is
\begin{small}\begin{verbatim}
val specializer : ('a, 'b) domain ->
  ('a, 'c, 'b) container2d -> fracfree:bool ->
  outputs:outchoice -> ('a, 'c -> ('c, 'b) outputs) code
\end{verbatim}\end{small}
\noindent which says that it outputs code for a 
function of type \verb+'c -> ('c, 'b) outputs+ with the 
constraints that \verb+'c+ be a container type and \verb+'b+ be
a domain type.  A complete listing, with a specialization, is
given in Appendix~ref{finalcode}.

\section{Related work}\label{related}

There is now quite an established line of work in scientific (numerical)
programming of using generative techniques (mainly in C++, see 
\cite{Czarnecki,Veldhuizen:1998:ISCOPE,ATLAS} and the work they cite).
However, their worries are somewhat different than ours: they mainly
aim for efficiency - they are able to generate code which is
usually faster than the handwritten Fortran programs they replace.
While they do not overlook maintainability issues, that is often a 
secondary concern.
This drive for efficiency usually means worrying about many low-level
details (cache sizes, cache misses, etc), whereas we are much more 
interested in writing very generic
programs which can hopefully still be quite efficient.
While clearly very successful, with FFTW \cite{FFTW98} being the most famous 
this work cannot claim to be type-safe.  In the case of FFTW, the work
of Kiselyov, Swadi, and Taha \cite{KiselyovTaha} is probably the first
that can claim type safety.

Our work is much more closely related to the work on generic programming
coming out of Symbolic Computation and Computer Algebra
\cite{musser89generic,Jenks:1992:ASC}.  Computer Algebra systems, starting
with ScratchPad \cite{ScratchPad}, on through Axiom \cite{Jenks:1992:ASC} and
Aldor \cite{Watt:2002:HCA} worked very hard at modelling the structures
encountered in mathematics (Groups, Rings, Fields, Modules, etc)
through abstract data types in various programming languages.  This
allowed them to code many algorithms very generically, allowing
extremely general mathematical structures to be defined and computed with.
Unfortunately, this very genericity came at an extremely high efficiency
cost; part of this is due to the fact that many mathematical structures
(with Fields and Matricies being good examples) are not inductive data
types.  In fact, any reasonable type system for properly reflecting 
mathematics requires dependent types (as was already known to the 
AUTOMATH people in 1968 \cite{Bruijn83}, but has been 
re-discovered many times).  In Maple itself, the \emph{Domains}
package (based on an older package called Gauss \cite{Gruntz:1994:IG})
successfully implements this strategy, but at a whopping factor of
30 efficiency cost over base Maple, which is already a factor of 60
slower than comparable C code.  This is why generative techiques must
be used to (at least) regain that last factor of 30.  What we show here
is that (in the context of MetaOCaml), this indeed works.

Others have used partial evaluation techniques to solve the
``configuration'' problem, closely related to what we have done here.
In particular, the work of Consel and his colleagues
\cite{lemeur-al:hosc03,Marlet-Thibault-Consel:jase99} is relevant.  They use
a declarative language to declare the specialization points of 
components; the components are written in a mainstream programming language
(here C), without the need for annotations.  While very pragmatic, this
inherits all of the defects of the host language - and in this case
this means that type-safety is completely lost.  Additionally, it is
clear that such an approach could not be used to express nearly as many
design points as we have been able to express in quite compact code.
Nevertheless, some of the underlying ideas of this work is very close
to ours.

Lastly, it should be noted that the programs we end up writing are very
reminiscent of hand-written language interpreters in cogen style
\cite{conf/pepm/BondorfD94,conf/icfp/Thiemann96}.  In hindsight,
this is not really surprising, as this is the style that MetaOCaml
forces upon us, by not having white-box code. Thus we have to be writing
code generating functions.

\section{Conclusion}\label{concl}

We have shown how to combine several techniques 
namely combinators, abstract interpretation, state-passing style, and 
dictionary passing (at staging time) can be 
carefully and successfuly combined to achieve our goals of 
writing a generic, efficient, type-safe implementation of the
Gaussian Elimination family of algorithms.  Furthermore, 
the generated code does not include unecessary artifacts introduced
by generalizing the code over several design points.

It should be easy to see this these techniques generalize immediately
to the larger family of algorithms containing LU decomposition, Cholesky 
factorization, forward as well as backward elimination, and of course
linear system solvers.

\subsection{Design points revisited}

Reviewing the design points listed in subsection~\ref{designpoints},
we have shown how to abstract out domains, whether the algorithm should
be fraction-free or not, matrix representations, the use of a size
measure (when available) for pivoting, output choices, and the
potential need for normalization.

On the other hand, we have not shown how to deal with domains where
zero-testing is undecidable, nor in fact on how to use a special
zero-testing routine when it is given; both of these are straightforward
using the techniques shown here.

Something else which has not been adequately addressed is whether
a particular domain requires that a fraction-free algorithm be used
or not -- in effect whether the input domain is in fact a Field or 
just a Division Ring.  We could simply add a new boolean field 
\texttt{is\_field} to our domain type, but this is not very elegant.
It would be much more pleasant if the compiler had some Theorem 
Proving capabilities, and we could instead let the compiler decide
this property.

Although we showed how to use a domain's length function, we are
hard-coding the \texttt{smaller\_than} function.  Akin to the
decision to use a fraction-free algorithm, it would be much better
to let the compiler somehow prove that we are either in a numerical
domain that suffers from instability issues, or in a symbolic domain
which suffers from coefficient growth issues, and conclude how
the length function should be used.

While we have shown how to deal with output choices, our approach
clearly does not scale: given something like the current implementations for 
LU Decomposition in Maple, we would have to use an algebraic type
with 100 constructors.  This is clearly not a realistic solution.
However, in these cases, a more sophisticated type system could
perhaps deal with the issue: the output type is quite simple at
stage 0.  Thus if there could be \emph{staged types} as well as
staged code, this problem could be easily handled.  Another intriguing
possibility would be to use GADTs.  Naturally, this could also be
handled via dependent types, but in this instance, this seems like
overkill.

Lastly, the container type could be abstracted out even more.  Our
algorithm right now is only really well-suited to dense rectangular
matrices.  Any kind of structured matrix, especially a sparse matrix,
could benefit from a differently structured traversal strategy. 
Simple issues like faster methods for row exchange would be quite
straightforward to handle, but handling sparsity is more challenging.
We believe that the techniques we show here are sufficient to handle
this as well, but would require a more significant restructuring of
the code generators.  Note that we have also completely ignored 
some issues
of frequent concern in implementations of Gaussian Elimination for
large matrices of floating point numbers, like full pivoting, 
blocking, etc.  Full pivoting would be quite straightforward to 
implement, whereas blocking would be more challenging.

\subsection{Future work}

We intend to deal with structured matrices next, as well as actually
implementing larger portions of the algorithm family centered around
Gaussian Elimination.  We then hope to continue generalizing this work --
as it is known that Gr\"{o}bner bases degenerate to Gaussian Elimination
in the case of linear systems, it would be interesting to see if we could
reproduce this via staging as well.  This is definitely considerably
more difficult, as linearity is a \emph{property} of the input system.
Dealing with the complete family of algorithms related to Gaussian 
Elimination, and linear algebra algorithms in general, should be
relatively straightforward, and we intend to complete this work to
ensure this is indeed the case.

We also want to pursue more aggressively the issue of abstract interpretation.
One item that intrigues us is how straightforward it is in our setting
to generalize the code shown in section 2 of~\cite{KiselyovTaha}:
\begin{small}
\begin{verbatim}
type ('b,'c) abstract = Elem of 'b | Any of 'c 
type ('a,'b) monoid = {one: 'b; op:'b->'b->'b;
    op_gen:('a,'b) code->('a,'b) code->('a,'b) code}

let concretize = function 
    | Elem x -> .<x>.
    | Any  x -> x
and generalize x = Elem x

let special_op dom a y = 
    if a=dom.one then 
        y 
    else 
        Any (dom.op_gen .<a>. .< .~(concretize y) >.)

let gen_op dom = fun x y ->
    match (x,y) with
    | (Elem a, Elem b) -> Elem (dom.op a b)
    | (Elem a, y)      -> special_op dom a y
    | (x, Elem b)      -> special_op dom b x
    | (x, y)           -> Any (dom.op_gen 
      .< .~(concretize x) >. .< .~(concretize y) >.)

\end{verbatim}
\end{small}
This code generates an abstract operator for any monoid.
We need to specify the (actual) identity for the monoid, as 
well as the monoid operation and a code generator for
the monoid information.  Then we use abstract interpretation
to evaluate the monoid operations as early as possible.  Note that
here we are able to evaluate any monoid operation, and not just
operations involving the identity.  
We can easily specialize this to the case covered in
\cite{KiselyovTaha} via
\begin{small}
\begin{verbatim}
let ( ** ) x y = gen {
    one = 1 ; op = (fun x y -> x * y);
    op_gen = fun x y -> .< .~x * .~y >. } x y 
let rec power n x = 
    if n=0 then generalize 1 else x ** (power (n-1) x)
let power3 = 
    .< fun x -> .~(concretize (power 3 (Any .<x>.))) >.
\end{verbatim}
\end{small}
and we get the expected result of
\begin{small}
\begin{verbatim}
val power3 = .< fun x -> x*x*x >.
\end{verbatim}
\end{small}
Better still, the version of power above, apart from the use of
\texttt{generalize}, looks very much like the power code that one would
na\"{\i}vely write (modifying it to do binary powering is straightforward).  
The implementation above assumes a commutative monoid, which
is frequently the case (and is the case here), but it can easily be
generalized to the non-commutative case.  On the other hand, the 
above version is easily generalized to do normalization of 
multivariate monomials over an arbitrary base monoid.

While we have used records here to encode all our abstract types,
we really should have used Ocaml's modules, as that is the proper way
to encode dictionaries in Ocaml.  If we then use Functors, for both
input and output configuration~\cite{KiselyovPersonal}, we can make
our implementation even clearer, as well as offering a solution to the
problem of configurable output types.

Lastly, we want to investigate the additional power that using a
combination of monadic and continuation passing style may bring.  It
would appear that staging certain complex aspects (like determinant
tracking) can be made considerably simpler by using this style.

\subsection{Acknowledgments}

This work started with a visit by the author to Walid Taha while the latter
was still at Yale.  Thanks to his gentle prodding, this paper finally 
got written.  This paper benefited from several useful conversations
with Bill Farmer on the topics of genericity, generative software, and
compilers as theorem provers.  Thanks go to David Linder for helping
the author catalogue all the implementations of Gaussian-Elimination
related algorithms in Maple, and Fr\'{e}d\'{e}ric Chyzak for drawing
the authors' attention (years ago) to this issue.  Not least, Oleg 
Kiselyov made some very very fruitful comments on a draft of this paper, for
which author is grateful.

\bibliographystyle{elsart-num}
\bibliography{ge}

\appendix
%Appendix A
\section{Maple code for \texttt{FractionFreeElimination}}
\label{FractionFreeElimination}
\begin{small}
\begin{verbatim}
FractionFreeElimination := proc(C,M,A,m,n) 
  local d,i,j,k,s,t,size;
  size := `if`(hasCategory(C,EuclideanDomain), 
               C[EuclideanNorm], length);
  s := 1; d := C[1];
  for k to m-1 do
      for i from k to m while A[i,k] = C[0] do end do;
      if i > m then return C[0] fi;
      for j from i+1 to m do
          if A[j,k] <> C[0] and 
              size(A[j,k]) < size(A[i,k]) then 
              i := j 
          end if
      end do;
      #  Pivot is A[i,k], interchange if necessary
      if i <> k then
        s := - s;
        for j from k to n do 
          t := A[k,j]; A[k,j] := A[i,j]; A[i,j] := t;
        end do;
      end if;
      #  Fraction-free elimination
      for i from k+1 to m do
          for j from k+1 to n do 
              t := C[`-`]( C[`*`](A[i,j],A[k,k]), 
                           C[`*`](A[k,j],A[i,k]));
              A[i,j] := C[Div](t, d)
          end do;
      end do;
     d := A[k,k]
  end do;
  s
end proc:
\end{verbatim}
\end{small}
%Appendix B
\section{Maple code for \texttt{`mod/rtable/Gausselim`}}
\label{modular}
\begin{small}
\begin{verbatim}
`mod/rtable/Gausselim` := proc(A::Matrix,rank::{integer,name},
                               det::{integer,name})
  local    B, d, nrows, ncols, right_margin, i,j,
      col, row, s, t, p, Q;

  Q := type( A, 'Matrix( rational )' );
  if not Q and not type( A, 'Matrix( ratpoly( algnum ) )' ) then
      error "first argument must be a matrix over a finite field"
  end if;
  ( nrows, ncols ) := LinearAlgebra:-Dimensions( A );

  p := args[-1];
  right_margin := ncols;
  if type( args[-2], 'integer' ) then
      right_margin := args[-2]
  end if;

  if nargs > 3 and type( det, 'name' ) 
               and nrows <> right_margin then
    error "cannot compute determinant of a non-square matrix"
  end if;

  if Q then
      B := map( s -> s mod p,  A );
  else
      B := map( s -> 'Normal'( s ) mod p, A );
  end if;

  row := 1; # initialise the row
  d := 1; # initialise the determinant

  for col to right_margin while row <= nrows do
      # Find the first nonzero entry in this column
      for i from row to nrows while B[ i, col ] = 0 do
          # empty
      end do;
      if i > nrows then
          # found a zero column so determinant is zero
          d := 0;
          next
      end if;
      if i <> row then
          # interchange row i with row row
          B[ i, 1..-1 ], B[ row, 1..-1 ] := 
              B[ row, 1..-1 ], B[ i, 1 .. -1 ];
          d := -d mod p; # keep track of determinant
      end if;

      if nargs = 4 then
          d := 'Normal'( d * B[ row, col ] ) mod p
      end if;

      # Make all the nonzero entries below this one 
      # are equal to 1
      if Q then
          s := 1/B[ row, col ] mod p;
          for i from row + 1 to nrows do
              # do i-th row
              if B[ i, col ] = 0 then
                  next
              end if;
              t := s * B[ i, col ] mod p;
              for j from col + 1 to ncols do
                  B[ i, j ] := B[ i, j ] - 
                               t * B[ row, j ] mod p
              end do;
              B[ i, col ] := 0
          end do
      else
          s := 'Normal'( 1 / B[ row, col ] ) mod p;
          for i from row + 1 to nrows do
              if B[ i, col ] = 0 then
                  next
              end if;
              t := 'Normal'( s * B[ i, col ] ) mod p;
              for j from col + 1 to ncols do
                B[i,j] := 'Normal'( B[ i, j ] - 
                                    t * B[ row, j ] ) mod p
              end do;
              B[ i, col ] := 0
          end do
      end if;

      row := 1 + row # go to next row
  end do; # go to next column

  if nargs > 2 and type( rank, 'name' ) then
      rank := row-1
  end if;
  if nargs > 3 and type( det, 'name' ) then
      det := d
  end if;
  B
end proc:
\end{verbatim}
\end{small}
%Appendix C
\section{Final MetaOCaml code}
\label{finalcode}
\begin{small}
\begin{verbatim}
type ('a,'c,'d) container2d = {
  get: ('a,'c) code -> ('a,int) code -> 
       ('a,int) code -> ('a,'d) code ;
  set: ('a,'c) code -> ('a,int) code -> 
       ('a,int) code -> ('a,'d) code -> ('a,unit) code;
  dim1: ('a,'c) code -> ('a,int) code;
  dim2: ('a,'c) code -> ('a,int) code;
  mapper: ('a, 'd->'d) code -> ('a,'c) code -> 
          ('a,'c) code;
  copy: ('a, 'c) code -> ('a,'c) code
} ;;

type ('a,'c,'d) state = {b:('a,'c) code; 
  r:('a,int ref) code; c:('a,int ref) code; 
  m:('a,int) code; n:('a,int) code;
  det:('a,'d ref) code; detsign:('a,'d ref) code};;

type ('a,'b) domain = {
  zero:('a,'b) code; one:('a,'b) code;
  minusone:('a,'b) code;
  plus:('a,'b) code -> ('a,'b) code -> ('a,'b) code; 
  times:('a,'b) code -> ('a,'b) code -> ('a,'b) code;
  minus:('a,'b) code -> ('a,'b) code -> ('a,'b) code; 
  div:('a,'b) code -> ('a,'b) code ->('a,'b) code;
  smaller_than:(('a,'b) code -> ('a,'b) code -> 
                ('a,bool) code) option;
  normalizerf:(('a,'b -> 'b) code ) option;
  normalizerg:(('a,'b) code -> ('a,'b) code ) option;
} ;;

type ('b,'c) outputs = 
    Matrix of 'b
  | MatrixRank of 'b * int
  | MatrixDet of 'b * 'c
  | MatrixRankDet of 'b * int * 'c ;;

type outchoice = JustMatrix | Rank | Det | RankDet;;
type dettrack = TrackNothing | TrackDet ;;

type ge_choices = 
  {fracfree:bool; track:dettrack; outputs:outchoice} ;;

let orcond_gen main_cond = function
  | Some alt_cond -> .< .~main_cond || .~alt_cond >.
  | None -> main_cond ;;

let seq a b = .< begin .~a ; .~b end >. ;;

let sapply2 g c1 c2 = match g with
  | Some f -> Some (f c1 c2)
  | None   -> None ;;

let dapply1 g c1 = match g with
  | Some f -> f c1
  | None   -> c1 ;;

let mdapply1 mapper g c1 = match g with
  | Some f -> .< .~(mapper f c1) >.
  | None   -> c1 ;;

let choose_output dom s = function
  | JustMatrix -> .< Matrix (.~s.b) >.
  | Rank       -> .< MatrixRank((.~s.b), ! .~(s.r)) >.
  | Det        -> .< MatrixDet((.~s.b), 
     .~(dom.times .<! .~(s.det)>. .<! .~(s.detsign)>. ))>.
  | RankDet    -> .<MatrixRankDet((.~s.b), 
    ! .~(s.r), .~(dom.times .<! .~(s.det)>.
    .< ! .~(s.detsign) >. )) >.

let ge_state_gen dom contr findpivot swap zerobelow choice=
  let main_loop s = 
  .< while !(.~s.c) < .~s.m && !(.~s.r) < .~s.n do
    begin
    match .~(findpivot s) with
    Some i -> begin
        if i <> !(.~s.r) then
            for j = !(.~s.c) to .~s.m-1 do
                .~(swap s .<i>. .<j>.);
            done;
        .~(zerobelow s) ;
        (.~s.r) := !(.~s.r) + 1;
        end;
    | None -> .~(match choice.track with
        | TrackNothing -> .< () >. ;
        | TrackDet     -> .< .~s.detsign := .~dom.zero >.;
        ) ;
    end;
    .~s.c := !(.~s.c) + 1;
  done >. in

  .< fun a -> 
  let b= .~(mdapply1 contr.mapper dom.normalizerf 
                     (contr.copy .<a>. )) and
      r=ref 0 and c=ref 0 and
      m= .~(contr.dim1 .<a>. ) and 
      n= .~(contr.dim2 .<a>. ) and
      det=ref .~(dom.one) and detsign=ref .~(dom.one) in
  .~(let st = {b= .<b>.; r= .<r>.; c= .<c>.; m= .<m>.; 
               n= .<n>.; det= .<det>.; 
               detsign= .<detsign>.} in
    seq
      (main_loop st )
      (choose_output dom st choice.outputs ) ) 
  >. ;;

let swapr_gen dom contr track s i j =
  let swap =
  .< let b = .~(s.b) and r = !(.~s.r) and i= .~i 
     and j= .~j in
      let t = .~(contr.get .<b>. .<i>. .<j>. ) in  begin
        .~(seq
            (contr.set .<b>. .<i>. .<j>. 
               (contr.get .<b>. .<r>. .<j>. ))
            (contr.set .<b>. .<r>. .<j>. .<t>. ) );
  end >. in
  let tds = .< .~s.detsign := 
        .~(dom.times .<! .~s.detsign>. dom.minusone) >. in
  match track with
  | TrackNothing -> swap
  | TrackDet     -> seq swap tds ;;

let findpivot_gen dom contr orcond_gen = fun s ->
  .< let i = ref (-1) in begin 
    for j = ! .~s.r to .~s.n-1 do
      if not ( .~(contr.get .< .~s.b>. .<j>. 
                 .<!( .~s.c)>. ) = .~(dom.zero)) then 
        if (.~(orcond_gen .< !i = (-1) >.
          (sapply2 dom.smaller_than 
            .< .~(contr.get .< .~s.b>. .<j>. .<! .~s.c>. )>.
            .< .~(contr.get .< .~s.b>. .<!i>. .<! .~s.c>. )>.
            ))) then
          i := j;
      done;
      if !i == -1 then None else Some !i;
  end >. ;;

let zerobelow_gen dom contr choice s =
  let inner_loop = 
  if not choice.fracfree then fun i s ->
    .< let t = .~(dom.div 
      (contr.get .<.~s.b>. .< .~i>. .<!(.~s.c)>. )
      (contr.get .<.~s.b>. .<!(.~s.r)>. .<!(.~s.c)>. ) ) in
    for j = !(.~s.c)+1 to .~s.m-1 do
      .~(contr.set .<.~s.b>. .<.~i>. .<j>. 
          (dapply1 dom.normalizerg 
          (dom.minus (contr.get .<.~s.b>. .<.~i>. .<j>.)
          (dom.times .<t>. 
              (contr.get .<.~s.b>. .<!(.~s.r)>. .<j>. )))))
    done; >.
  else fun i s -> 
    .< for j = !(.~s.c)+1 to .~s.m-1 do
      let t = .~(dapply1 dom.normalizerg (dom.minus 
        (dom.times 
          (contr.get .<.~s.b>. .<.~i>. .<j>. )
          (contr.get .<.~s.b>. .<! .~s.r>. .<! .~s.c>. ))
        (dom.times 
          (contr.get .<.~s.b>. .<!(.~s.r)>. .<j>. )
          (contr.get .<.~s.b>. .<.~i>. .<! .~s.r>. )))) in
      .~(contr.set .<.~s.b>. .<.~i>. .<j>. 
            (dom.div .<t>. .<! .~s.det>. ))
    done >. in
  let outer_loop =
  .< for i = !(.~s.r)+1 to .~s.n-1 do
    if not ( .~(contr.get .<.~s.b>. .<i>. .<! .~s.c>. ) 
             = .~dom.zero) then
      begin
        .~(inner_loop .<i>. s );
        .~(contr.set .<.~s.b>. .<i>. .<! .~s.c>. dom.zero)
      end;
  done >. in
  match choice.track with
  | TrackNothing -> outer_loop
  | TrackDet     -> seq outer_loop
    (if choice.fracfree then
      .< .~s.det := 
        .~(contr.get .<.~s.b>. .<! .~s.r>. .<! .~s.c>. )>.
    else
      .< .~s.det := .~(dom.times .<! .~s.det>. 
          (contr.get .<.~s.b>.
              .<! .~s.r>. .<! .~s.c>. ))>. );;

let specializer dom container ~fracfree ~outputs =
  let t = (if fracfree || outputs == Det 
                       || outputs == RankDet then 
      TrackDet else TrackNothing ) in
  let choice = 
      {fracfree=fracfree; track=t; outputs=outputs} in
  let fp = findpivot_gen dom container orcond_gen in
  let swap = swapr_gen dom container choice.track in
  let zb = zerobelow_gen dom container choice in
  ge_state_gen dom container fp swap zb choice;;

let dom_float = { 
   zero = .< 0. >. ; 
   one = .< 1. >. ;
   minusone = .< -1. >. ;
   plus = (fun x y -> .<.~x +. .~y>.) ;
   times = (fun x y -> .<.~x *. .~y>.) ;
   minus = (fun x y -> .<.~x -. .~y>.) ;
   div = (fun x y -> .<.~x /. .~y>.) ;
   smaller_than = 
       Some(fun x y -> .<abs_float .~x < abs_float .~y >.);
   normalizerf = None ;
   normalizerg = None };;

let array_container = {
  get = (fun x n m -> .< (.~x).(.~n).(.~m) >. );
  set = (fun x n m y -> .< (.~x).(.~n).(.~m) <- .~y >. );
  dim2 = (fun x -> .< Array.length .~x >.) ;
  dim1 = (fun x -> .< Array.length (.~x).(0) >. ) ;
  mapper = (fun f a -> .< Array.map 
      (fun x -> Array.map .~f x) .~a >.);
  copy = (fun a -> .<Array.map (fun x -> Array.copy x) 
                       (Array.copy .~a) >. )
};;

let spec_ge_float = 
    specializer dom_float array_container 
                ~fracfree:false ~outputs:RankDet ;;

let ge_float3 = .! spec_ge_float ;;
\end{verbatim}
\end{small}
\end{document}

### 2007-05-15 SCP2008/paper/metamonads.tex
move paper to here

--- 2005-06-27 paper/metamonads.tex
@@
       let-insertion, as we argue, is a control effect on its own.
-\begin{comment}
       : e.g.,\\
@@
       conditional branch.
-\end{comment}
-    \item Implementation of the |doM|-notation (patterned after
+    \item Implementation of the |perform|-notation (patterned after
       |do|-notation of Haskell) to make monadic code readable.
@@
 features: to construct a code expression, to combine them, and to
-execute them. Figure~\ref{easycode} shows the simplest code generator |one|,
-as well as more complex generators.
-
-\begin{figure}
+execute them. The following shows the simplest code generator |one|,
+and the simplest code combinators\footnote{%
+$\Longrightarrow$ under an expression shows the result of its evaluation}:
+
 \begin{code}
@@
 \evalresult{.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.}
-let simplest_param_code plus one = let gen x y = plus x (plus y one) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-let param_code1 plus one =
-  let gen x y = plus (plus y one) (plus x (plus y one)) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-let param_code1' plus one =
-  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
-  .<fun x y -> .~(gen .<x>. .<y>.)>.
-param_code1' plus one
-\evalresult{.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.}
-\end{code}
-\caption{Code generation and combinators. $\Longrightarrow$ under an
-  expression shows the result of its evaluation}\label{easycode}
-\end{figure}
+\end{code}
 
 We use MetaOCaml brackets |.<...>.| to generate code expressions,
-i.e., to construct future-stage computations. We use escapes |.~| to
-perform an immediate code generating computation \emph{while} we are
+i.e., to construct future-stage computations. MetaOCaml provides only
+one mechanism of combining code expressions, by inlining
+\oleg{splicing?} one into
+another. The power of that operation, called escape |.~|, comes from
+the fact that the expression to be spliced in (inlined) can be
+computed: escape lets us perform an arbitrary immediate code-generating
+computation \emph{while} we are
 building the future-stage computation. The immediate computation in
@@
 To see the benefit of code generation, we notice that we can easily
-parameterize our code, |simplest_param_code|, and use it to generate
-code that operates on integers, floating point numbers or booleans --
-in general, any domain that implements |plus| and |one|.
-
-The generator |param_code1| has two occurrences of |plus y one|,
+parameterize our code:
+
+\begin{code}
+let simplest_param_code plus one =
+  let gen x y = plus x (plus y one) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+and use it to generate code that operates on integers, floating point
+numbers or booleans -- in general, any domain that implements |plus|
+and |one|:
+\begin{code}
+let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
+  simplest_param_code plus one
+let plus x y = .<.~x || .~y>. and one = .<true>. in
+  simplest_param_code plus one
+\end{code}
+Running the former expression yields the function on |float|s, whereas
+the latter expression is the code expression for a boolean function.
+This clearly shows the separation of concerns, namely of that for domain
+operations.
+
+Let us consider a more complex expression:
+\begin{code}
+let param_code1 plus one =
+  let gen x y = plus (plus y one) (plus x (plus y one)) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+\end{code}
+with two occurrences of |plus y one|,
 which may be quite a complex computation and so we would rather not do
@@
 might not be able to determine if lifting them is sound. So, being
-conservative, the optimizer will leave the duplicates as they are. We
-may attempt to eliminate subexpressions as in |param_code1'|. However,
+conservative, the optimizer will leave the duplicates as they are. 
+We may attempt to eliminate subexpressions as follows: 
+\begin{code}
+let param_code1' plus one =
+  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
+  .<fun x y -> .~(gen .<x>. .<y>.)>.
+param_code1' plus one
+\evalresult{.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.}
+\end{code}
+However,
 the result of |param_code1' plus one| still exhibits duplicate
@@
 inserts the |let| expression in the generat\emph{ed} code. We need a
-combinator |letgen| to be used as\\
-|let ce = letgen (plus y one) in plus ce (plus x ce)|
-yielding the code like |.<let t = y + 1 in t + (x + t)>.| 
+combinator |letgen| to be used as
+\begin{code}
+let ce = letgen (plus y one) in plus ce (plus x ce)
+\end{code}
+yielding the code like 
+\begin{code}
+.<let t = y + 1 in t + (x + t)>.
+\end{code}
 But that seems impossible because |letgen exp| has to generate
@@
 
-\noindent Comparison of the code that did let-insertion at the generating stage\\
-|let ce = (plus y one) in  plus ce (plus x ce)|\\
-with the corresponding code inserting let at the generated code stage\\
-|letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))|\\
+\subsection{Monadic notation, making CPS code clear}\label{monadicnotation}
+
+Comparison of the code that did let-insertion at the generating stage
+\begin{code}
+let ce = (plus y one) in  plus ce (plus x ce)
+\end{code}
+with the corresponding code inserting let at the generated code stage
+\begin{code}
+letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
+\end{code}
 clearly shows the difference between  direct-style and CPS code.
-What was\\ |let ce = init in ...| in direct style became
+What was |let ce = init in ...| in direct style became
 |init' (fun ce -> ...)| in CPS. For one thing, |let| became
@@
 
-let l1 f = fun x -> doM \{ t <-- x; f t\}
-let l2 f = fun x y -> doM \{ tx <-- x; ty <-- y; f tx ty\}
+let l1 f = fun x -> perform \{ t <-- x; f t\}
+let l2 f = fun x y -> perform \{ tx <-- x; ty <-- y; f tx ty\}
 
@@
 example of the previous section as |param_code3|.
-
-\begin{figure}
 \begin{code}
@@
   in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
+\end{code}
+% param_code3 plus one;;
+%
+That does not seem like much of an improvement. With the help of
+camlp4 pre-processor, we introduce the |perform|-notation \cite{metamonadsURL},
+patterned after the |do|-notation of Haskell. 
+\begin{code}
 let param_code4 plus one =
-  let gen x y = doM \{ ce <-- retN (plus y one);
+  let gen x y = mdo \{ ce <-- retN (plus y one);
                       ret (plus ce (plus x ce)) \}
   in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
-let ifM' test th el = doM \{
+\end{code}
+The function
+|param_code4|, written in the |perform|-notation, is equivalent to
+|param_code3| -- in fact, the camlp4 preprocessor will convert the
+former into the latter. And yet, |param_code4| looks far more
+conventional, as if it were indeed in direct style.
+
+\subsection{Generating control statements}
+We can write operations that generate code other than let-statements,
+e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
+albeit straightforward, is not as general as we wish: its arguments are
+already generated pieces of code rather than monadic values. We
+``lift it'', see |ifM'|. 
+\begin{code}
+let ifM' test th el = perform \{
   testc <-- test; thc <-- th; elc <-- el;
@@
 \end{code}
-\end{figure}
-% param_code3 plus one;;
-
-That does not seem like much of an improvement. With the help of
-camlp4 pre-processor, we introduce the |doM|-notation \cite{metamonadsURL},
-patterned after the |do|-notation of Haskell. The function
-|param_code4|, written in the |doM|-notation, is equivalent to
-|param_code3| -- in fact, the camlp4 preprocessor will convert the
-former into the latter. And yet, |param_code4| looks far more
-conventional, as if it were indeed in direct style.
-
-We can write operations that generate code other than let-statements,
-e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
-albeit straightforward, is not as general as we wish: its arguments are
-already generated pieces of code rather than monadic values. We
-``lift it'', see |ifM'|. We define functions |l1|,
+%
+We define functions |l1|,
 |l2|, |l3| (analogues of |liftM|, |liftM2|, |liftM3| of Haskell) 
@@
 another |ifM| function, with the same
-interface (see Figure~\ref{ourmonad}). The difference between them is
+interface (see Figure~\ref{ourmonad}).
+The difference between them is
 apparent: in the code above with |ifM'|, the let-insertion
@@
 case-matching statements and |for|- and |while|-loops.
+\oleg{put them here?}
 
 \section{Aspects and Functors}\label{functors}
-
 The monad represents finer-scale code generation. We need tools for
@@
     \item \textbf{Augmented Matrices}: Whether all or only some
-      columns of the matrix participate in elimination. We currently
-        do not implement this aspect.
+      columns of the matrix participate in elimination.
 \end{enumerate}
@@
   module D = Det   module R = Rank   module P = DiscardPivot
-  let make_result b = doM \{ det  <-- D.fin ();  rank <-- R.fin ();
+  let make_result b = perform \{ det  <-- D.fin ();  rank <-- R.fin ();
     ret .< ( .~b, .~det, .~rank ) >. \}
@@
       let zerobelow b r c m n brc =
-        let innerbody i = doM \{
+        let innerbody i = perform \{
             bic <-- Ctr.get b i c;
@@
                       (Ctr.set b i c Dom.zero)) \} in 
-        doM \{
+        perform \{
               seqM (retLoopM (Idx.succ r) (Idx.pred n) innerbody) 
                    (Update.update_det brc) \} in
-      let dogen a = doM \{
+      let dogen a = perform \{
           r <-- Out.R.decl ();
@@
                                        (Idx.less (liftGet r) n) )
-               ( doM \{
+               ( perform \{
                rr <-- retN (liftGet r);
@@
 \bibliographystyle{plain}
+\section{Appendix 0}
+\oleg{grammar of our perform monad. Thank Lydia}
+
 \section{Appendix A}

### 2007-05-17 SCP2008/paper/metamonads.tex
change class files and adjust declaratiosn accordingly.  start editing introduction.

@@
-\documentclass{llncs}
+\documentclass[draft]{elsart}
 
@@
 \DefineShortVerb{\|}
-\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\mathindent,commandchars=\\\{\},fontsize=\small}
-\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\mathindent,commandchars=\\\{\},fontsize=\footnotesize}
+\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\small}
+\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\footnotesize}
 \newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
 % \setlength{\parskip}{0pt}
-\newlength{\mathindent}
-\setlength{\mathindent}{1em}
+\newlength{\ourmathindent}
+\setlength{\ourmathindent}{1em}
 
@@
 
-\parskip 0pt plus 1pt minus 2pt
+% \parskip 0pt plus 1pt minus 2pt
 \textfloatsep 4pt plus 2pt minus 3pt % Less space around figures
-\abovecaptionskip 0pt plus 0pt minus 2pt
-\belowcaptionskip 0pt plus 0pt minus 2pt
-
-\intextsep 2pt plus 0pt minus 1pt
+%\abovecaptionskip 0pt plus 0pt minus 2pt
+%\belowcaptionskip 0pt plus 0pt minus 2pt
+
+% \intextsep 2pt plus 0pt minus 1pt
 
@@
 
+\journal{Science of Computer Programming}
+
 \begin{document}
@@
 %code from abstract designs}
+\begin{frontmatter}
 \title{Multi-stage programming with functors and monads:
 eliminating abstraction overhead from generic code}
-\author{Jacques Carette\inst{1}
- \and
-Oleg Kiselyov\inst{2}
-}
-\institute{McMaster University,
-1280 Main St. West, Hamilton, Ontario Canada L8S 4K1
-\and
-FNMOC, Monterey, CA 93943}
-
-\maketitle
+\author{Jacques Carette\thanksref{1}}
+\address{McMaster University,
+1280 Main St. West, Hamilton, Ontario Canada L8S 4K1}
+\ead{carette@mcmaster.ca}
+\ead[url]{http://www.cas.mcmaster.ca/\textasciitilde carette}
+\author{Oleg Kiselyov}
+\address{FNMOC, Monterey, CA 93943}
+\ead{oleg@pobox.com}
+\ead[url]{http://pobox.com/\textasciitilde oleg/ftp/}
+
+\thanks[1]{Supported in part by NSERC Discovery Grant RPG262084-03.}
 
 \begin{abstract}
-With Gaussian Elimination as a representative family of numerical
-and symbolic
-algorithms, we use multi-stage programming, monads and Ocaml's
-advanced module system to demonstrate the complete elimination of the
-abstraction overhead while avoiding any inspection of the generated
-code.  We parameterize our Gaussian Elimination code to a great extent
+we use multi-stage programming, monads and Ocaml's
+advanced module system to demonstrate how to eliminate all
+abstraction overhead while avoiding any inspection of the resulting
+code.  We demonstrate this clearly with LU decomposition as a 
+representative family of symbolic and numerical algorithms, and also
+include an application to ODE solving (via a Runge-Kutta algorithm).
+We parameterize our code to a great extent
 (over domain, matrix representations, determinant tracking, 
@@
 
+\begin{keyword}
+MetaOcaml \sep linear algebra \sep genericity \sep generative \sep staging
+\sep Functor \sep symbolic.
+\end{keyword}
+% Should use either PACS or MSC scheme
+\end{frontmatter}
+
 % previous introduction is saved at the end
@@
 \emph{may} need to be addressed. For example, implementations of the
-widely used Gaussian Elimination (GE) algorithm --- the running
+widely used LU decomposition algorithm (which subsumes Gaussian 
+Elimination (GE)) --- the running
 example of our paper --- may need to account for the representation of
@@
 A survey \cite{carette04} of
-Gaussian elimination implementations in an industrial package Maple
+Gaussian elimination implementations in the industrial package Maple
 found 6 different aspects and 35 different implementations of the
 algorithm, as well as 45 implementations of directly related
-algorithms.  We can manually write each of these implementations
-optimizing for particular aspects and using cut-and-paste to ``share''
-similar pieces of code.
-We can write a very generic GE procedure that accounts for
+algorithms (LU decomposition, Cholesky decomposition, etc).  We can manually
+write each of these implementations optimizing for particular aspects and using
+cut-and-paste to ``share'' similar pieces of code.
+Or we can write a very generic procedure that accounts for
 all the aspects with appropriate abstractions \cite{Axiom}. The
@@
 generated code is well-formed. This is a challenge in string-based
-generation systems, which generally do not offer such guarantees and
+generation systems, which generally do not offer any guarantees and
 therefore make it very difficult to determine which part of the
@@
 \cite{HygienicMacros}) and ensuring the generated code is
-well-typed. Lisp-style macros, Scheme hygienic macros, camlp4
-preprocessor \cite{camlp4}, C++ template meta-programming, Template
+well-typed. Lisp-style macros, Scheme hygienic macros, the camlp4
+preprocessor \cite{camlp4}, C++ template meta-programming, and Template
 Haskell \cite{conf/dagstuhl/CzarneckiOST03} solve some of the above
@@
 MetaOCaml \cite{CTHL03,metaocaml-org}  solves all the above problems
-including the well-typing of the generated code \cite{TahaSheard97,TahaThesis}.
+including the well-typing of both the generator and 
+the generated code \cite{TahaSheard97,TahaThesis}.
 
 But more difficult problems remain. Is the generated code optimal? Do
-we still need post-processing to eliminate common subexpressions and
-fold constants, remove redundant bindings? Is the generator readable,
-resembling the original algorithm, and extensible? Are the aspects
+we still need post-processing to eliminate common subexpressions,
+fold constants, and remove redundant bindings? Is the generator readable,
+resembling the original algorithm? Is the generator extensible? Are the aspects
 truly modular? Can we add another aspect to it or another instance of
@@
 use full division when dealing with matrices of exact integers, nor is
-it worthwhile to use full pivoting on a matrix over $\mathbb Q $.
+it worthwhile to use full pivoting on a matrix over $\mathbb Q$.
 
@@
 
-In this paper we report on progress of solving these problems using GE
+In this paper, we report on our continued progress \cite{CaretteKiselyov05}
+solving these problems using linear solving and LU decomposition
 as our running example. Specifically, our contributions:
-\vspace*{-2pt}
+%\vspace*{-2pt}
 \begin{itemize}
@@
 \bibliography{metamonads}
-\bibliographystyle{plain}
+\bibliographystyle{elsart-num}
 \section{Appendix 0}

### 2007-05-17 SCP2008/paper/metamonads.tex
minor edits continued

@@
       places in the generated code.
-    \item Use functor type sharing constraints to encode domain-specific
-      knowledge.
+    \item Encode domain-specific knowledge in the generators which 
+      will catch domain-specific instantiation errors at generation
+      time.
+    \item Provide a thorough breakdown of the family of LU decomposition
+      algorithms.
 \end{itemize}
@@
 parametrized modules of OCaml to encode all of the aspects of the
-Gaussian Elimination algorithm family in completely separate,
+LU decomposition algorithm family in completely separate,
 independent modules.  We briefly discuss related work in
@@
 \section{Aspects and Functors}\label{functors}
-The monad represents finer-scale code generation. We need tools for
+The monad represents fine-scale code generation. We need tools for
 larger-scale modularization; we can use any abstraction
@@
 \begin{enumerate}
-\vspace*{-6pt}
+%\vspace*{-6pt}
     \item \textbf{Domain}: In which (algebraic) domain do the
@@
 use with MetaOCaml. Many helpful discussions with Walid Taha are very
-appreciated. We are grateful to anonymous reviewers for many
-helpful suggestions.
+appreciated.
 

### 2007-06-08 tagless/tagless-final.tex
added lost of comments, from the reviewers, etc. Go home to clean them up.

@@
 
-%\newcommand{\jacques}[1]{{\it [Jacques says: #1]}}
-%\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
-%\newcommand{\ccshan}[1]{{\it [Ken says: #1]}}
+\newcommand{\jacques}[1]{{\it [Jacques says: #1]}}
+\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
+\newcommand{\ccshan}[1]{{\it [Ken says: #1]}}
 
@@
 \begin{document}
+
+% http://flint.cs.yale.edu/aplas2007/
 
@@
 
+\oleg{Emphasize somewhere in the Intro that we solved the problem
+exactly as posed in Walid's, Xi's, and SPJ's papers. And we obtain the
+result in the form those papers declare desirable.
+We use their running example and achieve the result they
+call desirable.}
+
 A popular way to define and implement a language is to embed it in
@@
 
-We focus on embedding a typed object language into a typed metalanguage.
+We focus on embedding a \emph{typed} object language into a 
+\emph{typed} metalanguage.
 The benefit of types in this setting is to rule out meaningless object terms,
@@
 
-%\oleg{see tagless\_interp1.ml, module Tagfull for the complete code.
+% see tagless\_interp1.ml, module Tagfull for the complete code.
 % If you change the code in here, please adjust the .ml file
-%  accordingly. Let the paper and the accompanying code be in sync.}
+% accordingly. Let the paper and the accompanying code be in sync.
 
@@
 penalties.
+
+\oleg{We must stress that we do not Church-encode a universal type
+(or the solution in the previous section). The latter will not be
+patently total because we have to provide error continuatiuons and we
+have to argue why they will not be called. A compiler is mots likely
+incapable of such an argument and so has to compile them in. Here, we do
+not have any error continuations at all.}
 
@@
 
+\oleg{We must stress that: of course all parts of our solution are
+known. That was the whole point. The problem was claimed to be
+unsolvable with the existing means, and so new techniques have been
+proposed. We showed that regardless of the merits, power and convenience of
+those new techniques, the existing, old solutions were sufficient for
+the stated problem -- thus correcting the public record (of ICFP and
+POPL conferences). We don't say GADT, etc. are useless -- but we wish
+to show to what extent the existing old methods suffice.}
+\oleg{Our contribution that we make do with a very restricted polymorphism
+(HM plus the module system) which guarantees inference in ML and
+Haskell.}
+\oleg{So, our thrust is: there is public (ICFP, POPL) record of the problem
+being open and claims that solving it requires various
+extensions. We show that the problem as formulated in the literature
+can be solved with the existing means. We claim the novelty of this,
+essentially negative, result. We specifically disclaim any new
+inventions.}
+
+Again, the fact that our
+solution is elementary and uses only existing known tools to give a
+previously unknown solution to an important problem should not be
+counted as a disadvantage.
+We did not claim to invent a new method (of partial evaluation or of
+interpretation). Rather, we claim that existing, old methods, when
+properly combined, can solve problem that was publicly declared to be
+unsolvable with the existing, old means.
+Some parts of our solution have been known before; we mention that in
+the paper and reviewers point more references. However, none of the
+parts have been combined to solve the problem stated as open in
+the recent published literature Walid03 and WalidTagless.
+
 \section{The object language and its tagless interpreters}\label{language}
@@
 
-We have also implemented this compiler in Haskell. Since Haskell
+We have also implemented this compiler in Haskell. 
+\oleg{mention TH solution in footnote. ``I was rather hoping it could
+have been emulated via TemplateHaskell. I realise TH is untyped and a
+compile-time staging system. The claim would have been that the TH
+code produces will always be type correct. We could still map the
+typed ByteCode data type to untyped TH abstract syntax.''}
+Since Haskell
 has no (convenient, typed) staging facility, we had to emulate
@@
 \emph{syntactically} apparent.
+
+\oleg{Rev3:In 4.3, please carefully distinguish the limitation of GHC's
+implementation of GADTs from the type theory. The pattern match in 4.3
+is statically exhaustive, and can be shown so in other systems. If
+this code were implemented using inductive families Coq, for example,
+the user would be able to (and in fact required) to prove that the
+other cases could not occur.}
 
@@
 
+Stress typed.
+
+The new features may be deservingly
+exciting and suitable. We do not wish to stop progress. We merely wish to set
+the record straight as to which problem can be solved with which minimal and
+widely available tools.
+
+Even the idea of having multiple interpretations of the
+pervasives of a standard language (as in the Symantics structure) has
+been exploited by many (see e.g. the Lava and Hawk work, Augustsson's
+clever two-level re-interpretation of Haskell (to generate residual
+functions in Excel (!)), Rhiger's thesis, and many DSLs.
+
+Stress differences from phantom types.
+Use of phantom types is popular: see Lava 
+Lava: Hardware Design in Haskell (1998)) and Rhiger's thesis.
+(Lava: has universal type for Bit and NumSig and correspondent
+tag matching: it is not tagless).
+
+Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
+upon the Exp datatype. But that is cheating: phantom type means
+essentially we can easily do coerce. We use real types.  That's why he
+had to do tedious proofs in Sec 2 of soundness and completeness of
+embedding. Whereas our proofs are obvious.  His sec 3 is based on data
+representation of terms. They have type tags.  We do nothing of that
+kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
+type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
+cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
+thesis contains the definitions of the interpreter and the compiler,
+in the beginning. Use this in response to Rev1)
+OR:
+The simplicity of our encoding enables trivial proofs of
+soundness and completeness compared to Rhiger's.  It also makes
+self-interpretation simple, which is a strength of our work.
+
+first approach that gives typed CPS in a HM language. We also need to
+cite the paper by Stefan Monnier at PLPV2006 where they say, I think,
+that typed CPS is impossible without GADT or like.
+
+the basic idea of this paper (the encoding of the term
+GADT as its elimination form, i.e. as a coalgebraic structure) is not
+new to this work. It appeared many years ago in: Metacircularity in
+the polymorphic lambda-calculus, Frank Pfenning and Peter Lee.
+Theoretical Computer Science 89 (1991) 137-159. (Other encodings of
+GADTs have appeared since Pfenning and Lee's work, for example, in
+work on generic programming.)
+BTW,
+many GADT encodings I've seen (Sulzmann's et al) require existentials,
+which are not available in ML. The only GADT encodings I've seen in ML
+use unsafe operations.
+
+It seems that Pfenning and Lee embed $F_2$ with type constructions in 
+(pure) $F_3$.  We embed $F_1$ in (weak?) $F_2$, as I see it.  In a way, what 
+we do is very similar to what they do (Figure 1, p.152), except that we 
+do it in standard programming languges.  It is unclear if their work can 
+be implemented (yet) in any language.  And we preserve type-inference, 
+while their solution needs explcit types!
+The following line of their conclusion is worth citing: "... this does 
+not imply that the same language is also suitable for type 
+metaprogramming. ... such as partial evaluation... do not seem to be 
+expressible".
+I suspect you're right, but I'm still reading the paper.  See also page
+146: "for a term M in $F_1$ (a simply-typed term), the representation
+$\bar{M}$ will be in $F_2$".  The move from $F_2$ to $F_3$ 
+and beyond reminds me
+strongly of our attempts at self-interpretation without the notion of a
+syntactic hole.
+    Pfenning and Lee conclude that partial evaluation and program
+    transformations "do not seem to be expressible" even using their
+    extension to F-omega, perhaps because their avoidance of general
+    recursive types compels them to include the polymorphic lift that we
+    avoid in our Section 4.1.
+
+ Although GADTs
+may indeed be "more flexible and easier to use", they are not
+available in mainstream ML, and GHC as released does not implement
+them correctly (as Simon Peyton-Jones admits publicly).  Whereas
+Pasalic et al. have already solved the problem of tagless typed
+(staged) interpretation, our solution is the first to use only
+facilities available in mainstream ML and Haskell, namely
+Hindley-Milner and either simple modules or constructor classes.  The
+simplicity of these facilities is again a strength of our work: it
+makes embedded DSLs more practical to build.
+
+Type-indexed types have been around since at least Harper and
+Morrisett's Compiling Polymorphism Using Intensional Type Analysis,
+POPL 1995. They also were supported by Generic Haskell (Hinze,
+Juering, Loeh, Type Indexed Datatypes, SCP 2004).
+Generic Haskell is not mainstream.  Xi et al. point out why
+intensional type analysis cannot handle HOAS with fix.
+
 Our initial motivation came from several papers 
@@
 
+\oleg{From Rev3}This representation limited to primitive
+recursive folds over the datatype (as that is the elimination form
+provided.) That means that some operations are easy implement
+(interpretation, size calculation, etc.), some can be done with effort
+(one-pass cps, and other mutually recursive functions), and perhaps
+some cannot be implemented at all (those that are not primitive
+recursive). For example, can you implement a transformation that
+eliminates eta-expansions from a term?
+
 \acks%\subsection*{Acknowledgments}
@@
 kindly provided some useful references on coalgebraic structures in the
-$\lambda$\hyp calculus.
+$\lambda$\hyp calculus. We thank anonymous reviewers for pointing out
+related work.
 

### 2007-06-08 tagless/tagless-final.tex
Retarget for APLAS and JFP in parallel (the text is completely the same)

@@
-\listfiles
-\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar type-case}
-\documentclass[preprint]{sigplanconf}
-\usepackage{amsmath,amssymb}
-\usepackage{mathptmx}
-%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
-\usepackage[T1]{fontenc}
-\usepackage[override]{cmtt}
-\usepackage{comment}
-\usepackage{hyphenat}
-\usepackage{url}
-\usepackage{prooftree1} \proofrulebaseline=1.5ex \proofdotseparation=.75ex
-\usepackage{mdwlist}
-\usepackage{mdwtab}
-\usepackage{dblfloatfix}
-
-% Set pdftex paper size
-\ifx\pdfoutput\undefined\else \ifx\pdfoutput\relax\else \ifcase\pdfoutput\else
-    \setlength{\pdfpagewidth}{8.5in}
-    \setlength{\pdfpageheight}{11in}
-\fi \fi \fi
-
-\newcommand{\jacques}[1]{{\it [Jacques says: #1]}}
-\newcommand{\oleg}[1]{{\it [Oleg says: #1]}}
-\newcommand{\ccshan}[1]{{\it [Ken says: #1]}}
-
-\usepackage{natbib1}
-\let\cite=\citep
-
-\usepackage[compact]{fancyvrb1}
-\DefineShortVerb{\|}
-\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\parindent}
-
-\newtheorem{prop}{Proposition}
-
-\newcommand{\ZZ}{\mathbb{Z}}
-\newcommand{\BB}{\mathbb{B}}
-
-\newcommand{\fun}[1]{\mathopen{\lambda\mathord{#1}.\,}}
-\newcommand{\fix}[1]{\mathopen{\mathrm{fix\,}\mathord{#1}.\,}}
-\newcommand{\Forall}[1]{\mathopen{\forall\mathord{#1}.\,}}
-\newcommand{\Exists}[1]{\mathopen{\exists\mathord{#1}.\,}}
-\newcommand{\cond}[3]{\mathrm{if\ }#1\mathrm{\ then\ }#2\mathrm{\ else\ }#3}
-\newcommand{\be}[1]{\mathrm{let\ }#1\mathrm{\ in\ }}
-\newcommand{\True}{\mathinner{\mathrm{true}}}
-\newcommand{\False}{\mathinner{\mathrm{false}}}
-\newcommand{\Encode}[1]{\mathinner{\mathopen{\text{``}}#1\mathclose{\text{''}}}}
-\newcommand{\encode}[1]{\mathinner{\mathopen{\text{`}}#1\mathclose{\text{'}}}}
-\newcommand{\ident}[1]{\mathinner{\text{\textit{#1\/}}}}
-\newcommand{\subst}[3]{#1\left\{#2\mapsto#3\right\}}
-\DeclareMathOperator{\pe}{PE}
-\DeclareMathOperator{\si}{SI}
-
-% Make space around section headings no less than space around theorems
-\makeatletter
-\@tempdima\topsep \advance\@tempdima\parsep
-\ifdim\@sectionbelowskip   <\topsep \@sectionbelowskip   \topsep \fi
-\ifdim\@subsectionbelowskip<\topsep \@subsectionbelowskip\topsep \fi
-\makeatother
-
-% Allow things to flow a little more tightly around figures
-\renewcommand\floatpagefraction{.95}
-\renewcommand\topfraction{.95}
-\renewcommand\bottomfraction{.95}
-\renewcommand\textfraction{.05}
-
-% Match the rule inserted by \@makecaption in sigplanconf.cls
-\setproofrulebreadth .33pt
-\newenvironment{floatrule}
-    {\hrule width \hsize height .33pt \vspace{.5pc}}
-    {\par\addvspace{1ex}}
-
-% Make some Rel symbols into Bin symbols instead
-\DeclareMathSymbol{\to}{\mathbin}{symbols}{"21}
-\DeclareMathSymbol{:}{\mathbin}{operators}{"3A}
-
-\begin{document}
-
-% http://flint.cs.yale.edu/aplas2007/
-
-\conferenceinfo{ICFP '07}{Freiburg, Germany} 
-\copyrightyear{2007} 
-\copyrightdata{[to be supplied]}
-
-\makeatletter
-\titlebanner{Submitted to \@conferencename}        % These are ignored unless
-\preprintfooter{\rlap{\@titletext}}   % 'preprint' option specified.
-\def \@formatyear {\llap{\number\year/\number\month/\number\day}}
-\makeatother
-
-\title{Finally Tagless, Partially Evaluated}
-\subtitle{Tagless Staged Interpreters for Simpler Typed Languages}
-%Staged typed type-preserving tagless ``interpreters'', \emph{finally}
-
 \begin{comment}
@@
 \end{comment}
-\authorinfo
-{\hspace*{-15em}\smash{\normalsize\it\begin{minipage}[t]{30em}
-    It should also be possible to define languages with a highly
-    refined syntactic type structure. Ideally, such a treatment should
-    be metacircular, in the sense that the type structure used in the
-    defined language should be adequate for the defining language.
-    \hfill\rm\citep{reynolds-definitional}
-\end{minipage}}\hspace*{-15em}}{}{}
-
-\maketitle
 
@@
 \end{abstract}
+
+\begin{quote}
+    It should also be possible to define languages with a highly
+    refined syntactic type structure. Ideally, such a treatment should
+    be metacircular, in the sense that the type structure used in the
+    defined language should be adequate for the defining language.
+    \hfill\rm\citep{reynolds-definitional}
+\end{quote}
 
@@
 sense.
-\begin{prop}
+\begin{proposition}
 If an object term has the object type~$t$, then its
@@
 |forall repr.| |Symantics repr => repr |$t$.
-\end{prop}
+\end{proposition}
 Conversely, the type system of the metalanguage statically checks that the
@@
 due to interpreting |fix|.
-\begin{prop}
+\begin{proposition}
 If an object term~$e$ encoded in the metalanguage has type~$t$,
@@
 indefinitely or terminates with a value of the same type~$t$.
-\end{prop}
+\end{proposition}
 Generalizing from~|R| to all interpreters, we have the following
 broader and more useful, if also more obvious, proposition.
-\begin{prop}
+\begin{proposition}
   If an implementation of |Symantics| never gets stuck, then
@@
   that implementation.
-\end{prop}
+\end{proposition}
 These propositions follow immediately from the soundness of the
@@
 a self\hyp interpreter in the following sense.
-\begin{prop}
+\begin{proposition}
     $\si[\encode{e}]$ is observationally equivalent to~$e$.
-\end{prop}
+\end{proposition}
 As a corollary, we can pre-encode the self\hyp interpreter itself as
@@
 the self\hyp interpreter~$\si$.
-\begin{prop}
+\begin{proposition}
     Let $\pe$ be the partial evaluator~|P| in~\S\ref{S:PE-solution}.
@@
     equal up to $\alpha$\hyp conversion.
-\end{prop}
+\end{proposition}
 
@@
 
-\acks%\subsection*{Acknowledgments}
-We thank Martin Sulzmann and Walid Taha 
-for helpful discussions.  Sam Staton, Pieter Hofstra, and Bart Jacobs
-kindly provided some useful references on coalgebraic structures in the
-$\lambda$\hyp calculus. We thank anonymous reviewers for pointing out
-related work.
+\section*{Acknowledgments}
+\ourthanks
 

### 2007-06-09 tagless/tagless-final.tex
Improved according to teh comments from the reviewers

@@
 \section{Introduction}\label{intro}
-
-\oleg{Emphasize somewhere in the Intro that we solved the problem
-exactly as posed in Walid's, Xi's, and SPJ's papers. And we obtain the
-result in the form those papers declare desirable.
-We use their running example and achieve the result they
-call desirable.}
 
@@
 
-In sum, by Church\hyp encoding terms using ordinary functions,
-we achieve a tagless evaluator
-for a typed object language in a metalanguage with a simple type
-system \citep{hindley-principal,milner-theory}.  In this \emph{final}
-rather than \emph{initial} approach, both kinds of run-time errors
-in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
-term) are reported at compile time. 
-Because the new interpreter uses no
-universal type or
-pattern matching, it never results in a run-time error (and is in fact
-total).  Because this safety is obvious not just to us but also to the
-metalanguage implementation, we avoid the serious performance penalty
-\citep{WalidICFP02} of error checking.  \Citet{Gluck-jones-optimality}
-explains deeper technical reasons that inevitably lead to these performance
-penalties.
-
-\oleg{We must stress that we do not Church-encode a universal type
-(or the solution in the previous section). The latter will not be
-patently total because we have to provide error continuatiuons and we
-have to argue why they will not be called. A compiler is mots likely
-incapable of such an argument and so has to compile them in. Here, we do
-not have any error continuations at all.}
+In sum, by Church\hyp encoding terms using ordinary functions, we
+achieve a tagless evaluator for a typed object language in a
+metalanguage with a simple type system
+\citep{hindley-principal,milner-theory}.  In this \emph{final} rather
+than \emph{initial} approach, both kinds of run-time errors in
+\S\ref{tagproblem} (applying a nonfunction and evaluating an open
+term) are reported at compile time. Because the new interpreter
+uses no universal type or pattern matching, it never results in a
+run-time error (and is in fact total).  Because this safety is obvious
+not just to us but also to the metalanguage implementation, we avoid
+the serious performance penalty \citep{WalidICFP02} of error checking.
+\Citet{Gluck-jones-optimality} explains deeper technical reasons that
+inevitably lead to these performance penalties.
+
+Our solution is emphatically distinct from Church-encoding of the
+universal type. Church-encoding of the type |u|, \S\ref{tagproblem},
+requires two continuations; the interpreter like |app| above would
+have to provide both to the encoding of |e1|. The continuation
+corresponding to the |UB| branch of |u| must either raise an error or
+loop. For a well-typed object term that error continuation is never
+invoked, yet it must be supplied. Our interpreter has no error
+continuations at all.
 
@@
 \citet{metaocaml}) and standard Haskell. The complete code is available%
-\footnote{The URL will appear in the final version of the paper.}
-% connot give URL in the ICFP paper due to double-blind reviews
+\footnote{\oleg{put URL!}.}
 to supplement the paper. Our examples below switch between (Meta)OCaml
@@
 
-Our contributions are as follows.
+We take the problem of tagless (staged) typed-preserving
+interpretation exactly as it was posed in \citet{WalidICFP02} and
+\citet{xi-guarded}.  We use their running examples and achieve the
+result they call desirable.  Our contributions are as follows.
+
 \begin{enumerate*}
@@
 
-\oleg{We must stress that: of course all parts of our solution are
-known. That was the whole point. The problem was claimed to be
-unsolvable with the existing means, and so new techniques have been
-proposed. We showed that regardless of the merits, power and convenience of
-those new techniques, the existing, old solutions were sufficient for
-the stated problem -- thus correcting the public record (of ICFP and
-POPL conferences). We don't say GADT, etc. are useless -- but we wish
-to show to what extent the existing old methods suffice.}
-\oleg{Our contribution that we make do with a very restricted polymorphism
-(HM plus the module system) which guarantees inference in ML and
-Haskell.}
-\oleg{So, our thrust is: there is public (ICFP, POPL) record of the problem
-being open and claims that solving it requires various
-extensions. We show that the problem as formulated in the literature
-can be solved with the existing means. We claim the novelty of this,
-essentially negative, result. We specifically disclaim any new
-inventions.}
-
-Again, the fact that our
-solution is elementary and uses only existing known tools to give a
-previously unknown solution to an important problem should not be
-counted as a disadvantage.
-We did not claim to invent a new method (of partial evaluation or of
-interpretation). Rather, we claim that existing, old methods, when
-properly combined, can solve problem that was publicly declared to be
-unsolvable with the existing, old means.
-Some parts of our solution have been known before; we mention that in
-the paper and reviewers point more references. However, none of the
-parts have been combined to solve the problem stated as open in
-the recent published literature Walid03 and WalidTagless.
+We reiterate that we specifically do not propose any new language
+feature or technique. We use the existing, known approaches --
+Hindley-Milner type system with inference-preserving module system or
+constructor classes extensions, as realized in ML or Haskell98 -- to
+solve the problem that was stated in published record as lacking
+solution in ML or Haskell98 and most likely unimplementable in these
+languages without extensions. The simplicity of our solution and our
+use of only mainstream functional language features makes typed
+embedded DSLs more practical to build.
+\begin{comment}
+We may claim some contribution about type-preserving CPS. Check
+related work section in \citet{Guillemette-Monier-PLPV}, especially
+check the work of Shao on type-preserving CPS in Flint. The PLPV paper
+in related work shows other tasks, including closure conversion, which
+we may tackle in our approach. We may be able to write a
+type-preserving, assured compiler, whose properties and assured by HM.
+\end{comment}
 
@@
 Comparing |Symantics| with Figure~\ref{fig:object}
-shows how to represent every typed, closed object term in the
+shows how to represent \emph{every} typed, closed object term in the
 metalanguage. Moreover, the representation preserves types, in the following
@@
 We have also implemented this compiler in Haskell. 
-\oleg{mention TH solution in footnote. ``I was rather hoping it could
-have been emulated via TemplateHaskell. I realise TH is untyped and a
-compile-time staging system. The claim would have been that the TH
-code produces will always be type correct. We could still map the
-typed ByteCode data type to untyped TH abstract syntax.''}
 Since Haskell
@@
 it. To be precise, we defined a data type |ByteCode| with
-constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
+constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|%
+\footnote{\texttt{ByteCode} can be mapped to AST of Template Haskell
+  (TH). The output of our compiler will then be assuredly type-correct TH.}
 Whereas our representation of object terms uses HOAS,
@@
 construct values of type |P (a -> b)| as required by the type of |app|.
-Indeed the metalanguage implementation could reason thus, but it may not; GHC for one
-issues warnings.  Although this point may seem minor, it is the heart of
+Indeed the metalanguage implementation could reason thus, 
+but it may not; GHC for one issues warnings%
+\footnote{If we use inductive families (as those in Coq) or logical
+  frameworks with canonical forms, we can prove the pattern-matching
+  to be exhaustive. The coverage checker of Twelf does such proof.}. 
+Although this point may seem minor, it is the heart of
 the tagging problem and the purpose of tag elimination. A typed tagged
@@
 \emph{syntactically} apparent.
-
-\oleg{Rev3:In 4.3, please carefully distinguish the limitation of GHC's
-implementation of GADTs from the type theory. The pattern match in 4.3
-is statically exhaustive, and can be shown so in other systems. If
-this code were implemented using inductive families Coq, for example,
-the user would be able to (and in fact required) to prove that the
-other cases could not occur.}
 
@@
 
-Stress typed.
-
-The new features may be deservingly
-exciting and suitable. We do not wish to stop progress. We merely wish to set
-the record straight as to which problem can be solved with which minimal and
-widely available tools.
-
-Even the idea of having multiple interpretations of the
-pervasives of a standard language (as in the Symantics structure) has
-been exploited by many (see e.g. the Lava and Hawk work, Augustsson's
-clever two-level re-interpretation of Haskell (to generate residual
-functions in Excel (!)), Rhiger's thesis, and many DSLs.
-
-Stress differences from phantom types.
-Use of phantom types is popular: see Lava 
-Lava: Hardware Design in Haskell (1998)) and Rhiger's thesis.
-(Lava: has universal type for Bit and NumSig and correspondent
-tag matching: it is not tagless).
-
-Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
-upon the Exp datatype. But that is cheating: phantom type means
-essentially we can easily do coerce. We use real types.  That's why he
-had to do tedious proofs in Sec 2 of soundness and completeness of
-embedding. Whereas our proofs are obvious.  His sec 3 is based on data
-representation of terms. They have type tags.  We do nothing of that
-kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
-type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
-cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
-thesis contains the definitions of the interpreter and the compiler,
-in the beginning. Use this in response to Rev1)
-OR:
-The simplicity of our encoding enables trivial proofs of
-soundness and completeness compared to Rhiger's.  It also makes
-self-interpretation simple, which is a strength of our work.
-
-first approach that gives typed CPS in a HM language. We also need to
-cite the paper by Stefan Monnier at PLPV2006 where they say, I think,
-that typed CPS is impossible without GADT or like.
-
-the basic idea of this paper (the encoding of the term
-GADT as its elimination form, i.e. as a coalgebraic structure) is not
-new to this work. It appeared many years ago in: Metacircularity in
-the polymorphic lambda-calculus, Frank Pfenning and Peter Lee.
-Theoretical Computer Science 89 (1991) 137-159. (Other encodings of
-GADTs have appeared since Pfenning and Lee's work, for example, in
-work on generic programming.)
-BTW,
-many GADT encodings I've seen (Sulzmann's et al) require existentials,
-which are not available in ML. The only GADT encodings I've seen in ML
-use unsafe operations.
-
-It seems that Pfenning and Lee embed $F_2$ with type constructions in 
-(pure) $F_3$.  We embed $F_1$ in (weak?) $F_2$, as I see it.  In a way, what 
-we do is very similar to what they do (Figure 1, p.152), except that we 
-do it in standard programming languges.  It is unclear if their work can 
-be implemented (yet) in any language.  And we preserve type-inference, 
-while their solution needs explcit types!
-The following line of their conclusion is worth citing: "... this does 
-not imply that the same language is also suitable for type 
-metaprogramming. ... such as partial evaluation... do not seem to be 
-expressible".
-I suspect you're right, but I'm still reading the paper.  See also page
-146: "for a term M in $F_1$ (a simply-typed term), the representation
-$\bar{M}$ will be in $F_2$".  The move from $F_2$ to $F_3$ 
-and beyond reminds me
-strongly of our attempts at self-interpretation without the notion of a
-syntactic hole.
-    Pfenning and Lee conclude that partial evaluation and program
-    transformations "do not seem to be expressible" even using their
-    extension to F-omega, perhaps because their avoidance of general
-    recursive types compels them to include the polymorphic lift that we
-    avoid in our Section 4.1.
-
- Although GADTs
-may indeed be "more flexible and easier to use", they are not
-available in mainstream ML, and GHC as released does not implement
-them correctly (as Simon Peyton-Jones admits publicly).  Whereas
-Pasalic et al. have already solved the problem of tagless typed
-(staged) interpretation, our solution is the first to use only
-facilities available in mainstream ML and Haskell, namely
-Hindley-Milner and either simple modules or constructor classes.  The
-simplicity of these facilities is again a strength of our work: it
-makes embedded DSLs more practical to build.
-
-Type-indexed types have been around since at least Harper and
-Morrisett's Compiling Polymorphism Using Intensional Type Analysis,
-POPL 1995. They also were supported by Generic Haskell (Hinze,
-Juering, Loeh, Type Indexed Datatypes, SCP 2004).
-Generic Haskell is not mainstream.  Xi et al. point out why
-intensional type analysis cannot handle HOAS with fix.
-
 Our initial motivation came from several papers
 \citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} that used
-embedded interpreters to justify
-advanced type systems.  While we admire all this technical work,
-we were not convinced that their 
-motivating example actually needed all this machinery.
+embedded interpreters to justify advanced type systems, in particular,
+GADT. While we admire all this technical work, we were not convinced
+that their motivating example actually needed all this machinery.
+Although GADTs may indeed be more flexible and easier to use, they are
+not available in mainstream ML, and their current implementation in
+GHC has problems. We were also interested in discovering the minimal and
+widely available set of language features necessary for tagless
+type-preserving interpretation.
 
@@
 
-Our final approach seems related to \citearound{'s approach to HOAS via
-catamorphism and anamorphism}\citet{Washburn-Weirich-boxes}.
-
 A lot of effort has gone into ``typing dynamic typing'': to statically
@@
 
+The idea of multiple interpretations of the host language pervasives
+such as addition or application is quite common in implementing
+embedded DSL. It is also common to use phantom types to prevent
+forming of ill-typed DSL terms \citet{Lava,Rhiger-thesis}. However,
+the universal type still remains (|Bit| and |NumSig| in Lava
+\citet{Lava}, |Raw| and |Term| in \citet[Fig 2.2,
+  Sec. 3]{Rhiger-thesis}), along with the attendant pattern-matching
+overhead. These approaches are not tagless.  The universal type also
+greatly complicates the soundness and completeness proofs of embedding
+\citet{Rhiger-thesis}. In contrast, our proofs become trivial.
+The approach of \Citet[Sec 3.3.4]{Rhiger-thesis} does not 
+support typed CPS transformation.
+\begin{comment}
+Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
+upon the Exp datatype. But that is cheating: phantom type means
+essentially we can easily do coerce. We use real types.  That's why he
+had to do tedious proofs in Sec 2 of soundness and completeness of
+embedding. Whereas our proofs are obvious.  His sec 3 is based on data
+representation of terms. They have type tags.  We do nothing of that
+kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
+type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
+cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
+thesis contains the definitions of the interpreter and the compiler,
+in the beginning. Use this in response to Rev1)
+\end{comment}
+
 We are not the first to implement a typed interpreter for a typed
@@
 in~\S\ref{S:PE-solution}).
+
+Our approach is based on encoding a term in its elimination form, as a
+co-algebraic structure. That basic idea and its application to
+meta-circular interpertation first appeared in \citet{Pfenning-Lee}.
+Our approach however is implementable in mainstream ML and supports
+type inference and typed CPS and partial evaluation. In contrast,
+\citet{Pfenning-Lee} conclude that partial evaluation and program
+transformations "do not seem to be expressible" even using their
+extension to $F_{\omega}$, perhaps because their avoidance of general
+recursive types compels them to include the polymorphic lift that we
+avoid in \S\ref{S:PE-lift}.
+\begin{comment}
+It seems that Pfenning and Lee embed $F_2$ with type constructions in 
+(pure) $F_3$.  We embed $F_1$ in (weak?) $F_2$, as I see it.  In a way, what 
+we do is very similar to what they do (Figure 1, p.152), except that we 
+do it in standard programming languges.  It is unclear if their work can 
+be implemented (yet) in any language.  And we preserve type-inference, 
+while their solution needs explcit types!
+The following line of their conclusion is worth citing: "... this does 
+not imply that the same language is also suitable for type 
+metaprogramming. ... such as partial evaluation... do not seem to be 
+expressible".
+I suspect you're right, but I'm still reading the paper.  See also page
+146: "for a term M in $F_1$ (a simply-typed term), the representation
+$\bar{M}$ will be in $F_2$".  The move from $F_2$ to $F_3$ 
+and beyond reminds me
+strongly of our attempts at self-interpretation without the notion of a
+syntactic hole.
+\end{comment}
+
+Our encoding of the type function |repr_pe| in \S\ref{S:PE-solution}
+may be regarded as an emulation of type-indexed types and related to
+the intensional type analysis
+\citet{Morrisett-intensional,Generic-Haskell}. Our running HOAS
+example however, includes |fix|, which cannot be handled by the
+intensional type analysis \citet{xi-guarded}.  Our final approach
+seems related to \citearound{'s approach to HOAS via catamorphism and
+  anamorphism}\citet{Washburn-Weirich-boxes}.
 
@@
 
-\oleg{From Rev3}This representation limited to primitive
-recursive folds over the datatype (as that is the elimination form
-provided.) That means that some operations are easy implement
-(interpretation, size calculation, etc.), some can be done with effort
-(one-pass cps, and other mutually recursive functions), and perhaps
-some cannot be implemented at all (those that are not primitive
-recursive). For example, can you implement a transformation that
-eliminates eta-expansions from a term?
+Our representation of object terms in the elimination form encodes
+primitive recursive folds over the terms. This makes operations like
+interpretation and size calculations trivial to implement. It remains
+the subject of the current work to see if and how non-primitively 
+recursive operations can be supported in our approach.
 

### 2007-06-09 tagless/tagless-final.tex
lots of cutting for APLAS

@@
 We leave aside the solved problem of writing such a parser\slash type\hyp checker,
-whether using dependent types \citep{WalidICFP02} or not \citep{baars-typing}.
+whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}.
 
@@
 
-It is straightforward to create an algebraic data type, say in OCaml, to
-represent object terms such as those in Figure~\ref{fig:object}.
-For brevity, we elide treating integers, conditionals, and fixpoint in
-this section.
+\begin{figure}
+%
+(a)\vspace{-10pt}
 \begin{code}
 type var = VZ | VS of var
-type exp = V of var | B of bool
-         | L of exp | A of exp * exp
-\end{code}
-We represent each variable using a unary de Bruijn index.
-For example, we represent the object term $(\fun{x}x)\True$ as
+type exp = V of var | B of bool | L of exp | A of exp * exp
+\end{code}
+%
+(b)\vspace{-10pt}
 \begin{code}
@@
 \end{code}
-
-Following \citet{WalidICFP02},
-we try to implement an interpreter function |eval0|. It takes
-an object term such as |test1| above and gives us its value.
-The first argument to |eval0| is the environment, initially empty,
-which is the list of values bound to free variables in the
-interpreted code.
+%
+(c)\vspace{-10pt}
 \begin{code}
@@
 \end{code}
-
-If our OCaml-like metalanguage were untyped, the code above would be acceptable.
+%
+(d)\vspace{-10pt}
+\begin{code}
+type u = UB of bool | UA of (u -> u)
+\end{code}
+%
+(e)\vspace{-10pt}
+\begin{code}
+let rec eval env = function
+| V v       -> lookup env v
+| B b       -> UB b
+| L e       -> UA (fun x -> eval (x::env) e)
+| A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
+\end{code}
+%
+(f)\vspace{-10pt}
+\begin{code}
+let test1r = eval [] test1
+val test1r : u = UB true 
+\end{code}
+%
+(g)\vspace{-10pt}
+\begin{code}
+let test2 = A (B true, B false)
+let test2r = eval [] test2
+Exception: Match_failure in eval
+\end{code}
+%
+(h)\vspace{-10pt}
+\begin{code}
+let test3 = A (L (V (VS VZ)), B true)
+let test3r = eval [] test3
+Exception: Match_failure in lookup
+\end{code}
+%
+\caption{OCaml code illustrating the tag problem.}
+\label{fig:tag-problem}
+\end{figure}
+It is straightforward to create an algebraic data type, say in OCaml%
+%aplas-specific
+Fig.~\ref{fig:tag-problem}(a)%
+, to
+represent object terms such as those in Figure~\ref{fig:object}.
+% fig:tag-problem part (a) was here
+For brevity, we elide treating integers, conditionals, and fixpoint in
+this section.
+We represent each variable using a unary de Bruijn index.
+For example, we represent the object term $(\fun{x}x)\True$ as
+%aplas-specific
+in Fig.~\ref{fig:tag-problem}(b)%
+% fig:tag-problem part (b) was here
+
+Following \cite{WalidICFP02},
+we try to implement an interpreter function |eval0|%
+%aplas-specific
+, Fig.~\ref{fig:tag-problem}(c)%
+. It takes
+an object term such as |test1| above and gives us its value.
+The first argument to |eval0| is the environment, initially empty,
+which is the list of values bound to free variables in the
+interpreted code.
+% fig:tag-problem part (c) was here
+If our OCaml-like metalanguage were untyped, the code above would be 
+acceptable.
 The |L e| line exhibits interpretive overhead:
@@
 The usual solution is to introduce a universal type \citep[\S1.3]
-{WalidICFP02} containing both booleans and functions.
-\begin{code}
-type u = UB of bool | UA of (u -> u)
-\end{code}
-We can then write a typed interpreter
-\begin{code}
-let rec eval env = function
-| V v       -> lookup env v
-| B b       -> UB b
-| L e       -> UA (fun x -> eval (x::env) e)
-| A (e1,e2) -> match eval env e1 with UA f ->
-               f (eval env e2)
-\end{code}
+{WalidICFP02} containing both booleans and functions%
+%aplas-specific
+, Fig.~\ref{fig:tag-problem}(d)%
+.
+% fig:tag-problem part (d) was here
+We can then write a typed interpreter%
+%aplas-specific
+, Fig.~\ref{fig:tag-problem}(e)%
+% fig:tag-problem part (e) was here
 whose inferred type is |u list -> exp -> u|. Now we can evaluate
-our sample |test1| term.
-\begin{code}
-let test1r = eval [] test1
-val test1r : u = UB true 
-\end{code}
-
+our sample |test1| term%
+%aplas-specific
+, Fig.~\ref{fig:tag-problem}(f)%
+.
+% fig:tag-problem part (f) was here
 The unfortunate tag |UB| in the result reflects that |eval| is a partial
 function.  First, the pattern match |with UA f| in the line
-|A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
+|A (e1,e2)| is not exhaustive, so |eval| can fail%
+%aplas-specific
+, Fig.~\ref{fig:tag-problem}(g), 
+if we apply a boolean,
 as in the ill-typed term |A (B true, B false)|.
-\begin{code}
-let test2 = A (B true, B false)
-let test2r = eval [] test2
-Exception: Match_failure in eval
-\end{code}
+% fig:tag-problem part (g) was here
 Second, the |lookup|
 function assumes a nonempty environment, so |eval| can fail if we
-evaluate an open term.
-\begin{code}
-let test3 = A (L (V (VS VZ)), B true)
-let test3r = eval [] test3
-Exception: Match_failure in lookup
-\end{code}
+evaluate an open term%
+%aplas-specific
+, Fig.~\ref{fig:tag-problem}(h)%
+.
+% fig:tag-problem part (h) was here
 After all, the type |exp| represents object
@@
 nonexhaustive pattern matching in |lookup| and |eval| and the tags |UB|
-and |UA| \citep[\S1.4]{WalidICFP02}.  In other words, the algebraic data
+and |UA| \cite[\S1.4]{WalidICFP02}.  In other words, the algebraic data
 types above fail to express in the metalanguage that the object program
@@
 pattern\hyp matching operations that incur a performance penalty in
-interpretation \citep{WalidICFP02} and impair optimality in partial evaluation
-\citep{taha-tag}.  In short, the universal\hyp type solution is
+interpretation \cite{WalidICFP02} and impair optimality in partial evaluation
+\cite{taha-tag}.  In short, the universal\hyp type solution is
 unsatisfactory because it does not preserve typing.
@@
 GADTs or dependent types \citep{taha-tag}.  In fact, this problem
-motivated much work on GADTs \citep{xi-guarded,peyton-jones-simple} and
-on dependent types \citep{WalidICFP02,fogarty-concoqtion}.
+motivated much work on GADTs \cite{xi-guarded,peyton-jones-simple} and
+on dependent types \cite{WalidICFP02,fogarty-concoqtion}.
 
@@
 terms like |test3|
-\citep{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual},
+%aplas \cite{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual},
+\cite{WalidPOPL03,NanevskiJFP05,DaviesJACM01},
 so that |lookup| never receives an empty environment.  We discuss these
@@
 We represent object programs using ordinary functions rather than
-data constructors.  The interpreter below provides these functions.
+data constructors.  The interpreter below provides these functions%
+%aplas-specific
+, Fig.~\ref{fig:tag-final-proposal}(a)%
+.
+% fig:tag-final-proposal part (a) was here
+These functions comprise the entire interpreter.
+We now represent our sample term $(\fun{x}x)\True$ as
+%aplas-specific
+Fig.~\ref{fig:tag-final-proposal}(b)%
+.
+% fig:tag-final-proposal part (b) was here
+This representation is almost the same as in \S\ref{tagproblem}, only
+written with lowercase identifiers. To evaluate an object term is to
+apply its representation to the empty environment%
+%aplas-specific
+, Fig.~\ref{fig:tag-final-proposal}(c)%
+.
+% fig:tag-final-proposal part (c) was here
+The result has no tags: the interpreter patently uses no tags and no
+pattern matching. The term |b true| evaluates to a boolean and the term
+|lam varZ| evaluates to a function, both untagged. The |app| function
+applies |lam varZ| without pattern matching. What is more, evaluating an
+open term such as |testf3| below gives a type error rather than
+a run-time error
+%aplas-specific
+, Fig.~\ref{fig:tag-final-proposal}(d)%
+.
+% fig:tag-final-proposal part (d) was here
+The type error correctly complains
+that the initial environment should be a tuple rather than |()|.
+In other words, the term is open.
+\begin{figure}
+%
+(a)\vspace{-10pt}
 \begin{code}
@@
 \end{code}
-These functions comprise the entire interpreter.
-We now represent our sample term $(\fun{x}x)\True$ as
+%
+(b)\vspace{-10pt}
 \begin{code}
@@
 \end{code}
-This representation is almost the same as in \S\ref{tagproblem}, only
-written with lowercase identifiers. To evaluate an object term is to
-apply its representation to the empty environment.
+%
+(c)\vspace{-10pt}
 \begin{code}
@@
 \end{code}
-The result has no tags: the interpreter patently uses no tags and no
-pattern matching. The term |b true| evaluates to a boolean and the term
-|lam varZ| evaluates to a function, both untagged. The |app| function
-applies |lam varZ| without pattern matching. What is more, evaluating an
-open term such as |testf3| below gives a type error rather than
-a run-time error.
+%
+(d)\vspace{-10pt}
 \begin{code}
@@
 \end{code}
-The type error correctly complains
-that the initial environment should be a tuple rather than |()|.
-In other words, the term is open.
-
+%
+\caption{OCaml code illustrating our final proposal.}
+\label{fig:tag-final-proposal}
+\end{figure}
+%
 In sum, by Church\hyp encoding terms using ordinary functions, we
 achieve a tagless evaluator for a typed object language in a
-metalanguage with a simple type system
-\citep{hindley-principal,milner-theory}.  In this \emph{final} rather
+metalanguage with a simple 
+%aplas
+Hindley-Milner 
+type system
+%aplas \cite{hindley-principal,milner-theory}
+.  In this \emph{final} rather
 than \emph{initial} approach, both kinds of run-time errors in
@@
 not just to us but also to the metalanguage implementation, we avoid
-the serious performance penalty \citep{WalidICFP02} of error checking.
+the serious performance penalty \cite{WalidICFP02} of error checking.
 \Citet{Gluck-jones-optimality} explains deeper technical reasons that
@@
 We have programmed our interpreters in OCaml (and, for staging,
-\citet{metaocaml}) and standard Haskell. The complete code is available%
+\cite{metaocaml}) and standard Haskell. The complete code is available%
 \footnote{\oleg{put URL!}.}
@@
 We take the problem of tagless (staged) typed-preserving
-interpretation exactly as it was posed in \citet{WalidICFP02} and
-\citet{xi-guarded}.  We use their running examples and achieve the
+interpretation exactly as it was posed in \cite{WalidICFP02} and
+\cite{xi-guarded}.  We use their running examples and achieve the
 result they call desirable.  Our contributions are as follows.
@@
 \item We describe an approach to self\hyp interpretation compatible with the
-  above (\S\ref{selfinterp}).  Self\hyp interpretation turned out harder than expected.
+  above%
+%aplas: (\S\ref{selfinterp})
+.  Self\hyp interpretation turned out
+  harder than expected%
+%aplas
+\footnote{Due to the severe lack of space, we cannot describe
+self-interpretation and refer the reader to the accompanying code.}%
+.
 \end{enumerate*}
@@
 recently that some work on programming type-indexed values in ML
-\citep{yang-encoding} has given a hint of how such a function can be
+\cite{yang-encoding} has given a hint of how such a function can be
 expressed.''  We discuss related work in~\S\ref{related}.
@@
 We may claim some contribution about type-preserving CPS. Check
-related work section in \citet{Guillemette-Monier-PLPV}, especially
+related work section in \cite{Guillemette-Monier-PLPV}, especially
 check the work of Shao on type-preserving CPS in Flint. The PLPV paper
@@
 In contrast to \S\ref{intro}, we encode binding using higher-order
-abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
+abstract syntax (HOAS) \cite{miller-manipulating,pfenning-higher-order}
 rather than de Bruijn indices. This makes the encoding convenient and
@@
 the functions that construct object terms are methods in a type class
-|Symantics| (with a parameter |repr| of kind |* -> *|). The class is so named
+|Symantics| (with a parameter |repr| of kind |* -> *|)%
+%aplas-specific
+, Fig.~\ref{fig:symantics-haskell}(a)%
+. The class is so named
 because its interface gives the syntax of the object language and its
 instances give the semantics.
+\begin{figure}
+%
+(a)\vspace{-10pt}
 \begin{code}
@@
 \end{code}
+%
+(b)\vspace{-10pt}
+\begin{code}
+testpowfix () = lam (\x -> fix (\self -> lam (\n ->
+                 if_ (leq n (int 0)) (int 1)
+                     (mul x (app self (add n (int (-1))))))))
+\end{code}
+%
+(c)\vspace{-10pt}
+\begin{code}
+testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
+\end{code}
+%
+\caption{Symantics in Haskell.}
+\label{fig:symantics-haskell}
+\end{figure}
+% fig:symantics-haskell part (a) was here
 For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
@@
 For another example, the classical $\mathit{power}$ function is
-\begin{code}
-testpowfix () = 
-  lam (\x -> fix (\self -> lam (\n ->
-  if_ (leq n (int 0)) (int 1)
-      (mul x (app self (add n (int (-1))))))))
-\end{code}
+%aplas-specific
+in Fig.~\ref{fig:symantics-haskell}(b)
+% fig:symantics-haskell part (b) was here
 and the partial application $\fun{x} \mathit{power}\;x\;7$ is
-\begin{code}
-testpowfix7 () = 
- lam (\x -> app (app (testpowfix ()) x) (int 7))
-\end{code}
+in Fig.~\ref{fig:symantics-haskell}(c).
+% fig:symantics-haskell part (c) was here
 The dummy argument |()| above is to avoid the monomorphism
@@
 with a compiler flag.
-
 The methods |add|, |mul|, and |leq| are quite similar, and so are
@@
 encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
-Haskell's type checker also flags syntactically invalid object terms, such as if
-we forget |app| somewhere above.
-
-\begin{figure*}
+Haskell's type checker also flags syntactically invalid object terms, 
+such as if we forget |app| somewhere above.
+
+\begin{figure*}[t]
 \begin{floatrule}
@@
   val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
-  val if_ : ('c, bool) repr -> (unit -> ('c, 'da) repr) -> (unit -> ('c, 'da) repr) -> ('c, 'da) repr
+  val if_ : ('c, bool) repr-> (unit-> ('c, 'da) repr)-> (unit-> ('c, 'da) repr)-> ('c, 'da) repr
+end
+\end{code}
+\begin{code}
+module EX(S: Symantics) = struct
+  open S
+  let test1 () = app (lam (fun x -> x)) (bool true)
+  let testpowfix () =  lam (fun x -> fix (fun self -> lam (fun n ->
+                         if_ (leq n (int 0)) (fun () -> int 1)
+                             (fun () -> mul x (app self (add n (int (-1))))))))
+  let testpowfix7 =  lam (fun x -> app (app (testpowfix ()) x) (int 7))
 end
@@
 the additional type parameter |c|, an
-environment classifier \citep{WalidPOPL03} (required by MetaOCaml for
+environment classifier \cite{WalidPOPL03} (required by MetaOCaml for
 code generation in~\S\ref{compiler}); and the $\eta$-expanded type for
@@
 
-The functor below encodes our running examples |test1| and $\mathit{power}$.
-\begin{code}
-module EX(S: Symantics) = struct
-  open S
-  let test1 () =
-    app (lam (fun x -> x)) (bool true)
-  let testpowfix () = 
-    lam (fun x -> fix (fun self -> lam (fun n ->
-    if_ (leq n (int 0)) (fun () -> int 1)
-        (fun () -> mul x
-          (app self (add n (int (-1))))))))
-  let testpowfix7 = 
-    lam (fun x -> app (app (testpowfix ()) x)
-                      (int 7))
-end
-\end{code}
+The functor |EX|, Fig.~\ref{fig:ocaml-simple} encodes 
+our running examples |test1| and $\mathit{power}$.
 The dummy argument to |test1| and |testpowfix| is an artifact of
@@
 metalanguage.  The module below interprets each object\hyp language
-operation as the corresponding metalanguage operation.
+operation as the corresponding metalanguage operation%
+%aplas-specific
+, Fig.~\ref{fig:interpreter-R}(a)
+.
+% fig:interpreter-R part (a) was here
+%
+\begin{figure}
+%
+(a)\vspace{-10pt}
 \begin{code}
@@
   let app  e1 e2    = e1 e2
-  let fix  f        =
-    let rec self n = f self n in self
+  let fix  f        = let rec self n = f self n in self
 
@@
 \end{code}
-
+%
+(b)\vspace{-10pt}
+\begin{code}
+module EXR = EX(R)
+\end{code}
+%
+(c)\vspace{-10pt}
+\begin{code}
+newtype R a = R {unR::a}
+instance Symantics R where ...
+\end{code}
+%
+\caption{The tagless interpreter \texttt{R}.}
+\label{fig:interpreter-R}
+\end{figure}
+%
+%
 As in~\S\ref{ourapproach},
@@
 OCaml's addition, and |app| is OCaml's application. To run our
-examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|.
-\begin{code}
-module EXR = EX(R)
-\end{code}
+examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|%
+%aplas-specific
+, Fig.~\ref{fig:interpreter-R}(b)
+.
+% fig:interpreter-R part (b) was here
 Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
 In Haskell, we define
-\begin{code}
-newtype R a = R {unR::a}
-instance Symantics R where ...
-\end{code}
+%aplas-specific
+ Fig.~\ref{fig:interpreter-R}(c)
+.
+% fig:interpreter-R part (b) was here
 Although |R| looks like a tag, it is only
@@
 
+% aplas-specific
+For variety, the source code shows another interpreter, |L|, which measures
+the \emph{size} of each object term, defined as the number of term
+constructors. This interpreter is not only tagless but also total. It
+``evaluates'' even seemingly divergent terms.
+\begin{comment}
 For variety, we show another interpreter, which measures the \emph{size}
@@
 
-\begin{comment}
+%aplas \begin{comment}
 module EX1(S: Symantics) = struct
@@
 let module E =EX1(L) in E.tfix ();;
+%%aplas \end{comment}
 \end{comment}
@@
 \section{A tagless compiler (or, a staged interpreter)}\label{compiler}
-
+\vspace{-5pt}
 Besides immediate evaluation, we can compile our object language
@@
 present stage as values of type |('c,|$t$|) code| where |'c| is the
-environment classifier \citep{WalidPOPL03,calcagno-ml-like}. Code values are created
+environment classifier \cite{WalidPOPL03,calcagno-ml-like}. Code values are created
 by a \emph{bracket} form |.<|$e$|>.|, which specifies that the expression~$e$ is to be
@@
 Inserting brackets and escapes appropriately into the
-evaluator~|R| above yields the simple compiler~|C| below.
+evaluator~|R| above yields the simple compiler~|C| below%
+%aplas-specific
+, Fig.~\ref{fig:interpreter-C}(a)
+.
+% fig:interpreter-C part (a) was here
+%
+\begin{figure}[t]
+%
+(a)\vspace{-10pt}
 % this code does not have the 'sv parameter. It shows up later.
@@
   let app e1 e2     = .<.~e1 .~e2>.
-  let fix f = 
-    .<let rec self n = .~(f .<self>.) n in self>.
+  let fix f =  .<let rec self n = .~(f .<self>.) n in self>.
   let add e1 e2     = .<.~e1 + .~e2>.
@@
   let leq x y       = .<.~x <= .~y>.
-  let if_ eb et ee = 
-    .<if .~eb then .~(et ()) else .~(ee ())>.
+  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>.
 end
 \end{code}
+%
+(b)\vspace{-10pt}
+\begin{code}
+let module E = EX(C) in E.test1 ()
+\end{code}
+%
+(c)\vspace{-10pt}
+\begin{code}
+let module E = EX(C) in E.testpowfix7
+\end{code}
+%
+(d)\vspace{-10pt}
+\begin{code}
+.<fun x_1 -> (fun x_2 ->
+  let rec self_3 = fun n_4 ->
+   (fun x_5 -> if x_5 <= 0 then 1  else x_2 * (self_3 (x_5 + (-1))))
+   n_4 in self_3) x_1 7>.
+\end{code}
+%
+\caption{The tagless staged interpreter \texttt{C}.}
+\label{fig:interpreter-C}
+\end{figure}
+%
 This is a straightforward staging of
@@
 This compiler produces
-unoptimized code. For example, interpreting our |test1|
-\begin{code}
-let module E = EX(C) in E.test1 ()
-\end{code}
+unoptimized code. For example, interpreting our |test1|%
+%aplas-specific
+, Fig.~\ref{fig:interpreter-C}(b),
+% fig:interpreter-C part (b) was here
 gives the code value |.<(fun x_6 -> x_6) true>.|
@@
 with
-\begin{code}
-let module E = EX(C) in E.testpowfix7
-\end{code}
-gives a code value with many apparent $\beta$-redexes:
-\begin{code}
-.<fun x_1 -> (fun x_2 ->
-  let rec self_3 = fun n_4 ->
-   (fun x_5 -> if x_5 <= 0 then 1 
-               else x_2 * (self_3 (x_5 + (-1))))
-   n_4 in self_3) x_1 7>.
-\end{code}
+%aplas-specific
+Fig.~\ref{fig:interpreter-C}(c),
+% fig:interpreter-C part (c) was here
+gives a code value with many apparent $\beta$-redexes
+%aplas-specific
+Fig.~\ref{fig:interpreter-C}(d)
+% fig:interpreter-C part (d) was here
 
@@
 our bytecode uses integer-named
-variables to be realistic. We then define 
+variables to be realistic. 
+%aplas
+\begin{comment}
+We then define 
 \begin{code}
@@
 names. We define the compiler by making |C| an instance of the
-class |Symantics|. The implementation\footnote{The implementation uses
+class |Symantics|.
+\end{comment}
+The implementation\footnote{The implementation uses
 GADTs because we also wanted to write a typed interpreter for 
@@
 function, of type |a -> C a|, to send a value to the future stage
-\citep{xi-guarded,WalidPOPL03}.\footnote{We note in passing that, if we
+\cite{xi-guarded,WalidPOPL03}.\footnote{We note in passing that, if we
 were to add polymorphic \texttt{lift} to the type class
@@
     = int (unR (add n1 n2))
-  add e1 e2 = P1 Nothing
-    (add (abstr1 e1) (abstr1 e2))
+  add e1 e2 = P1 Nothing (add (abstr1 e1) (abstr1 e2))
   -- mul and leq are analogous and elided
-  if_ (P1 (Just s) _) et ef
-    = if unR s then et else ef
+  if_ (P1 (Just s) _) et ef = if unR s then et else ef
   if_ eb et ef = P1 Nothing
@@
 \begin{code}
-data P1 (a -> b) = S1 (P1 a -> P1 b)
-                 | E1 (C (a -> b))
-
-data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b))
-                      (C (a -> b))
+data P1 (a -> b) = S1 (P1 a -> P1 b) | E1 (C (a -> b))
+data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b)) (C (a -> b))
 \end{code}
@@
 GADTs and type-class functional dependencies
-\citep{oliveira-typecase}.  These
+\cite{oliveira-typecase}.  These
 methods are equivalent, and here we show the GADT way; |incope1.hs|
@@
 module P = struct
-  type ('c,'sv,'dv) repr = {st: 'sv option;
-                            dy: ('c,'dv) code}
+  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
   let abstr {dy = x} = x
@@
 
-  let int  (x:int)  = {st = Some (R.int x);
-                       dy = C.int x}
-  let bool (x:bool) = {st = Some (R.bool x);
-                       dy = C.bool x}
+  let int  (x:int)  = {st = Some (R.int x);  dy = C.int x}
+  let bool (x:bool) = {st = Some (R.bool x); dy = C.bool x}
 
@@
   | {st = Some b} -> if b then et () else ee ()
-  | _ -> pdyn (C.if_ (abstr eb) 
-                     (fun () -> abstr (et ()))
+  | _ -> pdyn (C.if_ (abstr eb) (fun () -> abstr (et ()))
                                 (fun () -> abstr (ee ())))
-  let lam f =
-  {st = Some f; 
+  let lam f = {st = Some f; 
                dy = C.lam (fun x -> abstr (f (pdyn x)))}
@@
 Many sophisticated approaches have been developed to make this trade-off
-\citep{jones-partial}, but this issue is orthogonal to our presentation.
+\cite{jones-partial}, but this issue is orthogonal to our presentation.
 A separate concern in our treatment of |fix| is possible code bloat in
 the residual program, which calls for let-insertion
-\citep{SwadiTahaKiselyovPasalic2006}.
+\cite{SwadiTahaKiselyovPasalic2006}.
 
@@
 \section{Continuation\hyp passing style}\label{variations}
+% aplas-specific
+\label{state}
+\label{S:CPS}
 
@@
 transformation.
-
+% Here's the text to replace the comment block below.
+This lets us decouple the evaluation strategy of the object language
+from that of the metalanguage. The accompanying code shows the CBN CPS
+interpreter (module |RCN| implementing |Symantics|) and a CBV CPS
+transformer |CPST|. The latter explicitly maps CPS interpretations to
+(direct) interpretations performed by the base interpreter~|S|. All
+these interpreters are typed, tagless and \emph{type-preserving} (as
+well as fully polymorphic in the answer-type). The type preservation
+is the consequence of the type soundness of the metalanguage.  We can
+modify the CBV CPS transformation to pass a piece of state along with
+the continuation. This technique lets us support mutable state.  Due
+to the severe lack of space we cannot describe these interpreters and
+refer the reader to the accompanying code.
+
+\begin{comment}
 \subsection{Call-by-name CPS interpreters}\label{S:CPS}
@@
 Haskell.  To represent a CBN object language in a CBV metalanguage,
-\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}
+\cite{reynolds-definitional,reynolds-relation} and \cite{PlotkinCBN}
 introduce CPS to make the evaluation strategy of a definitional
@@
 evaluates its argument before proceeding. This approach is in
-line with \citets{reynolds-relation}, albeit typed. The
+line with \cite{reynolds-relation}, albeit typed. The
 interpreter~|RCV| is useful for CBV evaluation of the object language
@@
 extensions.
-
+\end{comment}
+
+\begin{comment}
 \section{Self-interpretation}\label{selfinterp}
@@
 equivalent to the object application $\si\Encode{e}$
-\citep{jones-partial,taha-tag,Danvy-tagging-encoding}.
+\cite{jones-partial,taha-tag,Danvy-tagging-encoding}.
 A partial evaluator~$\pe$ maps object terms~$e$ to observationally
@@
 
-\begin{comment}
+%aplas \begin{comment}
 \jacques{But how 
@@
 hole below).}
-\end{comment}
-
-\begin{comment}
+%aplas \end{comment}
+
+%aplas \begin{comment}
 
@@
 
+%aplas \end{comment}
 \end{comment}
@@
 Even a simply typed $\lambda$-calculus obviously supports self\hyp
-interpretation, provided we use universal types \citep{taha-tag}.  The
-ensuing tagging overhead motivated \citet{taha-tag} to propose tag
+interpretation, provided we use universal types \cite{taha-tag}.  The
+ensuing tagging overhead motivated \cite{taha-tag} to propose tag
 elimination, which however does not statically guarantee that all tags
-will be removed \citep{WalidICFP02}.
+will be removed \cite{WalidICFP02}.
 
@@
 type-check dynamically\hyp typed values
-\citep{baars-typing,WalidICFP02,Guillemette-Monier-PLPV,haskell-list},
+\cite{baars-typing,WalidICFP02,Guillemette-Monier-PLPV,haskell-list},
 using the host language's type system to varying extents.
@@
 using an injection\hyp projection pair, as done for interpreters by
-\citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters},
-partial evaluation by \citet{Danvy-TDPE}, and type-level functions by
-\citet{oliveira-typecase}.  As explained in~\S\ref{S:PE-solution}, we
+\cite{Ramsey-ML-module-mania,Benton-embedded-interpreters},
+partial evaluation by \cite{Danvy-TDPE}, and type-level functions by
+\cite{oliveira-typecase}.  As explained in~\S\ref{S:PE-solution}, we
 avoid injection and projection at the type level by adding an argument
@@
 to combine online
-and offline partial evaluation.  Like \citet{SperberThiemann:TwoForOne},
+and offline partial evaluation.  Like \cite{SperberThiemann:TwoForOne},
 we strive for modularity by reusing interpreters for individual stages.  It would be interesting
-to try to derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
+to try to derive a \emph{cogen} \cite{Thiemann:cogeninsixlines}
 in the same manner.
@@
 embedded DSL. It is also common to use phantom types to prevent
-forming of ill-typed DSL terms \citet{Lava,Rhiger-thesis}. However,
+forming of ill-typed DSL terms \cite{Lava,Rhiger-thesis}. However,
 the universal type still remains (|Bit| and |NumSig| in Lava
-\citet{Lava}, |Raw| and |Term| in \citet[Fig 2.2,
+\cite{Lava}, |Raw| and |Term| in \cite[Fig 2.2,
   Sec. 3]{Rhiger-thesis}), along with the attendant pattern-matching
@@
 greatly complicates the soundness and completeness proofs of embedding
-\citet{Rhiger-thesis}. In contrast, our proofs become trivial.
+\cite{Rhiger-thesis}. In contrast, our proofs become trivial.
 The approach of \Citet[Sec 3.3.4]{Rhiger-thesis} does not 
@@
 type inference and typed CPS and partial evaluation. In contrast,
-\citet{Pfenning-Lee} conclude that partial evaluation and program
+\cite{Pfenning-Lee} conclude that partial evaluation and program
 transformations "do not seem to be expressible" even using their
@@
 the intensional type analysis
-\citet{Morrisett-intensional,Generic-Haskell}. Our running HOAS
+\cite{Morrisett-intensional,Generic-Haskell}. Our running HOAS
 example however, includes |fix|, which cannot be handled by the
-intensional type analysis \citet{xi-guarded}.  Our final approach
+intensional type analysis \cite{xi-guarded}.  Our final approach
 seems related to \citearound{'s approach to HOAS via catamorphism and
-  anamorphism}\citet{Washburn-Weirich-boxes}.
+  anamorphism}\cite{Washburn-Weirich-boxes}.
 
@@
 along this line.  In particular, they use
-contexts with a hole \citep[p.\,13]{honsell99coinductive} to define
+contexts with a hole \cite[p.\,13]{honsell99coinductive} to define
 \emph{observational equivalence}
-(see \S\ref{selfinterp}).
+%aplas
+% (see \S\ref{selfinterp}).
 \citearound{'s bibliography}\Citet{honsell99coinductive} refers to the
-foundational work in this important area.  Particularly intriguing is the link to the
+foundational work in this important area.  
+%aplas
+\begin{comment}
+Particularly intriguing is the link to the
 coinductive aspects of B\"{o}hm trees, as pointed out by
-\citet{berarducci-models} and \citet[Example 4.3.4]{jacobs-coalgebra}.
-
+\cite{berarducci-models} and \cite[Example 4.3.4]{jacobs-coalgebra}.
+\end{comment}
+
+%aplas
+\begin{comment}
 One way to understand our main idea is to eschew sum types and sum kinds
@@
 For the self\hyp interpreter, we then proceed to use a Church encoding for
-recursive data types \citep{bohm-automatic}.
-
-As \S\ref{S:clarify} observes, higher-rank and higher-kind
+recursive data types \cite{bohm-automatic}.
+\end{comment}
+
+%aplas: beforeAs \S\ref{S:clarify} observes, 
+We may observe that
+higher-rank and higher-kind
 polymorphism lets us type-check and compile object terms separately from
@@
 polymorphism in the separate compilation of modules
-\citep{shao-typed,shao-transparent}.
+\cite{shao-typed,shao-transparent}.
 

### 2007-06-11 tagless/tagless-final.tex
just some tricks to reduce space

@@
 %
-(a)\vspace{-10pt}
-\begin{code}
+(a) \begin{code2}
 type var = VZ | VS of var
 type exp = V of var | B of bool | L of exp | A of exp * exp
-\end{code}
-%
-(b)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(b) \begin{code2}
 let test1 = A (L (V VZ), B true)
-\end{code}
-%
-(c)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(c) \begin{code2}
 let rec lookup (x::env) = function
@@
 | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
-\end{code}
-%
-(d)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(d) \begin{code2}
 type u = UB of bool | UA of (u -> u)
-\end{code}
-%
-(e)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(e) \begin{code2}
 let rec eval env = function
@@
 | A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
-\end{code}
-%
-(f)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(f) \begin{code2}
 let test1r = eval [] test1
 val test1r : u = UB true 
-\end{code}
-%
-(g)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(g) \begin{code2}
 let test2 = A (B true, B false)
@@
 Exception: Match_failure in eval
-\end{code}
-%
-(h)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(h) \begin{code2}
 let test3 = A (L (V (VS VZ)), B true)
@@
 Exception: Match_failure in lookup
-\end{code}
-%
+\end{code2}
 \caption{OCaml code illustrating the tag problem.}
@@
 %
-(a)\vspace{-10pt}
-\begin{code}
+(a) \begin{code2}
 let varZ env        = fst env
@@
 let app e1 e2 env   = (e1 env) (e2 env)
-\end{code}
-%
-(b)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(b) \begin{code2}
 let testf1 = app (lam varZ) (b true)
-\end{code}
-%
-(c)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(c) \begin{code2}
 let testf1r = testf1 ()
 val testf1r : bool = true
-\end{code}
-%
-(d)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(d) \begin{code2}
 let testf3 = app (lam (varS varZ)) (b true)
 let testf3r = testf3 ()
-\end{code}
-%
+\end{code2}
 \caption{OCaml code illustrating our final proposal.}
@@
 %
-(a)\vspace{-10pt}
-\begin{code}
+(a) \begin{code2}
 class Symantics repr where
@@
   if_ :: repr Bool -> repr a -> repr a -> repr a
-\end{code}
-%
-(b)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(b) \begin{code2}
 testpowfix () = lam (\x -> fix (\self -> lam (\n ->
@@
                      (mul x (app self (add n (int (-1))))))))
-\end{code}
-%
-(c)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(c) \begin{code2}
 testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
-\end{code}
-%
+\end{code2}
 \caption{Symantics in Haskell.}
@@
 end
-\end{code}
-\begin{code}
+
 module EX(S: Symantics) = struct
@@
 \begin{figure}
-%
-(a)\vspace{-10pt}
-\begin{code}
+(a) \begin{code2}
 module R = struct
@@
 end
-\end{code}
-%
-(b)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(b) \begin{code2}
 module EXR = EX(R)
-\end{code}
-%
-(c)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(c) \begin{code2}
 newtype R a = R {unR::a}
 instance Symantics R where ...
-\end{code}
-%
+\end{code2}
 \caption{The tagless interpreter \texttt{R}.}
@@
 \begin{figure}[t]
-%
-(a)\vspace{-10pt}
 % this code does not have the 'sv parameter. It shows up later.
-\begin{code}
+(a) \begin{code2}
 module C = struct
@@
 end
-\end{code}
-%
-(b)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(b) \begin{code2}
 let module E = EX(C) in E.test1 ()
-\end{code}
-%
-(c)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(c) \begin{code2}
 let module E = EX(C) in E.testpowfix7
-\end{code}
-%
-(d)\vspace{-10pt}
-\begin{code}
+\end{code2}
+
+(d) \begin{code2}
 .<fun x_1 -> (fun x_2 ->
@@
    n_4 in self_3) x_1 7>.
-\end{code}
-%
+\end{code2}
 \caption{The tagless staged interpreter \texttt{C}.}

### 2007-06-12 tagless/tagless-final.tex
down to 19 pages. There is some work to be done in Sec 4.

@@
 \begin{quote}
+\small
     It should also be possible to define languages with a highly
@@
         including call-by-name (CBN) CPS, so to isolate the evaluation
-        order of the object language from that of the metalanguage.
+        order of the object language from that of the metalanguage
+%aplas
+\footnote{Due to the most severe lack of space, we cannot describe
+these features and refer the reader to the accompanying code.}%
+.
 \end{itemize*}
 We have programmed our interpreters in OCaml (and, for staging,
-\cite{metaocaml}) and standard Haskell. The complete code is available%
-\footnote{\oleg{put URL!}.}
+\cite{metaocaml}) and standard Haskell. The complete code is available
+\url{\oleg{put URL!}.}
 to supplement the paper. Our examples below switch between (Meta)OCaml
@@
     compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
-   in a typed metalanguage, in direct and continuation\hyp passing styles
-   (\S\ref{variations}).
+   in a typed metalanguage, in direct and continuation\hyp passing styles%
+% aplas
+%   (\S\ref{variations}).
+.
 \item All these interpreters use no type tags, patently never get stuck,
@@
 \item We show clean, comparable implementations in MetaOCaml and Haskell.
-\item We specify a functor signature that encompasses all our interpreters, from
-    evaluation and compilation (\S\ref{language}) to partial evaluation and CPS transformation (\S\ref{PE}).
+\item We specify a functor signature that encompasses all our
+  interpreters, from evaluation and compilation (\S\ref{language}) 
+   to partial evaluation%
+% aplas: commented
+%  and CPS transformation (\S\ref{PE}).
+.
 \item We point a clear way to extend the object language with more features
-    such as state (\S\ref{state}).
+    such as state%
+%aplas (\S\ref{state})%
+%aplas
+\footnotemark[\value{footnote}]
+.
 \item We describe an approach to self\hyp interpretation compatible with the
@@
 %aplas
-\footnote{Due to the severe lack of space, we cannot describe
-self-interpretation and refer the reader to the accompanying code.}%
+\footnotemark[\value{footnote}]
 .
@@
 metalanguage.  The module below interprets each object\hyp language
-operation as the corresponding metalanguage operation%
-%aplas-specific
-, Fig.~\ref{fig:interpreter-R}(a)
-.
-% fig:interpreter-R part (a) was here
-%
-\begin{figure}
-(a) \begin{code2}
+operation as the corresponding metalanguage operation
+\begin{code3}
 module R = struct
   type ('c,'dv) repr = 'dv (* no wrappers *)
-
-  let int  (x:int)  = x
-  let bool (b:bool) = b
-
-  let lam  f        = f
-  let app  e1 e2    = e1 e2
+  let int  (x:int)  = x         let bool (b:bool) = b
+  let lam  f        = f         let app  e1 e2    = e1 e2
   let fix  f        = let rec self n = f self n in self
-
-  let add  e1 e2    = e1 + e2
-  let mul  e1 e2    = e1 * e2
+  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
   let leq  x y      = x <= y
-  let if_  eb et ee = if eb then et () else ee ()
-end
-\end{code2}
-
-(b) \begin{code2}
-module EXR = EX(R)
-\end{code2}
-
-(c) \begin{code2}
-newtype R a = R {unR::a}
-instance Symantics R where ...
-\end{code2}
-\caption{The tagless interpreter \texttt{R}.}
-\label{fig:interpreter-R}
-\end{figure}
-%
+  let if_  eb et ee = if eb then et () else ee () end
+\end{code3}
 %
@@
 %aplas-specific
-, Fig.~\ref{fig:interpreter-R}(b)
+: \texttt{module EXR = EX(R)}%
 .
-% fig:interpreter-R part (b) was here
+%% \begin{code}
+%% module EXR = EX(R)
+%% \end{code}
 Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
+% aplas: commented out
+\begin{comment}
 In Haskell, we define
-%aplas-specific
- Fig.~\ref{fig:interpreter-R}(c)
-.
-% fig:interpreter-R part (b) was here
+\begin{code}
+newtype R a = R {unR::a}
+instance Symantics R where ...
+\end{code}
 Although |R| looks like a tag, it is only
@@
 In OCaml, too, it is obvious to the compiler that
+\end{comment}
+%aplas: the following line is for aplas only
+It is obvious to the compiler that
+%
 pattern matching cannot fail, because there is no
@@
   type ('c,'dv) repr = ('c,'dv) code
-
-  let int (x:int)   = .<x>.
-  let bool (b:bool) = .<b>.
-
+  let int (x:int)   = .<x>.           let bool (b:bool) = .<b>.
   let lam f         = .<fun x -> .~(f .<x>.)>.
@@
   let fix f =  .<let rec self n = .~(f .<self>.) n in self>.
-  let add e1 e2     = .<.~e1 + .~e2>.
-  let mul e1 e2     = .<.~e1 * .~e2>.
+  let add e1 e2     = .<.~e1 + .~e2>. let mul e1 e2     = .<.~e1 * .~e2>.
   let leq x y       = .<.~x <= .~y>.
-  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>.
-end
+  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>. end
 \end{code2}
@@
 (d) \begin{code2}
-.<fun x_1 -> (fun x_2 ->
-  let rec self_3 = fun n_4 ->
+.<fun x_1 -> (fun x_2 -> let rec self_3 = fun n_4 ->
    (fun x_5 -> if x_5 <= 0 then 1  else x_2 * (self_3 (x_5 + (-1))))
@@
 The resulting code obviously contains no tags and no pattern matching.
-
+%aplas: the line is for aplas. The original paragraph commented out
+The accompanying code shows the corresponding Haskell implemention. 
+\begin{comment}
 We have also implemented this compiler in Haskell. 
@@
 variables to be realistic. 
-%aplas
-\begin{comment}
 We then define 
@@
 class |Symantics|.
-\end{comment}
 The implementation\footnote{The implementation uses
@@
 accompanying code gives the full details.
+\end{comment}
 
@@
 |P1| to a function type |a -> b| should yield one of
-\begin{code}
+\begin{code3}
 data P1 (a -> b) = S1 (P1 a -> P1 b) | E1 (C (a -> b))
 data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b)) (C (a -> b))
-\end{code}
+\end{code3}
 That is, we need a nonparametric data type, something akin to
@@
 \citet{oliveira-typecase} dub the \emph{typecase} design pattern.
+%aplas-simplification
+Thus typed partial evaluation, like typed CPS transformation,
+inductively defines a map from source types to target types that
+performs case distinction on the source type.
+\begin{comment}
 
@@
 is not an accident, as we shall see in~\S\ref{S:CPS}.
+\end{comment}
 
@@
 in the accompanying source code shows the latter.
-
-\begin{figure*}
-\begin{floatrule}
-\begin{code}
-module type Symantics = sig
-  type ('c,'sv,'dv) repr
-
-  val int : int  -> ('c,int,int) repr
-  val bool: bool -> ('c,bool,bool) repr
-
-  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
-  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
-
-  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
-  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
-end
-\end{code}
-\end{floatrule}
-\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
-\label{fig:ocaml}
-\end{figure*}
-
+%aplas: the original text is in comments below. Here's a greatly
+% abbreaviated text
 We introduce a GADT with four data constructors.
-\begin{code}
+\begin{code3}
+data P t where
+  VI :: Int  -> P Int;                VB :: Bool -> P Bool
+  VF :: (P a -> P b) -> P (a -> b);   E  :: C t -> P t
+\end{code3}
+The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
+in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
+the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
+operand of type |P a -> P b| rather than |a -> b|. As before, we define a
+function |abstr| to extract a future-stage computation from a value of 
+type |P t|. The implementations of |int|, |bool|, and |add| are like
+in~\S\ref{S:PE-problem}.
+The interpretation of |lam f = VF f| just wraps the HOAS function |f|. 
+We can always compile |f| to a code value,
+but we delay it to apply |f| to concrete arguments. The interpretation of
+\begin{code3}
+app (VF f) ea = f ea;  app (E f)  ea = E (app f (abstr ea))
+\end{code3}
+checks to see if |ef| is such a delayed
+HOAS function |VF f|. If it is, we apply |f| to the
+concrete argument |ea|, giving us a chance to perform static
+computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
+dynamic value |E f|, we residualize.
+This solution using GADTs works but is not quite satisfactory. First, it
+cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
+the problem of nonexhaustive pattern\hyp matching reappears in
+|app|. Although Coq or logical
+frameworks with canonical forms could prove pattern-matching
+exhaustive, GHC cannot and issues warnings. Recall, our goal is to
+make this exhaustiveness \emph{syntactically} apparent.
+
+\begin{comment}
+We introduce a GADT with four data constructors.
+\begin{code3}
 data P t where
@@
   E  :: C t -> P t
-\end{code}
+\end{code3}
 The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
@@
 \emph{syntactically} apparent.
+\end{comment}
 
@@
 would ideally like is to write
+%aplas: inlining
+\texttt{data P t = P (Maybe (repr\_pe t)) (C t)}
+\begin{comment}
 \begin{code}
@@
 \end{code}
+\end{comment}
 where |repr_pe| is the type function defined
@@
 by
-\begin{code}
-repr_pe Int      = Int
-repr_pe Bool     = Bool
+\begin{code3}
+repr_pe Int      = Int          repr_pe Bool     = Bool
 repr_pe (a -> b) = P a -> P b
-\end{code}
+\end{code3}
 Although we can use type classes to define this type function
@@
 Figure~\ref{fig:ocaml} shows the new |Symantics| signature.
-
+% aplas: abbreviated figure
+\begin{figure*}
+\begin{floatrule}
+\begin{code}
+module type Symantics = sig
+  type ('c,'sv,'dv) repr
+  val int : int  -> ('c,int,int) repr
+  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
+  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+  val fix : ('x->'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
+  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val if_ : ('c,bool,bool) repr -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x)
+end
+\end{code}
+\end{floatrule}
+\caption{A (Meta)OCaml embedding of our object language that supports
+  partial evaluation (\texttt{bool}, \texttt{mul},
+\texttt{leq} are elided)}
+\label{fig:ocaml}
+\end{figure*}
+
+\begin{comment}
+\begin{figure*}
+\begin{floatrule}
+\begin{code}
+module type Symantics = sig
+  type ('c,'sv,'dv) repr
+
+  val int : int  -> ('c,int,int) repr
+  val bool: bool -> ('c,bool,bool) repr
+
+  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
+  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
+
+  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
+  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
+end
+\end{code}
+\end{floatrule}
+\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
+\label{fig:ocaml}
+\end{figure*}
+\end{comment}
+
+% aplas: abberviated figure here. The original is in teh comments
+% right down below
+\begin{figure}
+\begin{floatrule}
+\begin{code}
+module P = struct
+  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
+  let abstr {dy = x} = x    let pdyn x = {st = None; dy = x}
+  let int  (x:int)  = {st = Some (R.int x);  dy = C.int x}
+
+  let add e1 e2 = match e1, e2 with
+  | {st = Some 0}, e | e, {st = Some 0} -> e
+  | {st = Some m}, {st = Some n} -> int (R.add m n)
+  | _ -> pdyn (C.add (abstr e1) (abstr e2))
+
+  let if_ eb et ee = match eb with
+  | {st = Some b} -> if b then et () else ee ()
+  | _ -> pdyn (C.if_ (abstr eb) (fun () -> abstr (et ()))
+                                (fun () -> abstr (ee ())))
+  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
+
+  let app ef ea = match ef with {st = Some f} -> f ea
+                  | _ -> pdyn (C.app (abstr ef) (abstr ea)) end
+\end{code}
+\end{floatrule}
+\caption{Our partial evaluator (\texttt{bool}, \texttt{mul},
+  \texttt{leq} and \texttt{fix} are elided)}
+\label{fig:pe}
+\end{figure}
+
+\begin{comment}
 \begin{figure}
@@
 \end{figure}
+\end{comment}
 
@@
 argument~|'sv| to~|repr|.
+For example, the compiler |C| now begins \texttt{module C = struct
+  type ('c,'sv,'dv) repr = ('c,'dv) code} with the rest the same.
+In contrast, the partial evaluator~|P| relies on the type argument |'sv|.
+
+%aplas: here's the original text
+\begin{comment}
 For example, the compiler |C| now begins
@@
 In contrast, the partial evaluator~|P| relies on the type argument |'sv|.
+\end{comment}
 
@@
 
+%aplas: abbreviated paragraph. The original is commented below
+Any partial evaluator must decide how much to unfold recursion. The code in
+Figure~\ref{fig:pe} na\"\i vely unfolds |fix| whenever the
+argument is static.  In the accompanying source code is a conservative
+alternative |P.fix| that unrolls recursion only once, then residualizes.
+Many sophisticated approaches have been developed for this trade-off
+\cite{jones-partial}, but this issue is orthogonal to our presentation.
+\begin{comment}
 Any partial evaluator must decide how much to unfold recursion
@@
 \cite{SwadiTahaKiselyovPasalic2006}.
-
+\end{comment}
+
+%aplas: abbreviated paragraph. The original is in the comments below
 Given this implementation of~|P|, our running example
-\begin{code}
-let module E = EX(P) in E.test1 ()
-\end{code}
-evaluates to
-\begin{code}
-{P.st = Some true; P.dy = .<true>.}
-\end{code}
+\texttt{let module E = EX(P) in E.test1 ()} evaluates to
+\texttt{\{P.st = Some true; P.dy = .<true>.\}}
 of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
@@
 $\beta$-redexes in~\S\ref{compiler}, the partial evaluation
+\texttt{let module E = EX(P) in E.testpowfix7}
+gives the desired result
+\begin{code3}
+{P.st = Some <fun>;
+ P.dy = .<fun x_1 -> x_1 * (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * x_1)))))>.}
+\end{code3}
+
+\begin{comment}
+Given this implementation of~|P|, our running example
+\begin{code}
+let module E = EX(P) in E.test1 ()
+\end{code}
+evaluates to
+\begin{code}
+{P.st = Some true; P.dy = .<true>.}
+\end{code}
+of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
+a $\beta$-reduction has been statically performed to yield |true|.  More
+interestingly, whereas |testpowfix7| compiles to a code value with many
+$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
 \begin{code}
@@
 \end{code}
+\end{comment}
 
@@
 phases (with |Some| and |None|) but not object types.
-
+%aplas: commented out
+\begin{comment}
 Our typed partial evaluator is online and polyvariant.  It reuses the
@@
 evaluator and a compiler, but the general ideas are similar.
-
+\end{comment}
+
+% don't even have space for CPS
+\begin{comment}
 \section{Continuation\hyp passing style}\label{variations}
@@
 refer the reader to the accompanying code.
+\end{comment}
 
@@
 
+% aplas: this is the simplified paragraph. The original text is in the
+% comment right below.
+\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
+\citet{peyton-jones-simple} seem to argue that a self\hyp
+interpreter of a typed language cannot be tagless or Jones\hyp optimal:
+(1) One needs to encode a typed language in a typed language based on
+a sum type (at some level of the hierarchy);
+(2) It is not possible to write a \emph{direct} interpreter 
+for such an encoding of a typed language
+in a typed language, without either an
+advanced type system or the overhead of tagging and untagging;
+(3) Thus an indirect interpreter is necessary, which needs a universal
+  type (and hence tagging).
+While the logic is sound, we showed that the premise in the very first step
+was not valid.
+\begin{comment}
 \Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
@@
 was not valid.
+\end{comment}
 
@@
 type-check dynamically\hyp typed values
-\cite{baars-typing,WalidICFP02,Guillemette-Monier-PLPV,haskell-list},
+\cite{baars-typing,haskell-list},
 using the host language's type system to varying extents.
@@
 the \emph{typed} $\lambda$-calculus possesses a final coalgebra structure.
+% aplas: abbreviated. The original text is in comments
+\Citet{honsell99coinductive}
+investigate the untyped $\lambda$-calculus,
+along this line.
+% aplas: comment
+\begin{comment}
 \Citet{HonsellLenisa,honsell99coinductive}
 investigate the untyped $\lambda$-calculus,
-along this line.  In particular, they use
+along this line.  
+In particular, they use
 contexts with a hole \cite[p.\,13]{honsell99coinductive} to define
 \emph{observational equivalence}
-%aplas
-% (see \S\ref{selfinterp}).
+(see \S\ref{selfinterp}).
 \citearound{'s bibliography}\Citet{honsell99coinductive} refers to the
 foundational work in this important area.  
-%aplas
-\begin{comment}
 Particularly intriguing is the link to the
@@
 %aplas: beforeAs \S\ref{S:clarify} observes, 
-We may observe that
+We observe that
 higher-rank and higher-kind
-polymorphism lets us type-check and compile object terms separately from
-interpreters.  This observation is consistent with the role of
+polymorphism let us type-check and compile object terms separately from
+interpreters.  This is consistent with the role of
 polymorphism in the separate compilation of modules
-\cite{shao-typed,shao-transparent}.
+\cite{shao-typed}.
 
@@
 
-\section*{Acknowledgments}
-\ourthanks
+% aplas: acknowledgments are at the beginning
+% \section*{Acknowledgments}
+% \ourthanks
 
@@
 \bibsep=0pt
-\bibliography{tagless}
+% aplas
+\bibliography{aplas}
+% \bibliography{tagless}
 \end{document}

### 2007-06-12 tagless/tagless-final.tex
17.5 pages.

@@
 % accordingly. Let the paper and the accompanying code be in sync.
-
-\begin{figure}
-%
-(a) \begin{code2}
-type var = VZ | VS of var
-type exp = V of var | B of bool | L of exp | A of exp * exp
-\end{code2}
-
-(b) \begin{code2}
-let test1 = A (L (V VZ), B true)
-\end{code2}
-
-(c) \begin{code2}
-let rec lookup (x::env) = function
-| VZ   -> x
-| VS v -> lookup env v
-
-let rec eval0 env = function
-| V v       -> lookup env v
-| B b       -> b 
-| L e       -> fun x -> eval0 (x::env) e
-| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
-\end{code2}
-
-(d) \begin{code2}
-type u = UB of bool | UA of (u -> u)
-\end{code2}
-
-(e) \begin{code2}
-let rec eval env = function
-| V v       -> lookup env v
-| B b       -> UB b
-| L e       -> UA (fun x -> eval (x::env) e)
-| A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
-\end{code2}
-
-(f) \begin{code2}
-let test1r = eval [] test1
-val test1r : u = UB true 
-\end{code2}
-
-(g) \begin{code2}
-let test2 = A (B true, B false)
-let test2r = eval [] test2
-Exception: Match_failure in eval
-\end{code2}
-
-(h) \begin{code2}
-let test3 = A (L (V (VS VZ)), B true)
-let test3r = eval [] test3
-Exception: Match_failure in lookup
-\end{code2}
-\caption{OCaml code illustrating the tag problem.}
-\label{fig:tag-problem}
-\end{figure}
 It is straightforward to create an algebraic data type, say in OCaml%
@@
 %aplas-specific
-in Fig.~\ref{fig:tag-problem}(b)%
-% fig:tag-problem part (b) was here
-
+\texttt{let test1 = A (L (V VZ), B true)}.
+%aplas-else
+\begin{comment}
+\begin{code}
+let test1 = A (L (V VZ), B true)
+\end{code}
+\end{comment}
+
+\begin{figure}
+%
+(a) \begin{code2}
+type var = VZ | VS of var
+type exp = V of var | B of bool | L of exp | A of exp * exp
+\end{code2}
+
+(b) \begin{code2}
+let rec lookup (x::env) = function | VZ   -> x
+                                   | VS v -> lookup env v
+let rec eval0 env = function
+| V v       -> lookup env v
+| B b       -> b 
+| L e       -> fun x -> eval0 (x::env) e
+| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
+\end{code2}
+
+(c) \begin{code2}
+type u = UB of bool | UA of (u -> u)
+\end{code2}
+
+(d) \begin{code2}
+let rec eval env = function
+| V v       -> lookup env v
+| B b       -> UB b
+| L e       -> UA (fun x -> eval (x::env) e)
+| A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
+\end{code2}
+\caption{OCaml code illustrating the tag problem.}
+\label{fig:tag-problem}
+\end{figure}
+% 
 Following \cite{WalidICFP02},
@@
 %aplas-specific
-, Fig.~\ref{fig:tag-problem}(c)%
+, Fig.~\ref{fig:tag-problem}(b)%
 . It takes
@@
 interpreted code.
-% fig:tag-problem part (c) was here
+% fig:tag-problem part (b) was here
 If our OCaml-like metalanguage were untyped, the code above would be 
@@
 %aplas-specific
-, Fig.~\ref{fig:tag-problem}(d)%
+, Fig.~\ref{fig:tag-problem}(c)%
 .
-% fig:tag-problem part (d) was here
+% fig:tag-problem part (c) was here
 We can then write a typed interpreter%
 %aplas-specific
-, Fig.~\ref{fig:tag-problem}(e)%
-% fig:tag-problem part (e) was here
+, Fig.~\ref{fig:tag-problem}(d)
+% fig:tag-problem part (d) was here
 whose inferred type is |u list -> exp -> u|. Now we can evaluate
-our sample |test1| term%
 %aplas-specific
-, Fig.~\ref{fig:tag-problem}(f)%
-.
+\texttt{eval [] test1} obtaining |UB true|.
+%aplas-else
+\begin{comment}
+\begin{code}
+let test1r = eval [] test1
+val test1r : u = UB true 
+\end{code}
+\end{comment}
 % fig:tag-problem part (f) was here
 The unfortunate tag |UB| in the result reflects that |eval| is a partial
-function.  First, the pattern match |with UA f| in the line
-|A (e1,e2)| is not exhaustive, so |eval| can fail%
+function.  
 %aplas-specific
-, Fig.~\ref{fig:tag-problem}(g), 
-if we apply a boolean,
+First, the pattern match |with UA f| in the line |A (e1,e2)| is not 
+exhaustive, so |eval| can fail if we apply a boolean,
 as in the ill-typed term |A (B true, B false)|.
-% fig:tag-problem part (g) was here
 Second, the |lookup|
 function assumes a nonempty environment, so |eval| can fail if we
-evaluate an open term%
-%aplas-specific
-, Fig.~\ref{fig:tag-problem}(h)%
-.
-% fig:tag-problem part (h) was here
+evaluate an open term |A (L (V (VS VZ)), B true)|.
 After all, the type |exp| represents object
 terms both well-typed and ill-typed, both open and closed.
+%aplas-else
+\begin{comment}
+First, the pattern match |with UA f| in the line
+|A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
+as in the ill-typed term |A (B true, B false)|.
+\begin{code}
+let test2 = A (B true, B false)
+let test2r = eval [] test2
+Exception: Match_failure in eval
+\end{code}
+Second, the |lookup|
+function assumes a nonempty environment, so |eval| can fail if we
+evaluate an open term
+\begin{code}
+let test3 = A (L (V (VS VZ)), B true)
+let test3r = eval [] test3
+Exception: Match_failure in lookup
+\end{code}
+After all, the type |exp| represents object
+terms both well-typed and ill-typed, both open and closed.
+\end{comment}
 
@@
 
+%aplas-else
+\begin{comment}
 \subsection{Solutions using fancier types}
+\end{comment}
 
@@
 on dependent types \cite{WalidICFP02,fogarty-concoqtion}.
-
+%aplas-else
+\begin{comment}
 For a metalanguage's type system to allow the well-typed object term
 |test1| but disallow the ill-typed object term |test2|, fancier types
-such as GADTs or dependent types seem necessary.  Yet other type systems
+such as GADTs or dependent types seem necessary.  
+\end{comment}
+Yet other type systems
 have been proposed to distinguish closed terms like |test1| from open
@@
 so that |lookup| never receives an empty environment.  We discuss these
-proposals further in \S\ref{related}; here we just note that many
+proposals further in \S\ref{related}%
+%aplas-specific
+.
+%aplas-else
+\begin{comment}
+; here we just note that many
 advanced type systems have been devised to ensure statically that an
 object term is well-typed and closed.
+\end{comment}
 
@@
 We represent object programs using ordinary functions rather than
-data constructors.  The interpreter below provides these functions%
-%aplas-specific
-, Fig.~\ref{fig:tag-final-proposal}(a)%
-.
-% fig:tag-final-proposal part (a) was here
-These functions comprise the entire interpreter.
-We now represent our sample term $(\fun{x}x)\True$ as
-%aplas-specific
-Fig.~\ref{fig:tag-final-proposal}(b)%
-.
-% fig:tag-final-proposal part (b) was here
-This representation is almost the same as in \S\ref{tagproblem}, only
-written with lowercase identifiers. To evaluate an object term is to
-apply its representation to the empty environment%
-%aplas-specific
-, Fig.~\ref{fig:tag-final-proposal}(c)%
-.
-% fig:tag-final-proposal part (c) was here
-The result has no tags: the interpreter patently uses no tags and no
-pattern matching. The term |b true| evaluates to a boolean and the term
-|lam varZ| evaluates to a function, both untagged. The |app| function
-applies |lam varZ| without pattern matching. What is more, evaluating an
-open term such as |testf3| below gives a type error rather than
-a run-time error
-%aplas-specific
-, Fig.~\ref{fig:tag-final-proposal}(d)%
-.
-% fig:tag-final-proposal part (d) was here
-The type error correctly complains
-that the initial environment should be a tuple rather than |()|.
-In other words, the term is open.
-\begin{figure}
-%
-(a) \begin{code2}
+data constructors.  The interpreter below provides these functions
+\begin{code3}
 let varZ env        = fst env
@@
 let app e1 e2 env   = (e1 env) (e2 env)
-\end{code2}
-
-(b) \begin{code2}
+\end{code3}
+which comprise the entire interpreter.
+We now represent our sample term $(\fun{x}x)\True$ as
+%aplas-specific
+\texttt{let testf1 = app (lam varZ) (b true)}
+%aplas-else
+\begin{comment}
+\begin{code}
 let testf1 = app (lam varZ) (b true)
-\end{code2}
-
-(c) \begin{code2}
+\end{code}
+\end{comment}
+This representation is almost the same as in \S\ref{tagproblem}, only
+written with lowercase identifiers. To evaluate an object term is to
+apply its representation to the empty environment%
+%aplas-specific
+, |testf1 ()|, obtaining |true|.
+%aplas-else
+\begin{comment}
+\begin{code}
 let testf1r = testf1 ()
 val testf1r : bool = true
-\end{code2}
-
-(d) \begin{code2}
+\end{code}
+\end{comment}
+The result has no tags: the interpreter patently uses no tags and no
+pattern matching. The term |b true| evaluates to a boolean and the term
+|lam varZ| evaluates to a function, both untagged. The |app| function
+applies |lam varZ| without pattern matching. What is more, evaluating an
+%aplas-specific
+open term such as \texttt{app (lam (varS varZ)) (b true)}
+gives a type- rather than a run-time error.
+%aplas-else
+\begin{comment}
+\begin{code}
+open term such as |testf3| below gives a type error rather than
+a run-time error
 let testf3 = app (lam (varS varZ)) (b true)
 let testf3r = testf3 ()
-\end{code2}
-\caption{OCaml code illustrating our final proposal.}
-\label{fig:tag-final-proposal}
-\end{figure}
-%
+\end{code}
+\end{comment}
+The type error correctly complains
+that the initial environment should be a tuple rather than |()|.
+In other words, the term is open.
+
 In sum, by Church\hyp encoding terms using ordinary functions, we
@@
 \S\ref{tagproblem} above as |app (lam (\x -> x)) (bool True)|,
-whose inferred type is |Symantics repr => repr Bool|.
+whose inferred type is \texttt{Symantics repr => repr Bool}.
 For another example, the classical $\mathit{power}$ function is
@@
 due to interpreting |fix|.
+%aplas-specific
+(The source code shows a total interpreter |L| that measures
+the size of each object term).
+%
 \begin{proposition}
@@
 metalanguage's type system.
-
-% aplas-specific
-For variety, the source code shows another interpreter, |L|, which measures
-the \emph{size} of each object term, defined as the number of term
-constructors. This interpreter is not only tagless but also total. It
-``evaluates'' even seemingly divergent terms.
+% aplas-commented
 \begin{comment}
@@
 %aplas-specific
-Fig.~\ref{fig:interpreter-C}(d)
+Fig.~\ref{fig:interpreter-C}(d).
 % fig:interpreter-C part (d) was here
-
 This compiler does not incur
@@
 the Haskell data type 
+%aplas-specific
+\texttt{data P0 t = S0 (R t) $\mid$ E0 (C t)}.
+We then start to define |P0| as an instance of the |Symantics| class.
+Integer and boolean literals are immediate, present-stage
+values. Addition yields a static term (using~|R|) if and only if both operands
+are static; otherwise we extract the dynamic terms from the operands and
+add them using~|C|.
+Whereas |mul| and |leq| are as easy to define as |add|, we encounter
+a problem with |if_|.  If the first argument to |if_| is a dynamic term
+(type |C Bool|) and the second and third arguments are a static term
+(type |R a|) and a dynamic term (type |C a|), then we need to convert
+the static term to dynamic, but there is no polymorphic ``lift''
+function, of type |a -> C a|, to send a value to the future stage
+\cite{xi-guarded,WalidPOPL03}.
+%aplas-else
+\begin{comment}
 \begin{code}
@@
 f = app (lift f)}}
+\end{comment}
 
@@
 We switch to the Haskell data type
+%aplas-specific
+|data P1 t = P1 (Maybe (R t)) (C t)|
+%aplas-else
+\begin{comment}
 \begin{code}
@@
 \end{code}
+\end{comment}
 so that a partially evaluated term |P1 t| always contains a dynamic
@@
 can now define the polymorphic function
+%aplas-specific
+|abstr1 (P1 _ dyn) = dyn| of the type
+|abstr1 :: P1 t -> C t|
+%aplas-else
+\begin{comment}
 \begin{code}
@@
 \end{code}
+\end{comment}
 to extract it without requiring polymorphic lift into~|C|.  We then try
@@
 constructs of our object language, including |if_|.
-\begin{code}
+\begin{code3}
 instance Symantics P1 where
   int  x = P1 (Just (int x)) (int x)
-  bool b = P1 (Just (bool b)) (bool b)
-  add (P1 (Just n1) _) (P1 (Just n2) _)
-    = int (unR (add n1 n2))
+  add (P1 (Just n1) _) (P1 (Just n2) _) = int (unR (add n1 n2))
   add e1 e2 = P1 Nothing (add (abstr1 e1) (abstr1 e2))
-  -- mul and leq are analogous and elided
   if_ (P1 (Just s) _) et ef = if unR s then et else ef
-  if_ eb et ef = P1 Nothing
-    (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
-\end{code}
-
+  if_ eb et ef = P1 Nothing (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
+\end{code3}
 When we come to functions, however, we stumble.  According to our
@@
 
+%aplas-else
+\begin{comment}
   A lot of effort has gone into ``typing dynamic typing'': to statically
@@
   terms such as those read from a file.
+\end{comment}
 
@@
 partial evaluation (\S\ref{S:PE-solution}) and CPS
-transformation~(\S\ref{S:CPS}).  This shift also underlies fusion,
+transformation%
+%aplas
+.
+%aplas-else ~(\S\ref{S:CPS}).
+This shift also underlies fusion,
 functionalization, and amortized complexity analysis.
-
+%aplas-else
+\begin{comment}
 When the metalanguage does provide higher-rank and higher-kind
@@
 from any interpreters it may be plugged into.
+\end{comment}
 

### 2007-06-12 tagless/tagless-final.tex
shorten, shorten, shorten

@@
 Our main idea is to encode
-de Bruijn or higher-order abstract syntax
+% de Bruijn or higher-order abstract syntax
+HOAS
 using cogen functions rather than data constructors.
@@
     defined language should be adequate for the defining language.
-    \hfill\rm\citep{reynolds-definitional}
+    \hfill\rm John Reynolds~\citep{reynolds-definitional}
 \end{quote}
@@
 proposal and contributions.  
+%aplas
+\begin{comment}
 No matter how we represent the object language in the
@@
 whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}.
+\end{comment}
+%aplas-else
+We leave aside the solved problem of writing a parser\slash type\hyp checker,
+for embedding object language objects into the metalanguage
+(whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}),
+and just enter them by hand.
+%aplas
 
@@
 % accordingly. Let the paper and the accompanying code be in sync.
-It is straightforward to create an algebraic data type, say in OCaml%
+It is straightforward to create an algebraic data type, say in OCaml~%
 %aplas-specific
@@
 data constructors.  The interpreter below provides these functions
-\begin{code3}
-let varZ env        = fst env
-let varS vp env     = vp (snd env)
-let b (bv:bool) env = bv
+\begin{code4}
+let varZ env    = fst env       let b (bv:bool) env = bv
+let varS vp env = vp (snd env)  let app e1 e2 env   = (e1 env) (e2 env)
 let lam e env   = fun x -> e (x,env)
-let app e1 e2 env   = (e1 env) (e2 env)
-\end{code3}
+\end{code4}
 which comprise the entire interpreter.
@@
 %aplas
-Hindley-Milner 
-type system
+Hindley-Milner type system
 %aplas \cite{hindley-principal,milner-theory}
@@
 universal type. Church-encoding of the type |u|, \S\ref{tagproblem},
-requires two continuations; the interpreter like |app| above would
+requires two continuations; the interpreter (see |app| above) would
 have to provide both to the encoding of |e1|. The continuation
@@
 invoked, yet it must be supplied. Our interpreter has no error
-continuations at all.
+continuation at all.
 
 The evaluator above is wired directly into the
-functions |b|, |lam|, |app|, and so on.  In the rest of this paper, 
-we explain how to abstract the interpreter so as
-to process the same term in many other
-ways: compilation, partial evaluation, size
-computation, and so forth.
+functions |b|, |lam|, |app|, and so on.  %In the rest of this paper, 
+We explain how to abstract the interpreter so as
+to process the \emph{same} term in many other
+ways: compilation, partial evaluation, CPS
+conversion, and so forth.
 
@@
 %aplas
-\footnote{Due to the most severe lack of space, we cannot describe
-these features and refer the reader to the accompanying code.}%
+\footnote{Due to serious lack of space, 
+we refer the reader to the accompanying code for this.}%
 .
@@
 %aplas
-\footnotemark[\value{footnote}]
+\footnote{again, see code}
 .
@@
 power.
-
 In contrast to \S\ref{intro}, we encode binding using higher-order
@@
 \label{encoding}
-
 We embed our language in (Meta)OCaml and Haskell.  In Haskell,
@@
 class Symantics repr where
-  int  :: Int  -> repr Int
-  bool :: Bool -> repr Bool
-
+  int  :: Int  -> repr Int       bool :: Bool -> repr Bool
   lam :: (repr a -> repr b) -> repr (a -> b)
@@
 
-Comparing |Symantics| with Figure~\ref{fig:object}
+Comparing |Symantics| with Fig.~\ref{fig:object}
 shows how to represent \emph{every} typed, closed object term in the
-metalanguage. Moreover, the representation preserves types, in the following
-sense.
+metalanguage. Moreover, the representation preserves types.
 \begin{proposition}
@@
 \begin{code}
-module type Symantics = sig
-  type ('c, 'dv) repr
-
+module type Symantics = sig type ('c, 'dv) repr
   val int : int  -> ('c, int) repr
@@
 
-module EX(S: Symantics) = struct
-  open S
+module EX(S: Symantics) = struct open S
   let test1 () = app (lam (fun x -> x)) (bool true)
@@
 \end{figure*}
-
 To embed the same object language in (Meta)OCaml, we replace the type
@@
 \begin{code3}
-module R = struct
-  type ('c,'dv) repr = 'dv (* no wrappers *)
+module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
   let int  (x:int)  = x         let bool (b:bool) = b
@@
 (a) \begin{code2}
-module C = struct
-  type ('c,'dv) repr = ('c,'dv) code
+module C = struct type ('c,'dv) repr = ('c,'dv) code
   let int (x:int)   = .<x>.           let bool (b:bool) = .<b>.
@@
 % inductively because P below depends on repr_pe
-by
+by\vspace*{-0.8em}
 \begin{code3}
@@
 \end{code3}
+\vspace*{-0.8em}
 Although we can use type classes to define this type function
@@
 \begin{code}
-module type Symantics = sig
-  type ('c,'sv,'dv) repr
+module type Symantics = sig type ('c,'sv,'dv) repr
   val int : int  -> ('c,int,int) repr
@@
   let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
-
   let app ef ea = match ef with {st = Some f} -> f ea

### 2007-06-12 tagless/tagless-final.tex
16 pages

@@
 operation as the corresponding metalanguage operation
-\begin{code3}
+\vspace*{-0.8em}\begin{code3}
 module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
@@
 \end{code3}
+\vspace*{-0.8em}
 %
@@
 the size of each object term).
-%
+%aplas-specific (cont)
+We can also generalize from~|R| to all interpreters; the proof
+of these propositions follow immediately from the soundness of the
+metalanguage's type system.
+%end aplas-specific
 \begin{proposition}
@@
 \end{proposition}
-Generalizing from~|R| to all interpreters, we have the following
-broader and more useful, if also more obvious, proposition.
+% Generalizing from~|R| to all interpreters, we have the following
+% broader and more useful, if also more obvious, proposition.
 \begin{proposition}
@@
 \end{proposition}
-These propositions follow immediately from the soundness of the
-metalanguage's type system.
+% These propositions follow immediately from the soundness of the
+% metalanguage's type system.
 % aplas-commented
@@
 any interpretive overhead: the
-code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| and does not
-  call the interpreter, unlike the recursive calls to |eval0| and
-  |eval| in the |L e| lines in \S\ref{tagproblem}.
+code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| 
+% aplas: and does not
+% call the interpreter, unlike the recursive calls to |eval0| and
+% |eval| in the |L e| lines in \S\ref{tagproblem}.
 The resulting code obviously contains no tags and no pattern matching.
 %aplas: the line is for aplas. The original paragraph commented out
-The accompanying code shows the corresponding Haskell implemention. 
+The accompanying code shows the Haskell implementation. 
 \begin{comment}
@@
 Roughly, a partial evaluator interprets each object term to either
-a static (present-stage) term (using the evaluator~|R|) or
-a dynamic (future-stage) term (using the compiler~|C|).  To
+a static (present-stage) term (using~|R|) or
+a dynamic (future-stage) term (using~|C|).  To
 distinguish between static and dynamic terms, one might define
-the Haskell data type 
 %aplas-specific
@@
 
-Our |Symantics| class includes only separate lifting methods |bool| and
-|int|, not a parametrically polymorphic lifting method, for good reason:
+Our |Symantics| class only includes separate lifting methods |bool| and
+|int|, not a parametrically polymorphic lifting method:
 When compiling to a first-order target language such as machine code,
@@
 Thus polymorphic lift cannot be compiled without intensional type
-analysis.  To avoid the need for polymorphic lift, we turn to
+analysis.  To avoid this, we turn to
 \citearound{'s technique}\citet[see also \citealp{sumii-hybrid}]{asai-binding-time}:
@@
 alternative constructors of a |Maybe| value, |Just| and |Nothing|,
-tag each partially evaluated term with a phase: either present or
+tag each term with a phase: present or
 future.  This tag is not an object type tag: all pattern matching below
@@
 %aplas-specific
-|abstr1 (P1 _ dyn) = dyn| of the type
+|abstr1 (P1 _ dyn) = dyn| of type
 |abstr1 :: P1 t -> C t|
@@
 constructs of our object language, including |if_|.
+\vspace*{-0.7em}
 \begin{code3}
@@
 \end{code3}
+\vspace*{-0.7em}
 When we come to functions, however, we stumble.  According to our
@@
 |P1| to a function type |a -> b| should yield one of
+\vspace*{-0.7em}
 \begin{code3}
@@
 \end{code3}
+\vspace*{-0.7em}
 That is, we need a nonparametric data type, something akin to
@@
 \cite{oliveira-typecase}.  These
-methods are equivalent, and here we show the GADT way; |incope1.hs|
+methods are equivalent, and here we use GADTs; |incope1.hs|
 in the accompanying source code shows the latter.
@@
 % abbreaviated text
-We introduce a GADT with four data constructors.
+\vspace*{-0.7em}
 \begin{code3}
@@
 \end{code3}
+\vspace*{-0.7em}
 The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
@@
 but we delay it to apply |f| to concrete arguments. The interpretation of
+\vspace*{-0.7em}
 \begin{code3}
@@
 \end{code3}
+\vspace*{-0.7em}
 checks to see if |ef| is such a delayed
@@
 This solution using GADTs works but is not quite satisfactory. First, it
-cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
+cannot be ported to MetaOCaml.  Second,
 the problem of nonexhaustive pattern\hyp matching reappears in
@@
 We introduce a GADT with four data constructors.
-\begin{code3}
+\vspace*{-0.7em}
 data P t where
@@
 A simple and portable solution thus emerges: we bake |repr_pe| 
-into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
-portability, we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
+into the signature |Symantics|. Switching to MetaOCaml to demonstrate this,
+we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
 constructor took two arguments |'c| and~|'dv|. We add an argument
 |'sv| for the result of applying |repr_pe| to~|'dv|.
-Figure~\ref{fig:ocaml} shows the new |Symantics| signature.
+Figure~\ref{fig:ocaml} shows the new signature.
 % aplas: abbreviated figure
@@
 eliminated when one operand is statically~$0$; similarly for
-multiplication by~$0$ or~$1$.  Such algebraic simplifications are easy
+multiplication by~$0$ or~$1$.  
+\begin{comment}
+Such algebraic simplifications are easy
 to abstract over the specific domain (such as monoid or ring) where they
@@
 language with more base types and primitive operations.
+\end{comment}
+%aplas-specific
+In large languages, this can be quite effective.
 
@@
 gives the desired result
+\vspace*{-0.7em}
 \begin{code3}
@@
 \end{code3}
+\vspace*{-0.7em}
 
@@
 % aplas: abbreviated. The original text is in comments
-\Citet{honsell99coinductive}
-investigate the untyped $\lambda$-calculus,
-along this line.
+(see \Citet{honsell99coinductive} for the untyped case).
 % aplas: comment
@@
 because we represent object types as metalanguage types.  This work
-makes it safer and more efficient to embed domain\hyp specific languages
+makes it safer and more efficient to embed %domain\hyp specific languages
+DSLs
 in practical metalanguages such as Haskell and ML\@.
@@
 
-Our representation of object terms in the elimination form encodes
-primitive recursive folds over the terms. This makes operations like
-interpretation and size calculations trivial to implement. It remains
-the subject of the current work to see if and how non-primitively 
-recursive operations can be supported in our approach.
+Our representation of object terms in elimination form encodes
+primitive recursive folds over the terms. 
+%aplas This makes operations like interpretation trivial to implement. 
+We still
+have to understand if and how non-primitively 
+recursive operations can be supported.
 

### 2007-06-13 tagless/final_code/incope.hs
arranging the code

@@ -1,6 +1,9 @@
 {-# OPTIONS -fglasgow-exts -W #-}
 
 -- Interpreter, Compiler, Partial Evaluator
+-- Code accompanying the paper by
+--   Jacques Carette, Oleg Kiselyov, and Chung-chieh Shan
+
 
 module Incope where
 

### 2007-06-13 tagless/final_code/incope1.hs

@@ -2,6 +2,10 @@
 {-# OPTIONS -fallow-undecidable-instances #-}
 
 -- Interpreter, Compiler, Partial Evaluator
+-- The Partial Evaluator is implemented with typeclasses instead of GADTs
+--
+-- Code accompanying the paper by
+--   Jacques Carette, Oleg Kiselyov, and Chung-chieh Shan
 
 module Incope1 where
 

### 2007-06-13 tagless/tagless-final.tex
arranging the code
small typographical corrections

@@
 have been proposed to distinguish closed terms like |test1| from open
-terms like |test3|
+terms 
+%aplas-comment: like |test3|
 %aplas \cite{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual},
@@
 %aplas
-Hindley-Milner type system
+Hindley-Milner type system%
 %aplas \cite{hindley-principal,milner-theory}
@@
 \cite{metaocaml}) and standard Haskell. The complete code is available
-\url{\oleg{put URL!}.}
+\url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
 to supplement the paper. Our examples below switch between (Meta)OCaml
@@
 \item The partial evaluator avoids polymorphic lift and delays binding-time
-    analysis.  It bakes a type-to-type map into the interpreter
+    ana-lysis.  It bakes a type-to-type map into the interpreter
     interface to eliminate the need for GADTs and thus remain portable
@@
 class Symantics repr where
-  int  :: Int  -> repr Int       bool :: Bool -> repr Bool
+  int  :: Int  -> repr Int;       bool :: Bool -> repr Bool
   lam :: (repr a -> repr b) -> repr (a -> b)

### 2007-06-14 SCP2008/paper/metamonads.tex
make sure it latex cleanly
very minor edits, more to come

@@
 \begin{abstract}
-we use multi-stage programming, monads and Ocaml's
+We use multi-stage programming, monads and Ocaml's
 advanced module system to demonstrate how to eliminate all
@@
 code.  We demonstrate this clearly with LU decomposition as a 
-representative family of symbolic and numerical algorithms, and also
+representative family of symbolic and numeric algorithms, and also
 include an application to ODE solving (via a Runge-Kutta algorithm).
@@
 the resulting code is generated just right and not changed afterwards,
-we enjoy MetaOCaml's guaranty that the generated code is well-typed.
+MetaOCaml guarantees that the generated code is well-typed.
 We further demonstrate that various abstraction parameters (aspects)
 can be made orthogonal and compositional, even in the presence of
-name-generation for temporaries and other bindings and
+name-generation for temporaries, and 
 ``interleaving'' of aspects.  We also show how to encode some
 domain-specific knowledge so that ``clearly wrong'' compositions can
-be statically rejected by the compiler when processing the generator
-rather than the generated code.
+be rejected by the compiler at generation time rather than at a later
+stage.
 \end{abstract}
@@
 
-In high-performance, symbolic, and numeric computing, there is a
+In high-performance symbolic and numeric computing, there is a
 well-known issue of balancing between maximal performance and the
@@
 in linear algebra, there is a wealth of different aspects that
-\emph{may} need to be addressed. For example, implementations of the
+\emph{may} need to be addressed. Implementations of the
 widely used LU decomposition algorithm (which subsumes Gaussian 
@@
 the matrix, whether to compute and return the determinant or rank, how
-and whether search for pivot, etc. Furthermore, current architectures
+and whether search for a pivot, etc. Furthermore, current architectures
 demand more and more frequent tweaks which, in general, cannot be done by the
 compiler because the tweaking often involves domain knowledge.
+
 A survey \cite{carette04} of
 Gaussian elimination implementations in the industrial package Maple
-found 6 different aspects and 35 different implementations of the
-algorithm, as well as 45 implementations of directly related
-algorithms (LU decomposition, Cholesky decomposition, etc).  We can manually
-write each of these implementations optimizing for particular aspects and using
-cut-and-paste to ``share'' similar pieces of code.
+found 6 clearly identifiable aspects and 35 different implementations of the
+algorithm, as well as 45 implementations of directly related algorithms such as
+LU decomposition, Cholesky decomposition, and so on.  We could
+manually write each of these implementations optimizing for particular aspects
+and using cut-and-paste to ``share'' similar pieces of code.
 Or we can write a very generic procedure that accounts for
@@
 normalization, and potentially a relative size measure. The simplest
-case of such domain abstraction is |param_code1| in
-Fig.~\ref{easycode}. There, code-generators such as |plus| and |one|
+case of such domain abstraction is |param_code1|.
+There, code-generators such as |plus| and |one|
 were passed as arguments. We need far more than

### 2007-06-29 SCP2008/paper/metamonads.tex
Updated for the new code. Added notes

@@
 \begin{code}
-type ('v,'s,'w) monad = 's -> ('s -> 'v -> 'w) -> 'w
-let ret (a :'v) : ('v,'s,'w) monad = fun s k -> k s a
+type ('p,'v) monad = 's -> ('s -> 'v -> 'w) -> 'w
+    constraint 'p = <state : 's; answer : 'w; ..>
+
+let ret (a :'v) : ('p,'v) monad = fun s k -> k s a
 let bind a f = fun s k -> a s (fun s' b -> f b s' k)
-let fetch s k = k s s  and  store v s k = k (v::s) ()
-
-let k0 s v = v
-let runM m = m [] k0
-
-let l1 f = fun x -> perform \{ t <-- x; f t\}
-let l2 f = fun x y -> perform \{ tx <-- x; ty <-- y; f tx ty\}
+let fetch s k = k s s  and  store v _ k = k v ()
+
+let k0 _ v = v
+let runM m = fun s0 -> m s0 k0
+
+let l1 f = fun x     -> perform t <-- x; f t
+let l2 f = fun x y   -> perform tx <-- x; ty <-- y; f tx ty
 
@@
 let ifM test th el = fun s k -> 
-  k s .< if .~(test s k0) then .~(th s k0) else .~(el s k0) >.
+  k s .< if .~test then .~(th s k0) else .~(el s k0) >.
 \end{code}
@@
 construct computations that do have specific effects.  Operations |fetch| and
-|store v| construct computations that read and write the state. In our
-case the state is a list (of polymorphic variants), which models an
-open discriminated union, as we shall see later.
+|store v| construct computations that read and write the state.
 
@@
 
-Finally, |runM| runs our monad, that is, performs the computation of
+Finally, |runM| runs our monad, that is, given the initial state,
+performs the computation of
 the monad and returns its result, which in our case is the code
@@
                 ret (plus ce (plus x ce)))
-  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
+  in .<fun x y -> .~(runM (gen .<x>. .<y>.) ())>.
 \end{code}
@@
 let param_code4 plus one =
-  let gen x y = mdo \{ ce <-- retN (plus y one);
-                      ret (plus ce (plus x ce)) \}
-  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
+  let gen x y = perform ce <-- retN (plus y one);
+                        ret (plus ce (plus x ce))
+  in .<fun x y -> .~(runM (gen .<x>. .<y>.) ())>.
 \end{code}
@@
 \begin{code}
-let ifM' test th el = perform \{
+let ifM' test th el = perform
   testc <-- test; thc <-- th; elc <-- el;
-  ifL testc thc elc\}
+  ifL testc thc elc
 let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                    (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
- in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
+ in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
 \evalresult{.<fun a_1 i_2 -> \protect\newline
@@
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
- in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
+ in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
 \evalresult{.<fun a_1 i_2 -> 
@@
 case-matching statements and |for|- and |while|-loops.
-\oleg{put them here?}
+
+\begin{code}
+let loopM low high body = function
+  | UP -> (fun s k -> 
+      k s .< for j = .~low to .~high do .~(body .<j>. s k0) done >. )
+  | DOWN -> (fun s k -> 
+      k s .< for j = .~low downto .~high do .~(body .<j>. s k0) done >. )
+
+let whileM cond body = fun s k -> 
+  k s .< while .~(cond) do .~(body s k0) done >.
+
+let matchM x som non = fun s k -> k s .< match .~x with
+           | Some i -> .~(som .<i>. s k0)
+           | None   -> .~(non s k0) >.
+
+let genrecloop gen rtarg = fun s k -> 
+  k s .<let rec loop j = .~(gen .<loop>. .<j>. s k0) in loop .~rtarg>.
+\end{code}
+
+|loopM| is the analogue of the for-loop.
+
+(* We now define a lot of infrastructure.  We will be quite
+   thorough, and define base abstraction and lifted abstractions,
+   This will involve a lot of boilerplate code, which unfortunately 
+   cannot be so easily automated in MetaOCaml -- that would require 
+   introspection.  It could be done in camlp4, but that seems too 
+   much as well.  This 'base' could be elided, but when we decide
+   to make more use of Abstract Interpretation, we'll regret it,
+   so do it now. *)
+
+\begin{code}
+let lift x = .< x >.
+
+let liftRef x = .< ref .~x >. 
+let liftGet x = .< ! .~x >. 
+let unitL = fun s k -> k s .< () >.
+
+let liftPair x = (.< fst .~x >., .< snd .~x >.)
+
+(* logic code combinators - plain and monadic *)
+module Logic = struct
+  let notL a        = .< not .~a >.
+  let equalL a b    = .< .~a = .~ b >.
+  let notequalL a b = .< .~a <> .~ b >.
+  let andL a b     = .< .~a && .~b >. 
+end
+
+(* operations on code indices *)
+module Idx = struct
+  let zero = .< 0 >.
+  let one = .< 1 >.
+  let minusone = .< -1 >.
+  let succ a = .< .~a + 1 >.
+  let pred a = .< .~a - 1 >.
+  let less a b = .< .~a < .~b >.
+  let uminus a = .< - .~a >.
+  let add a b = .< .~a + .~b >.
+
+let cunit = .< () >.
+let update a f = let b = f (liftGet a) in .< .~a := .~b >.
+let assign a b = .< .~a := .~b >.
+let apply  f x = .< .~f .~x >.
+let updateM a f = ret (update a f)
+let assignM a b = ret (assign a b)
+let applyM  f x = ret (apply f x)
+\end{code}
+
+maintauining the state
+
+cf Atlas: no unbound variable.
+state-changing monad...
+extensible object... No worry about init
+lots of infrastructure...
+
+mention MLTon.
+The reason openrec are so challenging in Ocaml is
+because we want to store pieces of code, and exceptions
+can't be polymorphic...
+The benefit of explicit tags: no need to care about
+sharing of modules like Tracking, etc. fewer errors.
+
+In our
+case the state is a list (of polymorphic variants), which models an
+open discriminated union, as we shall see later.
 
@@
 \begin{code}
+type domain_kind = Domain_is_Ring | Domain_is_Field
+
+module S(T:
+  sig
+	(* Representation type of values,  to be specified *)
+    type ('a, 'b) rep
+  end) = struct
+
+open T
+
 module type DOMAIN = sig
-  type v    type 'a vc = ('a,v) code
-  type kind (* Field or Ring ? *)
-  val zero : 'a vc   val one : 'a vc
-  val plus : 'a vc -> 'a vc -> ('a vc, 's, 'w) monad
-  (* times, minus, uminus, div elided for brevity *)
-  val better_than : ('a vc -> 'a vc -> 
-      (('a,bool) code, 's, 'w) monad) option
-  val normalizerf : (('a,v -> v) code ) option
+  type v
+  val kind : domain_kind
+  val zero : v
+  val one : v
+  val plus : v -> v -> v
+  val times : v -> v -> v
+  val minus : v -> v -> v
+  val uminus : v -> v
+  val div : v -> v -> v
+  val better_than : (v -> v -> bool) option
+  val normalizer : (v -> v) option
 end
-module IntegerDomain : DOMAIN = struct
-  type v = int  type kind = domain_is_ring
+
+module type DOMAINL = sig
+  include DOMAIN
+  type 'a vc = ('a,v) rep
+  val zeroL : 'a vc
+  val oneL : 'a vc
+  val ( +^ ) : 'a vc -> 'a vc -> 'a vc
+  val ( *^ ) : 'a vc -> 'a vc -> 'a vc
+  val ( -^ ) : 'a vc -> 'a vc -> 'a vc
+  val uminusL : 'a vc -> 'a vc
+  val divL : 'a vc -> 'a vc -> 'a vc
+  val better_thanL : ('a vc -> 'a vc -> ('a,bool) rep) option
+  val normalizerL : ('a vc -> 'a vc) option
+end
+
+module IntegerDomain = struct
+    type v = int
+    let kind = Domain_is_Ring
+    let zero = 0
+    let one = 1
+    let plus x y = x + y
+    let times x y = x * y
+    let minus x y = x - y
+    let uminus x = -x
+    let div x y = x / y
+    let normalizer = None
+    let better_than = Some (fun x y -> abs x > abs y)
+end
+
+module T = Domains_sig.S(struct type ('a, 'b) rep = ('a, 'b) code end)
+open T
+
+open Domains_common
+(* because the operations are "syntactic" to a certain extent,
+   we have to repeat ourselves a lot *)
+module IntegerDomainL = struct
+    include IntegerDomain
     type 'a vc = ('a,v) code
-  let zero = .< 0 >.  and one = .< 1 >. 
-  let plus x y = ret .<.~x + .~y>. 
-  let better_than = Some (fun x y -> ret .<abs .~x > abs .~y >. )
-  let normalizerf = None
-  ...
+    let zeroL = .< 0 >.  
+    let oneL = .< 1 >. 
+    let (+^) x y = .<.~x + .~y>. 
+    let ( *^ ) x y = .<.~x * .~y>.
+    let ( -^ ) x y = .<.~x - .~y>.
+    let uminusL x = .<- .~x>.
+    let divL x y = .<.~x / .~y>. 
+    let normalizerL = None
+    let better_thanL = Some (fun x y -> .<abs .~x > abs .~y >. )
 end
+
 \end{code}
@@
 \begin{code2}
-module type OUTPUT = sig
-  type contr  type res
-  module D : DETERMINANT   module R : RANK   module P : TRACKPIVOT
-  val make_result : ('a,contr) code -> 
-    (('a,res) code,
-     [> `TDet of 'a D.lstate | `TRan of 'a R.lstate | `TPivot of 'a P.lstate]
-       list, ('a,'w) code) monad
+(* The `keyword' list of all the present internal features *)
+module type INTERNAL_FEATURES = sig
+  module R      : TrackRank.RANK
+  module P      : TRACKPIVOT
+  module L      : LOWER
 end
-module OutDetRank(Dom:DOMAIN)(C: CONTAINER2D)
-    (Det : DETERMINANT with type indet = Dom.v and type outdet = Dom.v)
-    (Rank : RANK) = struct
-  module Ctr = C(Dom)
-  type contr = Ctr.contr
-  type res = contr * Det.outdet * int
-  module D = Det   module R = Rank   module P = DiscardPivot
-  let make_result b = perform \{ det  <-- D.fin ();  rank <-- R.fin ();
-    ret .< ( .~b, .~det, .~rank ) >. \}
+
+module type OUTPUT = functor(OD : OUTPUTDEP) -> sig
+  module IF : INTERNAL_FEATURES
+  type res
+  val make_result : 'a wmatrix ->
+   (<classif : 'a; 
+     state : [> `TDet of 'a OD.Det.lstate |
+                'a IF.R.tag_lstate |
+                `TPivot of 'a IF.P.lstate |
+                `TLower of 'a IF.L.lstate]; ..>,res) cmonad
+   (*
+    ('a,res,[> 'a OD.Det.tag_lstate | 'a IF.R.tag_lstate 
+             | 'a IF.P.tag_lstate   | 'a IF.L.tag_lstate],'w) cmonad
+      *)
+end
+module OutDetRank(OD : OUTPUTDEP) = struct
+  module IF = struct
+      module R   = Rank
+      module P   = DiscardPivot
+      module L   = NoLower end
+  type res = C.contr * C.Dom.v * int
+  let make_result m = perform
+    det  <-- OD.Det.fin ();
+    rank <-- IF.R.fin ();
+    ret (Tuple.tup3 m.matrix det rank)
+  (* Initialization: check the preconditions of instantiation of this struct*)
+  let _ = OD.Det.fin ()
+  let _ = IF.R.fin ()
 end
@@
 
+\begin{code}
+(* The `keyword' list of all the present external features *)
+module type FEATURES = sig
+  module Det       : DETERMINANT
+  module PivotF    : PIVOT
+  module PivotRep  : PIVOTKIND
+  module Update    : UPDATE
+  module Input     : INPUT
+  module Output    : OUTPUT
+end
+\end{code}
+
 \begin{figure}
 \begin{code2}
-module Gen(Dom: DOMAIN)(C: CONTAINER2D)(PivotF: PIVOT)
-          (Update: UPDATE with type baseobj = Dom.v and type ctr = C(Dom).contr)
-          (Out: OUTPUT with type contr = C(Dom).contr and type D.indet = Dom.v 
-                        and type 'a D.lstate = 'a Update.D.lstate) = struct
-    module Ctr = C(Dom)
-    module Pivot = PivotF(Dom)(C)(Out.D)
-    let gen =
-      let zerobelow b r c m n brc =
-        let innerbody i = perform \{
-            bic <-- Ctr.get b i c;
-            whenM (l1 LogicCode.not (LogicCode.equal bic Dom.zero ))
-                (seqM (retLoopM (Idx.succ c) (Idx.pred m)
-                          (fun k -> Update.update b r c i k) )
-                      (Ctr.set b i c Dom.zero)) \} in 
-        perform \{
-              seqM (retLoopM (Idx.succ r) (Idx.pred n) innerbody) 
-                   (Update.update_det brc) \} in
-      let dogen a = perform \{
-          r <-- Out.R.decl ();
+module GenGE(F : FEATURES) = struct
+    module O = F.Output(F)
+
+    let wants_pack = O.IF.L.wants_pack
+    let can_pack   = 
+        let module U = F.Update(F.Det) in
+        (U.upd_kind = DivisionBased)
+    (* some more pre-flight tests *)
+    let _ = ensure ((not wants_pack) || can_pack) 
+           "Cannot return a packed L in this case"
+
+    let zerobelow mat pos = 
+        let module IF = O.IF in
+        let module U = F.Update(F.Det) in
+        let innerbody j bjc = perform
+            whenM (Logic.notequalL bjc C.Dom.zeroL ) (perform
+                det <-- F.Det.get ();
+                optSeqM (Iters.col_iter mat.matrix j (Idx.succ pos.p.colpos) 
+               (Idx.pred mat.numcol) C.getL
+                      (fun k bjk -> perform
+                      brk <-- ret (C.getL mat.matrix pos.p.rowpos k);
+                      U.update bjc pos.curval brk bjk 
+                          (fun ov -> C.col_head_set mat.matrix j k ov) det) UP )
+                      (IF.L.updt mat.matrix j pos.p.colpos C.Dom.zeroL 
+                          (* this makes no sense outside a field! *)
+                          (C.Dom.divL bjc pos.curval))) in
+        perform
+              seqM (Iters.row_iter mat.matrix pos.p.colpos
+              (Idx.succ pos.p.rowpos)
+              (Idx.pred mat.numrow) C.getL innerbody UP) 
+                   (U.update_det pos.curval)
+
+   let init input = perform
+        let module IF = O.IF in
+          (a,rmar,augmented) <-- F.Input.get_input input;
+          r <-- IF.R.decl ();
           c <-- retN (liftRef Idx.zero);
-          b <-- retN (Ctr.mapper Dom.normalizerf (Ctr.copy a));
-          m <-- retN (Ctr.dim1 a);
-          n <-- retN (Ctr.dim2 a);
-          () <-- Update.D.decl ();
-          () <-- Out.P.decl ();
-          seqM 
-            (retWhileM (LogicCode.and_ (Idx.less (liftGet c) m)
-                                       (Idx.less (liftGet r) n) )
-               ( perform \{
+          b <-- retN (C.mapper C.Dom.normalizerL (C.copy a));
+          m <-- retN (C.dim1 a);
+          rmar <-- retN rmar;
+          n <-- if augmented then retN (C.dim2 a) else ret rmar;
+          F.Det.decl ();
+          IF.P.decl rmar;
+          _ <-- IF.L.decl (if wants_pack then b else C.identity rmar m);
+          let mat = {matrix=b; numrow=n; numcol=m} in
+          ret (mat, r, c, rmar)
+
+   let forward_elim (mat, r, c, rmar) = perform
+        let module IF = O.IF in
+          whileM (Logic.andL (Idx.less (liftGet c) mat.numcol)
+                              (Idx.less (liftGet r) rmar) )
+             ( perform
              rr <-- retN (liftGet r);
              cc <-- retN (liftGet c);
-               pivot <-- l1 retN (Pivot.findpivot b rr m cc n);
-               seqM (retMatchM pivot (fun pv -> 
-                        seqM (zerobelow b rr cc m n pv)
-                             (Out.R.succ ()) )
-                        (Update.D.zero_sign () ))
-                    (Code.update c Idx.succ) \} ))
-            (Out.make_result b) \} in
-    .<fun a -> .~(runM (dogen .<a>.)) >.
+             let cp  = {rowpos=rr; colpos=cc} in
+             let module Pivot = F.PivotF(F.Det)(IF.P) in
+             pivot <-- l1 retN (Pivot.findpivot mat cp);
+             seqM (matchM pivot (fun pv -> 
+                      seqM (zerobelow mat {p=cp; curval=pv} )
+                           (IF.R.succ ()) )
+                      (F.Det.zero_sign () ))
+                  (updateM c Idx.succ) )
+
+   let gen input = perform
+          (mat, r, c, rmar) <-- init input;
+          seqM 
+            (forward_elim (mat, r, c, rmar))
+            (O.make_result mat)
 end
@@
 use with MetaOCaml. Many helpful discussions with Walid Taha are very
-appreciated.
+appreciated. The implementation of the monadic notation, |perform|,
+was the joint work with Lydia van Dijk.
 

### 2007-06-30 SCP2008/paper/metamonads.tex
more changes

@@
 \DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\small}
-\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\footnotesize}
+\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\small}
 \newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
@@
 \begin{abstract}
-We use multi-stage programming, monads and Ocaml's
+We use multi-stage programming, monads and OCaml's
 advanced module system to demonstrate how to eliminate all
@@
 representative family of symbolic and numeric algorithms, and also
-include an application to ODE solving (via a Runge-Kutta algorithm).
+include an application to solving ordinary differential equations 
+(via a Runge-Kutta algorithm).
 We parameterize our code to a great extent
@@
 domain-specific knowledge so that ``clearly wrong'' compositions can
-be rejected by the compiler at generation time rather than at a later
-stage.
+be rejected at or before generation time rather than during
+the compilation or running of the generated code.
 \end{abstract}
@@
 \begin{keyword}
-MetaOcaml \sep linear algebra \sep genericity \sep generative \sep staging
+MetaOCaml \sep linear algebra \sep genericity \sep generative \sep staging
 \sep Functor \sep symbolic.
@@
 \cite{KiselyovTaha}.  But more problems remain
-\cite{Padua:MetaOcaml:04}: generating binding statements (``names''),
-especially when generating loop bodies or conditional branches; making
+\cite{Padua:MetaOcaml:04}: generating binding forms (``names'')
+when generating loop bodies or conditional branches; making
 continuation-passing style (CPS) code clear.  Many authors
@@
 In this paper, we report on our continued progress \cite{CaretteKiselyov05}
+\oleg{should we say in the footnote: ``compared to the earlier,
+  conference version of this paper we describe a different
+  version of our generator dealing with the complete LU decomposition
+  and solving. We worked out previously missing aspects of in-place
+  updates, representaing permutation matrix, augmented input matrix,
+  and back-propagation. We have changed the representation of
+  domain-specific knowledge about permissible compositions of aspects.''}
+the code generating LU decomposition and solving 
 solving these problems using linear solving and LU decomposition
 as our running example. Specifically, our contributions:
-%\vspace*{-2pt}
 \begin{itemize}
@@
 LU decomposition algorithm family in completely separate,
-independent modules.  We briefly discuss related work in
+independent modules.  Section~\ref{s:ode} describes generating
+Runge-Kutta solvers for ordinary differential equations (ODE).
+We briefly discuss related work in
 section~\ref{related}. We then outline the future work and conclude.
@@
 let param_code2 plus one =
-  let gen x y k = letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
+  let gen x y k = letgen (plus y one)
+                         (fun ce -> k (plus ce (plus x ce)))
   and k0 x = x
@@
 
-Comparison of the code that did let-insertion at the generating stage
+Comparison of the let-insertion in the generator
 \begin{code}
@@
 \end{code}
-with the corresponding code inserting let at the generated code stage
+with the corresponding code generating let-insertion for the future
+stage
 \begin{code}
@@
 What was |let ce = init in ...| in direct style became
-|init' (fun ce -> ...)| in CPS. For one thing, |let| became
+|init' (fun ce -> ...)| in CPS. For one, |let| became
 ``inverted''. For another, what used to be an expression that yields
@@
 yields the answer of the type |w|.  The monad is polymorphic over the
-three type parameters.  Other implementations are
-possible. Except for the code in Figure~\ref{ourmonad}, the rest of our code
-treats the monad as a truly abstract data type. The implementation of
-the basic
-monadic operations |ret| and |bind| is conventional and clearly
-satisfies the monadic laws. Other monadic operations
-construct computations that do have specific effects.  Operations |fetch| and
-|store v| construct computations that read and write the state.
+three type parameters, which would require |monad| to be a type
+constructor with three arguments. When we use this monad for code
+generation, we will need yet another type variable, environment
+classifier \oleg{cite Walid03?}. With type constructors taking more
+and more arguments, it becomes more difficult to read and write
+types -- which we will be doing extensively when writing module
+signatures in Section XXX. The fact that OCaml renames all type
+variables when printing out types confuses matters further. An elegant
+solution to these kinds of problems has been suggested by 
+Garrigue on the Caml mailing list 
+(cited from
+\url{http://groups.google.com/group/fa.caml/msg/e80b1245702d6b24}
+no date, no exact ref). We use a single type parameter |'p| to
+represent all parameters of our monad (all parameters but the type of
+the monadic value |'v|). The type variable |'p| is constrained to be
+the type of an object with methods (fields) |state| and |answer|. The
+object may include more fields, represented by |..|. Values of that
+type are not part of our computations and need not exist. We merely
+use the object type, as an convenient way to specify extensible
+\emph{type-level} records in OCaml.
+
+Our monad could be implemented in other ways. Except for the code in
+Figure~\ref{ourmonad}, the rest of our code treats the monad as a
+truly abstract data type. The implementation of the basic monadic
+operations |ret| and |bind| is conventional and clearly satisfies the
+monadic laws. Other monadic operations construct computations that do
+have specific effects.  Operations |fetch| and |store v| construct
+computations that read and write the state.
 
@@
 camlp4 pre-processor, we introduce the |perform|-notation \cite{metamonadsURL},
-patterned after the |do|-notation of Haskell. 
+patterned after the |do|-notation of Haskell (see App.~\ref{app:perform}).
 \begin{code}
@@
 already generated pieces of code rather than monadic values. We
-``lift it'', see |ifM'|. 
+``lift it'':
 \begin{code}
@@
   ifL testc thc elc
-let gen a i = ifM' (ret .<(.~i) >= 0>.) 
-                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
- in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
-\evalresult{.<fun a_1 i_2 -> \protect\newline
-let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
-let gen a i = ifM (ret .<(.~i) >= 0>.) 
-                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
- in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
-\evalresult{.<fun a_1 i_2 -> 
-if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
-\end{code}
-%
+\end{code}
 We define functions |l1|,
@@
 The difference between them is
-apparent: in the code above with |ifM'|, the let-insertion
+apparent from the following example:
+\begin{code}
+let gen a i = ifM' (ret .<(.~i) >= 0>.) 
+                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
+ in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
+\evalresult{.<fun a_1 i_2 ->}
+\textcolor{red}{      let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
+let gen a i = ifM (ret .<(.~i) >= 0>.) 
+                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
+ in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
+\evalresult{.<fun a_1 i_2 ->}
+\textcolor{red}{      if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
+\end{code}
+%
+If we use |ifM'| to generate guarded array access code, the let-insertion
 happened \emph{before} the if-expression, that is, before the test that
@@
 effects when generating control constructions such as conditionals.
-The form |ifM| handles such effects correctly. We
-need similar operators for other OCaml control forms: for generating
-case-matching statements and |for|- and |while|-loops.
-
-\begin{code}
-let loopM low high body = function
-  | UP -> (fun s k -> 
-      k s .< for j = .~low to .~high do .~(body .<j>. s k0) done >. )
-  | DOWN -> (fun s k -> 
-      k s .< for j = .~low downto .~high do .~(body .<j>. s k0) done >. )
+The form |ifM| handles such effects correctly.
+
+We need similar operators for other OCaml control forms: for
+generating sequencing, case-matching statements and |for|- and |while|-loops.
+\begin{code}
+let seqM a b = fun s k -> 
+  k s .< begin .~(a s k0) ; .~(b s k0) end >.
 
@@
 \end{code}
-
-|loopM| is the analogue of the for-loop.
-
-(* We now define a lot of infrastructure.  We will be quite
-   thorough, and define base abstraction and lifted abstractions,
-   This will involve a lot of boilerplate code, which unfortunately 
-   cannot be so easily automated in MetaOCaml -- that would require 
-   introspection.  It could be done in camlp4, but that seems too 
-   much as well.  This 'base' could be elided, but when we decide
-   to make more use of Abstract Interpretation, we'll regret it,
-   so do it now. *)
-
-\begin{code}
-let lift x = .< x >.
-
-let liftRef x = .< ref .~x >. 
-let liftGet x = .< ! .~x >. 
-let unitL = fun s k -> k s .< () >.
-
-let liftPair x = (.< fst .~x >., .< snd .~x >.)
-
-(* logic code combinators - plain and monadic *)
-module Logic = struct
-  let notL a        = .< not .~a >.
-  let equalL a b    = .< .~a = .~ b >.
-  let notequalL a b = .< .~a <> .~ b >.
-  let andL a b     = .< .~a && .~b >. 
-end
-
-(* operations on code indices *)
-module Idx = struct
-  let zero = .< 0 >.
-  let one = .< 1 >.
-  let minusone = .< -1 >.
-  let succ a = .< .~a + 1 >.
-  let pred a = .< .~a - 1 >.
-  let less a b = .< .~a < .~b >.
-  let uminus a = .< - .~a >.
-  let add a b = .< .~a + .~b >.
-
-let cunit = .< () >.
-let update a f = let b = f (liftGet a) in .< .~a := .~b >.
-let assign a b = .< .~a := .~b >.
-let apply  f x = .< .~f .~x >.
-let updateM a f = ret (update a f)
-let assignM a b = ret (assign a b)
-let applyM  f x = ret (apply f x)
-\end{code}
-
-maintauining the state
-
-cf Atlas: no unbound variable.
-state-changing monad...
-extensible object... No worry about init
-lots of infrastructure...
-
-mention MLTon.
-The reason openrec are so challenging in Ocaml is
-because we want to store pieces of code, and exceptions
-can't be polymorphic...
-The benefit of explicit tags: no need to care about
-sharing of modules like Tracking, etc. fewer errors.
-
-In our
-case the state is a list (of polymorphic variants), which models an
-open discriminated union, as we shall see later.
 
@@
 \cite{EWD:EWD447}: information hiding and separation of concerns.  To
-apply these principles to the study of Gaussian Elimination, we need
+apply these principles to the study of LU decomposition, we need
 to understand what are the changes between different implementations, and 
@@
       pivoting matrix are to be returned. In the larger algorithm
-      family, routines like Maple's
+      family, routines like Maple's\\
       \texttt{LinearAlgebra:-LUDecomposition} have up to $2^6 +
@@
 different domains must satisfy:
-\begin{figure}[h]
+
 \begin{code}
 type domain_kind = Domain_is_Ring | Domain_is_Field
-
-module S(T:
-  sig
-	(* Representation type of values,  to be specified *)
-    type ('a, 'b) rep
-  end) = struct
-
-open T
 
@@
 
+module IntegerDomain : DOMAIN = struct
+    type v = int
+    let kind = Domain_is_Ring
+    let zero = 0
+    let one = 1
+    let plus x y = x + y
+    let times x y = x * y
+    let minus x y = x - y
+    let uminus x = -x
+    let div x y = x / y
+    let normalizer = None
+    let better_than = Some (fun x y -> abs x > abs y)
+end
+\end{code}
+
+\begin{code}
+module S(T:
+  sig
+	(* Representation type of values,  to be specified *)
+    type ('a, 'b) rep
+  end) = struct
+
+open T
 module type DOMAINL = sig
@@
   val normalizerL : ('a vc -> 'a vc) option
-end
-
-module IntegerDomain = struct
-    type v = int
-    let kind = Domain_is_Ring
-    let zero = 0
-    let one = 1
-    let plus x y = x + y
-    let times x y = x * y
-    let minus x y = x - y
-    let uminus x = -x
-    let div x y = x / y
-    let normalizer = None
-    let better_than = Some (fun x y -> abs x > abs y)
 end
@@
 end
-
-\end{code}
-\end{figure}
+\end{code}
+
+We now define a lot of infrastructure.  We will be quite thorough, and
+define base abstraction and lifted abstractions, This will involve a
+lot of boilerplate code, which unfortunately cannot be so easily
+automated in MetaOCaml -- that would require introspection.  It could
+be done in camlp4, but that seems too much as well.  This 'base' could
+be elided, but when we decide to make more use of Abstract
+Interpretation, we'll regret it, so do it now.
+\begin{code}
+let lift x = .< x >.
+
+let liftRef x = .< ref .~x >. 
+let liftGet x = .< ! .~x >. 
+let unitL = fun s k -> k s .< () >.
+
+let liftPair x = (.< fst .~x >., .< snd .~x >.)
+
+(* logic code combinators - plain and monadic *)
+module Logic = struct
+  let notL a        = .< not .~a >.
+  let equalL a b    = .< .~a = .~ b >.
+  let notequalL a b = .< .~a <> .~ b >.
+  let andL a b     = .< .~a && .~b >. 
+end
+
+(* operations on code indices *)
+module Idx = struct
+  let zero = .< 0 >.
+  let one = .< 1 >.
+  let minusone = .< -1 >.
+  let succ a = .< .~a + 1 >.
+  let pred a = .< .~a - 1 >.
+  let less a b = .< .~a < .~b >.
+  let uminus a = .< - .~a >.
+  let add a b = .< .~a + .~b >.
+
+let cunit = .< () >.
+let update a f = let b = f (liftGet a) in .< .~a := .~b >.
+let assign a b = .< .~a := .~b >.
+let apply  f x = .< .~f .~x >.
+let updateM a f = ret (update a f)
+let assignM a b = ret (assign a b)
+let applyM  f x = ret (apply f x)
+\end{code}
 
@@
 domain |'a vc|, and once more from values to monadic computations
-|('a vc, 's, 'w) monad|.
+|('p,'a vc) monad|.
 
@@
 methods with incorrect types will be caught statically, even
-\emph{before} any code generation takes place. The abstract type
-|domain_is_ring| encodes a semantic constraint that the full division
+\emph{before} any code generation takes place. The variant
+|domain_is_ring| of |domain_kind| encodes a semantic constraint 
+that the full division
 is not available. While the |DOMAIN| type may have looked daunting to
@@
 straightforward.  More interesting is the aspect of what to return
-from the GE algorithm.  One could create an algebraic data type (as
+from the LU algorithm.  One could create an algebraic data type (as
 was done in \cite{carette04}) to encode the various choices: the
@@
 the signature set the result type differently. Given below is this
-module type and one instantiation, which specifies the output of a GE
+module type and one instantiation, which specifies the output of a LU
 algorithm as a 3-tuple |contr * Det.outdet * int| of the U-factor, the
@@
 
+\subsection{Maintaining the state}
+another title: representing extensible state
+
+different aspects may need to keep their own state. We wish to assure
+modularity. One approach: using objects. The state is an object, each
+aspect is an instnace variable with some particular name. The problem
+is how to represent the initial `empty' state. Need to create the
+object with all fields set to None. But we need to know the names of
+all fields; as we add an aspect, our function that runs the generator
+must change to account for the different initial object with different
+slots.
+
+A better approach: property list (ref MLTon). Property list represents
+an extensible object. The initial object is just the empty list. Alas,
+the literal MLton's approach doesn't apply to us well. First of all,
+due to the environment classifiers, we need universes parameterized by
+classifiers. Also, the propoerty names are generated automatically in
+MLton approach, based on generativity of exceptions or reference
+cells. That means if some particular aspect happens to be included
+twice, the state will be incompatible. We will have to worry about
+sharing or the lack of sharing of our aspects. We would rather prefer
+our aspects to be stateless and work the same way whether two
+instances of the aspect are shared or not. So, we find polymorphic
+unions to be ideal. Manifest naming
+
+Encoding an object by its dual,
+list of polymorphic variants.
+
+Currently, we check at genertor-time that one cannot add a slot which
+already exists (already the part of the state) nor we can obtain the
+value of a non-existing slot. These problems may occur if our genetor
+is wrong (calls fin method before decl method of an aspect). It is
+possible to make these checks static (TFP2007, state-variable monad?)
+state-changing monad...
+
+\subsection{tracking}
+
 As is apparent from the output choices, several different quantities
-\emph{may} need to be tracked in a particular GE implementation.  We
+\emph{may} need to be tracked in a particular LU implementation.  We
 therefore need to be able to conditionally generate variables
@@
 value of the determinant, converting the tracking state
-to the final determinant value of the type |outdet|. GE of a
+to the final determinant value of the type |outdet|. LU of a
 floating-point matrix with no determinant tracking uses the
@@
 it. The determinant tracking aspect is complex because tracking
-variables, if any, are to be declared at the beginning of GE; the sign
+variables, if any, are to be declared at the beginning of LU; the sign
 of the determinant has to be updated on each row or column
@@
 
-The GE generator functor itself is 
+The LU generator functor itself is 
 parameterized by the domain, container, pivoting policy (full, row,
@@
 fully inlined; no extra functions are invoked and no tests other than those
-needed by the GE algorithm itself are performed. The GE function returns a
+needed by the LU algorithm itself are performed. The GE?? function returns a
 triple |int array * int * int| of the U-factor, determinant and the rank. The
@@
 contains benchmark code and timing comparisons.
+
+\section{Runge-Kutta solvers}
+\label{s:ode}
 
@@
 
-ATLAS \cite{ATLAS} is another successful project in this area.  However
-they use much simpler weaving technology, which leads them to note
-that \emph{generator complexity tends to go up along with flexibility, 
-so that these routines become almost insurmountable barriers to 
-outside contribution}.  Our results show how to surmount this barrier,
-by building modular, composable generators.
-SPIRAL \cite{Pueschel:05} is another such even more ambitious project.
-But SPIRAL does intentional code analysis, relying on a set of code
-transformation ``rules'' which make sense, but which are not proved 
-to be either complete or confluent.  The strength of both of these
-project relies on their platform-specific optimizations performed
-via search techniques, something we have not attempted here.
+ATLAS \cite{ATLAS} is another successful project in this area.
+However they use much simpler weaving technology, which leads them to
+note that \emph{generator complexity tends to go up along with
+  flexibility, so that these routines become almost insurmountable
+  barriers to outside contribution}. Our results show how to surmount
+this barrier, by building modular, composable generators. A
+significant part of ATLAS' complexity is that the generator is
+extremely error-prone and difficult to debug.  Indeed, when generating
+C code in C using |printf|, nothing prevents producing code that
+misses semicolons, open or close parentheses or variable
+bindings. MetaOCaml gives us assurance that these errors, and more
+subtle type errors, shall not occur in the generated code.  SPIRAL
+\cite{Pueschel:05} is another such even more ambitious project.  But
+SPIRAL does intentional code analysis, relying on a set of code
+transformation ``rules'' which make sense, but which are not proved to
+be either complete or confluent.  The strength of both of these
+project relies on their platform-specific optimizations performed via
+search techniques, something we have not attempted here.
 
@@
 \section{Appendix 0}
+\label{app:perform}
 \oleg{grammar of our perform monad. Thank Lydia}
@@
 \section{Appendix A}
-The code generated for |GenIV5|, fraction-free GE of the integer matrix
+The code generated for |GenIV5|, fraction-free LU of the integer matrix
 represented by a flat vector, full pivoting, returning the |U|-factor,
@@
 \section{Appendix B}
-The code generated for |GenFA1|, GE of the floating point matrix
+The code generated for |GenFA1|, LU of the floating point matrix
 represented by a 2D array, row pivoting, returning just the |U|-factor.

### 2007-07-03 SCP2008/paper/metamonads.tex
more writeup

@@
 
+One particular domain instance is |IntegerDomain|. The notation\\
+|module IntegerDomain : DOMAIN| makes the compiler verify that our
+|IntegerDomain| is indeed a |DOMAIN|, that is, satisfies the required
+signature. The constraint |DOMAIN| may be omitted; in that case, the
+compiler will verify the type when we try to use that structure as a
+|DOMAIN|. In any case, the errors such as missing ``methods'' or
+methods with incorrect types will be caught statically, even
+\emph{before} any code generation takes place. The variant
+|domain_is_ring| of |domain_kind| encodes a semantic constraint 
+that the full division
+is not available. While the |DOMAIN| type may have looked daunting to
+some, the implementation is quite straightforward.  Other domains such
+as |float| and arbitrary precision exact rational numbers |Num.num|
+are equally simple.
+
+A more complex domain: |Zp|
+\begin{code}
+module ZpMake(P:sig val p:int end) = struct
+    type v = int
+    let kind = Domain_is_Field
+    let zero = 0 and one = 1
+    let plus x y = (x + y) mod P.p
+    let times x y = (x * y) mod P.p
+    \dots
+    let normalizer = None and better_than = None
+    let () = assert (is_prime P.p)
+end
+\end{code}
+
+This domain is parametrized by a an integer number |p|, the
+(modulus?). Thus, ZpMake is a functor. It creates a domain that is a
+field, but with no defined order. Hence
+|normalizer| and |better_than| are set to |None|. However, |Zp| forms
+a field only when |p| is prime. Therefore, we must make this check,
+see the last line of that code. The expression 
+|assert (is_prime P.p)| is the initializing expression. It is executed
+when the correspo0nding module is instantaited
+\begin{code}
+module Z19 = ZpMakeL(struct let p = 19 end)
+\end{code}
+if we replace |19| with |9| above, we receive an error. That is a
+run-time error. It is raised however as we instantiate and
+combine modules that will eventually make the generator. Thus,
+although the error is reported at run-time rather than compile time as
+one might have hoped, the error is raised when \emph{generating the
+generator} -- well before the generation of the target code could
+begin. In our code we make extensive use of these so-called preflight 
+checks performed as part of module initialization. The checks seem to
+offer a good compromise: they are dynamic and so do not require
+complicated type systems; on the other hand, the checks are performed
+quite early, when buiding code generators, and so ensure that no 
+code violating the corresponding \emph{semantic} constraints 
+will be generated. Although some may frown on the use of the module
+initializing expressions (since that requires careful attention to the
+sharing and problems of multiple instantiation of a module), these
+concerns do not apply to our case: our preflight tests are all idempotent
+and maintain no state.
+
+Lifting:
+
+We now define a lot of infrastructure.  We will be quite thorough, and
+define base abstraction and lifted abstractions, This will involve a
+lot of boilerplate code, which unfortunately cannot be so easily
+automated in MetaOCaml -- that would require introspection.  It could
+be done in camlp4, but that seems too much as well.  This 'base' could
+be elided, but when we decide to make more use of Abstract
+Interpretation, we'll regret it, so do it now.
+\begin{code}
+let lift x = .< x >.
+
+let liftRef x = .< ref .~x >. 
+let liftGet x = .< ! .~x >. 
+let unitL = fun s k -> k s .< () >.
+
+let liftPair x = (.< fst .~x >., .< snd .~x >.)
+
+(* logic code combinators - plain and monadic *)
+module Logic = struct
+  let notL a        = .< not .~a >.
+  let equalL a b    = .< .~a = .~ b >.
+  let notequalL a b = .< .~a <> .~ b >.
+  let andL a b     = .< .~a && .~b >. 
+end
+
+(* operations on code indices *)
+module Idx = struct
+  let zero = .< 0 >.
+  let one = .< 1 >.
+  let minusone = .< -1 >.
+  let succ a = .< .~a + 1 >.
+  let pred a = .< .~a - 1 >.
+  let less a b = .< .~a < .~b >.
+  let uminus a = .< - .~a >.
+  let add a b = .< .~a + .~b >.
+
+let cunit = .< () >.
+let update a f = let b = f (liftGet a) in .< .~a := .~b >.
+let assign a b = .< .~a := .~b >.
+let apply  f x = .< .~f .~x >.
+let updateM a f = ret (update a f)
+let assignM a b = ret (assign a b)
+let applyM  f x = ret (apply f x)
+\end{code}
+
+\noindent  The types above are
+generally lifted twice: once from the value domain |v| to the code
+domain |'a vc|, and once more from values to monadic computations
+|('p,'a vc) monad|.
+
 \begin{code}
@@
 
-We now define a lot of infrastructure.  We will be quite thorough, and
-define base abstraction and lifted abstractions, This will involve a
-lot of boilerplate code, which unfortunately cannot be so easily
-automated in MetaOCaml -- that would require introspection.  It could
-be done in camlp4, but that seems too much as well.  This 'base' could
-be elided, but when we decide to make more use of Abstract
-Interpretation, we'll regret it, so do it now.
-\begin{code}
-let lift x = .< x >.
-
-let liftRef x = .< ref .~x >. 
-let liftGet x = .< ! .~x >. 
-let unitL = fun s k -> k s .< () >.
-
-let liftPair x = (.< fst .~x >., .< snd .~x >.)
-
-(* logic code combinators - plain and monadic *)
-module Logic = struct
-  let notL a        = .< not .~a >.
-  let equalL a b    = .< .~a = .~ b >.
-  let notequalL a b = .< .~a <> .~ b >.
-  let andL a b     = .< .~a && .~b >. 
-end
-
-(* operations on code indices *)
-module Idx = struct
-  let zero = .< 0 >.
-  let one = .< 1 >.
-  let minusone = .< -1 >.
-  let succ a = .< .~a + 1 >.
-  let pred a = .< .~a - 1 >.
-  let less a b = .< .~a < .~b >.
-  let uminus a = .< - .~a >.
-  let add a b = .< .~a + .~b >.
-
-let cunit = .< () >.
-let update a f = let b = f (liftGet a) in .< .~a := .~b >.
-let assign a b = .< .~a := .~b >.
-let apply  f x = .< .~f .~x >.
-let updateM a f = ret (update a f)
-let assignM a b = ret (assign a b)
-let applyM  f x = ret (apply f x)
-\end{code}
-
-\noindent  The types above are
-generally lifted twice: once from the value domain |v| to the code
-domain |'a vc|, and once more from values to monadic computations
-|('p,'a vc) monad|.
-
-One particular domain instance is |IntegerDomain|. The notation\\
-|module IntegerDomain : DOMAIN| makes the compiler verify that our
-|IntegerDomain| is indeed a |DOMAIN|, that is, satisfies the required
-signature. The constraint |DOMAIN| may be omitted; in that case, the
-compiler will verify the type when we try to use that structure as a
-|DOMAIN|. In any case, the errors such as missing ``methods'' or
-methods with incorrect types will be caught statically, even
-\emph{before} any code generation takes place. The variant
-|domain_is_ring| of |domain_kind| encodes a semantic constraint 
-that the full division
-is not available. While the |DOMAIN| type may have looked daunting to
-some, the implementation is quite straightforward.  Other domains such
-as |float| and arbitrary precision exact rational numbers |Num.num|
-are equally simple.
+\begin{code}
+module type CONTAINER2D = sig
+  module Dom:DOMAINL
+  type contr
+  type 'a vc = ('a,contr) rep
+  type 'a vo = ('a,Dom.v) rep
+  val getL : 'a vc -> ('a,int) rep -> ('a,int) rep -> 'a vo
+  val dim1 : 'a vc -> ('a,int) rep
+  val dim2 : 'a vc -> ('a,int) rep
+  val mapper : ('a vo -> 'a vo) option -> 'a vc -> 'a vc
+  val copy : 'a vc -> 'a vc
+  val init : ('a,int) rep -> ('a, int) rep -> 'a vc
+  val augment : 'a vc -> ('a,int) rep -> ('a, int) rep -> 'a vc ->
+                ('a, int) rep -> 'a vc
+  val identity : ('a,int) rep -> ('a, int) rep -> 'a vc
+  val swap_rows_stmt : 'a vc -> ('a, int) rep -> ('a, int) rep -> 
+                       ('a,unit) rep
+  val swap_cols_stmt : 'a vc -> ('a, int) rep -> ('a, int) rep -> 
+                       ('a,unit) rep
+  val row_head : 'a vc -> ('a, int) rep -> ('a, int) rep -> 'a vo
+  val col_head_set : 'a vc -> ('a,int) rep -> ('a,int) rep -> 'a vo -> 
+            ('a,unit) rep
+end
+\end{code}
 
 Parameterizing by the kind of container representing a matrix is
-almost as straightforward.  Our containers are  parametric
-over a |DOMAIN|, i.e., functors from a |DOMAIN| module
-to the actual implementation of a container. The functor signature
-|CONTAINER2D| specifies that a container must provide functions |dim1|
-and |dim2| to extract the dimensions, functions |get| and |set| to
-generate container getter and setters, the cloning generator |copy|
-and functions that generate code for row and column swapping. The
-inclusion of these functions in the signature of all containers makes
-it simpler to optimize the relevant functions depending on the actual
-representation of the container while not burdening the users of
-containers with efficiency details.
+almost as straightforward.  The type of our containers include the
+lifted domain |Dom| as one of the component. Particular containers are
+Our containers are  parametric
+over a |DOMAINL|, i.e., functors from a |DOMAINL| module
+to the actual implementation of a container. For example, the
+following functor defines a matrix container represented as arrays of
+rows.
+
+\begin{code}
+module GenericArrayContainer(Dom:DOMAINL) =
+  struct
+  module Dom = Dom
+  type contr = Dom.v array array (* Array of rows *)
+  type 'a vc = ('a,contr) code
+  type 'a vo = ('a,Dom.v) code
+  let getL x n m = .< (.~x).(.~n).(.~m) >.
+  let dim2 x = .< Array.length .~x >.       (* number of rows *)
+  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
+  \dots
+\end{code}
+%
+We also have containers wheer elements are stored in a 1D array, in a
+row-wise (C-like) or column-wise (Fortran-like) modes.
+
+The signature |CONTAINER2D| specifies that a container must provide
+functions |dim1| and |dim2| to extract the dimensions, functions |getL|
+to generate container getters, the cloning
+generator |copy| and functions that generate code for row and column
+swapping. The inclusion of these functions in the signature of all
+containers makes it simpler to optimize the relevant functions
+depending on the actual representation of the container while not
+burdening the users of containers with efficiency details.
 
@@
 \begin{code2}
+module type OUTPUTDEP = sig 
+    module PivotRep : PIVOTKIND 
+    module Det      : DETERMINANT
+end
 (* The `keyword' list of all the present internal features *)
@@
   type res
-  val make_result : 'a wmatrix ->
-   (<classif : 'a; 
-     state : [> `TDet of 'a OD.Det.lstate |
-                'a IF.R.tag_lstate |
-                `TPivot of 'a IF.P.lstate |
-                `TLower of 'a IF.L.lstate]; ..>,res) cmonad
-   (*
-    ('a,res,[> 'a OD.Det.tag_lstate | 'a IF.R.tag_lstate 
-             | 'a IF.P.tag_lstate   | 'a IF.L.tag_lstate],'w) cmonad
-      *)
+  val make_result : 'a wmatrix -> (\dots,res) monad
 end
@@
 \end{code2}
+
+We have several Rank aspects:
+\begin{code2}
+module Rank = struct
+  include TrackRank
+  let fin () = perform
+      r <-- mo_lookup ip;
+      ret (liftGet r)
+end
+
+module NoRank = struct
+  include TrackRank
+  let fin () = failwith "Rank is needed but is not computed"
+end
+\end{code2}
+
+Both contain the |fin ()| function to generate the code representing
+the computed rank. Only the |NoRank| does not track any rank, and so 
+|fin ()| raises an error. The code 
+|let _ = IF.R.fin ()| in |OutDetRank| invokes that fin function, which
+will produce the monadic code generating action, or raise teh
+error. We do not run the action at that time -- we only make sure
+there is an action to run. This is the preflight check again, to rule
+out the semantic error when the user specified that rank should be
+computed and returned and yet specified the |NoRank| aspects which
+computes no rank at all.
 

### 2007-07-04 SCP2008/paper/metamonads.tex
updating the text with the new version of the code

@@
 
-We have several Rank aspects:
-\begin{code2}
-module Rank = struct
-  include TrackRank
-  let fin () = perform
-      r <-- mo_lookup ip;
-      ret (liftGet r)
-end
-
-module NoRank = struct
-  include TrackRank
-  let fin () = failwith "Rank is needed but is not computed"
-end
-\end{code2}
-
-Both contain the |fin ()| function to generate the code representing
-the computed rank. Only the |NoRank| does not track any rank, and so 
-|fin ()| raises an error. The code 
-|let _ = IF.R.fin ()| in |OutDetRank| invokes that fin function, which
-will produce the monadic code generating action, or raise teh
-error. We do not run the action at that time -- we only make sure
-there is an action to run. This is the preflight check again, to rule
-out the semantic error when the user specified that rank should be
-computed and returned and yet specified the |NoRank| aspects which
-computes no rank at all.
+The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
+the pre-flight tests, to be explained in Section XXX.
 
@@
 module type DETERMINANT = sig
-  type indet  type outdet  type 'a lstate
-  type tdet = outdet ref   
-  val decl : unit -> 
-    (unit, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
-  val upd_sign : unit -> 
-    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
-  ...
+  type tdet = C.Dom.v ref
+  type 'a lstate
+  type ('pc,'v) lm = ('pc,'v) cmonad
+    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
+  type ('pc,'v) om = ('pc,'v) omonad
+    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
+  val decl : unit -> ('b,unit) lm (* could be unit rather than unit code...*)
+  val upd_sign  : unit -> ('b,unit) om
+  val zero_sign : unit -> ('b,unit) lm
+  val acc       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
+  val get       : unit -> ('b,tdet) lm
+  val set       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
+  val fin       : unit -> ('b,C.Dom.v) lm
 end
@@
 
+We have several instances of Det
+\begin{code}
+module NoDet =
+  struct
+  type tdet = C.Dom.v ref
+  type 'a lstate = unit
+  let decl () = unitL
+  let upd_sign () = ret None
+  let zero_sign () = unitL
+  let acc _ = unitL
+  let get () = ret (liftRef C.Dom.zeroL) (* hack alert! *)
+  let set _ = unitL
+  let fin () = failwith "Determinant is needed but not computed"
+  type ('pc,'v) lm = ('pc,'v) cmonad
+    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
+  type ('pc,'v) om = ('pc,'v) omonad
+    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
+end
+
+module AbstractDet =
+  struct
+  open C.Dom
+  type tdet = v ref
+  (* the first part of the state is an integer: which is +1, 0, -1:
+     the sign of the determinant *)
+  type 'a lstate = ('a,int ref) abstract * ('a,tdet) abstract
+  let fin = fun () -> perform
+      (det_sign,det) <-- mo_lookup ip;
+      ifM (Logic.equalL (liftGet det_sign) Idx.zero) (ret zeroL)
+      (ifM (Logic.equalL (liftGet det_sign) Idx.one) (ret (liftGet det))
+          (ret (uminusL (liftGet det))))
+end
+\end{code}
+
+Both contain the |fin ()| function to generate the code representing
+the computed rank. Only the |NoRank| does not track any rank, and so 
+|fin ()| raises an error. The code 
+|let _ = IF.R.fin ()| in |OutDetRank| invokes that fin function, which
+will produce the monadic code generating action, or raise teh
+error. We do not run the action at that time -- we only make sure
+there is an action to run. This is the preflight check again, to rule
+out the semantic error when the user specified that rank should be
+computed and returned and yet specified the |NoRank| aspects which
+computes no rank at all.
+
 \begin{code}
@@
 
-\begin{figure}
 \begin{code2}
@@
 \end{code2}
-\end{figure}
 
@@
 \begin{code}
-module GenIV5 = Gen(IntegerDomain)
-   (GenericVectorContainer)(FullPivot)
-   (FractionFreeUpdate(IntegerDomain)(GenericVectorContainer)(IDet))
-   (OutDetRank(IntegerDomain)(GenericVectorContainer)(IDet)(Rank))
-module GenFA1 = Gen(FloatDomain)
-   (GenericArrayContainer)(RowPivot)
-   (DivisionUpdate(FloatDomain)(GenericArrayContainer)(NoDet(FloatDomain)))
-   (OutJustMatrix(FloatDomain)(GenericArrayContainer)(NoDet(FloatDomain)))
-\end{code}
-\vspace*{-5pt}
+module GVC_I = GenericVectorContainer(IntegerDomainL)
+module G_GVC_I = GenLA(GVC_I)
+open G_GVC_I
+open G_GVC_I.GE
+module GenIV5 = GenGE(struct 
+    module Det = AbstractDet
+    module PivotF = FullPivot
+    module PivotRep = PermList
+    module Update = FractionFreeUpdate
+    module Input = InpJustMatrix
+    module Output = OutDetRank end)
+
+module GAC_F = GenericArrayContainer(FloatDomainL)
+module G_GAC_F = GenLA(GAC_F)
+open G_GAC_F
+open G_GAC_F.GE
+
+module GenFA1 = GenGE(struct
+    module Det = NoDet
+    module PivotF = RowPivot
+    module PivotRep = PermList
+    module Update = DivisionUpdate
+    module Input = InpJustMatrix 
+    module Output = OutJustMatrix end)
+module GenFA9 = GenGE(struct 
+    module Det = NoDet
+    module PivotF = RowPivot
+    module PivotRep = PermList
+    module Update = DivisionUpdate
+    module Input = InpJustMatrix
+    module Output = Out_LU_Packed end)
+
+let resIV5 = instantiate GenIV5.gen ;;
+let rIV5 = runit {pf =  resIV5 };;
+
+\end{code}
+|GenFA1| is no-frills GE; GenFA9 has LU part
+
 We can instantiate the |Gen| functor as shown above and inspect the generated
@@
 \label{app:perform}
-\oleg{grammar of our perform monad. Thank Lydia}
+Grammar of our perform monad
+We support four different constructs to introduce a monadic
+expressions.
+
+\begin{code}
+  perform exp
+  perform exp1; exp2
+  perform x <-- exp1; exp2
+  perform let x = foo in exp
+\end{code}
+
+which is almost literally the grammar of the Haskell's "do"-notation,
+with the differences that Haskell uses |do| and |<-| where we use
+|perform| and |<--|.
+
+We support not only |let x = foo in ...|  expressions but arbitrarily
+complex let-expressions, including |let rec| and |let module|.
+
+The actual bind function of the monad defaults to |bind| and the
+match-failure function to |failwith| (only used for refutable
+patterns; see below).  To select a different function, use the
+extended forms of |perform|. For example, use the function named 
+|bind| from module |Mod|.  In
+addition use the module's |failwith|-function in refutable patterns.
+\begin{code}
+        perform with module Mod in exp2
+\end{code}
+The code has full explanation...
 
@@
 \begin{code2}
-# val resIV5 : ('a,
-   Funct4.GenIV5.Ctr.contr ->
-   Funct4.OutDetRank(Funct4.IntegerDomain)(Funct4.GenericVectorContainer)
-                    (Funct4.IDet)(Funct4.Rank).res) code =
-  .<fun a_405 ->
-   let t_406 = (ref 0) in let t_407 = (ref 0) in
-   let t_408 = {arr = (Array.copy a_405.arr)} (a_405) in
-   let t_409 = a_405.m in let t_410 = a_405.n in
-   let t_411 = (ref 1) in let t_412 = (ref 1) in
-   while (((! t_407) < t_409) && ((! t_406) < t_410)) do
-    let t_413 = (! t_406) in let t_414 = (! t_407) in
-    let t_415 = (ref (None)) in
-    let t_435 =
+val resIV5 : ('a, GVC_I.contr -> GenIV5.O.res) code =
+  .<fun a_1 ->
+   let t_2 = (ref 0) in
+   let t_3 = (ref 0) in
+   let t_4 = {arr = (Array.copy a_1.arr)} (a_1) in
+   let t_5 = a_1.m in
+   let t_6 = a_1.n in
+   let t_7 = (ref 1) in
+   let t_8 = (ref 1) in
+   while (((! t_3) < t_5) && ((! t_2) < t_6)) do
+    let t_13 = (! t_2) in
+    let t_14 = (! t_3) in
+    let t_15 = (ref (None)) in
+    let t_34 =
      begin (* full pivoting *)
-      for j_431 = t_413 to (t_410 - 1) do
-       for j_432 = t_414 to (t_409 - 1) do
-        let t_433 = (t_408.arr).((j_431 * t_408.m) + j_432) in
-        if (not (t_433 = 0)) then
-         (match (! t_415) with
-          | Some (i_434) ->
-             if ((abs (snd i_434)) > (abs t_433)) then
-              (t_415 := (Some ((j_431, j_432), t_433))) else ()
-          | None -> (t_415 := (Some ((j_431, j_432), t_433))))
+      for j_30 = t_13 to (t_6 - 1) do
+       for j_31 = t_14 to (t_5 - 1) do
+        let t_32 = (t_4.arr).((j_30 * t_4.m) + j_31) in
+        if (t_32 <> 0) then
+         (match (! t_15) with
+          | Some (i_33) ->
+             if ((abs (snd i_33)) > (abs t_32)) then
+              (t_15 := (Some ((j_30, j_31), t_32)))
+             else ()
+          | None -> (t_15 := (Some ((j_30, j_31), t_32))))
         else ()
@@
       done;
-      (match (! t_415) with
-       | Some (i_416) ->  (* swapping of columns *)
-          if ((snd (fst i_416)) <> t_414) then begin
-           let a_424 = t_408.arr and nm_425 = (t_408.n * t_408.m)
-           and m_426 = t_408.m in
-           let rec loop_427 =
-            fun i1_428 -> fun i2_429 ->
-              if (i2_429 < nm_425) then
-               let t_430 = a_424.(i1_428) in
-               a_424.(i1_428) <- a_424.(i2_429);
-               a_424.(i2_429) <- t_430;
-               (loop_427 (i1_428 + m_426) (i2_429 + m_426))
+      (match (! t_15) with
+       | Some (i_16) -> (* swapping of columns *)
+          if ((snd (fst i_16)) <> t_14) then begin
+           let a_23 = t_4.arr
+           and nm_24 = (t_4.n * t_4.m)
+           and m_25 = t_4.m in
+           let rec loop_26 =
+            fun i1_27 ->
+             fun i2_28 ->
+              if (i2_28 < nm_24) then
+               let t_29 = a_23.(i1_27) in
+               a_23.(i1_27) <- a_23.(i2_28);
+               a_23.(i2_28) <- t_29;
+               (loop_26 (i1_27 + m_25) (i2_28 + m_25))
               else () in
-           (loop_427 t_414 (snd (fst i_416)));
-           (t_412 := (~- (! t_412))) (* adjust the sign of det *)
-          end else (); (* swapping of rows elided *)
-          (Some (snd i_416))
+           (loop_26 t_14 (snd (fst i_16)));
+           (t_8 := (~- (! t_8))) (* adjust the sign of det *)
+          end else ();
+          if ((fst (fst i_16)) <> t_13) then begin (* swapping of rows?*)
+           let a_17 = t_4.arr
+           and m_18 = t_4.m in
+           let i1_19 = (t_13 * m_18)
+           and i2_20 = ((snd (fst i_16)) * m_18) in
+           for i_21 = 0 to (m_18 - 1) do
+            let t_22 = a_17.(i1_19 + i_21) in
+            a_17.(i1_19 + i_21) <- a_17.(i2_20 + i_21);
+            a_17.(i2_20 + i_21) <- t_22
+           done;
+           (t_8 := (~- (! t_8)))
+          end else ();
+          (Some (snd i_16))
        | None -> (None))
      end in
-    (match t_435 with
-     | Some (i_436) ->
+    (match t_34 with
+     | Some (i_35) ->
         begin (* elimination loop *)
-         for j_437 = (t_413 + 1) to (t_410 - 1) do
-          if (not ((t_408.arr).((j_437 * t_408.m) + t_414) = 0)) then begin
-           for j_438 = (t_414 + 1) to (t_409 - 1) do
-            (t_408.arr).((j_437 * t_408.m) + j_438) <-
-             ((((t_408.arr).((j_437 * t_408.m) + j_438) * (* elided *)
+         for j_36 = (t_13 + 1) to (t_6 - 1) do
+          let t_37 = (t_4.arr).((j_36 * t_4.m) + t_14) in
+          if (t_37 <> 0) then begin
+           for j_38 = (t_14 + 1) to (t_5 - 1) do
+            (t_4.arr).((j_36 * t_4.m) + j_38) <-
+             ((((t_4.arr).((j_36 * t_4.m) + j_38) * i_35) -
+                ((t_4.arr).((t_13 * t_4.m) + j_38) * t_37)) / (! t_7))
            done;
-           (t_408.arr).((j_437 * t_408.m) + t_414) <- 0
+           (t_4.arr).((j_36 * t_4.m) + t_14) <- 0
           end else ()
-         done; (t_411 := i_436)
+         done;
+         (t_7 := i_35)
         end;
-        (t_406 := ((! t_406) + 1)) (* advance the rank *)
-     | None -> (t_412 := 0));
-    (t_407 := ((! t_407) + 1))
+        (t_2 := ((! t_2) + 1)) (* advance the rank *)
+     | None -> (t_8 := 0));
+    (t_3 := ((! t_3) + 1))
    done;
-   (t_408, 
-    if ((! t_412) = 0) then 0 (* adjust the sign of the determinant *)
-    else if ((! t_412) = 1) then (! t_411)
-    else (~- (! t_411)), (! t_406))>.
+   (t_4,
+    if ((! t_8) = 0) then 0 (* adjust the sign of the determinant *)
+    else if ((! t_8) = 1) then (! t_7)
+    else (~- (! t_7)), (! t_2))>.
 \end{code2}
@@
 \begin{code2}
-# val resFA1 : ('a,
-   Funct4.GenFA1.Ctr.contr ->
-   Funct4.OutJustMatrix(Funct4.FloatDomain)(Funct4.GenericArrayContainer)
-                       (Funct4.NoDet(Funct4.FloatDomain)).res) code =
+val resFA1 : ('a, GAC_F.contr -> GenFA1.O.res) code =
   .<fun a_1 ->
-   let t_2 = (ref 0) in let t_3 = (ref 0) in
+   let t_2 = (ref 0) in
+   let t_3 = (ref 0) in
    let t_5 = (Array.map (fun x_4 -> (Array.copy x_4)) (Array.copy a_1)) in
@@
    while (((! t_3) < t_6) && ((! t_2) < t_7)) do
-    let t_8 = (! t_2) in let t_9 = (! t_3) in
-    let t_10 = (ref (None)) in
-    let t_16 =
+    let t_10 = (! t_2) in
+    let t_11 = (! t_3) in
+    let t_12 = (ref (None)) in
+    let t_18 =
      begin  (* row pivoting *)
-      for j_13 = t_8 to (t_7 - 1) do
-       let t_14 = (t_5.(j_13)).(t_9) in
-       if (not (t_14 = 0.)) then
-        (match (! t_10) with
-         | Some (i_15) ->
-            if ((abs_float (snd i_15)) < (abs_float t_14)) then
-             (t_10 := (Some (j_13, t_14)))
+      for j_15 = t_10 to (t_7 - 1) do
+       let t_16 = (t_5.(j_15)).(t_11) in
+       if (t_16 <> 0.) then
+        (match (! t_12) with
+         | Some (i_17) ->
+            if ((abs_float (snd i_17)) < (abs_float t_16)) then
+             (t_12 := (Some (j_15, t_16)))
             else ()
-         | None -> (t_10 := (Some (j_13, t_14))))
+         | None -> (t_12 := (Some (j_15, t_16))))
        else ()
       done;
-      (match (! t_10) with
-       | Some (i_11) -> (* swapping of rows *)
-          if ((fst i_11) <> t_8) then begin
-           let t_12 = t_5.(t_8) in
-           t_5.(t_8) <- t_5.(fst i_11);
-           t_5.(fst i_11) <- t_12; () end else ();
-          (Some (snd i_11))
+      (match (! t_12) with (* swapping of rows *)
+       | Some (i_13) ->
+          if ((fst i_13) <> t_10) then
+           let t_14 = t_5.(t_10) in
+           t_5.(t_10) <- t_5.(fst i_13);
+           t_5.(fst i_13) <- t_14
+          else ();
+          (Some (snd i_13))
        | None -> (None))
      end in
-    (match t_16 with
-     | Some (i_17) ->
-        begin (* elimination loop, elided *) end;
+    (match t_18 with
+     | Some (i_19) -> (* elimination loop, elided?? *)
+        begin
+         for j_20 = (t_10 + 1) to (t_7 - 1) do
+          let t_21 = (t_5.(j_20)).(t_11) in
+          if (t_21 <> 0.) then begin
+           for j_22 = (t_11 + 1) to (t_6 - 1) do
+            (t_5.(j_20)).(j_22) <-
+             ((t_5.(j_20)).(j_22) -. ((t_21 /. i_19) *. (t_5.(t_10)).(j_22)))
+           done;
+           (t_5.(j_20)).(t_11) <- 0.
+          end else ()
+         done;
+         ()
+        end;
         (t_2 := ((! t_2) + 1))
@@
 \end{code2}
+
+\begin{code2}
+val resFA9 : ('a, GAC_F.contr -> GenFA9.O.res) code =
+  .<fun a_1 ->
+   let t_2 = (ref 0) in
+   let t_3 = (ref 0) in
+   let t_5 = (Array.map (fun x_4 -> (Array.copy x_4)) (Array.copy a_1)) in
+   let t_6 = (Array.length a_1.(0)) in
+   let t_7 = (Array.length a_1) in
+   let t_8 = (ref ([])) in
+   while (((! t_3) < t_6) && ((! t_2) < t_7)) do
+    let t_9 = (! t_2) in
+    let t_10 = (! t_3) in
+    let t_11 = (ref (None)) in
+    let t_17 =
+     begin
+      for j_14 = t_9 to (t_7 - 1) do
+       let t_15 = (t_5.(j_14)).(t_10) in
+       if (t_15 <> 0.) then
+        (match (! t_11) with
+         | Some (i_16) ->
+            if ((abs_float (snd i_16)) < (abs_float t_15)) then
+             (t_11 := (Some (j_14, t_15)))
+            else ()
+         | None -> (t_11 := (Some (j_14, t_15))))
+       else ()
+      done;
+      (match (! t_11) with
+       | Some (i_12) ->
+          if ((fst i_12) <> t_9) then begin
+           let t_13 = t_5.(t_9) in
+           t_5.(t_9) <- t_5.(fst i_12);
+           t_5.(fst i_12) <- t_13;
+           (t_8 := ((RowSwap ((fst i_12), t_9)) :: (! t_8)))
+          end else ();
+          (Some (snd i_12))
+       | None -> (None))
+     end in
+    (match t_17 with
+     | Some (i_18) ->
+        begin
+         for j_19 = (t_9 + 1) to (t_7 - 1) do
+          let t_20 = (t_5.(j_19)).(t_10) in
+          if (t_20 <> 0.) then
+           for j_21 = (t_10 + 1) to (t_6 - 1) do
+            (t_5.(j_19)).(j_21) <-
+             ((t_5.(j_19)).(j_21) -. ((t_20 /. i_18) *. (t_5.(t_9)).(j_21)))
+           done
+          else ()
+         done;
+         ()
+        end;
+        (t_2 := ((! t_2) + 1))
+     | None -> ());
+    (t_3 := ((! t_3) + 1))
+   done;
+   (t_5, (! t_8))>.
+\end{code2}
+
 \end{document}

### 2007-07-06 SCP2008/paper/metamonads.tex
partial new section layout out all the aspects in the algorithm family

@@
 \newcommand{\jacques}[1]{{\it [Jacques says: #1]}}
+
+%
+% Useful abbreviations
+%
+\newcommand{\floats}{\mathbb{F}}
+\newcommand{\reals}{\mathbb{R}}
 
@@
 
-% previous introduction is saved at the end
 \section{Introduction}
@@
 The rest of this paper is structured as follows: The next section
+gives an overview of the design space of LU decomposition algorithms,
+as well as the methodology we follow.  Then section~\ref{CPS}
 introduces code generation in MetaOCaml, the problem of name
 generation, and continuation-passing style (CPS) as a general
-solution.  We also introduce the monad and the issues of generating
+solution.  We also present a key monad and the issues of generating
 control statements. Section~\ref{functors} describes the use of
@@
 full at \cite{metamonadsURL}).
+
+\section{The design space}\label{design}
+
+Before studying the details of the technologies involved in the 
+implementation, it is worthwhile to carefully study the design space involved.
+A preliminary study~\cite{carette04} revealed a number of aspects of the 
+family of Gaussian Elimination algorithms.  In the present work, we outline
+a number of additional aspects involved in the faimily of LU decomposition 
+algorithms.  These will first be presented in a somewhat ad hoc manner, 
+roughly coresponding to the order in which they were ``discovered''.  These
+will then be reorganized into sets of semantically related aspects, which will
+form the basis of our design.  Finally, we will extract a methodology from
+our experience.
+
+Throughout, we assume that the reader is familiar with the basic LU
+decomposition algorithm, which factors a invertible matrix $A$ into a unit
+lower triangular matrix $L$ and (usually) an upper triangular matrix $U$,
+such that $A = LU$.  When pivoting is used, we get a unitary matrix $P$ such
+that the factorization is now $A = PLU$.  The case of numeric matrices is well
+covered in~\cite{Golub-vanLoan}.  When $A$ is singular, one can still get
+a $PLU$ decomposition, with $L$ remaining unit lower-triangular but with a
+$U$ which is no longer upper triangular but instead ``staggered''
+in the upper triangle.
+
+\subsection{Aspects}
+
+We reuse the english word ``aspect'' for the various facets of the family
+of LU decomposition algorithms.  While this use if very much in the same 
+spirit as in aspect-oriented programming (AOP)~\cite{Kiczales}, our 
+implementation methodology is radically different.  We firmly believe that
+our typed generative methodogy is much better suited to the functional
+programming paradigm than attempting to graft the program-trace-based
+methodology of object-oriented versions of AOP.
+
+At this point in time, it is better to think of aspects as purely
+design-time entities.  Here we are firmly influenced by Parnas' original
+view of modules and information hiding~\cite{journals/cacm/parnas72a} as well
+as his view of product families \cite{Parnas??}, and by Dijkstra's ideas on
+separation of concerns \cite{EWD:EWD447}.
+To apply these principles to the study of LU decomposition, we need
+to understand what are the changes between different implementations, and 
+what concerns need to be addressed.  We also need to study the degree
+to which these concerns are independent.
+
+The various aspects listed below all come from variations found in actual
+implementations (in various languages and settings).
+
+\begin{enumerate}
+	\item \textbf{Domain}: In which (algebraic) domain do the matrix
+		elements belong to.  Some implementations were very specific
+		($\mathbb{Z}, \mathbb{Q}, \mathbb{Z_p}, 
+		\mathbb{Z}_p\left[\alpha_1,\ldots,\alpha_n\right], 
+		\mathbb{Z}\left[x\right]$, $\mathbb{Q}\left(x\right)$, 
+		$\mathbb{Q}\left[\alpha\right]$, and floating point numbers 
+		($\floats$) for 
+		example), while others were generic for elements of a field,
+		multivariate polynomials over a field, or elements of a division ring
+		with possibly undecidable zero-equivalence.  In the roughly 85 pieces
+		of code we surveyed, 20 different domains were encountered.
+	\item \textbf{Representation of the matrix}: Whether the matrix
+		was represented as an array of arrays, a one-dimensional array,
+		a hash table, etc.  For the case of a one-dimensional array,
+		whether indexing was done in C or Fortran style.  Additionally,
+		if a particular representation had a special mechanism for efficient
+		row exchanges, this was sometimes used.
+	\item \textbf{Fraction-free}: Whether the 
+		algorithm is allowed to use unrestricted division, or only
+		exact (remainder-free) division.
+	\item \textbf{Length measure (for pivoting)}:  For stability reasons
+		(whether numerical or coefficient growth), if a domain possesses
+		an appropriate length measure, this was sometimes used to choose
+		an ``optimal'' pivot.  Not all domains have such a measure.
+	\item \textbf{Full division}: Whether the input domain supports full
+		division (i.e. is a \emph{field} or pretends to be ($\floats$)) 
+		or only exact division (i.e. a \emph{division ring}).
+	\item \textbf{Domain normalization}: Whether the arithmetic operations
+		of the base domain keep the results in normal form, or whether
+		an extra normalization step is required.  For example, some 
+		representations of polynomials require an extra step for
+		zero-testing.
+	\item \textbf{Output choices}:  Whether just the reduced matrix
+		(the `U' factor), both L and U, as well as
+		the rank, the determinant, and/or the sequence of 
+		pivots is to be returned.  For example, Maple's
+		\texttt{LinearAlgebra:-LUDecomposition} routine has
+		$2^6 + 2^5 + 2^2 = 100$ possible outputs, depending on whether
+		one chooses a $PLU$, $PLU^{1}R$ or \emph{Cholesky} 
+		decomposition.  We chose to only consider $PLU$ for now.
+	\item \textbf{Rank}: Whether to explicitly track the rank of the matrix
+		as the algorithm proceeds.
+	\item \textbf{Determinant}:  Whether to explicitly track the determinant
+		of the matrix as the algorithm proceeds.
+	\item \textbf{Code representation}: Whether we are actually generating
+		code or we are in fact representation the algorithm in ``direct
+		style'', albeit with a lot of abstraction overhead.
+	\item \textbf{Zero-equivalence}: Whether the 
+		arithmetic operations require a specialized zero-equivalence 
+		routine needs to be used.  It turns out that for certain classes
+		of \emph{expressions}, it is convenient to use a zero-equivalence
+		test which is separate from the domain normalization.  This is
+		usually the case when zero-equivalence is formally undecidable
+		but semi-algorithms or probabilistic algorithms do exist.
+		See~\cite{ZhCaJeMo07} for an example.
+    \item \textbf{Pivoting}: Whether to use no, 
+        column-wise, or full pivoting.
+    \item \textbf{Augmented Matrices}: Whether all or only some
+      columns of the matrix participate in elimination.
+	\item \textbf{Pivot representation}: Whether the pivot is represented
+	  as a list of row (and/or column) exchanges, as a unitary matrix,
+	  or as a permutation vector.
+  \item \textbf{Lower matrix}: whether the matrix $L$ should be tracked
+	  as the algorithm proceeds, reconstructed at the end of the
+	  algorithm, or not tracked at all.
+  \item \textbf{Input choices}: Whether the input matrix is an augmented
+	  matrix or not, and if so, where is the end of the main matrix.
+  \item \textbf{Packed}: Whether the output matrices $L$ and $U$ are
+	  packed into a single matrix for output.
+\end{enumerate}
+
+\noindent \jacques{Absorb this in the above} These are separated out from the
+others as they are cross-cutting concerns: in the case of the length measure, a
+property of the domain will influence the pivoting method \emph{if} pivoting is
+to be performed.\jacques{Also, should mention a few more aspects like
+inplace, errors-on-singular, logging, sparsity, as things that are not
+implemented}
+
+\subsection{Organizing aspects}
+
+\subsection{Methodology}
 
@@
 \section{Aspects and Functors}\label{functors}
+
 The monad represents fine-scale code generation. We need tools for
@@
 
-While the Object-Oriented Design community has acquired an extensive
-vocabulary for describing modularity ideas, the guiding principles for
-modular designs has not changed since they were first articulated by
-Parnas~\cite{journals/cacm/parnas72a} and Dijkstra~
-\cite{EWD:EWD447}: information hiding and separation of concerns.  To
-apply these principles to the study of LU decomposition, we need
-to understand what are the changes between different implementations, and 
-what concerns need to be addressed.  We also need to study the degree
-to which these concerns are independent.
-A study of Gaussian Elimination \cite{carette04} shows that
-the following variations occur:
-\begin{enumerate}
-%\vspace*{-6pt}
-    \item \textbf{Domain}: In which (algebraic) domain do the
-      matrix elements belong to.  Sometimes the domains are very
-      specific (e.g., $\mathbb{Z}, \mathbb{Q}, \mathbb{Z}_p$ and
-      floating point numbers), while in other cases the domains
-      were left generic, e.g., multivariate polynomials over a
-      field.  In the roughly 85 pieces of code surveyed
-      \cite{carette04} 20 different domains were encountered.
-    \item \textbf{Container}: Whether the matrix is represented as
-      an array of arrays, a one-dimensional array, a hash table, a
-      sparse matrix, etc., and whether indexing is done in C or
-      Fortran style.  Additionally, if a particular representation
-      had a special mechanism for efficient row exchanges.
-    \item \textbf{Output choices}: Whether just the reduced
-      matrix, or additionally the rank, the determinant, and the
-      pivoting matrix are to be returned. In the larger algorithm
-      family, routines like Maple's\\
-      \texttt{LinearAlgebra:-LUDecomposition} have up to $2^6 +
-      2^5 + 2^2 = 100$ outputs.
-    \item \textbf{Fraction-free}: Whether the Gaussian Elimination
-        algorithm is allowed to use unrestricted division, or only
-        exact (remainder-free) division.
-    \item \textbf{Pivoting}: Whether to use no, 
-        column-wise, or full pivoting.
-    \item \textbf{Augmented Matrices}: Whether all or only some
-      columns of the matrix participate in elimination.
-\end{enumerate}
-\noindent In addition to the above variations, there are two aspects that 
-recur frequently:
-\vspace*{-6pt}
-\begin{enumerate}
-    \item \textbf{Length measure}:  For stability reasons
-        (numerical or coefficient growth), if a domain
-      possesses
-        an appropriate length measure, this is sometimes used to choose
-        an ``optimal'' pivot.
-    \item \textbf{Normalization and zero-equivalence}: Whether the
-      arithmetic operations of the domain give normalized results, and
-      whether a specialized zero-equivalence routine is to be used.
-\end{enumerate}
-\noindent These are separated out from the others as they are cross-cutting
-concerns: in the case of the length measure, a property of the domain
-will influence the pivoting method \emph{if} pivoting is to be
-performed.
-
-The simplest parametrization is to make the domain abstract. As it
+Reviewing the various aspects outlined in section~\ref{aspects},
+the simplest parametrization is to make the domain abstract. 
+\jacques{This whole section needs to be edited to take into account
+the different organization of the material}.
+As it
 turns out, we need the following to exist in our domains: $0$, $1$,

### 2007-07-07 SCP2008/paper/metamonads.tex
editing of the latter sections

@@
 
-Reviewing the various aspects outlined in section~\ref{aspects},
-the simplest parametrization is to make the domain abstract. 
 \jacques{This whole section needs to be edited to take into account
 the different organization of the material}.
-As it
-turns out, we need the following to exist in our domains: $0$, $1$,
-$+$, $*$, (unary and binary) $-$, at least \emph{exact} division,
+
+\subsection{Domains}
+Reviewing the various aspects outlined in section~\ref{aspects}, the
+simplest parametrization is to make the domain abstract.  As it turns
+out, we need the following to exist in our domains: $0$, $1$, $+$,
+$*$, (unary and binary) $-$, at least \emph{exact} division,
 normalization, and potentially a relative size measure. The simplest
-case of such domain abstraction is |param_code1|.
-There, code-generators such as |plus| and |one|
-were passed as arguments. We need far more than
-two parameters, so we have to group them. Instead of the grouping
-offered by regular records, we use OCaml \emph{structures} (i.e.,
-modules)
-so we can take advantage of extensibility, type abstraction and constraints,
-and especially parameterized structures (\emph{functors}).
-We define the type of the domain, the signature |DOMAIN|, which
-different domains must satisfy:
+case of such domain abstraction is |param_code1|.  There,
+code-generators such as |plus| and |one| were passed as arguments. We
+need far more than two parameters, so we have to group them. Instead
+of the grouping offered by regular records, we use OCaml
+\emph{structures} (i.e., modules) so we can take advantage of
+extensibility, type abstraction and constraints, and especially
+parameterized structures (\emph{functors}).  We define the type of the
+domain, the signature |DOMAIN|, which different domains must satisfy:
 
@@
     let kind = Domain_is_Ring
-    let zero = 0
-    let one = 1
-    let plus x y = x + y
-    let times x y = x * y
-    let minus x y = x - y
+    let zero = 0 and one = 1
+    let plus x y = x + y  and  minus x y = x - y
+    let times x y = x * y and  div x y = x / y
     let uminus x = -x
-    let div x y = x / y
     let normalizer = None
@@
 \emph{before} any code generation takes place. The variant
-|domain_is_ring| of |domain_kind| encodes a semantic constraint 
+|Domain_is_Ring| of |IntegerDomain.domain_kind| encodes a semantic constraint 
 that the full division
@@
 
-A more complex domain: |Zp|
+A more complex domain is |Zp| of the field of integers in prime
+modulus:
 \begin{code}
@@
 \end{code}
-
-This domain is parametrized by a an integer number |p|, the
-(modulus?). Thus, ZpMake is a functor. It creates a domain that is a
-field, but with no defined order. Hence
-|normalizer| and |better_than| are set to |None|. However, |Zp| forms
+This domain is parametrized by an integer number |p|, the
+(modulus?). To be more precise, the structure |ZpMake| is
+parameterized over another structure of the type described by
+signature P; that argument structure has one field, the int value |p|.
+Such a parameterized structure (or, a function from structures to
+structures) is a \emph{functor}. The result of |ZpMake| is a domain that is a
+field with no defined order. Hence
+|normalizer| and |better_than| are set to |None|. |Zp| forms
 a field only when |p| is prime. Therefore, we must make this check,
-see the last line of that code. The expression 
-|assert (is_prime P.p)| is the initializing expression. It is executed
-when the correspo0nding module is instantaited
-\begin{code}
-module Z19 = ZpMakeL(struct let p = 19 end)
-\end{code}
-if we replace |19| with |9| above, we receive an error. That is a
-run-time error. It is raised however as we instantiate and
-combine modules that will eventually make the generator. Thus,
-although the error is reported at run-time rather than compile time as
-one might have hoped, the error is raised when \emph{generating the
-generator} -- well before the generation of the target code could
-begin. In our code we make extensive use of these so-called preflight 
-checks performed as part of module initialization. The checks seem to
-offer a good compromise: they are dynamic and so do not require
-complicated type systems; on the other hand, the checks are performed
-quite early, when buiding code generators, and so ensure that no 
-code violating the corresponding \emph{semantic} constraints 
-will be generated. Although some may frown on the use of the module
-initializing expressions (since that requires careful attention to the
-sharing and problems of multiple instantiation of a module), these
-concerns do not apply to our case: our preflight tests are all idempotent
-and maintain no state.
-
-Lifting:
-
-We now define a lot of infrastructure.  We will be quite thorough, and
-define base abstraction and lifted abstractions, This will involve a
-lot of boilerplate code, which unfortunately cannot be so easily
-automated in MetaOCaml -- that would require introspection.  It could
-be done in camlp4, but that seems too much as well.  This 'base' could
-be elided, but when we decide to make more use of Abstract
-Interpretation, we'll regret it, so do it now.
+in the last line of the above code. That line differs from the other 
+bindings in |ZpMake| in that it neither defines a function, such as
+|plus|, nor binds a value, such as |zero|. Therefore, the non-value
+expression |assert (is_prime P.p)| will be evaluated
+when the corresponding module is instantiated, like the following
+\begin{code}
+module Z19 = ZpMake(struct let p = 19 end)
+\end{code}
+We will call such an expression an initializing expression.  if we
+replace |p = 19| with |p = 9| above, we receive an error, a run-time
+error. It is raised however as we instantiate and combine modules that
+will eventually make the generator.  Although the error is reported at
+run-time rather than compile time as one might have hoped, the error
+is raised when \emph{generating the generator} -- well before the
+generation of the target code could begin. In our code we make
+extensive use of these so-called preflight checks performed as part of
+module initialization. The checks seem to offer a good compromise:
+they are dynamic and so do not require complicated type systems; on
+the other hand, the checks are performed quite early, when buiding
+code generators, and so ensure that no code violating the
+corresponding \emph{semantic} constraints will be generated. Although
+some may frown on the use of module initializing expressions (since
+that requires careful attention to sharing and multiple instantiations
+of a module), these concerns do not apply in our case: our preflight
+tests are all idempotent and maintain no state.
+
+\subsection{Abstract code and lifted domains}
+We will be using MetaOCaml operations to build code expressions, values
+of the type |('a,'v) code|. However, we wish to make our generators
+abstract over the particular code representations. Therefore, in the
+following Sections (in the rest of the paper?) rather than 
+MetaOCaml brackets and escapes, we will be using code combinators
+described below. These combinators are functions that produce and
+combine values of the type of abstract code.
+
+This abstraction lets us vary the output form of our generated LU
+programs. We may use thunks rather than MetaOCaml |('a,'v) code|
+values, for benchmarking purposes. We could use code combinators that
+produce C or Fortran code.
+
+Our code combinator library includes operations to create, dereference
+and mutate reference cells, compare values, manipulate values used
+for indexing arrays. The following is the implementation of a sample
+combinators, in the case of the abstract code being MetaOCaml code values. 
 \begin{code}
 let lift x = .< x >.
+
+let cunit = .< () >.
+let unitL = fun s k -> k s .< () >.
 
@@
 let liftGet x = .< ! .~x >. 
-let unitL = fun s k -> k s .< () >.
-
 let liftPair x = (.< fst .~x >., .< snd .~x >.)
 
-(* logic code combinators - plain and monadic *)
 module Logic = struct
@@
 
-(* operations on code indices *)
 module Idx = struct
-  let zero = .< 0 >.
-  let one = .< 1 >.
-  let minusone = .< -1 >.
+  let zero = .< 0 >. and one = .< 1 >. and minusone = .< -1 >.
   let succ a = .< .~a + 1 >.
@@
 
-let cunit = .< () >.
 let update a f = let b = f (liftGet a) in .< .~a := .~b >.
@@
 
-\begin{code}
-module S(T:
-  sig
-	(* Representation type of values,  to be specified *)
-    type ('a, 'b) rep
-  end) = struct
-
+We now convert the domains of values from teh previous section into
+lifted domains, of abstract code values. We define the signature
+|DOMAINL| 
+\begin{code}
+module S(T: sig type ('a, 'b) rep  end) = struct
 open T
@@
 end 
-
+\end{code}
+(part of the larger module |Domains_sig|) which is parameterised by
+the type of the abstract code, |('a, 'b) rep|. The line |include DOMAIN|
+says that lifted domains include all members of non-lifted domains,
+including the initializing expressions with pre-flight checks.
+non-lifted
+
+The following is a particular instance of |DOMAINL|, the lifted
+version of |IntegerDomain| of the previous section, in the particular
+case of abstract code being MetaOCaml code values |('a, 'b) code|:
+\begin{code}
 module T = Domains_sig.S(struct type ('a, 'b) rep = ('a, 'b) code end)
 open T
-
-open Domains_common
-(* because the operations are "syntactic" to a certain extent,
-   we have to repeat ourselves a lot *)
 module IntegerDomainL = struct
@@
 \end{code}
+
+\subsection{Containers}
 
@@
 
+\oleg{We should mention somewhere that we have implemented not only GE
+  but Solvers as well.}
+
 \section{Runge-Kutta solvers}
 \label{s:ode}
+\oleg{Drop this? no space}
 
@@
 There are many more aspects which can also be handled:
-Input variations (augmented
-matrices), error reporting (i.e. asking for the determinant of a 
+error reporting (i.e. asking for the determinant of a 
 non-square matrix), memory hierarchy issues, loop-unrolling
@@
 The actual bind function of the monad defaults to |bind| and the
-match-failure function to |failwith| (only used for refutable
-patterns; see below).  To select a different function, use the
+pattern-match--failure function to |failwith| (only used for refutable
+patterns).  To select a different function, use the
 extended forms of |perform|. For example, use the function named 
@@
    let t_3 = (ref 0) in
-   let t_4 = {arr = (Array.copy a_1.arr)} (a_1) in
-   let t_5 = a_1.m in
-   let t_6 = a_1.n in
+   let t_4 = (a_1) \{arr = (Array.copy a_1.arr)\} in
+   let t_5 = a_1.m in  (* magnitude of det *)
+   let t_6 = a_1.n in  (* sign of the det *)
    let t_7 = (ref 1) in
@@
     let t_34 =
-     begin (* full pivoting *)
+     begin (* full pivoting, search for the pivot *)
       for j_30 = t_13 to (t_6 - 1) do
@@
           end else ();
-          if ((fst (fst i_16)) <> t_13) then begin (* swapping of rows?*)
+          if ((fst (fst i_16)) <> t_13) then begin (* swapping of rows *)
            let a_17 = t_4.arr
@@
    done;
-   (t_4,
+   (t_4, (* matrix with the U factor *)
     if ((! t_8) = 0) then 0 (* adjust the sign of the determinant *)
@@
 \section{Appendix B}
-The code generated for |GenFA1|, LU of the floating point matrix
+The code generated for |GenFA9|, LU of the floating point matrix
 represented by a 2D array, row pivoting, returning just the |U|-factor.
-\begin{code2}
-val resFA1 : ('a, GAC_F.contr -> GenFA1.O.res) code =
-  .<fun a_1 ->
-   let t_2 = (ref 0) in
-   let t_3 = (ref 0) in
-   let t_5 = (Array.map (fun x_4 -> (Array.copy x_4)) (Array.copy a_1)) in
-   let t_6 = (Array.length a_1.(0)) in
-   let t_7 = (Array.length a_1) in
-   while (((! t_3) < t_6) && ((! t_2) < t_7)) do
-    let t_10 = (! t_2) in
-    let t_11 = (! t_3) in
-    let t_12 = (ref (None)) in
-    let t_18 =
-     begin  (* row pivoting *)
-      for j_15 = t_10 to (t_7 - 1) do
-       let t_16 = (t_5.(j_15)).(t_11) in
-       if (t_16 <> 0.) then
-        (match (! t_12) with
-         | Some (i_17) ->
-            if ((abs_float (snd i_17)) < (abs_float t_16)) then
-             (t_12 := (Some (j_15, t_16)))
-            else ()
-         | None -> (t_12 := (Some (j_15, t_16))))
-       else ()
-      done;
-      (match (! t_12) with (* swapping of rows *)
-       | Some (i_13) ->
-          if ((fst i_13) <> t_10) then
-           let t_14 = t_5.(t_10) in
-           t_5.(t_10) <- t_5.(fst i_13);
-           t_5.(fst i_13) <- t_14
-          else ();
-          (Some (snd i_13))
-       | None -> (None))
-     end in
-    (match t_18 with
-     | Some (i_19) -> (* elimination loop, elided?? *)
-        begin
-         for j_20 = (t_10 + 1) to (t_7 - 1) do
-          let t_21 = (t_5.(j_20)).(t_11) in
-          if (t_21 <> 0.) then begin
-           for j_22 = (t_11 + 1) to (t_6 - 1) do
-            (t_5.(j_20)).(j_22) <-
-             ((t_5.(j_20)).(j_22) -. ((t_21 /. i_19) *. (t_5.(t_10)).(j_22)))
-           done;
-           (t_5.(j_20)).(t_11) <- 0.
-          end else ()
-         done;
-         ()
-        end;
-        (t_2 := ((! t_2) + 1))
-     | None -> ());
-    (t_3 := ((! t_3) + 1))
-   done;
-   t_5>.
-\end{code2}
-
 \begin{code2}
@@
    let t_7 = (Array.length a_1) in
-   let t_8 = (ref ([])) in
+   let t_8 = (ref ([])) in  (* accumulate permutations in a list *)
    while (((! t_3) < t_6) && ((! t_2) < t_7)) do
@@
     let t_17 =
-     begin
+     begin   (* row pivoting *)
       for j_14 = t_9 to (t_7 - 1) do
@@
       done;
-      (match (! t_11) with
+      (match (! t_11) with  (* swapping of rows *)
        | Some (i_12) ->
@@
            t_5.(t_9) <- t_5.(fst i_12);
-           t_5.(fst i_12) <- t_13;
+           t_5.(fst i_12) <- t_13;  (* and accumulate permutations *)
            (t_8 := ((RowSwap ((fst i_12), t_9)) :: (! t_8)))
@@
      end in
-    (match t_17 with
+    (match t_17 with  (* elimination loop *)
      | Some (i_18) ->
@@
    done;
-   (t_5, (! t_8))>.
+   (t_5, (! t_8))>. (* return both L and U factors, list permutations *)
 \end{code2}

### 2007-07-07 SCP2008/paper/metamonads.tex
editing and rearranging of the final section.

@@
 
+The signature |CONTAINER2D| specifies that a container must provide
+functions |dim1| and |dim2| to extract the dimensions, functions |getL|
+to generate container getters, the cloning
+generator |copy| and functions that generate code for row and column
+swapping. The inclusion of these functions in the signature of all
+containers makes it simpler to optimize the relevant functions
+depending on the actual representation of the container while not
+burdening the users of containers with efficiency details.
+
 \begin{code}
@@
 
-Parameterizing by the kind of container representing a matrix is
-almost as straightforward.  The type of our containers include the
-lifted domain |Dom| as one of the component. Particular containers are
-Our containers are  parametric
-over a |DOMAINL|, i.e., functors from a |DOMAINL| module
-to the actual implementation of a container. For example, the
-following functor defines a matrix container represented as arrays of
-rows.
+The type of our containers include the lifted domain |Dom| as one of
+the component. This is quite convenient since operations on containers
+are usually accompanied by the operations on the retrieved and stored
+values. Our containers are parametric over a |DOMAINL|, i.e., functors
+from a |DOMAINL| module to the actual implementation of a
+container. For example, the following functor defines a matrix
+container represented as arrays of rows.
 
@@
 %
-We also have containers wheer elements are stored in a 1D array, in a
+The accompanying code includes containers whose
+elements are stored in a 1D array, in a
 row-wise (C-like) or column-wise (Fortran-like) modes.
 
-The signature |CONTAINER2D| specifies that a container must provide
-functions |dim1| and |dim2| to extract the dimensions, functions |getL|
-to generate container getters, the cloning
-generator |copy| and functions that generate code for row and column
-swapping. The inclusion of these functions in the signature of all
-containers makes it simpler to optimize the relevant functions
-depending on the actual representation of the container while not
-burdening the users of containers with efficiency details.
-
-The use of a |functor| for making a container parametric is fairly
-straightforward.  More interesting is the aspect of what to return
+\subsection{Maintaining the extensible state}
+
+Various aspects of our LU generator such as determinant
+computation, rank, permutation list may need to keep a state during
+the code generation. For example, the initializing section of the LU
+generator will invoke the |decl| of the determinant aspect so that the
+latter may generate let-binding for reference cells that track the
+magnitude and the sign of the determinant. The generator of the
+row-swapping code will notify the determinant aspect so that the
+latter may insert the code that flips the sign of the tracked
+determinant. Finally, the final section of the generator will ask the
+determinant aspect to generator code producing the final, signed value
+of the determinant. Clearly the determinant aspect needs to maintain
+state from one invocation of its methods to another; this state should
+include the `names' of mutable variables accumulating the sign and the
+magnitude of the determinant.
+
+The simplest method of keeping the state of the generator is by using
+mutable variables private to the Deteminant module (aspect). That,
+however, makes our aspects stateful. Although we are generating
+imperative code, we would like to keep our generators stateless and
+pure functional, for ease of comprehension and reasoning. Also, if a
+module is stateless, we do not have to worry about sharing, aliasing
+and multiple instances. Our main program may include several LU
+generators, which may refer to the Determinant aspect -- which may be
+present in one shared instance or in several. It is irrelevant for us
+which is which. With stateful modules, the issues of aliasing or
+separate instantiation are the source of very subtle problems.
+
+We chose therefore a different way of maintaining the generator state,
+using a monad. We need monad anyway for let-insertion; we add state as
+an additional monad parameter (Fig.~\ref{our-monad}) and monadic
+actions |fetch| and |store| to access that monadic state. The state is
+threaded throughout the entire code-generation computation.
+
+This monadic state has to accommodate several pieces of state, for
+various aspects. We should be able to add a new aspect, which may need
+to keep its own state as part of teh overall monadic state, without
+modifying or even recompiling the rest of the code. Thus our monadic
+state should be extensible. One way of achieving is via an extensible
+record, or OCaml object. Each aspect will have its own field; record
+subtyping will assure modularity. Alas, this approach makes it
+difficult to create the initial state, to pass to the monad's |runM|
+method. We should know the names of all fields and should know the
+proper initial value for these fields. This breaks the modularity.
+
+A better approach of property list has been suggested my the MLTon
+team \oleg{cite MLTon}.  Encoding an object by its dual, list of
+polymorphic variants.  Property list represents an extensible
+object. The initial object is just the empty list. Alas, the MLton's
+approach does not literally apply to us. First of all, due to
+the environment classifiers, we need universes parameterized by
+classifiers. Also, the property names are generated automatically
+in MLton approach, based on generativity of exceptions or reference
+cells. That means if some particular aspect happens to be included
+twice, the state will be incompatible. We will have to worry about
+sharing or the lack of sharing of our aspects. We would rather
+prefer our aspects to be stateless and work the same way whether two
+instances of the aspect are shared or not. Fortunately, OCaml offers a
+way to build open unions using polymorphic variants. Each variant is
+identified by a manifest tag, e.g., `Tdet, and may include a
+value. The tags are not generative and provide the form of manifest
+naming (somewhat similar to symbols of Lisp and Scheme).
+
+Below is the implementation of our open records with manifest
+naming, functions |orec_store| to add a new field to the open record
+and |orec_find| to obtain the value associated with a particular field
+name. Each `field' is characterized by a triple: injection, 
+projection functions and the string name. The latter is used for 
+printing error messages. For example, for the
+determinant tracking aspect, this triple has the form
+\begin{code}
+  let ip = (fun x -> `TDet x), 
+           (function `TDet x -> Some x | _ -> None), 
+           "Det"
+\end{code}
+We combine these functions with monadic
+actions to access monadic state and so obtain 
+|mo_extend| and |mo_lookup| to store and retrieve one particular value
+of the monadic state.
+\begin{code}
+let rec lookup ((_,prj,_) as ip) = 
+   function [] -> raise Not_found
+   | (h::t) -> (match prj h with Some x -> x | _ -> lookup ip t)
+
+let orec_store ((inj,_,name) as ip) v s =
+  let () = 
+    try let _ = lookup ip s in 
+        failwith ("The field of an open record is already present: " ^ name)
+    with Not_found -> () in
+  (inj v)::s
+
+let orec_find ((_,_,name) as ip) s =
+  try lookup ip s 
+  with Not_found -> failwith ("Failed to locate orec field: " ^ name)
+
+let mo_extend ip v = 
+  perform s <-- fetch; store (orec_store ip v s)
+
+let mo_lookup ip =
+  perform s <-- fetch; ret (orec_find ip s)
+\end{code}
+
+Currently, we check at generator-time that one cannot add a slot which
+already exists (already the part of the state) nor we can obtain the
+value of a non-existing slot. These problems may occur if our genetor
+is wrong (calls fin method before decl method of an aspect). It is
+possible to make these checks static (TFP2007, state-variable monad?)
+state-changing monad...
+
+\subsection{Determinant aspect}
+
+To track determinant we should be able to generate code
+for: defining variables used for tracking (|decl|),
+updating the sign or the absolute
+value of the determinant, converting the tracking state
+to the final determinant value of the type |outdet|. LU of a
+floating-point matrix with no determinant tracking uses the
+instantiation of |DETERMINANT| where |outdet| is |unit| and all the
+functions of that module generate no code. For integer matrices, we
+have to track some aspects of the determinant, even if we don't output
+it. The determinant tracking aspect is complex because tracking
+variables, if any, are to be declared at the beginning of LU; the sign
+of the determinant has to be updated on each row or column
+permutation; the value of the determinant should be updated per each
+pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
+code for the value of the type |Dom.v ref|, among
+various determinant-tracking functions. The |lstate| is a part of the
+overall monadic state.
+
+Other aspects follow a similar outline
+
+\begin{code}
+module type DETERMINANT = sig
+  type tdet = C.Dom.v ref
+  type 'a lstate
+  type ('pc,'v) lm = ('pc,'v) cmonad
+    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
+  type ('pc,'v) om = ('pc,'v) omonad
+    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
+  val decl : unit -> ('b,unit) lm (* could be unit rather than unit code...*)
+  val upd_sign  : unit -> ('b,unit) om
+  val zero_sign : unit -> ('b,unit) lm
+  val acc       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
+  val get       : unit -> ('b,tdet) lm
+  val set       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
+  val fin       : unit -> ('b,C.Dom.v) lm
+end
+\end{code}
+
+We have two instances of |Det|. The first one is used when the user
+requested that no determinant tracking is needed, and so no code is
+generated. The methods just return without executing any
+code-generation action.
+\begin{code}
+module NoDet =
+  struct
+  type tdet = C.Dom.v ref
+  type 'a lstate = unit
+  let decl () = unitL
+  let upd_sign () = ret None
+  let zero_sign () = unitL
+  let acc _ = unitL
+  let get () = ret (liftRef C.Dom.zeroL)
+  let set _ = unitL
+  let fin () = failwith "Determinant is needed but not computed"
+  type ('pc,'v) lm = ('pc,'v) cmonad
+    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
+  type ('pc,'v) om = ('pc,'v) omonad
+    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
+end
+\end{code}
+
+The following instance does track the determinant. The |decl| method
+generates two let-bindings for mutable variables tracking the
+magnitude and the sign of the determinant, and places the names of
+these variables in the monadic state. The |upd_sign| method, invoked
+from row- or column-swap generators, retrieves the name of the
+sign-accumulating variable and generates the code to update the
+sign. The |fin| method retrieves the names of both accumulators and
+generates code to compute the final, signed determinant value.
+\begin{code}
+module AbstractDet =
+  struct
+  open C.Dom
+  type tdet = v ref
+  (* the first part of the state is an integer: which is +1, 0, -1:
+     the sign of the determinant *)
+  type 'a lstate = ('a,int ref) abstract * ('a,tdet) abstract
+  let decl () = perform
+      ddecl <-- retN (liftRef oneL);
+      dsdecl <-- retN (liftRef Idx.one);
+      mo_extend ip (dsdecl,ddecl);
+      unitL
+  let upd_sign () = perform
+      det <-- mo_lookup ip;
+      det1 <-- ret (fst det);
+      ret (Some (assign det1 (Idx.uminus (liftGet det1))))
+  let fin = fun () -> perform
+      (det_sign,det) <-- mo_lookup ip;
+      ifM (Logic.equalL (liftGet det_sign) Idx.zero) (ret zeroL)
+      (ifM (Logic.equalL (liftGet det_sign) Idx.one) (ret (liftGet det))
+          (ret (uminusL (liftGet det))))
+  \dots
+end
+\end{code}
+
+\subsection{Output}
+
+More interesting is the aspect of what to return
 from the LU algorithm.  One could create an algebraic data type (as
@@
 
-\subsection{Maintaining the state}
-another title: representing extensible state
-
-different aspects may need to keep their own state. We wish to assure
-modularity. One approach: using objects. The state is an object, each
-aspect is an instnace variable with some particular name. The problem
-is how to represent the initial `empty' state. Need to create the
-object with all fields set to None. But we need to know the names of
-all fields; as we add an aspect, our function that runs the generator
-must change to account for the different initial object with different
-slots.
-
-A better approach: property list (ref MLTon). Property list represents
-an extensible object. The initial object is just the empty list. Alas,
-the literal MLton's approach doesn't apply to us well. First of all,
-due to the environment classifiers, we need universes parameterized by
-classifiers. Also, the propoerty names are generated automatically in
-MLton approach, based on generativity of exceptions or reference
-cells. That means if some particular aspect happens to be included
-twice, the state will be incompatible. We will have to worry about
-sharing or the lack of sharing of our aspects. We would rather prefer
-our aspects to be stateless and work the same way whether two
-instances of the aspect are shared or not. So, we find polymorphic
-unions to be ideal. Manifest naming
-
-Encoding an object by its dual,
-list of polymorphic variants.
-
-Currently, we check at genertor-time that one cannot add a slot which
-already exists (already the part of the state) nor we can obtain the
-value of a non-existing slot. These problems may occur if our genetor
-is wrong (calls fin method before decl method of an aspect). It is
-possible to make these checks static (TFP2007, state-variable monad?)
-state-changing monad...
-
-\subsection{tracking}
-
-As is apparent from the output choices, several different quantities
-\emph{may} need to be tracked in a particular LU implementation.  We
-therefore need to be able to conditionally generate variables
-representing the tracking state, and weave in corresponding tracking
-code. We may need to (independently) keep track of the rank, the
-determinant and the permutation list.  The tracking state variables
-then become part of the \emph{state} that is tracked by our monad.  To
-have all this choice when needed, and yet have our code be modular and
-composable as well as ensuring that the generated code does not
-contain any abstraction artifacts, it is important to make this state
-modular.  For example,
-\begin{code}
-module type DETERMINANT = sig
-  type tdet = C.Dom.v ref
-  type 'a lstate
-  type ('pc,'v) lm = ('pc,'v) cmonad
-    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
-  type ('pc,'v) om = ('pc,'v) omonad
-    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
-  val decl : unit -> ('b,unit) lm (* could be unit rather than unit code...*)
-  val upd_sign  : unit -> ('b,unit) om
-  val zero_sign : unit -> ('b,unit) lm
-  val acc       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
-  val get       : unit -> ('b,tdet) lm
-  val set       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
-  val fin       : unit -> ('b,C.Dom.v) lm
-end
-\end{code}
-\noindent  to track determinant we should be able to generate code
-for: defining variables used for tracking (|decl|),
-updating the sign or the absolute
-value of the determinant, converting the tracking state
-to the final determinant value of the type |outdet|. LU of a
-floating-point matrix with no determinant tracking uses the
-instantiation of |DETERMINANT| where |outdet| is |unit| and all the
-functions of that module generate no code. For integer matrices, we
-have to track some aspects of the determinant, even if we don't output
-it. The determinant tracking aspect is complex because tracking
-variables, if any, are to be declared at the beginning of LU; the sign
-of the determinant has to be updated on each row or column
-permutation; the value of the determinant should be updated per each
-pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
-code for the value of the type |Dom.v ref|, among
-various determinant-tracking functions. The |lstate| is a part of the
-overall monadic state. Other aspects, e.g., rank tracking, may use the
-monadic state for passing of rank tracking variables. To be able to
-compose determinant and rank tracking functors -- each of which may
-(or may not) use the monadic state for passing its own data -- we make
-extensive use of open records (a list of polymorphic variants
-appeared to be the easiest way to implement such a union, in a purely
-functional way). This lets us freely compose determinant-tracking,
-rank-tracking, and other aspects.
-
-We have several instances of Det
-\begin{code}
-module NoDet =
-  struct
-  type tdet = C.Dom.v ref
-  type 'a lstate = unit
-  let decl () = unitL
-  let upd_sign () = ret None
-  let zero_sign () = unitL
-  let acc _ = unitL
-  let get () = ret (liftRef C.Dom.zeroL) (* hack alert! *)
-  let set _ = unitL
-  let fin () = failwith "Determinant is needed but not computed"
-  type ('pc,'v) lm = ('pc,'v) cmonad
-    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
-  type ('pc,'v) om = ('pc,'v) omonad
-    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
-end
-
-module AbstractDet =
-  struct
-  open C.Dom
-  type tdet = v ref
-  (* the first part of the state is an integer: which is +1, 0, -1:
-     the sign of the determinant *)
-  type 'a lstate = ('a,int ref) abstract * ('a,tdet) abstract
-  let fin = fun () -> perform
-      (det_sign,det) <-- mo_lookup ip;
-      ifM (Logic.equalL (liftGet det_sign) Idx.zero) (ret zeroL)
-      (ifM (Logic.equalL (liftGet det_sign) Idx.one) (ret (liftGet det))
-          (ret (uminusL (liftGet det))))
-end
-\end{code}
-
-Both contain the |fin ()| function to generate the code representing
-the computed rank. Only the |NoRank| does not track any rank, and so 
-|fin ()| raises an error. The code 
-|let _ = IF.R.fin ()| in |OutDetRank| invokes that fin function, which
-will produce the monadic code generating action, or raise teh
-error. We do not run the action at that time -- we only make sure
-there is an action to run. This is the preflight check again, to rule
-out the semantic error when the user specified that rank should be
-computed and returned and yet specified the |NoRank| aspects which
-computes no rank at all.
-
-\begin{code}
-(* The `keyword' list of all the present external features *)
+Both instances of |Det| contain the |fin ()| function to generate the
+code representing the computed determinant. Only the |NoDet| does not
+track any determinant, and so |fin ()| raises an error. The code 
+|let _ = OD.Det.fin ()| in |OutDetRank| invokes that |fin| function, which will
+produce the monadic code generating action, or raise the error. We do
+not run the action at that time -- we only make sure there is an
+action to run. This is the preflight check again, to rule out the
+semantic error when the user specified that determinant should be computed
+and returned and yet specified the |NoDet| aspect which tracks no
+determinant at all.
+
+\subsection{Main generation}
+We combine all aspects in a record. This record is like a keyword
+argument list to the |GenGE| functor below.
+\begin{code}
 module type FEATURES = sig
@@
 end
+\end{code}
+
+\oleg{describe how generation proceeds}
+We first instantiate the container |GVC_I| for the vector over an
+integer domain. We pass the container to the main functor |GenLA|,
+since all the linear algebra aspects are parameterized by the
+container and the domain included in it. We open the |G_GVC_I| module
+to make all its components easily available. One of thos components is
+the GE generrator, |GE| (which we open as well). The module |G_GVC_I|
+also inlcudes the generator for GE solvers. The module |GE| includes
+the functor |GenGE| that is the generator. We instantiate the
+generator by passing the functor the `record' of various aspects (the
+order is irrelevant, but all aspects like |Det|, etc. must be
+specified). In the code below, we request generation with full pivot,
+fraction-free update, operate on non-augmented matrix and return the
+|U| factor, determinant and the rank. We run the monad passing the
+initial state |[]| and obtain the code in App A.
+
+\begin{code}
+module GVC_I = GenericVectorContainer(IntegerDomainL)
+module G_GVC_I = GenLA(GVC_I)
+open G_GVC_I
+open G_GVC_I.GE
+module GenIV5 = GenGE(struct 
+    module Det = AbstractDet
+    module PivotF = FullPivot
+    module PivotRep = PermList
+    module Update = FractionFreeUpdate
+    module Input = InpJustMatrix
+    module Output = OutDetRank end)
+let instantiate gen =
+    .<fun a -> .~(runM (gen .<a>.) []) >.;;
+let resIV5 = instantiate GenIV5.gen ;;
+let rIV5 = .! resIV5;;
+
+module GAC_F = GenericArrayContainer(FloatDomainL)
+module G_GAC_F = GenLA(GAC_F)
+open G_GAC_F
+open G_GAC_F.GE
+module GenFA9 = GenGE(struct 
+    module Det = NoDet
+    module PivotF = RowPivot
+    module PivotRep = PermList
+    module Update = DivisionUpdate
+    module Input = InpJustMatrix
+    module Output = Out_LU_Packed end)
 \end{code}
@@
 argument modules such as |PIVOT| are functors themselves (parameterized
-by the domain, the container, and the determinant functor). The sharing
-constraints express obvious constraints on the instantiation of |Gen|,
-for example, pivoting, determinant etc. components all use the same
-domain. It must be stressed that all structures (i.e., module
-instances) are stateless, and so we never have to worry that different
-aspect functors (such as |CONTAINER2D| and |PIVOT|) are instantiated
-with different but type-compatible instances of |DOMAIN|. That is, we
-are not concerned at all about value sharing. Aspects such as
-determinant tracking may be stateful so that the determinant update
-code have access to the determinant tracking variables declared
-previously. But that state is handled via the monadic state. As we
-have shown, open unions make the overall monadic state compositional
-with respect to the state of various aspects.
-
-In addition to the ``regular'' type sharing constraints shown in the
-|Gen| functor, there are also ``semantic'' sharing constraints, shown
+by the domain, the container, and the determinant functor).
+
+There are more pre-flight checks for various ``semantic'' 
+constraints, shown
 in the following structure of the |UPDATE| signature:
 \vspace*{-5pt}\begin{code}
-module DivisionUpdate
-  (Dom:DOMAIN with type kind = domain_is_field)
-  (C:CONTAINER2D)
-  (Det:DETERMINANT with type indet=Dom.v) = struct ... end
-\end{code}
-\vspace*{-4pt} 
+module DivisionUpdate(Det:DETERMINANT) =
+  struct
+  open C.Dom = struct 
+  let _ = assert (C.Dom.kind = Domains_sig.Domain_is_Field)
+end
+\end{code}
+%
 This structure implements an update policy of using
@@
 domain may still provide |Dom.div| of the same type, but that operation may
-only be used when we are sure that the division is exact. Our type
-sharing constraint expresses such domain-specific knowledge:
-instantiating |DivisionUpdate| with |IntegerDomain| leads to a
-compile-time error, when compiling the \emph{generator} code. Thus, in
-some cases we can use module types for ``semantic'' constraints
-that cannot normally be expressed via the types of module members.
-\vspace*{-5pt}
-\begin{code}
-module GVC_I = GenericVectorContainer(IntegerDomainL)
-module G_GVC_I = GenLA(GVC_I)
-open G_GVC_I
-open G_GVC_I.GE
-module GenIV5 = GenGE(struct 
-    module Det = AbstractDet
-    module PivotF = FullPivot
-    module PivotRep = PermList
-    module Update = FractionFreeUpdate
-    module Input = InpJustMatrix
-    module Output = OutDetRank end)
-
-module GAC_F = GenericArrayContainer(FloatDomainL)
-module G_GAC_F = GenLA(GAC_F)
-open G_GAC_F
-open G_GAC_F.GE
-
-module GenFA1 = GenGE(struct
-    module Det = NoDet
-    module PivotF = RowPivot
-    module PivotRep = PermList
-    module Update = DivisionUpdate
-    module Input = InpJustMatrix 
-    module Output = OutJustMatrix end)
-module GenFA9 = GenGE(struct 
-    module Det = NoDet
-    module PivotF = RowPivot
-    module PivotRep = PermList
-    module Update = DivisionUpdate
-    module Input = InpJustMatrix
-    module Output = Out_LU_Packed end)
-
-let resIV5 = instantiate GenIV5.gen ;;
-let rIV5 = runit {pf =  resIV5 };;
-
-\end{code}
-|GenFA1| is no-frills GE; GenFA9 has LU part
+only be used when we are sure that the division is exact.
 
 We can instantiate the |Gen| functor as shown above and inspect the generated
-code, e.g., by printing |GenFA1.gen|. The code can then be ``compiled'' as 
-|!. GenFA1.gen| or with off-shoring. The code for |GenIV5| (Appendix A) shows
+code, e.g., by printing |resIV5|. The code can then be ``compiled'' as 
+|!. resIV5| or with off-shoring. The code for |resIV5| (Appendix A) shows
 full pivoting, determinant and rank tracking. The code for all these aspects is
@@
 triple |int array * int * int| of the U-factor, determinant and the rank. The
-code generated by |GenFA1| (Appendix B) shows absolutely no traces of
+code generated by |GenFA9| (Appendix B) shows absolutely no traces of
 determinant tracking: no declaration of spurious variables, no extra tests,
@@
 \section{Appendix B}
-The code generated for |GenFA9|, LU of the floating point matrix
-represented by a 2D array, row pivoting, returning just the |U|-factor.
+The code generated for |GenFA9|, LU of the floating point
+non-augmented matrix
+represented by a 2D array, row pivoting, returning the complete
+factorization: |L| and |U|
+factors packed in a single matrix and the permutation matrix
+represented as the list of row number exchanges.
 \begin{code2}

### 2007-07-08 SCP2008/paper/metamonads.tex
Fix parts of the bibliography (and adjust some references). Minor edits.  Add some comments on things that still need to be fixed.

@@
 
-A survey \cite{carette04} of
+A survey \cite{Carette06} of
 Gaussian elimination implementations in the industrial package Maple
@@
 function call -- have a significant cost, especially for
-high-performance numerical computing \cite{carette04}.
+high-performance numerical computing \cite{Carette06}.
 
 A more appealing approach is generative programming
-\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser94algorithmoriented,BOOST,POOMA,ATLAS}.
+\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser89generic,musser94algorithmoriented,BOOST,POOMA,ATLAS}.
 The approach is not without problems, e.g., making sure that the
@@
 In this paper, we report on our continued progress \cite{CaretteKiselyov05}
-\oleg{should we say in the footnote: ``compared to the earlier,
-  conference version of this paper we describe a different
-  version of our generator dealing with the complete LU decomposition
-  and solving. We worked out previously missing aspects of in-place
-  updates, representaing permutation matrix, augmented input matrix,
-  and back-propagation. We have changed the representation of
-  domain-specific knowledge about permissible compositions of aspects.''}
-the code generating LU decomposition and solving 
-solving these problems using linear solving and LU decomposition
-as our running example. Specifically, our contributions:
+\footnote{We describe here an improved version of our generator
+  dealing with the complete LU decomposition algorithm, as well as
+  linear solving. We worked out previously missing aspects of in-place
+  updates, representing permutation matrices, dealing with augmented 
+  input matrix, and back-propagation. We have also changed the representation
+  of domain-specific knowledge about permissible compositions of aspects.
+  Also included is a careful description of all the aspects involved, as
+  well as documenting our development methodology for highly parametric
+  scientific software.}
+in using code generation for scientific (both numeric and symbolic)
+software.  We will use the algorithm family of LU decomposition 
+and linear system solving as our running examples to demonstrate our
+techniques.  Specifically, our contributions:
 \begin{itemize}
@@
     \item Implementation of the |perform|-notation (patterned after
-      |do|-notation of Haskell) to make monadic code readable.
+      the |do|-notation of Haskell) to make monadic code readable.
     \item Use of functors (including higher-order functors) to
       modularize the generator, express aspects (including results of
-      various types) and \emph{assure composability of aspects} even
-      for aspects that use state and have to be accounted in many
+      various types) and \emph{insure composability of aspects} even
+      for aspects that use state and have to be accounted for in many
       places in the generated code.
@@
 implementation, it is worthwhile to carefully study the design space involved.
-A preliminary study~\cite{carette04} revealed a number of aspects of the 
+A preliminary study~\cite{Carette06} revealed a number of aspects of the 
 family of Gaussian Elimination algorithms.  In the present work, we outline
@@
 of LU decomposition algorithms.  While this use if very much in the same 
-spirit as in aspect-oriented programming (AOP)~\cite{Kiczales}, our 
-implementation methodology is radically different.  We firmly believe that
+spirit as in aspect-oriented programming (AOP)~\cite{kiczales97aspectoriented},
+our implementation methodology is radically different\footnote{However it seems
+that we are closer to the original ideas on AOP~\cite{709568,mendhekar97rg}
+which were also concerned with scientific software}.  We firmly believe that
 our typed generative methodogy is much better suited to the functional
@@
 view of modules and information hiding~\cite{journals/cacm/parnas72a} as well
-as his view of product families \cite{Parnas??}, and by Dijkstra's ideas on
-separation of concerns \cite{EWD:EWD447}.
+as his view of product families \cite{journals/tse/Parnas76}, and by
+Dijkstra's ideas on separation of concerns \cite{EWD:EWD447}.
 To apply these principles to the study of LU decomposition, we need
@@
 		$2^6 + 2^5 + 2^2 = 100$ possible outputs, depending on whether
-		one chooses a $PLU$, $PLU^{1}R$ or \emph{Cholesky} 
+		one chooses a $PLU$, $PLUR$ or \emph{Cholesky} 
 		decomposition.  We chose to only consider $PLU$ for now.
@@
 		but semi-algorithms or probabilistic algorithms do exist.
-		See~\cite{ZhCaJeMo07} for an example.
+		See~\cite{ZhCaJeMo06a} for an example.
     \item \textbf{Pivoting}: Whether to use no, 
@@
 i.e., to construct future-stage computations. MetaOCaml provides only
-one mechanism of combining code expressions, by inlining
-\oleg{splicing?} one into
+one mechanism of combining code expressions, by splicing one
+piece of code into
 another. The power of that operation, called escape |.~|, comes from
@@
 evaluation, and extensively used by \cite{MSP:PADL04,KiselyovTaha} for
-code generation. Now, |param_code2 plus one| gives us the desired
-code.
+code generation. Like \cite{conf/pepm/BondorfD94}, we use this in the 
+context of writing a cogen by hand.  Now, |param_code2 plus one| gives us the
+desired code.
 
@@
 \subsection{Domains}
-Reviewing the various aspects outlined in section~\ref{aspects}, the
+Reviewing the various aspects outlined in section~\ref{design}, the
 simplest parametrization is to make the domain abstract.  As it turns
@@
 using a monad. We need monad anyway for let-insertion; we add state as
-an additional monad parameter (Fig.~\ref{our-monad}) and monadic
+an additional monad parameter (Fig.~\ref{ourmonad}) and monadic
 actions |fetch| and |store| to access that monadic state. The state is
@@
 from the LU algorithm.  One could create an algebraic data type (as
-was done in \cite{carette04}) to encode the various choices: the
+was done in \cite{Carette06}) to encode the various choices: the
 matrix, the matrix and the rank, the matrix and the determinant, the
@@
 \oleg{Drop this? no space}
+\jacques{We should drop this as a whole section, but I still think we 
+should mention it as something that we have done.  Mostly to say that
+things ``extend'' to other families too, and that no new difficulties 
+were encountered.}
 

### 2007-07-09 SCP2008/paper/metamonads.tex
finish the first sub-section describing the aspects

@@
 
+\newcounter{naspects}
 \begin{enumerate}
@@
 	  packed into a single matrix for output.
+\setcounter{naspects}{\value{enumi}}
 \end{enumerate}
 
-\noindent \jacques{Absorb this in the above} These are separated out from the
-others as they are cross-cutting concerns: in the case of the length measure, a
-property of the domain will influence the pivoting method \emph{if} pivoting is
-to be performed.\jacques{Also, should mention a few more aspects like
-inplace, errors-on-singular, logging, sparsity, as things that are not
-implemented}
+There are other aspects which we have not (yet) implemented, but that have
+been seen in different implementations.  In particular,
+\begin{enumerate}
+\setcounter{enumi}{\value{naspects}}
+\item \textbf{Logging}: This is a classical cross-cutting concern, which
+is also useful in this context.  Of course, there are many different
+parts of the algorithm which could be ``logged'', and ensuring fine control
+over this is a thorny issue.
+\item \textbf{Sparsity}: If a matrix is known to be sparse, the traversal
+(at least) should be sparse.  Additionally, it would better still if
+care was taken to minimize fill-in.
+\item \textbf{Other structure}: For example, if a matrix is known in
+advance to be real symmetric tri-diagonal, then LU decomposition can be
+done in $O(n^2)$ time (instead of $O(n^3)$; the main difficulty in 
+implementing this kind of reduction is the fact that this version of 
+LU requires $O(n)$ extra storage (unlike all other versions).
+\item \textbf{Warnings}: In the case of LU decomposition of a matrix over
+a domain for which we only have a heuristic algorithm for zero
+testing, it is customary to issue a warning (or otherwise log) when 
+a potentially zero pivot is chosen.
+\item \textbf{In-place}: When it is possible, it would be good to have the
+option to re-use the input matrix as the storage space for the output.
+\item \textbf{Error-on-singular}: There are instances where we are only
+interested in an LU decomposition for a non-singular matrix, and would
+prefer to get an exception when a singular matrix is encountered.
+\setcounter{naspects}{\value{enumi}}
+\end {enumerate}
+
+Clearly the above $\thenaspects$ aspects have a rather complex 
+dependency graph.  While some aspects are purely orthogonal to others, 
+most are inter-dependent.  For example, if one wants the determinant of
+the matrix as part of the output, then the determinant has to be tracked
+during the algorithm; a priori, it is possible to make incompatible choices,
+and our implementation has to insure that this is in fact impossible.
+More precisely, our goal is to make sure that this is detected long
+before the algorithm could be run.
+
+There are other kinds of dependencies: for example, whether or not a 
+domain has a length measure will influence pivoting, \emph{if} pivoting
+is to be performed.  While there are no circular dependencies, there is
+no clear directed graph of dependencies.  In other words, one cannot 
+simply use product, co-products, and compositions of the various aspects 
+to build the resulting algorithm.  We will get back to this point when
+we describe the details of the implementation.
 

### 2007-07-10 SCP2008/paper/metamonads.tex
added a couple of missing refs.

@@
 generation, we will need yet another type variable, environment
-classifier \oleg{cite Walid03?}. With type constructors taking more
+classifier \cite{taha-environment}. With type constructors taking more
 and more arguments, it becomes more difficult to read and write
@@
 
-A better approach of property list has been suggested my the MLTon
-team \oleg{cite MLTon}.  Encoding an object by its dual, list of
+A better approach of property list has been suggested my the MLton
+team \cite{mlton-proplist}.  Encoding an object by its dual, list of
 polymorphic variants.  Property list represents an extensible
@@
 were encountered.}
+\oleg{Ok, so we mention Runge-Kutta only in the introduction, and drop
+  this section then, right? Do you want to do that?}
 

### 2007-07-11 SCP2008/paper/metamonads.tex
Remove section on RK and mention in intro. Add a couple of references. Fix small part of future work section

@@
 Gaussian elimination implementations in the industrial package Maple
+\cite{Monagan:2001:M7P}
 found 6 clearly identifiable aspects and 35 different implementations of the
@@
 Or we can write a very generic procedure that accounts for
-all the aspects with appropriate abstractions \cite{Axiom}. The
+all the aspects with appropriate abstractions \cite{Gruntz:1994:IG,Axiom}. The
 abstraction mechanisms however -- be they procedure, method or a
@@
 
+We also used the same technology to implement a Runge-Kutta solver for
+ordinary differential equations, as well as a reimplementation of
+the FFT algorithm from~\cite{KiselyovTaha}.  The technology presented
+here was amply sufficient for these implementations.  Since our current
+implementations of these algorithms are rather straightforward compared to 
+our versions of LU Decomposition, these will not be mentioned further
+in this paper.
+
 The rest of this paper is structured as follows: The next section
@@
 LU decomposition algorithm family in completely separate,
-independent modules.  Section~\ref{s:ode} describes generating
-Runge-Kutta solvers for ordinary differential equations (ODE).
-We briefly discuss related work in
+independent modules.  We briefly discuss related work in
 section~\ref{related}. We then outline the future work and conclude.
@@
 
-\section{Runge-Kutta solvers}
-\label{s:ode}
-\oleg{Drop this? no space}
-\jacques{We should drop this as a whole section, but I still think we 
-should mention it as something that we have done.  Mostly to say that
-things ``extend'' to other families too, and that no new difficulties 
-were encountered.}
-\oleg{Ok, so we mention Runge-Kutta only in the introduction, and drop
-  this section then, right? Do you want to do that?}
-
 \section{Related and future work}\label{related}
@@
 
+\jacques{some of these items for future work seem to be out of date, they need
+to be reviewed}
+
 We are looking into encapsulating staging
@@
 warnings when zero-testing is undecidable and
-a value is only probabilistically non-zero, etc.  The larger program
-family of LU decompositions contains more aspects still.
+a value is only probabilistically non-zero, etc.
 

### 2007-07-11 SCP2008/paper/metamonads.tex
finish subsection on 'organizing aspects'

@@
 
+Further investigation of the aspects reveals that these fall into
+various categories.  This categorization has already been useful to us,
+as it revealed that parts of our implementation was not as well
+separated into distinct concerns as we had originally thought.  In
+particular, it revealed that our implementation of ``containers'' 
+mixed representation (i.e. data-structure) issues with abstraction
+issues (i.e. matrices and their natural representation-independent
+operations).  We hope to rectify this in the future.
+
+\begin{enumerate}
+	\item \textbf{Abstract domain}.  In other words, these are
+		firmly aspects which are domain-specific, and encode
+		requirements level entities.  In particular, items like
+		Domain, length measure, Matrix, permutation, full division
+		and structured matrices all belong here.
+	\item \textbf{Concrete representation}.  In other words, the
+		design-time issue of choosing a data-structure to represent
+		various abstract domains.  Items like whether the results 
+		should be packed, (part of) the representation of containers,
+		sparsity, and representation of permutations.
+	\item \textbf{Interface}.  In other words, what will be the
+		interface of each generated function.  This mostly deals
+		with the input and output choices, but also includes whether
+		the output should be in-place, and whether errors or warnings
+		should be issued in certain situations.  Logging is also
+		an interface issue.
+	\item \textbf{Algorithmic Strategy}.  This encodes the various
+		choices in the algorithms themselves.  For example, whether
+		to use fraction-free updates, which pivot strategy to use,
+		whether the matrix is augmented.
+	\item \textbf{Information tracking}.  What information should be
+		kept and/or computed while the algorithm proceeds, such
+		as whether to track determinant, rank, or pivot.
+	\item \textbf{Staging}.  This is the abstract interface to the
+		underlying programming language.  It (currently) encodes
+		whether to generate code that will be executed in a future
+		stage, or whether the code is direct and should be executed
+		now.  It should be possible to encode other strategies too.
+\end{enumerate}
+
+While the first two families of aspects are the most obvious, it is quite
+difficult to separate them out cleanly.  This seems to stem from the fact
+that in a non-staged setting, separating out these aspects leads to such
+fantastically inefficient code, especially in the setting of scientific
+computation, that our experience works \emph{against us}.  While it would
+be too much to speak of a paradigm shift, it is necessary to firmly think
+in terms of abstract ideal design while avoiding all thoughts of traditional
+efficiency.
+
+Interestingly, unlike traditional approaches \cite{journals/cacm/parnas72a}, we
+allow the interfaces to our (generated) routines to vary.  Compared to dynamic
+approaches in Object-Oriented language (late binding and dynamic dispatch) or
+functional languages (Haskell's dictionary-based \emph{type classes}), which
+offer some flexibility on interfaces, our interfaces are actually fixed at
+run-time, thus do incur any run-time inefficiency whatsoever.
+
+While we are by no means the first to speak of algorithmic strategy, to
+our knowledge our approach is one of the cleanest, in that the issue of
+strategy is completely separated into an independent entity (module), and
+yet we suffer no run-time inefficiency because of it.
+
+The information tracking aspects are in some ways to most intruiging ones:
+they are purely internal to the algorithms, and they are the ones which
+are most dependent on other choices.  The \emph{implementation} of these
+aspects is completely orthogonal to the other aspects, however their
+\emph{compositionality properties} are deeply enmeshed with that of other
+aspects.  While writing the code for the aspects was straightforward enough,
+this is where quite a bit of infrastructure was needed to encode
+various dependencies, so that it would be impossible to generate an
+instance of the LU Decomposition algorithm with information
+tracking aspects which were incompatible with the requirements of other
+aspects.
+
+Lastly, to ensure that we were indeed firmly in a \emph{cogen} setting,
+we abstracted out the underlying programming language completely.  This 
+allowed us to both generate (efficient) code and to write a directly
+runnable (highly inefficient) version of the algorithm from the same
+generator.  This brought out the issue that, once this was done, we had
+essentially abstracted out all of the syntactic sugar of the underlying
+programming language, and all we were left with was function application
+and monadic composition.  This means the code for our generator looks mostly
+like Scheme with added syntax for monads!
+
 \subsection{Methodology}
@@
 
+\jacques{Mention the relationship with the all-dynamic version of a 
+similar design, namely Maple's?  The fact that all choices there are
+dynamic, while all ours are completely static really shows that the
+design space along the dynamic--static dimension is quite large and
+versatile.}
+
 There is an interesting relation with aspect-oriented code
@@
 aspects are (comparatively) lightly typed, and are post-facto extensions of an
-existing piece of code.  Here aspects are weaved together ``from scratch'' to
+existing piece of code.  More precisely, they are post-facto extensions of
+potential program traces, specified in a particular language; these tend
+to be created to follow the operational behaviour of existing code, but
+is not restricted to such a setting.
+In our work, aspects are weaved together ``from scratch'' to
 make up a piece of code/functionality.  One can understand previous work to be
-more akin to dynamically typed aspect weaving, while we have started
-investigating statically typed one.
+more akin to dynamically typed, and dynamically specified aspect weaving, while
+we have started investigating statically typed and statically specified
+aspect weaving.
 

### 2007-07-11 SCP2008/paper/metamonads.tex
finish 2nd subsection, start 3rd.  Add references

@@
 \subsection{Methodology}
+
+Our overall approach can best be described a combination 
+of two approaches: hand-writing a code generator (cogen) suitable
+for multi-level specialization \cite{Gluck95,GluckJ97} and creating
+an embedded domain-specific language (EDSL) \cite{edsl,WhenHowDSL}.
+More specifically we start from a \emph{known} set of implementations
+of an algorithm, and extract commonalities and variation points.
+This is unlike~\cite{journals/tse/Parnas76} and most subsequent
+approaches to product families, as we do not over-engineer our design
+by imagining weird variations that are unlikely to come up in realistic
+situations, but only create variations when we notice them in actual use.
+This approach is quite well-suited to the development of scientific
+software, which has a rich history and where most useful variations have
+already appeared in some form.
+
+\jacques{need to describe actual process a bit more, maybe one or two
+paragraphs, and then this section is done}
 

### 2007-07-12 SCP2008/paper/metamonads.tex
a few small changes. Far more changes tomorrow

@@
 of the type |('a,'v) code|. However, we wish to make our generators
-abstract over the particular code representations. Therefore, in the
-following Sections (in the rest of the paper?) rather than 
-MetaOCaml brackets and escapes, we will be using code combinators
+abstract over particular code representations. Therefore, in the
+following Sections (in the rest of the paper?) instead of
+MetaOCaml brackets and escapes we will be using code combinators
 described below. These combinators are functions that produce and
@@
 
-This abstraction lets us vary the output form of our generated LU
-programs. We may use thunks rather than MetaOCaml |('a,'v) code|
-values, for benchmarking purposes. We could use code combinators that
-produce C or Fortran code.
+This encapsulation of staging annotations into a code combinators lets
+us easily change the output form of our generated LU programs. We may
+use single-stage thunks rather than MetaOCaml staged |('a,'v) code|
+values, for benchmarking and regression tests purposes. We could use
+code combinators producing C or Fortran code.
 
@@
 
-\jacques{some of these items for future work seem to be out of date, they need
-to be reviewed}
-
-We are looking into encapsulating staging
-annotations into just a few functors, so that the rest of the code (in
-particular, the |Gen| functor that puts it all together) should be
-annotation-free and thus can be used as is in a one-stage environment
-(pure OCaml) as well as in a multi-stage environment (generating
-extensions). The one-stage code is a good baseline for benchmarks and
-regression tests. Obtaining a generating extension from properly
-modularized OCaml code (along the lines of our |Gen|) is an exciting
-area of our future research.
-
 To the best of our knowledge, nobody has yet used functors to
@@
 continuations and our implementations of code generators like
-|ifM|.  As well, by using some additional syntactic sugar
-(for |ifM|, |whileM|, etc.), the available notation should be
-even more direct-style, and potentially clearer.
-We also would like to extend our monad to a monad transformer.
+|ifM|. The ultimate (and plausible) goal is to write a numerical
+algorithm in (almost) regular OCaml once, and being able to either
+run it as a regular OCaml program, or turn into a code generating
+aspect.
 

### 2007-07-12 SCP2008/paper/metamonads.tex
finish new section

@@
 code.  We demonstrate this clearly with LU decomposition as a 
-representative family of symbolic and numeric algorithms, and also
-include an application to solving ordinary differential equations 
-(via a Runge-Kutta algorithm).
+representative family of symbolic and numeric algorithms. 
 We parameterize our code to a great extent
@@
 
-\jacques{need to describe actual process a bit more, maybe one or two
-paragraphs, and then this section is done}
+Given a set of commonalities and variation points, the first task is to
+find \emph{semantic} reasons for these.  The underlying reason then
+forms the basis for the abstraction -- a (potentially higher-order)
+module is created to encapsulate the various concepts, and these are
+then implemented as generators.  It is very important at this stage to
+make sure to reify all available \emph{static} information, so that all
+of it is available to the generation process.  While we will see that
+the technical solutions exist to take advantage of such information,
+it is still a difficult design problem to properly encode this information.
+One important item is to try to keep the various pieces of generation-time
+information as orthogonal as possible.  This is unlike ordinary encodings
+of run-time information, where compression and elision frequently lead
+to increased efficiency.  Whenever dependencies between information is
+inevitable, then higher-order encodings should be sought.  When choices
+need to be made based on some (static) information, it is important to
+encode this information by using semantic concepts.
 

### 2007-07-12 SCP2008/paper/metamonads.tex
start first pass of whole paper edits

@@
 \ead[url]{http://pobox.com/\textasciitilde oleg/ftp/}
+
+\hyphenation{ne-ces-sa-ry}
 
@@
 domain-specific knowledge so that ``clearly wrong'' compositions can
-be rejected at or before generation time rather than during
+be rejected at or before generation time, rather than during
 the compilation or running of the generated code.
@@
 we still need post-processing to eliminate common subexpressions,
-fold constants, and remove redundant bindings? Is the generator readable,
-resembling the original algorithm? Is the generator extensible? Are the aspects
-truly modular? Can we add another aspect to it or another instance of
-the existing aspect without affecting the existing ones? Finally, can
-we express domain-specific knowledge, e.g., one should not attempt to
-use full division when dealing with matrices of exact integers, nor is
-it worthwhile to use full pivoting on a matrix over $\mathbb Q$.
-
-MetaOCaml is \emph{generative}: generated code can only be treated as
-a black box: it cannot be inspected and it cannot be post-processed
-(i.e., no intensional analysis). This approach gives a stronger
+fold constants, and remove redundant bindings? Is the generator readable?
+Does it bear as semblance to the original algorithm? Is the generator
+extensible? Are the aspects truly modular? Can we add another aspect to it or
+another instance of the existing aspect without affecting the existing ones?
+Finally, can we express domain-specific knowledge (for instance one should not
+attempt to use full division when dealing with matrices of exact integers, nor
+is it worthwhile to use full pivoting on a matrix over $\mathbb Q$)?
+
+MetaOCaml is purely \emph{generative}: generated code can only be treated as
+a black box: in other words, it cannot be inspected and it cannot be
+post-processed (i.e., no intensional analysis). This approach gives a stronger
 equational theory \cite{Taha2000}, and avoids the danger of creating
@@
 \cite{Padua:MetaOcaml:04}: generating binding forms (``names'')
-when generating loop bodies or conditional branches; making
+when generating loop bodies or conditional branches, and making
 continuation-passing style (CPS) code clear.  Many authors
@@
     \item Extending a let-insertion, memoizing monad of
-      \cite{MSP:PADL04,KiselyovTaha} for generating control structures
+      \cite{KiselyovTaha,MSP:PADL04} for generating control structures
       such as loops and conditionals. The extension is non-trivial
-      because of control dependencies and because
-      let-insertion, as we argue, is a control effect on its own.
-      : e.g.,\\
+	  because of control dependencies and because let-insertion, as we argue,
+	  is a control effect on its own: for example\\
       |let x = exp in ...| has a different \emph{effect} within a
@@
 generation, and continuation-passing style (CPS) as a general
-solution.  We also present a key monad and the issues of generating
+solution.  We also present a key monad and the issues around generating
 control statements. Section~\ref{functors} describes the use of
@@
 
-We build code generators out of primitive ones using code generation 
+We build large code generators out of primitive generators using 
 combinators. MetaOCaml, as an instance of a multi-stage
@@
 print out code expressions, so we can see the final generated code. It
-has no traces of |gen| and |plus|: their applications are done at the
+has no traces of |gen| or |plus|: their applications are done at the
 generation stage.
@@
 The final MetaOCaml feature, |.!| (pronounced ``run'') 
-executes the code expression: |.! simplest_code| is a function of two
+executes a code expression: |.! simplest_code| is a function of two
 integers, which we can apply: |(.! simplest_code) 1 2|. The original
 |simplest_code| is not a function on integers -- it is a code
-expression.
+expression which \emph{encodes} a function.
 
@@
 \end{code}
-Running the former expression yields the function on |float|s, whereas
+Running the former expression yields a function on |float|s, whereas
 the latter expression is the code expression for a boolean function.
-This clearly shows the separation of concerns, namely of that for domain
+This clearly shows the separation of concerns for domain
 operations.
@@
 the result of |param_code1' plus one| still exhibits duplicate
-sub-expressions.  Our |let|-insertion optimization saved the
+sub-expressions.  Our |let|-insertion optimization only saved the
 computation at the generating stage.  We need a combinator that
@@
 \end{code}
-yielding the code like 
+yielding code like 
 \begin{code}
@@
 But that seems impossible because |letgen exp| has to generate
-the expression |.<let t = exp in body>.| but |letgen| does not
-have the |body| yet. The body needs a temporary identifier |.<t>.|
+the expression |.<let t = exp in body>.| but |letgen| does not yet
+have the |body|. The body needs a temporary identifier |.<t>.|
 that is supposed to be the result of |letgen| itself.  Certainly
@@
 
-The key is to use continuation-passing style (CPS). Its benefits were
-first pointed out by \cite{Bondorf:92} in the context of partial
+The solution to this problem is to use continuation-passing style (CPS). Its
+benefits were first pointed out by \cite{Bondorf:92} in the context of partial
 evaluation, and extensively used by \cite{MSP:PADL04,KiselyovTaha} for
@@
 
-Comparison of the let-insertion in the generator
+Comparing the let-insertion in the generator
 \begin{code}
@@
 Not only can monadic code represent CPS \cite{Filinski:Representing},
-it also helps in composability by offering to add different
-layers of effects (state, exception, non-determinism, etc) to the
-basic monad \cite{liang-interpreter} in a controlled way.
+it also helps with composability by giving complete control over how
+different effects are layered
+(state, exception, non-determinism, etc) on top of the
+basic monad \cite{liang-interpreter}.
 
@@
 
-Our monad represents two kinds of computational effects: reading and
+Our monad encapsulates two kinds of computational effects: reading and
 writing a computation-wide state, and control effects. The latter are
-normally associated with exceptions, forking of computations, etc. --
+normally associated with exceptions, forking of computations, et
+cetera --
 in general, whenever a computation ends with something other than
@@
 
-In Figure~\ref{ourmonad}, the monad (yielding values of the type |v|)
+In Figure~\ref{ourmonad}, the monad (yielding values of type |v|)
 is implemented as a function of two
 arguments: the state (of type |s|) and the continuation. The
-continuation receives the current state and the value, and
-yields the answer of the type |w|.  The monad is polymorphic over the
+continuation receives the current state and a value, and
+yields an answer of type |w|.  The monad is polymorphic over the
 three type parameters, which would require |monad| to be a type
@@
 classifier \cite{taha-environment}. With type constructors taking more
-and more arguments, it becomes more difficult to read and write
+and more arguments, it becomes increasingly difficult to read and write
 types -- which we will be doing extensively when writing module
-signatures in Section XXX. The fact that OCaml renames all type
+signatures in Section~\ref{something}. The fact that OCaml renames all type
 variables when printing out types confuses matters further. An elegant
-solution to these kinds of problems has been suggested by 
-Garrigue on the Caml mailing list 
+solution to these sorts of problems has been suggested by 
+Jacques Garrigue on the Caml mailing list 
+\jacques{edited to here}
 (cited from
@@
 The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
-the pre-flight tests, to be explained in Section XXX.
+the pre-flight tests, to be explained in Section~\ref{something}.
 

### 2007-07-13 SCP2008/paper/metamonads.tex
many comments to Section 2. I believe it needs abbreviation, to keep it more technical.

@@
 \documentclass[draft]{elsart}
+\hyphenation{ana-lysis theo-ry}
 
@@
 % \setlength{\parskip}{0pt}
+\parskip=0pt
 \newlength{\ourmathindent}
@@
 representative family of symbolic and numeric algorithms. 
-We parameterize our code to a great extent
-(over domain, matrix representations, determinant tracking, 
-pivoting policies, result types, etc) at no run-time cost.  Because
+We parameterize our code to a great extent --
+over domain, input and permutation matrix representations, 
+determinant and rank tracking, 
+pivoting policies, result types, etc -- at no run-time cost.  Because
 the resulting code is generated just right and not changed afterwards,
@@
 \emph{may} need to be addressed. Implementations of the
-widely used LU decomposition algorithm (which subsumes Gaussian 
-Elimination (GE)) --- the running
+widely used LU decomposition algorithm and its particular case Gaussian 
+Elimination (GE) \oleg{can I say GE is an instance of LU?} --- the running
 example of our paper --- may need to account for the representation of
@@
 In this paper, we report on our continued progress \cite{CaretteKiselyov05}
-\footnote{We describe here an improved version of our generator
+\footnote{We describe here an new version of our generator
   dealing with the complete LU decomposition algorithm, as well as
@@
 	  because of control dependencies and because let-insertion, as we argue,
-	  is a control effect on its own: for example\\
-      |let x = exp in ...| has a different \emph{effect} within a
+	  is a control effect on its own: for example
+      \texttt{let x = exp in \dots} has a different \emph{effect} within a
       conditional branch.
@@
 our versions of LU Decomposition, these will not be mentioned further
-in this paper.
+in this paper (the code is available at \cite{metamonadsURL}).
 
@@
 
-Before studying the details of the technologies involved in the 
-implementation, it is worthwhile to carefully study the design space involved.
-A preliminary study~\cite{Carette06} revealed a number of aspects of the 
-family of Gaussian Elimination algorithms.  In the present work, we outline
-a number of additional aspects involved in the faimily of LU decomposition 
-algorithms.  These will first be presented in a somewhat ad hoc manner, 
-roughly coresponding to the order in which they were ``discovered''.  These
-will then be reorganized into sets of semantically related aspects, which will
-form the basis of our design.  Finally, we will extract a methodology from
-our experience.
+Before investigating implementation approaches, it is worthwhile to
+carefully study the design space involved.  A preliminary
+study~\cite{Carette06} revealed a number of aspects of the family of
+Gaussian Elimination algorithms.  In the present work, we outline a
+number of additional aspects involved in the faimily of LU
+decomposition algorithms.  These will first be presented in a somewhat
+ad hoc manner, roughly coresponding to the order in which they were
+``discovered''.  We then reorganize them into groups of
+semantically related aspects to form the basis of our design.
+Finally, we will extract a methodology from our experience.
 
@@
 covered in~\cite{Golub-vanLoan}.  When $A$ is singular, one can still get
-a $PLU$ decomposition, with $L$ remaining unit lower-triangular but with a
-$U$ which is no longer upper triangular but instead ``staggered''
+a $PLU$ decomposition with $L$ remaining unit lower-triangular. However, 
+$U$ is no longer upper triangular but rather ``staggered''
 in the upper triangle.
@@
 
-We reuse the english word ``aspect'' for the various facets of the family
-of LU decomposition algorithms.  While this use if very much in the same 
-spirit as in aspect-oriented programming (AOP)~\cite{kiczales97aspectoriented},
+We reuse the English word ``aspect'' for the various facets of the family
+of LU decomposition algorithms.  While our use shares the 
+spirit of aspect-oriented programming (AOP)~\cite{kiczales97aspectoriented},
 our implementation methodology is radically different\footnote{However it seems
-that we are closer to the original ideas on AOP~\cite{709568,mendhekar97rg}
+that we are closer to the original ideas of AOP~\cite{709568,mendhekar97rg}
 which were also concerned with scientific software}.  We firmly believe that
-our typed generative methodogy is much better suited to the functional
-programming paradigm than attempting to graft the program-trace-based
+our typed generative methodogy is much better suited to functional
+programming, compared to attempts to graft the program-trace-based
 methodology of object-oriented versions of AOP.
@@
 	\item \textbf{Representation of the matrix}: Whether the matrix
-		was represented as an array of arrays, a one-dimensional array,
-		a hash table, etc.  For the case of a one-dimensional array,
-		whether indexing was done in C or Fortran style.  Additionally,
-		if a particular representation had a special mechanism for efficient
-		row exchanges, this was sometimes used.
+		was represented as an array of arrays, a
+		one-dimensional array with C or Fortran indexing styles, 
+		a hash table, etc. Efficient row exchanges, if available for a
+		particular representation, were sometimes used.
 	\item \textbf{Fraction-free}: Whether the 
@@
 	\item \textbf{Length measure (for pivoting)}:  For stability reasons
-		(whether numerical or coefficient growth), if a domain possesses
-		an appropriate length measure, this was sometimes used to choose
+		(whether numerical or coefficient growth), if a domain 
+	  possesses
+		an appropriate length measure, it was sometimes used to choose
 		an ``optimal'' pivot.  Not all domains have such a measure.
@@
 		zero-testing.
-	\item \textbf{Output choices}:  Whether just the reduced matrix
-		(the `U' factor), both L and U, as well as
-		the rank, the determinant, and/or the sequence of 
-		pivots is to be returned.  For example, Maple's
+	\item \textbf{Output choices}:  Just the reduced matrix
+		(the `U' factor) or both L and U factors. The output
+	  choices also include
+		the rank, the determinant, and the sequence of 
+		pivots.  For example, Maple's
 		\texttt{LinearAlgebra:-LUDecomposition} routine has
@@
 		of the matrix as the algorithm proceeds.
-	\item \textbf{Code representation}: Whether we are actually generating
-		code or we are in fact representation the algorithm in ``direct
-		style'', albeit with a lot of abstraction overhead.
+	\item \textbf{Code representation}: the form and the language
+	  for the generated code (OCaml, C, Fortran, etc). A
+	  degenerate case, useful for testing, is for the generator
+	  to run the LU algorithm directly (albeit with great
+	  abstraction overhead).
 	\item \textbf{Zero-equivalence}: Whether the 
 		arithmetic operations require a specialized zero-equivalence 
-		routine needs to be used.  It turns out that for certain classes
-		of \emph{expressions}, it is convenient to use a zero-equivalence
-		test which is separate from the domain normalization.  This is
+		routine.  For certain classes
+		of \emph{expressions}, it turns out convenient to use 
+		a zero-equivalence
+		test that is separate from the domain normalization.  This is
 		usually the case when zero-equivalence is formally undecidable
@@
 	\item \textbf{Pivot representation}: Whether the pivot is represented
-	  as a list of row (and/or column) exchanges, as a unitary matrix,
+	  as a list of row and column exchanges, as a unitary matrix,
 	  or as a permutation vector.
@@
 	  matrix or not, and if so, where is the end of the main matrix.
+	  \oleg{Is this the same as Augmented matrices above?}
   \item \textbf{Packed}: Whether the output matrices $L$ and $U$ are
@@
 
-There are other aspects which we have not (yet) implemented, but that have
-been seen in different implementations.  In particular,
+The aspects in the following group have also been observed in
+practice but we have not yet implemented them:
 \begin{enumerate}
 \setcounter{enumi}{\value{naspects}}
-\item \textbf{Logging}: This is a classical cross-cutting concern, which
-is also useful in this context.  Of course, there are many different
-parts of the algorithm which could be ``logged'', and ensuring fine control
-over this is a thorny issue.
-\item \textbf{Sparsity}: If a matrix is known to be sparse, the traversal
-(at least) should be sparse.  Additionally, it would better still if
-care was taken to minimize fill-in.
-\item \textbf{Other structure}: For example, if a matrix is known in
-advance to be real symmetric tri-diagonal, then LU decomposition can be
-done in $O(n^2)$ time (instead of $O(n^3)$; the main difficulty in 
-implementing this kind of reduction is the fact that this version of 
-LU requires $O(n)$ extra storage (unlike all other versions).
-\item \textbf{Warnings}: In the case of LU decomposition of a matrix over
-a domain for which we only have a heuristic algorithm for zero
+\item \textbf{Logging}, a classical cross-cutting concern.
+\item \textbf{Sparsity}: If a matrix is known to be sparse, at least
+the traversal should be sparse. Maximal preservation of the sparsity
+is desirable. 
+\item \textbf{Other structure}: If a matrix is known in
+advance to be real symmetric tri-diagonal, LU decomposition can be
+done in $O(n^2)$ rather than $O(n^3)$ time, at the additional $O(n)$
+storage cost.
+\item \textbf{Warnings}: If a domain permits only heuristic zero
 testing, it is customary to issue a warning (or otherwise log) when 
 a potentially zero pivot is chosen.
-\item \textbf{In-place}: When it is possible, it would be good to have the
-option to re-use the input matrix as the storage space for the output.
-\item \textbf{Error-on-singular}: There are instances where we are only
-interested in an LU decomposition for a non-singular matrix, and would
-prefer to get an exception when a singular matrix is encountered.
+\item \textbf{In-place}: offering an option of in-place decomposition,
+  re-using the input matrix as the storage for the output.
+\item \textbf{Error-on-singular}: Raise an exception when the input
+  matrix is (near) singular.
 \setcounter{naspects}{\value{enumi}}
@@
 
-Clearly the above $\thenaspects$ aspects have a rather complex 
+Most of these aspects are inter-dependent. For example, if the
+determinant is the part of the output, the determinant should be
+tracked during the decomposition. Determinant should also be tracked
+if the fraction-free aspect is chosen. The availability of the length
+measure in the domain influences pivoting, \emph{if} pivoting
+is to be performed. One could therefore select aspects that turn out
+incompatible, and we are to prevent this. 
+More precisely, our goal is to detect the selection of
+incompatible aspects long before the generated code is run.
+
+\omitnow{Clearly the above $\thenaspects$ aspects have a rather complex 
 dependency graph.  While some aspects are purely orthogonal to others, 
@@
 and our implementation has to insure that this is in fact impossible.
-More precisely, our goal is to make sure that this is detected long
-before the algorithm could be run.
-
-There are other kinds of dependencies: for example, whether or not a 
-domain has a length measure will influence pivoting, \emph{if} pivoting
-is to be performed.  While there are no circular dependencies, there is
+There are other kinds of dependencies: for example, 
+  While there are no circular dependencies, there is
 no clear directed graph of dependencies.  In other words, one cannot 
@@
 to build the resulting algorithm.  We will get back to this point when
-we describe the details of the implementation.
+we describe the details of the implementation.}
 
@@
 
-Further investigation of the aspects reveals that these fall into
-various categories.  This categorization has already been useful to us,
-as it revealed that parts of our implementation was not as well
-separated into distinct concerns as we had originally thought.  In
-particular, it revealed that our implementation of ``containers'' 
+\oleg{I'm quite uneasy about this and the following sections. It
+  appears as `philosophy'. It's OK for us, because it is written in
+  the Perfect tense for us. We understand what each of these items
+  mean. OTH, readers will read this in the Future tense. They have not
+  seen our code and know nothing about our approach. To them, it may
+  appear just opaque philosophising.
+
+I'm in favor of drastically abbreviating these sections. Our message
+to the reader should be: please keep these groups of aspects as you
+read the description of our code below. We also point out that
+attempting to separate these aspects and letting the users
+independently change them leads -- in the unstaged case -- to such
+fantastically inefficient code so that such clean design is rarely
+(cite Aldor?) attempted. Staging let us remove the performance
+penalties of abstraction and so let us concentrate on the design. 
+The last phrase is actually more suitable for conclusion or related
+work. I think it would be much better to let related work make these
+points for us.}
+
+Further investigation revealed the following grouping of aspects:%
+\footnote{This grouping showed that some of our implementation did not
+separate distinct concerns well. In
+particular, our implementation of ``containers'' 
 mixed representation (i.e. data-structure) issues with abstraction
-issues (i.e. matrices and their natural representation-independent
-operations).  We hope to rectify this in the future.
+domain issues.  We hope to rectify this in the future.}
 
 \begin{enumerate}
-	\item \textbf{Abstract domain}.  In other words, these are
-		firmly aspects which are domain-specific, and encode
-		requirements level entities.  In particular, items like
-		Domain, length measure, Matrix, permutation, full division
-		and structured matrices all belong here.
-	\item \textbf{Concrete representation}.  In other words, the
-		design-time issue of choosing a data-structure to represent
-		various abstract domains.  Items like whether the results 
-		should be packed, (part of) the representation of containers,
-		sparsity, and representation of permutations.
-	\item \textbf{Interface}.  In other words, what will be the
-		interface of each generated function.  This mostly deals
-		with the input and output choices, but also includes whether
-		the output should be in-place, and whether errors or warnings
-		should be issued in certain situations.  Logging is also
-		an interface issue.
-	\item \textbf{Algorithmic Strategy}.  This encodes the various
-		choices in the algorithms themselves.  For example, whether
-		to use fraction-free updates, which pivot strategy to use,
-		whether the matrix is augmented.
-	\item \textbf{Information tracking}.  What information should be
-		kept and/or computed while the algorithm proceeds, such
-		as whether to track determinant, rank, or pivot.
-	\item \textbf{Staging}.  This is the abstract interface to the
-		underlying programming language.  It (currently) encodes
-		whether to generate code that will be executed in a future
-		stage, or whether the code is direct and should be executed
-		now.  It should be possible to encode other strategies too.
+	\item \textbf{Abstract domain}.  This group includes
+	  `mathematical' aspects: domain of matrix elements, length
+	  measure, full division, normalization, zero testing,
+	  symmetry and other matrix structure.
+	\item \textbf{Concrete representation}: choosing data
+	  structures for domains, containers, permutation
+	  matrices. This group also includes packing, in-place
+	  decomposition, and the
+	  representation of sparse matrices.
+	\item \textbf{Interface} of each generated function, including
+	  the input and output choices, logging and error reporting.
+		\oleg{In-place updates is also the issue of concrete
+		  representation! As well as of the abstract domains:
+		  for some domains we just can't run LU with no extra
+		  storage.} 
+	\item \textbf{Algorithmic Strategy}, such as fraction-free
+	  updates, pivoting strategies, augmented matrices.
+	\item \textbf{Tracking}, of determinant, rank, or pivot, etc.
+	\item \textbf{Staging}: whether the result if the program or a
+	  generator that will produce the program.
 \end{enumerate}
 
+\oleg{I'd rather replace this paragraph with my remark above.}
 While the first two families of aspects are the most obvious, it is quite
@@
 
-Interestingly, unlike traditional approaches \cite{journals/cacm/parnas72a}, we
-allow the interfaces to our (generated) routines to vary.  Compared to dynamic
-approaches in Object-Oriented language (late binding and dynamic dispatch) or
-functional languages (Haskell's dictionary-based \emph{type classes}), which
-offer some flexibility on interfaces, our interfaces are actually fixed at
-run-time, thus do incur any run-time inefficiency whatsoever.
-
-While we are by no means the first to speak of algorithmic strategy, to
-our knowledge our approach is one of the cleanest, in that the issue of
-strategy is completely separated into an independent entity (module), and
-yet we suffer no run-time inefficiency because of it.
-
+\oleg{I'd rather move this down where we discuss the output choices. I
+  want to get top the code quickly. Actually, the following paragraph
+  better belongs to related work.}  
+Unlike traditional approaches
+\cite{journals/cacm/parnas72a}, the interfaces of our generated
+routines vary depending on the choices of input and output
+aspects. Dynamic approaches in Object-Oriented language (late binding
+and dynamic dispatch) or functional languages (Haskell's
+dictionary-based \emph{type classes}), too, offer flexibility of
+interfaces. In our approach, however, it is the generator that
+produces code whose interfaces depend on the arguments of the
+generator. The interfaces of the generated routines are all fixed and
+hence effecient.
+
+\oleg{I'd drop this: the reader has not seen our implementation.}
 The information tracking aspects are in some ways to most intruiging ones:
@@
 
+\oleg{I'd move this to the next section.}
 Lastly, to ensure that we were indeed firmly in a \emph{cogen} setting,
@@
 an embedded domain-specific language (EDSL) \cite{edsl,WhenHowDSL}.
+\oleg{I'd keep only the above phrase and say that our approach of
+  using staging, monads and functors turns out to permit extensible
+  set of aspects and flexible enough for iterative improvement in the
+  design.}
+
+\oleg{I'm inclined then to drop the following two paragraphs. I'm
+  positive reviewers would be upset at philosophical poinst in the
+  middle of a technical paper.}
 More specifically we start from a \emph{known} set of implementations

### 2007-07-13 SCP2008/paper/metamonads.tex
the rest of my comments

@@
 mechanisms we want to structure our code generators, as long as none
-of those abstractions infiltrate the generated code.
-
-\jacques{This whole section needs to be edited to take into account
-the different organization of the material}.
+of those abstractions infiltrate the generated code. ML module system
+turns out most convenient.
 
 \subsection{Domains}
-Reviewing the various aspects outlined in section~\ref{design}, the
-simplest parametrization is to make the domain abstract.  As it turns
-out, we need the following to exist in our domains: $0$, $1$, $+$,
+
+We start by abstracting out domains. We have already seen the simplest
+case of domain abstraction in |param_code1|, which took
+code-generators such as |plus| and |one| as arguments.  We need far
+more than two parameters: our domains should include $0$, $1$, $+$,
 $*$, (unary and binary) $-$, at least \emph{exact} division,
-normalization, and potentially a relative size measure. The simplest
-case of such domain abstraction is |param_code1|.  There,
-code-generators such as |plus| and |one| were passed as arguments. We
-need far more than two parameters, so we have to group them. Instead
-of the grouping offered by regular records, we use OCaml
+normalization, and potentially a relative size measure. We could group
+these parameters in tuples or records.  Instead, we use OCaml
 \emph{structures} (i.e., modules) so we can take advantage of
@@
 \end{code}
-This domain is parametrized by an integer number |p|, the
-(modulus?). To be more precise, the structure |ZpMake| is
+This domain is parametrized by an integer number |p|. 
+To be more precise, the structure |ZpMake| is
 parameterized over another structure of the type described by
-signature P; that argument structure has one field, the int value |p|.
+the signature P; that argument structure has one field, the |int| value |p|.
 Such a parameterized structure (or, a function from structures to
@@
 expression |assert (is_prime P.p)| will be evaluated
-when the corresponding module is instantiated, like the following
+when the corresponding module is instantiated, as in the following
 \begin{code}
@@
 \end{code}
-We will call such an expression an initializing expression.  if we
-replace |p = 19| with |p = 9| above, we receive an error, a run-time
+We call such an expression an initializing expression. If we
+replace |p = 19| with |p = 9| above, we receive a run-time
 error. It is raised however as we instantiate and combine modules that
@@
 they are dynamic and so do not require complicated type systems; on
-the other hand, the checks are performed quite early, when buiding
+the other hand, the checks are run quite early, when buiding
 code generators, and so ensure that no code violating the
 corresponding \emph{semantic} constraints will be generated. Although
-some may frown on the use of module initializing expressions (since
+some may frown on the use of module initializing expressions since
 that requires careful attention to sharing and multiple instantiations
-of a module), these concerns do not apply in our case: our preflight
-tests are all idempotent and maintain no state.
+of a module, these concerns do not apply in our case: our preflight
+checks are all idempotent and maintain no state.
 
@@
 
-This encapsulation of staging annotations into a code combinators lets
+This encapsulation of staging annotations into code combinators lets
 us easily change the output form of our generated LU programs. We may
@@
 and mutate reference cells, compare values, manipulate values used
-for indexing arrays. The following is the implementation of a sample
-combinators, in the case of the abstract code being MetaOCaml code values. 
+for indexing arrays. The following is the implementation of sample
+combinators, in the case of the abstract code being MetaOCaml code. 
 \begin{code}
@@
 
-let cunit = .< () >.
-let unitL = fun s k -> k s .< () >.
-
-let liftRef x = .< ref .~x >. 
-let liftGet x = .< ! .~x >. 
+let cunit = .< () >.  and unitL = fun s k -> k s .< () >.
+
+let liftRef x = .< ref .~x >.  and   liftGet x = .< ! .~x >. 
 let liftPair x = (.< fst .~x >., .< snd .~x >.)
@@
   let zero = .< 0 >. and one = .< 1 >. and minusone = .< -1 >.
-  let succ a = .< .~a + 1 >.
-  let pred a = .< .~a - 1 >.
+  let succ a = .< .~a + 1 >.  and  pred a = .< .~a - 1 >.
   let less a b = .< .~a < .~b >.
-  let uminus a = .< - .~a >.
-  let add a b = .< .~a + .~b >.
+  let uminus a = .< - .~a >.  and  add a b = .< .~a + .~b >.
 
@@
 
-We now convert the domains of values from teh previous section into
+We now convert the domains of values from the previous section into
 lifted domains, of abstract code values. We define the signature
@@
 including the initializing expressions with pre-flight checks.
-non-lifted
-
 The following is a particular instance of |DOMAINL|, the lifted
@@
 
-The signature |CONTAINER2D| specifies that a container must provide
+The signature |CONTAINER2D| below specifies that a container must provide
 functions |dim1| and |dim2| to extract the dimensions, functions |getL|
@@
 The type of our containers include the lifted domain |Dom| as one of
-the component. This is quite convenient since operations on containers
+the components. This is quite convenient since operations on containers
 are usually accompanied by the operations on the retrieved and stored
@@
 
-Various aspects of our LU generator such as determinant
-computation, rank, permutation list may need to keep a state during
+Various aspects of our LU generator such as determinant, rank, 
+and permutation traking may need to keep a state during
 the code generation. For example, the initializing section of the LU
-generator will invoke the |decl| of the determinant aspect so that the
+generator will invoke the function |decl| of the determinant aspect so that the
 latter may generate let-binding for reference cells that track the
@@
 latter may insert the code that flips the sign of the tracked
-determinant. Finally, the final section of the generator will ask the
-determinant aspect to generator code producing the final, signed value
+determinant. The final section of the LU generator will ask the
+determinant aspect to generate code producing the final, signed value
 of the determinant. Clearly the determinant aspect needs to maintain
-state from one invocation of its methods to another; this state should
+state from one invocation of its functions to another; this state should
 include the `names' of mutable variables accumulating the sign and the
@@
 imperative code, we would like to keep our generators stateless and
-pure functional, for ease of comprehension and reasoning. Also, if a
-module is stateless, we do not have to worry about sharing, aliasing
-and multiple instances. Our main program may include several LU
-generators, which may refer to the Determinant aspect -- which may be
-present in one shared instance or in several. It is irrelevant for us
-which is which. With stateful modules, the issues of aliasing or
-separate instantiation are the source of very subtle problems.
+pure functional, for ease of comprehension and reasoning.  Our main
+program may include several LU generators referring to one Determinant
+aspect -- which may be present in one shared instance or in
+several. That is of no concern if the Determinant module is
+stateless. With stateful modules, the issues of aliasing or separate
+instantiation are the source of very subtle problems.
 
@@
 This monadic state has to accommodate several pieces of state, for
-various aspects. We should be able to add a new aspect, which may need
-to keep its own state as part of teh overall monadic state, without
+various aspects. We should be able to add a new aspect -- which may need
+to keep its own state as part of the overall monadic state -- without
 modifying or even recompiling the rest of the code. Thus our monadic
-state should be extensible. One way of achieving is via an extensible
-record, or OCaml object. Each aspect will have its own field; record
+state should be extensible. We could use an extensible
+record: an OCaml object. Each aspect will have its own field; record
 subtyping will assure modularity. Alas, this approach makes it
@@
 method. We should know the names of all fields and should know the
-proper initial value for these fields. This breaks the modularity.
-
-A better approach of property list has been suggested my the MLton
-team \cite{mlton-proplist}.  Encoding an object by its dual, list of
-polymorphic variants.  Property list represents an extensible
-object. The initial object is just the empty list. Alas, the MLton's
-approach does not literally apply to us. First of all, due to
-the environment classifiers, we need universes parameterized by
-classifiers. Also, the property names are generated automatically
-in MLton approach, based on generativity of exceptions or reference
-cells. That means if some particular aspect happens to be included
-twice, the state will be incompatible. We will have to worry about
-sharing or the lack of sharing of our aspects. We would rather
-prefer our aspects to be stateless and work the same way whether two
-instances of the aspect are shared or not. Fortunately, OCaml offers a
-way to build open unions using polymorphic variants. Each variant is
-identified by a manifest tag, e.g., `Tdet, and may include a
-value. The tags are not generative and provide the form of manifest
-naming (somewhat similar to symbols of Lisp and Scheme).
+proper initial value for these fields, which breaks modularity.
+
+A better approach of a property list has been suggested by the MLton
+team \cite{mlton-proplist}. A property list, too, represents an
+extensible object -- by its dual, a list of polymorphic variants.  The
+initial object is just the empty list. MLton's approach does not
+literally apply to us however. Environment classifiers of MetaOCaml
+require type universes parameterized by classifiers. The MLton
+approach uses generativity of exceptions or reference cells to
+generate property names, i.e., fields of an extensible record. That
+would make our aspects stateful modules, which we try to
+avoid. Fortunately, OCaml offers a way to build open unions with
+polymorphic variants. Each variant is identified by a manifest tag,
+e.g., |`Tdet|, and may include a value. The tags are not generative and
+provide a form of manifest naming, similar to symbols of
+Lisp and Scheme.
 
 Below is the implementation of our open records with manifest
-naming, functions |orec_store| to add a new field to the open record
+naming: functions |orec_store| to add a new field to the open record
 and |orec_find| to obtain the value associated with a particular field
@@
 \begin{code}
-  let ip = (fun x -> `TDet x), 
-           (function `TDet x -> Some x | _ -> None), 
-           "Det"
+let ip = 
+  (fun x -> `TDet x), (function `TDet x -> Some x | _ -> None), "Det"
 \end{code}
@@
 actions to access monadic state and so obtain 
-|mo_extend| and |mo_lookup| to store and retrieve one particular value
+|mo_extend| and |mo_lookup| to store and retrieve one component 
 of the monadic state.
@@
 
-let mo_lookup ip =
-  perform s <-- fetch; ret (orec_find ip s)
-\end{code}
-
-Currently, we check at generator-time that one cannot add a slot which
-already exists (already the part of the state) nor we can obtain the
-value of a non-existing slot. These problems may occur if our genetor
-is wrong (calls fin method before decl method of an aspect). It is
-possible to make these checks static (TFP2007, state-variable monad?)
-state-changing monad...
+let mo_lookup ip = perform s <-- fetch; ret (orec_find ip s)
+\end{code}
+%
+Currently, we check at generator-time that one should not add an
+already existing field to an open record, nor should one attempt to
+look up a field that does not exist. It is possible to make these
+checks static.
 
@@
 overall monadic state. 
-
 Other aspects follow a similar outline
@@
     constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
-  val decl : unit -> ('b,unit) lm (* could be unit rather than unit code...*)
+  val decl : unit -> ('b,unit) lm
   val upd_sign  : unit -> ('b,unit) om
@@
   let fin () = failwith "Determinant is needed but not computed"
-  type ('pc,'v) lm = ('pc,'v) cmonad
-    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
-  type ('pc,'v) om = ('pc,'v) omonad
-    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
 end
@@
 end
-(* The `keyword' list of all the present internal features *)
 module type INTERNAL_FEATURES = sig
@@
 end
-
 module type OUTPUT = functor(OD : OUTPUTDEP) -> sig
@@
 end
+
 module OutDetRank(OD : OUTPUTDEP) = struct
@@
     ret (Tuple.tup3 m.matrix det rank)
-  (* Initialization: check the preconditions of instantiation of this struct*)
   let _ = OD.Det.fin ()
@@
 The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
-the pre-flight tests, to be explained in Section~\ref{something}.
-
-Both instances of |Det| contain the |fin ()| function to generate the
+the pre-flight checks. As we saw in the previous section, 
+both instances of |Det| contain the |fin ()| function to generate the
 code representing the computed determinant. Only the |NoDet| does not
@@
 \subsection{Main generation}
-We combine all aspects in a record. This record is like a keyword
+We combine all aspects in a record, which serevs as a keyword
 argument list to the |GenGE| functor below.
@@
 \end{code}
-
-\oleg{describe how generation proceeds}
-We first instantiate the container |GVC_I| for the vector over an
-integer domain. We pass the container to the main functor |GenLA|,
-since all the linear algebra aspects are parameterized by the
-container and the domain included in it. We open the |G_GVC_I| module
-to make all its components easily available. One of thos components is
-the GE generrator, |GE| (which we open as well). The module |G_GVC_I|
-also inlcudes the generator for GE solvers. The module |GE| includes
-the functor |GenGE| that is the generator. We instantiate the
-generator by passing the functor the `record' of various aspects (the
-order is irrelevant, but all aspects like |Det|, etc. must be
-specified). In the code below, we request generation with full pivot,
-fraction-free update, operate on non-augmented matrix and return the
-|U| factor, determinant and the rank. We run the monad passing the
-initial state |[]| and obtain the code in App A.
 
@@
 let resIV5 = instantiate GenIV5.gen ;;
-let rIV5 = .! resIV5;;
-
+\end{code}
+%
+We first instantiate the container |GVC_I| for the vector over an
+integer domain. We pass the container to the main functor |GenLA|,
+since all the linear algebra aspects are parameterized by the
+container and the domain included in it. We open the |G_GVC_I| module
+to make all its components easily available. One of those components
+is the GE generator, |GE| (which we open as well). Besides |GE|, the
+module |G_GVC_I| includes the generator for GE solvers.  The main
+generator functor is |GenGE|, shown later.  We instantiate the
+generator by passing the functor the `record' of various aspects; the
+order is irrelevant, but all aspects like |Det|, etc. must be
+specified. In the code above, we request GE generation with full
+pivot, fraction-free update, operate on non-augmented matrix and
+return the |U| factor, determinant and the rank. We run the monad
+passing the initial state |[]| and obtain the code, which we can see
+by printing |resIV5|. The code can then be ``compiled'' as |!. resIV5|
+or with off-shoring. The code for |resIV5| (Appendix A) shows full
+pivoting, determinant and rank tracking. The code for all these
+aspects is fully inlined; no extra functions are invoked and no tests
+other than those needed by the LU algorithm itself are performed. The
+resulting function returns a triple |int array * int * int| of the
+U-factor, determinant and the rank.
+
+The following is another instantiation of the GE generator, with a
+different set of aspects.
+\begin{code}
 module GAC_F = GenericArrayContainer(FloatDomainL)
@@
 \end{code}
+The code generated by |GenFA9| (Appendix B) shows no traces of
+determinant tracking whatsoever: no declaration of spurious variables,
+no extra tests, etc. The code appears as if the determinant tracking
+aspect did not exist at all. The generated code for the above and
+other instantiations of |Gen| can be examined at
+\cite{metamonadsURL}. The website also contains benchmark code and
+timing comparisons.
+
+The LU generator functor itself is 
+parameterized by the domain, container, pivoting policy (full, row,
+nonzero, no pivoting), update policy (with either `fraction-less'
+or full division), and the result specification. Some of the
+argument modules such as |PIVOT| are functors themselves (parameterized
+by the domain, the container, and the determinant functor).
 
@@
                       U.update bjc pos.curval brk bjk 
-                          (fun ov -> C.col_head_set mat.matrix j k ov) det) UP )
+                          fun ov -> C.col_head_set mat.matrix j k ov) det) UP )
                       (IF.L.updt mat.matrix j pos.p.colpos C.Dom.zeroL 
@@
 
-The LU generator functor itself is 
-parameterized by the domain, container, pivoting policy (full, row,
-nonzero, no pivoting), update policy (with either `fraction-less'
-or full division), and the result specification. Some of the
-argument modules such as |PIVOT| are functors themselves (parameterized
-by the domain, the container, and the determinant functor).
-
 There are more pre-flight checks for various ``semantic'' 
@@
 domain may still provide |Dom.div| of the same type, but that operation may
-only be used when we are sure that the division is exact.
-
-We can instantiate the |Gen| functor as shown above and inspect the generated
-code, e.g., by printing |resIV5|. The code can then be ``compiled'' as 
-|!. resIV5| or with off-shoring. The code for |resIV5| (Appendix A) shows
-full pivoting, determinant and rank tracking. The code for all these aspects is
-fully inlined; no extra functions are invoked and no tests other than those
-needed by the LU algorithm itself are performed. The GE?? function returns a
-triple |int array * int * int| of the U-factor, determinant and the rank. The
-code generated by |GenFA9| (Appendix B) shows absolutely no traces of
-determinant tracking: no declaration of spurious variables, no extra tests,
-etc. The code appears as if the determinant tracking aspect did not exist
-at all. The generated code for the above and other instantiations of
-|Gen| can be examined at \cite{metamonadsURL}. The website also 
-contains benchmark code and timing comparisons.
-
-\oleg{We should mention somewhere that we have implemented not only GE
-  but Solvers as well.}
+only be used when we are sure that the division is exact. Thus if we
+specified |module Update = DivisionUpdate| when instantiating |GenIV5|
+above, we would have received an error because |IntegerDomainL| is not
+a field. That run-time error occurs before any code is produced (i.e.,
+before |resIV5| is computed).
 

### 2007-07-13 SCP2008/paper/metamonads.tex
Cut section 2 down to size.  Move some of the text from that section to other places when appropriate. A few minor edits in other sections as well.

@@
 widely used LU decomposition algorithm and its particular case Gaussian 
-Elimination (GE) \oleg{can I say GE is an instance of LU?} --- the running
+Elimination (GE) --- the running
 example of our paper --- may need to account for the representation of
@@
 The rest of this paper is structured as follows: The next section
-gives an overview of the design space of LU decomposition algorithms,
-as well as the methodology we follow.  Then section~\ref{CPS}
+gives an overview of the design space of LU decomposition algorithms.
+Then section~\ref{CPS}
 introduces code generation in MetaOCaml, the problem of name
@@
 independent modules.  We briefly discuss related work in
-section~\ref{related}. We then outline the future work and conclude.
+section~\ref{related}. We then outline future work. In our
+conclusion (section~\ref{conclusion}), we sum up our work, comment
+on programmings with aspects, and outline the overall methodology
+we used to structure our work.
 Appendices give samples of the generated code (which is available in
@@
 semantically related aspects to form the basis of our design.
-Finally, we will extract a methodology from our experience.
 
@@
 	  algorithm, or not tracked at all.
-  \item \textbf{Input choices}: Whether the input matrix is an augmented
-	  matrix or not, and if so, where is the end of the main matrix.
-	  \oleg{Is this the same as Augmented matrices above?}
+  \item \textbf{Input choices}: Grouping all the potential
+	  choices of inputs -- currently only augmented matrices require
+	  an extra input.
   \item \textbf{Packed}: Whether the output matrices $L$ and $U$ are
@@
 advance to be real symmetric tri-diagonal, LU decomposition can be
-done in $O(n^2)$ rather than $O(n^3)$ time, at the additional $O(n)$
+done in $O(n^2)$ rather than $O(n^3)$ time, at an additional $O(n)$
 storage cost.
@@
 
-\omitnow{Clearly the above $\thenaspects$ aspects have a rather complex 
-dependency graph.  While some aspects are purely orthogonal to others, 
-most are inter-dependent.  For example, if one wants the determinant of
-the matrix as part of the output, then the determinant has to be tracked
-during the algorithm; a priori, it is possible to make incompatible choices,
-and our implementation has to insure that this is in fact impossible.
-There are other kinds of dependencies: for example, 
-  While there are no circular dependencies, there is
-no clear directed graph of dependencies.  In other words, one cannot 
-simply use product, co-products, and compositions of the various aspects 
-to build the resulting algorithm.  We will get back to this point when
-we describe the details of the implementation.}
-
 \subsection{Organizing aspects}
-
-\oleg{I'm quite uneasy about this and the following sections. It
-  appears as `philosophy'. It's OK for us, because it is written in
-  the Perfect tense for us. We understand what each of these items
-  mean. OTH, readers will read this in the Future tense. They have not
-  seen our code and know nothing about our approach. To them, it may
-  appear just opaque philosophising.
-
-I'm in favor of drastically abbreviating these sections. Our message
-to the reader should be: please keep these groups of aspects as you
-read the description of our code below. We also point out that
-attempting to separate these aspects and letting the users
-independently change them leads -- in the unstaged case -- to such
-fantastically inefficient code so that such clean design is rarely
-(cite Aldor?) attempted. Staging let us remove the performance
-penalties of abstraction and so let us concentrate on the design. 
-The last phrase is actually more suitable for conclusion or related
-work. I think it would be much better to let related work make these
-points for us.}
 
@@
 	  the input and output choices, logging and error reporting.
-		\oleg{In-place updates is also the issue of concrete
-		  representation! As well as of the abstract domains:
-		  for some domains we just can't run LU with no extra
-		  storage.} 
 	\item \textbf{Algorithmic Strategy}, such as fraction-free
@@
 
-\oleg{I'd rather replace this paragraph with my remark above.}
-While the first two families of aspects are the most obvious, it is quite
-difficult to separate them out cleanly.  This seems to stem from the fact
-that in a non-staged setting, separating out these aspects leads to such
-fantastically inefficient code, especially in the setting of scientific
-computation, that our experience works \emph{against us}.  While it would
-be too much to speak of a paradigm shift, it is necessary to firmly think
-in terms of abstract ideal design while avoiding all thoughts of traditional
-efficiency.
-
-\oleg{I'd rather move this down where we discuss the output choices. I
-  want to get top the code quickly. Actually, the following paragraph
-  better belongs to related work.}  
-Unlike traditional approaches
-\cite{journals/cacm/parnas72a}, the interfaces of our generated
-routines vary depending on the choices of input and output
-aspects. Dynamic approaches in Object-Oriented language (late binding
-and dynamic dispatch) or functional languages (Haskell's
-dictionary-based \emph{type classes}), too, offer flexibility of
-interfaces. In our approach, however, it is the generator that
-produces code whose interfaces depend on the arguments of the
-generator. The interfaces of the generated routines are all fixed and
-hence effecient.
-
-\oleg{I'd drop this: the reader has not seen our implementation.}
-The information tracking aspects are in some ways to most intruiging ones:
-they are purely internal to the algorithms, and they are the ones which
-are most dependent on other choices.  The \emph{implementation} of these
-aspects is completely orthogonal to the other aspects, however their
-\emph{compositionality properties} are deeply enmeshed with that of other
-aspects.  While writing the code for the aspects was straightforward enough,
-this is where quite a bit of infrastructure was needed to encode
-various dependencies, so that it would be impossible to generate an
-instance of the LU Decomposition algorithm with information
-tracking aspects which were incompatible with the requirements of other
-aspects.
-
-\oleg{I'd move this to the next section.}
-Lastly, to ensure that we were indeed firmly in a \emph{cogen} setting,
-we abstracted out the underlying programming language completely.  This 
-allowed us to both generate (efficient) code and to write a directly
-runnable (highly inefficient) version of the algorithm from the same
-generator.  This brought out the issue that, once this was done, we had
-essentially abstracted out all of the syntactic sugar of the underlying
-programming language, and all we were left with was function application
-and monadic composition.  This means the code for our generator looks mostly
-like Scheme with added syntax for monads!
-
-\subsection{Methodology}
-
-Our overall approach can best be described a combination 
-of two approaches: hand-writing a code generator (cogen) suitable
-for multi-level specialization \cite{Gluck95,GluckJ97} and creating
-an embedded domain-specific language (EDSL) \cite{edsl,WhenHowDSL}.
-\oleg{I'd keep only the above phrase and say that our approach of
-  using staging, monads and functors turns out to permit extensible
-  set of aspects and flexible enough for iterative improvement in the
-  design.}
-
-\oleg{I'm inclined then to drop the following two paragraphs. I'm
-  positive reviewers would be upset at philosophical poinst in the
-  middle of a technical paper.}
-More specifically we start from a \emph{known} set of implementations
-of an algorithm, and extract commonalities and variation points.
-This is unlike~\cite{journals/tse/Parnas76} and most subsequent
-approaches to product families, as we do not over-engineer our design
-by imagining weird variations that are unlikely to come up in realistic
-situations, but only create variations when we notice them in actual use.
-This approach is quite well-suited to the development of scientific
-software, which has a rich history and where most useful variations have
-already appeared in some form.
-
-Given a set of commonalities and variation points, the first task is to
-find \emph{semantic} reasons for these.  The underlying reason then
-forms the basis for the abstraction -- a (potentially higher-order)
-module is created to encapsulate the various concepts, and these are
-then implemented as generators.  It is very important at this stage to
-make sure to reify all available \emph{static} information, so that all
-of it is available to the generation process.  While we will see that
-the technical solutions exist to take advantage of such information,
-it is still a difficult design problem to properly encode this information.
-One important item is to try to keep the various pieces of generation-time
-information as orthogonal as possible.  This is unlike ordinary encodings
-of run-time information, where compression and elision frequently lead
-to increased efficiency.  Whenever dependencies between information is
-inevitable, then higher-order encodings should be sought.  When choices
-need to be made based on some (static) information, it is important to
-encode this information by using semantic concepts.
+\oleg{In-place updates is also the issue of concrete
+  representation! As well as of the abstract domains:
+  for some domains we just can't run LU with no extra
+  storage.} 
+\jacques{Correct.  There are issues which are ``relational'' rather
+than being completely independent.  The above categories seem to 
+be independent, but that is too simplistic.  Some aspects, like
+in-place, have quite a number of pre-conditions on other aspects
+before they can be realized.}
 
@@
 |let _ = OD.Det.fin ()| in |OutDetRank| invokes that |fin| function, which will
-produce the monadic code generating action, or raise the error. We do
+produce the monadic code generating action, or raise an error. We do
 not run the action at that time -- we only make sure there is an
-action to run. This is the preflight check again, to rule out the
-semantic error when the user specified that determinant should be computed
-and returned and yet specified the |NoDet| aspect which tracks no
+action to run. This is another preflight check, to rule out the
+semantic error when a user specifies that the determinant should be computed
+and returned and yet specifies the |NoDet| aspect which tracks no
 determinant at all.
+
+It is also worth noticing the |INTERNAL_FEATURES| type, which bundles
+various information tracking aspects.  These are in some ways rather
+intruiging: they are purely internal to the algorithms, yet they are the
+ones which
+are most dependent on other choices.  The \emph{implementation} of these
+aspects is completely orthogonal to the other aspects, however their
+\emph{compositionality properties} are deeply enmeshed with that of other
+aspects.  While writing the code for these was straightforward enough,
+this required quite a bit of type-level infrastructure, as well as the
+preflight checks%
+\footnote{We had originally~\cite{CaretteKiselyov05} implemented all our
+preflight tests at the type level, but this did not scale.}
+to get everything to work properly.
 
@@
 
+\jacques{Mention the relationship with the all-dynamic version of a 
+similar design, namely Maple's?  The fact that all choices there are
+dynamic, while all ours are completely static really shows that the
+design space along the dynamic--static dimension is quite large and
+versatile.}
+
+Unlike traditional approaches
+\cite{journals/cacm/parnas72a}, the interfaces of our generated
+routines vary depending on the choices of input and output
+aspects. Dynamic approaches in Object-Oriented language (late binding
+and dynamic dispatch) or functional languages (Haskell's
+dictionary-based \emph{type classes}) also offer flexibility of
+interfaces. In our approach, however, it is the generator that
+produces code whose interfaces depend on the arguments of the
+generator. The interfaces of the generated routines are all fixed and
+hence efficient.
+
 To the best of our knowledge, nobody has yet used functors to
@@
 stateless functors and structures, and our monad with the
-compositional state makes aspects freely composable without having to
+compositional state makes aspects composable without having to
 worry about value aliasing. The only constraints to compositionality
@@
 
-\jacques{Mention the relationship with the all-dynamic version of a 
-similar design, namely Maple's?  The fact that all choices there are
-dynamic, while all ours are completely static really shows that the
-design space along the dynamic--static dimension is quite large and
-versatile.}
+\subsection{On aspects}
 
@@
 
+While the first two families of aspects (abstract domain and concrete
+representation) are the most obvious, it is quite
+difficult to separate them out cleanly.  This seems to stem from the fact
+that in a non-staged setting, separating out these aspects leads to such
+fantastically inefficient code, especially in the setting of scientific
+computation, that our experience works \emph{against us}.  While it would
+be too much to speak of a paradigm shift, it is necessary to firmly think
+in terms of abstract ideal design while avoiding all thoughts of traditional
+efficiency.  Interestingly, in conversations that the first author had
+with D. Parnas, we discovered this is apparently
+what~\cite{journals/cacm/parnas72a} was advocating.  We believe that we are
+taking the first steps towards a \emph{typed yet efficient} realization of
+these ideas, where the various design-time entities can be directly encoded in
+machine-checkable form.
+
+\subsection{Methodology}
+
+Our overall approach can best be described a combination 
+of two approaches: hand-writing a code generator (cogen) suitable
+for multi-level specialization \cite{Gluck95,GluckJ97} and creating
+an embedded domain-specific language (EDSL) \cite{edsl,WhenHowDSL}.
+Our approach, using staging, monads and functors, seems to permit 
+an extensible set of aspects and appears flexible enough for iterative
+improvement in the design.
+
+As we wanted to ensure that we were indeed firmly in a \emph{cogen} setting,
+we abstracted out the underlying programming language completely.  This 
+allowed us to both generate (efficient) code and to write a directly
+runnable (highly inefficient) version of the algorithm from the same
+generator.  This brought out the issue that, once this was done, we had
+essentially abstracted out all of the syntactic sugar of the underlying
+programming language, and all we were left with was function application
+and monadic composition.  This means the code for our generator looks mostly
+like Scheme with added syntax for monads!
+
+More specifically we start from a \emph{known} set of implementations
+of an algorithm, and extract commonalities and variation points.
+This is unlike~\cite{journals/tse/Parnas76} and most subsequent
+approaches to product families, as we do not over-engineer our design
+by imagining weird variations that are unlikely to come up in realistic
+situations, but only create variations when we notice them in actual use.
+This approach is quite well-suited to the development of scientific
+software, which has a rich history and where most useful variations have
+already appeared in some form.
+
+Given a set of commonalities and variation points, the first task is to
+find \emph{semantic} reasons for these.  The underlying reason then
+forms the basis for the abstraction -- a (potentially higher-order)
+module is created to encapsulate the various concepts, and these are
+then implemented as generators.  It is very important at this stage to
+make sure to reify all available \emph{static} information, so that all
+of it is available to the generation process.  While we will see that
+the technical solutions exist to take advantage of such information,
+it is still a difficult design problem to properly encode this information.
+One important item is to try to keep the various pieces of generation-time
+information as orthogonal as possible.  This is unlike ordinary encodings
+of run-time information, where compression and elision frequently lead
+to increased efficiency.  Whenever dependencies between information is
+inevitable, then higher-order encodings should be sought.  When choices
+need to be made based on some (static) information, it is important to
+encode this information by using semantic concepts.
+
 \subsection*{Acknowledgments}
@@
 appreciated. The implementation of the monadic notation, |perform|,
-was the joint work with Lydia van Dijk.
+was joint work with Lydia van Dijk.
 

### 2007-07-14 SCP2008/paper/metamonads.tex
more cleaning up

@@
 In this paper, we report on our continued progress \cite{CaretteKiselyov05}
-\footnote{We describe here an new version of our generator
+\footnote{We describe here a new version of our generator
   dealing with the complete LU decomposition algorithm, as well as
@@
 implementations of these algorithms are rather straightforward compared to 
-our versions of LU Decomposition, these will not be mentioned further
-in this paper (the code is available at \cite{metamonadsURL}).
+our versions of LU decomposition, we will not mention them further 
+(the code is available at \cite{metamonadsURL}).
 
@@
 gives an overview of the design space of LU decomposition algorithms.
-Then section~\ref{CPS}
+Section~\ref{CPS}
 introduces code generation in MetaOCaml, the problem of name
-generation, and continuation-passing style (CPS) as a general
-solution.  We also present a key monad and the issues around generating
+generation, and the continuation-passing style (CPS) as a general
+solution.  We also present the key monad and the issues of generating
 control statements. Section~\ref{functors} describes the use of
 parametrized modules of OCaml to encode all of the aspects of the
-LU decomposition algorithm family in completely separate,
-independent modules.  We briefly discuss related work in
-section~\ref{related}. We then outline future work. In our
-conclusion (section~\ref{conclusion}), we sum up our work, comment
-on programmings with aspects, and outline the overall methodology
-we used to structure our work.
-Appendices give samples of the generated code (which is available in
-full at \cite{metamonadsURL}).
+LU decomposition algorithm family as separate
+modules.  We discuss related work in
+Section~\ref{related} and outline future work. In our
+conclusion (Section~\ref{conclusion}) we comment
+on programmings with aspects and sum up our guiding methodology.
+Appendices give samples of the generated code, available in
+full at \cite{metamonadsURL}.
 
@@
 lower triangular matrix $L$ and (usually) an upper triangular matrix $U$,
-such that $A = LU$.  When pivoting is used, we get a unitary matrix $P$ such
+such that $A = LU$.  Pivoting adds a unitary matrix $P$ such
 that the factorization is now $A = PLU$.  The case of numeric matrices is well
@@
 \begin{enumerate}
-	\item \textbf{Domain}: In which (algebraic) domain do the matrix
-		elements belong to.  Some implementations were very specific
+	\item \textbf{Domain}: the (algebraic) domain of matrix
+		elements.  Some implementations were very specific
 		($\mathbb{Z}, \mathbb{Q}, \mathbb{Z_p}, 
@@
 storage cost.
-\item \textbf{Warnings}: If a domain permits only heuristic zero
+\item \textbf{Warnings}: In a domain with only heuristic zero
 testing, it is customary to issue a warning (or otherwise log) when 
@@
 Most of these aspects are inter-dependent. For example, if the
-determinant is the part of the output, the determinant should be
+determinant is part of the output, the determinant should be
 tracked during the decomposition. Determinant should also be tracked
@@
 	\item \textbf{Tracking}, of determinant, rank, or pivot, etc.
-	\item \textbf{Staging}: whether the result if the program or a
+	\item \textbf{Staging}: whether the result is the program or a
 	  generator that will produce the program.
@@
 
-\oleg{In-place updates is also the issue of concrete
-  representation! As well as of the abstract domains:
-  for some domains we just can't run LU with no extra
-  storage.} 
-\jacques{Correct.  There are issues which are ``relational'' rather
-than being completely independent.  The above categories seem to 
-be independent, but that is too simplistic.  Some aspects, like
-in-place, have quite a number of pre-conditions on other aspects
-before they can be realized.}
+%% \oleg{In-place updates is also the issue of concrete
+%%   representation! As well as of the abstract domains:
+%%   for some domains we just can't run LU with no extra
+%%   storage.} 
+%% \jacques{Correct.  There are issues which are ``relational'' rather
+%% than being completely independent.  The above categories seem to 
+%% be independent, but that is too simplistic.  Some aspects, like
+%% in-place, have quite a number of pre-conditions on other aspects
+%% before they can be realized.}
+
+\oleg{Perhaps we should just add the following phrase, paraphrased
+  from your remark:}
+The groupings are not entirely orthogonal (for example, in-place
+decomposition is possible only for specific domains), yet are useful
+as guidance in creating the modular LU generator discussed in 
+Section~\ref{functors}.
 
@@
 types -- which we will be doing extensively when writing module
-signatures in Section~\ref{something}. The fact that OCaml renames all type
+signatures in Section~\ref{functors}. The fact that OCaml renames all type
 variables when printing out types confuses matters further. An elegant
@@
 \jacques{edited to here}
-(cited from
-\url{http://groups.google.com/group/fa.caml/msg/e80b1245702d6b24}
-no date, no exact ref). We use a single type parameter |'p| to
+\oleg{I'd rather keep the following lines in the comments. I know
+  Jacques has been credited, but I don't have the exact reference
+  where he proposed it.}
+%% (cited from
+%% \url{http://groups.google.com/group/fa.caml/msg/e80b1245702d6b24}
+%% no date, no exact ref). 
+%
+We use a single type parameter |'p| to
 represent all parameters of our monad (all parameters but the type of
@@
 |plus|, nor binds a value, such as |zero|. Therefore, the non-value
-expression |assert (is_prime P.p)| will be evaluated
+expression |assert (is_prime P.p)|, to be called an initializing
+expression, will be evaluated
 when the corresponding module is instantiated, as in the following
@@
 \end{code}
-We call such an expression an initializing expression. If we
+If we
 replace |p = 19| with |p = 9| above, we receive a run-time
@@
 of the type |('a,'v) code|. However, we wish to make our generators
-abstract over particular code representations. Therefore, in the
-following Sections (in the rest of the paper?) instead of
+abstract over particular code representations. Therefore, when writing
+the LU generator (Section~\ref{sec:det} and below), instead of
 MetaOCaml brackets and escapes we will be using code combinators
@@
 lifted domains, of abstract code values. We define the signature
-|DOMAINL| 
+|DOMAINL|:
 \begin{code}
@@
 \end{code}
-(part of the larger module |Domains_sig|) which is parameterised by
-the type of the abstract code, |('a, 'b) rep|. The line |include DOMAIN|
+(part of a larger module |Domains_sig|) that is parameterised by
+the type of abstract code, |('a, 'b) rep|. The line |include DOMAIN|
 says that lifted domains include all members of non-lifted domains,
@@
     type 'a vc = ('a,v) code
-    let zeroL = .< 0 >.  
-    let oneL = .< 1 >. 
-    let (+^) x y = .<.~x + .~y>. 
-    let ( *^ ) x y = .<.~x * .~y>.
-    let ( -^ ) x y = .<.~x - .~y>.
+    let zeroL = .< 0 >.  and oneL = .< 1 >. 
+    let (+^) x y = .<.~x + .~y>. and ( -^ ) x y = .<.~x - .~y>.
+    let ( *^ ) x y = .<.~x * .~y>. and divL x y = .<.~x / .~y>. 
     let uminusL x = .<- .~x>.
-    let divL x y = .<.~x / .~y>. 
     let normalizerL = None
@@
 \subsection{Determinant aspect}
-
+\label{sec:det}
 To track determinant we should be able to generate code
@@
 functions of that module generate no code. For integer matrices, we
-have to track some aspects of the determinant, even if we don't output
+have to track some aspects of the determinant, even if we do not output
 it. The determinant tracking aspect is complex because tracking
@@
 overall monadic state. 
-Other aspects follow a similar outline
+Other aspects follow a similar outline.
 
@@
 
-We have two instances of |Det|. The first one is used when the user
-requested that no determinant tracking is needed, and so no code is
+We have two instances of |Det|. The first one is used to specify
+no determinant tracking, and so no code is
 generated. The methods just return without executing any
-code-generation action.
+code-generation actions.
 \begin{code}
@@
   type tdet = v ref
-  (* the first part of the state is an integer: which is +1, 0, -1:
-     the sign of the determinant *)
   type 'a lstate = ('a,int ref) abstract * ('a,tdet) abstract
   let decl () = perform
-      ddecl <-- retN (liftRef oneL);
+      ddecl <-- retN (liftRef oneL); (* track the sign: +1, 0, -1 *)
       dsdecl <-- retN (liftRef Idx.one);
@@
 
-It is also worth noticing the |INTERNAL_FEATURES| type, which bundles
-various information tracking aspects.  These are in some ways rather
-intruiging: they are purely internal to the algorithms, yet they are the
-ones which
-are most dependent on other choices.  The \emph{implementation} of these
-aspects is completely orthogonal to the other aspects, however their
-\emph{compositionality properties} are deeply enmeshed with that of other
-aspects.  While writing the code for these was straightforward enough,
-this required quite a bit of type-level infrastructure, as well as the
-preflight checks%
-\footnote{We had originally~\cite{CaretteKiselyov05} implemented all our
-preflight tests at the type level, but this did not scale.}
-to get everything to work properly.
+The module of the signature |INTERNAL_FEATURES| bundles information
+tracking aspects. The latter are not directly selectible by the
+user. Rather, they are \emph{function} of other user choices. The
+implementations of the tracking aspects such as |Rank|, |NoRank|,
+|PackedLower| are quite similar (and simpler) than the implementation
+of the |AbstractDet| and |NoDet| aspects in Section~\ref{sec:det}. The
+choice of a particular tracking aspect may depend on all other choices
+(e.g., it is not possible to extract the |L| factor if the domain is
+not a field). Currently we use pre-flight checks to ensure consistency
+of tracking aspects. Previously~\cite{CaretteKiselyov05} we tried to
+implement all our preflight tests at the (module) type level, using
+sharing constraints and module computations. That lead to obscure
+code, long impenetrable type error messages and made compilation very
+slow.
 
 \subsection{Main generation}
-We combine all aspects in a record, which serevs as a keyword
-argument list to the |GenGE| functor below.
+We combine all user-selectable aspects in a record, which serves as a `keyword
+argument' list to the |GenGE| functor below.
 \begin{code}
@@
 generator by passing the functor the `record' of various aspects; the
-order is irrelevant, but all aspects like |Det|, etc. must be
+order is irrelevant, but all aspects suc as |Det| must be
 specified. In the code above, we request GE generation with full
@@
 by printing |resIV5|. The code can then be ``compiled'' as |!. resIV5|
-or with off-shoring. The code for |resIV5| (Appendix A) shows full
+or with off-shoring. The code for |resIV5|
+(App.~\ref{app:code-GenIV5}) shows full
 pivoting, determinant and rank tracking. The code for all these
@@
 \end{code}
-The code generated by |GenFA9| (Appendix B) shows no traces of
+The code generated by |GenFA9| (App.~\ref{app:code-GenFA9}) shows no traces of
 determinant tracking whatsoever: no declaration of spurious variables,
@@
 nonzero, no pivoting), update policy (with either `fraction-less'
-or full division), and the result specification. Some of the
+or full division), and the output specification. Some of the
 argument modules such as |PIVOT| are functors themselves (parameterized
@@
 in the following structure of the |UPDATE| signature:
-\vspace*{-5pt}\begin{code}
+\begin{code}
 module DivisionUpdate(Det:DETERMINANT) =
@@
 %
-This structure implements an update policy of using
-|Dom.div| operation without restrictions -- which is possible only if
-the domain has such an unrestricted operation. A domain such as the integer
-domain may still provide |Dom.div| of the same type, but that operation may
-only be used when we are sure that the division is exact. Thus if we
+This structure implements an update policy relying on
+unrestricted |Dom.div|. Many domains provide |Dom.div|, for example,
+the integer domain. The latter however assumes that division is applied
+only if the dividend is the exact multiple of the divisor. Thus if we
 specified |module Update = DivisionUpdate| when instantiating |GenIV5|
@@
 versatile.}
+\oleg{yes, please!}
 
@@
 In our work, aspects are weaved together ``from scratch'' to
-make up a piece of code/functionality.  One can understand previous work to be
+make up a piece of code.  One can understand previous work to be
 more akin to dynamically typed, and dynamically specified aspect weaving, while
@@
 While the first two families of aspects (abstract domain and concrete
-representation) are the most obvious, it is quite
-difficult to separate them out cleanly.  This seems to stem from the fact
-that in a non-staged setting, separating out these aspects leads to such
-fantastically inefficient code, especially in the setting of scientific
-computation, that our experience works \emph{against us}.  While it would
-be too much to speak of a paradigm shift, it is necessary to firmly think
-in terms of abstract ideal design while avoiding all thoughts of traditional
-efficiency.  Interestingly, in conversations that the first author had
-with D. Parnas, we discovered this is apparently
-what~\cite{journals/cacm/parnas72a} was advocating.  We believe that we are
-taking the first steps towards a \emph{typed yet efficient} realization of
-these ideas, where the various design-time entities can be directly encoded in
-machine-checkable form.
+representation) are the most obvious, it is quite difficult to
+separate them out cleanly. Attempts at such a separation in a
+non-staged setting, have lead to fantastically inefficient code,
+unacceptable for scientific computation. Staging permitted us, perhaps
+for the first time, to think in terms of an ideal design without
+worrying about abstraction penalties.  Interestingly, in conversations
+of the first author with D. Parnas, we discovered this is
+apparently what~\cite{journals/cacm/parnas72a} was advocating.  We
+believe that we are taking the first steps towards a \emph{typed yet
+  efficient} realization of these ideas, where the various design-time
+entities can be directly encoded in machine-checkable form.
 
@@
 improvement in the design.
+
+\oleg{I'm inclined to suggest to omit the following, or at least
+  abbreviate. The omitting may be the best option as we don't have
+  much time for editing. The problem with explaining methodology is
+  like explaining Buddhism: it is all obvious. As I read about Zen
+  Buddhism: its simple truths are understandable if you already know
+  them.}
 
@@
 \bibliographystyle{elsart-num}
-\section{Appendix 0}
+\appendix
+\section{The \texttt{perform} monad notation}
 \label{app:perform}
-Grammar of our perform monad
 We support four different constructs to introduce a monadic
-expressions.
-
+expressions:
 \begin{code}
@@
 \end{code}
-
 which is almost literally the grammar of the Haskell's "do"-notation,
@@
 |perform| and |<--|.
-
 We support not only |let x = foo in ...|  expressions but arbitrarily
@@
 pattern-match--failure function to |failwith| (only used for refutable
-patterns).  To select a different function, use the
-extended forms of |perform|. For example, use the function named 
-|bind| from module |Mod|.  In
-addition use the module's |failwith|-function in refutable patterns.
+patterns).  Extended forms of |perform| let us override these
+defaults. For example, to use the function named 
+|bind| from module |Mod|, we write
 \begin{code}
@@
 \end{code}
-The code has full explanation...
-
-\section{Appendix A}
+
+\section{Code of the \texttt{GenIV5} algorithm}
+\label{app:code-GenIV5}
 The code generated for |GenIV5|, fraction-free LU of the integer matrix
@@
 \end{code2}
-\section{Appendix B}
+
+\section{Code of the \texttt{GenFA9} algorithm}
+\label{app:code-GenFA9}
 The code generated for |GenFA9|, LU of the floating point

### 2007-07-14 SCP2008/paper/metamonads.tex
final version

@@
-\documentclass[draft]{elsart}
-\hyphenation{ana-lysis theo-ry}
+\documentclass{elsart}
 
@@
 \DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\small}
-\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\small}
+\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\scriptsize}
 \newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
@@
 \hyphenation{ne-ces-sa-ry}
+\hyphenation{ana-lysis theo-ry}
 
@@
 \begin{abstract}
-We use multi-stage programming, monads and OCaml's
+We use multi-stage programming, monads and Ocaml's
 advanced module system to demonstrate how to eliminate all
@@
 Gaussian Elimination algorithms.  In the present work, we outline a
-number of additional aspects involved in the faimily of LU
+number of additional aspects involved in the family of LU
 decomposition algorithms.  These will first be presented in a somewhat
-ad hoc manner, roughly coresponding to the order in which they were
+ad hoc manner, roughly corresponding to the order in which they were
 ``discovered''.  We then reorganize them into groups of
@@
 which were also concerned with scientific software}.  We firmly believe that
-our typed generative methodogy is much better suited to functional
+our typed generative methodology is much better suited to functional
 programming, compared to attempts to graft the program-trace-based
@@
 		(whether numerical or coefficient growth), if a domain 
-	  possesses
-		an appropriate length measure, it was sometimes used to choose
-		an ``optimal'' pivot.  Not all domains have such a measure.
+		possesses an appropriate length measure, it was sometimes used to
+		choose an ``optimal'' pivot.  Not all domains have such a measure.
 	\item \textbf{Full division}: Whether the input domain supports full
@@
 is to be performed. One could therefore select aspects that turn out
-incompatible, and we are to prevent this. 
+incompatible, and we have to prevent this. 
 More precisely, our goal is to detect the selection of
@@
 
-\oleg{Perhaps we should just add the following phrase, paraphrased
-  from your remark:}
 The groupings are not entirely orthogonal (for example, in-place
@@
 execute them. The following shows the simplest code generator |one|,
-and the simplest code combinators\footnote{%
+and the simplest code combinator\footnote{%
 $\Longrightarrow$ under an expression shows the result of its evaluation}:
@@
 expression, which is inlined in the place of the escape. MetaOCaml can
-print out code expressions, so we can see the final generated code. It
-has no traces of |gen| or |plus|: their applications are done at the
-generation stage.
+print out code expressions, so we can examine the final generated code. It
+has no traces of |gen| or |plus|: they are generation stage computations.
 
@@
 |simplest_code| is not a function on integers -- it is a code
-expression which \emph{encodes} a function.
+expression which \emph{represents} (or encodes) a function.
 
@@
 with two occurrences of |plus y one|,
-which may be quite a complex computation and so we would rather not do
-it twice. We may be tempted to rely on the compiler's
+which may be a rather complex computation which we would rather not do
+twice. We may be tempted to rely on the compiler's
 common-subexpression elimination optimization. When the generated code is
 very complex, however, the compiler may overlook common subexpressions.  Or the
-subexpressions may occur in such an imperative context where the compiler
-might not be able to determine if lifting them is sound. So, being
+subexpressions may occur in an imperative context where the compiler
+might not be able to determine whether lifting them is sound. So, being
 conservative, the optimizer will leave the duplicates as they are. 
@@
 \end{code}
-with the corresponding code generating let-insertion for the future
+with the corresponding code generating let-insertion for a future
 stage
@@
 solution to these sorts of problems has been suggested by 
-Jacques Garrigue on the Caml mailing list 
-\jacques{edited to here}
-\oleg{I'd rather keep the following lines in the comments. I know
-  Jacques has been credited, but I don't have the exact reference
-  where he proposed it.}
+Jacques Garrigue on the Caml mailing list.
 %% (cited from
@@
 Finally, |runM| runs our monad, that is, given the initial state,
-performs the computation of
-the monad and returns its result, which in our case is the code
+it performs the computation of
+the monad and returns its result, which in our case is a code
 expression. We run the monad by passing it the initial state and the
@@
 The function
-|param_code4|, written in the |perform|-notation, is equivalent to
+|param_code4|, written using the |perform|-notation, is equivalent to
 |param_code3| -- in fact, the camlp4 preprocessor will convert the
@@
 already generated pieces of code rather than monadic values. We
-``lift it'':
+can ``lift it'':
 \begin{code}
@@
 when we have verified that the index is non-negative. This example
-makes it clear that the code generation (such as the one in |retN|) is 
-truly an effect and we have to be clear about the sequencing of
+makes it clear that code generation (such as the one in |retN|) is 
+truly an effect, and we have to be clear about the sequencing of
 effects when generating control constructions such as conditionals.
@@
 mechanisms we want to structure our code generators, as long as none
-of those abstractions infiltrate the generated code. ML module system
-turns out most convenient.
-
-\subsection{Domains}
+of those abstractions infiltrate the generated code. The ML module system
+turns out to be most convenient.
+
+\subsection{Domains}\label{sec:domains}
 
@@
 extensibility, type abstraction and constraints, and especially
-parameterized structures (\emph{functors}).  We define the type of the
-domain, the signature |DOMAIN|, which different domains must satisfy:
+parameterized structures (\emph{functors}).  We define the type,
+the signature |DOMAIN|, which different domains must satisfy:
 
@@
 |module IntegerDomain : DOMAIN| makes the compiler verify that our
-|IntegerDomain| is indeed a |DOMAIN|, that is, satisfies the required
-signature. The constraint |DOMAIN| may be omitted; in that case, the
+|IntegerDomain| is indeed a |DOMAIN|. 
+The constraint |DOMAIN| may be omitted; in that case, the
 compiler will verify the type when we try to use that structure as a
 |DOMAIN|. In any case, the errors such as missing ``methods'' or
-methods with incorrect types will be caught statically, even
+methods with incorrect types will be caught statically, 
 \emph{before} any code generation takes place. The variant
-|Domain_is_Ring| of |IntegerDomain.domain_kind| encodes a semantic constraint 
-that the full division
+|Domain_is_Ring| of |IntegerDomain.domain_kind| encodes a semantic constraint:
+that full division
 is not available. While the |DOMAIN| type may have looked daunting to
@@
 
-A more complex domain is |Zp| of the field of integers in prime
-modulus:
+A more complex domain is |Zp|, the field of integers in prime
+characteristic:
 \begin{code}
@@
 \end{code}
-This domain is parametrized by an integer number |p|. 
+This domain is parametrized by an integer |p|. 
 To be more precise, the structure |ZpMake| is
 parameterized over another structure of the type described by
-the signature P; that argument structure has one field, the |int| value |p|.
+the signature |P|, which has one field, the |int| value |p|.
 Such a parameterized structure (or, a function from structures to
-structures) is a \emph{functor}. The result of |ZpMake| is a domain that is a
+structures) is a \emph{functor}. The result of |ZpMake| is a domain which is a
 field with no defined order. Hence
 |normalizer| and |better_than| are set to |None|. |Zp| forms
-a field only when |p| is prime. Therefore, we must make this check,
+a field only when |p| is prime\footnote{or a prime power, a case we do
+not treat here.}.  Since we intend to make a field of prime characteristic, 
+we must check this, as is done
 in the last line of the above code. That line differs from the other 
 bindings in |ZpMake| in that it neither defines a function, such as
-|plus|, nor binds a value, such as |zero|. Therefore, the non-value
-expression |assert (is_prime P.p)|, to be called an initializing
+|plus|, nor binds a value, such as |zero|. This non-value
+expression |assert (is_prime P.p)|, which we will call an initializing
 expression, will be evaluated
-when the corresponding module is instantiated, as in the following
+when the corresponding module is instantiated.
 \begin{code}
@@
 replace |p = 19| with |p = 9| above, we receive a run-time
-error. It is raised however as we instantiate and combine modules that
-will eventually make the generator.  Although the error is reported at
-run-time rather than compile time as one might have hoped, the error
+error. However, it is raised as we instantiate and combine modules that
+will \emph{eventually} make the generator.  Although the error is reported at
+``run-time'' rather than during compilation as one might have hoped, the error
 is raised when \emph{generating the generator} -- well before the
 generation of the target code could begin.  In our code we make
-extensive use of these so-called preflight checks performed as part of
-module initialization. The checks seem to offer a good compromise:
-they are dynamic and so do not require complicated type systems; on
-the other hand, the checks are run quite early, when buiding
+extensive use of these ``preflight checks'', which are performed as part of
+module initialization. These checks seem to offer a good compromise:
+they are dynamic and so do not require a complicated type system; on
+the other hand, the checks are run quite early, when building
 code generators, and so ensure that no code violating the
 corresponding \emph{semantic} constraints will be generated. Although
-some may frown on the use of module initializing expressions since
-that requires careful attention to sharing and multiple instantiations
+some may frown on the use of module initializing expressions 
+as this requires careful attention to sharing and multiple instantiations
 of a module, these concerns do not apply in our case: our preflight
@@
 
-\subsection{Abstract code and lifted domains}
+\subsection{Abstract code and lifted domains}\label{sec:lifteddomains}
 We will be using MetaOCaml operations to build code expressions, values
-of the type |('a,'v) code|. However, we wish to make our generators
+of type |('a,'v) code|. However, we wish to make our generators
 abstract over particular code representations. Therefore, when writing
 the LU generator (Section~\ref{sec:det} and below), instead of
-MetaOCaml brackets and escapes we will be using code combinators
+MetaOCaml brackets and escapes we will be using the code combinators
 described below. These combinators are functions that produce and
-combine values of the type of abstract code.
+combine values of the type of \emph{abstract code}.
 
 This encapsulation of staging annotations into code combinators lets
-us easily change the output form of our generated LU programs. We may
-use single-stage thunks rather than MetaOCaml staged |('a,'v) code|
-values, for benchmarking and regression tests purposes. We could use
+us easily change the output form of our generated LU programs.  For example,
+We may use single-stage thunks rather than MetaOCaml staged |('a,'v) code|
+values, for benchmarking and regression tests purposes. We could also use
 code combinators producing C or Fortran code.
@@
 Our code combinator library includes operations to create, dereference
-and mutate reference cells, compare values, manipulate values used
+and mutate reference cells, compare values, and manipulate values used
 for indexing arrays. The following is the implementation of sample
 combinators, in the case of the abstract code being MetaOCaml code. 
+We basically need to ``lift'' most the underlying programming language
+into combinators.
 \begin{code}
@@
 generally lifted twice: once from the value domain |v| to the code
-domain |'a vc|, and once more from values to monadic computations
-|('p,'a vc) monad|.
+domain |('a,'v) code|, and once more from values to monadic computations
+|('p,'v) monad|.
 
@@
 says that lifted domains include all members of non-lifted domains,
-including the initializing expressions with pre-flight checks.
+including the initializing expressions with preflight checks.
 The following is a particular instance of |DOMAINL|, the lifted
@@
 \end{code}
+Such lifting is completely straightforward.  However, it is a
+program-text to program-text transformation over modules, and as such could
+only be automated (currently) by further use of camlp4.
 
@@
 
-\subsection{Maintaining the extensible state}
+\subsection{Maintaining an extensible state}
 
 Various aspects of our LU generator such as determinant, rank, 
-and permutation traking may need to keep a state during
-the code generation. For example, the initializing section of the LU
+and permutation tracking may need to keep a state during
+code generation. For example, the initializing section of the LU
 generator will invoke the function |decl| of the determinant aspect so that the
@@
 row-swapping code will notify the determinant aspect so that the
-latter may insert the code that flips the sign of the tracked
+latter may insert code that flips the sign of the tracked
 determinant. The final section of the LU generator will ask the
@@
 state from one invocation of its functions to another; this state should
-include the `names' of mutable variables accumulating the sign and the
+include the ``names'' of mutable variables accumulating the sign and the
 magnitude of the determinant.
@@
 The simplest method of keeping the state of the generator is by using
-mutable variables private to the Deteminant module (aspect). That,
+mutable variables, private to the Determinant module (aspect). That,
 however, makes our aspects stateful. Although we are generating
 imperative code, we would like to keep our generators stateless and
-pure functional, for ease of comprehension and reasoning.  Our main
+purely functional, for ease of comprehension and reasoning.  Our main
 program may include several LU generators referring to one Determinant
@@
 
-We chose therefore a different way of maintaining the generator state,
-using a monad. We need monad anyway for let-insertion; we add state as
+We therefore chose a different way of maintaining the generator state,
+using a monad. We already need a monad for let-insertion; we add state as
 an additional monad parameter (Fig.~\ref{ourmonad}) and monadic
@@
 
-This monadic state has to accommodate several pieces of state, for
+This monadic state has to accommodate several distinct pieces of state, for
 various aspects. We should be able to add a new aspect -- which may need
@@
 state should be extensible. We could use an extensible
-record: an OCaml object. Each aspect will have its own field; record
-subtyping will assure modularity. Alas, this approach makes it
-difficult to create the initial state, to pass to the monad's |runM|
-method. We should know the names of all fields and should know the
-proper initial value for these fields, which breaks modularity.
-
-A better approach of a property list has been suggested by the MLton
-team \cite{mlton-proplist}. A property list, too, represents an
+record: an OCaml object. Each aspect would have its own field; record
+subtyping would assure modularity. Alas, this approach makes it
+difficult to create an initial state, to pass to the monad's |runM|
+method. We would be required to know the names of all fields and should know
+the proper initial value for these fields, which breaks modularity.
+
+The MLton team suggests a better approach: property 
+lists~\cite{mlton-proplist}.  A property list, too, represents an
 extensible object -- by its dual, a list of polymorphic variants.  The
 initial object is just the empty list. MLton's approach does not
-literally apply to us however. Environment classifiers of MetaOCaml
+literally apply to us however. MetaOCaml's environment classifiers
 require type universes parameterized by classifiers. The MLton
@@
 and |orec_find| to obtain the value associated with a particular field
-name. Each `field' is characterized by a triple: injection, 
-projection functions and the string name. The latter is used for 
+name. Each ``field'' is characterized by a triple: an injection function, 
+a projection function and the string name. The latter is used for 
 printing error messages. For example, for the
@@
 %
-Currently, we check at generator-time that one should not add an
+Currently, we check at generation-time that one should not add an
 already existing field to an open record, nor should one attempt to
@@
 \label{sec:det}
-To track determinant we should be able to generate code
+To track the determinant, we should be able to generate code
 for: defining variables used for tracking (|decl|),
-updating the sign or the absolute
-value of the determinant, converting the tracking state
-to the final determinant value of the type |outdet|. LU of a
+updating the sign (|upd_sign|) or the absolute
+value of the determinant, and converting the tracking state
+to the final determinant value of type |outdet|. LU of a
 floating-point matrix with no determinant tracking uses the
@@
 
-\begin{code2}
+\begin{code}
 module type OUTPUTDEP = sig 
@@
 end
-
 module OutDetRank(OD : OUTPUTDEP) = struct
@@
 end
-\end{code2}
+\end{code}
 
 The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
-the pre-flight checks. As we saw in the previous section, 
-both instances of |Det| contain the |fin ()| function to generate the
+the preflight checks. As we saw in the previous section, 
+both instances of |Det| contain |fin ()| function to generate
 code representing the computed determinant. Only the |NoDet| does not
 track any determinant, and so |fin ()| raises an error. The code 
-|let _ = OD.Det.fin ()| in |OutDetRank| invokes that |fin| function, which will
+|let _ = OD.Det.fin ()| in |OutDetRank| invokes this |fin| function, which will
 produce the monadic code generating action, or raise an error. We do
@@
 action to run. This is another preflight check, to rule out the
-semantic error when a user specifies that the determinant should be computed
-and returned and yet specifies the |NoDet| aspect which tracks no
+semantic error where a user specifies that the determinant should be computed
+and returned, and yet specifies the |NoDet| aspect which tracks no
 determinant at all.
 
-The module of the signature |INTERNAL_FEATURES| bundles information
-tracking aspects. The latter are not directly selectible by the
-user. Rather, they are \emph{function} of other user choices. The
+The module of signature |INTERNAL_FEATURES| bundles information
+tracking aspects. The latter are not directly selectable by the
+user. Rather, they are \emph{functions} of other user choices. The
 implementations of the tracking aspects such as |Rank|, |NoRank|,
@@
 (e.g., it is not possible to extract the |L| factor if the domain is
-not a field). Currently we use pre-flight checks to ensure consistency
+not a field). Currently we use preflight checks to ensure consistency
 of tracking aspects. Previously~\cite{CaretteKiselyov05} we tried to
@@
 code, long impenetrable type error messages and made compilation very
-slow.
+slow.  Furthermore, at least in OCaml, type-level computations are not
+powerful enough to help us with tasks such as verifying that an integer
+is prime, as required in section~\ref{sec:domains}.
 
 \subsection{Main generation}
-We combine all user-selectable aspects in a record, which serves as a `keyword
-argument' list to the |GenGE| functor below.
+We combine all user-selectable aspects in a record, which serves as a ``keyword
+argument'' list to the |GenGE| functor below.
 \begin{code}
@@
 %
-We first instantiate the container |GVC_I| for the vector over an
+We first instantiate the container |GVC_I| for a vector over an
 integer domain. We pass the container to the main functor |GenLA|,
-since all the linear algebra aspects are parameterized by the
+since all linear algebra codes are parameterized by the
 container and the domain included in it. We open the |G_GVC_I| module
@@
 generator functor is |GenGE|, shown later.  We instantiate the
-generator by passing the functor the `record' of various aspects; the
-order is irrelevant, but all aspects suc as |Det| must be
+generator by passing to the functor the ``record'' of various aspects; the
+order is irrelevant, but all aspects such as |Det| must be
 specified. In the code above, we request GE generation with full
@@
 
-The LU generator functor itself is 
+The LU generator functor (called |GenGE| below) itself is 
 parameterized by the domain, container, pivoting policy (full, row,
@@
     (U.upd_kind = DivisionBased)
-    (* some more pre-flight tests *)
+  (* some more preflight tests *)
   let _ = ensure ((not wants_pack) || can_pack) 
@@
               U.update bjc pos.curval brk bjk 
-                          fun ov -> C.col_head_set mat.matrix j k ov) det) UP )
+                (fun ov -> C.col_head_set mat.matrix j k ov) det) UP )
               (IF.L.updt mat.matrix j pos.p.colpos C.Dom.zeroL 
@@
 
-There are more pre-flight checks for various ``semantic'' 
-constraints, shown
+A careful reading of this code will reveal that the core of the 
+Gaussian Elimination algorithm is still visible in this code generator:
+the |forward_elim| function iterates over the columns and rows of the
+matrix, finding a pivot, and zeroing the appropriate entries.
+
+There are more preflight checks for various ``semantic'' constraints, shown
 in the following structure of the |UPDATE| signature:
@@
 The monad in this paper is similar to the one described in
-\cite{MSP:PADL04,KiselyovTaha}.  However the latter papers used only
+\cite{MSP:PADL04,KiselyovTaha}.  However those papers used only
 |retN| and fixpoints (for generation-time iterations).  This paper
@@
 subtle type errors, shall not occur in the generated code.  SPIRAL
-\cite{Pueschel:05} is another such even more ambitious project.  But
+\cite{Pueschel:05} is another even more ambitious project.  But
 SPIRAL does intentional code analysis, relying on a set of code
-transformation ``rules'' which make sense, but which are not proved to
+transformation ``rules'' which make sense, but which are not proven to
 be either complete or confluent.  The strength of both of these
@@
 \cite{Axiom} and Aldor \cite{Watt:2002:HCA}.  Even though the Aldor
-compiler frequently can optimize away a lot of abstraction overhead, 
+compiler can frequently optimize away a lot of abstraction overhead, 
 it does not provide any guarantees that it will do so, unlike our
@@
 
-\jacques{Mention the relationship with the all-dynamic version of a 
-similar design, namely Maple's?  The fact that all choices there are
-dynamic, while all ours are completely static really shows that the
+Our code was initially motivated by trying to unify the various
+implementations in Maple.  Interestingly, when we compare our end 
+result with the options available from Maple's \texttt{LUDecomposition}
+algorithm, we notice a great deal of similarity.  The biggest difference
+is that in Maple, all the choices are done dynamically (and are dynamically
+typed), while ours choices are done statically, in a statically typed
+environment.  To us, this shows that the
 design space along the dynamic--static dimension is quite large and
-versatile.}
-\oleg{yes, please!}
+versatile.
 
@@
 
+It would be interesting to implement a camlp4 extension that 
+automates the lifting of (simple) modules as done in 
+section~\ref{sec:lifteddomains},
+first to the code level, and then to monadic values.  Even more
+interesting would be a (typed) extension to MetaOCaml that would 
+allow us to write such code with the same guarantees that the rest of
+MetaOCaml already affords us.
+
 We plan to further investigate the connection between delimited
@@
 by complex aspects at no run-time overhead.  The combination of
-stateless functors and structures, and our monad with the
+stateless functors and structures, and our monad with
 compositional state makes aspects composable without having to
@@
 \cite{kiczales97aspectoriented}: in AspectJ, 
-aspects are (comparatively) lightly typed, and are post-facto extensions of an
-existing piece of code.  More precisely, they are post-facto extensions of
+aspects are (comparatively) lightly typed, and are post-facto extensions of
 potential program traces, specified in a particular language; these tend
@@
 
-\oleg{I'm inclined to suggest to omit the following, or at least
-  abbreviate. The omitting may be the best option as we don't have
-  much time for editing. The problem with explaining methodology is
-  like explaining Buddhism: it is all obvious. As I read about Zen
-  Buddhism: its simple truths are understandable if you already know
-  them.}
+%\oleg{I'm inclined to suggest to omit the following, or at least
+%  abbreviate. The omitting may be the best option as we don't have
+%  much time for editing. The problem with explaining methodology is
+%  like explaining Buddhism: it is all obvious. As I read about Zen
+%  Buddhism: its simple truths are understandable if you already know
+%  them.}
+%
+%\jacques{All obvious, but worthwhile documenting anyways.  I really
+%  like reading papers where the ``obvious truth'' is explained in
+%  clear enough ways that I nevertheless learn something new.  Sometimes,
+%  all one really learns is a better vocabulary for describing something
+%  known, but that is already quite useful.  So I leave it in.}
 
@@
 allowed us to both generate (efficient) code and to write a directly
-runnable (highly inefficient) version of the algorithm from the same
-generator.  This brought out the issue that, once this was done, we had
+runnable (but highly inefficient) version of the algorithm from the same
+generator.  But this
 essentially abstracted out all of the syntactic sugar of the underlying
@@
 approaches to product families, as we do not over-engineer our design
-by imagining weird variations that are unlikely to come up in realistic
+by imagining variations that are unlikely to come up in realistic
 situations, but only create variations when we notice them in actual use.
@@
 module is created to encapsulate the various concepts, and these are
-then implemented as generators.  It is very important at this stage to
+implemented as generators.  It is very important at this stage to
 make sure to reify all available \emph{static} information, so that all
@@
 
-\bibliography{metamonads}
-\bibliographystyle{elsart-num}
+% Embed .bbl file directly to make document self-contained
+% \bibliography{metamonads}
+% \bibliographystyle{elsart-num}
+
+\begin{thebibliography}{10}
+\expandafter\ifx\csname url\endcsname\relax
+  \def\url#1{\texttt{#1}}\fi
+\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
+
+\bibitem{Carette06}
+J.~Carette, Gaussian {E}limination: a case study in efficient genericity with
+  {MetaOCaml}, Science of Computer Programming 62~(1) (2006) 3--24, special
+  Issue on the First MetaOCaml Workshop 2004.
+
+\bibitem{Monagan:2001:M7P}
+M.~B. Monagan, K.~O. Geddes, K.~M. Heal, G.~Labahn, S.~M. Vorkoetter,
+  J.~McCarron, P.~DeMarco, Maple 7 Programming Guide, Waterloo Maple Inc.,
+  2001.
+
+\bibitem{Gruntz:1994:IG}
+D.~Gruntz, M.~Monagan, Introduction to {Gauss}, \textsc{SIGSAM Bulletin}:
+  Communications on Computer Algebra 28~(2) (1994) 3--19.
+
+\bibitem{Axiom}
+R.~D. Jenks, R.~S. Sutor, {AXIOM}: The Scientific Computation System, Springer
+  Verlag, 1992.
+
+\bibitem{Czarnecki}
+K.~Czarnecki, U.~W. Eisenecker, Generative programming: methods, tools, and
+  applications, ACM Press/Addison-Wesley Publishing Co., 2000.
+
+\bibitem{Veldhuizen:1998:ISCOPE}
+T.~L. Veldhuizen, Arrays in {Blitz++}, in: Proceedings of the 2nd International
+  Scientific Computing in Object-Oriented Parallel Environments (ISCOPE'98),
+  Lecture Notes in Computer Science, Springer-Verlag, 1998.
+
+\bibitem{musser89generic}
+D.~R. Musser, A.~A. Stepanov, Generic programming, in: {ISSAC 1988}:
+  Proceedings of the International Symposium on Symbolic and Algebraic
+  Computation, Vol. 358 of Lecture Notes in Computer Science, Springer-Verlag,
+  1989, pp. 13--25.
+
+\bibitem{musser94algorithmoriented}
+D.~R. Musser, A.~A. Stepanov, Algorithm-oriented generic libraries, Software -
+  Practice and Experience 24~(7) (1994) 623--642.
+
+\bibitem{BOOST}
+J.~Siek, L.-Q. Lee, A.~Lumsdaine, The Boost Graph Library: User Guide and
+  Reference Manual, Addison-Wesley, 2002.
+
+\bibitem{POOMA}
+I.~John V.W.~Reynders, J.~C. Cummings, The {POOMA} framework, Comput. Phys.
+  12~(5) (1998) 453--459.
+
+\bibitem{ATLAS}
+R.~C. Whaley, A.~Petitet, J.~J. Dongarra, Automated empirical optimization of
+  software and the {ATLAS} project, Parallel Computing 27~(1--2) (2001) 3--35.
+
+\bibitem{HygienicMacros}
+E.~E. Kohlbecker, D.~P. Friedman, M.~Felleisen, B.~F. Duba, Hygienic macro
+  expansion., in: LISP and Functional Programming, 1986, pp. 151--161.
+
+\bibitem{camlp4}
+D.~de~Rauglaudre, Camlp4 reference manual,
+  \url{http://caml.inria.fr/camlp4/manual/} (Jan. 2002).
+
+\bibitem{conf/dagstuhl/CzarneckiOST03}
+K.~Czarnecki, J.~T. O'Donnell, J.~Striegnitz, W.~Taha, {DSL} implementation in
+  {MetaOCaml}, {Template Haskell}, and {C++}., in: C.~Lengauer, D.~S. Batory,
+  C.~Consel, M.~Odersky (Eds.), Domain-Specific Program Generation, Vol. 3016
+  of Lecture Notes in Computer Science, Springer, 2003, pp. 51--72.
+
+\bibitem{CTHL03}
+C.~Calcagno, W.~Taha, L.~Huang, X.~Leroy, Implementing multi-stage languages
+  using asts, gensym, and reflection, in: K.~Czarnecki, F.~Pfenning,
+  Y.~Smaragdakis (Eds.), Generative Programming and Component Engineering
+  (GPCE), LNCS, Springer-Verlag, 2003.
+
+\bibitem{metaocaml-org}
+{MetaOCaml}, \url{http://www.metaocaml.org}.
+
+\bibitem{TahaSheard97}
+W.~Taha, T.~Sheard, Multi-stage programming with explicit annotations, in:
+  Proceedings of the Symposium on Partial Evaluation and Semantic-Based Program
+  Manipulation {(PEPM)}, ACM Press, Amsterdam, 1997, pp. 203--217.
+
+\bibitem{TahaThesis}
+W.~Taha, Multi-stage programming: its theory and applications, Ph.D. thesis,
+  Oregon Graduate Institute of Science and Technology (1999).
+
+\bibitem{Taha2000}
+W.~Taha, A sound reduction semantics for untyped {CBN} multi-stage computation.
+  {Or}, the theory of {MetaML} is non-trival., in: PEPM, 2000, pp. 34--43.
+
+\bibitem{Pueschel:05}
+M.~P{\"u}schel, J.~M.~F. Moura, J.~Johnson, D.~Padua, M.~Veloso, B.~W. Singer,
+  J.~Xiong, F.~Franchetti, A.~Ga\v{c}i\'{c}, Y.~Voronenko, K.~Chen, R.~W.
+  Johnson, N.~Rizzolo, {SPIRAL}: Code generation for {DSP} transforms,
+  Proceedings of the IEEE, special issue on "Program Generation, Optimization,
+  and Adaptation" 93~(2).
+
+\bibitem{Kennedy01Telescoping}
+K.~Kennedy, B.~Broom, K.~Cooper, J.~Dongarra, R.~Fowler, D.~Gannon,
+  L.~Johnsson, J.~Mellor-Crummey, L.~Torczon, Telescoping languages: A strategy
+  for automatic generation of scientific problem-solving systems from annotated
+  libraries, Journal of Parallel and Distributed Computing 61~(12) (2001)
+  1803--1826.
+
+\bibitem{dongarra7}
+Z.~Chen, J.~Dongarra, P.~Luszczek, K.~Rothe, Lapack for clusters project: An
+  example of self adapting numerical software, Hawaii International Conference
+  on System Sciences HICSS-37.
+
+\bibitem{Veldhuizen:2004}
+T.~L. Veldhuizen, Active libraries and universal languages, Ph.D. thesis,
+  Indiana University Computer Science (May 2004).
+\newline\urlprefix\url{\url{http://osl.iu.edu/~tveldhui/papers/2004/dissertati%
+on.pdf}}
+
+\bibitem{KiselyovTaha}
+O.~Kiselyov, K.~N. Swadi, W.~Taha, A methodology for generating verified
+  combinatorial circuits, in: EMSOFT '04: Proceedings of the fourth ACM
+  international conference on Embedded software, ACM Press, New York, NY, USA,
+  2004, pp. 249--258.
+
+\bibitem{Padua:MetaOcaml:04}
+A.~Cohen, S.~Donadio, M.~J. Garzara'n, C.~Herrmann, D.~Padua, In search for a
+  program generator to implement generic transformations for high-performance
+  computing, metaOCaml Workshop, October 2004.
+
+\bibitem{CaretteKiselyov05}
+J.~Carette, O.~Kiselyov, Multi-stage programming with {F}unctors and {M}onads:
+  eliminating abstraction overhead from generic code, in: Generative
+  Programming and Component-based Engineering {GPCE}, 2005, pp. 256--274.
+
+\bibitem{MSP:PADL04}
+J.~L. Eckhardt, R.~Kaiabachev, K.~N. Swadi, W.~Taha, O.~Kiselyov, Practical
+  aspects of multi-stage programming, rice University Techical Report
+  {TR05}-451,
+  \url{http://www.cs.rice.edu/~taha/publications/preprints/2004-02-16.pdf}
+  (Feb. 2004).
+
+\bibitem{metamonadsURL}
+{Source code}, \url{http://www.cas.mcmaster.ca/~carette/metamonads/}.
+
+\bibitem{Golub-vanLoan}
+G.~H. {Golub}, C.~F. {Van Loan}, Matrix Computations, 3rd Edition, Johns
+  Hopkins University Press, Baltimore, MD, 1996.
+
+\bibitem{kiczales97aspectoriented}
+G.~Kiczales, J.~Lamping, A.~Menhdhekar, C.~Maeda, C.~Lopes, J.-M. Loingtier,
+  J.~Irwin, Aspect-oriented programming, in: M.~Ak\c{s}it, S.~Matsuoka (Eds.),
+  Proceedings European Conference on Object-Oriented Programming, Vol. 1241,
+  Springer-Verlag, Berlin, Heidelberg, and New York, 1997, pp. 220--242.
+
+\bibitem{709568}
+J.~Irwin, J.-M. Loingtier, J.~R. Gilbert, G.~Kiczales, J.~Lamping,
+  A.~Mendhekar, T.~Shpeisman, Aspect-oriented programming of sparse matrix
+  code, in: ISCOPE '97: Proceedings of the Scientific Computing in
+  Object-Oriented Parallel Environments, Springer-Verlag, London, UK, 1997, pp.
+  249--256.
+
+\bibitem{mendhekar97rg}
+A.~Mendhekar, G.~Kiczales, J.~Lamping, {RG}: {A} case-study for aspect-oriented
+  programming, Tech. Rep. SPL97-009 P9710044, Palo Alto, CA, USA (February
+  1997).
+
+\bibitem{journals/cacm/parnas72a}
+D.~L. Parnas, On the criteria to be used in decomposing systems into modules.,
+  Commun. ACM 15~(12) (1972) 1053--1058.
+
+\bibitem{journals/tse/Parnas76}
+D.~L. Parnas, On the design and development of program families., IEEE Trans.
+  Software Eng. 2~(1) (1976) 1--9.
+
+\bibitem{EWD:EWD447}
+E.~W. Dijkstra, On the role of scientific thought, published as
+  \cite{EWD:EWD447pub}.
+\newline\urlprefix\url{\url{http://www.cs.utexas.edu/users/EWD/ewd04xx/EWD447.%
+PDF}}
+
+\bibitem{ZhCaJeMo06a}
+W.~Zhou, J.~Carette, D.~J. Jeffrey, M.~B. Monagan, Hierarchical representations
+  with signatures for large expression management, in: Proceedings of
+  Artificial Intelligence and Symbolic Computation, Vol. 4120 of Lecture Notes
+  in Computer Science, 2006, pp. 254--268.
+
+\bibitem{Bondorf:92}
+A.~Bondorf, Improving binding times without explicit {CPS}-conversion, in: 1992
+  ACM Conference on Lisp and Functional Programming. San Francisco, California,
+  1992, pp. 1--10.
+
+\bibitem{conf/pepm/BondorfD94}
+A.~Bondorf, D.~Dussart, Improving {CPS}-based partial evaluation: Writing cogen
+  by hand., in: PEPM, 1994, pp. 1--9.
+
+\bibitem{Haskell98Report}
+S.~Peyton~Jones, et~al., The Revised Haskell 98 Report, Cambridge Univ.\null{}
+  Press, 2003, also on \url{http://haskell.org/}.
+
+\bibitem{moggi-notions}
+E.~Moggi, Notions of computation and monads, Information and Computation 93~(1)
+  (1991) 55--92.
+
+\bibitem{Filinski:Representing}
+A.~Filinski, Representing monads., in: POPL, 1994, pp. 446--457.
+
+\bibitem{liang-interpreter}
+S.~Liang, P.~Hudak, M.~Jones, Monad transformers and modular interpreters, in:
+  {POPL} '95: Conference Record of the Annual {ACM} Symposium on Principles of
+  Programming Languages, {ACM} {P}ress, {N}ew {Y}ork, 1995, pp. 333--343.
+
+\bibitem{taha-environment}
+W.~Taha, M.~F. Nielsen, Environment classifiers, in: {POPL} '03: Conference
+  Record of the Annual {ACM} Symposium on Principles of Programming Languages,
+  {ACM} {P}ress, {N}ew {Y}ork, 2003, pp. 26--37.
+
+\bibitem{mlton-proplist}
+{{MLton} team}, Property list, \url{http://mlton.org/PropertyList}.
+
+\bibitem{Watt:2002:HCA}
+S.~M. Watt, Aldor, in: J.~Grabmeier, E.~Kaltofen, V.~Weispfennig (Eds.),
+  Computer Algebra Handbook: Foundations, Applications, Systems, Springer
+  Verlag, 2003.
+
+\bibitem{Gluck95}
+R.~Gl{\"u}ck, R.~Nakashige, R.~Z{\"o}chling, Binding-time analysis applied to
+  mathematical algorithms, in: System Modelling and Optimization, 1995.
+
+\bibitem{GluckJ97}
+R.~Gl{\"u}ck, J.~J{\o}rgensen, An automatic program generator for multi-level
+  specialization., Lisp and Symbolic Computation 10~(2) (1997) 113--158.
+
+\bibitem{edsl}
+P.~Hudak, Building domain-specific embedded languages, ACM Comput. Surv.
+  28~(4es) (1996) 196.
+
+\bibitem{WhenHowDSL}
+M.~Mernik, J.~Heering, A.~M. Sloane, When and how to develop domain-specific
+  languages, ACM Comput. Surv. 37~(4) (2005) 316--344.
+
+\bibitem{EWD:EWD447pub}
+E.~W. Dijkstra, On the role of scientific thought, in: Selected Writings on
+  Computing: A Personal Perspective, Springer-Verlag, 1982, pp. 60--66.
+
+\end{thebibliography}
 \appendix

### 2007-07-18 tagless/tagless-final.tex
Reinflating the paper for JFP: done up to Section 2.1

--- 2007-06-12 tagless/tagless-final.tex
@@
 Our main idea is to encode
-% de Bruijn or higher-order abstract syntax
-HOAS
+\ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
 using cogen functions rather than data constructors.
@@
     defined language should be adequate for the defining language.
-    \hfill\rm John Reynolds~\citep{reynolds-definitional}
+    \rm \ifshort \hfill John Reynolds~\fi\citep{reynolds-definitional}
 \end{quote}
@@
 proposal and contributions.  
-%aplas
-\begin{comment}
+\ifshort
+We leave aside the solved problem of writing a parser\slash type\hyp checker,
+for embedding object language objects into the metalanguage
+(whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}),
+and just enter them by hand.
+\else
 No matter how we represent the object language in the
@@
 whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}.
-\end{comment}
-%aplas-else
-We leave aside the solved problem of writing a parser\slash type\hyp checker,
-for embedding object language objects into the metalanguage
-(whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}),
-and just enter them by hand.
-%aplas
+\fi
 
@@
 
-% see tagless\_interp1.ml, module Tagfull for the complete code.
-% If you change the code in here, please adjust the .ml file
-% accordingly. Let the paper and the accompanying code be in sync.
-It is straightforward to create an algebraic data type, say in OCaml~%
-%aplas-specific
-Fig.~\ref{fig:tag-problem}(a)%
-, to
-represent object terms such as those in Figure~\ref{fig:object}.
-% fig:tag-problem part (a) was here
-For brevity, we elide treating integers, conditionals, and fixpoint in
-this section.
-We represent each variable using a unary de Bruijn index.
-For example, we represent the object term $(\fun{x}x)\True$ as
-%aplas-specific
-\texttt{let test1 = A (L (V VZ), B true)}.
-%aplas-else
-\begin{comment}
-\begin{code}
-let test1 = A (L (V VZ), B true)
-\end{code}
-\end{comment}
-
-\begin{figure}
-%
-(a) \begin{code2}
+\begin{SaveVerbatim}{2a}
 type var = VZ | VS of var
 type exp = V of var | B of bool | L of exp | A of exp * exp
-\end{code2}
-
-(b) \begin{code2}
-let rec lookup (x::env) = function | VZ   -> x
-                                   | VS v -> lookup env v
+\end{SaveVerbatim}
+\begin{SaveVerbatim}[commandchars=\@\{\}]{2b}
+let rec lookup (x::env) = function VZ -> x | VS v -> lookup env v
 let rec eval0 env = function
@@
 | A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
-\end{code2}
-
-(c) \begin{code2}
+\end{SaveVerbatim}
+\begin{SaveVerbatim}{2c}
 type u = UB of bool | UA of (u -> u)
-\end{code2}
-
-(d) \begin{code2}
+\end{SaveVerbatim}
+\begin{SaveVerbatim}{2d}
 let rec eval env = function
@@
 | A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
-\end{code2}
+\end{SaveVerbatim}
+\begin{SaveVerbatim}{test1}
+let test1 = A (L (V VZ), B true)
+\end{SaveVerbatim}
+
+% see tagless\_interp1.ml, module Tagfull for the complete code.
+% If you change the code in here, please adjust the .ml file
+% accordingly. Let the paper and the accompanying code be in sync.
+It is straightforward to create an algebraic data type, say in OCaml\ifshort\
+Fig.~\ref{fig:tag-problem}(a)\fi, to
+represent object terms such as those in Figure~\ref{fig:object}.
+For brevity, we elide treating integers, conditionals, and fixpoint in
+this section.
+\ifshort\else\UseVerbatim{2a}\fi
+We represent each variable using a unary de Bruijn index.
+For example, we represent the object term $(\fun{x}x)\True$ as
+\ifshort \BUseVerbatim{test1}.\else \UseVerbatim{test1}\fi
+
+\ifshort
+\begin{figure}
+%
+(a) \BUseVerbatim{2a}
+
+(b) \BUseVerbatim{2b}
+
+(c) \BUseVerbatim{2c}
+
+(d) \BUseVerbatim{2d}
 \caption{OCaml code illustrating the tag problem.}
@@
 \end{figure}
-% 
+\fi
+
 Following \cite{WalidICFP02},
-we try to implement an interpreter function |eval0|%
-%aplas-specific
-, Fig.~\ref{fig:tag-problem}(b)%
-. It takes
+we try to implement an interpreter function |eval0|\ifshort,
+Fig.~\ref{fig:tag-problem}(b)\fi. It takes
 an object term such as |test1| above and gives us its value.
@@
 interpreted code.
-% fig:tag-problem part (b) was here
+\ifshort\else\UseVerbatim{2b}\fi
 If our OCaml-like metalanguage were untyped, the code above would be 
@@
 The usual solution is to introduce a universal type \citep[\S1.3]
-{WalidICFP02} containing both booleans and functions%
-%aplas-specific
-, Fig.~\ref{fig:tag-problem}(c)%
-.
-% fig:tag-problem part (c) was here
-We can then write a typed interpreter%
-%aplas-specific
-, Fig.~\ref{fig:tag-problem}(d)
-% fig:tag-problem part (d) was here
+{WalidICFP02} containing both booleans and functions\ifshort,
+Fig.~\ref{fig:tag-problem}(c)\fi.
+\ifshort\else\UseVerbatim{2c}\fi
+We can then write a typed interpreter\ifshort,
+Fig.~\ref{fig:tag-problem}(d), \else\UseVerbatim{2d}\fi
 whose inferred type is |u list -> exp -> u|. Now we can evaluate
-%aplas-specific
+\ifshort
 \texttt{eval [] test1} obtaining |UB true|.
-%aplas-else
-\begin{comment}
+\else
 \begin{code}
@@
 \end{code}
-\end{comment}
-% fig:tag-problem part (f) was here
+\fi
 The unfortunate tag |UB| in the result reflects that |eval| is a partial
 function.  
-%aplas-specific
-First, the pattern match |with UA f| in the line |A (e1,e2)| is not 
-exhaustive, so |eval| can fail if we apply a boolean,
-as in the ill-typed term |A (B true, B false)|.
-Second, the |lookup|
-function assumes a nonempty environment, so |eval| can fail if we
-evaluate an open term |A (L (V (VS VZ)), B true)|.
-After all, the type |exp| represents object
-terms both well-typed and ill-typed, both open and closed.
-%aplas-else
-\begin{comment}
 First, the pattern match |with UA f| in the line
@@
 as in the ill-typed term |A (B true, B false)|.
+\ifshort\else
 \begin{code}
@@
 \end{code}
+\fi
 Second, the |lookup|
@@
 evaluate an open term
+\ifshort
+|A (L (V (VS VZ)), B true)|.
+\else
 \begin{code}
@@
 \end{code}
+\fi
 After all, the type |exp| represents object
 terms both well-typed and ill-typed, both open and closed.
-\end{comment}
 
@@
 
-%aplas-else
-\begin{comment}
-\subsection{Solutions using fancier types}
-\end{comment}
+\ifshort\else\subsection{Solutions using fancier types}\fi
 
@@
 on dependent types \cite{WalidICFP02,fogarty-concoqtion}.
-%aplas-else
-\begin{comment}
+\ifshort\else
 For a metalanguage's type system to allow the well-typed object term
@@
 such as GADTs or dependent types seem necessary.  
-\end{comment}
+\fi
 Yet other type systems
 have been proposed to distinguish closed terms like |test1| from open
-terms like |test3|
-%aplas \cite{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual},
-\cite{WalidPOPL03,NanevskiJFP05,DaviesJACM01},
+terms 
+\ifshort\cite{WalidPOPL03,NanevskiJFP05,DaviesJACM01}\else
+like |test3|
+\cite{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual}\fi,
 so that |lookup| never receives an empty environment.  We discuss these
-proposals further in \S\ref{related}%
-%aplas-specific
-.
-%aplas-else
-\begin{comment}
-; here we just note that many
+proposals further in \S\ref{related}\ifshort\else;
+here we just note that many
 advanced type systems have been devised to ensure statically that an
-object term is well-typed and closed.
-\end{comment}
+object term is well-typed and closed\fi.
 
@@
 We represent object programs using ordinary functions rather than
-data constructors.  The interpreter below provides these functions
-\begin{code4}
+data constructors.  These functions comprise the entire interpreter, shown
+below.
+\ifshort
+\begin{code3}[xleftmargin=0pt]
 let varZ env    = fst env       let b (bv:bool) env = bv
@@
 let lam e env   = fun x -> e (x,env)
-\end{code4}
-which comprise the entire interpreter.
+\end{code3}
+\else
+\begin{code}
+let varZ env        = fst env
+let varS vp env     = vp (snd env)
+let b (bv:bool) env = bv
+let lam e env       = fun x -> e (x,env)
+let app e1 e2 env   = (e1 env) (e2 env)
+\end{code}
+\fi
 We now represent our sample term $(\fun{x}x)\True$ as
-%aplas-specific
-\texttt{let testf1 = app (lam varZ) (b true)}
-%aplas-else
-\begin{comment}
+\ifshort
+\texttt{let testf1 = app (lam varZ) (b true)}.
+\else
 \begin{code}
@@
 \end{code}
-\end{comment}
+\fi
 This representation is almost the same as in \S\ref{tagproblem}, only
 written with lowercase identifiers. To evaluate an object term is to
-apply its representation to the empty environment%
-%aplas-specific
-, |testf1 ()|, obtaining |true|.
-%aplas-else
-\begin{comment}
+apply its representation to the empty environment\ifshort
+, |testf1 ()|, obtaining |true|\fi.
+\ifshort\else
 \begin{code}
@@
 \end{code}
-\end{comment}
+\fi
 The result has no tags: the interpreter patently uses no tags and no
@@
 applies |lam varZ| without pattern matching. What is more, evaluating an
-%aplas-specific
-open term such as \texttt{app (lam (varS varZ)) (b true)}
-gives a type- rather than a run-time error.
-%aplas-else
-\begin{comment}
-\begin{code}
-open term such as |testf3| below gives a type error rather than
-a run-time error
+open term such as
+\ifshort
+\texttt{app (lam (varS varZ)) (b true)}
+\else
+|testf3| below
+\fi
+gives a type error rather than a run-time error.
+\ifshort\else
+\begin{code}
 let testf3 = app (lam (varS varZ)) (b true)
@@
 \end{code}
-\end{comment}
+\fi
 The type error correctly complains
-that the initial environment should be a tuple rather than |()|.
+that the initial environment should be a tuple rather than~|()|.
 In other words, the term is open.
@@
 metalanguage with a simple 
-%aplas
-Hindley-Milner type system
-%aplas \cite{hindley-principal,milner-theory}
+\ifshort
+Hindley-Milner \fi
+type system\ifshort\else\ \cite{hindley-principal,milner-theory}\fi
 .  In this \emph{final} rather
@@
 
-Our solution is emphatically distinct from Church-encoding of the
-universal type. Church-encoding of the type |u|, \S\ref{tagproblem},
-requires two continuations; the interpreter (see |app| above) would
-have to provide both to the encoding of |e1|. The continuation
-corresponding to the |UB| branch of |u| must either raise an error or
-loop. For a well-typed object term that error continuation is never
-invoked, yet it must be supplied. Our interpreter has no error
+Our solution is \emph{not} Church-encoding the
+universal type. The Church encoding of the type~|u| in \S\ref{tagproblem}
+requires two continuations; the function |app| in the interpreter above would
+have to provide both to the encoding of~|e1|. The continuation
+corresponding to the |UB| case of~|u| must either raise an error or
+loop. For a well-typed object term, that error continuation is never
+invoked, yet it must be supplied. In contrast, our interpreter has no error
 continuation at all.
@@
 The evaluator above is wired directly into the
-functions |b|, |lam|, |app|, and so on.  %In the rest of this paper, 
-We explain how to abstract the interpreter so as
+functions |b|, |lam|, |app|, and so on.  \ifshort We \else In the rest of this paper, we \fi
+explain how to abstract the interpreter so as
 to process the \emph{same} term in many other
@@
         including call-by-name (CBN) CPS, so to isolate the evaluation
-        order of the object language from that of the metalanguage
-%aplas
+        order of the object language from that of the metalanguage.\ifshort
 \footnote{Due to serious lack of space, 
-we refer the reader to the accompanying code for this.}%
-.
+we refer the reader to the accompanying code for this.}\fi
 \end{itemize*}
 We have programmed our interpreters in OCaml (and, for staging,
-\cite{metaocaml}) and standard Haskell. The complete code is available
-\url{\oleg{put URL!}.}
+MetaOCaml) and standard Haskell. The complete code is available at
+\url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
 to supplement the paper. Our examples below switch between (Meta)OCaml
@@
 
-We take the problem of tagless (staged) typed-preserving
-interpretation exactly as it was posed in \cite{WalidICFP02} and
-\cite{xi-guarded}.  We use their running examples and achieve the
+We attack the problem of tagless (staged) typed-preserving
+interpretation exactly as it was posed
+by \citet{WalidICFP02} and \citet{xi-guarded}.
+We use their running examples and achieve the
 result they call desirable.  Our contributions are as follows.
@@
     compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
-   in a typed metalanguage, in direct and continuation\hyp passing styles%
-% aplas
-%   (\S\ref{variations}).
-.
+   in a typed metalanguage, in direct and continuation\hyp passing
+   styles\ifshort\else\ (\S\ref{variations})\fi.
 \item All these interpreters use no type tags, patently never get stuck,
@@
   interpreters, from evaluation and compilation (\S\ref{language}) 
-   to partial evaluation%
-% aplas: commented
-%  and CPS transformation (\S\ref{PE}).
-.
+   to partial evaluation \ifshort\else and CPS transformation \fi(\S\ref{PE}).
 \item We point a clear way to extend the object language with more features
-    such as state%
-%aplas (\S\ref{state})%
-%aplas
-\footnote{again, see code}
-.
+    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
 \item We describe an approach to self\hyp interpretation compatible with the
-  above%
-%aplas: (\S\ref{selfinterp})
-.  Self\hyp interpretation turned out
-  harder than expected%
-%aplas
-\footnotemark[\value{footnote}]
-.
+  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned out
+  harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
 \end{enumerate*}
@@
 
-We reiterate that we specifically do not propose any new language
-feature or technique. We use the existing, known approaches --
-Hindley-Milner type system with inference-preserving module system or
-constructor classes extensions, as realized in ML or Haskell98 -- to
-solve the problem that was stated in published record as lacking
-solution in ML or Haskell98 and most likely unimplementable in these
-languages without extensions. The simplicity of our solution and our
-use of only mainstream functional language features makes typed
-embedded DSLs more practical to build.
+To reiterate, we do \emph{not} propose any new language
+feature or technique. We use features already present in
+mainstream functional languages---Hindley-Milner
+type system with either an inference\hyp preserving module system or
+constructor classes, as realized in ML and Haskell 98---to
+solve the problem that was stated in the published record as
+unsolved and likely unsolvable in ML or Haskell 98
+without extensions. The simplicity of our solution and our
+use of only mainstream features makes it more practical to build typed,
+embedded domain\hyp specific languages.
 \begin{comment}
@@
 \label{encoding}
-We embed our language in (Meta)OCaml and Haskell.  In Haskell,
-the functions that construct object terms are methods in a type class
-|Symantics| (with a parameter |repr| of kind |* -> *|)%
-%aplas-specific
-, Fig.~\ref{fig:symantics-haskell}(a)%
-. The class is so named
-because its interface gives the syntax of the object language and its
-instances give the semantics.
-\begin{figure}
-%
-(a) \begin{code2}
+
+\ifshort
+\begin{SaveVerbatim}{3a}
 class Symantics repr where
-  int  :: Int  -> repr Int       bool :: Bool -> repr Bool
+  int :: Int  -> repr Int;       bool :: Bool -> repr Bool
   lam :: (repr a -> repr b) -> repr (a -> b)
@@
   if_ :: repr Bool -> repr a -> repr a -> repr a
-\end{code2}
-
-(b) \begin{code2}
+\end{SaveVerbatim}
+\else
+\begin{SaveVerbatim}{3a}
+class Symantics repr where
+  int  :: Int  -> repr Int
+  bool :: Bool -> repr Bool
+
+  lam :: (repr a -> repr b) -> repr (a -> b)
+  app :: repr (a -> b) -> repr a -> repr b
+  fix :: (repr a -> repr a) -> repr a
+
+  add :: repr Int -> repr Int -> repr Int
+  mul :: repr Int -> repr Int -> repr Int
+  leq :: repr Int -> repr Int -> repr Bool
+  if_ :: repr Bool -> repr a -> repr a -> repr a
+\end{SaveVerbatim}
+\fi
+\begin{SaveVerbatim}{3b}
 testpowfix () = lam (\x -> fix (\self -> lam (\n ->
@@
                      (mul x (app self (add n (int (-1))))))))
-\end{code2}
-
-(c) \begin{code2}
+\end{SaveVerbatim}
+\begin{SaveVerbatim}{3c}
 testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
-\end{code2}
-\caption{Symantics in Haskell.}
-\label{fig:symantics-haskell}
-\end{figure}
-% fig:symantics-haskell part (a) was here
+\end{SaveVerbatim}
+
+We embed our language in (Meta)OCaml and Haskell.  In Haskell,
+the functions that construct object terms are methods in a type class
+|Symantics| (with a parameter |repr| of kind |* -> *|)\ifshort,
+Fig.~\ref{fig:symantics-haskell}(a)\fi. The class is so named
+because its interface gives the syntax of the object language and its
+instances give the semantics.
+\ifshort\else\UseVerbatim{3a}\fi
 For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
-\S\ref{tagproblem} above as |app (lam (\x -> x)) (bool True)|,
+\S\ref{tagproblem} above as \texttt{app (lam (\textbackslash x -> x)) (bool True)},
 whose inferred type is \texttt{Symantics repr => repr Bool}.
 For another example, the classical $\mathit{power}$ function is
-%aplas-specific
-in Fig.~\ref{fig:symantics-haskell}(b)
-% fig:symantics-haskell part (b) was here
+\ifshort in Fig.~\ref{fig:symantics-haskell}(b)
+\else\UseVerbatim{3b}\fi
 and the partial application $\fun{x} \mathit{power}\;x\;7$ is
-in Fig.~\ref{fig:symantics-haskell}(c).
-% fig:symantics-haskell part (c) was here
+\ifshort in Fig.~\ref{fig:symantics-haskell}(c).
+\else\UseVerbatim{3c}\fi
 The dummy argument |()| above is to avoid the monomorphism
@@
 accompanying code has the complete implementations.
+
+\ifshort
+\begin{figure}
+(a) \BUseVerbatim{3a}
+
+(b) \BUseVerbatim{3b}
+
+(c) \BUseVerbatim{3c}
+\caption{Symantics in Haskell}
+\label{fig:symantics-haskell}
+\end{figure}
+\fi
 
@@
   val app : ('c, 'da -> 'db) repr -> ('c, 'da) repr -> ('c, 'db) repr
-  val fix : (('c, 'da -> 'db) repr -> ('c, 'da -> 'db) repr) -> ('c, 'da -> 'db) repr
+  val fix : ('x -> 'x) -> (('c, 'da -> 'db) repr as 'x)
 
@@
   val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
-  val if_ : ('c, bool) repr-> (unit-> ('c, 'da) repr)-> (unit-> ('c, 'da) repr)-> ('c, 'da) repr
+  val if_ : ('c,bool) repr -> (unit -> ('c, 'da) repr)
+                           -> (unit -> ('c, 'da) repr) -> ('c, 'da) repr
 end
-
 module EX(S: Symantics) = struct open S
   let test1 () = app (lam (fun x -> x)) (bool true)
-  let testpowfix () =  lam (fun x -> fix (fun self -> lam (fun n ->
+  let testpowfix () =
+       lam (fun x -> fix (fun self -> lam (fun n ->
         if_ (leq n (int 0)) (fun () -> int 1)
@@
 \end{figure*}
-To embed the same object language in (Meta)OCaml, we replace the type
-class |Symantics| and its instances by a module signature |Symantics|
+To embed the same object language in (Meta)OCaml, we replace the \ifshort type
+class \fi |Symantics| \ifshort\else type class \fi and its instances by a module signature |Symantics|
 and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
 signature that suffices until~\S\ref{PE}.  The two differences are:
-the additional type parameter |c|, an
-environment classifier \cite{WalidPOPL03} (required by MetaOCaml for
-code generation in~\S\ref{compiler}); and the $\eta$-expanded type for
+the additional type parameter |'c|, an
+\emph{environment classifier} \cite{WalidPOPL03} required by MetaOCaml for
+code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
 |fix| and thunk types in |if_| since OCaml is a call-by-value
@@
 
-The functor |EX|, Fig.~\ref{fig:ocaml-simple} encodes 
+The functor |EX| in Fig.~\ref{fig:ocaml-simple} encodes 
 our running examples |test1| and $\mathit{power}$.
@@
 OCaml as a functor from |Symantics| to an appropriate semantic domain. This
-is essentially the same as the constraint |Symantics repr =>| in the
+is essentially the same as the constraint \texttt{Symantics repr =>} in the
 Haskell embedding.
@@
 % don't even have space for CPS
-\begin{comment}
+\ifshort\else
 \section{Continuation\hyp passing style}\label{variations}
@@
 refer the reader to the accompanying code.
-\end{comment}
+\fi
 
@@
 
-\begin{comment}
+\ifshort\else
 \section{Self-interpretation}\label{selfinterp}
@@
 
-%aplas \begin{comment}
 \jacques{But how 
@@
 hole below).}
-%aplas \end{comment}
-
-%aplas \begin{comment}
 
@@
 
-%aplas \end{comment}
-\end{comment}
+\fi
 
@@
 \cite{Pfenning-Lee} conclude that partial evaluation and program
-transformations "do not seem to be expressible" even using their
+transformations ``do not seem to be expressible'' even using their
 extension to $F_{\omega}$, perhaps because their avoidance of general

### 2007-07-19 tagless/tagless-final.tex
Done with reinflating the paper for JFP, except for related work

@@
 definitional interpreter is thus a promising way to build a domain\hyp specific
-language.
+language (DSL)\@.
 
@@
 \ifshort
-\begin{code3}[xleftmargin=0pt]
+\begin{code3}
 let varZ env    = fst env         let b (bv:bool) env = bv
-let varS vp env = vp (snd env)  let app e1 e2 env   = (e1 env) (e2 env)
-let lam e env   = fun x -> e (x,env)
+let varS vp env = vp (snd env)    let lam e env   = fun x -> e (x,env)
+let app e1 e2 env   = (e1 env) (e2 env)
 \end{code3}
@@
     \item transform the term to continuation\hyp passing style (CPS),
-        including call-by-name (CBN) CPS, so to isolate the evaluation
+        even call-by-name (CBN) CPS, so to isolate the evaluation
         order of the object language from that of the metalanguage.\ifshort
@@
 unsolved and likely unsolvable in ML or Haskell 98
-without extensions. The simplicity of our solution and our
+without extensions. The simplicity of our solution and its
 use of only mainstream features makes it more practical to build typed,
-embedded domain\hyp specific languages.
+embedded DSLs.
 \begin{comment}
@@
 \begin{floatrule}
-\begin{code}
+\begin{code2}
 module type Symantics = sig type ('c, 'dv) repr
@@
   val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
-  val if_ : ('c,bool) repr -> (unit -> ('c, 'da) repr)
-                           -> (unit -> ('c, 'da) repr) -> ('c, 'da) repr
+  val if_ : ('c, bool) repr
+            -> (unit -> 'x) -> (unit -> 'x) -> (('c, 'da) repr as 'x)
 end
@@
 end
-\end{code}
+\end{code2}
 \end{floatrule}
@@
 
-Thus we represent an object expression in
+Thus, we represent an object expression in
 OCaml as a functor from |Symantics| to an appropriate semantic domain. This
@@
 metalanguage.  The module below interprets each object\hyp language
-operation as the corresponding metalanguage operation
+operation as the corresponding metalanguage operation.
+\ifshort
 \vspace*{-0.8em}\begin{code3}
@@
 \vspace*{-0.8em}
+\else
+\begin{code}
+module R = struct
+  type ('c,'dv) repr = 'dv (* no wrappers *)
+
+  let int  (x:int)  = x
+  let bool (b:bool) = b
+  let lam  f        = f
+  let app  e1 e2    = e1 e2
+  let fix  f        = let rec self n = f self n in self
+  let add  e1 e2    = e1 + e2
+  let mul  e1 e2    = e1 * e2
+  let leq  x y      = x <= y
+  let if_  eb et ee = if eb then et () else ee ()
+end
+\end{code}
+\fi
 %
@@
 OCaml's addition, and |app| is OCaml's application. To run our
-examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|%
-%aplas-specific
-: \texttt{module EXR = EX(R)}%
-.
-%% \begin{code}
-%% module EXR = EX(R)
-%% \end{code}
-Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
-% aplas: commented out
-\begin{comment}
+examples, we instantiate the |EX| functor from~\S\ref{encoding}
+with~|R|\ifshort: \texttt{module EXR = EX(R)}\fi.
+\ifshort\else
+\begin{code}
+module EXR = EX(R)
+\end{code}
+\fi
+Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
+\ifshort It \else
 In Haskell, we define
@@
 cannot ever fail and is assuredly compiled away.
-In OCaml, too, it is obvious to the compiler that
-\end{comment}
-%aplas: the following line is for aplas only
-It is obvious to the compiler that
-%
+In OCaml, too, it
+\fi is obvious to the compiler that
 pattern matching cannot fail, because there is no
@@
 due to interpreting |fix|.
-%aplas-specific
+\ifshort
 (The source code shows a total interpreter |L| that measures
-the size of each object term).
-%aplas-specific (cont)
-We can also generalize from~|R| to all interpreters; the proof
-of these propositions follow immediately from the soundness of the
+the size of each object term.)
+\fi
+\ifshort
+We can also generalize from~|R| to all interpreters;
+these propositions follow immediately from the soundness of the
 metalanguage's type system.
-%end aplas-specific
+\fi
 \begin{proposition}
@@
 \end{proposition}
-% Generalizing from~|R| to all interpreters, we have the following
-% broader and more useful, if also more obvious, proposition.
+\ifshort\else
+Generalizing from~|R| to all interpreters, we have the following
+broader and more useful, if also more obvious, proposition.
+\fi
 \begin{proposition}
@@
 \end{proposition}
-% These propositions follow immediately from the soundness of the
-% metalanguage's type system.
-% aplas-commented
-\begin{comment}
+\ifshort\else
+These propositions follow immediately from the soundness of the
+metalanguage's type system.
+\fi
+\ifshort\else
 For variety, we show another interpreter, which measures the \emph{size}
@@
 \end{code}
-Now the expression
-\begin{code}
-let module E = EX(L) in E.test1 ()
-\end{code}
+Now the OCaml expression
+|let module E = EX(L) in E.test1 ()|
 evaluates to |3|. This interpreter is not only tagless but also
 total. It ``evaluates'' even seemingly divergent terms like
-\begin{code}
-app (fix (fun self -> self)) (int 1)
-\end{code}
-
-%aplas \begin{comment}
+|app (fix (fun self -> self)) (int 1)|.
+\fi
+
+\begin{comment}
+\begin{code}
 module EX1(S: Symantics) = struct
@@
 let module E =EX1(L) in E.tfix ();;
-%%aplas \end{comment}
+\end{code}
 \end{comment}
@@
 \section{A tagless compiler (or, a staged interpreter)}\label{compiler}
-\vspace{-5pt}
 Besides immediate evaluation, we can compile our object language
@@
 
-Inserting brackets and escapes appropriately into the
-evaluator~|R| above yields the simple compiler~|C| below%
-%aplas-specific
-, Fig.~\ref{fig:interpreter-C}(a)
-.
-% fig:interpreter-C part (a) was here
-%
-\begin{figure}[t]
 % this code does not have the 'sv parameter. It shows up later.
-(a) \begin{code2}
+\ifshort
+\begin{SaveVerbatim}{5a}
 module C = struct type ('c,'dv) repr = ('c,'dv) code
@@
   let if_ eb et ee = .<if .~eb then .~(et ()) else .~(ee ())>. end
-\end{code2}
-
-(b) \begin{code2}
+\end{SaveVerbatim}
+\else
+\begin{SaveVerbatim}{5a}
+module C = struct
+  type ('c,'dv) repr = ('c,'dv) code
+
+  let int (x:int)   = .<x>.
+  let bool (b:bool) = .<b>.
+  let lam f         = .<fun x -> .~(f .<x>.)>.
+  let app e1 e2     = .<.~e1 .~e2>.
+  let fix f         = .<let rec self n = .~(f .<self>.) n in self>.
+  let add e1 e2     = .<.~e1 + .~e2>.
+  let mul e1 e2     = .<.~e1 * .~e2>.
+  let leq x y       = .<.~x <= .~y>.
+  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>.
+end
+\end{SaveVerbatim}
+\fi
+\begin{SaveVerbatim}{5b}
 let module E = EX(C) in E.test1 ()
-\end{code2}
-
-(c) \begin{code2}
+\end{SaveVerbatim}
+\begin{SaveVerbatim}{5c}
 let module E = EX(C) in E.testpowfix7
-\end{code2}
-
-(d) \begin{code2}
+\end{SaveVerbatim}
+\begin{SaveVerbatim}{5d}
 .<fun x_1 -> (fun x_2 -> let rec self_3 = fun n_4 ->
-   (fun x_5 -> if x_5 <= 0 then 1  else x_2 * (self_3 (x_5 + (-1))))
+   (fun x_5 -> if x_5 <= 0 then 1 else x_2 * self_3 (x_5 + (-1)))
    n_4 in self_3) x_1 7>.
-\end{code2}
+\end{SaveVerbatim}
+
+Inserting brackets and escapes appropriately into the
+evaluator~|R| above yields the simple compiler~|C|
+\ifshort in Fig.~\ref{fig:interpreter-C}(a).\else below.
+\UseVerbatim{5a}
+\fi
+\ifshort
+\begin{figure}[t]
+(a) \BUseVerbatim{5a}
+
+(b) \BUseVerbatim{5b}
+
+(c) \BUseVerbatim{5c}
+
+(d) \BUseVerbatim{5d}
 \caption{The tagless staged interpreter \texttt{C}.}
@@
 \end{figure}
-%
+\fi
 This is a straightforward staging of
@@
 This compiler produces
-unoptimized code. For example, interpreting our |test1|%
-%aplas-specific
-, Fig.~\ref{fig:interpreter-C}(b),
-% fig:interpreter-C part (b) was here
-gives the code value |.<(fun x_6 -> x_6) true>.|
-of inferred type |('c, bool) C.repr|.  Interpreting |testpowfix7|
+unoptimized code. For example, interpreting our |test1| with
+\ifshort Fig.~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
+gives the code value |.<(fun|~|x_6|~|->|~|x_6)| |true>.|
+of inferred type |('c,|~|bool)| |C.repr|.  Interpreting |testpowfix7|
+\ifshort\pagebreak[3]\fi
 with
-%aplas-specific
-Fig.~\ref{fig:interpreter-C}(c),
-% fig:interpreter-C part (c) was here
-gives a code value with many apparent $\beta$-redexes
-%aplas-specific
-Fig.~\ref{fig:interpreter-C}(d).
-% fig:interpreter-C part (d) was here
+\ifshort Fig.~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
+gives a code value with many apparent $\beta$- and $\eta$-redexes\ifshort,
+Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
 This compiler does not incur
 any interpretive overhead: the
-code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| 
-% aplas: and does not
-% call the interpreter, unlike the recursive calls to |eval0| and
-% |eval| in the |L e| lines in \S\ref{tagproblem}.
+code produced for $\fun{x}x$ is simply |fun|~|x_6|~|->|~|x_6|\ifshort\else\
+and does not
+call the interpreter, unlike the recursive calls to |eval0| and
+|eval| in the |L e| lines in \S\ref{tagproblem}\fi.
 The resulting code obviously contains no tags and no pattern matching.
-%aplas: the line is for aplas. The original paragraph commented out
+\ifshort
 The accompanying code shows the Haskell implementation. 
-\begin{comment}
+\else
+
 We have also implemented this compiler in Haskell. 
@@
 has no (convenient, typed) staging facility, we had to emulate
-it. To be precise, we defined a data type |ByteCode| with
-constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|%
-\footnote{\texttt{ByteCode} can be mapped to AST of Template Haskell
-  (TH). The output of our compiler will then be assuredly type-correct TH.}
+it: we defined a data type |ByteCode| with
+constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.%
+\footnote{\texttt{ByteCode} can be mapped to the abstract syntax of Template
+    Haskell. The output of our compiler will then be assuredly type-correct
+    Template Haskell.}
 Whereas our representation of object terms uses HOAS,
@@
 accompanying code gives the full details.
-\end{comment}
+\fi
 
@@
 Roughly, a partial evaluator interprets each object term to either
-a static (present-stage) term (using~|R|) or
-a dynamic (future-stage) term (using~|C|).  To
+a static (present-stage) term (using\ifshort\else\ the evaluator\fi~|R|) or
+a dynamic (future-stage) term (using\ifshort\else\ the compiler\fi~|C|).  To
 distinguish between static and dynamic terms, one might define
-%aplas-specific
-\texttt{data P0 t = S0 (R t) $\mid$ E0 (C t)}.
+\ifshort
+\texttt{data P0~t~= S0~(R~t)~}\Verb+|+\texttt{ E0~(C~t)}.
+\else
+\begin{code}
+data P0 t = S0 (R t) | E0 (C t)
+\end{code}
+To extract a dynamic term from this type, we create the functions
+\begin{code}
+abstrI :: P0 Int -> C Int
+abstrI (S0 r) = int (unR r)
+abstrI (E0 c) = c
+
+abstrB :: P0 Bool -> C Bool
+abstrB (S0 r) = bool (unR r)
+abstrB (E0 c) = c
+\end{code}
+\fi
 We then start to define |P0| as an instance of the |Symantics| class.
@@
 add them using~|C|.
+\ifshort\else
+\begin{code}
+instance Symantics P0 where
+  int  x = S0 (int x)
+  bool x = S0 (bool x)
+  add (S0 e1) (S0 e2) = S0 (add e1 e2)
+  add e1 e2 = E0 (add (abstrI e1) (abstrI e2))
+\end{code}
+Thus, the two uses of |add| above refer to
+|add| for~|R| and |add| for~|C|.
+
+\fi
 Whereas |mul| and |leq| are as easy to define as |add|, we encounter
@@
 function, of type |a -> C a|, to send a value to the future stage
-\cite{xi-guarded,WalidPOPL03}.
-%aplas-else
-\begin{comment}
-\begin{code}
-data P0 t = S0 (R t) | E0 (C t)
-\end{code}
-To extract a dynamic term from this type, we create the functions
-\begin{code}
-abstrI :: P0 Int -> C Int
-abstrI (S0 r) = int (unR r)
-abstrI (E0 c) = c
-
-abstrB :: P0 Bool -> C Bool
-abstrB (S0 r) = bool (unR r)
-abstrB (E0 c) = c
-\end{code}
-We then start to define |P0| as an instance of the |Symantics| class.
-\begin{code}
-instance Symantics P0 where
-  int  x = S0 (int x)
-  bool x = S0 (bool x)
-  add (S0 e1) (S0 e2) = S0 (add e1 e2)
-  add e1 e2 = E0 (add (abstrI e1) (abstrI e2))
-\end{code}
-Integer and boolean literals are immediate, present-stage
-values. Addition yields a static term (using~|R|) if and only if both operands
-are static; otherwise we extract the dynamic terms from the operands and
-add them using~|C|.  Thus the two uses of |add| above refer to
-|add| for~|R| and |add| for~|C|.
-
-Whereas |mul| and |leq| are as easy to define as |add|, we encounter
-a problem with |if_|.  If the first argument to |if_| is a dynamic term
-(type |C Bool|) and the second and third arguments are a static term
-(type |R a|) and a dynamic term (type |C a|), then we need to convert
-the static term to dynamic, but there is no polymorphic ``lift''
-function, of type |a -> C a|, to send a value to the future stage
-\cite{xi-guarded,WalidPOPL03}.\footnote{We note in passing that, if we
+\cite{xi-guarded,WalidPOPL03}.\ifshort\else
+\footnote{If we
 were to add polymorphic \texttt{lift} to the type class
@@
 \texttt{Applicative} and thus \texttt{Functor}:\quad\texttt{fmap
-f = app (lift f)}}
-\end{comment}
+f = app (lift f)}}\fi
 
 Our |Symantics| class only includes separate lifting methods |bool| and
-|int|, not a parametrically polymorphic lifting method:
+|int|, not a parametrically polymorphic lifting method, for good reason:
 When compiling to a first-order target language such as machine code,
 booleans, integers, and functions may well be represented differently.
-Thus polymorphic lift cannot be compiled without intensional type
-analysis.  To avoid this, we turn to
-\citearound{'s technique}\citet[see also \citealp{sumii-hybrid}]{asai-binding-time}:
+Thus, compiling polymorphic lift requires intensional type
+analysis.  To avoid needing polymorphic lift, we turn to
+Asai's technique \cite{asai-binding-time,sumii-hybrid}:
 build a dynamic term
@@
 
+\begin{SaveVerbatim}{P1}
+data P1 t = P1 (Maybe (R t)) (C t)
+\end{SaveVerbatim}
+
 We switch to the Haskell data type
-%aplas-specific
-|data P1 t = P1 (Maybe (R t)) (C t)|
-%aplas-else
-\begin{comment}
-\begin{code}
-data P1 t = P1 (Maybe (R t)) (C t)
-\end{code}
-\end{comment}
+\ifshort \BUseVerbatim{P1} \else \UseVerbatim{P1} \fi
 so that a partially evaluated term |P1 t| always contains a dynamic
@@
 alternative constructors of a |Maybe| value, |Just| and |Nothing|,
-tag each term with a phase: present or
+tag each partially evaluated term with a phase: either present or
 future.  This tag is not an object type tag: all pattern matching below
@@
 can now define the polymorphic function
-%aplas-specific
+\ifshort
 |abstr1 (P1 _ dyn) = dyn| of type
 |abstr1 :: P1 t -> C t|
-%aplas-else
-\begin{comment}
+\else
 \begin{code}
@@
 \end{code}
-\end{comment}
+\fi
 to extract it without requiring polymorphic lift into~|C|.  We then try
@@
 constructs of our object language, including |if_|.
+\ifshort
 \vspace*{-0.7em}
@@
 \vspace*{-0.7em}
+\else
+\begin{code}
+instance Symantics P1 where
+  int  x = P1 (Just (int  x)) (int  x)
+  bool b = P1 (Just (bool b)) (bool b)
+
+  add (P1 (Just n1) _) (P1 (Just n2) _) = int (unR (add n1 n2))
+  add e1 e2 = P1 Nothing (add (abstr1 e1) (abstr1 e2))
+  -- mul and leq are analogous and elided
+
+  if_ (P1 (Just s) _) et ef = if unR s then et else ef
+  if_ eb et ef = P1 Nothing
+                    (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
+\end{code}
+\fi
 When we come to functions, however, we stumble.  According to our
@@
 
-To enable these simplifications, we delay binding-time analysis (BTA)
+To enable these simplifications, we delay binding-time analysis
 for a static function until it is applied, that is, until |lam f|
@@
 |P1| to a function type |a -> b| should yield one of
-\vspace*{-0.7em}
-\begin{code3}
+\ifshort\vspace*{-0.7em}\fi
+\begin{code}
 data P1 (a -> b) = S1 (P1 a -> P1 b) | E1 (C (a -> b))
 data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b)) (C (a -> b))
-\end{code3}
-\vspace*{-0.7em}
+\end{code}
+\ifshort\vspace*{-0.7em}\fi
 That is, we need a nonparametric data type, something akin to
@@
 \citet{oliveira-typecase} dub the \emph{typecase} design pattern.
-%aplas-simplification
-Thus typed partial evaluation, like typed CPS transformation,
+Thus, typed partial evaluation, like typed CPS transformation,
 inductively defines a map from source types to target types that
 performs case distinction on the source type.
-\begin{comment}
-
-Thus typed partial evaluation, like typed CPS transformation,
-inductively defines a map from source types to target types that
-performs case distinction on the source type.  The connection with CPS
-is not an accident, as we shall see in~\S\ref{S:CPS}.
-\end{comment}
+\ifshort\else The connection with CPS
+is not an accident, as we shall see in~\S\ref{S:CPS}.\fi
 
@@
 in the accompanying source code shows the latter.
-%aplas: the original text is in comments below. Here's a greatly
-% abbreaviated text
+\ifshort
 \vspace*{-0.7em}
@@
 \vspace*{-0.7em}
-The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
-in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
-the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
-operand of type |P a -> P b| rather than |a -> b|. As before, we define a
-function |abstr| to extract a future-stage computation from a value of 
-type |P t|. The implementations of |int|, |bool|, and |add| are like
-in~\S\ref{S:PE-problem}.
-The interpretation of |lam f = VF f| just wraps the HOAS function |f|. 
-We can always compile |f| to a code value,
-but we delay it to apply |f| to concrete arguments. The interpretation of
-\vspace*{-0.7em}
-\begin{code3}
-app (VF f) ea = f ea;  app (E f)  ea = E (app f (abstr ea))
-\end{code3}
-\vspace*{-0.7em}
-checks to see if |ef| is such a delayed
-HOAS function |VF f|. If it is, we apply |f| to the
-concrete argument |ea|, giving us a chance to perform static
-computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
-dynamic value |E f|, we residualize.
-This solution using GADTs works but is not quite satisfactory. First, it
-cannot be ported to MetaOCaml.  Second,
-the problem of nonexhaustive pattern\hyp matching reappears in
-|app|. Although Coq or logical
-frameworks with canonical forms could prove pattern-matching
-exhaustive, GHC cannot and issues warnings. Recall, our goal is to
-make this exhaustiveness \emph{syntactically} apparent.
-
-\begin{comment}
+\else
 We introduce a GADT with four data constructors.
-\vspace*{-0.7em}
+\begin{code}
 data P t where
@@
   E  :: C t -> P t
-\end{code3}
+\end{code}
+\fi
 The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
@@
 the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
-operand of type |P a -> P b| rather than |a -> b|. As before, we define a
-function to extract a future-stage computation from a value of type |P t|.
+operand of type |P a -> P b| rather than |a -> b|. We define a function
+\ifshort |abstr|, \fi like |abstr1| above\ifshort,\fi\
+to extract a future-stage computation from a value of type |P t|.
+\ifshort\else
 \begin{code}
@@
 \end{code}
+\fi
 The implementations of |int|, |bool|, and |add| are like
 in~\S\ref{S:PE-problem}.
-The interpretation of |lam f| just wraps the
-HOAS function |f|. We can always compile |f| to a code value,
+The interpretation of |lam f| is |VF f|, which just wraps the HOAS function |f|. 
+We can always compile |f| to a code value,
 but we delay it to apply |f| to concrete arguments. The interpretation of
-|app ef ea| checks to see if |ef| is such a delayed
+|app ef ea|
+\ifshort
+defined by
+\vspace*{-0.7em}
+\begin{code3}
+app (VF f) ea = f ea;  app (E f) ea = E (app f (abstr ea))
+\end{code3}
+\vspace*{-0.7em}
+\fi
+checks to see if |ef| is such a delayed
 HOAS function |VF f|. If it is, we apply |f| to the
@@
 dynamic value |E f|, we residualize.
-
+\ifshort\else
+
+\fi
 This solution using GADTs works but is not quite satisfactory. First, it
-cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
-the problem of nonexhaustive pattern\hyp matching reappears in |app|
-above: the type |P t| has four constructors, of which pattern\hyp
-matching in |app| uses only |VF| and~|E|. One may say that the
+cannot be ported to MetaOCaml\ifshort\else\ as GADTs are unavailable there\fi.  Second,
+the problem of nonexhaustive pattern\hyp matching reappears in
+|app|\ifshort. \else\
+above: the type |P t| has four constructors, of which the pattern in
+|app| matches only |VF| and~|E|. One may say that the
 constructors |VI| and |VB| obviously cannot occur because they do not
 construct values of type |P (a -> b)| as required by the type of |app|.
-Indeed the metalanguage implementation could reason thus, 
-but it may not; GHC for one issues warnings%
+\fi
+\ifshort
+Although Coq or logical
+frameworks with canonical forms could prove pattern\hyp matching
+exhaustive,
+\else
+Indeed the metalanguage implementation could reason thus,%
 \footnote{If we use inductive families (as those in Coq) or logical
-  frameworks with canonical forms, we can prove the pattern-matching
-  to be exhaustive. The coverage checker of Twelf does such proof.}. 
+  frameworks with canonical forms (as in Twelf with its coverage checker),
+  we can prove the pattern matching to be exhaustive.}
+but it may not;
+\fi
+GHC cannot and issues warnings.
+\ifshort
+Recall, our
+\else
 Although this point may seem minor, it is the heart of
@@
 interpreter contains many pattern\hyp matching forms that look partial
-but never fail in reality. The goal is to make this safety
-\emph{syntactically} apparent.
-\end{comment}
+but never fail in reality. The
+\fi
+goal is to make this exhaustiveness \emph{syntactically} apparent.
 
@@
 \label{S:PE-solution}
+
+\begin{SaveVerbatim}{P}
+data P t = P (Maybe (repr_pe t)) (C t)
+\end{SaveVerbatim}
+
 Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
 would ideally like is to write
-%aplas: inlining
-\texttt{data P t = P (Maybe (repr\_pe t)) (C t)}
-\begin{comment}
-\begin{code}
-data P t = P (Maybe (repr_pe t)) (C t)
-\end{code}
-\end{comment}
+\ifshort \BUseVerbatim{P} \else \UseVerbatim{P} \fi
 where |repr_pe| is the type function defined
 % inductively because P below depends on repr_pe
-by\vspace*{-0.8em}
+by
+\ifshort\vspace*{-0.8em}
 \begin{code3}
-repr_pe Int      = Int          repr_pe Bool     = Bool
-repr_pe (a -> b) = P a -> P b
+repr_pe Int = Int; repr_pe Bool = Bool; repr_pe (a -> b) = P a -> P b
 \end{code3}
 \vspace*{-0.8em}
+\else
+\begin{code}
+repr_pe Int      = Int
+repr_pe Bool     = Bool
+repr_pe (a -> b) = P a -> P b
+\end{code}
+\fi
 Although we can use type classes to define this type function
@@
 A simple and portable solution thus emerges: we bake |repr_pe| 
-into the signature |Symantics|. Switching to MetaOCaml to demonstrate this,
+into the signature |Symantics|. Switching to MetaOCaml to demonstrate
+this\ifshort\else\ portability\fi,
 we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
@@
 \begin{floatrule}
-\begin{code}
+\ifshort
+\begin{code2}
 module type Symantics = sig type ('c,'sv,'dv) repr
   val int : int -> ('c,int,int) repr
-  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
-  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-  val fix : ('x->'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
+  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+            -> ('c,'x,'da -> 'db) repr
+  val app : ('c,'x,'da -> 'db) repr
+            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
+                                'da -> 'db) repr as 'x)
   val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val if_ : ('c,bool,bool) repr -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x)
-end
-\end{code}
-\end{floatrule}
-\caption{A (Meta)OCaml embedding of our object language that supports
-  partial evaluation (\texttt{bool}, \texttt{mul},
-\texttt{leq} are elided)}
-\label{fig:ocaml}
-\end{figure*}
-
-\begin{comment}
-\begin{figure*}
-\begin{floatrule}
-\begin{code}
+  val if_ : ('c,bool,bool) repr
+            -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x) end
+\end{code2}
+\else
+\begin{code2}
 module type Symantics = sig
@@
 
-  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x) -> ('c,'x,'da->'db) repr
-  val app : ('c,'x,'da->'db) repr -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr, 'da->'db) repr as 'x)
+  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+            -> ('c,'x,'da -> 'db) repr
+  val app : ('c,'x,'da -> 'db) repr
+            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
+                                'da -> 'db) repr as 'x)
 
@@
   val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
-  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
+  val if_ : ('c,bool,bool) repr
+            -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
 end
-\end{code}
+\end{code2}
+\fi
 \end{floatrule}
-\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
+\caption{A (Meta)OCaml embedding of our object language that supports
+  partial evaluation \ifshort \protect\linebreak[2] (\texttt{bool},
+  \texttt{mul}, \texttt{leq} are elided)\fi}
 \label{fig:ocaml}
 \end{figure*}
-\end{comment}
-
-% aplas: abberviated figure here. The original is in teh comments
-% right down below
+
 \begin{figure}
 \begin{floatrule}
-\begin{code}
+\ifshort
+\begin{code2}
 module P = struct
@@
                   | _ -> pdyn (C.app (abstr ef) (abstr ea)) end
-\end{code}
-\end{floatrule}
-\caption{Our partial evaluator (\texttt{bool}, \texttt{mul},
-  \texttt{leq} and \texttt{fix} are elided)}
-\label{fig:pe}
-\end{figure}
-
-\begin{comment}
-\begin{figure}
-\begin{floatrule}
-\begin{code}
+\end{code2}
+\else
+\begin{code2}[commandchars=\@\[\]]
 module P = struct
@@
 
-  let int  (x:int)  = {st = Some (R.int x);  dy = C.int x}
+  let int  (x:int ) = {st = Some (R.int  x); dy = C.int  x}
   let bool (x:bool) = {st = Some (R.bool x); dy = C.bool x}
@@
                      | {st = Some b} -> if b then et () else ee ()
-  | _ -> pdyn (C.if_ (abstr eb) (fun () -> abstr (et ()))
-                                (fun () -> abstr (ee ())))
-  let lam f = {st = Some f; 
-               dy = C.lam (fun x -> abstr (f (pdyn x)))}
-
+                     | _ -> pdyn (C.if_ (abstr eb) (fun () @!->@! abstr (et @!()))
+                                                   (fun () @!->@! abstr (ee @!())))
+
+  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
   let app ef ea = match ef with
@@
 
-  let fix f = 
-    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
+  let fix f = let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
               in let rec self = function
-       | {st = Some _} as e -> app (f (lam self)) e
+                                | {st = Some @!_} as e -> app (f (lam self)) e
                                 | e -> pdyn (C.app fdyn (abstr e))
@@
 end
-\end{code}
+\end{code2}
+\fi
 \end{floatrule}
-\caption{Our partial evaluator (\texttt{mul} and \texttt{leq} are elided)}
+\caption{Our partial evaluator (\ifshort \texttt{bool}, \texttt{mul},
+  \texttt{leq} and \texttt{fix} \else \texttt{mul} and \texttt{leq} \fi
+  are elided)}
 \label{fig:pe}
 \end{figure}
-\end{comment}
 
@@
 type arguments |'c| and~|'dv|, which are treated by the new signature
-in the same way.  Hence all that needs to change in these interpreters
+in the same way.  Hence, all that needs to change in these interpreters
 to match the new signature is to add a phantom type
 argument~|'sv| to~|repr|.
-For example, the compiler |C| now begins \texttt{module C = struct
-  type ('c,'sv,'dv) repr = ('c,'dv) code} with the rest the same.
-In contrast, the partial evaluator~|P| relies on the type argument |'sv|.
-
-%aplas: here's the original text
-\begin{comment}
 For example, the compiler |C| now begins
+\ifshort
+\texttt{module C = struct
+  type ('c,'sv,'dv) repr = ('c,'dv) code}
+\else
 \begin{code}
@@
   type ('c,'sv,'dv) repr = ('c,'dv) code
-  (* the rest is the same *)
-\end{code}
+\end{code}
+\fi
+with the rest the same.
 In contrast, the partial evaluator~|P| relies on the type argument |'sv|.
-\end{comment}
 
@@
 multiplication by~$0$ or~$1$.  
-\begin{comment}
+\ifshort
+Such optimizations can be quite effective
+\else
 Such algebraic simplifications are easy
 to abstract over the specific domain (such as monoid or ring) where they
-apply.  These simplifications and abstractions help a lot in a large
-language with more base types and primitive operations.
-\end{comment}
-%aplas-specific
-In large languages, this can be quite effective.
-
-%aplas: abbreviated paragraph. The original is commented below
-Any partial evaluator must decide how much to unfold recursion. The code in
-Figure~\ref{fig:pe} na\"\i vely unfolds |fix| whenever the
+apply.  These simplifications and abstractions help a lot
+\fi
+in a large language with more base types and primitive operations.
+
+Any partial evaluator must decide how much to unfold recursion\ifshort.
+The code in Figure~\ref{fig:pe} na\"\i vely unfolds |fix|
+\else\
+statically: unfolding too little can degrade the residual code, whereas
+unfolding too much risks nontermination.  Our partial evaluator is no
+exception, because our object language includes |fix|.  The code in
+Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
+way'', that is, unfold |fix| rather than residualize
+\fi
+whenever the
 argument is static.  In the accompanying source code is a conservative
@@
 \cite{jones-partial}, but this issue is orthogonal to our presentation.
-\begin{comment}
-Any partial evaluator must decide how much to unfold recursion
-statically: unfolding too little can degrade the residual code, whereas
-unfolding too much risks nontermination.  Our partial evaluator is no
-exception, because our object language includes |fix|.  The code in
-Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
-way'', that is, unfold |fix| rather than residualize whenever the
-argument is static.  In the accompanying source code is a conservative
-alternative |P.fix| that unrolls recursion only once, then residualizes.
-Many sophisticated approaches have been developed to make this trade-off
-\cite{jones-partial}, but this issue is orthogonal to our presentation.
+\ifshort\else
 A separate concern in our treatment of |fix| is possible code bloat in
@@
 \cite{SwadiTahaKiselyovPasalic2006}.
-\end{comment}
-
-%aplas: abbreviated paragraph. The original is in the comments below
+\fi
+
 Given this implementation of~|P|, our running example
-\texttt{let module E = EX(P) in E.test1 ()} evaluates to
-\texttt{\{P.st = Some true; P.dy = .<true>.\}}
+\ifshort
+\texttt{let module E = EX(P) in E.test1 ()}
+\else
+\begin{code}
+let module E = EX(P) in E.test1 ()
+\end{code}
+\fi
+evaluates to
+\ifshort
+|{P.st = Some true; P.dy = .<true>.}|
+\else
+\begin{code}
+{P.st = Some true; P.dy = .<true>.}
+\end{code}
+\fi
 of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
@@
 $\beta$-redexes in~\S\ref{compiler}, the partial evaluation
+\ifshort
 \texttt{let module E = EX(P) in E.testpowfix7}
+\else
+\begin{code}
+let module E = EX(P) in E.testpowfix7
+\end{code}
+\fi
 gives the desired result
-\vspace*{-0.7em}
-\begin{code3}
+\ifshort\vspace*{-0.7em}\fi
+\begin{code}
 {P.st = Some <fun>;
- P.dy = .<fun x_1 -> x_1 * (x_1 * (x_1 * (x_1 * (x_1 * (x_1 * x_1)))))>.}
-\end{code3}
-\vspace*{-0.7em}
-
-\begin{comment}
-Given this implementation of~|P|, our running example
-\begin{code}
-let module E = EX(P) in E.test1 ()
-\end{code}
-evaluates to
-\begin{code}
-{P.st = Some true; P.dy = .<true>.}
-\end{code}
-of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
-a $\beta$-reduction has been statically performed to yield |true|.  More
-interestingly, whereas |testpowfix7| compiles to a code value with many
-$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
-\begin{code}
-let module E = EX(P) in E.testpowfix7
-\end{code}
-gives the desired result
-\begin{code}
-{P.st = Some <fun>;
- P.dy = .<fun x_1 -> x_1 * (x_1 * (x_1 * (x_1 *
-                    (x_1 * (x_1 * x_1)))))>.}
-\end{code}
-\end{comment}
+ P.dy = .<fun x -> x * (x * (x * (x * (x * (x * x)))))>.}
+\end{code}
+\ifshort\vspace*{-0.7em}\fi
 
@@
 phases (with |Some| and |None|) but not object types.
-%aplas: commented out
-\begin{comment}
+
+\ifshort\else
 Our typed partial evaluator is online and polyvariant.  It reuses the
@@
 evaluator and a compiler, but the general ideas are similar.
-\end{comment}
+\fi
 
@@
 \section{Continuation\hyp passing style}\label{variations}
-% aplas-specific
-\label{state}
-\label{S:CPS}
 
@@
 transformation.
-% Here's the text to replace the comment block below.
+\ifshort\label{state}\label{S:CPS}%
 This lets us decouple the evaluation strategy of the object language
@@
 refer the reader to the accompanying code.
-\fi
-
-\begin{comment}
+
+\else
+
 \subsection{Call-by-name CPS interpreters}\label{S:CPS}
@@
 Haskell.  To represent a CBN object language in a CBV metalanguage,
-\cite{reynolds-definitional,reynolds-relation} and \cite{PlotkinCBN}
+\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}
 introduce CPS to make the evaluation strategy of a definitional
@@
 let int (x:int) = fun k -> k x
-let add e1 e2 = fun k ->
-  e1 (fun v1 -> e2 (fun v2 -> k (v1 + v2)))
+let add e1 e2 = fun k -> e1 (fun v1 -> e2 (fun v2 -> k (v1 + v2)))
 \end{code}
@@
 \begin{code}
-((((int -> 'w1) -> 'w1) -> (int -> 'w1) -> 'w1)
- -> 'w2) -> 'w2
+((((int -> 'w1) -> 'w1) -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2
 \end{code}
@@
 module RCN = struct
-  type ('c,'sv,'dv) repr = 
-    {ko: 'w. ('sv -> 'w) -> 'w}
+  type ('c,'sv,'dv) repr = {ko: 'w. ('sv -> 'w) -> 'w}
   let int (x:int) = {ko = fun k -> k x}
-  let add e1 e2 = 
-    {ko = fun k -> e1.ko (fun v1 -> 
-                   e2.ko (fun v2 -> k (v1+v2)))}
-  let if_ eb et ee = 
-    {ko = fun k -> eb.ko 
-         (fun vb -> if vb then (et ()).ko k 
-                          else (ee ()).ko k)}
+  let add e1 e2 = {ko = fun k ->
+      e1.ko (fun v1 -> e2.ko (fun v2 -> k (v1 + v2)))}
+  let if_ eb et ee = {ko = fun k ->
+      eb.ko (fun vb -> if vb then (et ()).ko k else (ee ()).ko k)}
   let lam f = {ko = fun k -> k f}
-  let app e1 e2 = 
-    {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
-  let fix f = 
-    let rec fx f n = app (f (lam (fx f))) n in 
-    lam (fx f)
+  let app e1 e2 = {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
+  let fix f = let rec fx f n = app (f (lam (fx f))) n in lam (fx f)
   let run x = x.ko (fun v -> v)
@@
 module RCV = struct include RCN
-  let lam f = {ko = fun k -> k (fun e ->
-    e.ko (fun v -> f {ko = fun k -> k v}))}
+  let lam f = {ko = fun k -> k
+      (fun e -> e.ko (fun v -> f {ko = fun k -> k v}))}
 end
@@
 evaluates its argument before proceeding. This approach is in
-line with \cite{reynolds-relation}, albeit typed. The
+line with \citets{reynolds-relation}, albeit typed. The
 interpreter~|RCV| is useful for CBV evaluation of the object language
@@
   let int i = S.lam (fun k -> S.app k (S.int i))
-  let add e1 e2 = S.lam (fun k ->
-    S.app e1 (S.lam (fun v1 ->
+  let add e1 e2 = S.lam (fun k -> S.app e1 (S.lam (fun v1 ->
                                   S.app e2 (S.lam (fun v2 ->
                                   S.app k (S.add v1 v2))))))
-  let lam f = S.lam (fun k -> S.app k (S.lam
-    (fun x -> f (S.lam (fun k -> S.app k x)))))
-  let app e1 e2 = S.lam (fun k -> 
-    S.app e1 (S.lam (fun f ->
+  let lam f = S.lam (fun k -> S.app k
+              (S.lam (fun x -> f (S.lam (fun k -> S.app k x)))))
+  let app e1 e2 = S.lam (fun k -> S.app e1 (S.lam (fun f ->
                                   S.app e2 (S.lam (fun v ->
                                   S.app (S.app f v) k)))))
-  let fix f = S.fix (fun self -> (f self))
+  let fix = S.fix
 end
@@
 |('c, ..., (int -> 'w) -> 'w) S.repr|,
-the type of |lam (add (int 1))| is
-\begin{code}
-('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2)
-          -> 'w2) S.repr
+the type of \texttt{lam (add (int~1))} is
+\begin{code}
+('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr
 \end{code}
@@
   let testpowfix7 = (* same as before *)
-    lam (fun x -> app (app (testpowfix ()) x)
-                      (int 7))
+       lam (fun x -> app (app (testpowfix ()) x) (int 7))
 end
@@
 \begin{code}
-.<fun x_5 ->
-  (fun x_2 -> x_2 (fun x_3 x_4 -> x_4 x_3))
+.<fun x_5 -> (fun x_2 -> x_2 (fun x_3 x_4 -> x_4 x_3))
              (fun x_6 -> (fun x_1 -> x_1 true)
@@
 \begin{code}
-{P.st = Some <fun>;
- P.dy = .<fun x_5 -> x_5 true>.}
+{P.st = Some <fun>; P.dy = .<fun x_5 -> x_5 true>.}
 \end{code}
@@
 extensions.
-\end{comment}
+
+\fi
+\fi
 
@@
 \begin{align*}
-    \ident{int} &= \fun{x} x &
+    \ident{int} &= \fun{x} x \\
+    \ident{add} &= \fun{x} \fun{y} x+y \\
+    \ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0} \\
     \ident{lam} &= \fun{f} f \\
-    \ident{add} &= \fun{x} \fun{y} x+y &
     \ident{app} &= \fun{f} \fun{x} fx \\
-    \ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0} &
     \ident{fix} &= \fun{g} \fix{f} \fun{x} gfx
@@
 \begin{align*}
-    \encode{x} &= x &
+    \encode{x} &= x \\
+    \encode{n} &= \ident{int}\, n \\
+    \encode{e_1 + e_2} &= \ident{add} \encode{e_1} \encode{e_2} \\
+    \encode{\cond{b}{t}{e}} &= \rlap{$\ident{if\_} \encode{b}
+        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$} \\
     \encode{\fun{x}e} &= \ident{lam} (\fun{x} \encode{e}) \\
-    \encode{n} &= \ident{int}\, n &
     \encode{fx} &= \ident{app} \encode{f} \encode{x} \\
-    \encode{e_1 + e_2} &= \ident{add} \encode{e_1} \encode{e_2} &
-    \encode{\fix{f}e} &= \ident{fix} (\fun{f} \encode{e}) \\
-    \encode{\cond{b}{t}{e}} &= \rlap{$\ident{if\_} \encode{b}
-        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$}
+    \encode{\fix{f}e} &= \ident{fix} (\fun{f} \encode{e})
 \end{align*}
@@
 and apply it to the encoded term.  The result is the object term
-\begin{multline*}
+\begin{equation*}
     \bigl(\fun{e} e (\fun{f} \fun{x} fx) (\fun{f} f) (\fun{b} b)\bigr)
-\\
     \bigl(
@@
     \bigr),
-\end{multline*}
+\end{equation*}
 which partially evaluates to~$\True$ easily.
@@
 \begin{equation*}
+    \let\vcenter\vbox
     \begin{prooftree}
@@
 To interpret an object term~$e$ using an interpreter $I[~]$ is to
-evaluate the object term~$I[\encode{e}]$.  $\si$ is
+evaluate the object term~$I[\encode{e}]$.  $\si$~is
 a self\hyp interpreter in the following sense.
@@
 \end{proposition}
-As a corollary, we can pre-encode the self\hyp interpreter itself as
+As a corollary, we can pre-encode $\si$ itself as
 a context: the term $\si[\encode{\si[\encode{e}]}]$ is observationally
@@
 waiting to be plugged into a context as a higher-rank and higher-kind
-abstraction over the context.  Whereas our object language does not support
+abstraction over the context.  Even though our object language does not support
 higher abstraction, our metalanguages do, so they can type-check an object term
@@
 (in OCaml), or a value with a |Symantics| constraint over a type
-constructor (in Haskell).  Thus
+constructor (in Haskell).  Thus,
 ``context'' is a euphemism for a polymorphic argument, and ``plugging''
@@
 
+\begin{comment}
 \jacques{But how 
@@
 hole below).}
+\end{comment}
+
+\begin{comment}
 
@@
 
+\end{comment}
 \fi
@@
 \citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} that used
-embedded interpreters to justify advanced type systems, in particular,
-GADT. While we admire all this technical work, we were not convinced
+embedded interpreters to justify advanced type systems, in particular
+GADTs. While we admire all this technical work, we were not convinced
 that their motivating example actually needed all this machinery.
@@
 advanced type system or the overhead of tagging and untagging;
-(3) Thus an indirect interpreter is necessary, which needs a universal
+(3) Thus, an indirect interpreter is necessary, which needs a universal
   type (and hence tagging).
@@
 advanced type system or the overhead of tagging and untagging.
-\item Thus an indirect interpreter is necessary, which needs a universal
+\item Thus, an indirect interpreter is necessary, which needs a universal
   type (and hence tagging).
-\item Thus any self-interpreter must have tags and cannot be 
+\item Thus, any self-interpreter must have tags and cannot be 
   Jones-optimal.
@@
 because we represent object types as metalanguage types.  This work
-makes it safer and more efficient to embed %domain\hyp specific languages
-DSLs
+makes it safer and more efficient to embed DSLs
 in practical metalanguages such as Haskell and ML\@.
@@
 partial evaluation (\S\ref{S:PE-solution}) and CPS
-transformation%
-%aplas
-.
-%aplas-else ~(\S\ref{S:CPS}).
+transformation\ifshort\else~(\S\ref{S:CPS})\fi.
 This shift also underlies fusion,
 functionalization, and amortized complexity analysis.
-%aplas-else
-\begin{comment}
+\ifshort\else
 When the metalanguage does provide higher-rank and higher-kind
@@
 from any interpreters it may be plugged into.
-\end{comment}
+\fi
 
@@
 primitive recursive folds over the terms. 
-%aplas This makes operations like interpretation trivial to implement. 
+\ifshort\else
+This encoding makes operations like interpretation trivial to implement.
+\fi
 We still
@@
 recursive operations can be supported.
-
-% aplas: acknowledgments are at the beginning
-% \section*{Acknowledgments}
-% \ourthanks
-
-\bibliographystyle{mcbride}
-\bibsep=0pt
-% aplas
-\bibliography{aplas}
-% \bibliography{tagless}
-\end{document}

### 2007-07-19 tagless/tagless-final.tex
Done with resynchronizing JFP and APLAS versions

@@
     refined syntactic type structure. Ideally, such a treatment should
-    be metacircular, in the sense that the type structure used in the
+    be metacircular, in the sense that the type structure \linebreak[1] used in the
     defined language should be adequate for the defining language.
-    \rm \ifshort \hfill John Reynolds~\fi\citep{reynolds-definitional}
+    \rm \ifshort \hfill John Reynolds~\else\\\fi\citep{reynolds-definitional}
 \end{quote}
@@
 representation but varying the interpreter, we can
-\begin{itemize*}
+\begin{itemize}
     \item evaluate the term to a value in the metalanguage;
@@
 we refer the reader to the accompanying code for this.}\fi
-\end{itemize*}
+\end{itemize}
 We have programmed our interpreters in OCaml (and, for staging,
@@
 
-\begin{enumerate*}
+\begin{enumerate}
 \item We build interpreters that evaluate (\S\ref{language}),
@@
   harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
-\end{enumerate*}
+\end{enumerate}
 Our code is surprisingly simple and obvious in hindsight, but
@@
                   | _ -> pdyn (C.add (abstr e1) (abstr e2))
-
   let if_ eb et ee = match eb with
@@
                   | _ -> pdyn (C.app (abstr ef) (abstr ea))
-
   let fix f = let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
@@
     \end{tabular}
-\intertext{and the size\hyp measurer is the context}
+\intertext{and the size\hyp measurer at the end of~\S\ref{S:interpreter-RL} is the context}
     \mathrm{SZ}[~] &=
@@
 Although GADTs may indeed be more flexible and easier to use, they are
-not available in mainstream ML, and their current implementation in
-GHC has problems. We were also interested in discovering the minimal and
+not available in mainstream ML\@, and their current implementation in
+GHC has problems. We also wanted to find the minimal,
 widely available set of language features necessary for tagless
@@
 interpretation, provided we use universal types \cite{taha-tag}.  The
-ensuing tagging overhead motivated \cite{taha-tag} to propose tag
+ensuing tagging overhead motivated \citet{taha-tag} to propose tag
 elimination, which however does not statically guarantee that all tags
@@
 
-% aplas: this is the simplified paragraph. The original text is in the
-% comment right below.
-\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
-\citet{peyton-jones-simple} seem to argue that a self\hyp
-interpreter of a typed language cannot be tagless or Jones\hyp optimal:
-(1) One needs to encode a typed language in a typed language based on
-a sum type (at some level of the hierarchy);
-(2) It is not possible to write a \emph{direct} interpreter 
-for such an encoding of a typed language
-in a typed language, without either an
-advanced type system or the overhead of tagging and untagging;
-(3) Thus, an indirect interpreter is necessary, which needs a universal
-  type (and hence tagging).
-While the logic is sound, we showed that the premise in the very first step
-was not valid.
-\begin{comment}
 \Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
 \citet{peyton-jones-simple} seem to argue as follows that a self\hyp
-interpreter of a typed language cannot be tagless or Jones\hyp optimal.
-\begin{enumerate*}
+interpreter of a typed language cannot be tagless or Jones\hyp optimal:
+\begin{shortlist}
 \item One needs to encode a typed language in a typed language based on
-a sum type (at some level of the hierarchy).
+a sum type (at some level of the hierarchy)\\
 \item It is not possible to write a \emph{direct} interpreter 
@@
 in a typed language, without either an
-advanced type system or the overhead of tagging and untagging.
+advanced type system or the overhead of tagging and untagging\\
 \item Thus, an indirect interpreter is necessary, which needs a universal
-  type (and hence tagging).
+  type (and hence tagging)\ifshort\else\\
 \item Thus, any self-interpreter must have tags and cannot be 
-  Jones-optimal.
-\end{enumerate*}
+  Jones-optimal\fi.
+\end{shortlist}
 While the logic is sound, we showed that the premise in the very first step
-was not valid.
-\end{comment}
+is not valid.
 
@@
 
-%aplas-else
-\begin{comment}
+\ifshort\else
 A lot of effort has gone into ``typing dynamic typing'': to statically
 type-check dynamically\hyp typed values
-  \cite{baars-typing,haskell-list},
+\citep{baars-typing,WalidICFP02,Guillemette-Monier-PLPV,haskell-list},
 using the host language's type system to varying extents.
@@
 terms such as those read from a file.
-\end{comment}
+\fi
 
@@
 static and dynamic terms.  It is customary to implement such a bijection
-using an injection\hyp projection pair, as done for interpreters by
-\cite{Ramsey-ML-module-mania,Benton-embedded-interpreters},
-partial evaluation by \cite{Danvy-TDPE}, and type-level functions by
-\cite{oliveira-typecase}.  As explained in~\S\ref{S:PE-solution}, we
+using an injection\hyp projection pair, as done for interpreters
+\ifshort \cite{Ramsey-ML-module-mania,Benton-embedded-interpreters}\else
+by \citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters}\fi,
+partial evaluation \ifshort \cite{Danvy-TDPE}\else by \citet{Danvy-TDPE}\fi,
+and type-level functions \ifshort \cite{oliveira-typecase}\else by
+\citet{oliveira-typecase}\fi.  As explained in~\S\ref{S:PE-solution}, we
 avoid injection and projection at the type level by adding an argument
@@
 to combine online
-and offline partial evaluation.  Like \cite{SperberThiemann:TwoForOne},
-we strive for modularity by reusing interpreters for individual stages.  It would be interesting
+and offline partial evaluation.
+We strive for modularity by reusing interpreters for individual stages
+\cite{SperberThiemann:TwoForOne}.  It would be interesting
 to try to derive a \emph{cogen} \cite{Thiemann:cogeninsixlines}
@@
 
-The idea of multiple interpretations of the host language pervasives
-such as addition or application is quite common in implementing
-embedded DSL. It is also common to use phantom types to prevent
-forming of ill-typed DSL terms \cite{Lava,Rhiger-thesis}. However,
-the universal type still remains (|Bit| and |NumSig| in Lava
-\cite{Lava}, |Raw| and |Term| in \cite[Fig 2.2,
-  Sec. 3]{Rhiger-thesis}), along with the attendant pattern-matching
-overhead. These approaches are not tagless.  The universal type also
+It is common to implement an embedded DSL by providing multiple
+interpretations of host\hyp language pervasives such as addition and
+application.  It is also common to use phantom types to rule out
+ill-typed object terms, as Lava
+\cite{Lava} and \citet{Rhiger-thesis} do. However, these approaches
+are not tagless because they still use universal types, such as Lava's
+\texttt{Bit} and \texttt{NumSig} and Rhiger's \texttt{Raw} (his Fig.~2.2)
+and \texttt{Term} (his Chap.~3), and incur the attendant overhead of
+pattern matching.  The universal type also
 greatly complicates the soundness and completeness proofs of embedding
-\cite{Rhiger-thesis}. In contrast, our proofs become trivial.
-The approach of \Citet[Sec 3.3.4]{Rhiger-thesis} does not 
-support typed CPS transformation.
+\cite{Rhiger-thesis}, whereas our proofs are trivial.
+Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
 \begin{comment}
@@
 
-Our approach is based on encoding a term in its elimination form, as a
-co-algebraic structure. That basic idea and its application to
-meta-circular interpertation first appeared in \citet{Pfenning-Lee}.
-Our approach however is implementable in mainstream ML and supports
-type inference and typed CPS and partial evaluation. In contrast,
-\cite{Pfenning-Lee} conclude that partial evaluation and program
+We encode terms in elimination form, as a coalgebraic structure.
+\Citet{Pfenning-Lee} first described this basic idea and applied it to
+metacircular interpretation.
+Our approach, however, can be implemented in mainstream ML and supports
+type inference and typed CPS transformation and partial evaluation. In contrast,
+\citeauthor{Pfenning-Lee} conclude that partial evaluation and program
 transformations ``do not seem to be expressible'' even using their
-extension to $F_{\omega}$, perhaps because their avoidance of general
+extension to~$F_\omega$, perhaps because their avoidance of general
 recursive types compels them to include the polymorphic lift that we
-avoid in \S\ref{S:PE-lift}.
+avoid in~\S\ref{S:PE-lift}.
 \begin{comment}
@@
 we do is very similar to what they do (Figure 1, p.152), except that we 
-do it in standard programming languges.  It is unclear if their work can 
+do it in standard programming languages.  It is unclear if their work can 
 be implemented (yet) in any language.  And we preserve type-inference, 
-while their solution needs explcit types!
+while their solution needs explicit types!
 The following line of their conclusion is worth citing: "... this does 
@@
 Our encoding of the type function |repr_pe| in \S\ref{S:PE-solution}
-may be regarded as an emulation of type-indexed types and related to
-the intensional type analysis
-\cite{Morrisett-intensional,Generic-Haskell}. Our running HOAS
-example however, includes |fix|, which cannot be handled by the
-intensional type analysis \cite{xi-guarded}.  Our final approach
-seems related to \citearound{'s approach to HOAS via catamorphism and
-  anamorphism}\cite{Washburn-Weirich-boxes}.
+emulates type-indexed types and is related to intensional type analysis
+\cite{Morrisett-intensional,Generic-Haskell}. However, our object
+language and running examples in HOAS include |fix|,
+which intensional type analysis cannot handle
+\cite{xi-guarded}.  Our final approach
+seems related to \citearound{'s approach to HOAS using catamorphisms and
+  anamorphisms}\citet{Washburn-Weirich-boxes}.
 
 We could not find work that establishes that
-the \emph{typed} $\lambda$-calculus possesses a final coalgebra structure.
-% aplas: abbreviated. The original text is in comments
-(see \Citet{honsell99coinductive} for the untyped case).
-% aplas: comment
-\begin{comment}
+the \emph{typed} $\lambda$-calculus has a final coalgebra structure.
+\ifshort
+(See \Citet{honsell99coinductive} for the untyped case.)
+\else
 \Citet{HonsellLenisa,honsell99coinductive}
-investigate the untyped $\lambda$-calculus,
+investigate the untyped $\lambda$-calculus
 along this line.  
@@
 \emph{observational equivalence}
-(see \S\ref{selfinterp}).
+(see our~\S\ref{selfinterp}).
 \citearound{'s bibliography}\Citet{honsell99coinductive} refers to the
@@
 coinductive aspects of B\"{o}hm trees, as pointed out by
-\cite{berarducci-models} and \cite[Example 4.3.4]{jacobs-coalgebra}.
-\end{comment}
-
-%aplas
-\begin{comment}
+\citet{berarducci-models} and Jacobs \citeyearpar[Example 4.3.4]{jacobs-coalgebra}.
+\fi
+
+\ifshort\else
 One way to understand our main idea is to eschew sum types and sum kinds
@@
 recursive data types \cite{bohm-automatic}.
-\end{comment}
-
-%aplas: beforeAs \S\ref{S:clarify} observes, 
-We observe that
+\fi
+
+\ifshort We observe that \else As \S\ref{S:clarify} observes, \fi
 higher-rank and higher-kind
 polymorphism let us type-check and compile object terms separately from
-interpreters.  This is consistent with the role of
+interpreters.  This \ifshort\else observation \fi is consistent with the role of
 polymorphism in the separate compilation of modules

### 2007-07-21 tagless/tagless-final.tex
Cite Chen and Xi's typeful CPS in DML (PEPM 2003)

@@
   let abstr {dy = x} = x    let pdyn x = {st = None; dy = x}
+
   let int  (x:int)  = {st = Some (R.int x);  dy = C.int x}
-
   let add e1 e2 = match e1, e2 with
@@
   | _ -> pdyn (C.add (abstr e1) (abstr e2))
-
   let if_ eb et ee = match eb with
@@
 
-Our initial motivation came from several papers 
-\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} that used
-embedded interpreters to justify advanced type systems, in particular
-GADTs. While we admire all this technical work, we were not convinced
-that their motivating example actually needed all this machinery.
+Our initial motivation came from several papers that justify advanced
+type systems, in particular GADTs, by embedded interpreters
+\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} and
+CPS transformations \citep{Guillemette-Monier-PLPV,chen-typeful}.
+We admire all this technical machinery, but
+these motivating examples do not need it.
 Although GADTs may indeed be more flexible and easier to use, they are
-not available in mainstream ML\@, and their current implementation in
-GHC has problems. We also wanted to find the minimal,
-widely available set of language features necessary for tagless
+unavailable in mainstream ML and implemented problematically
+in GHC currently. We also wanted to find the minimal
+set of widespread language features needed for tagless
 type-preserving interpretation.
@@
 a sum type (at some level of the hierarchy)\\
-\item It is not possible to write a \emph{direct} interpreter 
+\item A \emph{direct} interpreter 
 for such an encoding of a typed language
-in a typed language, without either an
-advanced type system or the overhead of tagging and untagging\\
+in a typed language requires either
+advanced types or tagging overhead\\
 \item Thus, an indirect interpreter is necessary, which needs a universal
-  type (and hence tagging)\ifshort\else\\
+  type and hence tagging\ifshort\else\\
 \item Thus, any self-interpreter must have tags and cannot be 

### 2007-07-21 tagless/tagless-final.tex
Elide a bit less code from the journal version

@@
 \begin{abstract}
+\ifdim\parindent=0pt \parindent=1em \fi
 We have built the first family of tagless interpretations for a
@@
 The methods |add|, |mul|, and |leq| are quite similar, and so are
-|int| and |bool|. Therefore, we will frequently show implementations of
-only one method of each group and elide the rest, to save space. The
+|int| and |bool|. Therefore, we often show
+only one method of each group and elide the rest. The
 accompanying code has the complete implementations.
@@
 of each object term, defined as the number of term
-constructors. The following is slightly abbreviated code (see the
-accompanying source code for the complete definition).
+constructors.
 \begin{code}
@@
   type ('c,'dv) repr = int
+
   let int  (x:int)  = 1
+  let bool (b:bool) = 1
   let lam  f        = f 0 + 1
@@
   let fix  f        = f 0 + 1
+  let add  e1 e2    = e1 + e2 + 1
   let mul  e1 e2    = e1 + e2 + 1
+  let leq  x y      = x + y + 1
   let if_  eb et ee = eb + et () + ee () + 1
@@
 whenever the
-argument is static.  In the accompanying source code is a conservative
-alternative |P.fix| that unrolls recursion only once, then residualizes.
-Many sophisticated approaches have been developed for this trade-off
+argument is static.
+\ifshort
+In the accompanying source code is a conservative
+alternative |P.fix| that unfolds recursion only once, then residualizes.
+\else
+A conservative alternative is to unfold recursion only once, then residualize:
+\begin{code}
+let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
+\end{code}
+\fi
+Many sophisticated approaches have been developed to decide how much to unfold
 \cite{jones-partial}, but this issue is orthogonal to our presentation.

### 2007-07-26 tagless/tagless-final.tex
various edits, as per my email

@@
 postprocessing to eliminate tags.
-The interpretations include
+The statically type-preserving interpretations include
 an evaluator, a compiler (or staged evaluator), a partial
@@
 let testf3r = testf3 ()
+This expression has type unit but is here used with type 'a * 'b
 \end{code}
@@
     across Haskell 98 and ML.
+\item We show the first typed, tagless and statically type-preserving CPS
+    transformation with simple types, which is available for mainstream
+    functional languages (ML and Haskell). Type-preservation is assured by
+    the soundness of our embedding.
 \item We use the type system of the metalanguage
@@
 end
+
 module EX(S: Symantics) = struct open S
@@
 \end{floatrule}
-\caption{A simple (Meta)OCaml embedding of our object language}
+\caption{A simple (Meta)OCaml embedding of our object language, 
+    and examples of use}
 \label{fig:ocaml-simple}
@@
 The functor |EX| in Fig.~\ref{fig:ocaml-simple} encodes 
-our running examples |test1| and $\mathit{power}$.
+our running examples |test1| and the $\mathit{power}$ function
+(|testpowfix|).
 The dummy argument to |test1| and |testpowfix| is an artifact of
@@
 \else
+\begin{proposition}
+    The compiler |C| generates tagless, pattern-match free code.
+\end{proposition}
+This is syntactically clear from the body of |C|, and follows
+readily by structural induction and the properties of |.~| and
+|.< >.|.
 
@@
 references of the metalanguage.  The accompanying code illustrates these
-extensions.
+extensions.  No new techniques were needed in the implementation.
 

### 2007-07-26 tagless/tagless-final.tex
small edits to clarify things, a bit more to come

@@
 Our encoding of an object term abstracts over the various ways to
-interpret it yet statically assures that the interpreters never get
+interpret it, yet statically assures that the interpreters never get
 stuck.  To achieve self\hyp interpretation and show Jones\hyp
@@
 
-Following \cite{WalidICFP02},
+\noindent Following \cite{WalidICFP02},
 we try to implement an interpreter function |eval0|\ifshort,
@@
 uses no universal type or pattern matching, it never results in a
-run-time error (and is in fact total).  Because this safety is obvious
+run-time error, and is in fact total.  Because this safety is obvious
 not just to us but also to the metalanguage implementation, we avoid
@@
     \item compile the term, with staging support such as in MetaOCaml;
-    \item even partially evaluate the term, online; and
+    \item partially evaluate the term, online; and
     \item transform the term to continuation\hyp passing style (CPS),
-        even call-by-name (CBN) CPS, so to isolate the evaluation
+        even call-by-name (CBN) CPS, so as to isolate the evaluation
         order of the object language from that of the metalanguage.\ifshort
@@
 \item We describe an approach to self\hyp interpretation compatible with the
-  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned out
-  harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
+  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
+  out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
 \end{enumerate}
@@
 To reiterate, we do \emph{not} propose any new language
-feature or technique. We use features already present in
+feature or new technique. We use features already present in
 mainstream functional languages---Hindley-Milner
 type system with either an inference\hyp preserving module system or
-constructor classes, as realized in ML and Haskell 98---to
+constructor classes, as realized in ML and Haskell 98---and techniques
+which have all appeared in the literature, to
 solve the problem that was stated in the published record as
@@
 The language is close to \citets{xi-guarded}, without their polymorphic
-lift but with more constants so as to express Fibonacci, factorial, and
-power.
+lift but with more constants so as to more conveniently express Fibonacci,
+factorial, and power.
 In contrast to \S\ref{intro}, we encode binding using higher-order
@@
 evaluates to |3|. This interpreter is not only tagless but also
-total. It ``evaluates'' even seemingly divergent terms like
-|app (fix (fun self -> self)) (int 1)|.
+total. It ``evaluates'' even seemingly divergent terms, for instance
+|app (fix (fun self -> self)) (int 1)| evaluates to $3$.
 \fi
@@
 represents future-stage expressions of type~$t$ at the
-present stage as values of type |('c,|$t$|) code| where |'c| is the
+present stage as values of type |('c, |$t$|) code| where |'c| is the
 environment classifier \cite{WalidPOPL03,calcagno-ml-like}. Code values are created
@@
 a problem with |if_|.  If the first argument to |if_| is a dynamic term
-(type |C Bool|) and the second and third arguments are a static term
-(type |R a|) and a dynamic term (type |C a|), then we need to convert
+(of type |C Bool|), the second a static term (of type |R a|) and the third a
+dynamic term (of type |C a|), then we need to convert
 the static term to dynamic, but there is no polymorphic ``lift''
@@
 by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
-not statically known.  Even the same static function, applied to
+not statically known.  The same static function, applied to
 different static arguments, can give both static and dynamic results: we
@@
 The function |abstr|, as in~\S\ref{S:PE-GADT},
-extracts a future-stage code value from a result of
+extracts a future-stage code value from the result of
 partial evaluation.  Conversely, the function |pdyn| injects a
-code value into partial evaluation. As
+code value into the |repr| type. As
 in~\S\ref{S:PE-problem}, we build dynamic terms alongside
@@
 in a large language with more base types and primitive operations.
+\ifshort\else
+Incidentally, the code actually contains a more general implementation
+mechanism for such features, inspired in part by previous work in
+generative linear algebra~\cite{CaretteKiselyov05}.
+\fi
 
@@
 
-Unlike the GADT approach in~\S\ref{S:PE-GADT}, all pattern\hyp matching
-in~|P| is \emph{syntactically} exhaustive, so it is patent to the metalanguage
-implementation that |P| never gets stuck.  Further, all pattern\hyp matching occurs
+\noindent Unlike the GADT approach in~\S\ref{S:PE-GADT}, all pattern\hyp
+matching in~|P| is \emph{syntactically} exhaustive, so it is patent to the
+metalanguage implementation that |P| never gets stuck.  Further, all
+pattern\hyp matching occurs
 during partial evaluation, only to check if a value is known statically,
@@
 Because the evaluator above mostly consists of glorified identity
-functions, we expect our simple partial evaluator to reduce the
+functions, our simple partial evaluator reduces the
 result of this instantiation to~$e$.  In general, to
@@
 which partially evaluates to~$\True$ easily.
-However, encoding fails on an term with multiple $\lambda$\hyp
+However, encoding fails on a term with multiple $\lambda$\hyp
 abstractions at different types.  For example, the pre-encoding
@@
 
-We type-check a pre-encoded term~$\encode{e}$ and its interpreter~$I[~]$
-only together, never separately.  This treatment has the drawback that
+We always type-check a pre-encoded term~$\encode{e}$ and its interpreter~$I[~]$
+together, never separately.  This treatment has the drawback that
 we must duplicate a pre-encoded term in order to interpret it in
@@
 
+\begin{comment}
+Our solution could have been even more straightforward if we had
+total type-level functions at our disposal (|repr_pe| in
+\S\ref{S:PE-solution}) in MetaOCaml.  Simple type-level computations 
+certainly ought to become mainstream.
+\end{comment}
+
 \section{Conclusions}\label{conclusion}

### 2007-07-27 tagless/tagless-final.tex
added write-up about state-passing CPS

@@
 
-We can modify the CBV CPS transformation to pass a piece of state along
-with the continuation. This technique lets us support mutable state. We
-can also add mutable references to the object language using mutable
-references of the metalanguage.  The accompanying code illustrates these
-extensions.  No new techniques were needed in the implementation.
+We can modify the CBV CPS transformation to pass a piece of state
+along with the continuation. This technique lets us support mutable
+state and implement an extended object language, with operations
+|deref ()| to obtain the current state, |set v| to replace the current
+state with |v| returning the previous value of the state, and the
+let-form |let x = e1 in e2|. The latter enforces the order of
+evaluation: |e1| will always be evaluated before |e2| even if the
+expression |e2| does not need the value of |x|. The more familiar
+imperative sequencing form |begin e1; e2 end| is expressible in our
+object language as |let _ = e1 in e2|.  In HOAS, the let-form actually
+has the form |lapp e1 (fun x -> e2)|, which looks like the functional
+application with its two arguments swapped.  The extended object
+language corresponds to the following signature, which
+includes all of the original Symantics and adds imperative features.
+
+\begin{code}
+module type SymSI = sig
+  include Symantics
+  type state
+  type 'c states	(* static version of the state *)
+  val lapp : ('c,'sa,'da) repr -> 
+             (('c,'sa,'da) repr -> ('c,'sb,'db) repr) ->
+             ('c,'sb,'db) repr
+  val deref : unit -> ('c,'c states,state) repr
+  val set   : (('c,'c states,state) repr as 'x) -> 'x
+end
+\end{code}
+
+We can write a simple test: 
+\begin{code}
+let v0 = !state in state := 2; v0 + !state
+\end{code}
+which, in our language, takes the form
+\begin{code}
+module EXSI_INT(S: SymSI with type state = int and type 'c states = int) = 
+struct
+  open S
+  let test1 () = lapp (deref ()) (fun v0 -> 
+                  lapp (set (int 2)) (fun _ ->
+		   add v0 (deref ())))
+end;;
+\end{code}
+The accompanying source code shows several more tests, including the
+imperative power function that uses the state as the accumulator and
+the test of a higher-order state.
+
+The state-passing interpreter is the extension of the CBN CPS
+interpreter RCN of\S\ref{S:CPS}:
+\begin{code}
+module RCPS (ST: sig 
+  type state 
+  type 'c states 
+  type ('c,'sv,'dv) repr = 
+      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
+end) = struct
+  include ST
+  type ('c, 'sv, 'dv) result = 'c states -> 'sv
+  ... 
+  let get_res x = fun s0 -> x.ko (fun v s -> v) s0
+  let lapp e2 e1 = 
+    {ko = fun k -> e2.ko (fun v -> (app (lam e1) {ko = fun k -> k v}).ko k)}
+  let deref () = {ko = fun k s -> k s s}
+  let set e = {ko = fun k -> e.ko (fun v s -> k s v)}
+end
+\end{code}
+The implementations of |int|, |app|, |lam|, etc. functions is \emph{identical}
+to those of |RCN| and elided. New here is the extended type |repr|
+that now includes the state, and the `constructors' of the imperative
+forms of the object language: |lapp|, |deref|, and |set|. The
+interpreter is still CBN and so evaluation of |app ef ea| does not
+guarantee to evaluate |ea|. On the other hand, the implementation of
+the let-form |lapp ea ef| does assure that |ea| will always be
+evaluated. In the case of the first-order state such as |int| we 
+instantiate the interpreter as
+\begin{code}
+module RCPSI = RCPS(struct 
+  type state = int
+  type 'c states = int
+  type ('c,'sv,'dv) repr = 
+      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
+end);;
+\end{code}
+It fully implements the |Symantics| signature and so can run the pure
+code of the unextended object language: we can pass |RCPSI| as an
+argument to the functor |EX| (Fig.~\ref{fig:ocaml-simple}) and run 
+the example |test1| from there. The main use for |RCPSI| is to run the 
+extended object language:
+\begin{code}
+module EXPSI_INT = EXSI_INT(RCPSI)
+let cpsitesti1 = RCPSI.get_res (EXPSI_INT.test1 ()) 100 (* 102 *)
+\end{code}
+
+We can also add mutable references to the object language using
+mutable references of the metalanguage, as shown in the accompanying
+code.  No new techniques were needed in the implementation.
 

### 2007-07-27 tagless/tagless-final.tex
Small formatting changes to the JFP version that Jacques suggested: - for Figure 2, since we now have space. might it make sense to separate it into 2 figures (the embedding and the example)?  It certainly would make the figure clearer and simplify the text. - JFP discourages footnotes -- should we eliminate them all?

@@
 
-\begin{figure*}[t]
-\begin{floatrule}
-\begin{code2}
+\begin{SaveVerbatim}{ocaml-simple}
 module type Symantics = sig type ('c, 'dv) repr
@@
 end
-
+\end{SaveVerbatim}
+\begin{SaveVerbatim}{ocaml-examples}
 module EX(S: Symantics) = struct open S
@@
 end
-\end{code2}
+\end{SaveVerbatim}
+
+\ifshort
+\begin{figure}[t]
+\begin{floatrule}
+\begin{tabular}{@{}l@{}}
+\ifx\relax\normalbaselineskip\else\baselineskip\normalbaselineskip\fi
+\BUseVerbatim[baseline=b]{ocaml-simple}\\[\medskipamount]
+\ifx\relax\normalbaselineskip\else\baselineskip\normalbaselineskip\fi
+\BUseVerbatim[baseline=t]{ocaml-examples}
+\end{tabular}
 \end{floatrule}
@@
 \label{fig:ocaml-simple}
-\end{figure*}
+\label{fig:ocaml-examples}
+\end{figure}
+\else
+\begin{figure}[t]
+\begin{floatrule}
+\BUseVerbatim{ocaml-simple}
+\end{floatrule}
+\caption{A simple (Meta)OCaml embedding of our object language}
+\label{fig:ocaml-simple}
+\end{figure}
+\begin{figure}[t]
+\begin{floatrule}
+\BUseVerbatim{ocaml-examples}
+\end{floatrule}
+\caption{Examples using the embedding in Figure~\ref{fig:ocaml-simple} of our object language}
+\label{fig:ocaml-examples}
+\end{figure}
+\fi
+
 To embed the same object language in (Meta)OCaml, we replace the \ifshort type
@@
 
-The functor |EX| in Fig.~\ref{fig:ocaml-simple} encodes 
+The functor |EX| in Fig.~\ref{fig:ocaml-examples} encodes 
 our running examples |test1| and the $\mathit{power}$ function
@@
 \begin{proposition}
-    The compiler |C| generates tagless, pattern-match free code.
+    The compiler |C| generates tagless code free of pattern matching.
 \end{proposition}
@@
 it: we defined a data type |ByteCode| with
-constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.%
-\footnote{\texttt{ByteCode} can be mapped to the abstract syntax of Template
+constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
+(Alternatively, we could use Template Haskell as our staging facility:
+\texttt{ByteCode} can be mapped to the abstract syntax of Template
 Haskell. The output of our compiler will then be assuredly type-correct
-    Template Haskell.}
+Template Haskell.)
 Whereas our representation of object terms uses HOAS,
@@
 class |Symantics|.
-The implementation\footnote{The implementation uses
+The implementation is quite similar (but slightly more
+verbose) than the corresponding MetaOCaml code above. (The implementation uses
 GADTs because we also wanted to write a typed interpreter for 
-the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
-verbose) than the corresponding MetaOCaml code above. The
+the \texttt{ByteCode} \emph{data type}.) The
 accompanying code gives the full details.
@@
 \cite{xi-guarded,WalidPOPL03}.\ifshort\else
-\footnote{If we
+(By the way, if we
 were to add polymorphic \texttt{lift} to the type class
 \texttt{Symantics repr}, then \texttt{repr} would become an instance of
-\texttt{Applicative} and thus \texttt{Functor}:\quad\texttt{fmap
-f = app (lift f)}}\fi
+\texttt{Applicative} and thus \texttt{Functor}:\texttt{ fmap
+f = app (lift f)~}.)\fi
 
@@
 This solution using GADTs works but is not quite satisfactory. First, it
-cannot be ported to MetaOCaml\ifshort\else\ as GADTs are unavailable there\fi.  Second,
+cannot be ported to MetaOCaml\ifshort\else, as GADTs are unavailable there\fi.  Second,
 the problem of nonexhaustive pattern\hyp matching reappears in
@@
 \else
-Indeed the metalanguage implementation could reason thus,%
-\footnote{If we use inductive families (as those in Coq) or logical
+Indeed, the metalanguage implementation could reason thus:
+if we use inductive families (as in Coq) or logical
   frameworks with canonical forms (as in Twelf with its coverage checker),
-  we can prove the pattern matching to be exhaustive.}
-but it may not;
+  we can prove the pattern matching to be exhaustive.
+Then again, the metalanguage implementation may not reason thus:
 \fi
@@
 % aplas: abbreviated figure
-\begin{figure*}
+\begin{figure}
 \begin{floatrule}
@@
 \label{fig:ocaml}
-\end{figure*}
+\end{figure}
 

### 2007-07-27 tagless/tagless-final.tex
small edits here and there

@@
 along with the continuation. This technique lets us support mutable
-state and implement an extended object language, with operations
+state. This extended object language adds operations
 |deref ()| to obtain the current state, |set v| to replace the current
 state with |v| returning the previous value of the state, and the
-let-form |let x = e1 in e2|. The latter enforces the order of
+let-form |let x = e1 in e2|. The latter enforces an order of
 evaluation: |e1| will always be evaluated before |e2| even if the
@@
 object language as |let _ = e1 in e2|.  In HOAS, the let-form actually
-has the form |lapp e1 (fun x -> e2)|, which looks like the functional
+has the form |lapp e1 (fun x -> e2)|, which looks like functional
 application with its two arguments swapped.  The extended object
 language corresponds to the following signature, which
-includes all of the original Symantics and adds imperative features.
+includes all of the original |Symantics| and adds imperative features.
 
@@
 \end{code}
-
 We can write a simple test: 
@@
 \begin{code}
-module EXSI_INT(S: SymSI with type state = int and type 'c states = int) = 
+module EXSI_INT(S: SymSI with type state = int 
+                          and type 'c states = int) = 
 struct
@@
 \end{code}
-The accompanying source code shows several more tests, including the
-imperative power function that uses the state as the accumulator and
-the test of a higher-order state.
-
-The state-passing interpreter is the extension of the CBN CPS
+The accompanying source code shows several more tests, including an
+imperative power function that uses state as the accumulator as well as
+a test for a higher-order state.
+
+The state-passing interpreter is an extension of the CBN CPS
 interpreter RCN of\S\ref{S:CPS}:
@@
   let get_res x = fun s0 -> x.ko (fun v s -> v) s0
-  let lapp e2 e1 = 
-    {ko = fun k -> e2.ko (fun v -> (app (lam e1) {ko = fun k -> k v}).ko k)}
+  let lapp e2 e1 = {ko = fun k -> e2.ko 
+      (fun v -> (app (lam e1) {ko = fun k -> k v}).ko k)}
   let deref () = {ko = fun k s -> k s s}
@@
 \end{code}
-The implementations of |int|, |app|, |lam|, etc. functions is \emph{identical}
-to those of |RCN| and elided. New here is the extended type |repr|
-that now includes the state, and the `constructors' of the imperative
+The implementations of the |int|, |app|, |lam|, etc. functions is
+\emph{identical} to those of |RCN| and elided. New here is the extended type
+|repr|
+which now includes the state, and the ``constructors'' for the imperative
 forms of the object language: |lapp|, |deref|, and |set|. The
@@
 the let-form |lapp ea ef| does assure that |ea| will always be
-evaluated. In the case of the first-order state such as |int| we 
+evaluated. In the case of first-order state (such as |int|) we 
 instantiate the interpreter as
@@
 module EXPSI_INT = EXSI_INT(RCPSI)
-let cpsitesti1 = RCPSI.get_res (EXPSI_INT.test1 ()) 100 (* 102 *)
+let cpsitesti1 = RCPSI.get_res (EXPSI_INT.test1 ()) 100
+val cpsitesti1 : int = 102
 \end{code}
@@
 mutable references of the metalanguage, as shown in the accompanying
-code.  No new techniques were needed in the implementation.
+code.  It is worthwhile reiterating how close this implementation is to the CPS
+interpreter, and that adding state and imperative features needed no new
+techniques.
 
@@
 to |repr|.
+\begin{comment}
+Our solution could have been even more straightforward if we had
+total type-level functions at our disposal (|repr_pe| in
+\S\ref{S:PE-solution}) in MetaOCaml.  Simple type-level computations 
+certainly ought to become mainstream.
+\end{comment}
 
@@
 application.  It is also common to use phantom types to rule out
-ill-typed object terms, as Lava
-\cite{Lava} and \citet{Rhiger-thesis} do. However, these approaches
+ill-typed object terms, as done in Lava
+\cite{Lava} and by \citet{Rhiger-thesis}. However, these approaches
 are not tagless because they still use universal types, such as Lava's
-\texttt{Bit} and \texttt{NumSig} and Rhiger's \texttt{Raw} (his Fig.~2.2)
-and \texttt{Term} (his Chap.~3), and incur the attendant overhead of
+\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Fig.~2.2)
+and \texttt{Term} (his Chap.~3), which incur the attendant overhead of
 pattern matching.  The universal type also
@@
 Our approach, however, can be implemented in mainstream ML and supports
-type inference and typed CPS transformation and partial evaluation. In contrast,
+type inference, typed CPS transformation and partial evaluation. In contrast,
 \citeauthor{Pfenning-Lee} conclude that partial evaluation and program
@@
 higher-rank and higher-kind
-polymorphism let us type-check and compile object terms separately from
+polymorphism lets us type-check and compile object terms separately from
 interpreters.  This \ifshort\else observation \fi is consistent with the role of
@@
 \cite{shao-typed}.
-
-\begin{comment}
-Our solution could have been even more straightforward if we had
-total type-level functions at our disposal (|repr_pe| in
-\S\ref{S:PE-solution}) in MetaOCaml.  Simple type-level computations 
-certainly ought to become mainstream.
-\end{comment}
 

### 2007-07-27 tagless/tagless-final.tex
Small edits to Section 5.3 (state) Cite Jones "Challenging Problems" for optimality

@@
     across Haskell 98 and ML.
-\item We show the first typed, tagless and statically type-preserving CPS
-    transformation with simple types, which is available for mainstream
-    functional languages (ML and Haskell). Type-preservation is assured by
+\ifshort\else
+\item We show the first typed, tagless, and statically type-preserving CPS
+    transformation with simple types (\S\ref{variations}), which is available for mainstream
+    functional languages (ML and Haskell). Type preservation is assured by
     the soundness of our embedding.
+\fi
 \item We use the type system of the metalanguage
@@
 This output is a na\"{\i}ve CPS transformation of $(\fun{x}x)\True$,
-containing several apparent $\beta$-redexes.  To eliminate these
+containing several apparent $\beta$-redexes.  To reduce these
 redexes, we just change~|T| to instantiate |CPST| with |P| instead.
@@
 
-We can modify the CBV CPS transformation to pass a piece of state
+\begin{figure}
+    \begin{floatrule}
+    \begin{proofrules}
+        \[ \justifies \deref:t_s \]
+        \[ e:t_s \justifies \set e:t_s \]
+        \[ e_1:t_1 \quad \[ [x:t_1] \proofoverdots e_2:t_2 \] \justifies \lapp{e_1}{x}{e_2}:t_2 \]
+    \end{proofrules}
+    \end{floatrule}
+    \caption{Extending our typed object language with mutable state of type~$t_s$}
+    \label{fig:state}
+\end{figure}
+
+We can modify a CBN or CBV CPS transformation to pass a piece of state
 along with the continuation. This technique lets us support mutable
-state. This extended object language adds operations
-|deref ()| to obtain the current state, |set v| to replace the current
-state with |v| returning the previous value of the state, and the
-let-form |let x = e1 in e2|. The latter enforces an order of
-evaluation: |e1| will always be evaluated before |e2| even if the
-expression |e2| does not need the value of |x|. The more familiar
-imperative sequencing form |begin e1; e2 end| is expressible in our
-object language as |let _ = e1 in e2|.  In HOAS, the let-form actually
-has the form |lapp e1 (fun x -> e2)|, which looks like functional
-application with its two arguments swapped.  The extended object
-language corresponds to the following signature, which
-includes all of the original |Symantics| and adds imperative features.
-
+state. As Figure~\ref{fig:state} shows, we extend our object language
+with three imperative features.
+\begin{enumerate}
+    \item ``$\deref$'' gets the current state;
+    \item ``$\set e$'' sets the state to the value of~$e$ and returns
+        the previous value of the state;
+    \item the let-form ``$\lapp{e_1}{x}e_2$'' evaluates $e_1$
+        before~$e_2$ even if $e_2$ does not use~$x$.
+\end{enumerate}
+If $x$ does not appear in~$e_2$, then ``$\lapp{e_1}{x}e_2$'' is same as
+the more familiar sequencing form ``$e_1;e_2$''.
+We can embed this extended object language into OCaml by extending the
+|Symantics| signature in Figure~\ref{fig:ocaml}.
 \begin{code}
@@
   type 'c states       (* static version of the state *)
-  val lapp : ('c,'sa,'da) repr -> 
-             (('c,'sa,'da) repr -> ('c,'sb,'db) repr) ->
-             ('c,'sb,'db) repr
-  val deref : unit -> ('c,'c states,state) repr
-  val set   : (('c,'c states,state) repr as 'x) -> 'x
+  val lapp : (('c,'sa,'da) repr as 'x) -> ('x -> 'y)
+             -> (('c,'sb,'db) repr as 'y)
+  val deref : unit -> ('c, 'c states, state) repr
+  val set   : (('c, 'c states, state) repr as 'x) -> 'x
 end
 \end{code}
-We can write a simple test: 
-\begin{code}
-let v0 = !state in state := 2; v0 + !state
-\end{code}
-which, in our language, takes the form
-\begin{code}
-module EXSI_INT(S: SymSI with type state = int 
-                          and type 'c states = int) = 
-struct
-  open S
-  let test1 () = lapp (deref ()) (fun v0 -> 
-                  lapp (set (int 2)) (fun _ ->
-                   add v0 (deref ())))
-end;;
-\end{code}
-The accompanying source code shows several more tests, including an
-imperative power function that uses state as the accumulator as well as
-a test for a higher-order state.
-
-The state-passing interpreter is an extension of the CBN CPS
-interpreter RCN of\S\ref{S:CPS}:
-\begin{code}
-module RCPS (ST: sig 
+In HOAS, the type of |lapp| is that of function application with
+the two arguments swapped.  We can encode the term
+``$\lapp{\deref}{x} \set 2;\, x+\deref$''
+as the OCaml functor
+\begin{code}
+module EXSI_INT(S: SymSI
+  with type state = int and type 'c states = int) = struct open S
+  let test1 () = lapp (deref ()) (fun x -> 
+                  lapp (set (int 2)) (fun _ -> add x (deref ())))
+end
+\end{code}
+The accompanying source code shows several more tests, including
+a test for higher-order state and a power function that uses state
+as the accumulator.
+
+The state-passing interpreter extends the CBN CPS
+interpreter |RCN| of~\S\ref{S:CPS}.
+\begin{code}
+module RCPS(ST: sig 
   type state 
@@
   ... 
-  let get_res x = fun s0 -> x.ko (fun v s -> v) s0
-  let lapp e2 e1 = {ko = fun k -> e2.ko 
-      (fun v -> (app (lam e1) {ko = fun k -> k v}).ko k)}
+  let lapp e2 e1 = {ko = fun k ->
+      e2.ko (fun v -> (app (lam e1) {ko = fun k -> k v}).ko k)}
   let deref () = {ko = fun k s -> k s s}
   let set e = {ko = fun k -> e.ko (fun v s -> k s v)}
+  let get_res x = fun s0 -> x.ko (fun v s -> v) s0
 end
 \end{code}
-The implementations of the |int|, |app|, |lam|, etc. functions is
-\emph{identical} to those of |RCN| and elided. New here is the extended type
-|repr|
-which now includes the state, and the ``constructors'' for the imperative
-forms of the object language: |lapp|, |deref|, and |set|. The
-interpreter is still CBN and so evaluation of |app ef ea| does not
-guarantee to evaluate |ea|. On the other hand, the implementation of
-the let-form |lapp ea ef| does assure that |ea| will always be
-evaluated. In the case of first-order state (such as |int|) we 
+The implementations of |int|, |app|, |lam|, and so on are
+\emph{identical} to those of |RCN| and elided. New are the extended type
+|repr|,
+which now includes the state, and the functions
+|lapp|, |deref|, and |set| representing imperative features. The
+interpreter is still CBN, so evaluating |app ef ea| may not
+evaluate |ea|, but evaluating |lapp ea ef| does.
+For first-order state, such as of type~$\ZZ$, we 
 instantiate the interpreter as
@@
       {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
-end);;
-\end{code}
-It fully implements the |Symantics| signature and so can run the pure
-code of the unextended object language: we can pass |RCPSI| as an
-argument to the functor |EX| (Fig.~\ref{fig:ocaml-simple}) and run 
-the example |test1| from there. The main use for |RCPSI| is to run the 
-extended object language:
+end)
+\end{code}
+|RCPSI| matches the |Symantics| signature and implements the
+unextended object language: we can pass |RCPSI|
+to the functor |EX| (Fig.~\ref{fig:ocaml-simple}) and run 
+the example |test1| from there. The main use for |RCPSI| is to interpret the 
+extended object language.
 \begin{code}
@@
 equivalent object terms~$\pe(e)$.  It is said to be
-\emph{optimal} with respect to~|si| if $\pe(\si\Encode{e})$
+\emph{optimal} with respect to~|si| \citep{jones-challenging} if $\pe(\si\Encode{e})$
 is equal to~$e$ (up to $\alpha$\hyp conversion, or in some accounts, no
@@
     \ident{add} &= \fun{x} \fun{y} x+y \\
-    \ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0} \\
+    \ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}
+        \displaybreak[0] \\
     \ident{lam} &= \fun{f} f \\
@@
     \encode{\cond{b}{t}{e}} &= \rlap{$\ident{if\_} \encode{b}
-        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$} \\
+        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$}
+        \displaybreak[0] \\
     \encode{\fun{x}e} &= \ident{lam} (\fun{x} \encode{e}) \\

### 2007-07-27 tagless/tagless-final.tex
Extend related work

@@
 \citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} and
-CPS transformations \citep{Guillemette-Monier-PLPV,chen-typeful}.
+CPS transformations \citep{Guillemette-Monier-PLPV,shao-type-toplas,chen-typeful}.
 We admire all this technical machinery, but
@@
 to |repr|.
-\begin{comment}
-Our solution could have been even more straightforward if we had
-total type-level functions at our disposal (|repr_pe| in
-\S\ref{S:PE-solution}) in MetaOCaml.  Simple type-level computations 
-certainly ought to become mainstream.
-\end{comment}
+Our solution could have been even more straightforward if MetaOCaml
+provided total type-level functions such as |repr_pe| in
+\S\ref{S:PE-solution}---simple type-level computations 
+ought to become mainstream.
 
@@
 in~\S\ref{S:PE-solution}).
+
+Using Haskell, \citet{Guillemette-Monier-PLPV} implement a CPS transformation
+for HOAS terms and statically assure that it preserves object types.
+They represent proofs of type preservation as terms of a GADT, which is not
+sound (as they admit in \S4.2) without a separate totality check because
+any type is trivially inhabited by a nonterminating term in Haskell.
+In contrast, our CPS transformations use simpler types than GADTs and
+assure type preservation at the (terminating) type level rather than
+the term level of the metalanguage.
+\Citeauthor{Guillemette-Monier-PLPV} review other
+type\hyp preserving CPS transformations (mainly in the context of typed intermediate
+languages), in particular
+\citets{shao-type-toplas} and \citets{chen-typeful}.
+These approaches use de Bruijn indices and fancier
+type systems with type-level functions, GADTs, or type\hyp equality
+proofs.
 

### 2007-07-27 tagless/tagless-final.tex
Emphasize simplity in contributions

@@
 To reiterate, we do \emph{not} propose any new language
-feature or new technique. We use features already present in
+feature or \ifshort new \else even any new programming \fi technique.
+\ifshort
+We
+\else
+Rather, we solve an open problem by a novel combination of simple types
+and existing techniques.  More precisely, we
+\fi
+use features already present in
 mainstream functional languages---Hindley-Milner
@@
 which have all appeared in the literature, to
-solve the problem that was stated in the published record as
+solve a problem that was stated in the published record as
 unsolved and likely unsolvable in ML or Haskell 98
 without extensions. The simplicity of our solution and its
-use of only mainstream features makes it more practical to build typed,
+use of only mainstream features \ifshort\else are virtues that \fi make it more practical to build typed,
 embedded DSLs.

### 2007-07-28 tagless/tagless-final.tex
small corrections

@@
 \end{code}
-In HOAS, the type of |lapp| is that of function application with
+In HOAS, we write $\lapp{e_1}{x}e_2$ as |lapp e1 (fun x -> e2)|;
+the type of |lapp| is that of function application with
 the two arguments swapped.  We can encode the term
@@
 |lapp|, |deref|, and |set| representing imperative features. The
-interpreter is still CBN, so evaluating |app ef ea| may not
-evaluate |ea|, but evaluating |lapp ea ef| does.
+interpreter is still CBN, so evaluating |app ef ea| might not
+evaluate |ea|, but evaluating |lapp ea ef| always does.
 For first-order state, such as of type~$\ZZ$, we 
@@
 \end{code}
+If the state is higher-order, then the types |state| and |'c states|
+are no longer be the same; furthermore, |'s states| is mutually
+recursive with the type |('c,'sv,'dv) repr|, as demonstrated in the
+accompanying source code.
+
 |RCPSI| matches the |Symantics| signature and implements the
@@
 
+It is worthwhile reiterating how close this implementation is to the CPS
+interpreter, and that adding state and imperative features needed no new
+techniques.
 We can also add mutable references to the object language using
 mutable references of the metalanguage, as shown in the accompanying
-code.  It is worthwhile reiterating how close this implementation is to the CPS
-interpreter, and that adding state and imperative features needed no new
-techniques.
-
+code.
 \fi

### 2007-07-28 tagless/tagless-final.tex
Sorry, an even more miniscule change

@@
 \end{code}
-If the state is higher-order, then the types |state| and |'c states|
-are no longer be the same; furthermore, |'s states| is mutually
+If the state has a higher-order type, then the types |state| and |'c states|
+are no longer the same, and |'s states| is mutually
 recursive with the type |('c,'sv,'dv) repr|, as demonstrated in the

### 2007-08-28 tagless/tagless-final.tex
Get paper back down to 16 pages as a baseline to start from (by deleting text that got added to the APLAS version even though it was not in the original). Add a lone footnote to address a referee comment [other referee issues will be done in a subsequent checkin]

@@
 a typed object language in a typed metalanguage as an interesting
-problem.  The known solutions to this problem store object terms and
+problem.  The common solutions to this problem store object terms and
 values in the metalanguage in a universal type, a generalized algebraic
@@
 \ifshort\else\UseVerbatim{2a}\fi
-We represent each variable using a unary de Bruijn index.
+We represent each variable using a unary de Bruijn index
+\footnote{We use de Bruijn indices here to make the comparison with
+\citep{WalidICFP02} simpler}.
 For example, we represent the object term $(\fun{x}x)\True$ as
@@
 We have programmed our interpreters in OCaml (and, for staging,
-MetaOCaml) and standard Haskell. The complete code is available at
-\url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
+MetaOCaml \cite{metaocaml}) and standard Haskell. The complete code is
+available at \url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
 to supplement the paper. Our examples below switch between (Meta)OCaml
@@
 \caption{A simple (Meta)OCaml embedding of our object language, 
-    and examples of use}
+    and examples}
 \label{fig:ocaml-simple}
@@
 
+\ifshort
+Our initial motivation came from several papers 
+\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}
+that use embedded interpreters to justify advanced
+type systems, in particular GADTs.
+\else
 Our initial motivation came from several papers that justify advanced
@@
 CPS transformations \citep{Guillemette-Monier-PLPV,shao-type-toplas,chen-typeful}.
+\fi
 We admire all this technical machinery, but
@@
 
+% The following paragraph was not in the version submitted to APLAS,
+% yet was not commented out.  Of course, if we had space, we should add
+% this, but it is difficult to see how.
+\ifshort\else
 Using Haskell, \citet{Guillemette-Monier-PLPV} implement a CPS transformation
@@
 proofs.
+\fi
 

### 2007-08-28 tagless/tagless-final.tex
Soften the claim of "open problem" since [32] (and [38]) really do offer solutions, it's just that the authors of those papers justifying GADTs apparently did not understand those papers/techniques well enough at the time they did their work! [and fix a typo in a comment]

@@
 Our code is surprisingly simple and obvious in hindsight, but
-it has been an open problem to
-interpret a typed object language in a typed metalanguage without
-tagging or type\hyp system extensions.  For example, \citet{taha-tag}
+it has been cited as a difficult problem (\cite{sumii-hybrid}
+notwithstanding) to interpret a typed object language in a typed metalanguage
+without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
 say that ``expressing such an interpreter in a statically typed
@@
 \jacques{But how 
-to you create, in either Haskell or MetaOCaml, an untypechecked 
+do you create, in either Haskell or MetaOCaml, an untypechecked 
 interpreter-with-a-hole [UIH] ?}

### 2007-08-28 tagless/tagless-final.tex
deal with a number of issues.  Make extensive comments in the file containing the reviews.

@@
 so that a partially evaluated term |P1 t| always contains a dynamic
-component and sometimes contains a static component.  The two
+component and sometimes contains a static component.  By 
+distributivity, the two
 alternative constructors of a |Maybe| value, |Just| and |Nothing|,
@@
 Any partial evaluator must decide how much to unfold recursion\ifshort.
-The code in Figure~\ref{fig:pe} na\"\i vely unfolds |fix|
+Our code na\"\i vely unfolds |fix|
 \else\
@@
 \fi
-whenever the
-argument is static.
+whenever the argument is static.
 \ifshort
@@
 \end{shortlist}
-While the logic is sound, we showed that the premise in the very first step
-is not valid.
+While the logic is sound, we (following \citet{yang-encoding}) showed that the
+premise in the very first step is not valid.
 

### 2007-08-31 tagless/tagless-final.tex
Minor changes (including, unfortunately, replacing \cite by \citep)

@@
 for embedding object language objects into the metalanguage
-(whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}),
+(whether using dependent types \citep{WalidICFP02} or not \citep{baars-typing}),
 and just enter them by hand.
@@
 We leave aside the solved problem of writing such a parser\slash type\hyp checker,
-whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}.
+whether using dependent types \citep{WalidICFP02} or not \citep{baars-typing}.
 \fi
@@
 \ifshort\else\UseVerbatim{2a}\fi
-We represent each variable using a unary de Bruijn index
-\footnote{We use de Bruijn indices here to make the comparison with
-\citep{WalidICFP02} simpler}.
+We represent each variable using a unary de Bruijn index.%
+\footnote{We use de Bruijn indices to simplify the comparison with
+\citearound{'s work}\citet{WalidICFP02}.}
 For example, we represent the object term $(\fun{x}x)\True$ as
@@
 
-\noindent Following \cite{WalidICFP02},
+\noindent Following \citep{WalidICFP02},
 we try to implement an interpreter function |eval0|\ifshort,
@@
 nonexhaustive pattern matching in |lookup| and |eval| and the tags |UB|
-and |UA| \cite[\S1.4]{WalidICFP02}.  In other words, the algebraic data
+and |UA| \citep[\S1.4]{WalidICFP02}.  In other words, the algebraic data
 types above fail to express in the metalanguage that the object program
@@
 pattern\hyp matching operations that incur a performance penalty in
-interpretation \cite{WalidICFP02} and impair optimality in partial evaluation
-\cite{taha-tag}.  In short, the universal\hyp type solution is
+interpretation \citep{WalidICFP02} and impair optimality in partial evaluation
+\citep{taha-tag}.  In short, the universal\hyp type solution is
 unsatisfactory because it does not preserve typing.
@@
 GADTs or dependent types \citep{taha-tag}.  In fact, this problem
-motivated much work on GADTs \cite{xi-guarded,peyton-jones-simple} and
-on dependent types \cite{WalidICFP02,fogarty-concoqtion}.
+motivated much work on GADTs \citep{xi-guarded,peyton-jones-simple} and
+on dependent types \citep{WalidICFP02,fogarty-concoqtion}.
 \ifshort\else
@@
 terms 
-\ifshort\cite{WalidPOPL03,NanevskiJFP05,DaviesJACM01}\else
+\ifshort\citep{WalidPOPL03,NanevskiJFP05,DaviesJACM01}\else
 like |test3|
-\cite{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual}\fi,
+\citep{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual}\fi,
 so that |lookup| never receives an empty environment.  We discuss these
@@
 Hindley-Milner \fi
-type system\ifshort\else\ \cite{hindley-principal,milner-theory}\fi
+type system\ifshort\else\ \citep{hindley-principal,milner-theory}\fi
 .  In this \emph{final} rather
@@
 not just to us but also to the metalanguage implementation, we avoid
-the serious performance penalty \cite{WalidICFP02} of error checking.
+the serious performance penalty \citep{WalidICFP02} of error checking.
 \Citet{Gluck-jones-optimality} explains deeper technical reasons that
@@
 We have programmed our interpreters in OCaml (and, for staging,
-MetaOCaml \cite{metaocaml}) and standard Haskell. The complete code is
+\citet{metaocaml}) and standard Haskell. The complete code is
 available at \url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
@@
 recently that some work on programming type-indexed values in ML
-\cite{yang-encoding} has given a hint of how such a function can be
+\citep{yang-encoding} has given a hint of how such a function can be
 expressed.''  We discuss related work in~\S\ref{related}.
@@
 We may claim some contribution about type-preserving CPS. Check
-related work section in \cite{Guillemette-Monier-PLPV}, especially
+related work section in \citep{Guillemette-Monier-PLPV}, especially
 check the work of Shao on type-preserving CPS in Flint. The PLPV paper
@@
 In contrast to \S\ref{intro}, we encode binding using higher-order
-abstract syntax (HOAS) \cite{miller-manipulating,pfenning-higher-order}
+abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
 rather than de Bruijn indices. This makes the encoding convenient and
@@
 the additional type parameter |'c|, an
-\emph{environment classifier} \cite{WalidPOPL03} required by MetaOCaml for
+\emph{environment classifier} \citep{WalidPOPL03} required by MetaOCaml for
 code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
@@
 into OCaml code using MetaOCaml's staging facilities. MetaOCaml
-represents future-stage expressions of type~$t$ at the
-present stage as values of type |('c, |$t$|) code| where |'c| is the
-environment classifier \cite{WalidPOPL03,calcagno-ml-like}. Code values are created
-by a \emph{bracket} form |.<|$e$|>.|, which specifies that the expression~$e$ is to be
-evaluated at a future stage. The \emph{escape} |.~|$e$ must occur
+represents future-stage expressions of type~$t$
+as values of type |('c, |$t$|) code|, where |'c| is the
+environment classifier \citep{WalidPOPL03,calcagno-ml-like}. Code values are created
+by a \emph{bracket} form |.<|$e$|>.|, which quotes the expression~$e$ for
+evaluation at a future stage. The \emph{escape} |.~|$e$ must occur
 within a bracket and specifies that the expression~$e$ must be evaluated
@@
 spliced into the code being built by the enclosing bracket. The \emph{run} form |.!|$e$ evaluates
-the future-stage code value~$e$ by compiling and linking it at run-time.
-The bracket, escape, and run are akin to
+the future-stage code value~$e$ by compiling and linking it at run time.
+Bracket, escape, and run are akin to
 quasi-quotation, unquotation, and |eval| of Lisp.
@@
 of inferred type |('c,|~|bool)| |C.repr|.  Interpreting |testpowfix7|
-\ifshort\pagebreak[3]\fi
 with
@@
 Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
+\ifshort\pagebreak[3]\fi
 This compiler does not incur
@@
 The resulting code obviously contains no tags and no pattern matching.
+The environment classifiers here, like the tuple types in \S\ref{ourapproach},
+make it a type error to run an open expression.
 \ifshort
@@
 function, of type |a -> C a|, to send a value to the future stage
-\cite{xi-guarded,WalidPOPL03}.\ifshort\else
+\citep{xi-guarded,WalidPOPL03}.\ifshort\else
 (By the way, if we
@@
 analysis.  To avoid needing polymorphic lift, we turn to
-Asai's technique \cite{asai-binding-time,sumii-hybrid}:
+Asai's technique \citep{asai-binding-time,sumii-hybrid}:
 build a dynamic term
@@
 GADTs and type-class functional dependencies
-\cite{oliveira-typecase}.  These
+\citep{oliveira-typecase}.  These
 methods are equivalent, and here we use GADTs; |incope1.hs|
@@
 mechanism for such features, inspired in part by previous work in
-generative linear algebra~\cite{CaretteKiselyov05}.
+generative linear algebra~\citep{CaretteKiselyov05}.
 \fi
@@
 Many sophisticated approaches have been developed to decide how much to unfold
-\cite{jones-partial}, but this issue is orthogonal to our presentation.
+\citep{jones-partial}, but this issue is orthogonal to our presentation.
 \ifshort\else
@@
 the residual program, which calls for let-insertion
-\cite{SwadiTahaKiselyovPasalic2006}.
+\citep{SwadiTahaKiselyovPasalic2006}.
 \fi
@@
 equivalent to the object application $\si\Encode{e}$
-\cite{jones-partial,taha-tag,Danvy-tagging-encoding}.
+\citep{jones-partial,taha-tag,Danvy-tagging-encoding}.
 A partial evaluator~$\pe$ maps object terms~$e$ to observationally
@@
 Even a simply typed $\lambda$-calculus obviously supports self\hyp
-interpretation, provided we use universal types \cite{taha-tag}.  The
+interpretation, provided we use universal types \citep{taha-tag}.  The
 ensuing tagging overhead motivated \citet{taha-tag} to propose tag
 elimination, which however does not statically guarantee that all tags
-will be removed \cite{WalidICFP02}.
+will be removed \citep{WalidICFP02}.
 
@@
 While the logic is sound, we (following \citet{yang-encoding}) showed that the
-premise in the very first step is not valid.
+first step's premise is not valid.
 
@@
 using an injection\hyp projection pair, as done for interpreters
-\ifshort \cite{Ramsey-ML-module-mania,Benton-embedded-interpreters}\else
+\ifshort \citep{Ramsey-ML-module-mania,Benton-embedded-interpreters}\else
 by \citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters}\fi,
-partial evaluation \ifshort \cite{Danvy-TDPE}\else by \citet{Danvy-TDPE}\fi,
-and type-level functions \ifshort \cite{oliveira-typecase}\else by
+partial evaluation \ifshort \citep{Danvy-TDPE}\else by \citet{Danvy-TDPE}\fi,
+and type-level functions \ifshort \citep{oliveira-typecase}\else by
 \citet{oliveira-typecase}\fi.  As explained in~\S\ref{S:PE-solution}, we
@@
 We strive for modularity by reusing interpreters for individual stages
-\cite{SperberThiemann:TwoForOne}.  It would be interesting
-to try to derive a \emph{cogen} \cite{Thiemann:cogeninsixlines}
+\citep{SperberThiemann:TwoForOne}.  It would be interesting
+to try to derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
 in the same manner.
@@
 ill-typed object terms, as done in Lava
-\cite{Lava} and by \citet{Rhiger-thesis}. However, these approaches
+\citep{Lava} and by \citet{Rhiger-thesis}. However, these approaches
 are not tagless because they still use universal types, such as Lava's
@@
 greatly complicates the soundness and completeness proofs of embedding
-\cite{Rhiger-thesis}, whereas our proofs are trivial.
+\citep{Rhiger-thesis}, whereas our proofs are trivial.
 Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
@@
 emulates type-indexed types and is related to intensional type analysis
-\cite{Morrisett-intensional,Generic-Haskell}. However, our object
+\citep{Morrisett-intensional,Generic-Haskell}. However, our object
 language and running examples in HOAS include |fix|,
 which intensional type analysis cannot handle
-\cite{xi-guarded}.  Our final approach
+\citep{xi-guarded}.  Our final approach
 seems related to \citearound{'s approach to HOAS using catamorphisms and
@@
 In particular, they use
-contexts with a hole \cite[p.\,13]{honsell99coinductive} to define
+contexts with a hole \citep[p.\,13]{honsell99coinductive} to define
 \emph{observational equivalence}
@@
 For the self\hyp interpreter, we then proceed to use a Church encoding for
-recursive data types \cite{bohm-automatic}.
+recursive data types \citep{bohm-automatic}.
 \fi
@@
 polymorphism in the separate compilation of modules
-\cite{shao-typed}.
+\citep{shao-typed}.
 

### 2007-08-31 tagless/tagless-final.tex
More addressing APLAS reviewer comments

@@
 optimality, we relate this exemplar of higher-rank and higher-kind
-polymorphism (provided by ML functors and Haskell~98 constructor
-classes) to plugging a term into a context of let\hyp polymorphic
+polymorphism \ifshort\else (provided by ML functors and Haskell~98 constructor
+classes) \fi to plugging a term into a context of let\hyp polymorphic
 bindings.
@@
 \ifshort\vspace*{-0.7em}\fi
+unlike applying |P1| to |Int| or |Bool|.
 That is, we need a nonparametric data type, something akin to
@@
 The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
-in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
+in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  \ifshort\else However,
 the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
-operand of type |P a -> P b| rather than |a -> b|.  We define a function
+operand of type |P a -> P b| rather than |a -> b|.  \fi We define a function
 \ifshort |abstr|, \fi like |abstr1| above\ifshort,\fi\
@@
 Although GADTs may indeed be more flexible and easier to use, they are
-unavailable in mainstream ML and implemented problematically
-in GHC currently. We also wanted to find the minimal
+unavailable in mainstream ML, and their current implementation in GHC fails to
+detect exhaustive pattern matching.  We also wanted to find the minimal
 set of widespread language features needed for tagless
@@
 acceptable to MetaOCaml.
+
 \Citet{sumii-hybrid} also use Asai's method,
@@
 and offline partial evaluation.
+They predate us in deforesting the object term representation to enable tagless
+partial evaluation.
 We strive for modularity by reusing interpreters for individual stages
-\citep{SperberThiemann:TwoForOne}.  It would be interesting
+\citep{SperberThiemann:TwoForOne}: our partial evaluator~|P|
+reuses our tagless evaluator~|R| and tagless compiler~|C|, so it
+is patent that the \emph{output} of~|P| never gets
+stuck.  It would be interesting
 to try to derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}

### 2007-09-01 tagless/tagless-final.tex
small changes, preparing for the change in Sec 4.1

@@
 % accordingly. Let the paper and the accompanying code be in sync.
-It is straightforward to create an algebraic data type, say in OCaml\ifshort\
-Fig.~\ref{fig:tag-problem}(a)\fi, to
+It is straightforward to create an algebraic data type, say in OCaml%
+\ifshort, Fig.~\ref{fig:tag-problem}(a)\fi, to
 represent object terms such as those in Figure~\ref{fig:object}.
@@
 constructor classes, as realized in ML and Haskell 98---and techniques
-which have all appeared in the literature, to
+which have all appeared in the literature (in particular, \cite{yang-encoding,sumii-hybrid}), to
 solve a problem that was stated in the published record as
@@
 Although GADTs may indeed be more flexible and easier to use, they are
-unavailable in mainstream ML, and their current implementation in GHC fails to
+unavailable in mainstream ML, and their current implementation in GHC
+6.6.1 fails to
 detect exhaustive pattern matching.  We also wanted to find the minimal

### 2007-09-01 tagless/tagless-final.tex
removing section 4.3; all code in Sec 4 is OCaml

@@
 
+%------------------------------------------------------------------------
+% This is the version for the APLAS paper. No Haskell
+\ifshort
 \section{A tagless partial evaluator}\label{PE}
-
+Surprisingly, we can write a partial evaluator using the idea above,
+namely to build object terms using ordinary functions rather than data
+constructors.  We present this partial evaluator in a sequence of four
+attempts. It uses no universal type and no tags
+for object types.  We then discuss residualization and binding-time
+analysis.  Our partial evaluator is a modular extension of the evaluator
+in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
+that it uses the former to reduce static terms and the latter to build
+dynamic terms.
+
+\subsection{Avoiding polymorphic lift}
+\label{S:PE-lift}
+
+Roughly, a partial evaluator interprets each object term to either
+a static (present-stage) term (using\ifshort\else\ the evaluator\fi~|R|) or
+a dynamic (future-stage) term (using\ifshort\else\ the
+compiler\fi~|C|). To
+distinguish between static and dynamic terms, one might define
+|repr| in the first version of the partial evaluator |P0| as
+\texttt{type ('c,'dv) repr 
+	= S0 of ('c,'dv) R.repr~}\Verb+|+\texttt{~E0 of ('c,'dv) C.repr}.
+Integer and boolean literals are immediate, present-stage
+values. Addition yields a static term (using~|R.add|) if and only 
+if both operands are static; otherwise we extract the dynamic terms 
+from the operands and add them using~|C.add|. We use |C.int| to
+convert from the static term |('c,int) R.repr|, which is just |int|,
+to the dynamic term.
+Whereas |mul| and |leq| are as easy to define as |add|, we encounter
+a problem with |if_|.  If the first argument to |if_| is a dynamic term
+(of type |('c,bool) C.repr|), the second a static term 
+(of type |('c,'a) R.repr|) and the third a
+dynamic term (of type |('c,'a) C.repr|), then we need to convert
+the static term to dynamic, but there is no polymorphic ``lift''
+function, of type |'a -> ('c,'a) C.repr|, to send a value to the future stage
+\citep{xi-guarded,WalidPOPL03}.
+
+Our |Symantics| only includes separate lifting methods |bool| and
+|int|, not a parametrically polymorphic lifting method, for good reason:
+When compiling to a first-order target language such as machine code,
+booleans, integers, and functions may well be represented differently.
+Thus, compiling polymorphic lift requires intensional type
+analysis.  To avoid needing polymorphic lift, we turn to
+Asai's technique \citep{asai-binding-time,sumii-hybrid}:
+build a dynamic term
+alongside every static term.
+
+\subsection{Delaying binding-time analysis}
+\label{S:PE-problem}
+
+We switch to the data type
+\texttt{type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr}
+so that a partially evaluated term always contains a dynamic
+component and sometimes contains a static component.  By 
+distributivity, the two
+alternative constructors of an |option| value, |Some| and |None|,
+tag each partially evaluated term with a phase: either present or
+future.  This tag is not an object type tag: all pattern matching below
+is exhaustive.  Because the future-stage component is always present, we
+can now define the polymorphic function
+|let abstr1 (P1 (_,dyn)) = dyn| of type
+\texttt{abstr1 : ('c,'dv) repr -> ('c,'dv) C.repr}
+to extract it without requiring polymorphic lift into~|C|.  We then try
+to define the interpreter |P1|---and get as far as the first-order
+constructs of our object language, including |if_|.
+\vspace*{-0.7em}
+\begin{code3}
+module P1 : Symantics = struct
+  let int  (x:int)  = P1 (Some (R.int x), C.int x)
+  let add e1 e2 = match (e1,e2) with
+    | (P1 (Some n1,_),P1 (Some n2,_)) -> int (R.add n1 n2)
+    | _ -> P1 (None,(C.add (abstr1 e1) (abstr1 e2)))
+  let if_ = function
+    | P1 (Some s,_) -> fun et ee -> if s then et () else ee ()
+    | eb -> fun et ee -> P1 (None, C.if_ (abstr1 eb) 
+                                   (fun () -> abstr1 (et ()))
+                                   (fun () -> abstr1 (ee ())))
+\end{code3}
+\vspace*{-0.7em}
+When we come to functions, however, we stumble.  According to our
+definition of~|P1|, a partially evaluated object function, such as the
+identity $\fun{x}x$ embedded in OCaml as |lam (fun x -> x)|\texttt{ :
+}|('c,'a->'a) P1.repr|, consists of a dynamic part 
+(type |('c,'a->'a) C.repr|) and
+maybe a static part (type |('c,'a->'a) R.repr|).  The dynamic part is useful
+when this function is passed to another function that is only
+dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
+useful when this function is applied to a static argument, as in
+$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
+evaluate the function, that is, compute as much as possible statically
+when it is applied to a mix of static and dynamic inputs.  For example,
+the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
+by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
+not statically known.  The same static function, applied to
+different static arguments, can give both static and dynamic results: we
+want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
+to~$x$.
+
+To enable these simplifications, we delay binding-time analysis
+for a static function until it is applied, that is, until |lam f|
+appears as the argument of |app|.  To do so, we have to incorporate |f|
+as it is into the |P1.repr| data structure: the representation
+for a function type |'a->'b| should be either
+\vspace*{-0.7em}
+\begin{code}
+S1 of ('c,'a) repr -> ('c,'b) repr | E1 of ('c,'a->'b) C.repr
+P1 of (('c,'a) repr -> ('c,'b) repr) option * ('c,'a->'b) C.repr
+\end{code}
+\vspace*{-0.7em}
+unlike |P1.repr| of |int| or |bool|.
+That is, we need a nonparametric data type, something akin to
+type-indexed functions and type-indexed types, which
+\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
+Thus, typed partial evaluation, like typed CPS transformation,
+inductively defines a map from source types to target types that
+performs case distinction on the source type. In Haskell, typecase
+can be equivalently implemented either with GADTs or 
+type-class functional dependencies
+\citep{oliveira-typecase}. The accompanying code shows both
+approaches, neither of which is portable to OCaml. In addition,
+the problem of nonexhaustive pattern\hyp matching reappears in
+in the GADT approach because GHC 6.6.1 cannot see that a particular
+type of a GADT value precludes certain constructors.
+Recall, our
+goal is to make this exhaustiveness \emph{syntactically} apparent.
+
+\subsection{The ``final'' solution}
+\label{S:PE-solution}
+
+Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
+would ideally like is to write
+\texttt{type ('c,'dv) repr = P1 of repr\_pe ('c,'dv) R.repr option * ('c,'dv) C.repr}
+where |repr_pe| is the type function defined
+% inductively because P below depends on repr_pe
+by
+\vspace*{-0.8em}
+\begin{code3}
+repr_pe ('c,int) = ('c,int); repr_pe ('c,bool) = ('c,bool)
+repr_pe ('c,'a->'b) = ('c,'a) repr -> ('c,'b) repr
+\end{code3}
+\vspace*{-0.8em}
+Although we can use type classes to define this type function
+in Haskell, that is not portable to MetaOCaml. However,
+these three typecase alternatives are already present in existing
+methods of |Symantics|.
+A simple and portable solution thus emerges: we bake |repr_pe| 
+into the signature |Symantics|. 
+We recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
+constructor took two arguments |'c| and~|'dv|. We add an argument
+|'sv| for the result of applying |repr_pe| to~|'dv|.
+Figure~\ref{fig:ocaml} shows the new signature.
+% aplas: abbreviated figure
+\begin{figure}
+\begin{floatrule}
+\ifshort
+\begin{code2}
+module type Symantics = sig type ('c,'sv,'dv) repr
+  val int : int -> ('c,int,int) repr
+  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+            -> ('c,'x,'da -> 'db) repr
+  val app : ('c,'x,'da -> 'db) repr
+            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
+                                'da -> 'db) repr as 'x)
+  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val if_ : ('c,bool,bool) repr
+            -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x) end
+\end{code2}
+\else
+\begin{code2}
+module type Symantics = sig
+  type ('c,'sv,'dv) repr
+
+  val int : int  -> ('c,int,int) repr
+  val bool: bool -> ('c,bool,bool) repr
+
+  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+            -> ('c,'x,'da -> 'db) repr
+  val app : ('c,'x,'da -> 'db) repr
+            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
+  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
+                                'da -> 'db) repr as 'x)
+
+  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
+  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
+  val if_ : ('c,bool,bool) repr
+            -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
+end
+\end{code2}
+\fi
+\end{floatrule}
+\caption{A (Meta)OCaml embedding of our object language that supports
+  partial evaluation \ifshort \protect\linebreak[2] (\texttt{bool},
+  \texttt{mul}, \texttt{leq} are elided)\fi}
+\label{fig:ocaml}
+\end{figure}
+
+\begin{figure}
+\begin{floatrule}
+\ifshort
+\begin{code2}
+module P = struct
+  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
+  let abstr {dy = x} = x    let pdyn x = {st = None; dy = x}
+
+  let int  (x:int)  = {st = Some (R.int x);  dy = C.int x}
+  let add e1 e2 = match e1, e2 with
+  | {st = Some 0}, e | e, {st = Some 0} -> e
+  | {st = Some m}, {st = Some n} -> int (R.add m n)
+  | _ -> pdyn (C.add (abstr e1) (abstr e2))
+  let if_ eb et ee = match eb with
+  | {st = Some b} -> if b then et () else ee ()
+  | _ -> pdyn (C.if_ (abstr eb) (fun () -> abstr (et ()))
+                                (fun () -> abstr (ee ())))
+  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
+  let app ef ea = match ef with {st = Some f} -> f ea
+                  | _ -> pdyn (C.app (abstr ef) (abstr ea)) end
+\end{code2}
+\else
+\begin{code2}[commandchars=\@\[\]]
+module P = struct
+  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
+  let abstr {dy = x} = x
+  let pdyn x = {st = None; dy = x}
+
+  let int  (x:int ) = {st = Some (R.int  x); dy = C.int  x}
+  let bool (x:bool) = {st = Some (R.bool x); dy = C.bool x}
+
+  let add e1 e2 = match e1, e2 with
+                  | {st = Some 0}, e | e, {st = Some 0} -> e
+                  | {st = Some m}, {st = Some n} -> int (R.add m n)
+                  | _ -> pdyn (C.add (abstr e1) (abstr e2))
+  let if_ eb et ee = match eb with
+                     | {st = Some b} -> if b then et () else ee ()
+                     | _ -> pdyn (C.if_ (abstr eb) (fun () @!->@! abstr (et @!()))
+                                                   (fun () @!->@! abstr (ee @!())))
+
+  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
+  let app ef ea = match ef with
+                  | {st = Some f} -> f ea
+                  | _ -> pdyn (C.app (abstr ef) (abstr ea))
+  let fix f = let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
+              in let rec self = function
+                                | {st = Some @!_} as e -> app (f (lam self)) e
+                                | e -> pdyn (C.app fdyn (abstr e))
+                 in {st = Some self; dy = fdyn}
+end
+\end{code2}
+\fi
+\end{floatrule}
+\caption{Our partial evaluator (\ifshort \texttt{bool}, \texttt{mul},
+  \texttt{leq} and \texttt{fix} \else \texttt{mul} and \texttt{leq} \fi
+  are elided)}
+\label{fig:pe}
+\end{figure}
+
+The interpreters |R|, |L| and~|C| above only use the old
+type arguments |'c| and~|'dv|, which are treated by the new signature
+in the same way.  Hence, all that needs to change in these interpreters
+to match the new signature is to add a phantom type
+argument~|'sv| to~|repr|.
+For example, the compiler |C| now begins
+\ifshort
+\texttt{module C = struct
+  type ('c,'sv,'dv) repr = ('c,'dv) code}
+\else
+\begin{code}
+module C = struct
+  type ('c,'sv,'dv) repr = ('c,'dv) code
+\end{code}
+\fi
+with the rest the same.
+In contrast, the partial evaluator~|P| relies on the type argument |'sv|.
+
+Figure~\ref{fig:pe} shows the partial evaluator~|P|.
+Its type |repr| literally expresses the type equation for |repr_pe| above.
+The function |abstr|
+extracts a future-stage code value from the result of
+partial evaluation.  Conversely, the function |pdyn| injects a
+code value into the |repr| type. As
+in~\S\ref{S:PE-problem}, we build dynamic terms alongside
+any static ones to avoid polymorphic lift.
+
+The static portion of the interpretation of |lam f| is |Some f|, 
+which just wraps the HOAS
+function |f|. The interpretation of |app ef ea| 
+checks to see if |ef| has such a wrapped
+HOAS function. If it is, we apply |f| to the
+concrete argument |ea|, giving us a chance to perform static
+computations (see example the below). If
+|ef| has only a dynamic part, we residualize.
+
+To illustrate how to add optimizations, we improve |add| (and |mul|,
+elided) to simplify the generated code using the monoid (and ring)
+structure of~|int|: not only is addition performed statically
+(using~|R|) when both operands are statically known, but it is
+eliminated when one operand is statically~$0$; similarly for
+multiplication by~$0$ or~$1$.  
+\ifshort
+Such optimizations can be quite effective
+\else
+Such algebraic simplifications are easy
+to abstract over the specific domain (such as monoid or ring) where they
+apply.  These simplifications and abstractions help a lot
+\fi
+in a large language with more base types and primitive operations.
+\ifshort\else
+Incidentally, the code actually contains a more general implementation
+mechanism for such features, inspired in part by previous work in
+generative linear algebra~\citep{CaretteKiselyov05}.
+\fi
+
+Any partial evaluator must decide how much to unfold recursion\ifshort.
+Our code na\"\i vely unfolds |fix|
+\else\
+statically: unfolding too little can degrade the residual code, whereas
+unfolding too much risks nontermination.  Our partial evaluator is no
+exception, because our object language includes |fix|.  The code in
+Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
+way'', that is, unfold |fix| rather than residualize
+\fi
+whenever the argument is static.
+\ifshort
+In the accompanying source code is a conservative
+alternative |P.fix| that unfolds recursion only once, then residualizes.
+\else
+A conservative alternative is to unfold recursion only once, then residualize:
+\begin{code}
+let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
+\end{code}
+\fi
+Many sophisticated approaches have been developed to decide how much to unfold
+\citep{jones-partial}, but this issue is orthogonal to our presentation.
+\ifshort\else
+A separate concern in our treatment of |fix| is possible code bloat in
+the residual program, which calls for let-insertion
+\citep{SwadiTahaKiselyovPasalic2006}.
+\fi
+
+Given this implementation of~|P|, our running example
+\ifshort
+\texttt{let module E = EX(P) in E.test1 ()}
+\else
+\begin{code}
+let module E = EX(P) in E.test1 ()
+\end{code}
+\fi
+evaluates to
+\ifshort
+|{P.st = Some true; P.dy = .<true>.}|
+\else
+\begin{code}
+{P.st = Some true; P.dy = .<true>.}
+\end{code}
+\fi
+of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
+a $\beta$-reduction has been statically performed to yield |true|.  More
+interestingly, whereas |testpowfix7| compiles to a code value with many
+$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
+\ifshort
+\texttt{let module E = EX(P) in E.testpowfix7}
+\else
+\begin{code}
+let module E = EX(P) in E.testpowfix7
+\end{code}
+\fi
+gives the desired result
+\ifshort\vspace*{-0.7em}\fi
+\begin{code}
+{P.st = Some <fun>;
+ P.dy = .<fun x -> x * (x * (x * (x * (x * (x * x)))))>.}
+\end{code}
+\ifshort\vspace*{-0.7em}\fi
+
+\noindent All pattern\hyp
+matching in~|P| is \emph{syntactically} exhaustive, so it is patent to the
+metalanguage implementation that |P| never gets stuck.  Further, all
+pattern\hyp matching occurs
+during partial evaluation, only to check if a value is known statically,
+never what type it has.  In other words, our partial evaluator tags
+phases (with |Some| and |None|) but not object types.
+
+%------------------------------------------------------------------------
+% This is the longer version of the section
+\else
+
+\section{A tagless partial evaluator}\label{PE}
 Surprisingly, we can write a partial evaluator using the idea above,
@@
 
-\ifshort\else
 Our typed partial evaluator is online and polyvariant.  It reuses the

### 2007-09-01 tagless/tagless-final.tex
minor edits to clarify the text and playing around with spacing here and there

@@
 available at \url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
-to supplement the paper. Our examples below switch between (Meta)OCaml
+to supplement the paper. 
+\ifshort For simplicity, our main examples will be in MetaOCaml, but
+all examples have also been implemented in Haskell.
+\else Our examples below switch between (Meta)OCaml
 and Haskell even though we have implemented each example equivalently in
@@
 metalanguage than the other.  For example, MetaOCaml provides
-convenient, typed staging facilities.
+convenient, typed staging facilities.\fi
 
@@
 rather than de Bruijn indices. This makes the encoding convenient and
-also ensures that our object programs are closed.
+ensures that our object programs are closed.
 
@@
 restriction, to keep the type of |testpowfix| and |testpowfix7|
-polymorphic in |repr|. Instead of supplying this dummy
+polymorphic in |repr|. \ifshort\else Instead of supplying this dummy
 argument, we could have given the terms explicit polymorphic
@@
 avoid the dummy argument by switching off the monomorphism restriction
-with a compiler flag.
+with a compiler flag.\fi
 The methods |add|, |mul|, and |leq| are quite similar, and so are
@@
 \ifshort
-\vspace*{-0.8em}\begin{code3}
+\vspace*{-0.6em}\begin{code3}
 module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
@@
 \end{code3}
-\vspace*{-0.8em}
+\vspace*{-0.6em}
 \else
@@
 Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
-\ifshort\pagebreak[3]\fi
+%\ifshort\pagebreak[3]\fi
 This compiler does not incur
@@
 constructs of our object language, including |if_|.
-\vspace*{-0.7em}
+\vspace*{-0.6em}
 \begin{code3}
@@
 \end{code3}
-\vspace*{-0.7em}
+\vspace*{-0.6em}
 When we come to functions, however, we stumble.  According to our
@@
 for a function type |'a->'b| should be either
-\vspace*{-0.7em}
+\vspace*{-0.6em}
 \begin{code}
@@
 \end{code}
-\vspace*{-0.7em}
+\vspace*{-0.6em}
 unlike |P1.repr| of |int| or |bool|.
@@
 by
-\vspace*{-0.8em}
+\vspace*{-0.6em}
 \begin{code3}
@@
 \end{code3}
-\vspace*{-0.8em}
+\vspace*{-0.6em}
 Although we can use type classes to define this type function
@@
 \ifshort
-\vspace*{-0.7em}
+\vspace*{-0.6em}
 \begin{code3}
@@
 \end{code3}
-\vspace*{-0.7em}
+\vspace*{-0.6em}
 \else
@@
 |P1| to a function type |a -> b| should yield one of
-\ifshort\vspace*{-0.7em}\fi
+\ifshort\vspace*{-0.6em}\fi
 \begin{code}
@@
 \end{code}
-\ifshort\vspace*{-0.7em}\fi
+\ifshort\vspace*{-0.6em}\fi
 unlike applying |P1| to |Int| or |Bool|.
@@
 by
-\ifshort\vspace*{-0.8em}
+\ifshort\vspace*{-0.6em}
 \begin{code3}
@@
 these motivating examples do not need it.
-Although GADTs may indeed be more flexible and easier to use, they are
-unavailable in mainstream ML, and their current implementation in GHC
+Although GADTs may indeed be simpler and more flexible, they are
+unavailable in mainstream ML, and their implementation in GHC
 6.6.1 fails to

### 2007-09-04 tagless/tagless-final.tex
small changes

@@
 to supplement the paper. 
-\ifshort For simplicity, our main examples will be in MetaOCaml, but
+\ifshort For simplicity, main examples in the paper will be in MetaOCaml;
 all examples have also been implemented in Haskell.
@@
 the problem of nonexhaustive pattern\hyp matching reappears in
-in the GADT approach because GHC 6.6.1 cannot see that a particular
+the GADT approach because GHC 6.6.1 cannot see that a particular
 type of a GADT value precludes certain constructors.
@@
 function |f|. The interpretation of |app ef ea| 
-checks to see if |ef| has such a wrapped
+checks to see if |ef| is such a wrapped
 HOAS function. If it is, we apply |f| to the
 concrete argument |ea|, giving us a chance to perform static
-computations (see example the below). If
+computations (see the example below). If
 |ef| has only a dynamic part, we residualize.

### 2007-09-04 tagless/tagless-final.tex
Small final changes to the APLAS version, to use the newfound breathing room

@@
 
+\smallskip
 (b) \BUseVerbatim{2b}
 
+\smallskip
 (c) \BUseVerbatim{2c}
 
+\smallskip
 (d) \BUseVerbatim{2d}
-\caption{OCaml code illustrating the tag problem.}
+
+\medskip
+\caption{OCaml code illustrating the tag problem}
 \label{fig:tag-problem}
@@
 avoid the dummy argument by switching off the monomorphism restriction
-with a compiler flag.\fi
+with a compiler flag. \fi
 The methods |add|, |mul|, and |leq| are quite similar, and so are
@@
 
+\smallskip
 (b) \BUseVerbatim{3b}
 
+\smallskip
 (c) \BUseVerbatim{3c}
+
+\medskip
 \caption{Symantics in Haskell}
@@
 \begin{figure}[t]
-\begin{floatrule}
 \begin{tabular}{@{}l@{}}
 \ifx\relax\normalbaselineskip\else\baselineskip\normalbaselineskip\fi
-\BUseVerbatim[baseline=b]{ocaml-simple}\\[\medskipamount]
+\BUseVerbatim[baseline=b]{ocaml-simple}\\[\smallskipamount]
 \ifx\relax\normalbaselineskip\else\baselineskip\normalbaselineskip\fi
@@
 \end{tabular}
-\end{floatrule}
+
+\medskip
 \caption{A simple (Meta)OCaml embedding of our object language, 
@@
 the definitions of our object terms must look syntactically like
-values. Alternatively, we could have used
-the rank-2 record types of OCaml to maintain the necessary polymorphism.
-
+values. (Alternatively, we could have used
+the rank-2 record types of OCaml to maintain the necessary polymorphism.)
+\ifshort\else\par\fi
 Thus, we represent an object expression in
@@
 \ifshort
-\vspace*{-0.6em}\begin{code3}
+\begin{code3}
 module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
@@
 \end{code3}
-\vspace*{-0.6em}
 \else
@@
 
+\smallskip
 (b) \BUseVerbatim{5b}
 
+\smallskip
 (c) \BUseVerbatim{5c}
 
+\smallskip
 (d) \BUseVerbatim{5d}
-\caption{The tagless staged interpreter \texttt{C}.}
+
+\medskip
+\caption{The tagless staged interpreter \texttt{C}}
 \label{fig:interpreter-C}
@@
 Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
-%\ifshort\pagebreak[3]\fi
 This compiler does not incur
@@
 namely to build object terms using ordinary functions rather than data
-constructors.  We present this partial evaluator in a sequence of four
+constructors.  We present this partial evaluator in a sequence of three
 attempts. It uses no universal type and no tags
@@
 
-Roughly, a partial evaluator interprets each object term to either
+Roughly, a partial evaluator interprets each object term to yield either
 a static (present-stage) term (using\ifshort\else\ the evaluator\fi~|R|) or
@@
 compiler\fi~|C|). To
-distinguish between static and dynamic terms, one might define
-|repr| in the first version of the partial evaluator |P0| as
-\texttt{type ('c,'dv) repr 
-	= S0 of ('c,'dv) R.repr~}\Verb+|+\texttt{~E0 of ('c,'dv) C.repr}.
+distinguish between static and dynamic terms, we might try to define
+|repr| in the partial evaluator as
+\texttt{type ('c,'dv)\,repr 
+	= S0 of ('c,'dv)\,R.repr~}\Verb+|+\texttt{~E0 of ('c,'dv)\,C.repr}.
 Integer and boolean literals are immediate, present-stage
@@
 to the dynamic term.
+
 Whereas |mul| and |leq| are as easy to define as |add|, we encounter
-a problem with |if_|.  If the first argument to |if_| is a dynamic term
+a problem with |if_|.  Suppose that the first argument to |if_| is a dynamic term
 (of type |('c,bool) C.repr|), the second a static term 
-(of type |('c,'a) R.repr|) and the third a
-dynamic term (of type |('c,'a) C.repr|), then we need to convert
+(of type |('c,'a) R.repr|), and the third a
+dynamic term (of type |('c,'a) C.repr|). We then need to convert
 the static term to dynamic, but there is no polymorphic ``lift''
@@
 |let abstr1 (P1 (_,dyn)) = dyn| of type
-\texttt{abstr1 : ('c,'dv) repr -> ('c,'dv) C.repr}
+\texttt{('c,'dv) repr -> ('c,'dv) C.repr}
 to extract it without requiring polymorphic lift into~|C|.  We then try
@@
 constructs of our object language, including |if_|.
-\vspace*{-0.6em}
 \begin{code3}
@@
 \end{code3}
-\vspace*{-0.6em}
-When we come to functions, however, we stumble.  According to our
+However, we stumble on functions.  According to our
 definition of~|P1|, a partially evaluated object function, such as the
@@
 as it is into the |P1.repr| data structure: the representation
-for a function type |'a->'b| should be either
-\vspace*{-0.6em}
-\begin{code}
+for a function type |'a->'b| should be one of
+\begin{code3}
 S1 of ('c,'a) repr -> ('c,'b) repr | E1 of ('c,'a->'b) C.repr
 P1 of (('c,'a) repr -> ('c,'b) repr) option * ('c,'a->'b) C.repr
-\end{code}
-\vspace*{-0.6em}
+\end{code3}
 unlike |P1.repr| of |int| or |bool|.
@@
 performs case distinction on the source type. In Haskell, typecase
-can be equivalently implemented either with GADTs or 
+can be equivalently implemented either with GADTs or with
 type-class functional dependencies
 \citep{oliveira-typecase}. The accompanying code shows both
-approaches, neither of which is portable to OCaml. In addition,
+approaches, neither portable to OCaml. In addition,
 the problem of nonexhaustive pattern\hyp matching reappears in
@@
 type of a GADT value precludes certain constructors.
-Recall, our
-goal is to make this exhaustiveness \emph{syntactically} apparent.
+Thus GADTs fail to
+make it \emph{syntactically} apparent that pattern matching is exhaustive.
 
@@
 would ideally like is to write
-\texttt{type ('c,'dv) repr = P1 of repr\_pe ('c,'dv) R.repr option * ('c,'dv) C.repr}
+\begingroup\sloppy
+\texttt{type ('c,'dv) repr = P1 of (repr\_pe ('c,'dv)) R.repr option * ('c,'dv) C.repr}
 where |repr_pe| is the type function defined
@@
 by
-\vspace*{-0.6em}
 \begin{code3}
@@
 \end{code3}
-\vspace*{-0.6em}
+\endgroup\noindent
 Although we can use type classes to define this type function
@@
 \ifshort
+\smallskip
 \begin{code2}
@@
 \end{code2}
+\medskip
 \else
@@
 \ifshort
-\texttt{let module E = EX(P) in E.test1 ()}
+\texttt{let module E\,=\,EX(P) in E.test1 ()}
 \else

### 2007-09-05 tagless/final_code/incope.hs
code to answer Adam Chlipala question about the type of the parser.

@@ -745,3 +745,33 @@ test3_2 = unRR (test3_1)
 
 test3_3 :: (Symantics repr) => repr ((Int->Int)->Int)
 test3_3 = unRR (test3_1)
+
+-- ------------------------------------------------------------------------
+-- Representing parsed AST terms
+--
+-- The following DynTerm repr could well be the result of the
+-- parser/typechecker (created along the lines of `typing dynamic typing')
+-- that reads object terms from a string or a file.
+-- The type of the term is abstracted over; however, the parser does not commit
+-- to any interpretation of that term. As the following shows, we can
+-- interpret the same `dynamic' term in multiple ways.
+-- This test was inspired by the discussion with Adam Chlipala
+-- on Lambda-the-Ultimate on Sep 4-5, 2007.
+
+-- The Show constraint here is so that we could see the result of the
+-- computation...
+data DynTerm repr = forall a. Show a => DynTerm (repr a)
+
+-- A sample `dynamic' object term
+dynterm1 () = DynTerm (lam (\x -> x) `app` (bool True))
+
+-- Examples of several interpretations, using R, L, C, and P interpreters
+dynterm_R_show = case dynterm1 () of DynTerm t -> show $ compR t
+-- "True"
+dynterm_L_show = case dynterm1 () of DynTerm t -> show $ compL t
+-- "3"
+dynterm_C_show = case dynterm1 () of DynTerm t -> show $ compC t
+-- "((\\V0 -> V0) True)"
+dynterm_P_show = case dynterm1 () of DynTerm t -> show $ compP t
+-- "True"
+

### 2007-09-05 tagless/incope.hs

@@ -949,3 +949,33 @@ test3_2 = unRR (test3_1)
 
 test3_3 :: (Symantics repr) => repr ((Int->Int)->Int)
 test3_3 = unRR (test3_1)
+
+-- ------------------------------------------------------------------------
+-- Representing parsed AST terms
+--
+-- The following DynTerm repr could well be the result of the
+-- parser/typechecker (created along the lines of `typing dynamic typing')
+-- that reads object terms from a string or a file.
+-- The type of the term is abstracted over; however, the parser does not commit
+-- to any interpretation of that term. As the following shows, we can
+-- interpret the same `dynamic' term in multiple ways.
+-- This test was inspired by the discussion with Adam Chlipala
+-- on Lambda-the-Ultimate on Sep 4-5, 2007.
+
+-- The Show constraint here is so that we could see the result of the
+-- computation...
+data DynTerm repr = forall a. Show a => DynTerm (repr a)
+
+-- A sample `dynamic' object term
+dynterm1 () = DynTerm (lam (\x -> x) `app` (bool True))
+
+-- Examples of several interpretations, using R, L, C, and P interpreters
+dynterm_R_show = case dynterm1 () of DynTerm t -> show $ compR t
+-- "True"
+dynterm_L_show = case dynterm1 () of DynTerm t -> show $ compL t
+-- "3"
+dynterm_C_show = case dynterm1 () of DynTerm t -> show $ compC t
+-- "((\\V0 -> V0) True)"
+dynterm_P_show = case dynterm1 () of DynTerm t -> show $ compP t
+-- "True"
+

### 2007-11-14 tagless/incope.hs
typechecking experiments

@@ -4,6 +4,8 @@
 
 module Incope where
 
+import Data.Typeable  -- used ONLY for typechecking below
+
 {-
   The language is simply-typed lambda-calculus with fixpoint,
   integers, booleans and comparison.
@@ -964,7 +966,7 @@ test3_3 = unRR (test3_1)
 
 -- The Show constraint here is so that we could see the result of the
 -- computation...
-data DynTerm repr = forall a. Show a => DynTerm (repr a)
+data DynTerm repr = forall a. (Show a, Typeable a) => DynTerm (repr a)
 
 -- A sample `dynamic' object term
 dynterm1 () = DynTerm (lam (\x -> x) `app` (bool True))
@@ -978,6 +980,63 @@ dynterm_C_show = case dynterm1 () of DynTerm t -> show $ compC t
 -- "((\\V0 -> V0) True)"
 dynterm_P_show = case dynterm1 () of DynTerm t -> show $ compP t
 -- "True"
+
+
+data UExp = UInt Int | UBool Bool | UAdd UExp UExp | UIf UExp UExp UExp
+
+-- Typechecking is obviously partial
+-- Rather than use the Either data type to report errors (Error monad), 
+-- we report errors via 'error' for simplicity.
+typecheck :: Symantics repr => UExp -> DynTerm repr
+typecheck (UInt x) = DynTerm (int x)
+typecheck (UBool x) = DynTerm (bool x)
+typecheck (UAdd e1 e2) = tadd (typecheck e1) (typecheck e2)
+ where
+ tadd (DynTerm e1) (DynTerm e2) = DynTerm (add e1' e2')
+  where
+  Just e1' = gcast e1
+  Just e2' = gcast e2
+typecheck (UIf e1 e2 e3) = tif (typecheck e1) (typecheck e2) (typecheck e3)
+ where
+ tif (DynTerm e1) (DynTerm e2) (DynTerm e3) = DynTerm (if_ e1' e2 e3')
+  where
+  Just e1' = gcast e1
+  Just e3' = gcast e3
+
+
+rShow (DynTerm t) = show $ compR t
+lShow (DynTerm t) = show $ compL t
+cShow (DynTerm t) = show $ compC t
+
+dt1 () = typecheck (UAdd (UInt 1) (UInt 2))
+
+dt1rs = rShow $ dt1 () -- "3"
+dt1rc = cShow $ dt1 () -- "(1 + 2)"
+
[…]

### 2007-11-14 tagless/Incope.hs
Type-checking

@@ -4,8 +4,6 @@
 
 module Incope where
 
-import Data.Typeable  -- used ONLY for typechecking below
-
 {-
   The language is simply-typed lambda-calculus with fixpoint,
   integers, booleans and comparison.
@@ -952,91 +950,6 @@ test3_2 = unRR (test3_1)
 test3_3 :: (Symantics repr) => repr ((Int->Int)->Int)
 test3_3 = unRR (test3_1)
 
--- ------------------------------------------------------------------------
--- Representing parsed AST terms
---
--- The following DynTerm repr could well be the result of the
--- parser/typechecker (created along the lines of `typing dynamic typing')
--- that reads object terms from a string or a file.
--- The type of the term is abstracted over; however, the parser does not commit
--- to any interpretation of that term. As the following shows, we can
--- interpret the same `dynamic' term in multiple ways.
--- This test was inspired by the discussion with Adam Chlipala
--- on Lambda-the-Ultimate on Sep 4-5, 2007.
-
--- The Show constraint here is so that we could see the result of the
--- computation...
-data DynTerm repr = forall a. (Show a, Typeable a) => DynTerm (repr a)
-
--- A sample `dynamic' object term
-dynterm1 () = DynTerm (lam (\x -> x) `app` (bool True))
-
--- Examples of several interpretations, using R, L, C, and P interpreters
-dynterm_R_show = case dynterm1 () of DynTerm t -> show $ compR t
--- "True"
-dynterm_L_show = case dynterm1 () of DynTerm t -> show $ compL t
--- "3"
-dynterm_C_show = case dynterm1 () of DynTerm t -> show $ compC t
--- "((\\V0 -> V0) True)"
-dynterm_P_show = case dynterm1 () of DynTerm t -> show $ compP t
--- "True"
-
-
-data UExp = UInt Int | UBool Bool | UAdd UExp UExp | UIf UExp UExp UExp
-
--- Typechecking is obviously partial
--- Rather than use the Either data type to report errors (Error monad), 
--- we report errors via 'error' for simplicity.
-typecheck :: Symantics repr => UExp -> DynTerm repr
-typecheck (UInt x) = DynTerm (int x)
[…]

### 2007-11-14 tagless/IncopeTypecheck.hs

@@ -1,0 +1,154 @@
+{-# OPTIONS -fglasgow-exts -W #-}
+
+-- Interpreter, Compiler, Partial Evaluator 
+-- Typechecking into Symantics terms
+
+module IncopeTypecheck where
+
+import Incope
+import Data.Typeable
+import Control.Monad
+import Control.Monad.Error
+import Text.Show.Functions
+
+-- Representing parsed AST terms
+--
+-- The following DynTerm repr could well be the result of the
+-- parser/typechecker (created along the lines of `typing dynamic typing')
+-- that reads object terms from a string or a file.
+-- The type of the term is abstracted over; however, the parser does not commit
+-- to any interpretation of that term. As the following shows, we can
+-- interpret the same `dynamic' term in multiple ways.
+-- This test was inspired by the discussion with Adam Chlipala
+-- on Lambda-the-Ultimate on Sep 4-5, 2007.
+
+-- The Show constraint here is so that we could see the result of the
+-- computation...
+data DynTerm repr = forall a. (Show a, Typeable a) => DynTerm (repr a)
+
+-- A sample `dynamic' object term
+dynterm1 () = DynTerm (lam (\x -> x) `app` (bool True))
+
+-- Examples of several interpretations, using R, L, C, and P interpreters
+dynterm_R_show = case dynterm1 () of DynTerm t -> show $ compR t
+-- "True"
+dynterm_L_show = case dynterm1 () of DynTerm t -> show $ compL t
+-- "3"
+dynterm_C_show = case dynterm1 () of DynTerm t -> show $ compC t
+-- "((\\V0 -> V0) True)"
+dynterm_P_show = case dynterm1 () of DynTerm t -> show $ compP t
+-- "True"
+
+-- Typechecking
+-- Representing unparsed terms
+data UExp = UInt Int | UBool Bool | UAdd UExp UExp | UIf UExp UExp UExp
+	  | UApp UExp UExp
+	    -- functions are represented via combinators
+	  | UK UExp UExp -- K combinator, args give the types of args
[…]

### 2007-11-15 tagless/IncopeTypecheck.hs
small beautifications.

@@ -3,6 +3,11 @@
 -- Interpreter, Compiler, Partial Evaluator 
 -- Typechecking into Symantics terms
 
+-- The code does show how to convert untyped terms to 
+-- typed terms that can be interpreted by various typed interpreters. 
+-- Typechecking happens only once, regardless of the number of 
+-- interpretations of the term.
+
 module IncopeTypecheck where
 
 import Incope
@@ -40,7 +45,7 @@ dynterm_P_show = case dynterm1 () of DynTerm t -> show $ compP t
 -- "True"
 
 -- Typechecking
--- Representing unparsed terms
+-- Representing untyped terms (the AST produced by the parser, for example)
 data UExp = UInt Int | UBool Bool | UAdd UExp UExp | UIf UExp UExp UExp
 	  | UApp UExp UExp
 	    -- functions are represented via combinators
@@ -105,6 +110,7 @@ tc exp = either error id (typecheck exp)
 rShow (DynTerm t) = show $ compR t
 lShow (DynTerm t) = show $ compL t
 cShow (DynTerm t) = show $ compC t
+pShow (DynTerm t) = show $ compP t
 
 
 dt1:: Symantics repr => DynTerm repr
@@ -145,6 +151,7 @@ dt5 = tc (((ut2 `UApp` ut) `UApp` (UBool False)) `UApp` (UInt 20))
 dt5rl = lShow $ dt5  -- 9
 dt5rc = cShow $ dt5  
   -- "((((\\V0 -> (\\V1 -> V0)) (\\V2 -> (\\V3 -> V2))) False) 20)"
+dt5rp = pShow $ dt5  -- "(\\V0 -> 20)"
 
 dt6:: Symantics repr => DynTerm repr
 dt6 = tc (((ut2 `UApp` ut2) `UApp` (UBool False)) `UApp` (UInt 20))
@@ -152,3 +159,5 @@ dt6rl = lShow $ dt6
 -- "*** Exception: Bad App: UApp (UK (UK (UInt 1) (UBool True)) (UBool True)) (UK (UK (UInt 1) (UBool True)) (UBool True))
 
 -- Again, the error message could be better, but I am getting hungry...
+
+-- If we apply PE, it's like converting SK to normal lambdas...

### 2007-11-21 tagless/IncopeTypecheck.hs
working on the genuine typecheck

@@ -3,10 +3,13 @@
 -- Interpreter, Compiler, Partial Evaluator 
 -- Typechecking into Symantics terms
 
--- The code does show how to convert untyped terms to 
+-- The code demonstrates conversion of untyped terms into 
 -- typed terms that can be interpreted by various typed interpreters. 
 -- Typechecking happens only once, regardless of the number of 
--- interpretations of the term.
+-- interpretations of the term. Although typechecking function is
+-- necessarily partial (source terms may be ill-typed), the resulting
+-- terms are assured to be intertible without pattern-match errors.
+-- Furthermore, these assurances are _syntactically_ apparent.
 
 module IncopeTypecheck where
 
@@ -18,8 +21,7 @@ import Text.Show.Functions
 
 -- Representing parsed AST terms
 --
--- The following DynTerm repr could well be the result of the
--- parser/typechecker (created along the lines of `typing dynamic typing')
+-- The following |DynTerm repr| is the result of the parser/typechecker
 -- that reads object terms from a string or a file.
 -- The type of the term is abstracted over; however, the parser does not commit
 -- to any interpretation of that term. As the following shows, we can
@@ -30,6 +32,7 @@ import Text.Show.Functions
 -- The Show constraint here is so that we could see the result of the
 -- computation...
 data DynTerm repr = forall a. (Show a, Typeable a) => DynTerm (repr a)
+
 
 -- A sample `dynamic' object term
 dynterm1 () = DynTerm (lam (\x -> x) `app` (bool True))
@@ -46,12 +49,47 @@ dynterm_P_show = case dynterm1 () of DynTerm t -> show $ compP t
 
 -- Typechecking
 -- Representing untyped terms (the AST produced by the parser, for example)
+-- Whenever we need types (to annotate a bound variable), we use
+-- an closed expression, whose (inferred) type serves as the needed annotation
 data UExp = UInt Int | UBool Bool | UAdd UExp UExp | UIf UExp UExp UExp
[…]

### 2007-11-21 tagless/talk.tex
Start to work on the APLAS 2007 talk (nothing to see here yet)

\documentclass[ucs]{beamer}
\usepackage{helvet}
\usepackage[utf8x]{inputenc}
\usepackage{microtype}
\usepackage{calc}
%\usepackage{tikz}

\usepackage{textcomp}
\makeatletter
\uc@dclc{8249}{default}{\textlangle}
\uc@dclc{8250}{default}{\textrangle}
\makeatother

\usepackage[compact]{fancyvrb1}
\DefineVerbatimEnvironment{code}{Verbatim}{commandchars=\\\{\}}

\def\>{\leavevmode\rlap{\color{blue}$\blacktriangleright$}\ \ }
\newcommand\slide[2]{#2<#1>}

\setbeamertemplate{section in toc}{{\leavevmode\llap{$\blacktriangleright$ }\bfseries\inserttocsection}\par}
\setbeamertemplate{section in toc shaded}{{\color{black}\inserttocsection}\par}
\setbeamertemplate{subsection in toc shaded}[default][50]

\AtBeginSection[]
{
  \begin{frame}<beamer>
    \frametitle{Outline}
    \tableofcontents[currentsection]
  \end{frame}
}

\begin{document}

\newcommand{\authorstack}[3]{\begin{footnotesize}\begin{tabular}[t]{@{}c@{}}\begin{normalsize}#1\end{normalsize}\\[\jot]#2\\\texttt{#3}\end{tabular}\end{footnotesize}\ignorespaces}
\newsavebox\carette \sbox\carette{\authorstack{Jacques Carette}{McMaster University}{carette@mcmaster.ca}}
\newsavebox\oleg    \sbox\oleg   {\authorstack{Oleg Kiselyov}{FNMOC}{oleg@pobox.com}}
\newsavebox\ccshan  \sbox\ccshan {\authorstack{Chung-chieh Shan}{Rutgers University}{ccshan@rutgers.edu}}

\setbeamertemplate{sidebar right}{}

\title{Finally tagless, partially evaluated}
\subtitle{Tagless staged interpreters for simpler typed languages}
\author[Jacques Carette and Oleg Kiselyov and Chung-chieh Shan]{\usebox\carette \and \usebox\oleg \and \usebox\ccshan}
\date{30 November 2007}

\begin{frame}
    \titlepage
\end{frame}

\setbeamertemplate{background}{%
    \raisebox{0pt}[\paperheight][0pt]
        {\includegraphics[angle=90,width=\paperwidth+1pt]{White_pearl_necklace.jpg}}%
    \llap{\href{http://www.flickr.com/photos/28481088@N00/160781390/}
               {\scriptsize\color{gray}tanakawho on flickr\kern1pt}}}
\begin{frame}
\end{frame}
\setbeamertemplate{background}[default]

\setbeamertemplate{sidebar right}{\vfill\llap{\usebeamerfont{framesubtitle}\insertpagenumber\hskip\jot}\vskip\jot}

\begin{frame}
\end{frame}

\end{document}

### 2007-11-22 tagless/IncopeTypecheck.hs
Typechecker into HOAS, without staging but with totality guarantees

@@ -6,10 +6,99 @@
 -- The code demonstrates conversion of untyped terms into 
 -- typed terms that can be interpreted by various typed interpreters. 
 -- Typechecking happens only once, regardless of the number of 
--- interpretations of the term. Although typechecking function is
--- necessarily partial (source terms may be ill-typed), the resulting
--- terms are assured to be intertible without pattern-match errors.
+-- interpretations of the term. Although the typechecking function is
+-- necessarily partial (source terms may be ill-typed), we guarantee
+-- the resulting term can be interpreted without any pattern-match errors.
 -- Furthermore, these assurances are _syntactically_ apparent.
+
+-- That outcome is not trivial. Indeed, suppose we have a function
+-- tc :: UExp -> Gamma -> DynTerm repr, where UExp represents an 
+-- untyped expression (the result of parsing the source language)
+-- and DynTerm represents a typed expression of _some_ type, 
+-- interpreted by _any_ interpreter 'repr'.
+--
+-- For example, 
+--    tc (UAdd (UInt 1) (UInt 2)) []
+-- should give 
+--    DynTerm (add (int 1) (int 2))
+-- And it does, see the test dt1 below.
+--
+-- The trouble comes when we type-check a function, for example,
+--    UL "x" (UInt 100) (UAdd (UVar "x") (UInt 1))
+-- (here, (UInt 100) is a `type annotation', meaning that the bound variable
+-- "x" must have the type Int. All bound variables in the source language
+-- must be type annotated. The restriction is lifted if we implement
+-- the full HM typechecker with inference).
+-- The result of typechecking the function should be something like
+--    DynTerm (lam (\x -> ...))
+-- where ... is the result of type-checking the body assuming that "x"
+-- has the type Int. 
+--
+-- Naively, we would write
+--   DynTerm (lam (\x -> unDynTerm $ tc (UAdd (UVar "x") (UInt 1)) [("x",x)]))
+-- Although that approach works, it is horribly inefficient because
+-- the typechecking of the function body, 
+--   tc (UAdd (UVar "x") (UInt 1)) [("x",x)]
+-- will be run _every_ time our function is applied. We have the interpretation
[…]

### 2007-11-22 tagless/final_code/incope.hs
removing the type-checking stuff

@@ -746,32 +746,3 @@ test3_2 = unRR (test3_1)
 test3_3 :: (Symantics repr) => repr ((Int->Int)->Int)
 test3_3 = unRR (test3_1)
 
--- ------------------------------------------------------------------------
--- Representing parsed AST terms
---
--- The following DynTerm repr could well be the result of the
--- parser/typechecker (created along the lines of `typing dynamic typing')
--- that reads object terms from a string or a file.
--- The type of the term is abstracted over; however, the parser does not commit
--- to any interpretation of that term. As the following shows, we can
--- interpret the same `dynamic' term in multiple ways.
--- This test was inspired by the discussion with Adam Chlipala
--- on Lambda-the-Ultimate on Sep 4-5, 2007.
-
--- The Show constraint here is so that we could see the result of the
--- computation...
-data DynTerm repr = forall a. Show a => DynTerm (repr a)
-
--- A sample `dynamic' object term
-dynterm1 () = DynTerm (lam (\x -> x) `app` (bool True))
-
--- Examples of several interpretations, using R, L, C, and P interpreters
-dynterm_R_show = case dynterm1 () of DynTerm t -> show $ compR t
--- "True"
-dynterm_L_show = case dynterm1 () of DynTerm t -> show $ compL t
--- "3"
-dynterm_C_show = case dynterm1 () of DynTerm t -> show $ compC t
--- "((\\V0 -> V0) True)"
-dynterm_P_show = case dynterm1 () of DynTerm t -> show $ compP t
--- "True"
-

### 2007-11-23 tagless/talk.tex
Done with introduction part of talk

@@
+\pagenumbering{arabic}
 \documentclass[ucs]{beamer}
 \usepackage{helvet}
-\usepackage[utf8x]{inputenc}
+\usepackage{eulervm}
+%\usepackage[utf8x]{inputenc}
 \usepackage{microtype}
 \usepackage{calc}
-%\usepackage{tikz}
+
+\usepackage{tikz}
+\usepgflibrary{shapes}
+\usepgflibrary{arrows}
 
@@
 
+% To get the page numbers to stay constant on a frame [dpt]
+\mode<presentation>
+{
+  \setbeamertemplate{sidebar left}{\thispdfpagelabel{\insertframenumber}}
+}
+
+\title{Finally tagless, partially evaluated}
+\subtitle{Tagless staged interpreters for simpler typed languages}
+\author{Jacques Carette and Oleg Kiselyov and Chung-chieh Shan}
+
 \begin{document}
@@
 
-\title{Finally tagless, partially evaluated}
-\subtitle{Tagless staged interpreters for simpler typed languages}
 \author[Jacques Carette and Oleg Kiselyov and Chung-chieh Shan]{\usebox\carette \and \usebox\oleg \and \usebox\ccshan}
-\date{30 November 2007}
+\date{APLAS\\30 November 2007}
 
@@
 
-\setbeamertemplate{sidebar right}{\vfill\llap{\usebeamerfont{framesubtitle}\insertpagenumber\hskip\jot}\vskip\jot}
+\setbeamertemplate{sidebar right}{\vfill\llap{\usebeamerfont{framesubtitle}\insertframenumber\hskip\jot}\vskip\jot}
 
-\begin{frame}
+\tikzstyle{lang}=[ellipse,draw,anchor=west,text depth=3pt,text height=9pt]
+\tikzstyle{every picture}=[>=angle 60]
+
+\setbeamertemplate{background}{%
+    \raisebox{0pt}[\paperheight][0pt]
+        {\includegraphics[width=\paperwidth,trim=0 35 0 180,clip]{FC0836213122.jpg}}}
+\begin{frame}<1>[t,label=everywhere]{There's interpretation everywhere}
+    A fold on an inductive data type is an interpreter of a domain-specific
+    language.
+
+    \bigskip
+    \begin{tikzpicture}
+        \draw node [lang] (lang0) {contract};
+        \foreach \label/\prev/\cur in {grammar/0/1,music/1/2,$\lambda$-term/2/3}
+            \draw (lang\prev.east)+(.3,0) node [lang] (lang\cur) {\label};
+        \draw (lang3.east)+(.2,0) node [anchor=west] {\dots};
+    \onslide<2>
+        \draw (lang0)+(-110:2cm)   node (terp01) {value};
+        \draw (lang0)+( -85:2.5cm) node (terp02) {schedule};
+        \draw (lang1)+(-110:2cm)   node (terp11) {parse};
+        \draw (lang1)+( -85:2.5cm) node (terp12) {pretty-print};
+        \draw (lang2)+(-110:2cm)   node (terp21) {typeset};
+        \draw (lang2)+( -85:2.5cm) node (terp22) {perform};
+        \draw (lang3)+(-110:2cm)   node (terp31) {evaluate};
+        \draw (lang3)+( -85:2.5cm) node (terp32) {compile};
+        \foreach \lang in {0,...,3} {
+            \draw (lang\lang)+(-65:1.5cm) node {\dots};
+            \foreach \terp in {1,...,2}
+                \draw [->] (lang\lang) -- (terp\lang\terp);
+        }
+    \onslide<3>
+        \draw (lang3)+(-148:4.5cm)  node (terp31) {evaluate};
+        \draw (lang3)+(-133:4.25cm) node (terp32) {compile};
+        \draw (lang3)+(-113:4cm)    node (terp33) {specialize};
+        \draw (lang3)+( -93:4.4cm)  node (terp34) {CPS transform};
+        \draw (lang3)+( -73:4cm)    node (terp35) {pretty-print};
+        \draw (lang3)+( -58:3cm)    node {\dots};
+        \foreach \terp in {1,...,5} \draw [->] (lang3) -- (terp3\terp);
+    \end{tikzpicture}
+
+    \only<2>{The same language can be interpreted in many useful ways.}%
+    \only<3>{We focus on the $\lambda$-calculus as an example.}
+\end{frame}
+\setbeamertemplate{background}[default]
+\againframe<2->[t]{everywhere}
+
+\begin{frame}{How to preserve types?}
+    \def\bool{\onslide<5->{ \rlap{\alert{bool}}}}
+    \begin{tikzpicture}
+        \draw node [lang] (lang)
+          {\texttt{\alt<6->{\onslide<6>{\alert{leq} (\alert{lit} 3, }%
+                            \only<7>{\makebox[0pt]{\sffamily
+                                      Simply typed $\lambda$-calculus}}%
+                            \onslide<6>{\alert{lit} 4)}}
+                           {\onslide<2->{Leq (Lit 3, }%
+                            \only<1>{\makebox[0pt]{\sffamily
+                                      \alert{Typed} source language}}%
+                            \onslide<2->{Lit 4)}}:\
+                   \only<6>{\rlap{\alert{bool}}}%
+                   \onslide<2-5>{term\bool}}};
+        \draw [->] (lang) --
+            node [above,midway,anchor=east]
+              {\only<1>{\alert{Typed} metalanguage\hspace*{2em}}%
+               \only<2>{evaluate}%
+               \only<7>{\begin{tabular}{@{}r@{\qquad}}
+                        Haskell constructor classes\\
+                        or ML functors
+                        \end{tabular}}}
+            node [above,midway,anchor=east,text width=2.5in,inner xsep=0pt]
+                {\only<-2,7>{\color{bg}}\ttfamily
+                \alt<6->{\only<6>{\alert<6>{lit:\ int -> int}\\
+                lit i = i\\[1ex]
+                \alert<6>{leq:\ (int,int) -> bool}\\
+                leq (i,j) = i <= j}}
+                {\alert<5>{eval:\ term\only<5>{ bool} -> value\only<5>{ bool}}\\
+                eval (Lit i) = Int i\\
+                eval (Leq (m,n)) =\\
+                \quad \alert<4>{match (eval m, eval n) with\\
+                \quad (Int i, Int j) ->} Bool (i <= j)}}
+            +(-100:4cm)
+            node [anchor=north,text depth=0pt,text height=9pt]
+              {\texttt{\llap{\onslide<2-5>{Bool}
+                             \onslide<2-6>{true:}%
+                             \only<1>{\rlap{\sffamily
+                                            \alert{Typed} target languages}}%
+                             \only<7>{\rlap{\sffamily evaluate}}}\
+                       \only<6>{\rlap{\alert{bool}}}%
+                       \onslide<2-5>{value\bool}}};
+        \draw [->] (lang) -- +(-80:3.5cm)
+            node [anchor=north] {\only<7>{compile}};
+        \draw [->] (lang) -- +(-65:3cm)
+            node [anchor=north west,inner xsep=0pt] {\only<7>{specialize}};
+        \draw [->] (lang) -- +(-50:2.5cm)
+            node [anchor=north west,inner xsep=0pt] {\only<7>{CPS transform}};
+        \draw      (lang)    +(-35:2cm) node {\dots};
+    \end{tikzpicture}
+
+    \bigskip
+
+    \strut
+    \only<-3>{It should be obvious in the metalanguage that interpreting
+    a well-typed source term yields a well-typed target term.}%
+    \only<4>{The term should be well-typed (and closed), so pattern matching
+    in the metalanguage should always \textbf{obviously} succeed.}%
+    \only<5>{Previous solutions use (and motivate) fancier types:\\
+    GADT, dependent types.}%
+    \only<6>{Our simple solution is to be \textbf{finally tagless}:\\
+    replace term constructors by cogen functions.}%
+    \only<7>{To accommodate multiple interpretations,\\
+    the term abstracts over the cogen functions and their types.}%
+    \strut
 \end{frame}

### 2007-11-23 tagless/talk.tex
Fix some typos in the slides

@@
     \raisebox{0pt}[\paperheight][0pt]
-        {\includegraphics[angle=90,width=\paperwidth+1pt]{White_pearl_necklace.jpg}}%
+        {\includegraphics[width=\paperwidth+1pt,trim=0 100 0 0]{White_pearl_necklace.jpg}}%
     \llap{\href{http://www.flickr.com/photos/28481088@N00/160781390/}
@@
                                       Simply typed $\lambda$-calculus}}%
-                            \onslide<6>{\alert{lit} 4)}}
+                            \onslide<6>{\alert{lit} 4):}}
                            {\onslide<2->{Leq (Lit 3, }%
@@
                                       \alert{Typed} source language}}%
-                            \onslide<2->{Lit 4)}}:\
+                            \onslide<2->{Lit 4):}}\
                    \only<6>{\rlap{\alert{bool}}}%
@@
                 leq (i,j) = i <= j}}
-                {\alert<5>{eval:\ term\only<5>{ bool} -> value\only<5>{ bool}}\\
+                {\only<2-5>{\alert<5>{eval:\ term\only<5>{ 'a} -> value\only<5>{ 'a}}\\
                 eval (Lit i) = Int i\\
@@
                 \quad \alert<4>{match (eval m, eval n) with\\
-                \quad (Int i, Int j) ->} Bool (i <= j)}}
+                \quad (Int i, Int j) ->} Bool (i <= j)}}}
             +(-100:4cm)

### 2007-11-24 tagless/talk.tex
Some small fixes to be consistent about ML notation

@@
     \onslide<3>
-        \draw (lang3)+(-148:4.5cm)  node (terp31) {evaluate};
-        \draw (lang3)+(-133:4.25cm) node (terp32) {compile};
-        \draw (lang3)+(-113:4cm)    node (terp33) {specialize};
-        \draw (lang3)+( -93:4.4cm)  node (terp34) {CPS transform};
-        \draw (lang3)+( -73:4cm)    node (terp35) {pretty-print};
+        \draw (lang3)+(-148:4.5cm)  node [inner xsep=0pt] (terp31) {evaluate};
+        \draw (lang3)+(-133:4.25cm) node [inner xsep=0pt] (terp32) {compile};
+        \draw (lang3)+(-113:4cm)    node [inner xsep=0pt] (terp33) {specialize};
+        \draw (lang3)+( -93:4.4cm)  node [inner xsep=0pt] (terp34) {CPS transform};
+        \draw (lang3)+( -73:4cm)    node [inner xsep=0pt] (terp35) {pretty-print};
         \draw (lang3)+( -58:3cm)    node {\dots};
@@
 
-\begin{frame}{How to preserve types?}
-    \def\bool{\onslide<5->{ \rlap{\alert{bool}}}}
+\begin{frame}{Simple type preservation}
+    \def\bool{\only<5-6>{\ \alert{bool}}}
     \begin{tikzpicture}
@@
                             \onslide<6>{\alert{lit} 4):}}
-                           {\onslide<2->{Leq (Lit 3, }%
+                           {\onslide<2->{LEQ (LIT 3, }%
                             \only<1>{\makebox[0pt]{\sffamily
                                       \alert{Typed} source language}}%
-                            \onslide<2->{Lit 4):}}\
-                   \only<6>{\rlap{\alert{bool}}}%
-                   \onslide<2-5>{term\bool}}};
+                            \onslide<2->{LIT 4):}}%
+                   \makebox[3em][l]{\bool\only<2-5>{\ term}}}};
         \draw [->] (lang) --
@@
                 lit i = i\\[1ex]
-                \alert<6>{leq:\ (int,int) -> bool}\\
+                \alert<6>{leq:\ int * int -> bool}\\
                 leq (i,j) = i <= j}}
-                {\only<2-5>{\alert<5>{eval:\ term\only<5>{ 'a} -> value\only<5>{ 'a}}\\
-                eval (Lit i) = Int i\\
-                eval (Leq (m,n)) =\\
+                {\only<2-5>{\alert<5>{eval:\ \only<5>{'a }term -> \only<5>{'a }value}\\
+                eval (LIT i) = INT i\\
+                eval (LEQ (m,n)) =\\
                 \quad \alert<4>{match (eval m, eval n) with\\
-                \quad (Int i, Int j) ->} Bool (i <= j)}}}
+                \quad (INT i, INT j) ->} BOOL (i <= j)}}}
             +(-100:4cm)
             node [anchor=north,text depth=0pt,text height=9pt]
-              {\texttt{\llap{\onslide<2-5>{Bool}
+              {\texttt{\llap{\onslide<2-5>{BOOL}
                              \onslide<2-6>{true:}%
@@
                                             \alert{Typed} target languages}}%
-                             \only<7>{\rlap{\sffamily evaluate}}}\
-                       \only<6>{\rlap{\alert{bool}}}%
-                       \onslide<2-5>{value\bool}}};
+                             \only<7>{\rlap{\sffamily evaluate}}}%
+                       \makebox[3em][l]{\bool\only<2-5>{\ value}}}};
         \draw [->] (lang) -- +(-80:3.5cm)
-            node [anchor=north] {\only<7>{compile}};
+            node [anchor=north west,inner xsep=0pt] {\only<7>{\hspace*{-16pt}compile}};
         \draw [->] (lang) -- +(-65:3cm)
@@
         \draw [->] (lang) -- +(-50:2.5cm)
-            node [anchor=north west,inner xsep=0pt] {\only<7>{CPS transform}};
+            node [anchor=north west,inner xsep=0pt] {\only<7>{\hspace*{-3pt}CPS transform}};
         \draw      (lang)    +(-35:2cm) node {\dots};
@@
     \only<5>{Previous solutions use (and motivate) fancier types:\\
-    GADT, dependent types.}%
-    \only<6>{Our simple solution is to be \textbf{finally tagless}:\\
+    generalized abstract data types, dependent types.}%
+    \only<6>{Our simple solution is to be \textbf{finally tagless:}\\
     replace term constructors by cogen functions.}%
-    \only<7>{To accommodate multiple interpretations,\\
-    the term abstracts over the cogen functions and their types.}%
+    \only<7>{The term accommodates \textbf{multiple interpretations}\\
+    by abstracting over the cogen functions and their types.}%
     \strut

### 2007-11-25 tagless/talk.tex
Some more slides with not much content

@@
 \begin{frame}{Simple type preservation}
-    \def\bool{\only<5-6>{\ \alert{bool}}}
+    \def\bool{\only<6-7>{\ \alert{bool}}}
     \begin{tikzpicture}
         \draw node [lang] (lang)
-          {\texttt{\alt<6->{\onslide<6>{\alert{leq} (\alert{lit} 3, }%
-                            \only<7>{\makebox[0pt]{\sffamily
+          {\texttt{\alt<7->{\onslide<7>{\alert{leq} (\alert{lit} 3, }%
+                            \only<8>{\makebox[0pt]{\sffamily
                                       Simply typed $\lambda$-calculus}}%
-                            \onslide<6>{\alert{lit} 4):}}
-                           {\onslide<2->{LEQ (LIT 3, }%
+                            \onslide<7>{\alert{lit} 4):}}
+                           {\onslide<3->{LEQ (LIT 3, }%
                             \only<1>{\makebox[0pt]{\sffamily
                                       \alert{Typed} source language}}%
-                            \onslide<2->{LIT 4):}}%
-                   \makebox[3em][l]{\bool\only<2-5>{\ term}}}};
+                            \only<2>{\makebox[0pt]{$3 \le 4$}}%
+                            \onslide<3->{LIT 4):}}%
+                   \makebox[3em][l]{\bool\only<3-6>{\ term}}}};
         \draw [->] (lang) --
@@
               {\only<1>{\alert{Typed} metalanguage\hspace*{2em}}%
-               \only<2>{evaluate}%
-               \only<7>{\begin{tabular}{@{}r@{\qquad}}
-                        Haskell constructor classes\\
+               \only<2-3>{evaluate}%
+               \only<8>{\begin{tabular}{@{}r@{\qquad}}
+                        Haskell constructor instances\\
                         or ML functors
@@
             node [above,midway,anchor=east,text width=2.5in,inner xsep=0pt]
-                {\only<-2,7>{\color{bg}}\ttfamily
-                \alt<6->{\only<6>{\alert<6>{lit:\ int -> int}\\
+                {\only<-3,8>{\color{bg}}\ttfamily
+                \alt<7->{\only<7>{\alert<7>{lit:\ int -> int}\\
                 lit i = i\\[1ex]
-                \alert<6>{leq:\ int * int -> bool}\\
+                \alert<7>{leq:\ int * int -> bool}\\
                 leq (i,j) = i <= j}}
-                {\only<2-5>{\alert<5>{eval:\ \only<5>{'a }term -> \only<5>{'a }value}\\
+                {\only<3-6>{\alert<6>{eval:\ \only<6>{'a }term -> \only<6>{'a }value}\\
                 eval (LIT i) = INT i\\
                 eval (LEQ (m,n)) =\\
-                \quad \alert<4>{match (eval m, eval n) with\\
+                \quad \alert<5>{match (eval m, eval n) with\\
                 \quad (INT i, INT j) ->} BOOL (i <= j)}}}
@@
             node [anchor=north,text depth=0pt,text height=9pt]
-              {\texttt{\llap{\onslide<2-5>{BOOL}
-                             \onslide<2-6>{true:}%
+              {\texttt{\llap{\onslide<3-6>{BOOL}
+                             \onslide<3-7>{true:}%
                              \only<1>{\rlap{\sffamily
                                             \alert{Typed} target languages}}%
-                             \only<7>{\rlap{\sffamily evaluate}}}%
-                       \makebox[3em][l]{\bool\only<2-5>{\ value}}}};
+                             \only<2>{\rlap{\enspace$\mathsf{true}$}}%
+                             \only<8>{\rlap{\sffamily evaluate}}}%
+                       \makebox[3em][l]{\bool\only<3-6>{\ value}}}};
         \draw [->] (lang) -- +(-80:3.5cm)
-            node [anchor=north west,inner xsep=0pt] {\only<7>{\hspace*{-16pt}compile}};
+            node [anchor=north west,inner xsep=0pt] {\only<8>{\hspace*{-16pt}compile}};
         \draw [->] (lang) -- +(-65:3cm)
-            node [anchor=north west,inner xsep=0pt] {\only<7>{specialize}};
+            node [anchor=north west,inner xsep=0pt] {\only<8>{specialize}};
         \draw [->] (lang) -- +(-50:2.5cm)
-            node [anchor=north west,inner xsep=0pt] {\only<7>{\hspace*{-3pt}CPS transform}};
+            node [anchor=north west,inner xsep=0pt] {\only<8>{\hspace*{-3pt}CPS transform}};
         \draw      (lang)    +(-35:2cm) node {\dots};
@@
     \strut
-    \only<-3>{It should be obvious in the metalanguage that interpreting
+    \only<-4>{It should be obvious in the metalanguage that interpreting
     a well-typed source term yields a well-typed target term.}%
-    \only<4>{The term should be well-typed (and closed), so pattern matching
+    \only<5>{The term should be well-typed (and closed), so pattern matching
     in the metalanguage should always \textbf{obviously} succeed.}%
-    \only<5>{Previous solutions use (and motivate) fancier types:\\
+    \only<6>{Previous solutions use (and motivate) fancier types:\\
     generalized abstract data types, dependent types.}%
-    \only<6>{Our simple solution is to be \textbf{finally tagless:}\\
+    \only<7>{Our simple solution is to be \textbf{finally tagless:}\\
     replace term constructors by cogen functions.}%
-    \only<7>{The term accommodates \textbf{multiple interpretations}\\
+    \only<8>{The term accommodates \textbf{multiple interpretations}\\
     by abstracting over the cogen functions and their types.}%
@@
 
+\section{The object language}
+\subsection{As a constructor class in Haskell}
+\subsection{As a functor signature in ML}
+\section{Tagless interpretation}
+\subsection{Evaluation}
+\subsection{Compilation}
+\section{Type-indexed types}
+\subsection{Partial evaluation}
+\subsection{CPS transformation}
+
+\begin{frame}{Conclusion}
+    Initial type-checking can be done
+
+    Preserves sharing
+\end{frame}
+
 \end{document}

### 2007-11-25 tagless/talk.tex
Introduce the object language

@@
 \pagenumbering{arabic}
-\documentclass[ucs]{beamer}
+\documentclass[ucs,professionalfont]{beamer}
+\usepackage{concmath}
 \usepackage{helvet}
-\usepackage{eulervm}
+%\renewcommand{\rmdefault}{cmss}
 %\usepackage[utf8x]{inputenc}
+\usepackage[override]{cmtt}
+\usepackage{mdwtab}
+\usepackage{prooftree1} \proofrulebaseline=1.5ex \proofdotseparation=.75ex
 \usepackage{microtype}
@@
 \newcommand\slide[2]{#2<#1>}
+
+%\newcommand{\ZZ}{\mathbold{Z}}
+%\newcommand{\BB}{\mathbold{B}}
+\newcommand{\ZZ}{\textrm{int}}
+\newcommand{\BB}{\textrm{bool}}
+\newcommand{\fun}[1]{\mathopen{\lambda\mathord{#1}.\;}}
+\newcommand{\fix}[1]{\mathopen{\mathop{\textrm{fix}}\mathord{#1}.\;}}
+\newcommand{\cond}[2]{\mathopen{\textrm{if}}\;#1\mathrel{\textrm{then}}#2\mathrel{\textrm{else}}}
 
@@
 \begin{frame}{Simple type preservation}
-    \def\bool{\only<6-7>{\ \alert{bool}}}
+    \def\bool{\only<7-8>{\ \alert{bool}}}
     \begin{tikzpicture}
         \draw node [lang] (lang)
-          {\texttt{\alt<7->{\onslide<7>{\alert{leq} (\alert{lit} 3, }%
-                            \only<8>{\makebox[0pt]{\sffamily
-                                      Simply typed $\lambda$-calculus}}%
-                            \onslide<7>{\alert{lit} 4):}}
-                           {\onslide<3->{LEQ (LIT 3, }%
-                            \only<1>{\makebox[0pt]{\sffamily
-                                      \alert{Typed} source language}}%
-                            \only<2>{\makebox[0pt]{$3 \le 4$}}%
-                            \onslide<3->{LIT 4):}}%
-                   \makebox[3em][l]{\bool\only<3-6>{\ term}}}};
+          {\begin{overprint}[12em]
+               \onslide<1>\makebox[12em]{\alert{Typed} source language}
+               \onslide<2>\makebox[12em]{$3 \le 4$}
+               \onslide<3-7>\makebox[12em]{\texttt{LEQ (LIT 3, LIT 4):\makebox[4em][l]{\bool\ term}}}
+               \onslide<8>\makebox[12em]{\texttt{\alert{leq} (\alert{lit} 3, \alert{lit} 4):\makebox[4em][l]{\bool}}}
+               \onslide<9>\makebox[12em]{Simply typed $\lambda$-calculus}
+           \end{overprint}};
         \draw [->] (lang) --
@@
                \only<2-3>{evaluate}%
-               \only<8>{\begin{tabular}{@{}r@{\qquad}}
+               \only<9>{\begin{tabular}{@{}r@{\qquad}}
                         Haskell constructor instances\\
@@
             node [above,midway,anchor=east,text width=2.5in,inner xsep=0pt]
-                {\only<-3,8>{\color{bg}}\ttfamily
-                \alt<7->{\only<7>{\alert<7>{lit:\ int -> int}\\
+                {\only<-3,9>{\color{bg}}\ttfamily
+                \alt<8->{\only<8>{\alert<8>{lit:\ int -> int}\\
                 lit i = i\\[1ex]
-                \alert<7>{leq:\ int * int -> bool}\\
+                \alert<8>{leq:\ int * int -> bool}\\
                 leq (i,j) = i <= j}}
-                {\only<3-6>{\alert<6>{eval:\ \only<6>{'a }term -> \only<6>{'a }value}\\
+                {\only<3-7>{\alert<7>{eval:\ \only<7>{'a }term -> \only<7>{'a }value}\\
                 eval (LIT i) = INT i\\
@@
             node [anchor=north,text depth=0pt,text height=9pt]
-              {\texttt{\llap{\onslide<3-6>{BOOL}
-                             \onslide<3-7>{true:}%
-                             \only<1>{\rlap{\sffamily
-                                            \alert{Typed} target languages}}%
-                             \only<2>{\rlap{\enspace$\mathsf{true}$}}%
-                             \only<8>{\rlap{\sffamily evaluate}}}%
-                       \makebox[3em][l]{\bool\only<3-6>{\ value}}}};
+              {\llap{\onslide<3-7>{\texttt{BOOL }}%
+                     \onslide<3-8>{\texttt{true:}}%
+                     \only<1>{\rlap{\alert{Typed} target languages}}%
+                     \only<2>{\rlap{\enspace\textrm{true}}}%
+                     \only<9>{\rlap{evaluate}}}%
+               \makebox[3em][l]{\texttt{\bool\only<3-7>{\ value}}}};
         \draw [->] (lang) -- +(-80:3.5cm)
-            node [anchor=north west,inner xsep=0pt] {\only<8>{\hspace*{-16pt}compile}};
+            node [anchor=north west,inner xsep=0pt] {\only<9>{\hspace*{-16pt}compile}};
         \draw [->] (lang) -- +(-65:3cm)
-            node [anchor=north west,inner xsep=0pt] {\only<8>{specialize}};
+            node [anchor=north west,inner xsep=0pt] {\only<9>{specialize}};
         \draw [->] (lang) -- +(-50:2.5cm)
-            node [anchor=north west,inner xsep=0pt] {\only<8>{\hspace*{-3pt}CPS transform}};
+            node [anchor=north west,inner xsep=0pt] {\only<9>{\hspace*{-3pt}CPS transform}};
         \draw      (lang)    +(-35:2cm) node {\dots};
@@
     a well-typed source term yields a well-typed target term.}%
-    \only<5>{The term should be well-typed (and closed), so pattern matching
-    in the metalanguage should always \textbf{obviously} succeed.}%
-    \only<6>{Previous solutions use (and motivate) fancier types:\\
-    generalized abstract data types, dependent types.}%
-    \only<7>{Our simple solution is to be \textbf{finally tagless:}\\
+    \only<5-6>{The term should be \alert{\alt<5>{well-typed}{closed}}, so
+    \alert{\alt<5>{pattern matching}{environment lookup}} in the metalanguage should always \textbf{obviously} succeed.}%
+    \only<7>{Previous solutions use (and motivate) fancier types:\\
+    generalized abstract data types (GADT) and dependent types.}%
+    \only<8>{Our simple solution is to be \textbf{finally tagless:}\\
     replace term constructors by cogen functions.}%
-    \only<8>{The term accommodates \textbf{multiple interpretations}\\
+    \only<9>{The term accommodates \textbf{multiple interpretations}\\
     by abstracting over the cogen functions and their types.}%
@@
 \subsection{As a constructor class in Haskell}
+
+\begin{frame}[t,fragile]{The object language\only<3-8>{ as a constructor class}\only<9-16>{ as a functor signature}}
+\setbeamercovered{transparent}
+\begin{overprint}
+\onslide<1-2>
+\medskip
+\begin{proofrules}
+    \[ \[ [x:t_1] \proofoverdots e:t_2 \] \justifies \fun{x}e:t_1\to t_2 \]
+    \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
+    \[ e_1:t_1\to t_2 \quad e_2:t_1 \justifies e_1 e_2: t_2 \]
+    \[ \text{$n$ is an integer} \justifies n:\ZZ \]
+    \uncover<1>{\[ \text{$b$ is a boolean} \justifies b:\BB \]}
+    \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
+    \uncover<1>{\[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \times e_2:\ZZ \]}
+    \uncover<1>{\[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]}
+    \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
+\end{proofrules}
+\onslide<3-8>
+\begin{semiverbatim}
+class Symantics \alert<3>{repr} where
+     \alert<4,5>{int }\alert<5>{:: Int -> repr Int}
+     \alert<4,6>{lam }\alert<6>{:: (repr a -> repr b) -> repr (a -> b)}
+     \alert<4,6>{fix }\alert<6>{:: (repr a -> repr a) -> repr a}
+     \alert<4,7>{app }\alert<7>{:: repr (a -> b) -> repr a -> repr b}
+     \alert<4  >{add }:: repr Int -> repr Int -> repr Int
+     \alert<4,8>{if_ }\alert<8>{:: repr Bool -> repr a -> repr a -> repr a}
+\end{semiverbatim}
+\onslide<9-16>
+\begin{semiverbatim}
+module type Symantics = sig type \alert<9>{\smash{('c,'v)}\,repr}
+ val \alert<10,11>{int}\alert<11>{:\,int\,->\,('c,int)\,repr}
+ val \alert<10,12>{lam}\alert<12>{:\,(('c,'a)\,repr\,->\,('c,'b)\,repr)\,->\,('c,'a->'b)\,repr}
+ val \alert<10,12>{fix}\alert<12>{:\,('x\,->\,'x)\,->\,(('c,'a->'b)\,repr as 'x)}
+ val \alert<10,13>{app}\alert<13>{:\,('c,'a\,->\,'b)\,repr\,->\,('c,'a)\,repr\,->\,('c,'b)\,repr}
+ val \alert<10   >{add}:\,('c,int)\,repr\,->\,('c,int)\,repr\,->\,('c,int)\,repr
+ val \alert<10,14>{if_}\alert<14>{:\,('c,bool)\,repr\,->\,(unit\,->\,'x)\,->\,(unit\,->\,'x)}
+     \alert<10,14>{   }\alert<14>{ \,         \,    \,->\,(('c,'a)\,repr as 'x)}
+end
+\end{semiverbatim}
+\end{overprint}
+
+\onslide<1->
+\begin{tabular}{@{}Mll@{}}
+\only<5-8,11-14>{\text{\textcolor{structure.fg}{Object term}}}%
+\only<16>{\rlap{\textcolor{structure.fg}{ML higher-order functor}}}
+& \only<5-8,11-15>{\textcolor{structure.fg}{\only<-14>{\llap{\tikz\draw[->,semithick](0,0)--(1.5,0)node[anchor=mid,inner sep=0pt]{\vphantom{O}};\:}}\only<3-8>{Haskell term}\only<9-15>{ML higher-order functor}}} \\
+\multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{functor (S:Symantics) -> struct open S}}}} \\
+\only<16>{\rlap{\alert{\texttt{ let term () =}}}}%
+\visible<1-14>{\uncover<1-2,5-8,11-14>{\alert<6,12>{\fun{x} \fix{f} \fun{n}}}}
+& \ttfamily
+  \only<5-8>{\alert<6>{lam (\textbackslash x -> fix (\textbackslash f -> lam (\textbackslash n ->}}%
+  \only<11-16>{\alert<12>{lam\,(fun x\,-> fix\,(fun f\,-> lam\,(fun n\,->}} \\
+\visible<1-14>{\uncover<1-2,5-8,11-14>{\alert<8,14>{\cond{\textcolor{black}{n\le\alert<5,11>{0}}}{\textcolor{black}{1}}}}}
+& \ttfamily
+  \only<5-8>{\alert<8>{if\textunderscore} \alert<8>(leq n (\alert<5>{int 0})\alert<8>) \alert<8>(int 1\alert<8>)}%
+  \only<11-16>{\alert<14>{if\textunderscore} \alert<14>(leq n (\alert<11>{int 0})\alert<14>) \alert<14>{(fun\,()\,->}\,int 1\alert<14>)} \\
+\visible<1-14>{\uncover<1-2,5-8,11-14>{x\times \alert<7,13>{f(n-1)}}}
+& \ttfamily
+  \only<5-8>{\alert<8>(mul x (\alert<7>{app f (add n (int (-1)))})\alert<8>))))}%
+  \only<11-16>{\alert<14>{(fun\,()\,->}\,mul x\,(\alert<13>{app f\,(add n\,(int\,(-1)\!)\!)}\!)\!\alert<14>)\!)\!)\!)} \\
+\multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{end:\ functor (S:Symantics) -> sig}}}} \\
+\only<16>{\rlap{\alert{\texttt{ val term:\ unit\,->}}}}%
+\visible<1-14>{\uncover<1-2,5-8,11-14>{\mathord:\;\ZZ\to\ZZ}}
+& \ttfamily
+  \only<5-8>{::\ Symantics repr => repr (Int -> Int)}%
+  \only<11-16>{('c,int->int)\,S.repr} \\
+\multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{end}}}}
+\end{tabular}
+\end{frame}
+
 \subsection{As a functor signature in ML}

### 2007-11-28 tagless/talk.tex
More slides

@@
 %\renewcommand{\rmdefault}{cmss}
-%\usepackage[utf8x]{inputenc}
+\usepackage[utf8x]{inputenc}
 \usepackage[override]{cmtt}
@@
 \newcommand{\cond}[2]{\mathopen{\textrm{if}}\;#1\mathrel{\textrm{then}}#2\mathrel{\textrm{else}}}
+\newcommand{\ceil}[1]{\lceil#1\rceil}
 
@@
 \tikzstyle{lang}=[ellipse,draw,anchor=west,text depth=3pt,text height=9pt]
-\tikzstyle{every picture}=[>=angle 60]
+\tikzstyle{every picture}=[>=angle 60,semithick]
 
@@
                 eval (LIT i) = INT i\\
-                eval (LEQ (m,n)) =\\
-                \quad \alert<5>{match (eval m, eval n) with\\
+                eval (LEQ (e1,e2)) =\\
+                \quad \alert<5>{match (eval e1, eval e2) with\\
                 \quad (INT i, INT j) ->} BOOL (i <= j)}}}
@@
 \subsection{As a constructor class in Haskell}
+\subsection{As a functor signature in ML}
 
 \begin{frame}[t,fragile]{The object language\only<3-8>{ as a constructor class}\only<9-16>{ as a functor signature}}
-\setbeamercovered{transparent}
+\setbeamercovered{transparent=20}
 \begin{overprint}
@@
 class Symantics \alert<3>{repr} where
-     \alert<4,5>{int }\alert<5>{:: Int -> repr Int}
-     \alert<4,6>{lam }\alert<6>{:: (repr a -> repr b) -> repr (a -> b)}
-     \alert<4,6>{fix }\alert<6>{:: (repr a -> repr a) -> repr a}
-     \alert<4,7>{app }\alert<7>{:: repr (a -> b) -> repr a -> repr b}
-     \alert<4  >{add }:: repr Int -> repr Int -> repr Int
-     \alert<4,8>{if_ }\alert<8>{:: repr Bool -> repr a -> repr a -> repr a}
+     \alert<4,5>{int }\alert<5>{:\!: Int -> repr Int}
+     \alert<4,6>{lam }\alert<6>{:\!: (repr a -> repr b) -> repr (a -> b)}
+     \alert<4,6>{fix }\alert<6>{:\!: (repr a -> repr a) -> repr a}
+     \alert<4,7>{app }\alert<7>{:\!: repr (a -> b) -> repr a -> repr b}
+     \alert<4  >{add }:\!: repr Int -> repr Int -> repr Int
+     \alert<4,8>{if_ }\alert<8>{:\!: repr Bool -> repr a -> repr a -> repr a}
 \end{semiverbatim}
@@
 \begin{semiverbatim}
-module type Symantics = sig type \alert<9>{\smash{('c,'v)}\,repr}
+module type Symantics = sig type \alert<9>{\smash{('c,'a)}\,repr}
  val \alert<10,11>{int}\alert<11>{:\,int\,->\,('c,int)\,repr}
@@
 \only<16>{\rlap{\textcolor{structure.fg}{ML higher-order functor}}}
-& \only<5-8,11-15>{\textcolor{structure.fg}{\only<-14>{\llap{\tikz\draw[->,semithick](0,0)--(1.5,0)node[anchor=mid,inner sep=0pt]{\vphantom{O}};\:}}\only<3-8>{Haskell term}\only<9-15>{ML higher-order functor}}} \\
-\multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{functor (S:Symantics) -> struct open S}}}} \\
+& \only<5-8,11-15>{\textcolor{structure.fg}{\only<-14>{\llap{\tikz\draw[->](0,0)--(1.5,0)node[anchor=mid,inner sep=0pt]{\vphantom{O}};\:}}\only<3-8>{Haskell term}\only<9-15>{ML higher-order functor}}} \\
+\multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{module POWER (S:Symantics) = struct open S}}}} \\
 \only<16>{\rlap{\alert{\texttt{ let term () =}}}}%
@@
 \only<16>{\rlap{\alert{\texttt{ val term:\ unit\,->}}}}%
-\visible<1-14>{\uncover<1-2,5-8,11-14>{\mathord:\;\ZZ\to\ZZ}}
+\visible<1-14>{\uncover<1-2,5-8,11-14>{\mathord:\;\ZZ\to\ZZ\to\ZZ}}
 & \ttfamily
-  \only<5-8>{::\ Symantics repr => repr (Int -> Int)}%
-  \only<11-16>{('c,int->int)\,S.repr} \\
+  \only<5-8>{:\!:\ \!Symantics \!repr \!=>\,repr\,(Int\,->\,Int\,->\,Int)}%
+  \only<11-16>{('c,\,int\,->\,int\,->\,int)\,\only<16>{\alert{S.}}repr} \\
 \multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{end}}}}
@@
 
-\subsection{As a functor signature in ML}
+\begin{frame}[fragile]{Composing object programs as functors}
+\[
+    \visible<2->{\alert{\fun{x}}}
+    \bigl( \fun{x} \fix{f} \fun{n} \cond{n\le0}{1}{x\times f(n-1)} \bigr)
+    \visible<2->{\alert{\;x\;7}}
+\]
+\onslide<3->
+\begin{semiverbatim}
+module POWER7 (S:Symantics) = struct open S
+ module P = POWER(S)
+ let term () = \alert{lam (fun x -> app (app} (P.term ()) \alert{x)}
+                                 \alert{(int 7))}
+end: functor (S:Symantics) -> sig
+ val term: unit -> ('c,\,int\,->\,int)\,S.repr
+end
+\end{semiverbatim}
+\end{frame}
+
 \section{Tagless interpretation}
@@
 \subsection{Compilation}
+
+\begin{frame}[t,fragile]{Tagless interpretation: \alt<5->{Compilation}{Evaluation}}
+
+No worry about pattern matching or environment lookup!
+
+Well-typed source programs \textbf{obviously}
+\alt<5->{\alert{translate to well-typed target programs.}}
+        {don't go wrong.\\}%
+\strut
+\begin{semiverbatim}
+module \alt<5->{C}{R} = struct
+ \alert<1>{type ('c,'a) repr = \only<4->{\visible<5->{\alert<5>{('c,}}}'a\only<4->{\visible<5->{\alert<5>{) code}}}}
+ let int (x:int) = \only<4->{\visible<5->{\alert<5>{‹}}}x\only<4->{\visible<5->{\alert<5>{›}}}
+ let lam f       = \only<4->{\visible<5->{\alert<5>{‹}}}fun x -> \only<4->{\visible<5->{\alert<5>{~(}}}f \only<4->{\visible<5->{\alert<5>{‹}}}x\only<4->{\visible<5->{\alert<5>{›)›}}}
+ let fix g       = \only<4->{\visible<5->{\alert<5>{‹}}}let rec f n = \only<4->{\visible<5->{\alert<5>{~(}}}g \only<4->{\visible<5->{\alert<5>{‹}}}f\only<4->{\visible<5->{\alert<5>{›)}}} n in f\only<4->{\visible<5->{\alert<5>{›}}}
+ let app e1 e2   = \only<4->{\visible<5->{\alert<5>{‹}}}\only<4->{\visible<5->{\alert<5>{~}}}e1 \only<4->{\visible<5->{\alert<5>{~}}}e2\only<4->{\visible<5->{\alert<5>{›}}}
+ let add e1 e2   = \only<4->{\visible<5->{\alert<5>{‹}}}\only<4->{\visible<5->{\alert<5>{~}}}e1 + \only<4->{\visible<5->{\alert<5>{~}}}e2\only<4->{\visible<5->{\alert<5>{›}}}
+ let if_ e e1 e2 = \only<4->{\visible<5->{\alert<5>{‹}}}if \only<4->{\visible<5->{\alert<5>{~}}}e then \only<4->{\visible<5->{\alert<5>{~(}}}e1 ()\only<4->{\visible<5->{\alert<5>{)}}} else \only<4->{\visible<5->{\alert<5>{~(}}}e2 ()\only<4->{\visible<5->{\alert<5>{)›}}}
+end
+\only<2>{module POWER7R = \alert<2>{POWER7(R)}}\only<6->{module POWER7C = POWER7(C)}
+\only<2>{\>POWER7R.term () 2}\only<6->{\>POWER7C.term ()}
+  \only<2>{\rlap{128}}\onslide<6->{\alert{‹fun x -> (fun x -> let rec self = fun x ->
+    (fun x -> if x <= 0 then 1 else x * self (x + (-1)))
+    x in self) x 7›}}
+\end{semiverbatim}
+\end{frame}
+
 \section{Type-indexed types}
@@
 
+\begin{frame}[fragile]{\only<11->{Type-indexed types}\only<13->{: }\only<-10,13-15>{Partial evaluation}\only<16->{CPS transformation}}
+\setbeamercovered{transparent=20}
+\newcommand\remember[1]{\tikz[remember picture,baseline=0pt,text=structure.fg] \node[inner ysep=0pt,inner xsep=2pt,anchor=base] (#1) {#1};}
+\newcommand\remark[1]{\textcolor{structure.fg}{\footnotesize(* #1 *)}}
+\definecolor{dynamic}{rgb}{.7,0,.7}
+\definecolor{static}{rgb}{0,.7,.7}
+\newcommand\dynamic[1]{\textcolor{dynamic}{\texttt{‹#1›}}}
+\newcommand\static[1]{\textcolor{static}{\texttt{#1}}}
+\begin{overlayarea}{0pt}{3ex}
+\onslide+<1,13-15>{\rlap{\hspace*{-1pc}\texttt{module P = struct}}}%
+\onslide+<12>\vspace*{-1em}%
+\begin{semiverbatim}
+module type Symantics = sig type ('c,\alert<12>{'s},'a) repr
+ val int: int -> ('c,\alert<12>{int},int) repr
+ val lam: 'x\,-> ('c, \alert<12>{('c,'s,'a) repr ->
+            \,        ('c,'t,'b) repr} as\,'x, 'a\,->\,'b) repr
+\uncover<0>{ val fix: (('c, \alert<12>{('c,'s,'a) repr -> ('c,'t,'b) repr},
+                'a -> 'b) repr as 'x -> 'x) -> 'x}
+ val app: ('c, \alert<12>{('c,'s,'a) repr ->
+               ('c,'t,'b) repr} as\,'x, 'a\,->\,'b) repr ->\,'x
+\uncover<0>{ val add: 'x -> 'x -> (('c,\alert<12>{int},int) repr as\,'x)
+ val if_: ('c,\alert<12>{bool},bool) repr -> (unit->'x) -> (unit->'x)
+                              -> (('c,\alert<12>{'s},'a) repr as\,'x)}
+end
+\end{semiverbatim}
+\end{overlayarea}
+\begin{tabular}[L]{@{}>{\ttfamily\frenchspacing}l@{\hspace*{-.5em}}r@{ }cc@{}}
+\onslide+<1-10,13->
+type ('c,\alt<13->{\only<14->{\alert<14-15>{'s},}'a}{\alt<1>{ 'a}{\alert<2>{int}}}) repr
+                    & \onslide+<3-10>\remember{source term}
+                    & \onslide+<4-10>$3$
+                    & \onslide+<5-10>$x$ \\
+\onslide+<1-10,13->
+\ \ \ = \alt<13->{\alt<16->{\makebox[14pc][l]{\alert{('s\,->\,ans)\,->\,ans}}\remark{CBN CPS evaluator}}{('c,'a) code}}
+                 {\alt<1>{???}{('c,int) code}}
+                    & \onslide+<3-10>\remember{dynamic part}
+                    & \onslide+<4-10>\dynamic{3}
+                    & \onslide+<5-10>\dynamic{x} \\
+\onslide+<2-10,13->
+\ \ \ \alt<16->{= \makebox[14pc][l]{\alert{('c, ('s\,->\,ans)\,->\,ans) code}}\remark{CBN CPS compiler}}{* \alt<13->{\alert{\alt<14->{'s}{('c,'a) static}}}{int} option}
+                    & \onslide+<3-10>\remember{static part}
+                    & \onslide+<4-10>\static{Some 3}
+                    & \onslide+<5-10>\static{None} \\
+\onslide+<14-15>...\\
+\onslide+<14-15>\rlap{\hspace*{-1pc}end}%
+\onslide+<6-10>
+type ('c,\alert<6>{int->int}) repr
+                    && \onslide+<8-10>\hspace*{-4em}$\fun{x}x$
+                    & \onslide+<7-10>$f$ \\
+\onslide+<6-10>
+\ \ \ = ('c,int->int) code
+                    && \onslide+<8-10>\hspace*{-4em}\dynamic{fun \!x->x}
+                    & \onslide+<7-10>\dynamic{f} \\
+\onslide+<15>\rlap{\hspace*{-1pc}module POWER7P = POWER7(P)}%
+\onslide+<6-10>
+\ \ \ * \alert<8>{(('c,int) repr ->}
+                    && \onslide+<8-10>\hspace*{-4em}\static{Some (fun \!r->r)}
+                    & \onslide+<7-10>\static{None} \\
+\onslide+<15>\rlap{\hspace*{-1pc}\>POWER7P.term ()}%
+\onslide+<6-10>
+\ \ \ \ \ \ \alert<8>{('c,int) repr)} option \\
+\onslide+<15>\rlap{\hspace*{-1pc}\ \ (‹fun x -> x\,*\,x\,*\,x\,*\,x\,*\,x\,*\,x\,*\,x›, Some <fun>)}%
+\\
+\onslide+<9-10>
+type ('c,'a) \\
+\ \ \ = ('c,'a) code \\
+\ \ \ * \alert<9-10>{\alt<9>{???}{('c,'a) static}} option
+\end{tabular}
+\onslide+<10-13,14->
+\texttt{\begin{tabular}[L]{@{}l@{ }l@{}}
+    type ('c,\:int)        &static = int \\
+    type ('c,\:bool)       &static = bool \\
+    type ('c,\:'a\,->\,'b) &static = ('c,'a)\:repr -> ('c,'b)\:repr
+\end{tabular}}
+
+\onslide+<3-10>
+\begin{tikzpicture}[remember picture,overlay,->,rounded corners=1ex,draw=structure.fg,text=structure.fg]
+    \draw (source term) --
+        +(-2.4,0) |- (dynamic part);
+    \draw (source term) --
+        node [midway,above,inner sep=0pt] {\footnotesize interpret}
+        +(-2.4,0) |- (static part);
+\end{tikzpicture}
+\end{frame}
+
 \begin{frame}{Conclusion}
+    Early obvious safety
+
     Initial type-checking can be done
 
+    Accommodates interpretations with type-indexed types
+
     Preserves sharing
+
+    Acknowledge prior-work authors in slides
 \end{frame}

### 2008-01-23 SCP2008/paper/metamonads.tex
notes and some edits

@@
 generator is at fault when the generated code cannot be parsed. Other
-problems is preventing accidental variable capture (so-called hygiene
+problems are preventing accidental variable capture (so-called hygiene
 \cite{HygienicMacros}) and ensuring the generated code is
@@
 fold constants, and remove redundant bindings? Is the generator readable?
-Does it bear as semblance to the original algorithm? Is the generator
+Does it bear resemblance to the original algorithm? Is the generator
 extensible? Are the aspects truly modular? Can we add another aspect to it or
@@
 and linear system solving as our running examples to demonstrate our
-techniques.  Specifically, our contributions:
+techniques.  Specifically, our contributions are:
 \begin{itemize}
@@
 conclusion (Section~\ref{conclusion}) we comment
-on programmings with aspects and sum up our guiding methodology.
+on programming with aspects and sum up our guiding methodology.
 Appendices give samples of the generated code, available in
@@
   let uminus a = .< - .~a >.  and  add a b = .< .~a + .~b >.
+end
 
@@
 
-Our overall approach can best be described a combination 
+Our overall approach can best be described as a combination 
 of two approaches: hand-writing a code generator (cogen) suitable
@@
 of run-time information, where compression and elision frequently lead
-to increased efficiency.  Whenever dependencies between information is
+to increased efficiency.  Whenever dependency between various bits of
+information is
 inevitable, then higher-order encodings should be sought.  When choices
@@
   {TR05}-451,
-  \url{http://www.cs.rice.edu/~taha/publications/preprints/2004-02-16.pdf}
-  (Feb. 2004).
+  \url{http://www.cs.rice.edu/~taha/publications/preprints/2004-02-16.pdf}.
 

### 2008-01-26 SCP2008/paper/metamonads.tex
cleaned up the biblography, refreshed a few references.

@@
 \cite{KiselyovTaha}.  But more problems remain
-\cite{Padua:MetaOcaml:04}: generating binding forms (``names'')
+\cite{scp-CohenDGHKP06}: generating binding forms (``names'')
 when generating loop bodies or conditional branches, and making
@@
     \item Extending a let-insertion, memoizing monad of
-      \cite{KiselyovTaha,MSP:PADL04} for generating control structures
+      \cite{KiselyovTaha,SwadiMonadic06} for generating control structures
       such as loops and conditionals. The extension is non-trivial
@@
 benefits were first pointed out by \cite{Bondorf:92} in the context of partial
-evaluation, and extensively used by \cite{MSP:PADL04,KiselyovTaha} for
+evaluation, and extensively used by \cite{SwadiMonadic06,KiselyovTaha} for
 code generation. Like \cite{conf/pepm/BondorfD94}, we use this in the 
@@
 by printing |resIV5|. The code can then be ``compiled'' as |!. resIV5|
-or with off-shoring. The code for |resIV5|
+or with offshoring\cite{offshoring}. The code for |resIV5|
 (App.~\ref{app:code-GenIV5}) shows full
@@
 The monad in this paper is similar to the one described in
-\cite{MSP:PADL04,KiselyovTaha}.  However those papers used only
+\cite{SwadiMonadic06,KiselyovTaha}.  However those papers used only
 |retN| and fixpoints (for generation-time iterations).  This paper
@@
 non-square matrix), memory hierarchy issues, loop-unrolling
-\cite{Padua:MetaOcaml:04},
+\cite{scp-CohenDGHKP06},
 warnings when zero-testing is undecidable and
@@
 % Embed .bbl file directly to make document self-contained
-% \bibliography{metamonads}
-% \bibliographystyle{elsart-num}
-
-\begin{thebibliography}{10}
-\expandafter\ifx\csname url\endcsname\relax
-  \def\url#1{\texttt{#1}}\fi
-\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
-
-\bibitem{Carette06}
-J.~Carette, Gaussian {E}limination: a case study in efficient genericity with
-  {MetaOCaml}, Science of Computer Programming 62~(1) (2006) 3--24, special
-  Issue on the First MetaOCaml Workshop 2004.
-
-\bibitem{Monagan:2001:M7P}
-M.~B. Monagan, K.~O. Geddes, K.~M. Heal, G.~Labahn, S.~M. Vorkoetter,
-  J.~McCarron, P.~DeMarco, Maple 7 Programming Guide, Waterloo Maple Inc.,
-  2001.
-
-\bibitem{Gruntz:1994:IG}
-D.~Gruntz, M.~Monagan, Introduction to {Gauss}, \textsc{SIGSAM Bulletin}:
-  Communications on Computer Algebra 28~(2) (1994) 3--19.
-
-\bibitem{Axiom}
-R.~D. Jenks, R.~S. Sutor, {AXIOM}: The Scientific Computation System, Springer
-  Verlag, 1992.
-
-\bibitem{Czarnecki}
-K.~Czarnecki, U.~W. Eisenecker, Generative programming: methods, tools, and
-  applications, ACM Press/Addison-Wesley Publishing Co., 2000.
-
-\bibitem{Veldhuizen:1998:ISCOPE}
-T.~L. Veldhuizen, Arrays in {Blitz++}, in: Proceedings of the 2nd International
-  Scientific Computing in Object-Oriented Parallel Environments (ISCOPE'98),
-  Lecture Notes in Computer Science, Springer-Verlag, 1998.
-
-\bibitem{musser89generic}
-D.~R. Musser, A.~A. Stepanov, Generic programming, in: {ISSAC 1988}:
-  Proceedings of the International Symposium on Symbolic and Algebraic
-  Computation, Vol. 358 of Lecture Notes in Computer Science, Springer-Verlag,
-  1989, pp. 13--25.
-
-\bibitem{musser94algorithmoriented}
-D.~R. Musser, A.~A. Stepanov, Algorithm-oriented generic libraries, Software -
-  Practice and Experience 24~(7) (1994) 623--642.
-
-\bibitem{BOOST}
-J.~Siek, L.-Q. Lee, A.~Lumsdaine, The Boost Graph Library: User Guide and
-  Reference Manual, Addison-Wesley, 2002.
-
-\bibitem{POOMA}
-I.~John V.W.~Reynders, J.~C. Cummings, The {POOMA} framework, Comput. Phys.
-  12~(5) (1998) 453--459.
-
-\bibitem{ATLAS}
-R.~C. Whaley, A.~Petitet, J.~J. Dongarra, Automated empirical optimization of
-  software and the {ATLAS} project, Parallel Computing 27~(1--2) (2001) 3--35.
-
-\bibitem{HygienicMacros}
-E.~E. Kohlbecker, D.~P. Friedman, M.~Felleisen, B.~F. Duba, Hygienic macro
-  expansion., in: LISP and Functional Programming, 1986, pp. 151--161.
-
-\bibitem{camlp4}
-D.~de~Rauglaudre, Camlp4 reference manual,
-  \url{http://caml.inria.fr/camlp4/manual/} (Jan. 2002).
-
-\bibitem{conf/dagstuhl/CzarneckiOST03}
-K.~Czarnecki, J.~T. O'Donnell, J.~Striegnitz, W.~Taha, {DSL} implementation in
-  {MetaOCaml}, {Template Haskell}, and {C++}., in: C.~Lengauer, D.~S. Batory,
-  C.~Consel, M.~Odersky (Eds.), Domain-Specific Program Generation, Vol. 3016
-  of Lecture Notes in Computer Science, Springer, 2003, pp. 51--72.
-
-\bibitem{CTHL03}
-C.~Calcagno, W.~Taha, L.~Huang, X.~Leroy, Implementing multi-stage languages
-  using asts, gensym, and reflection, in: K.~Czarnecki, F.~Pfenning,
-  Y.~Smaragdakis (Eds.), Generative Programming and Component Engineering
-  (GPCE), LNCS, Springer-Verlag, 2003.
-
-\bibitem{metaocaml-org}
-{MetaOCaml}, \url{http://www.metaocaml.org}.
-
-\bibitem{TahaSheard97}
-W.~Taha, T.~Sheard, Multi-stage programming with explicit annotations, in:
-  Proceedings of the Symposium on Partial Evaluation and Semantic-Based Program
-  Manipulation {(PEPM)}, ACM Press, Amsterdam, 1997, pp. 203--217.
-
-\bibitem{TahaThesis}
-W.~Taha, Multi-stage programming: its theory and applications, Ph.D. thesis,
-  Oregon Graduate Institute of Science and Technology (1999).
-
-\bibitem{Taha2000}
-W.~Taha, A sound reduction semantics for untyped {CBN} multi-stage computation.
-  {Or}, the theory of {MetaML} is non-trival., in: PEPM, 2000, pp. 34--43.
-
-\bibitem{Pueschel:05}
-M.~P{\"u}schel, J.~M.~F. Moura, J.~Johnson, D.~Padua, M.~Veloso, B.~W. Singer,
-  J.~Xiong, F.~Franchetti, A.~Ga\v{c}i\'{c}, Y.~Voronenko, K.~Chen, R.~W.
-  Johnson, N.~Rizzolo, {SPIRAL}: Code generation for {DSP} transforms,
-  Proceedings of the IEEE, special issue on "Program Generation, Optimization,
-  and Adaptation" 93~(2).
-
-\bibitem{Kennedy01Telescoping}
-K.~Kennedy, B.~Broom, K.~Cooper, J.~Dongarra, R.~Fowler, D.~Gannon,
-  L.~Johnsson, J.~Mellor-Crummey, L.~Torczon, Telescoping languages: A strategy
-  for automatic generation of scientific problem-solving systems from annotated
-  libraries, Journal of Parallel and Distributed Computing 61~(12) (2001)
-  1803--1826.
-
-\bibitem{dongarra7}
-Z.~Chen, J.~Dongarra, P.~Luszczek, K.~Rothe, Lapack for clusters project: An
-  example of self adapting numerical software, Hawaii International Conference
-  on System Sciences HICSS-37.
-
-\bibitem{Veldhuizen:2004}
-T.~L. Veldhuizen, Active libraries and universal languages, Ph.D. thesis,
-  Indiana University Computer Science (May 2004).
-\newline\urlprefix\url{\url{http://osl.iu.edu/~tveldhui/papers/2004/dissertati%
-on.pdf}}
-
-\bibitem{KiselyovTaha}
-O.~Kiselyov, K.~N. Swadi, W.~Taha, A methodology for generating verified
-  combinatorial circuits, in: EMSOFT '04: Proceedings of the fourth ACM
-  international conference on Embedded software, ACM Press, New York, NY, USA,
-  2004, pp. 249--258.
-
-\bibitem{Padua:MetaOcaml:04}
-A.~Cohen, S.~Donadio, M.~J. Garzara'n, C.~Herrmann, D.~Padua, In search for a
-  program generator to implement generic transformations for high-performance
-  computing, metaOCaml Workshop, October 2004.
-
-\bibitem{CaretteKiselyov05}
-J.~Carette, O.~Kiselyov, Multi-stage programming with {F}unctors and {M}onads:
-  eliminating abstraction overhead from generic code, in: Generative
-  Programming and Component-based Engineering {GPCE}, 2005, pp. 256--274.
-
-\bibitem{MSP:PADL04}
-J.~L. Eckhardt, R.~Kaiabachev, K.~N. Swadi, W.~Taha, O.~Kiselyov, Practical
-  aspects of multi-stage programming, rice University Techical Report
-  {TR05}-451,
-  \url{http://www.cs.rice.edu/~taha/publications/preprints/2004-02-16.pdf}.
-
-\bibitem{metamonadsURL}
-{Source code}, \url{http://www.cas.mcmaster.ca/~carette/metamonads/}.
-
-\bibitem{Golub-vanLoan}
-G.~H. {Golub}, C.~F. {Van Loan}, Matrix Computations, 3rd Edition, Johns
-  Hopkins University Press, Baltimore, MD, 1996.
-
-\bibitem{kiczales97aspectoriented}
-G.~Kiczales, J.~Lamping, A.~Menhdhekar, C.~Maeda, C.~Lopes, J.-M. Loingtier,
-  J.~Irwin, Aspect-oriented programming, in: M.~Ak\c{s}it, S.~Matsuoka (Eds.),
-  Proceedings European Conference on Object-Oriented Programming, Vol. 1241,
-  Springer-Verlag, Berlin, Heidelberg, and New York, 1997, pp. 220--242.
-
-\bibitem{709568}
-J.~Irwin, J.-M. Loingtier, J.~R. Gilbert, G.~Kiczales, J.~Lamping,
-  A.~Mendhekar, T.~Shpeisman, Aspect-oriented programming of sparse matrix
-  code, in: ISCOPE '97: Proceedings of the Scientific Computing in
-  Object-Oriented Parallel Environments, Springer-Verlag, London, UK, 1997, pp.
-  249--256.
-
-\bibitem{mendhekar97rg}
-A.~Mendhekar, G.~Kiczales, J.~Lamping, {RG}: {A} case-study for aspect-oriented
-  programming, Tech. Rep. SPL97-009 P9710044, Palo Alto, CA, USA (February
-  1997).
-
-\bibitem{journals/cacm/parnas72a}
-D.~L. Parnas, On the criteria to be used in decomposing systems into modules.,
-  Commun. ACM 15~(12) (1972) 1053--1058.
-
-\bibitem{journals/tse/Parnas76}
-D.~L. Parnas, On the design and development of program families., IEEE Trans.
-  Software Eng. 2~(1) (1976) 1--9.
-
-\bibitem{EWD:EWD447}
-E.~W. Dijkstra, On the role of scientific thought, published as
-  \cite{EWD:EWD447pub}.
-\newline\urlprefix\url{\url{http://www.cs.utexas.edu/users/EWD/ewd04xx/EWD447.%
-PDF}}
-
-\bibitem{ZhCaJeMo06a}
-W.~Zhou, J.~Carette, D.~J. Jeffrey, M.~B. Monagan, Hierarchical representations
-  with signatures for large expression management, in: Proceedings of
-  Artificial Intelligence and Symbolic Computation, Vol. 4120 of Lecture Notes
-  in Computer Science, 2006, pp. 254--268.
-
-\bibitem{Bondorf:92}
-A.~Bondorf, Improving binding times without explicit {CPS}-conversion, in: 1992
-  ACM Conference on Lisp and Functional Programming. San Francisco, California,
-  1992, pp. 1--10.
-
-\bibitem{conf/pepm/BondorfD94}
-A.~Bondorf, D.~Dussart, Improving {CPS}-based partial evaluation: Writing cogen
-  by hand., in: PEPM, 1994, pp. 1--9.
-
-\bibitem{Haskell98Report}
-S.~Peyton~Jones, et~al., The Revised Haskell 98 Report, Cambridge Univ.\null{}
-  Press, 2003, also on \url{http://haskell.org/}.
-
-\bibitem{moggi-notions}
-E.~Moggi, Notions of computation and monads, Information and Computation 93~(1)
-  (1991) 55--92.
-
-\bibitem{Filinski:Representing}
-A.~Filinski, Representing monads., in: POPL, 1994, pp. 446--457.
-
-\bibitem{liang-interpreter}
-S.~Liang, P.~Hudak, M.~Jones, Monad transformers and modular interpreters, in:
-  {POPL} '95: Conference Record of the Annual {ACM} Symposium on Principles of
-  Programming Languages, {ACM} {P}ress, {N}ew {Y}ork, 1995, pp. 333--343.
-
-\bibitem{taha-environment}
-W.~Taha, M.~F. Nielsen, Environment classifiers, in: {POPL} '03: Conference
-  Record of the Annual {ACM} Symposium on Principles of Programming Languages,
-  {ACM} {P}ress, {N}ew {Y}ork, 2003, pp. 26--37.
-
-\bibitem{mlton-proplist}
-{{MLton} team}, Property list, \url{http://mlton.org/PropertyList}.
-
-\bibitem{Watt:2002:HCA}
-S.~M. Watt, Aldor, in: J.~Grabmeier, E.~Kaltofen, V.~Weispfennig (Eds.),
-  Computer Algebra Handbook: Foundations, Applications, Systems, Springer
-  Verlag, 2003.
-
-\bibitem{Gluck95}
-R.~Gl{\"u}ck, R.~Nakashige, R.~Z{\"o}chling, Binding-time analysis applied to
-  mathematical algorithms, in: System Modelling and Optimization, 1995.
-
-\bibitem{GluckJ97}
-R.~Gl{\"u}ck, J.~J{\o}rgensen, An automatic program generator for multi-level
-  specialization., Lisp and Symbolic Computation 10~(2) (1997) 113--158.
-
-\bibitem{edsl}
-P.~Hudak, Building domain-specific embedded languages, ACM Comput. Surv.
-  28~(4es) (1996) 196.
-
-\bibitem{WhenHowDSL}
-M.~Mernik, J.~Heering, A.~M. Sloane, When and how to develop domain-specific
-  languages, ACM Comput. Surv. 37~(4) (2005) 316--344.
-
-\bibitem{EWD:EWD447pub}
-E.~W. Dijkstra, On the role of scientific thought, in: Selected Writings on
-  Computing: A Personal Perspective, Springer-Verlag, 1982, pp. 60--66.
-
-\end{thebibliography}
+\bibliography{metamonads}
+\bibliographystyle{elsart-num}
+
+%% \begin{thebibliography}{10}
+%% \expandafter\ifx\csname url\endcsname\relax
+%%   \def\url#1{\texttt{#1}}\fi
+%% \expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
+
 \appendix

### 2008-01-26 SCP2008/paper/metamonads.tex
many small changes; addressing comments by rev2

@@
 determinant and rank tracking, 
-pivoting policies, result types, etc -- at no run-time cost.  Because
+pivoting policies, result types, etc. -- at no run-time cost.  Because
 the resulting code is generated just right and not changed afterwards,
@@
 well-known issue of balancing between maximal performance and the
-level of abstraction at which code is written.  Furthermore, already
-in linear algebra, there is a wealth of different aspects that
-\emph{may} need to be addressed. Implementations of the
+level of abstraction at which code is written. Widely used Gaussian
+Elimination (GE) -- the running example of our paper -- is typically
+presented in textbooks as a closely related family of algorithms for
+solving simultaneous linear equations, LU matrix decomposition, and
+computing the determinant and the rank of the matrix. All members of
+the family share the same pattern of applying elementary row
+operations to all rows of the matrix in order. The individual
+algorithms differ in the output, in application of pivoting, in
+algebraic domain and the use of full division. Modern architectures
+demand further splits \oleg{specializations? instantiations?
+  divisions?} of the family for particular matrix layouts, e.g.,
+sparsed or tiled.
+
+A survey \cite{Carette06} of
+Gaussian elimination implementations in the industrial package Maple
+\cite{Monagan:2001:M7P}
+found 6 clearly identifiable aspects and 35 different implementations of the
+algorithm, as well as 45 implementations of directly related
+algorithms such as Cholesky decomposition, and so on.  We could
+manually write each of these implementations optimizing for particular aspects
+and using cut-and-paste to ``share'' similar pieces of code.
+Or we can write a very generic procedure that accounts for
+all the aspects with appropriate abstractions \cite{Gruntz:1994:IG,Axiom}. The
+abstraction mechanisms however -- be they procedure, method or a
+function call -- have a significant cost, especially for
+high-performance numerical computing \cite{Carette06}. Eliminating
+this abstraction overhead involves either complex analyses or
+domain-specific knowledge and so cannot reliably \oleg{assuredly?}
+be done by the compiler \oleg{cite telescopic languages, active
+  libraries, or at least scp-CohenDGHKP06?}.
+
+\begin{comment}
+% Old text
+Furthermore, already
+in linear algebra, there is a wealth of different aspects 
+such as pivoting, algebraic domain, and matrix representation
+that \emph{may}\oleg{indeed, why may is emphasized?} need to be
+addressed. Implementations of the
 widely used LU decomposition algorithm and its particular case Gaussian 
@@
 compiler because the tweaking often involves domain knowledge. 
-
-A survey \cite{Carette06} of
-Gaussian elimination implementations in the industrial package Maple
-\cite{Monagan:2001:M7P}
-found 6 clearly identifiable aspects and 35 different implementations of the
-algorithm, as well as 45 implementations of directly related algorithms such as
-LU decomposition, Cholesky decomposition, and so on.  We could
-manually write each of these implementations optimizing for particular aspects
-and using cut-and-paste to ``share'' similar pieces of code.
-Or we can write a very generic procedure that accounts for
-all the aspects with appropriate abstractions \cite{Gruntz:1994:IG,Axiom}. The
-abstraction mechanisms however -- be they procedure, method or a
-function call -- have a significant cost, especially for
-high-performance numerical computing \cite{Carette06}.
+\end{comment}
 
@@
 Does it bear resemblance to the original algorithm? Is the generator
-extensible? Are the aspects truly modular? Can we add another aspect to it or
+extensible? Are the aspects truly modular? Can we add another aspect or
 another instance of the existing aspect without affecting the existing ones?
@@
 
-In this paper, we report on our continued progress \cite{CaretteKiselyov05}
+In this paper, we report on our continued progress \cite{CaretteKiselyov05}%
 \footnote{We describe here a new version of our generator
@@
       places in the generated code.
-    \item Encode domain-specific knowledge in the generators which 
-      will catch domain-specific instantiation errors at generation
+    \item Encode domain-specific knowledge in the generators so to
+      catch domain-specific instantiation errors at generation
       time.
-    \item Provide a thorough breakdown of the family of LU decomposition
+    \item Provide a thorough calssification of the family of LU decomposition
       algorithms.
@@
 		pivots.  For example, Maple's
-		\texttt{LinearAlgebra:-LUDecomposition} routine has
+		\texttt{LinearAlgebra:\hyp LUDecomposition} routine has
 		$2^6 + 2^5 + 2^2 = 100$ possible outputs, depending on whether
@@
 	\item \textbf{Code representation}: the form and the language
-	  for the generated code (OCaml, C, Fortran, etc). A
+	  for the generated code (OCaml, C, Fortran, etc.). A
 	  degenerate case, useful for testing, is for the generator
@@
 different effects are layered
-(state, exception, non-determinism, etc) on top of the
+(state, exception, non-determinism, etc.) on top of the
 basic monad \cite{liang-interpreter}.
@@
 
-let l1 f = fun x     -> perform t <-- x; f t
-let l2 f = fun x y   -> perform tx <-- x; ty <-- y; f tx ty
-
-let retN a = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
+let retN (a : ('c,'v)code) : 
+ (<classif: 'c; answer: ('c,'w)code; ..>,('c,'v)code) monad 
+   = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
 
@@
 writing a computation-wide state, and control effects. The latter are
-normally associated with exceptions, forking of computations, et
-cetera --
+normally associated with exceptions, forking of computations, etc. --
 in general, whenever a computation ends with something other than
@@
 \end{code}
-We define functions |l1|,
-|l2|, |l3| (analogues of |liftM|, |liftM2|, |liftM3| of Haskell) 
-to make such a lifting generic. However we also need
-another |ifM| function, with the same
+However we also need another |ifM| function, with the same
 interface (see Figure~\ref{ourmonad}).
-The difference between them is
-apparent from the following example:
+The difference between them is apparent from the following example:
 \begin{code}
@@
 
-One particular domain instance is |IntegerDomain|. The notation\\
-|module IntegerDomain : DOMAIN| makes the compiler verify that our
-|IntegerDomain| is indeed a |DOMAIN|. 
-The constraint |DOMAIN| may be omitted; in that case, the
-compiler will verify the type when we try to use that structure as a
+One particular domain instance is |IntegerDomain|. The type annotation
+|DOMAIN| in the definition of |IntegerDomain| makes the 
+compiler verify that the defined structure is indeed of a type |DOMAIN|. 
+The annotation may be omitted (see |ZpMake| below), in which
+case the compiler will verify the type when we try to use that structure as a
 |DOMAIN|. In any case, the errors such as missing ``methods'' or
@@
       _ <-- IF.L.decl (if wants_pack then b else C.identity rmar m);
-      let mat = {matrix=b; numrow=n; numcol=m} in
+      let mat = \{matrix=b; numrow=n; numcol=m\} in
       ret (mat, r, c, rmar)
@@
        cc <-- retN (liftGet c);
-       let cp  = {rowpos=rr; colpos=cc} in
+       let cp  = \{rowpos=rr; colpos=cc\} in
        let module Pivot = F.PivotF(F.Det)(IF.P) in
-       pivot <-- l1 retN (Pivot.findpivot mat cp);
+       pivot <-- bind (Pivot.findpivot mat cp) retN;
        seqM (matchM pivot (fun pv -> 
-              seqM (zerobelow mat {p=cp; curval=pv} )
+              seqM (zerobelow mat \{p=cp; curval=pv\} )
                    (IF.R.succ ()) )

### 2008-01-28 SCP2008/paper/metamonads.tex
minor edits of the already commited edits (plus some annoying diff noise due to space/tabs)

@@
 solving simultaneous linear equations, LU matrix decomposition, and
-computing the determinant and the rank of the matrix. All members of
+computing the determinant and the rank of a matrix. All members of
 the family share the same pattern of applying elementary row
-operations to all rows of the matrix in order. The individual
-algorithms differ in the output, in application of pivoting, in
+operations to rows of the matrix in a particular order. The individual
+algorithms differ in their output, in application of pivoting, in
 algebraic domain and the use of full division. Modern architectures
-demand further splits \oleg{specializations? instantiations?
-  divisions?} of the family for particular matrix layouts, e.g.,
+demand further divisions of the family for particular matrix layouts, e.g.,
 sparsed or tiled.
@@
 this abstraction overhead involves either complex analyses or
-domain-specific knowledge and so cannot reliably \oleg{assuredly?}
-be done by the compiler \oleg{cite telescopic languages, active
-  libraries, or at least scp-CohenDGHKP06?}.
-
-\begin{comment}
-% Old text
-Furthermore, already
-in linear algebra, there is a wealth of different aspects 
-such as pivoting, algebraic domain, and matrix representation
-that \emph{may}\oleg{indeed, why may is emphasized?} need to be
-addressed. Implementations of the
-widely used LU decomposition algorithm and its particular case Gaussian 
-Elimination (GE) --- the running
-example of our paper --- may need to account for the representation of
-the matrix, whether to compute and return the determinant or rank, how
-and whether search for a pivot, etc. Furthermore, current architectures
-demand more and more frequent tweaks which, in general, cannot be done by the
-compiler because the tweaking often involves domain knowledge. 
-\end{comment}
+domain-specific knowledge (or both!)
+\cite{Kennedy01Telescoping,Veldhuisen:2004,scp-CohenDGHKP06?},
+and so we can not rely on a general
+purpose compiler to assuredly perform such optimizations
 
@@
       places in the generated code.
-    \item Encode domain-specific knowledge in the generators so to
+    \item Encode domain-specific knowledge in the generators so as to
       catch domain-specific instantiation errors at generation
       time.
-    \item Provide a thorough calssification of the family of LU decomposition
+    \item Provide a thorough classification of the family of LU decomposition
       algorithms.
@@
 compiler verify that the defined structure is indeed of a type |DOMAIN|. 
+\jacques{In fact, it seems that this annotation cannot be put there
+because that will make type v abstract, which will result in compilation
+errors for the "lifted" domain which needs to have this type be
+available.}
 The annotation may be omitted (see |ZpMake| below), in which

### 2008-01-30 SCP2008/paper/metamonads.tex
deal with a few more comments

@@
 unreadable.  But this is needed for proper name generation.
-The problems of compositionality of code generators, expressing
-dependencies among them and domain-specific knowledge remain.
+To be able to build modular code generators, three important problems
+remain: compositionality, expressing dependencies, and integration of
+domain-specific knowledge.
 
@@
 that we are closer to the original ideas of AOP~\cite{709568,mendhekar97rg}
-which were also concerned with scientific software}.  We firmly believe that
+which were also concerned with scientific software.}.  We firmly believe that
 our typed generative methodology is much better suited to functional
@@
 
-To see the benefit of code generation, we notice that we can easily
-parameterize our code:
+By parametrizing our code, we can make the benefits of code generation
+evident:
 
@@
 
-let retN (a : ('c,'v)code) : 
- (<classif: 'c; answer: ('c,'w)code; ..>,('c,'v)code) monad 
+let retN (a : ('c,'v) code) : 
+ (<classif: 'c; answer: ('c,'w) code; ..>,('c,'v) code) monad 
    = fun s k -> .<let t = .~a in .~(k s .<t>.)>.
@@
 constructor with three arguments. When we use this monad for code
-generation, we will need yet another type variable, environment
-classifier \cite{taha-environment}. With type constructors taking more
+generation, we will need yet another type variable for the environment
+classifiers \cite{taha-environment}. With type constructors taking more
 and more arguments, it becomes increasingly difficult to read and write
@@
 result of invoking its continuation |k|. Rather, its result is a |let|
-code expression. Such a behavior is symptomatic of control operators
-(in particular, |abort|).
+code expression. Such behavior is symptomatic of control operators
+(in particular, |abort|).  The name can be taken to mean \emph{return
+a Named computation}; the name allows for proper sharing, but otherwise 
+|retN| is used in writing generators in the same way as |ret|.
 
@@
 us easily change the output form of our generated LU programs.  For example,
-We may use single-stage thunks rather than MetaOCaml staged |('a,'v) code|
+we may use single-stage thunks rather than MetaOCaml staged |('a,'v) code|
 values, for benchmarking and regression tests purposes. We could also use

### 2008-01-31 SCP2008/paper/metamonads.tex
answering all comments on Sec 3

@@
 domain-specific knowledge (or both!)
-\cite{Kennedy01Telescoping,Veldhuisen:2004,scp-CohenDGHKP06?},
+\cite{Kennedy01Telescoping,Veldhuizen:2004,scp-CohenDGHKP06},
 and so we can not rely on a general
@@
 let ret (a :'v) : ('p,'v) monad = fun s k -> k s a
-let bind a f = fun s k -> a s (fun s' b -> f b s' k)
+let bind (m : ('p,'v) monad) (f : 'v -> ('p,'u) monad) : ('p,'u) monad
+  = fun s k -> m s (fun s' b -> f b s' k)
 let fetch s k = k s s  and  store v _ k = k v ()
@@
 generation, we will need yet another type variable for the environment
-classifiers \cite{taha-environment}. With type constructors taking more
+classifiers \cite{taha-environment} (such as the type variable |'c| 
+in the type of |retN| in Figure~\ref{ourmonad}).
+With type constructors taking more
 and more arguments, it becomes increasingly difficult to read and write
@@
 The operation |retN a| is the let-insertion operation, whose simpler
-version we called |letgen| earlier. It is the first computation with
-a control effect: indeed, the result of |retN a| is \emph{not} the
+version we called |letgen| earlier. It is the first computation with a
+control effect: indeed, the result of |retN a| is \emph{not} the
 result of invoking its continuation |k|. Rather, its result is a |let|
-code expression. Such behavior is symptomatic of control operators
-(in particular, |abort|).  The name can be taken to mean \emph{return
-a Named computation}; the name allows for proper sharing, but otherwise 
-|retN| is used in writing generators in the same way as |ret|.
+code expression. Such behavior is symptomatic of control operators (in
+particular, |abort|).  The name can be taken to mean \emph{return a
+  Named computation}; the name allows for proper sharing, but
+otherwise |retN| is used in writing generators in the same way as
+|ret|. The type of |retN| is a specialization of the type of |ret|:
+the computation |retN| deals specifically with code values. The types
+of code values must include the environment classifier 
+(such as |'c|), which denotes
+the scope of free variables that may occur in the code value. The
+argument type of |retN| and the answer type of |retN| computation must
+have the same classifier -- which, informally, means that all free
+variables in the argument of |retN| are preserved in the answer of
+|retN|'s computation. When writing, for clarity, the type annotations
+of |retN| we see the benefits of type-level records: we introduce a
+new component of the record to specify the environment classifier
+|'c|, and we update the |answer| component of the record to specialize
+the answer type to be the type of code. The |state| component of the
+record is not affected, and so remains hidden in the ellipsis |..|. The
+number of parameters to the type constructor |monad| remains the same.
 

### 2008-02-01 SCP2008/paper/metamonads.tex
take care of a few more comments

@@
 \noindent  The types above are
-generally lifted twice: once from the value domain |v| to the code
+generally lifted in two ways: once from the value domain |v| to the code
 domain |('a,'v) code|, and once more from values to monadic computations
@@
 
+For our purposes, a container is an abstraction of an $n$-dimensional
+vector space, which we here specialize for $n=1,2$.  The $2$ dimensional
+case is our main interest, and its signature contains many functions particular
+to $n=2$.  For examples, we have rows and columns and operations specialized
+for them.  A container explicitly abstracts the underlying representation 
+of the data-structure, while offering an interface which is better-suited
+to linear algebra.
+
 The signature |CONTAINER2D| below specifies that a container must provide
@@
 depending on the actual representation of the container while not
-burdening the users of containers with efficiency details.
+burdening the users of containers with efficiency details.  Note that
+this module type is also abstract over the actual representation 
+|rep| of the generator data-structure.
 
@@
 %
-The accompanying code includes containers whose
+The accompanying code\cite{metamonadsURL} includes containers whose
 elements are stored in a 1D array, in a
@@
 code generation. For example, the initializing section of the LU
-generator will invoke the function |decl| of the determinant aspect so that the
+generator will invoke a function (|decl|) of the determinant aspect 
+which ``declares'' whether tracking the determinant is needed, 
+so that the
 latter may generate let-binding for reference cells that track the
@@
 The MLton team suggests a better approach: property 
-lists~\cite{mlton-proplist}.  A property list, too, represents an
-extensible object -- by its dual, a list of polymorphic variants.  The
-initial object is just the empty list. MLton's approach does not
-literally apply to us however. MetaOCaml's environment classifiers
+lists~\cite{mlton-proplist}.  A property list also represents an
+extensible object but by its dual, namely a list of polymorphic variants.  The
+initial object is just the empty list.  Unfortunately, we cannot apply
+MLton's approach litterally to our case.  MetaOCaml's environment classifiers
 require type universes parameterized by classifiers. The MLton
 approach uses generativity of exceptions or reference cells to
-generate property names, i.e., fields of an extensible record. That
-would make our aspects stateful modules, which we try to
-avoid. Fortunately, OCaml offers a way to build open unions with
+generate property names, i.e., fields of an extensible record. Using this
+technique would make our aspects stateful modules, something we do not
+want, as we wish our generator to be purely functional [even though it is
+generating imperative code].
+Fortunately, OCaml offers a way to build open unions with
 polymorphic variants. Each variant is identified by a manifest tag,

### 2008-02-04 SCP2008/paper/metamonads.tex
another batch of changes, more substantial this time

@@
 
+In fact, our implementation requires another level of indirection, and
+in fact reads
+\begin{code}
+module Make(CODE: Coderep.T) = struct
+  module D = Domains_sig.S(
+      struct type ('a, 'v) rep = ('a,'v) CODE.abstract 
+    end)
+open D
+open CODE
+...
+\end{code}
+where |CODE| is a module which implements a type |abstract| and various
+facilities over it, like the modules |Logic| and |Idx| mentioned above.
+For simplicity, we open |CODE| to allow access to the actual type |abstract|.
+
+To further trim down on the size of the signatures for various modules,
+we introduce two variants of |monad|, specialized for our purpuses.  The
+first is used when we are always generating (abstract) code, while the second
+helps when we might generate (abstract) code.
+\begin{code}
+type ('pc,'v) cmonad = 
+  ('p,('a,'v) abstract) monad
+    constraint
+      'p = <state : 's list; answer : ('a,'w) abstract>
+    constraint
+      'pc = <classif : 'a; answer : 'w; state : 's; ..>
+
+type ('pc,'v) omonad = 
+  ('p,('a,'v) abstract option) monad
+    constraint
+      'p = <state : 's list; answer : ('a,'w) abstract>
+    constraint
+      'pc = <classif : 'a; answer : 'w; state : 's; ..>
+\end{code}
 \subsection{Containers}
@@
 \begin{code}
-let rec lookup ((_,prj,_) as ip) = 
+type ('a,'b) open_rec = ('a -> 'b) * ('b -> 'a option)  * string
+
+let rec lookup ((_,prj,_) as ip:(('a,'b) open_rec) )
+   : 'b list -> 'a = 
    function [] -> raise Not_found
@@
 
-let orec_store ((inj,_,name) as ip) v s =
+let orec_store ((inj,_,name) as ip:(('a,'b) open_rec)) (v:'a) (s:'b list) 
+   : 'b list =
   let () = 
@@
 
-let orec_find ((_,_,name) as ip) s =
+let orec_find ((_,_,name) as ip:(('a,'b) open_rec)) (s:'b list) : 'a =
   try lookup ip s 
@@
 
-let mo_extend ip v = 
+let mo_extend (ip:('a,'b) open_rec) (v:'a) : ('c, 'd) monad = 
   perform s <-- fetch; store (orec_store ip v s)
 
-let mo_lookup ip = perform s <-- fetch; ret (orec_find ip s)
+let mo_lookup (ip:('a,'b) open_rec) : ('c, 'd) monad =
+  perform s <-- fetch; ret (orec_find ip s)
 \end{code}
@@
 floating-point matrix with no determinant tracking uses the
-instantiation of |DETERMINANT| where |outdet| is |unit| and all the
-functions of that module generate no code. For integer matrices, we
+instantiation of |DETERMINANT| (see definition below) where |outdet| is |unit|
+and all the
+functions of that module generate no code. For integer matrices, and
+in general whenever the matrix elements do not form a field, we
 have to track some aspects of the determinant, even if we do not output
-it. The determinant tracking aspect is complex because tracking
-variables, if any, are to be declared at the beginning of LU; the sign
+it, as the fraction-free update needs those aspects. The determinant tracking
+aspect is complex because tracking variables, if any, are to be declared at the
+beginning of LU; the sign
 of the determinant has to be updated on each row or column
@@
   val zero_sign : unit -> ('b,unit) lm
-  val acc       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
-  val get       : unit -> ('b,tdet) lm
-  val set       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
+  val acc_magn  : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
+  val get_magn  : unit -> ('b,tdet) lm
+  val set_magn  : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
   val fin       : unit -> ('b,C.Dom.v) lm
@@
 
-We have two instances of |Det|. The first one is used to specify
+We have two instances of the above. The first one is used to specify
 no determinant tracking, and so no code is
@@
   let zero_sign () = unitL
-  let acc _ = unitL
-  let get () = ret (liftRef C.Dom.zeroL)
-  let set _ = unitL
+  let acc_magn _ = unitL
+  let get_magn () = ret (liftRef C.Dom.zeroL)
+  let set_magn _ = unitL
   let fin () = failwith "Determinant is needed but not computed"
@@
 
-The following instance does track the determinant. The |decl| method
+The second instance does track the determinant. The |decl| method
 generates two let-bindings for mutable variables tracking the
@@
   let decl () = perform
-      ddecl <-- retN (liftRef oneL); (* track the sign: +1, 0, -1 *)
-      dsdecl <-- retN (liftRef Idx.one);
-      mo_extend ip (dsdecl,ddecl);
+      magn <-- retN (liftRef oneL);    (* track magnitude *)
+      sign <-- retN (liftRef Idx.one); (* track the sign: +1, 0, -1 *)
+      mo_extend ip (sign,magn);
       unitL
   let upd_sign () = perform
-      det <-- mo_lookup ip;
-      det1 <-- ret (fst det);
-      ret (Some (assign det1 (Idx.uminus (liftGet det1))))
+      (sign,_) <-- mo_lookup ip;
+      ret (Some (assign sign (Idx.uminus (liftGet sign))))
   let fin = fun () -> perform
-      (det_sign,det) <-- mo_lookup ip;
-      ifM (Logic.equalL (liftGet det_sign) Idx.zero) (ret zeroL)
-      (ifM (Logic.equalL (liftGet det_sign) Idx.one) (ret (liftGet det))
-          (ret (uminusL (liftGet det))))
+      (sign,magn) <-- mo_lookup ip;
+      ifM (Logic.equalL (liftGet sign) Idx.zero) (ret zeroL)
+      (ifM (Logic.equalL (liftGet sign) Idx.one) (ret (liftGet magn))
+          (ret (uminusL (liftGet magn))))
   \dots
@@
   type res
-  val make_result : 'a wmatrix -> (\dots,res) monad
+  val make_result : 'a wmatrix -> (<classif: 'a;\dots>,res) cmonad
 end
@@
 the preflight checks. As we saw in the previous section, 
-both instances of |Det| contain |fin ()| function to generate
+both instances of |DETERMINANT| contain a |fin ()| function to generate
 code representing the computed determinant. Only the |NoDet| does not
@@
 determinant at all.
+
+The type |wmatrix| is for convenience, a bundling of several pieces
+of data which are needed by both the |OUTPUT| aspect and the |PIVOT| 
+aspect.  It is defined as
+\begin{code}
+type 'a wmatrix = {matrix: 'a C.vc; numrow: ('a,int) abstract; 
+                   numcol: ('a,int) abstract}
+\end{code}
 
@@
 We first instantiate the container |GVC_I| for a vector over an
-integer domain. We pass the container to the main functor |GenLA|,
-since all linear algebra codes are parameterized by the
-container and the domain included in it. We open the |G_GVC_I| module
+integer domain.  |GenGE| is then one functor of many algorithm
+generators from a linear algebra library; since all our linear algebra codes
+are parameterized by the container and the domain included in it,
+We pass the container to the main functor |GenLA|.
+We open the resulting |G_GVC_I| module
 to make all its components easily available. One of those components
 is the GE generator, |GE| (which we open as well). Besides |GE|, the
-module |G_GVC_I| includes the generator for GE solvers.  The main
+module |G_GVC_I| includes the generator for GE-based solvers.  The main
 generator functor is |GenGE|, shown later.  We instantiate the
@@
 argument modules such as |PIVOT| are functors themselves (parameterized
-by the domain, the container, and the determinant functor).
+by the domain, the container, and the determinant functor). It is 
+worthwhile noting that we make sure of module subtyping -- |F| is of
+type |FEATURES| while |F.Output| is expecting an |OUTPUTDEP|, but
+|OUTPUTDEP| is a subtype of |FEATURES|.
 
@@
       whenM (Logic.notequalL bjc C.Dom.zeroL ) (perform
-        det <-- F.Det.get ();
+        det <-- F.Det.get_magn ();
         optSeqM (Iters.col_iter mat.matrix j (Idx.succ pos.p.colpos) 
@@
 routines vary depending on the choices of input and output
-aspects. Dynamic approaches in Object-Oriented language (late binding
+aspects. Dynamic approaches in Object-oriented languages (late binding
 and dynamic dispatch) or functional languages (Haskell's

### 2008-02-05 SCP2008/paper/metamonads.tex
partial changes, on the way to reorganize the paper

@@
 
+\subsection{Maintaining an extensible state}
+
+Various aspects of our generator will need to keep a state during
+code generation (see section~\ref{sec:det} for an example).
+
+The simplest method of keeping the state of the generator is by using
+mutable variables, private to each module (aspect). That would however make our
+aspects stateful. Although we are generating imperative code, we would like to
+keep our generators stateless and purely functional, for ease of comprehension
+and reasoning.  Our main
+program may include several generators referring to one particular
+aspect -- which may be present in one shared instance or in
+several. That is of no concern if the module is
+stateless, but with stateful modules, the issues of aliasing or separate
+instantiation are a source of very subtle problems.
+
+We therefore chose a different way of maintaining generator state,
+using a monad. We already need a monad for let-insertion; we now add state as
+an additional monad parameter (Fig.~\ref{ourmonad}) and monadic
+actions |fetch| and |store| to access that monadic state. The state is
+threaded throughout the entire code-generation computation.
+
+This monadic state has to accommodate several distinct pieces of state, for
+various aspects. We should be able to add a new aspect -- which may need
+to keep its own state as part of the overall monadic state -- without
+modifying or even recompiling the rest of the code. Thus our monadic
+state should be extensible. We could use an extensible
+record: an OCaml object. Each aspect would have its own field; record
+subtyping would insure modularity. Alas, this approach makes it
+difficult to create an initial state, to pass to the monad's |runM|
+method. We would be required to know the names of all fields and should know
+the proper initial value for these fields, which breaks modularity.
+
+The MLton team suggests a better approach: property 
+lists~\cite{mlton-proplist}.  A property list also represents an
+extensible object but by its dual, namely a list of polymorphic variants.  The
+initial object is just the empty list.  Unfortunately, we cannot apply
+MLton's approach litterally to our case.  MetaOCaml's environment classifiers
+require type universes parameterized by classifiers. The MLton
+approach uses generativity of exceptions or reference cells to
+generate property names, i.e., fields of an extensible record. Using this
+technique would make our aspects stateful modules, something we do not
+want, as we wish our generator to be purely functional [even though it is
+generating imperative code].
+Fortunately, OCaml offers a way to build open unions with
+polymorphic variants. Each variant is identified by a manifest tag,
+e.g., |`Tdet|, and may include a value. The tags are not generative and
+provide a form of manifest naming, similar to symbols of
+Lisp and Scheme.
+
+Below is the implementation of our open records with manifest
+naming: functions |orec_store| to add a new field to the open record
+and |orec_find| to obtain the value associated with a particular field
+name. Each ``field'' is characterized by a triple: an injection function, 
+a projection function and the string name. The latter is used for 
+printing error messages. For example, for the
+determinant tracking aspect, this triple has the form
+\begin{code}
+let ip = 
+  (fun x -> `TDet x), (function `TDet x -> Some x | _ -> None), "Det"
+\end{code}
+We combine these functions with monadic
+actions to access monadic state and so obtain 
+|mo_extend| and |mo_lookup| to store and retrieve one component 
+of the monadic state.
+\begin{code}
+type ('a,'b) open_rec = ('a -> 'b) * ('b -> 'a option)  * string
+
+let rec lookup ((_,prj,_) as ip:(('a,'b) open_rec) )
+   : 'b list -> 'a = 
+   function [] -> raise Not_found
+   | (h::t) -> (match prj h with Some x -> x | _ -> lookup ip t)
+
+let orec_store ((inj,_,name) as ip:(('a,'b) open_rec)) (v:'a) (s:'b list) 
+   : 'b list =
+  let () = 
+    try let _ = lookup ip s in 
+        failwith ("The field of an open record is already present: " ^ name)
+    with Not_found -> () in
+  (inj v)::s
+
+let orec_find ((_,_,name) as ip:(('a,'b) open_rec)) (s:'b list) : 'a =
+  try lookup ip s 
+  with Not_found -> failwith ("Failed to locate orec field: " ^ name)
+
+let mo_extend (ip:('a,'b) open_rec) (v:'a) : ('c, 'd) monad = 
+  perform s <-- fetch; store (orec_store ip v s)
+
+let mo_lookup (ip:('a,'b) open_rec) : ('c, 'd) monad =
+  perform s <-- fetch; ret (orec_find ip s)
+\end{code}
+%
+Currently, we check at generation-time that one should not add an
+already existing field to an open record, nor should one attempt to
+look up a field that does not exist. It is possible to make these
+checks static.  Our approach has the advantage of generating much
+clearer error messages.
+
 \section{Aspects and Functors}\label{functors}
@@
 The monad represents fine-scale code generation. We need tools for
-larger-scale modularization; we can use any abstraction
-mechanisms we want to structure our code generators, as long as none
-of those abstractions infiltrate the generated code. The ML module system
-turns out to be most convenient.
+larger-scale modularization; conveniently, we can use whatever abstraction
+mechanisms we want to structure our code generators, as long as these
+abstractions do not infiltrate the generated code. For our purposes, the
+ML module system turns out to be most convenient.
+
+\jacques{next paragraph will need rewritten when this section has
+finally taken shape}
+In this section, we will go through various aspects of one algorithm
+and explain how they are implemented.  In particular, and generally in
+order of increasing complexity, we will deal with abstracting the
+domain, the matrix representation, the code itself, determinant, etc.
+Then we put it all together (or in a new section?).
 
@@
 \subsection{Abstract code and lifted domains}\label{sec:lifteddomains}
+\jacques{I don't think I really want to delete the first half of this
+subsection.  I guess we can move that into section 3?  We need to be
+careful that section 3 does not become 1/2 the paper!  We might even
+consider splitting section 3 into 2 sections?}
 We will be using MetaOCaml operations to build code expressions, values
@@
 
-\subsection{Maintaining an extensible state}
-
-Various aspects of our LU generator such as determinant, rank, 
-and permutation tracking may need to keep a state during
-code generation. For example, the initializing section of the LU
-generator will invoke a function (|decl|) of the determinant aspect 
-which ``declares'' whether tracking the determinant is needed, 
-so that the
-latter may generate let-binding for reference cells that track the
-magnitude and the sign of the determinant. The generator of the
-row-swapping code will notify the determinant aspect so that the
-latter may insert code that flips the sign of the tracked
-determinant. The final section of the LU generator will ask the
-determinant aspect to generate code producing the final, signed value
-of the determinant. Clearly the determinant aspect needs to maintain
-state from one invocation of its functions to another; this state should
-include the ``names'' of mutable variables accumulating the sign and the
-magnitude of the determinant.
-
-The simplest method of keeping the state of the generator is by using
-mutable variables, private to the Determinant module (aspect). That,
-however, makes our aspects stateful. Although we are generating
-imperative code, we would like to keep our generators stateless and
-purely functional, for ease of comprehension and reasoning.  Our main
-program may include several LU generators referring to one Determinant
-aspect -- which may be present in one shared instance or in
-several. That is of no concern if the Determinant module is
-stateless. With stateful modules, the issues of aliasing or separate
-instantiation are the source of very subtle problems.
-
-We therefore chose a different way of maintaining the generator state,
-using a monad. We already need a monad for let-insertion; we add state as
-an additional monad parameter (Fig.~\ref{ourmonad}) and monadic
-actions |fetch| and |store| to access that monadic state. The state is
-threaded throughout the entire code-generation computation.
-
-This monadic state has to accommodate several distinct pieces of state, for
-various aspects. We should be able to add a new aspect -- which may need
-to keep its own state as part of the overall monadic state -- without
-modifying or even recompiling the rest of the code. Thus our monadic
-state should be extensible. We could use an extensible
-record: an OCaml object. Each aspect would have its own field; record
-subtyping would assure modularity. Alas, this approach makes it
-difficult to create an initial state, to pass to the monad's |runM|
-method. We would be required to know the names of all fields and should know
-the proper initial value for these fields, which breaks modularity.
-
-The MLton team suggests a better approach: property 
-lists~\cite{mlton-proplist}.  A property list also represents an
-extensible object but by its dual, namely a list of polymorphic variants.  The
-initial object is just the empty list.  Unfortunately, we cannot apply
-MLton's approach litterally to our case.  MetaOCaml's environment classifiers
-require type universes parameterized by classifiers. The MLton
-approach uses generativity of exceptions or reference cells to
-generate property names, i.e., fields of an extensible record. Using this
-technique would make our aspects stateful modules, something we do not
-want, as we wish our generator to be purely functional [even though it is
-generating imperative code].
-Fortunately, OCaml offers a way to build open unions with
-polymorphic variants. Each variant is identified by a manifest tag,
-e.g., |`Tdet|, and may include a value. The tags are not generative and
-provide a form of manifest naming, similar to symbols of
-Lisp and Scheme.
-
-Below is the implementation of our open records with manifest
-naming: functions |orec_store| to add a new field to the open record
-and |orec_find| to obtain the value associated with a particular field
-name. Each ``field'' is characterized by a triple: an injection function, 
-a projection function and the string name. The latter is used for 
-printing error messages. For example, for the
-determinant tracking aspect, this triple has the form
-\begin{code}
-let ip = 
-  (fun x -> `TDet x), (function `TDet x -> Some x | _ -> None), "Det"
-\end{code}
-We combine these functions with monadic
-actions to access monadic state and so obtain 
-|mo_extend| and |mo_lookup| to store and retrieve one component 
-of the monadic state.
-\begin{code}
-type ('a,'b) open_rec = ('a -> 'b) * ('b -> 'a option)  * string
-
-let rec lookup ((_,prj,_) as ip:(('a,'b) open_rec) )
-   : 'b list -> 'a = 
-   function [] -> raise Not_found
-   | (h::t) -> (match prj h with Some x -> x | _ -> lookup ip t)
-
-let orec_store ((inj,_,name) as ip:(('a,'b) open_rec)) (v:'a) (s:'b list) 
-   : 'b list =
-  let () = 
-    try let _ = lookup ip s in 
-        failwith ("The field of an open record is already present: " ^ name)
-    with Not_found -> () in
-  (inj v)::s
-
-let orec_find ((_,_,name) as ip:(('a,'b) open_rec)) (s:'b list) : 'a =
-  try lookup ip s 
-  with Not_found -> failwith ("Failed to locate orec field: " ^ name)
-
-let mo_extend (ip:('a,'b) open_rec) (v:'a) : ('c, 'd) monad = 
-  perform s <-- fetch; store (orec_store ip v s)
-
-let mo_lookup (ip:('a,'b) open_rec) : ('c, 'd) monad =
-  perform s <-- fetch; ret (orec_find ip s)
-\end{code}
-%
-Currently, we check at generation-time that one should not add an
-already existing field to an open record, nor should one attempt to
-look up a field that does not exist. It is possible to make these
-checks static.
-
 \subsection{Determinant aspect}
 \label{sec:det}
+% Various aspects of our LU generator such as determinant, rank, 
+% and permutation tracking may need to keep a state during
+% code generation. For example, the initializing section of the LU
+% generator will invoke a function (|decl|) of the determinant aspect 
+% which ``declares'' whether tracking the determinant is needed, 
+% so that the
+% latter may generate let-binding for reference cells that track the
+% magnitude and the sign of the determinant. The generator of the
+% row-swapping code will notify the determinant aspect so that the
+% latter may insert code that flips the sign of the tracked
+% determinant. The final section of the LU generator will ask the
+% determinant aspect to generate code producing the final, signed value
+% of the determinant. Clearly the determinant aspect needs to maintain
+% state from one invocation of its functions to another; this state should
+% include the ``names'' of mutable variables accumulating the sign and the
+% magnitude of the determinant.
+
 To track the determinant, we should be able to generate code

### 2008-02-14 tagless/tagless-final.tex
typechecking experiments
I took care of all simple comments.

@@
+%% TODO: add a ref 
+%%  Lars Birkedal's master thesis
+%%   http://www.itu.dk/people/birkedal/papers/paresm.ps.gz
+%% and thank Eijiro Sumii
+
 \begin{comment}
@@
 
-We attack the problem of tagless (staged) typed-preserving
+We attack the problem of tagless (staged) type-preserving
 interpretation exactly as it was posed
@@
 \item We build interpreters that evaluate (\S\ref{language}),
-    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
+   compile (or evaluate with staging) (\S\ref{S:compiler}), and 
+   partially evaluate (\S\ref{PE}) a typed higher-order object language
    in a typed metalanguage, in direct and continuation\hyp passing
@@
 \emph{environment classifier} \citep{WalidPOPL03} required by MetaOCaml for
-code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
+code generation in~\S\ref{S:compiler}; and the $\eta$-expanded type for
 |fix| and thunk types in |if_| since OCaml is a call-by-value
@@
 
-\section{A tagless compiler (or, a staged interpreter)}\label{compiler}
+\section{A tagless compiler (or, a staged interpreter)}\label{S:compiler}
+
 Besides immediate evaluation, we can compile our object language
@@
 analysis.  Our partial evaluator is a modular extension of the evaluator
-in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
+in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{S:compiler}, in
 that it uses the former to reduce static terms and the latter to build
@@
 future.  This tag is not an object type tag: all pattern matching below
-is exhaustive.  Because the future-stage component is always present, we
+is exhaustive.  Because the future-stage component is always available, we
 can now define the polymorphic function
@@
 
-Any partial evaluator must decide how much to unfold recursion\ifshort.
-Our code na\"\i vely unfolds |fix|
-\else\
+Any partial evaluator must decide how much to unfold recursion
 statically: unfolding too little can degrade the residual code, whereas
@@
 Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
-way'', that is, unfold |fix| rather than residualize
-\fi
-whenever the argument is static.
+way'', that is, whenever the 
+argument is static, it unfolds |fix| rather than residualize it.
 \ifshort
@@
 \fi
-of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
+of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{S:compiler},
 a $\beta$-reduction has been statically performed to yield |true|.  More
 interestingly, whereas |testpowfix7| compiles to a code value with many
-$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
+$\beta$-redexes in~\S\ref{S:compiler}, the partial evaluation
 \ifshort
@@
 Roughly, a partial evaluator interprets each object term to yield either
-a static (present-stage) term (using\ifshort\else\ the evaluator\fi~|R|) or
-a dynamic (future-stage) term (using\ifshort\else\ the compiler\fi~|C|).  To
+a static (present-stage) term (using the evaluator~|R|) or
+a dynamic (future-stage) term (using the compiler~|C|).  To
 distinguish between static and dynamic terms, one might define
-\ifshort
-\texttt{data P0~t~= S0~(R~t)~}\Verb+|+\texttt{ E0~(C~t)}.
-\else
 \begin{code}
@@
 \end{code}
-\fi
 We then start to define |P0| as an instance of the |Symantics| class.
@@
 add them using~|C|.
-\ifshort\else
 \begin{code}
@@
 
-\fi
 Whereas |mul| and |leq| are as easy to define as |add|, we encounter
@@
 function, of type |a -> C a|, to send a value to the future stage
-\citep{xi-guarded,WalidPOPL03}.\ifshort\else
-(By the way, if we
+\citep{xi-guarded,WalidPOPL03}. (By the way, if we
 were to add polymorphic \texttt{lift} to the type class
@@
 \texttt{Applicative} and thus \texttt{Functor}:\texttt{ fmap
-f = app (lift f)~}.)\fi
+f = app (lift f)~}.)
 
@@
 We switch to the Haskell data type
-\ifshort \BUseVerbatim{P1} \else \UseVerbatim{P1} \fi
+\UseVerbatim{P1}
 so that a partially evaluated term |P1 t| always contains a dynamic
@@
 future.  This tag is not an object type tag: all pattern matching below
-is exhaustive.  Because the future-stage component is always present, we
+is exhaustive.  Because the future-stage component is always available, we
 can now define the polymorphic function
-\ifshort
-|abstr1 (P1 _ dyn) = dyn| of type
-|abstr1 :: P1 t -> C t|
-\else
 \begin{code}
@@
 \end{code}
-\fi
 to extract it without requiring polymorphic lift into~|C|.  We then try
@@
 constructs of our object language, including |if_|.
-\ifshort
-\vspace*{-0.6em}
-\begin{code3}
-instance Symantics P1 where
-  int x = P1 (Just (int x)) (int x)
-  add (P1 (Just n1) _) (P1 (Just n2) _) = int (unR (add n1 n2))
-  add e1 e2 = P1 Nothing (add (abstr1 e1) (abstr1 e2))
-  if_ (P1 (Just s) _) et ef = if unR s then et else ef
-  if_ eb et ef = P1 Nothing (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
-\end{code3}
-\vspace*{-0.6em}
-\else
 \begin{code}
@@
 \end{code}
-\fi
 When we come to functions, however, we stumble.  According to our
@@
 |P1| to a function type |a -> b| should yield one of
-\ifshort\vspace*{-0.6em}\fi
 \begin{code}
@@
 \end{code}
-\ifshort\vspace*{-0.6em}\fi
 unlike applying |P1| to |Int| or |Bool|.
@@
 inductively defines a map from source types to target types that
-performs case distinction on the source type.
-\ifshort\else The connection with CPS
-is not an accident, as we shall see in~\S\ref{S:CPS}.\fi
+performs case distinction on the source type. The connection with CPS
+is not an accident, as we shall see in~\S\ref{S:CPS}.
 
@@
 in the accompanying source code shows the latter.
-\ifshort
-\vspace*{-0.7em}
-\begin{code3}
-data P t where
-  VI :: Int  -> P Int;                VB :: Bool -> P Bool
-  VF :: (P a -> P b) -> P (a -> b);   E  :: C t -> P t
-\end{code3}
-\vspace*{-0.7em}
-\else
 We introduce a GADT with four data constructors.
@@
 \end{code}
-\fi
 The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
-in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  \ifshort\else However,
+in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
 the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
-operand of type |P a -> P b| rather than |a -> b|.  \fi We define a function
-\ifshort |abstr|, \fi like |abstr1| above\ifshort,\fi\
-to extract a future-stage computation from a value of type |P t|.
-\ifshort\else
+operand of type |P a -> P b| rather than |a -> b|.  We define a function
+like |abstr1| above to extract a future-stage computation from a 
+value of type |P t|.
 \begin{code}
@@
 \end{code}
-\fi
 The implementations of |int|, |bool|, and |add| are like 
-in~\S\ref{S:PE-problem}.
-The interpretation of |lam f| is |VF f|, which just wraps the HOAS function |f|. 
+in~\S\ref{S:PE-problem}. The interpretation of |lam f| is |VF f|, 
+which just wraps the HOAS function |f|. 
 We can always compile |f| to a code value,
 but we delay it to apply |f| to concrete arguments. The interpretation of
-|app ef ea|
-\ifshort
-defined by
-\vspace*{-0.7em}
-\begin{code3}
-app (VF f) ea = f ea;  app (E f) ea = E (app f (abstr ea))
-\end{code3}
-\vspace*{-0.7em}
-\fi
-checks to see if |ef| is such a delayed
+|app ef ea| checks to see if |ef| is such a delayed
 HOAS function |VF f|. If it is, we apply |f| to the
@@
 dynamic value |E f|, we residualize.
-\ifshort\else
-
-\fi
+
 This solution using GADTs works but is not quite satisfactory. First, it
-cannot be ported to MetaOCaml\ifshort\else, as GADTs are unavailable there\fi.  Second,
+cannot be ported to MetaOCaml, as GADTs are unavailable there.  Second,
 the problem of nonexhaustive pattern\hyp matching reappears in
-|app|\ifshort. \else\
-above: the type |P t| has four constructors, of which the pattern in
+|app| above: the type |P t| has four constructors, of which the pattern in
 |app| matches only |VF| and~|E|. One may say that the
@@
 construct values of type |P (a -> b)| as required by the type of |app|.
-\fi
-\ifshort
-Although Coq or logical
-frameworks with canonical forms could prove pattern\hyp matching
-exhaustive,
-\else
 Indeed, the metalanguage implementation could reason thus:
@@
 Then again, the metalanguage implementation may not reason thus:
-\fi
 GHC cannot and issues warnings.
-\ifshort
-Recall, our
-\else
 Although this point may seem minor, it is the heart of
@@
 but never fail in reality. The
-\fi
 goal is to make this exhaustiveness \emph{syntactically} apparent.
@@
 
-\begin{SaveVerbatim}{P}
+Let us re-examine the problem in~\S\ref{S:PE-problem}. 
+\oleg{If we drop the previous Section, the phrase above should be changed.}
+What we
+would ideally like is to write
+\begin{code}
 data P t = P (Maybe (repr_pe t)) (C t)
-\end{SaveVerbatim}
-
-Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
-would ideally like is to write
-\ifshort \BUseVerbatim{P} \else \UseVerbatim{P} \fi
+\end{code}
 where |repr_pe| is the type function defined
@@
 by
-\ifshort\vspace*{-0.6em}
-\begin{code3}
-repr_pe Int = Int; repr_pe Bool = Bool; repr_pe (a -> b) = P a -> P b
-\end{code3}
-\vspace*{-0.6em}
-\else
 \begin{code}
@@
 \end{code}
-\fi
 Although we can use type classes to define this type function
-in Haskell, that is not portable to MetaOCaml. However,
+in Haskell, that solution is not portable to MetaOCaml. However,
 these three typecase alternatives are already present in existing
@@
 into the signature |Symantics|. Switching to MetaOCaml to demonstrate
-this\ifshort\else\ portability\fi,
-we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
+this portability, we recall from Figure~\ref{fig:ocaml-simple} 
+in~\S\ref{encoding} that the |repr| type
 constructor took two arguments |'c| and~|'dv|. We add an argument
@@
 Figure~\ref{fig:ocaml} shows the new signature.
-% aplas: abbreviated figure
 \begin{figure}
 \begin{floatrule}
-\ifshort
-\begin{code2}
-module type Symantics = sig type ('c,'sv,'dv) repr
-  val int : int -> ('c,int,int) repr
-  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-            -> ('c,'x,'da -> 'db) repr
-  val app : ('c,'x,'da -> 'db) repr
-            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
-                                'da -> 'db) repr as 'x)
-  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val if_ : ('c,bool,bool) repr
-            -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x) end
-\end{code2}
-\medskip
-\else
 \begin{code2}
@@
 \end{code2}
-\fi
 \end{floatrule}
 \caption{A (Meta)OCaml embedding of our object language that supports
-  partial evaluation \ifshort \protect\linebreak[2] (\texttt{bool},
-  \texttt{mul}, \texttt{leq} are elided)\fi}
+  partial evaluation}
 \label{fig:ocaml}
@@
 \begin{floatrule}
-\ifshort
-\begin{code2}
-module P = struct
-  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
-  let abstr {dy = x} = x    let pdyn x = {st = None; dy = x}
-
-  let int  (x:int)  = {st = Some (R.int x);  dy = C.int x}
-  let add e1 e2 = match e1, e2 with
-  | {st = Some 0}, e | e, {st = Some 0} -> e
-  | {st = Some m}, {st = Some n} -> int (R.add m n)
-  | _ -> pdyn (C.add (abstr e1) (abstr e2))
-  let if_ eb et ee = match eb with
-  | {st = Some b} -> if b then et () else ee ()
-  | _ -> pdyn (C.if_ (abstr eb) (fun () -> abstr (et ()))
-                                (fun () -> abstr (ee ())))
-  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
-  let app ef ea = match ef with {st = Some f} -> f ea
-                  | _ -> pdyn (C.app (abstr ef) (abstr ea)) end
-\end{code2}
-\medskip
-\else
 \begin{code2}[commandchars=\@\[\]]
@@
 \end{code2}
-\fi
 \end{floatrule}
-\caption{Our partial evaluator (\ifshort \texttt{bool}, \texttt{mul},
-  \texttt{leq} and \texttt{fix} \else \texttt{mul} and \texttt{leq} \fi
+\caption{Our partial evaluator \texttt{mul} and \texttt{leq}
   are elided)}
@@
 
-The interpreters |R|, |L| and~|C| above only use the old
+The interpreters |R|, |L| and~|C| in \S\ref{S:interpreter-RL},\S\ref{S:compiler}
+only use the old
 type arguments |'c| and~|'dv|, which are treated by the new signature
@@
 For example, the compiler |C| now begins
-\ifshort
-\texttt{module C = struct
-  type ('c,'sv,'dv) repr = ('c,'dv) code}
-\else
 \begin{code}
@@
 \end{code}
-\fi
 with the rest the same.
@@
 multiplication by~$0$ or~$1$.  
-\ifshort
-Such optimizations can be quite effective
-\else
 Such algebraic simplifications are easy
@@
 apply.  These simplifications and abstractions help a lot
-\fi
 in a large language with more base types and primitive operations.
-\ifshort\else
 Incidentally, the code actually contains a more general implementation
@@
 generative linear algebra~\citep{CaretteKiselyov05}.
-\fi
-
-Any partial evaluator must decide how much to unfold recursion\ifshort.
-Our code na\"\i vely unfolds |fix|
-\else\
+
+Any partial evaluator must decide how much to unfold recursion
 statically: unfolding too little can degrade the residual code, whereas
@@
 Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
-way'', that is, unfold |fix| rather than residualize
-\fi
-whenever the argument is static.
-\ifshort
-In the accompanying source code is a conservative
-alternative |P.fix| that unfolds recursion only once, then residualizes.
-\else
+way'', that is, whenever the 
+argument is static, it unfolds |fix| rather than residualize it.
 A conservative alternative is to unfold recursion only once, then residualize:
@@
 \end{code}
-\fi
 Many sophisticated approaches have been developed to decide how much to unfold
-\citep{jones-partial}, but this issue is orthogonal to our presentation.
-\ifshort\else
+\citep{Jones-Mix,jones-partial}, but this issue is 
+orthogonal to our presentation.
 A separate concern in our treatment of |fix| is possible code bloat in
 the residual program, which calls for let-insertion
-\citep{SwadiTahaKiselyovPasalic2006}.
-\fi
+\citep{BondorfDanvy,SwadiTahaKiselyovPasalic2006}.
 
 Given this implementation of~|P|, our running example
-\ifshort
-\texttt{let module E = EX(P) in E.test1 ()}
-\else
 \begin{code}
@@
 \end{code}
-\fi
 evaluates to
-\ifshort
-|{P.st = Some true; P.dy = .<true>.}|
-\else
 \begin{code}
@@
 \end{code}
-\fi
-of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
+of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{S:compiler},
 a $\beta$-reduction has been statically performed to yield |true|.  More
 interestingly, whereas |testpowfix7| compiles to a code value with many
-$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
-\ifshort
-\texttt{let module E = EX(P) in E.testpowfix7}
-\else
+$\beta$-redexes in~\S\ref{S:compiler}, the partial evaluation
 \begin{code}
@@
 \end{code}
-\fi
 gives the desired result
-\ifshort\vspace*{-0.7em}\fi
 \begin{code}
@@
 \end{code}
-\ifshort\vspace*{-0.7em}\fi
 
@@
 simulate a type function with a typecase distinction, using an extra
-type argument to |repr|. Happily, the type function |repr_pe| needed for
+type argument to |repr|. Luckily, the type function |repr_pe| needed for
 the partial evaluator 
@@
 turns any implementation of |Symantics| into a CPS interpreter, whether
-it is an evaluator.  This functor on interpreters performs a textbook
+it is an evaluator.\oleg{I don't understand this whether
+  classification: any interpreter is an evaluator, no? It all depends
+  on what the domain is. Anyway, the reviewer suggests rewriting the
+  clause as ``whether or not it is an evaluator''. I suggest to drop
+  it entirely.}
+This functor on interpreters performs a textbook
 CPS transformation on the object language.
@@
 
-To save space, we just
+For brevity, we just
 use the module returned by |CPST| as is. Because it does not
@@
 the two arguments swapped.  We can encode the term
-``$\lapp{\deref}{x} \set 2;\, x+\deref$''
+``$\lapp{\deref}{x} (\set 2;\, x+\deref)$''
 as the OCaml functor
@@
 in~\S\ref{S:interpreter-RL} may just as well be written in our object
-language.  In particular, the following \emph{object} functions implement an
-evaluator.  (We use the number~$0$ in lieu of a unit value.)
+language:~\Fig.\ref{fig:self-eval}.
+\begin{figure}
 \begin{align*}
@@
 \end{align*}
-We thus map each object terms~$e$ to an object term~$\encode{e}$ as follows.
+\caption{The \emph{object} functions implementing an
+evaluator.  (We use the number~$0$ in lieu of a unit value.)}
+\label{fig:self-eval}
+\end{figure}
+We thus map each object term~$e$ to an object term~$\encode{e}$ as follows.
 We call this mapping \emph{pre-encoding}.
@@
 instantiate the free variables in~$\encode{e}$ such as $\ident{int}$,
-$\ident{lam}$, and $\ident{add}$ by their definitions above.  For
+$\ident{lam}$, and $\ident{add}$ by their definitions given in
+Fig.~\ref{fig:self-eval}.  For
 example, the familiar object term $(\fun{x}x)\True$ pre-encodes to
@@
 \end{equation*}
-Because the evaluator above mostly consists of glorified identity
+Because the evaluator in Fig.~\ref{fig:self-eval} mostly consists 
+of glorified identity
 functions, our simple partial evaluator reduces the
@@
 interpretation, provided we use universal types \citep{taha-tag}.  The
-ensuing tagging overhead motivated \citet{taha-tag} to propose tag
+ensuing tagging overhead motivated \citet{Makholm-TagElim,taha-tag} 
+to propose tag
 elimination, which however does not statically guarantee that all tags
@@
 deforest the object term representation.  Put another way, we
-manual apply type-level partial evaluation to our type
+manually apply type-level partial evaluation to our type
 functions (see \S\ref{S:PE-solution}) to obtain simpler types 
@@
 metalanguage without using GADTs, dependent types, or a universal type.
-Our family of interpreters include an evaluator, a compiler, a partial
+Our family of interpreters includes an evaluator, a compiler, a partial
 evaluator, and CPS transformers.  It is patent that they never get stuck,

### 2008-02-16 tagless/tagless-final.tex
a few more comments

@@
 by a parser\slash type\hyp checker reading from a text string.
-We leave aside the solved problem of writing such a parser\slash type\hyp checker,
+We leave aside the solved problem of writing such a 
+parser\slash type\hyp checker,
 whether using dependent types \citep{WalidICFP02} or not \citep{baars-typing}.
+\oleg{Should we really leave it aside, given that we do have the code:
+  IncopeTypecheck, and given Sumii-san's advice to emphasize the fact
+  that we do have the code. Please see my comment in the related work
+  section in the paragraph that deals with typing dynamic typing.}
 \fi
@@
 \ifshort\else
-\begin{code}
+\begin{code}[commandchars=\\\{\}]
 let testf3 = app (lam (varS varZ)) (b true)
-let testf3r = testf3 ()
+let testf3r = testf3 \underline{()}
 This expression has type unit but is here used with type 'a * 'b
@@
 factorial, and power.
+\oleg{There is now IncopeDB.hs that defines at least R, L and C
+ interpreters (no PE at present though) using DeBruijn indices. So we
+ can say that (some of) our development is available in DeBruijn
+ style. We won't talk about it in the paper though.}
 In contrast to \S\ref{intro}, we encode binding using higher-order
@@
 terms such as those read from a file.
+\oleg{First, I suggest we remove the subjunctive mood: `we would
+  need...' We have the code already, IncopeTypecheck.hs. We know we
+  can typecheck untyped terms so that we do the typechekcing only
+  once, and interpret many times. All these interpretation is guranteed to
+  be free from any type errors. And this holds even for higher-order
+  terms encoded in HOAS. I also suggest we drop
+  \cite{haskell-list}. We included the citation solely to show how 
+  we could go on. Both links in haskell-list are not actually
+  appropriate: first, the object language is \emph{first-order}
+  and second they use GADTs. Our IncopeTypecheck uses only
+  Typeable and existentials, and no GADTs.
+  Finally, we need to find a way to talk about IncopeTypecheck more
+  directly, probably in the Introduction -- Sumii-san has directly
+  suggested we emphasize that result. For some reason, people (Walid
+  and Sumii-san, for ones) consider it important. Incidentally, I
+  have the comparison of IncopeTypecheck with `Typing Dynamic
+  Typing', which we can use if needed.}
 \fi

### 2008-02-19 tagless/tagless-final.tex
Removed section 4.3 with the description of GADT. That section was removed from the APLAS final paper too

@@
 \texttt{type ('c,'dv)\,repr 
-	= S0 of ('c,'dv)\,R.repr~}\Verb+|+\texttt{~E0 of ('c,'dv)\,C.repr}.
+	= S0 of ('c,'dv)\,R.repr~}\Verb+|+\texttt{~D0 of ('c,'dv)\,C.repr}.
 Integer and boolean literals are immediate, present-stage
@@
 Figure~\ref{fig:ocaml} shows the new signature.
-% aplas: abbreviated figure
 \begin{figure}
 \begin{floatrule}
-\ifshort
 \smallskip
@@
 \medskip
-\else
-\begin{code2}
-module type Symantics = sig
-  type ('c,'sv,'dv) repr
-
-  val int : int  -> ('c,int,int) repr
-  val bool: bool -> ('c,bool,bool) repr
-
-  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-            -> ('c,'x,'da -> 'db) repr
-  val app : ('c,'x,'da -> 'db) repr
-            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
-                                'da -> 'db) repr as 'x)
-
-  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
-  val if_ : ('c,bool,bool) repr
-            -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
-end
-\end{code2}
-\fi
 \end{floatrule}
@@
 \begin{floatrule}
-\ifshort
 \begin{code2}
@@
 \medskip
-\else
-\begin{code2}[commandchars=\@\[\]]
-module P = struct
-  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
-  let abstr {dy = x} = x
-  let pdyn x = {st = None; dy = x}
-
-  let int  (x:int ) = {st = Some (R.int  x); dy = C.int  x}
-  let bool (x:bool) = {st = Some (R.bool x); dy = C.bool x}
-
-  let add e1 e2 = match e1, e2 with
-                  | {st = Some 0}, e | e, {st = Some 0} -> e
-                  | {st = Some m}, {st = Some n} -> int (R.add m n)
-                  | _ -> pdyn (C.add (abstr e1) (abstr e2))
-  let if_ eb et ee = match eb with
-                     | {st = Some b} -> if b then et () else ee ()
-                     | _ -> pdyn (C.if_ (abstr eb) (fun () @!->@! abstr (et @!()))
-                                                   (fun () @!->@! abstr (ee @!())))
-
-  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
-  let app ef ea = match ef with
-                  | {st = Some f} -> f ea
-                  | _ -> pdyn (C.app (abstr ef) (abstr ea))
-  let fix f = let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
-              in let rec self = function
-                                | {st = Some @!_} as e -> app (f (lam self)) e
-                                | e -> pdyn (C.app fdyn (abstr e))
-                 in {st = Some self; dy = fdyn}
-end
-\end{code2}
-\fi
 \end{floatrule}
@@
 way'', that is, whenever the 
-argument is static, it unfolds |fix| rather than residualize it.
-\ifshort
+argument is static, we unfold |fix| rather than residualize it.
 In the accompanying source code is a conservative
 alternative |P.fix| that unfolds recursion only once, then residualizes.
-\else
-A conservative alternative is to unfold recursion only once, then residualize:
-\begin{code}
-let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
-\end{code}
-\fi
 Many sophisticated approaches have been developed to decide how much to unfold
@@
 a dynamic (future-stage) term (using the compiler~|C|).  To
-distinguish between static and dynamic terms, one might define
-\begin{code}
-data P0 t = S0 (R t) | E0 (C t)
-\end{code}
-To extract a dynamic term from this type, we create the functions
-\begin{code}
-abstrI :: P0 Int -> C Int
-abstrI (S0 r) = int (unR r)
-abstrI (E0 c) = c
-
-abstrB :: P0 Bool -> C Bool
-abstrB (S0 r) = bool (unR r)
-abstrB (E0 c) = c
-\end{code}
-We then start to define |P0| as an instance of the |Symantics| class.
+distinguish between static and dynamic terms, we might try to define
+|repr| in the partial evaluator (module |P0|) as
+\begin{code}
+type ('c,'dv) repr = S0 of ('c,'dv) R.repr | D0 of ('c,'dv) C.repr
+\end{code}
+(the digit zero in the names of the phase tags |S0| and |D0| is the
+version suffix).
+To extract a dynamic term from this type, we create the function
+\begin{code}
+let abstrI0 (e:('c,int) repr) : ('c,int) C.repr =
+  match e with
+  | S0 e -> C.int e
+  | D0 e -> e
+\end{code}
+and the similar function |abstrB0| for dynamic boolean terms. Here, 
+|C.int| is used to convert from the static term |('c,int) R.repr|, 
+which is just |int|, to the dynamic term. We can now define the following
+components of module |P0| required by the |Semantics| signature:
+\begin{code}
+let int  (x:int)  = S0 (R.int x)
+let bool (x:bool) = S0 (R.bool x)
+let add e1 e2 = 
+  match (e1,e2) with
+  | (S0 e1, S0 e2) -> S0 (R.add e1 e2)
+  | (e1, e2)       -> D0 (C.add (abstrI0 e1) (abstrI0 e2))
+\end{code}
 Integer and boolean literals are immediate, present-stage
-values. Addition yields a static term (using~|R|) if and only if both operands
-are static; otherwise we extract the dynamic terms from the operands and
-add them using~|C|.
-\begin{code}
-instance Symantics P0 where
-  int  x = S0 (int x)
-  bool x = S0 (bool x)
-  add (S0 e1) (S0 e2) = S0 (add e1 e2)
-  add     e1      e2  = E0 (add (abstrI e1) (abstrI e2))
-\end{code}
-Thus, the two uses of |add| above refer to
-|add| for~|R| and |add| for~|C|.
+values. Addition yields a static term (using~|R.add|) if and only 
+if both operands are static; otherwise we extract the dynamic terms 
+from the operands and add them using~|C.add|.
 
 Whereas |mul| and |leq| are as easy to define as |add|, we encounter
-a problem with |if_|.  If the first argument to |if_| is a dynamic term
-(of type |C Bool|), the second a static term (of type |R a|) and the third a
-dynamic term (of type |C a|), then we need to convert
+a problem with |if_|.  Suppose that the first argument to |if_| 
+is a dynamic term
+(of type |('c,bool) C.repr|), the second a static term 
+(of type |('c,'a) R.repr|), and the third a
+dynamic term (of type |('c,'a) C.repr|). We then need to convert
 the static term to dynamic, but there is no polymorphic ``lift''
-function, of type |a -> C a|, to send a value to the future stage
-\citep{xi-guarded,WalidPOPL03}. (By the way, if we
+function, of type |'a -> ('c,'a) C.repr|, to send a value to the future stage
+\citep{xi-guarded,WalidPOPL03}.
+\begin{comment}
+(By the way, if we
 were to add polymorphic \texttt{lift} to the type class
@@
 f = app (lift f)~}.)
-
-Our |Symantics| class only includes separate lifting methods |bool| and
+\end{comment}
+
+Our |Symantics| only includes separate lifting methods |bool| and
 |int|, not a parametrically polymorphic lifting method, for good reason:
@@
 Asai's technique \citep{asai-binding-time,sumii-hybrid}:
-build a dynamic term
-alongside every static term.
+build a dynamic term alongside every static term.
 
@@
 
-\begin{SaveVerbatim}{P1}
-data P1 t = P1 (Maybe (R t)) (C t)
-\end{SaveVerbatim}
-
-We switch to the Haskell data type
-\UseVerbatim{P1}
-so that a partially evaluated term |P1 t| always contains a dynamic
+We start building the partial evaluator anew (module |P1|) and
+switch to the data type
+\begin{code}
+type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr
+\end{code}
+so that a partially evaluated term always contains a dynamic
 component and sometimes contains a static component.  By 
 distributivity, the two
-alternative constructors of a |Maybe| value, |Just| and |Nothing|,
+alternative constructors of an |option| value, |Some| and |None|,
 tag each partially evaluated term with a phase: either present or
@@
 can now define the polymorphic function
-\begin{code}
-abstr1 :: P1 t -> C t
-abstr1 (P1 _ dyn) = dyn
-\end{code}
+|let abstr1 (P1 (_,dyn)) = dyn| of type
+\texttt{('c,'dv) repr -> ('c,'dv) C.repr}
 to extract it without requiring polymorphic lift into~|C|.  We then try
@@
 \begin{code}
-instance Symantics P1 where
-  int  x = P1 (Just (int  x)) (int  x)
-  bool b = P1 (Just (bool b)) (bool b)
-
-  add (P1 (Just n1) _) (P1 (Just n2) _) = int (unR (add n1 n2))
-  add e1 e2 = P1 Nothing (add (abstr1 e1) (abstr1 e2))
-  -- mul and leq are analogous and elided
-
-  if_ (P1 (Just s) _) et ef = if unR s then et else ef
-  if_ eb et ef = P1 Nothing
-                    (if_ (abstr1 eb) (abstr1 et) (abstr1 ef))
-\end{code}
-When we come to functions, however, we stumble.  According to our
+module P1 : Symantics = struct
+  let int (x:int) = P1 (Some (R.int x), C.int x)
+  let add e1 e2 = match (e1,e2) with
+    | (P1 (Some n1,_),P1 (Some n2,_)) -> int (R.add n1 n2)
+    | _ -> P1 (None,(C.add (abstr1 e1) (abstr1 e2)))
+  let if_ = function
+    | P1 (Some s,_) -> fun et ee -> if s then et () else ee ()
+    | eb -> fun et ee -> P1 (None, C.if_ (abstr1 eb) 
+                                   (fun () -> abstr1 (et ()))
+                                   (fun () -> abstr1 (ee ())))
+\end{code}
+However, we stumble on functions.  According to our
 definition of~|P1|, a partially evaluated object function, such as the
-identity $\fun{x}x$ embedded in Haskell as |lam (\x -> x)|\texttt{ ::
-}|P1 (a -> a)|, consists of a dynamic part (type |C (a -> a)|) and
-maybe a static part (type |R (a -> a)|).  The dynamic part is useful
+identity $\fun{x}x$ embedded in OCaml as |lam (fun x -> x)|\texttt{ :
+}|('c,'a->'a) P1.repr|, consists of a dynamic part 
+(type |('c,'a->'a) C.repr|) and
+maybe a static part (type |('c,'a->'a) R.repr|).  The dynamic part is useful
 when this function is passed to another function that is only
@@
 appears as the argument of |app|.  To do so, we have to incorporate |f|
-as it is into the |P1| data structure: applying the type constructor
-|P1| to a function type |a -> b| should yield one of
-\begin{code}
-data P1 (a -> b) = S1 (P1 a -> P1 b) | E1 (C (a -> b))
-data P1 (a -> b) = P1 (Maybe (P1 a -> P1 b)) (C (a -> b))
-\end{code}
-unlike applying |P1| to |Int| or |Bool|.
+as it is into the |P1.repr| data structure: the representation
+for a function type |'a->'b| should be one of
+\begin{code}
+S1 of ('c,'a) repr -> ('c,'b) repr | E1 of ('c,'a->'b) C.repr
+P1 of (('c,'a) repr -> ('c,'b) repr) option * ('c,'a->'b) C.repr
+\end{code}
+unlike |P1.repr| of |int| or |bool|.
 That is, we need a nonparametric data type, something akin to
@@
 \citet{oliveira-typecase} dub the \emph{typecase} design pattern.
-Thus, typed partial evaluation, like typed CPS transformation,
+Thus, typed partial evaluation, like typed CPS transformation
+(see \S\ref{S:CPS}),
 inductively defines a map from source types to target types that
-performs case distinction on the source type. The connection with CPS
-is not an accident, as we shall see in~\S\ref{S:CPS}.
-
-\subsection{Eliminating tags from typecase}
-\label{S:PE-GADT}
-
-Two common ways to provide typecase in Haskell are
-GADTs and type-class functional dependencies
-\citep{oliveira-typecase}.  These
-methods are equivalent, and here we use GADTs; |incope1.hs|
-in the accompanying source code shows the latter.
-We introduce a GADT with four data constructors.
-\begin{code}
-data P t where
-  VI :: Int  -> P Int
-  VB :: Bool -> P Bool
-  VF :: (P a -> P b) -> P (a -> b)
-  E  :: C t -> P t
-\end{code}
-The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
-in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
-the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
-operand of type |P a -> P b| rather than |a -> b|.  We define a function
-like |abstr1| above to extract a future-stage computation from a 
-value of type |P t|.
-\begin{code}
-abstr :: P t -> C t
-abstr (VI i) = int i
-abstr (VB b) = bool b
-abstr (VF f) = lam (abstr . f . E)
-abstr (E x)  = x
-\end{code}
-The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
-case uses the method |lam| of the |C| interpreter to compile~|f|.
-
-We may now make |P| an instance of
-|Symantics| and implement the partial evaluator as follows. We elide
-|mul|, |leq|, |if_|, and |fix|.
-\begin{code}
-instance Symantics P where
-  int x               = VI x
-  bool b              = VB b
-  add (VI n1) (VI n2) = VI (n1 + n2)
-  add e1 e2           = E (add (abstr e1) (abstr e2))
-  lam                 = VF
-  app (VF f) ea       = f ea
-  app (E f)  ea       = E (app f (abstr ea))
-\end{code}
-The implementations of |int|, |bool|, and |add| are like 
-in~\S\ref{S:PE-problem}. The interpretation of |lam f| is |VF f|, 
-which just wraps the HOAS function |f|. 
-We can always compile |f| to a code value,
-but we delay it to apply |f| to concrete arguments. The interpretation of
-|app ef ea| checks to see if |ef| is such a delayed
-HOAS function |VF f|. If it is, we apply |f| to the
-concrete argument |ea|, giving us a chance to perform static
-computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
-dynamic value |E f|, we residualize.
-
-This solution using GADTs works but is not quite satisfactory. First, it
-cannot be ported to MetaOCaml, as GADTs are unavailable there.  Second,
+performs case distinction on the source type. In Haskell, typecase
+can be equivalently implemented either with GADTs or with
+type-class functional dependencies
+\citep{oliveira-typecase}. The accompanying code shows both
+approaches (|Incope.hs| and |incope1.hs|), 
+neither portable to OCaml. In addition,
 the problem of nonexhaustive pattern\hyp matching reappears in
-|app| above: the type |P t| has four constructors, of which the pattern in
-|app| matches only |VF| and~|E|. One may say that the
-constructors |VI| and |VB| obviously cannot occur because they do not
-construct values of type |P (a -> b)| as required by the type of |app|.
-Indeed, the metalanguage implementation could reason thus:
-if we use inductive families (as in Coq) or logical
-  frameworks with canonical forms (as in Twelf with its coverage checker),
-  we can prove the pattern matching to be exhaustive.
-Then again, the metalanguage implementation may not reason thus:
-GHC cannot and issues warnings.
-Although this point may seem minor, it is the heart of
-the tagging problem and the purpose of tag elimination. A typed tagged
-interpreter contains many pattern\hyp matching forms that look partial
-but never fail in reality. The
-goal is to make this exhaustiveness \emph{syntactically} apparent.
+the GADT approach because GHC 6.6.1 cannot see that a particular
+type of a GADT value precludes certain constructors.
+Thus GADTs fail to
+make it \emph{syntactically} apparent that pattern matching is exhaustive.
+
+%% \subsection{Eliminating tags from typecase}
+%% \label{S:PE-GADT}
+
+%% Two common ways to provide typecase in Haskell are
+%% GADTs and type-class functional dependencies
+%% \citep{oliveira-typecase}.  These
+%% methods are equivalent, and here we use GADTs; |incope1.hs|
+%% in the accompanying source code shows the latter.
+%% We introduce a GADT with four data constructors.
+%% \begin{code}
+%% data P t where
+%%   VI :: Int  -> P Int
+%%   VB :: Bool -> P Bool
+%%   VF :: (P a -> P b) -> P (a -> b)
+%%   E  :: C t -> P t
+%% \end{code}
+%% The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
+%% in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |D0|).  However,
+%% the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
+%% operand of type |P a -> P b| rather than |a -> b|.  We define a function
+%% like |abstr1| above to extract a future-stage computation from a 
+%% value of type |P t|.
+%% \begin{code}
+%% abstr :: P t -> C t
+%% abstr (VI i) = int i
+%% abstr (VB b) = bool b
+%% abstr (VF f) = lam (abstr . f . E)
+%% abstr (E x)  = x
+%% \end{code}
+%% The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
+%% case uses the method |lam| of the |C| interpreter to compile~|f|.
+
+%% We may now make |P| an instance of
+%% |Symantics| and implement the partial evaluator as follows. We elide
+%% |mul|, |leq|, |if_|, and |fix|.
+%% \begin{code}
+%% instance Symantics P where
+%%   int x               = VI x
+%%   bool b              = VB b
+%%   add (VI n1) (VI n2) = VI (n1 + n2)
+%%   add e1 e2           = E (add (abstr e1) (abstr e2))
+%%   lam                 = VF
+%%   app (VF f) ea       = f ea
+%%   app (E f)  ea       = E (app f (abstr ea))
+%% \end{code}
+%% The implementations of |int|, |bool|, and |add| are like 
+%% in~\S\ref{S:PE-problem}. The interpretation of |lam f| is |VF f|, 
+%% which just wraps the HOAS function |f|. 
+%% We can always compile |f| to a code value,
+%% but we delay it to apply |f| to concrete arguments. The interpretation of
+%% |app ef ea| checks to see if |ef| is such a delayed
+%% HOAS function |VF f|. If it is, we apply |f| to the
+%% concrete argument |ea|, giving us a chance to perform static
+%% computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
+%% dynamic value |E f|, we residualize.
+
+%% This solution using GADTs works but is not quite satisfactory. First, it
+%% cannot be ported to MetaOCaml, as GADTs are unavailable there.  Second,
+%% the problem of nonexhaustive pattern\hyp matching reappears in
+%% |app| above: the type |P t| has four constructors, of which the pattern in
+%% |app| matches only |VF| and~|E|. One may say that the
+%% constructors |VI| and |VB| obviously cannot occur because they do not
+%% construct values of type |P (a -> b)| as required by the type of |app|.
+%% Indeed, the metalanguage implementation could reason thus:
+%% if we use inductive families (as in Coq) or logical
+%%   frameworks with canonical forms (as in Twelf with its coverage checker),
+%%   we can prove the pattern matching to be exhaustive.
+%% Then again, the metalanguage implementation may not reason thus:
+%% GHC cannot and issues warnings.
+%% Although this point may seem minor, it is the heart of
+%% the tagging problem and the purpose of tag elimination. A typed tagged
+%% interpreter contains many pattern\hyp matching forms that look partial
+%% but never fail in reality. The
+%% goal is to make this exhaustiveness \emph{syntactically} apparent.
 
@@
 
-Let us re-examine the problem in~\S\ref{S:PE-problem}. 
-\oleg{If we drop the previous Section, the phrase above should be changed.}
-What we
+Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
 would ideally like is to write
 \begin{code}
-data P t = P (Maybe (repr_pe t)) (C t)
+type ('c,'dv) repr = P1 of (repr_pe ('c,'dv)) R.repr option * ('c,'dv) C.repr
 \end{code}
@@
 \begin{code}
-repr_pe Int      = Int
-repr_pe Bool     = Bool
-repr_pe (a -> b) = P a -> P b
+repr_pe ('c,int)    = ('c,int)
+repr_pe ('c,bool)   = ('c,bool)
+repr_pe ('c,'a->'b) = ('c,'a) repr -> ('c,'b) repr
 \end{code}
 Although we can use type classes to define this type function
-in Haskell, that solution is not portable to MetaOCaml. However,
+in Haskell, that is not portable to MetaOCaml. However,
 these three typecase alternatives are already present in existing
@@
 A simple and portable solution thus emerges: we bake |repr_pe| 
-into the signature |Symantics|. Switching to MetaOCaml to demonstrate
-this portability, we recall from Figure~\ref{fig:ocaml-simple} 
-in~\S\ref{encoding} that the |repr| type
-constructor took two arguments |'c| and~|'dv|. We add an argument
-|'sv| for the result of applying |repr_pe| to~|'dv|.
+into the signature |Symantics|. 
+We recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that 
+the |repr| type constructor took two arguments |'c| and~|'dv|.
+We add an argument |'sv| for the result of applying |repr_pe| to~|'dv|.
 Figure~\ref{fig:ocaml} shows the new signature.
@@
 \end{floatrule}
-\caption{Our partial evaluator \texttt{mul} and \texttt{leq}
+\caption{Our partial evaluator (\texttt{mul} and \texttt{leq}
   are elided)}
@@
 Its type |repr| literally expresses the type equation for |repr_pe| above.
-The function |abstr|, as in~\S\ref{S:PE-GADT},
+The function |abstr|
 extracts a future-stage code value from the result of
@@
 any static ones to avoid polymorphic lift.
+
+The static portion of the interpretation of |lam f| is |Some f|, 
+which just wraps the HOAS
+function |f|. The interpretation of |app ef ea| 
+checks to see if |ef| is such a wrapped
+HOAS function. If it is, we apply |f| to the
+concrete argument |ea|, giving us a chance to perform static
+computations (see the example below). If
+|ef| has only a dynamic part, we residualize.
 
@@
 way'', that is, whenever the 
-argument is static, it unfolds |fix| rather than residualize it.
+argument is static, we unfold |fix| rather than residualize it.
 A conservative alternative is to unfold recursion only once, then residualize:
@@
 
-\noindent Unlike the GADT approach in~\S\ref{S:PE-GADT}, all pattern\hyp
+\noindent All pattern\hyp
 matching in~|P| is \emph{syntactically} exhaustive, so it is patent to the

### 2008-02-26 tagless/tagless-final.tex
Added ref and discussion of Lars Berkedal's work.

@@
 in~\S\ref{S:interpreter-RL} may just as well be written in our object
-language:~\Fig.\ref{fig:self-eval}.
+language:~Fig.~\ref{fig:self-eval}.
 \begin{figure}
@@
 
+\oleg{The following are almost unedited excerpts from our
+  correspondence with Sumii about Birkedal's thesis. We definitely
+  should include this discussion and the reference to his thesis.}
+\Citet{Birkedal-PE-ML} have hand-written a compiler generator (cogen)
+for the Core Standard ML, which transforms a program into its tagless
+generating extension (and therefore serves as a tagless partial
+evaluator, intepreter, and compiler in our sense). They still have to
+use the "universal type" at the cogen level, which is removed from
+compile-time, so that the end result is tagless indeed.  The main
+difference between the work of \Citet{Birkedal-PE-ML} and ours is how
+\emph{early} it becomes \emph{manifest} to the metalanguage
+implementation that our code is safe.  The fact that
+\Citet{Birkedal-PE-ML} 's generating extension never generates a
+specialized program that is ill-typed is not manifest to the SML
+compiler until the SML compiler sees each particular specialized
+program.  Moreover and analogously, the fact that
+\Citet{Birkedal-PE-ML} 's cogen never generates a generating extension
+that is ill-typed (or that may generate a specialized program that is
+ill-typed) is not manifest to the SML compiler until the SML compiler
+sees each particular generating extension (or each particular
+specialized program).  This early and manifest type safety is the
+payoff of our treating an object program not as a data structure but
+as calls to abstract-syntax constructor functions.  The tradeoff, as
+we currently see it, is that we have to swear off intensional analysis
+of code. \oleg{Unless we convert to our representation to GADT, which
+  is always possible, if GADTs are available.}
+
 % The following paragraph was not in the version submitted to APLAS,

### 2008-02-27 tagless/talk.tex
Final talk at APLAS 2007

@@
 \usepackage{prooftree1} \proofrulebaseline=1.5ex \proofdotseparation=.75ex
-\usepackage{microtype}
 \usepackage{calc}
@@
 
-% To get the page numbers to stay constant on a frame [dpt]
+% Record frame numbers as PDF page labels
 \mode<presentation>
@@
 \tikzstyle{every picture}=[>=angle 60,semithick]
+
+\begin{frame}{The goal of this talk}
+    \begin{center}
+        Write your interpreter by deforesting the object language,
+        to~exhibit more static safety in a simpler type system.
+    \end{center}
+\end{frame}
 
@@
 \only<5-8,11-14>{\text{\textcolor{structure.fg}{Object term}}}%
-\only<16>{\rlap{\textcolor{structure.fg}{ML higher-order functor}}}
-& \only<5-8,11-15>{\textcolor{structure.fg}{\only<-14>{\llap{\tikz\draw[->](0,0)--(1.5,0)node[anchor=mid,inner sep=0pt]{\vphantom{O}};\:}}\only<3-8>{Haskell term}\only<9-15>{ML higher-order functor}}} \\
+\only<16>{\rlap{\textcolor{structure.fg}{ML functor}}}
+& \only<5-8,11-15>{\textcolor{structure.fg}{\only<-14>{\llap{\tikz\draw[->](0,0)--(1.5,0)node[anchor=mid,inner sep=0pt]{\vphantom{O}};\:}}\only<3-8>{Haskell term}\only<9-15>{ML functor}}} \\
 \multicolumn{2}{@{}l@{}}{\only<16>{\alert{\texttt{module POWER (S:Symantics) = struct open S}}}} \\
@@
 \onslide+<9-10>
-type ('c,'a) \\
+type ('c,'a) repr \\
 \ \ \ = ('c,'a) code \\
@@
 
+\begin{frame}{CPS transformations}
+\begin{center}
+\begin{tikzpicture}[auto]
+    \node at (0,2) [lang] (lang1) {Simply typed $\lambda$-calculus};
+    \node at (0,0) [lang] (lang2) {Simply typed $\lambda$-calculus};
+    \draw [->] (lang1) to [bend right=30] node [swap] {CBV} (lang2);
+    \draw [->] (lang1) to [bend left=30]  node        {CBN} (lang2);
+    \draw [->] (lang2) -- +(-135:2cm) node [inner sep=0pt, anchor=north east] {evaluate};
+    \draw [->] (lang2) -- +(-105:2cm) node [inner xsep=0pt, anchor=north east] {compile\hspace*{-.75em}};
+    \draw [->] (lang2) -- +(-75:2cm)  node [inner xsep=0pt, anchor=north west] {\hspace*{-.75em}specialize};
+    \draw      (lang2)    +(-45:2cm)  node {\dots};
+\end{tikzpicture}
+\end{center}
+
+    Payoffs: evaluation order independence, mutable state
+\end{frame}
+
+\begin{frame}[fragile]{Other benefits}
+\setbeamercovered{transparent=20}
+\begin{block}{Supports initial type-checking}
+    Type-check once, even under~$\lambda$, then interpret many times.
+
+\begin{semiverbatim}
+  FilePath -> \alert{Maybe} (exists a. Typeable a =>
+                     forall repr. Symantics repr =>
+                     repr a)
+\end{semiverbatim}
+
+    ``Typing dynamic typing'' (ICFP 2002) works.  We have the code.
+\end{block}
+
+\pause
+\begin{block}{Preserves sharing in the metalanguage}
+    Compute the interpretation of a repeated object term once, then use it many
+    times.
+
+    \smallskip
+    $2\times3 + 2\times3$\hfill\texttt{let n = mul (int 2) (int 3) in add n n}
+\end{block}
+
+\pause
+\begin{block}{Embed one object language in another}
+\texttt{(Symantics repr, Symantics' repr') => repr (repr' Int)}\hspace*{-1em}
+\end{block}
+\end{frame}
+
 \begin{frame}{Conclusion}
-    Early obvious safety
-
-    Initial type-checking can be done
-
-    Accommodates interpretations with type-indexed types
-
-    Preserves sharing
-
-    Acknowledge prior-work authors in slides
+    Write your interpreter by deforesting the object language
+    \begin{itemize}
+    \item An abstract data type family
+    \item Type-indexed types
+    \end{itemize}
+    Exhibit more static safety in a simpler type system
+    \begin{itemize}
+    \item Early, obvious guarantees
+    \item Supports initial type-checking
+    \item Preserves sharing in the metalanguage
+    \item Embed one object language in another
+    \end{itemize}
 \end{frame}

### 2008-03-30 tagless/talk.tex
NJPLS talk

@@
 \newsavebox\oleg    \sbox\oleg   {\authorstack{Oleg Kiselyov}{FNMOC}{oleg@pobox.com}}
-\newsavebox\ccshan  \sbox\ccshan {\authorstack{Chung-chieh Shan}{Rutgers University}{ccshan@rutgers.edu}}
+\newsavebox\ccshan  \sbox\ccshan {\color{red}\authorstack{Chung-chieh Shan}{Rutgers University}{ccshan@rutgers.edu}}
 
@@
 \author[Jacques Carette and Oleg Kiselyov and Chung-chieh Shan]{\usebox\carette \and \usebox\oleg \and \usebox\ccshan}
-\date{APLAS\\30 November 2007}
+\date{NJPLS\\7 March 2008}
 
@@
                         Haskell constructor instances\\
-                        or ML functors
+                        or ML modules
                         \end{tabular}}}
@@
 \subsection{As a constructor class in Haskell}
-\subsection{As a functor signature in ML}
+\subsection{As a module signature in ML}
 
-\begin{frame}[t,fragile]{The object language\only<3-8>{ as a constructor class}\only<9-16>{ as a functor signature}}
+\begin{frame}[t,fragile]{The object language\only<3-8>{ as a constructor class}\only<9-16>{ as a module signature}}
 \setbeamercovered{transparent=20}

### 2008-06-03 SCP2008/paper/metamonads.tex
Added a section on GE

@@
 \ead{oleg@pobox.com}
-\ead[url]{http://pobox.com/\textasciitilde oleg/ftp/}
-
-\hyphenation{ne-ces-sa-ry}
-\hyphenation{ana-lysis theo-ry}
+\ead[url]{http://okmij.org/ftp/}
+
+\hyphenation{ne-ces-sa-ry ana-lysis theo-ry}
 
@@
 in the upper triangle.
+\oleg{Section XXX demonstrates one particular GE algorithm in detail,
+  for further reference.}
 
@@
 clearer error messages.
+
+\section{Gaussian Elimination}\label{GA-algorithm}
+
+For detailed reference, we present one particular Gaussian Elimination
+algorithm, for an in-place LU-decomposition of an integer matrix with
+full pivoting, returning the U-factor, determinant and rank. The
+$n\times m$-matrix is represented as a flat vector, with 0-indexed
+elements laid out in row-major format (C-style). The code below is
+hand-written, implementing the typical pseudo-code given in textbooks on
+Numerical Analysis \oleg{I used wikipedia; need a better references}.
+
+In OCaml, the matrix is represented by the value of the type
+|'a container2dfromvector|:
+\begin{code}
+type 'a container2dfromvector = {arr:('a array); n:int; m:int}
+
+let swap a i j =
+    let t = a.(i) in begin a.(i) <- a.(j); a.(j) <- t; end
+
+let swap_rows a (n,m) (r,c) i =
+  let row_r = r*m in			(* Beginning of row r *)
+  let row_i = i*m in 			(* Beginning of row i *)
+  for k = c to m-1 do
+      swap a (row_r + k) (row_i + k)
+  done
+
+let swap_cols a (n,m) c j =
+  let end_vector = n * m in
+  let rec loop col_c col_j =
+    if col_j < end_vector then
+      begin
+        swap a col_c col_j;
+        loop (col_c + m) (col_j + m)
+      end
+  in loop c j
+\end{code}
+We refer to the $a_{ij}$-the element of the matrix as |a.(i*m+j)|,
+where |a| is the array component of the record |container2dfromvector|.
+The figure defines three auxiliary functions, typically used in the
+textbook pseudo-code: |swap a i j| swaps two elements; 
+|swap_cols a (n,m) c j| swaps the column |c| with the column |j| in
+the |(n,m)|-matrix |a|. The function 
+|swap_rows a (n,m) (r,c) i| swaps the row |r| with the row |i| in 
+the non-yet examined portion of the matrix, 
+rectangular block |(r,c)-(n,m)|. We do not touch the 
+elements to the left of the column |c| because they are all zeros. 
+Since we know the layout of the matrix, we avoid 2D index
+computations.
+
+\begin{figure}
+\begin{code}[numbers=left,numberblanklines=false,fontsize=\scriptsize]
+let find_pivot a (n,m) r c =
+  let pivot = ref None in               (* ((i,j), pivot_val) option *)
+  begin
+  for i = r to n-1 do
+    for j = c to m-1 do
+      let cur = a.(i*m+j) in
+      if not (cur == 0) then
+          match !pivot with
+          | Some (_,oldpivot) ->
+              if abs oldpivot > abs cur then
+                 pivot := Some ((i,j), cur)
+          | None -> pivot := Some ((i,j),cur)
+  done; done;
+  !pivot
+  end
+
+let ge = fun a_orig ->
+    let r = ref 0 in                    (* current row index, 0-based *)
+    let c = ref 0 in                    (* current column index, 0-based *)
+    let a = Array.copy (a_orig.arr) in  (* to avoid clobbering A, save it *)
+    let m = a_orig.m in                 (* the number of columns *)
+    let n = a_orig.n in                 (* the number of rows *)
+    let det_sign = ref 1 in             (* Accumulate sign and magnitude *)
+    let det_magn = ref 1 in             (*   of the determinant *)
+    while !c < m && !r < n do
+        (* Look for a pivot *)
+        let pivot = find_pivot a (n,m) !r !c in
+        let piv_val = (match  pivot with
+        | Some ((piv_r, piv_c),piv_val) ->
+            if piv_c <> !c then
+               begin
+                 swap_cols a (n,m) !c piv_c;
+                 det_sign := - !det_sign (* flip the sign of the det *)
+               end;
+            if piv_r <> !r then
+               begin
+                 swap_rows a (n,m) (!r,!c) piv_r;
+                 det_sign := - !det_sign (* flip the sign of the det *)
+               end;
+            Some piv_val
+        | None -> None) in
+        (* now do the row-reduction over the (r,c)-(n,m) block *)
+        (match piv_val with
+        | Some a_rc -> begin
+            for ii = !r+1 to n-1 do
+                let cur = a.(ii*m + !c) in
+                if not (cur == 0)  then
+                  begin
+                    for j = !c+1 to m-1 do
+                      (* fraction-free elimination *)
+                      a.(ii*m+j) <- (a.(ii*m+j) * a_rc - a.(!r*m+j) * cur) 
+                                    / ! det_magn
+                    done;
+                    a.(ii*m+ !c) <- 0
+                  end;
+            done;
+            det_magn := a_rc;
+            r := !r + 1                 (* advance the rank only if pivot > 0*)
+          end
+        | None -> det_sign := 0);
+        c := !c + 1
+    done;
+    (* Final result *)
+    ({arr=a; n=n; m=m},                 (* The matrix now has the U factor*)
+    (if      !det_sign = 0 then  0      (* Compute the signed det *)
+     else if !det_sign = 1 then  !det_magn 
+     else                        (- !det_magn)),
+    !r)                                 (* Rank *)
+\end{code}
+\caption{Fraction-free in-place Gaussian-elimination over an integer
+  matrix}\label{fig:GE-algorithm}
+\end{figure}
+As typically presented in the textbooks, the main algorithm depends on
+the separately-defined function |pivot| to find a pivot. We use
+full-pivoting, i.e., searching the non-yet examined portion of the matrix, 
+the rectangular block |(r,c)-(n,m)|, for the element with the 
+maximum absolute value. The function returns the value of the
+found pivot and its location; the return value is an option type with
+|None| indicating that all examined elements are zeros. In the main
+algorithm, after we found the pivot, we swap the current column 
+with the pivot column, and swap the current row with the pivot row. 
+After the swap, the $a_{rc}$ element of the matrix is the pivot.
+Swapping two rows or two columns changes the sign of the determinant.
+After the swapping, the algorithm does so-called row-reduction
+over the |(r,c)-(n,m)| block of the matrix containing not-yet
+processed elements. We implement the \emph{fraction-free} algorithm:
+the division operation in line XXX assuredly divides two integers
+without remainder. The fraction-free algorithm requires accumulation
+of the determinant \oleg{Citation needed!}.
 

### 2008-06-03 SCP2008/paper/metamonads.tex
A few notes referencing the hand-written GE code

@@
 
-\section{Gaussian Elimination}\label{GA-algorithm}
+\section{Gaussian Elimination}\label{GE-algorithm}
+
+\oleg{I wonder should we always refer to GE rather than LU-elimination
+  in our paper? GE is a particulat instance of LU; one can accomplish
+  LU in other ways than GE. Also, GE can be used for other means that
+  LU, e.g., just to compute the determinant or to solve linear equations.
+  So, it seems more properly to say we do GE. If we use GE and LU
+  almost synonymously, as it seems to be the case in Mathematica, we
+  should note that in a footnote.}
 
@@
 
-We start by abstracting out domains. We have already seen the simplest
-case of domain abstraction in |param_code1|, which took
+We start by abstracting out domains. \oleg{Add: clearly the basic
+  structure of the algorithm, Fig.~\ref{fig:GE-algorithm} remains the
+  same for integer, float, etc. matrices.} We have already seen the
+simplest case of domain abstraction in |param_code1|, which took
 code-generators such as |plus| and |one| as arguments.  We need far
@@
 extensibility, type abstraction and constraints, and especially
-parameterized structures (\emph{functors}).  We define the type,
-the signature |DOMAIN|, which different domains must satisfy:
+parameterized structures (\emph{functors}).  We define the type, the
+signature |DOMAIN|, which different domains must satisfy:
 
@@
 
+\oleg{Say that CONTAINER2D is an abstraction of a particular
+  container, flat vector, used in the algorithm in
+  \S\ref{GE-algorithm}.}
 The signature |CONTAINER2D| below specifies that a container must provide
@@
 depending on the actual representation of the container while not
-burdening the users of containers with efficiency details.  Note that
+burdening the users of containers with efficiency details (see
+\S\ref{GE-algorithm} for an example of such an optimiation).  Note that
 this module type is also abstract over the actual representation 
@@
 value of the determinant, and converting the tracking state
-to the final determinant value of type |outdet|. LU of a
+to the final determinant value of type |outdet| (see
+Fig.~\ref{fig:GE-algorithm} for an example).
+LU of a
 floating-point matrix with no determinant tracking uses the
@@
 aspect is complex because tracking variables, if any, are to be declared at the
-beginning of LU; the sign
+beginning of the code (see lines 22-23 (XXX) of
+Fig.~\ref{fig:GE-algorithm}); the sign
 of the determinant has to be updated on each row or column
-permutation; the value of the determinant should be updated per each
-pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
+permutation (lines 32 and 37 of the figure XXX);
+the value of the determinant should be updated per each
+pivoting (line 56). We use |lstate| to pass the tracking state, e.g., a piece of
 code for the value of the type |Dom.v ref|, among
@@
 generates code to compute the final, signed determinant value.
+The generated code closely corresponds to the lines dealing with 
+|det_magn| and |det_sign| in Fig.~\ref{fig:GE-algorithm}.
 \begin{code}
@@
 \begin{code}
-type 'a wmatrix = {matrix: 'a C.vc; numrow: ('a,int) abstract; 
-                   numcol: ('a,int) abstract}
+type 'a wmatrix = \{matrix: 'a C.vc; numrow: ('a,int) abstract; 
+                   numcol: ('a,int) abstract\}
 \end{code}
@@
 resulting function returns a triple |int array * int * int| of the
-U-factor, determinant and the rank.
+U-factor, determinant and the rank. It is instructive to compare the
+generated code with the corresponding code in
+Fig.~\ref{fig:GE-algorithm}, the hand-written implementation of the
+textbook pseudo-code; the only difference is the naming of
+variables and the inlining of pivoting and swapping functions.
 
@@
 A careful reading of this code will reveal that the core of the 
-Gaussian Elimination algorithm is still visible in this code generator:
+Gaussian Elimination algorithm (Fig.~\ref{fig:GE-algorithm}) 
+is still visible in this code generator:
 the |forward_elim| function iterates over the columns and rows of the

### 2008-06-05 SCP2008/paper/metamonads.tex
a few remarks and beautifications

@@
 gives an overview of the design space of LU decomposition algorithms.
-Section~\ref{CPS}
+\S\ref{CPS}
 introduces code generation in MetaOCaml, the problem of name
@@
 solution.  We also present the key monad and the issues of generating
-control statements. Section~\ref{functors} describes the use of
+control statements. For the sake of reference, in \S\ref{GE-algorithm} 
+we present a particular Gaussian Elimination algorithm,
+a hand-written implementation of the standard textbook pseudo-code.
+\S\ref{functors} describes the use of
 parametrized modules of OCaml to encode all of the aspects of the
@@
 modules.  We discuss related work in
-Section~\ref{related} and outline future work. In our
-conclusion (Section~\ref{conclusion}) we comment
+\S\ref{related} and outline future work. In our
+conclusion (\S\ref{conclusion}) we comment
 on programming with aspects and sum up our guiding methodology.
@@
 in the upper triangle.
-\oleg{Section XXX demonstrates one particular GE algorithm in detail,
+\oleg{\S\ref{GE-algorithm} demonstrates one particular GE algorithm in detail,
   for further reference.}
@@
 as guidance in creating the modular LU generator discussed in 
-Section~\ref{functors}.
+\S\ref{functors}.
 
@@
 types -- which we will be doing extensively when writing module
-signatures in Section~\ref{functors}. The fact that OCaml renames all type
+signatures in \S\ref{functors}. The fact that OCaml renames all type
 variables when printing out types confuses matters further. An elegant
@@
 Various aspects of our generator will need to keep a state during
-code generation (see section~\ref{sec:det} for an example).
+code generation (see \S\ref{sec:det} for an example).
 
@@
 available.}
+\oleg{well, we can say IntegerDomain : DOMAIN with type v = int}
 The annotation may be omitted (see |ZpMake| below), in which
@@
 abstract over particular code representations. Therefore, when writing
-the LU generator (Section~\ref{sec:det} and below), instead of
+the LU generator (\S\ref{sec:det} and below), instead of
 MetaOCaml brackets and escapes we will be using the code combinators
@@
 container. For example, the following functor defines a matrix
-container represented as arrays of rows.
-
-\begin{code}
-module GenericArrayContainer(Dom:DOMAINL) =
+container as a single array, with elements stored in the
+row-major order -- the container used in Fig.\ref{fig:GE-algorithm}.
+\begin{code}
+module GenericVectorContainer(Dom:DOMAINL) =
   struct
   module Dom = Dom
-  type contr = Dom.v array array (* Array of rows *)
+  type contr = Dom.v container2dfromvector
   type 'a vc = ('a,contr) code
   type 'a vo = ('a,Dom.v) code
-  let getL x n m = .< (.~x).(.~n).(.~m) >.
-  let dim2 x = .< Array.length .~x >.       (* number of rows *)
-  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
+  let getL x i j = .< ((.~x).arr).(.~i* (.~x).m + .~j) >.
+  let dim2 x = .< (.~x).n >.      (* number of rows *)
+  let dim1 x = .< (.~x).m >.      (* number of cols *)
+  let mapper g a = match g with
   \dots
@@
 The accompanying code\cite{metamonadsURL} includes containers whose
-elements are stored in a 1D array, in a
-row-wise (C-like) or column-wise (Fortran-like) modes.
+elements are stored in a 1D array in a column-wise (Fortran-like)
+modes, and container for a matrix represented as arrays of rows.
 
@@
 |PackedLower| are quite similar (and simpler) than the implementation
-of the |AbstractDet| and |NoDet| aspects in Section~\ref{sec:det}. The
+of the |AbstractDet| and |NoDet| aspects in \S\ref{sec:det}. The
 choice of a particular tracking aspect may depend on all other choices
@@
 powerful enough to help us with tasks such as verifying that an integer
-is prime, as required in section~\ref{sec:domains}.
+is prime, as required in \S\ref{sec:domains}.
 
@@
 automates the lifting of (simple) modules as done in 
-section~\ref{sec:lifteddomains},
+\S\ref{sec:lifteddomains},
 first to the code level, and then to monadic values.  Even more
@@
 was joint work with Lydia van Dijk.
+Many helpful suggestions by the anonymous reviewers are gratefully
+acknowledged.
 

### 2008-06-05 tagless/tagless-final-aplas.tex
Ken starts to work on some reviewer comments

--- 2008-02-19 tagless/tagless-final.tex
@@
-%% TODO: add a ref 
-%%  Lars Birkedal's master thesis
-%%   http://www.itu.dk/people/birkedal/papers/paresm.ps.gz
-%% and thank Eijiro Sumii
-
-\begin{comment}
-\authorinfo{Jacques Carette}
-           {McMaster University}
-           {carette@mcmaster.ca}
-\authorinfo{Oleg Kiselyov}
-           {FNMOC}
-           {oleg@pobox.com}
-\authorinfo{Chung-chieh Shan}
-           {Rutgers University}
-           {ccshan@cs.rutgers.edu}
-\end{comment}
-
-\begin{abstract}
-\ifdim\parindent=0pt \parindent=1em \fi
-We have built the first family of tagless interpretations for a
-higher-order typed object
-language in a typed metalanguage (Haskell or ML) that
-require no dependent
-types, generalized algebraic data types, or
-postprocessing to eliminate tags.
-The statically type-preserving interpretations include
-an evaluator, a compiler (or staged evaluator), a partial
-evaluator, and call-by-name and
-call-by-value CPS transformers.
-
-Our main idea is to encode
-\ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
-using cogen functions rather than data constructors.
-In other words, we represent object terms not in an initial algebra
-but using the coalgebraic structure of the $\lambda$-calculus.
-Our representation also simulates inductive maps from types to
-types, which are required for typed partial evaluation and CPS transformations.
-
-Our encoding of an object term abstracts over the various ways to
-interpret it, yet statically assures that the interpreters never get
-stuck.  To achieve self\hyp interpretation and show Jones\hyp
-optimality, we relate this exemplar of higher-rank and higher-kind
-polymorphism \ifshort\else (provided by ML functors and Haskell~98 constructor
-classes) \fi to plugging a term into a context of let\hyp polymorphic
-bindings.
-\end{abstract}
-
-\begin{quote}
-\small
-    It should also be possible to define languages with a highly
-    refined syntactic type structure. Ideally, such a treatment should
-    be metacircular, in the sense that the type structure \linebreak[1] used in the
-    defined language should be adequate for the defining language.
-    \rm \ifshort \hfill John Reynolds~\else\\\fi\citep{reynolds-definitional}
-\end{quote}
-
-%\category{CR-number}{subcategory}{third-level}
-%\terms term1, term2
-%\keywords keyword1, keyword2
-
-\section{Introduction}\label{intro}
-
-A popular way to define and implement a language is to embed it in
-another \citep{reynolds-definitional}.  Embedding means to represent
-terms and values of the \emph{object language} as terms and values in the
-\emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
-specific object languages because it supports rapid prototyping and integration
-with the host environment \citep{hudak-building}.
-If the metalanguage supports \emph{staging}, then
-the embedding can compile object programs to the metalanguage and avoid the
-overhead of interpreting them on the fly \citep{WalidICFP02}.  A staged
-definitional interpreter is thus a promising way to build a domain\hyp specific
-language (DSL)\@.
-
-\begin{figure}
-    \begin{floatrule}
-    \begin{proofrules}
-        \[ \[ [x:t_1] \proofoverdots e:t_2 \] \justifies \fun{x}e:t_1\to t_2 \]
-        \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
-        \[ e_1:t_1\to t_2 \quad e_2:t_1 \justifies e_1 e_2: t_2 \]
-        \[ \text{$n$ is an integer} \justifies n:\ZZ \]
-        \[ \text{$b$ is a boolean} \justifies b:\BB \]
-        \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
-        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
-        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \times e_2:\ZZ \]
-        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]
-    \end{proofrules}
-    \end{floatrule}
-    \caption{Our typed object language}
-    \label{fig:object}
-\end{figure}
-
-We focus on embedding a \emph{typed} object language into a 
-\emph{typed} metalanguage.
-The benefit of types in this setting is to rule out meaningless object terms,
-thus enabling faster interpretation and assuring that our interpreters
-do not get stuck.
-To be concrete, we use the typed object language in
-Figure~\ref{fig:object} throughout this paper.  We aim not just for
-evaluation of object programs but also for
-compilation, partial evaluation, and other processing.
-
-\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
-a typed object language in a typed metalanguage as an interesting
-problem.  The common solutions to this problem store object terms and
-values in the metalanguage in a universal type, a generalized algebraic
-data type (GADT), or a dependent type.  In the remainder of this section,
-we discuss these solutions, identify their drawbacks, then summarize our
-proposal and contributions.  
-\ifshort
-We leave aside the solved problem of writing a parser\slash type\hyp checker,
-for embedding object language objects into the metalanguage
-(whether using dependent types \citep{WalidICFP02} or not \citep{baars-typing}),
-and just enter them by hand.
-\else
-No matter how we represent the object language in the
-metalanguage, the representation can be created either by hand (for example, by
-entering object terms at a metalanguage interpreter's prompt) or
-by a parser\slash type\hyp checker reading from a text string.
-We leave aside the solved problem of writing such a 
-parser\slash type\hyp checker,
-whether using dependent types \citep{WalidICFP02} or not \citep{baars-typing}.
-\oleg{Should we really leave it aside, given that we do have the code:
-  IncopeTypecheck, and given Sumii-san's advice to emphasize the fact
-  that we do have the code. Please see my comment in the related work
-  section in the paragraph that deals with typing dynamic typing.}
-\fi
-
-\subsection{The tag problem}\label{tagproblem}
-
-\begin{SaveVerbatim}{2a}
-type var = VZ | VS of var
-type exp = V of var | B of bool | L of exp | A of exp * exp
-\end{SaveVerbatim}
-\begin{SaveVerbatim}[commandchars=\@\{\}]{2b}
-let rec lookup (x::env) = function VZ -> x | VS v -> lookup env v
-let rec eval0 env = function
-| V v       -> lookup env v
-| B b       -> b 
-| L e       -> fun x -> eval0 (x::env) e
-| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
-\end{SaveVerbatim}
-\begin{SaveVerbatim}{2c}
-type u = UB of bool | UA of (u -> u)
-\end{SaveVerbatim}
-\begin{SaveVerbatim}{2d}
-let rec eval env = function
-| V v       -> lookup env v
-| B b       -> UB b
-| L e       -> UA (fun x -> eval (x::env) e)
-| A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
-\end{SaveVerbatim}
-\begin{SaveVerbatim}{test1}
-let test1 = A (L (V VZ), B true)
-\end{SaveVerbatim}
-
-% see tagless\_interp1.ml, module Tagfull for the complete code.
-% If you change the code in here, please adjust the .ml file
-% accordingly. Let the paper and the accompanying code be in sync.
-It is straightforward to create an algebraic data type, say in OCaml%
-\ifshort, Fig.~\ref{fig:tag-problem}(a)\fi, to
-represent object terms such as those in Figure~\ref{fig:object}.
-For brevity, we elide treating integers, conditionals, and fixpoint in
-this section.
-\ifshort\else\UseVerbatim{2a}\fi
-We represent each variable using a unary de Bruijn index.%
-\footnote{We use de Bruijn indices to simplify the comparison with
-\citearound{'s work}\citet{WalidICFP02}.}
-For example, we represent the object term $(\fun{x}x)\True$ as
-\ifshort \BUseVerbatim{test1}.\else \UseVerbatim{test1}\fi
-
-\ifshort
-\begin{figure}
-%
-(a) \BUseVerbatim{2a}
-
-\smallskip
-(b) \BUseVerbatim{2b}
-
-\smallskip
-(c) \BUseVerbatim{2c}
-
-\smallskip
-(d) \BUseVerbatim{2d}
-
-\medskip
-\caption{OCaml code illustrating the tag problem}
-\label{fig:tag-problem}
-\end{figure}
-\fi
-
-\noindent Following \citep{WalidICFP02},
-we try to implement an interpreter function |eval0|\ifshort,
-Fig.~\ref{fig:tag-problem}(b)\fi. It takes
-an object term such as |test1| above and gives us its value.
-The first argument to |eval0| is the environment, initially empty,
-which is the list of values bound to free variables in the
-interpreted code.
-\ifshort\else\UseVerbatim{2b}\fi
-If our OCaml-like metalanguage were untyped, the code above would be 
-acceptable.
-The |L e| line exhibits interpretive overhead:
-|eval0| traverses the function body~|e| every time (the result of
-evaluating) |L e| is applied. Staging can be used to remove this
-interpretive overhead \citep[\S1.1--2]{WalidICFP02}.
-
-However, the function |eval0| is ill-typed
-if we use OCaml or some other typed language as the metalanguage.
-The line |B b|
-says that |eval0| returns a boolean, whereas the next line |L e| says
-the result is a function, but all branches of a pattern-match form must
-yield values of the same type. 
-A related problem is the type of the environment |env|: a regular
-OCaml list cannot hold both boolean and function values.
-
-The usual solution is to introduce a universal type \citep[\S1.3]
-{WalidICFP02} containing both booleans and functions\ifshort,
-Fig.~\ref{fig:tag-problem}(c)\fi.
-\ifshort\else\UseVerbatim{2c}\fi
-We can then write a typed interpreter\ifshort,
-Fig.~\ref{fig:tag-problem}(d), \else\UseVerbatim{2d}\fi
-whose inferred type is |u list -> exp -> u|. Now we can evaluate
-\ifshort
-\texttt{eval [] test1} obtaining |UB true|.
-\else
-\begin{code}
-let test1r = eval [] test1
-val test1r : u = UB true 
-\end{code}
-\fi
-The unfortunate tag |UB| in the result reflects that |eval| is a partial
-function.  
-First, the pattern match |with UA f| in the line
-|A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
-as in the ill-typed term |A (B true, B false)|.
-\ifshort\else
-\begin{code}
-let test2 = A (B true, B false)
-let test2r = eval [] test2
-Exception: Match_failure in eval
-\end{code}
-\fi
-Second, the |lookup|
-function assumes a nonempty environment, so |eval| can fail if we
-evaluate an open term
-\ifshort
-|A (L (V (VS VZ)), B true)|.
-\else
-\begin{code}
-let test3 = A (L (V (VS VZ)), B true)
-let test3r = eval [] test3
-Exception: Match_failure in lookup
-\end{code}
-\fi
-After all, the type |exp| represents object
-terms both well-typed and ill-typed, both open and closed.
-
-If we evaluate only closed terms that have been type-checked, then
-|eval| would never fail. Alas, this soundness is not obvious to the
-metalanguage, whose type system we must still appease with the
-nonexhaustive pattern matching in |lookup| and |eval| and the tags |UB|
-and |UA| \citep[\S1.4]{WalidICFP02}.  In other words, the algebraic data
-types above fail to express in the metalanguage that the object program
-is well-typed.  This failure necessitates tagging and nonexhaustive
-pattern\hyp matching operations that incur a performance penalty in
-interpretation \citep{WalidICFP02} and impair optimality in partial evaluation
-\citep{taha-tag}.  In short, the universal\hyp type solution is
-unsatisfactory because it does not preserve typing.
-
-\ifshort\else\subsection{Solutions using fancier types}\fi
-
-It is commonly thought that to interpret a typed object language in
-a typed metalanguage while preserving types is difficult and requires
-GADTs or dependent types \citep{taha-tag}.  In fact, this problem
-motivated much work on GADTs \citep{xi-guarded,peyton-jones-simple} and
-on dependent types \citep{WalidICFP02,fogarty-concoqtion}.
-\ifshort\else
-For a metalanguage's type system to allow the well-typed object term
-|test1| but disallow the ill-typed object term |test2|, fancier types
-such as GADTs or dependent types seem necessary.  
-\fi
-Yet other type systems
-have been proposed to distinguish closed terms like |test1| from open
-terms 
-\ifshort\citep{WalidPOPL03,NanevskiJFP05,DaviesJACM01}\else
-like |test3|
-\citep{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual}\fi,
-so that |lookup| never receives an empty environment.  We discuss these
-proposals further in \S\ref{related}\ifshort\else;
-here we just note that many
-advanced type systems have been devised to ensure statically that an
-object term is well-typed and closed\fi.
-
-\subsection{Our final proposal}\label{ourapproach}
-
-We represent object programs using ordinary functions rather than
-data constructors.  These functions comprise the entire interpreter, shown
-below.
-\ifshort
-\begin{code3}
-let varZ env    = fst env         let b (bv:bool) env = bv
-let varS vp env = vp (snd env)    let lam e env   = fun x -> e (x,env)
-let app e1 e2 env   = (e1 env) (e2 env)
-\end{code3}
-\else
-\begin{code}
-let varZ env        = fst env
-let varS vp env     = vp (snd env)
-let b (bv:bool) env = bv
-let lam e env       = fun x -> e (x,env)
-let app e1 e2 env   = (e1 env) (e2 env)
-\end{code}
-\fi
-We now represent our sample term $(\fun{x}x)\True$ as
-\ifshort
-\texttt{let testf1 = app (lam varZ) (b true)}.
-\else
-\begin{code}
-let testf1 = app (lam varZ) (b true)
-\end{code}
-\fi
-This representation is almost the same as in \S\ref{tagproblem}, only
-written with lowercase identifiers. To evaluate an object term is to
-apply its representation to the empty environment\ifshort
-, |testf1 ()|, obtaining |true|\fi.
-\ifshort\else
-\begin{code}
-let testf1r = testf1 ()
-val testf1r : bool = true
-\end{code}
-\fi
-The result has no tags: the interpreter patently uses no tags and no
-pattern matching. The term |b true| evaluates to a boolean and the term
-|lam varZ| evaluates to a function, both untagged. The |app| function
-applies |lam varZ| without pattern matching. What is more, evaluating an
-open term such as
-\ifshort
-\texttt{app (lam (varS varZ)) (b true)}
-\else
-|testf3| below
-\fi
-gives a type error rather than a run-time error.
-\ifshort\else
-\begin{code}[commandchars=\\\{\}]
-let testf3 = app (lam (varS varZ)) (b true)
-let testf3r = testf3 \underline{()}
-This expression has type unit but is here used with type 'a * 'b
-\end{code}
-\fi
-The type error correctly complains
-that the initial environment should be a tuple rather than~|()|.
-In other words, the term is open.
-
-In sum, by Church\hyp encoding terms using ordinary functions, we
-achieve a tagless evaluator for a typed object language in a
-metalanguage with a simple 
-\ifshort
-Hindley-Milner \fi
-type system\ifshort\else\ \citep{hindley-principal,milner-theory}\fi
-.  In this \emph{final} rather
-than \emph{initial} approach, both kinds of run-time errors in
-\S\ref{tagproblem} (applying a nonfunction and evaluating an open
-term) are reported at compile time. Because the new interpreter
-uses no universal type or pattern matching, it never results in a
-run-time error, and is in fact total.  Because this safety is obvious
-not just to us but also to the metalanguage implementation, we avoid
-the serious performance penalty \citep{WalidICFP02} of error checking.
-\Citet{Gluck-jones-optimality} explains deeper technical reasons that
-inevitably lead to these performance penalties.
-
-Our solution is \emph{not} Church-encoding the
-universal type. The Church encoding of the type~|u| in \S\ref{tagproblem}
-requires two continuations; the function |app| in the interpreter above would
-have to provide both to the encoding of~|e1|. The continuation
-corresponding to the |UB| case of~|u| must either raise an error or
-loop. For a well-typed object term, that error continuation is never
-invoked, yet it must be supplied. In contrast, our interpreter has no error
-continuation at all.
-
-The evaluator above is wired directly into the
-functions |b|, |lam|, |app|, and so on.  \ifshort We \else In the rest of this paper, we \fi
-explain how to abstract the interpreter so as
-to process the \emph{same} term in many other
-ways: compilation, partial evaluation, CPS
-conversion, and so forth.
-
-\subsection{Contributions}\label{contributions}
-
-The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
-free in the encoding of an object term such as |testf1| above.  Defining
-these functions differently gives rise to different interpreters, that
-is, different folds on object programs.  Given the same term
-representation but varying the interpreter, we can
-\begin{itemize}
-    \item evaluate the term to a value in the metalanguage;
-    \item measure the size or depth of the term;
-    \item compile the term, with staging support such as in MetaOCaml;
-    \item partially evaluate the term, online; and
-    \item transform the term to continuation\hyp passing style (CPS),
-        even call-by-name (CBN) CPS, so as to isolate the evaluation
-        order of the object language from that of the metalanguage.\ifshort
-\footnote{Due to serious lack of space, 
-we refer the reader to the accompanying code for this.}\fi
-\end{itemize}
-We have programmed our interpreters in OCaml (and, for staging,
-\citet{metaocaml}) and standard Haskell. The complete code is
-available at \url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
-to supplement the paper. 
-\ifshort For simplicity, main examples in the paper will be in MetaOCaml;
-all examples have also been implemented in Haskell.
-\else Our examples below switch between (Meta)OCaml
-and Haskell even though we have implemented each example equivalently in
-both metalanguages, because some of our claims are more obvious in one
-metalanguage than the other.  For example, MetaOCaml provides
-convenient, typed staging facilities.\fi
-
-We attack the problem of tagless (staged) type-preserving
-interpretation exactly as it was posed
-by \citet{WalidICFP02} and \citet{xi-guarded}.
-We use their running examples and achieve the
-result they call desirable.  Our contributions are as follows.
-
-\begin{enumerate}
-\item We build interpreters that evaluate (\S\ref{language}),
-   compile (or evaluate with staging) (\S\ref{S:compiler}), and 
-   partially evaluate (\S\ref{PE}) a typed higher-order object language
-   in a typed metalanguage, in direct and continuation\hyp passing
-   styles\ifshort\else\ (\S\ref{variations})\fi.
-\item All these interpreters use no type tags, patently never get stuck,
-    and need no advanced type-system features such as GADTs, dependent types,
-    or intentional type analysis.
-\item The partial evaluator avoids polymorphic lift and delays binding-time
-    analysis.  It bakes a type-to-type map into the interpreter
-    interface to eliminate the need for GADTs and thus remain portable
-    across Haskell 98 and ML.
-\ifshort\else
-\item We show the first typed, tagless, and statically type-preserving CPS
-    transformation with simple types (\S\ref{variations}), which is available for mainstream
-    functional languages (ML and Haskell). Type preservation is assured by
-    the soundness of our embedding.
-\fi
-\item We use the type system of the metalanguage
-    to check statically that an object program is well-typed and closed.
-\item We show clean, comparable implementations in MetaOCaml and Haskell.
-\item We specify a functor signature that encompasses all our
-  interpreters, from evaluation and compilation (\S\ref{language}) 
-   to partial evaluation \ifshort\else and CPS transformation \fi(\S\ref{PE}).
-\item We point a clear way to extend the object language with more features
-    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
-\item We describe an approach to self\hyp interpretation compatible with the
-  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
-  out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
-\end{enumerate}
-Our code is surprisingly simple and obvious in hindsight, but
-it has been cited as a difficult problem (\cite{sumii-hybrid}
-notwithstanding) to interpret a typed object language in a typed metalanguage
-without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
-say that ``expressing such an interpreter in a statically typed
-programming language is a rather subtle matter. In fact, it is only
-recently that some work on programming type-indexed values in ML
-\citep{yang-encoding} has given a hint of how such a function can be
-expressed.''  We discuss related work in~\S\ref{related}.
-
-To reiterate, we do \emph{not} propose any new language
-feature or \ifshort new \else even any new programming \fi technique.
-\ifshort
-We
-\else
-Rather, we solve an open problem by a novel combination of simple types
-and existing techniques.  More precisely, we
-\fi
-use features already present in
-mainstream functional languages---Hindley-Milner
-type system with either an inference\hyp preserving module system or
-constructor classes, as realized in ML and Haskell 98---and techniques
-which have all appeared in the literature (in particular, \cite{yang-encoding,sumii-hybrid}), to
-solve a problem that was stated in the published record as
-unsolved and likely unsolvable in ML or Haskell 98
-without extensions. The simplicity of our solution and its
-use of only mainstream features \ifshort\else are virtues that \fi make it more practical to build typed,
-embedded DSLs.
-\begin{comment}
-We may claim some contribution about type-preserving CPS. Check
-related work section in \citep{Guillemette-Monier-PLPV}, especially
-check the work of Shao on type-preserving CPS in Flint. The PLPV paper
-in related work shows other tasks, including closure conversion, which
-we may tackle in our approach. We may be able to write a
-type-preserving, assured compiler, whose properties and assured by HM.
-\end{comment}
-
-\section{The object language and its tagless interpreters}\label{language}
-
-Figure~\ref{fig:object} shows our object language, a simply-typed
-$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
-The language is close to \citets{xi-guarded}, without their polymorphic
-lift but with more constants so as to more conveniently express Fibonacci,
-factorial, and power.
-\oleg{There is now IncopeDB.hs that defines at least R, L and C
- interpreters (no PE at present though) using DeBruijn indices. So we
- can say that (some of) our development is available in DeBruijn
- style. We won't talk about it in the paper though.}
-In contrast to \S\ref{intro}, we encode binding using higher-order
-abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
-rather than de Bruijn indices. This makes the encoding convenient and
-ensures that our object programs are closed.
-
-\subsection{How to make encoding flexible: abstract the interpreter}
-\label{encoding}
-
-\ifshort
-\begin{SaveVerbatim}{3a}
-class Symantics repr where
-  int :: Int  -> repr Int;       bool :: Bool -> repr Bool
-  lam :: (repr a -> repr b) -> repr (a -> b)
-  app :: repr (a -> b) -> repr a -> repr b
-  fix :: (repr a -> repr a) -> repr a
-
-  add :: repr Int -> repr Int -> repr Int
-  mul :: repr Int -> repr Int -> repr Int
-  leq :: repr Int -> repr Int -> repr Bool
-  if_ :: repr Bool -> repr a -> repr a -> repr a
-\end{SaveVerbatim}
-\else
-\begin{SaveVerbatim}{3a}
-class Symantics repr where
-  int  :: Int  -> repr Int
-  bool :: Bool -> repr Bool
-
-  lam :: (repr a -> repr b) -> repr (a -> b)
-  app :: repr (a -> b) -> repr a -> repr b
-  fix :: (repr a -> repr a) -> repr a
-
-  add :: repr Int -> repr Int -> repr Int
-  mul :: repr Int -> repr Int -> repr Int
-  leq :: repr Int -> repr Int -> repr Bool
-  if_ :: repr Bool -> repr a -> repr a -> repr a
-\end{SaveVerbatim}
-\fi
-\begin{SaveVerbatim}{3b}
-testpowfix () = lam (\x -> fix (\self -> lam (\n ->
-                 if_ (leq n (int 0)) (int 1)
-                     (mul x (app self (add n (int (-1))))))))
-\end{SaveVerbatim}
-\begin{SaveVerbatim}{3c}
-testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
-\end{SaveVerbatim}
-
-We embed our language in (Meta)OCaml and Haskell.  In Haskell,
-the functions that construct object terms are methods in a type class
-|Symantics| (with a parameter |repr| of kind |* -> *|)\ifshort,
-Fig.~\ref{fig:symantics-haskell}(a)\fi. The class is so named
-because its interface gives the syntax of the object language and its
-instances give the semantics.
-\ifshort\else\UseVerbatim{3a}\fi
-For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
-\S\ref{tagproblem} above as \texttt{app (lam (\textbackslash x -> x)) (bool True)},
-whose inferred type is \texttt{Symantics repr => repr Bool}.
-For another example, the classical $\mathit{power}$ function is
-\ifshort in Fig.~\ref{fig:symantics-haskell}(b)
-\else\UseVerbatim{3b}\fi
-and the partial application $\fun{x} \mathit{power}\;x\;7$ is
-\ifshort in Fig.~\ref{fig:symantics-haskell}(c).
-\else\UseVerbatim{3c}\fi
-The dummy argument |()| above is to avoid the monomorphism
-restriction, to keep the type of |testpowfix| and |testpowfix7|
-polymorphic in |repr|. \ifshort\else Instead of supplying this dummy
-argument, we could have given the terms explicit polymorphic
-signatures.  We however prefer for
-Haskell to infer the object types for us. We could also
-avoid the dummy argument by switching off the monomorphism restriction
-with a compiler flag. \fi
-The methods |add|, |mul|, and |leq| are quite similar, and so are
-|int| and |bool|. Therefore, we often show
-only one method of each group and elide the rest. The
-accompanying code has the complete implementations.
-
-\ifshort
-\begin{figure}
-(a) \BUseVerbatim{3a}
-
-\smallskip
-(b) \BUseVerbatim{3b}
-
-\smallskip
-(c) \BUseVerbatim{3c}
-
-\medskip
-\caption{Symantics in Haskell}
-\label{fig:symantics-haskell}
-\end{figure}
-\fi
-
-Comparing |Symantics| with Fig.~\ref{fig:object}
-shows how to represent \emph{every} typed, closed object term in the
-metalanguage. Moreover, the representation preserves types.
-\begin{proposition}
-If an object term has the object type~$t$, then its
-representation in the metalanguage has the type 
-|forall repr.| |Symantics repr => repr |$t$.
-\end{proposition}
-Conversely, the type system of the metalanguage statically checks that the
-represented object term is well-typed and closed.
-If we err, say replace |int 7| with |bool True| in
-|testpowfix7|, Haskell will complain there that the expected type |Int| does not
-match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
-encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
-Haskell's type checker also flags syntactically invalid object terms, 
-such as if we forget |app| somewhere above.
-
-\begin{SaveVerbatim}{ocaml-simple}
-module type Symantics = sig type ('c, 'dv) repr
-  val int : int  -> ('c, int) repr
-  val bool: bool -> ('c, bool) repr
-
-  val lam : (('c, 'da) repr -> ('c, 'db) repr) -> ('c, 'da -> 'db) repr
-  val app : ('c, 'da -> 'db) repr -> ('c, 'da) repr -> ('c, 'db) repr
-  val fix : ('x -> 'x) -> (('c, 'da -> 'db) repr as 'x)
-
-  val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
-  val mul : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
-  val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
-  val if_ : ('c, bool) repr
-            -> (unit -> 'x) -> (unit -> 'x) -> (('c, 'da) repr as 'x)
-end
-\end{SaveVerbatim}
-\begin{SaveVerbatim}{ocaml-examples}
-module EX(S: Symantics) = struct open S
-  let test1 () = app (lam (fun x -> x)) (bool true)
-  let testpowfix () =
-       lam (fun x -> fix (fun self -> lam (fun n ->
-        if_ (leq n (int 0)) (fun () -> int 1)
-            (fun () -> mul x (app self (add n (int (-1))))))))
-  let testpowfix7 = lam (fun x -> app (app (testpowfix ()) x) (int 7))
-end
-\end{SaveVerbatim}
-
-\ifshort
-\begin{figure}[t]
-\begin{tabular}{@{}l@{}}
-\ifx\relax\normalbaselineskip\else\baselineskip\normalbaselineskip\fi
-\BUseVerbatim[baseline=b]{ocaml-simple}\\[\smallskipamount]
-\ifx\relax\normalbaselineskip\else\baselineskip\normalbaselineskip\fi
-\BUseVerbatim[baseline=t]{ocaml-examples}
-\end{tabular}
-
-\medskip
-\caption{A simple (Meta)OCaml embedding of our object language, 
-    and examples}
-\label{fig:ocaml-simple}
-\label{fig:ocaml-examples}
-\end{figure}
-\else
-\begin{figure}[t]
-\begin{floatrule}
-\BUseVerbatim{ocaml-simple}
-\end{floatrule}
-\caption{A simple (Meta)OCaml embedding of our object language}
-\label{fig:ocaml-simple}
-\end{figure}
-\begin{figure}[t]
-\begin{floatrule}
-\BUseVerbatim{ocaml-examples}
-\end{floatrule}
-\caption{Examples using the embedding in Figure~\ref{fig:ocaml-simple} of our object language}
-\label{fig:ocaml-examples}
-\end{figure}
-\fi
-
-To embed the same object language in (Meta)OCaml, we replace the \ifshort type
-class \fi |Symantics| \ifshort\else type class \fi and its instances by a module signature |Symantics|
-and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
-signature that suffices until~\S\ref{PE}.  The two differences are:
-the additional type parameter |'c|, an
-\emph{environment classifier} \citep{WalidPOPL03} required by MetaOCaml for
-code generation in~\S\ref{S:compiler}; and the $\eta$-expanded type for
-|fix| and thunk types in |if_| since OCaml is a call-by-value
-language.
-
-The functor |EX| in Fig.~\ref{fig:ocaml-examples} encodes 
-our running examples |test1| and the $\mathit{power}$ function
-(|testpowfix|).
-The dummy argument to |test1| and |testpowfix| is an artifact of
-MetaOCaml, related to monomorphism: in order for us to run a
-piece of generated code, it must be polymorphic in its environment
-classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
-The value restriction dictates that
-the definitions of our object terms must look syntactically like
-values. (Alternatively, we could have used
-the rank-2 record types of OCaml to maintain the necessary polymorphism.)
-\ifshort\else\par\fi
-Thus, we represent an object expression in
-OCaml as a functor from |Symantics| to an appropriate semantic domain. This
-is essentially the same as the constraint \texttt{Symantics repr =>} in the
-Haskell embedding.
-
-\subsection{Two tagless interpreters}
-\label{S:interpreter-RL}
-
-Having abstracted our term representation over the interpreter, we are
-now ready to present a series of interpreters.  Each interpreter is an
-instance of the |Symantics| class in Haskell and a module implementing
-the |Symantics| signature in MetaOCaml.
-
-The first interpreter evaluates an object term to its value in the
-metalanguage.  The module below interprets each object\hyp language
-operation as the corresponding metalanguage operation.
-\ifshort
-\begin{code3}
-module R = struct type ('c,'dv) repr = 'dv (* no wrappers *)
-  let int  (x:int)  = x         let bool (b:bool) = b
-  let lam  f        = f         let app  e1 e2    = e1 e2
-  let fix  f        = let rec self n = f self n in self
-  let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
-  let leq  x y      = x <= y
-  let if_  eb et ee = if eb then et () else ee () end
-\end{code3}
-\else
-\begin{code}
-module R = struct
-  type ('c,'dv) repr = 'dv (* no wrappers *)
-
-  let int  (x:int)  = x
-  let bool (b:bool) = b
-  let lam  f        = f
-  let app  e1 e2    = e1 e2
-  let fix  f        = let rec self n = f self n in self
-  let add  e1 e2    = e1 + e2
-  let mul  e1 e2    = e1 * e2
-  let leq  x y      = x <= y
-  let if_  eb et ee = if eb then et () else ee ()
-end
-\end{code}
-\fi
-%
-As in~\S\ref{ourapproach},
-this interpreter is patently tagless, using neither a universal type nor
-any pattern matching: the operation |add| is really
-OCaml's addition, and |app| is OCaml's application. To run our
-examples, we instantiate the |EX| functor from~\S\ref{encoding}
-with~|R|\ifshort: \texttt{module EXR = EX(R)}\fi.
-\ifshort\else
-\begin{code}
-module EXR = EX(R)
-\end{code}
-\fi
-Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
-\ifshort It \else
-In Haskell, we define
-\begin{code}
-newtype R a = R {unR::a}
-instance Symantics R where ...
-\end{code}
-Although |R| looks like a tag, it is only
-a |newtype|.  The types |a| and |R a| are represented differently
-only at compile time, not at run time.  Pattern matching against~|R|
-cannot ever fail and is assuredly compiled away.
-In OCaml, too, it
-\fi is obvious to the compiler that
-pattern matching cannot fail, because there is no
-pattern matching. Evaluation can only fail to yield a value
-due to interpreting |fix|.
-\ifshort
-(The source code shows a total interpreter |L| that measures
-the size of each object term.)
-\fi
-\ifshort
-We can also generalize from~|R| to all interpreters;
-these propositions follow immediately from the soundness of the
-metalanguage's type system.
-\fi
-\begin{proposition}
-If an object term~$e$ encoded in the metalanguage has type~$t$,
-then evaluating~$e$ in the interpreter~|R| either continues
-indefinitely or terminates with a value of the same type~$t$.
-\end{proposition}
-\ifshort\else
-Generalizing from~|R| to all interpreters, we have the following
-broader and more useful, if also more obvious, proposition.
-\fi
-\begin{proposition}
-  If an implementation of |Symantics| never gets stuck, then
-  the type system of the object
-  language is sound with respect to the dynamic semantics defined by
-  that implementation.
-\end{proposition}
-\ifshort\else
-These propositions follow immediately from the soundness of the
-metalanguage's type system.
-\fi
-\ifshort\else
-For variety, we show another interpreter, which measures the \emph{size}
-of each object term, defined as the number of term
-constructors.
-\begin{code}
-module L = struct
-  type ('c,'dv) repr = int
-
-  let int  (x:int)  = 1
-  let bool (b:bool) = 1
-  let lam  f        = f 0 + 1
-  let app  e1 e2    = e1 + e2 + 1
-  let fix  f        = f 0 + 1
-  let add  e1 e2    = e1 + e2 + 1
-  let mul  e1 e2    = e1 + e2 + 1
-  let leq  x y      = x + y + 1
-  let if_  eb et ee = eb + et () + ee () + 1
-end
-\end{code}
-Now the OCaml expression
-|let module E = EX(L) in E.test1 ()|
-evaluates to |3|. This interpreter is not only tagless but also
-total. It ``evaluates'' even seemingly divergent terms, for instance
-|app (fix (fun self -> self)) (int 1)| evaluates to $3$.
-\fi
-
-\begin{comment}
-\begin{code}
-module EX1(S: Symantics) = struct
- open S
- let tfix () = app (fix (fun self -> self)) (int 1)
-end;;
-let module E =EX1(R) in E.tfix ();;
-let module E =EX1(L) in E.tfix ();;
-\end{code}
-\end{comment}
-
-\section{A tagless compiler (or, a staged interpreter)}\label{S:compiler}
-
-Besides immediate evaluation, we can compile our object language
-into OCaml code using MetaOCaml's staging facilities. MetaOCaml
-represents future-stage expressions of type~$t$
-as values of type |('c, |$t$|) code|, where |'c| is the
-environment classifier \citep{WalidPOPL03,calcagno-ml-like}. Code values are created
-by a \emph{bracket} form |.<|$e$|>.|, which quotes the expression~$e$ for
-evaluation at a future stage. The \emph{escape} |.~|$e$ must occur
-within a bracket and specifies that the expression~$e$ must be evaluated
-at the current stage; its result, which must be a code value, is
-spliced into the code being built by the enclosing bracket. The \emph{run} form |.!|$e$ evaluates
-the future-stage code value~$e$ by compiling and linking it at run time.
-Bracket, escape, and run are akin to
-quasi-quotation, unquotation, and |eval| of Lisp.
-
-% this code does not have the 'sv parameter. It shows up later.
-\ifshort
-\begin{SaveVerbatim}{5a}
-module C = struct type ('c,'dv) repr = ('c,'dv) code
-  let int (x:int)  = .<x>.            let bool (b:bool) = .<b>.
-  let lam f        = .<fun x -> .~(f .<x>.)>.
-  let app e1 e2    = .<.~e1 .~e2>.
-  let fix f =  .<let rec self n = .~(f .<self>.) n in self>.
-  let add e1 e2    = .<.~e1 + .~e2>.  let mul e1 e2 = .<.~e1 * .~e2>.
-  let leq x y      = .<.~x <= .~y>.
-  let if_ eb et ee = .<if .~eb then .~(et ()) else .~(ee ())>. end
-\end{SaveVerbatim}
-\else
-\begin{SaveVerbatim}{5a}
-module C = struct
-  type ('c,'dv) repr = ('c,'dv) code
-
-  let int (x:int)   = .<x>.
-  let bool (b:bool) = .<b>.
-  let lam f         = .<fun x -> .~(f .<x>.)>.
-  let app e1 e2     = .<.~e1 .~e2>.
-  let fix f         = .<let rec self n = .~(f .<self>.) n in self>.
-  let add e1 e2     = .<.~e1 + .~e2>.
-  let mul e1 e2     = .<.~e1 * .~e2>.
-  let leq x y       = .<.~x <= .~y>.
-  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>.
-end
-\end{SaveVerbatim}
-\fi
-\begin{SaveVerbatim}{5b}
-let module E = EX(C) in E.test1 ()
-\end{SaveVerbatim}
-\begin{SaveVerbatim}{5c}
-let module E = EX(C) in E.testpowfix7
-\end{SaveVerbatim}
-\begin{SaveVerbatim}{5d}
-.<fun x_1 -> (fun x_2 -> let rec self_3 = fun n_4 ->
-   (fun x_5 -> if x_5 <= 0 then 1 else x_2 * self_3 (x_5 + (-1)))
-   n_4 in self_3) x_1 7>.
-\end{SaveVerbatim}
-
-Inserting brackets and escapes appropriately into the
-evaluator~|R| above yields the simple compiler~|C|
-\ifshort in Fig.~\ref{fig:interpreter-C}(a).\else below.
-\UseVerbatim{5a}
-\fi
-\ifshort
-\begin{figure}[t]
-(a) \BUseVerbatim{5a}
-
-\smallskip
-(b) \BUseVerbatim{5b}
-
-\smallskip
-(c) \BUseVerbatim{5c}
-
-\smallskip
-(d) \BUseVerbatim{5d}
-
-\medskip
-\caption{The tagless staged interpreter \texttt{C}}
-\label{fig:interpreter-C}
-\end{figure}
-\fi
-This is a straightforward staging of
-|module R|.
-This compiler produces
-unoptimized code. For example, interpreting our |test1| with
-\ifshort Fig.~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
-gives the code value |.<(fun|~|x_6|~|->|~|x_6)| |true>.|
-of inferred type |('c,|~|bool)| |C.repr|.  Interpreting |testpowfix7|
-with
-\ifshort Fig.~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
-gives a code value with many apparent $\beta$- and $\eta$-redexes\ifshort,
-Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
-This compiler does not incur
-any interpretive overhead: the
-code produced for $\fun{x}x$ is simply |fun|~|x_6|~|->|~|x_6|\ifshort\else\
-and does not
-call the interpreter, unlike the recursive calls to |eval0| and
-|eval| in the |L e| lines in \S\ref{tagproblem}\fi.
-The resulting code obviously contains no tags and no pattern matching.
-The environment classifiers here, like the tuple types in \S\ref{ourapproach},
-make it a type error to run an open expression.
-\ifshort
-The accompanying code shows the Haskell implementation. 
-\else
-\begin{proposition}
-    The compiler |C| generates tagless code free of pattern matching.
-\end{proposition}
-This is syntactically clear from the body of |C|, and follows
-readily by structural induction and the properties of |.~| and
-|.< >.|.
-
-We have also implemented this compiler in Haskell. 
-Since Haskell
-has no (convenient, typed) staging facility, we had to emulate
-it: we defined a data type |ByteCode| with
-constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
-(Alternatively, we could use Template Haskell as our staging facility:
-\texttt{ByteCode} can be mapped to the abstract syntax of Template
-Haskell. The output of our compiler will then be assuredly type-correct
-Template Haskell.)
-Whereas our representation of object terms uses HOAS,
-our bytecode uses integer-named
-variables to be realistic. 
-We then define 
-\begin{code}
-newtype C t = C (Int -> (ByteCode t, Int)) 
-\end{code}
-where |Int| is the counter for creating fresh variable
-names. We define the compiler by making |C| an instance of the
-class |Symantics|.
-The implementation is quite similar (but slightly more
-verbose) than the corresponding MetaOCaml code above. (The implementation uses
-GADTs because we also wanted to write a typed interpreter for 
-the \texttt{ByteCode} \emph{data type}.) The
-accompanying code gives the full details.
-\fi
-
-%------------------------------------------------------------------------
-% This is the version for the APLAS paper. No Haskell
-\ifshort
 \section{A tagless partial evaluator}\label{PE}
@@
 phases (with |Some| and |None|) but not object types.
-
-%------------------------------------------------------------------------
-% This is the longer version of the section
-\else
-
-\section{A tagless partial evaluator}\label{PE}
-Surprisingly, we can write a partial evaluator using the idea above,
-namely to build object terms using ordinary functions rather than data
-constructors.  We present this partial evaluator in a sequence of three
-attempts. It uses no universal type and no tags
-for object types.  We then discuss residualization and binding-time
-analysis.  Our partial evaluator is a modular extension of the evaluator
-in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{S:compiler}, in
-that it uses the former to reduce static terms and the latter to build
-dynamic terms.
-
-\subsection{Avoiding polymorphic lift}
-\label{S:PE-lift}
-
-Roughly, a partial evaluator interprets each object term to yield either
-a static (present-stage) term (using the evaluator~|R|) or
-a dynamic (future-stage) term (using the compiler~|C|).  To
-distinguish between static and dynamic terms, we might try to define
-|repr| in the partial evaluator (module |P0|) as
-\begin{code}
-type ('c,'dv) repr = S0 of ('c,'dv) R.repr | D0 of ('c,'dv) C.repr
-\end{code}
-(the digit zero in the names of the phase tags |S0| and |D0| is the
-version suffix).
-To extract a dynamic term from this type, we create the function
-\begin{code}
-let abstrI0 (e:('c,int) repr) : ('c,int) C.repr =
-  match e with
-  | S0 e -> C.int e
-  | D0 e -> e
-\end{code}
-and the similar function |abstrB0| for dynamic boolean terms. Here, 
-|C.int| is used to convert from the static term |('c,int) R.repr|, 
-which is just |int|, to the dynamic term. We can now define the following
-components of module |P0| required by the |Semantics| signature:
-\begin{code}
-let int  (x:int)  = S0 (R.int x)
-let bool (x:bool) = S0 (R.bool x)
-let add e1 e2 = 
-  match (e1,e2) with
-  | (S0 e1, S0 e2) -> S0 (R.add e1 e2)
-  | (e1, e2)       -> D0 (C.add (abstrI0 e1) (abstrI0 e2))
-\end{code}
-Integer and boolean literals are immediate, present-stage
-values. Addition yields a static term (using~|R.add|) if and only 
-if both operands are static; otherwise we extract the dynamic terms 
-from the operands and add them using~|C.add|.
-
-Whereas |mul| and |leq| are as easy to define as |add|, we encounter
-a problem with |if_|.  Suppose that the first argument to |if_| 
-is a dynamic term
-(of type |('c,bool) C.repr|), the second a static term 
-(of type |('c,'a) R.repr|), and the third a
-dynamic term (of type |('c,'a) C.repr|). We then need to convert
-the static term to dynamic, but there is no polymorphic ``lift''
-function, of type |'a -> ('c,'a) C.repr|, to send a value to the future stage
-\citep{xi-guarded,WalidPOPL03}.
-\begin{comment}
-(By the way, if we
-were to add polymorphic \texttt{lift} to the type class
-\texttt{Symantics repr}, then \texttt{repr} would become an instance of
-\texttt{Applicative} and thus \texttt{Functor}:\texttt{ fmap
-f = app (lift f)~}.)
-\end{comment}
-
-Our |Symantics| only includes separate lifting methods |bool| and
-|int|, not a parametrically polymorphic lifting method, for good reason:
-When compiling to a first-order target language such as machine code,
-booleans, integers, and functions may well be represented differently.
-Thus, compiling polymorphic lift requires intensional type
-analysis.  To avoid needing polymorphic lift, we turn to
-Asai's technique \citep{asai-binding-time,sumii-hybrid}:
-build a dynamic term alongside every static term.
-
-\subsection{Delaying binding-time analysis}
-\label{S:PE-problem}
-
-We start building the partial evaluator anew (module |P1|) and
-switch to the data type
-\begin{code}
-type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr
-\end{code}
-so that a partially evaluated term always contains a dynamic
-component and sometimes contains a static component.  By 
-distributivity, the two
-alternative constructors of an |option| value, |Some| and |None|,
-tag each partially evaluated term with a phase: either present or
-future.  This tag is not an object type tag: all pattern matching below
-is exhaustive. Because the future-stage component is always available, we
-can now define the polymorphic function
-|let abstr1 (P1 (_,dyn)) = dyn| of type
-\texttt{('c,'dv) repr -> ('c,'dv) C.repr}
-to extract it without requiring polymorphic lift into~|C|.  We then try
-to define the interpreter |P1|---and get as far as the first-order
-constructs of our object language, including |if_|.
-\begin{code}
-module P1 : Symantics = struct
-  let int (x:int) = P1 (Some (R.int x), C.int x)
-  let add e1 e2 = match (e1,e2) with
-    | (P1 (Some n1,_),P1 (Some n2,_)) -> int (R.add n1 n2)
-    | _ -> P1 (None,(C.add (abstr1 e1) (abstr1 e2)))
-  let if_ = function
-    | P1 (Some s,_) -> fun et ee -> if s then et () else ee ()
-    | eb -> fun et ee -> P1 (None, C.if_ (abstr1 eb) 
-                                   (fun () -> abstr1 (et ()))
-                                   (fun () -> abstr1 (ee ())))
-\end{code}
-However, we stumble on functions.  According to our
-definition of~|P1|, a partially evaluated object function, such as the
-identity $\fun{x}x$ embedded in OCaml as |lam (fun x -> x)|\texttt{ :
-}|('c,'a->'a) P1.repr|, consists of a dynamic part 
-(type |('c,'a->'a) C.repr|) and
-maybe a static part (type |('c,'a->'a) R.repr|).  The dynamic part is useful
-when this function is passed to another function that is only
-dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
-useful when this function is applied to a static argument, as in
-$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
-evaluate the function, that is, compute as much as possible statically
-when it is applied to a mix of static and dynamic inputs.  For example,
-the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
-by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
-not statically known.  The same static function, applied to
-different static arguments, can give both static and dynamic results: we
-want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
-to~$x$.
-
-To enable these simplifications, we delay binding-time analysis
-for a static function until it is applied, that is, until |lam f|
-appears as the argument of |app|.  To do so, we have to incorporate |f|
-as it is into the |P1.repr| data structure: the representation
-for a function type |'a->'b| should be one of
-\begin{code}
-S1 of ('c,'a) repr -> ('c,'b) repr | E1 of ('c,'a->'b) C.repr
-P1 of (('c,'a) repr -> ('c,'b) repr) option * ('c,'a->'b) C.repr
-\end{code}
-unlike |P1.repr| of |int| or |bool|.
-That is, we need a nonparametric data type, something akin to
-type-indexed functions and type-indexed types, which
-\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
-Thus, typed partial evaluation, like typed CPS transformation
-(see \S\ref{S:CPS}),
-inductively defines a map from source types to target types that
-performs case distinction on the source type. In Haskell, typecase
-can be equivalently implemented either with GADTs or with
-type-class functional dependencies
-\citep{oliveira-typecase}. The accompanying code shows both
-approaches (|Incope.hs| and |incope1.hs|), 
-neither portable to OCaml. In addition,
-the problem of nonexhaustive pattern\hyp matching reappears in
-the GADT approach because GHC 6.6.1 cannot see that a particular
-type of a GADT value precludes certain constructors.
-Thus GADTs fail to
-make it \emph{syntactically} apparent that pattern matching is exhaustive.
-
-%% \subsection{Eliminating tags from typecase}
-%% \label{S:PE-GADT}
-
-%% Two common ways to provide typecase in Haskell are
-%% GADTs and type-class functional dependencies
-%% \citep{oliveira-typecase}.  These
-%% methods are equivalent, and here we use GADTs; |incope1.hs|
-%% in the accompanying source code shows the latter.
-%% We introduce a GADT with four data constructors.
-%% \begin{code}
-%% data P t where
-%%   VI :: Int  -> P Int
-%%   VB :: Bool -> P Bool
-%%   VF :: (P a -> P b) -> P (a -> b)
-%%   E  :: C t -> P t
-%% \end{code}
-%% The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
-%% in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |D0|).  However,
-%% the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
-%% operand of type |P a -> P b| rather than |a -> b|.  We define a function
-%% like |abstr1| above to extract a future-stage computation from a 
-%% value of type |P t|.
-%% \begin{code}
-%% abstr :: P t -> C t
-%% abstr (VI i) = int i
-%% abstr (VB b) = bool b
-%% abstr (VF f) = lam (abstr . f . E)
-%% abstr (E x)  = x
-%% \end{code}
-%% The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
-%% case uses the method |lam| of the |C| interpreter to compile~|f|.
-
-%% We may now make |P| an instance of
-%% |Symantics| and implement the partial evaluator as follows. We elide
-%% |mul|, |leq|, |if_|, and |fix|.
-%% \begin{code}
-%% instance Symantics P where
-%%   int x               = VI x
-%%   bool b              = VB b
-%%   add (VI n1) (VI n2) = VI (n1 + n2)
-%%   add e1 e2           = E (add (abstr e1) (abstr e2))
-%%   lam                 = VF
-%%   app (VF f) ea       = f ea
-%%   app (E f)  ea       = E (app f (abstr ea))
-%% \end{code}
-%% The implementations of |int|, |bool|, and |add| are like 
-%% in~\S\ref{S:PE-problem}. The interpretation of |lam f| is |VF f|, 
-%% which just wraps the HOAS function |f|. 
-%% We can always compile |f| to a code value,
-%% but we delay it to apply |f| to concrete arguments. The interpretation of
-%% |app ef ea| checks to see if |ef| is such a delayed
-%% HOAS function |VF f|. If it is, we apply |f| to the
-%% concrete argument |ea|, giving us a chance to perform static
-%% computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
-%% dynamic value |E f|, we residualize.
-
-%% This solution using GADTs works but is not quite satisfactory. First, it
-%% cannot be ported to MetaOCaml, as GADTs are unavailable there.  Second,
-%% the problem of nonexhaustive pattern\hyp matching reappears in
-%% |app| above: the type |P t| has four constructors, of which the pattern in
-%% |app| matches only |VF| and~|E|. One may say that the
-%% constructors |VI| and |VB| obviously cannot occur because they do not
-%% construct values of type |P (a -> b)| as required by the type of |app|.
-%% Indeed, the metalanguage implementation could reason thus:
-%% if we use inductive families (as in Coq) or logical
-%%   frameworks with canonical forms (as in Twelf with its coverage checker),
-%%   we can prove the pattern matching to be exhaustive.
-%% Then again, the metalanguage implementation may not reason thus:
-%% GHC cannot and issues warnings.
-%% Although this point may seem minor, it is the heart of
-%% the tagging problem and the purpose of tag elimination. A typed tagged
-%% interpreter contains many pattern\hyp matching forms that look partial
-%% but never fail in reality. The
-%% goal is to make this exhaustiveness \emph{syntactically} apparent.
-
-\subsection{The ``final'' solution}
-\label{S:PE-solution}
-
-Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
-would ideally like is to write
-\begin{code}
-type ('c,'dv) repr = P1 of (repr_pe ('c,'dv)) R.repr option * ('c,'dv) C.repr
-\end{code}
-where |repr_pe| is the type function defined
-% inductively because P below depends on repr_pe
-by
-\begin{code}
-repr_pe ('c,int)    = ('c,int)
-repr_pe ('c,bool)   = ('c,bool)
-repr_pe ('c,'a->'b) = ('c,'a) repr -> ('c,'b) repr
-\end{code}
-Although we can use type classes to define this type function
-in Haskell, that is not portable to MetaOCaml. However,
-these three typecase alternatives are already present in existing
-methods of |Symantics|.
-A simple and portable solution thus emerges: we bake |repr_pe| 
-into the signature |Symantics|. 
-We recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that 
-the |repr| type constructor took two arguments |'c| and~|'dv|.
-We add an argument |'sv| for the result of applying |repr_pe| to~|'dv|.
-Figure~\ref{fig:ocaml} shows the new signature.
-\begin{figure}
-\begin{floatrule}
-\begin{code2}
-module type Symantics = sig
-  type ('c,'sv,'dv) repr
-
-  val int : int  -> ('c,int,int) repr
-  val bool: bool -> ('c,bool,bool) repr
-
-  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-            -> ('c,'x,'da -> 'db) repr
-  val app : ('c,'x,'da -> 'db) repr
-            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
-                                'da -> 'db) repr as 'x)
-
-  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
-  val if_ : ('c,bool,bool) repr
-            -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
-end
-\end{code2}
-\end{floatrule}
-\caption{A (Meta)OCaml embedding of our object language that supports
-  partial evaluation}
-\label{fig:ocaml}
-\end{figure}
-
-\begin{figure}
-\begin{floatrule}
-\begin{code2}[commandchars=\@\[\]]
-module P = struct
-  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
-  let abstr {dy = x} = x
-  let pdyn x = {st = None; dy = x}
-
-  let int  (x:int ) = {st = Some (R.int  x); dy = C.int  x}
-  let bool (x:bool) = {st = Some (R.bool x); dy = C.bool x}
-
-  let add e1 e2 = match e1, e2 with
-                  | {st = Some 0}, e | e, {st = Some 0} -> e
-                  | {st = Some m}, {st = Some n} -> int (R.add m n)
-                  | _ -> pdyn (C.add (abstr e1) (abstr e2))
-  let if_ eb et ee = match eb with
-                     | {st = Some b} -> if b then et () else ee ()
-                     | _ -> pdyn (C.if_ (abstr eb) (fun () @!->@! abstr (et @!()))
-                                                   (fun () @!->@! abstr (ee @!())))
-
-  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
-  let app ef ea = match ef with
-                  | {st = Some f} -> f ea
-                  | _ -> pdyn (C.app (abstr ef) (abstr ea))
-  let fix f = let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
-              in let rec self = function
-                                | {st = Some @!_} as e -> app (f (lam self)) e
-                                | e -> pdyn (C.app fdyn (abstr e))
-                 in {st = Some self; dy = fdyn}
-end
-\end{code2}
-\end{floatrule}
-\caption{Our partial evaluator (\texttt{mul} and \texttt{leq}
-  are elided)}
-\label{fig:pe}
-\end{figure}
-
-The interpreters |R|, |L| and~|C| in \S\ref{S:interpreter-RL},\S\ref{S:compiler}
-only use the old
-type arguments |'c| and~|'dv|, which are treated by the new signature
-in the same way.  Hence, all that needs to change in these interpreters
-to match the new signature is to add a phantom type
-argument~|'sv| to~|repr|.
-For example, the compiler |C| now begins
-\begin{code}
-module C = struct
-  type ('c,'sv,'dv) repr = ('c,'dv) code
-\end{code}
-with the rest the same.
-In contrast, the partial evaluator~|P| relies on the type argument |'sv|.
-
-Figure~\ref{fig:pe} shows the partial evaluator~|P|.
-Its type |repr| literally expresses the type equation for |repr_pe| above.
-The function |abstr|
-extracts a future-stage code value from the result of
-partial evaluation.  Conversely, the function |pdyn| injects a
-code value into the |repr| type. As
-in~\S\ref{S:PE-problem}, we build dynamic terms alongside
-any static ones to avoid polymorphic lift.
-
-The static portion of the interpretation of |lam f| is |Some f|, 
-which just wraps the HOAS
-function |f|. The interpretation of |app ef ea| 
-checks to see if |ef| is such a wrapped
-HOAS function. If it is, we apply |f| to the
-concrete argument |ea|, giving us a chance to perform static
-computations (see the example below). If
-|ef| has only a dynamic part, we residualize.
-
-To illustrate how to add optimizations, we improve |add| (and |mul|,
-elided) to simplify the generated code using the monoid (and ring)
-structure of~|int|: not only is addition performed statically
-(using~|R|) when both operands are statically known, but it is
-eliminated when one operand is statically~$0$; similarly for
-multiplication by~$0$ or~$1$.  
-Such algebraic simplifications are easy
-to abstract over the specific domain (such as monoid or ring) where they
-apply.  These simplifications and abstractions help a lot
-in a large language with more base types and primitive operations.
-Incidentally, the code actually contains a more general implementation
-mechanism for such features, inspired in part by previous work in
-generative linear algebra~\citep{CaretteKiselyov05}.
-
-Any partial evaluator must decide how much to unfold recursion
-statically: unfolding too little can degrade the residual code, whereas
-unfolding too much risks nontermination.  Our partial evaluator is no
-exception, because our object language includes |fix|.  The code in
-Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
-way'', that is, whenever the 
-argument is static, we unfold |fix| rather than residualize it.
-A conservative alternative is to unfold recursion only once, then residualize:
-\begin{code}
-let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
-\end{code}
-Many sophisticated approaches have been developed to decide how much to unfold
-\citep{Jones-Mix,jones-partial}, but this issue is 
-orthogonal to our presentation.
-A separate concern in our treatment of |fix| is possible code bloat in
-the residual program, which calls for let-insertion
-\citep{BondorfDanvy,SwadiTahaKiselyovPasalic2006}.
-
-Given this implementation of~|P|, our running example
-\begin{code}
-let module E = EX(P) in E.test1 ()
-\end{code}
-evaluates to
-\begin{code}
-{P.st = Some true; P.dy = .<true>.}
-\end{code}
-of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{S:compiler},
-a $\beta$-reduction has been statically performed to yield |true|.  More
-interestingly, whereas |testpowfix7| compiles to a code value with many
-$\beta$-redexes in~\S\ref{S:compiler}, the partial evaluation
-\begin{code}
-let module E = EX(P) in E.testpowfix7
-\end{code}
-gives the desired result
-\begin{code}
-{P.st = Some <fun>;
- P.dy = .<fun x -> x * (x * (x * (x * (x * (x * x)))))>.}
-\end{code}
-
-\noindent All pattern\hyp
-matching in~|P| is \emph{syntactically} exhaustive, so it is patent to the
-metalanguage implementation that |P| never gets stuck.  Further, all
-pattern\hyp matching occurs
-during partial evaluation, only to check if a value is known statically,
-never what type it has.  In other words, our partial evaluator tags
-phases (with |Some| and |None|) but not object types.
-
-Our typed partial evaluator is online and polyvariant.  It reuses the
-compiler~|C| and the evaluator~|R| by composing them.  This situation is
-simpler than \citets{SperberThiemann:TwoForOne} composition of a partial
-evaluator and a compiler, but the general ideas are similar.
-\fi
-
-% don't even have space for CPS
-\ifshort\else
-\section{Continuation\hyp passing style}\label{variations}
-
-Our approach accommodates
-several variants, including
-a call-by-name CPS interpreter and a call-by-value CPS
-transformation.
-\ifshort\label{state}\label{S:CPS}%
-This lets us decouple the evaluation strategy of the object language
-from that of the metalanguage. The accompanying code shows the CBN CPS
-interpreter (module |RCN| implementing |Symantics|) and a CBV CPS
-transformer |CPST|. The latter explicitly maps CPS interpretations to
-(direct) interpretations performed by the base interpreter~|S|. All
-these interpreters are typed, tagless and \emph{type-preserving} (as
-well as fully polymorphic in the answer-type). The type preservation
-is the consequence of the type soundness of the metalanguage.  We can
-modify the CBV CPS transformation to pass a piece of state along with
-the continuation. This technique lets us support mutable state.  Due
-to the severe lack of space we cannot describe these interpreters and
-refer the reader to the accompanying code.
-
-\else
-
-\subsection{Call-by-name CPS interpreters}\label{S:CPS}
-
-The object language generally inherits the evaluation strategy from
-the metalanguage---call-by-value (CBV) in OCaml, call-by-name (CBN) in
-Haskell.  To represent a CBN object language in a CBV metalanguage,
-\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}
-introduce CPS to make the evaluation strategy of a definitional
-interpreter indifferent to that of the metalanguage. To achieve the same
-indifference in the typed setting, we build a CBN CPS interpreter for
-our object language in OCaml.
-
-The interpretation of an object term is a function
-mapping a continuation~|k| to the answer
-returned by~|k|.
-\begin{code}
-let int (x:int) = fun k -> k x
-let add e1 e2 = fun k -> e1 (fun v1 -> e2 (fun v2 -> k (v1 + v2)))
-\end{code}
-In both |int| and |add|, the interpretation has type 
-|(int -> 'w)|\texttt{ }|-> 'w|, where |'w| is the (polymorphic) answer type.
-
-Unlike CBV CPS, the CBN CPS interprets
-abstraction and application as follows:
-\begin{code}
-let lam f = fun k -> k f
-let app e1 e2 = fun k -> e1 (fun f -> f e2 k)
-\end{code}
-Characteristic of CBN, |app e1 e2|
-does not evaluate the argument~|e2| by applying it to the
-continuation~|k|. Rather, it passes |e2| unevaluated to the abstraction.
-Interpreting $\fun{x} x+1$ yields type
-\begin{code}
-((((int -> 'w1) -> 'w1) -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2
-\end{code}
-
-We would like to collect those interpretation functions into a module
-with signature |Symantics|, to include the CBN CPS interpreter within our
-general framework. Alas, as in~\S\ref{S:PE-problem}, the type of
-an object term inductively determines the type of its interpretation:
-the interpretation of an object term of type~$t$ may not have type
-|(|$t$|->'w)->'w|, because $t$ may be a function type.  Again we
-simulate a type function with a typecase distinction, using an extra
-type argument to |repr|. Luckily, the type function |repr_pe| needed for
-the partial evaluator 
-in~\S\ref{S:PE-solution} is precisely the same type function we
-need for CBN CPS\@.
-\begin{code}
-module RCN = struct
-  type ('c,'sv,'dv) repr = {ko: 'w. ('sv -> 'w) -> 'w}
-  let int (x:int) = {ko = fun k -> k x}
-  let add e1 e2 = {ko = fun k ->
-      e1.ko (fun v1 -> e2.ko (fun v2 -> k (v1 + v2)))}
-  let if_ eb et ee = {ko = fun k ->
-      eb.ko (fun vb -> if vb then (et ()).ko k else (ee ()).ko k)}
-  let lam f = {ko = fun k -> k f}
-  let app e1 e2 = {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
-  let fix f = let rec fx f n = app (f (lam (fx f))) n in lam (fx f)
-  let run x = x.ko (fun v -> v)
-end
-\end{code}
-
-This interpreter~|RCN| is fully polymorphic over the answer type,
-using higher-rank polymorphism through OCaml record types.
-It could also be a functor parameterized over
-the answer type.
-
-Because |RCN| has the signature |Symantics|, we can instantiate our previous
-examples with it, and all works as expected.  More interesting
-is the example $(\fun{x}1)\bigl((\fix{f}f)\mathinner2\bigr)$, which terminates
-under CBN but not CBV\@.
-\begin{code}
-module EXS(S: Symantics) = struct open S
- let diverg () = app (lam (fun x -> int 1)) 
-                     (app (fix (fun f->f)) (int 2))
-end
-\end{code}
-Interpreting |EXS| with the |R| interpreter of
-\S\ref{S:interpreter-RL} does not terminate.
-\begin{code}
-let module M = EXS(R) in M.diverg ()
-\end{code}
-In contrast, the CBN interpreter gives the result~|1|.
-\begin{code}
-let module M = EXS(RCN) in RCN.run (M.diverg ())
-\end{code}
-
-\subsection{CBV CPS transformers}
-
-Changing one definition turns our CBN CPS interpreter into CBV\@.
-\begin{code}
-module RCV = struct include RCN
-  let lam f = {ko = fun k -> k
-      (fun e -> e.ko (fun v -> f {ko = fun k -> k v}))}
-end
-\end{code}
-Now an applied abstraction
-evaluates its argument before proceeding. This approach is in
-line with \citets{reynolds-relation}, albeit typed. The
-interpreter~|RCV| is useful for CBV evaluation of the object language
-whether the metalanguage is CBV or CBN\@.
-
-We turn to a more general approach to CBV CPS: a CPS transformer that
-turns any implementation of |Symantics| into a CPS interpreter, whether
-it is an evaluator.\oleg{I don't understand this whether
-  classification: any interpreter is an evaluator, no? It all depends
-  on what the domain is. Anyway, the reviewer suggests rewriting the
-  clause as ``whether or not it is an evaluator''. I suggest to drop
-  it entirely.}
-This functor on interpreters performs a textbook
-CPS transformation on the object language.
-\begin{code}
-module CPST(S: Symantics) = struct
-  let int i = S.lam (fun k -> S.app k (S.int i))
-  let add e1 e2 = S.lam (fun k -> S.app e1 (S.lam (fun v1 ->
-                                  S.app e2 (S.lam (fun v2 ->
-                                  S.app k (S.add v1 v2))))))
-  let lam f = S.lam (fun k -> S.app k
-              (S.lam (fun x -> f (S.lam (fun k -> S.app k x)))))
-  let app e1 e2 = S.lam (fun k -> S.app e1 (S.lam (fun f ->
-                                  S.app e2 (S.lam (fun v ->
-                                  S.app (S.app f v) k)))))
-  let fix = S.fix
-end
-\end{code}
-This (abbreviated) code explicitly maps CPS interpretations to
-(direct) interpretations performed by 
-the base interpreter~|S|.
-
-The module returned by |CPST| does not define |repr|
-and thus does not have signature |Symantics|.
-The reason is again the type of |lam f|. Whereas
-|int| and |add| return the (abbreviated) type
-|('c, ..., (int -> 'w) -> 'w) S.repr|,
-the type of \texttt{lam (add (int~1))} is
-\begin{code}
-('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr
-\end{code}
-Hence, to write the type equation defining |CPST.repr| we again need
-a type function with a typecase distinction, similar to |repr_pe|
-in~\S\ref{S:PE-solution}. Alas, the type function we need is not
-identical to |repr_pe|, so we need to add another type argument to
-|repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
-terms in previous implementations of |Symantics| stay unchanged, but the
-|repr| type equations in those implementations have to take a new
-(phantom) type argument.
-
-For brevity, we just
-use the module returned by |CPST| as is. Because it does not
-match the signature |Symantics|, we cannot apply the |EX| functor to it.
-Nevertheless, we can write the tests.
-\begin{code}
-module T = struct
-  module M = CPST(C)
-  open M
-  let test1 () = (* same as before *)
-       app (lam (fun x -> x)) (bool true)
-  let testpowfix () = ... (* same as before *)
-  let testpowfix7 = (* same as before *)
-       lam (fun x -> app (app (testpowfix ()) x) (int 7))
-end
-\end{code}
-We instantiate |CPST| with the desired base interpreter~|C|,
-then use the result |M| to
-interpret object terms. Those terms are \emph{exactly} as before.
-Having to textually copy the terms is the
-price we pay for this simplified treatment.
-Our discussion of self\hyp interpretation in~\S\ref{selfinterp} shows
-that this copying is not frivolous but represents plugging a term into
-a context, which is one of the many faces of polymorphism.
-
-With 
-|CPST| instantiated by the compiler~|C| above,
-|T.test1| gives
-\begin{code}
-.<fun x_5 -> (fun x_2 -> x_2 (fun x_3 x_4 -> x_4 x_3))
-             (fun x_6 -> (fun x_1 -> x_1 true)
-                         (fun x_7 -> x_6 x_7 x_5))>.
-\end{code}
-This output is a na\"{\i}ve CPS transformation of $(\fun{x}x)\True$,
-containing several apparent $\beta$-redexes.  To reduce these
-redexes, we just change~|T| to instantiate |CPST| with |P| instead.
-\begin{code}
-{P.st = Some <fun>; P.dy = .<fun x_5 -> x_5 true>.}
-\end{code}
-
-\subsection{State and imperative features}
-\label{state}
-
-\begin{figure}
-    \begin{floatrule}
-    \begin{proofrules}
-        \[ \justifies \deref:t_s \]
-        \[ e:t_s \justifies \set e:t_s \]
-        \[ e_1:t_1 \quad \[ [x:t_1] \proofoverdots e_2:t_2 \] \justifies \lapp{e_1}{x}{e_2}:t_2 \]
-    \end{proofrules}
-    \end{floatrule}
-    \caption{Extending our typed object language with mutable state of type~$t_s$}
-    \label{fig:state}
-\end{figure}
-
-We can modify a CBN or CBV CPS transformation to pass a piece of state
-along with the continuation. This technique lets us support mutable
-state. As Figure~\ref{fig:state} shows, we extend our object language
-with three imperative features.
-\begin{enumerate}
-    \item ``$\deref$'' gets the current state;
-    \item ``$\set e$'' sets the state to the value of~$e$ and returns
-        the previous value of the state;
-    \item the let-form ``$\lapp{e_1}{x}e_2$'' evaluates $e_1$
-        before~$e_2$ even if $e_2$ does not use~$x$.
-\end{enumerate}
-If $x$ does not appear in~$e_2$, then ``$\lapp{e_1}{x}e_2$'' is same as
-the more familiar sequencing form ``$e_1;e_2$''.
-We can embed this extended object language into OCaml by extending the
-|Symantics| signature in Figure~\ref{fig:ocaml}.
-\begin{code}
-module type SymSI = sig
-  include Symantics
-  type state
-  type 'c states       (* static version of the state *)
-  val lapp : (('c,'sa,'da) repr as 'x) -> ('x -> 'y)
-             -> (('c,'sb,'db) repr as 'y)
-  val deref : unit -> ('c, 'c states, state) repr
-  val set   : (('c, 'c states, state) repr as 'x) -> 'x
-end
-\end{code}
-In HOAS, we write $\lapp{e_1}{x}e_2$ as |lapp e1 (fun x -> e2)|;
-the type of |lapp| is that of function application with
-the two arguments swapped.  We can encode the term
-``$\lapp{\deref}{x} (\set 2;\, x+\deref)$''
-as the OCaml functor
-\begin{code}
-module EXSI_INT(S: SymSI
-  with type state = int and type 'c states = int) = struct open S
-  let test1 () = lapp (deref ()) (fun x -> 
-                  lapp (set (int 2)) (fun _ -> add x (deref ())))
-end
-\end{code}
-The accompanying source code shows several more tests, including
-a test for higher-order state and a power function that uses state
-as the accumulator.
-
-The state-passing interpreter extends the CBN CPS
-interpreter |RCN| of~\S\ref{S:CPS}.
-\begin{code}
-module RCPS(ST: sig 
-  type state 
-  type 'c states 
-  type ('c,'sv,'dv) repr = 
-      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
-end) = struct
-  include ST
-  type ('c, 'sv, 'dv) result = 'c states -> 'sv
-  ... 
-  let lapp e2 e1 = {ko = fun k ->
-      e2.ko (fun v -> (app (lam e1) {ko = fun k -> k v}).ko k)}
-  let deref () = {ko = fun k s -> k s s}
-  let set e = {ko = fun k -> e.ko (fun v s -> k s v)}
-  let get_res x = fun s0 -> x.ko (fun v s -> v) s0
-end
-\end{code}
-The implementations of |int|, |app|, |lam|, and so on are
-\emph{identical} to those of |RCN| and elided. New are the extended type
-|repr|,
-which now includes the state, and the functions
-|lapp|, |deref|, and |set| representing imperative features. The
-interpreter is still CBN, so evaluating |app ef ea| might not
-evaluate |ea|, but evaluating |lapp ea ef| always does.
-For first-order state, such as of type~$\ZZ$, we 
-instantiate the interpreter as
-\begin{code}
-module RCPSI = RCPS(struct 
-  type state = int
-  type 'c states = int
-  type ('c,'sv,'dv) repr = 
-      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
-end)
-\end{code}
-If the state has a higher-order type, then the types |state| and |'c states|
-are no longer the same, and |'s states| is mutually
-recursive with the type |('c,'sv,'dv) repr|, as demonstrated in the
-accompanying source code.
-
-|RCPSI| matches the |Symantics| signature and implements the
-unextended object language: we can pass |RCPSI|
-to the functor |EX| (Fig.~\ref{fig:ocaml-simple}) and run 
-the example |test1| from there. The main use for |RCPSI| is to interpret the 
-extended object language.
-\begin{code}
-module EXPSI_INT = EXSI_INT(RCPSI)
-let cpsitesti1 = RCPSI.get_res (EXPSI_INT.test1 ()) 100
-val cpsitesti1 : int = 102
-\end{code}
-
-It is worthwhile reiterating how close this implementation is to the CPS
-interpreter, and that adding state and imperative features needed no new
-techniques.
-We can also add mutable references to the object language using
-mutable references of the metalanguage, as shown in the accompanying
-code.
-\fi
-\fi
-
-\ifshort\else
-\section{Self-interpretation}\label{selfinterp}
-
-We turn to interpreting the object language in the object language, to
-clarify how expressive our typed object language can be and to argue that our
-partial evaluator is Jones\hyp optimal.
-
-Given an \emph{encoding} of each object term~$e$ as an \emph{object} term~$\Encode{e}$,
-a \emph{self\hyp interpreter} is usually defined as an object
-function~$\si$ such that any object term~$e$ is observationally
-equivalent to the object application $\si\Encode{e}$
-\citep{jones-partial,taha-tag,Danvy-tagging-encoding}.
-A partial evaluator~$\pe$ maps object terms~$e$ to observationally
-equivalent object terms~$\pe(e)$.  It is said to be
-\emph{optimal} with respect to~|si| \citep{jones-challenging} if $\pe(\si\Encode{e})$
-is equal to~$e$ (up to $\alpha$\hyp conversion, or in some accounts, no
-less efficient than~$e$).
-
-Intuitively, self\hyp interpretation is straightforward in our
-framework: the functions comprising the interpreters
-in~\S\ref{S:interpreter-RL} may just as well be written in our object
-language:~\Fig.\ref{fig:self-eval}.
-\begin{figure}
-\begin{align*}
-    \ident{int} &= \fun{x} x \\
-    \ident{add} &= \fun{x} \fun{y} x+y \\
-    \ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}
-        \displaybreak[0] \\
-    \ident{lam} &= \fun{f} f \\
-    \ident{app} &= \fun{f} \fun{x} fx \\
-    \ident{fix} &= \fun{g} \fix{f} \fun{x} gfx
-\end{align*}
-\caption{The \emph{object} functions implementing an
-evaluator.  (We use the number~$0$ in lieu of a unit value.)}
-\label{fig:self-eval}
-\end{figure}
-We thus map each object term~$e$ to an object term~$\encode{e}$ as follows.
-We call this mapping \emph{pre-encoding}.
-\begin{align*}
-    \encode{x} &= x \\
-    \encode{n} &= \ident{int}\, n \\
-    \encode{e_1 + e_2} &= \ident{add} \encode{e_1} \encode{e_2} \\
-    \encode{\cond{b}{t}{e}} &= \rlap{$\ident{if\_} \encode{b}
-        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$}
-        \displaybreak[0] \\
-    \encode{\fun{x}e} &= \ident{lam} (\fun{x} \encode{e}) \\
-    \encode{fx} &= \ident{app} \encode{f} \encode{x} \\
-    \encode{\fix{f}e} &= \ident{fix} (\fun{f} \encode{e})
-\end{align*}
-The metavariables $x$ and~$n$ stand for a variable and an integer,
-respectively.
-This pre-encoding is just like how we represent object terms in the
-metalanguage in the preceding sections, but it produces
-terms in the object language rather than the metalanguage.
-
-To evaluate~$\encode{e}$, then, we
-instantiate the free variables in~$\encode{e}$ such as $\ident{int}$,
-$\ident{lam}$, and $\ident{add}$ by their definitions given in
-Fig.~\ref{fig:self-eval}.  For
-example, the familiar object term $(\fun{x}x)\True$ pre-encodes to
-\begin{equation*}
-    \encode{(\fun{x}x)\True} = \ident{app}
-    (\ident{lam} (\fun{x} x))\, (\ident{bool} \True),
-\end{equation*}
-and to evaluate this pre-encoded term is to evaluate the object term
-\begin{equation*}
-    (\fun{f} \fun{x} fx) \,
-    ((\fun{f} f) (\fun{x} x))\, ((\fun{b} b) \True).
-\end{equation*}
-Because the evaluator in Fig.~\ref{fig:self-eval} mostly consists 
-of glorified identity
-functions, our simple partial evaluator reduces the
-result of this instantiation to~$e$.  In general, to
-interpret~$\encode{e}$ using an interpreter is to instantiate its free
-variables by that interpreter's definitions.
-
-\subsection{Avoiding higher polymorphism}
-
-Any approach to self\hyp interpretation needs to spell out first how to
-encode object terms~$e$ to object terms~$\Encode{e}$, and then how to
-interpret~$\Encode{e}$ in the object language.  For our approach, we
-want to define encoding in terms of pre-encoding, and interpretation
-using some notion of instantiation.  Unfortunately, the simple type
-structure of our object language hinders both tasks.  To continue with
-the example term above, we could try to define
-\begin{equation*}
-    \Encode{e} =
-    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}} \encode{e},
-\end{equation*}
-in particular
-\begin{equation*}
-    \Encode{(\fun{x}x)\True} =
-    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
-    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool} \True).
-\end{equation*}
-To type-check this encoded term, we give the variable $\ident{lam}$ the
-simple type $(\BB\to\BB)\to\BB\to\BB$.
-We then define the self\hyp interpreter
-\begin{equation*}
-    \si = \fun{e} e
-    (\fun{f} \fun{x} fx)
-    (\fun{f} f)
-    (\fun{b} b)
-\end{equation*}
-and apply it to the encoded term.  The result is the object term
-\begin{equation*}
-    \bigl(\fun{e} e (\fun{f} \fun{x} fx) (\fun{f} f) (\fun{b} b)\bigr)
-    \bigl(
-    \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
-    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool} \True)
-    \bigr),
-\end{equation*}
-which partially evaluates to~$\True$ easily.
-However, encoding fails on a term with multiple $\lambda$\hyp
-abstractions at different types.  For example, the pre-encoding
-\begin{equation*}
-    \encode{\fun{f}\fun{x}fx}
-    = \ident{lam} (\fun{f} \ident{lam} (\fun{x} \ident{app} f x))
-\end{equation*}
-does not type-check in any typing environment, because $\ident{lam}$ needs
-to take two incompatible types.  In sum, we need more polymorphism in the
-object type system to type $\ident{lam}$, $\ident{app}$, $\ident{fix}$,
-and~$\ident{if\_}$ (and $\Encode{e}$ and~$\si$).
-(The polytypes in |Symantics| given by Haskell's type classes and OCaml's
-modules supply this polymorphism.)  Moreover, we need to encode any
-polymorphism of the object language \emph{into} the object language to achieve
-self\hyp interpretation.
-
-\subsection{Introducing let-bound polymorphism}
-
-Instead of adding higher-rank and higher-kind polymorphism to our object
-language (along with polymorphism over kinds!), we add let-bound polymorphism.
-As usual,
-we can add a new typing rule
-\begin{equation*}
-    \let\vcenter\vbox
-    \begin{prooftree}
-        e_1:t_1 \quad \subst{e_2}{x}{e_1}:t_2
-        \justifies \be{x=e_1} e_2 : t_2
-    \end{prooftree}
-    .
-\end{equation*}
-The pre-encoding of a let\hyp expression is trivial.
-\begin{align*}
-    \encode{\be{x=e_1}e_2} \quad &= \quad \be{x=\encode{e_1}} \encode{e_2}
-\intertext{A \emph{context} is an object term with a hole~$[~]$.  The
-hole may occur under a binder, so plugging a term into the context may
-capture free variables of the term.  By pre-encoding a hole to a hole,
-we extend pre-encoding from a translation on terms to one on
-contexts.}
-    \encode{[~]} \quad &= \quad [~]
-\end{align*}
-We define an interpreter in the object language to be not a term but
-a context.  For example, the evaluator is the context
-\begin{align*}
-    \si[~] &=
-    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
-        \be{\ident{int} &= \fun{x} x&} \\
-        \be{\ident{add} &= \fun{x} \fun{y} x+y&} \\
-        \be{\ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}&} \\
-        \be{\ident{lam} &= \fun{f} f&} \\
-        \be{\ident{app} &= \fun{f} \fun{x} fx&} \\
-        \be{\ident{fix} &= \fun{g} \fix{f} \fun{x} gfx&} [~],
-    \end{tabular}
-\intertext{and the size\hyp measurer at the end of~\S\ref{S:interpreter-RL} is the context}
-    \mathrm{SZ}[~] &=
-    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
-        \be{\ident{int} &= \fun{x} 1&} \\
-        \be{\ident{add} &= \fun{x} \fun{y} x+y+1&} \\
-        \be{\ident{if\_}&= b + t\,0 + e\,0&} \\
-        \be{\ident{lam} &= \fun{f} f\,0 + 1&} \\
-        \be{\ident{app} &= \fun{f} \fun{x} f + x + 1&} \\
-        \be{\ident{fix} &= \fun{g} g\,0 + 1&} [~].
-    \end{tabular}
-\end{align*}
-To interpret an object term~$e$ using an interpreter $I[~]$ is to
-evaluate the object term~$I[\encode{e}]$.  $\si$~is
-a self\hyp interpreter in the following sense.
-\begin{proposition}
-    $\si[\encode{e}]$ is observationally equivalent to~$e$.
-\end{proposition}
-As a corollary, we can pre-encode $\si$ itself as
-a context: the term $\si[\encode{\si[\encode{e}]}]$ is observationally
-equivalent to $\si[\encode{e}]$, and in turn to~$e$.  In other words,
-$\si$ can interpret itself.  Our partial evaluator is optimal with respect to
-the self\hyp interpreter~$\si$.
-\begin{proposition}
-    Let $\pe$ be the partial evaluator~|P| in~\S\ref{S:PE-solution}.
-    Then the object terms $\pe(\si[\encode{e}])$ and~$\pe(e)$ are
-    either both undefined or both defined and
-    equal up to $\alpha$\hyp conversion.
-\end{proposition}
-
-\subsection{Contexts clarify polymorphism}
-\label{S:clarify}
-
-We always type-check a pre-encoded term~$\encode{e}$ and its interpreter~$I[~]$
-together, never separately.  This treatment has the drawback that
-we must duplicate a pre-encoded term in order to interpret it in
-multiple ways.  The meta-notions of contexts and plugging may seem ad
-hoc, but in fact they just reflect the type-class and module
-machinery that we have been using in this paper all along.
-
-In the presence of let-bound polymorphism, we can understand a term
-waiting to be plugged into a context as a higher-rank and higher-kind
-abstraction over the context.  Even though our object language does not support
-higher abstraction, our metalanguages do, so they can type-check an object term
-separately from its interpreter---either as a functor from a |Symantics| module
-containing a type constructor
-(in OCaml), or a value with a |Symantics| constraint over a type
-constructor (in Haskell).  Thus,
-``context'' is a euphemism for a polymorphic argument, and ``plugging''
-is a euphemism for application.
-
-\begin{comment}
-\jacques{But how 
-do you create, in either Haskell or MetaOCaml, an untypechecked 
-interpreter-with-a-hole [UIH] ?}
-\oleg{Well, one can make an argument that we already have such an
-interpreter with polymorphic let and the hole: incope. In Haskell, the
-declaration of an instance of Symantics is like the sequence of
-polymorphic lets. We construct terms where lam, add, etc, are free
-variables. We apply the interpreter to the semantics (plug the hole)
-by instantiating these terms (binding the free variables lam, etc. to
-the particular instance of Symantics). The unRR construction does
-this plugging in explicitly.}
-Again, what is different from the above is that we can
-typecheck terms separately, without inserting them first within the
-hole of a particular interpreter. Rank-2 type of |repr| helps. It lets
-enough of the type information out so the typechecking can
-proceed. So, |repr| is the representation of the polymorphic
-interpreter context with the hole, which permits separately
-typecheckable terms (the evaluation still entails `duplication' so to
-speak -- which is one way how polymorphism is resolved).
-
-\oleg{Mention the RR interpreter: for any term E, (RR E) is equivalent to E.
-RR is not an identity: it is an interpreter that encapsulates another
-interpreter. The RR interpreter can be made self if we treat RR as a
-special form and interpret it always with itself (similar to let and
-hole below).}
-\end{comment}
-
-\begin{comment}
-
-The crucial role of the higher-order type parameter r
-
-The type constructor "r" above represents a particular interpreter.  The
-meta-type "r tau" hides how the interpreter represents the object type
-"tau" yet exposes enough of the type information so we can type-check
-the encoding of an object term without knowing what "r" is.  The checked
-term is then well-typed in any interpreter.  Each instance of Symantics
-instantiates "r" to interpret terms in a particular way. The L
-interpreter is quite illustrative. We need the above semantics to be
-able to represent both R and L (the latter returns only Int as the
-values) in the same framework.
-
-We encode a term like |add 1 2| as
-\texttt{app \_add (app \_int 1) (app \_int 2)} where |_add| and |_int| are just
-`free variables'. Now, how to typecheck such a term? Some type should
-be assigned to these free variables. The goal is to complete the work
-without needing any type annotations (so we don't have to introduce any
-type language), with all types inferred and all terms typed. It seems
-the second-order type R neatly separates the typechecking part from
-the representation of R: it hides aspects that depend on the
-particular interpreter, and yet lets enough type information through
-(via its type argument) to permit the typechecking of terms, and infer
-all the types.
-
-The only approach that does seem to work
-is the one in incope.hs or incope.ml. If we de-sugar away records and
-type-classes, the type of a term L of the inferred type tau is
-$$ 
-  (\ZZ \rightarrow r \ZZ) \rightarrow
-  (\BB \rightarrow r \BB) \rightarrow
-  \forall \alpha \beta. (r \alpha \rightarrow r \beta)
-      \rightarrow r (\alpha\rightarrow\beta) \rightarrow ... r \tau
-$$
-% (Int -> r Int) ->
-% (Bool -> r Bool) ->
-% (forall alpha beta. (r alpha -> r beta) -> r (alpha->beta)) -> ... r tau
-
-or, if we denote the sequence of initial arguments as |S r|, terms have
-the type |S r -> r tau|
-The interpreter has the type
-|(forall r. S r -> r tau) -> r' tau|
-
-The higher-order type (variable) of kind |*->*| seems essential. So, at
-least we need some fragment of Fw (somehow our OCaml code manages to
-avoid the full Fw; probably because the module language is separated
-from the term language). Thus, we seem to need a fragment of Fw. It
-seems the inference is possible, as our Haskell and OCaml code
-constructively illustrates. Perhaps we need to characterize our
-fragment.
-
-\end{comment}
-\fi
-
-\section{Related work}\label{related}
-
-\ifshort
-Our initial motivation came from several papers 
-\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}
-that use embedded interpreters to justify advanced
-type systems, in particular GADTs.
-\else
-Our initial motivation came from several papers that justify advanced
-type systems, in particular GADTs, by embedded interpreters
-\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} and
-CPS transformations \citep{Guillemette-Monier-PLPV,shao-type-toplas,chen-typeful}.
-\fi
-We admire all this technical machinery, but
-these motivating examples do not need it.
-Although GADTs may indeed be simpler and more flexible, they are
-unavailable in mainstream ML, and their implementation in GHC
-6.6.1 fails to
-detect exhaustive pattern matching.  We also wanted to find the minimal
-set of widespread language features needed for tagless
-type-preserving interpretation.
-
-Even a simply typed $\lambda$-calculus obviously supports self\hyp
-interpretation, provided we use universal types \citep{taha-tag}.  The
-ensuing tagging overhead motivated \citet{Makholm-TagElim,taha-tag} 
-to propose tag
-elimination, which however does not statically guarantee that all tags
-will be removed \citep{WalidICFP02}.
-
-\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
-\citet{peyton-jones-simple} seem to argue as follows that a self\hyp
-interpreter of a typed language cannot be tagless or Jones\hyp optimal:
-\begin{shortlist}
-\item One needs to encode a typed language in a typed language based on
-a sum type (at some level of the hierarchy)\\
-\item A \emph{direct} interpreter 
-for such an encoding of a typed language
-in a typed language requires either
-advanced types or tagging overhead\\
-\item Thus, an indirect interpreter is necessary, which needs a universal
-  type and hence tagging\ifshort\else\\
-\item Thus, any self-interpreter must have tags and cannot be 
-  Jones-optimal\fi.
-\end{shortlist}
-While the logic is sound, we (following \citet{yang-encoding}) showed that the
-first step's premise is not valid.
-
-\citet{Danvy-tagging-encoding} discuss Jones optimality at length and
-apply HOAS to typed self\hyp interpretation.  However, their source
-language is untyped.  Therefore, their object\hyp term encoding has
-tags, and their interpreter can raise run-time errors.
-Nevertheless, HOAS lets the partial
-evaluator remove all the tags. In contrast, our object encoding and
-interpreters do not have tags to start with and obviously cannot
-raise run-time errors.
-
-\ifshort\else
-A lot of effort has gone into ``typing dynamic typing'': to statically
-type-check dynamically\hyp typed values
-\citep{baars-typing,WalidICFP02,Guillemette-Monier-PLPV,haskell-list},
-using the host language's type system to varying extents.
-Our object terms are statically typed, so we would
-need one of these techniques to interpret dynamically\hyp typed
-terms such as those read from a file.
-\oleg{First, I suggest we remove the subjunctive mood: `we would
-  need...' We have the code already, IncopeTypecheck.hs. We know we
-  can typecheck untyped terms so that we do the typechekcing only
-  once, and interpret many times. All these interpretation is guranteed to
-  be free from any type errors. And this holds even for higher-order
-  terms encoded in HOAS. I also suggest we drop
-  \cite{haskell-list}. We included the citation solely to show how 
-  we could go on. Both links in haskell-list are not actually
-  appropriate: first, the object language is \emph{first-order}
-  and second they use GADTs. Our IncopeTypecheck uses only
-  Typeable and existentials, and no GADTs.
-  Finally, we need to find a way to talk about IncopeTypecheck more
-  directly, probably in the Introduction -- Sumii-san has directly
-  suggested we emphasize that result. For some reason, people (Walid
-  and Sumii-san, for ones) consider it important. Incidentally, I
-  have the comparison of IncopeTypecheck with `Typing Dynamic
-  Typing', which we can use if needed.}
-\fi
-
-Our partial evaluator establishes a bijection |repr_pe| between static
-and dynamic types (the valid values of |'sv| and |'dv|), and between
-static and dynamic terms.  It is customary to implement such a bijection
-using an injection\hyp projection pair, as done for interpreters
-\ifshort \citep{Ramsey-ML-module-mania,Benton-embedded-interpreters}\else
-by \citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters}\fi,
-partial evaluation \ifshort \citep{Danvy-TDPE}\else by \citet{Danvy-TDPE}\fi,
-and type-level functions \ifshort \citep{oliveira-typecase}\else by
-\citet{oliveira-typecase}\fi.  As explained in~\S\ref{S:PE-solution}, we
-avoid injection and projection at the type level by adding an argument
-to |repr|.
-Our solution could have been even more straightforward if MetaOCaml
-provided total type-level functions such as |repr_pe| in
-\S\ref{S:PE-solution}---simple type-level computations 
-ought to become mainstream.
-
-At the term level, we also avoid converting between static and dynamic
-terms by building them in parallel, using \citearound{'s
-method}\citet{asai-binding-time}.
-This method type-checks in Hindley-Milner once we
-deforest the object term representation.  Put another way, we
-manually apply type-level partial evaluation to our type
-functions (see \S\ref{S:PE-solution}) to obtain simpler types 
-acceptable to MetaOCaml.
-
-\Citet{sumii-hybrid} also use Asai's method,
-to combine online
-and offline partial evaluation.
-They predate us in deforesting the object term representation to enable tagless
-partial evaluation.
-We strive for modularity by reusing interpreters for individual stages
-\citep{SperberThiemann:TwoForOne}: our partial evaluator~|P|
-reuses our tagless evaluator~|R| and tagless compiler~|C|, so it
-is patent that the \emph{output} of~|P| never gets
-stuck.  It would be interesting
-to try to derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
-in the same manner.
-
-It is common to implement an embedded DSL by providing multiple
-interpretations of host\hyp language pervasives such as addition and
-application.  It is also common to use phantom types to rule out
-ill-typed object terms, as done in Lava
-\citep{Lava} and by \citet{Rhiger-thesis}. However, these approaches
-are not tagless because they still use universal types, such as Lava's
-\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Fig.~2.2)
-and \texttt{Term} (his Chap.~3), which incur the attendant overhead of
-pattern matching.  The universal type also
-greatly complicates the soundness and completeness proofs of embedding
-\citep{Rhiger-thesis}, whereas our proofs are trivial.
-Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
-\begin{comment}
-Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
-upon the Exp datatype. But that is cheating: phantom type means
-essentially we can easily do coerce. We use real types.  That's why he
-had to do tedious proofs in Sec 2 of soundness and completeness of
-embedding. Whereas our proofs are obvious.  His sec 3 is based on data
-representation of terms. They have type tags.  We do nothing of that
-kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
-type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
-cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
-thesis contains the definitions of the interpreter and the compiler,
-in the beginning. Use this in response to Rev1)
-\end{comment}
-
-We are not the first to implement a typed interpreter for a typed
-language.  \Citet{laod93} use type classes to implement a metacircular
-interpreter (rather than a self\hyp interpreter) of a
-typed version of the SK language, which is quite different from our
-object language.  Their interpreter
-appears to be tagless, but they could not have implemented a
-compiler or partial evaluator in the same way, since they rely
-heavily on injection\hyp projection pairs.
-
-\Citet{fiore:nbe-ppdp2002} and \citet{balat:tdpe-popl2004} also build
-a tagless partial evaluator, using delimited control operators.  It is
-type-directed, so the user must represent, as a term, the type of every
-term to be partially evaluated.  We shift this work to the type checker
-of the metalanguage.  By avoiding term-level type representations, our
-approach makes it easier to perform algebraic simplifications (as
-in~\S\ref{S:PE-solution}).
-
-% The following paragraph was not in the version submitted to APLAS,
-% yet was not commented out.  Of course, if we had space, we should add
-% this, but it is difficult to see how.
-\ifshort\else
-Using Haskell, \citet{Guillemette-Monier-PLPV} implement a CPS transformation
-for HOAS terms and statically assure that it preserves object types.
-They represent proofs of type preservation as terms of a GADT, which is not
-sound (as they admit in \S4.2) without a separate totality check because
-any type is trivially inhabited by a nonterminating term in Haskell.
-In contrast, our CPS transformations use simpler types than GADTs and
-assure type preservation at the (terminating) type level rather than
-the term level of the metalanguage.
-\Citeauthor{Guillemette-Monier-PLPV} review other
-type\hyp preserving CPS transformations (mainly in the context of typed intermediate
-languages), in particular
-\citets{shao-type-toplas} and \citets{chen-typeful}.
-These approaches use de Bruijn indices and fancier
-type systems with type-level functions, GADTs, or type\hyp equality
-proofs.
-\fi
-
-We encode terms in elimination form, as a coalgebraic structure.
-\Citet{Pfenning-Lee} first described this basic idea and applied it to
-metacircular interpretation.
-Our approach, however, can be implemented in mainstream ML and supports
-type inference, typed CPS transformation and partial evaluation. In contrast,
-\citeauthor{Pfenning-Lee} conclude that partial evaluation and program
-transformations ``do not seem to be expressible'' even using their
-extension to~$F_\omega$, perhaps because their avoidance of general
-recursive types compels them to include the polymorphic lift that we
-avoid in~\S\ref{S:PE-lift}.
-\begin{comment}
-It seems that Pfenning and Lee embed $F_2$ with type constructions in 
-(pure) $F_3$.  We embed $F_1$ in (weak?) $F_2$, as I see it.  In a way, what 
-we do is very similar to what they do (Figure 1, p.152), except that we 
-do it in standard programming languages.  It is unclear if their work can 
-be implemented (yet) in any language.  And we preserve type-inference, 
-while their solution needs explicit types!
-The following line of their conclusion is worth citing: "... this does 
-not imply that the same language is also suitable for type 
-metaprogramming. ... such as partial evaluation... do not seem to be 
-expressible".
-I suspect you're right, but I'm still reading the paper.  See also page
-146: "for a term M in $F_1$ (a simply-typed term), the representation
-$\bar{M}$ will be in $F_2$".  The move from $F_2$ to $F_3$ 
-and beyond reminds me
-strongly of our attempts at self-interpretation without the notion of a
-syntactic hole.
-\end{comment}
-
-Our encoding of the type function |repr_pe| in \S\ref{S:PE-solution}
-emulates type-indexed types and is related to intensional type analysis
-\citep{Morrisett-intensional,Generic-Haskell}. However, our object
-language and running examples in HOAS include |fix|,
-which intensional type analysis cannot handle
-\citep{xi-guarded}.  Our final approach
-seems related to \citearound{'s approach to HOAS using catamorphisms and
-  anamorphisms}\citet{Washburn-Weirich-boxes}.
-
-We could not find work that establishes that
-the \emph{typed} $\lambda$-calculus has a final coalgebra structure.
-\ifshort
-(See \Citet{honsell99coinductive} for the untyped case.)
-\else
-\Citet{HonsellLenisa,honsell99coinductive}
-investigate the untyped $\lambda$-calculus
-along this line.  
-In particular, they use
-contexts with a hole \citep[p.\,13]{honsell99coinductive} to define
-\emph{observational equivalence}
-(see our~\S\ref{selfinterp}).
-\citearound{'s bibliography}\Citet{honsell99coinductive} refers to the
-foundational work in this important area.  
-Particularly intriguing is the link to the
-coinductive aspects of B\"{o}hm trees, as pointed out by
-\citet{berarducci-models} and Jacobs \citeyearpar[Example 4.3.4]{jacobs-coalgebra}.
-\fi
-
-\ifshort\else
-One way to understand our main idea is to eschew sum types and sum kinds
-for their dual, record types and record kinds.
-For the self\hyp interpreter, we then proceed to use a Church encoding for
-recursive data types \citep{bohm-automatic}.
-\fi
-
-\ifshort We observe that \else As \S\ref{S:clarify} observes, \fi
-higher-rank and higher-kind
-polymorphism lets us type-check and compile object terms separately from
-interpreters.  This \ifshort\else observation \fi is consistent with the role of
-polymorphism in the separate compilation of modules
-\citep{shao-typed}.
-
-\section{Conclusions}\label{conclusion}
-
-We solve the problem of embedding a typed object language in a typed
-metalanguage without using GADTs, dependent types, or a universal type.
-Our family of interpreters includes an evaluator, a compiler, a partial
-evaluator, and CPS transformers.  It is patent that they never get stuck,
-because we represent object types as metalanguage types.  This work
-makes it safer and more efficient to embed DSLs
-in practical metalanguages such as Haskell and ML\@.
-
-Our main idea is to represent object programs not in an initial algebra
-but using the existing coalgebraic structure of the $\lambda$-calculus.
-More generally, to squeeze more invariants out of a type system as
-simple as Hindley-Milner, we shift the burden of representation and
-computation from consumers to producers: encoding object terms as calls
-to metalanguage functions (\S\ref{ourapproach}); build dynamic terms
-alongside static ones (\S\ref{S:PE-lift}); simulating type functions for
-partial evaluation (\S\ref{S:PE-solution}) and CPS
-transformation\ifshort\else~(\S\ref{S:CPS})\fi.
-This shift also underlies fusion,
-functionalization, and amortized complexity analysis.
-\ifshort\else
-When the metalanguage does provide higher-rank and higher-kind
-polymorphism, we can type-check and compile an object term separately
-from any interpreters it may be plugged into.
-\fi
-
-Our representation of object terms in elimination form encodes
-primitive recursive folds over the terms. 
-\ifshort\else
-This encoding makes operations like interpretation trivial to implement.
-\fi
-We still
-have to understand if and how non-primitively 
-recursive operations can be supported.

### 2008-06-05 tagless/tagless-final.tex
Ken starts to work on some reviewer comments

@@
 \ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
-using cogen functions rather than data constructors.
+using compiler-generator (cogen) functions rather than data constructors.
 In other words, we represent object terms not in an initial algebra
@@
 
-Our solution is \emph{not} Church-encoding the
+Unlike \citearound{'s HOAS-based term representation}\citet{JFP-Mogensen},
+our solution is \emph{not} Church-encoding the
 universal type. The Church encoding of the type~|u| in \S\ref{tagproblem}
@@
     such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
+    \ Our term encoding is contravariant in the object language, so
+    extending the language does not invalidate terms already encoded.
 \item We describe an approach to self\hyp interpretation compatible with the
@@
 \else
-Rather, we solve an open problem by a novel combination of simple types
-and existing techniques.  More precisely, we
-\fi
-use features already present in
-mainstream functional languages---Hindley-Milner
-type system with either an inference\hyp preserving module system or
-constructor classes, as realized in ML and Haskell 98---and techniques
-which have all appeared in the literature (in particular, \cite{yang-encoding,sumii-hybrid}), to
-solve a problem that was stated in the published record as
-unsolved and likely unsolvable in ML or Haskell 98
-without extensions. The simplicity of our solution and its
+Rather, we
+\fi
+solve a problem that was stated in the published record as open and
+likely unsolvable in ML or Haskell 98 without extensions, by a novel
+combination of simple types and techniques already described in the
+literature that use features present in mainstream functional languages.
+In particular, we follow \citearound{'s encoding of type-indexed
+values}\citet{yang-encoding}, \citearound{'s construction of dynamic
+terms alongside static terms}\citet{asai-binding-time}, and
+\citearound{'s deforestation of syntax
+constructors}\citet{sumii-hybrid}.  These techniques require just
+a Hindley-Milner type system with either a module system that supports
+functors or constructor classes, as realized in all variants of ML and
+Haskell.
+The simplicity of our solution and its
 use of only mainstream features \ifshort\else are virtues that \fi make it more practical to build typed,
@@
 language.
+We shorten some of the types using OCaml's |as| syntax.
 
@@
 
-Inserting brackets and escapes appropriately into the
-evaluator~|R| above yields the simple compiler~|C|
+To turn the evaluator~|R| into a simple compiler, we bracket the
+computation on values to be performed at run time, then escape the code
+generation from terms to be performed at compile time.  Adding these
+stage annotations yields the compiler~|C|
 \ifshort in Fig.~\ref{fig:interpreter-C}(a).\else below.
@@
 \ifshort
-\section{A tagless partial evaluator}\label{PE}
-Surprisingly, we can write a partial evaluator using the idea above,
-namely to build object terms using ordinary functions rather than data
-constructors.  We present this partial evaluator in a sequence of three
-attempts. It uses no universal type and no tags
-for object types.  We then discuss residualization and binding-time
-analysis.  Our partial evaluator is a modular extension of the evaluator
-in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{S:compiler}, in
-that it uses the former to reduce static terms and the latter to build
-dynamic terms.
-
-\subsection{Avoiding polymorphic lift}
-\label{S:PE-lift}
-
-Roughly, a partial evaluator interprets each object term to yield either
-a static (present-stage) term (using\ifshort\else\ the evaluator\fi~|R|) or
-a dynamic (future-stage) term (using\ifshort\else\ the
-compiler\fi~|C|). To
-distinguish between static and dynamic terms, we might try to define
-|repr| in the partial evaluator as
-\texttt{type ('c,'dv)\,repr 
-	= S0 of ('c,'dv)\,R.repr~}\Verb+|+\texttt{~D0 of ('c,'dv)\,C.repr}.
-Integer and boolean literals are immediate, present-stage
-values. Addition yields a static term (using~|R.add|) if and only 
-if both operands are static; otherwise we extract the dynamic terms 
-from the operands and add them using~|C.add|. We use |C.int| to
-convert from the static term |('c,int) R.repr|, which is just |int|,
-to the dynamic term.
-
-Whereas |mul| and |leq| are as easy to define as |add|, we encounter
-a problem with |if_|.  Suppose that the first argument to |if_| is a dynamic term
-(of type |('c,bool) C.repr|), the second a static term 
-(of type |('c,'a) R.repr|), and the third a
-dynamic term (of type |('c,'a) C.repr|). We then need to convert
-the static term to dynamic, but there is no polymorphic ``lift''
-function, of type |'a -> ('c,'a) C.repr|, to send a value to the future stage
-\citep{xi-guarded,WalidPOPL03}.
-
-Our |Symantics| only includes separate lifting methods |bool| and
-|int|, not a parametrically polymorphic lifting method, for good reason:
-When compiling to a first-order target language such as machine code,
-booleans, integers, and functions may well be represented differently.
-Thus, compiling polymorphic lift requires intensional type
-analysis.  To avoid needing polymorphic lift, we turn to
-Asai's technique \citep{asai-binding-time,sumii-hybrid}:
-build a dynamic term
-alongside every static term.
-
-\subsection{Delaying binding-time analysis}
-\label{S:PE-problem}
-
-We switch to the data type
-\texttt{type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr}
-so that a partially evaluated term always contains a dynamic
-component and sometimes contains a static component.  By 
-distributivity, the two
-alternative constructors of an |option| value, |Some| and |None|,
-tag each partially evaluated term with a phase: either present or
-future.  This tag is not an object type tag: all pattern matching below
-is exhaustive.  Because the future-stage component is always available, we
-can now define the polymorphic function
-|let abstr1 (P1 (_,dyn)) = dyn| of type
-\texttt{('c,'dv) repr -> ('c,'dv) C.repr}
-to extract it without requiring polymorphic lift into~|C|.  We then try
-to define the interpreter |P1|---and get as far as the first-order
-constructs of our object language, including |if_|.
-\begin{code3}
-module P1 : Symantics = struct
-  let int (x:int) = P1 (Some (R.int x), C.int x)
-  let add e1 e2 = match (e1,e2) with
-    | (P1 (Some n1,_),P1 (Some n2,_)) -> int (R.add n1 n2)
-    | _ -> P1 (None,(C.add (abstr1 e1) (abstr1 e2)))
-  let if_ = function
-    | P1 (Some s,_) -> fun et ee -> if s then et () else ee ()
-    | eb -> fun et ee -> P1 (None, C.if_ (abstr1 eb) 
-                                   (fun () -> abstr1 (et ()))
-                                   (fun () -> abstr1 (ee ())))
-\end{code3}
-However, we stumble on functions.  According to our
-definition of~|P1|, a partially evaluated object function, such as the
-identity $\fun{x}x$ embedded in OCaml as |lam (fun x -> x)|\texttt{ :
-}|('c,'a->'a) P1.repr|, consists of a dynamic part 
-(type |('c,'a->'a) C.repr|) and
-maybe a static part (type |('c,'a->'a) R.repr|).  The dynamic part is useful
-when this function is passed to another function that is only
-dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
-useful when this function is applied to a static argument, as in
-$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
-evaluate the function, that is, compute as much as possible statically
-when it is applied to a mix of static and dynamic inputs.  For example,
-the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
-by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
-not statically known.  The same static function, applied to
-different static arguments, can give both static and dynamic results: we
-want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
-to~$x$.
-
-To enable these simplifications, we delay binding-time analysis
-for a static function until it is applied, that is, until |lam f|
-appears as the argument of |app|.  To do so, we have to incorporate |f|
-as it is into the |P1.repr| data structure: the representation
-for a function type |'a->'b| should be one of
-\begin{code3}
-S1 of ('c,'a) repr -> ('c,'b) repr | E1 of ('c,'a->'b) C.repr
-P1 of (('c,'a) repr -> ('c,'b) repr) option * ('c,'a->'b) C.repr
-\end{code3}
-unlike |P1.repr| of |int| or |bool|.
-That is, we need a nonparametric data type, something akin to
-type-indexed functions and type-indexed types, which
-\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
-Thus, typed partial evaluation, like typed CPS transformation,
-inductively defines a map from source types to target types that
-performs case distinction on the source type. In Haskell, typecase
-can be equivalently implemented either with GADTs or with
-type-class functional dependencies
-\citep{oliveira-typecase}. The accompanying code shows both
-approaches, neither portable to OCaml. In addition,
-the problem of nonexhaustive pattern\hyp matching reappears in
-the GADT approach because GHC 6.6.1 cannot see that a particular
-type of a GADT value precludes certain constructors.
-Thus GADTs fail to
-make it \emph{syntactically} apparent that pattern matching is exhaustive.
-
-\subsection{The ``final'' solution}
-\label{S:PE-solution}
-
-Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
-would ideally like is to write
-\begingroup\sloppy
-\texttt{type ('c,'dv) repr = P1 of (repr\_pe ('c,'dv)) R.repr option * ('c,'dv) C.repr}
-where |repr_pe| is the type function defined
-% inductively because P below depends on repr_pe
-by
-\begin{code3}
-repr_pe ('c,int) = ('c,int); repr_pe ('c,bool) = ('c,bool)
-repr_pe ('c,'a->'b) = ('c,'a) repr -> ('c,'b) repr
-\end{code3}
-\endgroup\noindent
-Although we can use type classes to define this type function
-in Haskell, that is not portable to MetaOCaml. However,
-these three typecase alternatives are already present in existing
-methods of |Symantics|.
-A simple and portable solution thus emerges: we bake |repr_pe| 
-into the signature |Symantics|. 
-We recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
-constructor took two arguments |'c| and~|'dv|. We add an argument
-|'sv| for the result of applying |repr_pe| to~|'dv|.
-Figure~\ref{fig:ocaml} shows the new signature.
-\begin{figure}
-\begin{floatrule}
-\smallskip
-\begin{code2}
-module type Symantics = sig type ('c,'sv,'dv) repr
-  val int : int -> ('c,int,int) repr
-  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-            -> ('c,'x,'da -> 'db) repr
-  val app : ('c,'x,'da -> 'db) repr
-            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
-  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
-                                'da -> 'db) repr as 'x)
-  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
-  val if_ : ('c,bool,bool) repr
-            -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x) end
-\end{code2}
-\medskip
-\end{floatrule}
-\caption{A (Meta)OCaml embedding of our object language that supports
-  partial evaluation \ifshort \protect\linebreak[2] (\texttt{bool},
-  \texttt{mul}, \texttt{leq} are elided)\fi}
-\label{fig:ocaml}
-\end{figure}
-
-\begin{figure}
-\begin{floatrule}
-\begin{code2}
-module P = struct
-  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
-  let abstr {dy = x} = x    let pdyn x = {st = None; dy = x}
-
-  let int  (x:int)  = {st = Some (R.int x);  dy = C.int x}
-  let add e1 e2 = match e1, e2 with
-  | {st = Some 0}, e | e, {st = Some 0} -> e
-  | {st = Some m}, {st = Some n} -> int (R.add m n)
-  | _ -> pdyn (C.add (abstr e1) (abstr e2))
-  let if_ eb et ee = match eb with
-  | {st = Some b} -> if b then et () else ee ()
-  | _ -> pdyn (C.if_ (abstr eb) (fun () -> abstr (et ()))
-                                (fun () -> abstr (ee ())))
-  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
-  let app ef ea = match ef with {st = Some f} -> f ea
-                  | _ -> pdyn (C.app (abstr ef) (abstr ea)) end
-\end{code2}
-\medskip
-\end{floatrule}
-\caption{Our partial evaluator (\ifshort \texttt{bool}, \texttt{mul},
-  \texttt{leq} and \texttt{fix} \else \texttt{mul} and \texttt{leq} \fi
-  are elided)}
-\label{fig:pe}
-\end{figure}
-
-The interpreters |R|, |L| and~|C| above only use the old
-type arguments |'c| and~|'dv|, which are treated by the new signature
-in the same way.  Hence, all that needs to change in these interpreters
-to match the new signature is to add a phantom type
-argument~|'sv| to~|repr|.
-For example, the compiler |C| now begins
-\ifshort
-\texttt{module C = struct
-  type ('c,'sv,'dv) repr = ('c,'dv) code}
-\else
-\begin{code}
-module C = struct
-  type ('c,'sv,'dv) repr = ('c,'dv) code
-\end{code}
-\fi
-with the rest the same.
-In contrast, the partial evaluator~|P| relies on the type argument |'sv|.
-
-Figure~\ref{fig:pe} shows the partial evaluator~|P|.
-Its type |repr| literally expresses the type equation for |repr_pe| above.
-The function |abstr|
-extracts a future-stage code value from the result of
-partial evaluation.  Conversely, the function |pdyn| injects a
-code value into the |repr| type. As
-in~\S\ref{S:PE-problem}, we build dynamic terms alongside
-any static ones to avoid polymorphic lift.
-
-The static portion of the interpretation of |lam f| is |Some f|, 
-which just wraps the HOAS
-function |f|. The interpretation of |app ef ea| 
-checks to see if |ef| is such a wrapped
-HOAS function. If it is, we apply |f| to the
-concrete argument |ea|, giving us a chance to perform static
-computations (see the example below). If
-|ef| has only a dynamic part, we residualize.
-
-To illustrate how to add optimizations, we improve |add| (and |mul|,
-elided) to simplify the generated code using the monoid (and ring)
-structure of~|int|: not only is addition performed statically
-(using~|R|) when both operands are statically known, but it is
-eliminated when one operand is statically~$0$; similarly for
-multiplication by~$0$ or~$1$.  
-\ifshort
-Such optimizations can be quite effective
-\else
-Such algebraic simplifications are easy
-to abstract over the specific domain (such as monoid or ring) where they
-apply.  These simplifications and abstractions help a lot
-\fi
-in a large language with more base types and primitive operations.
-\ifshort\else
-Incidentally, the code actually contains a more general implementation
-mechanism for such features, inspired in part by previous work in
-generative linear algebra~\citep{CaretteKiselyov05}.
-\fi
-
-Any partial evaluator must decide how much to unfold recursion
-statically: unfolding too little can degrade the residual code, whereas
-unfolding too much risks nontermination.  Our partial evaluator is no
-exception, because our object language includes |fix|.  The code in
-Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
-way'', that is, whenever the 
-argument is static, we unfold |fix| rather than residualize it.
-In the accompanying source code is a conservative
-alternative |P.fix| that unfolds recursion only once, then residualizes.
-Many sophisticated approaches have been developed to decide how much to unfold
-\citep{jones-partial}, but this issue is orthogonal to our presentation.
-\ifshort\else
-A separate concern in our treatment of |fix| is possible code bloat in
-the residual program, which calls for let-insertion
-\citep{SwadiTahaKiselyovPasalic2006}.
-\fi
-
-Given this implementation of~|P|, our running example
-\ifshort
-\texttt{let module E\,=\,EX(P) in E.test1 ()}
-\else
-\begin{code}
-let module E = EX(P) in E.test1 ()
-\end{code}
-\fi
-evaluates to
-\ifshort
-|{P.st = Some true; P.dy = .<true>.}|
-\else
-\begin{code}
-{P.st = Some true; P.dy = .<true>.}
-\end{code}
-\fi
-of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{S:compiler},
-a $\beta$-reduction has been statically performed to yield |true|.  More
-interestingly, whereas |testpowfix7| compiles to a code value with many
-$\beta$-redexes in~\S\ref{S:compiler}, the partial evaluation
-\ifshort
-\texttt{let module E = EX(P) in E.testpowfix7}
-\else
-\begin{code}
-let module E = EX(P) in E.testpowfix7
-\end{code}
-\fi
-gives the desired result
-\ifshort\vspace*{-0.7em}\fi
-\begin{code}
-{P.st = Some <fun>;
- P.dy = .<fun x -> x * (x * (x * (x * (x * (x * x)))))>.}
-\end{code}
-\ifshort\vspace*{-0.7em}\fi
-
-\noindent All pattern\hyp
-matching in~|P| is \emph{syntactically} exhaustive, so it is patent to the
-metalanguage implementation that |P| never gets stuck.  Further, all
-pattern\hyp matching occurs
-during partial evaluation, only to check if a value is known statically,
-never what type it has.  In other words, our partial evaluator tags
-phases (with |Some| and |None|) but not object types.
-
+\input{tagless-final-aplas}
 %------------------------------------------------------------------------
@@
 analysis.  To avoid needing polymorphic lift, we turn to
-Asai's technique \citep{asai-binding-time,sumii-hybrid}:
-build a dynamic term alongside every static term.
+Asai's technique \citep{asai-binding-time,sumii-hybrid} of
+building a dynamic term alongside every static term.
 
@@
 the metalanguage---call-by-value (CBV) in OCaml, call-by-name (CBN) in
-Haskell.  To represent a CBN object language in a CBV metalanguage,
+Haskell.\footnote{To be more precise, most Haskell implementations
+use call-by-need, which is observationally equivalent to call-by-name
+because sharing is not observable \citep{ariola-call-by-need-popl}.}
+To represent a CBN object language in a CBV metalanguage,
 \citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}
@@
 \end{code}
-In HOAS, we write $\lapp{e_1}{x}e_2$ as |lapp e1 (fun x -> e2)|;
+In HOAS\@, we write the term ``$\lapp{e_1}{x}e_2$'' as |lapp e1 (fun x -> e2)|;
 the type of |lapp| is that of function application with
@@
 as the accumulator.
+Because the |SymSI| signature extends |Symantics|, any encoding of
+a term in the pure object language (that is, any functor that takes
+a |Symantics| module as argument) can also be used as a term in the
+extended object language (for example, applied to an implementation of
+|SymSI|).
 
@@
 At the term level, we also avoid converting between static and dynamic
-terms by building them in parallel, using \citearound{'s
-method}\citet{asai-binding-time}.
+terms by building them in parallel, as \citet{asai-binding-time} does
+in his partial evaluator.
 This method type-checks in Hindley-Milner once we
@@
 
-\Citet{sumii-hybrid} also use Asai's method,
+\Citets{JFP-Mogensen} self-reducer for the untyped
+$\lambda$-calculus preceded Asai's partial evaluator in constructing
+static and dynamic terms in parallel.  However, this self-reducer
+constructs a static term for every object term, even a bound variable,
+so (in terms of Fig.~\ref{fig:pe}) it moves some work from |app|
+to~|pdyn|.  In contrast, we follow Asai and construct the static term
+optionally, so as to perform type-directed lifting without requiring the
+encoding of an object term to juggle explicit type indices.
+
+\Citet{sumii-hybrid} also use Asai's construction,
 to combine online

### 2008-06-12 SCP2008/paper/metamonads.tex
Lots of work, especially complete rewriting of Sec on lifted domains

@@
 
-\usepackage{fancyvrb}
+\usepackage[compact]{fancyvrb1}
 \DefineShortVerb{\|}
-\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\small}
-\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\ourmathindent,commandchars=\\\{\},fontsize=\scriptsize}
+\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\ourmathindent,fontsize=\small}
+\DefineVerbatimEnvironment{code2}{Verbatim}{xleftmargin=\ourmathindent,fontsize=\scriptsize}
 \newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
@@
 
-\subsection{Aspects}
+\subsection{Aspects}\label{sec:aspects}
 
@@
 
-\subsection{Organizing aspects}
+\subsection{Organizing aspects}\label{sec:aspect-groups}
 
@@
     \item \textbf{Tracking}, of determinant, rank, or pivot, etc.
-    \item \textbf{Staging}: whether the result is the program or a
+    \item \textbf{Interpretation}: whether the result is the program or a
       generator that will produce the program.
@@
 
-\begin{code}
+\begin{code}[commandchars=\\\{\}]
 let one = .<1>. and plus x y = .<.~x + .~y>.
@@
 We may attempt to eliminate subexpressions as follows: 
-\begin{code}
+\begin{code}[commandchars=\\\{\}]
 let param_code1' plus one =
@@
 
-\begin{code}
+\begin{code}[commandchars=\\\{\}]
 let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
@@
 The difference between them is apparent from the following example:
-\begin{code}
+\begin{code}[commandchars=\\\{\}]
 let gen a i = ifM' (ret .<(.~i) >= 0>.) 
@@
 
-Various aspects of our generator will need to keep a state during
-code generation (see \S\ref{sec:det} for an example).
+Various aspects of our generator need to keep a state during
+code generation (e.g., the name of the variable for the sign of
+the determinant, \S\ref{sec:det}).
 
 The simplest method of keeping the state of the generator is by using
-mutable variables, private to each module (aspect). That would however make our
-aspects stateful. Although we are generating imperative code, we would like to
-keep our generators stateless and purely functional, for ease of comprehension
-and reasoning.  Our main
-program may include several generators referring to one particular
-aspect -- which may be present in one shared instance or in
-several. That is of no concern if the module is
-stateless, but with stateful modules, the issues of aliasing or separate
-instantiation are a source of very subtle problems.
+mutable variables, private to each module (aspect). That would however
+make our aspects stateful. Although we are generating imperative code,
+we would like to keep our generators stateless and purely functional,
+for ease of comprehension and reasoning.  Our main program may include
+several generators referring to one particular aspect -- which may be
+present in one shared instance or in several. That is of no concern if
+the module is stateless, but with stateful modules, the issues of
+aliasing or separate instantiation are a source of very subtle
+problems.
 
@@
 lists~\cite{mlton-proplist}.  A property list also represents an
-extensible object but by its dual, namely a list of polymorphic variants.  The
+extensible object but by its dual, namely a list of fields.  The
 initial object is just the empty list.  Unfortunately, we cannot apply
-MLton's approach litterally to our case.  MetaOCaml's environment classifiers
-require type universes parameterized by classifiers. The MLton
+MLton's approach literally to our case. The MLton
 approach uses generativity of exceptions or reference cells to
-generate property names, i.e., fields of an extensible record. Using this
-technique would make our aspects stateful modules, something we do not
-want, as we wish our generator to be purely functional [even though it is
-generating imperative code].
-Fortunately, OCaml offers a way to build open unions with
+generate property names, i.e., fields of an extensible record. This
+technique would make our aspects stateful modules, which we described
+earlier as undesirable. MLton's approach to generating field
+names also makes it difficult to store in the fields MetaOCaml's code
+values, whose type contains a generic type variable, the environment
+classifier. 
+Fortunately, OCaml lets us build open unions with
 polymorphic variants. Each variant is identified by a manifest tag,
@@
 elements laid out in row-major format (C-style). The code below is
-hand-written, implementing the typical pseudo-code given in textbooks on
-Numerical Analysis \oleg{I used wikipedia; need a better references}.
-
-In OCaml, the matrix is represented by the value of the type
-|'a container2dfromvector|:
+a hand-written implementation of the typical pseudo-code in
+Numerical Analysis textbooks \oleg{I used wikipedia; need a better references}.
+
+In OCaml, the matrix is represented by the value of the following type:
 \begin{code}
@@
 where |a| is the array component of the record |container2dfromvector|.
-The figure defines three auxiliary functions, typically used in the
+The above code defines three auxiliary functions, typically used in the
 textbook pseudo-code: |swap a i j| swaps two elements; 
@@
 the |(n,m)|-matrix |a|. The function 
-|swap_rows a (n,m) (r,c) i| swaps the row |r| with the row |i| in 
+|swap_rows| |a| |(n,m)| |(r,c) i| swaps the row |r| with the row |i| in 
 the non-yet examined portion of the matrix, 
@@
 over the |(r,c)-(n,m)| block of the matrix containing not-yet
-processed elements. We implement the \emph{fraction-free} algorithm:
+processed elements. We implement the \emph{fraction-free}
+algorithm \oleg{Citation needed!}:
 the division operation in line XXX assuredly divides two integers
-without remainder. The fraction-free algorithm requires accumulation
-of the determinant \oleg{Citation needed!}.
+with no remainder, which requires the accumulation
+of the determinant .
 
@@
 
-\jacques{next paragraph will need rewritten when this section has
-finally taken shape}
-In this section, we will go through various aspects of one algorithm
-and explain how they are implemented.  In particular, and generally in
-order of increasing complexity, we will deal with abstracting the
-domain, the matrix representation, the code itself, determinant, etc.
-Then we put it all together (or in a new section?).
+In this section, we use the the sample code in
+Fig.~\ref{fig:GE-algorithm} to identify (some of the) aspects
+discussed in \S\ref{sec:aspects}. We abstract these aspects and
+describe their implementation as modules of the code generator.  We
+then present the generic Gaussian Elimination algorithm -- what is
+left after all the aspects are abstracted away. Finally, we show an
+instantiation of the generic generator, whose execution yields code
+for the algorithm of Fig.~\ref{fig:GE-algorithm}, this time
+automatically generated rather than manually written.
+
+We describe here the following aspects, in the general
+order of increasing complexity: domain (of the group `Abstract
+domain', see \S\ref{sec:aspect-groups}), code generation combinators
+(of the group `Interpretation'), matrix representation (of the group
+`Concrete representation'), determinant (the group `Tracking'), and
+output (the group `Interface').
 
@@
 
-We start by abstracting out domains. \oleg{Add: clearly the basic
-  structure of the algorithm, Fig.~\ref{fig:GE-algorithm} remains the
-  same for integer, float, etc. matrices.} We have already seen the
-simplest case of domain abstraction in |param_code1|, which took
-code-generators such as |plus| and |one| as arguments.  We need far
-more than two parameters: our domains should include $0$, $1$, $+$,
-$*$, (unary and binary) $-$, at least \emph{exact} division,
-normalization, and potentially a relative size measure. We could group
-these parameters in tuples or records.  Instead, we use OCaml
-\emph{structures} (i.e., modules) so we can take advantage of
-extensibility, type abstraction and constraints, and especially
-parameterized structures (\emph{functors}).  We define the type, the
-signature |DOMAIN|, which different domains must satisfy:
+Clearly the basic structure of the algorithm,
+Fig.~\ref{fig:GE-algorithm}, remains the same for integer, float,
+etc. matrices and so can be abstracted over the numeric domain.  We
+have already seen the simplest case of domain abstraction in
+|param_code1|, which took code-generators such as |plus| and |one| as
+arguments.  We need far more than two parameters: our domains should
+include $0$, $1$, $+$, $*$, (unary and binary) $-$, at least
+\emph{exact} division, normalization, and potentially a relative size
+measure. We could group these parameters in tuples or records.
+Instead, we use OCaml \emph{structures} (i.e., modules) so we can take
+advantage of extensibility, type abstraction and constraints, and
+especially parameterized structures (\emph{functors}).  We define the
+type, the signature |DOMAIN|, which different domains must satisfy:
 
@@
 
-module IntegerDomain : DOMAIN = struct
+module IntegerDomain : DOMAIN with type v = int  = struct
   type v = int
@@
 compiler verify that the defined structure is indeed of a type |DOMAIN|. 
-\jacques{In fact, it seems that this annotation cannot be put there
-because that will make type v abstract, which will result in compilation
-errors for the "lifted" domain which needs to have this type be
-available.}
-\oleg{well, we can say IntegerDomain : DOMAIN with type v = int}
 The annotation may be omitted (see |ZpMake| below), in which
@@
 characteristic:
-\begin{code}
+\begin{code}[commandchars=\\\{\}]
 module ZpMake(P:sig val p:int end) = struct
@@
 \subsection{Abstract code and lifted domains}\label{sec:lifteddomains}
-\jacques{I don't think I really want to delete the first half of this
-subsection.  I guess we can move that into section 3?  We need to be
-careful that section 3 does not become 1/2 the paper!  We might even
-consider splitting section 3 into 2 sections?}
-We will be using MetaOCaml operations to build code expressions, values
-of type |('a,'v) code|. However, we wish to make our generators
-abstract over particular code representations. Therefore, when writing
-the LU generator (\S\ref{sec:det} and below), instead of
-MetaOCaml brackets and escapes we will be using the code combinators
-described below. These combinators are functions that produce and
-combine values of the type of \emph{abstract code}.
-
-This encapsulation of staging annotations into code combinators lets
-us easily change the output form of our generated LU programs.  For example,
-we may use single-stage thunks rather than MetaOCaml staged |('a,'v) code|
-values, for benchmarking and regression tests purposes. We could also use
-code combinators producing C or Fortran code.
-
-Our code combinator library includes operations to create, dereference
-and mutate reference cells, compare values, and manipulate values used
-for indexing arrays. The following is the implementation of sample
-combinators, in the case of the abstract code being MetaOCaml code. 
-We basically need to ``lift'' most the underlying programming language
-into combinators.
-\begin{code}
-let lift x = .< x >.
-
-let cunit = .< () >.  and unitL = fun s k -> k s .< () >.
-
-let liftRef x = .< ref .~x >.  and   liftGet x = .< ! .~x >. 
-let liftPair x = (.< fst .~x >., .< snd .~x >.)
-
-module Logic = struct
-  let notL a        = .< not .~a >.
-  let equalL a b    = .< .~a = .~ b >.
-  let notequalL a b = .< .~a <> .~ b >.
-  let andL a b     = .< .~a && .~b >. 
-end
-
-module Idx = struct
-  let zero = .< 0 >. and one = .< 1 >. and minusone = .< -1 >.
-  let succ a = .< .~a + 1 >.  and  pred a = .< .~a - 1 >.
-  let less a b = .< .~a < .~b >.
-  let uminus a = .< - .~a >.  and  add a b = .< .~a + .~b >.
-end
-
-let update a f = let b = f (liftGet a) in .< .~a := .~b >.
-let assign a b = .< .~a := .~b >.
-let apply  f x = .< .~f .~x >.
-let updateM a f = ret (update a f)
-let assignM a b = ret (assign a b)
-let applyM  f x = ret (apply f x)
-\end{code}
-
-\noindent  The types above are
-generally lifted in two ways: once from the value domain |v| to the code
-domain |('a,'v) code|, and once more from values to monadic computations
-|('p,'v) monad|.
-
-We now convert the domains of values from the previous section into
-lifted domains, of abstract code values. We define the signature
-|DOMAINL|:
-\begin{code}
-module S(T: sig type ('a, 'b) rep  end) = struct
-open T
+
+In our sample GE code, Fig.~\ref{fig:GE-algorithm}, the operation |not|
+|(curr == 0)|, line 46, compares two integers (or, generally, two
+domain elements); the operation |det_magn := a_rc|, line 56, assigns
+the domain element to the corresponding reference cell; the operation
+|r := !r + 1| on the next line increments the rank, the cardinal
+number. One can easily imagine a different interpretation of the same
+program, where |not (curr == 0)| \emph{generates} code to compare two
+domain elements, |det_magn := a_rc| and |r := !r + 1| generate code
+for the assignment and the in-place increment. The structure of the GE
+algorithm is clearly invariant upon this change in
+interpretation. This lets us abstract the algorithm over the
+interpretation of basic operations, so that the same GE code, given
+different concrete interpretations, can LU-factorize a given matrix,
+can generate LU-factorization programs in OCaml, C or Fortran, or can
+pretty-print the factorization procedure.
+
+The interpretation aspect is too implemented as an OCaml module.  The
+aspect is quite large -- there are many basic operations to abstract
+over -- and so the module is structured into several
+sub-modules. First we introduce an abstract type |('a,'b) rep| which
+describes what sort of objects the interpretation may produce (e.g.,
+strings with C code, AST, etc). The type has two parameters: the
+second specifies the type of the object, and the first is the
+`placeholder' for all other information that may need to be tracked
+about the object in a particular interpretation. The interpretation as
+MetaOCaml code values uses the first parameter of |rep| to track the
+environment classifier.
+\oleg{I'm describing in this section coderep.mli and one of its
+  instances, code.ml. The only change is that I use ('a,'b) rep rather
+  than ('a,'b) abstract, for uniformity and clarity. Why did we even
+  need to introduce this additional distinction, `rep' vs
+  `abstract'. Also, the name `abstract' is a bit too
+  general. Anyway, the explanation does not have to follow the code to
+  the letter, as Reviewer 2 has suggested.}
+
+The first sub-module of the interpretation aspect is the numeric
+domain. The signature |DOMAIN| of \S\ref{sec:domains} defined the set
+of operations on numeric objects of some type |v|. We now
+generalize, or `lift', |DOMAIN| into |DOMAINL| so we can likewise
+operate on interpretations of numeric objects, of the type
+|('a,v) rep|:
+\begin{code}
 module type DOMAINL = sig
@@
 \end{code}
-(part of a larger module |Domains_sig|) that is parameterised by
-the type of abstract code, |('a, 'b) rep|. The line |include DOMAIN|
-says that lifted domains include all members of non-lifted domains,
-including the initializing expressions with preflight checks.
+The line |include DOMAIN| says that lifted domains include all members
+of non-lifted domains, specifically including the initializing
+expressions with preflight checks.
+
+An interpretation also needs to specify how to compare objects
+and to manipulate objects representing integers, typically used as
+indices of matrix elements. We group the operations into the
+following two structures:
+\begin{code}
+module Logic : sig
+  val notL      : ('a, bool) rep -> ('a, bool) rep
+  val equalL    : ('a, 'b) rep   -> ('a, 'b) rep   -> ('a, bool) rep
+  val notequalL : ('a, 'b) rep   -> ('a, 'b) rep   -> ('a, bool) rep
+  val andL      : ('a, bool) rep -> ('a, bool) rep -> ('a, bool) rep
+end
+module Idx : sig
+  val zero : ('a, int) rep
+  val one  : ('a, int) rep
+  val minusone : ('a, int) rep
+  val succ : ('a, int) rep -> ('a, int) rep
+  val pred : ('a, int) rep -> ('a, int) rep
+  val less : ('a, 'b) rep  -> ('a, 'b) rep -> ('a, bool) rep
+  val uminus : ('a, int) rep -> ('a, int) rep
+  val add : ('a, int) rep   -> ('a, int) rep -> ('a, int) rep
+  val minusoneL : 'a -> ('a -> ('b, int) rep -> 'c) -> 'c
+end
+\end{code}
+
+As we argued in \S\ref{CPS}, we will be writing our generic GE
+algorithm in the monadic style. For convenience we define the
+following type synonyms for our |('p,'v) monad| of Fig.\ref{ourmonad}:
+The first is used when we are always generating (abstract) code, while
+the second helps when we might generate (abstract) code.
+\begin{code}
+type ('pc,'v) cmonad = 
+  ('p,('a,'v) rep) monad
+    constraint
+      'p = <state : 's list; answer : ('a,'w) rep>
+    constraint
+      'pc = <classif : 'a; answer : 'w; state : 's; ..>
+
+type ('pc,'v) omonad = 
+  ('p,('a,'v) rep option) monad
+    constraint
+      'p = <state : 's list; answer : ('a,'w) rep>
+    constraint
+      'pc = <classif : 'a; answer : 'w; state : 's; ..>
+\end{code}
+
+Finally the interpretation aspect must specify the very basic
+operations: injection of literal values, function
+applications, statement sequencing, conditional, loops, creating,
+dereferencing and assigining to reference cells, etc:
+\begin{code}
+val lift  : 'b -> ('a, 'b) rep
+val unitL : ('pc,unit) cmonad
+
+val apply  : ('a, 'b -> 'c) rep -> ('a, 'b) rep -> ('a, 'c) rep
+val applyM : ('a, 'b -> 'c) rep -> ('a, 'b) rep -> 
+             (<classif: 'a; ..>,'c) monad
+
+val seqM :
+   (<classif: 'a; state: 's; answer: 'b; ..>,'b) cmonad ->
+   (<classif: 'a; state: 's; answer: 'c; ..>,'c) cmonad ->
+   (<classif: 'a; state: 's; ..>,'c) cmonad
+val optSeqM : 
+   (<classif: 'a; state: 's; answer: 'b; ..>,'b) cmonad ->
+   (<classif: 'a; state: 's; answer: 'b; ..>,'b) cmonad option ->
+   (<classif: 'a; state: 's; ..>,'c) cmonad
+
+val ifM : ('a, bool) rep ->
+   (<classif: 'a; state: 's; answer: 'b; ..>,'b) cmonad ->
+   (<classif: 'a; state: 's; answer: 'b; ..>,'b) cmonad ->
+   (<classif: 'a; state: 's; ..>,'b) cmonad
+
+val liftRef : ('a, 'b) rep -> ('a, 'b ref) rep
+val liftGet : ('a, 'b ref) rep -> ('a, 'b) rep
+val assign  : ('a, 'b ref) rep -> ('a, 'b) rep -> ('a, unit) rep
+val assignM : ('a, 'b ref) rep -> ('a, 'b) rep -> 
+              (<classif: 'a; ..>,unit) cmonad
+\end{code}
+
+The `pure' operations of the interpretation produce interpretation
+objects (of the type |('a,'b)| rep) as the result. We can trivially
+`lift' these operations into the monad by composing them with |ret| of
+the monad. Some other operations, like |seqM| and |ifM| are present
+only in the monadic form: these are \emph{control} operations.
+
+We provide two concrete instances of the interpretation aspect: one
+uses thunks (for benchmarking and regression tests purposes) and the
+other uses MetaOCaml's code values |('a,'v) code| as the realization
+of |('a,'v) rep|. We could also use interpretations producing C or 
+Fortran code. The following is a sample implementation of the
+interpretation aspect for MetaOCaml code values:
+\begin{code}
+let lift x = .< x >.   and  unitL = fun s k -> k s .< () >.
+
+let liftRef x = .< ref .~x >.  and   liftGet x = .< ! .~x >. 
+let liftPair x = (.< fst .~x >., .< snd .~x >.)
+
+module Logic = struct
+  let notL a        = .< not .~a >.
+  let equalL a b    = .< .~a = .~ b >.
+  let notequalL a b = .< .~a <> .~ b >.
+  let andL a b      = .< .~a && .~b >. 
+end
+
+module Idx = struct
+  let zero = .< 0 >. and one = .< 1 >. and minusone = .< -1 >.
+  let succ a = .< .~a + 1 >.  and  pred a = .< .~a - 1 >.
+  let less a b = .< .~a < .~b >.
+  let uminus a = .< - .~a >.  and  add a b = .< .~a + .~b >.
+end
+
+let update a f = let b = f (liftGet a) in .< .~a := .~b >.
+let assign a b = .< .~a := .~b >.
+let apply  f x = .< .~f .~x >.
+let updateM a f = ret (update a f)
+let assignM a b = ret (assign a b)
+let applyM  f x = ret (apply f x)
+\end{code}
 The following is a particular instance of |DOMAINL|, the lifted
-version of |IntegerDomain| of the previous section, in the particular
-case of abstract code being MetaOCaml code values |('a, 'b) code|:
-\begin{code}
-module T = Domains_sig.S(struct type ('a, 'b) rep = ('a, 'b) code end)
-open T
+version of |IntegerDomain| of the previous section, again
+for the MetaOCaml code value interpretation.
+\begin{code}
 module IntegerDomainL = struct
@@
 
-In fact, our implementation requires another level of indirection, and
-in fact reads
-\begin{code}
-module Make(CODE: Coderep.T) = struct
-  module D = Domains_sig.S(
-      struct type ('a, 'v) rep = ('a,'v) CODE.abstract 
-    end)
-open D
-open CODE
-...
-\end{code}
-where |CODE| is a module which implements a type |abstract| and various
-facilities over it, like the modules |Logic| and |Idx| mentioned above.
-For simplicity, we open |CODE| to allow access to the actual type |abstract|.
-
-To further trim down on the size of the signatures for various modules,
-we introduce two variants of |monad|, specialized for our purpuses.  The
-first is used when we are always generating (abstract) code, while the second
-helps when we might generate (abstract) code.
-\begin{code}
-type ('pc,'v) cmonad = 
-  ('p,('a,'v) abstract) monad
-    constraint
-      'p = <state : 's list; answer : ('a,'w) abstract>
-    constraint
-      'pc = <classif : 'a; answer : 'w; state : 's; ..>
-
-type ('pc,'v) omonad = 
-  ('p,('a,'v) abstract option) monad
-    constraint
-      'p = <state : 's list; answer : ('a,'w) abstract>
-    constraint
-      'pc = <classif : 'a; answer : 'w; state : 's; ..>
-\end{code}
+We will consider the interpretation aspect ``open'' and so we can
+refer to all operations such as |apply|, |seqM|, etc, without further
+qualification.
+
 \subsection{Containers}
@@
 For our purposes, a container is an abstraction of an $n$-dimensional
-vector space, which we here specialize for $n=1,2$.  The $2$ dimensional
+vector space, which we here specialize for $n=1,2$.  The $2$-dimensional
 case is our main interest, and its signature contains many functions particular
@@
 burdening the users of containers with efficiency details (see
-\S\ref{GE-algorithm} for an example of such an optimiation).  Note that
+\S\ref{GE-algorithm} for an example of such an optimization).  Note that
 this module type is also abstract over the actual representation 
@@
 row-major order -- the container used in Fig.\ref{fig:GE-algorithm}.
-\begin{code}
+\begin{code}[commandchars=\\\{\}]
 module GenericVectorContainer(Dom:DOMAINL) =
@@
 %
-The accompanying code\cite{metamonadsURL} includes containers whose
+The accompanying code \cite{metamonadsURL} includes containers whose
 elements are stored in a 1D array in a column-wise (Fortran-like)
@@
   val zero_sign : unit -> ('b,unit) lm
-  val acc_magn  : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
+  val acc_magn  : ('a,C.Dom.v) rep -> (<classif : 'a; ..>,unit) lm
   val get_magn  : unit -> ('b,tdet) lm
-  val set_magn  : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
+  val set_magn  : ('a,C.Dom.v) rep -> (<classif : 'a; ..>,unit) lm
   val fin       : unit -> ('b,C.Dom.v) lm
@@
   type tdet = v ref
-  type 'a lstate = ('a,int ref) abstract * ('a,tdet) abstract
+  type 'a lstate = ('a,int ref) rep * ('a,tdet) rep
   let decl () = perform
@@
           (ret (uminusL (liftGet magn))))
-  \dots
+  ...
 end
@@
 \begin{code}
-type 'a wmatrix = \{matrix: 'a C.vc; numrow: ('a,int) abstract; 
-                   numcol: ('a,int) abstract\}
+type 'a wmatrix = {matrix: 'a C.vc; numrow: ('a,int) rep; 
+                   numcol: ('a,int) rep}
 \end{code}
@@
       _ <-- IF.L.decl (if wants_pack then b else C.identity rmar m);
-      let mat = \{matrix=b; numrow=n; numcol=m\} in
+      let mat = {matrix=b; numrow=n; numcol=m} in
       ret (mat, r, c, rmar)
@@
        cc <-- retN (liftGet c);
-       let cp  = \{rowpos=rr; colpos=cc\} in
+       let cp  = {rowpos=rr; colpos=cc} in
        let module Pivot = F.PivotF(F.Det)(IF.P) in
@@
        seqM (matchM pivot (fun pv -> 
-              seqM (zerobelow mat \{p=cp; curval=pv\} )
+              seqM (zerobelow mat {p=cp; curval=pv} )
                    (IF.R.succ ()) )
@@
    let t_3 = (ref 0) in
-   let t_4 = (a_1) \{arr = (Array.copy a_1.arr)\} in
+   let t_4 = (a_1) {arr = (Array.copy a_1.arr)} in
    let t_5 = a_1.m in  (* magnitude of det *)

### 2008-06-12 SCP2008/paper/metamonads.tex
More writing and explanation, up to Section Output

@@
 For our purposes, a container is an abstraction of an $n$-dimensional
-vector space, which we here specialize for $n=1,2$.  The $2$-dimensional
-case is our main interest, and its signature contains many functions particular
-to $n=2$.  For examples, we have rows and columns and operations specialized
-for them.  A container explicitly abstracts the underlying representation 
-of the data-structure, while offering an interface which is better-suited
-to linear algebra.
-
-\oleg{Say that CONTAINER2D is an abstraction of a particular
-  container, flat vector, used in the algorithm in
-  \S\ref{GE-algorithm}.}
+vector space, which we here specialize for $n=1,2$.  The
+$2$-dimensional case is our main interest, and its signature contains
+many functions particular to $n=2$.  For examples, we have rows and
+columns and operations specialized for them.  A container explicitly
+abstracts the underlying representation of the data-structure, while
+offering an interface which is better-suited to linear algebra. In
+particular, a container is an abstraction of the flat vector
+|container2dfromvector| of our sample algorithm in
+\S\ref{GE-algorithm}.
+
 The signature |CONTAINER2D| below specifies that a container must provide
@@
 
+The generic GE algorithm described later is a functor over the
+container.  Given a particular container aspect, the functor yields a
+module containing various algorithmic aspects for the user to choose,
+as well as the main GE driver. The given container instance is
+available to all these aspects under the name of |C| (so that the type
+of the container can be referred to as |C.contr| and the dimensions
+can be obtained using |C.dim1| and |C.dim2|). The numeric domain, which
+is part of the container, can be referred to as |D.Dom| and the type
+of the domain elements is |C.Dom.v|. We shall see many such references
+as we describe particular algorithmic aspects below.
+
 \subsection{Determinant aspect}
@@
 
-To track the determinant, we should be able to generate code
-for: defining variables used for tracking (|decl|),
-updating the sign (|upd_sign|) or the absolute
-value of the determinant, and converting the tracking state
-to the final determinant value of type |outdet| (see
-Fig.~\ref{fig:GE-algorithm} for an example).
-LU of a
-floating-point matrix with no determinant tracking uses the
-instantiation of |DETERMINANT| (see definition below) where |outdet| is |unit|
-and all the
-functions of that module generate no code. For integer matrices, and
-in general whenever the matrix elements do not form a field, we
-have to track some aspects of the determinant, even if we do not output
-it, as the fraction-free update needs those aspects. The determinant tracking
-aspect is complex because tracking variables, if any, are to be declared at the
-beginning of the code (see lines 22-23 (XXX) of
-Fig.~\ref{fig:GE-algorithm}); the sign
-of the determinant has to be updated on each row or column
-permutation (lines 32 and 37 of the figure XXX);
-the value of the determinant should be updated per each
-pivoting (line 56). We use |lstate| to pass the tracking state, e.g., a piece of
-code for the value of the type |Dom.v ref|, among
-various determinant-tracking functions. The |lstate| is a part of the
-overall monadic state. 
-Other aspects follow a similar outline.
-
+The determinant aspect is one of several tracking and strategic
+aspects. The main function of the algorithm invokes the functions of
+these aspects at some interesting points during the computation (e.g.,
+pivot is required, two rows have to be permuted, the final answer to the
+user has to be built). The instance of the aspect may do something (or
+nothing) at each point, for example, find pivot according to the
+pivoting strategy (such as row pivoting, any non-zero element), update
+the current value of the determinant, etc.
+
+Our sample Gaussian Elimination algorithm, Fig.\ref{fig:GE-algorithm},
+demonstrates that tracking determinant is complex: first we define the
+variables |det_sign| and |det_magn| used for tracking (lines 22 and
+23), then we have to flip the sign when swapping two rows or two
+columns or when the matrix found to be singular (lines 32, 37, 59).
+The value of the determinant should be updated per each pivoting (line
+56). Finally, we convert the tracking state to the resulting
+determinant value (lines 64-66). We observe these lines are not
+contiguous: the determinant aspect is consulted at several separate
+places in the GE algorithm, and the aspect is supposed to keep
+state. As other aspects, the determinant aspect is a module of the
+following signature:
 \begin{code}
@@
 \end{code}
-
-We have two instances of the above. The first one is used to specify
-no determinant tracking, and so no code is
-generated. The methods just return without executing any
-code-generation actions.
+The type |'a lstate| denotes the state being kept as part of the
+overall monadic state (tagged as |`TDet|). The signature specifies the
+operations of the aspect, such as |decl| for defining the tracking
+state, |zero_sign| to note singularity, |set_magn| to set the
+magnitude and |fin| to convert the tracking state to the resulting
+value, of  the type |C.Dom.v|, the type of the container
+elements. Other aspects follow a similar outline.
+
+We have two instances of |DETERMINANT|. The first corresponds to
+no determinant tracking, and so all the functions are dummy.
 \begin{code}
@@
 
-The second instance does track the determinant. The |decl| method
-generates two let-bindings for mutable variables tracking the
-magnitude and the sign of the determinant, and places the names of
-these variables in the monadic state. The |upd_sign| method, invoked
-from row- or column-swap generators, retrieves the name of the
-sign-accumulating variable and generates the code to update the
-sign. The |fin| method retrieves the names of both accumulators and
-generates code to compute the final, signed determinant value.
-The generated code closely corresponds to the lines dealing with 
-|det_magn| and |det_sign| in Fig.~\ref{fig:GE-algorithm}.
+The second instance does track the determinant. For integer matrices,
+and in general whenever the matrix elements do not form a field, 
+the fraction-free update requires tracking some facets of the
+determinant, even if we do not output it.
 \begin{code}
@@
 \end{code}
+The |decl| method generates two let-bindings for mutable variables
+tracking the magnitude and the sign of the determinant, and places the
+names of these variables in the monadic state. The |upd_sign| method,
+invoked from row- or column-swap generators, retrieves the name of the
+sign-accumulating variable and generates the code to update the
+sign. The |fin| method retrieves the names of both accumulators and
+generates code to compute the final, signed determinant value.  The
+generated code closely corresponds to the lines dealing with
+|det_magn| and |det_sign| in Fig.~\ref{fig:GE-algorithm}.
 
@@
 
-More interesting is the aspect of what to return
-from the LU algorithm.  One could create an algebraic data type (as
-was done in \cite{Carette06}) to encode the various choices: the
-matrix, the matrix and the rank, the matrix and the determinant, the
-matrix, rank and determinant, and so on. This is wholly unsatisfying
-as we know that for any single use, only one of the choices is ever
-possible, yet any routine which calls the generated code must deal
-with these unreachable options.  Instead we use a module type with an
+More interesting is the aspect of what to return from the LU
+algorithm.  One could create an algebraic data type (as was done in
+\cite{Carette06}) to encode the various choices: the matrix, the
+matrix and the rank, the matrix and the determinant, the matrix, rank
+and determinant, and so on. This is wholly unsatisfying as we know
+that for any single use, only one of the choices is ever possible, yet
+any routine which calls the generated code must deal with these
+unreachable options.  Instead we use a module type with an
 \emph{abstract} type |res| for the result type; different instances of
 the signature set the result type differently. Given below is this
-module type and one instantiation, which specifies the output of a LU
-algorithm as a 3-tuple |contr * Det.outdet * int| of the U-factor, the
-determinant and the rank.
+module type and one instantiation, |OutDetRank|, which specifies the
+output of a GE algorithm as a 3-tuple |contr * Det.outdet * int| of
+the U-factor, the determinant and the rank. That choice of output
+corresponds to our sample algorithm in Fig.~\ref{fig:GE-algorithm}.
 
@@
 pivot, fraction-free update, operate on non-augmented matrix and
-return the |U| factor, determinant and the rank. We run the monad
+return the |U| factor, determinant and the rank. This choice
+corresponds to our sample algorithm of Fig.\ref{fig:GE-algorithm}.
+We run the monad
 passing the initial state |[]| and obtain the code, which we can see
@@
 parameterized by the domain, container, pivoting policy (full, row,
-nonzero, no pivoting), update policy (with either `fraction-less'
-or full division), and the output specification. Some of the
-argument modules such as |PIVOT| are functors themselves (parameterized
-by the domain, the container, and the determinant functor). It is 
-worthwhile noting that we make sure of module subtyping -- |F| is of
-type |FEATURES| while |F.Output| is expecting an |OUTPUTDEP|, but
-|OUTPUTDEP| is a subtype of |FEATURES|.
+nonzero, no pivoting), update policy (with either `fraction-less' or
+full division), and the output specification. Some of the argument
+modules such as |PIVOT| are functors themselves (parameterized by the
+domain, the container, and the determinant functor).  We rely on
+module subtyping: For example, |F.Output| of the type |OUTPUT| is a
+functor requiring an argument structure of the signature
+|OUTPUTDEP|. The fact that the signature |FEATURES| contains all the
+fields of |OUTPUTDEP| and then some lets us pass |F| (of the type
+|FEATURES|) as an argument to instantiate |F.Output|.
 

### 2008-06-13 SCP2008/paper/metamonads.tex
Have to go through the main generation section

@@
 The first is used when we are always generating (abstract) code, while
-the second helps when we might generate (abstract) code.
-\begin{code}
-type ('pc,'v) cmonad = 
-  ('p,('a,'v) rep) monad
+the second helps when we might generate (abstract) code. The helper
+type synonym |('pc,'p,'a) cmonad_constraint| is effectively the
+abbreviation for a set of constraints imposed on its arguments.
+\begin{code}
+type ('pc,'p,'a) cmonad_constraint = unit
       constraint
-      'p = <state : 's list; answer : ('a,'w) rep>
+	  'p = <state : 's list; answer : ('a,'w) abstract>
       constraint
@@
 
-type ('pc,'v) omonad = 
-  ('p,('a,'v) rep option) monad
-    constraint
-      'p = <state : 's list; answer : ('a,'w) rep>
-    constraint
-      'pc = <classif : 'a; answer : 'w; state : 's; ..>
+type ('pc,'v) cmonad = ('p,('a,'v) abstract) monad
+      constraint _ = ('pc,'p,'a) cmonad_constraint
+
+type ('pc,'v) omonad = ('p,('a,'v) abstract option) monad
+      constraint _ = ('pc,'p,'a) cmonad_constraint 
 \end{code}
@@
 
-The signature |CONTAINER2D| below specifies that a container must provide
-functions |dim1| and |dim2| to extract the dimensions, functions |getL|
-to generate container getters, the cloning
-generator |copy| and functions that generate code for row and column
+The signature |CONTAINER2D| below specifies that a container must
+provide functions |dim1| and |dim2| to extract the dimensions,
+functions |getL| to generate container getters, the cloning generator
+|copy| and functions that generate code for row and column
 swapping. The inclusion of these functions in the signature of all
@@
 burdening the users of containers with efficiency details (see
-\S\ref{GE-algorithm} for an example of such an optimization).  Note that
-this module type is also abstract over the actual representation 
-|rep| of the generator data-structure.
-
+\S\ref{GE-algorithm} for an example of such an optimization).  We
+could have made the type |CONTAINER2D| a functor, parameterized over
+|DOMAINL|. The current design simplifies the signature of the generic
+GE algorithm at a cost of an extra instantiation step (The
+instantiation of the generic algorithm is described in
+\S\ref{sec:main-generation}. The expense is worthwhile: currently
+OCaml is quite slow when processing functors with complex signatures).
 \begin{code}
@@
   type 'a lstate
+  type 'pc pc_constraint = unit
+    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
   type ('pc,'v) lm = ('pc,'v) cmonad
-    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
+    constraint _  = 'pc pc_constraint
   type ('pc,'v) om = ('pc,'v) omonad
-    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
-  val decl      : unit -> ('b,unit) lm
+    constraint _  = 'pc pc_constraint
+  type 'pc nm = ('p,unit) monad
+    constraint _ = ('pc,'p,_) cmonad_constraint
+    constraint _ = 'pc pc_constraint
+  val decl : unit -> 'b nm (* no code is generated *)
   val upd_sign  : unit -> ('b,unit) om
   val zero_sign : unit -> ('b,unit) lm
-  val acc_magn  : ('a,C.Dom.v) rep -> (<classif : 'a; ..>,unit) lm
+  val acc_magn  : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
   val get_magn  : unit -> ('b,tdet) lm
-  val set_magn  : ('a,C.Dom.v) rep -> (<classif : 'a; ..>,unit) lm
+  val set_magn  : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
   val fin       : unit -> ('b,C.Dom.v) lm
@@
 overall monadic state (tagged as |`TDet|). The signature specifies the
-operations of the aspect, such as |decl| for defining the tracking
-state, |zero_sign| to note singularity, |set_magn| to set the
+operations of the aspect, such as |decl| for initializing the tracking
+state, |zero_sign| to note matrix singularity, |set_magn| to set the
 magnitude and |fin| to convert the tracking state to the resulting
@@
 elements. Other aspects follow a similar outline.
+\oleg{The following is perhaps too particular to mention in the
+  paper. AT least we can keep it in comments:
+  We stress the difference in return types of \texttt{decl},
+  \texttt{upd\_sign} and \texttt{zero\_sign} functions. The former says
+  that \texttt{decl} is an action which is executed only for its
+  side-effect. It yields no interpretation object (e.g., produces no
+  code value). The function \texttt{zero\_sign} always produces a code
+  value -- an expression such as assignment that has the type unit. 
+  The function
+  \texttt{upd\_sign} is the generator that may produce code, or may
+  not. The option type lets us avoid generating code such as 
+  \texttt{a := b; ()} with useless dummies.}
 
 We have two instances of |DETERMINANT|. The first corresponds to
-no determinant tracking, and so all the functions are dummy.
-\begin{code}
-module NoDet =
-  struct
+no determinant tracking, and so all functions are dummy.
+\begin{code}
+module NoDet = struct
   type tdet = C.Dom.v ref
   type 'a lstate = unit
-  let decl ()      = unitL
+  let decl ()      = ret ()
   let upd_sign ()  = ret None
@@
 \begin{code}
-module AbstractDet =
-  struct
+module AbstractDet = struct
   open C.Dom
@@
       sign <-- retN (liftRef Idx.one); (* track the sign: +1, 0, -1 *)
-      mo_extend ip (sign,magn);
-      unitL
+      mo_extend ip (sign,magn)
   let upd_sign () = perform
@@
 
+\subsection{Other aspects}
+
+For completeness, we briefly describe and show the signatures of the
+other aspects. The |RANK| aspect tracks rank; the aspect is a simpler
+version of the determinant tracking aspect.  |PIVOTKIND| describes the
+representation of the permutation matrix to track permutations
+executed in pivoting; |TRACKPIVOT| specifies if these permutations are
+tracked at all. |LOWER| tracks the |L| factor when GE algorithm is
+used for the in-place LU-decomposition. |PIVOT| is a higher-order
+functor abstracting over the pivoting algorithm (full pivoting, row
+pivoting, no pivoting, etc). If pivot is found, the aspect should swap
+rows and columns appropriately. The swapping may need to be tracked if
+we wish to return the permutation matrix. If the determinant is
+tracked, the swapping must update the sign of the determinant. The
+wapiing also affects the L factor being accumulated in-place. The
+|UPDATE| aspect abstracts over the algorithm for updating matrix
+elements during row-reductions: full division update or fraction-free
+update. Finally, the |INPUT| aspect is an interface aspect abstracting
+over the choice of augmented matrix.
+
+\begin{code2}
+module type RANK = sig
+  type 'a tag_lstate = 'a tag_lstate_
+  val decl   : unit -> ('b, int ref) lm
+  val succ   : unit -> ('b, unit) lm
+  val fin    : unit -> ('b, int) lm
+end
+
+module type PIVOTKIND = sig
+  type perm_rep
+  type 'a ira = ('a, int) rep
+  type 'a fra
+  type 'a pra = ('a, perm_rep) rep
+  val add    : 'a fra -> 'a pra -> 'a pra
+  val empty  : 'a ira -> 'a pra
+  val rowrep : 'a ira -> 'a ira -> 'a fra
+  val colrep : 'a ira -> 'a ira -> 'a fra
+end
+
+module type TRACKPIVOT = sig
+  type perm_rep
+  type 'a ira = ('a, int) rep
+  type 'a fra   type 'a pra   type 'a lstate
+  type 'pc pc_constraint = unit
+    constraint 'pc = <state : [> `TPivot of 'a lstate ]; classif : 'a; ..>
+  type ('pc,'v) lm = ('pc,'v) cmonad constraint _  = 'pc pc_constraint
+  type ('pc,'a) nm = ('p,unit) monad
+    constraint _ = ('pc,'p,'a) cmonad_constraint
+    constraint _ = 'pc pc_constraint
+  val rowrep : 'a ira -> 'a ira -> 'a fra
+  val colrep : 'a ira -> 'a ira -> 'a fra
+  val decl   : ('a, int) rep -> ('pc,'a) nm
+  val add : 'a fra -> (<classif : 'a; state : [> `TPivot of 'a lstate ]; ..>, unit) omonad
+  val fin    : unit -> ('b,perm_rep) lm
+end
+
+module type LOWER = sig
+  type 'a lstate = ('a, C.contr) rep
+  type ('pc,'v) lm = ('pc,'v) cmonad
+    constraint 'pc = <state : [> `TLower of 'a lstate ]; classif : 'a; ..>
+  val decl   : ('a, C.contr) rep -> (<classif : 'a; ..>, C.contr) lm
+  val updt   : 'a C.vc -> ('a,int) rep -> ('a,int) rep -> 'a C.vo -> 
+            'a C.Dom.vc -> (<classif : 'a;..>, unit) lm option
+  val fin    : unit -> ('a,  C.contr) lm
+  val wants_pack : bool
+end
+
+module type PIVOT = functor (D: DETERMINANT) -> functor (P: TRACKPIVOT) -> 
+          functor (L: LOWER) -> sig
+ val findpivot : 'a wmatrix -> 'a curpos -> 
+   (<classif : 'a;  state : [> `TDet of 'a D.lstate | `TPivot of 'a P.lstate ]; ..>, 
+    C.Dom.v option) cmonad
+end
+
+type update_kind = FractionFree | DivisionBased
+module type UPDATE = functor(D:DETERMINANT) -> sig
+  type 'a in_val = 'a C.Dom.vc
+  val update :  'a in_val -> 'a in_val -> 'a in_val -> 'a in_val -> 
+      ('a in_val -> ('a, unit) rep) -> ('a, C.Dom.v ref) rep -> 
+      (<classif : 'a; ..>, unit) cmonad
+  val update_det : 'a in_val -> (<classif : 'a; ..>,unit) D.lm
+  val upd_kind : update_kind
+end
+
+module type INPUT = sig
+  type inp
+  val get_input : ('a, inp) rep ->
+         (<classif : 'a; ..>, ('a, C.contr) rep * ('a, int) rep * bool) monad
+end 
+\end{code2}
+
 \subsection{Output}
 
-More interesting is the aspect of what to return from the LU
+More interesting is the aspect of what to return from the GE
 algorithm.  One could create an algebraic data type (as was done in
@@
   type res
-  val make_result : 'a wmatrix -> (<classif: 'a;\dots>,res) cmonad
+  val make_result : 'a wmatrix -> (<classif: 'a;...>,res) cmonad
 end
@@
 The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
-the preflight checks. As we saw in the previous section, 
-both instances of |DETERMINANT| contain a |fin ()| function to generate
-code representing the computed determinant. Only the |NoDet| does not
-track any determinant, and so |fin ()| raises an error. The code 
-|let _ = OD.Det.fin ()| in |OutDetRank| invokes this |fin| function, which will
-produce the monadic code generating action, or raise an error. We do
-not run the action at that time -- we only make sure there is an
-action to run. This is another preflight check, to rule out the
-semantic error where a user specifies that the determinant should be computed
-and returned, and yet specifies the |NoDet| aspect which tracks no
-determinant at all.
+the preflight checks. As we saw in the previous section, both
+instances of |DETERMINANT| contain a |fin ()| function to generate
+code representing the computed determinant. The instance |NoDet|
+however does no tracking, and so |fin ()| raises an error. The code
+|let _ = OD.Det.fin ()| in |OutDetRank| invokes this |fin| function,
+which will produce the monadic code generating action, or raise an
+error. We do not run the action at that time -- we only make sure
+there is an action to run. This is another preflight check, to rule
+out the semantic error where a user specifies that the determinant
+should be computed and returned, and yet specifies the |NoDet| aspect
+which tracks no determinant at all.
 
@@
 
-\subsection{Main generation}
+\subsection{Main generation}\label{sec:main-generation}
 We combine all user-selectable aspects in a record, which serves as a ``keyword

### 2008-06-14 SCP2008/paper/metamonads.tex
various small changes throughout the text

@@
 algorithm \oleg{Citation needed!}:
-the division operation in line XXX assuredly divides two integers
+the division operation in line 51 assuredly divides two integers
 with no remainder, which requires the accumulation
-of the determinant .
+of the determinant.
 
@@
 describe their implementation as modules of the code generator.  We
-then present the generic Gaussian Elimination algorithm -- what is
-left after all the aspects are abstracted away. Finally, we show an
-instantiation of the generic generator, whose execution yields code
-for the algorithm of Fig.~\ref{fig:GE-algorithm}, this time
-automatically generated rather than manually written.
+then present in \S\ref{sec:main-generation} the generic Gaussian
+Elimination algorithm -- what is left after all the aspects are
+abstracted away. Finally, we show an instantiation of the generic
+generator, whose execution yields code for the algorithm of
+Fig.~\ref{fig:GE-algorithm}, this time automatically generated rather
+than manually written.
 
@@
 
-\subsection{Abstract code and lifted domains}\label{sec:lifteddomains}
+\subsection{Abstracting interpretations}\label{sec:lifteddomains}
 
@@
 describes what sort of objects the interpretation may produce (e.g.,
-strings with C code, AST, etc). The type has two parameters: the
+AST, strings with C code, etc). The type has two parameters: the
 second specifies the type of the object, and the first is the
@@
 algorithm in the monadic style. For convenience we define the
-following type synonyms for our |('p,'v) monad| of Fig.\ref{ourmonad}:
-The first is used when we are always generating (abstract) code, while
-the second helps when we might generate (abstract) code. The helper
+following type synonyms for our |('p,'v) monad| of Fig.~\ref{ourmonad}:
+The first is used for a monadic action that always produces a useful
+value; the |omonad| synonym describes a monadic action that may
+produce an interpretation object. The helper
 type synonym |('pc,'p,'a) cmonad_constraint| is effectively the
@@
 type ('pc,'p,'a) cmonad_constraint = unit
-      constraint
-	  'p = <state : 's list; answer : ('a,'w) abstract>
-      constraint
-          'pc = <classif : 'a; answer : 'w; state : 's; ..>
-
-type ('pc,'v) cmonad = ('p,('a,'v) abstract) monad
+  constraint 'p = <state : 's list; answer : ('a,'w) rep>
+  constraint 'pc = <classif : 'a; answer : 'w; state : 's; ..>
+
+type ('pc,'v) cmonad = ('p,('a,'v) rep) monad
   constraint _ = ('pc,'p,'a) cmonad_constraint
 
-type ('pc,'v) omonad = ('p,('a,'v) abstract option) monad
+type ('pc,'v) omonad = ('p,('a,'v) rep option) monad
   constraint _ = ('pc,'p,'a) cmonad_constraint 
@@
 
-We will consider the interpretation aspect ``open'' and so we can
+We consider the interpretation aspect ``open'' and so we can
 refer to all operations such as |apply|, |seqM|, etc, without further
@@
 burdening the users of containers with efficiency details (see
-\S\ref{GE-algorithm} for an example of such an optimization).  We
-could have made the type |CONTAINER2D| a functor, parameterized over
-|DOMAINL|. The current design simplifies the signature of the generic
-GE algorithm at a cost of an extra instantiation step (The
-instantiation of the generic algorithm is described in
-\S\ref{sec:main-generation}. The expense is worthwhile: currently
-OCaml is quite slow when processing functors with complex signatures).
+\S\ref{GE-algorithm} for an example of such an optimization).
 \begin{code}
@@
 
-The type of our containers include the lifted domain |Dom| as one of
-the components. This is quite convenient since operations on containers
-are usually accompanied by the operations on the retrieved and stored
-values. Our containers are parametric over a |DOMAINL|, i.e., functors
-from a |DOMAINL| module to the actual implementation of a
-container. For example, the following functor defines a matrix
-container as a single array, with elements stored in the
-row-major order -- the container used in Fig.\ref{fig:GE-algorithm}.
+The type of our containers includes the lifted domain |Dom| as one of
+the components. This is quite convenient since operations on
+containers are usually accompanied by the operations on the retrieved
+and stored values. The particular instances of the containers are
+parametric over a |DOMAINL|, i.e., functors from a |DOMAINL| module to
+the actual implementation of a container.  For example, the following
+functor defines a matrix container as a single array, with elements
+stored in the row-major order -- the container used in
+Fig.~\ref{fig:GE-algorithm}.
 \begin{code}[commandchars=\\\{\}]
@@
   let dim1 x = .< (.~x).m >.      (* number of cols *)
-  let mapper g a = match g with
   \dots
 \end{code}
-%
-The accompanying code \cite{metamonadsURL} includes containers whose
-elements are stored in a 1D array in a column-wise (Fortran-like)
-modes, and container for a matrix represented as arrays of rows.
-
-The generic GE algorithm described later is a functor over the
-container.  Given a particular container aspect, the functor yields a
-module containing various algorithmic aspects for the user to choose,
-as well as the main GE driver. The given container instance is
-available to all these aspects under the name of |C| (so that the type
-of the container can be referred to as |C.contr| and the dimensions
-can be obtained using |C.dim1| and |C.dim2|). The numeric domain, which
-is part of the container, can be referred to as |D.Dom| and the type
-of the domain elements is |C.Dom.v|. We shall see many such references
-as we describe particular algorithmic aspects below.
+The accompanying code \cite{metamonadsURL} includes the container with
+elements stored in a 1D array in a column-wise (Fortran-like)
+mode, and the container for a matrix represented as an array of rows.
+
+We could have defined the type |CONTAINER2D| to be a functor with
+|DOMAINL| as an argument. The type |CONTAINER2D| is used in the
+signatures of other functors such as |GenLA| of
+\S\ref{sec:main-generation} below. If the type |CONTAINER2D| were a
+functor, |GenLA| would have been a higher-order functor, and we would
+have to pass to |GenLA| two arguments: the container functor and a
+|DOMAINL| to apply the container functor to. In the current design,
+the user first builds a particular container instance by applying the
+functor such as |GenericVectorContainer| to the desired domain. The
+user then passes this container instance to |GenLA| as a single
+argument. The current design simplifies module signatures at
+the expense of making the instantiation of the GE algorithm
+`multi-stage'. The stepwise instantiation seems more intuitive
+however; it is certainly faster since currently OCaml is quite slow
+when instantiating functors with complex signatures.
+
+As mentioned above, the generic GE algorithm |GenLA| is paremeterized
+over |CONTAINER2D|.  Given a particular container instance, the
+functor |GenLA| yields a module containing various algorithmic aspects
+for the user to choose, as well as the main GE driver. The given
+container instance is available to all these aspects under the name of
+|C| (so that the type of the container can be referred to as |C.contr|
+and the dimensions can be obtained using |C.dim1| and |C.dim2|). The
+numeric domain, which is part of the container, can be referred to as
+|D.Dom| and the type of the domain elements is |C.Dom.v|. We shall see
+many such references as we describe particular algorithmic aspects
+below.
 
@@
 The determinant aspect is one of several tracking and strategic
-aspects. The main function of the algorithm invokes the functions of
-these aspects at some interesting points during the computation (e.g.,
-pivot is required, two rows have to be permuted, the final answer to the
-user has to be built). The instance of the aspect may do something (or
-nothing) at each point, for example, find pivot according to the
-pivoting strategy (such as row pivoting, any non-zero element), update
-the current value of the determinant, etc.
-
-Our sample Gaussian Elimination algorithm, Fig.\ref{fig:GE-algorithm},
+aspects. The main GE procedure invokes various functions of these
+aspects at some interesting points, e.g., the pivot is required, two
+rows have to be permuted, the final answer to the user has to be
+built. An aspect may do something at each point, for
+example, find pivot according to a pivoting strategy, update the
+current value of the determinant, etc.
+
+Our sample Gaussian Elimination algorithm, Fig.~\ref{fig:GE-algorithm},
 demonstrates that tracking determinant is complex: first we define the
 variables |det_sign| and |det_magn| used for tracking (lines 22 and
-23), then we have to flip the sign when swapping two rows or two
+23), then we have to change the sign when swapping two rows or two
 columns or when the matrix found to be singular (lines 32, 37, 59).
@@
 For completeness, we briefly describe and show the signatures of the
-other aspects. The |RANK| aspect tracks rank; the aspect is a simpler
-version of the determinant tracking aspect.  |PIVOTKIND| describes the
-representation of the permutation matrix to track permutations
-executed in pivoting; |TRACKPIVOT| specifies if these permutations are
-tracked at all. |LOWER| tracks the |L| factor when GE algorithm is
+other aspects. |RANK| tracks rank; this aspect is a simpler
+version of determinant tracking. |PIVOTKIND| defines the
+representation of the permutation matrix accumulating pivoting
+permutations; |TRACKPIVOT| specifies if these permutations are
+tracked at all. |LOWER| tracks the |L| factor when the GE algorithm is
 used for the in-place LU-decomposition. |PIVOT| is a higher-order
@@
 pivoting, no pivoting, etc). If pivot is found, the aspect should swap
-rows and columns appropriately. The swapping may need to be tracked if
-we wish to return the permutation matrix. If the determinant is
-tracked, the swapping must update the sign of the determinant. The
-wapiing also affects the L factor being accumulated in-place. The
+rows and columns appropriately. The swapping may need to be tracked in
+the permutation matrix. If we track the determinant,
+the swapping must update its sign. The
+swapping also affects the |L| factor being accumulated in-place. The
 |UPDATE| aspect abstracts over the algorithm for updating matrix
 elements during row-reductions: full division update or fraction-free
-update. Finally, the |INPUT| aspect is an interface aspect abstracting
-over the choice of augmented matrix.
+update. Finally, |INPUT| is an interface aspect letting us handle both
+ordinary and augmented matrices.
 
@@
 module type RANK = sig
-  type 'a tag_lstate = 'a tag_lstate_
+  type 'a tag_lstate
   val decl   : unit -> ('b, int ref) lm
@@
 
-The type |wmatrix| is for convenience, a bundling of several pieces
-of data which are needed by both the |OUTPUT| aspect and the |PIVOT| 
-aspect.  It is defined as
+The type |wmatrix| is for convenience, a bundling of several pieces of
+data needed by both |OUTPUT| and |PIVOT| aspects:
 \begin{code}
@@
 sharing constraints and module computations. That lead to obscure
-code, long impenetrable type error messages and made compilation very
-slow.  Furthermore, at least in OCaml, type-level computations are not
-powerful enough to help us with tasks such as verifying that an integer
-is prime, as required in \S\ref{sec:domains}.
+code, long impenetrable type error messages and very slow compilation.
+Furthermore, type-level computations in OCaml are not
+powerful enough for the tasks such as verifying that an
+integer is prime, see \S\ref{sec:domains}.
 

### 2008-06-14 SCP2008/paper/metamonads.tex
mostly done

@@
 
-The type |wmatrix| is for convenience, a bundling of several pieces of
-data needed by both |OUTPUT| and |PIVOT| aspects:
+The type |wmatrix| denotes the |(0,0)-(numrow-1,numcol-1)|
+rectangular block of |matrix|:
 \begin{code}
@@
 \end{code}
+The type is used in |INPUT|, |OUTPUT| and |PIVOT| aspects to refer to
+the non-augmented part of the matrix.
 
@@
 \subsection{Main generation}\label{sec:main-generation}
-We combine all user-selectable aspects in a record, which serves as a ``keyword
-argument'' list to the |GenGE| functor below.
+
+This section presents the core GE algorithm |GenGE|, after all aspects
+have been factored out. We also describe how to instantiate the core
+algorithm with sample aspects, to obtain particular GE procedures such
+as our running example, Fig.~\ref{fig:GE-algorithm}. The instantiation
+process is multi-step, mainly for the sake of speed of OCaml
+compilation. Currently OCaml is quite slow compiling source code with
+many complex modules. The benefit of the multi-step process, simple
+module signatures, may help make instantiation process more
+comprehensible.
+
+The core GE algorithm and all the aspects are part of one large
+functor, |Ge.LAMake|, parameterized by the interpretation aspect,
+\S\ref{sec:lifteddomains}. If we select as an interpretation
+generating code in the form of MetaOCaml code values, we write
+\begin{code}
+module GEF = Ge.LAMake(Code)
+open GEF
+open Domains_code
+\end{code}
+Here, |Code| is the interpretation with |('a,'b) rep| being 
+|('a,b) code|; the module |Domains_code| contains various instances of |DOMAINL|
+(such as \linebreak|FloatDomainL|, |IntegerDomainL|) for that particular
+choice of |('a,'b) rep|. Opening |GEF| makes available various
+container functors and the functor |GenLA|. Combining
+the container functors with domain structures gives containers
+with particular elements and particular representation. For example, in
+\begin{code}
+module GVC_I = GenericVectorContainer(IntegerDomainL)
+module G_GVC_I = GenLA(GVC_I)
+open G_GVC_I
+open G_GVC_I.GE
+\end{code}
+we define |GVC_I| to represent a matrix with integer elements arranged
+in a flat vector in the row-major order. Instantiating the functor |GenLA|
+with this container makes available all algorithmic and
+tracking aspects of GE (such as |AbstractDet|, |NoDet|, |FullPivot|,
+etc) as well as the module |GE| with the core GE functor |GenGE|
+(besides |GE|, |GenLA| contains modules for GE-based solvers). All
+these components already incorporate the choices for the
+container, domain and interpretation aspects we have made earlier. We
+may now select particular features of the desired GE algorithm and
+instantiate and run |GenGE|.
+
+We combine all user-selectable aspects in a record, which serves as a
+``keyword argument'' list to the |GenGE| functor, shown later.
 \begin{code}
@@
 
-\begin{code}
-module GVC_I = GenericVectorContainer(IntegerDomainL)
-module G_GVC_I = GenLA(GVC_I)
-open G_GVC_I
-open G_GVC_I.GE
+We instantiate |GenGE| by passing to the functor the ``record'' of
+various aspects; the order is irrelevant, but all aspects such as
+|Det| must be specified. In the following code, we request GE generation
+with full pivot, fraction-free update, operate on non-augmented matrix
+and return the |U| factor, determinant and the rank. This choice
+corresponds to our sample algorithm of Fig.~\ref{fig:GE-algorithm}.
+\begin{code}
 module GenIV5 = GenGE(struct 
@@
 \end{code}
-%
-We first instantiate the container |GVC_I| for a vector over an
-integer domain.  |GenGE| is then one functor of many algorithm
-generators from a linear algebra library; since all our linear algebra codes
-are parameterized by the container and the domain included in it,
-We pass the container to the main functor |GenLA|.
-We open the resulting |G_GVC_I| module
-to make all its components easily available. One of those components
-is the GE generator, |GE| (which we open as well). Besides |GE|, the
-module |G_GVC_I| includes the generator for GE-based solvers.  The main
-generator functor is |GenGE|, shown later.  We instantiate the
-generator by passing to the functor the ``record'' of various aspects; the
-order is irrelevant, but all aspects such as |Det| must be
-specified. In the code above, we request GE generation with full
-pivot, fraction-free update, operate on non-augmented matrix and
-return the |U| factor, determinant and the rank. This choice
-corresponds to our sample algorithm of Fig.\ref{fig:GE-algorithm}.
-We run the monad
-passing the initial state |[]| and obtain the code, which we can see
-by printing |resIV5|. The code can then be ``compiled'' as |!. resIV5|
-or with offshoring\cite{offshoring}. The code for |resIV5|
-(App.~\ref{app:code-GenIV5}) shows full
-pivoting, determinant and rank tracking. The code for all these
-aspects is fully inlined; no extra functions are invoked and no tests
-other than those needed by the LU algorithm itself are performed. The
-resulting function returns a triple |int array * int * int| of the
-U-factor, determinant and the rank. It is instructive to compare the
-generated code with the corresponding code in
-Fig.~\ref{fig:GE-algorithm}, the hand-written implementation of the
-textbook pseudo-code; the only difference is the naming of
-variables and the inlining of pivoting and swapping functions.
+We run the monad passing the initial state |[]| and obtain the code,
+which we can see by printing |resIV5|. The code can then be
+``compiled'' as |!. resIV5| or with offshoring \cite{offshoring}. The
+code for |resIV5| (App.~\ref{app:code-GenIV5}) shows full pivoting,
+determinant and rank tracking. The code for all these aspects is fully
+inlined; no extra functions are invoked and no tests other than those
+needed by the GE algorithm itself are performed. The resulting
+function returns a triple |int array * int * int| of the U-factor,
+determinant and the rank. It is instructive to compare the generated
+code with the corresponding code in Fig.~\ref{fig:GE-algorithm}, the
+hand-written implementation of the textbook pseudo-code; the only
+difference is the naming of variables and the inlining of pivoting and
+swapping functions.
 
@@
 
-The LU generator functor (called |GenGE| below) itself is
-parameterized by the domain, container, pivoting policy (full, row,
-nonzero, no pivoting), update policy (with either `fraction-less' or
-full division), and the output specification. Some of the argument
-modules such as |PIVOT| are functors themselves (parameterized by the
-domain, the container, and the determinant functor).  We rely on
-module subtyping: For example, |F.Output| of the type |OUTPUT| is a
-functor requiring an argument structure of the signature
-|OUTPUTDEP|. The fact that the signature |FEATURES| contains all the
-fields of |OUTPUTDEP| and then some lets us pass |F| (of the type
-|FEATURES|) as an argument to instantiate |F.Output|.
+The core GE algorithm |GenGE|, as part of |Ge.LAMake| and |GenLA|, is
+already parameterized by the domain, container and interpretation.
+The algorithm is further parameterized by |FEATURES|, i.e., pivoting
+policy (full, row, nonzero, no pivoting), update policy (with either
+`fraction-less' or full division), determinant, permutation matrix,
+input and output specifications. Some of the argument modules such as
+|PIVOT| are functors themselves (parameterized by the domain, the
+container, and the determinant).  We rely on module subtyping: For
+example, |F.Output| of the type |OUTPUT| is a functor requiring an
+argument of the signature |OUTPUTDEP|. The fact that the signature
+|FEATURES| contains all the fields of |OUTPUTDEP| and then some lets
+us pass |F| (of the type |FEATURES|) as an argument to instantiate
+|F.Output|.
 
@@
 \begin{code}
-module DivisionUpdate(Det:DETERMINANT) =
-  struct
-  open C.Dom = struct 
+module DivisionUpdate(Det:DETERMINANT) = struct
+  open C.Dom
+  type 'a in_val = 'a vc
+  let update bic brc brk bik setter _ = perform
+      y <-- ret (bik -^ ((divL bic brc) *^ brk));
+      ret (setter (applyMaybe normalizerL y))
+  let update_det v = Det.acc_magn v
+  let upd_kind = DivisionBased
   let _ = assert (C.Dom.kind = Domains_sig.Domain_is_Field)
@@
 This structure implements an update policy relying on
-unrestricted |Dom.div|. Many domains provide |Dom.div|, for example,
+unrestricted |Dom.divL|. Many domains provide |divL|, for example,
 the integer domain. The latter however assumes that division is applied
@@
 to be created to follow the operational behaviour of existing code, but
-is not restricted to such a setting.
+are not restricted to such a setting.
 In our work, aspects are weaved together ``from scratch'' to
@@
 appreciated. The implementation of the monadic notation, |perform|,
-was joint work with Lydia van Dijk.
-Many helpful suggestions by the anonymous reviewers are gratefully
-acknowledged.
+was joint work with Lydia van Dijk.  We gratefully acknowledge
+numerous, helpful suggestions by the anonymous reviewers.
 

### 2008-06-16 SCP2008/paper/metamonads.tex
check in of some updates, mostly in pages 1-7

@@
 advanced module system to demonstrate how to eliminate all
-abstraction overhead while avoiding any inspection of the resulting
-code.  We demonstrate this clearly with LU decomposition as a 
+abstraction overhead from generic programs while avoiding any inspection of
+the resulting code.  We demonstrate this clearly with Gaussian Elimination as a 
 representative family of symbolic and numeric algorithms. 
@@
 demand further divisions of the family for particular matrix layouts, e.g.,
-sparsed or tiled.
+sparse or tiled.
 
@@
 algorithm, as well as 45 implementations of directly related
-algorithms such as Cholesky decomposition, and so on.  We could
-manually write each of these implementations optimizing for particular aspects
-and using cut-and-paste to ``share'' similar pieces of code.
+algorithms such as LU decomposition, Cholesky decomposition, and so on.  We
+could manually write each of these implementations, optimizing for particular
+aspects and using cut-and-paste to ``share'' similar pieces of code.
 Or we can write a very generic procedure that accounts for
@@
 and so we can not rely on a general
-purpose compiler to assuredly perform such optimizations
+purpose compiler to assuredly perform such optimizations.
 
@@
 problems. Of the widely available maintainable languages, only
-MetaOCaml \cite{CTHL03,metaocaml-org}  solves all the above problems
+MetaOCaml \cite{CTHL03,metaocaml-org}  solves all of the above problems,
 including the well-typing of both the generator and 
@@
 MetaOCaml is purely \emph{generative}: generated code can only be treated as
-a black box: in other words, it cannot be inspected and it cannot be
+a black box -- in other words, it cannot be inspected nor can it be
 post-processed (i.e., no intensional analysis). This approach gives a stronger
@@
 \cite{TahaThesis} for many simple examples).  For more complex
-examples, new techniques are necessary, e.g., abstract interpretation
+examples, new techniques are necessary, for example abstract interpretation
 \cite{KiselyovTaha}.  But more problems remain
@@
   updates, representing permutation matrices, dealing with augmented 
-  input matrix, and back-propagation. We have also changed the representation
+  input matrix, and back-propagation. We have changed the representation
   of domain-specific knowledge about permissible compositions of aspects.
@@
 in using code generation for scientific (both numeric and symbolic)
-software.  We will use the algorithm family of LU decomposition 
-and linear system solving as our running examples to demonstrate our
+software.  We will use the algorithm family of Gaussian elimination, 
+applied to perform LU decomposition 
+and linear system solving, as our running examples to demonstrate our
 techniques.  Specifically, our contributions are:
@@
       time.
-    \item Provide a thorough classification of the family of LU decomposition
-      algorithms.
+    \item Provide a thorough classification of the family of Gaussian
+    elimination algorithms.
 \end{itemize}
@@
 The rest of this paper is structured as follows: The next section
-gives an overview of the design space of LU decomposition algorithms.
+gives an overview of the design space of Gaussian Elimination algorithms
+(and their application to LU and linear system solving).
 \S\ref{CPS}
@@
 \S\ref{functors} describes the use of
-parametrized modules of OCaml to encode all of the aspects of the
-LU decomposition algorithm family as separate
-modules.  We discuss related work in
+the parametrized modules of OCaml to encode all of the aspects of our
+algorithm family as separate modules.  We discuss related work in
 \S\ref{related} and outline future work. In our
@@
 Gaussian Elimination algorithms.  In the present work, we outline a
-number of additional aspects involved in the family of LU
+number of additional aspects involved in the (related) family of LU
 decomposition algorithms.  These will first be presented in a somewhat
@@
 in the upper triangle.
-\oleg{\S\ref{GE-algorithm} demonstrates one particular GE algorithm in detail,
-  for further reference.}
 
@@
 We reuse the English word ``aspect'' for the various facets of the family
-of LU decomposition algorithms.  While our use shares the 
+of Gaussian Elimination algorithms.  While our use shares the 
 spirit of aspect-oriented programming (AOP)~\cite{kiczales97aspectoriented},
@@
 which were also concerned with scientific software.}.  We firmly believe that
-our typed generative methodology is much better suited to functional
+our typed generative methodology is better suited to functional
 programming, compared to attempts to graft the program-trace-based
@@
 Dijkstra's ideas on separation of concerns \cite{EWD:EWD447}.
-To apply these principles to the study of LU decomposition, we need
+To apply these principles, we need
 to understand what are the changes between different implementations, and 
@@
         multivariate polynomials over a field, or elements of a division ring
-        with possibly undecidable zero-equivalence.  In the roughly 85 pieces
-        of code we surveyed, 20 different domains were encountered.
+        with possibly undecidable zero-equivalence.  In the roughly $85$ pieces
+        of code we surveyed, $20$ different domains were encountered.
     \item \textbf{Representation of the matrix}: Whether the matrix
@@
       degenerate case, useful for testing, is for the generator
-      to run the LU algorithm directly (albeit with great
+      to run the algorithm directly (albeit with great
       abstraction overhead).
@@
         routine.  For certain classes
-        of \emph{expressions}, it turns out convenient to use 
+        of \emph{expressions}, it turns out to be convenient to use 
         a zero-equivalence
@@
 is to be performed. One could therefore select aspects that turn out
-incompatible, and we have to prevent this. 
+to be incompatible, and we have to prevent this. 
 More precisely, our goal is to detect the selection of
@@
 decomposition is possible only for specific domains), yet are useful
-as guidance in creating the modular LU generator discussed in 
+as guidance in creating the modular generator discussed in 
 \S\ref{functors}.
 
-\section{Generating binding statements, CPS, and monad}\label{CPS}
+\section{Techniques for typed code generation}\label{CPS}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+This section  outlines various necessary techniques for typed code
+generation in MetaOCaml.  We start with an introduction to code generation
+with MetaOCaml, where our examples are chosen to illustrate issues of
+direct concern to generic programs.  Next we introduce monads and our
+monadic notation, as a means to make writing programs in continuation
+passing style (CPS) more palatable.  Generation of control statements can
+lead to various subtle issues, and solutions are covered in 
+\S\ref{ssec:control}.  Finally, during generation we need to keep track of
+various aspects, and we use an extensible state for this purpose, described
+in \S\ref{ssec:state}.
+
+\subsection{MetaOCaml}
 
@@
 combinators. MetaOCaml, as an instance of a multi-stage
-programming system \cite{TahaThesis}, provides exactly the needed
+programming system \cite{TahaThesis}, provides exactly the necessary
 features: to construct a code expression, to combine them, and to
@@
 computation \emph{while} we are
-building the future-stage computation. The immediate computation in
+building a future-stage computation. The immediate computation in
 |simplest_code| is the evaluation of the function |gen|, which in turn
@@
 code expressions as (opaque) values. The function |gen| returns a code
-expression, which is inlined in the place of the escape. MetaOCaml can
+expression, which is inlined in the location of the escape. MetaOCaml can
 print out code expressions, so we can examine the final generated code. It
@@
 
-\subsection{Generating control statements}
+\subsection{Generating control statements}\label{ssec:control}
 We can write operations that generate code other than let-statements,
@@
 
-\subsection{Maintaining an extensible state}
+\subsection{Maintaining an extensible state}\label{ssec:state}
 
@@
 
-\oleg{I wonder should we always refer to GE rather than LU-elimination
-  in our paper? GE is a particulat instance of LU; one can accomplish
-  LU in other ways than GE. Also, GE can be used for other means that
-  LU, e.g., just to compute the determinant or to solve linear equations.
-  So, it seems more properly to say we do GE. If we use GE and LU
-  almost synonymously, as it seems to be the case in Mathematica, we
-  should note that in a footnote.}
-
-For detailed reference, we present one particular Gaussian Elimination
+For detailed reference, we present one particular Gaussian Elimination%
+\footnote{For the purposes of this paper, we will use Gaussian Elimination (GE)
+and LU decomposition (LU) as quasi-synonyms, even though we are well-aware
+that GE can be used for other purposes, and LU can be performed by other
+means than GE.}
 algorithm, for an in-place LU-decomposition of an integer matrix with
@@
 a hand-written implementation of the typical pseudo-code in
-Numerical Analysis textbooks \oleg{I used wikipedia; need a better references}.
+Numerical Analysis textbooks (see for example~\cite{BurdenFaires89}).
 
@@
 processed elements. We implement the \emph{fraction-free}
-algorithm \oleg{Citation needed!}:
+algorithm\footnote{sometimes also called the Gauss-Bareiss algorithm}
+\cite{Bareiss68}
 the division operation in line 51 assuredly divides two integers
@@
 environment classifier.
-\oleg{I'm describing in this section coderep.mli and one of its
-  instances, code.ml. The only change is that I use ('a,'b) rep rather
-  than ('a,'b) abstract, for uniformity and clarity. Why did we even
-  need to introduce this additional distinction, `rep' vs
-  `abstract'. Also, the name `abstract' is a bit too
-  general. Anyway, the explanation does not have to follow the code to
-  the letter, as Reviewer 2 has suggested.}
 
@@
 elements. Other aspects follow a similar outline.
-\oleg{The following is perhaps too particular to mention in the
-  paper. AT least we can keep it in comments:
-  We stress the difference in return types of \texttt{decl},
+%\oleg{The following is perhaps too particular to mention in the
+%  paper. }
+% \jacques{Actually, I liked it, so after slight editing, I kept it}
+It is instructive the examine the difference in return types of \texttt{decl},
 \texttt{upd\_sign} and \texttt{zero\_sign} functions. The former says
@@
 \texttt{upd\_sign} is the generator that may produce code, or may
-  not. The option type lets us avoid generating code such as 
-  \texttt{a := b; ()} with useless dummies.}
+not. The option type lets us avoid generating pointless code such as 
+\texttt{a := b; ()}.
 

### 2008-06-17 SCP2008/paper/metamonads.tex
finish complete readthrough and spell check

@@
 pivoting policies, result types, etc. -- at no run-time cost.  Because
-the resulting code is generated just right and not changed afterwards,
+the resulting code is generated just right and not changed afterward,
 MetaOCaml guarantees that the generated code is well-typed.
@@
 
-\subsection{MetaOCaml}
-
-We build large code generators out of primitive generators using 
+\subsection{MetaOCaml and basic abstraction}\label{ssec:metaocaml}
+
+We wish to build large code generators out of primitive generators using 
 combinators. MetaOCaml, as an instance of a multi-stage
@@
 features: to construct a code expression, to combine them, and to
-execute them. The following shows the simplest code generator |one|,
-and the simplest code combinator\footnote{%
+execute them. The following shows a simple code generator |one|,
+and a simple code combinator\footnote{%
 $\Longrightarrow$ under an expression shows the result of its evaluation}:
@@
 We use MetaOCaml brackets |.<...>.| to generate code expressions,
-i.e., to construct future-stage computations. MetaOCaml provides only
-one mechanism of combining code expressions, by splicing one
-piece of code into
-another. The power of that operation, called escape |.~|, comes from
+i.e. to construct future-stage computations. MetaOCaml provides only
+one mechanism for combining code expressions, by splicing one
+piece of code into another. The power of that operation, called escape and
+denoted |.~|, comes from
 the fact that the expression to be spliced in (inlined) can be
-computed: escape lets us perform an arbitrary immediate code-generating
+computed: escape lets us perform an arbitrary \emph{code-generating}
 computation \emph{while} we are
@@
 code expressions as (opaque) values. The function |gen| returns a code
-expression, which is inlined in the location of the escape. MetaOCaml can
+expression, which is inlined in the location of the escape. MetaOCaml 
+conveniently can
 print out code expressions, so we can examine the final generated code. It
-has no traces of |gen| or |plus|: they are generation stage computations.
+has no traces of |gen| or |plus|: those are purely generation stage
+computations.
 
@@
 
-By parametrizing our code, we can make the benefits of code generation
+By parameterizing our code, we can make the benefits of code generation
 evident:
@@
 Running the former expression yields a function on |float|s, whereas
-the latter expression is the code expression for a boolean function.
-This clearly shows the separation of concerns for domain
-operations.
+the latter expression is a code expression for a boolean function.
+This simple technique clearly shows how we can be polymorphic over
+domain operations and yet still generate efficient domain-specific code,
+thus achieving a proper separation of concerns.
 
@@
 which may be a rather complex computation which we would rather not do
-twice. We may be tempted to rely on the compiler's
+twice. We might be tempted to rely on the compiler's
 common-subexpression elimination optimization. When the generated code is
@@
 \end{code}
-However,
-the result of |param_code1' plus one| still exhibits duplicate
-sub-expressions.  Our |let|-insertion optimization only saved the
-computation at the generating stage.  We need a combinator that
-inserts the |let| expression in the generat\emph{ed} code. We need a
+The result of |param_code1' plus one| still exhibits duplicate
+sub-expressions.  This is because our |let|-insertion optimization only saved
+the computation at the generating stage.  We need a combinator that
+inserts the |let| expression in the generat\emph{ed} code, in a other words a
 combinator |letgen| to be used as
@@
 |init' (fun ce -> ...)| in CPS. For one, |let| became
-``inverted''. For another, what used to be an expression that yields
+``inverted''. Secondly, what used to be an expression that yields
 a value, |init|, became an expression that takes an extra argument,
@@
 the above example. In larger expressions with many let-forms, the
-number of parentheses around |fun| increases, the need to add and
+number of parentheses around |fun| increases considerably, the need to add and
 then invoke the |k| continuation argument become increasingly annoying. The
 inconvenience is great enough for some people to explicitly avoid CPS
-or claim that numerical programmers (our users) cannot or will not
+or claim that programmers of scientific software (our users) cannot or will not
 program in CPS. Clearly a better notation is needed.
@@
   = fun s k -> m s (fun s' b -> f b s' k)
+
 let fetch s k = k s s  and  store v _ k = k v ()
@@
 \end{code}
-\caption{Our monad}\label{ourmonad}
+\caption{Our monad, helper functions and uses}\label{ourmonad}
 \end{figure}
@@
 %
-We use a single type parameter |'p| to
-represent all parameters of our monad (all parameters but the type of
-the monadic value |'v|). The type variable |'p| is constrained to be
+We use a single type parameter |'p| to represent all parameters of our monad,
+more precisely all parameters but the type of
+the monadic value |'v|. The type variable |'p| is constrained to be
 the type of an object with methods (fields) |state| and |answer|. The
@@
 type are not part of our computations and need not exist. We merely
-use the object type, as an convenient way to specify extensible
+use the object type as an convenient way to specify extensible
 \emph{type-level} records in OCaml.
@@
 |'c|, and we update the |answer| component of the record to specialize
-the answer type to be the type of code. The |state| component of the
+the answer type to be a code type. The |state| component of the
 record is not affected, and so remains hidden in the ellipsis |..|. The
@@
 %
-That does not seem like much of an improvement. With the help of
-camlp4 pre-processor, we introduce the |perform|-notation \cite{metamonadsURL},
-patterned after the |do|-notation of Haskell (see App.~\ref{app:perform}).
+That may not seem like much of an improvement, but with the help of the
+camlp4 pre-processor, we can introduce the |perform|-notation
+\cite{metamonadsURL}, patterned after the |do|-notation of Haskell (see
+App.~\ref{app:perform}).
 \begin{code}
@@
 |param_code4|, written using the |perform|-notation, is equivalent to
-|param_code3| -- in fact, the camlp4 preprocessor will convert the
+|param_code3| -- in fact, the camlp4 preprocessor converts the
 former into the latter. And yet, |param_code4| looks far more
@@
 albeit straightforward, is not as general as we wish: its arguments are
-already generated pieces of code rather than monadic values. We
-can ``lift it'':
+pieces of code rather than monadic values. We can ``lift it'':
 \begin{code}
@@
 If we use |ifM'| to generate guarded array access code, the let-insertion
-happened \emph{before} the if-expression, that is, before the test that
+happens \emph{before} the if-expression, that is, before the test that
 the index |i| is positive. If |i| turned out
@@
 error. On the other hand, the code with |ifM| accesses the array only
-when we have verified that the index is non-negative. This example
-makes it clear that code generation (such as the one in |retN|) is 
-truly an effect, and we have to be clear about the sequencing of
+after we have verified that the index is non-negative. This example
+should make it clear that code generation (such as the one in |retN|) is 
+truly an effect, and that we have to be clear about the sequencing of
 effects when generating control constructions such as conditionals.
@@
 
+One can think of this particular use of continuation as delimiting the 
+``current scope'' of a block of code.  When constructing blocks
+with a new scope, we use a fresh continuation; this allows us to generate
+all named computations at the ``top'' of the current scope but no farther.
+
 \subsection{Maintaining an extensible state}\label{ssec:state}
@@
 Various aspects of our generator need to keep a state during
-code generation (e.g., the name of the variable for the sign of
+code generation (for example the name of the variable for the sign of
 the determinant, \S\ref{sec:det}).
-
-The simplest method of keeping the state of the generator is by using
+The simplest method of keeping such state is by using
 mutable variables, private to each module (aspect). That would however
@@
 We therefore chose a different way of maintaining generator state,
-using a monad. We already need a monad for let-insertion; we now add state as
-an additional monad parameter (Fig.~\ref{ourmonad}) and monadic
-actions |fetch| and |store| to access that monadic state. The state is
+using a monad. We already saw that for let-insertions we could use
+a continuation monad;  we now show how to use the state component
+of our monad (Fig.~\ref{ourmonad}). The monadic actions |fetch| and
+|store| are used to access that monadic state, which is
 threaded throughout the entire code-generation computation.
@@
 lists~\cite{mlton-proplist}.  A property list also represents an
-extensible object but by its dual, namely a list of fields.  The
+extensible object but via its dual, namely a list of fields.  The
 initial object is just the empty list.  Unfortunately, we cannot apply
@@
 approach uses generativity of exceptions or reference cells to
-generate property names, i.e., fields of an extensible record. This
+generate property names, i.e. the fields of an extensible record. This
 technique would make our aspects stateful modules, which we described
-earlier as undesirable. MLton's approach to generating field
-names also makes it difficult to store in the fields MetaOCaml's code
-values, whose type contains a generic type variable, the environment
-classifier. 
+earlier as undesirable. MLton's approach to generating field names also makes
+it difficult to store code values in those fields, as a code value has a type
+which contains a generic type variable, the environment classifier. 
 Fortunately, OCaml lets us build open unions with
 polymorphic variants. Each variant is identified by a manifest tag,
-e.g., |`Tdet|, and may include a value. The tags are not generative and
-provide a form of manifest naming, similar to symbols of
-Lisp and Scheme.
+e.g. |`Tdet|, and may include a value. The tags are not generative and
+provide a form of manifest naming, similar to symbols in Lisp and Scheme.
 
 Below is the implementation of our open records with manifest
-naming: functions |orec_store| to add a new field to the open record
+naming: functions |orec_store| to add a new field to an open record
 and |orec_find| to obtain the value associated with a particular field
 name. Each ``field'' is characterized by a triple: an injection function, 
-a projection function and the string name. The latter is used for 
+a projection function and the string name. The latter is (only) used for 
 printing error messages. For example, for the
-determinant tracking aspect, this triple has the form
+determinant tracking aspect (\S\ref{sec:det}), this triple has the form
 \begin{code}
@@
 
-let rec lookup ((_,prj,_) as ip:(('a,'b) open_rec) )
-   : 'b list -> 'a = 
+let rec lookup ((_,prj,_) as ip:(('a,'b) open_rec)) : 'b list -> 'a = 
    function [] -> raise Not_found
@@
     try let _ = lookup ip s in 
-        failwith ("The field of an open record is already present: " ^ name)
+        failwith ("Field "^name^" of an open record is already present."
     with Not_found -> () in
@@
 look up a field that does not exist. It is possible to make these
-checks static.  Our approach has the advantage of generating much
+checks static.  Our approach has the advantage of generating \emph{much}
 clearer error messages.
@@
 means than GE.}
-algorithm, for an in-place LU-decomposition of an integer matrix with
+algorithm, for an in-place LU decomposition of an integer matrix with
 full pivoting, returning the U-factor, determinant and rank. The
 $n\times m$-matrix is represented as a flat vector, with 0-indexed
-elements laid out in row-major format (C-style). The code below is
+elements laid out in row-major format (C-style). The code in 
+Fig.~\ref{fig:GE-algorithm} is
 a hand-written implementation of the typical pseudo-code in
@@
 
-In OCaml, the matrix is represented by the value of the following type:
+In OCaml, the matrix is represented by a value of the following type:
 \begin{code}
@@
 where |a| is the array component of the record |container2dfromvector|.
-The above code defines three auxiliary functions, typically used in the
+The code above defines three auxiliary functions, typically used in 
 textbook pseudo-code: |swap a i j| swaps two elements; 
-|swap_cols a (n,m) c j| swaps the column |c| with the column |j| in
+|swap_cols a (n,m) c j| swaps column |c| with column |j| in
 the |(n,m)|-matrix |a|. The function 
-|swap_rows| |a| |(n,m)| |(r,c) i| swaps the row |r| with the row |i| in 
+|swap_rows| |a| |(n,m)| |(r,c) i| swaps row |r| with row |i| in 
 the non-yet examined portion of the matrix, 
@@
   end
-
 let ge = fun a_orig ->
@@
                 (* fraction-free elimination *)
-                      a.(ii*m+j) <- (a.(ii*m+j) * a_rc - a.(!r*m+j) * cur) 
-                                    / ! det_magn
+                a.(ii*m+j) <- (a.(ii*m+j) * a_rc - a.(!r*m+j) * cur) / ! det_magn
               done;
@@
 \end{figure}
-As typically presented in the textbooks, the main algorithm depends on
-the separately-defined function |pivot| to find a pivot. We use
-full-pivoting, i.e., searching the non-yet examined portion of the matrix, 
-the rectangular block |(r,c)-(n,m)|, for the element with the 
+As is typical of textbook presentations, the main algorithm depends on
+a separately-defined function |pivot| to find a pivot. Here we use
+full-pivoting, i.e. searching the completel as yet unexamined portion of the
+matrix, the rectangular block |(r,c)-(n,m)|, for the element with the 
 maximum absolute value. The function returns the value of the
-found pivot and its location; the return value is an option type with
-|None| indicating that all examined elements are zeros. In the main
-algorithm, after we found the pivot, we swap the current column 
-with the pivot column, and swap the current row with the pivot row. 
+pivot thus found, and its location; the return value is an option type with
+|None| indicating that all examined elements are zeroes. In the main
+algorithm, after we have found the pivot, we swap the current column 
+with the pivot column, and swap the current row with the pivot row (if
+necessary). 
 After the swap, the $a_{rc}$ element of the matrix is the pivot.
 Swapping two rows or two columns changes the sign of the determinant.
-After the swapping, the algorithm does so-called row-reduction
-over the |(r,c)-(n,m)| block of the matrix containing not-yet
-processed elements. We implement the \emph{fraction-free}
+After the swaps, the algorithm performs so-called row-reduction
+over the |(r,c)-(n,m)| block of the matrix.
+We implement the \emph{fraction-free} version of the
 algorithm\footnote{sometimes also called the Gauss-Bareiss algorithm}
-\cite{Bareiss68}
-the division operation in line 51 assuredly divides two integers
+\cite{Bareiss68},
+where the division operation in line 50 assuredly divides two integers
 with no remainder, which requires the accumulation
@@
 
-The monad represents fine-scale code generation. We need tools for
-larger-scale modularization; conveniently, we can use whatever abstraction
-mechanisms we want to structure our code generators, as long as these
-abstractions do not infiltrate the generated code. For our purposes, the
+Our monad gives us the tools to implement fine-scale code generation. We need
+tools for larger-scale modularization; conveniently, we can use whatever
+abstraction mechanisms we want to structure our code generators, as long as
+these abstractions do not infiltrate the generated code. For our purposes, the
 ML module system turns out to be most convenient.
@@
 abstracted away. Finally, we show an instantiation of the generic
-generator, whose execution yields code for the algorithm of
+generator, whose execution yields (exactly) the code for the algorithm of
 Fig.~\ref{fig:GE-algorithm}, this time automatically generated rather
@@
 
-We describe here the following aspects, in the general
+We describe will the following aspects, generally in
 order of increasing complexity: domain (of the group `Abstract
@@
 Fig.~\ref{fig:GE-algorithm}, remains the same for integer, float,
-etc. matrices and so can be abstracted over the numeric domain.  We
+polynomial, 
+etc. matrices and so can be abstracted over the domain.  We
 have already seen the simplest case of domain abstraction in
-|param_code1|, which took code-generators such as |plus| and |one| as
-arguments.  We need far more than two parameters: our domains should
-include $0$, $1$, $+$, $*$, (unary and binary) $-$, at least
+|param_code1| (\S\ref{ssec:metaocaml}), which took code-generators such as
+|plus| and |one| as arguments.  We need far more than two parameters: our
+domains should include $0$, $1$, $+$, $*$, (unary and binary) $-$, at least
 \emph{exact} division, normalization, and potentially a relative size
 measure. We could group these parameters in tuples or records.
-Instead, we use OCaml \emph{structures} (i.e., modules) so we can take
+It is instead more convenient to use OCaml \emph{structures} (i.e., modules) so
+that we can take
 advantage of extensibility, type abstraction and constraints, and
-especially parameterized structures (\emph{functors}).  We define the
+especially parameterized structures (\emph{functors}).  We define a
 type, the signature |DOMAIN|, which different domains must satisfy:
@@
 case the compiler will verify the type when we try to use that structure as a
-|DOMAIN|. In any case, the errors such as missing ``methods'' or
-methods with incorrect types will be caught statically, 
-\emph{before} any code generation takes place. The variant
+|DOMAIN| (typically in a functor instantiation). In any case, the errors such
+as missing ``methods'' or methods with incorrect types will be caught
+statically, \emph{before} any code generation takes place. The variant
 |Domain_is_Ring| of |IntegerDomain.domain_kind| encodes a semantic constraint:
@@
 not treat here.}.  Since we intend to make a field of prime characteristic, 
-we must check this, as is done
+we must check this, which is done
 in the last line of the above code. That line differs from the other 
@@
 If we
-replace |p = 19| with |p = 9| above, we receive a run-time
+replace |p = 19| with |p = 9| above, we receive a ``run-time''
 error. However, it is raised as we instantiate and combine modules that
@@
 generation of the target code could begin.  In our code we make
-extensive use of these ``preflight checks'', which are performed as part of
+extensive use of these ``preflight checks'' which are performed as part of
 module initialization. These checks seem to offer a good compromise:
@@
 corresponding \emph{semantic} constraints will be generated. Although
-some may frown on the use of module initializing expressions 
-as this requires careful attention to sharing and multiple instantiations
-of a module, these concerns do not apply in our case: our preflight
-checks are all idempotent and maintain no state.
+some may frown on the use of module initializing expressions,
+as in general this requires careful attention to sharing and multiple
+instantiations of a module, these concerns do not apply in our case: our
+preflight checks are all idempotent and maintain no state.
 
@@
 |(curr == 0)|, line 46, compares two integers (or, generally, two
-domain elements); the operation |det_magn := a_rc|, line 56, assigns
+domain elements); the operation |det_magn := a_rc|, line 55, assigns
 the domain element to the corresponding reference cell; the operation
@@
 interpretation of basic operations, so that the same GE code, given
-different concrete interpretations, can LU-factorize a given matrix,
-can generate LU-factorization programs in OCaml, C or Fortran, or can
+different concrete interpretations, can LU factorize a given matrix,
+can generate LU-factorization programs in OCaml as well as C or Fortran, or can
 pretty-print the factorization procedure.
 
-The interpretation aspect is too implemented as an OCaml module.  The
+The interpretation aspect is also implemented as an OCaml module.  This
 aspect is quite large -- there are many basic operations to abstract
@@
 describes what sort of objects the interpretation may produce (e.g.,
-AST, strings with C code, etc). The type has two parameters: the
+ASTs, strings with C code, etc). The type has two parameters: the
 second specifies the type of the object, and the first is the
@@
 
-The first sub-module of the interpretation aspect is the numeric
+The first sub-module of the interpretation aspect is the base
 domain. The signature |DOMAIN| of \S\ref{sec:domains} defined the set
-of operations on numeric objects of some type |v|. We now
+of operations on base objects of some type |v|. We now
 generalize, or `lift', |DOMAIN| into |DOMAINL| so we can likewise
-operate on interpretations of numeric objects, of the type
-|('a,v) rep|:
+operate on other interpretations of these objects, of type |('a,v) rep|:
 \begin{code}
@@
 An interpretation also needs to specify how to compare objects
-and to manipulate objects representing integers, typically used as
-indices of matrix elements. We group the operations into the
+and to manipulate objects representing integers (typically used as
+indices of matrix elements). We group the operations into the
 following two structures:
@@
 As we argued in \S\ref{CPS}, we will be writing our generic GE
-algorithm in the monadic style. For convenience we define the
-following type synonyms for our |('p,'v) monad| of Fig.~\ref{ourmonad}:
-The first is used for a monadic action that always produces a useful
+algorithm in a monadic style. For convenience we define the
+following type synonyms for our |('p,'v) monad| of Fig.~\ref{ourmonad}: The
+first (|cmonad|) is used for a monadic action that always produces a useful
 value; the |omonad| synonym describes a monadic action that may
@@
 
-Finally the interpretation aspect must specify the very basic
+Finally the interpretation aspect must specify the following very basic
 operations: injection of literal values, function
 applications, statement sequencing, conditional, loops, creating,
-dereferencing and assigining to reference cells, etc:
+dereferencing and assigning to reference cells, etc:
 \begin{code}
@@
 The `pure' operations of the interpretation produce interpretation
-objects (of the type |('a,'b)| rep) as the result. We can trivially
-`lift' these operations into the monad by composing them with |ret| of
+objects (of type |('a,'b)| rep) as their result. We can trivially
+`lift' these operations into the monad by composing them with |ret| from
 the monad. Some other operations, like |seqM| and |ifM| are present
@@
 For our purposes, a container is an abstraction of an $n$-dimensional
-vector space, which we here specialize for $n=1,2$.  The
+vector space, which we specialize here for $n=1,2$.  The
 $2$-dimensional case is our main interest, and its signature contains
@@
 the components. This is quite convenient since operations on
-containers are usually accompanied by the operations on the retrieved
-and stored values. The particular instances of the containers are
-parametric over a |DOMAINL|, i.e., functors from a |DOMAINL| module to
+containers are usually accompanied by operations on retrieved values,
+which are subsequently stored again.
+The particular instances of the containers are
+parametric over a |DOMAINL|, i.e. functors from a |DOMAINL| module to
 the actual implementation of a container.  For example, the following
 functor defines a matrix container as a single array, with elements
-stored in the row-major order -- the container used in
+stored in row-major order -- the container used in
 Fig.~\ref{fig:GE-algorithm}.
-\begin{code}[commandchars=\\\{\}]
+\begin{code}
 module GenericVectorContainer(Dom:DOMAINL) =
@@
 \end{code}
-The accompanying code \cite{metamonadsURL} includes the container with
+The accompanying code \cite{metamonadsURL} includes an implementation with
 elements stored in a 1D array in a column-wise (Fortran-like)
-mode, and the container for a matrix represented as an array of rows.
+mode, and another for a matrix represented as an array of rows.
 
@@
 signatures of other functors such as |GenLA| of
-\S\ref{sec:main-generation} below. If the type |CONTAINER2D| were a
+\S\ref{sec:main-generation}. If the type |CONTAINER2D| were a
 functor, |GenLA| would have been a higher-order functor, and we would
@@
 the expense of making the instantiation of the GE algorithm
-`multi-stage'. The stepwise instantiation seems more intuitive
+``multi-stage''. The stepwise instantiation seems more intuitive
 however; it is certainly faster since currently OCaml is quite slow
@@
 
-As mentioned above, the generic GE algorithm |GenLA| is paremeterized
+As mentioned above, the generic GE algorithm |GenLA| is parametrized
 over |CONTAINER2D|.  Given a particular container instance, the
@@
 and the dimensions can be obtained using |C.dim1| and |C.dim2|). The
-numeric domain, which is part of the container, can be referred to as
-|D.Dom| and the type of the domain elements is |C.Dom.v|. We shall see
+domain, which is part of the container, can be referred to as
+|C.Dom| and the type of the domain elements is |C.Dom.v|. We shall see
 many such references as we describe particular algorithmic aspects
@@
 The determinant aspect is one of several tracking and strategic
-aspects. The main GE procedure invokes various functions of these
-aspects at some interesting points, e.g., the pivot is required, two
-rows have to be permuted, the final answer to the user has to be
-built. An aspect may do something at each point, for
-example, find pivot according to a pivoting strategy, update the
+aspects. The main GE procedure invokes various functions of these aspects at
+some interesting points, for example when the pivot is required, when two rows
+have to be permuted, or when the final answer to the user has to be
+built. An aspect may do something at each one of these times, for
+example, find a pivot according to a pivoting strategy, update the
 current value of the determinant, etc.
 
-Our sample Gaussian Elimination algorithm, Fig.~\ref{fig:GE-algorithm},
-demonstrates that tracking determinant is complex: first we define the
-variables |det_sign| and |det_magn| used for tracking (lines 22 and
+Our sample Gaussian Elimination algorithm (Fig.~\ref{fig:GE-algorithm}),
+demonstrates that tracking the determinant is quite complex: first we define
+the variables |det_sign| and |det_magn| used for tracking (lines 22 and
 23), then we have to change the sign when swapping two rows or two
-columns or when the matrix found to be singular (lines 32, 37, 59).
-The value of the determinant should be updated per each pivoting (line
-56). Finally, we convert the tracking state to the resulting
-determinant value (lines 64-66). We observe these lines are not
-contiguous: the determinant aspect is consulted at several separate
+columns or when the matrix found to be singular (lines 32, 37, 58).
+The value of the determinant should be updated for each pivoting (line
+55). Finally, we convert the tracking state to the resulting
+determinant value (lines 63-65). Most importantly, we observe that these lines
+are not contiguous: the determinant aspect is consulted at several separate
 places in the GE algorithm, and the aspect is supposed to keep
-state. As other aspects, the determinant aspect is a module of the
-following signature:
+state. As with other aspects, the determinant aspect is a module,
+and has the following signature:
 \begin{code}
@@
 operations of the aspect, such as |decl| for initializing the tracking
-state, |zero_sign| to note matrix singularity, |set_magn| to set the
+state, |zero_sign| to record matrix singularity, |set_magn| to set the
 magnitude and |fin| to convert the tracking state to the resulting
@@
 % \jacques{Actually, I liked it, so after slight editing, I kept it}
-It is instructive the examine the difference in return types of \texttt{decl},
-\texttt{upd\_sign} and \texttt{zero\_sign} functions. The former says
-that \texttt{decl} is an action which is executed only for its
+It is instructive the examine the difference in the return types of the
+\texttt{decl}, \texttt{upd\_sign} and \texttt{zero\_sign} functions. The first
+says that \texttt{decl} is an action which is executed only for its
 side-effect. It yields no interpretation object (e.g., produces no
 code value). The function \texttt{zero\_sign} always produces a code
-value -- an expression such as assignment that has the type unit. 
-The function
-\texttt{upd\_sign} is the generator that may produce code, or may
+value -- an expression such as assignment that has type unit. 
+The function \texttt{upd\_sign} is a generator that may produce code, or may
 not. The option type lets us avoid generating pointless code such as 
@@
       mo_extend ip (sign,magn)
-  let upd_sign () = perform
+  let upd_sign () = perform            (* flip sign *)
       (sign,_) <-- mo_lookup ip;
       ret (Some (assign sign (Idx.uminus (liftGet sign))))
-  let fin = fun () -> perform
-      (sign,magn) <-- mo_lookup ip;
+  let fin = fun () -> perform          (* reconstruct det and *)
+      (sign,magn) <-- mo_lookup ip;    (* generate code       *)
       ifM (Logic.equalL (liftGet sign) Idx.zero) (ret zeroL)
@@
 tracked at all. |LOWER| tracks the |L| factor when the GE algorithm is
-used for the in-place LU-decomposition. |PIVOT| is a higher-order
+used for in-place LU-decomposition. |PIVOT| is a higher-order
 functor abstracting over the pivoting algorithm (full pivoting, row
-pivoting, no pivoting, etc). If pivot is found, the aspect should swap
+pivoting, no pivoting, etc). If a pivot is found, the aspect should swap
 rows and columns appropriately. The swapping may need to be tracked in
-the permutation matrix. If we track the determinant,
-the swapping must update its sign. The
-swapping also affects the |L| factor being accumulated in-place. The
+a permutation matrix (or some other representation). If we track the
+determinant, swapping must update its sign. The
+swapping also affects the |L| factor being accumulated. The
 |UPDATE| aspect abstracts over the algorithm for updating matrix
@@
 \emph{abstract} type |res| for the result type; different instances of
-the signature set the result type differently. Given below is this
+the signature set the result type differently. Below we show this
 module type and one instantiation, |OutDetRank|, which specifies the
-output of a GE algorithm as a 3-tuple |contr * Det.outdet * int| of
+output of a GE algorithm as a $3$-tuple |contr * Det.outdet * int| of
 the U-factor, the determinant and the rank. That choice of output
@@
 The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
-the preflight checks. As we saw in the previous section, both
+preflight checks. As we saw in the previous section, both
 instances of |DETERMINANT| contain a |fin ()| function to generate
@@
 out the semantic error where a user specifies that the determinant
-should be computed and returned, and yet specifies the |NoDet| aspect
-which tracks no determinant at all.
+should be computed and returned, and yet specifies the |NoDet| aspect.
 
@@
 \end{code}
-The type is used in |INPUT|, |OUTPUT| and |PIVOT| aspects to refer to
+The type is used in the |INPUT|, |OUTPUT| and |PIVOT| aspects to refer to
 the non-augmented part of the matrix.
@@
 choice of a particular tracking aspect may depend on all other choices
-(e.g., it is not possible to extract the |L| factor if the domain is
+(e.g. it is not possible to extract the |L| factor if the domain is
 not a field). Currently we use preflight checks to ensure consistency
@@
 powerful enough for the tasks such as verifying that an
-integer is prime, see \S\ref{sec:domains}.
+integer is prime (see \S\ref{sec:domains}).
 
@@
 algorithm with sample aspects, to obtain particular GE procedures such
-as our running example, Fig.~\ref{fig:GE-algorithm}. The instantiation
+as our running example (Fig.~\ref{fig:GE-algorithm}). The instantiation
 process is multi-step, mainly for the sake of speed of OCaml
-compilation. Currently OCaml is quite slow compiling source code with
-many complex modules. The benefit of the multi-step process, simple
-module signatures, may help make instantiation process more
+compilation.  The benefit of the multi-step process, simple
+module signatures, may also help make the instantiation process more
 comprehensible.
@@
 we define |GVC_I| to represent a matrix with integer elements arranged
-in a flat vector in the row-major order. Instantiating the functor |GenLA|
+in a flat vector in row-major order. Instantiating the functor |GenLA|
 with this container makes available all algorithmic and
 tracking aspects of GE (such as |AbstractDet|, |NoDet|, |FullPivot|,
-etc) as well as the module |GE| with the core GE functor |GenGE|
+etc) as well as the module |GE| with the core GE generator functor |GenGE|
 (besides |GE|, |GenLA| contains modules for GE-based solvers). All
@@
 
-We combine all user-selectable aspects in a record, which serves as a
+We combine all user-selectable aspects in a ``record'', which serves as a
 ``keyword argument'' list to the |GenGE| functor, shown later.
@@
 |Det| must be specified. In the following code, we request GE generation
-with full pivot, fraction-free update, operate on non-augmented matrix
-and return the |U| factor, determinant and the rank. This choice
+with full pivot, fraction-free update, operating on non-augmented matrix
+and returning the |U| factor, determinant and the rank. This choice
 corresponds to our sample algorithm of Fig.~\ref{fig:GE-algorithm}.
@@
 \end{code}
-We run the monad passing the initial state |[]| and obtain the code,
-which we can see by printing |resIV5|. The code can then be
+We run the monad, passing in the initial state |[]| and thus obtain code,
+which we can see by printing |resIV5|. This code can then be
 ``compiled'' as |!. resIV5| or with offshoring \cite{offshoring}. The
@@
 Gaussian Elimination algorithm (Fig.~\ref{fig:GE-algorithm}) 
-is still visible in this code generator:
+is still visible in this code generator: for example,
 the |forward_elim| function iterates over the columns and rows of the
-matrix, finding a pivot, and zeroing the appropriate entries.
+matrix, finding a pivot, and zeroing the appropriate entries.  With
+sufficient added syntactic sugar, we could indeed make the generator look
+like the algorithm.
 
@@
 the integer domain. The latter however assumes that division is applied
-only if the dividend is the exact multiple of the divisor. Thus if we
+only if the dividend is an exact multiple of the divisor. Thus if we
 specified |module Update = DivisionUpdate| when instantiating |GenIV5|
 above, we would have received an error because |IntegerDomainL| is not
-a field. That run-time error occurs before any code is produced (i.e.,
+a field. That error occurs before any code is produced (i.e.
 before |resIV5| is computed).
@@
 \cite{SwadiMonadic06,KiselyovTaha}.  However those papers used only
-|retN| and fixpoints (for generation-time iterations).  This paper
+|retN| and fixpoints (for generation-time iterations).  Our work
 does not involve monadic fixpoints because the generator is not
@@
 
-|Blitz++| \cite{Veldhuizen:1998:ISCOPE} and {C++} template
-meta-programming in general similarly eliminate levels
+|Blitz++| \cite{Veldhuizen:1998:ISCOPE}, and {C++} template
+meta-programming in general, similarly eliminate levels
 of abstraction.  With traits and concepts, some domain-specific
@@
 composition errors) are detected only when compiling the generated
-code. It is immensely difficult to correlate errors (e.g., line
+code. It is immensely difficult to correlate errors (e.g. line
 numbers) to the ones in the generator itself.
@@
 C code in C using |printf|, nothing prevents producing code that
-misses semicolons, open or close parentheses or variable
+is missing semicolons, open or close parentheses or variable
 bindings. MetaOCaml gives us assurance that these errors, and more
-subtle type errors, shall not occur in the generated code.  SPIRAL
+subtle type errors, shall never occur in the generated code.  SPIRAL
 \cite{Pueschel:05} is another even more ambitious project.  But
@@
 specialization of mathematical algorithms. Although it can eliminate
-some overhead from a very generic implementation (e.g., by inlining
+some overhead from a very generic implementation (e.g. by inlining
 aspects implemented as higher-order functions), specialization cannot
@@
 Our code was initially motivated by trying to unify the various
-implementations in Maple.  Interestingly, when we compare our end 
+implementations found in Maple.  Interestingly, when we compare our end 
 result with the options available from Maple's \texttt{LUDecomposition}
@@
 allow us to write such code with the same guarantees that the rest of
-MetaOCaml already affords us.
+MetaOCaml already affords us.  Unfortunately, as modules are not first-class
+objects in OCaml, this currently seems out of reach.
 
@@
 continuations and our implementations of code generators like
-|ifM|. The ultimate (and plausible) goal is to write a numerical
-algorithm in (almost) regular OCaml once, and being able to either
-run it as a regular OCaml program, or turn into a code generating
+|ifM|. The ultimate (and plausible) goal is to write an
+algorithm in (almost) regular OCaml once, and be able to either
+run it as a regular OCaml program, or turn it into a code generating
 aspect.
@@
 \section{Conclusion}\label{conclusion}
-In this paper we have demonstrated numerical code extensively parameterized
+In this paper we have demonstrated code extensively parameterized
 by complex aspects at no run-time overhead.  The combination of
@@
 are the typing ones plus the constraints we specifically
-impose, including semantic constraints (e.g., rings do not have full
-division).
+impose, including semantic constraints.
 
@@
 make up a piece of code.  One can understand previous work to be
-more akin to dynamically typed, and dynamically specified aspect weaving, while
+more akin to dynamically typed and dynamically specified aspect weaving, while
 we have started investigating statically typed and statically specified
@@
 separate them out cleanly. Attempts at such a separation in a
-non-staged setting, have lead to fantastically inefficient code,
+non-staged setting have lead to fantastically inefficient code,
 unacceptable for scientific computation. Staging permitted us, perhaps

### 2008-06-20 SCP2008/paper/metamonads.tex
all the changes that Oleg recommended

@@
 the latter expression is a code expression for a boolean function.
-This simple technique clearly shows how we can be polymorphic over
+This simple technique clearly shows how we can abstract over
 domain operations and yet still generate efficient domain-specific code,
@@
 after we have verified that the index is non-negative. This example
-should make it clear that code generation (such as the one in |retN|) is 
+demonstrates that code generation (such as the one in |retN|) is 
 truly an effect, and that we have to be clear about the sequencing of
@@
 using a monad. We already saw that for let-insertions we could use
-a continuation monad;  we now show how to use the state component
+a continuation monad;  we now demonstrate the state component
 of our monad (Fig.~\ref{ourmonad}). The monadic actions |fetch| and
@@
     try let _ = lookup ip s in 
-        failwith ("Field "^name^" of an open record is already present."
+        failwith ("Field "^name^" of an open record is already present.")
     with Not_found -> () in
@@
 
-We describe will the following aspects, generally in
+We will describe the following aspects, generally in
 order of increasing complexity: domain (of the group `Abstract
@@
 Fig.~\ref{fig:GE-algorithm}.
-\begin{code}
+\begin{code}[commandchars=\\\{\}]
 module GenericVectorContainer(Dom:DOMAINL) =
@@
 % \jacques{Actually, I liked it, so after slight editing, I kept it}
-It is instructive the examine the difference in the return types of the
+It is instructive to examine the difference in the return types of the
 \texttt{decl}, \texttt{upd\_sign} and \texttt{zero\_sign} functions. The first
@@
 The function \texttt{upd\_sign} is a generator that may produce code, or may
-not. The option type lets us avoid generating pointless code such as 
-\texttt{a := b; ()}.
+not. The option type lets us avoid generating code such as 
+\texttt{a := b; ()} with a pointless |()|.
 
@@
           if ((fst (fst i_16)) <> t_13) then begin (* swapping of rows *)
-           let a_17 = t_4.arr
-           and m_18 = t_4.m in
-           let i1_19 = (t_13 * m_18)
-           and i2_20 = ((snd (fst i_16)) * m_18) in
+           let a_17 = t_4.arr and m_18 = t_4.m in
+           let i1_19 = (t_13 * m_18) and i2_20 = ((snd (fst i_16)) * m_18) in
            for i_21 = 0 to (m_18 - 1) do

### 2008-06-21 SCP2008/paper/reply-to-reviews.txt
draft reply to the reviewers

	We are very grateful to the reviewers for their comments and
so helpful suggestions. We have implemented them all. Below is the
detailed response (omitting suggestions about spelling and grammar,
which we have corrected as recommended).

Reviewer #1:

The summary of our work given by the reviewer is profound; we couldn't
have said better ourselves.

Reviewer #2:

> Two larger changes I suggest are restructuring section 4 and performing
> stronger evaluation of the work.
We have implemented them, as described below.

> Restructuring section 4
> The subsections of section 4 are rather random and disjoint. There
> needs to be a more cohesive plan about what is going on in this
> section. Better yet a concrete example that gets built upon in each
> subsection and motivates the subsection's functionality. Sections 4.1,
> 4.3, 4.5, 4.6 seem to be about defining modules for particular
> aspects. Part of Section 4.2 seems superfluous from the comment
> above. The "DOMAIN" module seems superfluous too: you never need to do
> real operations on values, but only generate code for operations on
> values. Hence "DOMAINL" is what you need and "DOMAIN" is not
> needed. Also I don't think "CONTAINER2D" is ever seen again after
> being introduced in Section 4.2 -- this seems troubling.

We added a paragraph at the end of the container section clarifying
that the `current' container is available to the algorithmic aspects,
under the name of |C|. We describe in more detail the `multi-stage'
nature of instantiating the generic algorithm, pointing out which
module is a part of which.

> Section 4.4 is about programming in OCaml and how to maintain the
> algorithm's state.  Section 4.7 is about putting it all together.

> From my understanding of your work here is an idea to make your paper more
> precise and concise.
>  - Move Section 4.4 into Section 3 because that section seems to deal with
>    low-level programming issues.
Moved.

> - In the intro of Section 4 make a link to the list in Section 2 and say your
>   going to pick some of those aspects and explain how they are implemented -
>   one per subsection.
Done

> - Delete the first half of 4.2.
> - Combine 4.1 with the second half of 4.2, dropping the use of "DOMAIN" and
>   just do everything in terms of "DOMAINL".

It seems DOMAIN is a nice and simple abstraction to build upon later;
it also corresponds to our introductory example in Sec 2. We have
completely re-written Sec 4.2 (which is now Sec 5.2), but decided to 
leave DOMAIN in.

> - Sections 4.3, 4.5, 4.6 then become Sections 4.2, 4.3, 4.4.
> - Section 4.7 could be the last subsection of 4 
> or a new whole section with more detail.

This plan has being followed through. Hopefully the introduction to
(now) Sec 5 describes the cohesive plan of the section.

> Strengthening evaluation
> I understand that program generators are complex programs, that
> explaining them can be difficult and that providing concise and
> meaningful code snippets about them can also be hard. Nevertheless a
> good paper on programming program generators needs to do these
> things. At the end of your paper I thought you had laid down all the
> pieces of the machinery well, but I didn't have much of a sense of how
> the machine runs. So I kind of have to take your word that everything
> works, especially since the code on p24-5 is not explained in detail
> and the appendix code is not readable (by itself it's not very
> convincing: how can I tell this code demonstrates something good?). In
> general, and this would strengthen the merit of the paper, code
> snippets need to be explained better - too often I see a new
> identifier in the code that is not explained.

The comparison of the generated code with the hand-written code
hopefully tells now that the generated code does do the right
thing. We have expanded many explanations for the code snippets,
explicitly relating them to the hand-written sample code -- at the
expense of the length of the paper.

> Showing and explaining the code for the unstaged algorithm might help
> and would give a point of reference to compare the generated code
> with. Explaining the generated code could also be helpful, but maybe
> too hard. Maybe it would be possible to focus on one or more small
> areas of the algorithm and show: this is what the area looks like in
> unstaged program, this is the part of the code generator responsible
> for producing code for this area, and this is the area in the
> generated code. Something like this would give the reader a greater
> understanding of how logic flows through your generation process and
> how your functor technique is ultimately in control of fine-grained
> details of code building.

We have done exactly that. We added Sec 4 with the hand-written
unstaged code, and refer to it when discussing various aspects,
especially the determinant aspect.

> p23, "off-shoring": give relevant citation
Done. Incidentally, Walid Taha spells offshoring without the hyphen,
so we have adjusted the spelling.

> p2, line 1, For an introduction paragraph this could be clearer and stronger.
> What are the "aspects of linear algebra"? How are they addressed and by what?
> (I know they are explained later on.) Why is "may" emphasized?
We re-wrote the paragraph.

> p2, line 6, LU decomposition is a particular case of Gaussian Elimination, not
> vice-versa
We have made that clear all throughout the text (noting that it is
quite common, especially in the naming of library functions, to treat
GE and LU as quasi-synonyms)

> p2, line 10, this paper doesn't not discuss
> hardware-architecture-specific optimization, in fact native code
> generation seems beyond the scope of this paper, I think all you want
> to say is that generative programming allows the programmer to express
> domain-specific optimizations
Yes, we did

> p4, "The design space", this decomposition of the design space is very
> interesting and an important result, but it would be good if this
> section was more compact. E.g., maybe present the aspects in their
> groups straight up instead of two lists.
That is true. However, as this is important, it needs to
be explained in detail, so that all the relevant information is recorded,
in writing. We do consider this identification of aspects as one of
the contributions.

> p10, Figure 1, I don't follow the intuition behind naming the
> operation "retN"; I don't see how the operation is similar to ordinary
> monadic "ret", or what the "N" means; wouldn't a name like "letM" be
> better?
We have added the explanation of the name (on p11 now). Also we chose
retN for consistency with Swadi et al, where RetN was first
introduced.

> p10, Figure 1, showing the type of "retN" and maybe some other values
> would be helpful
Done for ret, retN and bind.

> p10, the type variable 'p and its constraint: could you explain the
> problem with classifiers in more detail? if the type was "type ('s,
> 'w, 'v) monad" what's wrong with having types like "('a state, 'w,
> ('a, 'v) code) monad"?
There is more explanation at the end of p11 and beginning of
p12. There is nothing wrong, in principle, with the type 
"('a state, 'w,  ('a, 'v) code) monad" other than the fact the types
become bigger and bigger; furthermore, only a couple of components of
the type are really needed when giving a type signature.

> p14, line 4, don't you want to assert that "ZpMake" is a "DOMAIN" as was done
> for "IntegerDomain"?
Actually no: we omitted the annotation on purpose, to demonstrate that
we can do it. This is further explained in the text (the end of p18).

> p15, shouldn't the code here appear as the "code type" implementation
> of an "abstract code" module?
> p15, what are "Logic" and "Idx" used for?
> p16, what is the point of the "S" module? why can't "DOMAINL" be
> parameterized by "T" (containing "rep") directly?
> p16, Section 4.2, none of the abstract code combinators were used in
> the second half of this section: was the first half necessary? what is
> the relationship between these two halves that warrants them being in
> the same subsection?  p16, Section 4.2, I understand the purposes of
> abstracting code representation.  Additionally it adds to the
> genericity of the code generator, which is good, but doesn't incur any
> abstraction overhead, which is also good. However for the purposes of
> this paper it is perhaps an abstraction too far: the paper never needs
> to generate anything other than MetaOCaml code and it increases the
> complexity of code snippets. Mentioning that you do abstract code
> representation in the real implementation but elide it for the paper
> might make the paper more succinct and clearer.

The section on abstract code and lifted domain has been completely
re-written, hopefully the above questions are all answered, in
particular about the meaning of Logic and Idx and the merits of
abstracting over interpretations.

> p17, line 4, where does the identifier "rep" come from in this code?
> p17, line -2, what accompanying code?
> p18, line 4, what is the "decl" function?
> p19, showing the types of these functions would be helpful
> p19, line -3, what is the "DETERMINANT" module? it hasn't been introduced yet
> p20, what are "cmonad" and "omonad"?

Sec 5.2 has been completely re-written; hopefully it now answers all
these questions,

> p17, line 4, what is the design decision for making "Dom" a definition
> inside "CONTAINER2D" instead of a parameter of "CONTAINER2D"? For
> readers not familiar with the OCaml module system it might be helpful
> to provide a brief introduction somewhere.
> p17, line -12, now in the module implementation "Dom" is made functor
> parameter, why? and why can you just replace "rep" with "code"?
We added a new paragraph to explain this, the last but one paragraph 
of Sec 5.3.

> p18, although most readers are probably familiar with simple LU
> decomposition, a brief overview that shows how determinant and rank
> are computed in the algorithm with an unstaged code example showing
> the state needed for tracking would be helpful to demonstrate the
> problems stated here
We added a new section, Sec 4.

> p20, shouldn't there be an "answer" method in the state objects?
This is now covered by proper definition of cmonad/omonad.

> p19, line -2, why do you need to track the determinant for integer
> matrices
We explained that in the new Sec 4.

> p20, what is "classif" used for? initially I thought it meant "class
> if", but I just worked out it means "classifier", so this relates to
> an earlier comment, so I think it is important you explain in detail
> why you need to put the classifier in this object.
That is now explained on p11.

> p20, line 5, "e.g., a piece of code...": the type of "lstate" depends on the
> implementation and neither of the implementations shown set "lstate" to "v
> ref".
AbstractDet contains a tdet, which is defined to be v ref
one line above which, by the open statement right above it, it Dom.v ref.

> p21, keep the names of the sign and magnitude(?) state consistent
> across "decl", "upd_sign" and "fin"; also keep the code to retrieve
> the sign and magnitude state consistent across "upd_sign" and "fin"
We have done that.

> p20, what do "acc", "get", "set" do?
Changed.

> p21, what is "wmatrix"? shouldn't this be an instance of "CONTAINER2D"?
It is now defined. We explained its differences from the CONTAINER2D
(the very end of p29).

> p21, "(...,res) monad" what does "..." mean?
We made it more precise.

> p21, the "unitL" in "decl" is superfluous, right? "mo_extend" will
> already return a monadic unit result
Actually no. mo_extend has the result type ('a,unit) monad whereas
unitL has the result type ('a,unit) cmonad, which corresponds to
('a,('c,unit) code) monad. In other words, the former describes an
action that generates no code whereas the latter is an action that
generates code (such as an assignment operation) that has the type
unit. However, we did mean that decl is an action to be executed
solely for its side effects, and it was used as such in the code (its
result was disregarded). So, the type ('a,unit) monad for it is far
better. We have changed the code and the paper to reflect that. There
is no longer `superfluous' unitL. Thank you for making this point.

> p23, what does "GenLA" do?
It is explained now, p30.

> p23, it would be nice to see the module signatures for the other
> features not discussed thus far
We added a new section, Sec 5.5, with the description of the other
features and their signatures.  Perhaps the code can be moved to an
appendix though?

> p24, missing "perform" in "init"
Actually, it is on the first line.

> p24, "module O = F.Output(F)" the functor "Output" takes a "OUTPUTDEP" not a
> "FEATURES", right?
Right, we explained that in more detail (see the paragraph before code
on p32).

### 2008-06-23 SCP2008/paper/metamonads.tex
take care of the comments I got from Pouya

@@
 A survey \cite{Carette06} of
-Gaussian elimination implementations in the industrial package Maple
+Gaussian Elimination implementations in the industrial package Maple
 \cite{Monagan:2001:M7P}
@@
 in using code generation for scientific (both numeric and symbolic)
-software.  We will use the algorithm family of Gaussian elimination, 
+software.  We will use the algorithm family of Gaussian Elimination, 
 applied to perform LU decomposition 
@@
     \item Provide a thorough classification of the family of Gaussian
-    elimination algorithms.
+    Elimination algorithms.
 \end{itemize}
@@
 Throughout, we assume that the reader is familiar with the basic LU
-decomposition algorithm, which factors a invertible matrix $A$ into a unit
+decomposition algorithm, which factors an invertible matrix $A$ into a unit
 lower triangular matrix $L$ and (usually) an upper triangular matrix $U$,
@@
       `mathematical' aspects: domain of matrix elements, length
-      measure, full division, normalization, zero testing,
+      measure, full division, zero testing,
       symmetry and other matrix structure.
@@
       matrices. This group also includes packing, in-place
-      decomposition, and the
+      decomposition, normalization, and the
       representation of sparse matrices.
@@
 the computation at the generating stage.  We need a combinator that
-inserts the |let| expression in the generat\emph{ed} code, in a other words a
+inserts the |let| expression in the generat\emph{ed} code, in other words a
 combinator |letgen| to be used as
@@
 \end{code}
-\caption{Fraction-free in-place Gaussian-elimination over an integer
+\caption{Fraction-free in-place Gaussian Elimination over an integer
   matrix}\label{fig:GE-algorithm}
@@
 a separately-defined function |pivot| to find a pivot. Here we use
-full-pivoting, i.e. searching the completel as yet unexamined portion of the
+full-pivoting, i.e. searching the complete as yet unexamined portion of the
 matrix, the rectangular block |(r,c)-(n,m)|, for the element with the 
@@
 represented by a flat vector, full pivoting, returning the |U|-factor,
-the determinant and the rank.
+the determinant and the rank.  The comments, however, were inserted by
+hand.
 \begin{code2}

### 2008-06-23 SCP2008/paper/metamonads.tex
include .bbl directly into document for submission

@@
 
+% \bibliography{metamonads}
+% \bibliographystyle{elsart-num}
+
 % Embed .bbl file directly to make document self-contained
-\bibliography{metamonads}
-\bibliographystyle{elsart-num}
-
-%% \begin{thebibliography}{10}
-%% \expandafter\ifx\csname url\endcsname\relax
-%%   \def\url#1{\texttt{#1}}\fi
-%% \expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
+\begin{thebibliography}{10}
+\expandafter\ifx\csname url\endcsname\relax
+  \def\url#1{\texttt{#1}}\fi
+\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
+
+\bibitem{Carette06}
+J.~Carette, Gaussian {E}limination: a case study in efficient genericity with
+  {MetaOCaml}, Science of Computer Programming 62~(1) (2006) 3--24, special
+  Issue on the First MetaOCaml Workshop 2004.
+
+\bibitem{Monagan:2001:M7P}
+M.~B. Monagan, K.~O. Geddes, K.~M. Heal, G.~Labahn, S.~M. Vorkoetter,
+  J.~McCarron, P.~DeMarco, Maple 7 Programming Guide, Waterloo Maple Inc.,
+  2001.
+
+\bibitem{Gruntz:1994:IG}
+D.~Gruntz, M.~Monagan, Introduction to {Gauss}, \textsc{SIGSAM Bulletin}:
+  Communications on Computer Algebra 28~(2) (1994) 3--19.
+
+\bibitem{Axiom}
+R.~D. Jenks, R.~S. Sutor, {AXIOM}: The Scientific Computation System, Springer
+  Verlag, 1992.
+
+\bibitem{Kennedy01Telescoping}
+K.~Kennedy, B.~Broom, K.~Cooper, J.~Dongarra, R.~Fowler, D.~Gannon,
+  L.~Johnsson, J.~Mellor-Crummey, L.~Torczon, Telescoping languages: A strategy
+  for automatic generation of scientific problem-solving systems from annotated
+  libraries, Journal of Parallel and Distributed Computing 61~(12) (2001)
+  1803--1826.
+
+\bibitem{Veldhuizen:2004}
+T.~L. Veldhuizen, Active libraries and universal languages, Ph.D. thesis,
+  Indiana University Computer Science (May 2004).
+\newline\urlprefix\url{http://osl.iu.edu/~tveldhui/papers/2004/dissertation.pd%
+f}
+
+\bibitem{scp-CohenDGHKP06}
+A.~Cohen, S.~Donadio, M.~J. Garzar{\'a}n, C.~A. Herrmann, O.~Kiselyov, D.~A.
+  Padua, In search of a program generator to implement generic transformations
+  for high-performance computing, Sci. {C}omput. {P}rogram 62~(1) (2006)
+  25--46.
+
+\bibitem{Czarnecki}
+K.~Czarnecki, U.~W. Eisenecker, Generative programming: methods, tools, and
+  applications, ACM Press/Addison-Wesley Publishing Co., 2000.
+
+\bibitem{Veldhuizen:1998:ISCOPE}
+T.~L. Veldhuizen, Arrays in {Blitz++}, in: Proceedings of the 2nd International
+  Scientific Computing in Object-Oriented Parallel Environments (ISCOPE'98),
+  Lecture Notes in Computer Science, Springer-Verlag, 1998, pp. 223--230.
+
+\bibitem{musser89generic}
+D.~R. Musser, A.~A. Stepanov, Generic programming, in: {ISSAC 1988}:
+  Proceedings of the International Symposium on Symbolic and Algebraic
+  Computation, Vol. 358 of Lecture Notes in Computer Science, Springer-Verlag,
+  1989, pp. 13--25.
+
+\bibitem{musser94algorithmoriented}
+D.~R. Musser, A.~A. Stepanov, Algorithm-oriented generic libraries, Software -
+  Practice and Experience 24~(7) (1994) 623--642.
+
+\bibitem{BOOST}
+J.~Siek, L.-Q. Lee, A.~Lumsdaine, The Boost Graph Library: User Guide and
+  Reference Manual, Addison-Wesley, 2002.
+
+\bibitem{POOMA}
+I.~John V.W.~Reynders, J.~C. Cummings, The {POOMA} framework, Comput. Phys.
+  12~(5) (1998) 453--459.
+
+\bibitem{ATLAS}
+R.~C. Whaley, A.~Petitet, J.~J. Dongarra, Automated empirical optimization of
+  software and the {ATLAS} project, Parallel Computing 27~(1--2) (2001) 3--35.
+
+\bibitem{HygienicMacros}
+E.~E. Kohlbecker, D.~P. Friedman, M.~Felleisen, B.~F. Duba, Hygienic macro
+  expansion, in: LISP and Functional Programming, 1986, pp. 151--161.
+
+\bibitem{camlp4}
+D.~de~Rauglaudre, Camlp4 reference manual,
+  \url{http://caml.inria.fr/camlp4/manual/} (Jan. 2002).
+
+\bibitem{conf/dagstuhl/CzarneckiOST03}
+K.~Czarnecki, J.~T. O'Donnell, J.~Striegnitz, W.~Taha, {DSL} implementation in
+  {MetaOCaml}, {Template Haskell}, and {C++}, in: C.~Lengauer, D.~S. Batory,
+  C.~Consel, M.~Odersky (Eds.), Domain-Specific Program Generation, Vol. 3016
+  of Lecture Notes in Computer Science, Springer, 2003, pp. 51--72.
+
+\bibitem{CTHL03}
+C.~Calcagno, W.~Taha, L.~Huang, X.~Leroy, Implementing multi-stage languages
+  using asts, gensym, and reflection, in: F.~Pfenning, Y.~Smaragdakis (Eds.),
+  GPCE, Vol. 2830 of Lecture Notes in Computer Science, Springer, 2003, pp.
+  57--76.
+
+\bibitem{metaocaml-org}
+{MetaOCaml}, \url{http://www.metaocaml.org}.
+
+\bibitem{TahaSheard97}
+W.~Taha, T.~Sheard, Multi-stage programming with explicit annotations, in:
+  Proceedings of the Symposium on Partial Evaluation and Semantic-Based Program
+  Manipulation {(PEPM)}, ACM, Amsterdam, 1997, pp. 203--217.
+
+\bibitem{TahaThesis}
+W.~Taha, Multi-stage programming: its theory and applications, Ph.D. thesis,
+  Oregon Graduate Institute of Science and Technology (1999).
+
+\bibitem{Taha2000}
+W.~Taha, A sound reduction semantics for untyped {CBN} multi-stage computation.
+  {Or}, the theory of {MetaML} is non-trival, in: PEPM, 2000, pp. 34--43.
+
+\bibitem{Pueschel:05}
+M.~P{\"u}schel, J.~M.~F. Moura, J.~Johnson, D.~Padua, M.~Veloso, B.~W. Singer,
+  J.~Xiong, F.~Franchetti, A.~Ga\v{c}i\'{c}, Y.~Voronenko, K.~Chen, R.~W.
+  Johnson, N.~Rizzolo, {SPIRAL}: Code generation for {DSP} transforms,
+  Proceedings of the IEEE, special issue on "Program Generation, Optimization,
+  and Adaptation" 93~(2).
+
+\bibitem{dongarra7}
+Z.~Chen, J.~Dongarra, P.~Luszczek, K.~Rothe, Lapack for clusters project: An
+  example of self adapting numerical software, Hawaii International Conference
+  on System Sciences HICSS-37.
+
+\bibitem{KiselyovTaha}
+O.~Kiselyov, K.~N. Swadi, W.~Taha, A methodology for generating verified
+  combinatorial circuits, in: EMSOFT '04: Proceedings of the fourth ACM
+  international conference on Embedded software, ACM, 2004, pp. 249--258.
+
+\bibitem{CaretteKiselyov05}
+J.~Carette, O.~Kiselyov, Multi-stage programming with {F}unctors and {M}onads:
+  Eliminating abstraction overhead from generic code, in: Gl{\"u}ck and Lowry
+  \cite{DBLP:conf/gpce/2005}, pp. 256--274.
+
+\bibitem{SwadiMonadic06}
+K.~N. Swadi, W.~Taha, O.~Kiselyov, E.~Pasalic, A monadic approach for avoiding
+  code duplication when staging memoized functions, in: J.~Hatcliff, F.~Tip
+  (Eds.), PEPM, ACM, 2006, pp. 160--169.
+
+\bibitem{metamonadsURL}
+{Source code}, \url{http://www.cas.mcmaster.ca/~carette/metamonads/}.
+
+\bibitem{Golub-vanLoan}
+G.~H. {Golub}, C.~F. {Van Loan}, Matrix Computations, 3rd Edition, Johns
+  Hopkins University Press, Baltimore, MD, 1996.
+
+\bibitem{kiczales97aspectoriented}
+G.~Kiczales, J.~Lamping, A.~Menhdhekar, C.~Maeda, C.~Lopes, J.-M. Loingtier,
+  J.~Irwin, Aspect-oriented programming, in: M.~Ak\c{s}it, S.~Matsuoka (Eds.),
+  Proceedings European Conference on Object-Oriented Programming, Vol. 1241,
+  Springer-Verlag, 1997, pp. 220--242.
+
+\bibitem{709568}
+J.~Irwin, J.-M. Loingtier, J.~R. Gilbert, G.~Kiczales, J.~Lamping,
+  A.~Mendhekar, T.~Shpeisman, Aspect-oriented programming of sparse matrix
+  code, in: ISCOPE '97: Proceedings of the Scientific Computing in
+  Object-Oriented Parallel Environments, Springer-Verlag, London, UK, 1997, pp.
+  249--256.
+
+\bibitem{mendhekar97rg}
+A.~Mendhekar, G.~Kiczales, J.~Lamping, {RG}: {A} case-study for aspect-oriented
+  programming, Tech. Rep. SPL97-009 P9710044, {X}erox {PARC}, Palo Alto, CA,
+  USA (February 1997).
+
+\bibitem{journals/cacm/parnas72a}
+D.~L. Parnas, On the criteria to be used in decomposing systems into modules,
+  Commun. ACM 15~(12) (1972) 1053--1058.
+
+\bibitem{journals/tse/Parnas76}
+D.~L. Parnas, On the design and development of program families, IEEE Trans.
+  Software Eng. 2~(1) (1976) 1--9.
+
+\bibitem{EWD:EWD447}
+E.~W. Dijkstra, On the role of scientific thought, published as
+  \cite{EWD:EWD447pub}.
+\newline\urlprefix\url{http://www.cs.utexas.edu/users/EWD/ewd04xx/EWD447.PDF}
+
+\bibitem{ZhCaJeMo06a}
+W.~Zhou, J.~Carette, D.~J. Jeffrey, M.~B. Monagan, Hierarchical representations
+  with signatures for large expression management, in: Proceedings of
+  Artificial Intelligence and Symbolic Computation, Vol. 4120 of Lecture Notes
+  in Computer Science, 2006, pp. 254--268.
+
+\bibitem{Bondorf:92}
+A.~Bondorf, Improving binding times without explicit {CPS}-conversion, in: 1992
+  ACM Conference on Lisp and Functional Programming. San Francisco, California,
+  1992, pp. 1--10.
+
+\bibitem{conf/pepm/BondorfD94}
+A.~Bondorf, D.~Dussart, Improving {CPS-}based partial evaluation: Writing cogen
+  by hand, in: {ACM} {SIGPLAN} Workshop on Partial Evaluation and
+  Semantics-Based Program Manipulation, Orlando, Florida, 1994, pp. 1--9.
+
+\bibitem{Haskell98Report}
+S.~Peyton~Jones, et~al., The Revised Haskell 98 Report, Cambridge Univ.\null{}
+  Press, 2003, also on \url{http://haskell.org/}.
+
+\bibitem{moggi-notions}
+E.~Moggi, Notions of computation and monads, Information and Computation 93~(1)
+  (1991) 55--92.
+
+\bibitem{Filinski:Representing}
+A.~Filinski, Representing monads, in: POPL, 1994, pp. 446--457.
+
+\bibitem{liang-interpreter}
+S.~Liang, P.~Hudak, M.~Jones, Monad transformers and modular interpreters, in:
+  {POPL} '95: Conference Record of the Annual {ACM} Symposium on Principles of
+  Programming Languages, {ACM} {P}ress, {N}ew {Y}ork, 1995, pp. 333--343.
+
+\bibitem{taha-environment}
+W.~Taha, M.~F. Nielsen, Environment classifiers, in: {POPL} '03: Conference
+  Record of the Annual {ACM} Symposium on Principles of Programming Languages,
+  {ACM} {P}ress, {N}ew {Y}ork, 2003, pp. 26--37.
+
+\bibitem{mlton-proplist}
+{{MLton} team}, Property list, \url{http://mlton.org/PropertyList}.
+
+\bibitem{BurdenFaires89}
+R.~L. Burden, J.~D. Faires, Numerical analysis: 4th ed, PWS Publishing Co.,
+  Boston, MA, USA, 1989.
+
+\bibitem{Bareiss68}
+E.~H. Bareiss, {S}ylvester's identity and multistep {G}aussian {Elimination},
+  Math. Comput. 22~(103) (1968) 565--579.
+
+\bibitem{offshoring}
+J.~Eckhardt, R.~Kaiabachev, E.~Pasalic, K.~N. Swadi, W.~Taha, Implicitly
+  heterogeneous multi-stage programming, in: Gl{\"u}ck and Lowry
+  \cite{DBLP:conf/gpce/2005}, pp. 275--292.
+
+\bibitem{Watt:2002:HCA}
+S.~M. Watt, Aldor, in: J.~Grabmeier, E.~Kaltofen, V.~Weispfennig (Eds.),
+  Computer Algebra Handbook: Foundations, Applications, Systems, Springer
+  Verlag, 2003, pp. 265--270.
+
+\bibitem{Gluck95}
+R.~Gl{\"u}ck, R.~Nakashige, R.~Z{\"o}chling, Binding-time analysis applied to
+  mathematical algorithms, in: System Modelling and Optimization, Chapman and
+  Hall, 1995, pp. 137--146.
+
+\bibitem{GluckJ97}
+R.~Gl{\"u}ck, J.~J{\o}rgensen, An automatic program generator for multi-level
+  specialization, Lisp and Symbolic Computation 10~(2) (1997) 113--158.
+
+\bibitem{edsl}
+P.~Hudak, Building domain-specific embedded languages, ACM Comput. Surv.
+  28~(4es) (1996) 196.
+
+\bibitem{WhenHowDSL}
+M.~Mernik, J.~Heering, A.~M. Sloane, When and how to develop domain-specific
+  languages, ACM Comput. Surv. 37~(4) (2005) 316--344.
+
+\bibitem{DBLP:conf/gpce/2005}
+R.~Gl{\"u}ck, M.~R. Lowry (Eds.), Generative Programming and Component
+  Engineering, 4th Internation Conference, GPCE 2005, Tallinn, Estonia,
+  September 29 - October 1, 2005, Proceedings, Vol. 3676 of Lecture Notes in
+  Computer Science, Springer, 2005.
+
+\bibitem{EWD:EWD447pub}
+E.~W. Dijkstra, On the role of scientific thought, in: Selected Writings on
+  Computing: A Personal Perspective, Springer-Verlag, 1982, pp. 60--66.
+
+\end{thebibliography}
 

### 2008-08-13 tagless/tagless-final.tex
mostly some minor wording changes (in recent edits) and adding a comment

@@
 Unlike \citearound{'s HOAS-based term representation}\citet{JFP-Mogensen},
-our solution is \emph{not} Church-encoding the
+our solution does \emph{not} involve Church-encoding the
 universal type. The Church encoding of the type~|u| in \S\ref{tagproblem}
@@
 |C.int| is used to convert from the static term |('c,int) R.repr|, 
-which is just |int|, to the dynamic term. We can now define the following
+which is just |int|, to a dynamic term. We can now define the following
 components of module |P0| required by the |Semantics| signature:
@@
 the static term to dynamic, but there is no polymorphic ``lift''
-function, of type |'a -> ('c,'a) C.repr|, to send a value to the future stage
+function, of type |'a -> ('c,'a) C.repr|, to send a value to a future stage
 \citep{xi-guarded,WalidPOPL03}.
@@
 booleans, integers, and functions may well be represented differently.
-Thus, compiling polymorphic lift requires intensional type
+Compiling a polymorphic lift function thus requires intensional type
 analysis.  To avoid needing polymorphic lift, we turn to
@@
 performs case distinction on the source type. In Haskell, typecase
-can be equivalently implemented either with GADTs or with
+can be implemented using either GADTs or
 type-class functional dependencies
@@
 approaches (|Incope.hs| and |incope1.hs|), 
-neither portable to OCaml. In addition,
-the problem of nonexhaustive pattern\hyp matching reappears in
+neither of which is portable to OCaml. In addition,
+the problem of non\hyp exhaustive pattern\hyp matching reappears in
 the GADT approach because GHC 6.6.1 cannot see that a particular
@@
 make it \emph{syntactically} apparent that pattern matching is exhaustive.
+\jacques{what about GHC 6.8?}
 
@@
 methods of |Symantics|.
-A simple and portable solution thus emerges: we bake |repr_pe| 
-into the signature |Symantics|. 
+A simple and portable solution\footnote{though somewhat long-winded} thus
+emerges: we bake |repr_pe| into the signature |Symantics|. 
 We recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that 
@@
 with the rest the same.
-In contrast, the partial evaluator~|P| relies on the type argument |'sv|.
+Predictably, the partial evaluator~|P| crucially relies on the type argument
+|'sv|.
 
@@
 We turn to a more general approach to CBV CPS: a CPS transformer that
-turns any implementation of |Symantics| into a CPS interpreter, whether
-it is an evaluator.\oleg{I don't understand this whether
-  classification: any interpreter is an evaluator, no? It all depends
-  on what the domain is. Anyway, the reviewer suggests rewriting the
-  clause as ``whether or not it is an evaluator''. I suggest to drop
-  it entirely.}
+turns any implementation of |Symantics| into a CPS version of that
+evaluator.
 This functor on interpreters performs a textbook
@@
 \Citet{Birkedal-PE-ML} have hand-written a compiler generator (cogen)
-for the Core Standard ML, which transforms a program into its tagless
+for Core Standard ML, which transforms a program into its tagless
 generating extension (and therefore serves as a tagless partial
@@
 \emph{early} it becomes \emph{manifest} to the metalanguage
-implementation that our code is safe.  The fact that
+that our code is safe.  The fact that
 \Citet{Birkedal-PE-ML} 's generating extension never generates a
@@
 compiler until the SML compiler sees each particular specialized
-program.  Moreover and analogously, the fact that
+program.  Moreover, and in a similar vein, the fact that
 \Citet{Birkedal-PE-ML} 's cogen never generates a generating extension

### 2008-08-13 tagless/tagless-final.tex
for completeness, add a Functor version of the CPS code, we might want to refer to it

@@
 the answer type.
+\begin{code}
+module RCN(W : sig type w end)= struct
+  type ('c,'sv,'dv) repr = ('sv -> W.w) -> W.w
+  ...
+\end{code}
+This allows a translation of these techniques to, for example, Standard ML.
+Unfortunately, the loss of polymorphism in the answer type makes the 
+CPS evaluator less pleasant to use than the other implementations.
 

### 2008-08-14 tagless/tagless-final.tex
Various changes -- I'm in the middle of clarifying the self-interpretation section

@@
 %%   http://www.itu.dk/people/birkedal/papers/paresm.ps.gz
-%% and thank Eijiro Sumii
 
@@
 \ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
-using compiler-generator (cogen) functions rather than data constructors.
+using combinator functions rather than data constructors.
 In other words, we represent object terms not in an initial algebra
@@
 factorial, and power.
+\begin{comment}
 \oleg{There is now IncopeDB.hs that defines at least R, L and C
- interpreters (no PE at present though) using DeBruijn indices. So we
- can say that (some of) our development is available in DeBruijn
+ interpreters (no PE at present though) using de Bruijn indices. So we
+ can say that (some of) our development is available in de Bruijn
  style. We won't talk about it in the paper though.}
-In contrast to \S\ref{intro}, we encode binding using higher-order
+\end{comment}
+In contrast to \S\ref{intro}, we use higher-order
 abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
-rather than de Bruijn indices. This makes the encoding convenient and
+rather than de Bruijn indices to encode binding and
 ensures that our object programs are closed.
+We find HOAS to be more convenient, though we could have continued to
+use de Bruijn indices.
 
@@
 using higher-rank polymorphism through OCaml record types.
-It could also be a functor parameterized over
+To avoid this higher-rank polymorphism in the core language, which
+Standard ML does not support, we could also define |RCN| as
+a functor parameterized over
 the answer type.
@@
 \end{code}
-This allows a translation of these techniques to, for example, Standard ML.
+This allows a translation of these techniques to, for example, Standard ML\@.
 Unfortunately, the loss of polymorphism in the answer type makes the 
@@
 (phantom) type argument.
+The verbosity of these types is the only difficulty in defining a
+new signature to replace |Symantics| that the output of |CPST| matches.
 
@@
   open M
-  let test1 () = (* same as before *)
-       app (lam (fun x -> x)) (bool true)
+  let test1 () =
+       app (lam (fun x -> x)) (bool true)  (* same as before *)
   let testpowfix () = ...                  (* same as before *)
@@
 as the accumulator.
-Because the |SymSI| signature extends |Symantics|, any encoding of
-a term in the pure object language (that is, any functor that takes
-a |Symantics| module as argument) can also be used as a term in the
-extended object language (for example, applied to an implementation of
-|SymSI|).
 
@@
 
+Because the |SymSI| signature extends |Symantics|, any encoding of
+a term in the pure object language (that is, any functor that takes
+a |Symantics| module as argument) can also be used as a term in the
+extended object language (for example, applied to an implementation of
+|SymSI|).  In particular,
 |RCPSI| matches the |Symantics| signature and implements the
@@
 
-It is worthwhile reiterating how close this implementation is to the CPS
-interpreter, and that adding state and imperative features needed no new
+We reiterate that this implementation adding state and imperative
+features is very close to the CPS interpreter and uses no new
 techniques.
@@
 \emph{optimal} with respect to~|si| \citep{jones-challenging} if $\pe(\si\Encode{e})$
-is equal to~$e$ (up to $\alpha$\hyp conversion, or in some accounts, no
+is $\alpha$\hyp equivalent to~$e$ (or in some accounts, no
 less efficient than~$e$).
 
-Intuitively, self\hyp interpretation is straightforward in our
-framework: the functions comprising the interpreters
-in~\S\ref{S:interpreter-RL} may just as well be written in our object
-language:~Fig.~\ref{fig:self-eval}.
+Self\hyp interpretation in our framework is straightforward at the
+term level: the functions comprising the interpreters
+in~\S\ref{S:interpreter-RL} may as well be written in our object
+language, as in Fig.~\ref{fig:self-eval}.
 \begin{figure}
@@
 We call this mapping \emph{pre-encoding}.
-\begin{align*}
+\begin{equation}
+\begin{split}
     \encode{x} &= x \\
     \encode{n} &= \ident{int}\, n \\
+    \encode{\fun{x}e} &= \ident{lam} (\fun{x} \encode{e}) \\
     \encode{e_1 + e_2} &= \ident{add} \encode{e_1} \encode{e_2} \\
     \encode{\cond{b}{t}{e}} &= \rlap{$\ident{if\_} \encode{b}
-        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$}
-        \displaybreak[0] \\
-    \encode{\fun{x}e} &= \ident{lam} (\fun{x} \encode{e}) \\
+        \left(\fun{\_}\encode{t}\right) \left(\fun{\_}\encode{e}\right)$} \\
     \encode{fx} &= \ident{app} \encode{f} \encode{x} \\
     \encode{\fix{f}e} &= \ident{fix} (\fun{f} \encode{e})
-\end{align*}
+\end{split}
+\end{equation}
 The metavariables $x$ and~$n$ stand for a variable and an integer,
@@
 example, the familiar object term $(\fun{x}x)\True$ pre-encodes to
-\begin{equation*}
+\begin{equation}
     \encode{(\fun{x}x)\True} = \ident{app}
-    (\ident{lam} (\fun{x} x))\, (\ident{bool} \True),
-\end{equation*}
+    (\ident{lam} (\fun{x} x))\, (\ident{bool}\, \True),
+\end{equation}
 and to evaluate this pre-encoded term is to evaluate the object term
-\begin{equation*}
+\begin{equation}
     (\fun{f} \fun{x} fx) \,
     ((\fun{f} f) (\fun{x} x))\, ((\fun{b} b) \True).
-\end{equation*}
+\end{equation}
 Because the evaluator in Fig.~\ref{fig:self-eval} mostly consists 
@@
 the example term above, we could try to define
-\begin{equation*}
+\begin{equation}
     \Encode{e} =
     \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}} \encode{e},
-\end{equation*}
+\end{equation}
 in particular
-\begin{equation*}
+\begin{equation}
     \Encode{(\fun{x}x)\True} =
     \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
-    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool} \True).
-\end{equation*}
-To type-check this encoded term, we give the variable $\ident{lam}$ the
+    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool}\, \True).
+\end{equation}
+To type-check this encoded term, we give the bound variable $\ident{lam}$ the
 simple type $(\BB\to\BB)\to\BB\to\BB$.
 We then define the self\hyp interpreter
-\begin{equation*}
+\begin{equation}
     \si = \fun{e} e
@@
     (\fun{b} b)
-\end{equation*}
+\end{equation}
 and apply it to the encoded term.  The result is the object term
-\begin{equation*}
+\begin{equation}
+\kern-\mintagsep
     \bigl(\fun{e} e (\fun{f} \fun{x} fx) (\fun{f} f) (\fun{b} b)\bigr)
@@
     \fun{\ident{app}} \fun{\ident{lam}} \fun{\ident{bool}}
-    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool} \True)
+    \ident{app} (\ident{lam} (\fun{x} x))\, (\ident{bool}\, \True)
     \bigr),
-\end{equation*}
+\end{equation}
 which partially evaluates to~$\True$ easily.
@@
 abstractions at different types.  For example, the pre-encoding
-\begin{equation*}
+\begin{equation}
+\label{e:pre-encoding-example}
     \encode{\fun{f}\fun{x}fx}
     = \ident{lam} (\fun{f} \ident{lam} (\fun{x} \ident{app} f x))
-\end{equation*}
+\end{equation}
 does not type-check in any typing environment, because $\ident{lam}$ needs
@@
 polymorphism of the object language \emph{into} the object language to achieve
-self\hyp interpretation.
+\emph{self}\hyp interpretation.
 
@@
 we can add a new typing rule
-\begin{equation*}
+\begin{equation}
+\label{e:let}
     \let\vcenter\vbox
@@
     .
-\end{equation*}
+\end{equation}
 The pre-encoding of a let\hyp expression is trivial.
-\begin{align*}
+\begin{align}
     \encode{\be{x=e_1}e_2} \quad &= \quad \be{x=\encode{e_1}} \encode{e_2}
@@
     \encode{[~]} \quad &= \quad [~]
-\end{align*}
+\end{align}
 We define an interpreter in the object language to be not a term but
 a context.  For example, the evaluator is the context
-\begin{align*}
+\begin{align}
+\label{e:evaluator}
     \si[~] &=
@@
 \intertext{and the size\hyp measurer at the end of~\S\ref{S:interpreter-RL} is the context}
+\label{e:size-measurer}
     \mathrm{SZ}[~] &=
@@
     \end{tabular}
-\end{align*}
+\end{align}
 To interpret an object term~$e$ using an interpreter $I[~]$ is to
@@
 
-We always type-check a pre-encoded term~$\encode{e}$ and its interpreter~$I[~]$
-together, never separately.  This treatment has the drawback that
-we must duplicate a pre-encoded term in order to interpret it in
-multiple ways.  The meta-notions of contexts and plugging may seem ad
-hoc, but in fact they just reflect the type-class and module
-machinery that we have been using in this paper all along.
+We always type-check a pre-encoded term~$\encode{e}$ in the context of
+a particular interpreter~$I[~]$, never alone.  For example, to
+type-check the pre-encoded term~\eqref{e:pre-encoding-example}, we must
+plug it into an interpreter, such as the evaluator~\eqref{e:evaluator}.
+This treatment has the drawback that we must duplicate the pre-encoding
+of a term in order to interpret it in multiple ways, that is, to plug it
+into multiple interpreters such as the evaluator~\eqref{e:evaluator} and
+the size\hyp measurer~\eqref{e:size-measurer}.  In return, we avoid
+adding polymorphism to the object language's type system, because we can
+state the let rule~\eqref{e:let} in terms of substitution rather than by
+generalizing and instantiating types.  In other words, we use contexts
+and plugging in the object language in place of the type-class and
+module machinery in the metalanguages.
 
@@
 
-It is common to implement an embedded DSL by providing multiple
-interpretations of host\hyp language pervasives such as addition and
-application.  It is also common to use phantom types to rule out
+Our separation between the |Symantics| interface and its many
+implementations codifies the common practice of implementing an embedded
+DSL by specifying an abstract syntax of object\hyp language pervasives,
+such as addition and application, then providing multiple
+interpretations of them.  \Citet{JonesNielsen-D58} prefigured this
+separation when they decomposed a denotational definition of an untyped
+object language into a core semantics (which we call abstract syntax)
+and multiple interpretations.  When implementing an embedded DSL in this
+way, it is also common practice to use phantom types to rule out
 ill-typed object terms, as done in Lava
-\citep{Lava} and by \citet{Rhiger-thesis}. However, these approaches
+\citep{Lava} and by \citet{Rhiger-thesis}. However, these two approaches
 are not tagless because they still use universal types, such as Lava's
@@
 compile-time, so that the end result is tagless indeed.  The main
-difference between the work of \Citet{Birkedal-PE-ML} and ours is how
+difference between the work of \citet{Birkedal-PE-ML} and ours is how
 \emph{early} it becomes \emph{manifest} to the metalanguage
 that our code is safe.  The fact that
-\Citet{Birkedal-PE-ML} 's generating extension never generates a
+\citet{Birkedal-PE-ML}'s generating extension never generates a
 specialized program that is ill-typed is not manifest to the SML
@@
 program.  Moreover, and in a similar vein, the fact that
-\Citet{Birkedal-PE-ML} 's cogen never generates a generating extension
+\citet{Birkedal-PE-ML}'s cogen never generates a generating extension
 that is ill-typed (or that may generate a specialized program that is

### 2008-08-28 tagless/tagless-final.tex
mostly comments in the notes; while we had figured out what to do about most of Rev2's comments in SF, it has been too long and I have forgotten.
many more updates to the paper - although unfortunately even more notes.

@@
 \citep{taha-tag}.  In short, the universal\hyp type solution is
-unsatisfactory because it does not preserve typing.
+unsatisfactory because it does not reflect the type of the encoded term.
 
@@
 all examples have also been implemented in Haskell.
-\else Our examples below switch between (Meta)OCaml
-and Haskell even though we have implemented each example equivalently in
-both metalanguages, because some of our claims are more obvious in one
-metalanguage than the other.  For example, MetaOCaml provides
-convenient, typed staging facilities.\fi
+\else Except for the fundamental definitions in \S\ref{encoding},
+our examples will be shown in (Meta)OCaml, 
+even though we have implemented each example equivalently in
+both (Meta)OCaml and Haskell%
+\footnote{an earlier version of this paper switched liberally between 
+both languages, as we believed that our claims were made more obvious
+by a judicious choice of metalanguage, but the reviewers disagreed,
+pointing out that few people are equally conversant in both.}.
+We have chosen to use mainly MetaOCaml because it provides
+convenient, typed staging facilities.  We will comment on the significant
+differences between the implementations.
+\fi
 
@@
 
-Comparing |Symantics| with Fig.~\ref{fig:object}
-shows how to represent \emph{every} typed, closed object term in the
-metalanguage. Moreover, the representation preserves types.
-\begin{proposition}
-If an object term has the object type~$t$, then its
-representation in the metalanguage has the type 
-|forall repr.| |Symantics repr => repr |$t$.
-\end{proposition}
-Conversely, the type system of the metalanguage statically checks that the
-represented object term is well-typed and closed.
-If we err, say replace |int 7| with |bool True| in
-|testpowfix7|, Haskell will complain there that the expected type |Int| does not
-match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
-encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
-Haskell's type checker also flags syntactically invalid object terms, 
-such as if we forget |app| somewhere above.
-
 \begin{SaveVerbatim}{ocaml-simple}
@@
 To embed the same object language in (Meta)OCaml, we replace the \ifshort type
-class \fi |Symantics| \ifshort\else type class \fi and its instances by a module signature |Symantics|
-and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
+class \fi |Symantics| \ifshort\else type class \fi and its instances by a
+module signature |Symantics| and its implementations.
+Figure~\ref{fig:ocaml-simple} shows a simple
 signature that suffices until~\S\ref{PE}.  The two differences are:
@@
 
+Comparing |Symantics| with Fig.~\ref{fig:object}
+shows how to represent \emph{every} typed, closed object term in the
+metalanguage. Moreover, the representation preserves types.
+\begin{proposition}\label{prop:typing}
+If an object term has object type~$t$, then its
+representation in the metalanguage has type 
+|forall repr.| |Symantics repr => repr |$t$.
+\end{proposition}
+
+\begin{proof}
+The rules for deriving a type $t$ from a term $T$ of the object language
+correspond \emph{bijectively} to the rules encoded in the signature
+|Symantics| (when the latter is read as a set of implications, \`{a} la
+Curry-Howard).  In other words, each of the rules from Figure~\ref{fig:object}
+corresponds bijectively to one line of |Symantics|.
+
+It is worth noting that, as we use HOAS, rules that normally depend on 
+context, in other words rules which are themselves an encoding of a function
+from the context to a typing relation, (like those for |lam| and |fix|)
+correspond to actual metalanguage functions.
+\end{proof}
+Conversely, the type system of the metalanguage statically checks that the
+represented object term is well-typed and closed.
+If we err, say replace |int 7| with |bool True| in
+|testpowfix7|, the type checker will complain there that the expected type
+|Int| does not match the inferred |Bool|.  Similarly, the object term
+$\fun{x}xx$ and its
+encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
+Both Haskell's and MetaOCaml's type checker also flags syntactically invalid
+object terms, such as if we forget |app| somewhere above.
+
+For simplicity, all examples in the rest of this paper will be in
+(Meta)OCaml, but the accompanying code shows a full Haskell implementation
+as well.
+
 \subsection{Two tagless interpreters}
@@
 
-Having abstracted our term representation over the interpreter, we are
-now ready to present a series of interpreters.  Each interpreter is an
+Now that our term representation is independent of any particular interpreter,
+we are now ready to present a series of interpreters.  Each interpreter is an
 instance of the |Symantics| class in Haskell and a module implementing
@@
 Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
+It
+\begin{comment}
+%% commenting out this Haskell in JFP version as well
 \ifshort It \else
@@
 In OCaml, too, it
-\fi is obvious to the compiler that
+\fi 
+\enc{comment}
+is obvious to the compiler that
 pattern matching cannot fail, because there is no
@@
 \begin{proposition}
-If an object term~$e$ encoded in the metalanguage has type~$t$,
-then evaluating~$e$ in the interpreter~|R| either continues
+If an object term~$e$ encoded (as $e'$) in the metalanguage  has (object)
+type~$t$, then evaluating~$e'$ in the interpreter~|R| either continues
 indefinitely or terminates with a value of the same type~$t$.
@@
 \begin{proposition}
-  If an implementation of |Symantics| never gets stuck, then
+  If, for an implementation |I| of |Symantics|, if for all object
+  terms $e$ (encoded as $e'$), evaluating $e'$ with |I| always
+  terminates, then 
   the type system of the object
@@
   that implementation.
+  In other words, the (total) relation induced by |I| is a sub-relation
+  of the one expressed by the type system of the object language.
+  \jacques{Please check that I have this right}
 \end{proposition}
@@
 For variety, we show another interpreter, which measures the \emph{size}
+(or \emph{length}, thus the mnemonic |L|)
 of each object term, defined as the number of term
@@
 \begin{proposition}
-    The compiler |C| generates tagless code free of pattern matching.
+    For all object terms $e$ of object type $t$ (encoded as $e'$),
+    evaluating $e'$ via the compiler |C| always terminates and
+    generates tagless code (i.e. of type $t$ |code|) free of pattern matching.
 \end{proposition}
 This is syntactically clear from the body of |C|, and follows
-readily by structural induction and the properties of |.~| and
+readily by structural induction and the properties%
+\footnote{which are not easy to establish \citep{MetaML}} of |.~| and
 |.< >.|.
@@
 %------------------------------------------------------------------------
-% This is the version for the APLAS paper. No Haskell
+% This is the version for the APLAS paper.
 \ifshort
@@
 Figure~\ref{fig:pe} shows the partial evaluator~|P|.
-Its type |repr| literally expresses the type equation for |repr_pe| above.
+Its type |repr| expresses the type definition for |repr_pe| given
+as the start of this section, with |'sv| taking the place of |repr_pe t|.
 The function |abstr|
@@
 in a large language with more base types and primitive operations.
-Incidentally, the code actually contains a more general implementation
-mechanism for such features, inspired in part by previous work in
-generative linear algebra~\citep{CaretteKiselyov05}.
+Incidentally, the accompanying code actually contains a more general
+implementation mechanism for such features, inspired in part by previous work
+in generative linear algebra~\citep{CaretteKiselyov05}.
 

### 2008-09-03 tagless/tagless-final.tex
some more work on JFP paper (some real advance but still too many notes)
A few comments and changes

@@
   out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
+\item We show how the module system of Ocaml and Haskell's type class 
+   system can be used to parametrize the semantics of (a particular style of
+   syntax encoding of) lamda terms.
 \end{enumerate}
@@
 
+A referee aptly summarized our paper as ``proposing a way to write a 
+typed fold function over a typed term''.
+
 \section{The object language and its tagless interpreters}\label{language}
@@
 factorial, and power.
-\begin{comment}
-\oleg{There is now IncopeDB.hs that defines at least R, L and C
- interpreters (no PE at present though) using de Bruijn indices. So we
- can say that (some of) our development is available in de Bruijn
- style. We won't talk about it in the paper though.}
-\end{comment}
-In contrast to \S\ref{intro}, we use higher-order
+In contrast to \S\ref{intro}, in the rest of the paper 
+we use higher-order
 abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
@@
 ensures that our object programs are closed.
-We find HOAS to be more convenient, though we could have continued to
-use de Bruijn indices.
+We find HOAS to be more convenient. We do present the version of our
+developement using de Bruijn indices in the accompanying source code, file
+|incope-dB.ml|.
 
@@
 Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
-It
 \begin{comment}
 %% commenting out this Haskell in JFP version as well
-\ifshort It \else
 In Haskell, we define
@@
 cannot ever fail and is assuredly compiled away.
-In OCaml, too, it
-\fi 
-\enc{comment}
-is obvious to the compiler that
+\end{comment}
+It is obvious to the compiler that
 pattern matching cannot fail, because there is no
@@
 the problem of non\hyp exhaustive pattern\hyp matching reappears in
-the GADT approach because GHC 6.6.1 cannot see that a particular
-type of a GADT value precludes certain constructors.
-Thus GADTs fail to
+the GADT approach because GHC 6.8 and prior cannot see that a particular
+type of a GADT value precludes certain constructors. Although this
+is the implementation problem of GHC,
+it indicates that assuring exhaustive pattern match with GADTs
+requires non-trivial reasoning (beyond the abilities of GHC at the moment);
+thus GADTs fail to
 make it \emph{syntactically} apparent that pattern matching is exhaustive.
-\jacques{what about GHC 6.8?}
 
@@
 Its type |repr| expresses the type definition for |repr_pe| given
-as the start of this section, with |'sv| taking the place of |repr_pe t|.
+at the start of this section, with |'sv| taking the place of |repr_pe t|.
 The function |abstr|
@@
 
-Given an \emph{encoding} of each object term~$e$ as an \emph{object} term~$\Encode{e}$,
-a \emph{self\hyp interpreter} is usually defined as an object
-function~$\si$ such that any object term~$e$ is observationally
-equivalent to the object application $\si\Encode{e}$
-\citep{jones-partial,taha-tag,Danvy-tagging-encoding}.
-A partial evaluator~$\pe$ maps object terms~$e$ to observationally
-equivalent object terms~$\pe(e)$.  It is said to be
-\emph{optimal} with respect to~|si| \citep{jones-challenging} if $\pe(\si\Encode{e})$
-is $\alpha$\hyp equivalent to~$e$ (or in some accounts, no
-less efficient than~$e$).
+Given an \emph{encoding} of each object term~$e$ as an \emph{object}
+term~$\Encode{e}$, a \emph{self\hyp interpreter} is usually defined as
+an object function~$\si$ such that any object term~$e$ is
+observationally equivalent to the object application $\si\Encode{e}$
+\citep{jones-partial,taha-tag,Danvy-tagging-encoding}.  A particular
+use of self-interpreters is in defining the notion of optimality of
+partial evaluation.  A partial evaluator~$\pe$ maps object terms~$e$
+to observationally equivalent object terms~$\pe(e)$.  Recall that
+$\si\Encode{e}$ is too observationally equivalent to $e$ but less
+efficient (``observing'' $\si\Encode{e}$ takes more resources). A
+partial evaluator is said to be \emph{optimal} with respect to~|si|
+\citep{jones-challenging} if the partial evaluator removes all the 
+ineffiency in $\si\Encode{e}$ added by the self-interpreter: that is,
+for all $e$, $\pe(\si\Encode{e})$ is $\alpha$\hyp
+equivalent to~$e$ (or in some accounts, no less efficient than~$e$).
 

### 2008-09-03 tagless/Incope.hs

@@ -438,7 +438,8 @@ apparent.
 -- that Asai's code type-checks in Hindley-Milner as soon as we
 -- deforest the static code representation.
 
-data Rep dynamic static = Rep { dynamic :: C dynamic, static :: Maybe static }
+data Rep repr dynamic static = 
+    Rep { dynamic :: repr dynamic, static :: Maybe static }
 
 zE dynamic = Rep dynamic Nothing
 
@@ -447,7 +448,6 @@ zbool b = Rep (bool b) (Just b)
 zlam f = Rep (lam (dynamic . f . zE)) (Just f)
 zapp (Rep _ (Just f)) = f
 zapp (Rep f _       ) = zE . app f . dynamic
-zfix f = f (zE (fix (dynamic . f . zE)))
 zadd (Rep _ (Just n1)) (Rep _ (Just n2)) = zint (n1 + n2)
 zadd (Rep n1 _       ) (Rep n2 _       ) = zE (add n1 n2)
 zmul (Rep _ (Just n1)) (Rep _ (Just n2)) = zint (n1 * n2)
@@ -457,19 +457,39 @@ zleq (Rep n1 _       ) (Rep n2 _       ) = zE (leq n1 n2)
 zif_ (Rep _ (Just b1)) et ee = if b1 then et else ee
 zif_ (Rep be _       ) et ee = zE (if_ be (dynamic et) (dynamic ee))
 
-ztestgib = zlam (\x -> zlam (\y ->
+zfix f = case f (zfix f)  of
+	  Rep _ (Just g) -> Rep dfix
+			        (Just (\x -> 
+				       case x of
+			                Rep cde Nothing -> zE (app dfix cde)
+			                x     -> g x))
+	  Rep _ Nothing -> zE dfix
+ where dfix = fix (dynamic . f . zE)
+
+-- unit to supress the monomorphism restriction
+
+ztestgib () = zlam (\x -> zlam (\y ->
                 zfix (\self -> zlam (\n ->
                     zif_ (zleq n (zint 0)) x
                       (zif_ (zleq n (zint 1)) y
                        (zadd (zapp self (zadd n (zint (-1))))
                              (zapp self (zadd n (zint (-2))))))))))
 
-ztestgib1 = zapp (zapp (zapp ztestgib (zint 1)) (zint 1)) (zint 5)
+ztestgib5 () = zlam (\x -> zlam (\y ->
[…]

### 2008-09-03 tagless/tagless-final.tex
minor edit of the english, more substantial changes later

@@
 the GADT approach because GHC 6.8 and prior cannot see that a particular
-type of a GADT value precludes certain constructors. Although this
-is the implementation problem of GHC,
+type of GADT value precludes certain constructors. Although this
+is an implementation issue of GHC,
 it indicates that assuring exhaustive pattern match with GADTs
 requires non-trivial reasoning (beyond the abilities of GHC at the moment);
-thus GADTs fail to
+certainly GADTs fail to
 make it \emph{syntactically} apparent that pattern matching is exhaustive.
@@
 to observationally equivalent object terms~$\pe(e)$.  Recall that
-$\si\Encode{e}$ is too observationally equivalent to $e$ but less
-efficient (``observing'' $\si\Encode{e}$ takes more resources). A
+$\si\Encode{e}$ is also observationally equivalent to $e$ but less
+efficient (reducing $\si\Encode{e}$ takes more resources). A
 partial evaluator is said to be \emph{optimal} with respect to~|si|

### 2008-09-05 tagless/tagless-final.tex
Some clarifications to the introduction, especially the contributions

@@
 and just enter them by hand.
-\else
-No matter how we represent the object language in the
-metalanguage, the representation can be created either by hand (for example, by
-entering object terms at a metalanguage interpreter's prompt) or
-by a parser\slash type\hyp checker reading from a text string.
-We leave aside the solved problem of writing such a 
-parser\slash type\hyp checker,
-whether using dependent types \citep{WalidICFP02} or not \citep{baars-typing}.
-\oleg{Should we really leave it aside, given that we do have the code:
-  IncopeTypecheck, and given Sumii-san's advice to emphasize the fact
-  that we do have the code. Please see my comment in the related work
-  section in the paragraph that deals with typing dynamic typing.}
 \fi
@@
 \ifshort \BUseVerbatim{test1}.\else \UseVerbatim{test1}\fi
-
 \ifshort
+
 \begin{figure}
@@
 \end{figure}
-\fi
-
-\noindent Following \citep{WalidICFP02},
+
+\fi
+Following \citet{WalidICFP02},
 we try to implement an interpreter function |eval0|\ifshort,
@@
 
-If we evaluate only closed terms that have been type-checked, then
-|eval| would never fail. Alas, this soundness is not obvious to the
+Although |eval| never fails on
+well-typed closed terms, this soundness is not obvious to the
 metalanguage, whose type system we must still appease with the
@@
 \citep{taha-tag}.  In short, the universal\hyp type solution is
-unsatisfactory because it does not reflect the type of the encoded term.
+unsatisfactory because it does not preserve the type of the encoded term.
 
@@
 
-It is commonly thought that to interpret a typed object language in
-a typed metalanguage while preserving types is difficult and requires
+It is commonly thought that the type-preserving interpretation of
+a typed object language in a typed metalanguage is difficult and requires
 GADTs or dependent types \citep{taha-tag}.  In fact, this problem
 motivated much work on GADTs \citep{xi-guarded,peyton-jones-simple} and
-on dependent types \citep{WalidICFP02,fogarty-concoqtion}.
-\ifshort\else
-For a metalanguage's type system to allow the well-typed object term
-|test1| but disallow the ill-typed object term |test2|, fancier types
-such as GADTs or dependent types seem necessary.  
-\fi
+on dependent types \citep{WalidICFP02,fogarty-concoqtion}\ifshort\else,
+in order for the metalanguage's type system to allow the well-typed object term
+|test1| but disallow the ill-typed object term |test2|\fi.
 Yet other type systems
@@
 here we just note that many
-advanced type systems have been devised to ensure statically that an
+fancy type systems have been devised to ensure statically that an
 object term is well-typed and closed\fi.
@@
 
-We represent object programs using ordinary functions rather than
-data constructors.  These functions comprise the entire interpreter, shown
-below.
+Following an old idea of \citet{reynolds-user-defined},
+we represent object programs using ordinary functions rather than
+data constructors.  These functions comprise the entire interpreter:
 \ifshort
@@
 
-In sum, by Church\hyp encoding terms using ordinary functions, we
-achieve a tagless evaluator for a typed object language in a
-metalanguage with a simple 
+In sum, using ordinary functions rather than data constructors
+to represent well-typed terms, we achieve a tagless evaluator
+for a typed object language in a metalanguage with a simple 
 \ifshort
 Hindley-Milner \fi
-type system\ifshort\else\ \citep{hindley-principal,milner-theory}\fi
-.  In this \emph{final} rather
-than \emph{initial} approach, both kinds of run-time errors in
-\S\ref{tagproblem} (applying a nonfunction and evaluating an open
-term) are reported at compile time. Because the new interpreter
-uses no universal type or pattern matching, it never results in a
+type system\ifshort\else\ \citep{hindley-principal,milner-theory}\fi.
+We call this approach \emph{final} (in contrast to \emph{initial}),
+because we represent each object term not by its abstract syntax
+but by its denotation in a semantic algebra.  This representation
+makes it trivial to implement a primitive recursive function over
+object terms, such as an evaluator.  Or, as a referee puts it aptly, our
+proposal is ``a way to write a typed fold function over a typed term.''
+
+We emphasize ``typed'' and ``fold'' in the previous sentence.
+Our encoding of the recursive type of terms \citep{bohm-automatic}
+makes folds over terms much easier to write than functions over
+terms that are not primitive recursive (or, compositional).
+In contrast, \citearound{'s encoding of the sum type of
+terms}\citet{JFP-Mogensen} does not privilege folds.
+In exchange, we statically express object types in the metalanguage
+and prevent both kinds of run-time errors in
+\S\ref{tagproblem}, due to evaluating ill-typed or open terms.
+Because the new interpreter
+uses no universal type or pattern matching, it never gives a
 run-time error, and is in fact total.  Because this safety is obvious
 not just to us but also to the metalanguage implementation, we avoid
-the serious performance penalty \citep{WalidICFP02} of error checking.
-\Citet{Gluck-jones-optimality} explains deeper technical reasons that
-inevitably lead to these performance penalties.
-
-Unlike \citearound{'s HOAS-based term representation}\citet{JFP-Mogensen},
-our solution does \emph{not} involve Church-encoding the
+the serious performance penalty \citep{WalidICFP02}
+that inevitably arises \citep{Gluck-jones-optimality}
+of error checking at run time.
+
+Our solution does \emph{not} involve Church-encoding the
 universal type. The Church encoding of the type~|u| in \S\ref{tagproblem}
@@
 
-The evaluator above is wired directly into the
-functions |b|, |lam|, |app|, and so on.  \ifshort We \else In the rest of this paper, we \fi
-explain how to abstract the interpreter so as
-to process the \emph{same} term in many other
-ways: compilation, partial evaluation, CPS
-conversion, and so forth.
-
-\subsection{Contributions}\label{contributions}
-
-The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
-free in the encoding of an object term such as |testf1| above.  Defining
-these functions differently gives rise to different interpreters, that
-is, different folds on object programs.  Given the same term
-representation but varying the interpreter, we can
+The evaluator above is wired directly into
+functions such as |b|, |lam|, and |app|, whose names appear
+free in |testf1| above.
+\ifshort We \else In the rest of this paper, we \fi
+explain how to abstract over these functions' definitions
+and apply different folds to the same object language, so as
+to process the same term using many other
+interpreters: we can
 \begin{itemize}
@@
     \item transform the term to continuation\hyp passing style (CPS),
-        even call-by-name (CBN) CPS, so as to isolate the evaluation
+        even call-by-name (CBN) CPS in a call-by-value (CBV) metalanguage,
+        so as to isolate the evaluation
         order of the object language from that of the metalanguage.\ifshort
@@
 \end{itemize}
-We have programmed our interpreters in OCaml (and, for staging,
+We have programmed all our interpreters and examples in OCaml (and, for staging,
 \citet{metaocaml}) and standard Haskell. The complete code is
@@
 all examples have also been implemented in Haskell.
-\else Except for the fundamental definitions in \S\ref{encoding},
-our examples will be shown in (Meta)OCaml, 
-even though we have implemented each example equivalently in
-both (Meta)OCaml and Haskell%
-\footnote{an earlier version of this paper switched liberally between 
-both languages, as we believed that our claims were made more obvious
-by a judicious choice of metalanguage, but the reviewers disagreed,
-pointing out that few people are equally conversant in both.}.
-We have chosen to use mainly MetaOCaml because it provides
-convenient, typed staging facilities.  We will comment on the significant
-differences between the implementations.
-\fi
+\else Except for the basic definitions in \S\ref{encoding},
+we show our examples in (Meta)OCaml even though some of our
+claims are more obvious in Haskell, for consistency and because
+MetaOCaml provides convenient, typed staging facilities.
+\fi
+
+\subsection{Contributions}\label{contributions}
 
@@
 result they call desirable.  Our contributions are as follows.
-
 \begin{enumerate}
-\item We build interpreters that evaluate (\S\ref{language}),
-   compile (or evaluate with staging) (\S\ref{S:compiler}), and 
+\item We build the first \emph{family} of interpreters, each instantiating the
+   \emph{same} signature, that evaluate (\S\ref{language}),
+   compile (\S\ref{S:compiler}), and 
    partially evaluate (\S\ref{PE}) a typed higher-order object language
@@
    styles\ifshort\else\ (\S\ref{variations})\fi.
-\item All these interpreters use no type tags, patently never get stuck,
+\item These interpreters use no type tags
     and need no advanced type-system features such as GADTs, dependent types,
     or intentional type analysis.
-\item The partial evaluator avoids polymorphic lift and delays binding-time
-    analysis.  It bakes a type-to-type map into the interpreter
-    interface to eliminate the need for GADTs and thus remain portable
-    across Haskell 98 and ML.
-\ifshort\else
-\item We show the first typed, tagless, and statically type-preserving CPS
-    transformation with simple types (\S\ref{variations}), which is available for mainstream
-    functional languages (ML and Haskell). Type preservation is assured by
-    the soundness of our embedding.
-\fi
-\item We use the type system of the metalanguage
-    to check statically that an object program is well-typed and closed.
-\item We show clean, comparable implementations in MetaOCaml and Haskell.
-\item We specify a functor signature that encompasses all our
-  interpreters, from evaluation and compilation (\S\ref{language}) 
-   to partial evaluation \ifshort\else and CPS transformation \fi(\S\ref{PE}).
+    Yet the type system of the metalanguage
+    assures statically that each object program is well-typed and closed,
+    and that each interpreter preserves types and never gets stuck.
+    In particular, our (online) partial evaluator and CPS transformers
+    avoid GADTs in their implementation and stay portable across
+    Haskell 98 and ML, by baking into their interface an inductive
+    map from input types to output types.
+\item Our clean, comparable implementations using OCaml modules and Haskell
+    type classes show how to parametrize our final representation of object
+    terms over multiple ways to assign them meanings.
 \item We point a clear way to extend the object language with more features
@@
   out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
-\item We show how the module system of Ocaml and Haskell's type class 
-   system can be used to parametrize the semantics of (a particular style of
-   syntax encoding of) lamda terms.
 \end{enumerate}
@@
 literature that use features present in mainstream functional languages.
-In particular, we follow \citearound{'s encoding of type-indexed
-values}\citet{yang-encoding}, \citearound{'s construction of dynamic
-terms alongside static terms}\citet{asai-binding-time}, and
-\citearound{'s deforestation of syntax
-constructors}\citet{sumii-hybrid}.  These techniques require just
-a Hindley-Milner type system with either a module system that supports
+In particular, we follow \citets{yang-encoding} encoding of type-indexed
+values, \citets{Sperber-SelfApplicable} and \citets{asai-binding-time}
+construction of dynamic terms alongside static terms, and
+\citets{sumii-hybrid} deforestation of syntax
+constructors.  These techniques require just
+a Hindley-Milner type system with either module
 functors or constructor classes, as realized in all variants of ML and
@@
 embedded DSLs.
-\begin{comment}
-We may claim some contribution about type-preserving CPS. Check
-related work section in \citep{Guillemette-Monier-PLPV}, especially
-check the work of Shao on type-preserving CPS in Flint. The PLPV paper
-in related work shows other tasks, including closure conversion, which
-we may tackle in our approach. We may be able to write a
-type-preserving, assured compiler, whose properties and assured by HM.
-\end{comment}
-
-A referee aptly summarized our paper as ``proposing a way to write a 
-typed fold function over a typed term''.
+
+\ifshort\else
+However we represent an object term,
+the representation can be created either by hand (for example, by
+entering object terms at a metalanguage interpreter's prompt) or
+by parsing and type-checking text.
+It is known how to write such a parser and type-checker,
+whether using fancy types \citep{Guillemette-Monier-PLPV,haskell-list,WalidICFP02} or not \citep{baars-typing}.
+We have ourselves implemented a parser and type-checker for our object
+language, which returns (either a parse or type error or) a finally
+encoded object term that can then be interpreted in multiple ways
+without repeated type-checking.  We leave this problem aside in the
+rest of this paper.
+\fi
 
@@
 \section{A tagless partial evaluator}\label{PE}
-Surprisingly, we can write a partial evaluator using the idea above,
-namely to build object terms using ordinary functions rather than data
-constructors.  We present this partial evaluator in a sequence of three
-attempts. It uses no universal type and no tags
+Surprisingly, the |Symantics| interface presented above extends to
+encompass an online partial evaluator.  We present this partial evaluator
+in a sequence of three attempts. It uses no universal type and no tags
 for object types.  We then discuss residualization and binding-time
@@
 analysis.  To avoid needing polymorphic lift, we turn to
-Asai's technique \citep{asai-binding-time,sumii-hybrid} of
-building a dynamic term alongside every static term.
+\citets{Sperber-SelfApplicable} and \citets{asai-binding-time} technique of
+building a dynamic term alongside every static term \citep{sumii-hybrid}.
 
@@
 in~\S\ref{S:PE-problem}, we build dynamic terms alongside
-any static ones to avoid polymorphic lift.
+any static ones to express how the lift function is indexed
+by the dynamic type.  What is new is that we analogously
+build a static type alongside the dynamic type to express
+how the static type is indexed by the dynamic type.
 
@@
 At the term level, we also avoid converting between static and dynamic
-terms by building them in parallel, as \citet{asai-binding-time} does
-in his partial evaluator.
+terms by building them in parallel, as \citet{Sperber-SelfApplicable}
+and \citet{asai-binding-time} do in their partial evaluators.
 This method type-checks in Hindley-Milner once we
@@
 \Citets{JFP-Mogensen} self-reducer for the untyped
-$\lambda$-calculus preceded Asai's partial evaluator in constructing
+$\lambda$-calculus preceded Sperber's and Asai's partial evaluators in constructing
 static and dynamic terms in parallel.  However, this self-reducer
@@
 so (in terms of Fig.~\ref{fig:pe}) it moves some work from |app|
-to~|pdyn|.  In contrast, we follow Asai and construct the static term
+to~|pdyn|.  In contrast, we follow Sperber and Asai in constructing the static term
 optionally, so as to perform type-directed lifting without requiring the
@@
 
-\Citet{sumii-hybrid} also use Asai's construction,
+\Citet{sumii-hybrid} also use Sperber and Asai's construction,
 to combine online
@@
 
-\ifshort\else
-One way to understand our main idea is to eschew sum types and sum kinds
-for their dual, record types and record kinds.
-For the self\hyp interpreter, we then proceed to use a Church encoding for
-recursive data types \citep{bohm-automatic}.
-\fi
-
 \ifshort We observe that \else As \S\ref{S:clarify} observes, \fi
@@
 \fi
-
-Our representation of object terms in elimination form encodes
-primitive recursive folds over the terms. 
-\ifshort\else
-This encoding makes operations like interpretation trivial to implement.
-\fi
-We still
-have to understand if and how non-primitively 
-recursive operations can be supported.

### 2008-09-07 tagless/tagless-final.tex
Start to formalize our claims further

@@
     avoid GADTs in their implementation and stay portable across
-    Haskell 98 and ML, by baking into their interface an inductive
+    Haskell 98 and ML, by expressing in their interface an inductive
     map from input types to output types.
@@
 ensures that our object programs are closed.
-We find HOAS to be more convenient. We do present the version of our
-developement using de Bruijn indices in the accompanying source code, file
-|incope-dB.ml|.
+We find HOAS to be more convenient, but we have also implemented our
+approach using de Bruijn indices (in the accompanying source file
+|incope-dB.ml|).
 
@@
 The methods |add|, |mul|, and |leq| are quite similar, and so are
-|int| and |bool|. Therefore, we often show
-only one method of each group and elide the rest. The
+|int| and |bool|. Therefore, we often elide all but
+one method of each group. The
 accompanying code has the complete implementations.
@@
 The dummy argument to |test1| and |testpowfix| is an artifact of
-MetaOCaml, related to monomorphism: in order for us to run a
+MetaOCaml: in order for us to run a
 piece of generated code, it must be polymorphic in its environment
-classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
-The value restriction dictates that
-the definitions of our object terms must look syntactically like
-values. (Alternatively, we could have used
-the rank-2 record types of OCaml to maintain the necessary polymorphism.)
+classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}),
+so we must define our object terms as syntactic values to satisfy
+the value restriction.
+(Alternatively, we could have used
+OCaml's rank-2 record types to maintain the necessary polymorphism.)
 \ifshort\else\par\fi
 Thus, we represent an object expression in
-OCaml as a functor from |Symantics| to an appropriate semantic domain. This
+OCaml as a functor from |Symantics| to a semantic domain. This
 is essentially the same as the constraint \texttt{Symantics repr =>} in the
@@
 Comparing |Symantics| with Fig.~\ref{fig:object}
-shows how to represent \emph{every} typed, closed object term in the
-metalanguage. Moreover, the representation preserves types.
+shows how to represent \emph{every} well-typed object term in the
+metalanguage.
+We formalize this representation by defining $H$ and~$M$,
+two inductive maps from terms and types in our object language
+to terms and types in Haskell and OCaml:
+\begin{align}
+    H(\ZZ) &= \texttt{Int} &
+    M(\ZZ) &= \texttt{int} \notag\\
+    H(\BB) &= \texttt{Bool} &
+    M(\BB) &= \texttt{bool} \notag\\
+    H(t_1 \to t_2) &= H(t_1) \mathbin{\texttt{->}} H(t_2) &
+    M(t_1 \to t_2) &= M(t_1) \mathbin{\texttt{->}} M(t_2) \displaybreak[0] \\
+    H(x) &= x &
+    M(x) &= x \notag\\
+    H(\fun{x}e) &= \texttt{lam (\textbackslash x -> $H(e)$)} &
+    M(\fun{x}e) &= \texttt{lam (fun x -> $M(e)$)} \notag\\
+    H(\fix{f}e) &= \texttt{fix (\textbackslash x -> $H(e)$)} &
+    M(\fix{x}e) &= \texttt{fix (fun x -> $M(e)$)} \notag\\
+    H(e_1 e_2) &= \texttt{app $H(e_1)$ $H(e_2)$} &
+    M(e_1 e_2) &= \texttt{app $M(e_1)$ $M(e_2)$} \notag\\
+    H(n) &= \texttt{int $n$} &
+    M(n) &= \texttt{int $n$} \notag\\
+    H(\True) &= \texttt{bool True} &
+    M(\True) &= \texttt{bool true} \notag\\
+    H(\False) &= \texttt{bool False} &
+    M(\False) &= \texttt{bool false} \notag\\
+    \!H(\cond{e}{e_1}{e_2}) &= \rlap{\texttt{if\_ $H(e)$ $H(e_1)$ $H(e_2)$}} \notag\\
+    \!M(\cond{e}{e_1}{e_2}) &= \rlap{\texttt{if\_ $M(e)$ (fun () -> $M(e_1)$) (fun () -> $M(e_2)$)}} \notag\\
+    H(e_1 + e_2) &= \texttt{add $H(e_1)$ $H(e_2)$} &
+    M(e_1 + e_2) &= \texttt{add $M(e_1)$ $M(e_2)$} \notag\\
+    H(e_1 \times e_2) &= \texttt{mul $H(e_1)$ $H(e_2)$} &
+    M(e_1 \times e_2) &= \texttt{mul $M(e_1)$ $M(e_2)$} \notag\\
+    H(e_1 \le e_2) &= \texttt{leq $H(e_1)$ $H(e_2)$} &
+    M(e_1 \le e_2) &= \texttt{leq $M(e_1)$ $M(e_2)$}
+\end{align}
+These definitions assume that our object language, Haskell, and OCaml
+use the same variable names~$x$ and integer literals~$n$.
+If $\Gamma$ is a typing context $x_1:t_1,\dotsc,x_n:t_n$ in the object
+language, then we define the metalanguage contexts
+\begin{align}
+    \texttt{repr\;$H(\Gamma)$}
+    &= x_1:\texttt{repr}\;H(t_1),\dotsc,x_n:\texttt{repr}\;H(t_n) \text, \notag\\
+    \texttt{('c,\:$M(\Gamma)$)\;repr}
+    &= x_1:\texttt{('c,\:$M(t_1)$)\;repr},\dotsc,x_n:\texttt{('c,\:$M(t_n)$)\;repr} \text.
+\end{align}
+
+The following proposition states the trivial but fundamental fact that
+this representation preserves types.
 \begin{proposition}\label{prop:typing}
-If an object term has object type~$t$, then its
-representation in the metalanguage has type 
-|forall repr.| |Symantics repr => repr |$t$.
+If an object term~$e$ has the type~$t$ in the context~$\Gamma$,
+then the Haskell term~$H(e)$ has the type \texttt{repr\;$H(t)$}
+in the context
+\[
+    \texttt{repr}:\star\to\star,\quad
+    \texttt{Symantics repr},\quad
+    \texttt{repr}\;H(\Gamma)
+    \text,
+\]
+and the OCaml term~$M(e)$ has the type \texttt{('c,\:$M(t)$)\;repr}
+in the context
+\[
+    \texttt{S:Symantics},\quad
+    \texttt{open S},\quad
+    \texttt{'c}:\star,\quad
+    \texttt{('c,\:$M(\Gamma)$)\;repr}
+    \text.
+\]
 \end{proposition}
-
 \begin{proof}
-The rules for deriving a type $t$ from a term $T$ of the object language
-correspond \emph{bijectively} to the rules encoded in the signature
-|Symantics| (when the latter is read as a set of implications, \`{a} la
-Curry-Howard).  In other words, each of the rules from Figure~\ref{fig:object}
-corresponds bijectively to one line of |Symantics|.
-
-It is worth noting that, as we use HOAS, rules that normally depend on 
-context, in other words rules which are themselves an encoding of a function
-from the context to a typing relation, (like those for |lam| and |fix|)
-correspond to actual metalanguage functions.
+By structural induction on the derivation in the object language
+that $e$ has type~$t$ in~$\Gamma$.
 \end{proof}
-Conversely, the type system of the metalanguage statically checks that the
+\begin{corollary}
+If a closed object term~$e$ has the type~$t$,
+then the Haskell term~$H(e)$ has the type
+\[
+    \texttt{forall repr. Symantics repr => repr $H(t)$}
+\]
+and the OCaml functor
+\[
+    \texttt{functor (S:Symantics) -> struct open S let term () = $M(e)$ end}
+\]
+has the signature
+\[
+    \texttt{functor (S:Symantics) -> sig val term: unit -> ('c,\:$M(t)$)\;S.repr end}
+    \text{.}
+\]
+\end{corollary}
+
+Conversely, the type system of the metalanguage checks that the
 represented object term is well-typed and closed.
@@
 
-For simplicity, all examples in the rest of this paper will be in
+For simplicity, all examples in the rest of this paper are in
 (Meta)OCaml, but the accompanying code shows a full Haskell implementation
@@
 Now that our term representation is independent of any particular interpreter,
-we are now ready to present a series of interpreters.  Each interpreter is an
+we are ready to present a series of interpreters.  Each interpreter is an
 instance of the |Symantics| class in Haskell and a module implementing
-the |Symantics| signature in MetaOCaml.
+the |Symantics| signature in OCaml.
 
 The first interpreter evaluates an object term to its value in the
-metalanguage.  The module below interprets each object\hyp language
+metalanguage.  The module below is metacircular in that it
+interprets each object\hyp language
 operation as the corresponding metalanguage operation.
@@
 
-\ifshort\else
-A lot of effort has gone into ``typing dynamic typing'': to statically
-type-check dynamically\hyp typed values
-\citep{baars-typing,WalidICFP02,Guillemette-Monier-PLPV,haskell-list},
-using the host language's type system to varying extents.
-Our object terms are statically typed, so we would
-need one of these techniques to interpret dynamically\hyp typed
-terms such as those read from a file.
-\oleg{First, I suggest we remove the subjunctive mood: `we would
-  need...' We have the code already, IncopeTypecheck.hs. We know we
-  can typecheck untyped terms so that we do the typechekcing only
-  once, and interpret many times. All these interpretation is guranteed to
-  be free from any type errors. And this holds even for higher-order
-  terms encoded in HOAS. I also suggest we drop
-  \cite{haskell-list}. We included the citation solely to show how 
-  we could go on. Both links in haskell-list are not actually
-  appropriate: first, the object language is \emph{first-order}
-  and second they use GADTs. Our IncopeTypecheck uses only
-  Typeable and existentials, and no GADTs.
-  Finally, we need to find a way to talk about IncopeTypecheck more
-  directly, probably in the Introduction -- Sumii-san has directly
-  suggested we emphasize that result. For some reason, people (Walid
-  and Sumii-san, for ones) consider it important. Incidentally, I
-  have the comparison of IncopeTypecheck with `Typing Dynamic
-  Typing', which we can use if needed.}
-\fi
-
 Our partial evaluator establishes a bijection |repr_pe| between static

### 2008-09-10 tagless/tagless-final.tex
Don't cite haskell-list

@@
 by parsing and type-checking text.
-It is known how to write such a parser and type-checker,
-whether using fancy types \citep{Guillemette-Monier-PLPV,haskell-list,WalidICFP02} or not \citep{baars-typing}.
-We have ourselves implemented a parser and type-checker for our object
-language, which returns (either a parse or type error or) a finally
+It is known how to write such a type checker
+for a higher-order object language such as ours,
+whether using fancy types \citep{Guillemette-Monier-PLPV,WalidICFP02} or not \citep{baars-typing}.
+We have ourselves implemented a type checker for our object
+language (in the accompanying source file |IncopeTypecheck.hs|),
+which maps an ordinary syntax tree to (either a type error or) a finally
 encoded object term that can then be interpreted in multiple ways

### 2008-09-10 tagless/tagless-final.tex
Fixed formalization in Section 2

@@
 \end{proof}
-\begin{corollary}
+\begin{corollary}\label{cor:typing}
 If a closed object term~$e$ has the type~$t$,
@@
 
-For simplicity, all examples in the rest of this paper are in
-(Meta)OCaml, but the accompanying code shows a full Haskell implementation
-as well.
-
 \subsection{Two tagless interpreters}
@@
 due to interpreting |fix|.
+The soundness of the object language's type system with respect to
+the dynamic semantics specified by a definitional interpreter
+follows from the soundness of the metalanguage's type system.
 \ifshort
@@
 \begin{proposition}
-If an object term~$e$ encoded (as $e'$) in the metalanguage  has (object)
-type~$t$, then evaluating~$e'$ in the interpreter~|R| either continues
-indefinitely or terminates with a value of the same type~$t$.
+If a closed object term~$e$ has type~$t$,
+and the OCaml module |I| implements the signature |Symantics|,
+then under the OCaml module definition
+\[
+    \texttt{\begin{tabular}{@{}l@{}}
+    module RESULT = \\
+    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
+    \quad (I)
+    \end{tabular}}
+\]
+evaluating the expression \texttt{RESULT.term ()} never gets stuck: it either
+does not terminate or evaluates to a value of type
+\texttt{('c,\:$M(t)$)\;I.repr} (polymorphic over~|'c|).
 \end{proposition}
-\ifshort\else
-Generalizing from~|R| to all interpreters, we have the following
-broader and more useful, if also more obvious, proposition.
-\fi
-\begin{proposition}
-  If, for an implementation |I| of |Symantics|, if for all object
-  terms $e$ (encoded as $e'$), evaluating $e'$ with |I| always
-  terminates, then 
-  the type system of the object
-  language is sound with respect to the dynamic semantics defined by
-  that implementation.
-  In other words, the (total) relation induced by |I| is a sub-relation
-  of the one expressed by the type system of the object language.
-  \jacques{Please check that I have this right}
-\end{proposition}
-\ifshort\else
-These propositions follow immediately from the soundness of the
-metalanguage's type system.
-\fi
+\begin{proof}
+By Corollary~\ref{cor:typing} and the type soundness of (this fragment of) OCaml.
+\end{proof}
+\begin{corollary}
+If a closed object term~$e$ has type~$t$,
+then under the OCaml module definition
+\[
+    \texttt{\begin{tabular}{@{}l@{}}
+    module RESULT = \\
+    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
+    \quad (R)
+    \end{tabular}}
+\]
+evaluating the expression \texttt{RESULT.term ()} never gets stuck: it either
+does not terminate or evaluates to a value of type~$M(t)$.
+\end{corollary}
 \ifshort\else
@@
 evaluates to |3|. This interpreter is not only tagless but also
-total. It ``evaluates'' even seemingly divergent terms, for instance
+total. It ``evaluates'' even seemingly divergent terms; for instance,
 |app (fix (fun self -> self)) (int 1)| evaluates to $3$.

### 2008-09-10 tagless/tagless-final.tex
Fixed formalization in Section 3

@@
     M(x) &= x \notag\\
-    H(\fun{x}e) &= \texttt{lam (\textbackslash x -> $H(e)$)} &
-    M(\fun{x}e) &= \texttt{lam (fun x -> $M(e)$)} \notag\\
-    H(\fix{f}e) &= \texttt{fix (\textbackslash x -> $H(e)$)} &
-    M(\fix{x}e) &= \texttt{fix (fun x -> $M(e)$)} \notag\\
+    H(\fun{x}e) &= \texttt{lam (\textbackslash $x$ -> $H(e)$)} &
+    M(\fun{x}e) &= \texttt{lam (fun $x$ -> $M(e)$)} \notag\\
+    H(\fix{f}e) &= \texttt{fix (\textbackslash $x$ -> $H(e)$)} &
+    M(\fix{x}e) &= \texttt{fix (fun $x$ -> $M(e)$)} \notag\\
     H(e_1 e_2) &= \texttt{app $H(e_1)$ $H(e_2)$} &
@@
     \texttt{repr\;$H(\Gamma)$}
-    &= x_1:\texttt{repr}\;H(t_1),\dotsc,x_n:\texttt{repr}\;H(t_n) \text, \notag\\
+    &= x_1:\texttt{repr}\;H(t_1),\dotsc,x_n:\texttt{repr}\;H(t_n) \text, \\
     \texttt{('c,\:$M(\Gamma)$)\;repr}
@@
 \begin{proposition}
-    For all object terms $e$ of object type $t$ (encoded as $e'$),
-    evaluating $e'$ via the compiler |C| always terminates and
-    generates tagless code (i.e. of type $t$ |code|) free of pattern matching.
+If an object term~$e$ has the type~$t$
+in the context $x_1:t_1,\dotsc,x_n:t_n$,
+then in a MetaOCaml environment
+\[
+    \texttt{open C},\quad
+    x_1 \mapsto \texttt{.<$y_1$>.},\enspace
+    \dotsc,\enspace
+    x_n \mapsto \texttt{.<$y_n$>.}
+\]
+where each $y_i$ is a future-stage variable of type $M(t_i)$,
+the MetaOCaml term~$M(e)$ evaluates to a code value,
+of type~\texttt{('c,\:$M(t)$)\;code} (polymorphic over~|'c|),
+that contains no pattern-matching operations.
 \end{proposition}
-This is syntactically clear from the body of |C|, and follows
-readily by structural induction and the properties%
-\footnote{which are not easy to establish \citep{MetaML}} of |.~| and
-|.< >.|.
+\begin{proof}
+    By structural induction on the typing derivation of~$e$.
+\end{proof}
+\begin{corollary}
+If a closed object term~$e$ has type~$t$,
+then under the OCaml module definition
+\[
+    \texttt{\begin{tabular}{@{}l@{}}
+    module RESULT = \\
+    \quad (functor (S:Symantics) -> struct open S let term () = $M(e)$ end) \\
+    \quad (C)
+    \end{tabular}}
+\]
+the expression \texttt{RESULT.term ()} evaluates to a code value,
+of type~\texttt{('c,\:$M(t)$)\;code} (polymorphic over~|'c|),
+that contains no pattern-matching operations.
+\end{corollary}
 
@@
 Since Haskell
-has no (convenient, typed) staging facility, we had to emulate
-it: we defined a data type |ByteCode| with
+has no convenient facility for typed staging, we emulate
+it by defining a data type |ByteCode| with
 constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
-(Alternatively, we could use Template Haskell as our staging facility:
+(Alternatively, we could use Template Haskell \citep{sheard-template} as our staging facility:
 \texttt{ByteCode} can be mapped to the abstract syntax of Template
-Haskell. The output of our compiler will then be assuredly type-correct
+Haskell. The output of our compiler would then be assuredly type-correct
 Template Haskell.)
@@
 The implementation is quite similar (but slightly more
-verbose) than the corresponding MetaOCaml code above. (The implementation uses
+verbose) than the MetaOCaml code above. (The implementation uses
 GADTs because we also wanted to write a typed interpreter for 

### 2008-09-12 tagless/tagless-final.tex
Rename x and y to e1 and e2 for consistency

@@
   let add  e1 e2    = e1 + e2   let mul  e1 e2    = e1 * e2
-  let leq  x y      = x <= y
+  let leq  e1 e2    = e1 <= e2
   let if_  eb et ee = if eb then et () else ee () end
@@
   let mul  e1 e2    = e1 * e2
-  let leq  x y      = x <= y
+  let leq  e1 e2    = e1 <= e2
   let if_  eb et ee = if eb then et () else ee ()
@@
   let mul  e1 e2    = e1 + e2 + 1
-  let leq  x y      = x + y + 1
+  let leq  e1 e2    = e1 + e2 + 1
   let if_  eb et ee = eb + et () + ee () + 1
@@
   let add e1 e2    = .<.~e1 + .~e2>.  let mul e1 e2 = .<.~e1 * .~e2>.
-  let leq x y      = .<.~x <= .~y>.
+  let leq e1 e2    = .<.~e1 <= .~e2>.
   let if_ eb et ee = .<if .~eb then .~(et ()) else .~(ee ())>. end
@@
   let mul e1 e2     = .<.~e1 * .~e2>.
-  let leq x y       = .<.~x <= .~y>.
+  let leq e1 e2     = .<.~e1 <= .~e2>.
   let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>.

### 2008-09-12 tagless/tagless-final.tex
Add code using de Bruijn indices

@@
 We find HOAS to be more convenient, but we have also implemented our
-approach using de Bruijn indices (in the accompanying source file
+approach using de Bruijn indices
+(in Section~\ref{S:de-Bruijn} and the accompanying source file
 |incope-dB.ml|).
@@
 \end{comment}
+
+\subsection{Higher-order abstract syntax versus de Bruijn indices}
+\label{S:de-Bruijn}
+
+\begin{figure}
+\begin{floatrule}
+\begin{BVerbatim}
+module type Symantics = sig
+  type ('c,'h,'dv) repr
+  type ('c,'dv) vr			(* variable representation *)
+
+  val vz  : ('c, ('c,'d) vr * 'h, 'd) repr
+  val vs  : ('c, 'h, 'd) repr -> ('c, _ * 'h, 'd) repr
+
+  val int : int  -> ('c,'h,int) repr
+  val bool: bool -> ('c,'h,bool) repr
+
+  val lam : ('c, ('c,'da) vr * 'h, 'db) repr -> ('c,'h,'da->'db) repr
+  val app : ('c,'h,'da->'db) repr -> ('c,'h,'da) repr -> ('c,'h,'db) repr
+  val fix : ('c, ('c,'da->'db) vr * 'h, 'da->'db) repr
+            -> ('c, 'h, 'da->'db) repr
+
+  val add : ('c,'h,int) repr -> ('c,'h,int) repr -> ('c,'h,int) repr
+  val mul : ('c,'h,int) repr -> ('c,'h,int) repr -> ('c,'h,int) repr
+  val leq : ('c,'h,int) repr -> ('c,'h,int) repr -> ('c,'h,bool) repr
+  val if_ : ('c,'h,bool) repr
+            -> (unit -> 'x) -> (unit -> 'x) -> (('c,'h,'da) repr as 'x)
+end
+
+module R = struct
+  type ('c,'h,'dv) repr = 'h -> 'dv
+  type ('c,'d) vr = 'd
+
+  let vz   (x,_) = x
+  let vs v (_,h) = v h
+
+  let int  (x:int)  h = x
+  let bool (b:bool) h = b
+  let lam  f        h = fun x -> f (x,h)
+  let app  e1 e2    h = (e1 h) (e2 h)
+  let fix  f        h = let rec self n = f (self,h) n in self
+  let add  e1 e2    h = e1 h + e2 h
+  let mul  e1 e2    h = e1 h * e2 h
+  let leq  e1 e2    h = e1 h <= e2 h
+  let if_  eb et ee h = if eb h then et () h else ee () h
+end
+\end{BVerbatim}
+\end{floatrule}
+\caption{Embedding and evaluating our object language using de Bruijn indices}
+\label{fig:de-Bruijn}
+\end{figure}
+
+Because Haskell and ML allow case analysis on $\lambda$-bound variables,
+one might worry that our HOAS representation of the object language
+allows \emph{exotic terms} and is thus inadequate.  To the contrary,
+because the representation of an object term is \emph{parametrically}
+polymorphic over the type constructor |repr| of the interpreter,
+$\lambda$-bound object variables cannot be case-analyzed.
+We thus follow \citet{washburn-boxes-jfp} in ``enforcing term
+parametricity with type parametricity'' to represent and fold over
+abstract syntax.
+
+Although the rest of this paper continues to represent binding
+using HOAS\@, our approach is compatible with de Bruijn indices.
+The accompanying source file |incope-dB.ml| implements this
+alternative, starting with the |Symantics| signature and the |R|
+evaluator in Fig.~\ref{fig:de-Bruijn}.
 
@@
 which intensional type analysis cannot handle
-\citep{xi-guarded}.  Our final approach
-seems related to \citearound{'s approach to HOAS using catamorphisms and
-  anamorphisms}\citet{Washburn-Weirich-boxes}.
+\citep{xi-guarded}.
 

### 2008-09-12 tagless/tagless-final.tex
Revised Section 4

@@
 construction of dynamic terms alongside static terms, and
-\citets{sumii-hybrid} deforestation of syntax
+\citets{Thiemann-combinators} deforestation of syntax
 constructors.  These techniques require just
@@
 \section{A tagless partial evaluator}\label{PE}
-Surprisingly, the |Symantics| interface presented above extends to
-encompass an online partial evaluator.  We present this partial evaluator
-in a sequence of three attempts. It uses no universal type and no tags
-for object types.  We then discuss residualization and binding-time
+Surprisingly, this |Symantics| interface extends to
+encompass an online partial evaluator that uses no universal type and no tags
+for object types.  We present this partial evaluator
+in a sequence of three attempts to express the types
+of residualization and binding-time
 analysis.  Our partial evaluator is a modular extension of the evaluator
@@
 distinguish between static and dynamic terms, we might try to define
-|repr| in the partial evaluator (module |P0|) as
+|repr| in the partial evaluator as follows.
+In the phase tags |S0| and~|D0|, the digit zero indicates
+our initial attempt.
 \begin{code}
@@
 \end{code}
-(the digit zero in the names of the phase tags |S0| and |D0| is the
-version suffix).
 To extract a dynamic term from this type, we create the function
 \begin{code}
-let abstrI0 (e:('c,int) repr) : ('c,int) C.repr =
-  match e with
-  | S0 e -> C.int e
-  | D0 e -> e
+let abstrI0 (e : ('c,int) repr) : ('c,int) C.repr =
+  match e with S0 e -> C.int e | D0 e -> e
 \end{code}
 and the similar function |abstrB0| for dynamic boolean terms. Here, 
-|C.int| is used to convert from the static term |('c,int) R.repr|, 
-which is just |int|, to a dynamic term. We can now define the following
-components of module |P0| required by the |Semantics| signature:
+|C.int| is used to convert a static term (of type |('c,int) R.repr|, 
+which is just |int|) to a dynamic term. We can now define the following
+components required by the |Semantics| signature:
 \begin{code}
@@
 let bool (x:bool) = S0 (R.bool x)
-let add e1 e2 = 
-  match (e1,e2) with
+let add e1 e2 = match (e1,e2) with
   | (S0 e1, S0 e2) -> S0 (R.add e1 e2)
-  | (e1, e2)       -> D0 (C.add (abstrI0 e1) (abstrI0 e2))
+  | _ -> D0 (C.add (abstrI0 e1) (abstrI0 e2))
 \end{code}
@@
 
-Our |Symantics| only includes separate lifting methods |bool| and
-|int|, not a parametrically polymorphic lifting method, for good reason:
+Our |Symantics| signature only includes separate lifting methods |bool| and
+|int|, not a polymorphic lifting method, for good reason:
 When compiling to a first-order target language such as machine code,
@@
 
-We start building the partial evaluator anew (module |P1|) and
+We start building the partial evaluator anew and
 switch to the data type
@@
 so that a partially evaluated term always contains a dynamic
-component and sometimes contains a static component.  By 
-distributivity, the two
+component and sometimes contains a static component.
+The two
 alternative constructors of an |option| value, |Some| and |None|,
-tag each partially evaluated term with a phase: either present or
-future.  This tag is not an object type tag: all pattern matching below
-is exhaustive. Because the future-stage component is always available, we
-can now define the polymorphic function
-|let abstr1 (P1 (_,dyn)) = dyn| of type
-\texttt{('c,'dv) repr -> ('c,'dv) C.repr}
-to extract it without requiring polymorphic lift into~|C|.  We then try
-to define the interpreter |P1|---and get as far as the first-order
-constructs of our object language, including |if_|.
-\begin{code}
-module P1 : Symantics = struct
+tag each partially evaluated term to indicate whether its value is
+known statically at the present stage.
+This tag is not an object type tag: all pattern matching below
+is exhaustive. Now that the future-stage component is always available, we
+can define the polymorphic function
+\begin{code}
+let abstr1 (P1 (_,dyn) : ('c,'dv) repr) : ('c,'dv) C.repr = dyn
+\end{code}
+to extract it without needing polymorphic lift into~|C|.  We then try
+to define the term combinators\nobreak\hspace{0pt}---and get as far as
+the first-order constructs of our object language, including |if_|.
+\begin{code}
 let int (x:int) = P1 (Some (R.int x), C.int x)
 let add e1 e2 = match (e1,e2) with
-    | (P1 (Some n1,_),P1 (Some n2,_)) -> int (R.add n1 n2)
-    | _ -> P1 (None,(C.add (abstr1 e1) (abstr1 e2)))
-  let if_ = function
-    | P1 (Some s,_) -> fun et ee -> if s then et () else ee ()
-    | eb -> fun et ee -> P1 (None, C.if_ (abstr1 eb) 
-                                   (fun () -> abstr1 (et ()))
+  | (P1 (Some n1, _), P1 (Some n2, _)) -> int (R.add n1 n2)
+  | _ -> P1 (None, C.add (abstr1 e1) (abstr1 e2))
+let if_ eb et ee = match eb with
+  | P1 (Some s, _) -> if s then et () else ee ()
+  | _ -> P1 (None, C.if_ (abstr1 eb) (fun () -> abstr1 (et ()))
                                      (fun () -> abstr1 (ee ())))
 \end{code}
-However, we stumble on functions.  According to our
-definition of~|P1|, a partially evaluated object function, such as the
-identity $\fun{x}x$ embedded in OCaml as |lam (fun x -> x)|\texttt{ :
-}|('c,'a->'a) P1.repr|, consists of a dynamic part 
-(type |('c,'a->'a) C.repr|) and
-maybe a static part (type |('c,'a->'a) R.repr|).  The dynamic part is useful
+However, we stumble on functions.  Given how we just
+defined~|repr|, a partially evaluated object function, such as the
+identity $\fun{x}x$ (of type $\ZZ\to\ZZ$) embedded in OCaml as
+|lam (fun x -> x)| (of type |('c,int->int) repr|), consists of
+a dynamic part (of type |('c,int->int) C.repr|) and optionally
+a static part (of type |('c,int->int) R.repr|).  The dynamic part is useful
 when this function is passed to another function that is only
@@
 appears as the argument of |app|.  To do so, we have to incorporate |f|
-as it is into the |P1.repr| data structure: the representation
-for a function type |'a->'b| should be one of
+as is into |lam f|: the type |('c,'a->'b) repr| should be one of
 \begin{code}
@@
 \end{code}
-unlike |P1.repr| of |int| or |bool|.
+unlike |('c,int) repr| or |('c,bool) repr|.
 That is, we need a nonparametric data type, something akin to
@@
 
-Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
-would ideally like is to write
-\begin{code}
-type ('c,'dv) repr = P1 of (repr_pe ('c,'dv)) R.repr option * ('c,'dv) C.repr
-\end{code}
-where |repr_pe| is the type function defined
-% inductively because P below depends on repr_pe
+The problem in the last section is that we want to write
+\begin{code}
+type ('c,'dv) repr = P1 of ('c,'dv) static option * ('c,'dv) C.repr
+\end{code}
+where |static| is the type function defined
+% inductively because P below depends on static
 by
 \begin{code}
-repr_pe ('c,int)    = ('c,int)
-repr_pe ('c,bool)   = ('c,bool)
-repr_pe ('c,'a->'b) = ('c,'a) repr -> ('c,'b) repr
+('c,int)    static = ('c,int) R.repr
+('c,bool)   static = ('c,bool) R.repr
+('c,'a->'b) static = ('c,'a) repr -> ('c,'b) repr
 \end{code}
 Although we can use type classes to define this type function
-in Haskell, that is not portable to MetaOCaml. However,
-these three typecase alternatives are already present in existing
+in Haskell, that is not portable to OCaml. However,
+the three typecase alternatives of |static| are already present in existing
 methods of |Symantics|.
-A simple and portable solution\footnote{though somewhat long-winded} thus
-emerges: we bake |repr_pe| into the signature |Symantics|. 
-We recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that 
-the |repr| type constructor took two arguments |'c| and~|'dv|.
-We add an argument |'sv| for the result of applying |repr_pe| to~|'dv|.
-Figure~\ref{fig:ocaml} shows the new signature.
+Thus emerges
+a simple and portable solution, if a long-winded one:
+we bake |static| into the signature |Symantics|. 
+In Figure~\ref{fig:ocaml-simple},
+the |repr| type constructor took two arguments |('c,'dv)|;
+in Figure~\ref{fig:ocaml},
+we add an argument |'sv| for the type |('c,'dv) static|.
 \begin{figure}
@@
 
-The interpreters |R|, |L| and~|C| in \S\ref{S:interpreter-RL},\S\ref{S:compiler}
+The interpreters |R|, |L| and~|C| in \S\ref{S:interpreter-RL} and~\S\ref{S:compiler}
 only use the old
@@
 with the rest the same.
-Predictably, the partial evaluator~|P| crucially relies on the type argument
-|'sv|.
 
 Figure~\ref{fig:pe} shows the partial evaluator~|P|.
-Its type |repr| expresses the type definition for |repr_pe| given
-at the start of this section, with |'sv| taking the place of |repr_pe t|.
+Its type |repr| expresses the definition for |static| given
+at the start of this section, with |'sv| taking the crucial place of |('c,'dv) static|.
 The function |abstr|
@@
 HOAS function. If it is, we apply |f| to the
-concrete argument |ea|, giving us a chance to perform static
+concrete argument |ea|, so as to perform static
 computations (see the example below). If
@@
 multiplication by~$0$ or~$1$.  
-Such algebraic simplifications are easy
-to abstract over the specific domain (such as monoid or ring) where they
+Although our basic machinery for partial evaluation is independent of
+such algebraic simplifications, it makes them easy to add and
+to abstract over the specific domains (such as monoid or ring) where they
 apply.  These simplifications and abstractions help a lot
@@
 
-\noindent All pattern\hyp
+All pattern\hyp
 matching in~|P| is \emph{syntactically} exhaustive, so it is patent to the
-metalanguage implementation that |P| never gets stuck.  Further, all
-pattern\hyp matching occurs
-during partial evaluation, only to check if a value is known statically,
-never what type it has.  In other words, our partial evaluator tags
-phases (with |Some| and |None|) but not object types.
-
-Our typed partial evaluator is online and polyvariant.  It reuses the
+metalanguage implementation that |P| never gets stuck.  Further,
+|P| uses pattern\hyp matching
+only to check if a value is known statically,
+never to check what type a value has dynamically.
+In other words, our partial evaluator tags
+phases (with |Some| and |None|) but not object types, so it
+is patent that the \emph{output} of~|P| never gets stuck.
+
+Predating us, \citet{Thiemann-combinators} and \citet{sumii-hybrid} deforested
+the object term representation and expressed a partial evaluator as a collection
+of term combinators in a typed metalanguage.  Like us, \citeauthor{sumii-hybrid}
+follow \citet{Sperber-SelfApplicable} and \citet{asai-binding-time} in building
+static and dynamic terms in tandem, to combine offline and online partial
+evaluation.
+\citearound{'s earlier self-reducer for the untyped
+$\lambda$-calculus}\Citet{JFP-Mogensen} also
+builds static and dynamic terms in tandem.  However, it
+builds a static term for every object term, even a bound variable,
+so it moves some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
+and remains untyped.  In contrast, we follow Sperber, Asai,
+and \citeauthor{sumii-hybrid} in leaving the static term
+optional, so as to perform lifting without
+juggling explicit type indices in the encoding of an object term.
+
+Our contribution to the literature on partial evaluation is to use mere
+Hindley-Milner types in the metalanguage to assure statically and patently that
+partially evaluating a well-typed object program not only never gets stuck but
+also, if it terminates, produces a well-typed output program that never gets
+stuck.  Moreover, these types form an instance of a general |Symantics|
+signature that encompasses other interpreters such as evaluation and
+compilation.  This early and manifest assurance of type safety contrasts, for
+example, with \citearound{'s compiler generator (cogen) for
+ML}\citet{Birkedal-PE-ML}, which transforms a program into its tagless
+generating extension.  Because that cogen uses a universal type, the fact that
+it never generates an ill-typed generating extension from a well-typed input
+program is only manifest when each generating extension is type-checked, and the
+fact that the generating extension never generates an ill-typed residual program
+from well-typed static input is only manifest when each residual program is
+type-checked.  Similarly, the fact that \citet{fiore:nbe-ppdp2002} and
+\citets{balat:tdpe-popl2004} partial evaluator using delimited control
+operators never turns well-typed code into ill-typed code is not assured by the
+metalanguage, whether or not as part of a typed family of interpreter modules.
+\begin{comment}
+% Removed the following comment in view of Section 4.5 of Danvy's
+% original TDPE paper (POPL 1996)
+It is
+type-directed, so the user must represent, as a term, the type of every
+term to be partially evaluated.  We shift this work to the type checker
+of the metalanguage.
+By avoiding term-level type representations, our
+approach makes it easier to perform algebraic simplifications (as
+in~\S\ref{S:PE-solution}).
+\end{comment}
+
+Our partial evaluator reuses the
 compiler~|C| and the evaluator~|R| by composing them.  This situation is
@@
 simulate a type function with a typecase distinction, using an extra
-type argument to |repr|. Luckily, the type function |repr_pe| needed for
+type argument to |repr|. Luckily, the type function |static| needed for
 the partial evaluator 
@@
 Hence, to write the type equation defining |CPST.repr| we again need
-a type function with a typecase distinction, similar to |repr_pe|
+a type function with a typecase distinction, similar to |static|
 in~\S\ref{S:PE-solution}. Alas, the type function we need is not
-identical to |repr_pe|, so we need to add another type argument to
+identical to |static|, so we need to add another type argument to
 |repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
@@
 
-Our partial evaluator establishes a bijection |repr_pe| between static
+Our partial evaluator establishes a bijection |static| between static
 and dynamic types (the valid values of |'sv| and |'dv|), and between
@@
 Our solution could have been even more straightforward if MetaOCaml
-provided total type-level functions such as |repr_pe| in
+provided total type-level functions such as |static| in
 \S\ref{S:PE-solution}---simple type-level computations 
 ought to become mainstream.
-
-At the term level, we also avoid converting between static and dynamic
-terms by building them in parallel, as \citet{Sperber-SelfApplicable}
-and \citet{asai-binding-time} do in their partial evaluators.
-This method type-checks in Hindley-Milner once we
-deforest the object term representation.  Put another way, we
-manually apply type-level partial evaluation to our type
-functions (see \S\ref{S:PE-solution}) to obtain simpler types 
-acceptable to MetaOCaml.
-
-\Citets{JFP-Mogensen} self-reducer for the untyped
-$\lambda$-calculus preceded Sperber's and Asai's partial evaluators in constructing
-static and dynamic terms in parallel.  However, this self-reducer
-constructs a static term for every object term, even a bound variable,
-so (in terms of Fig.~\ref{fig:pe}) it moves some work from |app|
-to~|pdyn|.  In contrast, we follow Sperber and Asai in constructing the static term
-optionally, so as to perform type-directed lifting without requiring the
-encoding of an object term to juggle explicit type indices.
-
-\Citet{sumii-hybrid} also use Sperber and Asai's construction,
-to combine online
-and offline partial evaluation.
-They predate us in deforesting the object term representation to enable tagless
-partial evaluation.
-We strive for modularity by reusing interpreters for individual stages
-\citep{SperberThiemann:TwoForOne}: our partial evaluator~|P|
-reuses our tagless evaluator~|R| and tagless compiler~|C|, so it
-is patent that the \emph{output} of~|P| never gets
-stuck.  It would be interesting
-to try to derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
-in the same manner.
 
@@
 
-\Citet{fiore:nbe-ppdp2002} and \citet{balat:tdpe-popl2004} also build
-a tagless partial evaluator, using delimited control operators.  It is
-type-directed, so the user must represent, as a term, the type of every
-term to be partially evaluated.  We shift this work to the type checker
-of the metalanguage.  By avoiding term-level type representations, our
-approach makes it easier to perform algebraic simplifications (as
-in~\S\ref{S:PE-solution}).
-
-\oleg{The following are almost unedited excerpts from our
-  correspondence with Sumii about Birkedal's thesis. We definitely
-  should include this discussion and the reference to his thesis.}
-\Citet{Birkedal-PE-ML} have hand-written a compiler generator (cogen)
-for Core Standard ML, which transforms a program into its tagless
-generating extension (and therefore serves as a tagless partial
-evaluator, intepreter, and compiler in our sense). They still have to
-use the "universal type" at the cogen level, which is removed from
-compile-time, so that the end result is tagless indeed.  The main
-difference between the work of \citet{Birkedal-PE-ML} and ours is how
-\emph{early} it becomes \emph{manifest} to the metalanguage
-that our code is safe.  The fact that
-\citet{Birkedal-PE-ML}'s generating extension never generates a
-specialized program that is ill-typed is not manifest to the SML
-compiler until the SML compiler sees each particular specialized
-program.  Moreover, and in a similar vein, the fact that
-\citet{Birkedal-PE-ML}'s cogen never generates a generating extension
-that is ill-typed (or that may generate a specialized program that is
-ill-typed) is not manifest to the SML compiler until the SML compiler
-sees each particular generating extension (or each particular
-specialized program).  This early and manifest type safety is the
-payoff of our treating an object program not as a data structure but
-as calls to abstract-syntax constructor functions.  The tradeoff, as
-we currently see it, is that we have to swear off intensional analysis
-of code. \oleg{Unless we convert to our representation to GADT, which
-  is always possible, if GADTs are available.}
-
 % The following paragraph was not in the version submitted to APLAS,
@@
 
-Our encoding of the type function |repr_pe| in \S\ref{S:PE-solution}
+Our encoding of the type function |static| in \S\ref{S:PE-solution}
 emulates type-indexed types and is related to intensional type analysis

### 2008-09-12 tagless/tagless-final.tex
Minor changes to make notation more uniform

@@
 It is straightforward to create an algebraic data type, say in OCaml%
-\ifshort, Fig.~\ref{fig:tag-problem}(a)\fi, to
+\ifshort, Figure~\ref{fig:tag-problem}(a)\fi, to
 represent object terms such as those in Figure~\ref{fig:object}.
@@
 we try to implement an interpreter function |eval0|\ifshort,
-Fig.~\ref{fig:tag-problem}(b)\fi. It takes
+Figure~\ref{fig:tag-problem}(b)\fi. It takes
 an object term such as |test1| above and gives us its value.
@@
 {WalidICFP02} containing both booleans and functions\ifshort,
-Fig.~\ref{fig:tag-problem}(c)\fi.
+Figure~\ref{fig:tag-problem}(c)\fi.
 \ifshort\else\UseVerbatim{2c}\fi
 We can then write a typed interpreter\ifshort,
-Fig.~\ref{fig:tag-problem}(d), \else\UseVerbatim{2d}\fi
+Figure~\ref{fig:tag-problem}(d), \else\UseVerbatim{2d}\fi
 whose inferred type is |u list -> exp -> u|. Now we can evaluate
@@
     \item evaluate the term to a value in the metalanguage;
-    \item measure the size or depth of the term;
+    \item measure the length of the term;
     \item compile the term, with staging support such as in MetaOCaml;
@@
 approach using de Bruijn indices
-(in Section~\ref{S:de-Bruijn} and the accompanying source file
+(in \S\ref{S:de-Bruijn} and the accompanying source file
 |incope-dB.ml|).
@@
 |Symantics| (with a parameter |repr| of kind |* -> *|)\ifshort,
-Fig.~\ref{fig:symantics-haskell}(a)\fi. The class is so named
+Figure~\ref{fig:symantics-haskell}(a)\fi. The class is so named
 because its interface gives the syntax of the object language and its
@@
 For another example, the classical $\mathit{power}$ function is
-\ifshort in Fig.~\ref{fig:symantics-haskell}(b)
+\ifshort in Figure~\ref{fig:symantics-haskell}(b)
 \else\UseVerbatim{3b}\fi
 and the partial application $\fun{x} \mathit{power}\;x\;7$ is
-\ifshort in Fig.~\ref{fig:symantics-haskell}(c).
+\ifshort in Figure~\ref{fig:symantics-haskell}(c).
 \else\UseVerbatim{3c}\fi
@@
 
-The functor |EX| in Fig.~\ref{fig:ocaml-examples} encodes 
+The functor |EX| in Figure~\ref{fig:ocaml-examples} encodes 
 our running examples |test1| and the $\mathit{power}$ function
@@
 
-Comparing |Symantics| with Fig.~\ref{fig:object}
+Comparing |Symantics| with Figure~\ref{fig:object}
 shows how to represent \emph{every} well-typed object term in the
@@
 The first interpreter evaluates an object term to its value in the
-metalanguage.  The module below is metacircular in that it
-interprets each object\hyp language
-operation as the corresponding metalanguage operation.
+metalanguage.  The module |R| below is metacircular in that it
+\emph{runs} each object\hyp language
+operation by executing the corresponding metalanguage operation.
 \ifshort
@@
 (The source code shows a total interpreter |L| that measures
-the size of each object term.)
+the length of each object term.)
 \fi
@@
 \ifshort\else
-For variety, we show another interpreter, which measures the \emph{size}
-(or \emph{length}, thus the mnemonic |L|)
+For variety, we show another interpreter~|L|, which measures the
+\emph{length}
 of each object term, defined as the number of term
@@
 alternative, starting with the |Symantics| signature and the |R|
-evaluator in Fig.~\ref{fig:de-Bruijn}.
+evaluator in Figure~\ref{fig:de-Bruijn}.
 
@@
 stage annotations yields the compiler~|C|
-\ifshort in Fig.~\ref{fig:interpreter-C}(a).\else below.
+\ifshort in Figure~\ref{fig:interpreter-C}(a).\else below.
 \UseVerbatim{5a}
@@
 unoptimized code. For example, interpreting our |test1| with
-\ifshort Fig.~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
+\ifshort Figure~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
 gives the code value |.<(fun|~|x_6|~|->|~|x_6)| |true>.|
@@
 with
-\ifshort Fig.~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
+\ifshort Figure~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
 gives a code value with many apparent $\beta$- and $\eta$-redexes\ifshort,
-Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
+Figure~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
 This compiler does not incur
@@
 unextended object language: we can pass |RCPSI|
-to the functor |EX| (Fig.~\ref{fig:ocaml-simple}) and run 
+to the functor |EX| (Figure~\ref{fig:ocaml-simple}) and run 
 the example |test1| from there. The main use for |RCPSI| is to interpret the 
@@
 in~\S\ref{S:interpreter-RL} may as well be written in our object
-language, as in Fig.~\ref{fig:self-eval}.
+language, as in Figure~\ref{fig:self-eval}.
 \begin{figure}
@@
 $\ident{lam}$, and $\ident{add}$ by their definitions given in
-Fig.~\ref{fig:self-eval}.  For
+Figure~\ref{fig:self-eval}.  For
 example, the familiar object term $(\fun{x}x)\True$ pre-encodes to
@@
 \end{equation}
-Because the evaluator in Fig.~\ref{fig:self-eval} mostly consists 
+Because the evaluator in Figure~\ref{fig:self-eval} mostly consists 
 of glorified identity
@@
     \end{tabular}
-\intertext{and the size\hyp measurer at the end of~\S\ref{S:interpreter-RL} is the context}
-\label{e:size-measurer}
-    \mathrm{SZ}[~] &=
+\intertext{and the length\hyp measurer at the end of~\S\ref{S:interpreter-RL} is the context}
+\label{e:length-measurer}
+    \mathrm{L}[~] &=
     \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
@@
 into multiple interpreters such as the evaluator~\eqref{e:evaluator} and
-the size\hyp measurer~\eqref{e:size-measurer}.  In return, we avoid
+the length\hyp measurer~\eqref{e:length-measurer}.  In return, we avoid
 adding polymorphism to the object language's type system, because we can
@@
 are not tagless because they still use universal types, such as Lava's
-\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Fig.~2.2)
+\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Figure~2.2)
 and \texttt{Term} (his Chap.~3), which incur the attendant overhead of

### 2008-09-12 tagless/tagless-final.tex
Explain the de Bruijn representation a bit more

@@
 evaluator in Figure~\ref{fig:de-Bruijn}.
+In this encoding of the object language, |vz| represents the
+innermost variable, |vs vz| represents the second-to-innermost variable,
+and so on.  The new type argument |'h| to |repr| tracks the type of
+the environment as a nested tuple, each of whose components is a
+value of type |('c,'dv) vr| representing a variable of type~|'dv|.
+The evaluator~|R| interprets each object term as a function
+from its environment to its value.
 

### 2008-09-16 tagless/tagless-final.tex
rephrase a sentence to ensure that it is simple to check that it is grammatically correct

@@
 from well-typed static input is only manifest when each residual program is
-type-checked.  Similarly, the fact that \citet{fiore:nbe-ppdp2002} and
-\citets{balat:tdpe-popl2004} partial evaluator using delimited control
-operators never turns well-typed code into ill-typed code is not assured by the
+type-checked.  Similarly, the fact that the partial evaluator of
+\citet{fiore:nbe-ppdp2002}, nor that of
+\citet{balat:tdpe-popl2004}, both of which use delimited control
+operators, never turn well-typed code into ill-typed code is not assured by the
 metalanguage, whether or not as part of a typed family of interpreter modules.

### 2008-09-17 tagless/tagless-final.tex
a fair number of changes, but mostly dealing with bibliographic issues as well as making our contributions clearer.  Removed section 6 in its entirety but added no comments about the topic.

@@
-%% TODO: add a ref 
-%%  Lars Birkedal's master thesis
-%%   http://www.itu.dk/people/birkedal/papers/paresm.ps.gz
-
 \begin{comment}
@@
 \ifdim\parindent=0pt \parindent=1em \fi
-We have built the first family of tagless interpretations for a
+We have built the first \emph{family} of tagless interpretations for a
 higher-order typed object
@@
 
-Our main idea is to encode
+Our principal technique is to encode
 \ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
@@
 interpret it, yet statically assures that the interpreters never get
-stuck.  To achieve self\hyp interpretation and show Jones\hyp
-optimality, we relate this exemplar of higher-rank and higher-kind
-polymorphism \ifshort\else (provided by ML functors and Haskell~98 constructor
-classes) \fi to plugging a term into a context of let\hyp polymorphic
-bindings.
+stuck.  
+%% To achieve self\hyp interpretation and show Jones\hyp
+%% optimality, we relate this exemplar of higher-rank and higher-kind
+%% polymorphism \ifshort\else (provided by ML functors and Haskell~98 
+%% constructor classes) \fi to plugging a term into a context of 
+%% let\hyp polymorphic bindings.
+
+As a side-effect, we provide another convincing application of the 
+usefulness of being able to abstract over higher-kinded types.
 \end{abstract}
@@
 the serious performance penalty \citep{WalidICFP02}
-that inevitably arises \citep{Gluck-jones-optimality}
-of error checking at run time.
+that arises from error checking at run time.
 
@@
     extending the language does not invalidate terms already encoded.
-\item We describe an approach to self\hyp interpretation compatible with the
-  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
-  out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
+%% \item We describe an approach to self\hyp interpretation compatible with the
+%%   above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
+%%   out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
+\item Show how higher-kinded abstraction can be useful for building
+    embedded DSLs.
 \end{enumerate}
@@
 it has been cited as a difficult problem (\cite{sumii-hybrid}
+and \cite{Thiemann-combinators}
 notwithstanding) to interpret a typed object language in a typed metalanguage
@@
 price we pay for this simplified treatment.
+\begin{comment}
 Our discussion of self\hyp interpretation in~\S\ref{selfinterp} shows
@@
 a context, which is one of the many faces of polymorphism.
+\end{comment}
 
@@
 \ifshort\else
+\begin{comment}
+\jacques{Should we really eliminate this entirely?}
 \section{Self-interpretation}\label{selfinterp}
@@
 is a euphemism for application.
+\end{comment}
 
@@
 
+\jacques{From what other people have written (like those who refer to our
+work), it seems these comments of ours were prescient and not altogether
+obvious.  We should consider adding some comments about the crucial role
+of 'repr' (r below) somewhere in the paper}
+
 The crucial role of the higher-order type parameter r
@@
 unavailable in mainstream ML, and their implementation in GHC
-6.6.1 fails to
+6.8 fails to
 detect exhaustive pattern matching.  We also wanted to find the minimal
@@
 
+That it is possible to regard our various evaluators as forming a 
+family is also an old idea \citep{Holst-AMIX}. \Citeauthor{Holst-AMIX}
+introduces \emph{language triplets} as an interesting mechanism to 
+relate (untyped) languages in much the same way we do.  Seeds of 
+several of the techniques we use here can also be found in this paper.
+
 Our partial evaluator establishes a bijection |static| between static
@@
 such as addition and application, then providing multiple
-interpretations of them.  \Citet{JonesNielsen-D58} prefigured this
-separation when they decomposed a denotational definition of an untyped
-object language into a core semantics (which we call abstract syntax)
+interpretations of them.  \Citet{JonesNielson:94:AbstractInterpretation}
+prefigured this separation when they decomposed a denotational definition of an
+untyped object language into a core semantics (which we call abstract syntax)
 and multiple interpretations.  When implementing an embedded DSL in this
@@
 thesis contains the definitions of the interpreter and the compiler,
-in the beginning. Use this in response to Rev1)
+in the beginning. Use this in response to Rev1
 \end{comment}
+
+In many ways, our work builds on that of \citep{Thiemann-combinators}.
+We use the same basic techniques, but our partial evaluator is online
+as this is the only way to make partial evaluation ``fit'' in our
+family.  In other words, by moving to an online partial evaluator and
+using staging instead of a $2$-level language, we are able to show that
+\citeauthor{Thiemann-combinators}'s techniques are highly parametric
+and typeable using mainstream features.  Using many different type
+classes, including constructor classes, \citep{Thiemann-GenClass}
+presents a Gofer implementation of set of \emph{binding-time polymorphic}
+combinators.  By gathering all their classes into one and dropping the
+requirement for a polymorphic |lift|, they could have invented
+|Symantics| themselves.
 
@@
 language.  \Citet{laod93} use type classes to implement a metacircular
-interpreter (rather than a self\hyp interpreter) of a
+interpreter (a special kind of self\hyp interpreter) of a
 typed version of the SK language, which is quite different from our
@@
 along this line.  
+\begin{comment}
 In particular, they use
@@
 (see our~\S\ref{selfinterp}).
+\end{comment}
 \citearound{'s bibliography}\Citet{honsell99coinductive} refers to the
@@
 
-\ifshort We observe that \else As \S\ref{S:clarify} observes, \fi
+We observe that
 higher-rank and higher-kind

### 2008-10-03 tagless/tagless-final.tex
Repack related work

@@
 partial evaluation.  Conversely, the function |pdyn| injects a
-code value into the |repr| type. As
+code value into the |repr| type.
+Thus, |abstr| and |pdyn| are like the \emph{reify} and
+\emph{reflect} functions defined in normalization by evaluation
+\citep{Danvy-TDPE}, but as
 in~\S\ref{S:PE-problem}, we build dynamic terms alongside
 any static ones to express how the lift function is indexed
-by the dynamic type.  What is new is that we analogously
+by the dynamic type.  Analogously, we now
 build a static type alongside the dynamic type to express
 how the static type is indexed by the dynamic type.
+Thus we establish a bijection |static| between static and dynamic types,
+without defining at the type level the injection\hyp projection pairs
+customarily used to establish such bijections for interpreters
+\citep{Ramsey-ML-module-mania,Benton-embedded-interpreters},
+partial evaluation \citep{Danvy-TDPE},
+and type-level functions \citep{oliveira-typecase}.
+This emulation of type-indexed types is related to intensional type analysis
+\citep{Morrisett-intensional,Generic-Haskell}, but
+intensional type analysis cannot handle our |fix|
+\citep{xi-guarded}.
 
@@
 
-Predating us, \citet{Thiemann-combinators} and \citet{sumii-hybrid} deforested
+Our partial evaluator owes much to
+\citet{Thiemann-combinators} and \citet{sumii-hybrid}, who deforested
 the object term representation and expressed a partial evaluator as a collection
@@
 also, if it terminates, produces a well-typed output program that never gets
-stuck.  Moreover, these types form an instance of a general |Symantics|
+stuck.  Moreover, thanks to the online binding-time analysis performed by
+our partial evaluator (in contrast to \citeauthor{Thiemann-combinators}'s),
+these types form an instance of a general |Symantics|
 signature that encompasses other interpreters such as evaluation and
@@
 type-checked.  Similarly, the fact that the partial evaluator of
-\citet{fiore:nbe-ppdp2002}, nor that of
+\citet{fiore:nbe-ppdp2002} and that of
 \citet{balat:tdpe-popl2004}, both of which use delimited control
@@
 
-Even a simply typed $\lambda$-calculus obviously supports self\hyp
-interpretation, provided we use universal types \citep{taha-tag}.  The
+The simply typed $\lambda$-calculus can be interpreted even just in
+itself, provided we use universal types \citep{taha-tag}.  The
 ensuing tagging overhead motivated \citet{Makholm-TagElim,taha-tag} 
-to propose tag
-elimination, which however does not statically guarantee that all tags
+to propose \emph{tag
+elimination}, which however does not statically guarantee that all tags
 will be removed \citep{WalidICFP02}.
@@
 \Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
-\citet{peyton-jones-simple} seem to argue as follows that a self\hyp
-interpreter of a typed language cannot be tagless or Jones\hyp optimal:
-\begin{shortlist}
-\item One needs to encode a typed language in a typed language based on
-a sum type (at some level of the hierarchy)\\
-\item A \emph{direct} interpreter 
-for such an encoding of a typed language
-in a typed language requires either
-advanced types or tagging overhead\\
-\item Thus, an indirect interpreter is necessary, which needs a universal
-  type and hence tagging\ifshort\else\\
-\item Thus, any self-interpreter must have tags and cannot be 
-  Jones-optimal\fi.
-\end{shortlist}
+\citet{peyton-jones-simple} seem to argue that a typed
+interpreter of a typed language cannot be tagless without
+advanced types, based on the premise that the only way to
+encode a typed language in a typed language is to use
+a sum type (at some level of the hierarchy).
 While the logic is sound, we (following \citet{yang-encoding}) showed that the
-first step's premise is not valid.
+premise is not valid.
 
@@
 
-That it is possible to regard our various evaluators as forming a 
-family is also an old idea \citep{Holst-AMIX}. \Citeauthor{Holst-AMIX}
-introduces \emph{language triplets} as an interesting mechanism to 
-relate (untyped) languages in much the same way we do.  Seeds of 
-several of the techniques we use here can also be found in this paper.
-
-Our partial evaluator establishes a bijection |static| between static
-and dynamic types (the valid values of |'sv| and |'dv|), and between
-static and dynamic terms.  It is customary to implement such a bijection
-using an injection\hyp projection pair, as done for interpreters
-\ifshort \citep{Ramsey-ML-module-mania,Benton-embedded-interpreters}\else
-by \citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters}\fi,
-partial evaluation \ifshort \citep{Danvy-TDPE}\else by \citet{Danvy-TDPE}\fi,
-and type-level functions \ifshort \citep{oliveira-typecase}\else by
-\citet{oliveira-typecase}\fi.  As explained in~\S\ref{S:PE-solution}, we
-avoid injection and projection at the type level by adding an argument
-to |repr|.
-Our solution could have been even more straightforward if MetaOCaml
-provided total type-level functions such as |static| in
-\S\ref{S:PE-solution}---simple type-level computations 
-ought to become mainstream.
-
 Our separation between the |Symantics| interface and its many
@@
 such as addition and application, then providing multiple
-interpretations of them.  \Citet{JonesNielson:94:AbstractInterpretation}
+interpretations of them.
+The techniques we use to form such a family of interpreters
+find their seeds in \citearound{'s \emph{language
+triplets}}\citet{Holst-AMIX}, though in an untyped setting.
+\Citet{JonesNielson:94:AbstractInterpretation} also
 prefigured this separation when they decomposed a denotational definition of an
@@
 
-In many ways, our work builds on that of \citep{Thiemann-combinators}.
-We use the same basic techniques, but our partial evaluator is online
-as this is the only way to make partial evaluation ``fit'' in our
-family.  In other words, by moving to an online partial evaluator and
-using staging instead of a $2$-level language, we are able to show that
-\citeauthor{Thiemann-combinators}'s techniques are highly parametric
-and typeable using mainstream features.  Using many different type
-classes, including constructor classes, \citep{Thiemann-GenClass}
-presents a Gofer implementation of set of \emph{binding-time polymorphic}
-combinators.  By gathering all their classes into one and dropping the
-requirement for a polymorphic |lift|, they could have invented
-|Symantics| themselves.
+\Citet{Thiemann-GenClass}
+implemented a set of \emph{binding-time polymorphic}
+combinators in Gofer, using many constructor classes.
+By merging all their classes into one and dropping
+polymorphic lift, they could have invented
+|Symantics|.
 
@@
 
-Our encoding of the type function |static| in \S\ref{S:PE-solution}
-emulates type-indexed types and is related to intensional type analysis
-\citep{Morrisett-intensional,Generic-Haskell}. However, our object
-language and running examples in HOAS include |fix|,
-which intensional type analysis cannot handle
-\citep{xi-guarded}.
-
 We could not find work that establishes that

### 2008-10-03 tagless/tagless-final.tex
State some correctness properties of our partial evaluator

@@
 \end{code}
+If the object program does not use |fix|, then the output of |P|
+is $\beta$-normal.
+Also, |P| is correct in that, if interpreting an
+object term using~|P| terminates, then the |dy| component of the output is
+equivalent to the interpretation of the same object term using~|C|, modulo
+$\alpha$-renaming, $\beta$-reduction, and algebraic simplification.  To prove
+this correctness by structural induction on the object term, we need to
+strengthen the induction hypothesis to assert that the |st| component, if not
+|None|, is consistent with the |dy| component.
 

### 2008-10-07 SCP2008/paper/metamonads.tex
changes to code and paper (tex and pdf), as per latest email.

@@
 
-let mo_extend (ip:('a,'b) open_rec) (v:'a) : ('c, 'd) monad = 
+let mo_extend (ip:('a,'b) open_rec) (v:'a) : ('c, unit) monad = 
   perform s <-- fetch; store (orec_store ip v s)
 
-let mo_lookup (ip:('a,'b) open_rec) : ('c, 'd) monad =
+let mo_lookup (ip:('a,'b) open_rec) : ('c, 'a) monad =
   perform s <-- fetch; ret (orec_find ip s)
@@
 \end{code}
-We refer to the $a_{ij}$-the element of the matrix as |a.(i*m+j)|,
+We refer to the $a_{ij}$\hyp th element of the matrix as |a.(i*m+j)|,
 where |a| is the array component of the record |container2dfromvector|.
@@
 matrix, the rectangular block |(r,c)-(n,m)|, for the element with the 
-maximum absolute value. The function returns the value of the
+nonzero minimum absolute value. The function returns the value of the
 pivot thus found, and its location; the return value is an option type with

### 2008-10-09 tagless/tagless-final.tex
take care of one more comment, and further adjust jfp-notes to reflect current status
Abstracting over an inductive map over object types Minor fixes also

@@
 types, which are required for typed partial evaluation and CPS transformations.
-
-Our encoding of an object term abstracts over the various ways to
+Our encoding of an object term abstracts uniformly over the family of ways to
 interpret it, yet statically assures that the interpreters never get
@@
 %% let\hyp polymorphic bindings.
-
-As a side-effect, we provide another convincing application of the 
-usefulness of being able to abstract over higher-kinded types.
+This family of interpreters thus demonstrates again that it is
+useful to abstract over higher-kinded types.
 \end{abstract}
@@
 %%   out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
-\item Show how higher-kinded abstraction can be useful for building
+\item We show how to use higher-kinded abstraction to build
     embedded DSLs.
@@
 transformation.
+Of course, CPS is a well-studied topic, and \citearound{'s work on
+program generation}\citet{Thiemann-combinators} already includes a
+CPS evaluator expressed
+using combinator functions rather than data constructors.
+We focus here on expressing CPS transformations as part of a larger,
+typed family of interpreters.
 \ifshort\label{state}\label{S:CPS}%
@@
 |(|$t$|->'w)->'w|, because $t$ may be a function type.  Again we
-simulate a type function with a typecase distinction, using an extra
-type argument to |repr|. Luckily, the type function |static| needed for
+simulate a type function with a typecase distinction, by changing the
+type arguments to |repr|. Luckily, the type function |static| needed for
 the partial evaluator 
 in~\S\ref{S:PE-solution} is precisely the same type function we
-need for CBN CPS\@.
+need for CBN CPS\@, so our CBN interpreter can match the
+|Symantics| signature in~\S\ref{S:PE-solution}, without even
+using the |'dv| argument to |repr|.
 \begin{code}
@@
 using higher-rank polymorphism through OCaml record types.
-To avoid this higher-rank polymorphism in the core language, which
-Standard ML does not support, we could also define |RCN| as
+To avoid this higher-rank polymorphism in the core language,
+we could also define |RCN| as
 a functor parameterized over
@@
 \begin{code}
-module RCN(W : sig type w end)= struct
-  type ('c,'sv,'dv) repr = ('sv -> W.w) -> W.w
-  ...
-\end{code}
-This allows a translation of these techniques to, for example, Standard ML\@.
-Unfortunately, the loss of polymorphism in the answer type makes the 
-CPS evaluator less pleasant to use than the other implementations.
+module RCN(W: sig type w end) = struct
+  type ('c,'sv,'dv) repr = ('sv -> W.w) -> W.w   ...
+\end{code}
+This alternative is more cumbersome to use because the functor
+needs to applied once for each answer type, but it translates
+to, for example, Standard ML\@, whose core language does not
+support higher-rank polymorphism.
 
@@
 Now an applied abstraction
-evaluates its argument before proceeding. This approach is in
-line with \citets{reynolds-relation}, albeit typed. The
+evaluates its argument before proceeding.  The
 interpreter~|RCV| is useful for CBV evaluation of the object language
 whether the metalanguage is CBV or CBN\@.
+To match the same |Symantics| signature as |RCN| above, |RCV| uses
+\citets{reynolds-relation} CBV CPS transformation, in which variables
+denote computations (that is, functions from continuations),
+rather than \citets{PlotkinCBN}, in which variables denote values.
 
@@
 evaluator.
-This functor on interpreters performs a textbook
+This functor on interpreters performs \citets{PlotkinCBN} textbook
 CPS transformation on the object language.
@@
 in~\S\ref{S:PE-solution}. Alas, the type function we need is not
-identical to |static|, so we need to add another type argument to
+identical to |static|, so again we need to change the type arguments to
 |repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
@@
 \end{code}
+
+\subsection{Abstracting over an inductive map on object types}
+
+\begin{figure}[t]
+\begin{floatrule}
+\begin{code}
+module type Symantics1 = sig
+  type 'c dint
+  type 'c dbool
+  type ('c,'da,'db) darr
+  type ('c,'dv) repr
+  val int : int  -> ('c, 'c dint) repr
+  val bool: bool -> ('c, 'c dbool) repr
+  val lam : (('c,'da) repr -> ('c,'db) repr) -> ('c, ('c,'da,'db) darr) repr
+  val app : ('c, ('c,'da,'db) darr) repr -> ('c, 'da) repr -> ('c, 'db) repr
+  val fix : ('x -> 'x) -> (('c, ('c,'da,'db) darr) repr as 'x)
+  val add : ('c, 'c dint) repr -> ('c, 'c dint) repr -> ('c, 'c dint) repr
+  val mul : ('c, 'c dint) repr -> ('c, 'c dint) repr -> ('c, 'c dint) repr
+  val leq : ('c, 'c dint) repr -> ('c, 'c dint) repr -> ('c, 'c dbool) repr
+  val if_ : ('c, 'c dbool) repr
+            -> (unit -> 'x) -> (unit -> 'x) -> (('c, 'da) repr as 'x)
+end
+\end{code}
+\end{floatrule}
+\caption{A (Meta)OCaml embedding that abstracts over an inductive map on object types}
+\label{fig:inductive}
+\end{figure}
+
+Having seen that each CPS interpreter above matches a differently
+modified |Symantics| signature, one may wonder whether |Symantics|
+can be generalized to encompass them all.  The answer is yes: the
+|Symantics1| signature in Figure~\ref{fig:inductive} abstracts our
+representation of object terms not only over the type constructor
+|repr| but also over the three branches that make up an inductive
+map, such as |static| in~\S\ref{S:PE-solution}, from object types
+to metalanguage types.  The first two branches (for the object
+types $\ZZ$ and~$\BB$) become the abstract types |dint| and
+|dbool|, whereas the third branch (for object types $t_1\to t_2$)
+becomes the new abstract type constructor |darr|.
+
+Almost every interpreter in this paper can be made to match the
+|Symantics1| signature without changing any terms, by defining
+|dint|, |dbool|, |darr|, and |repr| suitably.  For example,
+the types in the evaluator~|R| and the CBV CPS transformer |CPST|
+should be changed as follows.
+\begin{code}[commandchars=\@\[\]]
+module R = struct
+  type 'c dint = int
+  type 'c dbool = bool
+  type ('c,'da,'db) darr = 'da -> 'db
+  type ('c,'dv) repr = 'dv   ...
+
+module CPST(S: Symantics1)(W: sig type 'c dw end) = struct open W
+  type 'c dint = 'c S.dint
+  type 'c dbool = 'c S.dbool
+  type ('c,'da,'db) darr =
+    ('c, 'da, ('c, ('c, 'db, 'c dw) S.darr, 'c dw) S.darr) S.darr
+  type ('c,'dv) repr =
+    ('c, ('c, ('c, 'dv, 'c dw) S.darr, 'c dw) S.darr) S.repr   ...
+\end{code}
+Modified thus, |CPST| produces modules that match |Symantics1| and
+thus can be not only evaluated or compiled but also transformed
+using |CPST| again.  The accompanying source file |inco.ml| shows
+the details, including one-pass CPS transformations in the
+higher-order style of \citet{danvy-representing}.
+
+The only interpreter in this paper that does not fit |Symantics1|
+is the partial evaluator~|P|.  It does not fit because it uses
+\emph{two} inductive maps on object types---both |'sv| and |'dv|
+in Figure~\ref{fig:ocaml}.  We could define a |Symantics2|
+signature to abstract over \emph{two} inductive maps over object
+types; it would include 4 abstract types and 2 abstract type
+constructors in addition to |repr|.  It would then be easy to
+make~|P| match |Symantics2|, but the output of~|P| can still only
+match |Symantics1|.  This escalation points to a need for either
+record-kind polymorphism (so that |'dv| in
+Figure~\ref{fig:inductive} may be more than just one type) or
+type-indexed types (so that we do not need to emulate them in the
+first place).
 
@@
 along with the continuation. This technique lets us support mutable
-state. As Figure~\ref{fig:state} shows, we extend our object language
+state (or more generally any monadic effect) by representing it
+using continuations \citep{filinski-representing}.
+As Figure~\ref{fig:state} shows, we extend our object language
 with three imperative features.
@@
     \item the let-form ``$\lapp{e_1}{x}e_2$'' evaluates $e_1$
-        before~$e_2$ even if $e_2$ does not use~$x$.
+        before~$e_2$ even if $e_2$ does not use~$x$ and even if
+        evaluation is CBN\@.
 \end{enumerate}
-If $x$ does not appear in~$e_2$, then ``$\lapp{e_1}{x}e_2$'' is same as
+The form ``$\lapp{e_1}{x}e_2$'' is equivalent to
+``let'' in \citearound{'s monadic metalanguage}\citet{moggi-notions}.
+If $x$ does not appear in~$e_2$, then it is same as
 the more familiar sequencing form ``$e_1;e_2$''.
-We can embed this extended object language into OCaml by extending the
+We embed this extended object language into OCaml by extending the
 |Symantics| signature in Figure~\ref{fig:ocaml}.
@@
       {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
-end) = struct
-  include ST
-  type ('c, 'sv, 'dv) result = 'c states -> 'sv
-  ... 
+end) = struct include ST ... 
   let lapp e2 e1 = {ko = fun k ->
@@
 the example |test1| from there. The main use for |RCPSI| is to interpret the 
-extended object language.
+extended language.
 \begin{code}
@@
 code.
+Yet another way to add side effects to the object language is to
+write a monadic interpreter (for a specific monad or a general
+class of monads), which can be structured as a module matching the
+|Symantics1| signature in Figure~\ref{fig:inductive}.
 \fi
@@
 \jacques{Should we really eliminate this entirely?}
+\section{Polymorphism over interpreters}
+
+It is crucial to our achievements above that the same object term can be
+interpreted by multiple interpreters, whether an evaluator, compiler, partial
+evaluator, or CPS transformer.  In the typical \emph{initial} approach to DSL
+embedding, interpretation is represented in the metalanguage by applying the
+interpreter to the object term.  Flipping this application around, our
+\emph{final} approach applies the object term to the interpreter.  In this
+section, we explain how this flip amounts to Church-encoding the recursive GADT
+of well-typed object terms and how it affects the notion of self-interpretation
+and Jones optimality.
+
+At the type level, an interpreter is represented by its type constructor
+|repr|.  For example, a |Symantics| module or instance that matches the simple
+encoding in~\S\ref{language} is in System~$F_\omega$ an interpreter
+
+, over which an object term is a polymorphic function---either an ML
+functor from a |Symantics| module containing |repr| or a Haskell value with
+a |Symantics| constraint over |repr|.
+
 \section{Self-interpretation}\label{selfinterp}
@@
 find their seeds in \citearound{'s \emph{language
-triplets}}\citet{Holst-AMIX}, though in an untyped setting.
+triplets}}\citet{Holst-AMIX}, though in an untyped setting.  In the typed
+setting, \citet{Nielson88} uses a typed $\lambda$-calculus as a metalanguage
+for expressing families of program analyses; however the embedding of the
+object language into the metalanguage, unlike our |Symantics| is not typed.
 \Citet{JonesNielson:94:AbstractInterpretation} also

### 2008-10-12 tagless/tagless-final.tex
Try to clarify Nielson88

@@
 find their seeds in \citearound{'s \emph{language
-triplets}}\citet{Holst-AMIX}, though in an untyped setting.  In the typed
-setting, \citet{Nielson88} uses a typed $\lambda$-calculus as a metalanguage
-for expressing families of program analyses; however the embedding of the
-object language into the metalanguage, unlike our |Symantics| is not typed.
+triplets}}\citet{Holst-AMIX}, though in an untyped setting.
 \Citet{JonesNielson:94:AbstractInterpretation} also
@@
 untyped object language into a core semantics (which we call abstract syntax)
-and multiple interpretations.  When implementing an embedded DSL in this
-way, it is also common practice to use phantom types to rule out
+and multiple interpretations.
+
+In the typed
+setting, \citet{Nielson88} expressed families of program analyses on
+typed object languages using a typed $\lambda$-calculus as
+a metalanguage; however, the embeddings of the
+object language and the analyses are not type-checked in the
+metalanguage, unlike with our |Symantics| signature.
+When implementing a typed, embedded DSL\@,
+it is also common practice to use phantom types to rule out
 ill-typed object terms, as done in Lava

### 2008-10-15 tagless/tagless-final.tex
add missing word and remove use of 'thus' twice in same sentence

@@
 This alternative is more cumbersome to use because the functor
-needs to applied once for each answer type, but it translates
+needs to be applied once for each answer type, but it translates
 to, for example, Standard ML\@, whose core language does not
@@
 Modified thus, |CPST| produces modules that match |Symantics1| and
-thus can be not only evaluated or compiled but also transformed
+can be not only evaluated or compiled but also transformed
 using |CPST| again.  The accompanying source file |inco.ml| shows

### 2008-10-20 tagless/online-pe.hs
An online partial evaluator for the pure lambda calculus, in Haskell and in Scheme
From: Chung-chieh Shan <ccshan@rutgers.edu>
Date: Mon, 20 Oct 2008 18:38:10 +0200
To: Jacques Carette <carette@mcmaster.ca>
Cc: Neil Jones <neil@diku.dk>, Oleg Kiselyov <oleg@okmij.org>
Subject: Re: Progress

@@ -1,0 +1,17 @@
+{-# LANGUAGE Rank2Types #-}
+-- [ccshan 2008-10-20]
+
+type Repr r sv dv = (Maybe sv, r dv)
+
+appPE :: (forall a b. r (a->b) -> r a -> r b) ->
+         Repr r (Repr r sa da -> Repr r sb db) (da->db) ->
+         Repr r sa da ->
+         Repr r sb db
+appPE app m n = case fst m of
+    Nothing -> (Nothing, app (snd m) (snd n))
+    Just g -> g n
+
+lamPE :: (forall a b. (r a -> r b) -> r (a->b)) ->
+         (Repr r sa da -> Repr r sb db) ->
+         Repr r (Repr r sa da -> Repr r sb db) (da->db)
+lamPE lam g = (Just g, lam (\x -> snd (g (Nothing, x))))

### 2008-10-23 tagless/tagless-final.tex
Say something about how much metalanguage polymorphism we need

@@
 higher-order style of \citet{danvy-representing}.
+
+The abstract type constructors in Figure~\ref{fig:inductive} exemplify
+the amount of polymorphism that our technique requires of the
+metalanguage in order to represent a given object language.  Generally,
+our technique represents term constructions (such as~$+$) by applying
+abstract functions (such as \texttt{add}) and represents type
+constructions (such as~$\to$) and typing judgments (namely `$:$') by
+applying abstract type constructors (such as \texttt{darr} and
+\texttt{repr}).  Therefore, it requires the metalanguage to support
+enough polymorphism to abstract over the interpretation of each
+inference rule for well-typed terms and well-kinded types.  For example,
+to encode System~F's term generalization rule
+\begin{equation*}
+    \begin{prooftree}
+        \[ [\alpha:\star] \proofoverdots e : t \]
+        \justifies \Lambda\alpha.\,e : \forall\alpha.\,t \text,
+    \end{prooftree}
+\end{equation*}
+the metalanguage must let terms (representing $\Lambda\alpha.\,e$)
+abstract over terms (interpreting~$\Lambda$) that are polymorphic both
+over type constructors of kind $\star\to\star$ (representing $t$ with
+$\alpha$ free) and over polymorphic terms of type
+$\forall\alpha\colon\star\dotso$ (representing $e$ with $\alpha$ free).
+These uses of
+higher-rank and higher-kind
+polymorphism let us type-check and compile object terms separately from
+interpreters.  This observation is consistent with the role of
+polymorphism in the separate compilation of modules
+\citep{shao-typed}.
 
@@
 
-We observe that
-higher-rank and higher-kind
-polymorphism lets us type-check and compile object terms separately from
-interpreters.  This \ifshort\else observation \fi is consistent with the role of
-polymorphism in the separate compilation of modules
-\citep{shao-typed}.
-
 \section{Conclusions}\label{conclusion}

### 2008-10-23 tagless/tagless-final.tex
Refer to Mogensen-SelfApplicable, which uses the same fold encoding as we do, but untyped

@@
 We emphasize ``typed'' and ``fold'' in the previous sentence.
-Our encoding of the recursive type of terms \citep{bohm-automatic}
+We use a typed version of \citets{Mogensen-SelfApplicable}
+encoding of the recursive type of terms \citep{bohm-automatic}, which
 makes folds over terms much easier to write than functions over
 terms that are not primitive recursive (or, compositional).
-In contrast, \citearound{'s encoding of the sum type of
+In contrast, \citearound{'s earlier encoding of the sum type of
 terms}\citet{JFP-Mogensen} does not privilege folds.
@@
 evaluation.
-\citearound{'s earlier self-reducer for the untyped
-$\lambda$-calculus}\Citet{JFP-Mogensen} also
-builds static and dynamic terms in tandem.  However, it
-builds a static term for every object term, even a bound variable,
-so it moves some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
-and remains untyped.  In contrast, we follow Sperber, Asai,
+\citearound{'s earlier self-reducers for the untyped
+$\lambda$-calculus}\Citet{JFP-Mogensen,Mogensen-SelfApplicable} also
+build static and dynamic terms in tandem.  However, they
+build a static term for every object term, even a bound variable,
+so they move some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
+and remain untyped.  In contrast, we follow Sperber, Asai,
 and \citeauthor{sumii-hybrid} in leaving the static term

### 2008-10-24 tagless/tagless-final.tex
add a bit to the related work and clean up the notes some more

@@
 
+Very recently others have also begun to realize the usefulness of being
+able to abstract over higher-kinded types, like our |repr|.
+\Citet{moors-generics-higher-kind} illustrate how they have leveraged the 
+same power within Scala.  \Citet{HORM08} also uses Scala, yet are clearly
+influenced by our work \citep{carette-kiselyov-shan-aplas}
+(as they themselves note); where we have concentrated
+on multiple efficient interpretations of the same language, they have 
+concentrated on composability.
+
 \section{Conclusions}\label{conclusion}

### 2008-10-31 tagless/tagless-final.tex
add ref to Consel
just a few minor edits

@@
 \end{code}
-and the similar function |abstrB0| for dynamic boolean terms. Here, 
+and a similar function |abstrB0| for dynamic boolean terms. Here, 
 |C.int| is used to convert a static term (of type |('c,int) R.repr|, 
 which is just |int|) to a dynamic term. We can now define the following
-components required by the |Semantics| signature:
+components required by the |Symantics| signature:
 \begin{code}
@@
 so they move some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
-and remain untyped.  In contrast, we follow Sperber, Asai,
+and remain untyped.  
+The idea of generating both static and dynamic components alongside each
+other seems to be in found in an early form in \citet[\S5]{consel-schism-93}. 
+In contrast, we follow Sperber, Asai,
 and \citeauthor{sumii-hybrid} in leaving the static term
@@
 The verbosity of these types is the only difficulty in defining a
-new signature to replace |Symantics| that the output of |CPST| matches.
+replacement signature for |Symantics| which captures that of |CPST|
+as well.
 
@@
 The techniques we use to form such a family of interpreters
-find their seeds in \citearound{'s \emph{language
+find their origins in \citearound{'s \emph{language
 triplets}}\citet{Holst-AMIX}, though in an untyped setting.

### 2008-10-31 tagless/IncoSelf.hs
Separating self-interpreter part out of Incope.hs

@@ -1,0 +1,495 @@
+{-# OPTIONS -fglasgow-exts -W #-}
+
+-- Experimenting with Self-interperter in the tagless-final style
+
+module IncoSelf where
+
+{-
+  The language is simply-typed lambda-calculus with fixpoint,
+  integers, booleans and comparison.
+
+  Lam hoas_fn | App e e | Fix hoas_fn |
+  I Int | B Bool | Add ie1 ie2 | Mul ie1 ie2 | Leq ie1 ie2 |
+  IF b e-then e-else
+  
+  The language is just expressive enough for the Gibonacci function
+  and the power function.
+
+  The compiler, the interpreter and the source and target languages
+  are *all* typed. The interpreter and the compiler use no tags.
+  There is no pattern-match failure possible: the evaluators never
+  get stuck.
+
+-}
+
+import Incope
+
+-- ---------------------------------------------------------------------------
+-- Self-interpretation
+
+twice_inc_3_e () =
+    -- The evaluation context in the following four lines is a self-interpreter
+    -- of the object language, encoded in the metalanguage.
+    let lam_ = lam (\f -> f) in
+    let app_ = lam (\f -> lam (\x -> app f x)) in
+    let add_ = lam (\m -> lam (\n -> add m n)) in
+    let int_ = lam (\i -> i) in
+    -- The term in the following three lines is the object term
+    --      let twice_ f x = f (f x) in let inc_ n = n + 1 in twice_ inc_ 3
+    -- encoded in the object language then encoded in the metalanguage.
+    let twice_ = app lam_ (lam (\f -> 
+	   app lam_ (lam (\x -> app (app app_ f) (app (app app_ f) x))))) in
+    let inc_ = app lam_ (lam (\n -> app (app add_ n) (app int_ (int 1)))) in
+    app (app app_ (app (app app_ twice_) inc_)) (app int_ (int 3))
+test_inc3e_r = compR $ twice_inc_3_e ()
+test_inc3e_c = compC $ twice_inc_3_e ()
+test_inc3e_p = compP $ twice_inc_3_e ()
+-- to verify the Prop 5
+test_inc3p = compP $ (app (app
+                       (lam (\twice -> lam (\inc_ -> 
+                                              app (app twice inc_) (int 3))))
[…]

### 2008-10-31 tagless/Incope.hs

@@ -1,4 +1,5 @@
-{-# OPTIONS -fglasgow-exts -W #-}
+{-# LANGUAGE GADTs, TypeOperators, ScopedTypeVariables #-}
+{-# OPTIONS_GHC -W #-}
 
 -- Interpreter, Compiler, Partial Evaluator
 
@@ -6,7 +7,7 @@ module Incope where
 
 {-
   The language is simply-typed lambda-calculus with fixpoint,
-  integers, booleans and comparison.
+  integers, booleans and comparison: essentially, PCF.
 
   Lam hoas_fn | App e e | Fix hoas_fn |
   I Int | B Bool | Add ie1 ie2 | Mul ie1 ie2 | Leq ie1 ie2 |
@@ -19,14 +20,6 @@ module Incope where
   are *all* typed. The interpreter and the compiler use no tags.
   There is no pattern-match failure possible: the evaluators never
   get stuck.
-
-  The implementation below is *almost* possible in OCaml. Indeed,
-  one can emulate Haskell typeclasses using explicit dictionary passing.
-  The dictionary will by a polymorphic (rank-2) record -- but it is OK,
-  OCaml permits it. Alas, The typeclass below uses higher-order type
-  constructors: repr is of a kind * -> *. OCaml does not support
-  higher-order polymorphism.
-
 -}
 
 -- This class defines syntax (and its instances, semantics) of our language
@@ -205,30 +198,6 @@ instance Show (ByteCode t) where
     show (IF be et ee)
         = "(if " ++ show be ++ 
           " then " ++ show et ++ " else " ++ show ee ++ ")"
-
--- An evaluator for the ByteCode: the virtual machine
--- The evaluator is partial: if we attempt to evaluate an open code
--- (a variable not found in the env), we will get an error.
--- We use Dynamic to `assure' that variables are properly typed,
--- that is, that the variable name (counter) witnesses its type.
--- A better assurance is to use STRef. Strange correspondence:
--- mutation and lambda application. The essence of both is sharing.
-{-
-type BEnv = [(Int,Dynamic)]
-
-eval :: Typeable t => BEnv -> ByteCode t -> t
-eval env (Var n) | Just dv <- lookup n env,
-		   Just v  <- fromDynamic dv = v
-eval env v@Var{} = error $ "Open code? variable not found: " ++ show v
[…]

### 2008-10-31 tagless/Incope.hs
Defining Symantics2 and injecting Symantics into Symantics2. Now we we can variously interpret a term after PE (and before PE, too).

@@ -25,21 +25,17 @@ module Incope where
 -- This class defines syntax (and its instances, semantics) of our language
 -- This class is Haskell98!
 class Symantics repr where
-    int :: Int -> repr Int                -- int literal
+    int  :: Int  -> repr Int              -- int literal
     bool :: Bool -> repr Bool             -- bool literal
 
     lam :: (repr a -> repr b) -> repr (a->b)
     app :: repr (a->b) -> repr a -> repr b
     fix :: (repr a -> repr a) -> repr a
 
-    add :: repr Int -> repr Int -> repr Int
-    mul :: repr Int -> repr Int -> repr Int
+    add :: repr Int  -> repr Int -> repr Int
+    mul :: repr Int  -> repr Int -> repr Int
+    leq :: repr Int  -> repr Int -> repr Bool
     if_ :: repr Bool -> repr a -> repr a -> repr a
-    leq :: repr Int -> repr Int -> repr Bool
-
--- The following `projection' function is specific to repr.
--- It is like `run' of the monad
---    comp :: repr a -> something a
 
 test1 () = add (int 1) (int 2)
 test2 () = lam (\x -> add x x)
@@ -378,7 +374,7 @@ ptestpw7  = compP . testpowfix7 $ ()
 ptestpw72 = compP  (app (testpowfix7 ()) (int 2))
 
 
-{- GADTs, although work, are still not quite satsifactory:
+{- GADTs, although work, are still not quite satisfactory:
 let us look at the following lines in incope.hs
 
     app ef ea = case ef of
@@ -407,35 +403,65 @@ apparent.
 -- that Asai's code type-checks in Hindley-Milner as soon as we
 -- deforest the static code representation.
 
-data Rep repr dynamic static = 
-    Rep { dynamic :: repr dynamic, static :: Maybe static }
+-- Generalized Semantics, with an inductive type map
 
-zE dynamic = Rep dynamic Nothing
+class Symantics2 rep where
+    zint  :: Int  -> rep Int Int              -- int literal
[…]

### 2008-11-03 tagless/tagless-final.tex
A few small corrections; new link to the code

@@
 encoding of the recursive type of terms \citep{bohm-automatic}, which
-makes folds over terms much easier to write than functions over
-terms that are not primitive recursive (or, compositional).
+makes it much easier to write folds over terms than term functions
+that are not primitive recursive (or, compositional).
 In contrast, \citearound{'s earlier encoding of the sum type of
@@
 \citet{metaocaml}) and standard Haskell. The complete code is
-available at \url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
+available at \url{http://okmij.org/ftp/tagless-final/}
 to supplement the paper. 
@@
 rather than de Bruijn indices to encode binding and
-ensures that our object programs are closed.
+ensure that our object programs are closed.
 We find HOAS to be more convenient, but we have also implemented our
@@
 encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
-Both Haskell's and MetaOCaml's type checker also flags syntactically invalid
+Both Haskell's and MetaOCaml's type checkers also flag syntactically invalid
 object terms, such as if we forget |app| somewhere above.
@@
 the future-stage code value~$e$ by compiling and linking it at run time.
-Bracket, escape, and run are akin to
+Bracket, escape, and run are akin (modulo hygiene) to
 quasi-quotation, unquotation, and |eval| of Lisp.
@@
 useful when this function is applied to a static argument, as in
-$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
+$(\fun{x}x)1$.  Neither part, however, lets us \emph{partially}
 evaluate the function, that is, compute as much as possible statically

### 2008-11-03 tagless/Incope.hs
Beautifying the code for publishing.

@@ -2,6 +2,8 @@
 {-# OPTIONS_GHC -W #-}
 
 -- Interpreter, Compiler, Partial Evaluator
+-- Code accompanying the paper by
+--   Jacques Carette, Oleg Kiselyov, and Chung-chieh Shan
 
 module Incope where
 
@@ -62,8 +64,6 @@ testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
 -- It is a typed, tagless interpreter: R is not a tag. The interpreter
 -- never gets stuck, because it evaluates typed terms only
 
--- Note that everything going on in the interpreter is straight out of
--- "Boxes go Bananas" by Washburn and Weirich (intended or otherwise)
 newtype R a = R a deriving Show
 unR (R x) = x
 
@@ -158,12 +158,11 @@ ltestpw72 = compL (app (testpowfix7 ()) (int 2)) -- 17
 -- (typed bytecode). The GADT does _not_ use the higher-order abstract
 -- syntax. We could have used template Haskell. Alas, its expressions
 -- are untyped.
--- Note how ByteCode represents MetaOCaml's `code'. Thus the compiler
+-- ByteCode represents MetaOCaml's `code'. Thus the compiler
 -- below neatly maps to the MetaOCaml (with no GADTs).
--- Also note, that like MetaOCaml `code', we never pattern-match on
--- ByteCode!
--- Note how the compiler never raises any exception and matches no tags
--- (no generated code has any tags)
+-- Like MetaOCaml `code', we never pattern-match on ByteCode!
+-- The compiler never raises any exception and matches no tags:
+-- generated code has no tags at all.
 
 -- The LIFT bytecode operation is used only during evaluation and for fmap
 -- it corresponds to a CSP in MetaOCaml.
@@ -201,6 +200,7 @@ newtype C t = C (Int -> (ByteCode t, Int))
 unC (C t) vc0 = t vc0
 
 {-
+-- If we had a polymorphic LIFT function, we could do:
 instance Functor C where
     fmap f = app (C(\vc -> (LIFT f, vc)))
 -}
@@ -323,7 +323,7 @@ instance Symantics P where
     -- residualize
     fix f = f (E (fix (abstr . f . E)))
     -}
-    -- Now, we just go all the way (see Jacques' point)
+    -- Now, we just go all the way
     -- provided `fixing' produces static results...
 
     fix f = pfix f -- need this charade for GADTs sake
[…]

### 2008-11-03 tagless/final_code/Incope.hs

@@ -1,15 +1,15 @@
-{-# OPTIONS -fglasgow-exts -W #-}
+{-# LANGUAGE GADTs, TypeOperators, ScopedTypeVariables #-}
+{-# OPTIONS_GHC -W #-}
 
 -- Interpreter, Compiler, Partial Evaluator
 -- Code accompanying the paper by
 --   Jacques Carette, Oleg Kiselyov, and Chung-chieh Shan
 
-
 module Incope where
 
 {-
   The language is simply-typed lambda-calculus with fixpoint,
-  integers, booleans and comparison.
+  integers, booleans and comparison: essentially, PCF.
 
   Lam hoas_fn | App e e | Fix hoas_fn |
   I Int | B Bool | Add ie1 ie2 | Mul ie1 ie2 | Leq ie1 ie2 |
@@ -27,21 +27,17 @@ module Incope where
 -- This class defines syntax (and its instances, semantics) of our language
 -- This class is Haskell98!
 class Symantics repr where
-    int :: Int -> repr Int                -- int literal
+    int  :: Int  -> repr Int              -- int literal
     bool :: Bool -> repr Bool             -- bool literal
 
     lam :: (repr a -> repr b) -> repr (a->b)
     app :: repr (a->b) -> repr a -> repr b
     fix :: (repr a -> repr a) -> repr a
 
-    add :: repr Int -> repr Int -> repr Int
-    mul :: repr Int -> repr Int -> repr Int
+    add :: repr Int  -> repr Int -> repr Int
+    mul :: repr Int  -> repr Int -> repr Int
+    leq :: repr Int  -> repr Int -> repr Bool
     if_ :: repr Bool -> repr a -> repr a -> repr a
-    leq :: repr Int -> repr Int -> repr Bool
-
--- The following `projection' function is specific to repr.
--- It is like `run' of the monad
---    comp :: repr a -> something a
 
 test1 () = add (int 1) (int 2)
 test2 () = lam (\x -> add x x)
@@ -147,25 +143,29 @@ ltest3 = compL . test3 $ ()
 
 
 ltestgib  = compL . testgib  $ ()
-ltestgib1 = compL . testgib1 $ ()
+ltestgib1 = compL . testgib1 $ () -- 23
 ltestgib2 = compL . testgib2 $ ()
 
 ltestpw   = compL . testpowfix $ ()
 ltestpw7  = compL . testpowfix7 $ ()
[…]

### 2008-11-03 tagless/final_code/incope1.hs

@@ -2,8 +2,8 @@
 {-# OPTIONS -fallow-undecidable-instances #-}
 
 -- Interpreter, Compiler, Partial Evaluator
--- The Partial Evaluator is implemented with typeclasses instead of GADTs
---
+-- This is the typeclass version of the code in Incope.hs
+
 -- Code accompanying the paper by
 --   Jacques Carette, Oleg Kiselyov, and Chung-chieh Shan
 
@@ -17,9 +17,7 @@ module Incope1 where
   I Int | Add ie1 ie2 |
   IFEQ ie1 ie2 e-then e-else
   
-  The language is just expressive enough for the Fibonacci function.
-
-  This is the typeclass version of the code in incope.hs
+  The language is just expressive enough for the Gibonacci function.
 -}
 
 -- This class defines syntax (and its instances, semantics) of our language
@@ -36,6 +34,7 @@ class SymanticsA ar a ci ca | ar -> ci, ar a -> ci ca  where
 class SymanticsAB ar a b ca cb cab | ar a->ca, ar b->cb, ar a b -> ca cb cab  where
     lam :: (ar ca a -> ar cb b) -> ar cab (a->b)
     app :: ar cab (a->b) -> ar ca a -> ar cb b
+
 
 test1 () = add (int 1) (int 2)
 test2 () = lam (\x -> add x x)
@@ -85,10 +84,6 @@ itestg = mkitest testgib1
 -- (typed bytecode). The GADT does _not_ use the higher-order abstract
 -- syntax. We could have used template Haskell. Alas, its expressions
 -- are untyped.
--- Note how ByteCode represents MetaOCaml's `code'. Thus the compiler
--- below neatly maps to the MetaOCaml (with no GADTs).
--- Note how the compiler never raises any exception and matches no tags
--- (no generated code has any tags)
 
 data ByteCode t where
     Var :: Int -> ByteCode t                -- variables identified by numbers
@@ -99,7 +94,6 @@ data ByteCode t where
     Add :: ByteCode Int -> ByteCode Int -> ByteCode Int
     IFEQ :: ByteCode Int -> ByteCode Int -> ByteCode t -> ByteCode t ->
             ByteCode t
-
 instance Show (ByteCode t) where
     show (Var n) = "V" ++ show n
     show (Lam n b) = "(\\V" ++ show n ++ " -> " ++ show b ++ ")"
@@ -157,14 +151,15 @@ ctestg = compC . testgib1 $ ()
 
[…]

### 2008-11-03 tagless/incope1.hs

@@ -2,6 +2,10 @@
 {-# OPTIONS -fallow-undecidable-instances #-}
 
 -- Interpreter, Compiler, Partial Evaluator
+-- This is the typeclass version of the code in Incope.hs
+
+-- Code accompanying the paper by
+--   Jacques Carette, Oleg Kiselyov, and Chung-chieh Shan
 
 module Incope1 where
 
@@ -14,19 +18,6 @@ module Incope1 where
   IFEQ ie1 ie2 e-then e-else
   
   The language is just expressive enough for the Gibonacci function.
-
-  The compiler, the interpreter and the source and target languages
-  are *all* typed. The interpreter and the compiler use no tags.
-  There is no pattern-match failure possible: the evaluators never
-  get stuck.
-
-  The implementation below is *almost* possible in OCaml. Indeed,
-  one can emulate Haskell typeclasses using explicit dictionary passing.
-  The dictionary will by a polymorphic (rank-2) record -- but it is OK,
-  OCaml permits it. Alas, The typeclass below uses higher-order type
-  constructors: repr and res are of a kind * -> *. OCaml does not support
-  higher-order polymorphism.
-
 -}
 
 -- This class defines syntax (and its instances, semantics) of our language
@@ -44,10 +35,6 @@ class SymanticsAB ar a b ca cb cab | ar a->ca, ar b->cb, ar a b -> ca cb cab  wh
     lam :: (ar ca a -> ar cb b) -> ar cab (a->b)
     app :: ar cab (a->b) -> ar ca a -> ar cb b
 
-
--- The following `projection' function is specific to repr.
--- It is like `run' of the monad
---    comp :: repr a -> something a
 
 test1 () = add (int 1) (int 2)
 test2 () = lam (\x -> add x x)
@@ -68,8 +55,6 @@ testgib1 () = app (app (app (testgib ()) (int 1)) (int 1)) (int 5)
 -- It is typed, tagless interpreter: R is not a tag. The interpreter
 -- never gets stuck, because it evaluates typed terms only
 
--- Note that everything going on here is straight out of
--- "Boxes go Bananas" by Washburn and Weirich (intended or otherwise)
 newtype R t a = R a deriving Show
 unR (R x) = x
 
@@ -99,10 +84,6 @@ itestg = mkitest testgib1
[…]

### 2008-11-03 tagless/tagless-final.tex
Revise introduction

@@
 A popular way to define and implement a language is to embed it in
-another \citep{reynolds-definitional}.  Embedding means to represent
+another \citep{landin-next}.  Embedding means to represent
 terms and values of the \emph{object language} as terms and values in the
-\emph{metalanguage}.  Embedding is especially appropriate for domain\hyp
-specific object languages because it supports rapid prototyping and integration
+\emph{metalanguage}, so as to interpret the former in the latter
+\citep{reynolds-definitional}.
+Embedding is especially appropriate for domain\hyp
+specific languages (DSLs) because it supports rapid prototyping and integration
 with the host environment \citep{hudak-building}.
-If the metalanguage supports \emph{staging}, then
-the embedding can compile object programs to the metalanguage and avoid the
-overhead of interpreting them on the fly \citep{WalidICFP02}.  A staged
-definitional interpreter is thus a promising way to build a domain\hyp specific
-language (DSL)\@.
+
+Most interpreters suffer from various kinds of overhead, making it less
+efficient to run object programs via the metalanguage than to implement
+the object language directly on the machine running the metalanguage
+\citep{jones-partial}.  Two major sources of overhead are dispatching on
+the syntax of object terms and checking the types of object values.  If
+the metalanguage supports code generation
+\citep{nielson-two-level,nielson-automatic,gomard-partial,bawden-quasiquotation,taha-sound},
+then the embedding can avoid the dispatching overhead by compiling
+object programs, that is, by specializing an interpreter to object
+programs \citep{futamura-partial}.  Specializing an interpreter is thus
+a promising way to build a DSL\@.  However, the type-checking overhead
+remains, especially if the object language and the metalanguage both
+have a sound type system.  This overhead has motivated much work on
+typed specialization
+\citep{danvy-simple,Danvy-tagging-encoding,hughes-type,Birkedal-PE-ML,taha-tag,Makholm-TagElim}
+and type systems (see~\S\ref{s:fancier}).
 
@@
 
-We focus on embedding a \emph{typed} object language into a 
-\emph{typed} metalanguage.
-The benefit of types in this setting is to rule out meaningless object terms,
-thus enabling faster interpretation and assuring that our interpreters
-do not get stuck.
-To be concrete, we use the typed object language in
-Figure~\ref{fig:object} throughout this paper.  We aim not just for
-evaluation of object programs but also for
-compilation, partial evaluation, and other processing.
-
-\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
-a typed object language in a typed metalanguage as an interesting
-problem.  The common solutions to this problem store object terms and
-values in the metalanguage in a universal type, a generalized algebraic
-data type (GADT), or a dependent type.  In the remainder of this section,
-we discuss these solutions, identify their drawbacks, then summarize our
-proposal and contributions.  
+This paper shows how to eliminate type-checking overhead, whether in the
+context of code generation and whether in the presence of dispatching
+overhead.  We use metalanguage types, without such fancy features as
+generalized algebraic data types (GADTs) or dependent types, to rule out
+ill-typed object terms statically,
+thus speeding up interpretation and assuring that our interpreters
+do not get stuck.  We illustrate the problem of type-checking overhead
+in this section using a simple evaluator as example.  We apply our
+solution first to evaluation, then to code\hyp generation tasks such as
+partial evaluation.
+
 \ifshort
@@
 % accordingly. Let the paper and the accompanying code be in sync.
+To be concrete, we use the typed object language in
+Figure~\ref{fig:object} throughout this paper.
 It is straightforward to create an algebraic data type, say in OCaml%
@@
 \fi
-Following \citet{WalidICFP02},
-we try to implement an interpreter function |eval0|\ifshort,
+Let us try to implement an interpreter function |eval0|\ifshort,
 Figure~\ref{fig:tag-problem}(b)\fi. It takes
@@
 |eval0| traverses the function body~|e| every time (the result of
-evaluating) |L e| is applied. Staging can be used to remove this
-interpretive overhead \citep[\S1.1--2]{WalidICFP02}.
+evaluating) |L e| is applied. Code generation can be used to remove this
+interpretive overhead \citep{jones-partial,futamura-partial,WalidICFP02}.
 
@@
 
-The usual solution is to introduce a universal type \citep[\S1.3]
-{WalidICFP02} containing both booleans and functions\ifshort,
+The usual solution is to introduce a universal type
+containing booleans and functions\ifshort,
 Figure~\ref{fig:tag-problem}(c)\fi.
@@
 nonexhaustive pattern matching in |lookup| and |eval| and the tags |UB|
-and |UA| \citep[\S1.4]{WalidICFP02}.  In other words, the algebraic data
+and |UA|.  In other words, the algebraic data
 types above fail to express in the metalanguage that the object program
@@
 pattern\hyp matching operations that incur a performance penalty in
-interpretation \citep{WalidICFP02} and impair optimality in partial evaluation
-\citep{taha-tag}.  In short, the universal\hyp type solution is
+interpretation and impair optimality in partial evaluation
+\citep{jones-partial,taha-tag}.  In short, the universal\hyp type solution is
 unsatisfactory because it does not preserve the type of the encoded term.
 
-\ifshort\else\subsection{Solutions using fancier types}\fi
+\ifshort\else\subsection{Solutions using fancier types}\label{s:fancier}\fi
 
@@
 |test1| but disallow the ill-typed object term |test2|\fi.
-Yet other type systems
+Yet other fancy type systems
 have been proposed to distinguish closed terms like |test1| from open
@@
 \citep{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual}\fi,
-so that |lookup| never receives an empty environment.  We discuss these
-proposals further in \S\ref{related}\ifshort\else;
-here we just note that many
-fancy type systems have been devised to ensure statically that an
-object term is well-typed and closed\fi.
+so that |lookup| never receives an empty environment.
 

### 2008-11-03 tagless/tagless-final.tex
Address some of Olivier's comments and at least record the rest to deal with shortly

@@
         \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
-        \[ e_1:t_1\to t_2 \quad e_2:t_1 \justifies e_1 e_2: t_2 \]
+        \[ e_0:t_1\to t \quad e_1:t_1 \justifies e_0 e_1: t \]
         \[ \text{$n$ is an integer} \justifies n:\ZZ \]
@@
     and need no advanced type-system features such as GADTs, dependent types,
-    or intentional type analysis.
+    or intensional type analysis.
     Yet the type system of the metalanguage
@@
 object terms, such as if we forget |app| somewhere above.
+\ccshan{Make the point that the error messages are nice because we also
+metacircularly take advantage of any niceness in the error messages of
+the metalanguage implementation!}
 
@@
 \citep{BondorfDanvy,SwadiTahaKiselyovPasalic2006}.
+\ccshan{Fix the citations in the preceding sentence.}
 
@@
 and remain untyped.  
-The idea of generating both static and dynamic components alongside each
-other seems to be in found in an early form in \citet[\S5]{consel-schism-93}. 
 In contrast, we follow Sperber, Asai,
@@
 juggling explicit type indices in the encoding of an object term.
+The idea of generating static and dynamic components alongside each
+other is found in an early form in Schism \citep[\S5]{consel-schism-93}. 
+\ccshan{Refer to PE tradition and (earlier, even) the category-theory
+concept of \emph{gluing}.}
 
@@
 
-The simply typed $\lambda$-calculus can be interpreted even just in
-itself, provided we use universal types \citep{taha-tag}.  The
+The simply typed $\lambda$-calculus can interpret itself,
+provided we use universal types \citep{taha-tag}.  The
 ensuing tagging overhead motivated \citet{Makholm-TagElim,taha-tag} 
@@
 
+\ccshan{The following should be far from the first mention of optimality
+in this paper.}
 \citet{Danvy-tagging-encoding} discuss Jones optimality at length and
@@
 language.  \Citet{laod93} use type classes to implement a metacircular
-interpreter (a special kind of self\hyp interpreter) of a
+interpreter of a
 typed version of the SK language, which is quite different from our
@@
 because we represent object types as metalanguage types.  This work
-makes it safer and more efficient to embed DSLs
+improves the safety and reduces the overhead of embedding DSLs
 in practical metalanguages such as Haskell and ML\@.

### 2008-11-03 tagless/tagless-final.tex
Small fixes

@@
 constructors in addition to |repr|.  It would then be easy to
-make~|P| match |Symantics2|, but the output of~|P| can still only
+turn~|P| into a functor that returns a |Symantics2| module, but
+the input to~|P| can still only
 match |Symantics1|.  This escalation points to a need for either
@@
 
-Very recently others have also begun to realize the usefulness of being
-able to abstract over higher-kinded types, like our |repr|.
-\Citet{moors-generics-higher-kind} illustrate how they have leveraged the 
-same power within Scala.  \Citet{HORM08} also uses Scala, yet are clearly
-influenced by our work \citep{carette-kiselyov-shan-aplas}
-(as they themselves note); where we have concentrated
+Other researchers have very recently realized that it is useful to
+abstract over higher-kinded types, like our |repr|.
+\Citet{moors-generics-higher-kind} put the
+same power to work in Scala.  \Citet{HORM08} also use Scala and note
+that they are influenced by our work
+\citep{carette-kiselyov-shan-aplas}.  Where we have concentrated
 on multiple efficient interpretations of the same language, they have 
-concentrated on composability.
+concentrated on composing the languages and interpretations.
 

### 2008-11-05 tagless/tagless-final.tex
Finish addressing Olivier's comments

@@
 \citep{jones-partial}.  Two major sources of overhead are dispatching on
-the syntax of object terms and checking the types of object values.  If
+the syntax of object terms and tagging the types of object values.  If
 the metalanguage supports code generation
@@
 programs \citep{futamura-partial}.  Specializing an interpreter is thus
-a promising way to build a DSL\@.  However, the type-checking overhead
+a promising way to build a DSL\@.  However, the tagging overhead
 remains, especially if the object language and the metalanguage both
-have a sound type system.  This overhead has motivated much work on
+have a sound type system.
+The quest to remove all interpretive overhead, in particular by
+specializing the interpreter using a \emph{Jones-optimal} partial
+evaluator \citep{jones-partial}, has motivated much work on
 typed specialization
@@
 
-This paper shows how to eliminate type-checking overhead, whether in the
+This paper shows how to eliminate tagging overhead, whether in the
 context of code generation and whether in the presence of dispatching
@@
 thus speeding up interpretation and assuring that our interpreters
-do not get stuck.  We illustrate the problem of type-checking overhead
+do not get stuck.  We illustrate the problem of tagging overhead
 in this section using a simple evaluator as example.  We apply our
@@
 object terms, such as if we forget |app| somewhere above.
-\ccshan{Make the point that the error messages are nice because we also
-metacircularly take advantage of any niceness in the error messages of
-the metalanguage implementation!}
+Because our encoding of terms and types are so straightforward and
+metacircular, these error messages from the metalanguage implementation
+are just about as readable as those for ``native'' type errors such as
+|fun x -> x x|.
 
@@
 the residual program, which calls for let-insertion
-\citep{BondorfDanvy,SwadiTahaKiselyovPasalic2006}.
-\ccshan{Fix the citations in the preceding sentence.}
+\citep{BondorfDanvy}.
 
@@
 The idea of generating static and dynamic components alongside each
-other is found in an early form in Schism \citep[\S5]{consel-schism-93}. 
-\ccshan{Refer to PE tradition and (earlier, even) the category-theory
-concept of \emph{gluing}.}
+other is part of the tradition that developed partial evaluators such as
+Schism \citep[\S5]{consel-schism-93}.
 
@@
 
-\ccshan{The following should be far from the first mention of optimality
-in this paper.}
 \citet{Danvy-tagging-encoding} discuss Jones optimality at length and

### 2008-11-06 tagless/tagless-final.tex
Mention PCF

@@
 $\lambda$-calculus with fixpoint, integers, booleans, and comparison.
-The language is close to \citets{xi-guarded}, without their polymorphic
-lift but with more constants so as to more conveniently express Fibonacci,
-factorial, and power.
+The language is similar to \citearound{'s PCF}\citet{plotkin-lcf}.
+It is also close to \citets{xi-guarded}, without their polymorphic
+lift but with more constants so as to more conveniently express examples.
 In contrast to \S\ref{intro}, in the rest of the paper 

### 2008-11-06 tagless/final_code/Incope.hs
add missing files and fix up some comments

@@ -396,10 +396,10 @@ apparent.
 -- ------------------------------------------------------------------------
 -- A partial evaluator that does not use GADTs (except in using C)
 
--- Ken thinks that, without GADTs, it's not possible to explain in Haskell or
--- MetaOCaml how PE is an instance of Symantics (unless Symantics is 
--- generalized to carry both `static' and `dynamic' parts). But we can 
--- still explain PE by itself. In some sense we're just observing 
+-- At first, it seemed that, without GADTs, it was not possible to explain
+-- in Haskell or MetaOCaml how PE is an instance of Symantics (unless 
+-- Symantics is generalized to carry both `static' and `dynamic' parts). But 
+-- we can still explain PE by itself. In some sense we're just observing 
 -- that Asai's code type-checks in Hindley-Milner as soon as we
 -- deforest the static code representation.
 

### 2008-11-06 tagless/final_code/IncopeTypecheck.hs

@@ -1,0 +1,435 @@
+{-# OPTIONS -fglasgow-exts -W #-}
+
+-- Interpreter, Compiler, Partial Evaluator 
+-- Typechecking into Symantics terms
+
+-- The code demonstrates conversion of untyped terms into 
+-- typed terms that can be interpreted by various typed interpreters. 
+-- Typechecking happens only once, regardless of the number of 
+-- interpretations of the term. Although the typechecking function is
+-- necessarily partial (source terms may be ill-typed), we guarantee
+-- the resulting term can be interpreted without any pattern-match errors.
+-- Furthermore, these assurances are _syntactically_ apparent.
+
+-- That outcome is not trivial. Indeed, suppose we have a function
+-- tc :: UExp -> Gamma -> DynTerm repr, where UExp represents an 
+-- untyped expression (the result of parsing the source language)
+-- and DynTerm represents a typed expression of _some_ type, 
+-- interpreted by _any_ interpreter 'repr'.
+--
+-- For example, 
+--    tc (UAdd (UInt 1) (UInt 2)) []
+-- should give 
+--    DynTerm (add (int 1) (int 2))
+-- And it does, see the test dt1 below.
+--
+-- The trouble comes when we type-check a function, for example,
+--    UL "x" (UInt 100) (UAdd (UVar "x") (UInt 1))
+-- (here, (UInt 100) is a `type annotation', meaning that the bound variable
+-- "x" must have the type Int. All bound variables in the source language
+-- must be type annotated. The restriction is lifted if we implement
+-- the full HM typechecker with inference).
+-- The result of typechecking the function should be something like
+--    DynTerm (lam (\x -> ...))
+-- where ... is the result of type-checking the body assuming that "x"
+-- has the type Int. 
+--
+-- Naively, we would write
+--   DynTerm (lam (\x -> unDynTerm $ tc (UAdd (UVar "x") (UInt 1)) [("x",x)]))
+-- Although that approach works, it is horribly inefficient because
+-- the typechecking of the function body, 
+--   tc (UAdd (UVar "x") (UInt 1)) [("x",x)]
+-- will be run _every_ time our function is applied. We have the interpretation
+-- overhead at the typechecking stage. We have, however, a far more serious
+-- problem: the function |tc| cannot be total. Indeed, an untyped expression
[…]

### 2008-11-14 tagless/tagless-final.tex
Go over letter to the editor

@@
 $\alpha$ free) and over polymorphic terms of type
-$\forall\alpha\colon\star\dotso$ (representing $e$ with $\alpha$ free).
+$\forall\alpha\mathord:\mathord\star\dotso$ (representing $e$ with $\alpha$ free).
 These uses of

### 2009-03-03 tagless/stack-symantics.hs
added stack-symantics

@@ -1,0 +1,98 @@
+{-# LANGUAGE NoMonomorphismRestriction #-}
+{-# LANGUAGE Rank2Types #-}
+
+-- A stack machine as an instance of restricted Symantics 
+-- (restricted to arithmetics at present)
+
+module Forth where
+
+import Prelude hiding ((>>=), (>>), return, fail, drop)
+import qualified Prelude
+
+-- Abstract Stack machine
+class StackMachine stk where
+    push :: Show a => a -> stk s -> stk (a,s)
+    sadd :: stk (Int,(Int,s)) -> stk (Int,s)
+    swap :: stk (a,(b,s)) -> stk (b,(a,s))
+    drop :: stk (a,s) -> stk s
+
+(>>) = flip (.)
+
+newtype R a = R a
+
+-- An interepreter
+instance StackMachine R where
+    push x = \ (R s) -> R (x,s)
+    sadd   = \ (R (x,(y,s))) -> R (x+y,s)
+    swap   = \ (R (x,(y,s))) -> R (y,(x,s))
+    drop   = \ (R (x,s)) -> R s
+
+runR m = let R r = m (R ()) in r
+
+-- `Compiler'
+newtype C a = C String
+liftC f (C x) = C $ f x
+
+instance StackMachine C where
+    push x = liftC (++ (" push " ++ show x))
+    sadd   = liftC (++ " add")
+    swap   = liftC (++ " swap")
+    drop   = liftC (++ " drop")
+
+runC m = let C r = m (C "") in r
+
+tst1 = push 2 >>
+       push 3 >>
+       push 7 >>
+       sadd >>
+       sadd
+
+tst1R = runR tst1
+tst1C = runC tst1
+
+-- Type error
+-- tstadd = runR add
+
+tst2 = push 2 >>
+       push 3 >>
+       sadd    >>
+       push "hello" >>
+       swap >>
+       push 7 >>
+       sadd >>
+       swap >>
+       drop
+
+tst2R = runR tst2
+tst2C = runC tst2
+
+
+-- From our paper, taken verbatim
+class Symantics repr where
+    int  :: Int  -> repr Int              -- int literal
+    bool :: Bool -> repr Bool             -- bool literal
+
+    lam :: (repr a -> repr b) -> repr (a->b)
+    app :: repr (a->b) -> repr a -> repr b
+    fix :: (repr a -> repr a) -> repr a
+
[…]

### 2009-03-04 tagless/stack-symantics.hs
just extend with the obvious definitions

@@ -13,8 +13,11 @@ import qualified Prelude
 class StackMachine stk where
     push :: Show a => a -> stk s -> stk (a,s)
     sadd :: stk (Int,(Int,s)) -> stk (Int,s)
+    smul :: stk (Int,(Int,s)) -> stk (Int,s)
+    sleq :: Ord a => stk (a,(a,s)) -> stk (Bool,s)
     swap :: stk (a,(b,s)) -> stk (b,(a,s))
     drop :: stk (a,s) -> stk s
+    sapp :: stk ((a->b),(a,s)) -> stk (b,s)
 
 (>>) = flip (.)
 
@@ -24,8 +27,11 @@ newtype R a = R a
 instance StackMachine R where
     push x = \ (R s) -> R (x,s)
     sadd   = \ (R (x,(y,s))) -> R (x+y,s)
+    smul   = \ (R (x,(y,s))) -> R (x*y,s)
+    sleq   = \ (R (x,(y,s))) -> R (x<=y,s)
     swap   = \ (R (x,(y,s))) -> R (y,(x,s))
     drop   = \ (R (x,s)) -> R s
+    sapp   = \ (R (f,(a,s))) -> R ((f a), s)
 
 runR m = let R r = m (R ()) in r
 
@@ -36,8 +42,11 @@ liftC f (C x) = C $ f x
 instance StackMachine C where
     push x = liftC (++ (" push " ++ show x))
     sadd   = liftC (++ " add")
+    smul   = liftC (++ " mul")
+    sleq   = liftC (++ " <=")
     swap   = liftC (++ " swap")
     drop   = liftC (++ " drop")
+    sapp   = liftC (++ " apply")
 
 runC m = let C r = m (C "") in r
 
@@ -87,6 +96,9 @@ instance StackMachine c => Symantics (RR c) where
     int x  = RR (push x)
     bool x = RR (push x)
     add x y = RR (sadd . unRR y . unRR x)
+    leq x y = RR (sleq . unRR x . unRR y) -- note the flip!
+    mul x y = RR (smul . unRR y . unRR x)
+    app f x = RR (sapp . unRR f . unRR x)
 
 runRRR m = runR (unRR m)
 runRRC m = runC (unRR m)
@@ -96,3 +108,9 @@ test3R = runRRR test3
 -- (10,())
 test3C = runRRC test3
 -- " push 1 push 2 add push 3 push 4 add add"
+
[…]

### 2009-03-06 tagless/tagless-final.tex
Fixing typos.

@@
 \begin{SaveVerbatim}{ocaml-simple}
-module type Symantics = sig type ('c, 'dv) repr
+module type Symantics = sig
+  type ('c, 'dv) repr
   val int : int  -> ('c, int) repr
@@
 \begin{SaveVerbatim}{ocaml-examples}
-module EX(S: Symantics) = struct open S
+module EX(S: Symantics) = struct
+  open S
   let test1 () = app (lam (fun x -> x)) (bool true)
@@
 two inductive maps from terms and types in our object language
-to terms and types in Haskell and OCaml:
+to terms and types in Haskell and OCaml:\pagebreak
 \begin{align}
@@
     M(\fun{x}e) &= \texttt{lam (fun $x$ -> $M(e)$)} \notag\\
-    H(\fix{f}e) &= \texttt{fix (\textbackslash $x$ -> $H(e)$)} &
-    M(\fix{x}e) &= \texttt{fix (fun $x$ -> $M(e)$)} \notag\\
+    H(\fix{f}e) &= \texttt{fix (\textbackslash $f$ -> $H(e)$)} &
+    M(\fix{f}e) &= \texttt{fix (fun $f$ -> $M(e)$)} \notag\\
     H(e_1 e_2) &= \texttt{app $H(e_1)$ $H(e_2)$} &
@@
 \end{code}
-with the rest the same.
+with the rest being the same.
 
@@
 If the state has a higher-order type, then the types |state| and |'c states|
-are no longer the same, and |'s states| is mutually
+are no longer the same, and |'c states| is mutually
 recursive with the type |('c,'sv,'dv) repr|, as demonstrated in the
